### 2014-04-22 src/impliedVolModels.tex
Primera version

\chapter{Implied volatility models}

\section{Black-Scholes formula}

In finance, an option is a derivative, that is, a contract which gives the owner the right, but not the obligation to buy or sell an underlying asset at a given price called strike price. An option can be executed at any time before an expiration date previously defined no matter what price the underlying asset has.

The Black-Scholes formula~\cite{black1973}, developed in the early 1970's, Myron Scholes, Robert Merton and Fisher Black,  allows to determine an option value $V$ based on the underlying asset price $S(t)$ at a time $t$ and the following constant parameters:

\begin{description}
\item [$\sigma$:] underlying asset price volatility which measures the standard deviation of the returns
\item [$\mu$:] underlying asset drift which is a measure of the average rate of growth of the stock
\item[$E$:] option strike or excersice price
\item[$T$:] option date of expiry
\item[$t$:] current time
\item[$r$:] risk-free interest rate
\end{description}

\subsection{Stock price model}

The Black-Scholes model assume that the underlying price $S$ follows a lognormal random walk:

\begin{equation}\label{eq:stockprice}
dS = \mu S dt + \sigma S dB
\end{equation}

\noindent where $B$ is a Brownian motion. This stochastic differential equation has two components: a deterministic term given by $\mu S dt$ and a random term given by $\sigma S dX$.

\subsection{Brownian motion}

A brownian motion $B$ (also called a Wiener process) is a stochastic process characterized by the three following properties:

\begin{description}
\item[Continuity:] $B(t)$ is a continuos function
\item[Normal increments:]  $B(t)-B(s)$ has a normal distribution with mean $0$ and variance $t-s$.
\item[Independence of increments:] for every choice of nonnegative real numbers $0 \leq s_1 <  t_1 \leq \cdots \leq s_n < t_n < \infty$, the increment random variables $W_{t_1} - W_{s_1}, \cdots, W_{t_n} - W_{s_n}$ are jointly independent.
\end{description}

\subsection{Stochastic differential equation}

An stochastic differential integral has the form:

\begin{equation}
W(T)=\int_0^T f(t)dB(t) \, .
\end{equation}

\noindent This equations is also expressed in an abbreviate form:

\begin{equation}
dW = f(t) dB \, .
\end{equation}

\noindent Therefore, the integral form of the stock price model shown in equation (\ref{eq:stockprice}) is:

\begin{equation}
S(T)=\int_0^T \mu S(t) dt + \int_0^T \sigma S(t) dB(t)
\end{equation}

\subsection{Ito's Lemma}

Ito's lemma is used to find the differential of a time dependent function of a stochastic process. In option pricing we need to find the option price $V(S(t))$ which depends on a stochastic stock price model $S(t)$. 
$V(S,t)$ is required to be  differentiable function of $S$ and once differentiable function of $t$.

%These are known or can be easily obtained from the market, excepting by volatility which must be estimated. However, rather than assuming a volatility a priori and computing option prices from it, the model can be used to estimate volatility at given prices, time to expiration and strike price. This obtained volatility is called the implied volatility of an option. Additionally, some models obtain implied volatility from futures (other derivative from prices).

### 2014-04-22 src/introduction.tex
Primera version

\chapter{Introduction}

The computational complexity of machine learning algorithms has become a limiting factor for problems that require processing large volumes of data and where the response time is crucial. An example of this is financial volatility time series forecast.

Currently, it is known that volatility is a process where many factors influence its behaviour. However, many models of volatility made forecasts based only on price returns. Examples are the ARFIMA, GARCH and stochastic volatility models. Machines learning methods have allowed a greater number of explanatory variables to be included, although in many cases, these increase calculation time.

Algorithms that process large amounts of data in a short time and are able to detect changes in the financial market are required to achieve fast and accurate responses. Online learning algorithms have been developed to solve large-scale problems since they process an instance at a time, updating the model at each step incrementally. This is opposed to the batch algorithms where a phase of training that uses a large collection of historical data is first required and the resulting model is later used to predict. 
This thesis proposal is to design and develop an online learning algorithm that meets or improves the performance of current volatility algorithms and ensures rapid responses to large-scale problems. Model comparisons will be made by the SPA (superior predictive ability) test proposed by Hansen and
Lunde~\cite{hansen+lunde2006}  that allows comparing of forecast models giving a relative measure of model performance with respect to one or more models.

%Keywords: Volatility, Machine learning, Online learning, large-scale learning

\section{The problem}

In financial markets, volatility is one of the key elements to model the stochastic dynamic behaviour of financial assets. This is mainly because volatility gives a measure of uncertainty about future returns and it is often viewed as an indicator of the vulnerability of financial markets. It is also important as an input parameter in problems like derivative pricing, hedging and portfolio management. For example, in option pricing we need to known first the volatility of the underlying asset before pricing an option.

Besides, volatility provides important information for studying asset returns because the latter are largely uncorrelated and nonlinearly dependent. However, it has been found that volatility exhibits significant autocorrelation and predictable patterns~\cite{poon+granger2003} and  many models have been proposed to forecast its behaviour (see section ~\ref{sec:stylizedfacts}).

The volatility of the data is due to a large number of factors affecting the market, with some of them directly measurable such as historical prices, trends, supply and demand. However, others such as monetary policies and news are not directly measurable and are included in other models out of the scope of this thesis. Figure~\ref{fig:VolFactorNews} shows the VIX index, the daily US implied stock market volatility, commonly known as the "financial fear factor" which is clearly affected by historical events.

\begin{figure}[h]
 \centering
 \includegraphics[width=0.7\textwidth]{img/vol-factor-thenews.jpg}
 % vol-factor-thenews.jpg: 710x506 pixel, 72dpi, 25.05x17.85 cm, bb=0 0 710 506
 \caption{The news as volatility factor}
 \label{fig:VolFactorNews}
\end{figure}

Volatility models have been extensively studied~\cite{gatheral2006,poon+granger2003,knight2002} and some of  them include machine learning methods~\cite{hamidetal2004,donaldsonetal1997,shiyietal2008,shiyietal2010,gavrishchaka2006,vasilios2012}.  
The stochastic behaviour of volatility and the large number of factors affecting it result in models with a large number of parameters and slow response time due to the computationally expensive routines. These expensive calculations could be avoided using online learning algorithms, which allow response time reduction and are therefore suitable for high frequency trading applications.

In this thesis, a model will be proposed based on online learning algorithms to forecast volatility in high frequency financial time series.
The main problems to be deal with in this thesis are:

\begin{itemize}
   \item Although volatility is often predictable~\cite{poon+granger2003}, there are non-linear features~\cite{gatheral2006} to be considered in this proposal.
    \item The number of factors affecting volatility has been studied~\cite{srinivasan2009} and some other unknown factors affect volatility and they should also be included.
    \item The response time needs to be very fast and in order to ensure this, the proposed model will be presented as an online machine learning algorithm. 
\end{itemize}

\section{Volatility definition}

\begin{itemize}

\item Is the size of the price movement.

\item Variance is a measure of distribution of returns and is not neccesarily bound by any time period.
Volatility is a measure of the standard deviation (square root of the variance) over a certain time interval. In finance, variance and volatility both gives you a sense of an asset's risk. Variance gives you a sense of the risk in the asset over its lifetime, while volatility gives you a sense of the movement of the asset in eg. the past month or the past year.

\item The main underlying difference is in their definition. Variance has a fixed mathematical definition, however volatility does not as such. Volatility is said to be the measure of fluctuations of a process.

Volatility is a subjective term, whereas variance is an objective term i.e. given the data you can definitely find the variance, while you can't find volatility just having the data. Volatility is associated with the process, and not with the data.

In order to know the volatility you need to have an idea of the process i.e you need to have an observation of the dispersion of the process. All the different processes will have different methods to compute volatilities based on the underlying assumptions of the process.
\end{itemize}

\subsubsection{Types of volatility}

The volatility of a stock is not directly observable~\cite{tsay2005,engle1993}. For example, daily volatility is not directly observable from only daily returns because there is only one observation in a trading day.  If intraday data is available, then volatility could be estimated. However, intraday returns are not the only explanatory variables for volatility and several estimators have been proposed. These estimators are observable variables that are related to the latent variable of interest called volatility proxies~\cite{devilderetal2007}. Examples of volatility proxies are the following:

\begin{description}%[leftmargin=0.4cm]\itemsep4pt
\item[realized volatility:] is also known as historic volatility and
it is the actual variance in the price of a stock over time.
Realized volatility is measured in terms of the standard deviation
using the historical stock prices. It is commonly calculated based on
intraday price returns:

\begin{equation}
\label{eq:retintra}
r_{t,n}=100(\ln(p_{t,n}) - \ln(p_{t,n-1}))
\end{equation}

\noindent where $p_{t,n}$ is the price observed at day $t=1,\dots,T$ and
intraday sample $n=2,\dots,N$. Realized volatility is defined as:

\begin{equation}
\label{eq:rv}
    \hat{\sigma}(t) = \sum_{n=1}^N r_{t,n}^2 \, , 
\end{equation}

\noindent where $N$ is the number of intraday samples and $T$ is the
number of days.

In order to include overnight returns, Hansen and
Lunde~\cite{hansen+lunde2005} introduced a scaling version of
realized volatility using the following definitions:

\begin{eqnarray}
r_{t}&=&100(\ln(p_{t,N}) - \ln(p_{t-1,N})) \label{eq: ret 1} \\
\bar{\rho}(t) &=& \sum_{t=1}^T r_{t}^2  \label{eq: ret 2} \, .
\end{eqnarray}

\noindent where equation~(\ref{eq: ret 1}) represents overnight 
returns and the volatility as 
equation~(\ref{eq: ret 2}), where $p_{t,N}$ is the last intraday 
sample at day $t$. The scaled realized volatility $\rho(t)$ is 
defined as:

\begin{eqnarray}
\label{eq:srv}
\rho(t) = \gamma \hat{\rho}(t) \, , \qquad & \qquad \gamma = \displaystyle \frac{\bar{\rho}(t)}{\displaystyle\sum_{t=1}^T \hat{\rho}(t)}
\end{eqnarray}

Realized volatility has also been defined as the absolute value return or as the mean of the sum of intraday squared returns at short intervals of time. The majority of research carried out in the literature obtain the daily volatility as the daily squared returns as is shown in equation~(\ref{eq: ret 2}).  However, it has been proven that this measurement noise is too high for observing the true volatility process~\cite{andersen+bollerslev1998}. Hansen and Lunde~\cite{hansen+lunde2006} stated that the use of a noisy proxy could result in an inferior model being chosen as the best one. The realized volatility, as calculated by the cumulative sum of squared intraday returns and shown in equation (\ref{eq:rv}), is less noisy and doesn't lead to choosing an inferior model.

\item[implied volatility:] volatility not only can be extracted from returns but it can also be derived from option or future pricing models.  The volatility obtained corresponds to the market's prediction of future volatility. In finance, an option is a derivative, that is, a contract which gives the owner the right, but not the obligation to buy or sell an underlying asset at a given price called strike price. An option can be executed at any time before an expiration date previously defined no matter what price the underlying asset has. For example, the Black-Scholes model~\cite{black1973} determines the fair option value based on stock price, strike price, time to option expiration, the interest rate and volatility. These are known or can be easily obtained from the market, excepting by volatility which must be estimated. However, rather than assuming a volatility a priori and computing option prices from it, the model can be used to estimate volatility at given prices, time to expiration and strike 
price. This obtained volatility is called the implied volatility of an option. Additionally, some models obtain implied volatility from futures (other derivative from prices). For instance, the Barone-Adesi and Whaley futures option model~\cite{baroneetal1987} is also used to determine future volatilities~\cite{hamidetal2004}. Higher implied volatility is indicative of greater price fluctuation in either direction. Implied volatility is found by determining the value which makes theoretical prices equal to market prices. In this way volatility is ``implied'' by the current market price of the stock.

\end{description}

For trading strategies, the interest is centred in forecasting realized volatility over the life of an option and to take advantage when this volatility differs from the implied volatility. This is called volatility arbitrage. For example, a trader will buy an option and hedge the underlying asset if the implied volatility is under the realized volatility.

\section{Volatility methods}

In the existing literature, there are four main classes of asset
return volatility models: the general autoregressive conditional
heteroskedasticity (GARCH) models, the stochastic volatility (SV)
models, the realized volatility models and the machine learning based
models. A comparison of the first three models can be found
in~\cite{wei2012}.

For many years the most popular methods for estimating financial
volatility were the autoregressive conditional heteroskedasticity
(ARCH) models~\cite{engle1982} and the general ARCH (GARCH)
models~\cite{bollerslev1986}. For instance, the GARCH(1,1) defines
returns $y_t$ and volatility $\sigma_t$ as:

\begin{eqnarray*}
    y_t &=& \sigma_t \epsilon_t \\
     \sigma_t^2 &=& \alpha_0 + \alpha_1 \epsilon_{t-1}^2 + \beta_1
     \sigma_{t-1}^2
\end{eqnarray*}

\noindent where $\epsilon_t$ is standard Gaussian white noise,
$\alpha_0,\alpha_1,\beta_1 \geq 0$ are required to ensure that the
variance will never be negative and $\alpha_1+\beta_1 <1$ is needed to
guarantee a weakly stationary process~\cite{nelson1990}.

Since the introduction of the GARCH models, several extensions have been proposed, but none of them seems to beat the GARCH(1,1) model~\cite{lunde+hansen2005}. Despite its popularity, GARCH models have several limitations: firstly, a time series model may be non-linear in mean and/or non-linear in variance, but ARCH and GARCH models are non-linear in variance, but not in mean. Besides, GARCH models often fail to capture highly irregular phenomena, like wild market fluctuations.

SV models explain how volatility varies in a random fashion. These models are useful because they explain why options with different strikes and expirations dates have different Black-Scholes implied volatilities, phenomenon known as the volatility smile. This is useful because the Black-Scholes model assumes that the volatility of the underlying asset is constant which is not always true. There are several SV models and the most well-known and popular is the Heston model~\cite{heston1993}. Additional information about SV models can be found in~\cite{shephard1995}.

The realized volatility constructed from high frequency intraday returns gave rise to the realized volatility models mainly because the realized volatility series is much more homoskedastic and seems to be a long memory process~\cite{andersonetal2003}. For realized volatility, the autoregressive fractionally integrated moving average (ARFIMA) process emerged as a standard model~\cite{chenetal2010} and many variations have been studied, but all of them produce similar forecasting results to the ARFIMA(1,d,1) model~\cite{koopmanetal2005}.

On the other hand, machine learning based models, especially artificial neural networks (ANN) and support vector machines (SVM) have arisen as an alternative to forecast volatility. ANN is a statistical technique inspired by biological neural networks which is capable of changing its structure based on external or internal information during a training phase~\cite{sammut2011}. SVM are supervised learning models for classification analysis which recognize patterns finding a separating hyperplane. An extension for regression analysis is known as support vector regression (SVR).

Since machine learning models and in particular ANN do not require assumptions about the data (gaussianity for example) and allow more explanatory variables than returns to be included, they have become widely used in solving financial problems, specially volatility forecasting~\cite{hamidetal2004,donaldsonetal1997}. There are also many works focused on the using of SVM in volatility forecasting~\cite{shiyietal2008,shiyietal2010,gavrishchaka2006,vasilios2012}.

However, just as with ANN, SVMs have scalability problems because their training process is computationally intensive and it is done in batch mode. The scalability problem worsens when new additional training data is available and a re-training process from scratch needs to be done. This problem can be avoided using online machine learning algorithms that allow one instance at a time to be processed with low computationally expensive calculations.

\subsection{Online learning}

Classic statistical theory of sequential prediction enforces strong assumptions on the statistical properties of the input sequence (for example, stationary stochastic process). However, these assumptions can be unknown or change over time. In online learning there is no previous assumption about the data and the sequence is allowed to be deterministic, stochastic or even adaptive.

Moreover, in case we receive data streams, ANN or SVM cannot introduce new information into the model without a re-training process, so we will have to use the same non-updated model until we decide to compute another one if it is possible.  Online learning algorithms allow one example at a time to be introduced into an existing model incrementally~\cite{vovk2005}. This is extremely important when the problem has large data streams and real-time forecasting must be done.  This is the most common scenario when we want to forecast a wide range of data such as stock prices and volatilities, electricity power, intrusion detection, web-mining, server load, etc.  Besides, many problems of high interest in machine learning can be treated as online ones and they can also use these types of algorithms.

The online learning framework was first introduced in the perceptron algorithm~\cite{rosenblatt58}. There are several other widely used online methods such as passive-aggressive~\cite{crammerETall2006}, stochastic gradient descent~\cite{zhang2004}, aggregating algorithm~\cite{vovk2001} and the second order perceptron~\cite{cesa-bianchi2005}.  In~\cite{cesa-bianchi2006} an in-depth analysis of online learning is provided.

The motivation for online learning is to obtain computational efficiency and tackle the shifting problem, i.e. that the distribution of the data is unknown or changes over time. Online learning algorithms can deal with this problem because they have a tracking ability which is a strategy based on retaining weak dependence on past examples by using two types of models:

\textit{a)} \textbf{memory boundedness:} consists of limiting the number of support vectors in order to improve computational efficiency. One example of this is the budget perceptron~\cite{crammeretal2004} which reduces the number of examples used for prediction. Alternatively, in the forgetron algorithm~\cite{dekeletal2008} the damage caused by removing old examples is discussed, which can be avoided by removing samples with small influences. Other examples are the sliding window kernel (RLS)~\cite{vanvaerenberghetal2006}, which only considers a sliding window of the most recent data, and in \cite{arce+salinas2012} is shown a variant of aggregating algorithm for regression~\cite{vovk2001} considering only a sliding window of the most recent data, optimising also common matrix operations.

\textit{b)} \textbf{weight decay:} one example of this is the shifting perceptron algorithm which implements an exponential decaying scheme for the examples~\cite{cavallantietal2007}.
Performance of an online learning algorithm is measured by the cumulative loss it suffers along its run on a sequence of examples. In order to minimize this loss, the learner may update the hypothesis after each round so as to be more accurate in later rounds.

\section{Thesis proposal}

This thesis proposal is to build an algorithm to forecast realized volatility in financial markets considering not only intraday returns but also other explanatory volatility variables. Because of the amount of high frequency data to be considered and short response times required in financial applications, the algorithm will be formulated in online mode. The online algorithm will also allow the shifting problem to be tackled while being adaptive to wild changes in the volatility process.
The online learning model will be formulated considering the following:

\begin{description}%[leftmargin=0.4cm]\itemsep4pt
    \item[large volumes of data:] high frequency data will be used to estimate realized volatility. The frequency to be considered will be 10 to 15 minutes. This frequency will be analysed using signal processing theory. 
    \item[non-stationary data:] volatility is a time series known because of its non-stationary feature, so statistics will be studied before using the model.
    \item[stylized facts:] there are several well-documented features about financial time series volatility called stylized facts~\cite{poon+granger2003} that can be included in the model.
    \item[low response time:] the model has to be simple but powerful in order to achieve the response time needed. If the model is extended to real time data, ticks can arrive at rates of milliseconds and the model should respond at the same rate or less.
    \item[financial data:] the model can be constructed using data of stocks or different derivatives such as options and futures. The chosen implicit volatility model will depend on the data type.
    \item[input vector:] currently, many models are only based on expected returns, but volatility is a complex variable which depends also on many factors studied in the literature such as trading volume, the tick rate, time of the day and the behaviour of the market as a hole~\cite{gatheral2006}. All these factors can improve the design of the proposed online algorithm.
    \item[kernel version:] the online algorithm will be later kernelised in order to map the input data into a high dimensional feature space (kernel trick) which will provide better properties to design the solution.
\end{description}

The model proposed has to be compared with the current popular financial models of financial volatility such as: ARFIMA, GARCH, SV models and machine learning models. Performance will be measured using the realized volatility shown in equation~(\ref{eq:srv}) as a benchmark.

In order to compare different models, apart from using the standard performance measurements like MAE and MSE, the superior predictive ability (SPA) methodology proposed by Hansen and Lunde~\cite{hansen+lunde2006} will also be studied. SPA is a test to compare two or more forecasting models at the same time giving a relative performance measurement of a base model in comparison with alternative models.

The model will be compared against traditional financial volatility models such as Black-Scholes, SV, GARCH and ARFIMA using different forecast errors measurements. Although the best way to measure forecast errors should be the investor returns, this implies knowing how the model information is used as a strategy, so this information is not going to be included in this thesis. Therefore, in order to compare the different volatility forecast models, popular evaluation measurements will be used. In the literature the following are commonly used: {\em Mean Error} (ME), {\em Mean Square Error} (MSE), {\em Mean Absolute Error} (MAE) and {\em Mean Absolute Percent Error} {MAPE} and others are less commonly used such as LINEX loss function which weight differently positive and negative errors. In ~\cite{granger1999} Granger describes a variety of non-symmetric loss function including LINEX. The superior predictive ability (SPA) methodology proposed by Hansen and Lunde~\cite{hansen+lunde2006} will also be studied 
which allows comparison of two or more forecasting models at the same time giving a relative performance measurement of a base model compared with alternative models.  All these measurements will be out-of-sample in order to show their predictive power.

\section{Hypotheses}

Volatility forecasting has been a largely studied and different approaches have arisen to describe its stochastic behaviour. It is well-known that volatility depends on many factors, but most of the existing models only consider a few of them. In order to introduce more factors easily, classic machine learning models have been studied and their performance have shown to be successfull in comparison with most used volatility models. However, their training process is computationally expensive and it is not possible to update the model with new market information before new information arrives. Online learning methods can also include more factors into a model, but they process one example at a time to update an existing model incrementally without having a training phase, thus they improve response times and are more suitable for financial applications.

The thesis hypotheses are the following:

\begin{enumerate} 
\item Machine learning based models that consider more volatility explanatory factors will adapt and perform better than popular volatility forecasting models based only on returns or a few factors.
\item Stylized facts can be efficiently introduced as explanatory factors and improve model performance.
\item The use of online learning methods will allow inclusion of as many variables as needed to ensure low response times.  
\end{enumerate}

\section{General Objective}
To design and implement an online learning model for volatility forecasting capable of adapting to their non-stationary behaviour using volatility factors as model input ensuring low response times.  
\section{Specific Objectives}
The specific objectives to be followed in this thesis are:
\begin{enumerate}
  \item To determine the best volatility factors based on their stylized facts and their performance in machine learning models.
  \item To determine the best sampling of high frequency time series.
  \item To define the best realized volatility proxy based on the literature and experimental results.
  \item To design and implement a novel online learning algorithm with large input factors capable of ensuring low response time.
  \item To modify the original algorithm using kernel trick and evaluate its performance.
  \item To compare performance of the most popular algorithms for volatility forecasting such as ARFIMA, GARCH and SV models.
  \item To assess the predictive capacity of the proposed model using the SPA test.
  \item To compare the performance of the most popular models and the proposed model using real financial data.
\end{enumerate}

\section{Results}

The contributions expected at the end of this thesis are:
\begin{itemize}
   \item Main financial volatility concepts will be explained as well
   as a detailed state of the art of volatility models that allow to
   others to introduce this topic easily.
   \item A well documented and implemented online algorithm for forecasting
   volatility which will also be useful for other forecasting
   problems.
\end{itemize}
The results expected at the end of this thesis are:
\begin{itemize}
  \item A set of volatility explanatory variables to create a descriptive input vector.
   \item A new online learning method for volatility forecasting with low response times.
    \item Partial results publicated in national and international conferences.
    \item Publication in a scientific journal.
\end{itemize}

\newpage
\section{Stylized facts}
\label{sec:stylizedfacts}

There are several known features exhibited by financial instruments called \textit{stylized facts}, which have been empirically studied and some of them have been documented only recently. The importance of \textit{stylized facts} is that they allow to improve volatility models showing that it is possible to model volatility dynamic.

\begin{description}
 \item[Dependence] Volatility is said to be a long-memory process because its autocorrelation function exhibits long-range dependence. The returns autocorrelation function is shown in figure \ref{fig:returnacf} and it shows how the correlations are significant even for very long lags, this imply that the dependence between events that are far apart diminishes very slowly with increasing time.
 Moreover, volatility tends to cluster, i.e. high volatility periods are generally followed by high volatility periods.

 \begin{figure}[h]
 \centering
 \includegraphics[scale=0.5]{plots/spy_returns_acf.pdf}
 % spy_returns_dist.pdf: 504x504 pixel, 72dpi, 17.78x17.78 cm, bb=0 0 504 504
 \caption{SPY returns ACF}
 \label{fig:returnacf}
\end{figure}

 \item[Returns distribution] The returns distribution has fat tails, i.e. the number of extreme events (either positive or negative) returns is larger than what is hypothesized by common data generation process (generally normal distribution assumption). In the markets, fat tails are an undesirable feature because of the additional risk they imply.
 In the figure \ref{fig:returndist} is shown the SPY returns distribution based on daily dates from the period 1st July 1998 to 4th April 2013. The distribution was compared against the normal distribution which clearly doesn't fit the data.

 \begin{figure}[h]
 \centering
 \includegraphics[scale=0.5]{plots/spy_returns_dist.pdf}
 % spy_returns_dist.pdf: 504x504 pixel, 72dpi, 17.78x17.78 cm, bb=0 0 504 504
 \caption{SPY returns distribution}
 \label{fig:returndist}
\end{figure}

 \item[Volume] refers to the level of trading activity in a market.
 \item[Calendar effects] days of week, weekend, holidays.
 \item[Intraday effects] time of the day.
 \item[Overnight effects]
\end{description}

%Autocorrelation
%Agregar gráficos de ACF de retornos y volatilities

### 2014-08-25 src/thesis_title.tex
adecuando template doctorado USM

\thispagestyle{empty}
\vspace{1.5in}
\begin{center}
{\textsc{Universidad T\'ecnica Federico Santa Mar\'ia}}\\
{\textsc{Departamento de Inform\'atica}}\\
\medskip
\includegraphics[scale=0.3]{img/logo_usm.eps}
\bigskip
\bigskip
\bigskip
\bigskip \\
%Titulo
{\textsc{\Large{{INSERT TITLE}}}}\\%

%\medskip
%{\textsc{\Large{{based on Coreset}}}}%
\bigskip
\bigskip
\bigskip
\bigskip
\bigskip

{\MakeUppercase{Dissertation}\\
\bigskip
\bigskip
Submitted in partial satisfaction of the requirements\\
for the degree of\\
\bigskip
\bigskip
\MakeUppercase{Doctor en Ingenier\'ia Inform\'atica}\\
\bigskip
\bigskip
by\\
\bigskip
{\textsc{\large{John Nash}}}\\
\bigskip
\bigskip
In Valpara\'iso, Chile}
\end{center}

\begin{center}
{\normalsize August 2014.}
\end{center}

\pagebreak

### 2014-10-29 src/1_Introduction.tex
Template CValles

\chapter{Introduction}\label{ch:introduction}

\vspace{0.5cm}

"An individual economic variable, views as a time series, can wander extensively
and yet some pairs of series may be expected to move so that they do not drift
far apart."-Robert F. Engle and Clive W.J. Granger~\cite{engle1987}.\\
In this introduction we present the scope, objectives, hypothesis and organization of this thesis.

\section{Scope of this Research}

Cointegration concept was introduced by~\cite{engle1987} and implies that one or
more linear combinations of non-stationary variables are stationary even though
individually they are not.  Moreover ~\cite{stock+watson1988} observed that
cointegration reflects the common stochastic trends providing a useful way to
understand cointegration relationships. These common stochastic trends can be
also interpreted as a long-run equilibrium relationships.

This idea was inmediatly adopted in finance since it could represent their
long-run relationship implied by economic theory~\cite{laietAl1991},
\cite{lence+falk2005}.  Economic theory suggest that economic time series are
mean-reverting process and therefore, it reflects the idea of that some set of
variables cannot wander too far from each other.

On the other hand, the efficient markets hypothesis, also known as the random
walk theory states that current stock prices fully reflect available
information related to its value and there is no way to earn excess
profits~\cite{fama1970}.  This means that if we have stock prices from a
jointly efficient market, they cannot be
cointegrated~\cite{granger1986},\cite{dwyer1992}. However, \cite{richards1995}
claims that cointegration is directly at odds with market efficiency, even
though, there is no evidende that cointegration among asset prices have
implications about market efficiency~\cite{lence+falk2005}.

Despite the fact that cointegration on closing daily rates of currency pairs has
not been found \cite{coleman1990}, \cite{copeland1991}, different time series
frequencies can have different behaviours~\cite{aldridge2009}. Pair trading is a
very common example of cointegration application~\cite{herlemont2003} 
but cointegration can also be extended to a larger set of
variables~\cite{mukherjee1995},\cite{engle2004}.

%So why do we care about cointegration? Someone else can probably give more
%econometric applications, but in quantitative finance, cointegration forms the
%basis of the pairs trading strategy: suppose we have two cointegrated stocks X
%and Y, with the particular (for concreteness) cointegrating relationship X - 2Y
%= Z, where Z is a stationary series of zero mean. For example, X could be
%McDonald's, Y could be Burger King, and the cointegration relationship would
%mean that X tends to be priced twice as high as Y, so that when X is more than
%twice the price of Y, we expect X to move down or Y to move up in the near
%future (and analogously, if X is less than twice the price of Y, we expect X to
%move up or Y to move down). This suggests the following trading strategy: if X -
%2Y > d, for some positive threshold d, then we should sell X and buy Y (since we
%expect X to decrease in price and Y to increase), and similarly, if X - 2Y < -d,
%then we should buy X and sell Y.

Vector error correction model (VECM) introduces this long-run relationship
among a set of cointegrated variables as an error correction term. VECM is a
special case of the vector autorregresive model (VAR) model. VAR model
expresses future values as a linear combination of variables past values.
However, VAR model cannot be used with non-stationary variables. VECM is a
linear model but in terms of variable differences. If cointegration exists,
variable differences are stationary and they introduce an error correction term
which adjusts coefficients to bring the variables back to equilibrium. In
finance, many economic time series turn to be stationary when they are
differentiated and cointegration restrictions often improves
forecasting~\cite{duy1998}. Therefore, VECM has been widely adopted.

Both VECM and VAR model parameters are obtained using ordinary least squares
(OLS) method. OLS has two main problems: is sensitive to errors on input data
and involves many calculations. The former problem is commonly solved using
Ridge Regression (RR) \cite{hoerl1970} which introduces a regularization
parameter that leads to an unbiased estimation with better generalization
capability. The second problem of computational complexity depends on the number
of past values and observations considered.  Recently, online learning
algorithms have been proposed to solve problems with large data sets because of
their simplicity and their ability to update the model when new data is
available.

In this thesis, we propose an online formulation of the VECM called Onlive VECM
(OVECM) based on consideration of only a sliding window of the most recent data.
The algorithm introduces matrix optimizations in order to reduce the number of
operations and also takes into account the fact that cointegration vector space
doesn't experience large changes with small changes in the input data. Moreover,
RR instead of OLS is used to solve VECM. Our method is later tested using four
currency rates from the foreign exchange market with different frequencies.  Model
efectiveness is focused on out-of-sample forecast rather than in-sample fitting.
This criteria allows the OVECM prediction capability to be expressed rather than
just explaining data history. Our method performance is compared with its
optimal offline algorithm.

\section{Research Objectives}
The main motivation for this research is ...

The specific objectives of this research are,
\begin{itemize}
\item $\Large \mathcal{O}_1$: \emph{Especific Objective 1.}
\item $\Large \mathcal{O}_2$: \emph{Especific Objective 2.}
\item $\Large \mathcal{O}_3$: \emph{Especific Objective 3.}
\item $\Large \mathcal{O}_4$: \emph{Especific Objective 4.}
\item $\Large \mathcal{O}_5$: \emph{Especific Objective 5.}
\end{itemize}

\section{Research Hypothesis}

The main research hypothesis explored in this dissertation is the following \\

\textit{Hypothesis llll . }

\section{Organization of this Thesis}

Chapter 1 contains ...

### 2014-10-29 src/firmas.tex
Template CValles

\thispagestyle{empty}
\cleardoublepage%
\bigskip
\bigskip
\bigskip
\bigskip
\bigskip
\vspace{1cm}
\begin{center}
\includegraphics[scale=0.6]{fig/logo_usm}\\
\cambiaFont{ppl}{\Large \textsc{Examining Committee}}\\
\bigskip
\end{center}
\vspace{2cm}
\begin{tabular}[c]{lc}
{\textsc Prof. Dr. Luis Salinas} & \begin{picture}(150,5) \line(1,0){150} \end{picture}\\
{Universidad Técnica Federico Santa María} & {Advisor}\\
Chile & \\
& \\
& \\
{\textsc Prof. Dr. } &  \begin{picture}(150,5) \line(1,0){150} \end{picture}\\
{Universidad Técnica Federico Santa María} & {Committee Member} \\
Chile & \\
& \\
& \\
{Prof. Dr. } & \begin{picture}(150,5) \line(1,0){150} \end{picture}\\
{University} & {External Peer Reviewer}\\
{Chile} & \\
\end{tabular}

\cleardoublepage%

### 2014-10-29 src/0_Abstract.tex
adding chapters

\chapter*{Abstract}

\emph{Financial time series are known for their non-stationary behaviour. However,
sometimes they exhibit some stationary linear combinations. When this happens,
it is said that those time series are cointegrated. Vector error correction
model (VECM) is an econometric model which characterizes the joint dynamic
behaviour of a set of cointegrated variables in terms of forces pulling towards
equilibrium.}

\emph{VECM parameters are obtained using the ordinary least squares (OLS)
method.  Even though OLS is extensively used, it has over-fitting
issues and is computationally expensive. Ridge regression is commonly used
instead of OLS since they include a regularization parameter which could improve
prediction error.}

\emph{In this study, we propose an online VECM which optimizes how model parameters
are obtained by using only relevant past data and RR instead of OLS.  Our
proposal also takes advantage of the long-run relationship between the time
series in order to obtain improved execution times. Our proposed method is
tested using synthetic data and foreign exchange rates.
Results show that our proposal enables execution time to be reduced by at least
50\% while maintaining good accuracy levels.}

\vspace{0.8cm}

\emph{\textbf{Keywords:} Computational finance  - Cointegration  - Time Series - 
Online algorithms - Resampling  - Regression}

### 2014-10-29 src/3_TimeSeries.tex
adding chapters

--- 2014-04-22 src/introduction.tex
@@
-\chapter{Introduction}
+\chapter{Time Series}
+\section{Implied volatility models}
 
-The computational complexity of machine learning algorithms has become a limiting factor for problems that require processing large volumes of data and where the response time is crucial.
-An example of this is financial volatility time series forecast.
+\section{Black-Scholes formula}
 
-Currently, it is known that volatility is a process where many factors influence its behaviour.
-However, many models of volatility made forecasts based only on price returns.
-Examples are the ARFIMA, GARCH and stochastic volatility models.
-Machines learning methods have allowed a greater number of explanatory variables to be included, although in many cases, these increase calculation time.
+In finance, an option is a derivative, that is, a contract which gives the owner the right, but not the obligation to buy or sell an underlying asset at a given price called strike price.
+An option can be executed at any time before an expiration date previously defined no matter what price the underlying asset has.
 
-Algorithms that process large amounts of data in a short time and are able to detect changes in the financial market are required to achieve fast and accurate responses.
-Online learning algorithms have been developed to solve large-scale problems since they process an instance at a time, updating the model at each step incrementally.
-This is opposed to the batch algorithms where a phase of training that uses a large collection of historical data is first required and the resulting model is later used to predict.
-This thesis proposal is to design and develop an online learning algorithm that meets or improves the performance of current volatility algorithms and ensures rapid responses to large-scale problems.
-Model comparisons will be made by the SPA (superior predictive ability) test proposed by Hansen and
-Lunde~\cite{hansen+lunde2006}  that allows comparing of forecast models giving a relative measure of model performance with respect to one or more models.
+The Black-Scholes formula~\cite{black1973}, developed in the early 1970's, Myron Scholes, Robert Merton and Fisher Black,  allows to determine an option value $V$ based on the underlying asset price $S(t)$ at a time $t$ and the following constant parameters:
 
-%Keywords: Volatility, Machine learning, Online learning, large-scale learning
+\begin{description}
+\item [$\sigma$:] underlying asset price volatility which measures the standard deviation of the returns
+\item [$\mu$:] underlying asset drift which is a measure of the average rate of growth of the stock
+\item[$E$:] option strike or excersice price
+\item[$T$:] option date of expiry
+\item[$t$:] current time
+\item[$r$:] risk-free interest rate
+\end{description}
 
-\section{The problem}
+\subsection{Stock price model}
 
-In financial markets, volatility is one of the key elements to model the stochastic dynamic behaviour of financial assets.
-This is mainly because volatility gives a measure of uncertainty about future returns and it is often viewed as an indicator of the vulnerability of financial markets.
-It is also important as an input parameter in problems like derivative pricing, hedging and portfolio management.
-For example, in option pricing we need to known first the volatility of the underlying asset before pricing an option.
+The Black-Scholes model assume that the underlying price $S$ follows a lognormal random walk:
 
-Besides, volatility provides important information for studying asset returns because the latter are largely uncorrelated and nonlinearly dependent.
-However, it has been found that volatility exhibits significant autocorrelation and predictable patterns~\cite{poon+granger2003} and  many models have been proposed to forecast its behaviour (see section ~\ref{sec:stylizedfacts}).
+\begin{equation}\label{eq:stockprice}
+dS = \mu S dt + \sigma S dB
+\end{equation}
 
-The volatility of the data is due to a large number of factors affecting the market, with some of them directly measurable such as historical prices, trends, supply and demand.
-However, others such as monetary policies and news are not directly measurable and are included in other models out of the scope of this thesis.
-Figure~\ref{fig:VolFactorNews} shows the VIX index, the daily US implied stock market volatility, commonly known as the "financial fear factor" which is clearly affected by historical events.
+\noindent where $B$ is a Brownian motion. This stochastic differential equation has two components: a deterministic term given by $\mu S dt$ and a random term given by $\sigma S dX$.
 
-\begin{figure}[h]
- \centering
- \includegraphics[width=0.7\textwidth]{img/vol-factor-thenews.jpg}
- % vol-factor-thenews.jpg: 710x506 pixel, 72dpi, 25.05x17.85 cm, bb=0 0 710 506
- \caption{The news as volatility factor}
- \label{fig:VolFactorNews}
-\end{figure}
+\subsection{Brownian motion}
 
-Volatility models have been extensively studied~\cite{gatheral2006,poon+granger2003,knight2002} and some of  them include machine learning methods~\cite{hamidetal2004,donaldsonetal1997,shiyietal2008,shiyietal2010,gavrishchaka2006,vasilios2012}.
-The stochastic behaviour of volatility and the large number of factors affecting it result in models with a large number of parameters and slow response time due to the computationally expensive routines.
-These expensive calculations could be avoided using online learning algorithms, which allow response time reduction and are therefore suitable for high frequency trading applications.
+A brownian motion $B$ (also called a Wiener process) is a stochastic process characterized by the three following properties:
 
-In this thesis, a model will be proposed based on online learning algorithms to forecast volatility in high frequency financial time series.
-The main problems to be deal with in this thesis are:
+\begin{description}
+\item[Continuity:] $B(t)$ is a continuos function
+\item[Normal increments:]  $B(t)-B(s)$ has a normal distribution with mean $0$ and variance $t-s$.
+\item[Independence of increments:] for every choice of nonnegative real numbers $0 \leq s_1 <  t_1 \leq \cdots \leq s_n < t_n < \infty$, the increment random variables $W_{t_1} - W_{s_1}, \cdots, W_{t_n} - W_{s_n}$ are jointly independent.
+\end{description}
 
-\begin{itemize}
-   \item Although volatility is often predictable~\cite{poon+granger2003}, there are non-linear features~\cite{gatheral2006} to be considered in this proposal.
-    \item The number of factors affecting volatility has been studied~\cite{srinivasan2009} and some other unknown factors affect volatility and they should also be included.
-    \item The response time needs to be very fast and in order to ensure this, the proposed model will be presented as an online machine learning algorithm. 
-\end{itemize}
+\subsection{Stochastic differential equation}
+
+An stochastic differential integral has the form:
+
+\begin{equation}
+W(T)=\int_0^T f(t)dB(t) \, .
+\end{equation}
+
+\noindent This equations is also expressed in an abbreviate form:
+
+\begin{equation}
+dW = f(t) dB \, .
+\end{equation}
+
+\noindent Therefore, the integral form of the stock price model shown in equation (\ref{eq:stockprice}) is:
+
+\begin{equation}
+S(T)=\int_0^T \mu S(t) dt + \int_0^T \sigma S(t) dB(t)
+\end{equation}
+
+\subsection{Ito's Lemma}
+
+Ito's lemma is used to find the differential of a time dependent function of a stochastic process.
+In option pricing we need to find the option price $V(S(t))$ which depends on a stochastic stock price model $S(t)$.
+$V(S,t)$ is required to be  differentiable function of $S$ and once differentiable function of $t$.
+
+%These are known or can be easily obtained from the market, excepting by volatility which must be estimated.
+However, rather than assuming a volatility a priori and computing option prices from it, the model can be used to estimate volatility at given prices, time to expiration and strike price.
+This obtained volatility is called the implied volatility of an option.
+Additionally, some models obtain implied volatility from futures (other derivative from prices).
+
+\section{Realized Volatility Models}
+
+\section{GARCH}
+test
+\section{ARFIMA}
+\section{Stochastic Volatility}
 
@@
 
-\subsection{Online learning}
-
-Classic statistical theory of sequential prediction enforces strong assumptions on the statistical properties of the input sequence (for example, stationary stochastic process).
-However, these assumptions can be unknown or change over time.
-In online learning there is no previous assumption about the data and the sequence is allowed to be deterministic, stochastic or even adaptive.
-
-Moreover, in case we receive data streams, ANN or SVM cannot introduce new information into the model without a re-training process, so we will have to use the same non-updated model until we decide to compute another one if it is possible.
-Online learning algorithms allow one example at a time to be introduced into an existing model incrementally~\cite{vovk2005}.
-This is extremely important when the problem has large data streams and real-time forecasting must be done.
-This is the most common scenario when we want to forecast a wide range of data such as stock prices and volatilities, electricity power, intrusion detection, web-mining, server load, etc.
-Besides, many problems of high interest in machine learning can be treated as online ones and they can also use these types of algorithms.
-
-The online learning framework was first introduced in the perceptron algorithm~\cite{rosenblatt58}.
-There are several other widely used online methods such as passive-aggressive~\cite{crammerETall2006}, stochastic gradient descent~\cite{zhang2004}, aggregating algorithm~\cite{vovk2001} and the second order perceptron~\cite{cesa-bianchi2005}.
-In~\cite{cesa-bianchi2006} an in-depth analysis of online learning is provided.
-
-The motivation for online learning is to obtain computational efficiency and tackle the shifting problem, i.e.
-that the distribution of the data is unknown or changes over time.
-Online learning algorithms can deal with this problem because they have a tracking ability which is a strategy based on retaining weak dependence on past examples by using two types of models:
-
-\textit{a)} \textbf{memory boundedness:} consists of limiting the number of support vectors in order to improve computational efficiency.
-One example of this is the budget perceptron~\cite{crammeretal2004} which reduces the number of examples used for prediction.
-Alternatively, in the forgetron algorithm~\cite{dekeletal2008} the damage caused by removing old examples is discussed, which can be avoided by removing samples with small influences.
-Other examples are the sliding window kernel (RLS)~\cite{vanvaerenberghetal2006}, which only considers a sliding window of the most recent data, and in \cite{arce+salinas2012} is shown a variant of aggregating algorithm for regression~\cite{vovk2001} considering only a sliding window of the most recent data, optimising also common matrix operations.
-
-\textit{b)} \textbf{weight decay:} one example of this is the shifting perceptron algorithm which implements an exponential decaying scheme for the examples~\cite{cavallantietal2007}.
-Performance of an online learning algorithm is measured by the cumulative loss it suffers along its run on a sequence of examples.
-In order to minimize this loss, the learner may update the hypothesis after each round so as to be more accurate in later rounds.
-
-\section{Thesis proposal}
-
-This thesis proposal is to build an algorithm to forecast realized volatility in financial markets considering not only intraday returns but also other explanatory volatility variables.
-Because of the amount of high frequency data to be considered and short response times required in financial applications, the algorithm will be formulated in online mode.
-The online algorithm will also allow the shifting problem to be tackled while being adaptive to wild changes in the volatility process.
-The online learning model will be formulated considering the following:
-
-\begin{description}%[leftmargin=0.4cm]\itemsep4pt
-    \item[large volumes of data:] high frequency data will be used to estimate realized volatility.
-The frequency to be considered will be 10 to 15 minutes.
-This frequency will be analysed using signal processing theory.
-    \item[non-stationary data:] volatility is a time series known because of its non-stationary feature, so statistics will be studied before using the model.
-    \item[stylized facts:] there are several well-documented features about financial time series volatility called stylized facts~\cite{poon+granger2003} that can be included in the model.
-    \item[low response time:] the model has to be simple but powerful in order to achieve the response time needed.
-If the model is extended to real time data, ticks can arrive at rates of milliseconds and the model should respond at the same rate or less.
-    \item[financial data:] the model can be constructed using data of stocks or different derivatives such as options and futures. The chosen implicit volatility model will depend on the data type.
-    \item[input vector:] currently, many models are only based on expected returns, but volatility is a complex variable which depends also on many factors studied in the literature such as trading volume, the tick rate, time of the day and the behaviour of the market as a hole~\cite{gatheral2006}.
-All these factors can improve the design of the proposed online algorithm.
-    \item[kernel version:] the online algorithm will be later kernelised in order to map the input data into a high dimensional feature space (kernel trick) which will provide better properties to design the solution.
-\end{description}
-
-The model proposed has to be compared with the current popular financial models of financial volatility such as: ARFIMA, GARCH, SV models and machine learning models.
-Performance will be measured using the realized volatility shown in equation~(\ref{eq:srv}) as a benchmark.
-
-In order to compare different models, apart from using the standard performance measurements like MAE and MSE, the superior predictive ability (SPA) methodology proposed by Hansen and Lunde~\cite{hansen+lunde2006} will also be studied.
-SPA is a test to compare two or more forecasting models at the same time giving a relative performance measurement of a base model in comparison with alternative models.
-
-The model will be compared against traditional financial volatility models such as Black-Scholes, SV, GARCH and ARFIMA using different forecast errors measurements.
-Although the best way to measure forecast errors should be the investor returns, this implies knowing how the model information is used as a strategy, so this information is not going to be included in this thesis.
-Therefore, in order to compare the different volatility forecast models, popular evaluation measurements will be used.
-In the literature the following are commonly used: {\em Mean Error} (ME), {\em Mean Square Error} (MSE), {\em Mean Absolute Error} (MAE) and {\em Mean Absolute Percent Error} {MAPE} and others are less commonly used such as LINEX loss function which weight differently positive and negative errors.
-In ~\cite{granger1999} Granger describes a variety of non-symmetric loss function including LINEX.
-The superior predictive ability (SPA) methodology proposed by Hansen and Lunde~\cite{hansen+lunde2006} will also be studied
-which allows comparison of two or more forecasting models at the same time giving a relative performance measurement of a base model compared with alternative models.
-All these measurements will be out-of-sample in order to show their predictive power.
-
-\section{Hypotheses}
-
-Volatility forecasting has been a largely studied and different approaches have arisen to describe its stochastic behaviour.
-It is well-known that volatility depends on many factors, but most of the existing models only consider a few of them.
-In order to introduce more factors easily, classic machine learning models have been studied and their performance have shown to be successfull in comparison with most used volatility models.
-However, their training process is computationally expensive and it is not possible to update the model with new market information before new information arrives.
-Online learning methods can also include more factors into a model, but they process one example at a time to update an existing model incrementally without having a training phase, thus they improve response times and are more suitable for financial applications.
-
-The thesis hypotheses are the following:
-
-\begin{enumerate} 
-\item Machine learning based models that consider more volatility explanatory factors will adapt and perform better than popular volatility forecasting models based only on returns or a few factors.
-\item Stylized facts can be efficiently introduced as explanatory factors and improve model performance.
-\item The use of online learning methods will allow inclusion of as many variables as needed to ensure low response times.  
-\end{enumerate}
-
-\section{General Objective}
-To design and implement an online learning model for volatility forecasting capable of adapting to their non-stationary behaviour using volatility factors as model input ensuring low response times.  
-\section{Specific Objectives}
-The specific objectives to be followed in this thesis are:
-\begin{enumerate}
-  \item To determine the best volatility factors based on their stylized facts and their performance in machine learning models.
-  \item To determine the best sampling of high frequency time series.
-  \item To define the best realized volatility proxy based on the literature and experimental results.
-  \item To design and implement a novel online learning algorithm with large input factors capable of ensuring low response time.
-  \item To modify the original algorithm using kernel trick and evaluate its performance.
-  \item To compare performance of the most popular algorithms for volatility forecasting such as ARFIMA, GARCH and SV models.
-  \item To assess the predictive capacity of the proposed model using the SPA test.
-  \item To compare the performance of the most popular models and the proposed model using real financial data.
-\end{enumerate}
-
-\section{Results}
-
-The contributions expected at the end of this thesis are:
-\begin{itemize}
-   \item Main financial volatility concepts will be explained as well
-   as a detailed state of the art of volatility models that allow to
-   others to introduce this topic easily.
-   \item A well documented and implemented online algorithm for forecasting
-   volatility which will also be useful for other forecasting
-   problems.
-\end{itemize}
-The results expected at the end of this thesis are:
-\begin{itemize}
-  \item A set of volatility explanatory variables to create a descriptive input vector.
-   \item A new online learning method for volatility forecasting with low response times.
-    \item Partial results publicated in national and international conferences.
-    \item Publication in a scientific journal.
-\end{itemize}
-
-\newpage
 \section{Stylized facts}
@@
 \end{description}
-
-%Autocorrelation
-%Agregar gráficos de ACF de retornos y volatilities

### 2014-10-29 src/4_OnlineLearning.tex
adding chapters

--- 2014-04-22 src/introduction.tex
@@
-\chapter{Introduction}
-
-The computational complexity of machine learning algorithms has become a limiting factor for problems that require processing large volumes of data and where the response time is crucial.
-An example of this is financial volatility time series forecast.
-
-Currently, it is known that volatility is a process where many factors influence its behaviour.
-However, many models of volatility made forecasts based only on price returns.
-Examples are the ARFIMA, GARCH and stochastic volatility models.
-Machines learning methods have allowed a greater number of explanatory variables to be included, although in many cases, these increase calculation time.
-
-Algorithms that process large amounts of data in a short time and are able to detect changes in the financial market are required to achieve fast and accurate responses.
-Online learning algorithms have been developed to solve large-scale problems since they process an instance at a time, updating the model at each step incrementally.
-This is opposed to the batch algorithms where a phase of training that uses a large collection of historical data is first required and the resulting model is later used to predict.
-This thesis proposal is to design and develop an online learning algorithm that meets or improves the performance of current volatility algorithms and ensures rapid responses to large-scale problems.
-Model comparisons will be made by the SPA (superior predictive ability) test proposed by Hansen and
-Lunde~\cite{hansen+lunde2006}  that allows comparing of forecast models giving a relative measure of model performance with respect to one or more models.
-
-%Keywords: Volatility, Machine learning, Online learning, large-scale learning
-
-\section{The problem}
-
-In financial markets, volatility is one of the key elements to model the stochastic dynamic behaviour of financial assets.
-This is mainly because volatility gives a measure of uncertainty about future returns and it is often viewed as an indicator of the vulnerability of financial markets.
-It is also important as an input parameter in problems like derivative pricing, hedging and portfolio management.
-For example, in option pricing we need to known first the volatility of the underlying asset before pricing an option.
-
-Besides, volatility provides important information for studying asset returns because the latter are largely uncorrelated and nonlinearly dependent.
-However, it has been found that volatility exhibits significant autocorrelation and predictable patterns~\cite{poon+granger2003} and  many models have been proposed to forecast its behaviour (see section ~\ref{sec:stylizedfacts}).
-
-The volatility of the data is due to a large number of factors affecting the market, with some of them directly measurable such as historical prices, trends, supply and demand.
-However, others such as monetary policies and news are not directly measurable and are included in other models out of the scope of this thesis.
-Figure~\ref{fig:VolFactorNews} shows the VIX index, the daily US implied stock market volatility, commonly known as the "financial fear factor" which is clearly affected by historical events.
-
-\begin{figure}[h]
- \centering
- \includegraphics[width=0.7\textwidth]{img/vol-factor-thenews.jpg}
- % vol-factor-thenews.jpg: 710x506 pixel, 72dpi, 25.05x17.85 cm, bb=0 0 710 506
- \caption{The news as volatility factor}
- \label{fig:VolFactorNews}
-\end{figure}
-
-Volatility models have been extensively studied~\cite{gatheral2006,poon+granger2003,knight2002} and some of  them include machine learning methods~\cite{hamidetal2004,donaldsonetal1997,shiyietal2008,shiyietal2010,gavrishchaka2006,vasilios2012}.
-The stochastic behaviour of volatility and the large number of factors affecting it result in models with a large number of parameters and slow response time due to the computationally expensive routines.
-These expensive calculations could be avoided using online learning algorithms, which allow response time reduction and are therefore suitable for high frequency trading applications.
-
-In this thesis, a model will be proposed based on online learning algorithms to forecast volatility in high frequency financial time series.
-The main problems to be deal with in this thesis are:
-
-\begin{itemize}
-   \item Although volatility is often predictable~\cite{poon+granger2003}, there are non-linear features~\cite{gatheral2006} to be considered in this proposal.
-    \item The number of factors affecting volatility has been studied~\cite{srinivasan2009} and some other unknown factors affect volatility and they should also be included.
-    \item The response time needs to be very fast and in order to ensure this, the proposed model will be presented as an online machine learning algorithm. 
-\end{itemize}
-
-\section{Volatility definition}
-
-\begin{itemize}
-
-\item Is the size of the price movement.
-
-\item Variance is a measure of distribution of returns and is not neccesarily bound by any time period.
-Volatility is a measure of the standard deviation (square root of the variance) over a certain time interval.
-In finance, variance and volatility both gives you a sense of an asset's risk.
-Variance gives you a sense of the risk in the asset over its lifetime, while volatility gives you a sense of the movement of the asset in eg.
-the past month or the past year.
-
-\item The main underlying difference is in their definition.
-Variance has a fixed mathematical definition, however volatility does not as such.
-Volatility is said to be the measure of fluctuations of a process.
-
-Volatility is a subjective term, whereas variance is an objective term i.e.
-given the data you can definitely find the variance, while you can't find volatility just having the data.
-Volatility is associated with the process, and not with the data.
-
-In order to know the volatility you need to have an idea of the process i.e you need to have an observation of the dispersion of the process.
-All the different processes will have different methods to compute volatilities based on the underlying assumptions of the process.
-\end{itemize}
-
-\subsubsection{Types of volatility}
-
-The volatility of a stock is not directly observable~\cite{tsay2005,engle1993}.
-For example, daily volatility is not directly observable from only daily returns because there is only one observation in a trading day.
-If intraday data is available, then volatility could be estimated.
-However, intraday returns are not the only explanatory variables for volatility and several estimators have been proposed.
-These estimators are observable variables that are related to the latent variable of interest called volatility proxies~\cite{devilderetal2007}.
-Examples of volatility proxies are the following:
-
-\begin{description}%[leftmargin=0.4cm]\itemsep4pt
-\item[realized volatility:] is also known as historic volatility and
-it is the actual variance in the price of a stock over time.
-Realized volatility is measured in terms of the standard deviation
-using the historical stock prices. It is commonly calculated based on
-intraday price returns:
-
-\begin{equation}
-\label{eq:retintra}
-r_{t,n}=100(\ln(p_{t,n}) - \ln(p_{t,n-1}))
-\end{equation}
-
-\noindent where $p_{t,n}$ is the price observed at day $t=1,\dots,T$ and
-intraday sample $n=2,\dots,N$. Realized volatility is defined as:
-
-\begin{equation}
-\label{eq:rv}
-    \hat{\sigma}(t) = \sum_{n=1}^N r_{t,n}^2 \, , 
-\end{equation}
-
-\noindent where $N$ is the number of intraday samples and $T$ is the
-number of days.
-
-In order to include overnight returns, Hansen and
-Lunde~\cite{hansen+lunde2005} introduced a scaling version of
-realized volatility using the following definitions:
-
-\begin{eqnarray}
-r_{t}&=&100(\ln(p_{t,N}) - \ln(p_{t-1,N})) \label{eq: ret 1} \\
-\bar{\rho}(t) &=& \sum_{t=1}^T r_{t}^2  \label{eq: ret 2} \, .
-\end{eqnarray}
-
-\noindent where equation~(\ref{eq: ret 1}) represents overnight 
-returns and the volatility as 
-equation~(\ref{eq: ret 2}), where $p_{t,N}$ is the last intraday 
-sample at day $t$. The scaled realized volatility $\rho(t)$ is 
-defined as:
-
-\begin{eqnarray}
-\label{eq:srv}
-\rho(t) = \gamma \hat{\rho}(t) \, , \qquad & \qquad \gamma = \displaystyle \frac{\bar{\rho}(t)}{\displaystyle\sum_{t=1}^T \hat{\rho}(t)}
-\end{eqnarray}
-
-Realized volatility has also been defined as the absolute value return or as the mean of the sum of intraday squared returns at short intervals of time.
-The majority of research carried out in the literature obtain the daily volatility as the daily squared returns as is shown in equation~(\ref{eq: ret 2}).
-However, it has been proven that this measurement noise is too high for observing the true volatility process~\cite{andersen+bollerslev1998}.
-Hansen and Lunde~\cite{hansen+lunde2006} stated that the use of a noisy proxy could result in an inferior model being chosen as the best one.
-The realized volatility, as calculated by the cumulative sum of squared intraday returns and shown in equation (\ref{eq:rv}), is less noisy and doesn't lead to choosing an inferior model.
-
-\item[implied volatility:] volatility not only can be extracted from returns but it can also be derived from option or future pricing models.
-The volatility obtained corresponds to the market's prediction of future volatility.
-In finance, an option is a derivative, that is, a contract which gives the owner the right, but not the obligation to buy or sell an underlying asset at a given price called strike price.
-An option can be executed at any time before an expiration date previously defined no matter what price the underlying asset has.
-For example, the Black-Scholes model~\cite{black1973} determines the fair option value based on stock price, strike price, time to option expiration, the interest rate and volatility.
-These are known or can be easily obtained from the market, excepting by volatility which must be estimated.
-However, rather than assuming a volatility a priori and computing option prices from it, the model can be used to estimate volatility at given prices, time to expiration and strike
-price.
-This obtained volatility is called the implied volatility of an option.
-Additionally, some models obtain implied volatility from futures (other derivative from prices).
-For instance, the Barone-Adesi and Whaley futures option model~\cite{baroneetal1987} is also used to determine future volatilities~\cite{hamidetal2004}.
-Higher implied volatility is indicative of greater price fluctuation in either direction.
-Implied volatility is found by determining the value which makes theoretical prices equal to market prices.
-In this way volatility is ``implied'' by the current market price of the stock.
-
-\end{description}
-
-For trading strategies, the interest is centred in forecasting realized volatility over the life of an option and to take advantage when this volatility differs from the implied volatility.
-This is called volatility arbitrage.
-For example, a trader will buy an option and hedge the underlying asset if the implied volatility is under the realized volatility.
-
-\section{Volatility methods}
-
-In the existing literature, there are four main classes of asset
-return volatility models: the general autoregressive conditional
-heteroskedasticity (GARCH) models, the stochastic volatility (SV)
-models, the realized volatility models and the machine learning based
-models. A comparison of the first three models can be found
-in~\cite{wei2012}.
-
-For many years the most popular methods for estimating financial
-volatility were the autoregressive conditional heteroskedasticity
-(ARCH) models~\cite{engle1982} and the general ARCH (GARCH)
-models~\cite{bollerslev1986}. For instance, the GARCH(1,1) defines
-returns $y_t$ and volatility $\sigma_t$ as:
-
-\begin{eqnarray*}
-    y_t &=& \sigma_t \epsilon_t \\
-     \sigma_t^2 &=& \alpha_0 + \alpha_1 \epsilon_{t-1}^2 + \beta_1
-     \sigma_{t-1}^2
-\end{eqnarray*}
-
-\noindent where $\epsilon_t$ is standard Gaussian white noise,
-$\alpha_0,\alpha_1,\beta_1 \geq 0$ are required to ensure that the
-variance will never be negative and $\alpha_1+\beta_1 <1$ is needed to
-guarantee a weakly stationary process~\cite{nelson1990}.
-
-Since the introduction of the GARCH models, several extensions have been proposed, but none of them seems to beat the GARCH(1,1) model~\cite{lunde+hansen2005}.
-Despite its popularity, GARCH models have several limitations: firstly, a time series model may be non-linear in mean and/or non-linear in variance, but ARCH and GARCH models are non-linear in variance, but not in mean.
-Besides, GARCH models often fail to capture highly irregular phenomena, like wild market fluctuations.
-
-SV models explain how volatility varies in a random fashion.
-These models are useful because they explain why options with different strikes and expirations dates have different Black-Scholes implied volatilities, phenomenon known as the volatility smile.
-This is useful because the Black-Scholes model assumes that the volatility of the underlying asset is constant which is not always true.
-There are several SV models and the most well-known and popular is the Heston model~\cite{heston1993}.
-Additional information about SV models can be found in~\cite{shephard1995}.
-
-The realized volatility constructed from high frequency intraday returns gave rise to the realized volatility models mainly because the realized volatility series is much more homoskedastic and seems to be a long memory process~\cite{andersonetal2003}.
-For realized volatility, the autoregressive fractionally integrated moving average (ARFIMA) process emerged as a standard model~\cite{chenetal2010} and many variations have been studied, but all of them produce similar forecasting results to the ARFIMA(1,d,1) model~\cite{koopmanetal2005}.
-
-On the other hand, machine learning based models, especially artificial neural networks (ANN) and support vector machines (SVM) have arisen as an alternative to forecast volatility.
-ANN is a statistical technique inspired by biological neural networks which is capable of changing its structure based on external or internal information during a training phase~\cite{sammut2011}.
-SVM are supervised learning models for classification analysis which recognize patterns finding a separating hyperplane.
-An extension for regression analysis is known as support vector regression (SVR).
-
-Since machine learning models and in particular ANN do not require assumptions about the data (gaussianity for example) and allow more explanatory variables than returns to be included, they have become widely used in solving financial problems, specially volatility forecasting~\cite{hamidetal2004,donaldsonetal1997}.
-There are also many works focused on the using of SVM in volatility forecasting~\cite{shiyietal2008,shiyietal2010,gavrishchaka2006,vasilios2012}.
-
-However, just as with ANN, SVMs have scalability problems because their training process is computationally intensive and it is done in batch mode.
-The scalability problem worsens when new additional training data is available and a re-training process from scratch needs to be done.
-This problem can be avoided using online machine learning algorithms that allow one instance at a time to be processed with low computationally expensive calculations.
-
+\chapter{Machine Learning Models}
+\section{Batch learning}
+\section{Online learning}
 \subsection{Online learning}
@@
 In order to minimize this loss, the learner may update the hypothesis after each round so as to be more accurate in later rounds.
-
-\section{Thesis proposal}
-
-This thesis proposal is to build an algorithm to forecast realized volatility in financial markets considering not only intraday returns but also other explanatory volatility variables.
-Because of the amount of high frequency data to be considered and short response times required in financial applications, the algorithm will be formulated in online mode.
-The online algorithm will also allow the shifting problem to be tackled while being adaptive to wild changes in the volatility process.
-The online learning model will be formulated considering the following:
-
-\begin{description}%[leftmargin=0.4cm]\itemsep4pt
-    \item[large volumes of data:] high frequency data will be used to estimate realized volatility.
-The frequency to be considered will be 10 to 15 minutes.
-This frequency will be analysed using signal processing theory.
-    \item[non-stationary data:] volatility is a time series known because of its non-stationary feature, so statistics will be studied before using the model.
-    \item[stylized facts:] there are several well-documented features about financial time series volatility called stylized facts~\cite{poon+granger2003} that can be included in the model.
-    \item[low response time:] the model has to be simple but powerful in order to achieve the response time needed.
-If the model is extended to real time data, ticks can arrive at rates of milliseconds and the model should respond at the same rate or less.
-    \item[financial data:] the model can be constructed using data of stocks or different derivatives such as options and futures. The chosen implicit volatility model will depend on the data type.
-    \item[input vector:] currently, many models are only based on expected returns, but volatility is a complex variable which depends also on many factors studied in the literature such as trading volume, the tick rate, time of the day and the behaviour of the market as a hole~\cite{gatheral2006}.
-All these factors can improve the design of the proposed online algorithm.
-    \item[kernel version:] the online algorithm will be later kernelised in order to map the input data into a high dimensional feature space (kernel trick) which will provide better properties to design the solution.
-\end{description}
-
-The model proposed has to be compared with the current popular financial models of financial volatility such as: ARFIMA, GARCH, SV models and machine learning models.
-Performance will be measured using the realized volatility shown in equation~(\ref{eq:srv}) as a benchmark.
-
-In order to compare different models, apart from using the standard performance measurements like MAE and MSE, the superior predictive ability (SPA) methodology proposed by Hansen and Lunde~\cite{hansen+lunde2006} will also be studied.
-SPA is a test to compare two or more forecasting models at the same time giving a relative performance measurement of a base model in comparison with alternative models.
-
-The model will be compared against traditional financial volatility models such as Black-Scholes, SV, GARCH and ARFIMA using different forecast errors measurements.
-Although the best way to measure forecast errors should be the investor returns, this implies knowing how the model information is used as a strategy, so this information is not going to be included in this thesis.
-Therefore, in order to compare the different volatility forecast models, popular evaluation measurements will be used.
-In the literature the following are commonly used: {\em Mean Error} (ME), {\em Mean Square Error} (MSE), {\em Mean Absolute Error} (MAE) and {\em Mean Absolute Percent Error} {MAPE} and others are less commonly used such as LINEX loss function which weight differently positive and negative errors.
-In ~\cite{granger1999} Granger describes a variety of non-symmetric loss function including LINEX.
-The superior predictive ability (SPA) methodology proposed by Hansen and Lunde~\cite{hansen+lunde2006} will also be studied
-which allows comparison of two or more forecasting models at the same time giving a relative performance measurement of a base model compared with alternative models.
-All these measurements will be out-of-sample in order to show their predictive power.
-
-\section{Hypotheses}
-
-Volatility forecasting has been a largely studied and different approaches have arisen to describe its stochastic behaviour.
-It is well-known that volatility depends on many factors, but most of the existing models only consider a few of them.
-In order to introduce more factors easily, classic machine learning models have been studied and their performance have shown to be successfull in comparison with most used volatility models.
-However, their training process is computationally expensive and it is not possible to update the model with new market information before new information arrives.
-Online learning methods can also include more factors into a model, but they process one example at a time to update an existing model incrementally without having a training phase, thus they improve response times and are more suitable for financial applications.
-
-The thesis hypotheses are the following:
-
-\begin{enumerate} 
-\item Machine learning based models that consider more volatility explanatory factors will adapt and perform better than popular volatility forecasting models based only on returns or a few factors.
-\item Stylized facts can be efficiently introduced as explanatory factors and improve model performance.
-\item The use of online learning methods will allow inclusion of as many variables as needed to ensure low response times.  
-\end{enumerate}
-
-\section{General Objective}
-To design and implement an online learning model for volatility forecasting capable of adapting to their non-stationary behaviour using volatility factors as model input ensuring low response times.  
-\section{Specific Objectives}
-The specific objectives to be followed in this thesis are:
-\begin{enumerate}
-  \item To determine the best volatility factors based on their stylized facts and their performance in machine learning models.
-  \item To determine the best sampling of high frequency time series.
-  \item To define the best realized volatility proxy based on the literature and experimental results.
-  \item To design and implement a novel online learning algorithm with large input factors capable of ensuring low response time.
-  \item To modify the original algorithm using kernel trick and evaluate its performance.
-  \item To compare performance of the most popular algorithms for volatility forecasting such as ARFIMA, GARCH and SV models.
-  \item To assess the predictive capacity of the proposed model using the SPA test.
-  \item To compare the performance of the most popular models and the proposed model using real financial data.
-\end{enumerate}
-
-\section{Results}
-
-The contributions expected at the end of this thesis are:
-\begin{itemize}
-   \item Main financial volatility concepts will be explained as well
-   as a detailed state of the art of volatility models that allow to
-   others to introduce this topic easily.
-   \item A well documented and implemented online algorithm for forecasting
-   volatility which will also be useful for other forecasting
-   problems.
-\end{itemize}
-The results expected at the end of this thesis are:
-\begin{itemize}
-  \item A set of volatility explanatory variables to create a descriptive input vector.
-   \item A new online learning method for volatility forecasting with low response times.
-    \item Partial results publicated in national and international conferences.
-    \item Publication in a scientific journal.
-\end{itemize}
-
-\newpage
-\section{Stylized facts}
-\label{sec:stylizedfacts}
-
-There are several known features exhibited by financial instruments called \textit{stylized facts}, which have been empirically studied and some of them have been documented only recently.
-The importance of \textit{stylized facts} is that they allow to improve volatility models showing that it is possible to model volatility dynamic.
-
-\begin{description}
- \item[Dependence] Volatility is said to be a long-memory process because its autocorrelation function exhibits long-range dependence.
-The returns autocorrelation function is shown in figure \ref{fig:returnacf} and it shows how the correlations are significant even for very long lags, this imply that the dependence between events that are far apart diminishes very slowly with increasing time.
- Moreover, volatility tends to cluster, i.e. high volatility periods are generally followed by high volatility periods.
-
- \begin{figure}[h]
- \centering
- \includegraphics[scale=0.5]{plots/spy_returns_acf.pdf}
- % spy_returns_dist.pdf: 504x504 pixel, 72dpi, 17.78x17.78 cm, bb=0 0 504 504
- \caption{SPY returns ACF}
- \label{fig:returnacf}
-\end{figure}
-
- \item[Returns distribution] The returns distribution has fat tails, i.e.
-the number of extreme events (either positive or negative) returns is larger than what is hypothesized by common data generation process (generally normal distribution assumption).
-In the markets, fat tails are an undesirable feature because of the additional risk they imply.
- In the figure \ref{fig:returndist} is shown the SPY returns distribution based on daily dates from the period 1st July 1998 to 4th April 2013.
-The distribution was compared against the normal distribution which clearly doesn't fit the data.
-
- \begin{figure}[h]
- \centering
- \includegraphics[scale=0.5]{plots/spy_returns_dist.pdf}
- % spy_returns_dist.pdf: 504x504 pixel, 72dpi, 17.78x17.78 cm, bb=0 0 504 504
- \caption{SPY returns distribution}
- \label{fig:returndist}
-\end{figure}
-
- \item[Volume] refers to the level of trading activity in a market.
- \item[Calendar effects] days of week, weekend, holidays.
- \item[Intraday effects] time of the day.
- \item[Overnight effects]
-\end{description}
-
-%Autocorrelation
-%Agregar gráficos de ACF de retornos y volatilities

### 2014-10-29 src/5_Proposal.tex
adding chapters

\chapter{Online VECM proposal}
\section{\uppercase{Methodology}}
\label{sec:methodology}
\noindent 
\subsection{Online VECM} \label{sec:proposal}

Since VECM parameter estimation is computationally expensive, we propose an
online version of VECM (OVECM).  OVECM considers only a sliding window of the
most recent data. Moreover, since cointegration vectors represent long-run
relationships which vary little in time, OVECM determines firstly if they require calculation.

OVECM also implements matrix optimizations in order to reduce execution time,
such as updating matrices with new data, removing old data and introducing new
cointegration vectors.

Algorithm~\ref{alg:proposal} shows our OVECM proposal which considers the
following:

\begin{itemize}
\item The function \texttt{getJohansen} returns cointegration vectors given by
the Johansen method considering the trace statistic test at 95\%
significance level.
\item The function \texttt{vecMatrix} returns the matrices~(\ref{eq:vecA})
and~(\ref{eq:vecB}) that allows VECM to be solved.
\item The function \texttt{vecMatrixOnline} returns the
matrices~(\ref{eq:vecA}) and~(\ref{eq:vecB}) aggregating new data and removing
the old one, avoiding calculation of the matrix $\mathbf{A}$ from scratch.
\item Out-of-sample outputs are saved in the variables 
$\mathbf{Y}_{\text{true}}$ and $\mathbf{Y}_{\text{pred}}$.
\item The model is solved using OLS.
\item In-sample outputs are saved in the variables $\Delta
\mathbf{y}_{\text{true}}$ and $\Delta \mathbf{y}_{\text{pred}}$.
\item The function \texttt{mape} gets the in-sample MAPE for the $l$ time
series.
\item Cointegration vectors are obtained at the beginning and when they are required to be updated. This updating is decided based on the in-sample MAPE of the last $n$ inputs. The average of all
MAPEs are stored in the variable $e$. If the average of MAPEs
($\texttt{mean}(e)$) is above a certain error given by the mean\_error threshold, cointegration vectors are updated.
\item If new cointegration vectors are required, the function
\texttt{vecMatrixUpdate} only updates the corresponding columns of matrix
$\mathbf{A}$ affected by those vectors (see equation~\ref{eq:vecA}).
\end{itemize}

\begin{algorithm}[ht]
\begin{algorithmic}[1]
\REQUIRE $\,$ \\
$\mathbf{y}$: matrix with $N$ input vectors and $l$ time series\\
$p$: number of past values \\
$L$: sliding window size ($L<N$) \\
$\text{mean\_error}$: MAPE threshold \\
$n$: interpolation points to obtain MAPE \\
\ENSURE  $\,$ \\
$\{\mathbf{y}_{\text{pred}}[L+1],\dots,\mathbf{y}_{\text{pred}}[N]\}$: model predictions 
\FOR { $i =0$ to $N-L$ }
    \STATE $\mathbf{y}_i \gets \mathbf{y}[i:i+L]$
	\IF {i = 0 }
	    \STATE{$v \gets \texttt{getJohansen}(\mathbf{y}_i,p)$}
	    \STATE{$[\mathbf{A} \quad \mathbf{B}] \gets
        \texttt{vecMatrix}(\mathbf{y}_i,p,v)$}
	\ELSE
	    \STATE{$[\mathbf{A} \quad \mathbf{B}] \gets
        \texttt{vecMatrixOnline}(\mathbf{y}_i,p,v,\mathbf{A},\mathbf{B})$}
        %\STATE $\Delta \mathbf{Y}_{\text{true}}[i] \gets \mathbf{B}[-1,:]$
        \STATE $\Delta \mathbf{Y}_{\text{pred}}[i] \gets \mathbf{A}[-1,:] \times \mathbf{X}$
    \ENDIF
    \STATE $\mathbf{X} \gets \texttt{OLS} (\mathbf{A},\mathbf{B})$%\mathbf{(A^\top A)^{-1}A^\top B}$
    %\STATE $\Delta \mathbf{y}_{\text{true}} \gets \mathbf{B}[-n,:]$
    %\STATE $\Delta \mathbf{y}_{\text{pred}} \gets \mathbf{A}[-n,:] \times \mathbf{X}$
    \STATE $e \gets \texttt{mape}(\mathbf{B}[-n,:],\mathbf{A}[-n,:] \times \mathbf{X})$
    %\STATE $e \gets \texttt{mape}(\Delta \mathbf{y}_{\text{true}}, \Delta
    %\mathbf{y}_{\text{pred}})$
    \IF {$\texttt{mean}(e) > \text{mean\_error}$}
	    \STATE{$v \gets \texttt{getJohansen}(\mathbf{y}_i,p)$}
	    \STATE{$\mathbf{A} \gets
        \texttt{vecMatrixUpdate}(\mathbf{y}_i,p,v,\mathbf{A})$}
        \STATE $\mathbf{X} \gets \texttt{OLS} (\mathbf{A},\mathbf{B})$%\mathbf{(A^\top A)^{-1}A^\top B}$
        %\STATE $\mathbf{X} \gets \mathbf{(A^\top A)^{-1}A^\top B}$
    \ENDIF
\ENDFOR
\STATE $\mathbf{Y}_{\text{true}} \gets \mathbf{Y}[L+1:N]$
\STATE $\mathbf{Y}_{\text{pred}}\gets 
\mathbf{Y}[L:N-1] +\Delta \mathbf{Y}_{\text{pred}}$
%\STATE $\text{MAPE} \gets \texttt{mape}(\Delta \mathbf{Y}_{\text{true}}, \Delta
%\mathbf{Y}_{\text{pred}})$
\end{algorithmic}
\caption{OVECM: Online VECM}
\label{alg:proposal}
\end{algorithm}

Our proposal was compared against VECM and ARIMA. Both algorithms were adapted
to an online context in order to get a reasonable comparison with our proposal
(see algorithms \ref{alg:SLVECM} and \ref{alg:SLARIMA}). VECM and ARIMA are
called with a sliding window of the most recent data, whereby the models are
updated at every time step.

\begin{algorithm}[ht]
\begin{algorithmic}[1]
\REQUIRE $\,$ \\
$\mathbf{y}$: matrix with $N$ input vectors and $l$ time series\\
$p$: number of past values \\
$L$: sliding window size ($L<N$) \\
\ENSURE  $\,$ \\
$\{ \mathbf{y}_{\text{pred}}[L+1],\dots,\mathbf{y}_{\text{pred}}[N]\}$: model predictions 
\FOR { $i =0$ to $N-L$ }
    \STATE $\mathbf{y}_i \gets \mathbf{y}[i:i+L+1]$
        \STATE $model = VECM(\mathbf{y}_i, p)$
        \STATE $\mathbf{Y}_{\text{pred}}[i] = model.predict(\mathbf{y}[i+L])$
\ENDFOR
\STATE $\mathbf{Y}_{\text{true}} = \mathbf{y}[i+L+1:N] $
\end{algorithmic}
\caption{SLVECM: Sliding window VECM}
\label{alg:SLVECM}
\end{algorithm}

Since we know out time series are I(1) SLARIMA is called with $d=1$. ARIMA is
executed for every time series.

\begin{algorithm}[ht]
\begin{algorithmic}[1]
\REQUIRE $\,$ \\
$\mathbf{y}$: matrix with $N$ input vectors and $l$ time series\\
$p$: autoregressive order \\
$d$: integrated order\\
$q$: moving average order\\
$L$: sliding window size ($L<N$) \\
\ENSURE  $\,$ \\
$\{ \mathbf{y}_{\text{pred}}[L+1],\dots,\mathbf{y}_{\text{pred}}[N]\}$: model predictions 
\FOR { $i =0$ to $N-L$ }
\FOR { $j =0$ to $l-1$ }
    \STATE $\mathbf{y}_i \gets \mathbf{y}[i:i+L+1,j]$
        \STATE $model = ARIMA(\mathbf{y}_i, (p,d,q))$
        \STATE $\mathbf{Y}_{\text{pred}}[i,j] = model.predict(\mathbf{y}[i+L,j])$
\ENDFOR
\ENDFOR
\STATE $\mathbf{Y}_{\text{true}} = \mathbf{y}[i+L+1:N,:] $
\end{algorithmic}
\caption{SLARIMA: Sliding window ARIMA}
\label{alg:SLARIMA}
\end{algorithm}

Both OVECM and SLVECM time complexity is dominated by Johansen method which is
$O(n^3)$. Thus, both algorithms order is $O(Cn^3)$ where $C$ is the number of
iterations.

\subsection{Evaluation methods} \label{sec:evaluation}

Forecast performance is evaluated using different methods. We have chosen three
measures usually used:
\begin{description}
\item[MAPE] Mean Average Percent Error which presents forecast errors as a
percentage.
\begin{equation}\label{eq:MAPE}
\text{MAPE} = \frac{1}{N} \sum_{t=1}^{N} 
\frac{\left|\mathbf{y}_t-\hat{\mathbf{y}}_t\right|}{\left|\mathbf{y}_t\right|}
 \times 100 
\end{equation}
\item[MAE] Mean Average Error which measures the distance between forecasts to the
true value.
\begin{equation}\label{eq:MAE}
\text{MAE} = \frac{1}{N} \sum_{t=1}^{N} 
\left| 
\mathbf{y}_t-\hat{\mathbf{y}}_t
\right| 
\end{equation}
\item[RMSE] Root Mean Square Error also measures the distance between forecasts
to the true values but, unlike MAE, large deviations from the true value have a
large impact on RMSE due to squaring forecast error.
\begin{equation}\label{eq:RMSE}
\text{RMSE} = \sqrt{
\frac{\displaystyle \sum_{t=1}^{N} (\mathbf{y}_t-\hat{\mathbf{y}}_t)^2}{N}}
\end{equation}
\end{description}

\subsection{Model selection} \label{sec:pselection}
Akaike Information Criterion (AIC) is often used in model selection where AIC
with smaller values are preferred since they represent a trade-off between bias
and variance.  AIC is obtained as follows:

\begin{equation}
\label{eq:aicformula}
AIC = \underset{\text{bias}}{-\frac{2l}{N}} + 
\underset{\text{variance}}{\frac{2k}{N}}
\end{equation}

\noindent where

\begin{description}
\item[l] is the loglikelihood function
\item[k] number of estimated parameters
\item[N] number of observations
\end{description}

Loglikelihood function is obtained from the Residual Sum of Squares (RSS):

\begin{equation}
\label{eq:ll}
l = -\frac{N}{2} \left(1 + ln(2\pi) + ln\left(\frac{RSS}{N}\right)\right) 
\end{equation}

### 2014-10-29 src/1_Introduction.tex
new chapters

@@
 In this introduction we present the scope, objectives, hypothesis and organization of this thesis.
+
+\section{Introduction}
+A time series is a sequence S of historical measurements y t of an observable
+variable y at equal time intervals. Time series are studied for several purposes
+such as the forecasting of the future based on knowledge of the past, the un-
+derstanding of the phenomenon underlying the measures, or simply a succinct
+description of the salient features of the series. In this chapter we shall
+confine
+ourselves to the problem of forecasting. Forecasting future values of an
+observed
+time series plays an important role in nearly all fields of science and
+engineering,
+such as economics, finance, business intelligence, meteorology and telecommu-
+nication [43]. An important aspect of the forecasting task is represented by the
+size of the horizon. If the one-step forecasting of a time series is already a
+chal-
+lenging task, performing multi-step forecasting is more difficult [53] because
+of
+
+additional complications, like accumulation of errors, reduced accuracy, and in-
+creased uncertainty [58,49].
+The forecasting domain has been influenced, for a long time, by linear sta-
+tistical methods such as ARIMA models. However, in the late 1970s and early
+1980s, it became increasingly clear that linear models are not adapted to many
+real applications [25]. In the same period, several useful nonlinear time series
+models were proposed such as the bilinear model [44], the threshold autoregres-
+sive model [56,54,55] and the autoregressive conditional heteroscedastic (ARCH)
+model [22] (see [25] and [26] for a review). However, the analytical study of
+non-
+linear time series analysis and forecasting is still in its infancy compared to
+linear
+time series [25].
+In the last two decades, machine learning models have drawn attention and
+have established themselves as serious contenders to classical statistical
+models
+in the forecasting community [1,43,61]. These models, also called black-box or
+data-driven models [40], are examples of nonparametric nonlinear models which
+use only historical data to learn the stochastic dependency between the past and
+the future. For instance, Werbos found that Artificial Neural Networks (ANNs)
+outperform the classical statistical methods such as linear regression and Box-
+Jenkins approaches [59,60]. A similar study has been conducted by Lapedes and
+Farber [33] who conclude that ANNs can be successfully used for modeling and
+forecasting nonlinear time series. Later, other models appeared such as decision
+trees, support vector machines and nearest neighbor regression [29,3]. Moreover,
+the empirical accuracy of several machine learning models has been explored in a
+number of forecasting competitions under different data conditions (e.g. the
+NN3,
+NN5, and the annual ESTSP competitions [19,20,34,35]) creating interesting
+scientific debates in the area of data mining and forecasting [28,45,21].
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+The computational complexity of machine learning algorithms has become a limiting factor for problems that require processing large volumes of data and where the response time is crucial.
+An example of this is financial volatility time series forecast.
+
+Currently, it is known that volatility is a process where many factors influence its behaviour.
+However, many models of volatility made forecasts based only on price returns.
+Examples are the ARFIMA, GARCH and stochastic volatility models.
+Machines learning methods have allowed a greater number of explanatory variables to be included, although in many cases, these increase calculation time.
+
+Algorithms that process large amounts of data in a short time and are able to detect changes in the financial market are required to achieve fast and accurate responses.
+Online learning algorithms have been developed to solve large-scale problems since they process an instance at a time, updating the model at each step incrementally.
+This is opposed to the batch algorithms where a phase of training that uses a large collection of historical data is first required and the resulting model is later used to predict.
+This thesis proposal is to design and develop an online learning algorithm that meets or improves the performance of current volatility algorithms and ensures rapid responses to large-scale problems.
+Model comparisons will be made by the SPA (superior predictive ability) test proposed by Hansen and
+Lunde~\cite{hansen+lunde2006}  that allows comparing of forecast models giving a relative measure of model performance with respect to one or more models.
+
+\section{The problem}
+
+In financial markets, volatility is one of the key elements to model the stochastic dynamic behaviour of financial assets.
+This is mainly because volatility gives a measure of uncertainty about future returns and it is often viewed as an indicator of the vulnerability of financial markets.
+It is also important as an input parameter in problems like derivative pricing, hedging and portfolio management.
+For example, in option pricing we need to known first the volatility of the underlying asset before pricing an option.
+
+Besides, volatility provides important information for studying asset returns because the latter are largely uncorrelated and nonlinearly dependent.
+However, it has been found that volatility exhibits significant autocorrelation and predictable patterns~\cite{poon+granger2003} and  many models have been proposed to forecast its behaviour (see section ~\ref{sec:stylizedfacts}).
+
+The volatility of the data is due to a large number of factors affecting the market, with some of them directly measurable such as historical prices, trends, supply and demand.
+However, others such as monetary policies and news are not directly measurable and are included in other models out of the scope of this thesis.
+Figure~\ref{fig:VolFactorNews} shows the VIX index, the daily US implied stock market volatility, commonly known as the "financial fear factor" which is clearly affected by historical events.
+
+\begin{figure}[h]
+ \centering
+ \includegraphics[width=0.7\textwidth]{img/vol-factor-thenews.jpg}
+ % vol-factor-thenews.jpg: 710x506 pixel, 72dpi, 25.05x17.85 cm, bb=0 0 710 506
+ \caption{The news as volatility factor}
+ \label{fig:VolFactorNews}
+\end{figure}
+
+Volatility models have been extensively studied~\cite{gatheral2006,poon+granger2003,knight2002} and some of  them include machine learning methods~\cite{hamidetal2004,donaldsonetal1997,shiyietal2008,shiyietal2010,gavrishchaka2006,vasilios2012}.
+The stochastic behaviour of volatility and the large number of factors affecting it result in models with a large number of parameters and slow response time due to the computationally expensive routines.
+These expensive calculations could be avoided using online learning algorithms, which allow response time reduction and are therefore suitable for high frequency trading applications.
+
+In this thesis, a model will be proposed based on online learning algorithms to forecast volatility in high frequency financial time series.
+The main problems to be deal with in this thesis are:
+
+\begin{itemize}
+   \item Although volatility is often predictable~\cite{poon+granger2003}, there are non-linear features~\cite{gatheral2006} to be considered in this proposal.
+    \item The number of factors affecting volatility has been studied~\cite{srinivasan2009} and some other unknown factors affect volatility and they should also be included.
+    \item The response time needs to be very fast and in order to ensure this, the proposed model will be presented as an online machine learning algorithm. 
+\end{itemize}
 
@@
 Chapter 1 contains ...
+
+%\section{Hypotheses}
+%
+%
+%Volatility forecasting has been a largely studied and different approaches have arisen to describe its stochastic behaviour.
+It is well-known that volatility depends on many factors, but most of the existing models only consider a few of them.
+In order to introduce more factors easily, classic machine learning models have been studied and their performance have shown to be successfull in comparison with most used volatility models.
+However, their training process is computationally expensive and it is not possible to update the model with new market information before new information arrives.
+Online learning methods can also include more factors into a model, but they process one example at a time to update an existing model incrementally without having a training phase, thus they improve response times and are more suitable for financial applications.
+%
+%
+%The thesis hypotheses are the following:
+%
+%\begin{enumerate} 
+%\item Machine learning based models that consider more volatility explanatory factors will adapt and perform better than popular volatility forecasting models based only on returns or a few factors.
+%\item Stylized facts can be efficiently introduced as explanatory factors and improve model performance.
+%\item The use of online learning methods will allow inclusion of as many variables as needed to ensure low response times.  
+%\end{enumerate}
+%
+%\section{General Objective}
+%To design and implement an online learning model for volatility forecasting capable of adapting to their non-stationary behaviour using volatility factors as model input ensuring low response times.
+%\section{Specific Objectives}
+%The specific objectives to be followed in this thesis are:
+%\begin{enumerate}
+%  \item To determine the best volatility factors based on their stylized facts and their performance in machine learning models.
+%  \item To determine the best sampling of high frequency time series.
+%  \item To define the best realized volatility proxy based on the literature and experimental results.
+%  \item To design and implement a novel online learning algorithm with large input factors capable of ensuring low response time.
+%  \item To modify the original algorithm using kernel trick and evaluate its performance.
+%  \item To compare performance of the most popular algorithms for volatility forecasting such as ARFIMA, GARCH and SV models.
+%  \item To assess the predictive capacity of the proposed model using the SPA test.
+%  \item To compare the performance of the most popular models and the proposed model using real financial data.
+%\end{enumerate}
+
+%\section{Thesis proposal}
+%
+%This thesis proposal is to build an algorithm to forecast realized volatility in financial markets considering not only intraday returns but also other explanatory volatility variables.
+Because of the amount of high frequency data to be considered and short response times required in financial applications, the algorithm will be formulated in online mode.
+The online algorithm will also allow the shifting problem to be tackled while being adaptive to wild changes in the volatility process.
+%The online learning model will be formulated considering the following:
+%
+%\begin{description}%[leftmargin=0.4cm]\itemsep4pt
+%    \item[large volumes of data:] high frequency data will be used to estimate realized volatility.
+The frequency to be considered will be 10 to 15 minutes.
+This frequency will be analysed using signal processing theory.
+%    \item[non-stationary data:] volatility is a time series known because of its non-stationary feature, so statistics will be studied before using the model.
+%    \item[stylized facts:] there are several well-documented features about financial time series volatility called stylized facts~\cite{poon+granger2003} that can be included in the model.
+%    \item[low response time:] the model has to be simple but powerful in order to achieve the response time needed.
+If the model is extended to real time data, ticks can arrive at rates of milliseconds and the model should respond at the same rate or less.
+%    \item[financial data:] the model can be constructed using data of stocks or different derivatives such as options and futures. The chosen implicit volatility model will depend on the data type.
+%    \item[input vector:] currently, many models are only based on expected returns, but volatility is a complex variable which depends also on many factors studied in the literature such as trading volume, the tick rate, time of the day and the behaviour of the market as a hole~\cite{gatheral2006}.
+All these factors can improve the design of the proposed online algorithm.
+%    \item[kernel version:] the online algorithm will be later kernelised in order to map the input data into a high dimensional feature space (kernel trick) which will provide better properties to design the solution.
+%\end{description}
+%
+%The model proposed has to be compared with the current popular financial models of financial volatility such as: ARFIMA, GARCH, SV models and machine learning models.
+Performance will be measured using the realized volatility shown in equation~(\ref{eq:srv}) as a benchmark.
+%
+%In order to compare different models, apart from using the standard performance measurements like MAE and MSE, the superior predictive ability (SPA) methodology proposed by Hansen and Lunde~\cite{hansen+lunde2006} will also be studied.
+SPA is a test to compare two or more forecasting models at the same time giving a relative performance measurement of a base model in comparison with alternative models.
+%
+%The model will be compared against traditional financial volatility models such as Black-Scholes, SV, GARCH and ARFIMA using different forecast errors measurements.
+Although the best way to measure forecast errors should be the investor returns, this implies knowing how the model information is used as a strategy, so this information is not going to be included in this thesis.
+Therefore, in order to compare the different volatility forecast models, popular evaluation measurements will be used.
+In the literature the following are commonly used: {\em Mean Error} (ME), {\em Mean Square Error} (MSE), {\em Mean Absolute Error} (MAE) and {\em Mean Absolute Percent Error} {MAPE} and others are less commonly used such as LINEX loss function which weight differently positive and negative errors.
+In ~\cite{granger1999} Granger describes a variety of non-symmetric loss function including LINEX.
+The superior predictive ability (SPA) methodology proposed by Hansen and Lunde~\cite{hansen+lunde2006} will also be studied
+%which allows comparison of two or more forecasting models at the same time giving a relative performance measurement of a base model compared with alternative models.
+All these measurements will be out-of-sample in order to show their predictive power.

### 2014-11-19 src/0_Abstract.tex
comenzando a juntar

@@
 
-\emph{Financial time series are known for their non-stationary behaviour. However,
-sometimes they exhibit some stationary linear combinations. When this happens,
-it is said that those time series are cointegrated. Vector error correction
-model (VECM) is an econometric model which characterizes the joint dynamic
-behaviour of a set of cointegrated variables in terms of forces pulling towards
-equilibrium.}
+\emph{Financial time series are known for their non-stationary behaviour and
+this has motivated its study using different techniques. 
+One relevant observation is that sometimes time series exhibit some stationary
+linear combinations. When this happens, it is said that those time series are
+cointegrated.  Cointegration has been then one of the main features studied.
+Vector error correction model (VECM) is an econometric model which characterizes
+the joint dynamic behaviour of a set of cointegrated variables in terms of
+forces pulling towards equilibrium.}
 
@@
 
-\emph{In this study, we propose an online VECM which optimizes how model parameters
-are obtained by using only relevant past data and RR instead of OLS.  Our
-proposal also takes advantage of the long-run relationship between the time
-series in order to obtain improved execution times. Our proposed method is
-tested using synthetic data and foreign exchange rates.
-Results show that our proposal enables execution time to be reduced by at least
-50\% while maintaining good accuracy levels.}
+\emph{In this thesis, financial time series features are studied and different
+algorithms were developed in order to optimize parameters an increase performance.
+In particular, we propose an online VECM which optimizes how model parameters
+are obtained considering only past data and different ways of solving the model
+using machine learning techniques.  
+The long-run relationship between the time series is used in order to make
+optimizations and obtain improved execution times. However, due to the large
+amount of data available and the need of quick response, algorithms presented
+were optimized using high performance computing.
+Our experiments were tested using synthetic data and foreign exchange rates.
+Results show that cointegration and high performance computing allow to obtain
+better models and execution times.}
 

### 2014-11-19 src/1_Introduction.tex
comenzando a juntar

@@
 far apart."-Robert F. Engle and Clive W.J. Granger~\cite{engle1987}.\\
-In this introduction we present the scope, objectives, hypothesis and organization of this thesis.
+In this introduction we present the scope, objectives, hypothesis and
+organization of this thesis.
 
-\section{Introduction}
-A time series is a sequence S of historical measurements y t of an observable
-variable y at equal time intervals. Time series are studied for several purposes
-such as the forecasting of the future based on knowledge of the past, the un-
-derstanding of the phenomenon underlying the measures, or simply a succinct
-description of the salient features of the series. In this chapter we shall
-confine
-ourselves to the problem of forecasting. Forecasting future values of an
-observed
-time series plays an important role in nearly all fields of science and
-engineering,
-such as economics, finance, business intelligence, meteorology and telecommu-
-nication [43]. An important aspect of the forecasting task is represented by the
-size of the horizon. If the one-step forecasting of a time series is already a
-chal-
-lenging task, performing multi-step forecasting is more difficult [53] because
-of
+\section{Scope of this research}
 
-additional complications, like accumulation of errors, reduced accuracy, and in-
-creased uncertainty [58,49].
-The forecasting domain has been influenced, for a long time, by linear sta-
-tistical methods such as ARIMA models. However, in the late 1970s and early
-1980s, it became increasingly clear that linear models are not adapted to many
-real applications [25]. In the same period, several useful nonlinear time series
-models were proposed such as the bilinear model [44], the threshold autoregres-
-sive model [56,54,55] and the autoregressive conditional heteroscedastic (ARCH)
-model [22] (see [25] and [26] for a review). However, the analytical study of
-non-
-linear time series analysis and forecasting is still in its infancy compared to
-linear
-time series [25].
-In the last two decades, machine learning models have drawn attention and
-have established themselves as serious contenders to classical statistical
-models
-in the forecasting community [1,43,61]. These models, also called black-box or
-data-driven models [40], are examples of nonparametric nonlinear models which
-use only historical data to learn the stochastic dependency between the past and
-the future. For instance, Werbos found that Artificial Neural Networks (ANNs)
-outperform the classical statistical methods such as linear regression and Box-
-Jenkins approaches [59,60]. A similar study has been conducted by Lapedes and
-Farber [33] who conclude that ANNs can be successfully used for modeling and
-forecasting nonlinear time series. Later, other models appeared such as decision
-trees, support vector machines and nearest neighbor regression [29,3]. Moreover,
-the empirical accuracy of several machine learning models has been explored in a
-number of forecasting competitions under different data conditions (e.g. the
-NN3,
-NN5, and the annual ESTSP competitions [19,20,34,35]) creating interesting
-scientific debates in the area of data mining and forecasting [28,45,21].
+The stochastic behaviour of financial time series, its incrementing
+amount of data available and the need of performing accurate forecasting in
+short periods of time has motivated researchers to create efficient and fast
+algorithms. This study involves interdisciplinary knowledge such as: finance,
+scientific computing, high performance computing, machine learning, physics
+among others.
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+Forecasting financial time series have been modeled using classical statistical
+models. More recently , machine learning models have been extensively used in
+forecasting. However, getting model parameters is a computational challenge.
+The computational complexity of machine learning algorithms has become a
+limiting factor for problems that require processing large volumes of data and
+where the response time is crucial.
 
-The computational complexity of machine learning algorithms has become a limiting factor for problems that require processing large volumes of data and where the response time is crucial.
-An example of this is financial volatility time series forecast.
+Therefore, algorithms that process large amount of data in a short period of
+time are required.  Online learning algorithms have been developed to solve
+large-scale problems since they process an instance at a time, updating the
+model at each step incrementally. This is opposed to the batch algorithms where
+a phase of training that uses a large collection of historical data is first
+required and the resulting model is later used to predict.
 
-Currently, it is known that volatility is a process where many factors influence its behaviour.
-However, many models of volatility made forecasts based only on price returns.
-Examples are the ARFIMA, GARCH and stochastic volatility models.
-Machines learning methods have allowed a greater number of explanatory variables to be included, although in many cases, these increase calculation time.
-
-Algorithms that process large amounts of data in a short time and are able to detect changes in the financial market are required to achieve fast and accurate responses.
-Online learning algorithms have been developed to solve large-scale problems since they process an instance at a time, updating the model at each step incrementally.
-This is opposed to the batch algorithms where a phase of training that uses a large collection of historical data is first required and the resulting model is later used to predict.
-This thesis proposal is to design and develop an online learning algorithm that meets or improves the performance of current volatility algorithms and ensures rapid responses to large-scale problems.
-Model comparisons will be made by the SPA (superior predictive ability) test proposed by Hansen and
-Lunde~\cite{hansen+lunde2006}  that allows comparing of forecast models giving a relative measure of model performance with respect to one or more models.
-
-\section{The problem}
-
-In financial markets, volatility is one of the key elements to model the stochastic dynamic behaviour of financial assets.
-This is mainly because volatility gives a measure of uncertainty about future returns and it is often viewed as an indicator of the vulnerability of financial markets.
-It is also important as an input parameter in problems like derivative pricing, hedging and portfolio management.
-For example, in option pricing we need to known first the volatility of the underlying asset before pricing an option.
-
-Besides, volatility provides important information for studying asset returns because the latter are largely uncorrelated and nonlinearly dependent.
-However, it has been found that volatility exhibits significant autocorrelation and predictable patterns~\cite{poon+granger2003} and  many models have been proposed to forecast its behaviour (see section ~\ref{sec:stylizedfacts}).
-
-The volatility of the data is due to a large number of factors affecting the market, with some of them directly measurable such as historical prices, trends, supply and demand.
-However, others such as monetary policies and news are not directly measurable and are included in other models out of the scope of this thesis.
-Figure~\ref{fig:VolFactorNews} shows the VIX index, the daily US implied stock market volatility, commonly known as the "financial fear factor" which is clearly affected by historical events.
-
-\begin{figure}[h]
- \centering
- \includegraphics[width=0.7\textwidth]{img/vol-factor-thenews.jpg}
- % vol-factor-thenews.jpg: 710x506 pixel, 72dpi, 25.05x17.85 cm, bb=0 0 710 506
- \caption{The news as volatility factor}
- \label{fig:VolFactorNews}
-\end{figure}
-
-Volatility models have been extensively studied~\cite{gatheral2006,poon+granger2003,knight2002} and some of  them include machine learning methods~\cite{hamidetal2004,donaldsonetal1997,shiyietal2008,shiyietal2010,gavrishchaka2006,vasilios2012}.
-The stochastic behaviour of volatility and the large number of factors affecting it result in models with a large number of parameters and slow response time due to the computationally expensive routines.
-These expensive calculations could be avoided using online learning algorithms, which allow response time reduction and are therefore suitable for high frequency trading applications.
-
-In this thesis, a model will be proposed based on online learning algorithms to forecast volatility in high frequency financial time series.
-The main problems to be deal with in this thesis are:
-
-\begin{itemize}
-   \item Although volatility is often predictable~\cite{poon+granger2003}, there are non-linear features~\cite{gatheral2006} to be considered in this proposal.
-    \item The number of factors affecting volatility has been studied~\cite{srinivasan2009} and some other unknown factors affect volatility and they should also be included.
-    \item The response time needs to be very fast and in order to ensure this, the proposed model will be presented as an online machine learning algorithm. 
-\end{itemize}
-
-\section{Scope of this Research}
-
-Cointegration concept was introduced by~\cite{engle1987} and implies that one or
-more linear combinations of non-stationary variables are stationary even though
-individually they are not.  Moreover ~\cite{stock+watson1988} observed that
-cointegration reflects the common stochastic trends providing a useful way to
-understand cointegration relationships. These common stochastic trends can be
-also interpreted as a long-run equilibrium relationships.
-
-This idea was inmediatly adopted in finance since it could represent their
-long-run relationship implied by economic theory~\cite{laietAl1991},
-\cite{lence+falk2005}.  Economic theory suggest that economic time series are
-mean-reverting process and therefore, it reflects the idea of that some set of
-variables cannot wander too far from each other.
-
-On the other hand, the efficient markets hypothesis, also known as the random
-walk theory states that current stock prices fully reflect available
-information related to its value and there is no way to earn excess
-profits~\cite{fama1970}.  This means that if we have stock prices from a
-jointly efficient market, they cannot be
-cointegrated~\cite{granger1986},\cite{dwyer1992}. However, \cite{richards1995}
-claims that cointegration is directly at odds with market efficiency, even
-though, there is no evidende that cointegration among asset prices have
-implications about market efficiency~\cite{lence+falk2005}.
-
-Despite the fact that cointegration on closing daily rates of currency pairs has
-not been found \cite{coleman1990}, \cite{copeland1991}, different time series
-frequencies can have different behaviours~\cite{aldridge2009}. Pair trading is a
-very common example of cointegration application~\cite{herlemont2003} 
-but cointegration can also be extended to a larger set of
-variables~\cite{mukherjee1995},\cite{engle2004}.
-
-%So why do we care about cointegration? Someone else can probably give more
-%econometric applications, but in quantitative finance, cointegration forms the
-%basis of the pairs trading strategy: suppose we have two cointegrated stocks X
-%and Y, with the particular (for concreteness) cointegrating relationship X - 2Y
-%= Z, where Z is a stationary series of zero mean. For example, X could be
-%McDonald's, Y could be Burger King, and the cointegration relationship would
-%mean that X tends to be priced twice as high as Y, so that when X is more than
-%twice the price of Y, we expect X to move down or Y to move up in the near
-%future (and analogously, if X is less than twice the price of Y, we expect X to
-%move up or Y to move down). This suggests the following trading strategy: if X -
-%2Y > d, for some positive threshold d, then we should sell X and buy Y (since we
-%expect X to decrease in price and Y to increase), and similarly, if X - 2Y < -d,
-%then we should buy X and sell Y.
-
-Vector error correction model (VECM) introduces this long-run relationship
-among a set of cointegrated variables as an error correction term. VECM is a
-special case of the vector autorregresive model (VAR) model. VAR model
-expresses future values as a linear combination of variables past values.
-However, VAR model cannot be used with non-stationary variables. VECM is a
-linear model but in terms of variable differences. If cointegration exists,
-variable differences are stationary and they introduce an error correction term
-which adjusts coefficients to bring the variables back to equilibrium. In
-finance, many economic time series turn to be stationary when they are
-differentiated and cointegration restrictions often improves
-forecasting~\cite{duy1998}. Therefore, VECM has been widely adopted.
-
-Both VECM and VAR model parameters are obtained using ordinary least squares
-(OLS) method. OLS has two main problems: is sensitive to errors on input data
-and involves many calculations. The former problem is commonly solved using
-Ridge Regression (RR) \cite{hoerl1970} which introduces a regularization
-parameter that leads to an unbiased estimation with better generalization
-capability. The second problem of computational complexity depends on the number
-of past values and observations considered.  Recently, online learning
-algorithms have been proposed to solve problems with large data sets because of
-their simplicity and their ability to update the model when new data is
-available.
+The specific scope of this study is to use financial time series features in
+order to design a forecasting algorithm which ensures accuracy and low response
+times. Cointegration is the main feature studied and it refers that one or more
+linear combinations of these variables are stationary even though individually
+they are not~\cite{engle87}.  Some models, such as the Vector Error Correction
+(VECM), take advantage of this property and describe the joint behaviour of
+several cointegrated variables.
 
@@
 \section{Research Objectives}
-The main motivation for this research is ...
+The main motivation for this research is the develpment of efficient methods for
+forecasting financial time series.
 
@@
 \begin{itemize}
-\item $\Large \mathcal{O}_1$: \emph{Especific Objective 1.}
-\item $\Large \mathcal{O}_2$: \emph{Especific Objective 2.}
-\item $\Large \mathcal{O}_3$: \emph{Especific Objective 3.}
-\item $\Large \mathcal{O}_4$: \emph{Especific Objective 4.}
-\item $\Large \mathcal{O}_5$: \emph{Especific Objective 5.}
+\item $\Large \mathcal{O}_1$: \emph{A review of the literature on time series
+analysis models including machine learning techniques.}
+\item $\Large \mathcal{O}_2$: \emph{Development of a set of know features of the
+studied time series and applied them to improve forecasting.}
+\item $\Large \mathcal{O}_3$: \emph{Development of parallel and efficient
+algorithms to ensure quick time responses.}
+\item $\Large \mathcal{O}_4$: \emph{Deep mathematical analysis of the proposal
+and financial concepts involved.}
+\item $\Large \mathcal{O}_5$: \emph{Design and implement representative set of
+experiments in order to show when and why proposal performs better.}
 \end{itemize}

### 2014-11-20 src/1_Introduction.tex
avance

--- 2014-10-29 src/1_Introduction.tex
@@
 far apart."-Robert F. Engle and Clive W.J. Granger~\cite{engle1987}.\\
-In this introduction we present the scope, objectives, hypothesis and organization of this thesis.
+In this introduction we present the scope, objectives, hypothesis and
+organization of this thesis.
 
-\section{Scope of this Research}
+\section{Scope of this research}
+
+The stochastic behaviour of financial time series, its incrementing
+amount of data available and the need of performing accurate forecasting in
+short periods of time has motivated researchers to create efficient and fast
+algorithms. This study involves interdisciplinary knowledge such as: finance,
+scientific computing, high performance computing, machine learning, physics
+among others.
+
+Forecasting financial time series have been modeled using classical statistical
+models. More recently , machine learning models have been extensively used in
+forecasting. However, getting model parameters is a computational challenge.
+The computational complexity of machine learning algorithms has become a
+limiting factor for problems that require processing large volumes of data and
+where the response time is crucial.
+
+Therefore, algorithms that process large amount of data in a short period of
+time are required.  Online learning algorithms have been developed to solve
+large-scale problems since they process an instance at a time, updating the
+model at each step incrementally. This is opposed to the batch algorithms where
+a phase of training that uses a large collection of historical data is first
+required and the resulting model is later used to predict.
+
+The specific scope of this study is to use financial time series features in
+order to design a forecasting algorithm which ensures accuracy and low response
+times. Cointegration is the main feature studied and it refers that one or more
+linear combinations of these variables are stationary even though individually
+they are not~\cite{engle87}.  Some models, such as the Vector Error Correction
+(VECM), take advantage of this property and describe the joint behaviour of
+several cointegrated variables.
+
+In this thesis, we propose an online formulation of the VECM called Onlive VECM
+(OVECM) based on consideration of only a sliding window of the most recent data.
+The algorithm introduces matrix optimizations in order to reduce the number of
+operations and also takes into account the fact that cointegration vector space
+doesn't experience large changes with small changes in the input data. Moreover,
+RR instead of OLS is used to solve VECM. Our method is later tested using four
+currency rates from the foreign exchange market with different frequencies.  Model
+efectiveness is focused on out-of-sample forecast rather than in-sample fitting.
+This criteria allows the OVECM prediction capability to be expressed rather than
+just explaining data history. Our method performance is compared with its
+optimal offline algorithm.
+
+\section{Research Objectives}
+The main motivation for this research is the develpment of efficient methods for
+forecasting financial time series.
+
+The specific objectives of this research are,
+\begin{itemize}
+\item $\Large \mathcal{O}_1$: \emph{A review of the literature on time series
+analysis models including machine learning techniques.}
+\item $\Large \mathcal{O}_2$: \emph{Development of a set of know features of the
+studied time series and applied them to improve forecasting.}
+\item $\Large \mathcal{O}_3$: \emph{Development of parallel and efficient
+algorithms to ensure quick time responses.}
+\item $\Large \mathcal{O}_4$: \emph{Deep mathematical analysis of the proposal
+and financial concepts involved.}
+\item $\Large \mathcal{O}_5$: \emph{Design and implement representative set of
+experiments in order to show when and why proposal performs better.}
+\end{itemize}
+
+\section{Research Hypothesis}
 
@@
 also interpreted as a long-run equilibrium relationships.
-
-This idea was inmediatly adopted in finance since it could represent their
-long-run relationship implied by economic theory~\cite{laietAl1991},
-\cite{lence+falk2005}.  Economic theory suggest that economic time series are
-mean-reverting process and therefore, it reflects the idea of that some set of
-variables cannot wander too far from each other.
-
-On the other hand, the efficient markets hypothesis, also known as the random
-walk theory states that current stock prices fully reflect available
-information related to its value and there is no way to earn excess
-profits~\cite{fama1970}.  This means that if we have stock prices from a
-jointly efficient market, they cannot be
-cointegrated~\cite{granger1986},\cite{dwyer1992}. However, \cite{richards1995}
-claims that cointegration is directly at odds with market efficiency, even
-though, there is no evidende that cointegration among asset prices have
-implications about market efficiency~\cite{lence+falk2005}.
-
-Despite the fact that cointegration on closing daily rates of currency pairs has
-not been found \cite{coleman1990}, \cite{copeland1991}, different time series
-frequencies can have different behaviours~\cite{aldridge2009}. Pair trading is a
-very common example of cointegration application~\cite{herlemont2003} 
-but cointegration can also be extended to a larger set of
-variables~\cite{mukherjee1995},\cite{engle2004}.
-
-%So why do we care about cointegration? Someone else can probably give more
-%econometric applications, but in quantitative finance, cointegration forms the
-%basis of the pairs trading strategy: suppose we have two cointegrated stocks X
-%and Y, with the particular (for concreteness) cointegrating relationship X - 2Y
-%= Z, where Z is a stationary series of zero mean. For example, X could be
-%McDonald's, Y could be Burger King, and the cointegration relationship would
-%mean that X tends to be priced twice as high as Y, so that when X is more than
-%twice the price of Y, we expect X to move down or Y to move up in the near
-%future (and analogously, if X is less than twice the price of Y, we expect X to
-%move up or Y to move down). This suggests the following trading strategy: if X -
-%2Y > d, for some positive threshold d, then we should sell X and buy Y (since we
-%expect X to decrease in price and Y to increase), and similarly, if X - 2Y < -d,
-%then we should buy X and sell Y.
 
@@
 
-In this thesis, we propose an online formulation of the VECM called Onlive VECM
-(OVECM) based on consideration of only a sliding window of the most recent data.
-The algorithm introduces matrix optimizations in order to reduce the number of
-operations and also takes into account the fact that cointegration vector space
-doesn't experience large changes with small changes in the input data. Moreover,
-RR instead of OLS is used to solve VECM. Our method is later tested using four
-currency rates from the foreign exchange market with different frequencies.  Model
-efectiveness is focused on out-of-sample forecast rather than in-sample fitting.
-This criteria allows the OVECM prediction capability to be expressed rather than
-just explaining data history. Our method performance is compared with its
-optimal offline algorithm.
+The main research hypothesis explored in this dissertation is the following:
+\\
 
-\section{Research Objectives}
-The main motivation for this research is ...
-
-The specific objectives of this research are,
-\begin{itemize}
-\item $\Large \mathcal{O}_1$: \emph{Especific Objective 1.}
-\item $\Large \mathcal{O}_2$: \emph{Especific Objective 2.}
-\item $\Large \mathcal{O}_3$: \emph{Especific Objective 3.}
-\item $\Large \mathcal{O}_4$: \emph{Especific Objective 4.}
-\item $\Large \mathcal{O}_5$: \emph{Especific Objective 5.}
-\end{itemize}
-
-\section{Research Hypothesis}
-
-The main research hypothesis explored in this dissertation is the following \\
-
-\textit{Hypothesis llll . }
+\textit{An online learning algorithms based on cointegration and high
+performance computing will allow to obtain better and faster forecasting
+algorithms for financial time series maintaining good accuracy levels.}
 
@@
 
-Chapter 1 contains ...
+Chapter 1 contains relevant finance concepts required to understand financial
+time series models such as market hypothesis, frequency, order book generation,
+market microestructure.
+Chapter 2 discuss main time series models including classic ones such as ARMA,
+GARCH and more recent ones such as VECM, VAR and volatility models. %mejorar
+Chapter 3 gives a brief introduction to machine learning and its variant online
+learning. 
+Chapter 4 presents different algorithms propose to forecast financial time
+series using machine learning techniques.
+In Chapter 5 a deep study of foreign exchange market features is presented and
+experiments of cointegration are carried out in order to show its dynamic.
+Chapter 6 presents a discussion of found results and summarize the main
+conlusions of this thesis. It also presents some research directions for future
+studied.

### 2014-11-20 src/2_Finance.tex
avance

\chapter{High Frequency Trading}

High frequency trading (HFT) strategies are based on buy and sell assets in
short periods of time gaining a small profit in every transaction, the key is
the amount of transactions that algorithms are capable to execute. 
In 2012 more than 50\% of US market share of exchange trades were HFT.  HFT has
motivated computer-driven strategies capable of processing large amount of data
in short periods of time.

\section{High Frequency Trading}

HFT is not a strategy but a technology which allow to implement particular
trading strategies. The aim of HFT is to benefit from market short-term pricing
inefficiencies~\cite{chlistalla2011}.
High frequency trades are short-term positions that commonly end the trading
day avoiding leaving opened positions to the next business day. HFT is
frequently associate to Algorithm trading strategies, but the former is focused
in to reduce the market impact of large institutionals positions.

In 2014 HFT represents nearly 50\% of equity trades in the US and more than 20\%
in Europe showing a consistent fall since 2009 which was its best year. Revenues
have fallen dramatically.
Figure~\ref{fig:HFTmarket} shows HFT trades percentage of US equity shares.

\begin{figure}[!h]
  %\vspace{-0.8cm}
  \centering
  \includegraphics[width=\textwidth]{img/HFTmarket}
  \caption{Left figure shows HFT market share in the US and Europe. Right figure
  shows revenue in the US.}
  \label{fig:HFTmarket}
\end{figure}

One of the reasons of this fall is that exchange markets have adapted, having
now faster and more transparent and efficient market structures than before.
Even though this fall, HFT is still a major component of regulated markets and
will probably remain a topic of interest for researchers in the near future.

Despite the fact HFT has been criticized on qualitative issues concerning fariness and
systemic risk, empirical research shows that HFT has lead to beneficial impacts
such as tighter spreads, increased liquidity, more efficient price formation,
reduced transaction costs and lower market volatility.

\subsection{Financial Markets}

A financial market is any marketplace where buyers and sellers participate
trading different assets such as equities, bonds, currencies and derivatives
(future or options). One of the main objectives of financial markets is to set
prices for global trade.
A financial market has many components but the most commonly used are money
markets and capital markets. Money markets are used for short-term basis,
usually for assets up to one year, for greater periods capital markets are used.
Capital markets include the stock or equity market and the bond or debt market
and it their movements are the most widely followed.

Figure~\ref{fig:capitalmarket} shows the typical structure of capital markets
existed from the early 1929s through much of the 1990s where the broker-dealers
played the central and most profitable role.  At the core are the exchanges or
inter-dealer networks (foreign exchange trading). Exchanges are the centralized
marketplaces for transacting.  Broker-dealers perform two functions: trading for
their own accounts and transacting for their customers. Broker-dealers use
inter-dealer brokers to quickly find the best price for a particular asset among
the network of other broker-dealers. Occasionally, broker-dealers also deal
directly with other broker-dealers, particularly for less liquid instruments.
Broker-dealers clients are institutional investors, corporate clients,
commercial clients, and high-net-worth individuals.

\begin{figure}[!h]
  %\vspace{-0.8cm}
  \centering
  \includegraphics[width=0.75\textwidth]{img/capitalmarkets}
  \caption{Old structure of capital markets.}
  \label{fig:capitalmarket}
\end{figure}

This centralized structure existed until computer technology allowed a better
communication structure. Today financial markets are more descentralized
providing more liquidity.

### 2014-11-20 src/3_TimeSeries.tex
avance

@@
+\chapter{Time Series}
+
+Cointegration concept was introduced by~\cite{engle1987} and implies that one or
+more linear combinations of non-stationary variables are stationary even though
+individually they are not.  Moreover ~\cite{stock+watson1988} observed that
+cointegration reflects the common stochastic trends providing a useful way to
+understand cointegration relationships. These common stochastic trends can be
+also interpreted as a long-run equilibrium relationships.
+
+This idea was inmediatly adopted in finance since it could represent their
+long-run relationship implied by economic theory~\cite{laietAl1991},
+\cite{lence+falk2005}.  Economic theory suggest that economic time series are
+mean-reverting process and therefore, it reflects the idea of that some set of
+variables cannot wander too far from each other.
+
+On the other hand, the efficient markets hypothesis, also known as the random
+walk theory states that current stock prices fully reflect available
+information related to its value and there is no way to earn excess
+profits~\cite{fama1970}.  This means that if we have stock prices from a
+jointly efficient market, they cannot be
+cointegrated~\cite{granger1986},\cite{dwyer1992}. However, \cite{richards1995}
+claims that cointegration is directly at odds with market efficiency, even
+though, there is no evidende that cointegration among asset prices have
+implications about market efficiency~\cite{lence+falk2005}.
+
+Despite the fact that cointegration on closing daily rates of currency pairs has
+not been found \cite{coleman1990}, \cite{copeland1991}, different time series
+frequencies can have different behaviours~\cite{aldridge2009}. Pair trading is a
+very common example of cointegration application~\cite{herlemont2003} 
+but cointegration can also be extended to a larger set of
+variables~\cite{mukherjee1995},\cite{engle2004}.
 \chapter{Time Series}

### 2014-11-21 src/2_Finance.tex
avance finanzas

@@
 trading strategies. The aim of HFT is to benefit from market short-term pricing
-inefficiencies~\cite{chlistalla2011}.
+inefficiencies~\cite{chlistalla2011}. HFT is characterized for the use of
+high-speed and sophisticated quantitative and algorithmic computer applications
+for modeling and executing orders efficiently. In order to make fast decisions,
+HFT firms require speed access to trading center servers and sometimes they are
+physically near minimizing network latencies.
+
 High frequency trades are short-term positions that commonly end the trading
@@
 in to reduce the market impact of large institutionals positions.
+Figure~\ref{fig:HFTtimes} shows a survey done to traders about the holding time
+of a position opened. Most of them agreed that HFT refers to positions between
+1-second and 10-minutes. Overnight positions are avoided since they are more
+risky and also more expensives fees, HFT firms end the trading day closing all
+opened positions.
+
+\begin{figure}[!h]
+  %\vspace{-0.8cm}
+  \centering
+  \includegraphics[width=\textwidth]{img/HFTtradestime}
+  \caption{Holding time of an opened position of a high frequency trade.}
+  \label{fig:HFTtimes}
+\end{figure}
 
@@
 Figure~\ref{fig:HFTmarket} shows HFT trades percentage of US equity shares.
+
+One of the reasons of this fall is that exchange markets have adapted, having
+now faster and more transparent and efficient market structures than before.
+This has been possible due to its investment in technology enhancing
+reliability and stability of transactions of transactions.
+Even though this fall, HFT is still a major component of regulated markets and
+will probably remain a topic of interest for researchers in the near future.
 
@@
 
-One of the reasons of this fall is that exchange markets have adapted, having
-now faster and more transparent and efficient market structures than before.
-Even though this fall, HFT is still a major component of regulated markets and
-will probably remain a topic of interest for researchers in the near future.
-
-Despite the fact HFT has been criticized on qualitative issues concerning fariness and
-systemic risk, empirical research shows that HFT has lead to beneficial impacts
-such as tighter spreads, increased liquidity, more efficient price formation,
+Despite the fact HFT has been criticized on qualitative issues concerning
+fairness and systemic risk, empirical research shows that HFT has lead to
+beneficial impacts such as reducing spreads (difference between buyers and
+sellers prices), increased liquidity, allowing more efficient price formation,
 reduced transaction costs and lower market volatility.
@@
 
+\begin{figure}[!h]
+  %\vspace{-0.8cm}
+  \centering
+  \includegraphics[width=0.7\textwidth]{img/capitalmarkets}
+  \caption{Old structure of capital markets.}
+  \label{fig:capitalmarket}
+\end{figure}
+
 Figure~\ref{fig:capitalmarket} shows the typical structure of capital markets
@@
 
+This centralized structure existed until computer technology allowed a better
+communication structure. Today financial markets are more descentralized
+providing more liquidity. Exchanges and inter-dealer brokers were replaced by
+liquidity pools or Electronic Communication Networks (ECNs) which are able to
+transmit order quickly matching buyers and sellers optimally. There are also
+dark liquidity pools where trader identity and orders remain anonymous.
+
+Figure~\ref{fig:capitalmarketnow} shows current structucture of capital markets
+including ECNs and darl pools. In this structure, ECNs, Exchanges, dark pools,
+broker-dealers and retail brokerages can execute orders. However, there are some
+institutional clients have also become broke-dealers.
+
 \begin{figure}[!h]
@@
   \centering
-  \includegraphics[width=0.75\textwidth]{img/capitalmarkets}
-  \caption{Old structure of capital markets.}
-  \label{fig:capitalmarket}
+  \includegraphics[width=0.8\textwidth]{img/capitalmarketsnow}
+  \caption{Actual structure of capital markets.}
+  \label{fig:capitalmarketnow}
 \end{figure}
 
-This centralized structure existed until computer technology allowed a better
-communication structure. Today financial markets are more descentralized
-providing more liquidity.
+Equity market and foreign exchange market (Forex) are the most populat markets
+for high frequency trading strategies. In equity markets can be traded stocks
+such as futures and options, exchange-traded funds (ETFs) among others financial
+instruments.
+
+Forex market allows the trading of interest rates denominated pair currencies
+such as EURUSD or USDJPY. Traders of this market are diverse, some of them are
+high frequency traders, long-term investors and corporations. The Forex market
+used to be centralized, only commercial banks had exclusive access to
+inter-dealer networks. Today, Forex market is descentralized and has become the
+biggest market in terms of volume of trading. The main participants are
+international banks geographically dispersed with continuous 24 hours operations
+excepting weekends. Figure~\ref{fig:Forextimes} shows different market trading
+hours in GMT for London, New York, Sydney and Tokyo.
+
+\begin{figure}[!h]
+  %\vspace{-0.8cm}
+  \centering
+  \includegraphics[width=0.8\textwidth]{img/forex-trading-hours}
+  \caption{Forex market trading hours (GMT).}
+  \label{fig:Forextimes}
+\end{figure}
+
+\subsection{Price formation process}
+
+\subsection{Efficient market hypothesis}

### 2014-11-21 src/2_Finance.tex
avance 20

@@
 
+A trade consists at least of two orders: one to enter the market (buy order) and
+exit the market (sell order). Traders have different types of order they can
+execute:
+
+\begin{description}
+\item[Market order:] allow to buy or sell an asset at the current best price
+whatever that price is. The transaction price is determined by the market
+considering previous executed orders and volume requested.
+\item[Limit order:] allow to specify the price you want to buy or sell an asset.
+Depending on the market conditions, this type of orders can neven be executed.
+\item[Stop order:] are suitable for investors who are unable to monitor their
+investments for a period of time. Stop order allow to specify the price that a
+position should be closed called stop price. When the stop price is reached a
+market order is executed, this means that market price won't be exactly the same
+specified in the stop order.
+\end{description}
+
+In terms of commissions, limit order are more expensive than market orders and
+there is no certainty of execution.
+
+Moreover, all orders can specify other parameters such as:
+
+\begin{description}
+\item[Fill or Kill:] is an order that must be executed inmediately or being
+cancelled, no partial fulfillments are allowed. 
+\item[Day:] the order is only valid during the day.
+\item[Good til canceled:] a order is active until the investor decides to cancel
+ir or the trade is executed.
+\end{description}
+
+In order to determine the execution price, buyers and sellers orders are placed
+in an order book which help to determine which order can be fulfilled.
+Figure~\ref{fig:orderbook} illustrates how buyers and sellers are ordered. The
+ask or offer price is the current lower price a seller is willing to accept for
+a good. The bid price corresponds to the current highest price a buyer is
+willing to pay for a good. Orders with the same price are prioritorized by
+arrival time and placed on top of the book.  The difference between ask and bid
+price is called spread.
+
+\begin{figure}[!h]
+  %\vspace{-0.8cm}
+  \centering
+  \includegraphics[width=0.5\textwidth]{img/orderbook}
+  \caption{Order book. Buyers and sellers are ordered according to the bid or
+  ask price and the market determines the mid-price and transacted volume.}
+  \label{fig:orderbook}
+\end{figure}
+
+Today, order books are available and they are a popular source of study among
+researchers. To determine the mid-price and spread modeling are some of the
+problems studied.
+
 \subsection{Efficient market hypothesis}

### 2014-11-27 src/5_Proposal.tex
var y vec matrices

@@
 \end{equation}
+
+    \begin{tikzpicture}[node distance=2cm]
+    %\node (start) [startstop] {Start};
+    %\node (in1) [io, below of=start] {Input};
+    \node (in1) [io] {Input: y, p, L, $\epsilon$, n};
+    \node (pro1) [process, below of=in1] {$y_i = y[i:i+L]$};
+    \node (dec1) [decision, below of=pro1, yshift=-0.5cm] {i = 0};
+    \node (pro2a) [process, below of=dec1, yshift=-0.5cm] {$[\mathbf{A} \quad \mathbf{B}] =\texttt{ovecm}(\mathbf{y}_i,p,v,\mathbf{A},\mathbf{B})$
+                                                        $\Delta \mathbf{Y}_{\text{true}}[i] = \mathbf{B}[-1,:]$
+                                                        $\Delta \mathbf{Y}_{\text{pred}}[i] = \mathbf{A}[-1,:] \times \mathbf{X}$};
+
+    \node (pro2b) [process, right of=dec1, xshift=4cm] {$v = \texttt{Johansen}(\mathbf{y}_i,p)$ $[\mathbf{A} \quad 
+                                                                \mathbf{B}] = \texttt{vecm}(\mathbf{y}_i,p,v)$};
+    \node (pro3) [process, below of=pro2a, yshift=-0.5cm] {$\mathbf{X} = \text{OLS} (\mathbf{A},\mathbf{B})$ \\
+                                                        $\mathbf{y}_{\text{true}}[i] = \mathbf{y}_i[-n,:]$ \\
+                                                        $\mathbf{y}_{\text{pred}}[i] = \mathbf{A}[-n,:] \times \mathbf{X} +$ \\$ \mathbf{y}_i[-n-1,:-1]$ \\
+                                                        $e = \texttt{mape}(\mathbf{y}_{\text{true}}, \mathbf{y}_{\text{pred}})$};
+
+    \node (dec2) [decision, below of=pro3, yshift=-0.5cm] { $e > \epsilon $ };
+
+    \node (pro4a) [process, right of=dec2, xshift=4cm] {$v = \texttt{Johansen}(\mathbf{y}_i,p)$ \\
+                                                        $\mathbf{A} = \texttt{uvecm}(\mathbf{y}_i,p,v,\mathbf{A})$ \\
+                                                        $\mathbf{X} = \texttt{OLS} (\mathbf{A},\mathbf{B})$};
+
+    \node (pro5) [process, left of=pro2a, xshift=-4cm] {$ i++ $};
+
+    \node (dec3) [decision, below of=dec2, yshift=-0.5cm] { $i > it $ };
+
+    \node (out1) [io, below of=dec3] {Output: $\Delta \mathbf{y}_{true}, \Delta \mathbf{y}_{pred}$};
+
+    %\node (stop) [startstop, below of=out1] {Stop};
+
+    %\draw [arrow] (start) -- (in1);
+    \draw [arrow] (in1) -- (pro1);
+    \draw [arrow] (pro1) -- (dec1);
+    \draw [arrow] (dec1) -- node[anchor=east] {no} (pro2a);
+    \draw [arrow] (dec1) -- node[anchor=south] {si} (pro2b);
+    \draw [arrow] (pro2a) -- (pro3);
+    \draw [arrow] (pro2b) |- (pro3);
+    \draw [arrow] (pro3) -- (dec2);
+    \draw [arrow] (dec2) --  node[anchor=south] {si} (pro4a);
+    \draw [arrow] (pro4a) |- (dec3);
+
+    \draw [arrow] (dec2) --  node[anchor=east] {no} (dec3);
+    \draw [arrow] (dec3.west) -| node[anchor=east] {no} (pro5);
+    \draw [arrow] (pro5) |- (pro1);
+    \draw [arrow] (dec3) --  node[anchor=east] {si} (out1);
+    %\draw [arrow] (out1) -- (stop);
+    \end{tikzpicture}

### 2014-12-17 src/2_Finance.tex
avance time series

@@
 opened positions.
-
 \begin{figure}[!h]
@@
 \end{figure}
-
 In 2014 HFT represents nearly 50\% of equity trades in the US and more than 20\%
@@
 
-\subsection{Financial Markets}
+\section{Financial Markets}
 
@@
 
-\subsection{Price formation process}
+\section{Price formation process}
 
@@
 
-\subsection{Efficient market hypothesis}
+\section{Stylized facts of asset returns}
+\label{sec:stylizedfacts}
+
+There are several known features exhibited by financial instruments called
+stylized facts, which have been empirically studied and some of them have been
+documented only recently and accepted as truth. Stylized facts are usually
+formulated in terms of qualitative properties of asset returns and may not be
+precise enought to distinguish among different parametric models.
+\cite{cont2001}.
+
+Returns of a time series $y$ is defined as:
+\begin{equation}
+\label{eq:returndef}
+r_t = y_t - y_{t-1}  \, ,
+\end{equation}
+
+Some stylized facts which are common to a wide set of financial assets
+\cite{sewell2011} are:
+
+\begin{description}
+\item[Dependence] autocorrelation in returns is largely insignificant. However,
+autocorrelation in the absolute and squared returns is always positive and
+significant and decays slowly. In addition, the autocorrelation in the absolute
+returns is generally higher than the autocorrelation in the corresponding
+squared returns.
+\item[Distribution] The distribution of returns is approximately symmetric and
+has high kurtosis (i.e fat tails and a peaked centre compared with the normal
+distribution). However, distribution of returns whose were obtained from higher
+frequencies looks more like a normal distribution.
+\item[Heterogeneity] despite the fact that financial returns are non-stationary,
+stationary periods can be observed. Economist speaks in terms of a structural
+break \cite{stock1994}, in machine learning this is known as drift which means
+that the statistical properties of the target variable change over time
+\cite{widmer1996}, \cite{tsymbal2004}.
+\item[Non-Linearity] financial returns may be non-linear in mean and/or
+non-linear in variance.
+\item[Calendar effects] are also called seasonal effects and they are cyclical
+anomalies in returns where the cycle is based on the calendar. Some of known
+calendar effects are: intraday effect, weekend effect, Monday effect, intramonth
+effect, the January effect and Holiday effect. The most important calendar
+anomalies are the January effect and the weekend effect. Figure
+\ref{fig:mondayeffect} shows the Monday effect from 1928 to 2007. It shows that
+in average Monday has negative returns. However, this is changing in time,
+figure \ref{fig:mondayeffect2} shows returns only for year 2007 and its last
+three months and it shows that Mondays and Wednesdays has positive returns in
+average.
+
+\begin{figure}[h!]
+\centering
+\begin{subfigure}[b]{0.45\textwidth}
+ \includegraphics[width=\textwidth]{img/average_daily_change_1928_2007}
+ \caption{Average daily return: 1928-2007}
+ \label{fig:mondayeffect}
+\end{subfigure}
+\begin{subfigure}[b]{0.45\textwidth}
+ \includegraphics[width=\textwidth]{img/thumb-average_daily_change_2007}
+ \caption{Average daily return: 2007}
+ \label{fig:mondayeffect2}
+\end{subfigure}
+\end{figure}
+
+\end{description}
+
+\begin{description}
+ \item[Dependence]
+
+ The returns
+ autocorrelation function (ACF) of SPY is shown in figure
+ \ref{fig:returnacf}.The figure shows how the correlations are significant even
+ for very long lags, this imply that a long-memory process because of its
+ long-range dependence. 
+ Moreover, volatility tends to cluster, i.e. high volatility
+ periods are generally followed by high volatility periods.
+
+ \begin{figure}[h]
+ \centering
+ \includegraphics[width=0.9\textwidth]{plots/spy_returns_acf.pdf}
+ % spy_returns_dist.pdf: 504x504 pixel, 72dpi, 17.78x17.78 cm, bb=0 0 504 504
+ \caption{SPY returns ACF}
+ \label{fig:returnacf}
+\end{figure}
+
+ \item[Returns distribution] The returns distribution has fat tails, i.e. the
+ number of extreme events (either positive or negative) returns is larger than
+ what is hypothesized by common data generation process (generally normal
+ distribution assumption). In the markets, fat tails are an undesirable feature
+ because of the additional risk they imply.  In the figure \ref{fig:returndist}
+ is shown the SPY returns distribution based on daily dates from the period 1st
+ July 1998 to 4th April 2013. The distribution was compared against the normal
+ distribution which clearly doesn't fit the data.  
+ \begin{figure}[h]
+ \centering
+ \includegraphics[scale=0.5]{plots/spy_returns_dist.pdf}
+ % spy_returns_dist.pdf: 504x504 pixel, 72dpi, 17.78x17.78 cm, bb=0 0 504 504
+ \caption{SPY returns distribution}
+ \label{fig:returndist}
+\end{figure}
+ \item[Volume] refers to the level of trading activity in a market.
+ \item[Calendar effects] days of week, weekend, holidays.
+ \item[Intraday effects] time of the day.
+ \item[Overnight effects]
+\end{description}
+
+\section{Efficient market hypothesis}
+
+\section{Volatility in the financial markets}
+
+In financial markets, volatility is one of the key elements to model the
+stochastic dynamic behaviour of financial assets. This is mainly because
+volatility gives a measure of uncertainty about future returns and it is often
+viewed as an indicator of the vulnerability of financial markets. It is also
+important as an input parameter in problems like derivative pricing, hedging
+and portfolio management.  For example, in option pricing we need to known
+first the volatility of the underlying asset before pricing an option.
+
+Besides, volatility provides important information for studying asset returns
+because the latter are largely uncorrelated and nonlinearly dependent. However,
+it has been found that volatility exhibits significant autocorrelation and
+predictable patterns~\cite{poon+granger2003} and  many models have been
+proposed to forecast its behaviour.
+
+The volatility of the data is due to a large number of factors affecting the
+market, with some of them directly measurable such as historical prices,
+trends, supply and demand. However, others such as monetary policies and news
+are not directly measurable and are included in other models out of the scope
+of this thesis.
+
+\subsubsection{Types of volatility}
+
+The volatility of a stock is not directly observable~\cite{tsay2005,engle1993}.
+For example, daily volatility is not directly observable from only daily returns because there is only one observation in a trading day.
+If intraday data is available, then volatility could be estimated.
+However, intraday returns are not the only explanatory variables for volatility and several estimators have been proposed.
+These estimators are observable variables that are related to the latent variable of interest called volatility proxies~\cite{devilderetal2007}.
+Examples of volatility proxies are the following:
+
+\begin{description}
+\item[realized volatility:] is also known as historic volatility and
+it is the actual variance in the price of a stock over time.
+Realized volatility is measured in terms of the standard deviation
+using the historical stock prices. It is commonly calculated based on
+intraday price returns:
+\begin{equation}
+\label{eq:retintra}
+r_{t,n}=100(\ln(p_{t,n}) - \ln(p_{t,n-1}))
+\end{equation}
+\noindent where $p_{t,n}$ is the price observed at day $t=1,\dots,T$ and
+intraday sample $n=2,\dots,N$. Realized volatility is defined as:
+\begin{equation}
+\label{eq:rv}
+    \hat{\sigma}(t) = \sum_{n=1}^N r_{t,n}^2 \, , 
+\end{equation}
+\noindent where $N$ is the number of intraday samples and $T$ is the
+number of days. 
+In order to include overnight returns, Hansen and
+Lunde~\cite{hansen+lunde2005} introduced a scaling version of
+realized volatility using the following definitions:
+\begin{eqnarray}
+r_{t}&=&100(\ln(p_{t,N}) - \ln(p_{t-1,N})) \label{eq: ret 1} \\
+\bar{\rho}(t) &=& \sum_{t=1}^T r_{t}^2  \label{eq: ret 2} \, .
+\end{eqnarray}
+\noindent where equation~(\ref{eq: ret 1}) represents overnight 
+returns and the volatility as 
+equation~(\ref{eq: ret 2}), where $p_{t,N}$ is the last intraday 
+sample at day $t$. The scaled realized volatility $\rho(t)$ is 
+defined as:
+\begin{eqnarray}
+\label{eq:srv}
+\rho(t) = \gamma \hat{\rho}(t) \, , \qquad & \qquad \gamma = \displaystyle \frac{\bar{\rho}(t)}{\displaystyle\sum_{t=1}^T \hat{\rho}(t)}
+\end{eqnarray}
+Realized volatility has also been defined as the absolute value return or as
+the mean of the sum of intraday squared returns at short intervals of time. The
+majority of research carried out in the literature obtain the daily volatility
+as the daily squared returns as is shown in equation~(\ref{eq: ret 2}).
+However, it has been proven that this measurement noise is too high for
+observing the true volatility process~\cite{andersen+bollerslev1998}. Hansen
+and Lunde~\cite{hansen+lunde2006} stated that the use of a noisy proxy could
+result in an inferior model being chosen as the best one. The realized
+volatility, as calculated by the cumulative sum of squared intraday returns and
+shown in equation (\ref{eq:rv}), is less noisy and doesn't lead to choosing an
+inferior model.   
+\item[implied volatility:] volatility not only can be extracted from returns
+but it can also be derived from option or future pricing models.  The
+volatility obtained corresponds to the market's prediction of future
+volatility. In finance, an option is a derivative, that is, a contract which
+gives the owner the right, but not the obligation to buy or sell an underlying
+asset at a given price called strike price. An option can be executed at any
+time before an expiration date previously defined no matter what price the
+underlying asset has. For example, the Black-Scholes model~\cite{black1973}
+determines the fair option value based on stock price, strike price, time to
+option expiration, the interest rate and volatility. These are known or can be
+easily obtained from the market, excepting by volatility which must be
+estimated. However, rather than assuming a volatility a priori and computing
+option prices from it, the model can be used to estimate volatility at given
+prices, time to expiration and strike price. This obtained volatility is called
+the implied volatility of an option. Additionally, some models obtain implied
+volatility from futures (other derivative from prices). For instance, the
+Barone-Adesi and Whaley futures option model~\cite{baroneetal1987} is also used
+to determine future volatilities~\cite{hamidetal2004}. Higher implied
+volatility is indicative of greater price fluctuation in either direction.
+Implied volatility is found by determining the value which makes theoretical
+prices equal to market prices. In this way volatility is ``implied'' by the
+current market price of the stock.
+\end{description}
+
+For trading strategies, the interest is centred in forecasting realized
+volatility over the life of an option and to take advantage when this
+volatility differs from the implied volatility. This is called volatility
+arbitrage. For example, a trader will buy an option and hedge the underlying
+asset if the implied volatility is under the realized volatility.
+
+\subsubsection{Volatility methods}
+
+In the existing literature, there are four main classes of asset
+return volatility models: the general autoregressive conditional
+heteroskedasticity (GARCH) models, the stochastic volatility (SV)
+models, the realized volatility models and the machine learning based
+models. A comparison of the first three models can be found
+in~\cite{wei2012}.
+
+For many years the most popular methods for estimating financial
+volatility were the autoregressive conditional heteroskedasticity
+(ARCH) models~\cite{engle1982} and the general ARCH (GARCH)
+models~\cite{bollerslev1986}. For instance, the GARCH(1,1) defines
+returns $y_t$ and volatility $\sigma_t$ as:
+
+\begin{eqnarray*}
+    y_t &=& \sigma_t \epsilon_t \\
+     \sigma_t^2 &=& \alpha_0 + \alpha_1 \epsilon_{t-1}^2 + \beta_1
+     \sigma_{t-1}^2
+\end{eqnarray*}
+
+\noindent where $\epsilon_t$ is standard Gaussian white noise,
+$\alpha_0,\alpha_1,\beta_1 \geq 0$ are required to ensure that the
+variance will never be negative and $\alpha_1+\beta_1 <1$ is needed to
+guarantee a weakly stationary process~\cite{nelson1990}.
+
+Since the introduction of the GARCH models, several extensions have been
+proposed, but none of them seems to beat the GARCH(1,1)
+model~\cite{lunde+hansen2005}. Despite its popularity, GARCH models have
+several limitations: firstly, a time series model may be non-linear in mean
+and/or non-linear in variance, but ARCH and GARCH models are non-linear in
+variance, but not in mean. Besides, GARCH models often fail to capture highly
+irregular phenomena, like wild market fluctuations.
+
+SV models explain how volatility varies in a random fashion. These models are
+useful because they explain why options with different strikes and expirations
+dates have different Black-Scholes implied volatilities, phenomenon known as
+the volatility smile. This is useful because the Black-Scholes model assumes
+that the volatility of the underlying asset is constant which is not always
+true. There are several SV models and the most well-known and popular is the
+Heston model~\cite{heston1993}. Additional information about SV models can be
+found in~\cite{shephard1995}.
+
+The realized volatility constructed from high frequency intraday returns gave
+rise to the realized volatility models mainly because the realized volatility
+series is much more homoskedastic and seems to be a long memory
+process~\cite{andersonetal2003}. For realized volatility, the autoregressive
+fractionally integrated moving average (ARFIMA) process emerged as a standard
+model~\cite{chenetal2010} and many variations have been studied, but all of
+them produce similar forecasting results to the ARFIMA(1,d,1)
+model~\cite{koopmanetal2005}.
+
+On the other hand, machine learning based models, especially artificial neural
+networks (ANN) and support vector machines (SVM) have arisen as an alternative
+to forecast volatility. ANN is a statistical technique inspired by biological
+neural networks which is capable of changing its structure based on external or
+internal information during a training phase~\cite{sammut2011}. SVM are
+supervised learning models for classification analysis which recognize patterns
+finding a separating hyperplane. An extension for regression analysis is known
+as support vector regression (SVR).
+
+Since machine learning models and in particular ANN do not require assumptions
+about the data (gaussianity for example) and allow more explanatory variables
+than returns to be included, they have become widely used in solving financial
+problems, specially volatility
+forecasting~\cite{hamidetal2004,donaldsonetal1997}. There are also many works
+focused on the using of SVM in volatility
+forecasting~\cite{shiyietal2008,shiyietal2010,gavrishchaka2006,vasilios2012}.
+
+However, just as with ANN, SVMs have scalability problems because their
+training process is computationally intensive and it is done in batch mode. The
+scalability problem worsens when new additional training data is available and
+a re-training process from scratch needs to be done. This problem can be
+avoided using online machine learning algorithms that allow one instance at a
+time to be processed with low computationally expensive calculations.
+
+%Volatility models have been extensively
+%studied~\cite{gatheral2006,poon+granger2003,knight2002} and some of  them
+%include machine learning
+%methods~\cite{hamidetal2004,donaldsonetal1997,shiyietal2008,shiyietal2010,gavrishchaka2006,vasilios2012}.  
+%

### 2014-12-17 src/3_TimeSeries.tex
var y vec matrices
avance time series

@@
-\chapter{Time Series}
+\chapter{Time Series models}
 
+\vspace{0.5cm}
+
+A time series is a collection of observations in time (discrete or continuous).
+The analysis of time series main objective is to find possible internal
+structure in the data such as autocorrelation, trend or seasonal variation.
+Some of application and uses of time series analysis are data compression,
+explanatory variables (relationships with other variables, seasonal factors,
+etc.), signal processing, forecasting (predict future values). This chapter
+reviews the more relevant techniques in the rich and rapidly growing field of
+time series analysis.
+
+\section{Characteristics of Time Series}
+
+\subsection{Stationary }
+A strictly stationary times series $y_t$ is one for which the probabilistic behavior
+of every collection of values $\{y_{t_1},y_{t_2},\dots,y_{t_L}\}$ is identical
+to that of the time shifted set, more precisely: \[ P\{y_{t_1} \leq
+c_1,\dots,y_{t_L} \leq c_L\} = P\{y_{t_1+h} \leq c_1,\dots,y_{t_L+h} \leq c_k\}
+\quad \forall L \in \mathbb{N}, \forall h \in \mathbb{Z}\] \noindent where
+$c_1,\dots,c_L$ are constants.  This definition is too strong and difficult to
+assess it from a single data set. The weak version of this definition imposes
+conditions only on the two first two moments.
+
+A weakly stationary time series is a process which mean, variance and auto
+covariance do not change over time: \begin{eqnarray*} E(Y_t) &=& \mu  \quad
+\forall t \in \mathbb{N} \\ E(Y^2_t) &=& \sigma^2  \quad \forall t \in
+\mathbb{N} \\ \lambda(s,t)&=&\lambda(s+h,t+h) \quad \forall s,t \in \mathbb{N},
+\forall h \in \mathbb{Z} \end{eqnarray*}
+
+\noindent with $\lambda(s,t) = E[(y_s-\mu)(y_t - \mu)]$
+
+\subsection{Integration}
+
+Following Johansen \cite{johansen1995} we shall say that a stochastic process
+$Y_t$ which satisfies $Y_t-E(Y_t) = \sum_{i=0}^\infty C_i\,\varepsilon_{t-i}$ is
+called $I(0)$, and then we shall write $Y_t\sim I(0)$, whenever
+$\sum_{i=0}^\infty C_i \neq 0$ and $\sum_{i=0}^\infty C_i\,z^i$ converges for
+$z\in\mathbb{C}$ with $|z|<1$.  It is understood that the condition
+$\varepsilon_t\sim iid(0,\sigma^2)$ holds.
+
+A (vector) time series $\mathbf{y}_t$ is said to be {\em integrated of order\/}
+$d$, and then we shall write $\mathbf{y}_t\sim I(d)$, whenever after $d$ times
+(discrete) differentiation an stationary process is
+obtained~\cite{banerjee1993};
+more precisely, whenever
+$(1-L)^d\,\mathbf{y}_t\sim\text{I(0)}$, where $L$ is the usual lag operator:
+$(1-L)\,\mathbf{y}_t = \Delta\mathbf{y}_t = \mathbf{y}_t-\mathbf{y}_{t-1}$ for
+all $t$.
+
+Note that this definition includes the scalar case as time series of
+vectors of dimension 1; in this scalar case we will write the time series in
+nonbold format.
+
+\subsection{Cointegration}
 Cointegration concept was introduced by~\cite{engle1987} and implies that one or
@@
 
-This idea was inmediatly adopted in finance since it could represent their
-long-run relationship implied by economic theory~\cite{laietAl1991},
-\cite{lence+falk2005}.  Economic theory suggest that economic time series are
-mean-reverting process and therefore, it reflects the idea of that some set of
-variables cannot wander too far from each other.
+Let $\mathbf{y}_t^\nu$, $\nu=1,\dots,p$, be a set of $p$ vector time series of
+order $I(1)$.  They are said to be {\em cointegrated\/} if a vector
+$\beta=[\beta(1),\dots,\beta(p)]^\top \in \mathbb{R}^p$ exists, such that the
+time series,
+\begin{equation}
+\mathbf{Z}_t:= 
+\sum_{\nu=1}^p \beta(\nu)\,\mathbf{y}_t^\nu\,\sim\,\text{I(0)}\,.
+\end{equation}
+
+In other words, a set of $I(1)$ time series is said to be cointegrated if a
+linear combination of them exists, which is I(0).
+
+The idea of cointegration was inmediatly adopted in finance since it could
+represent their long-run relationship implied by economic
+theory~\cite{laietAl1991}, \cite{lence+falk2005}.  Economic theory suggest that
+economic time series are mean-reverting process and therefore, it reflects the
+idea of that some set of variables cannot wander too far from each other.
 
 On the other hand, the efficient markets hypothesis, also known as the random
-walk theory states that current stock prices fully reflect available
-information related to its value and there is no way to earn excess
-profits~\cite{fama1970}.  This means that if we have stock prices from a
-jointly efficient market, they cannot be
-cointegrated~\cite{granger1986},\cite{dwyer1992}. However, \cite{richards1995}
-claims that cointegration is directly at odds with market efficiency, even
-though, there is no evidende that cointegration among asset prices have
-implications about market efficiency~\cite{lence+falk2005}.
+walk theory states that current stock prices fully reflect available information
+related to its value and there is no way to earn excess profits~\cite{fama1970}.
+This means that if we have stock prices from a jointly efficient market, they
+cannot be cointegrated \cite{granger1986}, \cite{dwyer1992}. However,
+\cite{richards1995} claims that cointegration is directly at odds with market
+efficiency, even though, there is no evidende that cointegration among asset
+prices have implications about market efficiency~\cite{lence+falk2005}.
 
@@
 frequencies can have different behaviours~\cite{aldridge2009}. Pair trading is a
-very common example of cointegration application~\cite{herlemont2003} 
-but cointegration can also be extended to a larger set of
+very common example of cointegration application~\cite{herlemont2003} but
+cointegration can also be extended to a larger set of
 variables~\cite{mukherjee1995},\cite{engle2004}.
-\chapter{Time Series}
-\section{Implied volatility models}
 
-\section{Black-Scholes formula}
-
-In finance, an option is a derivative, that is, a contract which gives the owner the right, but not the obligation to buy or sell an underlying asset at a given price called strike price.
-An option can be executed at any time before an expiration date previously defined no matter what price the underlying asset has.
-
-The Black-Scholes formula~\cite{black1973}, developed in the early 1970's, Myron Scholes, Robert Merton and Fisher Black,  allows to determine an option value $V$ based on the underlying asset price $S(t)$ at a time $t$ and the following constant parameters:
-
-\begin{description}
-\item [$\sigma$:] underlying asset price volatility which measures the standard deviation of the returns
-\item [$\mu$:] underlying asset drift which is a measure of the average rate of growth of the stock
-\item[$E$:] option strike or excersice price
-\item[$T$:] option date of expiry
-\item[$t$:] current time
-\item[$r$:] risk-free interest rate
-\end{description}
-
-\subsection{Stock price model}
-
-The Black-Scholes model assume that the underlying price $S$ follows a lognormal random walk:
-
-\begin{equation}\label{eq:stockprice}
-dS = \mu S dt + \sigma S dB
-\end{equation}
-
-\noindent where $B$ is a Brownian motion. This stochastic differential equation has two components: a deterministic term given by $\mu S dt$ and a random term given by $\sigma S dX$.
-
-\subsection{Brownian motion}
-
-A brownian motion $B$ (also called a Wiener process) is a stochastic process characterized by the three following properties:
-
-\begin{description}
-\item[Continuity:] $B(t)$ is a continuos function
-\item[Normal increments:]  $B(t)-B(s)$ has a normal distribution with mean $0$ and variance $t-s$.
-\item[Independence of increments:] for every choice of nonnegative real numbers $0 \leq s_1 <  t_1 \leq \cdots \leq s_n < t_n < \infty$, the increment random variables $W_{t_1} - W_{s_1}, \cdots, W_{t_n} - W_{s_n}$ are jointly independent.
-\end{description}
-
-\subsection{Stochastic differential equation}
-
-An stochastic differential integral has the form:
-
-\begin{equation}
-W(T)=\int_0^T f(t)dB(t) \, .
-\end{equation}
-
-\noindent This equations is also expressed in an abbreviate form:
-
-\begin{equation}
-dW = f(t) dB \, .
-\end{equation}
-
-\noindent Therefore, the integral form of the stock price model shown in equation (\ref{eq:stockprice}) is:
-
-\begin{equation}
-S(T)=\int_0^T \mu S(t) dt + \int_0^T \sigma S(t) dB(t)
-\end{equation}
-
-\subsection{Ito's Lemma}
-
-Ito's lemma is used to find the differential of a time dependent function of a stochastic process.
-In option pricing we need to find the option price $V(S(t))$ which depends on a stochastic stock price model $S(t)$.
-$V(S,t)$ is required to be  differentiable function of $S$ and once differentiable function of $t$.
-
-%These are known or can be easily obtained from the market, excepting by volatility which must be estimated.
-However, rather than assuming a volatility a priori and computing option prices from it, the model can be used to estimate volatility at given prices, time to expiration and strike price.
-This obtained volatility is called the implied volatility of an option.
-Additionally, some models obtain implied volatility from futures (other derivative from prices).
-
-\section{Realized Volatility Models}
-
-\section{GARCH}
-test
-\section{ARFIMA}
-\section{Stochastic Volatility}
-
-\section{Volatility definition}
-
-\begin{itemize}
-
-\item Is the size of the price movement.
-
-\item Variance is a measure of distribution of returns and is not neccesarily bound by any time period.
-Volatility is a measure of the standard deviation (square root of the variance) over a certain time interval.
-In finance, variance and volatility both gives you a sense of an asset's risk.
-Variance gives you a sense of the risk in the asset over its lifetime, while volatility gives you a sense of the movement of the asset in eg.
-the past month or the past year.
-
-\item The main underlying difference is in their definition.
-Variance has a fixed mathematical definition, however volatility does not as such.
-Volatility is said to be the measure of fluctuations of a process.
-
-Volatility is a subjective term, whereas variance is an objective term i.e.
-given the data you can definitely find the variance, while you can't find volatility just having the data.
-Volatility is associated with the process, and not with the data.
-
-In order to know the volatility you need to have an idea of the process i.e you need to have an observation of the dispersion of the process.
-All the different processes will have different methods to compute volatilities based on the underlying assumptions of the process.
-\end{itemize}
-
-\subsubsection{Types of volatility}
+\subsection{Volatility}
 
 The volatility of a stock is not directly observable~\cite{tsay2005,engle1993}.
-For example, daily volatility is not directly observable from only daily returns because there is only one observation in a trading day.
-If intraday data is available, then volatility could be estimated.
-However, intraday returns are not the only explanatory variables for volatility and several estimators have been proposed.
-These estimators are observable variables that are related to the latent variable of interest called volatility proxies~\cite{devilderetal2007}.
+For example, daily volatility is not directly observable from only daily returns
+because there is only one observation in a trading day.  If intraday data is
+available, then volatility could be estimated. However, intraday returns are not
+the only explanatory variables for volatility and several estimators have been
+proposed. These estimators are observable variables that are related to the
+latent variable of interest called volatility proxies~\cite{devilderetal2007}.
 Examples of volatility proxies are the following:
@@
 
-Realized volatility has also been defined as the absolute value return or as the mean of the sum of intraday squared returns at short intervals of time.
-The majority of research carried out in the literature obtain the daily volatility as the daily squared returns as is shown in equation~(\ref{eq: ret 2}).
-However, it has been proven that this measurement noise is too high for observing the true volatility process~\cite{andersen+bollerslev1998}.
-Hansen and Lunde~\cite{hansen+lunde2006} stated that the use of a noisy proxy could result in an inferior model being chosen as the best one.
-The realized volatility, as calculated by the cumulative sum of squared intraday returns and shown in equation (\ref{eq:rv}), is less noisy and doesn't lead to choosing an inferior model.
+Realized volatility has also been defined as the absolute value return or as the
+mean of the sum of intraday squared returns at short intervals of time. The
+majority of research carried out in the literature obtain the daily volatility
+as the daily squared returns as is shown in equation~(\ref{eq: ret 2}).
+However, it has been proven that this measurement noise is too high for
+observing the true volatility process~\cite{andersen+bollerslev1998}. Hansen and
+Lunde~\cite{hansen+lunde2006} stated that the use of a noisy proxy could result
+in an inferior model being chosen as the best one. The realized volatility, as
+calculated by the cumulative sum of squared intraday returns and shown in
+equation (\ref{eq:rv}), is less noisy and doesn't lead to choosing an inferior
+model.
 
-\item[implied volatility:] volatility not only can be extracted from returns but it can also be derived from option or future pricing models.
-The volatility obtained corresponds to the market's prediction of future volatility.
-In finance, an option is a derivative, that is, a contract which gives the owner the right, but not the obligation to buy or sell an underlying asset at a given price called strike price.
-An option can be executed at any time before an expiration date previously defined no matter what price the underlying asset has.
-For example, the Black-Scholes model~\cite{black1973} determines the fair option value based on stock price, strike price, time to option expiration, the interest rate and volatility.
-These are known or can be easily obtained from the market, excepting by volatility which must be estimated.
-However, rather than assuming a volatility a priori and computing option prices from it, the model can be used to estimate volatility at given prices, time to expiration and strike
-price.
-This obtained volatility is called the implied volatility of an option.
-Additionally, some models obtain implied volatility from futures (other derivative from prices).
-For instance, the Barone-Adesi and Whaley futures option model~\cite{baroneetal1987} is also used to determine future volatilities~\cite{hamidetal2004}.
-Higher implied volatility is indicative of greater price fluctuation in either direction.
-Implied volatility is found by determining the value which makes theoretical prices equal to market prices.
-In this way volatility is ``implied'' by the current market price of the stock.
+\item[implied volatility:] volatility not only can be extracted from returns but
+it can also be derived from option or future pricing models.  The volatility
+obtained corresponds to the market's prediction of future volatility. In
+finance, an option is a derivative, that is, a contract which gives the owner
+the right, but not the obligation to buy or sell an underlying asset at a given
+price called strike price. An option can be executed at any time before an
+expiration date previously defined no matter what price the underlying asset
+has. For example, the Black-Scholes model~\cite{black1973} determines the fair
+option value based on stock price, strike price, time to option expiration, the
+interest rate and volatility. These are known or can be easily obtained from the
+market, excepting by volatility which must be estimated. However, rather than
+assuming a volatility a priori and computing option prices from it, the model
+can be used to estimate volatility at given prices, time to expiration and
+strike price. This obtained volatility is called the implied volatility of an
+option. Additionally, some models obtain implied volatility from futures (other
+derivative from prices). For instance, the Barone-Adesi and Whaley futures
+option model~\cite{baroneetal1987} is also used to determine future
+volatilities~\cite{hamidetal2004}. Higher implied volatility is indicative of
+greater price fluctuation in either direction. Implied volatility is found by
+determining the value which makes theoretical prices equal to market prices. In
+this way volatility is ``implied'' by the current market price of the stock.
 
@@
 
-For trading strategies, the interest is centred in forecasting realized volatility over the life of an option and to take advantage when this volatility differs from the implied volatility.
-This is called volatility arbitrage.
-For example, a trader will buy an option and hedge the underlying asset if the implied volatility is under the realized volatility.
+For trading strategies, the interest is centred in forecasting realized
+volatility over the life of an option and to take advantage when this volatility
+differs from the implied volatility. This is called volatility arbitrage. For
+example, a trader will buy an option and hedge the underlying asset if the
+implied volatility is under the realized volatility.
 
-\section{Volatility methods}
+\subsubsection{Volatility methods}
 
@@
 
-Since the introduction of the GARCH models, several extensions have been proposed, but none of them seems to beat the GARCH(1,1) model~\cite{lunde+hansen2005}.
-Despite its popularity, GARCH models have several limitations: firstly, a time series model may be non-linear in mean and/or non-linear in variance, but ARCH and GARCH models are non-linear in variance, but not in mean.
-Besides, GARCH models often fail to capture highly irregular phenomena, like wild market fluctuations.
+Since the introduction of the GARCH models, several extensions have been
+proposed, but none of them seems to beat the GARCH(1,1)
+model~\cite{lunde+hansen2005}. Despite its popularity, GARCH models have several
+limitations: firstly, a time series model may be non-linear in mean and/or
+non-linear in variance, but ARCH and GARCH models are non-linear in variance,
+but not in mean. Besides, GARCH models often fail to capture highly irregular
+phenomena, like wild market fluctuations.
 
-SV models explain how volatility varies in a random fashion.
-These models are useful because they explain why options with different strikes and expirations dates have different Black-Scholes implied volatilities, phenomenon known as the volatility smile.
-This is useful because the Black-Scholes model assumes that the volatility of the underlying asset is constant which is not always true.
-There are several SV models and the most well-known and popular is the Heston model~\cite{heston1993}.
-Additional information about SV models can be found in~\cite{shephard1995}.
+SV models explain how volatility varies in a random fashion. These models are
+useful because they explain why options with different strikes and expirations
+dates have different Black-Scholes implied volatilities, phenomenon known as the
+volatility smile. This is useful because the Black-Scholes model assumes that
+the volatility of the underlying asset is constant which is not always true.
+There are several SV models and the most well-known and popular is the Heston
+model~\cite{heston1993}. Additional information about SV models can be found
+in~\cite{shephard1995}.
 
-The realized volatility constructed from high frequency intraday returns gave rise to the realized volatility models mainly because the realized volatility series is much more homoskedastic and seems to be a long memory process~\cite{andersonetal2003}.
-For realized volatility, the autoregressive fractionally integrated moving average (ARFIMA) process emerged as a standard model~\cite{chenetal2010} and many variations have been studied, but all of them produce similar forecasting results to the ARFIMA(1,d,1) model~\cite{koopmanetal2005}.
+The realized volatility constructed from high frequency intraday returns gave
+rise to the realized volatility models mainly because the realized volatility
+series is much more homoskedastic and seems to be a long memory
+process~\cite{andersonetal2003}. For realized volatility, the autoregressive
+fractionally integrated moving average (ARFIMA) process emerged as a standard
+model~\cite{chenetal2010} and many variations have been studied, but all of them
+produce similar forecasting results to the ARFIMA(1,d,1)
+model~\cite{koopmanetal2005}.
 
-On the other hand, machine learning based models, especially artificial neural networks (ANN) and support vector machines (SVM) have arisen as an alternative to forecast volatility.
-ANN is a statistical technique inspired by biological neural networks which is capable of changing its structure based on external or internal information during a training phase~\cite{sammut2011}.
-SVM are supervised learning models for classification analysis which recognize patterns finding a separating hyperplane.
-An extension for regression analysis is known as support vector regression (SVR).
+On the other hand, machine learning based models, especially artificial neural
+networks (ANN) and support vector machines (SVM) have arisen as an alternative
+to forecast volatility. ANN is a statistical technique inspired by biological
+neural networks which is capable of changing its structure based on external or
+internal information during a training phase~\cite{sammut2011}. SVM are
+supervised learning models for classification analysis which recognize patterns
+finding a separating hyperplane. An extension for regression analysis is known
+as support vector regression (SVR).
 
-Since machine learning models and in particular ANN do not require assumptions about the data (gaussianity for example) and allow more explanatory variables than returns to be included, they have become widely used in solving financial problems, specially volatility forecasting~\cite{hamidetal2004,donaldsonetal1997}.
-There are also many works focused on the using of SVM in volatility forecasting~\cite{shiyietal2008,shiyietal2010,gavrishchaka2006,vasilios2012}.
+Since machine learning models and in particular ANN do not require assumptions
+about the data (gaussianity for example) and allow more explanatory variables
+than returns to be included, they have become widely used in solving financial
+problems, specially volatility
+forecasting~\cite{hamidetal2004,donaldsonetal1997}. There are also many works
+focused on the using of SVM in volatility
+forecasting~\cite{shiyietal2008,shiyietal2010,gavrishchaka2006,vasilios2012}.
 
-However, just as with ANN, SVMs have scalability problems because their training process is computationally intensive and it is done in batch mode.
-The scalability problem worsens when new additional training data is available and a re-training process from scratch needs to be done.
-This problem can be avoided using online machine learning algorithms that allow one instance at a time to be processed with low computationally expensive calculations.
+However, just as with ANN, SVMs have scalability problems because their training
+process is computationally intensive and it is done in batch mode. The
+scalability problem worsens when new additional training data is available and a
+re-training process from scratch needs to be done. This problem can be avoided
+using online machine learning algorithms that allow one instance at a time to be
+processed with low computationally expensive calculations.
 
-\section{Stylized facts}
-\label{sec:stylizedfacts}
+%\section{Black-Scholes formula}
+%
+%In finance, an option is a derivative, that is, a contract which gives the owner
+%the right, but not the obligation to buy or sell an underlying asset at a given
+%price called strike price. An option can be executed at any time before an
+%expiration date previously defined no matter what price the underlying asset
+%has. 
+%
+%The Black-Scholes formula~\cite{black1973}, developed in the early 1970's, Myron
+%Scholes, Robert Merton and Fisher Black,  allows to determine an option value
+%$V$ based on the underlying asset price $S(t)$ at a time $t$ and the following
+%constant parameters: 
+%
+%\begin{description}
+%\item [$\sigma$:] underlying asset price volatility which measures the standard
+%deviation of the returns
+%\item [$\mu$:] underlying asset drift which is a measure of the average rate of
+%growth of the stock
+%\item[$E$:] option strike or excersice price
+%\item[$T$:] option date of expiry
+%\item[$t$:] current time
+%\item[$r$:] risk-free interest rate
+%\end{description}
+%
+%\subsection{Stock price model}
+%
+%The Black-Scholes model assume that the underlying price $S$ follows a lognormal random walk:
+%
+%\begin{equation}\label{eq:stockprice}
+%dS = \mu S dt + \sigma S dB
+%\end{equation}
+%
+%\noindent where $B$ is a Brownian motion. This stochastic differential equation
+%has two components: a deterministic term given by $\mu S dt$ and a random term
+%given by $\sigma S dX$.
+%
+%\subsection{Brownian motion}
+%
+%A brownian motion $B$ (also called a Wiener process) is a stochastic process
+%characterized by the three following properties:
+%
+%\begin{description}
+%\item[Continuity:] $B(t)$ is a continuos function
+%\item[Normal increments:]  $B(t)-B(s)$ has a normal distribution with mean $0$
+%and variance $t-s$.
+%\item[Independence of increments:] for every choice of nonnegative real numbers
+%$0 \leq s_1 <  t_1 \leq \cdots \leq s_n < t_n < \infty$, the increment random
+%variables $W_{t_1} - W_{s_1}, \cdots, W_{t_n} - W_{s_n}$ are jointly
+%independent.
+%\end{description}
+%
+%\subsection{Stochastic differential equation}
+%
+%An stochastic differential integral has the form:
+%
+%
+%\begin{equation}
+%W(T)=\int_0^T f(t)dB(t) \, .
+%\end{equation}
+%
+%\noindent This equations is also expressed in an abbreviate form:
+%
+%\begin{equation}
+%dW = f(t) dB \, .
+%\end{equation}
+%
+%\noindent Therefore, the integral form of the stock price model shown in
+%equation (\ref{eq:stockprice}) is:
+%
+%\begin{equation}
+%S(T)=\int_0^T \mu S(t) dt + \int_0^T \sigma S(t) dB(t)
+%\end{equation}
+%
+%\subsection{Ito's Lemma}
+%
+%Ito's lemma is used to find the differential of a time dependent function of a
+%stochastic process. In option pricing we need to find the option price $V(S(t))$
+%which depends on a stochastic stock price model $S(t)$.  $V(S,t)$ is required to
+%be  differentiable function of $S$ and once differentiable function of $t$.
+%
+%
+%
+%
+%
+%%These are known or can be easily obtained from the market, excepting by
+%volatility which must be estimated. However, rather than assuming a volatility
+%a priori and computing option prices from it, the model can be used to estimate
+%volatility at given prices, time to expiration and strike price. This obtained
+%volatility is called the implied volatility of an option. Additionally, some
+%models obtain implied volatility from futures (other derivative from prices). 
+%
+%
+% 
+%\section{Realized Volatility Models}
+%
+%\section{GARCH} test \section{ARFIMA} \section{Stochastic Volatility}
+%
+%\section{Volatility definition}
+%
+%\begin{itemize}
+%
+%\item Is the size of the price movement.
+%
+%\item Variance is a measure of distribution of returns and is not neccesarily
+%bound by any time period.  Volatility is a measure of the standard deviation
+%(square root of the variance) over a certain time interval. In finance,
+%variance and volatility both gives you a sense of an asset's risk. Variance
+%gives you a sense of the risk in the asset over its lifetime, while volatility
+%gives you a sense of the movement of the asset in eg. the past month or the
+%past year.
+%
+%
+%\item The main underlying difference is in their definition. Variance has a
+%fixed mathematical definition, however volatility does not as such. Volatility
+%is said to be the measure of fluctuations of a process.
+%
+%Volatility is a subjective term, whereas variance is an objective term i.e.
+%given the data you can definitely find the variance, while you can't find
+%volatility just having the data. Volatility is associated with the process, and
+%not with the data.
+%
+%In order to know the volatility you need to have an idea of the process i.e you
+%need to have an observation of the dispersion of the process. All the different
+%processes will have different methods to compute volatilities based on the
+%underlying assumptions of the process.  \end{itemize}
+%
 
-There are several known features exhibited by financial instruments called \textit{stylized facts}, which have been empirically studied and some of them have been documented only recently.
-The importance of \textit{stylized facts} is that they allow to improve volatility models showing that it is possible to model volatility dynamic.
+\subsection{Vector Autorregresive Models}\label{sec:varvec}
 
-\begin{description}
- \item[Dependence] Volatility is said to be a long-memory process because its autocorrelation function exhibits long-range dependence.
-The returns autocorrelation function is shown in figure \ref{fig:returnacf} and it shows how the correlations are significant even for very long lags, this imply that the dependence between events that are far apart diminishes very slowly with increasing time.
- Moreover, volatility tends to cluster, i.e. high volatility periods are generally followed by high volatility periods.
+%VECM is a special case of VAR model and both describe the joint behaviour
+%of a set of variables.
 
- \begin{figure}[h]
- \centering
- \includegraphics[scale=0.5]{plots/spy_returns_acf.pdf}
- % spy_returns_dist.pdf: 504x504 pixel, 72dpi, 17.78x17.78 cm, bb=0 0 504 504
- \caption{SPY returns ACF}
- \label{fig:returnacf}
-\end{figure}
+The VAR($p$) model is a general framework describing the behaviour of a
+set of $l$ endogenous variables as a linear combination of their last $p$
+values, where $l,p\in\mathbb{N}$. 
+In our case, each one of these $l$ variables is a scalar time series
+$y_{\lambda,t}$, $\lambda=1,\dots,l$, and we represent them all together
+at time $t$ by the the vector time series:
+\begin{equation}
+\label{eq:variables}
+\mathbf{y}_t = 
+\begin{bmatrix} y_{1,t} & y_{2,t} & \dots & y_{l,t} \end{bmatrix}^\top.
+\end{equation}
+\noindent
+Notice that the vectors $\mathbf{y}_t$ are assumed to be $l$-dimensional.
 
- \item[Returns distribution] The returns distribution has fat tails, i.e.
-the number of extreme events (either positive or negative) returns is larger than what is hypothesized by common data generation process (generally normal distribution assumption).
-In the markets, fat tails are an undesirable feature because of the additional risk they imply.
- In the figure \ref{fig:returndist} is shown the SPY returns distribution based on daily dates from the period 1st July 1998 to 4th April 2013.
-The distribution was compared against the normal distribution which clearly doesn't fit the data.
+The VAR($p$) model describes the behaviour of a dependent variable in terms of
+its own lagged values and the lags of the others variables in the system. The
+model with $p$ lags is formulated as the system of $N$:
+\begin{align}
+\label{eq:var}
+\mathbf{y}_t 
+= \boldsymbol{\Phi}_1 \mathbf{y}_{t-1} +
+  \boldsymbol{\Phi}_2 \mathbf{y}_{t-2} + \dots +
+  \boldsymbol{\Phi}_p\mathbf{y}_{t-p} +
+  \mathbf{c} + \boldsymbol{\epsilon}_t \nonumber \\
+t=p+1,\dots,N,
+\end{align}
+\noindent where 
+$\boldsymbol{\Phi}_1, \boldsymbol{\Phi}_2,\dots,\boldsymbol{\Phi}_p$
+are $l\times l$-matrices of real coefficients,
+$\boldsymbol{\epsilon}_{p+1},
+ \boldsymbol{\epsilon}_{p+2}, \dots, \boldsymbol{\epsilon}_N$ 
+are error terms, $\mathbf{c}$ is a constant vector and $N$ is the total
+number of samples.
 
- \begin{figure}[h]
- \centering
- \includegraphics[scale=0.5]{plots/spy_returns_dist.pdf}
- % spy_returns_dist.pdf: 504x504 pixel, 72dpi, 17.78x17.78 cm, bb=0 0 504 504
- \caption{SPY returns distribution}
- \label{fig:returndist}
-\end{figure}
+Notice that, regarding our notation of section (\ref{sec:coint}),
+we have here 
+$\mathbf{y}_t^0 = \mathbf{y}_t$,
+$\mathbf{y}_t^\nu = \mathbf{y}_{t-\nu}$ and
+the $\lambda$-th component of the vector time series $\mathbf{y}_t^\nu$
+is the scalar time series $y_{\lambda,t}^\nu$, where $\nu=1,\dots,p$ and
+$\lambda=1,\dots,l$.
+Transposing each equation of the system (\ref{eq:var}) we can write
+the VAR($p$) model in block-matrix form as:
+\begin{equation}\label{eq:vareq}
+\mathbf{B} = \mathbf{A} \mathbf{X} + \mathbf{E} \, , 
+\end{equation}
+%%
+\noindent where:
+%%
+\begin{alignat}{2}
+\mathbf{B}
+&= \begin{bmatrix}
+   \mathbf{y}_{p+1}^\top \\
+   \mathbf{y}_{p+2}^\top \\
+   \vdots \\
+   \mathbf{y}_N^\top
+   \end{bmatrix}_{(N-p)\times l}
+&\quad
+\mathbf{A}
+&= \begin{pmat}[{...|}]
+   \mathbf{y}_p^\top & \mathbf{y}_{p-1}^\top & \dots 
+                    & \mathbf{y}_1^\top & 1 \cr
+   \mathbf{y}_{p+1}^\top & \mathbf{y}_p^\top & \dots
+                       & \mathbf{y}_2^\top & 1 \cr
+   \vdots & \vdots & \ddots & \vdots & \vdots \cr
+   \mathbf{y}_{N-1}^\top & \mathbf{y}_{N-2}^\top & \dots 
+                       & \mathbf{y}_{N-p}^\top & 1 \cr
+   \end{pmat}_{(N-p)\times(pl+1)} \\
+\mathbf{X}
+&= \begin{bmatrix}
+   \boldsymbol{\Phi}_1^\top \\
+   \boldsymbol{\Phi}_2^\top \\
+   \vdots \\
+   \boldsymbol{\Phi}_p^\top \\
+   \mathbf{c}^\top
+   \end{bmatrix}_{(pl+1)\times l}
+&\quad
+\mathbf{E}
+&= \begin{bmatrix}
+   \boldsymbol{\epsilon}_{p+1}^\top \\
+   \boldsymbol{\epsilon}_{p+2}^\top \\
+   \vdots \\
+   \boldsymbol{\epsilon}_N^\top \\
+   \end{bmatrix}_{(N-p)\times l}
+\end{alignat}
+Taking into account the error term $\mathbf{E}$, equation~(\ref{eq:vareq}) 
+can be solved with respect to $\mathbf{X}$ using the ordinary least
+squares estimation.
 
- \item[Volume] refers to the level of trading activity in a market.
- \item[Calendar effects] days of week, weekend, holidays.
- \item[Intraday effects] time of the day.
- \item[Overnight effects]
-\end{description}
+VECM is a special form of a VAR model for I(1) variables that are also
+cointegrated~\cite{banerjee1993}.
+
+It is obtained re-writing equation (\ref{eq:var}) in terms of the new
+variable $\Delta\mathbf{y}_t=\mathbf{y}_t-\mathbf{y}_{t-1}$.
+The VECM model, expressed in terms those differences, takes the form:
+\begin{equation}\label{eq:vec}
+\Delta \mathbf{y}_t 
+= \boldsymbol{\Omega}\,\mathbf{y}_{t-1}
+  + \sum_{i=1}^{p-1} \boldsymbol{\Phi}_i^*\,\Delta\mathbf{y}_{t-i}
+  + \mathbf{c} + \boldsymbol{\epsilon}_t\,,
+\end{equation}
+\noindent
+where the coefficients matrices $\boldsymbol{\Phi}_i^*$ and 
+$\boldsymbol{\Omega}$, expressed in terms of the matrices
+$\boldsymbol{\Phi}_i$ of (\ref{eq:var}), are:
+\begin{align*}
+\boldsymbol{\Phi}_i^* 
+&:= -\sum_{j=i+1}^{p}\boldsymbol{\Phi}_j\,, \\
+\boldsymbol{\Omega}
+&:= -\left( \mathbb{I} - \boldsymbol{\Phi}_1 - \dots 
+    - \boldsymbol{\phi}_p \right)\,. 
+\end{align*}
+The following well known properties of the matrix $\boldsymbol{\Omega}$
+\cite{johansen1995} will be useful in the sequel:
+\begin{itemize}
+\item
+If $\boldsymbol{\Omega} = \mathbf{0}$, there is no cointegration.
+\item 
+If $rank(\boldsymbol{\Omega})=l$, i.e., if $\boldsymbol{\Omega}$ has
+full rank, then the time series are not I(1) but stationary.
+\item
+If $rank(\boldsymbol{\Omega})=r$, $0<r<l$, then there is cointegration
+and the matrix $\boldsymbol{\Omega}$ can be expressed as
+$\boldsymbol{\Omega}=\boldsymbol{\alpha\beta}^\top$, where $\boldsymbol{\alpha}$
+and $\boldsymbol{\beta}$ are
+$l\times r$ matrices and
+$\text{rank}(\boldsymbol{\alpha})=\text{rank}(\boldsymbol{\beta})=r$.
+\item
+The columns of $\boldsymbol{\beta}$ contains the cointegration vectors and the rows of
+$\boldsymbol{\alpha}$ correspond with the adjusted vectors. 
+$\boldsymbol{\beta}$ is obtained by Johansen procedure~\cite{johansen1988},
+whereas $\boldsymbol{\alpha}$ has to be determined as a variable in the VECM.
+\end{itemize}
+It is worth noticing that the factorization of the matrix
+$\boldsymbol\Omega$ is not unique, since for any $r \times r$
+nonsingular matrix $\mathbf{H}$, $\boldsymbol{\alpha}^*:=\boldsymbol{\alpha}\mathbf{H}$,
+and $\boldsymbol{\beta}^*=\boldsymbol{\beta}(\mathbf{H}^{-1})^\top$ we have
+$\boldsymbol{\alpha\beta}^\top=\boldsymbol{\alpha}^*(\boldsymbol{\beta}^*)^\top$.
+If cointegration exists, then equation (\ref{eq:vec}) can be written
+as follows:
+\begin{equation}\label{eq:vecfull}
+\Delta\mathbf{y}_t 
+= \boldsymbol{\alpha\beta}^\top\mathbf{y}_{t-1} 
+  + \sum_{i=1}^{p-1}\boldsymbol{\Phi}_i^*\,\Delta\mathbf{y}_{t-i}
+  + \mathbf{c} + \boldsymbol{\epsilon}_t\,,
+\end{equation}
+\noindent
+which is a VAR model but for time series differences.
+
+Transposing each equation of the system (\ref{eq:vecfull}) we can write
+the VECM($p$) model in block-matrix form as:
+\begin{equation}\label{eq:vareq}
+\mathbf{B} = 
+\mathbf{A} \mathbf{X} + 
+\mathbf{E} \, , 
+\end{equation}
+%
+\noindent where $\mathbf{B}$ dimension is $((N-p)\times l)$, $\mathbf{A}$
+dimension is $((N-p)\times(r+(p-1)l +1))$, $\mathbf{X}$ dimension is $((r+(p-1)l
++1)\times l)$ and $\mathbf{E}$ dimension is $((N-p)\times l)$:
+%
+\begin{alignat}{3}
+\mathbf{B}
+&= \begin{bmatrix}
+   \Delta\mathbf{y}_{p+1}^\top \\
+   \Delta\mathbf{y}_{p+2}^\top \\
+   \vdots \\
+   \Delta\mathbf{y}_N^\top
+   \end{bmatrix}
+&\quad
+\mathbf{X}
+&= \begin{bmatrix}
+   \boldsymbol{\alpha}^\top \\
+   \boldsymbol{\Phi}_1^{*\top} \\
+   \boldsymbol{\Phi}_2^{*\top} \\
+   \vdots \\
+   \boldsymbol{\Phi}_{p-1}^{*\top} \\
+   \mathbf{c}^\top
+   \end{bmatrix}
+&\quad
+\mathbf{E}
+&= \begin{bmatrix}
+   \boldsymbol{\epsilon}_{p+1}^\top \\
+   \boldsymbol{\epsilon}_{p+2}^\top \\
+   \vdots \\
+   \boldsymbol{\epsilon}_N^\top \\
+   \end{bmatrix}
+\end{alignat}
+\noindent and 
+\begin{align}
+\mathbf{A}
+&= \begin{pmat}[{....|}]
+   \mathbf{y}_p^\top \boldsymbol{\beta} & \Delta \mathbf{y}_p^\top & \Delta\mathbf{y}_{p-1}^\top & \dots 
+                    & \Delta\mathbf{y}_2^\top & 1 \cr
+   \mathbf{y}_{p+1}^\top  \boldsymbol{\beta} &\Delta\mathbf{y}_{p+1}^\top & \Delta\mathbf{y}_p^\top & \dots
+                       & \Delta\mathbf{y}_3^\top & 1 \cr
+   \vdots & \vdots & \vdots & \ddots & \vdots & \vdots \cr
+   \mathbf{y}_{N-1}^\top  \boldsymbol{\beta} &\Delta\mathbf{y}_{N-1}^\top & \Delta\mathbf{y}_{N-2}^\top & \dots 
+                       & \Delta\mathbf{y}_{N-p-1}^\top & 1 \cr
+   \end{pmat}\, .
+\end{align}
+Taking into account the error term $\mathbf{E}$, equation~(\ref{eq:vareq}) 
+can be solved with respect to $\mathbf{X}$ using the ordinary least
+squares estimation.
+
+\subsection{Ordinary Least Squares method}
+
+The solution $\widehat{\mathbf{A}}$ to
+equation~(\ref{eq:vareq}) can be obtained by the ordinary least squares (OLS)
+method. $\widehat{\mathbf{X}}$ is the solution of the quadratic optimization problem
+\begin{equation*}
+\widehat{\mathbf{X}} = \underset{\mathbf{X}}{\text{Arg\;min}}
+\|\mathbf{A}\mathbf{X}-\mathbf{B}\|_2^2
+\end{equation*}
+\noindent for which the solution $\widehat{\mathbf{X}}$ is well-known:
+\begin{equation*}
+\label{eq:MP}
+\widehat{\mathbf{X}}=\mathbf{A}^{\!\!+}\,\mathbf{B}
+\end{equation*}
+\noindent where $\mathbf{A}^{\!\!+}$ is the Moore-Penrose pseudo-inverse
+which, when $\mathbf{A}$ is full rank, can be written as follows: 
+\begin{equation}
+\label{eq:pseudoinverse}
+\mathbf{A}^{\!\!+}= (\mathbf{A}^{\!\!\top} \mathbf{A})^{-1}\mathbf{A}^{\!\!\top} \, .
+\end{equation}
+when $\mathbf{A}$ is not full rank, i.e
+$rank(\mathbf{A})=k <  n \leq m$, $\mathbf{A}^\top \mathbf{A}$ is
+always singular and equation~(\ref{eq:pseudoinverse}) cannot be used.
+More generally, the pseudo-inverse is best computed using the compact
+singular value decomposition (SVD) of $\mathbf{A}$ which is:
+%\begin{equation}
+%    \label{eq:compactsvd}
+%    \underset{m \times n}{\mathbf{A}}=
+%    \underset{m \times k}{\mathbf{U_1}} \enskip
+%    \underset{k \times k}{\Sigma_1} \enskip
+%    \underset{k \times n}{\mathbf{V}_1^{\top}} \, .
+%\end{equation}
+\begin{equation*}
+    \label{eq:compactsvd}
+    \mathbf{A}=
+    \mathbf{U_1}
+    \boldsymbol \Sigma_1
+    \mathbf{V}_1^{\top} \, .
+\end{equation*}
+Pseudo-inverse can then be written as follows:
+\begin{equation*}
+\label{eq:pseudoinversesvd}
+\mathbf{A}^{\!\!+} = \mathbf{V}_1 \boldsymbol \Sigma_1^{-1} \mathbf{U}_1^\top \, .
+\end{equation*}
+
+\begin{equation}
+ \label{eq:varmatrix}
+               \underbrace{ \left[ \begin{array}{c}
+                \mathbf{y}_{p+1} \\
+                \mathbf{y}_{p+2} \\
+                \vdots           \\
+                \mathbf{y}_N  
+               \end{array} \right]  }_{\substack{ \mathbf{B}^\top\\ (N-p) \times l}}   
+= 
+\underbrace{
+  \begin{bmatrix}
+    \phi_1 & \dots & \phi_p & \mathbf{c}   
+  \end{bmatrix}}_{\substack{ \mathbf{X}^\top\\ l \times  (l \times p+1)}}
+\underbrace{\begin{bmatrix}
+   \mathbf{y}_p     & \mathbf{y}_{p+1}  & \dots     & \mathbf{y}_{N-1}\\
+   \mathbf{y}_{p-1} & \mathbf{y}_{p}  & \dots     & \mathbf{y}_{N-2} \\
+   \vdots           & \vdots            & \ddots    & \vdots  \\
+   \mathbf{y}_{1}   & \mathbf{y}_{2}& \dots      & \mathbf{y}_{N-p} \\
+   1                &  1                & \dots      &             1 
+   \end{bmatrix}}_{\substack{ \mathbf{A}^\top\\  (l \times p+1) \times (N-p)}}
++
+\underbrace{\begin{bmatrix}
+              \mathbf{\epsilon}_{p+1}  \\ 
+              \mathbf{\epsilon}_{p+2}  \\ 
+               \vdots                 \\
+              \mathbf{\epsilon}_N     
+             \end{bmatrix}}_{\substack{\mathbf{E}^\top\\ (N-p) \times l }} 
+\end{equation}
+
+%\begin{equation}
+% \label{eq:vecmatrix}
+%\underbrace{
+%      \begin{bmatrix}
+%       \mathbf{\Delta y}_{p+1}  \\ 
+%       \mathbf{\Delta y}_{p+2}  \\ 
+%       \vdots                   \\ 
+%       \mathbf{\Delta y}_N      
+%      \end{bmatrix}}_{\mathbf{B} } =
+%\arraycolsep=1.4pt  
+%\underbrace{\begin{bmatrix}
+%   \alpha^\top & \phi^*_1 & \dots & \phi^*_{p-1} &\mathbf{c}   
+%  \end{bmatrix}}_{\mathbf{X}^\top}
+%\underbrace{\begin{bmatrix}
+% \beta\mathbf{y}_p      & \beta\mathbf{y}_{p+1}   & \dots&   \beta\mathbf{y}_{N-1}   \\
+% \mathbf{\Delta y}_{p}   & \mathbf{\Delta y}_{p+1}& \dots &  \mathbf{\Delta y}_{N-1} \\
+% \mathbf{\Delta y}_{p-1} & \mathbf{\Delta y}_{p}  & \dots &  \mathbf{\Delta y}_{N-2}   \\
+% \vdots                  & \vdots                 & \ddots&  \vdots                   \\
+% \mathbf{\Delta y}_{2}   & \mathbf{\Delta y}_{3} & \dots &   \mathbf{\Delta y}_{N-p+1} \\
+% 1                      & 1                       & \dots     & 1   
+% \end{bmatrix}}_{\mathbf{A}^\top}
+%+
+%\underbrace{\begin{bmatrix}
+%              \mathbf{\epsilon}_{p+1} \\ 
+%              \mathbf{\epsilon}_{p+2} \\ 
+%              \vdots \\ 
+%              \mathbf{\epsilon}_N
+%             \end{bmatrix}}_{\mathbf{E}^\top}
+%\end{equation}
+
+%VECMs may be estimated by Stata’s vec command. These models
+%
+%are employed because many economic time series appear to be
+%
+%‘first-difference stationary,’ with their levels exhibiting unit root or
+%
+%nonstationary behavior. Conventional regression estimators, including
+%
+%VARs, have good properties when applied to covariance-stationary
+%
+%time series, but encounter difficulties when applied to nonstationary or
+%
+%integrated processes.
+%
+%These difficulties were illustrated by Granger and Newbold
+%
+%(J. Econometrics, 1974) when they introduced the concept of spurious
+%
+%regressions. If you have two independent random walk processes, a
+%
+%regression of one on the other will yield a significant coefficient, even
+%
+%though they are not related in any way.
+%
+%
+%This insight, and Nelson and Plosser’s findings (J. Mon. Ec., 1982) that
+%
+%unit roots might be present in a wide variety of macroeconomic series
+%
+%in levels or logarithms, gave rise to the industry of unit root testing, and
+%
+%the implication that variables should be rendered stationary by
+%
+%differencing before they are included in an econometric model.
+%
+%Further theoretical developments by Granger and Engle in their
+%
+%celebrated paper (Econometrica, 1987) raised the possibility that two
+%
+%or more integrated, nonstationary time series might be cointegrated, so
+%
+%that some linear combination of these series could be stationary even
+%
+%though each series is not.
+%
+%
+%
+%If two series are both integrated (of order one, or I(1)) we could model
+%
+%their interrelationship by taking first differences of each series and
+%
+%including the differences in a VAR or a structural model.
+%
+%However, this approach would be suboptimal if it was determined that
+%
+%these series are indeed cointegrated. In that case, the VAR would only
+%
+%express the short-run responses of these series to innovations in each
+%
+%series. This implies that the simple regression in first differences is
+%
+%misspecified.
+%
+%If the series are cointegrated, they move together in the long run. A
+%
+%VAR in first differences, although properly specified in terms of
+%
+%covariance-stationary series, will not capture those long-run
+%
+%tendences.
+%
+%Accordingly, the VAR concept may be extended to the vector
+%
+%error-correction model, or VECM, where there is evidence of
+%
+%cointegration among two or more series. The model is fit to the first
+%
+%differences of the nonstationary variables, but a lagged error-correction
+%
+%term is added to the relationship.
+%
+%In the case of two variables, this term is the lagged residual from the
+%
+%cointegrating regression, of one of the series on the other in levels. It
+%
+%expresses the prior disequilibrium from the long-run relationship, in
+%
+%which that residual would be zero.
+%
+%In the case of multiple variables, there is a vector of error-correction
+%
+%terms, of length equal to the number of cointegrating relationships, or
+%
+%cointegrating vectors, among the series.

### 2015-01-26 src/0_Abstract.tex
Avance dia 1

@@
 cointegrated.  Cointegration has been then one of the main features studied.
-Vector error correction model (VECM) is an econometric model which characterizes
+Vector error correction model (VECM) is an econometric model which characterises
 the joint dynamic behaviour of a set of cointegrated variables in terms of
@@
 issues and is computationally expensive. Ridge regression is commonly used
-instead of OLS since they include a regularization parameter which could improve
+instead of OLS since they include a regularisation parameter which could improve
 prediction error.}
@@
 \emph{In this thesis, financial time series features are studied and different
-algorithms were developed in order to optimize parameters an increase performance.
-In particular, we propose an online VECM which optimizes how model parameters
-are obtained considering only past data and different ways of solving the model
-using machine learning techniques.  
-The long-run relationship between the time series is used in order to make
-optimizations and obtain improved execution times. However, due to the large
-amount of data available and the need of quick response, algorithms presented
-were optimized using high performance computing.
+algorithms were developed in order to optimise parameters and increase performance.
+In particular, I propose an online algorithm based on VECM which optimises how model parameters are obtained and reduces execution times.
+This is achieved considering only a sliding window of the last historical data and using machine learning techniques to solve the model.
+Moreover, the long-run relationship between the time series is used in order to make
+optimisations and obtain improved execution times. 
+Due to the large amount of financial data available and the need of quick response, 
+the algorithm presented
+was optimised using high performance computing.
 Our experiments were tested using synthetic data and foreign exchange rates.
 Results show that cointegration and high performance computing allow to obtain
-better models and execution times.}
+models with better performance accuracy and reduced execution times.}
 

### 2015-01-26 src/2_Finance.tex
Avance dia 1

@@
 short periods of time gaining a small profit in every transaction, the key is
-the amount of transactions that algorithms are capable to execute. 
+the amount of transactions that algorithms are capable to execute obtaining profit considering transaction costs. 
 In 2012 more than 50\% of US market share of exchange trades were HFT.  HFT has
@@
 trading strategies. The aim of HFT is to benefit from market short-term pricing
-inefficiencies~\cite{chlistalla2011}. HFT is characterized for the use of
+inefficiencies~\cite{chlistalla2011}. HFT is characterised for the use of
 high-speed and sophisticated quantitative and algorithmic computer applications
-for modeling and executing orders efficiently. In order to make fast decisions,
+for modelling and executing orders efficiently. In order to make fast decisions,
 HFT firms require speed access to trading center servers and sometimes they are
-physically near minimizing network latencies.
+physically near minimising network latencies.
 
@@
 day avoiding leaving opened positions to the next business day. HFT is
-frequently associate to Algorithm trading strategies, but the former is focused
-in to reduce the market impact of large institutionals positions.
+frequently associate to algorithm trading strategies, but the former is focused
+in to reduce the market impact of large institutional positions and the later refers to trade execution strategies that are typically used by fund managers to buy or sell large amounts of assets.
+The duration of the positions in HFT it is not well-defined.
 Figure~\ref{fig:HFTtimes} shows a survey done to traders about the holding time
@@
 1-second and 10-minutes. Overnight positions are avoided since they are more
-risky and also more expensives fees, HFT firms end the trading day closing all
+risky and also have more expensive fees, HFT firms end the trading day closing all
 opened positions.
+
 \begin{figure}[!h]
@@
 \end{figure}
-In 2014 HFT represents nearly 50\% of equity trades in the US and more than 20\%
-in Europe showing a consistent fall since 2009 which was its best year. Revenues
+In 2014 HFT represented nearly 50\% of equity trades in the US and more than 20\%
+in Europe showing a consistent fall since 2009 which was its best year. US market Revenues
 have fallen dramatically.
-Figure~\ref{fig:HFTmarket} shows HFT trades percentage of US equity shares.
-
-One of the reasons of this fall is that exchange markets have adapted, having
-now faster and more transparent and efficient market structures than before.
-This has been possible due to its investment in technology enhancing
-reliability and stability of transactions of transactions.
-Even though this fall, HFT is still a major component of regulated markets and
-will probably remain a topic of interest for researchers in the near future.
+Figure~\ref{fig:HFTmarket} on the left shows HFT trades percentage of US and Europe equity shares. The right figure shows how revenues in US stocks have fallen the last years.
 
@@
 \end{figure}
+One of the reasons of this fall is that exchange markets have adapted, having
+now faster and more transparent and efficient market structures than before.
+This has been possible due to its investment in technology enhancing
+reliability and stability of transactions. Even though this fall, HFT is still a major component of regulated markets and
+will probably remain a topic of interest for researchers in the near future.
 
-Despite the fact HFT has been criticized on qualitative issues concerning
-fairness and systemic risk, empirical research shows that HFT has lead to
+On the other hand, HFT has been criticised on qualitative issues concerning
+fairness and systemic risk. However, empirical research shows that HFT has lead to
 beneficial impacts such as reducing spreads (difference between buyers and
 sellers prices), increased liquidity, allowing more efficient price formation,
-reduced transaction costs and lower market volatility.
+reduced transaction costs and lower market volatility. HFT makes the stock market more efficient and helps small investors who trade at random times over the day.
 
@@
 \end{figure}
-
 Figure~\ref{fig:capitalmarket} shows the typical structure of capital markets

### 2015-01-26 src/2_Finance.tex
Falta hipotesis mercado eficiente

--- 2014-11-21 src/2_Finance.tex
@@
 1-second and 10-minutes. Overnight positions are avoided since they are more
-risky and also more expensives fees, HFT firms end the trading day closing all
+risky and also have more expensive fees, HFT firms end the trading day closing all
 opened positions.
@@
 \end{figure}
-
-In 2014 HFT represents nearly 50\% of equity trades in the US and more than 20\%
-in Europe showing a consistent fall since 2009 which was its best year. Revenues
+In 2014 HFT represented nearly 50\% of equity trades in the US and more than 20\%
+in Europe showing a consistent fall since 2009 which was its best year. US market Revenues
 have fallen dramatically.
-Figure~\ref{fig:HFTmarket} shows HFT trades percentage of US equity shares.
-
-One of the reasons of this fall is that exchange markets have adapted, having
-now faster and more transparent and efficient market structures than before.
-This has been possible due to its investment in technology enhancing
-reliability and stability of transactions of transactions.
-Even though this fall, HFT is still a major component of regulated markets and
-will probably remain a topic of interest for researchers in the near future.
+Figure~\ref{fig:HFTmarket} on the left shows HFT trades percentage of US and Europe equity shares. The right figure shows how revenues in US stocks have fallen the last years.
 
@@
 \end{figure}
+One of the reasons of this fall is that exchange markets have adapted, having
+now faster and more transparent and efficient market structures than before.
+This has been possible due to its investment in technology enhancing
+reliability and stability of transactions. Even though this fall, HFT is still a major component of regulated markets and
+will probably remain a topic of interest for researchers in the near future.
 
-Despite the fact HFT has been criticized on qualitative issues concerning
-fairness and systemic risk, empirical research shows that HFT has lead to
+On the other hand, HFT has been criticised on qualitative issues concerning
+fairness and systemic risk. However, empirical research shows that HFT has lead to
 beneficial impacts such as reducing spreads (difference between buyers and
 sellers prices), increased liquidity, allowing more efficient price formation,
-reduced transaction costs and lower market volatility.
+reduced transaction costs and lower market volatility. HFT makes the stock market more efficient and helps small investors who trade at random times over the day.
 
-\subsection{Financial Markets}
+\section{Financial Markets}
 
@@
 played the central and most profitable role.  At the core are the exchanges or
-inter-dealer networks (foreign exchange trading). Exchanges are the centralized
+inter-dealer networks (foreign exchange trading). Exchanges are the centralised
 marketplaces for transacting.  Broker-dealers perform two functions: trading for
@@
 
-This centralized structure existed until computer technology allowed a better
-communication structure. Today financial markets are more descentralized
+This centralised structure existed until computer technology allowed a better
+communication structure. Today financial markets are more decentralised
 providing more liquidity. Exchanges and inter-dealer brokers were replaced by
@@
 
-Figure~\ref{fig:capitalmarketnow} shows current structucture of capital markets
-including ECNs and darl pools. In this structure, ECNs, Exchanges, dark pools,
+Figure~\ref{fig:capitalmarketnow} shows current structure of capital markets
+including ECNs and dark pools. In this structure, ECNs, Exchanges, dark pools,
 broker-dealers and retail brokerages can execute orders. However, there are some
-institutional clients have also become broke-dealers.
+institutional clients that have also become broke-dealers.
 
@@
 \end{figure}
-
-Equity market and foreign exchange market (Forex) are the most populat markets
+Equity market and foreign exchange market (Forex) are the most popular markets
 for high frequency trading strategies. In equity markets can be traded stocks
@@
 Forex market allows the trading of interest rates denominated pair currencies
-such as EURUSD or USDJPY. Traders of this market are diverse, some of them are
+such as EURUSD (Euro to UD dollar) or USDJPY (US dollar to Japanese yen). Traders of this market are diverse, some of them are
 high frequency traders, long-term investors and corporations. The Forex market
-used to be centralized, only commercial banks had exclusive access to
-inter-dealer networks. Today, Forex market is descentralized and has become the
+used to be centralised, only commercial banks had exclusive access to
+inter-dealer networks. Today, Forex market is decentralised and has become the
 biggest market in terms of volume of trading. The main participants are
@@
 \begin{description}
-\item[Market order:] allow to buy or sell an asset at the current best price
-whatever that price is. The transaction price is determined by the market
+\item[Market order:] allows to buy or sell an asset at the current best price
+whatever the price is. The transaction price is determined by the market
 considering previous executed orders and volume requested.
 \item[Limit order:] allow to specify the price you want to buy or sell an asset.
-Depending on the market conditions, this type of orders can neven be executed.
+Depending on the market conditions, this type of orders can never be executed.
 \item[Stop order:] are suitable for investors who are unable to monitor their
-investments for a period of time. Stop order allow to specify the price that a
+investments for a period of time. Stop order allows to specify the price that a
 position should be closed called stop price. When the stop price is reached a
@@
 
-In terms of commissions, limit order are more expensive than market orders and
+In terms of commissions, limit and stop order are more expensive than market orders and
 there is no certainty of execution.
@@
 \begin{description}
-\item[Fill or Kill:] is an order that must be executed inmediately or being
+\item[Fill or Kill:] is an order that must be executed immediately or being
 cancelled, no partial fulfillments are allowed. 
@@
 \item[Good til canceled:] a order is active until the investor decides to cancel
-ir or the trade is executed.
+it or the trade is executed.
 \end{description}
@@
 a good. The bid price corresponds to the current highest price a buyer is
-willing to pay for a good. Orders with the same price are prioritorized by
-arrival time and placed on top of the book.  The difference between ask and bid
-price is called spread.
+willing to pay for a good. Orders with the same price are prioritized by
+arrival time and placed on top of the book.  The difference between current ask and bid
+price is called spread and their average is called mid-price.
 
@@
 \end{figure}
+Today, order books are available and they are a popular source of study among
+researchers. The mid-price and spread modelling are some of the
+problems related with this area.
 
-Today, order books are available and they are a popular source of study among
-researchers. To determine the mid-price and spread modeling are some of the
-problems studied.
-
-\subsection{Efficient market hypothesis}
+\section{Efficient market hypothesis}

### 2015-01-26 src/3_TimeSeries.tex
Falta hipotesis mercado eficiente

@@
 time series analysis.
+
+\section{Stylized facts of asset returns}
+\label{sec:stylizedfacts}
+
+There are several known features exhibited by financial instruments called
+stylized facts, which have been empirically studied and some of them have been
+documented only recently and accepted as truth. Stylized facts are usually
+formulated in terms of qualitative properties of asset returns and may not be
+precise enought to distinguish among different parametric models.
+\cite{cont2001}.
+
+Returns of a time series $y$ is defined as:
+\begin{equation}
+\label{eq:returndef}
+r_t = y_t - y_{t-1}  \, ,
+\end{equation}
+
+Some stylized facts which are common to a wide set of financial assets
+\cite{sewell2011} are:
+
+\begin{description}
+\item[Dependence] autocorrelation in returns is largely insignificant. However,
+autocorrelation in the absolute and squared returns is always positive and
+significant and decays slowly. In addition, the autocorrelation in the absolute
+returns is generally higher than the autocorrelation in the corresponding
+squared returns.
+\item[Distribution] The distribution of returns is approximately symmetric and
+has high kurtosis (i.e fat tails and a peaked centre compared with the normal
+distribution). However, distribution of returns whose were obtained from higher
+frequencies looks more like a normal distribution.
+\item[Heterogeneity] despite the fact that financial returns are non-stationary,
+stationary periods can be observed. Economist speaks in terms of a structural
+break \cite{stock1994}, in machine learning this is known as drift which means
+that the statistical properties of the target variable change over time
+\cite{widmer1996}, \cite{tsymbal2004}.
+\item[Non-Linearity] financial returns may be non-linear in mean and/or
+non-linear in variance.
+\item[Calendar effects] are also called seasonal effects and they are cyclical
+anomalies in returns where the cycle is based on the calendar. Some of known
+calendar effects are: intraday effect, weekend effect, Monday effect, intramonth
+effect, the January effect and Holiday effect. The most important calendar
+anomalies are the January effect and the weekend effect. Figure
+\ref{fig:mondayeffect} shows the Monday effect from 1928 to 2007. It shows that
+in average Monday has negative returns. However, this is changing in time,
+figure \ref{fig:mondayeffect2} shows returns only for year 2007 and its last
+three months and it shows that Mondays and Wednesdays has positive returns in
+average.
+
+\begin{figure}[h!]
+\centering
+\begin{subfigure}[b]{0.45\textwidth}
+ \includegraphics[width=\textwidth]{img/average_daily_change_1928_2007}
+ \caption{Average daily return: 1928-2007}
+ \label{fig:mondayeffect}
+\end{subfigure}
+\begin{subfigure}[b]{0.45\textwidth}
+ \includegraphics[width=\textwidth]{img/thumb-average_daily_change_2007}
+ \caption{Average daily return: 2007}
+ \label{fig:mondayeffect2}
+\end{subfigure}
+\end{figure}
+
+\end{description}
+
+\begin{description}
+ \item[Dependence]
+
+ The returns
+ autocorrelation function (ACF) of SPY is shown in figure
+ \ref{fig:returnacf}.The figure shows how the correlations are significant even
+ for very long lags, this imply that a long-memory process because of its
+ long-range dependence. 
+ Moreover, volatility tends to cluster, i.e. high volatility
+ periods are generally followed by high volatility periods.
+
+ \begin{figure}[h]
+ \centering
+ \includegraphics[width=0.9\textwidth]{plots/spy_returns_acf.pdf}
+ % spy_returns_dist.pdf: 504x504 pixel, 72dpi, 17.78x17.78 cm, bb=0 0 504 504
+ \caption{SPY returns ACF}
+ \label{fig:returnacf}
+\end{figure}
+
+ \item[Returns distribution] The returns distribution has fat tails, i.e. the
+ number of extreme events (either positive or negative) returns is larger than
+ what is hypothesized by common data generation process (generally normal
+ distribution assumption). In the markets, fat tails are an undesirable feature
+ because of the additional risk they imply.  In the figure \ref{fig:returndist}
+ is shown the SPY returns distribution based on daily dates from the period 1st
+ July 1998 to 4th April 2013. The distribution was compared against the normal
+ distribution which clearly doesn't fit the data.  
+ \begin{figure}[h]
+ \centering
+ \includegraphics[scale=0.5]{plots/spy_returns_dist.pdf}
+ % spy_returns_dist.pdf: 504x504 pixel, 72dpi, 17.78x17.78 cm, bb=0 0 504 504
+ \caption{SPY returns distribution}
+ \label{fig:returndist}
+\end{figure}
+ \item[Volume] refers to the level of trading activity in a market.
+ \item[Calendar effects] days of week, weekend, holidays.
+ \item[Intraday effects] time of the day.
+ \item[Overnight effects]
+\end{description}
+
+\section{Volatility in the financial markets}
+
+In financial markets, volatility is one of the key elements to model the
+stochastic dynamic behaviour of financial assets. This is mainly because
+volatility gives a measure of uncertainty about future returns and it is often
+viewed as an indicator of the vulnerability of financial markets. It is also
+important as an input parameter in problems like derivative pricing, hedging
+and portfolio management.  For example, in option pricing we need to known
+first the volatility of the underlying asset before pricing an option.
+
+Besides, volatility provides important information for studying asset returns
+because the latter are largely uncorrelated and nonlinearly dependent. However,
+it has been found that volatility exhibits significant autocorrelation and
+predictable patterns~\cite{poon+granger2003} and  many models have been
+proposed to forecast its behaviour.
+
+The volatility of the data is due to a large number of factors affecting the
+market, with some of them directly measurable such as historical prices,
+trends, supply and demand. However, others such as monetary policies and news
+are not directly measurable and are included in other models out of the scope
+of this thesis.
+
+\subsubsection{Types of volatility}
+
+The volatility of a stock is not directly observable~\cite{tsay2005,engle1993}.
+For example, daily volatility is not directly observable from only daily returns because there is only one observation in a trading day.
+If intraday data is available, then volatility could be estimated.
+However, intraday returns are not the only explanatory variables for volatility and several estimators have been proposed.
+These estimators are observable variables that are related to the latent variable of interest called volatility proxies~\cite{devilderetal2007}.
+Examples of volatility proxies are the following:
+
+\begin{description}
+\item[realized volatility:] is also known as historic volatility and
+it is the actual variance in the price of a stock over time.
+Realized volatility is measured in terms of the standard deviation
+using the historical stock prices. It is commonly calculated based on
+intraday price returns:
+\begin{equation}
+\label{eq:retintra}
+r_{t,n}=100(\ln(p_{t,n}) - \ln(p_{t,n-1}))
+\end{equation}
+\noindent where $p_{t,n}$ is the price observed at day $t=1,\dots,T$ and
+intraday sample $n=2,\dots,N$. Realized volatility is defined as:
+\begin{equation}
+\label{eq:rv}
+    \hat{\sigma}(t) = \sum_{n=1}^N r_{t,n}^2 \, , 
+\end{equation}
+\noindent where $N$ is the number of intraday samples and $T$ is the
+number of days. 
+In order to include overnight returns, Hansen and
+Lunde~\cite{hansen+lunde2005} introduced a scaling version of
+realized volatility using the following definitions:
+\begin{eqnarray}
+r_{t}&=&100(\ln(p_{t,N}) - \ln(p_{t-1,N})) \label{eq: ret 1} \\
+\bar{\rho}(t) &=& \sum_{t=1}^T r_{t}^2  \label{eq: ret 2} \, .
+\end{eqnarray}
+\noindent where equation~(\ref{eq: ret 1}) represents overnight 
+returns and the volatility as 
+equation~(\ref{eq: ret 2}), where $p_{t,N}$ is the last intraday 
+sample at day $t$. The scaled realized volatility $\rho(t)$ is 
+defined as:
+\begin{eqnarray}
+\label{eq:srv}
+\rho(t) = \gamma \hat{\rho}(t) \, , \qquad & \qquad \gamma = \displaystyle \frac{\bar{\rho}(t)}{\displaystyle\sum_{t=1}^T \hat{\rho}(t)}
+\end{eqnarray}
+Realized volatility has also been defined as the absolute value return or as
+the mean of the sum of intraday squared returns at short intervals of time. The
+majority of research carried out in the literature obtain the daily volatility
+as the daily squared returns as is shown in equation~(\ref{eq: ret 2}).
+However, it has been proven that this measurement noise is too high for
+observing the true volatility process~\cite{andersen+bollerslev1998}. Hansen
+and Lunde~\cite{hansen+lunde2006} stated that the use of a noisy proxy could
+result in an inferior model being chosen as the best one. The realized
+volatility, as calculated by the cumulative sum of squared intraday returns and
+shown in equation (\ref{eq:rv}), is less noisy and doesn't lead to choosing an
+inferior model.   
+\item[implied volatility:] volatility not only can be extracted from returns
+but it can also be derived from option or future pricing models.  The
+volatility obtained corresponds to the market's prediction of future
+volatility. In finance, an option is a derivative, that is, a contract which
+gives the owner the right, but not the obligation to buy or sell an underlying
+asset at a given price called strike price. An option can be executed at any
+time before an expiration date previously defined no matter what price the
+underlying asset has. For example, the Black-Scholes model~\cite{black1973}
+determines the fair option value based on stock price, strike price, time to
+option expiration, the interest rate and volatility. These are known or can be
+easily obtained from the market, excepting by volatility which must be
+estimated. However, rather than assuming a volatility a priori and computing
+option prices from it, the model can be used to estimate volatility at given
+prices, time to expiration and strike price. This obtained volatility is called
+the implied volatility of an option. Additionally, some models obtain implied
+volatility from futures (other derivative from prices). For instance, the
+Barone-Adesi and Whaley futures option model~\cite{baroneetal1987} is also used
+to determine future volatilities~\cite{hamidetal2004}. Higher implied
+volatility is indicative of greater price fluctuation in either direction.
+Implied volatility is found by determining the value which makes theoretical
+prices equal to market prices. In this way volatility is ``implied'' by the
+current market price of the stock.
+\end{description}
+
+For trading strategies, the interest is centred in forecasting realized
+volatility over the life of an option and to take advantage when this
+volatility differs from the implied volatility. This is called volatility
+arbitrage. For example, a trader will buy an option and hedge the underlying
+asset if the implied volatility is under the realized volatility.
+
+\subsubsection{Volatility methods}
+
+In the existing literature, there are four main classes of asset
+return volatility models: the general autoregressive conditional
+heteroskedasticity (GARCH) models, the stochastic volatility (SV)
+models, the realized volatility models and the machine learning based
+models. A comparison of the first three models can be found
+in~\cite{wei2012}.
+
+For many years the most popular methods for estimating financial
+volatility were the autoregressive conditional heteroskedasticity
+(ARCH) models~\cite{engle1982} and the general ARCH (GARCH)
+models~\cite{bollerslev1986}. For instance, the GARCH(1,1) defines
+returns $y_t$ and volatility $\sigma_t$ as:
+
+\begin{eqnarray*}
+    y_t &=& \sigma_t \epsilon_t \\
+     \sigma_t^2 &=& \alpha_0 + \alpha_1 \epsilon_{t-1}^2 + \beta_1
+     \sigma_{t-1}^2
+\end{eqnarray*}
+
+\noindent where $\epsilon_t$ is standard Gaussian white noise,
+$\alpha_0,\alpha_1,\beta_1 \geq 0$ are required to ensure that the
+variance will never be negative and $\alpha_1+\beta_1 <1$ is needed to
+guarantee a weakly stationary process~\cite{nelson1990}.
+
+Since the introduction of the GARCH models, several extensions have been
+proposed, but none of them seems to beat the GARCH(1,1)
+model~\cite{lunde+hansen2005}. Despite its popularity, GARCH models have
+several limitations: firstly, a time series model may be non-linear in mean
+and/or non-linear in variance, but ARCH and GARCH models are non-linear in
+variance, but not in mean. Besides, GARCH models often fail to capture highly
+irregular phenomena, like wild market fluctuations.
+
+SV models explain how volatility varies in a random fashion. These models are
+useful because they explain why options with different strikes and expirations
+dates have different Black-Scholes implied volatilities, phenomenon known as
+the volatility smile. This is useful because the Black-Scholes model assumes
+that the volatility of the underlying asset is constant which is not always
+true. There are several SV models and the most well-known and popular is the
+Heston model~\cite{heston1993}. Additional information about SV models can be
+found in~\cite{shephard1995}.
+
+The realized volatility constructed from high frequency intraday returns gave
+rise to the realized volatility models mainly because the realized volatility
+series is much more homoskedastic and seems to be a long memory
+process~\cite{andersonetal2003}. For realized volatility, the autoregressive
+fractionally integrated moving average (ARFIMA) process emerged as a standard
+model~\cite{chenetal2010} and many variations have been studied, but all of
+them produce similar forecasting results to the ARFIMA(1,d,1)
+model~\cite{koopmanetal2005}.
+
+On the other hand, machine learning based models, especially artificial neural
+networks (ANN) and support vector machines (SVM) have arisen as an alternative
+to forecast volatility. ANN is a statistical technique inspired by biological
+neural networks which is capable of changing its structure based on external or
+internal information during a training phase~\cite{sammut2011}. SVM are
+supervised learning models for classification analysis which recognize patterns
+finding a separating hyperplane. An extension for regression analysis is known
+as support vector regression (SVR).
+
+Since machine learning models and in particular ANN do not require assumptions
+about the data (gaussianity for example) and allow more explanatory variables
+than returns to be included, they have become widely used in solving financial
+problems, specially volatility
+forecasting~\cite{hamidetal2004,donaldsonetal1997}. There are also many works
+focused on the using of SVM in volatility
+forecasting~\cite{shiyietal2008,shiyietal2010,gavrishchaka2006,vasilios2012}.
+
+However, just as with ANN, SVMs have scalability problems because their
+training process is computationally intensive and it is done in batch mode. The
+scalability problem worsens when new additional training data is available and
+a re-training process from scratch needs to be done. This problem can be
+avoided using online machine learning algorithms that allow one instance at a
+time to be processed with low computationally expensive calculations.
+
+%Volatility models have been extensively
+%studied~\cite{gatheral2006,poon+granger2003,knight2002} and some of  them
+%include machine learning
+%methods~\cite{hamidetal2004,donaldsonetal1997,shiyietal2008,shiyietal2010,gavrishchaka2006,vasilios2012}.  
+%
 

### 2015-02-02 src/2_Finance.tex
Agregando hipotesis mercado eficiente

@@
 
+\section{Technical analysis}
+
+\section{Fundamental analysis}
+
 \section{Efficient market hypothesis}
+
+Efficient Market Hypothesis (EMH) was developed independently by  Paul Samuelson and Eugene Fama in the 1960s.
+EMH also known as the random walk theory states that current stock prices fully reflect available information related to its value and there is no way to earn excess profits
+\cite{fama1970}.
+EMH requires that all the participants have rational expectations, and investors reactions be random and follow a normal distribution pattern.
+Thus, any one can be wrong about the market, but the market is always right as a whole.
+There are three common forms of EMH: weak-form efficiency, semi-strong efficiency and strong-form efficiency.
+
+The weak form claims that prices already reflect all past publicly available information.
+Then strategies based on technical analysis will not be able to produce excess returns.
+However, the weak form accepts that some forms of fundamental analysis may still provide excess return.
+Therefore, future prices cannot be predicted based on analysis of historical data.
+This implies that future prices movements are determined entirely by information not contained in the past prices and participants are unable to systematically profit from market inefficiencies.
+However, many studies have shown a marked tendency for the stock markets to trend over time periods.
+Various explanations for such large and apparently non-random price movements have been promulgated.
+Even Fama has accepted price anomalies which do not follow the weak-form efficiency hypothesis \cite{fama+french2008}.
+
+The semi-strong form of the EMH claims that prices instantly change to reflect new public (not private) information such that no excess return can be obtained.
+In this form neither technical or fundamental analysis can produce excess return.
+
+The strong form of the EMH additionally claims that prices instantly reflect even hidden, private or "insider" information.
+
+EMH is based on rational human behaviour and its validity has been criticised by
+psychologists and behavioural economists who argue that the EMH is based
+on counterfactual assumptions regarding human behaviour, that is,
+rationality. Recent advances in evolutionary psychology and the cognitive
+neurosciences may be able to reconcile the EMH with behavioural
+anomalies.
+On the other hand, if the EMH is true, the market really walks randomly and therefore there shouldn't be any difference between experienced and novice traders.
+Kim Man Lui proved the contrary in a controlled experiment \cite{man2013}.

### 2015-02-02 src/2_Finance.tex
Finance chapter ready

@@
 a good. The bid price corresponds to the current highest price a buyer is
-willing to pay for a good. Orders with the same price are prioritized by
+willing to pay for a good. Orders with the same price are prioritised by
 arrival time and placed on top of the book.  The difference between current ask and bid
@@
 
-\section{Technical analysis}
-
-\section{Fundamental analysis}
-
 \section{Efficient market hypothesis}
@@
 The weak form claims that prices already reflect all past publicly available information.
-Then strategies based on technical analysis will not be able to produce excess returns.
-However, the weak form accepts that some forms of fundamental analysis may still provide excess return.
 Therefore, future prices cannot be predicted based on analysis of historical data.
@@
 The semi-strong form of the EMH claims that prices instantly change to reflect new public (not private) information such that no excess return can be obtained.
-In this form neither technical or fundamental analysis can produce excess return.
 
 The strong form of the EMH additionally claims that prices instantly reflect even hidden, private or "insider" information.
+
+EMH is related with two approaches to investment analysis: fundamental and technical analysis.
+Fundamental analysts base their predictions of stock price behaviour on fundamental factors such as internal information of a company, its industry or the economy.
+Technical analysts, by contrast, consider that all this financial information is already included in the prices and believe that future stock prices can be predicted studying the historical market behaviour.
+A market technician bases its predictions on historical patterns of prices of volume changes.
+
+Under the weak form of EMH, strategies based on technical analysis will not be able to produce excess returns.
+However, the weak form accepts that some forms of fundamental analysis may still provide excess return.
+Similarly, the semi-strong form of EMH  neither technical or fundamental analysis can produce excess return.
+
+If the stock market efficiently digests all available information, there is little justification for seeking excess returns gains from investing.
+However, EMH doesn't lessen the the importance of investing only change its philosophy.
+The only way an investor can possibly obtain higher returns is by purchasing riskier investments.
+
+ Researches can determine now how efficient a financial market is, i.e how efficiently information is processed.
 

### 2015-02-05 src/3_TimeSeries.tex
Avance capitulo 2

@@
-\chapter{Time Series models}
+\chapter{Financial Time Series}
 
@@
 
+\section{Characteristics of Time Series}
+
+\subsection{Stationary }
+A strictly stationary times series $Y_t$ is one for which the probabilistic behaviour
+of every collection of values $\{Y_{t_1},Y_{t_2},\dots,Y_{t_L}\}$ is identical
+to that of the time shifted set, more precisely: \[ P\{Y_{t_1} \leq
+c_1,\dots,Y_{t_L} \leq c_L\} = P\{Y_{t_1+h} \leq c_1,\dots,Y_{t_L+h} \leq c_L\}
+\quad \forall L \in \mathbb{N}, \forall h \in \mathbb{Z}\] \noindent where
+$c_1,\dots,c_L$ are constants.  This definition is too strong and difficult to
+assess it from a single data set. The weak version of this definition imposes
+conditions only on the two first two moments.
+
+A weakly stationary time series is a process which mean, variance and auto
+covariance do not change over time: \begin{eqnarray*} E(Y_t) &=& \mu  \quad
+\forall t \in \mathbb{N} \\ E(Y^2_t) &=& \sigma^2  \quad \forall t \in
+\mathbb{N} \\ \lambda(s,t)&=&\lambda(s+h,t+h) \quad \forall s,t \in \mathbb{N},
+\forall h \in \mathbb{Z} \end{eqnarray*}
+
+\noindent with $\lambda(s,t) = E[(Y_s-\mu)(Y_t - \mu)]$
+
+In finance, a shock represents an unexpected change in a variable or a in its error term in a particular time period.
+For stationary time series, shocks to the system will gradually die away.
+That is, a shock during time $t$ will have a smaller effect in time $t+1$, a smaller effect on $t+2$ and so on.
+For non-stationary data, the persistence of shocks will always be infinite.
+The use of non-stationary data can lead to spurious regressions.
+If two stationary variables are generated as independent random series and are trending over time, when one of those variables is regressed on the other, they could have a high $R^2$ even if the two are totally unrelated.
+So, if standard regression techniques are applied to non-stationary data, it could look good under standard measures but valueless \cite{brooks2002}.
+
+\subsection{Integration}
+
+Following Johansen \cite{johansen1995} we shall say that a stochastic process
+$Y_t$ which satisfies $Y_t-E(Y_t) = \sum_{i=0}^\infty C_i\,\varepsilon_{t-i}$ is
+called $I(0)$, and then we shall write $Y_t\sim I(0)$, whenever
+$\sum_{i=0}^\infty C_i \neq 0$ and $\sum_{i=0}^\infty C_i\,z^i$ converges for
+$z\in\mathbb{C}$ with $|z|<1$.  It is understood that the condition
+$\varepsilon_t\sim iid(0,\sigma^2)$ holds.
+
+A (vector) time series $\mathbf{y}_t$ is said to be {\em integrated of order\/}
+$d$, and then we shall write $\mathbf{y}_t\sim I(d)$, whenever after $d$ times
+(discrete) differentiation an stationary process is
+obtained~\cite{banerjee1993};
+more precisely, whenever
+$(1-L)^d\,\mathbf{y}_t\sim\text{I(0)}$, where $L$ is the usual lag operator:
+$(1-L)\,\mathbf{y}_t = \Delta\mathbf{y}_t = \mathbf{y}_t-\mathbf{y}_{t-1}$ for
+all $t$.
+
+Note that this definition includes the scalar case as time series of
+vectors of dimension 1; in this scalar case we will write the time series in
+non-bold format.
+
+\subsection{Cointegration}
+Cointegration concept was introduced by Engle in 1987 \cite{engle1987} and implies that one or
+more linear combinations of non-stationary variables are stationary even though
+individually they are not.
+
+Let $\mathbf{y}_t^\nu$, $\nu=1,\dots,p$, be a set of $p$ vector time series of
+order $I(1)$.  They are said to be {\em cointegrated\/} if a vector
+$\beta=[\beta(1),\dots,\beta(p)]^\top \in \mathbb{R}^p$ exists, such that the
+time series,
+\begin{equation}
+\mathbf{Z}_t:= 
+\sum_{\nu=1}^p \beta(\nu)\,\mathbf{y}_t^\nu\,\sim\,\text{I(0)}\,.
+\end{equation}
+
+In other words, a set of $I(1)$ time series is said to be cointegrated if a
+linear combination of them exists, which is I(0).
+
+Stock and Watson in 1988 \cite{stock+watson1988} observed that
+cointegration reflects the common stochastic trends providing a useful way to
+understand cointegration relationships. These common stochastic trends can be
+also interpreted as a long-run equilibrium relationships.
+
+The idea of cointegration was immediately adopted in finance since it could
+represent their long-run relationship implied by economic
+theory~\cite{laietAl1991}, \cite{lence+falk2005}.  Economic theory suggest that
+economic time series are mean-reverting process and therefore, it reflects the
+idea of that some set of variables cannot wander too far from each other.
+
+On the other hand, the efficient markets hypothesis, also known as the random
+walk theory states that current stock prices fully reflect available information
+related to its value and there is no way to earn excess profits~\cite{fama1970}.
+This means that if we have stock prices from a jointly efficient market, they
+cannot be cointegrated \cite{granger1986}, \cite{dwyer1992}. However,
+\cite{richards1995} claims that cointegration is directly at odds with market
+efficiency, even though, there is no evidence that cointegration among asset
+prices have implications about market efficiency~\cite{lence+falk2005}.
+
+Despite the fact that cointegration on closing daily rates of currency pairs has
+not been found \cite{coleman1990}, \cite{copeland1991}, different time series
+frequencies can have different behaviours~\cite{aldridge2009}. Pair trading is a
+very common example of cointegration application~\cite{herlemont2003} but
+cointegration can also be extended to a larger set of
+variables~\cite{mukherjee1995},\cite{engle2004}.
+
 \section{Stylized facts of asset returns}
@@
 formulated in terms of qualitative properties of asset returns and may not be
-precise enought to distinguish among different parametric models.
+precise enough to distinguish among different parametric models.
 \cite{cont2001}.
@@
 returns is generally higher than the autocorrelation in the corresponding
-squared returns.
+squared returns. The returns autocorrelation function (ACF) of SPY is shown in figure
+ \ref{fig:returnacf}. The figure shows how the correlations are significant even
+ for very long lags, this imply that a long-memory process because of its
+ long-range dependence. 
+ \begin{figure}[h]
+ \centering
+ \includegraphics[width=0.7\textwidth]{plots/spy_returns_acf.pdf}
+ \caption{SPY returns ACF}
+ \label{fig:returnacf}
+\end{figure}
 \item[Distribution] The distribution of returns is approximately symmetric and
@@
 frequencies looks more like a normal distribution.
+The returns distribution tails are larger than
+ what is hypothesised by common data generation process (generally normal
+ distribution assumption). In the markets, fat tails are an undesirable feature
+ because of the additional risk they imply.  In the figure \ref{fig:returndist}
+ is shown the SPY returns distribution based on daily dates from the period 1st
+ July 1998 to 4th April 2013. The distribution was compared against the normal
+ distribution which clearly doesn't fit the data.  
+ \begin{figure}[h]
+ \centering
+ \includegraphics[scale=0.5]{plots/spy_returns_dist.pdf}
+ % spy_returns_dist.pdf: 504x504 pixel, 72dpi, 17.78x17.78 cm, bb=0 0 504 504
+ \caption{SPY returns distribution}
+ \label{fig:returndist}
+\end{figure}
 \item[Heterogeneity] despite the fact that financial returns are non-stationary,
@@
 \end{figure}
-
-\end{description}
-
-\begin{description}
- \item[Dependence]
-
- The returns
- autocorrelation function (ACF) of SPY is shown in figure
- \ref{fig:returnacf}.The figure shows how the correlations are significant even
- for very long lags, this imply that a long-memory process because of its
- long-range dependence. 
- Moreover, volatility tends to cluster, i.e. high volatility
- periods are generally followed by high volatility periods.
-
- \begin{figure}[h]
- \centering
- \includegraphics[width=0.9\textwidth]{plots/spy_returns_acf.pdf}
- % spy_returns_dist.pdf: 504x504 pixel, 72dpi, 17.78x17.78 cm, bb=0 0 504 504
- \caption{SPY returns ACF}
- \label{fig:returnacf}
-\end{figure}
-
- \item[Returns distribution] The returns distribution has fat tails, i.e. the
- number of extreme events (either positive or negative) returns is larger than
- what is hypothesized by common data generation process (generally normal
- distribution assumption). In the markets, fat tails are an undesirable feature
- because of the additional risk they imply.  In the figure \ref{fig:returndist}
- is shown the SPY returns distribution based on daily dates from the period 1st
- July 1998 to 4th April 2013. The distribution was compared against the normal
- distribution which clearly doesn't fit the data.  
- \begin{figure}[h]
- \centering
- \includegraphics[scale=0.5]{plots/spy_returns_dist.pdf}
- % spy_returns_dist.pdf: 504x504 pixel, 72dpi, 17.78x17.78 cm, bb=0 0 504 504
- \caption{SPY returns distribution}
- \label{fig:returndist}
-\end{figure}
- \item[Volume] refers to the level of trading activity in a market.
- \item[Calendar effects] days of week, weekend, holidays.
- \item[Intraday effects] time of the day.
- \item[Overnight effects]
 \end{description}
@@
 \section{Volatility in the financial markets}
-
 In financial markets, volatility is one of the key elements to model the
@@
 
-\subsubsection{Types of volatility}
+Despite the fact variance and volatility are related, they are not the same concept. Variance is a measure of distribution of returns and is not necessarily
+bound by any time period.  Volatility is a measure of the standard deviation
+(square root of the variance) over a certain time interval. In finance,
+variance and volatility both gives you a sense of an asset's risk. Variance
+gives you a sense of the risk in the asset over its lifetime, while volatility
+gives you a sense of the movement of the asset in, for example, the past month or the
+past year. The main underlying difference is in their definition. Variance has a
+fixed mathematical definition, however volatility does not as such. Volatility
+is said to be the measure of fluctuations of a process.
+
+Volatility is a subjective term, whereas variance is an objective term i.e.
+given the data you can definitely find the variance, while you can't find
+volatility just having the data. Volatility is associated with the process, and
+not with the data. In order to know the volatility you need to have an idea of the process i.e you
+need to have an observation of the dispersion of the process. All the different
+processes will have different methods to compute volatilities based on the
+underlying assumptions of the process.
+
+\subsection{Types of volatility}
 
@@
 
-\begin{description}
-\item[realized volatility:] is also known as historic volatility and
+\subsubsection{realized volatility} is also known as historic volatility and
 it is the actual variance in the price of a stock over time.
@@
 inferior model.   
-\item[implied volatility:] volatility not only can be extracted from returns
+\subsubsection{implied volatility volatility} not only can be extracted from returns
 but it can also be derived from option or future pricing models.  The
@@
 current market price of the stock.
+
+In finance, an option is a derivative, that is, a contract which gives the owner
+the right, but not the obligation to buy or sell an underlying asset at a given
+price called strike price. An option can be executed at any time before an
+expiration date previously defined no matter what price the underlying asset
+has.
+
+The Black-Scholes formula~\cite{black1973}, developed in the early 1970's, Myron
+Scholes, Robert Merton and Fisher Black,  allows to determine an option value
+$V$ based on the underlying asset price $S(t)$ at a time $t$ and the following
+constant parameters:
+
+\begin{description}
+\item [$\sigma$:] underlying asset price volatility which measures the standard
+deviation of the returns
+\item [$\mu$:] underlying asset drift which is a measure of the average rate of
+growth of the stock
+\item[$E$:] option strike or excersice price
+\item[$T$:] option date of expiry
+\item[$t$:] current time
+\item[$r$:] risk-free interest rate
 \end{description}
+
+The Black-Scholes model assume that the underlying price $S$ follows a lognormal random walk:
+
+\begin{equation}\label{eq:stockprice}
+dS = \mu S dt + \sigma S dB
+\end{equation}
+
+\noindent where $B$ is a Brownian motion. This stochastic differential equation
+has two components: a deterministic term given by $\mu S dt$ and a random term
+given by $\sigma S dX$.
+
+A brownian motion $B$ (also called a Wiener process) is a stochastic process
+characterized by the three following properties:
+
+\begin{description}
+\item[Continuity:] $B(t)$ is a continuos function
+\item[Normal increments:]  $B(t)-B(s)$ has a normal distribution with mean $0$
+and variance $t-s$.
+\item[Independence of increments:] for every choice of nonnegative real numbers
+$0 \leq s_1 <  t_1 \leq \cdots \leq s_n < t_n < \infty$, the increment random
+variables $W_{t_1} - W_{s_1}, \cdots, W_{t_n} - W_{s_n}$ are jointly
+independent.
+\end{description}
+
+An stochastic differential integral has the form:
+
+\begin{equation}
+W(T)=\int_0^T f(t)dB(t) \, .
+\end{equation}
+
+\noindent This equations is also expressed in an abbreviate form:
+
+\begin{equation}
+dW = f(t) dB \, .
+\end{equation}
+
+\noindent Therefore, the integral form of the stock price model shown in
+equation (\ref{eq:stockprice}) is:
+
+\begin{equation}
+S(T)=\int_0^T \mu S(t) dt + \int_0^T \sigma S(t) dB(t)
+\end{equation}
+
+Ito's lemma is used to find the differential of a time dependent function of a
+stochastic process. In option pricing we need to find the option price $V(S(t))$
+which depends on a stochastic stock price model $S(t)$.  $V(S,t)$ is required to
+be  differentiable function of $S$ and once differentiable function of $t$.
+
+These are known or can be easily obtained from the market, excepting by
+volatility which must be estimated. However, rather than assuming a volatility
+a priori and computing option prices from it, the model can be used to estimate
+volatility at given prices, time to expiration and strike price. This obtained
+volatility is called the implied volatility of an option. Additionally, some
+models obtain implied volatility from futures (other derivative from prices).
 
@@
 
-\subsubsection{Volatility methods}
+\subsection{Volatility methods}
 
@@
 
-%Volatility models have been extensively
-%studied~\cite{gatheral2006,poon+granger2003,knight2002} and some of  them
-%include machine learning
-%methods~\cite{hamidetal2004,donaldsonetal1997,shiyietal2008,shiyietal2010,gavrishchaka2006,vasilios2012}.  
-%
+\section{Vector Autoregressive Models}\label{sec:varvec}
 
-\section{Characteristics of Time Series}
-
-\subsection{Stationary }
-A strictly stationary times series $y_t$ is one for which the probabilistic behavior
-of every collection of values $\{y_{t_1},y_{t_2},\dots,y_{t_L}\}$ is identical
-to that of the time shifted set, more precisely: \[ P\{y_{t_1} \leq
-c_1,\dots,y_{t_L} \leq c_L\} = P\{y_{t_1+h} \leq c_1,\dots,y_{t_L+h} \leq c_k\}
-\quad \forall L \in \mathbb{N}, \forall h \in \mathbb{Z}\] \noindent where
-$c_1,\dots,c_L$ are constants.  This definition is too strong and difficult to
-assess it from a single data set. The weak version of this definition imposes
-conditions only on the two first two moments.
-
-A weakly stationary time series is a process which mean, variance and auto
-covariance do not change over time: \begin{eqnarray*} E(Y_t) &=& \mu  \quad
-\forall t \in \mathbb{N} \\ E(Y^2_t) &=& \sigma^2  \quad \forall t \in
-\mathbb{N} \\ \lambda(s,t)&=&\lambda(s+h,t+h) \quad \forall s,t \in \mathbb{N},
-\forall h \in \mathbb{Z} \end{eqnarray*}
-
-\noindent with $\lambda(s,t) = E[(y_s-\mu)(y_t - \mu)]$
-
-\subsection{Integration}
-
-Following Johansen \cite{johansen1995} we shall say that a stochastic process
-$Y_t$ which satisfies $Y_t-E(Y_t) = \sum_{i=0}^\infty C_i\,\varepsilon_{t-i}$ is
-called $I(0)$, and then we shall write $Y_t\sim I(0)$, whenever
-$\sum_{i=0}^\infty C_i \neq 0$ and $\sum_{i=0}^\infty C_i\,z^i$ converges for
-$z\in\mathbb{C}$ with $|z|<1$.  It is understood that the condition
-$\varepsilon_t\sim iid(0,\sigma^2)$ holds.
-
-A (vector) time series $\mathbf{y}_t$ is said to be {\em integrated of order\/}
-$d$, and then we shall write $\mathbf{y}_t\sim I(d)$, whenever after $d$ times
-(discrete) differentiation an stationary process is
-obtained~\cite{banerjee1993};
-more precisely, whenever
-$(1-L)^d\,\mathbf{y}_t\sim\text{I(0)}$, where $L$ is the usual lag operator:
-$(1-L)\,\mathbf{y}_t = \Delta\mathbf{y}_t = \mathbf{y}_t-\mathbf{y}_{t-1}$ for
-all $t$.
-
-Note that this definition includes the scalar case as time series of
-vectors of dimension 1; in this scalar case we will write the time series in
-nonbold format.
-
-\subsection{Cointegration}
-Cointegration concept was introduced by~\cite{engle1987} and implies that one or
-more linear combinations of non-stationary variables are stationary even though
-individually they are not.  Moreover ~\cite{stock+watson1988} observed that
-cointegration reflects the common stochastic trends providing a useful way to
-understand cointegration relationships. These common stochastic trends can be
-also interpreted as a long-run equilibrium relationships.
-
-Let $\mathbf{y}_t^\nu$, $\nu=1,\dots,p$, be a set of $p$ vector time series of
-order $I(1)$.  They are said to be {\em cointegrated\/} if a vector
-$\beta=[\beta(1),\dots,\beta(p)]^\top \in \mathbb{R}^p$ exists, such that the
-time series,
-\begin{equation}
-\mathbf{Z}_t:= 
-\sum_{\nu=1}^p \beta(\nu)\,\mathbf{y}_t^\nu\,\sim\,\text{I(0)}\,.
-\end{equation}
-
-In other words, a set of $I(1)$ time series is said to be cointegrated if a
-linear combination of them exists, which is I(0).
-
-The idea of cointegration was inmediatly adopted in finance since it could
-represent their long-run relationship implied by economic
-theory~\cite{laietAl1991}, \cite{lence+falk2005}.  Economic theory suggest that
-economic time series are mean-reverting process and therefore, it reflects the
-idea of that some set of variables cannot wander too far from each other.
-
-On the other hand, the efficient markets hypothesis, also known as the random
-walk theory states that current stock prices fully reflect available information
-related to its value and there is no way to earn excess profits~\cite{fama1970}.
-This means that if we have stock prices from a jointly efficient market, they
-cannot be cointegrated \cite{granger1986}, \cite{dwyer1992}. However,
-\cite{richards1995} claims that cointegration is directly at odds with market
-efficiency, even though, there is no evidende that cointegration among asset
-prices have implications about market efficiency~\cite{lence+falk2005}.
-
-Despite the fact that cointegration on closing daily rates of currency pairs has
-not been found \cite{coleman1990}, \cite{copeland1991}, different time series
-frequencies can have different behaviours~\cite{aldridge2009}. Pair trading is a
-very common example of cointegration application~\cite{herlemont2003} but
-cointegration can also be extended to a larger set of
-variables~\cite{mukherjee1995},\cite{engle2004}.
-
-\subsection{Volatility}
-
-The volatility of a stock is not directly observable~\cite{tsay2005,engle1993}.
-For example, daily volatility is not directly observable from only daily returns
-because there is only one observation in a trading day.  If intraday data is
-available, then volatility could be estimated. However, intraday returns are not
-the only explanatory variables for volatility and several estimators have been
-proposed. These estimators are observable variables that are related to the
-latent variable of interest called volatility proxies~\cite{devilderetal2007}.
-Examples of volatility proxies are the following:
-
-\begin{description}%[leftmargin=0.4cm]\itemsep4pt
-\item[realized volatility:] is also known as historic volatility and
-it is the actual variance in the price of a stock over time.
-Realized volatility is measured in terms of the standard deviation
-using the historical stock prices. It is commonly calculated based on
-intraday price returns:
-
-\begin{equation}
-\label{eq:retintra}
-r_{t,n}=100(\ln(p_{t,n}) - \ln(p_{t,n-1}))
-\end{equation}
-
-\noindent where $p_{t,n}$ is the price observed at day $t=1,\dots,T$ and
-intraday sample $n=2,\dots,N$. Realized volatility is defined as:
-
-\begin{equation}
-\label{eq:rv}
-    \hat{\sigma}(t) = \sum_{n=1}^N r_{t,n}^2 \, , 
-\end{equation}
-
-\noindent where $N$ is the number of intraday samples and $T$ is the
-number of days.
-
-In order to include overnight returns, Hansen and
-Lunde~\cite{hansen+lunde2005} introduced a scaling version of
-realized volatility using the following definitions:
-
-\begin{eqnarray}
-r_{t}&=&100(\ln(p_{t,N}) - \ln(p_{t-1,N})) \label{eq: ret 1} \\
-\bar{\rho}(t) &=& \sum_{t=1}^T r_{t}^2  \label{eq: ret 2} \, .
-\end{eqnarray}
-
-\noindent where equation~(\ref{eq: ret 1}) represents overnight 
-returns and the volatility as 
-equation~(\ref{eq: ret 2}), where $p_{t,N}$ is the last intraday 
-sample at day $t$. The scaled realized volatility $\rho(t)$ is 
-defined as:
-
-\begin{eqnarray}
-\label{eq:srv}
-\rho(t) = \gamma \hat{\rho}(t) \, , \qquad & \qquad \gamma = \displaystyle \frac{\bar{\rho}(t)}{\displaystyle\sum_{t=1}^T \hat{\rho}(t)}
-\end{eqnarray}
-
-Realized volatility has also been defined as the absolute value return or as the
-mean of the sum of intraday squared returns at short intervals of time. The
-majority of research carried out in the literature obtain the daily volatility
-as the daily squared returns as is shown in equation~(\ref{eq: ret 2}).
-However, it has been proven that this measurement noise is too high for
-observing the true volatility process~\cite{andersen+bollerslev1998}. Hansen and
-Lunde~\cite{hansen+lunde2006} stated that the use of a noisy proxy could result
-in an inferior model being chosen as the best one. The realized volatility, as
-calculated by the cumulative sum of squared intraday returns and shown in
-equation (\ref{eq:rv}), is less noisy and doesn't lead to choosing an inferior
-model.
-
-\item[implied volatility:] volatility not only can be extracted from returns but
-it can also be derived from option or future pricing models.  The volatility
-obtained corresponds to the market's prediction of future volatility. In
-finance, an option is a derivative, that is, a contract which gives the owner
-the right, but not the obligation to buy or sell an underlying asset at a given
-price called strike price. An option can be executed at any time before an
-expiration date previously defined no matter what price the underlying asset
-has. For example, the Black-Scholes model~\cite{black1973} determines the fair
-option value based on stock price, strike price, time to option expiration, the
-interest rate and volatility. These are known or can be easily obtained from the
-market, excepting by volatility which must be estimated. However, rather than
-assuming a volatility a priori and computing option prices from it, the model
-can be used to estimate volatility at given prices, time to expiration and
-strike price. This obtained volatility is called the implied volatility of an
-option. Additionally, some models obtain implied volatility from futures (other
-derivative from prices). For instance, the Barone-Adesi and Whaley futures
-option model~\cite{baroneetal1987} is also used to determine future
-volatilities~\cite{hamidetal2004}. Higher implied volatility is indicative of
-greater price fluctuation in either direction. Implied volatility is found by
-determining the value which makes theoretical prices equal to market prices. In
-this way volatility is ``implied'' by the current market price of the stock.
-
-\end{description}
-
-For trading strategies, the interest is centred in forecasting realized
-volatility over the life of an option and to take advantage when this volatility
-differs from the implied volatility. This is called volatility arbitrage. For
-example, a trader will buy an option and hedge the underlying asset if the
-implied volatility is under the realized volatility.
-
-\subsubsection{Volatility methods}
-
-In the existing literature, there are four main classes of asset
-return volatility models: the general autoregressive conditional
-heteroskedasticity (GARCH) models, the stochastic volatility (SV)
-models, the realized volatility models and the machine learning based
-models. A comparison of the first three models can be found
-in~\cite{wei2012}.
-
-For many years the most popular methods for estimating financial
-volatility were the autoregressive conditional heteroskedasticity
-(ARCH) models~\cite{engle1982} and the general ARCH (GARCH)
-models~\cite{bollerslev1986}. For instance, the GARCH(1,1) defines
-returns $y_t$ and volatility $\sigma_t$ as:
-
-\begin{eqnarray*}
-    y_t &=& \sigma_t \epsilon_t \\
-     \sigma_t^2 &=& \alpha_0 + \alpha_1 \epsilon_{t-1}^2 + \beta_1
-     \sigma_{t-1}^2
-\end{eqnarray*}
-
-\noindent where $\epsilon_t$ is standard Gaussian white noise,
-$\alpha_0,\alpha_1,\beta_1 \geq 0$ are required to ensure that the
-variance will never be negative and $\alpha_1+\beta_1 <1$ is needed to
-guarantee a weakly stationary process~\cite{nelson1990}.
-
-Since the introduction of the GARCH models, several extensions have been
-proposed, but none of them seems to beat the GARCH(1,1)
-model~\cite{lunde+hansen2005}. Despite its popularity, GARCH models have several
-limitations: firstly, a time series model may be non-linear in mean and/or
-non-linear in variance, but ARCH and GARCH models are non-linear in variance,
-but not in mean. Besides, GARCH models often fail to capture highly irregular
-phenomena, like wild market fluctuations.
-
-SV models explain how volatility varies in a random fashion. These models are
-useful because they explain why options with different strikes and expirations
-dates have different Black-Scholes implied volatilities, phenomenon known as the
-volatility smile. This is useful because the Black-Scholes model assumes that
-the volatility of the underlying asset is constant which is not always true.
-There are several SV models and the most well-known and popular is the Heston
-model~\cite{heston1993}. Additional information about SV models can be found
-in~\cite{shephard1995}.
-
-The realized volatility constructed from high frequency intraday returns gave
-rise to the realized volatility models mainly because the realized volatility
-series is much more homoskedastic and seems to be a long memory
-process~\cite{andersonetal2003}. For realized volatility, the autoregressive
-fractionally integrated moving average (ARFIMA) process emerged as a standard
-model~\cite{chenetal2010} and many variations have been studied, but all of them
-produce similar forecasting results to the ARFIMA(1,d,1)
-model~\cite{koopmanetal2005}.
-
-On the other hand, machine learning based models, especially artificial neural
-networks (ANN) and support vector machines (SVM) have arisen as an alternative
-to forecast volatility. ANN is a statistical technique inspired by biological
-neural networks which is capable of changing its structure based on external or
-internal information during a training phase~\cite{sammut2011}. SVM are
-supervised learning models for classification analysis which recognize patterns
-finding a separating hyperplane. An extension for regression analysis is known
-as support vector regression (SVR).
-
-Since machine learning models and in particular ANN do not require assumptions
-about the data (gaussianity for example) and allow more explanatory variables
-than returns to be included, they have become widely used in solving financial
-problems, specially volatility
-forecasting~\cite{hamidetal2004,donaldsonetal1997}. There are also many works
-focused on the using of SVM in volatility
-forecasting~\cite{shiyietal2008,shiyietal2010,gavrishchaka2006,vasilios2012}.
-
-However, just as with ANN, SVMs have scalability problems because their training
-process is computationally intensive and it is done in batch mode. The
-scalability problem worsens when new additional training data is available and a
-re-training process from scratch needs to be done. This problem can be avoided
-using online machine learning algorithms that allow one instance at a time to be
-processed with low computationally expensive calculations.
-
-%\section{Black-Scholes formula}
-%
-%In finance, an option is a derivative, that is, a contract which gives the owner
-%the right, but not the obligation to buy or sell an underlying asset at a given
-%price called strike price. An option can be executed at any time before an
-%expiration date previously defined no matter what price the underlying asset
-%has. 
-%
-%The Black-Scholes formula~\cite{black1973}, developed in the early 1970's, Myron
-%Scholes, Robert Merton and Fisher Black,  allows to determine an option value
-%$V$ based on the underlying asset price $S(t)$ at a time $t$ and the following
-%constant parameters: 
-%
-%\begin{description}
-%\item [$\sigma$:] underlying asset price volatility which measures the standard
-%deviation of the returns
-%\item [$\mu$:] underlying asset drift which is a measure of the average rate of
-%growth of the stock
-%\item[$E$:] option strike or excersice price
-%\item[$T$:] option date of expiry
-%\item[$t$:] current time
-%\item[$r$:] risk-free interest rate
-%\end{description}
-%
-%\subsection{Stock price model}
-%
-%The Black-Scholes model assume that the underlying price $S$ follows a lognormal random walk:
-%
-%\begin{equation}\label{eq:stockprice}
-%dS = \mu S dt + \sigma S dB
-%\end{equation}
-%
-%\noindent where $B$ is a Brownian motion. This stochastic differential equation
-%has two components: a deterministic term given by $\mu S dt$ and a random term
-%given by $\sigma S dX$.
-%
-%\subsection{Brownian motion}
-%
-%A brownian motion $B$ (also called a Wiener process) is a stochastic process
-%characterized by the three following properties:
-%
-%\begin{description}
-%\item[Continuity:] $B(t)$ is a continuos function
-%\item[Normal increments:]  $B(t)-B(s)$ has a normal distribution with mean $0$
-%and variance $t-s$.
-%\item[Independence of increments:] for every choice of nonnegative real numbers
-%$0 \leq s_1 <  t_1 \leq \cdots \leq s_n < t_n < \infty$, the increment random
-%variables $W_{t_1} - W_{s_1}, \cdots, W_{t_n} - W_{s_n}$ are jointly
-%independent.
-%\end{description}
-%
-%\subsection{Stochastic differential equation}
-%
-%An stochastic differential integral has the form:
-%
-%
-%\begin{equation}
-%W(T)=\int_0^T f(t)dB(t) \, .
-%\end{equation}
-%
-%\noindent This equations is also expressed in an abbreviate form:
-%
-%\begin{equation}
-%dW = f(t) dB \, .
-%\end{equation}
-%
-%\noindent Therefore, the integral form of the stock price model shown in
-%equation (\ref{eq:stockprice}) is:
-%
-%\begin{equation}
-%S(T)=\int_0^T \mu S(t) dt + \int_0^T \sigma S(t) dB(t)
-%\end{equation}
-%
-%\subsection{Ito's Lemma}
-%
-%Ito's lemma is used to find the differential of a time dependent function of a
-%stochastic process. In option pricing we need to find the option price $V(S(t))$
-%which depends on a stochastic stock price model $S(t)$.  $V(S,t)$ is required to
-%be  differentiable function of $S$ and once differentiable function of $t$.
-%
-%
-%
-%
-%
-%%These are known or can be easily obtained from the market, excepting by
-%volatility which must be estimated. However, rather than assuming a volatility
-%a priori and computing option prices from it, the model can be used to estimate
-%volatility at given prices, time to expiration and strike price. This obtained
-%volatility is called the implied volatility of an option. Additionally, some
-%models obtain implied volatility from futures (other derivative from prices). 
-%
-%
-% 
-%\section{Realized Volatility Models}
-%
-%\section{GARCH} test \section{ARFIMA} \section{Stochastic Volatility}
-%
-%\section{Volatility definition}
-%
-%\begin{itemize}
-%
-%\item Is the size of the price movement.
-%
-%\item Variance is a measure of distribution of returns and is not neccesarily
-%bound by any time period.  Volatility is a measure of the standard deviation
-%(square root of the variance) over a certain time interval. In finance,
-%variance and volatility both gives you a sense of an asset's risk. Variance
-%gives you a sense of the risk in the asset over its lifetime, while volatility
-%gives you a sense of the movement of the asset in eg. the past month or the
-%past year.
-%
-%
-%\item The main underlying difference is in their definition. Variance has a
-%fixed mathematical definition, however volatility does not as such. Volatility
-%is said to be the measure of fluctuations of a process.
-%
-%Volatility is a subjective term, whereas variance is an objective term i.e.
-%given the data you can definitely find the variance, while you can't find
-%volatility just having the data. Volatility is associated with the process, and
-%not with the data.
-%
-%In order to know the volatility you need to have an idea of the process i.e you
-%need to have an observation of the dispersion of the process. All the different
-%processes will have different methods to compute volatilities based on the
-%underlying assumptions of the process.  \end{itemize}
-%
-
-\subsection{Vector Autorregresive Models}\label{sec:varvec}
-
-%VECM is a special case of VAR model and both describe the joint behaviour
-%of a set of variables.
-
-The VAR($p$) model is a general framework describing the behaviour of a
+The vector autoregressive VAR($p$) model is one of the most easy to use, successful and flexible models for the analysis of multivariate time series.
+It is a natural extension of the univariate autoregressive (AR) model.
+The VAR model has proven to be useful for describing the dynamic behaviour of economic and financial time series.
+VAR is a general framework describing the behaviour of a
 set of $l$ endogenous variables as a linear combination of their last $p$
@@
 
-VECM is a special form of a VAR model for I(1) variables that are also
+\section{Vector error correction model}
+Vector error correction model (VECM) is a special form of a VAR model for I(1) variables that are also
 cointegrated~\cite{banerjee1993}.
@@
 
+\section{Unit root tests}
+
+\section{Johansen method}
+
 %\begin{equation}

### 2015-02-06 src/3_TimeSeries.tex
Spurious regression ready

@@
 
-In finance, a shock represents an unexpected change in a variable or a in its error term in a particular time period.
-For stationary time series, shocks to the system will gradually die away.
-That is, a shock during time $t$ will have a smaller effect in time $t+1$, a smaller effect on $t+2$ and so on.
-For non-stationary data, the persistence of shocks will always be infinite.
-The use of non-stationary data can lead to spurious regressions.
-If two stationary variables are generated as independent random series and are trending over time, when one of those variables is regressed on the other, they could have a high $R^2$ even if the two are totally unrelated.
-So, if standard regression techniques are applied to non-stationary data, it could look good under standard measures but valueless \cite{brooks2002}.
+In finance, a shock represents an unexpected change in a variable or a in its
+error term in a particular time period. For stationary time series, shocks to
+the system will gradually die away. That is, a shock during time $t$ will have a
+smaller effect in time $t+1$, a smaller effect on $t+2$ and so on. For
+non-stationary data, the persistence of shocks will always be infinite. The use
+of non-stationary data can lead to spurious regressions. Spurious regression
+dates back to Yule in 1926 \cite{yule1926} If two stationary
+variables are generated as independent random series and are trending over time,
+when one of those variables is regressed on the other, they could have a high
+$R^2$ even if the two are totally unrelated. So, if standard regression
+techniques are applied to non-stationary data, it could look good under standard
+measures but valueless \cite{brooks2002}.
+
+\texttt{Example of spurious regression} An example is shown below. Two random walks series $x_1t$ and $x_2t$ with a small drift ($\alpha=0.09$) are generated and
+then regressed one on the other:
+ \begin{equation}
+ x_i,t = \alpha + x_{i,t-1} + \epsilon_i,t  \qquad \text{where} \qquad \mathcal{N}(0,1), \quad i=1,2
+ \end{equation}
+
+\begin{figure}[!h]
+  \centering
+  \includegraphics[width=0.8\textwidth]{img/spurious1}
+  \caption{Two random walks time series $x_1,t$ and $x_2,t$}
+  \label{fig:spurious1}
+\end{figure}
+
+\begin{figure}[!h]
+  \centering
+  \includegraphics[width=0.8\textwidth]{img/spurious2}
+  \caption{Regression between two random walks time series}
+  \label{fig:spurious2}
+\end{figure}
+
+These two random walks, which are independent by construction, appear to be related with a $R^2=0.883$.
+However, what it is happening is that they are drifting in the same direction.
+Drift can be removed using returns or first differences.
+If after taking returns or first differences the regression fit is still good we will say that the variables are cointegrated, this means that both series may drift but their residuals will not.
+A more formal definition of cointegration is given in the section \ref{sec:cointegration}.
+
+%                            OLS Regression Results                            
+%==============================================================================
+%Dep. Variable:                      y   R-squared:                       0.883
+%Model:                            OLS   Adj. R-squared:                  0.882
+%Method:                 Least Squares   F-statistic:                     742.2
+%Date:                Fri, 06 Feb 2015   Prob (F-statistic):           1.61e-47
+%Time:                        06:07:47   Log-Likelihood:                -218.03
+%No. Observations:                 100   AIC:                             440.1
+%Df Residuals:                      98   BIC:                             445.3
+%Df Model:                           1                                         
+%==============================================================================
+%                 coef    std err          t      P>|t|      [95.0% Conf. Int.]
+%------------------------------------------------------------------------------
+%x1             0.8883      0.033     27.243      0.000         0.824     0.953
+%const          2.6726      0.405      6.606      0.000         1.870     3.475
+%==============================================================================
+%Omnibus:                        4.133   Durbin-Watson:                   0.363
+%Prob(Omnibus):                  0.127   Jarque-Bera (JB):                4.145
+%Skew:                          -0.469   Prob(JB):                        0.126
+%Kurtosis:                       2.662   Cond. No.                         23.3
+%==============================================================================
 
@@
 
-\subsection{Cointegration}
+\subsection{Cointegration} \label{sec:cointegration}
 Cointegration concept was introduced by Engle in 1987 \cite{engle1987} and implies that one or

### 2015-02-09 src/3_TimeSeries.tex
Cointegration example

@@
 smaller effect in time $t+1$, a smaller effect on $t+2$ and so on. For
-non-stationary data, the persistence of shocks will always be infinite. The use
+non-stationary data, the persistence of shocks will always be infinite.
+
+\subsection{Non-stationary processes}
+
+There are different types of non-stationary time series models often found in economics:
+
+\subsubsection{Deterministic trend}
+
+Deterministic trend or trend stationary processes have the following form:
+
+\[
+X_t = f(t) + \epsilon{t} \quad ,
+\]
+
+\noindent where $t$ is the time trend and $\epsilon{t}$ represents a stationary error term (with mean 0 and variance $\sigma^2$) and $f(t)$ is a deterministic function of time:
+\begin{itemize}
+\item If $f(t) = \alpha + \beta t$ we have a linear trend model which is widely used. 
+\item If $f(t)= \alpha \exp^rt$ we have an exponential growth curve.
+\item If $f(t) = c_1 + c_2t + c_3 t^2$ we have a quadratic trend model
+\item If $f(t) = \frac{1}{k+\alpha \beta ^t }$ we have a logistic curve
+\end{itemize}
+
+\subsubsection{Stochastic trend}
+Stochastic trend processes are also called unit root or difference stationarity processes and have the following form:
+
+\[
+X_t = \mu + X_{t-1} + \epsilon_t
+\]
+\noindent where $\epsilon_t$ is a stationary process. When $\mu = 0 $ the process is called pure random walk and when $\mu \neq 0$  the process is called random walk with drift.
+
+Alternatively this process can be expressed using the lag operator $L$ such as:
+
+\[
+(1-L) X_t = \mu  + \epsilon_t
+\]
+
+This process is also called unit root because the root of the characteristic equation ($1-z = 0$) is the unity.
+
+A random walk with or without drift can be transformed to a stationary process by differencing the time serie once.
+The disadvantage of differencing is that the process loses one observation each time the time series is differentiated.
+
+Apart from a stochastic trend, many economic financial time series seem to involve an exponential trend, this is the reason why researchers often take the logarithmic transformation before doing analysis.
+
+Other less common forms of non-stationarity are structural break in mean and structural break in variance.
+
+\newpage
+\subsection{Spurious regression}
+The use
 of non-stationary data can lead to spurious regressions. Spurious regression
@@
 
+\subsection{Unit root tests}
+
+\newpage
 \subsection{Integration}
@@
 variables~\cite{mukherjee1995},\cite{engle2004}.
+
+\subsection{Johansen method}
+Johansen in 1988 \cite{johansen1988} suggests a method for determining cointegration vectors.
+
+There are two statistics for cointegration under Johansen approach: the trace (shown in equation \ref{eq:johansentrace}) and the maximum-eigenvalue statistic (shown in equation \ref{eq:johansenmax}).
+
+\begin{equation}
+\label{eq:johansentrace}
+\lambda_{\text{trace}} (r) = -T \sum_{i=r+1}^g \ln(1-\hat{\lambda_i})
+\end{equation}
+
+\begin{equation}
+\label{eq:johansenmax}
+\lambda_{\text{max}} (r,r+1) = -T \ln(1-\hat{\lambda_{r+1}})
+\end{equation}
+
+Where $r$ is the number of cointegration vectors and $\hat]{\lambda}_i$ is the estimated value for the ith ordered eigenvalue.
+$\lambda_{\text{trace}}$ null hypothesis is that the number of cointegration vectors is less than or equal to $r$ against that there are more than $r$.
+$\lambda_{\text{max}}$ has the null hypothesis that the number of cointegration vectors is $r$ against an alternative of $r
++1$.
+
+\newpage
+\texttt{Cointegration example}
+
+If we have two-dimensional process $\mathbf{X}_t$, $t=1,\dots,T$ by:
+
+\begin{eqnarray*}
+\mathbf{X}_{1t} &=& \sum_{i=1}^t \epsilon_{1i} + \epsilon_{2t} \\
+\mathbf{X}_{2t} &=& a \sum_{i=1}^t \epsilon_{1i} + \epsilon_{3t} 
+\end{eqnarray*}
+
+Since $\mathbf{X}_{1t}$ and $\mathbf{X}_{2t}$ are I(1) processes and there
+exist a vector $\beta = [a -1]$ such that:
+
+\[
+\beta^\top \mathbf{X}_t = a \mathbf{X}_{1t} -\mathbf{X}_{2t} = 
+a\epsilon_{2t} - \epsilon_{3t} \sim \text{I(0)}
+\]
+
+then, both processes are said to be cointegrated.
+
+If we add a I(0) process
+$\mathbf{X}_{3t} = \epsilon_{4t}$  we find that there exists two cointegration
+vectors now: $\begin{bmatrix}a &-1& 0\end{bmatrix}$ and $\begin{bmatrix}0
+&0&1\end{bmatrix}$ since:
+
+\[
+\beta^\top \mathbf{X}_t = 
+\begin{bmatrix}
+a & -1 & 0 \\
+0 & 0 & 1
+\end{bmatrix} 
+\begin{bmatrix} 
+\mathbf{X}_{1t} \\
+\mathbf{X}_{2t} \\
+\mathbf{X}_{3t}
+\end{bmatrix} = 
+\begin{bmatrix}
+a\epsilon_{2t} - \epsilon_{3t} \\
+\epsilon_{4t}
+\end{bmatrix}
+\]
+
+\begin{figure}[!h]
+  \centering
+  \includegraphics[width=\textwidth]{img/coint-example}
+  \caption{Cointegration example: two processes I(1)(blue and green) and one process I(0) (red)}
+  \label{fig:cointex}
+\end{figure}
+
+This example shows how cointegration vectors describes the stable relations
+between the processes by linear relations that are more stationary than the
+original process.
+Cointegration vectors can be obtained using the Johansen method which gives with a certain probability the number of significant vectors.
+Figure \cite{fig:cointvectors1} shows the linear combinations obtained using cointegration vectors given by Johansen method.
+The method confirm that only two vectors are found and the third one given doesn't result in a I(0) process.
+
+\begin{figure}[!h]
+  \centering
+ \begin{minipage}{\textwidth} %[b][\textheight][s]
+  \centering
+  \includegraphics[width=0.8\textwidth]{img/coint-example2} 
+   \caption{Time series linear combination using the first cointegration vector }
+   \includegraphics[width=0.8\textwidth]{img/coint-example3} 
+    \caption{Time series linear combination using the second cointegration vector}
+    \includegraphics[width=0.8\textwidth]{img/coint-example4}
+ \caption{Time series linear combination using the third cointegration vector}
+  \label{fig:cointvectors1}
+\end{minipage}
+\end{figure}
+
+\newpage
 

### 2015-02-10 src/3_TimeSeries.tex
Adding Unit root tests

@@
 
-\subsection{Unit root tests}
-
 \newpage
@@
 
-\newpage
+\section{Unit root tests}
+
+Many economic and financial time series exhibit trending behaviour or non-stationary behaviour in the mean or variance.
+An important econometric task is to determine the type of non-stationary.
+For trend stationary I(0) time series a time-trend regression is appropriate.
+For I(1) time series taking first differences transforms it into a stationary one.
+For cointegration to determine if time series are I(1) is mandatory to model their long-run relationship.
+In order to determine if a time series is an a I(1) or I(0) process unit root tests are used.
+Several tests have been developed but the most widely used is an extension of the Dickey and Fuller test \cite{dickey1979} called the Augmented Dickey-Fuller (ADF) test developed in 1984 by Said and Dickey in 1984\cite{said1984}.
+
+The ADF test tests the null hypothesis that a time series $y_t$ is I(1) against the alternative that it is I(0), assuming that the dynamics in the data have an ARMA structure.
+The ADF test is based on estimating the test regression:
+
+\[
+y_t = \boldsymbol{\beta}^\top
+\]
 
@@
 
-\section{Unit root tests}
-
-\section{Johansen method}
-
 %\begin{equation}

### 2015-02-10 src/3_TimeSeries.tex
unit rood finished

@@
 
+\begin{equation}
+\label{eq:adftest}
+y_t = \boldsymbol{\beta}^\top \mathbb{D}_t + \phi y_{t-1} + \sum_{j=1}^p \Psi_j \Delta y_{t-j} + \epsilon_t
+\end{equation}
+
+\noindent where $\mathbb{D}_t$ is a vector of deterministic terms. Lag length $p$ has to be determined when applying the test, Akaike Information Criteria (AIC) is commonly used to determine it.
+The p-lagged difference terms $\Delta y_{t-j}$ are used to approximate an ARMA process and the value of $p$ is set so that the error $\epsilon_t$ is serially uncorrelated. 
+Under the null hypothesis, $y_t$ is I(1) which implies that $\phi=1$. 
+The test statistic is the following:
+
 \[
-y_t = \boldsymbol{\beta}^\top
+\text{ADF}_t = \frac{\hat{\phi} -1}{SE(\phi)}
 \]
+
+The $\text{ADF}_t$ statistic is based on the least squares estimates of equation \ref{eq:adftest}.
 

### 2015-02-10 src/3_TimeSeries.tex
Chapter 3 finished

@@
 \newpage
-\subsection{Spurious regression}
+\subsection{Spurious regression} \label{sec:spurious}
 The use
@@
 
-\subsection{Cointegration} \label{sec:cointegration}
+\subsection{Cointegration} \label{sec:coint}
 Cointegration concept was introduced by Engle in 1987 \cite{engle1987} and implies that one or
@@
 
-This example shows how cointegration vectors describes the stable relations
-between the processes by linear relations that are more stationary than the
-original process.
-Cointegration vectors can be obtained using the Johansen method which gives with a certain probability the number of significant vectors.
-Figure \cite{fig:cointvectors1} shows the linear combinations obtained using cointegration vectors given by Johansen method.
-The method confirm that only two vectors are found and the third one given doesn't result in a I(0) process.
-
 \begin{figure}[!h]
@@
   \centering
-  \includegraphics[width=0.8\textwidth]{img/coint-example2} 
+  \includegraphics[width=0.9\textwidth]{img/coint-example2} 
    \caption{Time series linear combination using the first cointegration vector }
-   \includegraphics[width=0.8\textwidth]{img/coint-example3} 
+    \label{fig:cointvectors3}
+   \includegraphics[width=0.9\textwidth]{img/coint-example3} 
     \caption{Time series linear combination using the second cointegration vector}
-    \includegraphics[width=0.8\textwidth]{img/coint-example4}
+     \label{fig:cointvectors4}
+    \includegraphics[width=0.9\textwidth]{img/coint-example4}
  \caption{Time series linear combination using the third cointegration vector}
-  \label{fig:cointvectors1}
+   \label{fig:cointvectors5}
 \end{minipage}
@@
 
+This example shows how cointegration vectors describes the stable relations
+between the processes by linear relations that are more stationary than the
+original process. Cointegration vectors can be obtained using the Johansen method which gives with a certain probability the number of significant vectors. Figures    
+\ref{fig:cointvectors3}, \ref{fig:cointvectors4}  and \ref{fig:cointvectors5}  shows the linear combinations obtained using cointegration vectors given by Johansen method.
+The method confirm that only two vectors are found and the third one given doesn't result in a I(0) process.
+
 \section{Unit root tests}
-
 Many economic and financial time series exhibit trending behaviour or non-stationary behaviour in the mean or variance.
@@
 \]
+\noindent where SE is the standard error.
 
@@
 can be solved with respect to $\mathbf{X}$ using the ordinary least
-squares estimation.
-
-\subsection{Ordinary Least Squares method}
+squares estimation (see section \ref{sec:OLS}).
+
+VECM models are employed because many economic time series appear to be first-difference stationary, i.e. I(1)
+
+Conventional regression estimators, including VARs, have good properties when applied to covariance-stationary time series, but encounter difficulties when applied to non stationary or integrated processes.
+
+These difficulties were illustrated by Granger and Newbold in 1974 \cite{granger1974} when they introduced the concept of spurious regressions (see section \ref{sec:spurious}).
+In 1982, Nelson and Plosser  \cite{nelson1982} showed that unit roots might be present in a wide variety of macroeconomic series
+in levels or logarithms. This finding gave rise to the industry of unit root testing, and
+the implication that variables should be rendered stationary by
+differencing before they are included in an econometric model.
+Further theoretical developments by Granger and Engle in 1987 \cite{engle87} raised the possibility that two or more integrated, non stationary time series might be cointegrated, so that some linear combination of these series could be stationary even though each series were not.
+
+If two series are both integrated (of order one, or I(1)) we could model
+their interrelationship by taking first differences of each series and
+including the differences in a VAR or a structural model.
+However, this approach would be suboptimal if it was determined that
+these series are indeed cointegrated. In that case, the VAR only
+express the short-run responses of these series to innovations in each
+series. This implies that the simple regression in first differences is
+misspecified.
+If the series are cointegrated, they move together in the long-run. A
+VAR in first differences will not capture those long-run tendences.
+VECM fit to the first differences of the non stationary variables, but a lagged error-correction term is added to the relationship representing the long-run relationship.
+
+\subsection{Ordinary Least Squares method} \label{sec:OLS}
 
@@
 \end{equation*}
-
-\begin{equation}
- \label{eq:varmatrix}
-               \underbrace{ \left[ \begin{array}{c}
-                \mathbf{y}_{p+1} \\
-                \mathbf{y}_{p+2} \\
-                \vdots           \\
-                \mathbf{y}_N  
-               \end{array} \right]  }_{\substack{ \mathbf{B}^\top\\ (N-p) \times l}}   
-= 
-\underbrace{
-  \begin{bmatrix}
-    \phi_1 & \dots & \phi_p & \mathbf{c}   
-  \end{bmatrix}}_{\substack{ \mathbf{X}^\top\\ l \times  (l \times p+1)}}
-\underbrace{\begin{bmatrix}
-   \mathbf{y}_p     & \mathbf{y}_{p+1}  & \dots     & \mathbf{y}_{N-1}\\
-   \mathbf{y}_{p-1} & \mathbf{y}_{p}  & \dots     & \mathbf{y}_{N-2} \\
-   \vdots           & \vdots            & \ddots    & \vdots  \\
-   \mathbf{y}_{1}   & \mathbf{y}_{2}& \dots      & \mathbf{y}_{N-p} \\
-   1                &  1                & \dots      &             1 
-   \end{bmatrix}}_{\substack{ \mathbf{A}^\top\\  (l \times p+1) \times (N-p)}}
-+
-\underbrace{\begin{bmatrix}
-              \mathbf{\epsilon}_{p+1}  \\ 
-              \mathbf{\epsilon}_{p+2}  \\ 
-               \vdots                 \\
-              \mathbf{\epsilon}_N     
-             \end{bmatrix}}_{\substack{\mathbf{E}^\top\\ (N-p) \times l }} 
-\end{equation}
-
-%\begin{equation}
-% \label{eq:vecmatrix}
-%\underbrace{
-%      \begin{bmatrix}
-%       \mathbf{\Delta y}_{p+1}  \\ 
-%       \mathbf{\Delta y}_{p+2}  \\ 
-%       \vdots                   \\ 
-%       \mathbf{\Delta y}_N      
-%      \end{bmatrix}}_{\mathbf{B} } =
-%\arraycolsep=1.4pt  
-%\underbrace{\begin{bmatrix}
-%   \alpha^\top & \phi^*_1 & \dots & \phi^*_{p-1} &\mathbf{c}   
-%  \end{bmatrix}}_{\mathbf{X}^\top}
-%\underbrace{\begin{bmatrix}
-% \beta\mathbf{y}_p      & \beta\mathbf{y}_{p+1}   & \dots&   \beta\mathbf{y}_{N-1}   \\
-% \mathbf{\Delta y}_{p}   & \mathbf{\Delta y}_{p+1}& \dots &  \mathbf{\Delta y}_{N-1} \\
-% \mathbf{\Delta y}_{p-1} & \mathbf{\Delta y}_{p}  & \dots &  \mathbf{\Delta y}_{N-2}   \\
-% \vdots                  & \vdots                 & \ddots&  \vdots                   \\
-% \mathbf{\Delta y}_{2}   & \mathbf{\Delta y}_{3} & \dots &   \mathbf{\Delta y}_{N-p+1} \\
-% 1                      & 1                       & \dots     & 1   
-% \end{bmatrix}}_{\mathbf{A}^\top}
-%+
-%\underbrace{\begin{bmatrix}
-%              \mathbf{\epsilon}_{p+1} \\ 
-%              \mathbf{\epsilon}_{p+2} \\ 
-%              \vdots \\ 
-%              \mathbf{\epsilon}_N
-%             \end{bmatrix}}_{\mathbf{E}^\top}
-%\end{equation}
-
-%VECMs may be estimated by Stata’s vec command. These models
-%
-%are employed because many economic time series appear to be
-%
-%‘first-difference stationary,’ with their levels exhibiting unit root or
-%
-%nonstationary behavior. Conventional regression estimators, including
-%
-%VARs, have good properties when applied to covariance-stationary
-%
-%time series, but encounter difficulties when applied to nonstationary or
-%
-%integrated processes.
-%
-%These difficulties were illustrated by Granger and Newbold
-%
-%(J. Econometrics, 1974) when they introduced the concept of spurious
-%
-%regressions. If you have two independent random walk processes, a
-%
-%regression of one on the other will yield a significant coefficient, even
-%
-%though they are not related in any way.
-%
-%
-%This insight, and Nelson and Plosser’s findings (J. Mon. Ec., 1982) that
-%
-%unit roots might be present in a wide variety of macroeconomic series
-%
-%in levels or logarithms, gave rise to the industry of unit root testing, and
-%
-%the implication that variables should be rendered stationary by
-%
-%differencing before they are included in an econometric model.
-%
-%Further theoretical developments by Granger and Engle in their
-%
-%celebrated paper (Econometrica, 1987) raised the possibility that two
-%
-%or more integrated, nonstationary time series might be cointegrated, so
-%
-%that some linear combination of these series could be stationary even
-%
-%though each series is not.
-%
-%
-%
-%If two series are both integrated (of order one, or I(1)) we could model
-%
-%their interrelationship by taking first differences of each series and
-%
-%including the differences in a VAR or a structural model.
-%
-%However, this approach would be suboptimal if it was determined that
-%
-%these series are indeed cointegrated. In that case, the VAR would only
-%
-%express the short-run responses of these series to innovations in each
-%
-%series. This implies that the simple regression in first differences is
-%
-%misspecified.
-%
-%If the series are cointegrated, they move together in the long run. A
-%
-%VAR in first differences, although properly specified in terms of
-%
-%covariance-stationary series, will not capture those long-run
-%
-%tendences.
-%
-%Accordingly, the VAR concept may be extended to the vector
-%
-%error-correction model, or VECM, where there is evidence of
-%
-%cointegration among two or more series. The model is fit to the first
-%
-%differences of the nonstationary variables, but a lagged error-correction
-%
-%term is added to the relationship.
-%
-%In the case of two variables, this term is the lagged residual from the
-%
-%cointegrating regression, of one of the series on the other in levels. It
-%
-%expresses the prior disequilibrium from the long-run relationship, in
-%
-%which that residual would be zero.
-%
-%In the case of multiple variables, there is a vector of error-correction
-%
-%terms, of length equal to the number of cointegrating relationships, or
-%
-%cointegrating vectors, among the series.

### 2015-02-17 src/4_OnlineLearning.tex
Inicio capitulo ML

@@
 \chapter{Machine Learning Models}
+Machine learning is a scientific discipline focused on the development of algorithms focused on learning from examples.
+This idea has become central to the design of search engines, robots systems and forecasts applications which process large data sets.
+However, machines require an extended training period when developing algorithms to predict future behaviour such as financial time series forecasting where new data is available in a short period of time.
+Online machine learning techniques tackle this problems and allow to update the model with new data or to compute a new model using less data and in a short period of time.
+
+\vspace{0.5cm}
+
+\section{Introduction}
+
+There are three types of Machine Learning (ML) classified depending on the nature of the learning input or output available to a learning system.
+\begin{description}
+\item[supervised learning] also known as classification or regression \cite{bishop2006}, the input data  is a pair which contains the example input and its desired outputs (also called labels).
+The goal is to learn a general rule that maps inputs to outputs.
+ Its two mainstream approaches are: support vector machines (SVMs) \cite{vapnik1998} and ensemble learning \cite{breiman1998}.
+\item[unsupervised learning] also known as clustering \cite{ben2005} and reinforcement learning \cite{sutton1998}.
+\end{description}
+
+All three types of problems can be viewed as optimization problems. The ML core task is to define a learning criterion, i.e the function to be optimized.
+
 \section{Batch learning}

### 2015-02-17 src/4_OnlineLearning.tex
bath learning tomorrow

@@
 However, machines require an extended training period when developing algorithms to predict future behaviour such as financial time series forecasting where new data is available in a short period of time.
-Online machine learning techniques tackle this problems and allow to update the model with new data or to compute a new model using less data and in a short period of time.
+Online machine learning techniques tackle this problem and allow to update the model with new data or to compute a new model using less data allowing to give a response in a short period of time.
 
@@
 \section{Introduction}
+Machine Learning (ML) studies computer algorithms for learning
+something such as to complete a task, to make accurate predictions or to behave intelligently.
+The learning is always based on samples and the objective is about to do better in the future based on the past experiences in an automatic way.
+ML is a subarea of artificial intelligence and broadly intersects with other fields such as statistics, mathematics, physics and computer science.
+There are many examples of machine learning problems: time series forecasting, image processing, face detection, spam filtering, weather prediction, search engines, among many others.
 
-There are three types of Machine Learning (ML) classified depending on the nature of the learning input or output available to a learning system.
+ML models are often more accurate than what can be created through direct programming. The reason is that ML models are data driven and are able to examine large amounts of data.
+
+There are three types of ML classified depending on the nature of the learning input or output available to a learning system.
 \begin{description}
-\item[supervised learning] also known as classification or regression \cite{bishop2006}, the input data  is a pair which contains the example input and its desired outputs (also called labels).
-The goal is to learn a general rule that maps inputs to outputs.
+\item[supervised learning]  the input data is a tuple which contains the example input and its desired outputs (also called labels).
+The list of tuples is called the training set.
+The concept of supervised learning comes from the supervisor, acting as a teacher in the learning process.
+The goal is to learn a general rule that maps inputs to outputs optimising a target function.
+There are two related problem types in supervised learning: classification and regression problems \cite{bishop2006}.
 Its two mainstream approaches are: support vector machines (SVMs) \cite{vapnik1998} and ensemble learning \cite{breiman1998}.
-\item[unsupervised learning] also known as clustering \cite{ben2005} and reinforcement learning \cite{sutton1998}.
+Furthermore supervised learning can be categorised into offline or batch learning and online learning (see section \ref{sec:batch} and \ref{sec:online}).
+\item[unsupervised learning] also known as clustering \cite{ben2005}.
+In this type of learning no labels are given and the system has to find a structure on its own, discovering hidden patterns in data.
+\item[reinforcement learning] is the problem faced by an agent that must learn 
+through trial-and-error interactions with a dynamic environment. It is based on programming agents by reward and punishment without needing to specify how the task is to be achieved \cite{sutton1998}.
 \end{description}
@@
 
-\section{Batch learning}
-\section{Online learning}
+\section{Batch learning} \label{sec:batch}
+
+Batch learning, also called statistical or offline learning, is a supervised machine learning framework.
+In batch learning, there is a data set available where the learner can build the internal model without any limits in accessing the data.
+There is time enough to carefully analyse the dataset, build large predictive models and combine them in a sophisticated way.
+
+%%% Learning problem. Ac;a ma;ana, formulas y descripcion de tallada del problema de aprendizaje
+
+\section{Online learning} \label{sec:online}
+
+In contrast to offline learning, online learning has access to a sample only once.
+The goal is the same, predicting targets as accurate as possible.
+For example, stock market prediction can be seen as online learning.
+The algorithm makes a prediction of the stock, little time after the real stock price is available, this information can be incorporated to the learner to further improve the prediction accuracy.
+In general, there is very much data available in an online learning setup, the data set grows continuously.
+Offline learning has equal or superior accuracy compared to online learning when the same amount of data is used.
+
 \subsection{Online learning}
@@
 Performance of an online learning algorithm is measured by the cumulative loss it suffers along its run on a sequence of examples.
-In order to minimize this loss, the learner may update the hypothesis after each round so as to be more accurate in later rounds.
+In order to minimise this loss, the learner may update the hypothesis after each round so as to be more accurate in later rounds.
+
+%%% Measure error

### 2015-02-18 src/4_OnlineLearning.tex
Learning theory

@@
 Its two mainstream approaches are: support vector machines (SVMs) \cite{vapnik1998} and ensemble learning \cite{breiman1998}.
-Furthermore supervised learning can be categorised into offline or batch learning and online learning (see section \ref{sec:batch} and \ref{sec:online}).
+Furthermore supervised learning can be categorised into offline or batch learning and online learning (see section \ref{sec:onoffline}).
 \item[unsupervised learning] also known as clustering \cite{ben2005}.
@@
 
-All three types of problems can be viewed as optimization problems. The ML core task is to define a learning criterion, i.e the function to be optimized.
+%%% Learning problem. Ac;a ma;ana, formulas y descripcion de tallada del problema de aprendizaje
 
-\section{Batch learning} \label{sec:batch}
+\section{Statistical learning theory} \label{sec:mltheory}
+
+All ML problems can be viewed as optimisation problems. The ML core task is to define a learning criterion, i.e the function to be optimised.
+
+In this thesis we will emphasised the supervised learning problem since it is the most common way of modelling financial problems.
+
+Supervised learning consists in to find a learning function $f: X \rightarrow Y$ which models a set of examples called training set $S$ which have been drawn randomly and independently according to a unknown joint distribution function $p(x,y)$ on $X \times Y$:
+
+$$
+S= \{(x_1,y_1),\,(x_2,y_2),\dots\,,\,(x_n,y_n)\} \in \mathcal{S}, (x_i,y_i) \in X\times Y \quad \forall i = 1\dots n
+$$
+
+\noindent where $X \subseteq \mathbb{R}^m$ and $Y = \mathbb{R}$ in case of regression problems and $Y = \{1,-1\}$  if it is a classification problem.
+
+A learning algorithm $\mathcal{A}$ takes as input a data set $S \in \mathcal{S}$ and output a function $f_S$:
+
+\begin{eqnarray*}
+\mathcal{A}: \mathcal{S} &\rightarrow & \mathcal{H} \\
+S &\rightarrow &\mathcal{A}(S) = f_S
+\end{eqnarray*}
+
+\section{Batch and Online learning} \label{sec:onoffline}
 
@@
 There is time enough to carefully analyse the dataset, build large predictive models and combine them in a sophisticated way.
-
-%%% Learning problem. Ac;a ma;ana, formulas y descripcion de tallada del problema de aprendizaje
-
-\section{Online learning} \label{sec:online}
 

### 2015-02-20 src/4_OnlineLearning.tex
generalisation concept

@@
 
+\noindent where $\mathcal{H}$ is called the hypothesis space, is the space of functions that the algorithm is allowed to search.
+The selection of $f_S$ is based on a loss function $V(f(x),y)$ which expected risk has to be minimised.
+
+\[
+E[V(f(x),y)] = \int V(f(x),y) dp(x,y)
+\]
+
+\noindent $V(f(x),y)$ denote the price paid for mistakes. Therefore, $V(f(x),y)=0$ if$f(x)=y$.
+
+For regression, the most common loss function is square loss or L2 loss function:
+\begin{equation}
+\label{eq:l2loss}
+V(f(x),y) = (f(x)-y)^2
+\end{equation}
+\noindent another option is the L1 loss
+\begin{equation}
+\label{eq:l1loss}
+V(f(x),y) = |f(x)-y| \, .
+\end{equation}
+\noindent the choice of loss function here gives rise to several well-known learning algorithms such as regularised least squares and support vector machines. %verificar esto7776
+
+Since the true distribution is unknown and only training samples are available, the objective is to estimate a function $\hat{f}$ through empirical risk (training error) minimisation (ERM):
+
+\begin{equation} 
+\label{eq:erm}
+R_{\text{emp}}[f] = \frac{1}{n} \sum_{i=1}^n V(f(x),y)
+\end{equation}
+
+Since ERM is an ill-posed problem, Tikhonov introduced a regularisation which ensures well-posedness and generalisation of ERM by constraining the hypothesis space $\mathcal{H}$ usually called regularised ERM or Tikhonov regularisation.
+
+Tikhonov regularisation minimised over the hypothesis space $\mathcal{H}$ for a fixed positive parameter $\lambda$ the following:
+
+\begin{equation} 
+\label{eq:rerm}
+R_{\text{emp}}[f] = \frac{1}{n} \sum_{i=1}^n V(f(x),y) + \lambda \mathcal{R}(f)
+\end{equation}
+
+\noindent where $\mathcal{R}(f)$ is the regulariser, a penalisation on $f$.
+
+In all learning algorithms that are trained from example data, there is a tradeoff \cite{dietterich2003} between three factors:
+\begin{description}
+\item[Complexity of the hypothesis class] for best generalisation, we should adjust the complexity of the learner model $\hat{f}$ to the complexity of the data $f$.
+In polynomial regression, the complexity parameter is the order of the fitted polynomial, and therefore we need to find a way to choose the best order that minimises the generalisation error, that is, tune the complexity of the model to best fit the complexity of the function inherent in the data.
+\item[Generalisation accuracy on new examples] is the capability of the algorithm to generate the right output for an input instance outside the training set.
+\item[The amount of training data]  generally generalisation accuracy increases as the amount of training data increases.
+\end{description}
+
+Very complex models will fit the training data better (low bias) but it could overfit it and generalise poorly (high variance), this is also called the bias-variance tradeoff.
+Generalisation accuracy can sometimes be improved with more training data.
+The figure \ref{fig:tradeoff} illustrates how generalisation accuracy depends on the complexity of the model and the amount of training data.
+
+\begin{figure}[!h]
+  \centering
+  \includegraphics[width=0.8\textwidth]{img/3tradeoff}
+  \caption{Triple tradeoff in empirical learning}
+  \label{fig:tradeoff}
+\end{figure}
+
+% Revisar!!!! _========================================
+
+The bias-variance tradeoff
+
+Finding an \hat{f} that generalizes to points outside of the training set can be done with any of the countless algorithms used for supervised learning.
+It turns out that whichever function \hat{f} we select, we can decompose its expected error on an unseen sample x as follows:[3]:34[4]:223
+
+\begin{align}
+\mathrm{E}\Big[\big(y - \hat{f}(x)\big)^2\Big]
+ & = \mathrm{Bias}\big[\hat{f}(x)\big]^2 + \mathrm{Var}\big[\hat{f}(x)\big] + \sigma^2 \\
+\end{align}
+Where:
+
+\begin{align}
+ \mathrm{Bias}\big[\hat{f}(x)\big] = \mathrm{E}\big[\hat{f}(x)\big] - f(x)
+\end{align}
+and
+
+\begin{align}
+\mathrm{Var}\big[\hat{f}(x)\big] = \mathrm{E}\Big[ \big( \hat{f}(x) - \mathrm{E}[\hat{f}(x)] \big)^2 \Big] 
+\end{align}
+
+% Revisar!!!! _========================================
+
 \section{Batch and Online learning} \label{sec:onoffline}

### 2015-02-23 src/4_OnlineLearning.tex
Bias-variance tradeoff

@@
 
-The bias-variance tradeoff
+\subsection{The bias-variance tradeoff}
 
-Finding an \hat{f} that generalizes to points outside of the training set can be done with any of the countless algorithms used for supervised learning.
-It turns out that whichever function \hat{f} we select, we can decompose its expected error on an unseen sample x as follows:[3]:34[4]:223
+Models learning error can be split into two main components: error due to bias and error due to variance.
+Bias measures how far off in general these models' predictions are from the correct value.
+Variance is taken as the variability of a model prediction for a given data point.
+
+If we have a learning model $\hat{f}$, its expected error on an unseen sample $x$ can be decomposed as follows:
 
@@
 \mathrm{E}\Big[\big(y - \hat{f}(x)\big)^2\Big]
- & = \mathrm{Bias}\big[\hat{f}(x)\big]^2 + \mathrm{Var}\big[\hat{f}(x)\big] + \sigma^2 \\
+ & = \mathrm{Bias}\big[\hat{f}(x)\big]^2 + \mathrm{Var}\big[\hat{f}(x)\big] + \sigma^2
 \end{align}
-Where:
+\noindent where:
+\begin{align*}
+ \mathrm{Bias}\big[\hat{f}(x)\big] &= \mathrm{E}\big[\hat{f}(x)\big] - f(x) \\
+ \mathrm{Var}\big[\hat{f}(x)\big] &= \mathrm{E}\Big[ \big( \hat{f}(x) - \mathrm{E}[\hat{f}(x)] \big)^2 \Big] 
+\end{align*}
 
-\begin{align}
- \mathrm{Bias}\big[\hat{f}(x)\big] = \mathrm{E}\big[\hat{f}(x)\big] - f(x)
-\end{align}
-and
+Bias is introduced by the model selection.
+Therefore the model building process is repeated (through resampling) and substantially different averages of prediction values are obtained, bias will be high.
+The error due to variance is the amount by which the prediction, over one training set, differs from the expected predicted value, over all the training sets.
+Variance measures how inconsistent are the predictions from one another, over different training sets, not whether they are accurate or not.
 
-\begin{align}
-\mathrm{Var}\big[\hat{f}(x)\big] = \mathrm{E}\Big[ \big( \hat{f}(x) - \mathrm{E}[\hat{f}(x)] \big)^2 \Big] 
-\end{align}
-
-% Revisar!!!! _========================================
+The objective is to simultaneously reduce bias and variance as much as possible in order to obtain as accurate model as is feasible.
+However, there is a tradeoff to be made when selecting models.
+Models that exhibit small variance and high bias underfit the truth target.
+Models that exhibit high variance and low bias overfit the truth target.
 

### 2015-02-23 src/4_OnlineLearning.tex
regularisation ready
now ridge regression

@@
 
-Since ERM is an ill-posed problem, Tikhonov introduced a regularisation which ensures well-posedness and generalisation of ERM by constraining the hypothesis space $\mathcal{H}$ usually called regularised ERM or Tikhonov regularisation.
-
-Tikhonov regularisation minimised over the hypothesis space $\mathcal{H}$ for a fixed positive parameter $\lambda$ the following:
-
-\begin{equation} 
-\label{eq:rerm}
-R_{\text{emp}}[f] = \frac{1}{n} \sum_{i=1}^n V(f(x),y) + \lambda \mathcal{R}(f)
-\end{equation}
-
-\noindent where $\mathcal{R}(f)$ is the regulariser, a penalisation on $f$.
-
+\subsection{Learning algorithms}
 In all learning algorithms that are trained from example data, there is a tradeoff \cite{dietterich2003} between three factors:
@@
 
-% Revisar!!!! _========================================
-
 \subsection{The bias-variance tradeoff}
@@
 
-If we have a learning model $\hat{f}$, its expected error on an unseen sample $x$ can be decomposed as follows:
+If we have a learning model $\hat{f}$, its expected generalisation error on an unseen or testing sample $x$ can be decomposed as follows:
 
@@
 
+This generalisation error is unknown, but it can be estimated using a training sample, a testing sample or by estimating the model complexity.
+Figure \ref{fig:traintesterror} shows prediction error for training and testing sample.
+In general, in the training sample prediction error is decreasing with more complex models but it overfit the data and it is not a good model for the test data.
+Figure \ref{fig:bvtradeoff} shows how testing error can be decomposed into bias and variance components.
+The best model will be the one has a balance between bias and variance.
+
+\begin{figure}[!h]
+  \centering
+  \includegraphics[width=0.8\textwidth]{img/model_complexity}
+  \caption{Training and test error}
+  \label{fig:traintesterror}
+\end{figure}
+
 The objective is to simultaneously reduce bias and variance as much as possible in order to obtain as accurate model as is feasible.
@@
 Models that exhibit high variance and low bias overfit the truth target.
+
+This bias-variance tradeoff is the reason why in order to obtain the best model, data is usually broken in three subsets: training set, validation set and testing set.
+Training set is used to determine the model, validation set is used to estimate the generalisation error and finally testing set is used to estimate the accuracy of the model.
+Usually the partition is 50\% for training set, 25\% for validation and 25\% for testing purposes.
+This procedure can be extended to a Cross Validation (CV) procedure \cite{geisser1975}, usually used when the amount of data is limited.
+Various splitting strategies lead to various CV estimates.
+In K-fold cross-validation the training data is divided randomly into K distinct subsets, then the network is trained using K-1 subsets, and tested on the remaining subset.
+The process of training and testing is then repeated for each of the K possible choices of the subset omitted from the training.
+The average performance on the K omitted subsets is then our estimate of the generalisation performance.
+
+\begin{figure}[!h]
+  \centering
+  \includegraphics[width=0.8\textwidth]{img/biasvariancetradeoff}
+  \caption{Bias variance tradeoff}
+  \label{fig:bvtradeoff}
+\end{figure}
+
+Another way to control the complexity of the model is to include in the optimisation function not only the generalisation error but a penalisation of high complex models.
+
+\subsection{Regularization}
+Since ERM is an ill-posed problem, Tikhonov introduced a regularisation which ensures well-posedness and generalisation of ERM, i.e prevents overfit, by constraining the hypothesis space $\mathcal{H}$ usually called regularised ERM or Tikhonov regularisation.
+
+Tikhonov regularisation minimised over the hypothesis space $\mathcal{H}$ for a fixed positive parameter $\lambda$ the following:
+
+\begin{equation} 
+\label{eq:rerm}
+R_{\text{emp}}[f] = \frac{1}{n} \sum_{i=1}^n V(f(x),y) + \lambda \mathcal{R}(f)
+\end{equation}
+
+\noindent where $\mathcal{R}(f)$ is the regulariser, a penalisation on $f$.
+
+One example of regularisation in linear models is ridge regression, which is a regularised least squares method.
+The least squares (LS) method is a well known way to solve a regression
+problem. 
+LS method consists of minimising the sum of squared errors:
+
+%\begin{eqnarray*}
+%\label{eq:problem2}
+% J(\mathbf{\phi}) &=& \sum_{t=1}^N (f(\mathbf{x}_t)-y_t)^2 
+% = \sum_{t=1}^N (\mathbf{\phi}^\intercal {\mathbf{x}}_t-y_t)^2 
+% = \| \mathbf{X}\mathbf{\phi} - \mathbf{Y} \|_2^2 
+%\end{eqnarray*}
+
+% Aca seguir agregando slides.tex
 

### 2015-02-24 src/4_OnlineLearning.tex
ann and svm

@@
 
-One example of regularisation in linear models is ridge regression, which is a regularised least squares method.
+\subsection{Ridge Regression}
+One example of regularisation in linear models is Ridge Regression (RR), which is a regularised least squares method.
 The least squares (LS) method is a well known way to solve a regression
@@
 
-%\begin{eqnarray*}
-%\label{eq:problem2}
-% J(\mathbf{\phi}) &=& \sum_{t=1}^N (f(\mathbf{x}_t)-y_t)^2 
-% = \sum_{t=1}^N (\mathbf{\phi}^\intercal {\mathbf{x}}_t-y_t)^2 
-% = \| \mathbf{X}\mathbf{\phi} - \mathbf{Y} \|_2^2 
-%\end{eqnarray*}
+\begin{eqnarray*}
+\label{eq:lsm}
+ J(\mathbf{\phi}) =& \sum_{t=1}^N (f(\mathbf{x}_t)-y_t)^2 \\
+ =& \sum_{t=1}^N (\mathbf{\phi}^\top {\mathbf{x}}_t-y_t)^2 \\
+ =& \| \mathbf{X}\mathbf{\phi} - \mathbf{Y} \|_2^2 
+\end{eqnarray*}
 
-% Aca seguir agregando slides.tex
+The optimal solution for $\mathbf{\phi}$ is:
+\begin{equation*}
+\mathbf{\phi}=(\mathbf{X}^\top
+\mathbf{X})^{-1}\mathbf{X}^\top \mathbf{Y}
+\end{equation*}
 
-\section{Batch and Online learning} \label{sec:onoffline}
+In order to avoid the singularity of the matrix $\mathbf{X}^\top \mathbf{X}$, a regularisation term is introduced:
 
-Batch learning, also called statistical or offline learning, is a supervised machine learning framework.
-In batch learning, there is a data set available where the learner can build the internal model without any limits in accessing the data.
-There is time enough to carefully analyse the dataset, build large predictive models and combine them in a sophisticated way.
+\begin{equation*}
+\label{eq:problem} 
+J(\mathbf{\phi}) =  \| \mathbf{X}\mathbf{\phi} - \mathbf{Y} \|_2^2  + \lambda
+ \| \mathbf{\phi}\| ^2
+\end{equation*}
 
-In contrast to offline learning, online learning has access to a sample only once.
-The goal is the same, predicting targets as accurate as possible.
+\noindent which optimal solution $\mathbf{\phi}_*$ is well known:
+
+\begin{equation*}
+\label{eq:optsolRR}
+\mathbf{\phi}_*=(\mathbf{X}^\top \mathbf{X}+\lambda \mathbb{I})^{-1}\mathbf{X}^\top y \, ,
+\end{equation*}
+
+\texttt{Demo}
+\begin{eqnarray*}
+J(\mathbf{\phi}) &=&  \| \mathbf{X}\mathbf{\phi} - \mathbf{Y} \|_2^2  + \lambda
+ \| \mathbf{\phi}\| ^2 \\
+ &=&  \sum_{t=1}^N (\mathbf{\phi}^\top {\bf x}_t-y_t)^2 + \lambda \sum_{i=1}^p \phi_i^2 \\
+ &=& (\phi^\top x_1-y_1)^2 + \dots + (\phi^\top x_N-y_N)^2 + \lambda (\phi_1^2+\dots+\phi_p^2)
+\end{eqnarray*}
+\noindent taking derivatives
+ \begin{eqnarray*}
+ \frac{\partial J(\mathbf{\phi})}{\partial \phi_1}&=& 
+ 2(\phi^\top \mathbf{x}_1-y_1)\mathbf{x}_{11} + \dots + 2(\phi^\top \mathbf{x}_N-y_N)\mathbf{x}_{N1} + 2\lambda \phi_1 \\
+ &=& 2\mathbf{x}_1^\top(\mathbf{X}\phi-\mathbf{Y}) + 2\lambda\phi_1\\
+& \vdots &\\
+  \frac{\partial J(\mathbf{\phi})}{\partial \phi_p}&=& 
+ 2\mathbf{x}_p^\top(\mathbf{X}\phi-\mathbf{Y}) + 2\lambda\phi_p 
+\end{eqnarray*}
+
+Then we have that:
+\begin{eqnarray*}
+\frac{\partial J(\mathbf{\phi})}{\partial \phi}&=& 
+ 2\mathbf{X}^\top(\mathbf{X}\phi-\mathbf{Y}) + 2\lambda\phi
+\end{eqnarray*}
+Since $\frac{\partial J(\mathbf{\phi})}{\partial \phi}=0$ we have:
+\begin{eqnarray*}
+2\mathbf{X}^\top(\mathbf{X}\phi-\mathbf{Y}) + 2\lambda\phi&=&0 \\
+\mathbf{X}^\top\mathbf{X}\phi - \mathbf{X}^\top\mathbf{Y} + \lambda\phi &=& 0\\
+(\mathbf{X}^\top\mathbf{X}+\lambda\mathbb{I})\phi &=&  \mathbf{X}^\top\mathbf{Y} \\
+\phi &=& (\mathbf{X}^\top\mathbf{X}+\lambda\mathbb{I})^{-1}  \mathbf{X}^\top\mathbf{Y}
+\end{eqnarray*}
+
+When $\mathbf{X}$ is orthonormal then $\mathbf{X}^\top \mathbf{X} =\mathbb{I}$ and there is a simple relation between the ridge estimator and the OLS estimator:
+\begin{eqnarray*}
+\phi_* (\lambda) &=& (\mathbf{X}^\top \mathbf{X}+\lambda \mathbb{I})^{-1}\mathbf{X}^\top \mathbf{Y} \\
+ &=& (\mathbb{I} + \lambda \mathbb{I})^{-1} \mathbf{X}^\top \mathbf{Y} \\
+ &=&(1+\lambda)^{-1} \mathbb{I} \mathbf{X}^\top \mathbf{Y} \\
+ &=&(1+\lambda)^{-1} (\mathbf{X}^\top \mathbf{X})^{-1}\mathbf{X}^\top \mathbf{Y} \\
+ &=&(1+\lambda)^{-1} \phi
+\end{eqnarray*}
+
+\subsection{Machine learning algorithms}
+In recent years many successful machine learning applications have been developed.
+Artificial neural network (ANN) and Support Vector Machines (SVM) have been some of the most popularly used machine learning algorithm.
+Historically, SVMs emerged after ANN.
+
+ANN is inspired by biological learning systems even though it does not mimic it fully.
+ANN are organised into layers and have unidirectional connections between the layers, feed-forward network is the most used.
+
+\begin{figure}[!h]
+  \centering
+  \includegraphics[width=0.8\textwidth]{img/ffn}
+  \caption{Feed-forward neural network}
+  \label{fig:ffn}
+\end{figure}
+
+SVM on the other hand is a relatively new learning algorithm. It can be
+similarly used to learn target functions. However, unlike ANN, it is very well founded
+based on theory in statistical learning [2]. The major difference between SVM and ANN
+is in the error optimization. In ANN, the aim of learning is to obtain a set of weight
+values which minimize the training error while in SVM the training error is set to
+minimum while training adjust the capacity of the machine. During training, SVM
+learned the parameters a?s and the number of support vectors which is equivalent to the
+number of hidden units in ANN.
+the nodes are artificial
+neurons and directed edges are connections between neuron outputs and neuron inputs.
+The main characteristics of neural networks are that they have the ability to learn
+complex nonlinear input-output relationships, they use sequential training procedures and
+they adapt themselves to the data.
+
+Support Vector Machine (SVM) was introduced in 1992 by Vapnik and his coworkers
+[3]. In its original form, SVM is a training algorithm for linear classification.
+Only later it was used for regression, principal component analysis, novelty detection and
+also for non-linear case. SVM tunes the capacity of the classification function by
+maximizing the margin between the training patterns and the decision boundary. The
+solution is expressed as a linear combination of supporting patterns, which are the subset
+of training patterns close to the decision boundary, called the support vectors.
+
+For nonlinear
+case, SVM mapped the data sets of input space into a higher dimensional feature
+space, which is linear and the large-margin learning algorithm is then applied. However,
+the mapping can be implicitly done by kernel functions. In the high dimensional feature
+space, simpler and linear hyper plane classifiers that have maximal margin between the
+classes can be obtained.
+
+\subsection{Kernel methods}
+
+Kernel functions may be used to introduce a nonlinear separation
+in input space despite the linear classifier in decision space
+
+\section{Online learning} \label{sec:onoffline}
+
+Online learning is a supervised machine learning framework that is useful when we have sequential access to a sample only once.
+This differs from the classical batch learning, where there is an entire data set available and the learner can build the internal model without any limits in accessing the data.
+In batch learning, there is time enough to carefully analyse the dataset, build large predictive models and combine them in a sophisticated way.
+
+In only learning the goal is the same as batch learning, predicting targets as accurate as possible.
 For example, stock market prediction can be seen as online learning.
 The algorithm makes a prediction of the stock, little time after the real stock price is available, this information can be incorporated to the learner to further improve the prediction accuracy.
-In general, there is very much data available in an online learning setup, the data set grows continuously.
+In general, there is too much data available in an online learning setup, the data set grows continuously.
 Offline learning has equal or superior accuracy compared to online learning when the same amount of data is used.
-
-\subsection{Online learning}
 

### 2015-02-27 src/3_TimeSeries.tex
cambios de notación

@@
 
-\texttt{Example of spurious regression} An example is shown below. Two random walks series $x_1t$ and $x_2t$ with a small drift ($\alpha=0.09$) are generated and
+\textbf{Example of spurious regression} \quad 
+An example is shown below. Two random walks series $x_1t$ and $x_2t$ with a small drift ($\alpha=0.09$) are generated and
 then regressed one on the other:
@@
 %==============================================================================
+$\blacksquare$
 
@@
 \newpage
-\texttt{Cointegration example}
+\textbf{Cointegration example} \quad
 
@@
 The method confirm that only two vectors are found and the third one given doesn't result in a I(0) process.
+
+$\blacksquare$
 
@@
 \begin{equation}\label{eq:vareq}
-\mathbf{B} = \mathbf{A} \mathbf{X} + \mathbf{E} \, , 
+\mathbf{Y} = \mathbf{A} \mathbf{X} + \mathbf{E} \, , 
 \end{equation}
@@
 \begin{alignat}{2}
-\mathbf{B}
+\mathbf{Y}
 &= \begin{bmatrix}
@@
 \begin{equation}\label{eq:vareq}
-\mathbf{B} = 
+\mathbf{Y} = 
 \mathbf{A} \mathbf{X} + 
@@
 %
-\noindent where $\mathbf{B}$ dimension is $((N-p)\times l)$, $\mathbf{A}$
+\noindent where $\mathbf{Y}$ dimension is $((N-p)\times l)$, $\mathbf{A}$
 dimension is $((N-p)\times(r+(p-1)l +1))$, $\mathbf{X}$ dimension is $((r+(p-1)l
@@
 \begin{alignat}{3}
-\mathbf{B}
+\mathbf{Y}
 &= \begin{bmatrix}
@@
 VECM fit to the first differences of the non stationary variables, but a lagged error-correction term is added to the relationship representing the long-run relationship.
-
-\subsection{Ordinary Least Squares method} \label{sec:OLS}
-
-The solution $\widehat{\mathbf{A}}$ to
-equation~(\ref{eq:vareq}) can be obtained by the ordinary least squares (OLS)
-method. $\widehat{\mathbf{X}}$ is the solution of the quadratic optimization problem
-\begin{equation*}
-\widehat{\mathbf{X}} = \underset{\mathbf{X}}{\text{Arg\;min}}
-\|\mathbf{A}\mathbf{X}-\mathbf{B}\|_2^2
-\end{equation*}
-\noindent for which the solution $\widehat{\mathbf{X}}$ is well-known:
-\begin{equation*}
-\label{eq:MP}
-\widehat{\mathbf{X}}=\mathbf{A}^{\!\!+}\,\mathbf{B}
-\end{equation*}
-\noindent where $\mathbf{A}^{\!\!+}$ is the Moore-Penrose pseudo-inverse
-which, when $\mathbf{A}$ is full rank, can be written as follows: 
-\begin{equation}
-\label{eq:pseudoinverse}
-\mathbf{A}^{\!\!+}= (\mathbf{A}^{\!\!\top} \mathbf{A})^{-1}\mathbf{A}^{\!\!\top} \, .
-\end{equation}
-when $\mathbf{A}$ is not full rank, i.e
-$rank(\mathbf{A})=k <  n \leq m$, $\mathbf{A}^\top \mathbf{A}$ is
-always singular and equation~(\ref{eq:pseudoinverse}) cannot be used.
-More generally, the pseudo-inverse is best computed using the compact
-singular value decomposition (SVD) of $\mathbf{A}$ which is:
-%\begin{equation}
-%    \label{eq:compactsvd}
-%    \underset{m \times n}{\mathbf{A}}=
-%    \underset{m \times k}{\mathbf{U_1}} \enskip
-%    \underset{k \times k}{\Sigma_1} \enskip
-%    \underset{k \times n}{\mathbf{V}_1^{\top}} \, .
-%\end{equation}
-\begin{equation*}
-    \label{eq:compactsvd}
-    \mathbf{A}=
-    \mathbf{U_1}
-    \boldsymbol \Sigma_1
-    \mathbf{V}_1^{\top} \, .
-\end{equation*}
-Pseudo-inverse can then be written as follows:
-\begin{equation*}
-\label{eq:pseudoinversesvd}
-\mathbf{A}^{\!\!+} = \mathbf{V}_1 \boldsymbol \Sigma_1^{-1} \mathbf{U}_1^\top \, .
-\end{equation*}

### 2015-02-27 src/4_OnlineLearning.tex
cambios de notación

--- 2015-02-23 src/4_OnlineLearning.tex
@@
 \end{equation}
-\noindent the choice of loss function here gives rise to several well-known learning algorithms such as regularised least squares and support vector machines. %verificar esto7776
+\noindent the choice of loss function here gives rise to several well-known learning algorithms such as regularised least squares and support vector machines. %verificar esto
 
@@
 
-One example of regularisation in linear models is ridge regression, which is a regularised least squares method.
-The least squares (LS) method is a well known way to solve a regression
-problem. 
-LS method consists of minimising the sum of squared errors:
+\subsection{Machine learning algorithms}
+In recent years many successful machine learning applications have been developed.
+Artificial neural networks (ANN) and Support Vector Machines (SVM) have been some of the most popularly used machine learning algorithm \cite{haykin1998}.
+Historically, SVMs emerged after ANN.
 
-%\begin{eqnarray*}
-%\label{eq:problem2}
-% J(\mathbf{\phi}) &=& \sum_{t=1}^N (f(\mathbf{x}_t)-y_t)^2 
-% = \sum_{t=1}^N (\mathbf{\phi}^\intercal {\mathbf{x}}_t-y_t)^2 
-% = \| \mathbf{X}\mathbf{\phi} - \mathbf{Y} \|_2^2 
-%\end{eqnarray*}
+The main characteristics of neural networks are that they have the ability to learn
+complex nonlinear input-output relationships, they use sequential training procedures 
+they adapt themselves to the data.
 
-% Aca seguir agregando slides.tex
+ANN is inspired by biological learning systems.
+ANN are organised into layers and have unidirectional connections between the layers, feed-forward networks  (FFN) are the most used.
+FFN consist of a series of layers.
+The first layer has a connection from the network input.
+Each subsequent layer has a connection from the previous layer.
+The final layer produces the network's output.
+Figure \ref{fig:ffn} shows a FFN with its different layers.
 
-\section{Batch and Online learning} \label{sec:onoffline}
+\begin{figure}[!h]
+  \centering
+  \includegraphics[width=0.8\textwidth]{img/ffn}
+  \caption{Feed-forward neural network}
+  \label{fig:ffn}
+\end{figure}
 
-Batch learning, also called statistical or offline learning, is a supervised machine learning framework.
-In batch learning, there is a data set available where the learner can build the internal model without any limits in accessing the data.
-There is time enough to carefully analyse the dataset, build large predictive models and combine them in a sophisticated way.
+Support Vector Machine (SVM) was introduced in 1992 by Vapnik and his coworkers
+\cite{boser1992}.
+In its original form, SVM is a training algorithm for linear classification.
+Only later it was used for regression, principal component analysis, novelty detection and also for non-linear case.
+However, unlike ANN, it is very well
+based on theory in statistical learning \cite{cortes1995}.  SVM tunes the capacity of 
+maximising the margin between the training patterns and the decision boundary. The
+solution is expressed as a linear combination of supporting patterns, which are the 
+of training patterns close to the decision boundary, called the support vectors.
 
-In contrast to offline learning, online learning has access to a sample only once.
-The goal is the same, predicting targets as accurate as possible.
+The major difference between SVM and ANN is in the error optimisation. In ANN, the aim of learning is to obtain a set of weight values which minimise the training error 
+minimum while training adjust the capacity of the machine.
+
+For nonlinear case, SVM mapped the data sets of input space into a higher dimensional feature space, which is linear and the large margin learning algorithm is then applied.
+However, the mapping can be implicitly done by kernel functions.
+In the high dimensional feature space, simpler and linear hyper plane classifiers that have maximal margin between the classes can be obtained.
+
+%\subsection{Kernel methods}
+%Kernel functions may be used to introduce a nonlinear separation
+%in input space despite the linear classifier in decision space
+
+\section{Online learning} \label{sec:onoffline}
+
+Online learning is a supervised machine learning framework that is useful when we have sequential access to a sample only once.
+This differs from the classical batch learning, where there is an entire data set available and the learner can build the internal model without any limits in accessing the data.
+In batch learning, there is time enough to carefully analyse the dataset, build large predictive models and combine them in a sophisticated way.
+
+In only learning the goal is the same as batch learning, predicting targets as accurate as possible.
 For example, stock market prediction can be seen as online learning.
 The algorithm makes a prediction of the stock, little time after the real stock price is available, this information can be incorporated to the learner to further improve the prediction accuracy.
-In general, there is very much data available in an online learning setup, the data set grows continuously.
+In general, there is too much data available in an online learning setup, the data set grows continuously.
 Offline learning has equal or superior accuracy compared to online learning when the same amount of data is used.
-
-\subsection{Online learning}
 
@@
 
+\subsection{Online Ridge Regression}
+
+Online RR is the online formulation of the regularised Least Squares method
+and is based on the following equivalent formulation of the RR optimal solution.
+
+\begin{eqnarray*}
+\mathbf{\phi}_*&=&(\mathbf{X}^\top \mathbf{X}+\lambda
+\mathbb{I})^{-1}\mathbf{X}^\top y \\
+\mathbf{\phi}_* &=& \displaystyle \big (\sum_{t=1}^m
+\mathbf{x}_t \mathbf{x}_t  ^\top + \lambda
+\mathbb{I}\big)^{-1} \sum_{t=1}^m y_t
+\mathbf{x}_t \\ \\
+\mathbf{\phi}_*&=& \mathbf{A}^{-1}\mathbf{b}
+\end{eqnarray*}
+
+\noindent where $\mathbf{A}=\mathbf{X}^\top \mathbf{X} +\lambda
+\mathbb{I}$ and $\mathbf{b}=\displaystyle  \sum_{t=1}^m y_t \mathbf{x}_t$.
+
+\begin{algorithm}[H]
+\begin{algorithmic}[1]
+\REQUIRE $\,$ \\
+$\{\mathbf{x}_1,\dots,\mathbf{x}_{m} \}$: $m$ input vectors \\
+$\{y_1,\dots,y_{m} \}$: $m$ targets \\
+$\lambda$: regularization parameter \\
+\ENSURE  $\,$ \\
+$\{f(\mathbf{x}_1),\dots,f(\mathbf{x}_{m}) \}$: model predictions \\
+\STATE Initialize $\mathbf{A}=\lambda \mathbb{I}$
+and $\mathbf{b}=0$
+\FOR { $t = 1$ to $m$ }
+	\STATE read new $\mathbf{x}_t$
+	\STATE output prediction $f(\mathbf{x}_t) =  \mathbf{b}^\top \mathbf{A}^{-1}\mathbf{x}_t$
+   	\STATE $\mathbf{A} = \mathbf{A} + \mathbf{x}_L
+        \mathbf{x}_L^\top -\mathbf{x}_1 \mathbf{x}_1^\top $
+   	\STATE Read new $y_t$
+    	\STATE $\mathbf{b} = \mathbf{b} + y_t \mathbf{x}_t$
+\ENDFOR
+
+\end{algorithmic}
+\caption{Ridge Regression}
+\label{alg:RR}
+\end{algorithm}
+
+\subsection{Aggregating algorithm regression}
+
 %%% Measure error

### 2015-02-27 src/4_OnlineLearning.tex
demos listas

@@
 
+\subsection{Ridge Regression}
+
+One example of regularisation in linear models is Ridge Regression (RR), which is a regularised least squares method.
+The least squares (LS) method is a well known way to solve a regression
+problem. 
+LS method consists of minimising the sum of squared errors:
+
+\begin{eqnarray*}
+\label{eq:lsm}
+ J(\mathbf{X}) =& \displaystyle \sum_{t=1}^N (f(\mathbf{x}_t)-y_t)^2 \\
+ =&\displaystyle \sum_{t=1}^N (\mathbf{\mathbf{X}}^\top {\mathbf{x}}_t-y_t)^2 \\
+ =& \| \mathbf{A}\mathbf{\mathbf{X}} - \mathbf{Y} \|_2^2 
+\end{eqnarray*}
+
+\noindent which is equivalent to find a solution to 
+\begin{equation}
+\label{eq:regproblem}
+\underset{m \times n}{\mathbf{A}} \enskip \underset{n \times
+l}{\mathbf{\mathbf{X}}} = \underset{m \times l}{\mathbf{Y}}
+\end{equation}
+
+The optimal solution for $\mathbf{\mathbf{X}}$ is:
+\begin{equation*}
+\mathbf{\mathbf{X}}= \mathbf{A}^+ \mathbf{Y}
+%(\mathbf{X}^\top\mathbf{X}
+%\mathbf{X})^{-1}\mathbf{X}^\top \mathbf{Y}
+\end{equation*}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\noindent where $\mathbf{A}^+$ is the Moore-Penrose pseudo-inverse
+which can be written as follows:
+
+\begin{equation}
+\label{eq:pseudoinverse}
+\mathbf{A}^+= (\mathbf{A}^\top \mathbf{A})^{-1}\mathbf{A}^\top \, .
+\end{equation}
+
+\textbf{Demo}\quad
+
+To solve equation~(\ref{eq:pseudoinverse}) is equivalent to solve the following
+optimisation problem:
+
+\begin{equation}
+\label{eq:RRproblem2}
+\underset{\mathbf{X}(\lambda)}{\text{min}} \quad \|
+\mathbf{C}\mathbf{\mathbf{X}} - \mathbf{F} \|_2^2
+\end{equation}
+
+\noindent where
+
+\begin{equation*}
+	\mathbf{C}=
+	\begin{bmatrix} \quad &\mathbf{A}& \quad \\ \hdashline \quad
+& \lambda\mathbb{I}&\quad  
+\end{bmatrix} 
+\quad \text{and} \quad
+	\mathbf{F}=\begin{bmatrix} \quad &\mathbf{Y}& \quad \\ \hdashline \quad
+& \mathbf{0}&\quad  \end{bmatrix} \, .
+\end{equation*}
+
+Applying equation~(\ref{eq:MP}) and considering that $\mathbf{C}^\top
+\mathbf{C} = \mathbf{A}^\top \mathbf{A} + \lambda \mathbb{I}$ and 
+$\mathbf{C}^\top \mathbf{F}=\mathbf{A}^\top \mathbf{Y} $ we have:
+
+\begin{eqnarray*}
+\mathbf{X}(\lambda)&=&(\mathbf{C}^\top
+\mathbf{C})^{-1}\mathbf{C}^\top \mathbf{F} \\
+&=& (\mathbf{A}^\top \mathbf{A} + \lambda \mathbb{I})^{-1} \mathbf{A}^\top \mathbf{Y}
+\end{eqnarray*}
+
+$\blacksquare$
+
+However, when $\mathbf{A}$ is not full rank, i.e
+$rank(\mathbf{A})=k <  n \leq m$, $\mathbf{A}^\top \mathbf{A}$ is
+always singular and equation~(\ref{eq:pseudoinverse}) cannot be used.
+More generally, the pseudo-inverse is best computed using the compact
+singular value decomposition (SVD) of $\mathbf{A}$:
+
+\begin{equation}
+    \label{eq:compactsvd}
+    \underset{m \times n}{\mathbf{A}}=
+    \underset{m \times k}{\mathbf{U_1}} \enskip
+    \underset{k \times k}{\mathbf{\Sigma_1}} \enskip
+    \underset{k \times n}{\mathbf{V_1}^\top}
+\end{equation}
+
+\noindent as follows
+
+\begin{equation}
+\label{eq:pseudoinversesvd}
+\mathbf{A}^+ = \mathbf{V_1\Sigma_1^{-1}U_1^\top}
+\end{equation}
+
+\textbf{Demo}\quad
+
+In the case matrix $\mathbf{A}$ is non singular, its solution is
+straight forward:
+
+\begin{equation}
+\label{OLSsolution}
+    \mathbf{\mathbf{X}}=\mathbf{A}^{-1}\mathbf{Y}
+\end{equation}
+
+Since the problem shown in equation~(\ref{eq:regproblem}) has not
+solution, the minimum norm given by equation~(\ref{OLSsolution}) is
+obtained by solving the equivalent problem:
+
+\begin{equation*}
+\label{eq:proyectorsol}
+\mathbf{X \hat{\mathbf{X}} = PY} 
+\end{equation*}
+
+\noindent where $\mathbf{P=U_1 U_1^\top}$ is the projection onto the
+Col($\mathbf{A}$).
+
+Since $\mathbf{V} = [\underset{(n \times k)}{\mathbf{V_1}} |
+\underset{(n \times k)}{\mathbf{V_2}}]$ and $\mathbf{V_1^\top V_2 =
+0}$ we can express $\mathbf{\hat{\mathbf{X}}} = \mathbf{V_1 \mathbf{X}_1 + V_2 \mathbf{X}_2}$
+with $\mathbf{\mathbf{X}_2=0}$ because $\mathbf{\hat{\mathbf{X}}}$ lives in the
+$\text{Row}(\mathbf{A})$ given by $\mathbf{V_1}$, so we have:
+
+\begin{eqnarray*}
+\mathbf{X \hat{\mathbf{X}}} &=& \mathbf{PB} \\
+\mathbf{U_1 \Sigma_1 V_1^\top \hat{\mathbf{X}}} &=& \mathbf{U_1 U_1^\top B} \\
+\mathbf{ V_1^\top \hat{\mathbf{X}}} &=&  \mathbf{\Sigma_1^{-1} U_1^\top B} \\ 
+\mathbf{ V_1^\top V_1 \mathbf{X}_1} &=& \mathbf{\Sigma_1^{-1}
+U_1^\top B} \\
+\mathbf{\mathbf{X}_1}&=& \mathbf{\Sigma_1^{-1} U_1^\top B}
+\end{eqnarray*}
+
+\noindent from this result we can obtain $\mathbf{\hat{\mathbf{X}}}$ and
+therefore the pseudo-inverse expression:
+
+\begin{eqnarray*}
+\mathbf{\hat{\mathbf{X}}} &=& \mathbf{V_1 \mathbf{X}_1} \\
+                &=& \mathbf{V_1 \Sigma_1^{-1} U_1^\top Y} \\
+\mathbf{X^+} &=& \mathbf{V_1 \Sigma_1^{-1} U_1^\top} \, .
+\end{eqnarray*}
+
+$\blacksquare$
+
+In order to avoid the singularity of the matrix $\mathbf{A}^\top \mathbf{A}$, a regularisation term is introduced:
+
+\begin{equation*}
+\label{eq:problem} 
+J(\mathbf{\mathbf{X}}) =  \| \mathbf{A}\mathbf{\mathbf{X}} - \mathbf{Y} \|_2^2  + \lambda
+ \| \mathbf{\mathbf{X}}\| ^2
+\end{equation*}
+
+\noindent which optimal solution $\mathbf{\mathbf{X}}_*$ is well known:
+
+\begin{equation*}
+\label{eq:optsolRR}
+\mathbf{\mathbf{X}}_*=(\mathbf{A}^\top \mathbf{A}+\lambda \mathbb{I})^{-1}\mathbf{A}^\top y \, ,
+\end{equation*}
+
+\textbf{Demo}\quad
+
+\begin{eqnarray*}
+J(\mathbf{\mathbf{X}}) &=&  \| \mathbf{A}\mathbf{\mathbf{X}} - \mathbf{Y} \|_2^2  + \lambda
+ \| \mathbf{\mathbf{X}}\| ^2 \\
+ &=&  \sum_{t=1}^N (\mathbf{\mathbf{X}}^\top {\bf a}_t-y_t)^2 + \lambda \sum_{i=1}^p \mathbf{X}_i^2 \\
+ &=& (\mathbf{X}^\top a_1-y_1)^2 + \dots + (\mathbf{X}^\top a_N-y_N)^2 + \lambda (\mathbf{X}_1^2+\dots+\mathbf{X}_p^2)
+\end{eqnarray*}
+\noindent taking derivatives
+ \begin{eqnarray*}
+ \frac{\partial J(\mathbf{\mathbf{X}})}{\partial \mathbf{X}_1}&=& 
+ 2(\mathbf{X}^\top \mathbf{a}_1-y_1)\mathbf{a}_{11} + \dots + 2(\mathbf{X}^\top \mathbf{a}_N-y_N)\mathbf{a}_{N1} + 2\lambda \mathbf{X}_1 \\
+ &=& 2\mathbf{a}_1^\top(\mathbf{A}\mathbf{X}-\mathbf{Y}) + 2\lambda\mathbf{X}_1\\
+& \vdots &\\
+  \frac{\partial J(\mathbf{\mathbf{X}})}{\partial \mathbf{X}_p}&=& 
+ 2\mathbf{a}_p^\top(\mathbf{A}\mathbf{X}-\mathbf{Y}) + 2\lambda\mathbf{X}_p 
+\end{eqnarray*}
+
+Then we have that:
+\begin{eqnarray*}
+\frac{\partial J(\mathbf{\mathbf{X}})}{\partial \mathbf{X}}&=& 
+ 2\mathbf{A}^\top(\mathbf{A}\mathbf{X}-\mathbf{Y}) + 2\lambda\mathbf{X}
+\end{eqnarray*}
+Since $\frac{\partial J(\mathbf{\mathbf{X}})}{\partial \mathbf{X}}=0$ we have:
+\begin{eqnarray*}
+2\mathbf{A}^\top(\mathbf{A}\mathbf{X}-\mathbf{Y}) + 2\lambda\mathbf{X}&=&0 \\
+\mathbf{A}^\top\mathbf{A}\mathbf{X} - \mathbf{A}^\top\mathbf{Y} + \lambda\mathbf{X} &=& 0\\
+(\mathbf{A}^\top\mathbf{A}+\lambda\mathbb{I})\mathbf{X} &=&  \mathbf{A}^\top\mathbf{Y} \\
+\mathbf{X} &=& (\mathbf{A}^\top\mathbf{A}+\lambda\mathbb{I})^{-1}  \mathbf{A}^\top\mathbf{Y}
+\end{eqnarray*}
+
+$\blacksquare$
+
+When $\mathbf{A}$ is orthonormal then $\mathbf{A}^\top \mathbf{A} =\mathbb{I}$ and there is a simple relation between the ridge estimator and the OLS estimator:
+\begin{eqnarray*}
+\mathbf{X}_* (\lambda) &=& (\mathbf{A}^\top \mathbf{A}+\lambda \mathbb{I})^{-1}\mathbf{A}^\top \mathbf{Y} \\
+ &=& (\mathbb{I} + \lambda \mathbb{I})^{-1} \mathbf{A}^\top \mathbf{Y} \\
+ &=&(1+\lambda)^{-1} \mathbb{I} \mathbf{A}^\top \mathbf{Y} \\
+ &=&(1+\lambda)^{-1} (\mathbf{A}^\top \mathbf{A})^{-1}\mathbf{A}^\top \mathbf{Y} \\
+ &=&(1+\lambda)^{-1} \mathbf{X}
+\end{eqnarray*}
+
 \subsection{Machine learning algorithms}
@@
 \begin{eqnarray*}
-\mathbf{\phi}_*&=&(\mathbf{X}^\top \mathbf{X}+\lambda
-\mathbb{I})^{-1}\mathbf{X}^\top y \\
-\mathbf{\phi}_* &=& \displaystyle \big (\sum_{t=1}^m
+\mathbf{\mathbf{X}}_*&=&(\mathbf{A}^\top \mathbf{A}+\lambda
+\mathbb{I})^{-1}\mathbf{A}^\top y \\
+\mathbf{\mathbf{X}}_* &=& \displaystyle \big (\sum_{t=1}^m
 \mathbf{x}_t \mathbf{x}_t  ^\top + \lambda
@@
 \mathbf{x}_t \\ \\
-\mathbf{\phi}_*&=& \mathbf{A}^{-1}\mathbf{b}
+\mathbf{\mathbf{X}}_*&=& \mathbf{A}^{-1}\mathbf{b}
 \end{eqnarray*}
 
-\noindent where $\mathbf{A}=\mathbf{X}^\top \mathbf{X} +\lambda
+\noindent where $\mathbf{A}=\mathbf{A}^\top \mathbf{A} +\lambda
 \mathbb{I}$ and $\mathbf{b}=\displaystyle  \sum_{t=1}^m y_t \mathbf{x}_t$.

### 2015-02-27 src/4_OnlineLearning.tex
demo bias-variance tradeoff

@@
 \mathrm{E}\Big[\big(y - \hat{f}(x)\big)^2\Big]
- & = \mathrm{Bias}\big[\hat{f}(x)\big]^2 + \mathrm{Var}\big[\hat{f}(x)\big] + \sigma^2
+ & = \mathrm{Bias}(\hat{f}(x))^2 + \mathrm{Var}(\hat{f}(x)) + \sigma^2
 \end{align}
@@
 \begin{align*}
- \mathrm{Bias}\big[\hat{f}(x)\big] &= \mathrm{E}\big[\hat{f}(x)\big] - f(x) \\
- \mathrm{Var}\big[\hat{f}(x)\big] &= \mathrm{E}\Big[ \big( \hat{f}(x) - \mathrm{E}[\hat{f}(x)] \big)^2 \Big] 
+ \mathrm{Bias}(\hat{f}(x)) &= \mathrm{E}\big[\hat{f}(x)\big] - f(x) \\
+ \mathrm{Var}(\hat{f}(x)) &= \mathrm{E}\Big[ \big( \hat{f}(x) - \mathrm{E}[\hat{f}(x)] \big)^2 \Big] 
 \end{align*}
+
+\textbf{Demo}\quad
+
+\begin{eqnarray*}
+E[(y-\hat{f}(\mathbf{x}))^2] & =&
+E[(y-f(\mathbf{x})+f(\mathbf{x})-\hat{f}(\mathbf{x}))^2] \\
+&=& E[(B-f(\mathbf{x}))^2] +
+E[(f(\mathbf{x})-\hat{f}(\mathbf{x}))^2] + \dots \\
+& &
+2\cancelto{0}{E[y-f(\mathbf{x})]}E[f(\mathbf{x})-\hat{f}(\mathbf{x})] \\
+&=& \sigma^2 + MSE(\hat{f}(\mathbf{x}))
+\end{eqnarray*}
+
+\noindent where
+
+\begin{eqnarray*}
+    MSE(\hat{f}(\mathbf{x})) &=&
+    E[f(\mathbf{x})-\hat{f}(\mathbf{x}))]^2 \\
+    &=& E[f(\mathbf{x})-E[\hat{f}(\mathbf{x})] +
+    E[\hat{f}(\mathbf{x})]-\hat{f}(\mathbf{x}))]^2 \\
+    &=&  E[(f(\mathbf{x})-E[\hat{f}(\mathbf{x})])^2] +
+      E[(\hat{f}(\mathbf{x})-E[\hat{f}(\mathbf{x})])^2] + \dots\\
+    & & E[(f(\mathbf{x})-E[\hat{f}(\mathbf{x})])] 
+    \cancelto{0}{E[(\hat{f}(\mathbf{x})-E[\hat{f}(\mathbf{x})])]} \\
+    &=& Bias(\hat{f}(\mathbf{x}))^2 + Var(\hat{f}(\mathbf{x}))
+\end{eqnarray*}
+
+$\blacksquare$
 
@@
 The optimal solution for $\mathbf{\mathbf{X}}$ is:
-\begin{equation*}
+\begin{equation}
+\label{eq:MP}
 \mathbf{\mathbf{X}}= \mathbf{A}^+ \mathbf{Y}
-%(\mathbf{X}^\top\mathbf{X}
-%\mathbf{X})^{-1}\mathbf{X}^\top \mathbf{Y}
-\end{equation*}
+\end{equation}
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \noindent where $\mathbf{A}^+$ is the Moore-Penrose pseudo-inverse
@@
 
-\begin{equation*}
-\label{eq:problem} 
+\begin{equation}
+\label{eq:RRproblem} 
 J(\mathbf{\mathbf{X}}) =  \| \mathbf{A}\mathbf{\mathbf{X}} - \mathbf{Y} \|_2^2  + \lambda
  \| \mathbf{\mathbf{X}}\| ^2
-\end{equation*}
+\end{equation}
 
@@
 
+\subsection{Selection of Lambda}
+
+The additional term $\lambda \|\mathbf{\mathbf{X}}\|_2^2$ in the optimisation
+problem shown in equation~(\ref{eq:RRproblem}) has two effects on the solution:
+shrinks the coefficients towards zero and improves the conditioning of the
+problem.
+
 When $\mathbf{A}$ is orthonormal then $\mathbf{A}^\top \mathbf{A} =\mathbb{I}$ and there is a simple relation between the ridge estimator and the OLS estimator:
@@
 \end{eqnarray*}
+
+Figure~\ref{fig:shrinks} shows a visual example of the shrinking of the
+coefficients:
+
+\begin{figure}[h!]
+\includegraphics[width=0.5\linewidth]{img/shrinks}
+\caption{Shrink of regression coefficients}
+\label{fig:shrinks}
+\end{figure}
+
+On the other hand, the effect of adding the term $\lambda \mathbb{I}$
+to the matrix $\mathbf{A}^\top \mathbf{A}$
+(equation~(\ref{eq:optsolRR})) improves its condition number since it
+increases its diagonal values when $\lambda > 0 $.  The matrix
+$\mathbf{A}^\top \mathbf{A}$ is symmetrical ($(\mathbf{A}^\top
+\mathbf{A})^\top = \mathbf{A}^\top \mathbf{A}$) and therefore
+diagonalizable.  If we know the eigenvalue decomposition of $\mathbf{A}
+= \mathbf{U\Sigma V^\top}$, then:
+
+\begin{eqnarray*}
+\mathbf{A}^\top \mathbf{A}+\lambda \mathbb{I}&=&\mathbf{V\Sigma^2
+V}^\top + \lambda \mathbf{V} \mathbf{V}^\top\\ &=&\mathbf{V}
+(\mathbf{\Sigma}^2+\lambda\mathbb{I}) \mathbf{V}^\top \, ,
+\end{eqnarray*}
+
+\noindent where
+
+\begin{equation*}
+\mathbf{\Sigma}^2+\lambda\mathbb{I}=
+\begin{bmatrix}
+\sigma^2_1 + \lambda & \, & \, \\
+\, & \sigma^2_2 +\lambda & \, \\
+\, & \, & \ddots & \, \\
+\, & \, & \, & \sigma^2_n +\lambda \, .
+\end{bmatrix}
+\end{equation*}
+
+where $\sigma_1 \geq \sigma_2 \geq \dots \geq \sigma_n$.
+
+Since the condition number of a matrix $\mathbf{A}$ is defined as:
+
+\begin{equation*}
+	\kappa = \|\mathbf{A}\| \|\mathbf{A}^{-1}\|
+\end{equation*}
+
+If matrix $\mathbf{A}$ is non-singular, its condition number can be
+expressed in terms of its singular values. The effect of adding the
+regularization term affects the condition number as follows:
+
+\begin{eqnarray*}
+\kappa_{ols} &=& \|\mathbf{A}\| \|\mathbf{A}^{-1}\|=\frac{\sigma_1}{\sigma_n} \\
+\kappa_{ridge} &=& \|\mathbf{A}^\intercal \mathbf{A} + \lambda \mathbb{I}\| 
+\|(\mathbf{A}^\intercal \mathbf{A} + \lambda \mathbb{I})^{-1}\|=\frac{\sigma_1+\lambda}{\sigma_n + \lambda} \,
+\end{eqnarray*}
+
+It is easy to see that the term $\lambda$ improves the condition number:
+
+\begin{equation*}
+        \frac{\sigma_1+\lambda}{\sigma_n + \lambda} <
+        \frac{\sigma_1}{\sigma_n} \,  \qquad \forall \quad \lambda > 0
+\end{equation*}
+
+However, $\lambda$ cannot be too large. Tipically $\lambda$ is small and its
+magnitude depends on the matrix $\mathbf{A}$.
+
+For rank deficient matrices we know that $\text{det}(\mathbf{A A^\top})=0$, adding the
+term $\lambda \mathbb{I}$ we have that $\text{det}(\mathbf{A A^\top}+\lambda \mathbb{I}) =
+p(\lambda)$ where $p(\lambda)$ is a polynomial of degree $n$ ($\mathbf{A}$ is
+$m \times n$). The zeros of $p(\lambda)$ are discretes, so it can be represented
+as:
+
+\[
+p(\lambda) =
+\lambda(\lambda-\lambda_1)^{n_1}(\lambda-\lambda_2)^{n_2}\dots(\lambda-\lambda_s)^{n_s}
+\]
+
+\noindent where $n_1 + n_2 + \dots + n_s = n$.
+
+This means that $\lambda$ must be small in order to ensure that $p(\lambda)$
+does not vanish.
+
+%\subsection{The bias-variance tradeoff}
+%
+%To determine parameter $\lambda$ is crucial for ridge regression since it could
+%reduce the expected prediction error by reducing variance (bias-variance trafeoff) considering a biased
+%estimator. It is known that the prediction error is
+%obtained as:
+%
+%\begin{equation}
+% \label{eq:prederror}
+%E[(B-\hat{f}(\mathbf{X}))^2] = \sigma^2 + Bias(\hat{f}(\mathbf{X}))^2 + Var(\hat{f}(\mathbf{X}))
+%\end{equation}
+%
+%\noindent where $\hat{f}(\mathbf{X})=\mathbf{AX}(\lambda)$ and
+%$f(\mathbf{X})=\mathbf{AX}$.
+%
+%
+%
+%The bias of OLS is:
+%
+%\begin{eqnarray*}
+%Bias(\hat{\mathbf{X}}) &=& E[\hat{\mathbf{X}}] - \mathbf{X} \\
+%&=& E[ (\mathbf{A}^\top \mathbf{A})^{-1}\mathbf{A}^\top \mathbf{B}] - \mathbf{X} \\
+%&=& E[ (\mathbf{A}^\top \mathbf{A})^{-1}\mathbf{A}^\top (\mathbf{AX})] - \mathbf{X}  \\
+%&=& \mathbf{X}  - \mathbf{X}  \\
+%&=&  0
+%\end{eqnarray*}
+%
+%
+%The bias of ridge regression when $\mathbf{A A^\top}$ is non-singular
+%can be obtained expressing ridge regression solution
+%$\mathbf{\lambda}$ in terms of OLS solution $\hat{\mathbf{X}}$:
+%
+%\begin{eqnarray*}
+%\mathbf{X}(\lambda) &=&( \mathbf{A}^\top \mathbf{A} + \lambda \mathbb{I})^{-1}\mathbf{B} \\
+%&=& (\mathbb{I} + \lambda (\mathbf{A}^\top \mathbf{A})^{-1})^{-1} (\mathbf{A}^\top \mathbf{A})^{-1}\mathbf{A}^\top \mathbf{B} \\
+%&=&  (\mathbb{I} + \lambda (\mathbf{A}^\top \mathbf{A})^{-1})^{-1}  \hat{\mathbf{X}} \\
+%&=& \mathbf{W} \hat{\mathbf{X}} 
+%\end{eqnarray*}
+%
+%\noindent where $\mathbf{W}  = (\mathbb{I} + \lambda (\mathbf{A}^\top
+%\mathbf{A})^{-1})^{-1}  $ it is defined for simplicity. Ridge
+%regression bias is then obtained as:
+%
+%\begin{eqnarray*}
+%Bias(\mathbf{X}(\lambda)) &=& E[\mathbf{X}(\lambda)] - \mathbf{X} \\
+%&=& E[\mathbf{W}\hat{\mathbf{X}}] - \mathbf{X} \\
+%&=&  \mathbf{W} \mathbf{X} - \mathbf{X} \neq 0 
+%\end{eqnarray*}
+%
+%
+%
+%The variance of OLS is:
+%
+%\begin{equation*}
+%Var(\hat{\mathbf{X}}) = \sigma^2 (\mathbf{A}^\top \mathbf{A} )^{-1}
+%\end{equation*}
+%
+%\noindent and the variance of ridge regression is:
+%
+%\begin{eqnarray*}
+%Var(\mathbf{X}(\lambda)) &=& Var(\mathbf{W}\hat{\mathbf{X}}) \\ 
+%&=& \mathbf{W}Var(\hat{\mathbf{X}})\mathbf{W}^\top \\
+%&=& \sigma^2 \mathbf{W}(\mathbf{A}^\top \mathbf{A} )^{-1}\mathbf{W}^\top
+%\end{eqnarray*}
+%
+%
+%\begin{figure}[h!]
+%\includegraphics[width=0.8\linewidth]{plots/biasvariance}
+%\caption{Bias-variance tradeoff}
+%\label{fig:biasvariance}
+%\end{figure}
+%
+%The figure~\ref{fig:biasvariance} shows the bias-variance tradeoff
+%given by equation~(\ref{eq:prederror}). Ridge regression shows an
+%increasing bias and a decreasing variance. Despite OLS has zero bias,
+%its variance is greater than ridge for small values of $\lambda$.
+%It can be shown that, in terms of prediction error, ridge (black line)
+%is lower than OLS (dotted line)~\cite{hoerl+kennard1970}.
+%
+%
+%
 

### 2015-03-02 src/4_OnlineLearning.tex
Bias variance on lambda ready

@@
 
-\subsection{The bias-variance tradeoff}
+\subsection{The bias-variance tradeoff} \label{sec:biasvar}
 
@@
 \begin{align}
+\label{eq:bvtrade}
 \mathrm{E}\Big[\big(y - \hat{f}(x)\big)^2\Big]
@@
 
-\subsection{Selection of Lambda}
+\subsection{The Lambda parameter}
 
@@
 
-%\subsection{The bias-variance tradeoff}
-%
-%To determine parameter $\lambda$ is crucial for ridge regression since it could
-%reduce the expected prediction error by reducing variance (bias-variance trafeoff) considering a biased
-%estimator. It is known that the prediction error is
-%obtained as:
-%
-%\begin{equation}
-% \label{eq:prederror}
-%E[(B-\hat{f}(\mathbf{X}))^2] = \sigma^2 + Bias(\hat{f}(\mathbf{X}))^2 + Var(\hat{f}(\mathbf{X}))
-%\end{equation}
-%
-%\noindent where $\hat{f}(\mathbf{X})=\mathbf{AX}(\lambda)$ and
-%$f(\mathbf{X})=\mathbf{AX}$.
-%
-%
-%
-%The bias of OLS is:
-%
-%\begin{eqnarray*}
-%Bias(\hat{\mathbf{X}}) &=& E[\hat{\mathbf{X}}] - \mathbf{X} \\
-%&=& E[ (\mathbf{A}^\top \mathbf{A})^{-1}\mathbf{A}^\top \mathbf{B}] - \mathbf{X} \\
-%&=& E[ (\mathbf{A}^\top \mathbf{A})^{-1}\mathbf{A}^\top (\mathbf{AX})] - \mathbf{X}  \\
-%&=& \mathbf{X}  - \mathbf{X}  \\
-%&=&  0
-%\end{eqnarray*}
-%
-%
-%The bias of ridge regression when $\mathbf{A A^\top}$ is non-singular
-%can be obtained expressing ridge regression solution
-%$\mathbf{\lambda}$ in terms of OLS solution $\hat{\mathbf{X}}$:
-%
-%\begin{eqnarray*}
-%\mathbf{X}(\lambda) &=&( \mathbf{A}^\top \mathbf{A} + \lambda \mathbb{I})^{-1}\mathbf{B} \\
-%&=& (\mathbb{I} + \lambda (\mathbf{A}^\top \mathbf{A})^{-1})^{-1} (\mathbf{A}^\top \mathbf{A})^{-1}\mathbf{A}^\top \mathbf{B} \\
-%&=&  (\mathbb{I} + \lambda (\mathbf{A}^\top \mathbf{A})^{-1})^{-1}  \hat{\mathbf{X}} \\
-%&=& \mathbf{W} \hat{\mathbf{X}} 
-%\end{eqnarray*}
-%
-%\noindent where $\mathbf{W}  = (\mathbb{I} + \lambda (\mathbf{A}^\top
-%\mathbf{A})^{-1})^{-1}  $ it is defined for simplicity. Ridge
-%regression bias is then obtained as:
-%
-%\begin{eqnarray*}
-%Bias(\mathbf{X}(\lambda)) &=& E[\mathbf{X}(\lambda)] - \mathbf{X} \\
-%&=& E[\mathbf{W}\hat{\mathbf{X}}] - \mathbf{X} \\
-%&=&  \mathbf{W} \mathbf{X} - \mathbf{X} \neq 0 
-%\end{eqnarray*}
-%
-%
-%
-%The variance of OLS is:
-%
-%\begin{equation*}
-%Var(\hat{\mathbf{X}}) = \sigma^2 (\mathbf{A}^\top \mathbf{A} )^{-1}
-%\end{equation*}
-%
-%\noindent and the variance of ridge regression is:
-%
-%\begin{eqnarray*}
-%Var(\mathbf{X}(\lambda)) &=& Var(\mathbf{W}\hat{\mathbf{X}}) \\ 
-%&=& \mathbf{W}Var(\hat{\mathbf{X}})\mathbf{W}^\top \\
-%&=& \sigma^2 \mathbf{W}(\mathbf{A}^\top \mathbf{A} )^{-1}\mathbf{W}^\top
-%\end{eqnarray*}
-%
-%
-%\begin{figure}[h!]
-%\includegraphics[width=0.8\linewidth]{plots/biasvariance}
-%\caption{Bias-variance tradeoff}
-%\label{fig:biasvariance}
-%\end{figure}
-%
-%The figure~\ref{fig:biasvariance} shows the bias-variance tradeoff
-%given by equation~(\ref{eq:prederror}). Ridge regression shows an
-%increasing bias and a decreasing variance. Despite OLS has zero bias,
-%its variance is greater than ridge for small values of $\lambda$.
-%It can be shown that, in terms of prediction error, ridge (black line)
-%is lower than OLS (dotted line)~\cite{hoerl+kennard1970}.
-%
-%
-%
+\subsection{Selection of Lambda}
+
+One of the ways to determine parameter $\lambda$ is using the bias-variance tradeoff (see section \ref{sec:biasvar}). This parameter is crucial for ridge regression since it could
+reduce the expected prediction error by reducing variance, considering a biased
+estimator. 
+It is known that the prediction error can be express as a decomposition between bias and variance. The prediction error is
+obtained as:
+
+\begin{equation}
+ \label{eq:prederror}
+E[(Y-\hat{f}(\mathbf{X}))^2] = \sigma^2 + Bias(\hat{f}(\mathbf{X}))^2 + Var(\hat{f}(\mathbf{X}))
+\end{equation}
+
+\noindent where $\hat{f}(\mathbf{X})=\mathbf{AX}(\lambda)$ for RR and
+$\hat{f}(\mathbf{X})=\mathbf{AX}$ in the case of OLS.
+
+For ridge regression we will use the following notation: $\hat{f}(\mathbf{X})=\mathbf{AX}(\lambda)$ and
+for OLS $f(\mathbf{X})=\mathbf{AX}$. The solution of OLS is well known $\hat{\mathbf{X}}=(\mathbf{A}^\top \mathbf{A})^{-1}\mathbf{A}^\top \mathbf{Y}$, and its bias is:
+
+\begin{eqnarray*}
+Bias(\hat{\mathbf{X}}) &=& E[\hat{\mathbf{X}}] - \mathbf{X} \\
+&=& E[ (\mathbf{A}^\top \mathbf{A})^{-1}\mathbf{A}^\top \mathbf{Y}] - \mathbf{X} \\
+&=& E[ (\mathbf{A}^\top \mathbf{A})^{-1}\mathbf{A}^\top (\mathbf{AX})] - \mathbf{X}  \\
+&=& \mathbf{X}  - \mathbf{X}  \\
+&=&  0
+\end{eqnarray*}
+
+The bias of ridge regression when $\mathbf{A A^\top}$ is non-singular
+can be obtained expressing ridge regression solution
+$\mathbf{\lambda}$ in terms of OLS solution $\hat{\mathbf{X}}$:
+
+\begin{eqnarray*}
+\mathbf{X}(\lambda) &=&( \mathbf{A}^\top \mathbf{A} + \lambda \mathbb{I})^{-1}\mathbf{Y} \\
+&=& (\mathbb{I} + \lambda (\mathbf{A}^\top \mathbf{A})^{-1})^{-1} (\mathbf{A}^\top \mathbf{A})^{-1}\mathbf{A}^\top \mathbf{Y} \\
+&=&  (\mathbb{I} + \lambda (\mathbf{A}^\top \mathbf{A})^{-1})^{-1}  \hat{\mathbf{X}} \\
+&=& \mathbf{W} \hat{\mathbf{X}} 
+\end{eqnarray*}
+
+\noindent where $\mathbf{W}  = (\mathbb{I} + \lambda (\mathbf{A}^\top
+\mathbf{A})^{-1})^{-1}  $ it is defined for simplicity. Ridge
+regression bias is then obtained as:
+
+\begin{eqnarray*}
+Bias(\mathbf{X}(\lambda)) &=& E[\mathbf{X}(\lambda)] - \mathbf{X} \\
+&=& E[\mathbf{W}\hat{\mathbf{X}}] - \mathbf{X} \\
+&=&  \mathbf{W} \mathbf{X} - \mathbf{X} \neq 0 
+\end{eqnarray*}
+
+\begin{figure}[h!]
+\includegraphics[width=\linewidth]{img/biasvariance}
+\caption{Bias-variance tradeoff depending on lambda}
+\label{fig:biasvariance}
+\end{figure}
+
+The variance of OLS is:
+
+\begin{equation*}
+Var(\hat{\mathbf{X}}) = \sigma^2 (\mathbf{A}^\top \mathbf{A} )^{-1}
+\end{equation*}
+
+\noindent and the variance of ridge regression is:
+
+\begin{eqnarray*}
+Var(\mathbf{X}(\lambda)) &=& Var(\mathbf{W}\hat{\mathbf{X}}) \\ 
+&=& \mathbf{W}Var(\hat{\mathbf{X}})\mathbf{W}^\top \\
+&=& \sigma^2 \mathbf{W}(\mathbf{A}^\top \mathbf{A} )^{-1}\mathbf{W}^\top
+\end{eqnarray*}
+
+The figure~\ref{fig:biasvariance} shows the bias-variance tradeoff
+given by equation~(\ref{eq:prederror}). Ridge regression shows an
+increasing bias and a decreasing variance. Despite OLS has zero bias,
+its variance is greater than ridge for small values of $\lambda$.
+It can be shown that, in terms of prediction error, ridge (black line)
+is lower than OLS (dotted line)~\cite{hoerl1970}.
 

### 2015-03-02 src/4_OnlineLearning.tex
AAR

@@
 
+Since 
+\begin{equation}
+\label{eq:notation}
+	\mathbf{A} = 
+\left[
+  \begin{tabular}{c>{$}c<{$}c}
+    --- & \mathbf{a}^{\top}_{1} & ---\\
+    --- & \mathbf{a}^{\top}_{2} & ---\\
+    & \vdots & \\
+    --- & \mathbf{a}^{\top}_{m} & ---
+  \end{tabular}
+\right]
+\quad \text{and} \quad
+\mathbf{B} =
+\left[
+  \begin{tabular}{c>{$}c<{$}c}
+    --- & \mathbf{b}_{1} & ---\\
+    --- & \mathbf{b}_{2} & ---\\
+    & \vdots & \\
+    --- & \mathbf{b}_{m} & ---
+  \end{tabular}
+\right]
+\end{equation}
+
+Equation~(\ref{eq:optsolRR}) can also be written as:
 \begin{eqnarray*}
-\mathbf{\mathbf{X}}_*&=&(\mathbf{A}^\top \mathbf{A}+\lambda
-\mathbb{I})^{-1}\mathbf{A}^\top y \\
-\mathbf{\mathbf{X}}_* &=& \displaystyle \big (\sum_{t=1}^m
-\mathbf{x}_t \mathbf{x}_t  ^\top + \lambda
-\mathbb{I}\big)^{-1} \sum_{t=1}^m y_t
-\mathbf{x}_t \\ \\
-\mathbf{\mathbf{X}}_*&=& \mathbf{A}^{-1}\mathbf{b}
+\label{eq:RReapand}
+\mathbf{\mathbf{X}}_{\text{ridge}}&=&(\mathbf{A}^\top \mathbf{A}+ \lambda
+\mathbb{I})^{-1}\mathbf{A}^\top \mathbf{Y} \\
+&=& \displaystyle \big (\sum_{t=1}^m
+\mathbf{a}_t \mathbf{a}_t  ^\top + \lambda \mathbb{I}\big )^{-1}
+\sum_{t=1}^m \mathbf{a}_t \mathbf{y}_t \, .
 \end{eqnarray*}
 
-\noindent where $\mathbf{A}=\mathbf{A}^\top \mathbf{A} +\lambda
-\mathbb{I}$ and $\mathbf{b}=\displaystyle  \sum_{t=1}^m y_t \mathbf{x}_t$.
+Lets define $\displaystyle\mathbf{S}= \sum_{t=1}^m \mathbf{a}_t
+\mathbf{a}_t  ^\top + \lambda \mathbb{I} $ and $\mathbf{W}=
+\displaystyle\sum_{t=1}^m \mathbf{a}_t \mathbf{y}_t$, so the
+algorithm~\ref{alg:RR} shows the iterative formulation:
 
@@
 \REQUIRE $\,$ \\
-$\{\mathbf{x}_1,\dots,\mathbf{x}_{m} \}$: $m$ input vectors \\
-$\{y_1,\dots,y_{m} \}$: $m$ targets \\
+$\{\mathbf{a}_1,\dots,\mathbf{a}_{m} \}$: $m$ input vectors \\
+$\{\mathbf{y}_1,\dots,\mathbf{y}_{m} \}$: $m$ target vectors \\
 $\lambda$: regularization parameter \\
 \ENSURE  $\,$ \\
-$\{f(\mathbf{x}_1),\dots,f(\mathbf{x}_{m}) \}$: model predictions \\
-\STATE Initialize $\mathbf{A}=\lambda \mathbb{I}$
-and $\mathbf{b}=0$
+$\{f(\mathbf{a}_1),\dots,f(\mathbf{a}_{m}) \}$: model predictions \\
+\STATE Initialize $\mathbf{S}=\lambda \mathbb{I}$
+and $\mathbf{W}=0$
 \FOR { $t = 1$ to $m$ }
-	\STATE read new $\mathbf{x}_t$
-	\STATE output prediction $f(\mathbf{x}_t) =  \mathbf{b}^\top \mathbf{A}^{-1}\mathbf{x}_t$
-   	\STATE $\mathbf{A} = \mathbf{A} + \mathbf{x}_L
-        \mathbf{x}_L^\top -\mathbf{x}_1 \mathbf{x}_1^\top $
+	\STATE read new $\mathbf{a}_t$
+	\STATE $\mathbf{X}=\mathbf{S}^{-1}\mathbf{W}$
+	\STATE output prediction $f(\mathbf{a}_t) = \mathbf{X}^\top \mathbf{a}_t$
+   	\STATE $\mathbf{S} = \mathbf{S} + \mathbf{a}_t \mathbf{a}_t^\top$
    	\STATE Read new $y_t$
-    	\STATE $\mathbf{b} = \mathbf{b} + y_t \mathbf{x}_t$
+    	\STATE $\mathbf{W} = \mathbf{W} + \mathbf{a}_t \mathbf{y}_t$
 \ENDFOR
-
 \end{algorithmic}
-\caption{Ridge Regression}
+\caption{Online Ridge Regression}
 \label{alg:RR}
@@
 
-\subsection{Aggregating algorithm regression}
+\section{The Aggregating Algorithm for Regression}
+
+The AAR, proposed by~\cite{vovk2001}, is an application of the aggregating
+algorithm to the problem of regression. The idea is introduce the new input
+vector $\mathbf{x}_{m+1}$ to solve the model parameters:
+
+\begin{equation}
+\label{eq:AARexpand}
+\mathbf{X}_{aar} = \displaystyle \big (\sum_{t=1}^{m+1}
+\mathbf{a}_t \mathbf{a}_t  ^\intercal + \gamma \mathbb{I}\big )^{-1}
+\sum_{t=1}^m \mathbf{a}_t \mathbf{b}_t \, .
+\end{equation}
+
+If we define $\displaystyle\mathbf{S}= \sum_{t=1}^{m+1} \mathbf{a}_t
+\mathbf{a}_t  ^\intercal + \gamma \mathbb{I} $ and $\mathbf{W}=
+\displaystyle\sum_{t=1}^m \mathbf{a}_t \mathbf{b}_t$, the
+algorithm~\ref{alg:AAR} is slightly different to the algorithm~\ref{alg:RR}, 
+which updated matrix $\mathbf{S}$ before making the prediction:
+
+\begin{algorithm}[ht]
+\begin{algorithmic}[1]
+\REQUIRE $\,$ \\
+$\{\mathbf{a}_1,\dots,\mathbf{a}_{m} \}$: $m$ input vectors \\
+$\{\mathbf{b}_1,\dots,\mathbf{b}_{m} \}$: $m$ target vectors \\
+$\lambda$: regularization parameter \\
+\ENSURE  $\,$ \\
+$\{f(\mathbf{a}_1),\dots,f(\mathbf{a}_{m}) \}$: model predictions \\
+\STATE Initialize $\mathbf{S}=\lambda \mathbb{I}$
+and $\mathbf{W}=0$
+\FOR { $t = 1$ to $m$ }
+	\STATE read new $\mathbf{a}_t$
+   	\STATE $\mathbf{S} = \mathbf{S} + \mathbf{a}_t \mathbf{a}_t^\intercal$
+	\STATE $\mathbf{X}=\mathbf{S}^{-1}\mathbf{W}$
+	\STATE output prediction $f(\mathbf{a}_t) = \mathbf{X}^\intercal \mathbf{a}_t$
+   	\STATE Read new $\mathbf{y}_t$
+    	\STATE $\mathbf{W} = \mathbf{W} + \mathbf{a}_t \mathbf{b}_t$
+\ENDFOR
+\end{algorithmic}
+\caption{{\em The aggregating algorithm for regression}}
+\label{alg:AAR}
+\end{algorithm}
 

### 2015-03-02 src/4_OnlineLearning.tex
efficient computation

@@
 %in input space despite the linear classifier in decision space
-
+\newpage
 \section{Online learning} \label{sec:onoffline}
@@
 \mathbf{a}_t \mathbf{a}_t  ^\intercal + \gamma \mathbb{I}\big )^{-1}
-\sum_{t=1}^m \mathbf{a}_t \mathbf{b}_t \, .
+\sum_{t=1}^m \mathbf{a}_t \mathbf{y}_t \, .
 \end{equation}
@@
 \mathbf{a}_t  ^\intercal + \gamma \mathbb{I} $ and $\mathbf{W}=
-\displaystyle\sum_{t=1}^m \mathbf{a}_t \mathbf{b}_t$, the
+\displaystyle\sum_{t=1}^m \mathbf{a}_t \mathbf{y}_t$, the
 algorithm~\ref{alg:AAR} is slightly different to the algorithm~\ref{alg:RR}, 
@@
 $\{\mathbf{a}_1,\dots,\mathbf{a}_{m} \}$: $m$ input vectors \\
-$\{\mathbf{b}_1,\dots,\mathbf{b}_{m} \}$: $m$ target vectors \\
-$\lambda$: regularization parameter \\
+$\{\mathbf{y}_1,\dots,\mathbf{y}_{m} \}$: $m$ target vectors \\
+$\lambda$: regularisation parameter \\
 \ENSURE  $\,$ \\
@@
    	\STATE Read new $\mathbf{y}_t$
-    	\STATE $\mathbf{W} = \mathbf{W} + \mathbf{a}_t \mathbf{b}_t$
+    	\STATE $\mathbf{W} = \mathbf{W} + \mathbf{a}_t \mathbf{y}_t$
 \ENDFOR
@@
 
+\section{Efficient computation}
+
+In order to solve OLS in a efficient way, Coleman and
+Sun~\cite{coleman+sun2010} presented an iterative algorithm which
+uses $\mathbf{X}(\lambda)$ to approximate $\mathbf{X}(0)$.
+
+Using the compact SVD (shown in equation~(\ref{eq:compactsvd}))
+$\mathbf{X}(\lambda)$ is expressed as follows:
+
+\begin{eqnarray}
+\label{eq:optsolRRsvd}
+\mathbf{X}(\lambda) & = & (\mathbf{A}^\top \mathbf{A}+ \lambda
+\mathbb{I})^{-1}\mathbf{A}^\top \mathbf{Y} \nonumber \\
+& = &\mathbf{V}_1(\Sigma_1^2+\lambda \mathbb{I})^{-1}\mathbf{\Sigma_1
+U_1^\top Y}
+\end{eqnarray}
+
+\noindent where is easy to see that $\mathbf{X}(\lambda) \rightarrow
+\mathbf{\hat{X}}=\mathbf{V}_1 \mathbf{\Sigma}_1^{-1}\mathbf{U}_1^\top
+\mathbf{Y}$ as $\lambda \rightarrow 0$.
+
+The method consists in obtaining $\mathbf{X}(\lambda)$ and then refine
+by adding more terms of its Taylor expansion to approximate
+$\mathbf{X}(0) = \mathbf{\hat{X}}$. The Taylor expansion about $\lambda_0$ is:
+
+\begin{equation}
+\label{eq:taylor}
+    \mathbf{X}(\lambda)=\mathbf{X}(\lambda_0) + \sum_{k=1}^\infty
+    \mathbf{s}_k(\lambda-\lambda_0)^{k}
+\end{equation}
+
+\noindent where $\mathbf{s}_k=\frac{1}{k!}\mathbf{X}(\lambda)^{(k)}$
+and $\mathbf{X}(\lambda)^{(k)}$ is obtained by taking differences 
+$\frac{\partial}{\partial \lambda}$ of:
+
+\begin{eqnarray*}
+(\mathbf{A}^\top \mathbf{A}+ \lambda\mathbb{I}) \mathbf{X}(\lambda) & = & \mathbf{A}^\top \mathbf{Y}\\
+(\mathbf{A}^\top \mathbf{A}+ \lambda\mathbb{I}) \mathbf{X}(\lambda)^{(1)} + \mathbf{X}(\lambda)& = & 0 \\
+\mathbf{X}(\lambda)^{(1)}  &=& -(\mathbf{A}^\top \mathbf{A}+ \lambda\mathbb{I}) ^{-1} \mathbf{X}(\lambda) \\
+\mathbf{X}(\lambda)^{(2)}  &=& -2(\mathbf{A}^\top \mathbf{A}+ \lambda\mathbb{I}) ^{-1} \mathbf{X}(\lambda)^{(1)} \\
+& \vdots & \\
+\mathbf{X}(\lambda)^{(k)}  &=& -k(\mathbf{A}^\top \mathbf{A}+ \lambda\mathbb{I}) ^{-1} \mathbf{X}(\lambda)^{(k-1)} 
+\end{eqnarray*}
+
+\noindent for $\lambda=0$ we have:
+
+\begin{equation}
+\label{eq:taylor}
+    \mathbf{X}(0)=\mathbf{X}(\lambda_0) + \sum_{k=1}^\infty
+     (-1)^k \mathbf{s}_k \lambda_0^k
+\end{equation}
+
+\noindent therefore in order to ensure convergence, we can see that$\lambda_0$
+selection cannot be large.
+
+The algorithm for computing $\mathbf{X}(0)$ is the following:
+
+\begin{algorithm}[H]
+\begin{algorithmic}[1]
+\REQUIRE $\,$ \\
+$\mathbf{A}$: design matrix \\
+$\mathbf{Y}$: response matrix \\
+$\lambda$: rank deficient parameter \\
+\ENSURE  $\,$ \\
+$\mathbf{X}$: parameters \\
+\STATE $\mathbf{M}=\mathbf{A^\top A}$ \\
+\STATE Initialize $\mathbf{Q R}=\mathbf{M}$ \\
+\STATE $\mathbf{X} = \mathbf{R^{-1}Q^\top Y}$ \\
+\STATE $\mathbf{s} = \mathbf{X}$ \\
+\FOR { $i = 1,2,3,\dots$ }
+	\STATE $\mathbf{s} =
+        -(\mathbf{M}+\lambda\mathbb{I})^{-1}\mathbf{s}$\\
+	\STATE $\mathbf{X}=\mathbf{X} + (-1)^i {\color{red}\mathbf{s}
+        \lambda^i}$
+\ENDFOR
+\end{algorithmic}
+\caption{Algorithm for handling rank deficient matrices}
+\label{alg:coleman}
+\end{algorithm}
+
+The algorithm~\ref{alg:coleman} solves equation~(\ref{eq:taylor}). However, this
+version is unstable since typically $\|\mathbf{s}\|$ is very large and
+$\lambda^i$ is very small ($\lambda$ is small).
+
+The following algorithm shows a more stable version of
+algorithm~\ref{alg:coleman}.
+
+\begin{algorithm}[H]
+\begin{algorithmic}[1]
+\REQUIRE $\,$ \\
+$\mathbf{A}$: design matrix \\
+$\mathbf{Y}$: response matrix \\
+$\lambda$: rank deficient parameter \\
+\ENSURE  $\,$ \\
+$\mathbf{X}$: parameters \\
+\STATE $\mathbf{M}=\mathbf{A^\top A}$ \\
+\STATE Initialize $\mathbf{Q R}=\mathbf{M}$ \\
+\STATE $\mathbf{X} = \mathbf{R^{-1}Q^\top Y}$ \\
+\STATE $\mathbf{t} = \mathbf{X}$ \\
+\FOR { $i = 1,2,3,\dots$ }
+        \STATE $\mathbf{t} =\lambda \mathbf{t}$  
+        \STATE $\mathbf{t} =  -(\mathbf{M}+\lambda\mathbb{I})^{-1}\mathbf{t}$
+	\STATE $\mathbf{X}=\mathbf{X} + \mathbf{t}$
+\ENDFOR
+\end{algorithmic}
+\caption{Algorithm for handling rank deficient matrices improved}
+\label{alg:colemanimproved}
+\end{algorithm}
+
+Both algorithms are equivalent, but algorithm~\ref{alg:colemanimproved} is more
+stable and converges in typically less than 10 steps.
+It is important to notice that the QR factorisation is computed only once and it
+is computationally less expensive than the SVD.
+
 %%% Measure error

### 2015-03-02 src/4_OnlineLearning.tex
efficient computation ready

@@
 
+Despite the bias-variance tradeoff provides a conceptual framework for
+determining a good model, is not directly useful. Some popular methods
+for determine model selection are:
+
+\begin{itemize}
+    \item Akaike information criterion (AIC) (Akaike, 1974)
+    \item Bootstrap-based selection (Efron and Tibshirani, 1997)
+    \item Cross-validation (Stone, 1974)
+\end{itemize}
+
+On the other hand, $\lambda$ can also be used to reduce computational time as it
+is shown in equation~(\ref{eq:taylor}). Golub (~\cite{golub1965}) suggested that
+$\lambda$ be chosen so that:
+
+\[
+\frac{\lambda}{\lambda + \delta^2} < 0.1
+\]
+
+\noindent where $\delta$ is a lower bound of the smallest non-zero singular
+value $\sigma_k$. One approach is to choose the greatest lower bound for
+$\delta=\sigma_k$. Hence:
+
+\[
+\frac{\lambda}{\lambda + \sigma_k^2} < 0.1
+\]
+
+\noindent if we choose $\lambda=\beta \sigma_k^2$ then we get the following
+relation:
+
+\[
+\frac{\beta \sigma_k^2}{\beta \sigma_k^2 + \sigma_k^2} =
+\frac{\beta}{\beta+1} < 0.1 \Rightarrow \beta < \frac{1}{9}  
+\]
+
+Experiments done by ~\cite{coleman+sun2010} have shown that $\beta = 0.01$
+produces satisfactory results. This means that $\lambda = 0.01 \sigma_k^2$.
+
+However, get $\sigma_k$ imply obtaining first the SVD, which is
+computational expensive.
+
+Since the algorithm presented by ~\cite{coleman+sun2010} already compute the QR
+factorization of the matrix $\mathbf{A}$ they show a procedure for getting
+a $\lambda$ approximation using QR.
+
+\begin{enumerate}
+\item Compute QR factorization: $\mathbf{A} = \mathbf{Q_1R_1}$
+\item Let $\mathbf{W}$ denote the set of absolute values of the nonzero diagonal
+elements of $\mathbf{R_1}$. Let $w_{\text{min}}$ and $w_{\text{max}}$ denote the
+smallest and largest elements of $\mathbf{W}$ respectively. Both 
+$\lambda_1 = \hat{\beta} w_{\text{min}}^2$
+$\lambda_2 = \hat{\beta} \frac{w_{\text{min}}^2}{w_{\text{max}}^2}$ where
+$\hat{\beta}=0.00025$ produce satisfactory results.
+\item The $\lambda_{\text{QR}}$ approximation is obtained as:
+$\lambda_{\text{QR}} = \frac{\lambda_1 + \lambda_2}{2}$  
+\end{enumerate}
+
 %%% Measure error

### 2015-03-05 src/4_OnlineLearning.tex
Agregando paper OVECM

@@
 
-\section{The Aggregating Algorithm for Regression}
+\subsection{The Aggregating Algorithm for Regression}
 
@@
 
-\section{Efficient computation}
+\subsection{Efficient computation}
 
@@
 
-%%% Measure error
+\section{Evaluation methods} \label{sec:evaluation}
+
+Forecast performance is evaluated using different methods. We have chosen three
+measures usually used:
+\begin{description}
+\item[MAPE] Mean Average Percent Error which presents forecast errors as a
+percentage.
+\begin{equation}\label{eq:MAPE}
+\text{MAPE} = \frac{1}{N} \sum_{t=1}^{N} 
+\frac{\left|\mathbf{y}_t-\hat{\mathbf{y}}_t\right|}{\left|\mathbf{y}_t\right|}
+ \times 100 
+\end{equation}
+\item[MAE] Mean Average Error which measures the distance between forecasts to the
+true value.
+\begin{equation}\label{eq:MAE}
+\text{MAE} = \frac{1}{N} \sum_{t=1}^{N} 
+\left| 
+\mathbf{y}_t-\hat{\mathbf{y}}_t
+\right| 
+\end{equation}
+\item[RMSE] Root Mean Square Error also measures the distance between forecasts
+to the true values but, unlike MAE, large deviations from the true value have a
+large impact on RMSE due to squaring forecast error.
+\begin{equation}\label{eq:RMSE}
+\text{RMSE} = \sqrt{
+\frac{\displaystyle \sum_{t=1}^{N} (\mathbf{y}_t-\hat{\mathbf{y}}_t)^2}{N}}
+\end{equation}
+\end{description}
+
+\section{Model selection} \label{sec:pselection}
+Akaike Information Criterion (AIC) is often used in model selection where AIC
+with smaller values are preferred since they represent a trade-off between bias
+and variance.  AIC is obtained as follows:
+
+\begin{equation}
+\label{eq:aicformula}
+AIC = \underset{\text{bias}}{-\frac{2l}{N}} + 
+\underset{\text{variance}}{\frac{2k}{N}}
+\end{equation}
+
+\noindent where
+
+\begin{description}
+\item[l] is the loglikelihood function
+\item[k] number of estimated parameters
+\item[N] number of observations
+\end{description}
+
+Loglikelihood function is obtained from the Residual Sum of Squares (RSS):
+
+\begin{equation}
+\label{eq:ll}
+l = -\frac{N}{2} \left(1 + ln(2\pi) + ln\left(\frac{RSS}{N}\right)\right) 
+\end{equation}

### 2015-03-05 src/5_Proposal.tex
Agregando paper OVECM

@@
 \chapter{Online VECM proposal}
-\section{\uppercase{Methodology}}
-\label{sec:methodology}
-\noindent 
+Financial time series are known for their non-stationary behaviour. However,
+sometimes they exhibit some stationary linear combinations. When this happens,
+it is said that those time series are cointegrated.The Vector Error Correction
+Model (VECM) is an econometric model which characterizes the joint dynamic
+behaviour of a set of cointegrated variables in terms of forces pulling towards
+equilibrium. In this study, we propose an Online VEC model (OVECM) which
+optimizes how model parameters are obtained using a sliding window of the most
+recent data. Our proposal also takes advantage of the long-run relationship
+between the time series in order to obtain improved execution times. Our
+proposed method is tested using four foreign exchange rates with a frequency of
+1-minute, all related to the USD currency base. OVECM is compared with VECM and
+ARIMA models in terms of forecasting accuracy and execution times. We show that
+OVECM outperforms ARIMA forecasting and enables execution time to
+be reduced considerably while maintaining good accuracy levels compared with
+VECM.
+
+\vspace{0.5cm}
+
+\section{The problem}
+In finance, it is common to find variables with a long-run and/or a short-run
+equilibrium relationship. This relationship is called cointegration and it
+reflects the idea of that some set of variables cannot wander too far away from
+each other.
+
+Cointegration means that one or more combinations of these variables is
+stationary even though individually they are not.
+
+Some financial models, such as  vector error correction (VEC), take advantage of
+this property and describe the joint behaviour of several cointegrated
+financial instruments.
+
+VEC is a special case of the vector autorregressive model (VAR). VAR is a linear
+model which express future values in terms of its own history and other
+variables past values. If cointegration exists, VEC model allows to introduce an
+error correction term due to its cointegration estimation error.
+
+VEC as well as VAR model parameters are obtained using ordinary least squares
+method (OLS). Since OLS involves many calculations, the parameter estimation
+method is computationally expensive when the number of past values and
+observations considered increases. Moreover, OLS is an ill-posed problem which
+admits an infinite number of solutions.
+
+Ridge regression method (RR) tackles this ill-posed problem and is usually
+formulated instead of OLS. RR includes a regularisation parameter that leads to
+better generalisation capability. However, RR is still computationally
+expensive dealing with large datasets. Recently, online learning algorithms
+have been proposed to solve problems with large data sets because of their
+simplicity and their ability of updating the model when new data is available.
+
+There are several popular online methods such as
+perceptron~\cite{rosenblatt58}, passive-aggressive~\cite{crammerETall2006},
+stochastic gradient descent~\cite{zhang2004}, aggregating
+algorithm~\cite{vovk2001} and the second order
+perceptron~\cite{cesa-bianchi2005}.  In~\cite{cesa-bianchi2006}, an in-deph
+analysis of online learning is provided.  In particular, the {\em aggregating
+algorithm for regression} (AAR) method is a recursive formulation of RR
+suitable for using in an online context.
+
+We propose an online formulation of the VEC model based on the
+AAR method considering only a sliding window of the most recent data. The
+algorithm introduces matrix optimisations in order to reduce the number of
+operations. Our method is later tested with financial data from the foreign
+exchange market.
+
+\section{Methodology} \label{sec:methodology}
+
 \subsection{Online VECM} \label{sec:proposal}
@@
         \texttt{vecMatrixOnline}(\mathbf{y}_i,p,v,\mathbf{A},\mathbf{B})$}
-        %\STATE $\Delta \mathbf{Y}_{\text{true}}[i] \gets \mathbf{B}[-1,:]$
         \STATE $\Delta \mathbf{Y}_{\text{pred}}[i] \gets \mathbf{A}[-1,:] \times \mathbf{X}$
     \ENDIF
-    \STATE $\mathbf{X} \gets \texttt{OLS} (\mathbf{A},\mathbf{B})$%\mathbf{(A^\top A)^{-1}A^\top B}$
-    %\STATE $\Delta \mathbf{y}_{\text{true}} \gets \mathbf{B}[-n,:]$
-    %\STATE $\Delta \mathbf{y}_{\text{pred}} \gets \mathbf{A}[-n,:] \times \mathbf{X}$
+    \STATE $\mathbf{X} \gets \texttt{OLS} (\mathbf{A},\mathbf{B})$
     \STATE $e \gets \texttt{mape}(\mathbf{B}[-n,:],\mathbf{A}[-n,:] \times \mathbf{X})$
-    %\STATE $e \gets \texttt{mape}(\Delta \mathbf{y}_{\text{true}}, \Delta
-    %\mathbf{y}_{\text{pred}})$
     \IF {$\texttt{mean}(e) > \text{mean\_error}$}
@@
         \texttt{vecMatrixUpdate}(\mathbf{y}_i,p,v,\mathbf{A})$}
-        \STATE $\mathbf{X} \gets \texttt{OLS} (\mathbf{A},\mathbf{B})$%\mathbf{(A^\top A)^{-1}A^\top B}$
-        %\STATE $\mathbf{X} \gets \mathbf{(A^\top A)^{-1}A^\top B}$
+        \STATE $\mathbf{X} \gets \texttt{OLS} (\mathbf{A},\mathbf{B})$
     \ENDIF
@@
 \mathbf{Y}[L:N-1] +\Delta \mathbf{Y}_{\text{pred}}$
-%\STATE $\text{MAPE} \gets \texttt{mape}(\Delta \mathbf{Y}_{\text{true}}, \Delta
-%\mathbf{Y}_{\text{pred}})$
 \end{algorithmic}
@@
 
-\subsection{Evaluation methods} \label{sec:evaluation}
-
-Forecast performance is evaluated using different methods. We have chosen three
-measures usually used:
-\begin{description}
-\item[MAPE] Mean Average Percent Error which presents forecast errors as a
-percentage.
-\begin{equation}\label{eq:MAPE}
-\text{MAPE} = \frac{1}{N} \sum_{t=1}^{N} 
-\frac{\left|\mathbf{y}_t-\hat{\mathbf{y}}_t\right|}{\left|\mathbf{y}_t\right|}
- \times 100 
-\end{equation}
-\item[MAE] Mean Average Error which measures the distance between forecasts to the
-true value.
-\begin{equation}\label{eq:MAE}
-\text{MAE} = \frac{1}{N} \sum_{t=1}^{N} 
-\left| 
-\mathbf{y}_t-\hat{\mathbf{y}}_t
-\right| 
-\end{equation}
-\item[RMSE] Root Mean Square Error also measures the distance between forecasts
-to the true values but, unlike MAE, large deviations from the true value have a
-large impact on RMSE due to squaring forecast error.
-\begin{equation}\label{eq:RMSE}
-\text{RMSE} = \sqrt{
-\frac{\displaystyle \sum_{t=1}^{N} (\mathbf{y}_t-\hat{\mathbf{y}}_t)^2}{N}}
-\end{equation}
-\end{description}
-
-\subsection{Model selection} \label{sec:pselection}
-Akaike Information Criterion (AIC) is often used in model selection where AIC
-with smaller values are preferred since they represent a trade-off between bias
-and variance.  AIC is obtained as follows:
-
-\begin{equation}
-\label{eq:aicformula}
-AIC = \underset{\text{bias}}{-\frac{2l}{N}} + 
-\underset{\text{variance}}{\frac{2k}{N}}
-\end{equation}
-
-\noindent where
-
-\begin{description}
-\item[l] is the loglikelihood function
-\item[k] number of estimated parameters
-\item[N] number of observations
-\end{description}
-
-Loglikelihood function is obtained from the Residual Sum of Squares (RSS):
-
-\begin{equation}
-\label{eq:ll}
-l = -\frac{N}{2} \left(1 + ln(2\pi) + ln\left(\frac{RSS}{N}\right)\right) 
-\end{equation}
-
     \begin{tikzpicture}[node distance=2cm]
@@
     \end{tikzpicture}
+
+\section{Results} \label{sec:results}
+
+\subsection{Data} \label{sec:unitroot}
+Tests of SLVECM, SLARIMA and our proposal OVECM were carried out using foreign
+four exchange data rates: EURUSD, GBPUSD, USDCHF and USDJPY. This data was
+collected from the free database Dukascopy which gives access to the Swiss
+Foreign Exchange Marketplace ~\cite{Dukascopy2014}.
+
+The reciprocal of the last two rates (CHFUSD, JPYUSD) were used in order to
+obtain the same base currency for all rates.  The tests were done using
+1-minute frequency from ask prices which corresponded to 1.440 data points per
+day from the 11th to the 15th of August 2014.
+
+\subsection{Unit root tests} \label{sec:unitroot}
+Before running the tests, we firstly checked if they were I(1) time series
+using the Augmented Dickey Fuller (ADF) test.
+
+\begin{table}[h!]
+\caption{Unit roots tests}
+\label{tab:adf}
+\begin{center}
+\begin{tabular}{|l|c|c|c|c|c|}
+\hline
+& \textbf{Statistic} & \textbf{Critical value} & \textbf{Result}\\
+\hline
+EURUSD          & -0.64 & -1.94 & True       \\
+$\Delta$EURUSD & -70.45   & -1.94 & False       \\
+GBPUSD          & -0.63   & -1.94 & True          \\
+$\Delta$GBPUSD & -54.53   & -1.94 & False       \\
+CHFUSD          & -0.88   & -1.94 & True         \\
+$\Delta$CHFUSD & -98.98   & -1.94 & False       \\
+JPYUSD          & -0.65 & -1.94 & True        \\
+$\Delta$JPYUSD & -85.78 & -1.94 & False     \\ 
+\hline
+\end{tabular}
+\end{center}
+\end{table}
+
+Table~\ref{tab:adf} shows that all currency rates cannot reject the unit root
+test but they rejected it with their first differences. This means that all of
+them are I(1) time series and we are allowed to use VECM and therefore OVECM.
+
+\subsection{Parameter selection} \label{sec:paramselection}
+
+In order to set OVECM parameters: windows size $L$ and lag order $p$,
+we propose to use several window sizes: $L = 100, 400, 700, 1000$. For every window size $L$ we chose lag order with minimum AIC.
+
+ARIMA parameters were also obtained using AIC.
+Parameters optimisation is presented in table~\ref{tab:params}:
+
+\begin{table}[ht]
+\caption{Parameters optimisation. VECM order and ARIMA parameters were selected
+using AIC.}
+\label{tab:params}
+\begin{center}
+\begin{tabular}{|c|c|c|}
+\hline
+Windows size $L$ & VECM & ARIMA\\
+ $L$ & order $(p)$ & order $(p,d,q)$ \\
+\hline
+100 & 2 & p=2,d=1,q=1\\
+400 & 5 & p=1,d=1,q=1\\
+700 & 3 &p=2,d=1,q=1\\
+1000 &3 & p=2,d=1,q=1\\
+\hline
+\end{tabular}
+\end{center}
+\end{table}
+
+In OVECM we also define a mean\_error variable, which was defined based on the
+in-sample MAPEs. Figure \ref{fig:mapes} shows how MAPE moves and how mean\_error
+variable help us to decide whether new cointegration vectors are needed.
+
+\begin{figure}[!h]
+  \centering
+   \includegraphics[width = 7.5cm]{img/mapes} 
+  \caption{In-sample MAPEs example for 50 minutes. The average of them is
+  considered to obtain new cointegration vectors.}
+  \label{fig:mapes}
+ \end{figure}
+
+\subsection{Execution times} \label{sec:exectimes}
+We ran OVECM and SLVECM 400 iterations. SLARIMA execution time is excluded
+because its is not comparable with OVECM and SLVECM. SLARIMA was created based
+on statsmodels library routine ARIMA.
+
+The execution times are shown in the table~\ref{tab:extimes}.
+
+\begin{table}[h!]
+\caption{Execution times}
+\label{tab:extimes}
+\begin{center}
+\begin{tabular}{|l|c|c|c|c|c|}
+\hline
+& $L$ & order & e  & Time[s] \\
+\hline
+OVECM & 100 &p=2  & 0      & 2.492\\
+OVECM & 100 &p=2  & 0.0026  & 1.606\\
+SLVECM & 100 &p=2& -- & 2.100\\
+\hline
+OVECM & 400 & p=5  & 0      & 3.513\\
+OVECM & 400 &p=5  & 0.0041  & 2.569\\
+SLVECM & 400 & p=5 & -- & 3.222\\
+\hline
+OVECM & 700 &p=3  & 0      & 3.296\\
+OVECM & 700 &p=3  & 0.0032  & 2.856\\
+SLVECM & 700 &p=3 & -- & 3.581\\
+\hline
+OVECM & 1000 & p=3 & 0      & 4.387\\
+OVECM & 1000 & p=3  & 0.0022  & 2.408\\
+SLVECM & 1000 & p=3  & -- & 3.609\\
+\hline
+\end{tabular}
+\end{center}
+\end{table}
+
+It is clear that execution time depends directly on $L$ and $p$ since they
+determine the size of matrix $\mathbf{A}$ and therefore affect the OLS function
+execution time.
+It is worthy of note that execution time also depends on mean\_error because it
+determines how many times OVECM will recalculate cointegration vectors which is
+an expensive routine.
+
+Figure~\ref{fig:mapes} shows an example of the in-sample MAPE for 50 iterations.
+When the average of the in-sample MAPEs is above mean\_error new cointegration
+vectors are obtained. In consequence, OVECM performance increases when
+mean\_error increases. However, this could affect accuracy, but
+table~\ref{tab:stats} shows that using an appropriate mean\_error doesn't affect
+accuracy considerable.
+
+\subsection{Performance accuracy} \label{sec:performacc}
+
+Table~\ref{tab:stats} shows in-sample and out-of-sample performance measures:
+MAPE, MAE and RMSE for OVECM, SLVECM and SLARIMA. Test were done using the
+parameters defined in table \ref{tab:params}.  We can see that OVECM has very
+similar performance than SLVECM and this support the theory that cointegration
+vectors vary little in time. Moreover, OVECM also out performed SLARIMA based on
+these three performance measures.
+
+\begin{table*}[ht!]
+\caption{Model measures}
+\label{tab:stats}
+\begin{center}
+\begin{tabular}{|l|l|c|c|c|c|c|c|c|c|}
+\hline
+\multicolumn{4}{|c|}{Model} & \multicolumn{3}{|c|}{In-sample} &
+\multicolumn{3}{|c|}{Out-of-sample} \\ 
+\hline
+\hline
+Method & $L$ & Parameters & e & 
+MAPE & MAE& RMSE& 
+MAPE & MAE& RMSE \\
+\hline
+ OVECM  &   100  &  P=2& 0.0026  &  0.00263&  0.00085&  0.00114&  0.00309&  0.00094&  0.00131\\
+ OVECM  &   400  &  P=5& 0.0041  &  0.00378&  0.00095&  0.00127&  0.00419&  0.00103&  0.00143\\
+ OVECM  &   700  &  P=3& 0.0032  &  0.00323&  0.00099&  0.00130&  0.00322&  0.00097&  0.00132\\
+ OVECM  &   1000 &  P=3& 0.0022  &  
+ \textbf{0.00175}&  \textbf{0.00062}&  \textbf{0.00087} &  
+ \textbf{0.00172}&  \textbf{0.00061}&  \textbf{0.00090}\\
+\hline
+ SLVECM  &   100 &  P=2& -  &  0.00262&  0.00085&  0.00113&  0.00310&  0.00095&  0.00132\\
+ SLVECM  &   400 &  P=5& -  &  0.00375&  0.00095&  0.00126&  0.00419&  0.00103&  0.00143\\
+ SLVECM  &   700 &  P=3& -  &  0.00324&  0.00099&  0.00130&  0.00322&  0.00098&  0.00132\\
+ SLVECM  &   1000 &  P=3& -  &  
+ \textbf{0.00174}&  \textbf{0.00061}&  \textbf{0.00087}&  
+ \textbf{0.00172}&  \textbf{0.00061}&  \textbf{0.00090}\\
+\hline
+SLARIMA & 100  &p=2, d=1, q=1 & - & 0.00285  &  0.00110  &  0.00308  &  0.00312  &0.00098  &  0.00144 \\
+SLARIMA & 400  &p=1, d=1, q=1 & - & 0.00377  &  0.00101  &  0.00128  &  0.00418 & 0.00106 & 0.00145 \\
+SLARIMA & 700  &p=2, d=1, q=1 & - & 0.00329  &  0.00102  &  0.00136  &  0.00324  & 0.00097  &  0.00133 \\
+SLARIMA & 1000 &p=2, d=1, q=1 & - & \textbf{0.00281}  & \textbf{0.00074}  &  \textbf{0.00105}  &  \textbf{0.00177} & \textbf{0.00063}  &  \textbf{0.00092} \\
+\hline
+\end{tabular}
+\end{center}
+\end{table*}
+
+We can also notice that in-sample performance in OVECM and SLVECM is related
+with the out-of-sample performance.  This differs with SLARIMA which models with
+good in-sample performance are not necessarily good out-of-sample models.
+Moreover OVECM outperformed SLARIMA using the same window size.
+
+Figure ~\ref{fig:accuracy} shows the out-of-sample forecasts made by our
+proposal OVECM with the best parameters found based on table~\ref{tab:stats}
+which follows the time series very well.
+
+\begin{figure}[h]
+  \vspace{-0.2cm}
+  \centering
+   \includegraphics[width = 7.5cm]{img/accuracy} 
+  \caption{OVECM forecasting accuracy example for 50 minutes using $L = 1000$ and $p =3$}
+  \label{fig:accuracy}
+ \end{figure}
+
+\section{Conclusions} \label{sec:conclusions}
+
+A new online vector error correction method was presented. 
+We have shown that our proposed OVECM considerably reduces execution times
+without compromising solution accuracy.  OVECM was compared with VECM and ARIMA
+with the same sliding window sizes and OVECM outperformed both in terms of
+execution time. Traditional VECM slightly outperformed our proposal but the
+OVECM execution time is lower.
+This reduction of execution time is mainly because OVECM avoids the cointegration vector
+calculation using the Johansen method. The condition for getting new vectors is
+given by the mean\_error variable which controls how many times the Johansen method is
+called. Additionally, OVECM introduces matrix optimization in order to get the
+new model in an iterative way.
+We could see that our algorithm took much less than a minute at every
+step. This means that it could also be used with higher frequency data and would
+still provide responses before new data arrives.
+
+For future study, it would be interesting to improve the out-of-sample forecast
+by considering more explicative variables, to increase window sizes or trying new
+conditions to obtain new cointegration vectors.
+
+Since OVECM is an online algorithm which optimizes processing time, it could be
+used by investors as an input for strategy robots. Moreover, some technical
+analysis methods could be based on its output.

### 2015-03-11 src/3_TimeSeries.tex
adding AVECM proposal

@@
 
-The vector autoregressive VAR($p$) model is one of the most easy to use, successful and flexible models for the analysis of multivariate time series.
+The vector autoregressive VAR($p$) model \cite{sims1980} is one of the most easy to use, successful and flexible models for the analysis of multivariate time series.
 It is a natural extension of the univariate autoregressive (AR) model.
@@
 \section{Vector error correction model}
-Vector error correction model (VECM) is a special form of a VAR model for I(1) variables that are also
+Vector error correction model (VECM)  \cite{engle87} is a special form of a VAR model for I(1) variables that are also
 cointegrated~\cite{banerjee1993}.
@@
 
-VECM models are employed because many economic time series appear to be first-difference stationary, i.e. I(1)
-
 Conventional regression estimators, including VARs, have good properties when applied to covariance-stationary time series, but encounter difficulties when applied to non stationary or integrated processes.
@@
 If the series are cointegrated, they move together in the long-run. A
-VAR in first differences will not capture those long-run tendences.
+VAR in first differences will not capture those long-run tendencies.
 VECM fit to the first differences of the non stationary variables, but a lagged error-correction term is added to the relationship representing the long-run relationship.
+
+In finance, many economic time series are revealed to be stationary when they
+are differentiated and moreover cointegration restrictions often improve
+forecasting \cite{duy1998}. Therefore, VECM has been widely adopted in
+finance: \cite{mukherjee1995}, \cite{maysami2000}, \cite{arestis2001} and
+\cite{seong2013} to name a few. Pair
+trading is a very common example of a cointegration
+application \cite{herlemont2003} but also can also be extended to a larger set
+of variables \cite{mukherjee1995}, \cite{engle2004}.

### 2015-03-11 src/5_2_Proposal.tex
adding AVECM proposal

\chapter{Adaptive parallel algorithm for VECM}

VECM parameters are those which maximise cointegration relations in the past. However, this search is computationally expensive since the Johansen
method~\cite{johansen1995} is used, which is of order $O(n^3)$. We propose
to parallelise this search in order to get new parameters at every iteration.

\vspace{0.5cm}

\section{Introduction}
Both VECM and VAR model parameters are obtained using ordinary least squares
(OLS) method \cite{golub1980}. Since OLS involves many calculations, the parameter estimation
method is computationally expensive when the number of past values and
observations increases. Moreover, obtaining cointegration vectors 
is also an expensive routine because of the use of the Johansen method.

Model effectiveness is focused on out-of-sample forecast rather than in-sample
fitting. This criterion allows our proposal AVECM prediction capability to be
expressed rather than just explaining data history.

The forecast capability of our method was measured using MAPE, MAE and
RMSE \cite{armstrong1992}. Tests were run using four currency rates: Euro (EUR) to
United States Dollar (USD) (EURUSD), British Pound (GBP) to USD (GBPUSD), USD to Swiss
Franc (CHF) (USDCHF) and USD to Japanese Yen (JPY) (USDJPY) with a 10-seconds frequency.

\def\rot#1{{\color{red}{#1}}}
\def\gruen#1{{\color{green}{#1}}}
\def\blau#1{{\color{blue}{#1}}}

\section{Background}
\label{sec:background}

VECM is obtained re-writing equation (\ref{eq:var}) in terms of the new
variable $\Delta\mathbf{y}_t=\mathbf{y}_t-\mathbf{y}_{t-1}$.
The VECM model, expressed in terms those differences, takes the form:
\begin{equation}\label{eq:vec}
\Delta \mathbf{y}_t 
= \boldsymbol{\Omega}\,\mathbf{y}_{t-1}
  + \sum_{i=1}^{p-1} \boldsymbol{\Phi}_i^*\,\Delta\mathbf{y}_{t-i}
  + \mathbf{c} + \boldsymbol{\epsilon}_t\,,
\end{equation}
\noindent
where the coefficients matrices $\boldsymbol{\Phi}_i^*$ and 
$\boldsymbol{\Omega}$, expressed in terms of the matrices
$\boldsymbol{\Phi}_i$ of (\ref{eq:var}), are:
\begin{align*}
\boldsymbol{\Phi}_i^* 
&:= -\sum_{j=i+1}^{p}\boldsymbol{\Phi}_j\,, \\
\boldsymbol{\Omega}
&:= -\left( \mathbb{I} - \boldsymbol{\Phi}_1 - \dots 
    - \boldsymbol{\phi}_p \right)\,. 
\end{align*}
The following well known properties of the matrix $\boldsymbol{\Omega}$
\cite{johansen1995} will be useful in the sequel:
\begin{itemize}
\item
If $\boldsymbol{\Omega} = \mathbf{0}$, there is no cointegration.
\item 
If $rank(\boldsymbol{\Omega})=l$, i.e., if $\boldsymbol{\Omega}$ has
full rank, then the time series are not I(1) but stationary.
\item
If $rank(\boldsymbol{\Omega})=r$, $0<r<l$, then there is cointegration
and the matrix $\boldsymbol{\Omega}$ can be expressed as
$\boldsymbol{\Omega}=\boldsymbol{\alpha\beta}^\top$, where $\boldsymbol{\alpha}$
and $\boldsymbol{\beta}$ are
$l\times r$ matrices and
$\text{rank}(\boldsymbol{\alpha})=\text{rank}(\boldsymbol{\beta})=r$.
\item
The columns of $\boldsymbol{\beta}$ contains the cointegration vectors and the rows of
$\boldsymbol{\alpha}$ correspond with the adjusted vectors. 
$\boldsymbol{\beta}$ is obtained by Johansen procedure~\cite{johansen1988},
whereas $\boldsymbol{\alpha}$ has to be determined as a variable in the VECM.
\end{itemize}

If cointegration exists, then equation (\ref{eq:vec}) can be written
as follows:
\begin{equation}\label{eq:vecfull}
\Delta\mathbf{y}_t 
= \boldsymbol{\alpha\beta}^\top\mathbf{y}_{t-1} 
  + \sum_{i=1}^{p-1}\boldsymbol{\Phi}_i^*\,\Delta\mathbf{y}_{t-i}
  + \mathbf{c} + \boldsymbol{\epsilon}_t\,,
\end{equation}
\noindent
which is a VAR model but for time series differences.

Transposing each equation of the system (\ref{eq:vecfull}) we can write
the VECM($p$) model in block-matrix form as:
\begin{equation}\label{eq:vareq}
\mathbf{B} = 
\mathbf{A} \mathbf{X} + 
\mathbf{E} \, , 
\end{equation}
%
\noindent where $\mathbf{B}$ dimension is $((N-p)\times l)$, $\mathbf{A}$
dimension is $((N-p)\times(r+(p-1)l +1))$, $\mathbf{X}$ dimension is $((r+(p-1)l
+1)\times l)$ and $\mathbf{E}$ dimension is $((N-p)\times l)$:
%
\begin{alignat}{3}
\mathbf{B}
&= \begin{bmatrix}
   \Delta\mathbf{y}_{p+1}^\top \\
   \Delta\mathbf{y}_{p+2}^\top \\
   \vdots \\
   \Delta\mathbf{y}_N^\top
   \end{bmatrix}
&\quad
\mathbf{X}
&= \begin{bmatrix}
   \boldsymbol{\alpha}^\top \\
   \boldsymbol{\Phi}_1^{*\top} \\
   \boldsymbol{\Phi}_2^{*\top} \\
   \vdots \\
   \boldsymbol{\Phi}_{p-1}^{*\top} \\
   \mathbf{c}^\top
   \end{bmatrix}
&\quad
\mathbf{E}
&= \begin{bmatrix}
   \boldsymbol{\epsilon}_{p+1}^\top \\
   \boldsymbol{\epsilon}_{p+2}^\top \\
   \vdots \\
   \boldsymbol{\epsilon}_N^\top \\
   \end{bmatrix}
\end{alignat}
\noindent and 
\begin{align}
\mathbf{A} 
&= \begin{pmat}[{....|}]
   \mathbf{y}_p^\top \boldsymbol{\beta} & \Delta \mathbf{y}_p^\top & \Delta\mathbf{y}_{p-1}^\top & \dots 
                    & \Delta\mathbf{y}_2^\top & 1 \cr
   \mathbf{y}_{p+1}^\top  \boldsymbol{\beta} &\Delta\mathbf{y}_{p+1}^\top & \Delta\mathbf{y}_p^\top & \dots
                       & \Delta\mathbf{y}_3^\top & 1 \cr
   \vdots & \vdots & \vdots & \ddots & \vdots & \vdots \cr
   \mathbf{y}_{N-1}^\top  \boldsymbol{\beta} &\Delta\mathbf{y}_{N-1}^\top & \Delta\mathbf{y}_{N-2}^\top & \dots 
                       & \Delta\mathbf{y}_{N-p-1}^\top & 1 \cr
   \end{pmat}\, .
\end{align}
Taking into account the error term $\mathbf{E}$, equation~(\ref{eq:vareq}) 
can be solved with respect to $\mathbf{X}$ using the ordinary least
squares estimation.

\section{Motivation} \label{sec:proposal}

Cointegration vectors can be found applying the Johansen method which uses a
sample of the last historical data. However, one of the main limitations of this
methodology is that it assumes cointegration vectors do not change in time.
In fact, the long-run relationships between the time series might change due to
several economic factors that can lead to structural breaks in the cointegrating
relation \cite{gregoryETal1996}. 
Experiments show that cointegration depends on the amount $L$ of historical data
considered, and how many lags $p$ we use.

Figure~\ref{fig:hists} shows the number $r$ of cointegration vectors given by
the Johansen method for different values of the parameters $L$ and $p$, using
four currencies data, i.e., $l=4$, and considering 1000 iterations $it$.
We know that when there is no cointegration vectors, i.e $r=0$, 
there is no cointegration. In the same way, when $r=l$ reveals that no process is I(1) but,
instead, they are all stationary.
The interesting cases of cointegration are those, where $r$ lies strictly
between $0$ and $l$, i.e. $0<r<l$,

Experiments were run with data starting at 9:00 GMT of the 11th of August 2014,
which corresponds to the opening of the New York financial market.

\begin{figure}[!h]
  %\vspace{-0.8cm}
  \centering
  \includegraphics[width=\textwidth]{img/histCointVectors}
  \caption{Histogram of the number of cointegration vectors using $p=1$. Four
  possible values for $L$ are shown (100, 1900, 6100 and 10700).
  In the cases of $L=100$ and $L=6100$, most of the time there is little
  cointegration (0 or 4 cointegration vectors). However, for $L=1900$ and
  $L=10700$ significant cointegration is found.}
  \label{fig:hists}
\end{figure}

\begin{figure}[!h]
  %\vspace{-0.8cm}
  \centering
  \includegraphics[width=\textwidth]{img/histCointVectorsp5}
  \caption{Histogram of the number of cointegration vectors using $p=5$. 
   Despite the change in the distribution of cointegration vectors, 
   the percentage of cointegration is nearly the same as in the case $p=1$. 
   For example, for $L=10700$ the sum of 1 and 2 number of cointegration
   vectors is very similar to 1 cointegration vector for the case $L=10700$
   in figure \ref{fig:hists}.}
   \label{fig:histsp5}
\end{figure}

The same experiments, using now $p=5$, were carried out in order to see the
effect on the number of cointegration vectors found. 
Figure~\ref{fig:histsp5} shows that, despite the change in the distribution
of the number of cointegration vectors, the sum of their number with $0<r<4$,
remain nearly the same. 
Therefore we can say that the number of lags doesn't significantly affect
the extent of cointegration found.

In order to measure the extent of cointegration, we introduce a
{\em percentage of cointegration\/} as following:
\begin{equation} \label{eq:pcoint}
PC = 
\frac{\#\{ it \mid \text{$it$ has $r$ c.v. with $0<r<l$}\}}
     {\#it}\times 100
\end{equation}
where c.v. stands for cointegration vectors and $it$ is the number of iterations.

The goal of our next experiment was to find a relation between this ratio
$PC$ and the performance of the accuracy measure MAPE (see equation 
\ref{eq:MAPE}). 
The experiments suggest that there is indeed a relation between the $PC$ ratio
and MAPE.
Figure~\ref{fig:cointvsmape} shows that MAPE rapidly decreases with increasing
$PC$.

\begin{figure}[!h]
  %\vspace{-0.8cm}
  \centering
  \includegraphics[width=\textwidth]{img/MAPEvsCoint-offset21600-p-2-freq-10s}
  \caption{MAPE versus percentage of cointegration. MAPE was obtained from 1000
  iterations considering several possible $L=[100,10000,400]$ and $p=[1,5]$.
  Best $L$ found was 8900.}
  \label{fig:cointvsmape}
\end{figure}

From these results we observe that cointegration changes with different values
for $L$ and $p$ and also that better cointegration percentage leads to better
accuracy performance.
Therefore we propose to choose best $L$ and $p$ parameters determined at every
step. 
The criterion for choosing $L$ and $p$ is to maximise the percentage of
cointegration $PC$ considering a number of past iterations $it$.

Algorithm ~\ref{alg:AVECM} summarises our proposal. 
The function {\bf get\_best\_params} makes a grid search on the two vector
lists $Ls$ and $ps$ and returns the parameters $L$ and $p$ which maximise
the percentage of cointegration $PC$ (see equation~\ref{eq:pcoint}) for a
pre-defined number of iterations.
Since this search is computational expensive, the number of cointegration
vectors at each iteration is computed in parallel, thus ensuring a response
before new data is available. 
This parallel routine also allows to refine our grid search leading to obtain
improved forecasting performance.
After that optimum $L$ and $p$ parameters are found, VECM is built and used
to forecast next data.

\begin{algorithm}[ht]
\begin{algorithmic}[1]
\REQUIRE $\,$ \\
$\mathbf{y}$: matrix with $N$ input vectors and $l$ time series\\
$ps$: vector with the number of past values range \\
$Ls$: vector with window sizes range ($L<N$) \\
$outit$: Number of out-of-sample iterations($outit < N-L$)
$it$: Number of iterations of testing (in-sample)\\
$of$: Starting point of Testing \\
\ENSURE  $\,$ \\
$\{ \mathbf{y}_{\text{pred}}[1],\dots,\mathbf{y}_{\text{pred}}[N-L]\}$: model predictions 
\FOR { $i =of$ to $of + outit$ }
   \STATE $\mathbf{Y} \gets \mathbf{y}[i-\texttt{max}(Ls):i]$
    \STATE $L,p \gets
    \texttt{get\_best\_params}(Ls,ps,it,Y)$
    \STATE $\mathbf{y}_i \gets \mathbf{y}[i-L:i]$
        \STATE $model = VECM(\mathbf{y}_i, p)$
        \STATE $\mathbf{y}_{\text{pred}}[i] = model.predict()$
\ENDFOR
\end{algorithmic}
\caption{AVECM: Adaptive VECM.}
\label{alg:AVECM}
\end{algorithm}

%
%
%\subsection{Evaluation methods} \label{sec:evaluation}
%
%Forecast performance was evaluated using different methods. We chose three
%measures that are frequently used:
%\begin{description}
%\item
%{\bf MAPE}, Mean Average Percent Error, which presents forecast errors as a
%percentage:
%\begin{equation}\label{eq:MAPE}
%\text{MAPE} = \frac{1}{N} \sum_{t=1}^{N} 
%\frac{\left|\mathbf{y}_t-\hat{\mathbf{y}}_t\right|}{\left|\mathbf{y}_t\right|}
% \times 100 
%\end{equation}
%\item
%{\bf MAE}, Mean Average Error, which measures the distance between forecasts and the
%true value.
%\begin{equation}\label{eq:MAE}
%\text{MAE} = \frac{1}{N} \sum_{t=1}^{N} 
%\left| 
%\mathbf{y}_t-\hat{\mathbf{y}}_t
%\right| 
%\end{equation}
%\item
%{\bf RMSE}, Root Mean Square Error, also measures the distance between forecasts
%and the true values but, unlike MAE, large deviations from the true value have a
%large impact on RMSE due to squaring forecast error.
%\begin{equation}\label{eq:RMSE}
%\text{RMSE} = \sqrt{
%\frac{\displaystyle \sum_{t=1}^{N} (\mathbf{y}_t-\hat{\mathbf{y}}_t)^2}{N}}
%\end{equation}
%\end{description}
%
%
%
\section{Experimental results}
\label{sec:results}
\subsection{Data} \label{sec:unitroot}
AVECM tests were carried out using four foreign exchange rates all related to
USD: EURUSD, GBPUSD, USDCHF and USDJPY. This data was collected from the free
database Dukascopy which gives access to the Swiss Foreign Exchange marketplace
~\cite{Dukascopy2014}.

The tests were done using 10-seconds frequency from ask prices which
corresponded to 8640 data points per day from the 11th to the 15th of August
2014. Since data datetimes were in GMT, opening market times for London, New
York, Sidney and Tokyo corresponded with data points 1440, 3240, 6120 and 6840
respectively. Our tests were made considering two hours after the opening time
of the 13th of August. Therefore, in order to obtain these times, we added 2
days and 2 hours to all the opening times ($2\times 8640 + 2 \times 360$) to obtain:
19440, 21240, 24120 and 24840. We called offset to these times in
table~\ref{tab:stats}, which allowed us to obtain performance at different
times of the day.

\subsection{Unit root tests} \label{sec:unitroot}
Before running the tests, we firstly checked whether the time series were
I(1) using the Augmented Dickey Fuller (ADF) test at 95\% significance level.
Table~\ref{tab:adf} shows that all currency rates cannot reject the unit root
test but they rejected it with their first differences. This means that all of
them are I(1) time series and we are allowed to use VECM and therefore AVECM.

\begin{table}[h!]
\begin{center}
\begin{tabular}{|l|c|c|c|c|c|}
\hline
& \textbf{Statistic} & \textbf{Critical value} & \textbf{Result}\\
\hline
EURUSD          &  -0.05398   & -1.94101 & True       \\
$\Delta$ EURUSD & -89.12344   & -1.94101 & False       \\
GBPUSD          &  -0.88208   & -1.94101 & True          \\
$\Delta$ GBPUSD & -82.56501   & -1.94101 & False       \\
CHFUSD          & -0.49941    & -1.94101 & True         \\
$\Delta$ CHFUSD & -74.21940   & -1.94101 & False       \\
JPYUSD          &  0.36721    & -1.94101 & True        \\
$\Delta$ JPYUSD & -101.15589  & -1.94101 & False     \\ 
\hline
\end{tabular}
\end{center}
\caption{Unit roots tests for EURUSD, GBPUSD, USDCHF and USDJPY at 10-seconds
frequency.}
\label{tab:adf}
\end{table}

\section{Parallel implementation} \label{sec:paralell}

To determine optimal $L$ and $p$ parameters require the using of the Johansen
method which is a computationally expensive routine. 
In order to improve the execution time of this search, our proposal included
a parallel search of VECM parameters using high performance computing.
The main objective is to obtain a response before a new data arrives in
the next 10 seconds.

The Johansen method is implemented in the Python Statsmodels
library~\cite{seabold2010} and the parallel implementation was done using
MPI in Python. 
MPI was chosen because it allows large-scale parallel applications
with wide portability to be built, being able to run in large clusters
or on local computers.
Our tests were done in a cluster with 2 servers Xeon E5-2667 (2.90GHz)
of 24 cores each (48 cores in total) and 24GB RAM.

Since financial market are constantly changing, we ran our
algorithm using different times of the day. We chose to use the algorithm
starting on the third day and two hours after the market opening times. These
times corresponded to data points 19440 (5 GMT London), 21240 (10 GMT New
York), 24120 (19 GMT Sydney) and 24840 (9 GMT Tokyo).

In order to determine the percentage of cointegration vectors, we considered
different number of iterations: 10, 50 and 100. The number of possible
combinations ($nparam$) of $L$ and $p$ is shown for 12, 24 and 47.

\subsection{Performance accuracy} \label{sec:performacc}

Table~\ref{tab:stats} shows the out-of-sample performance measures: MAPE, MAE
and RMSE for AVECM for different offsets, iterations and number of parameters.
We observed that performance measures differ for different offsets having even
in some cases different orders. London and New York offsets are order $10^{-3}$
while Sidney and Tokyo vary between $10^{-3}$ and $10^{-4}$. Despite this
difference in the order, they all achieve their minimum MAPE, MAE and RMSE
measurements at 100 iterations. In the case of Tokyo, it also reduces its MAPE
order. In all cases, we can see that performance measures are improving with
the number of parameters.  This is because there are more possible options to find
optimal cointegration parameters.

Figure ~\ref{fig:accuracy} shows part of the out-of-sample forecasts made by
our proposed AVECM using 100 iterations and offset New York.

\begin{figure}[!h]
  \centering
  \includegraphics[width=\textwidth]{img/accuracy}
  \caption{AVECM out-of-sample forecasting for four currency rates.}
  \label{fig:accuracy}
\end{figure}

\begin{table*}[ht!]
\tiny
\caption{Performance and execution times for min MAPE criteria}
\label{tab:stats}
\begin{center}
\begin{tabular}{|l|l|l|c|c|c|c|c|}
\hline
\multicolumn{3}{|c}{Parameters} & 
\multicolumn{3}{|c}{Out-of-sample Performance} &
\multicolumn{2}{|c|}{Time[s]} \\ 
\hline

\hline
Offset & Iterations & $nparam$ & 
MAPE & MAE& RMSE& 
Sequencial & Parallel \\
\hline
 \multirow{12}{*}{19440 (London)} &
 \multirow{3}{*}{10} 
  &  12 & 1.8481E-03 & 5.1974E-04 & 7.9714E-04 & 2.64 & 2.61    \\
 &&  24 & 1.8793E-03 & 5.2935E-04 & 8.0128E-04 & 7.38 & 3.03 \\ 
 &&  47 & 1.8804E-03 & 5.2477E-04 & 7.9274E-04 & 18.86 & 3.07 \\ 
 \cline{2-8} 
 & \multirow{3}{*}{50} 
 &  12&  1.5760E-03 & 3.8559E-04 & 6.5579E-04 & 12.49 & 3.99 \\
 && 24 & 1.5645E-03 & 3.7379E-04 & 6.3128E-04 & 36.69 & 6.45 \\
 && 47 & 1.5757E-03 & 3.7336E-04 & \textbf{6.2828E-04} & 92.30 & 5.75 \\
 \cline{2-8} 
 & \multirow{3}{*}{100} 
 &  12&  1.5760E-03 & 3.8559E-04 & 6.5579E-04 & 25.00 & 5.90 \\ 
 && 24 & \textbf{1.5501E-03} & \textbf{3.6575E-04} & 6.4058E-04 & 70.58 & 9.74 \\ 
 && 47 & 1.5812E-03 & 3.8773E-04 & 6.4740E-04 & 185.05 & 11.09 \\
 \cline{2-8} 
 & \multirow{3}{*}{150} 
 &  12&  1.5760E-03 & 3.8559E-04 & 6.5579E-04 & 37.09 & 7.55 \\ 
 && 24 & 1.5501E-03 & 3.6575E-04 & 6.4058E-04 & 104.52 & 14.33 \\
 && 47 & 1.5812E-03 & 3.8773E-04 & 6.4740E-04 & 292.84 & 15.44 \\
\hline
\hline
 \multirow{12}{*}{21240 (New York)} &
 \multirow{3}{*}{10} 
  &  12 & 1.8358E-03 & 3.4906E-04 & 4.9314E-04 & 2.64 & 2.45 \\
 &&  24 & 1.8287E-03 & 3.4661E-04 & 4.9140E-04 & 7.38 & 3.03 \\
 &&  47 & 1.8402E-03 & 3.4937E-04 & 4.9298E-04 & 18.86 & 3.00 \\
 \cline{2-8} 
 & \multirow{3}{*}{50} 
 &  12&  1.6911E-03 & 3.3614E-04 & 4.8197E-04 & 12.49 & 4.00 \\
 && 24 & 1.6871E-03 & 3.3546E-04 & 4.8198E-04 & 36.69 & 6.15 \\
 && 47 & 1.6991E-03 & 3.3790E-04 & \textbf{4.8141E-04} & 92.30 & 6.17 \\
 \cline{2-8} 
 & \multirow{3}{*}{100} 
 &  12&  1.6911E-03 & 3.3614E-04 & 4.8197E-04 & 25.00 & 6.03 \\ 
 && 24 & \textbf{1.6857E-03} & \textbf{3.3510E-04} & 4.8308E-04 & 70.58 & 9.34 \\ 
 && 47 & 1.6978E-03 & 3.3766E-04 & 4.8285E-04 & 185.05 & 10.36 \\
 \cline{2-8} 
 & \multirow{3}{*}{150} 
 &  12& 1.6911E-03 & 3.3614E-04 & 4.8197E-04 & 37.09 & 7.77 \\   
 && 24 &1.6857E-03 & 3.3510E-04 & 4.8308E-04 & 104.52 & 13.84 \\
 && 47 &1.6978E-03 & 3.3766E-04 & 4.8285E-04 & 292.84 & 13.06 \\
\hline
\hline
 \multirow{12}{*}{24120 (Sidney)} &
 \multirow{3}{*}{10} 
  &  12 & 8.5847E-04 & 2.2284E-04 & 3.1202E-04 & 2.64 & 2.41 \\ 
 &&  24 & 7.9127E-04 & 1.9526E-04 & 2.9580E-04 & 7.38 & 2.86 \\ 
 &&  47 & 8.0491E-04 & 1.9578E-04 & 2.9669E-04 & 18.86 & 3.13 \\ 
 \cline{2-8} 
 & \multirow{3}{*}{50} 
 &  12& 9.1471E-04 & 2.2555E-04 & 3.1371E-04 & 12.49 & 4.02 \\ 
 && 24 &7.3172E-04 & 1.8017E-04 & 2.8054E-04 & 36.69 & 5.45 \\
 && 47 &7.3172E-04 & 1.8017E-04 & 2.8054E-04 & 92.30 & 6.48 \\
 \cline{2-8} 
 & \multirow{3}{*}{100} 
 &  12& 8.0458E-04 & 2.0706E-04 & 2.9542E-04 & 25.00 & 5.90 \\ 
 && 24 &\textbf{6.2547E-04} & 1.6451E-04 & 2.5945E-04 & 70.58 & 9.23 \\ 
 && 47 &6.3424E-04 & \textbf{1.6400E-04} & \textbf{2.5849E-04} & 185.05 & 11.13 \\
 \cline{2-8} 
 & \multirow{3}{*}{150} 
 &  12&  7.2437E-04 & 1.8782E-04 & 2.8016E-04 & 37.09 & 7.83 \\ 
 && 24 & 6.2547E-04 & 1.6451E-04 & 2.5945E-04 & 104.52 & 12.42 \\
 && 47 & 6.2547E-04 & 1.6451E-04 & 2.5945E-04 & 292.84 & 13.95 \\
\hline
\hline
 \multirow{12}{*}{24840 (Tokyo)} &
 \multirow{3}{*}{10} 
  &  12 & 1.0787E-03 & 3.3948E-04 & 5.5145E-04 & 2.64 & 2.45 \\ 
 &&  24 & 1.0823E-03 & 3.3992E-04 & 5.4697E-04 & 7.38 & 2.96 \\ 
 &&  47 & 1.0885E-03 & 3.4462E-04 & 5.5231E-04 & 18.86 & 3.05 \\
 \cline{2-8} 
 & \multirow{3}{*}{50} 
 &  12& 1.0753E-03 & 3.4955E-04 & 5.7100E-04 & 12.49 & 3.97 \\ 
 && 24 &1.1042E-03 & 3.5395E-04 & 5.6747E-04 & 36.69 & 6.08 \\ 
 && 47 &1.1238E-03 & 3.5794E-04 & 5.7330E-04 & 92.30 & 6.67 \\ 
 \cline{2-8} 
 & \multirow{3}{*}{100} 
 &  12& \textbf{8.4691E-04} & \textbf{2.8979E-04} & \textbf{5.2553E-04} & 25.00 & 5.74 \\ 
 && 24 &8.6649E-04 & 2.9405E-04 & 5.2874E-04 & 70.58 & 10.29 \\
 && 47 &8.8877E-04 & 2.9821E-04 & 5.3796E-04 & 185.05 & 9.00 \\
 \cline{2-8} 
 & \multirow{3}{*}{150} 
 &  12& 8.5021E-04 & 2.9579E-04 & 5.2707E-04 & 37.09 & 7.54 \\ 
 && 24 &8.6350E-04 & 2.9183E-04 & 5.2837E-04 & 104.52 & 13.05 \\
 && 47 &8.8597E-04 & 2.9594E-04 & 5.3759E-04 & 292.84 & 13.82 \\
\hline
\end{tabular}
\end{center}
\end{table*}

\subsection{Execution times} \label{sec:exectimes}
We ran AVECM 100 ($outit$ in algorithm~\ref{alg:AVECM}) iterations.
Table~\ref{tab:stats} shows execution times and performance measures for
different times of the market, in-sample iterations and number of parameters used.
The number of parameters represents the possible combinations of values for $L$ and
$P$. The $L$ parameter was always chosen between 100 and 4000 and $nparams$
determines how many values in this range were considered. $p$ always took values
between 1 and 5. Execution times were measured using the Time python library.

Execution time depends directly on $L$ and $p$ since they determine the size of
matrix $\mathbf{A}$ and therefore affects the OLS function execution time.
Therefore, if we try more combinations of $L$ and $p$ (increasing $nparam$) the
algorithm will take longer. However, the idea is to improve accuracy measures
ensuring execution times below 10 seconds (the time series frequency).

Table ~\ref{tab:stats} shows that best performance accuracy measures are
achieved in times near or below 10 seconds in the parallel version. Contrarily,
sequential times are high, above 10 seconds in most of the cases.

Despite there being some parallel execution times above 10 seconds, the
performance measures didn't improve and so they can be dismissed.

Figure~\ref{fig:extimes} shows sequential and parallel time and speed-up
computing time for VECM with 100 iterations.

\begin{figure}[!h]
  \centering
  \includegraphics[width=0.7\textwidth]{img/extimes}
  \caption{Computing time of sequential and parallel algorithm is shown in the
  upper figure. Speed-up is shown below.}
  \label{fig:extimes}
\end{figure}

Execution times do not consider the loading data time, just the finding of
best parameters and the building the VECM model. The time MPI spends
transferring data and synchronising process is about two seconds independently
of the number of process considered.

\section{Conclusions}
\label{sec:conclusions}
Cointegration in financial time series has been largely studied and the 
Johansen method is commonly used to obtain it. 
In practice it has been found that cointegration relations change with time. 
However, cointegration model-based such as VECM assumes that cointegration
remains unchanged in time. 
We empirically showed that the Johansen method is sensitive to the number
of lags but also to the amount of data considered.

Moreover, we introduced the notion of {\em percentage of cointegration\/} and
have found that out-of-sample forecast performance is related to the value of
this figure in the last samples.  We used this information to set the model
parameters.  Our proposal AVECM consists then on an adaptive algorithm to update VECM
parameters every time that new data is available. These parameters are found by
maximising the percentage of cointegration of the last samples or iterations.

Determining VECM parameters was the most expensive routine and it was run
using parallel processes using MPI which allowed a grid search within a range of
values for $L$ and $p$ to be made.
Tests were done using real currency rates data. 
Our proposal was tested in several scenarios using different times of
the day to show its behaviour related to the opening times of
financial markets.

Results showed that our proposed AVECM improves performance measures by finding
parameters of $L$ and $p$ maximising the percentage of cointegration.
These facts were consistent for different times of the day. 
We found that increasing the number of parameters will always lead to
better performance measures. Furthermore, in the case of Tokyo 
financial market opening time, MAPE reduced an order from $10^{-3}$ to $10^{-4}$.

The parallel implementation allowed the execution times to be reduced
more than ten times and therefore a response time was obtain before 10
seconds. Since we used 10-second frequency we can say that our proposal is suitable
for use in an online context for real applications because response times
were less than this frequency.

For future study, it would be interesting to explore the relation between
cointegration and performance in order to propose new criteria for
improving VECM parameters.

### 2015-03-11 src/4_OnlineLearning.tex
AVECM

@@
 Moreover, in case we receive data streams, ANN or SVM cannot introduce new information into the model without a re-training process, so we will have to use the same non-updated model until we decide to compute another one if it is possible.
-Online learning algorithms allow one example at a time to be introduced into an existing model incrementally~\cite{vovk2005}.
+Online learning algorithms allow one example at a time to be introduced into an existing model incrementally \cite{vovk2005}.
 This is extremely important when the problem has large data streams and real-time forecasting must be done.
@@
 
-The online learning framework was first introduced in the perceptron algorithm~\cite{rosenblatt58}.
-There are several other widely used online methods such as passive-aggressive~\cite{crammerETall2006}, stochastic gradient descent~\cite{zhang2004}, aggregating algorithm~\cite{vovk2001} and the second order perceptron~\cite{cesa-bianchi2005}.
-In~\cite{cesa-bianchi2006} an in-depth analysis of online learning is provided.
+The online learning framework was first introduced in the perceptron algorithm \cite{rosenblatt58}.
+There are several other widely used online methods such as passive-aggressive \cite{crammerETall2006}, stochastic gradient descent \cite{zhang2004}, aggregating algorithm \cite{vovk2001} and the second order perceptron \cite{cesa-bianchi2005}.
+In \cite{cesa-bianchi2006} an in-depth analysis of online learning is provided.
 

### 2015-03-11 src/5_1_Proposal.tex
AVECM

--- 2015-03-05 src/5_Proposal.tex
@@
 sometimes they exhibit some stationary linear combinations. When this happens,
-it is said that those time series are cointegrated.The Vector Error Correction
-Model (VECM) is an econometric model which characterizes the joint dynamic
+it is said that those time series are cointegrated. The Vector Error Correction
+Model (VECM) is an econometric model which characterises the joint dynamic
 behaviour of a set of cointegrated variables in terms of forces pulling towards
 equilibrium. In this study, we propose an Online VEC model (OVECM) which
-optimizes how model parameters are obtained using a sliding window of the most
+optimises how model parameters are obtained using a sliding window of the most
 recent data. Our proposal also takes advantage of the long-run relationship
@@
 
-Some financial models, such as  vector error correction (VEC), take advantage of
+Some financial models, such as  vector error correction model (VECM), take advantage of
 this property and describe the joint behaviour of several cointegrated
@@
 
-VEC is a special case of the vector autorregressive model (VAR). VAR is a linear
+VECM is a special case of the vector autoregressive model (VAR). VAR is a linear
 model which express future values in terms of its own history and other
-variables past values. If cointegration exists, VEC model allows to introduce an
+variables past values. If cointegration exists, VECM allows to introduce an
 error correction term due to its cointegration estimation error.
 
-VEC as well as VAR model parameters are obtained using ordinary least squares
+VECM as well as VAR model parameters are obtained using ordinary least squares
 method (OLS). Since OLS involves many calculations, the parameter estimation
@@
 
-There are several popular online methods such as
-perceptron~\cite{rosenblatt58}, passive-aggressive~\cite{crammerETall2006},
-stochastic gradient descent~\cite{zhang2004}, aggregating
-algorithm~\cite{vovk2001} and the second order
-perceptron~\cite{cesa-bianchi2005}.  In~\cite{cesa-bianchi2006}, an in-deph
-analysis of online learning is provided.  In particular, the {\em aggregating
-algorithm for regression} (AAR) method is a recursive formulation of RR
-suitable for using in an online context.
-
 We propose an online formulation of the VEC model based on the
@@
 
-\subsection{Online VECM} \label{sec:proposal}
-
 Since VECM parameter estimation is computationally expensive, we propose an
@@
 
-OVECM also implements matrix optimizations in order to reduce execution time,
+OVECM also implements matrix optimisations in order to reduce execution time,
 such as updating matrices with new data, removing old data and introducing new
@@
 \item The function \texttt{vecMatrix} returns the matrices~(\ref{eq:vecA})
-and~(\ref{eq:vecB}) that allows VECM to be solved.
+and~(\ref{eq:vecY}) that allows VECM to be solved.
 \item The function \texttt{vecMatrixOnline} returns the
-matrices~(\ref{eq:vecA}) and~(\ref{eq:vecB}) aggregating new data and removing
+matrices~(\ref{eq:vecA}) and~(\ref{eq:vecY}) aggregating new data and removing
 the old one, avoiding calculation of the matrix $\mathbf{A}$ from scratch.
@@
 	    \STATE{$v \gets \texttt{getJohansen}(\mathbf{y}_i,p)$}
-	    \STATE{$[\mathbf{A} \quad \mathbf{B}] \gets
+	    \STATE{$[\mathbf{A} \quad \mathbf{Y}] \gets
         \texttt{vecMatrix}(\mathbf{y}_i,p,v)$}
 	\ELSE
-	    \STATE{$[\mathbf{A} \quad \mathbf{B}] \gets
-        \texttt{vecMatrixOnline}(\mathbf{y}_i,p,v,\mathbf{A},\mathbf{B})$}
+	    \STATE{$[\mathbf{A} \quad \mathbf{Y}] \gets
+        \texttt{vecMatrixOnline}(\mathbf{y}_i,p,v,\mathbf{A},\mathbf{Y})$}
         \STATE $\Delta \mathbf{Y}_{\text{pred}}[i] \gets \mathbf{A}[-1,:] \times \mathbf{X}$
     \ENDIF
-    \STATE $\mathbf{X} \gets \texttt{OLS} (\mathbf{A},\mathbf{B})$
-    \STATE $e \gets \texttt{mape}(\mathbf{B}[-n,:],\mathbf{A}[-n,:] \times \mathbf{X})$
+    \STATE $\mathbf{X} \gets \texttt{OLS} (\mathbf{A},\mathbf{Y})$
+    \STATE $e \gets \texttt{mape}(\mathbf{Y}[-n,:],\mathbf{A}[-n,:] \times \mathbf{X})$
     \IF {$\texttt{mean}(e) > \text{mean\_error}$}
@@
         \texttt{vecMatrixUpdate}(\mathbf{y}_i,p,v,\mathbf{A})$}
-        \STATE $\mathbf{X} \gets \texttt{OLS} (\mathbf{A},\mathbf{B})$
+        \STATE $\mathbf{X} \gets \texttt{OLS} (\mathbf{A},\mathbf{Y})$
     \ENDIF
@@
 \end{algorithm}
+
+A pseudocode of the algorithm \ref{alg:proposal} is detailed in the appendix \ref{fig:pseudocodigo}.
 
@@
 iterations.
-
-    \begin{tikzpicture}[node distance=2cm]
-    %\node (start) [startstop] {Start};
-    %\node (in1) [io, below of=start] {Input};
-    \node (in1) [io] {Input: y, p, L, $\epsilon$, n};
-    \node (pro1) [process, below of=in1] {$y_i = y[i:i+L]$};
-    \node (dec1) [decision, below of=pro1, yshift=-0.5cm] {i = 0};
-    \node (pro2a) [process, below of=dec1, yshift=-0.5cm] {$[\mathbf{A} \quad \mathbf{B}] =\texttt{ovecm}(\mathbf{y}_i,p,v,\mathbf{A},\mathbf{B})$
-                                                        $\Delta \mathbf{Y}_{\text{true}}[i] = \mathbf{B}[-1,:]$
-                                                        $\Delta \mathbf{Y}_{\text{pred}}[i] = \mathbf{A}[-1,:] \times \mathbf{X}$};
-
-    \node (pro2b) [process, right of=dec1, xshift=4cm] {$v = \texttt{Johansen}(\mathbf{y}_i,p)$ $[\mathbf{A} \quad 
-                                                                \mathbf{B}] = \texttt{vecm}(\mathbf{y}_i,p,v)$};
-    \node (pro3) [process, below of=pro2a, yshift=-0.5cm] {$\mathbf{X} = \text{OLS} (\mathbf{A},\mathbf{B})$ \\
-                                                        $\mathbf{y}_{\text{true}}[i] = \mathbf{y}_i[-n,:]$ \\
-                                                        $\mathbf{y}_{\text{pred}}[i] = \mathbf{A}[-n,:] \times \mathbf{X} +$ \\$ \mathbf{y}_i[-n-1,:-1]$ \\
-                                                        $e = \texttt{mape}(\mathbf{y}_{\text{true}}, \mathbf{y}_{\text{pred}})$};
-
-    \node (dec2) [decision, below of=pro3, yshift=-0.5cm] { $e > \epsilon $ };
-
-    \node (pro4a) [process, right of=dec2, xshift=4cm] {$v = \texttt{Johansen}(\mathbf{y}_i,p)$ \\
-                                                        $\mathbf{A} = \texttt{uvecm}(\mathbf{y}_i,p,v,\mathbf{A})$ \\
-                                                        $\mathbf{X} = \texttt{OLS} (\mathbf{A},\mathbf{B})$};
-
-    \node (pro5) [process, left of=pro2a, xshift=-4cm] {$ i++ $};
-
-    \node (dec3) [decision, below of=dec2, yshift=-0.5cm] { $i > it $ };
-
-    \node (out1) [io, below of=dec3] {Output: $\Delta \mathbf{y}_{true}, \Delta \mathbf{y}_{pred}$};
-
-    %\node (stop) [startstop, below of=out1] {Stop};
-
-    %\draw [arrow] (start) -- (in1);
-    \draw [arrow] (in1) -- (pro1);
-    \draw [arrow] (pro1) -- (dec1);
-    \draw [arrow] (dec1) -- node[anchor=east] {no} (pro2a);
-    \draw [arrow] (dec1) -- node[anchor=south] {si} (pro2b);
-    \draw [arrow] (pro2a) -- (pro3);
-    \draw [arrow] (pro2b) |- (pro3);
-    \draw [arrow] (pro3) -- (dec2);
-    \draw [arrow] (dec2) --  node[anchor=south] {si} (pro4a);
-    \draw [arrow] (pro4a) |- (dec3);
-
-    \draw [arrow] (dec2) --  node[anchor=east] {no} (dec3);
-    \draw [arrow] (dec3.west) -| node[anchor=east] {no} (pro5);
-    \draw [arrow] (pro5) |- (pro1);
-    \draw [arrow] (dec3) --  node[anchor=east] {si} (out1);
-    %\draw [arrow] (out1) -- (stop);
-    \end{tikzpicture}
 

### 2015-03-13 src/1_Introduction.tex
intro fix

@@
 algorithms. This study involves interdisciplinary knowledge such as: finance,
-scientific computing, high performance computing, machine learning, physics
+scientific computing, high performance computing, machine learning
 among others.
 
-Forecasting financial time series have been modeled using classical statistical
-models. More recently , machine learning models have been extensively used in
-forecasting. However, getting model parameters is a computational challenge.
+Forecasting financial time series have been modelled using classical statistical
+approaches. More recently, machine learning models have been extensively used in
+forecasting. However, their main disadvantage is that getting model parameters is a computational challenge.
 The computational complexity of machine learning algorithms has become a
 limiting factor for problems that require processing large volumes of data and
-where the response time is crucial.
+where response time is crucial.
 
-Therefore, algorithms that process large amount of data in a short period of
-time are required.  Online learning algorithms have been developed to solve
+Therefore, algorithms that process large amount of data in a short periods of
+time are required. Recently, online learning algorithms have been developed to solve
 large-scale problems since they process an instance at a time, updating the
 model at each step incrementally. This is opposed to the batch algorithms where
-a phase of training that uses a large collection of historical data is first
-required and the resulting model is later used to predict.
+the forecast model is built using a large collection of historical data in
+a training phase.
 
@@
 times. Cointegration is the main feature studied and it refers that one or more
-linear combinations of these variables are stationary even though individually
-they are not~\cite{engle87}.  Some models, such as the Vector Error Correction
+linear combinations of these time series are stationary even though individually
+they are not~\cite{engle87}.  Some models, such as the Vector Error Correction Model
 (VECM), take advantage of this property and describe the joint behaviour of
@@
 
-In this thesis, we propose an online formulation of the VECM called Onlive VECM
-(OVECM) based on consideration of only a sliding window of the most recent data.
-The algorithm introduces matrix optimizations in order to reduce the number of
+%%
+In this thesis, an online formulation of the VECM called Online VECM
+(OVECM) is proposed. OVECM is based on the consideration of a sliding window of the most recent data.
+The algorithm introduces matrix optimisations in order to reduce the number of
 operations and also takes into account the fact that cointegration vector space
-doesn't experience large changes with small changes in the input data. Moreover,
-RR instead of OLS is used to solve VECM. Our method is later tested using four
-currency rates from the foreign exchange market with different frequencies.  Model
-efectiveness is focused on out-of-sample forecast rather than in-sample fitting.
-This criteria allows the OVECM prediction capability to be expressed rather than
-just explaining data history. Our method performance is compared with its
-optimal offline algorithm.
+doesn't experience large changes with small changes in the input data. Moreover, VECM parameters are obtained using machine learning methods. Our method is later tested using four
+currency rates from the foreign exchange market with different frequencies.  
+On the other hand, in order to improve VECM parameters, an adaptive VECM algorithm is presented called AVECM.
+AVECM allows VECM parameters to be found by maximising the number of cointegration relations for a given set of parameters on a grid search.
+This grid search is done in parallel.
+
+Models effectiveness were focused on the out-of-sample forecast rather than on the in-sample fitting.
+This criteria allows the OVECM and AVECM prediction capability to be expressed rather than just explaining data history.
+Our method performance is compared with its optimal offline algorithm.
 
 \section{Research Objectives}
-The main motivation for this research is the develpment of efficient methods for
+The main motivation for this research is the development of efficient methods for
 forecasting financial time series.
@@
 analysis models including machine learning techniques.}
-\item $\Large \mathcal{O}_2$: \emph{Development of a set of know features of the
+\item $\Large \mathcal{O}_2$: \emph{Development of a set of known features of the
 studied time series and applied them to improve forecasting.}
 \item $\Large \mathcal{O}_3$: \emph{Development of parallel and efficient
-algorithms to ensure quick time responses.}
+algorithms to ensure quick response times .}
 \item $\Large \mathcal{O}_4$: \emph{Deep mathematical analysis of the proposal
@@
 
-Cointegration concept was introduced by~\cite{engle1987} and implies that one or
+Cointegration concept was introduced by Engle and Granger in 1987 \cite{engle1987} and implies that one or
 more linear combinations of non-stationary variables are stationary even though
-individually they are not.  Moreover ~\cite{stock+watson1988} observed that
+individually they are not.  Moreover Stock and Watson in 1988 \cite{stock+watson1988} observed that
 cointegration reflects the common stochastic trends providing a useful way to
@@
 among a set of cointegrated variables as an error correction term. VECM is a
-special case of the vector autorregresive model (VAR) model. VAR model
+special case of the vector autoregresive model (VAR) model. VAR model
 expresses future values as a linear combination of variables past values.
@@
 Ridge Regression (RR) \cite{hoerl1970} which introduces a regularization
-parameter that leads to an unbiased estimation with better generalization
+parameter that leads to an unbiased estimation with better generalisation
 capability. The second problem of computational complexity depends on the number
@@
 \textit{An online learning algorithms based on cointegration and high
-performance computing will allow to obtain better and faster forecasting
-algorithms for financial time series maintaining good accuracy levels.}
+performance computing will allow faster forecasting
+algorithms for financial time series to be obtained while maintaining good accuracy levels.}
 
-\section{Organization of this Thesis}
+\section{Organisation of this Thesis}
 
@@
 time series models such as market hypothesis, frequency, order book generation,
-market microestructure.
+market microstructure.
 Chapter 2 discuss main time series models including classic ones such as ARMA,
@@
 experiments of cointegration are carried out in order to show its dynamic.
-Chapter 6 presents a discussion of found results and summarize the main
-conlusions of this thesis. It also presents some research directions for future
+Chapter 6 presents a discussion of found results and summarise the main
+conclusions of this thesis. It also presents some research directions for future
 studied.

### 2015-04-01 src/3_TimeSeries.tex

@@
 \end{equation}
-%
+
 \noindent where $\mathbf{Y}$ dimension is $((N-p)\times l)$, $\mathbf{A}$
@@
 +1)\times l)$ and $\mathbf{E}$ dimension is $((N-p)\times l)$:
-%
-\begin{alignat}{3}
+
+\begin{align}
+\label{eq:vecY}
 \mathbf{Y}
-&= \begin{bmatrix}
+= \begin{bmatrix}
    \Delta\mathbf{y}_{p+1}^\top \\
@@
    \end{bmatrix}
-&\quad
+\end{align}
+
+\begin{align}
+\label{eq:vecX}
 \mathbf{X}
-&= \begin{bmatrix}
+= \begin{bmatrix}
    \boldsymbol{\alpha}^\top \\
@@
    \end{bmatrix}
-&\quad
+\end{align}
+
+\begin{align}
+\label{eq:vecX}
 \mathbf{E}
-&= \begin{bmatrix}
+= \begin{bmatrix}
    \boldsymbol{\epsilon}_{p+1}^\top \\
@@
    \end{bmatrix}
-\end{alignat}
+\end{align}
+
 \noindent and
+
 \begin{align}
+\label{eq:vecA}
 \mathbf{A}
@@
 \end{align}
+
 Taking into account the error term $\mathbf{E}$, equation~(\ref{eq:vareq}) 

### 2015-04-01 src/6_Conclusion.tex

\chapter{Conclusion}

\section{Conclusions of this thesis}

\section{Contributions of this thesis}

\section{Future Work}

### 2015-04-09 src/5_2_Proposal.tex
ordenando propuesta 2

@@
 \chapter{Adaptive parallel algorithm for VECM}
-
-VECM parameters are those which maximise cointegration relations in the past. However, this search is computationally expensive since the Johansen
-method~\cite{johansen1995} is used, which is of order $O(n^3)$. We propose
-to parallelise this search in order to get new parameters at every iteration.
 
@@
 
-\section{Introduction}
-Both VECM and VAR model parameters are obtained using ordinary least squares
-(OLS) method \cite{golub1980}. Since OLS involves many calculations, the parameter estimation
-method is computationally expensive when the number of past values and
-observations increases. Moreover, obtaining cointegration vectors 
-is also an expensive routine because of the use of the Johansen method.
+VECM parameters are obtained using ordinary least squares (OLS) method
+\cite{golub1980}, which is computationally expensive when the number of past
+values $L$ and observations $p$ increases. Moreover, obtaining cointegration
+vectors is also an expensive routine because of the use of the Johansen method.
 
-Model effectiveness is focused on out-of-sample forecast rather than in-sample
-fitting. This criterion allows our proposal AVECM prediction capability to be
-expressed rather than just explaining data history.
+In this study we found that for some combinations of $L$ and $p$, the number of
+cointegration relations is increased and this leads to better performance. We
+propose to parallelise this search in order to achive faster execution times.
+
+\section{The problem} \label{sec:proposal}
+
+Cointegration vectors can be found applying the Johansen method which uses a
+sample of the most recent historical data. However, one of the main limitations of this
+methodology is that it assumes cointegration vectors do not change in time.  In
+fact, the long-run relationships between the time series might change due to
+several economic factors that can lead to structural breaks in the cointegrating
+relation \cite{gregoryETal1996}.
+
+In this study, we show that cointegration relations depend on the amount $L$ of
+historical data considered, and how many lags $p$ were used. This search is
+computationally expensive, considering that the Johansen
+method~\cite{johansen1995} (which is of order $O(n^3)$) is used and these
+parameters change at every iteration.
+
+Our proposal consists on to parallelise this search in order to get new
+parameters at every iteration.
 
@@
 
-\def\rot#1{{\color{red}{#1}}}
-\def\gruen#1{{\color{green}{#1}}}
-\def\blau#1{{\color{blue}{#1}}}
+\section{Methodology}
+
+The relation between the number of cointegration relations and parameters $L$
+and $p$ is studied and then a relation between cointegration and performance is
+stated.
+
+Figure~\ref{fig:hists} shows the number $r$ of cointegration vectors given by
+the Johansen method for different values of the parameters $L$ and $p$, using
+four currencies data, i.e., $l=4$, and considering 1000 iterations $it$.
+We know that when there is no cointegration vectors, i.e $r=0$, 
+there is no cointegration. In the same way, when $r=l$ reveals that no process is I(1) but,
+instead, they are all stationary.
+The interesting cases of cointegration are those, where $r$ lies strictly
+between $0$ and $l$, i.e. $0<r<l$,
+
+Experiments were run with data starting at 9:00 GMT of the 11th of August 2014,
+which corresponds to the opening of the New York financial market.
+
+\begin{figure}[!h]
+  %\vspace{-0.8cm}
+  \centering
+  \includegraphics[width=\textwidth]{img/histCointVectors}
+  \caption{Histogram of the number of cointegration vectors using $p=1$. Four
+  possible values for $L$ are shown (100, 1900, 6100 and 10700).
+  In the cases of $L=100$ and $L=6100$, most of the time there is little
+  cointegration (0 or 4 cointegration vectors). However, for $L=1900$ and
+  $L=10700$ significant cointegration is found.}
+  \label{fig:hists}
+\end{figure}
+
+\begin{figure}[!h]
+  %\vspace{-0.8cm}
+  \centering
+  \includegraphics[width=\textwidth]{img/histCointVectorsp5}
+  \caption{Histogram of the number of cointegration vectors using $p=5$. 
+   Despite the change in the distribution of cointegration vectors, 
+   the percentage of cointegration is nearly the same as in the case $p=1$. 
+   For example, for $L=10700$ the sum of 1 and 2 number of cointegration
+   vectors is very similar to 1 cointegration vector for the case $L=10700$
+   in figure \ref{fig:hists}.}
+   \label{fig:histsp5}
+\end{figure}
+
+The same experiments, using now $p=5$, were carried out in order to see the
+effect on the number of cointegration vectors found. 
+Figure~\ref{fig:histsp5} shows that, despite the change in the distribution
+of the number of cointegration vectors, the sum of their number with $0<r<4$,
+remain nearly the same. 
+Therefore we can say that the number of lags doesn't significantly affect
+the extent of cointegration found.
+
+In order to measure the extent of cointegration, we introduce a
+{\em percentage of cointegration\/} as following:
+\begin{equation} \label{eq:pcoint}
+PC = 
+\frac{\#\{ it \mid \text{$it$ has $r$ c.v. with $0<r<l$}\}}
+     {\#it}\times 100
+\end{equation}
+where c.v. stands for cointegration vectors and $it$ is the number of iterations.
+
+The goal of our next experiment was to find a relation between this ratio
+$PC$ and the performance of the accuracy measure MAPE (see equation 
+\ref{eq:MAPE}). 
+The experiments suggest that there is indeed a relation between the $PC$ ratio
+and MAPE.
+Figure~\ref{fig:cointvsmape} shows that MAPE rapidly decreases with increasing
+$PC$.
+
+\begin{figure}[!h]
+  %\vspace{-0.8cm}
+  \centering
+  \includegraphics[width=0.8\textwidth]{img/MAPEvsCoint-offset21600-p-2-freq-10s}
+  \caption{MAPE versus percentage of cointegration. MAPE was obtained from 1000
+  iterations considering several possible $L=[100,10000,400]$ and $p=[1,5]$.
+  Best $L$ found was 8900.}
+  \label{fig:cointvsmape}
+\end{figure}
+
+From these results we observe that cointegration changes with different values
+for $L$ and $p$ and also that better cointegration percentage leads to better
+accuracy performance.
+Therefore we propose to choose best $L$ and $p$ parameters determined at every
+step. 
+The criterion for choosing $L$ and $p$ is to maximise the percentage of
+cointegration $PC$ considering a number of past iterations $it$.
+
+Algorithm ~\ref{alg:AVECM} summarises our proposal. 
+The function {\bf get\_best\_params} makes a grid search on the two vector
+lists $Ls$ and $ps$ and returns the parameters $L$ and $p$ which maximise
+the percentage of cointegration $PC$ (see equation~\ref{eq:pcoint}) for a
+pre-defined number of iterations.
+Since this search is computational expensive, the number of cointegration
+vectors at each iteration is computed in parallel, thus ensuring a response
+before new data is available. 
+This parallel routine also allows to refine our grid search leading to obtain
+improved forecasting performance.
+After that optimum $L$ and $p$ parameters are found, VECM is built and used
+to forecast next data.
+
+\begin{algorithm}[ht]
+\begin{algorithmic}[1]
+\REQUIRE $\,$ \\
+$\mathbf{y}$: matrix with $N$ input vectors and $l$ time series\\
+$ps$: vector with the number of past values range \\
+$Ls$: vector with window sizes range ($L<N$) \\
+$outit$: Number of out-of-sample iterations($outit < N-L$)
+$it$: Number of iterations of testing (in-sample)\\
+$of$: Starting point of Testing \\
+\ENSURE  $\,$ \\
+$\{ \mathbf{y}_{\text{pred}}[1],\dots,\mathbf{y}_{\text{pred}}[N-L]\}$: model predictions 
+\FOR { $i =of$ to $of + outit$ }
+   \STATE $\mathbf{Y} \gets \mathbf{y}[i-\texttt{max}(Ls):i]$
+    \STATE $L,p \gets
+    \texttt{get\_best\_params}(Ls,ps,it,Y)$
+    \STATE $\mathbf{y}_i \gets \mathbf{y}[i-L:i]$
+        \STATE $model = VECM(\mathbf{y}_i, p)$
+        \STATE $\mathbf{y}_{\text{pred}}[i] = model.predict()$
+\ENDFOR
+\end{algorithmic}
+\caption{AVECM: Adaptive VECM.}
+\label{alg:AVECM}
+\end{algorithm}
 
@@
 squares estimation.
-
-\section{Motivation} \label{sec:proposal}
-
-Cointegration vectors can be found applying the Johansen method which uses a
-sample of the last historical data. However, one of the main limitations of this
-methodology is that it assumes cointegration vectors do not change in time.
-In fact, the long-run relationships between the time series might change due to
-several economic factors that can lead to structural breaks in the cointegrating
-relation \cite{gregoryETal1996}. 
-Experiments show that cointegration depends on the amount $L$ of historical data
-considered, and how many lags $p$ we use.
-
-Figure~\ref{fig:hists} shows the number $r$ of cointegration vectors given by
-the Johansen method for different values of the parameters $L$ and $p$, using
-four currencies data, i.e., $l=4$, and considering 1000 iterations $it$.
-We know that when there is no cointegration vectors, i.e $r=0$, 
-there is no cointegration. In the same way, when $r=l$ reveals that no process is I(1) but,
-instead, they are all stationary.
-The interesting cases of cointegration are those, where $r$ lies strictly
-between $0$ and $l$, i.e. $0<r<l$,
-
-Experiments were run with data starting at 9:00 GMT of the 11th of August 2014,
-which corresponds to the opening of the New York financial market.
-
-\begin{figure}[!h]
-  %\vspace{-0.8cm}
-  \centering
-  \includegraphics[width=\textwidth]{img/histCointVectors}
-  \caption{Histogram of the number of cointegration vectors using $p=1$. Four
-  possible values for $L$ are shown (100, 1900, 6100 and 10700).
-  In the cases of $L=100$ and $L=6100$, most of the time there is little
-  cointegration (0 or 4 cointegration vectors). However, for $L=1900$ and
-  $L=10700$ significant cointegration is found.}
-  \label{fig:hists}
-\end{figure}
-
-\begin{figure}[!h]
-  %\vspace{-0.8cm}
-  \centering
-  \includegraphics[width=\textwidth]{img/histCointVectorsp5}
-  \caption{Histogram of the number of cointegration vectors using $p=5$. 
-   Despite the change in the distribution of cointegration vectors, 
-   the percentage of cointegration is nearly the same as in the case $p=1$. 
-   For example, for $L=10700$ the sum of 1 and 2 number of cointegration
-   vectors is very similar to 1 cointegration vector for the case $L=10700$
-   in figure \ref{fig:hists}.}
-   \label{fig:histsp5}
-\end{figure}
-
-The same experiments, using now $p=5$, were carried out in order to see the
-effect on the number of cointegration vectors found. 
-Figure~\ref{fig:histsp5} shows that, despite the change in the distribution
-of the number of cointegration vectors, the sum of their number with $0<r<4$,
-remain nearly the same. 
-Therefore we can say that the number of lags doesn't significantly affect
-the extent of cointegration found.
-
-In order to measure the extent of cointegration, we introduce a
-{\em percentage of cointegration\/} as following:
-\begin{equation} \label{eq:pcoint}
-PC = 
-\frac{\#\{ it \mid \text{$it$ has $r$ c.v. with $0<r<l$}\}}
-     {\#it}\times 100
-\end{equation}
-where c.v. stands for cointegration vectors and $it$ is the number of iterations.
-
-The goal of our next experiment was to find a relation between this ratio
-$PC$ and the performance of the accuracy measure MAPE (see equation 
-\ref{eq:MAPE}). 
-The experiments suggest that there is indeed a relation between the $PC$ ratio
-and MAPE.
-Figure~\ref{fig:cointvsmape} shows that MAPE rapidly decreases with increasing
-$PC$.
-
-\begin{figure}[!h]
-  %\vspace{-0.8cm}
-  \centering
-  \includegraphics[width=\textwidth]{img/MAPEvsCoint-offset21600-p-2-freq-10s}
-  \caption{MAPE versus percentage of cointegration. MAPE was obtained from 1000
-  iterations considering several possible $L=[100,10000,400]$ and $p=[1,5]$.
-  Best $L$ found was 8900.}
-  \label{fig:cointvsmape}
-\end{figure}
-
-From these results we observe that cointegration changes with different values
-for $L$ and $p$ and also that better cointegration percentage leads to better
-accuracy performance.
-Therefore we propose to choose best $L$ and $p$ parameters determined at every
-step. 
-The criterion for choosing $L$ and $p$ is to maximise the percentage of
-cointegration $PC$ considering a number of past iterations $it$.
-
-Algorithm ~\ref{alg:AVECM} summarises our proposal. 
-The function {\bf get\_best\_params} makes a grid search on the two vector
-lists $Ls$ and $ps$ and returns the parameters $L$ and $p$ which maximise
-the percentage of cointegration $PC$ (see equation~\ref{eq:pcoint}) for a
-pre-defined number of iterations.
-Since this search is computational expensive, the number of cointegration
-vectors at each iteration is computed in parallel, thus ensuring a response
-before new data is available. 
-This parallel routine also allows to refine our grid search leading to obtain
-improved forecasting performance.
-After that optimum $L$ and $p$ parameters are found, VECM is built and used
-to forecast next data.
-
-\begin{algorithm}[ht]
-\begin{algorithmic}[1]
-\REQUIRE $\,$ \\
-$\mathbf{y}$: matrix with $N$ input vectors and $l$ time series\\
-$ps$: vector with the number of past values range \\
-$Ls$: vector with window sizes range ($L<N$) \\
-$outit$: Number of out-of-sample iterations($outit < N-L$)
-$it$: Number of iterations of testing (in-sample)\\
-$of$: Starting point of Testing \\
-\ENSURE  $\,$ \\
-$\{ \mathbf{y}_{\text{pred}}[1],\dots,\mathbf{y}_{\text{pred}}[N-L]\}$: model predictions 
-\FOR { $i =of$ to $of + outit$ }
-   \STATE $\mathbf{Y} \gets \mathbf{y}[i-\texttt{max}(Ls):i]$
-    \STATE $L,p \gets
-    \texttt{get\_best\_params}(Ls,ps,it,Y)$
-    \STATE $\mathbf{y}_i \gets \mathbf{y}[i-L:i]$
-        \STATE $model = VECM(\mathbf{y}_i, p)$
-        \STATE $\mathbf{y}_{\text{pred}}[i] = model.predict()$
-\ENDFOR
-\end{algorithmic}
-\caption{AVECM: Adaptive VECM.}
-\label{alg:AVECM}
-\end{algorithm}
 

### 2015-04-09 src/5_2_Proposal.tex
propuesta ordenada

@@
 \section{Methodology}
-
-The relation between the number of cointegration relations and parameters $L$
-and $p$ is studied and then a relation between cointegration and performance is
-stated.
-
-Figure~\ref{fig:hists} shows the number $r$ of cointegration vectors given by
-the Johansen method for different values of the parameters $L$ and $p$, using
-four currencies data, i.e., $l=4$, and considering 1000 iterations $it$.
-We know that when there is no cointegration vectors, i.e $r=0$, 
-there is no cointegration. In the same way, when $r=l$ reveals that no process is I(1) but,
-instead, they are all stationary.
-The interesting cases of cointegration are those, where $r$ lies strictly
-between $0$ and $l$, i.e. $0<r<l$,
-
-Experiments were run with data starting at 9:00 GMT of the 11th of August 2014,
-which corresponds to the opening of the New York financial market.
-
-\begin{figure}[!h]
-  %\vspace{-0.8cm}
-  \centering
-  \includegraphics[width=\textwidth]{img/histCointVectors}
-  \caption{Histogram of the number of cointegration vectors using $p=1$. Four
-  possible values for $L$ are shown (100, 1900, 6100 and 10700).
-  In the cases of $L=100$ and $L=6100$, most of the time there is little
-  cointegration (0 or 4 cointegration vectors). However, for $L=1900$ and
-  $L=10700$ significant cointegration is found.}
-  \label{fig:hists}
-\end{figure}
-
-\begin{figure}[!h]
-  %\vspace{-0.8cm}
-  \centering
-  \includegraphics[width=\textwidth]{img/histCointVectorsp5}
-  \caption{Histogram of the number of cointegration vectors using $p=5$. 
-   Despite the change in the distribution of cointegration vectors, 
-   the percentage of cointegration is nearly the same as in the case $p=1$. 
-   For example, for $L=10700$ the sum of 1 and 2 number of cointegration
-   vectors is very similar to 1 cointegration vector for the case $L=10700$
-   in figure \ref{fig:hists}.}
-   \label{fig:histsp5}
-\end{figure}
-
-The same experiments, using now $p=5$, were carried out in order to see the
-effect on the number of cointegration vectors found. 
-Figure~\ref{fig:histsp5} shows that, despite the change in the distribution
-of the number of cointegration vectors, the sum of their number with $0<r<4$,
-remain nearly the same. 
-Therefore we can say that the number of lags doesn't significantly affect
-the extent of cointegration found.
-
-In order to measure the extent of cointegration, we introduce a
-{\em percentage of cointegration\/} as following:
-\begin{equation} \label{eq:pcoint}
-PC = 
-\frac{\#\{ it \mid \text{$it$ has $r$ c.v. with $0<r<l$}\}}
-     {\#it}\times 100
-\end{equation}
-where c.v. stands for cointegration vectors and $it$ is the number of iterations.
-
-The goal of our next experiment was to find a relation between this ratio
-$PC$ and the performance of the accuracy measure MAPE (see equation 
-\ref{eq:MAPE}). 
-The experiments suggest that there is indeed a relation between the $PC$ ratio
-and MAPE.
-Figure~\ref{fig:cointvsmape} shows that MAPE rapidly decreases with increasing
-$PC$.
-
-\begin{figure}[!h]
-  %\vspace{-0.8cm}
-  \centering
-  \includegraphics[width=0.8\textwidth]{img/MAPEvsCoint-offset21600-p-2-freq-10s}
-  \caption{MAPE versus percentage of cointegration. MAPE was obtained from 1000
-  iterations considering several possible $L=[100,10000,400]$ and $p=[1,5]$.
-  Best $L$ found was 8900.}
-  \label{fig:cointvsmape}
-\end{figure}
-
-From these results we observe that cointegration changes with different values
-for $L$ and $p$ and also that better cointegration percentage leads to better
-accuracy performance.
-Therefore we propose to choose best $L$ and $p$ parameters determined at every
-step. 
-The criterion for choosing $L$ and $p$ is to maximise the percentage of
-cointegration $PC$ considering a number of past iterations $it$.
-
-Algorithm ~\ref{alg:AVECM} summarises our proposal. 
-The function {\bf get\_best\_params} makes a grid search on the two vector
-lists $Ls$ and $ps$ and returns the parameters $L$ and $p$ which maximise
-the percentage of cointegration $PC$ (see equation~\ref{eq:pcoint}) for a
-pre-defined number of iterations.
-Since this search is computational expensive, the number of cointegration
-vectors at each iteration is computed in parallel, thus ensuring a response
-before new data is available. 
-This parallel routine also allows to refine our grid search leading to obtain
-improved forecasting performance.
-After that optimum $L$ and $p$ parameters are found, VECM is built and used
-to forecast next data.
-
-\begin{algorithm}[ht]
-\begin{algorithmic}[1]
-\REQUIRE $\,$ \\
-$\mathbf{y}$: matrix with $N$ input vectors and $l$ time series\\
-$ps$: vector with the number of past values range \\
-$Ls$: vector with window sizes range ($L<N$) \\
-$outit$: Number of out-of-sample iterations($outit < N-L$)
-$it$: Number of iterations of testing (in-sample)\\
-$of$: Starting point of Testing \\
-\ENSURE  $\,$ \\
-$\{ \mathbf{y}_{\text{pred}}[1],\dots,\mathbf{y}_{\text{pred}}[N-L]\}$: model predictions 
-\FOR { $i =of$ to $of + outit$ }
-   \STATE $\mathbf{Y} \gets \mathbf{y}[i-\texttt{max}(Ls):i]$
-    \STATE $L,p \gets
-    \texttt{get\_best\_params}(Ls,ps,it,Y)$
-    \STATE $\mathbf{y}_i \gets \mathbf{y}[i-L:i]$
-        \STATE $model = VECM(\mathbf{y}_i, p)$
-        \STATE $\mathbf{y}_{\text{pred}}[i] = model.predict()$
-\ENDFOR
-\end{algorithmic}
-\caption{AVECM: Adaptive VECM.}
-\label{alg:AVECM}
-\end{algorithm}
-
-\section{Background}
-\label{sec:background}
 
@@
 
+\subsection{Motivation}
+The relation between the number of cointegration relations
+and parameters $L$ and $p$ is studied and then a relation between cointegration
+and performance is stated.
+
+Figure~\ref{fig:hists} shows the number $r$ of cointegration vectors given by
+the Johansen method for different values of the parameters $L$ and $p$, using
+four currencies data, i.e., $l=4$, and 1000 considering iterations ($it=1000$).
+\begin{figure}[!h]
+  %\vspace{-0.8cm}
+  \centering
+  \includegraphics[width=\textwidth]{img/histCointVectors}
+  \caption{Histogram of the number of cointegration vectors using $p=1$. Four
+  possible values for $L$ are shown (100, 1900, 6100 and 10700).
+  In the cases of $L=100$ and $L=6100$, most of the time there is little
+  cointegration (0 or 4 cointegration vectors). However, for $L=1900$ and
+  $L=10700$ significant cointegration is found.}
+  \label{fig:hists}
+\end{figure}
+We know that when there is no cointegration vectors, i.e $r=0$, 
+there is no cointegration. In the same way, when $r=l$ reveals that no process is I(1) but,
+instead, they are all stationary.
+The interesting cases of cointegration are those, where $r$ lies strictly
+between $0$ and $l$, i.e. $0<r<l$,
+
+Experiments were run with data starting at 9:00 GMT of the 11th of August 2014,
+which corresponds to the opening of the New York financial market.
+
+\begin{figure}[!h]
+  %\vspace{-0.8cm}
+  \centering
+  \includegraphics[width=\textwidth]{img/histCointVectorsp5}
+  \caption{Histogram of the number of cointegration vectors using $p=5$. 
+   Despite the change in the distribution of cointegration vectors, 
+   the percentage of cointegration is nearly the same as in the case $p=1$. 
+   For example, for $L=10700$ the sum of 1 and 2 number of cointegration
+   vectors is very similar to 1 cointegration vector for the case $L=10700$
+   in figure \ref{fig:hists}.}
+   \label{fig:histsp5}
+\end{figure}
+
+The same experiments, using now $p=5$, were carried out in order to see the
+effect on the number of cointegration vectors found. 
+Figure~\ref{fig:histsp5} shows that, despite the change in the distribution
+of the number of cointegration vectors, the sum of their number with $0<r<4$,
+remain nearly the same. 
+Therefore we can say that the number of lags doesn't significantly affect
+the extent of cointegration found.
+
+In order to measure the extent of cointegration, we introduce a
+{\em percentage of cointegration\/} as following:
+\begin{equation} \label{eq:pcoint}
+PC = 
+\frac{\#\{ it \mid \text{$it$ has $r$ c.v. with $0<r<l$}\}}
+     {\#it}\times 100
+\end{equation}
+where c.v. stands for cointegration vectors and $it$ is the number of iterations.
+
+The goal of our next experiment was to find a relation between this ratio
+$PC$ and the performance of the accuracy measure MAPE (see equation 
+\ref{eq:MAPE}). 
+The experiments suggest that there is indeed a relation between the $PC$ ratio
+and MAPE.
+Figure~\ref{fig:cointvsmape} shows that MAPE rapidly decreases with increasing
+$PC$.
+
+\begin{figure}[!h]
+  %\vspace{-0.8cm}
+  \centering
+  \includegraphics[width=0.7\textwidth]{img/MAPEvsCoint-offset21600-p-2-freq-10s}
+  \caption{MAPE versus percentage of cointegration. MAPE was obtained from 1000
+  iterations considering several possible $L=[100,10000,400]$ and $p=[1,5]$.
+  Best $L$ found was 8900.}
+  \label{fig:cointvsmape}
+\end{figure}
+
+From these results we observe that cointegration changes with different values
+for $L$ and $p$ and also that better cointegration percentage leads to better
+accuracy performance.  Therefore we propose to choose best $L$ and $p$
+parameters determined at every step.  The criterion for choosing $L$ and $p$ is
+to maximise the percentage of cointegration $PC$ considering a number of past
+iterations $it$.
+
+Algorithm ~\ref{alg:AVECM} solves equation \ref{eq:vareq} defining first
+parameters $L$ and $p$ using the function {\bf get\_best\_params} which makes a
+grid search on the two vector lists $Ls$ and $ps$ and returns the parameters $L$
+and $p$ which maximise the percentage of cointegration $PC$ (see
+equation~\ref{eq:pcoint}) for a pre-defined number of iterations.  Since this
+search is computational expensive, the number of cointegration vectors at each
+iteration is computed in parallel, thus ensuring a response before new data is
+available.  This parallel routine also allows to refine our grid search leading
+to obtain improved forecasting performance.  After that optimum $L$ and $p$
+parameters are found, VECM is built and used to forecast next data.
+
+\begin{algorithm}[ht]
+\begin{algorithmic}[1]
+\REQUIRE $\,$ \\
+$\mathbf{y}$: matrix with $N$ input vectors and $l$ time series\\
+$ps$: vector with the number of past values range \\
+$Ls$: vector with window sizes range ($L<N$) \\
+$outit$: Number of out-of-sample iterations($outit < N-L$)
+$it$: Number of iterations of testing (in-sample)\\
+$of$: Starting point of Testing \\
+\ENSURE  $\,$ \\
+$\{ \mathbf{y}_{\text{pred}}[1],\dots,\mathbf{y}_{\text{pred}}[N-L]\}$: model predictions 
+\FOR { $i =of$ to $of + outit$ }
+   \STATE $\mathbf{Y} \gets \mathbf{y}[i-\texttt{max}(Ls):i]$
+    \STATE $L,p \gets
+    \texttt{get\_best\_params}(Ls,ps,it,Y)$
+    \STATE $\mathbf{y}_i \gets \mathbf{y}[i-L:i]$
+        \STATE $model = VECM(\mathbf{y}_i, p)$
+        \STATE $\mathbf{y}_{\text{pred}}[i] = model.predict()$
+\ENDFOR
+\end{algorithmic}
+\caption{AVECM: Adaptive VECM.}
+\label{alg:AVECM}
+\end{algorithm}
 %

### 2015-04-10 src/5_2_Proposal.tex
Agregar análisis de la frequencia de las series

@@
 Figure~\ref{fig:hists} shows the number $r$ of cointegration vectors given by
-the Johansen method for different values of the parameters $L$ and $p$, using
-four currencies data, i.e., $l=4$, and 1000 considering iterations ($it=1000$).
+the Johansen method for different values of the parameters $L$ leaving $p$ set
+as $1$, using four currencies data, i.e., $l=4$, and 1000 considering iterations
+($it=1000$).
 \begin{figure}[!h]
@@
 \end{figure}
-We know that when there is no cointegration vectors, i.e $r=0$, 
-there is no cointegration. In the same way, when $r=l$ reveals that no process is I(1) but,
-instead, they are all stationary.
-The interesting cases of cointegration are those, where $r$ lies strictly
-between $0$ and $l$, i.e. $0<r<l$,
+We know that when no cointegration vectors can be found, i.e $r=0$, implies that
+the variables are not cointegrated. In the same way, when $r=l$ reveals that
+none of the processes is I(1) but, they are all stationary.  The interesting cases of
+cointegration are those, where $r$ lies strictly between $0$ and $l$, i.e.
+$0<r<l$,
 
@@
 
-The same experiments, using now $p=5$, were carried out in order to see the
+The same experiments, setting $p=5$, were carried out in order to see the
 effect on the number of cointegration vectors found. 
 Figure~\ref{fig:histsp5} shows that, despite the change in the distribution
-of the number of cointegration vectors, the sum of their number with $0<r<4$,
-remain nearly the same. 
-Therefore we can say that the number of lags doesn't significantly affect
+of the number of cointegration vectors, the sum of the cases where $0<r<4$,
+remains nearly the same. 
+Therefore, we can say that the number of lags doesn't significantly affect
 the extent of cointegration found.

### 2015-04-15 src/3_TimeSeries.tex
agregando tercera propuesta

@@
 explanatory variables (relationships with other variables, seasonal factors,
-etc.), signal processing, forecasting (predict future values). This chapter
-reviews the more relevant techniques in the rich and rapidly growing field of
-time series analysis.
+etc.), signal processing, forecasting (predict future values).  There are two
+main approaches to study financial time series: to study directions of financial
+rates and to explain financial rate volatility.  This chapter reviews the more
+relevant techniques in the rich and rapidly growing field of time series
+analysis considering these two approaches.
 

### 2015-04-15 src/4_OnlineLearning.tex
agregando tercera propuesta

@@
 
-Online learning is a supervised machine learning framework that is useful when we have sequential access to a sample only once.
-This differs from the classical batch learning, where there is an entire data set available and the learner can build the internal model without any limits in accessing the data.
-In batch learning, there is time enough to carefully analyse the dataset, build large predictive models and combine them in a sophisticated way.
-
-In only learning the goal is the same as batch learning, predicting targets as accurate as possible.
-For example, stock market prediction can be seen as online learning.
-The algorithm makes a prediction of the stock, little time after the real stock price is available, this information can be incorporated to the learner to further improve the prediction accuracy.
-In general, there is too much data available in an online learning setup, the data set grows continuously.
-Offline learning has equal or superior accuracy compared to online learning when the same amount of data is used.
-
-Classic statistical theory of sequential prediction enforces strong assumptions on the statistical properties of the input sequence (for example, stationary stochastic process).
-However, these assumptions can be unknown or change over time.
-In online learning there is no previous assumption about the data and the sequence is allowed to be deterministic, stochastic or even adaptive.
-
-Moreover, in case we receive data streams, ANN or SVM cannot introduce new information into the model without a re-training process, so we will have to use the same non-updated model until we decide to compute another one if it is possible.
-Online learning algorithms allow one example at a time to be introduced into an existing model incrementally \cite{vovk2005}.
-This is extremely important when the problem has large data streams and real-time forecasting must be done.
-This is the most common scenario when we want to forecast a wide range of data such as stock prices and volatilities, electricity power, intrusion detection, web-mining, server load, etc.
-Besides, many problems of high interest in machine learning can be treated as online ones and they can also use these types of algorithms.
-
-The online learning framework was first introduced in the perceptron algorithm \cite{rosenblatt58}.
-There are several other widely used online methods such as passive-aggressive \cite{crammerETall2006}, stochastic gradient descent \cite{zhang2004}, aggregating algorithm \cite{vovk2001} and the second order perceptron \cite{cesa-bianchi2005}.
-In \cite{cesa-bianchi2006} an in-depth analysis of online learning is provided.
-
-The motivation for online learning is to obtain computational efficiency and tackle the shifting problem, i.e.
-that the distribution of the data is unknown or changes over time.
-Online learning algorithms can deal with this problem because they have a tracking ability which is a strategy based on retaining weak dependence on past examples by using two types of models:
-
-\textit{a)} \textbf{memory boundedness:} consists of limiting the number of support vectors in order to improve computational efficiency.
-One example of this is the budget perceptron~\cite{crammeretal2004} which reduces the number of examples used for prediction.
-Alternatively, in the forgetron algorithm~\cite{dekeletal2008} the damage caused by removing old examples is discussed, which can be avoided by removing samples with small influences.
-Other examples are the sliding window kernel (RLS)~\cite{vanvaerenberghetal2006}, which only considers a sliding window of the most recent data, and in \cite{arce+salinas2012} is shown a variant of aggregating algorithm for regression~\cite{vovk2001} considering only a sliding window of the most recent data, optimising also common matrix operations.
-
-\textit{b)} \textbf{weight decay:} one example of this is the shifting perceptron algorithm which implements an exponential decaying scheme for the examples~\cite{cavallantietal2007}.
-Performance of an online learning algorithm is measured by the cumulative loss it suffers along its run on a sequence of examples.
-In order to minimise this loss, the learner may update the hypothesis after each round so as to be more accurate in later rounds.
+Online learning is a supervised machine learning framework that is useful when
+we have sequential access to a sample only once.  This differs from the
+classical batch learning, where there is an entire data set available and the
+learner can build the internal model without any limits in accessing the data.
+In batch learning, there is time enough to carefully analyse the dataset, build
+large predictive models and combine them in a sophisticated way. However, when
+execution time is critical and new data is required to be included in the model,
+online learning algorithms are a better option.
+
+In batch learning, the training phase is commonly a computationally expensive
+process. Therefore, when new data arrives it can't be included easily to the
+model. Moreover, it could happen that we won't have enought time to process the
+new data before more data arrives. In online learning algorithms there is no
+training phase, but the model is updated and evaluated at every time step. This
+model updating is computationally less expensive than a training phase.
+Online algorithms allow incremental learning by processing one instance at a
+time. This is done updating the current model instead of building the model from
+scratch.
+
+Online learning is also useful when some past data may be irrelevant
+or we want to improve computational efficiency. However, the use of less
+historical data could affect accuracy.
+
+The goal of online learning is the same as batch learning, predicting targets as
+accurate as possible. For example, stock market prediction can be seen as online
+learning. The algorithm makes a prediction of the stock, little time after the
+real stock price is available, this information can be incorporated to the
+learner to further improve the prediction accuracy. In general, there is too
+much data available in an online learning setup, the data set grows
+continuously. Offline learning has equal or superior accuracy compared to online
+learning when the same amount of data is used.
+
+Classic statistical theory of sequential prediction enforces strong assumptions
+on the statistical properties of the input sequence (for example, stationary
+stochastic process). However, these assumptions can be unknown or change over
+time. In online learning there is no previous assumption about the data and the
+sequence is allowed to be deterministic, stochastic or even adaptive.
+
+Moreover, in case we receive data streams, ANN or SVM cannot introduce new
+information into the model without a re-training process, so we will have to use
+the same non-updated model until we decide to compute another one if it is
+possible.  Online learning algorithms allow one example at a time to be
+introduced into an existing model incrementally \cite{vovk2005}. This is
+extremely important when the problem has large data streams and real-time
+forecasting must be done.  This is the most common scenario when we want to
+forecast a wide range of data such as stock prices and volatilities, electricity
+power, intrusion detection, web-mining, server load, etc.  Besides, many
+problems of high interest in machine learning can be treated as online ones and
+they can also use these types of algorithms.
+
+The online learning framework was first introduced in the perceptron algorithm
+\cite{rosenblatt58}. There are several other widely used online methods such as
+passive-aggressive \cite{crammerETall2006}, stochastic gradient descent
+\cite{zhang2004}, aggregating algorithm \cite{vovk2001} and the second order
+perceptron \cite{cesa-bianchi2005}.  In \cite{cesa-bianchi2006} an in-depth
+analysis of online learning is provided.
+
+The motivation for online learning is to obtain computational efficiency and
+tackle the shifting problem, i.e. that the distribution of the data is unknown
+or changes over time. Online learning algorithms can deal with this problem
+because they have a tracking ability which is a strategy based on retaining weak
+dependence on past examples by using two types of models:
+
+\textit{a)} \textbf{memory boundedness:} consists of limiting the number of
+support vectors in order to improve computational efficiency. One example of
+this is the budget perceptron~\cite{crammeretal2004} which reduces the number of
+examples used for prediction. Alternatively, in the forgetron
+algorithm~\cite{dekeletal2008} the damage caused by removing old examples is
+discussed, which can be avoided by removing samples with small influences. Other
+examples are the sliding window kernel (RLS)~\cite{vanvaerenberghetal2006},
+which only considers a sliding window of the most recent data, and in
+\cite{arce+salinas2012} is shown a variant of aggregating algorithm for
+regression~\cite{vovk2001} considering only a sliding window of the most recent
+data, optimising also common matrix operations.
+
+\textit{b)} \textbf{weight decay:} one example of this is the shifting
+perceptron algorithm which implements an exponential decaying scheme for the
+examples~\cite{cavallantietal2007}.  Performance of an online learning algorithm
+is measured by the cumulative loss it suffers along its run on a sequence of
+examples. In order to minimise this loss, the learner may update the hypothesis
+after each round so as to be more accurate in later rounds.
+
+There is sometimes confusion about online and incremental learning concepts.
+Incremental learning refers to any online learning process that learns the same
+model as would be learnt by a batch learning algorithm.
+
+Incremental learning is useful when the input to a learning process is stream
+data, with the need or desire to be able to use the result of learning at any
+point in time, based on the input observations received so far.
+
+In principle, the stream of observations may be infinitely long, or the next
+observation long delayed, precluding any hope of waiting until all the
+observations have been received. One would rather not simply accumulate and
+store all the inputs and, upon receipt of each new one, apply a batch learning
+algorithm to the entire sequence of inputs received so far. It would be
+preferable compu- tationally if the existing hypothesis or other artifact of
+learning could be updated in response to each newly received input observation.
+
+Incremental learning is very useful when there is no need to record fundamental
+data and only a summary needs to be retained. Due to this, incremental
+algorithms are often characterized as memoryless, because no memory of past data
+is required.  The algorithm is online but not incremental if it doesn't produce
+the same result for all observations that the corresponding batch algorithm
+would for these same observations.
+
+Algorithm~\ref{alg:onlinealg} shows the online learning algorithm structure:
+
+\begin{algorithm}[ht]
+\begin{algorithmic}[1]
+    \STATE Receives input $\mathbf{x}_t$
+    \STATE Makes prediction $\mathbf{\hat{y}}_t$
+    \STATE Receives response $\mathbf{y}_t$
+    \STATE Incurs loss $l_t(\mathbf{y}_t,\mathbf{\hat{y}}_t)$
+\end{algorithmic}
+\caption{Structure of a Learning System}
+\label{alg:onlinealg}
+\end{algorithm}
+
+\noindent where $l$ is some loss function. Performance is later measured after
+$T$ trials as:
+
+\begin{equation*}
+L_T = \sum_{t=1}^T l_t(\mathbf{y}_t,\mathbf{\hat{y}}_t)
+\end{equation*}
+
+The objective is minimize this loss function for all instances.
+The quality of online learning algorithms is measured by a quantity known as
+regret which is the difference between the performance of the online algorithm
+and its optimal predictor $E^* \in \Theta$ given by:
+
+\begin{equation*}
+L^*_T= \text{min}_{E \in \Theta} L_T^E \, ,
+\end{equation*}
+
+\noindent where $L_T^E = \sum_{t=1}^T
+l_t(\mathbf{y}_t,\mathbf{y}^E_t)$ and $\mathbf{y}^E_t$ is the expert estimation.
+
+Therefore regret is defined as:
+
+\begin{equation*}
+R_T = L_T - L^*_T
+\end{equation*}
 

### 2015-04-15 src/5_3_Proposal.tex
agregando tercera propuesta

\chapter{Real-time intraday Forex rates forecasting using a regularized Vector Error Correction model}

### 2015-04-16 src/5_2_Proposal.tex
mejorando propuestas

@@
-\chapter{Adaptive parallel algorithm for VECM}
+\chapter{Adaptive parallel algorithm for Vector error correction model}
 
@@
 
-VECM parameters are obtained using ordinary least squares (OLS) method
-\cite{golub1980}, which is computationally expensive when the number of past
-values $L$ and observations $p$ increases. Moreover, obtaining cointegration
-vectors is also an expensive routine because of the use of the Johansen method.
+Vector error correction model (VECM) parameters are obtained using ordinary
+least squares (OLS) method \cite{golub1980}, which is computationally expensive
+when the number of past values $L$ and observations $p$ increases. Moreover,
+obtaining cointegration vectors is also an expensive routine because of the use
+of the Johansen method.
 
@@
 
-The forecast capability of our method was measured using MAPE, MAE and
-RMSE \cite{armstrong1992}. Tests were run using four currency rates: Euro (EUR) to
-United States Dollar (USD) (EURUSD), British Pound (GBP) to USD (GBPUSD), USD to Swiss
-Franc (CHF) (USDCHF) and USD to Japanese Yen (JPY) (USDJPY) with a 10-seconds frequency.
+In finance, little academic reasearch has been done using high-frequency Forex
+rates \cite{Bekiros2015}. In our study, tests were run using four currency
+rates: Euro (EUR) to United States Dollar (USD) (EURUSD), British Pound (GBP) to
+USD (GBPUSD), USD to Swiss Franc (CHF) (USDCHF) and USD to Japanese Yen (JPY)
+(USDJPY) with a 10-seconds frequency.  The forecast capability of our method was
+later measured using MAPE, MAE and RMSE \cite{armstrong1992}.
 
@@
 
-VECM is obtained re-writing equation (\ref{eq:var}) in terms of the new
-variable $\Delta\mathbf{y}_t=\mathbf{y}_t-\mathbf{y}_{t-1}$.
+VECM is obtained re-writing the VAR model, shown in equation (\ref{eq:var}), in
+terms of differences defines as: $\Delta\mathbf{y}_t=\mathbf{y}_t-\mathbf{y}_{t-1}$.
 The VECM model, expressed in terms those differences, takes the form:
@@
 \end{algorithm}
-%
-%
-%\subsection{Evaluation methods} \label{sec:evaluation}
-%
-%Forecast performance was evaluated using different methods. We chose three
-%measures that are frequently used:
-%\begin{description}
-%\item
-%{\bf MAPE}, Mean Average Percent Error, which presents forecast errors as a
-%percentage:
-%\begin{equation}\label{eq:MAPE}
-%\text{MAPE} = \frac{1}{N} \sum_{t=1}^{N} 
-%\frac{\left|\mathbf{y}_t-\hat{\mathbf{y}}_t\right|}{\left|\mathbf{y}_t\right|}
-% \times 100 
-%\end{equation}
-%\item
-%{\bf MAE}, Mean Average Error, which measures the distance between forecasts and the
-%true value.
-%\begin{equation}\label{eq:MAE}
-%\text{MAE} = \frac{1}{N} \sum_{t=1}^{N} 
-%\left| 
-%\mathbf{y}_t-\hat{\mathbf{y}}_t
-%\right| 
-%\end{equation}
-%\item
-%{\bf RMSE}, Root Mean Square Error, also measures the distance between forecasts
-%and the true values but, unlike MAE, large deviations from the true value have a
-%large impact on RMSE due to squaring forecast error.
-%\begin{equation}\label{eq:RMSE}
-%\text{RMSE} = \sqrt{
-%\frac{\displaystyle \sum_{t=1}^{N} (\mathbf{y}_t-\hat{\mathbf{y}}_t)^2}{N}}
-%\end{equation}
-%\end{description}
-%
-%
-%
+
 \section{Experimental results}

### 2015-04-16 src/5_3_Proposal.tex
mejorando propuestas

@@
-\chapter{Real-time intraday Forex rates forecasting using a regularized Vector Error Correction model}
+\chapter{Real-time intraday Forex rates forecasting using a regularized Vector
+Error Correction model}
+
+Vector error correction model (VECM) parameters are commonly obtained using the
+ordinary least squares method. Since this parameters could be sensitive to the
+data, this proposal is to get these parameters using ridge regression which
+could lead to a better generalization capability. In order to give with a
+real-time response, an online version of VECM (OVECM) is proposed. OVECM solves
+VECM considering only a sliding window of historical data and therefore
+better response times could be achieved.
+
+\section{The problem}
+
+VECM introduces the long-run relationship among a set of cointegrated variables
+as an error correction term. VECM is a special case of the vector autorregresive
+model (VAR) model. VAR model expresses future values as a linear combination of
+variables past values.  However, VAR model cannot be used with non-stationary
+variables, which is a common feature of financial assets. VECM is a linear model
+but in terms of variable differences. If cointegration exists, variable
+differences are stationary and they introduce an error correction term which
+adjusts coefficients to bring the variables back to equilibrium. In finance,
+many economic time series turn to be stationary when they are differentiated and
+cointegration restrictions often improves forecasting~\cite{duy1998}. Therefore,
+VECM has been widely adopted.
+
+Both VECM and VAR model parameters are obtained using ordinary least squares
+(OLS) method. OLS has two main problems: is sensitive to errors on input data
+and involves many calculations. The former problem is commonly solved using
+Ridge Regression (RR) \cite{hoerl1970} which introduces a regularization
+parameter that leads to an unbiased estimation with better generalization
+capability. The second problem of computational complexity depends on the number
+of past values and observations considered.  Recently, online learning
+algorithms have been proposed to solve problems with large data sets because of
+their simplicity and their ability to update the model when new data is
+available.
+
+Our proposal is an online formulation of the VECM called Online VECM (OVECM)
+based on consideration of only a sliding window of the historical data.  OVECM
+introduces matrix optimizations in order to reduce the number of operations and
+also takes into account the fact that cointegration vector space doesn't
+experience large changes with small changes in the input data. Moreover, OVECM
+uses RR instead of OLS to obtain VECM parameters. Our method is later tested using four
+currency rates from the foreign exchange market with a frequency of 10 seconds.
+Model efectiveness is focused on out-of-sample forecast rather than in-sample
+fitting.  This criteria allows the OVECM prediction capability to be expressed
+rather than just explaining data history. Our method performance is compared
+with its optimal offline algorithm.
+
+%The next sections are organized as follows: section~\ref{sec:background}
+%presents the VAR and VECM, the OVECM algorithm proposed is presented in
+%section~\ref{sec:methodology}. Section~\ref{sec:results} gives a description of
+%the data used and the tests carried on to show accuracy and time comparison of
+%our proposal against the traditional VECM and
+%section~\ref{sec:conclusions} includes conclusions and a proposal for future
+%tudy.
+
+\section{\uppercase{Methodology}}
+\label{sec:methodology}
+
+Since financial problems are stream data problems, it is unfeasible to include
+all input data in a VECM model. Our proposal consists on an online version of
+VECM (OVECM) capable of updating the model with new arrival data and give
+responses in a short period of time. OVECM only considers a time varying window
+of historical data and the using of RR or its variant AAR (Aggregating Algorithm
+for Regression) as an alternative of OLS to get model parameters.
+
+On the other hand, getting cointegration vectors using Johansen method is also
+an expensive procedure. However, since cointegration
+vectors represent the long-run relationship between the time series, they
+vary little in time. Our proposal obtains new cointegration vectors only when
+this long-term relationship changes. This change is detected by tracking 
+the Mean Absolute Percentage Error (MAPE) of the last $n$ in-sample forecasts.
+
+%Since VECM is a model based on time series differences, the MAPE is obtained
+%from $\Delta \mathbf{y}$ as following:
+% 
+%\begin{equation}\label{eq:MAPE}
+%\text{MAPE}[t] = \frac{1}{n} \sum_{i=1}^{n} \left| 
+%\frac{\Delta \mathbf{y}_{\text{true}}[t-i]-\Delta
+%\mathbf{y}_{\text{pred}}[t-i]}{\Delta \mathbf{y}_{\text{true}[t-i]}}
+%\right| \, , 
+%\end{equation}
+%
+%\noindent where $\Delta \mathbf{y}_{\text{true}}$ is the actual value of the
+%time series $\mathbf{y}$ differences and $\Delta \mathbf{y}_{\text{pred}}$ is
+%their forecast value.
+
+%The method proposed consists of
+%a modification of RR considering a sliding window which contains only
+%the last $L$ samples %and the new input $\mathbf{x}_t$, 
+%i.e. $\{\mathbf{x}_i\}_{i=t-L+1}^{t-1}$.
+
+In order to formulate this algorithm we need to define first the
+following matrices which considers only the last $L$ instances:
+
+\begin{equation}
+\label{eq:notation}
+	\mathbf{A}(t) = 
+\left[
+  \begin{tabular}{c>{$}c<{$}c}
+    --- & \mathbf{a}^{\top}_{t-L} & ---\\
+    --- & \mathbf{a}^{\top}_{t-L+1} & ---\\
+    & \vdots & \\
+    --- & \mathbf{a}^{\top}_{t} & ---
+  \end{tabular}
+\right]
+\quad \text{and} \quad
+\mathbf{B}(t) =
+\left[
+  \begin{tabular}{c>{$}c<{$}c}
+    --- & \mathbf{b}_{t-L} & ---\\
+    --- & \mathbf{b}_{t-L+1} & ---\\
+    & \vdots & \\
+    --- & \mathbf{b}_{t} & ---
+  \end{tabular}
+\right] \, ,
+\
+\end{equation}
+
+The RR solution $\mathbf{X}(t)$ using the sliding window matrices defined above
+is:
+
+\begin{equation}
+\label{eq:optsolSLAAR}
+%\mathbf{X}(t)&=&()^{-1}{\bf A}(t)^\top{\bf B}(t) \\
+\mathbf{X}(t)=\mathbf{S}(t)^{-1} \mathbf{W}(t) \, ,
+\end{equation}
+
+\noindent where $\mathbf{S}{t}$ and $\mathbf{W}{t}$ are define as:
+
+\begin{eqnarray*}
+\mathbf{S}(t) =&{\bf A}(t)^\top{\bf A}(t)+ \gamma \mathbb{I} &=
+\sum_{i=0}^L \mathbf{a}_{t-i}\mathbf{a}_{t-i}^\top + \gamma \mathbb{I}
+\label{eq:S} \\
+\mathbf{W}(t) =& {\bf A}(t)^\top{\bf B}(t) &= 
+\sum_{i=0}^L \mathbf{a}_{t-i}\mathbf{b}_{t-i} 
+\label{eq:W}
+\end{eqnarray*}
+
+It is worth noticing that, at the next time step, the matrices $\mathbf{S}(t+1)$
+and $\mathbf{W}(t+1)$ are slightly different to $\mathbf{S}(t)$ and
+$\mathbf{W}(t)$:
+
+\begin{eqnarray*}
+\mathbf{S}(t+1)&=&
+\mathbf{S}(t) +
+\mathbf{a}_{t+1}
+\mathbf{a}_{t+1}^\top -
+\mathbf{a}_{t-L} \mathbf{a}_{t-L}^\top \\
+\mathbf{W}(t+1)&=&
+\mathbf{W}(t) +
+\mathbf{a}_{t+1}
+\mathbf{b}_{t+1} -
+\mathbf{a}_{t-L} \mathbf{b}_{t-L} \, .
+\end{eqnarray*}
+
+Therefore Sherman-Morrison-Woodbury formula can be used to reduce inverse matrix
+calculations.
+
+The algorithm~\ref{alg:proposal} shows our online VECM using three different
+methods for geting model parameters: OLS, RR, AAR.
+
+\begin{algorithm}[ht]
+\begin{algorithmic}[1]
+\REQUIRE $\,$ \\
+$\mathbf{y}$: matrix with $N$ input vectors and $l$ time series\\
+$p$: number of past values \\
+method: \{'OLS','RR','AAR'\} \\
+$\lambda$: regularization parameter (only requires for RR and AAR methods) \\
+$L$: window size of data($L<N$) \\
+$\text{mean\_error}$: MAPE threshold \\
+$n$: windows size to obtain in-sample MAPE \\
+\ENSURE  $\,$ \\
+$\{\mathbf{y}_{\text{pred}}[L+1],\dots, \mathbf{y}_{\text{pred}}[N]\}$: model predictions 
+\STATE solver = new \texttt{Solver}(method,$\lambda$) \\
+\FOR { $t =1$ to $N-L$ }
+    \STATE $\mathbf{Y} \gets \mathbf{y}[t:t+L-1]$
+    \STATE $\mathbf{y}_t \gets \mathbf{y}[t+L]$
+	\IF {$t = 1$}
+	    \STATE{$v \gets \texttt{getJohansen}(\mathbf{Y},p)$}
+	    \STATE{$[\mathbf{A} \quad \mathbf{B}] \gets
+        \texttt{vecMatrix}(\mathbf{Y},p,v)$}
+        \STATE \texttt{solver.init\_model($\mathbf{A},\mathbf{B}$)} 
+    \ENDIF
+	\STATE{$[\mathbf{a}_t \quad \mathbf{b}_t] \gets
+    \texttt{vecMatrixOnline}(\mathbf{Y},\mathbf{y}_t,p,v)$}
+    \STATE $[\mathbf{X} \quad \mathbf{y}_{\text{pred}}[t]] \gets \texttt{solver.regression}
+    (\mathbf{A},\mathbf{B},\mathbf{a}_t,\mathbf{b}_t)$
+    \STATE $\mathbf{Y}_{\text{pred}} = \mathbf{AX}[-n:]+\mathbf{Y}[-n-1:-1]$
+    \STATE $\mathbf{e} = \texttt{mape}(\mathbf{Y}_{\text{pred}},\mathbf{Y}[-n:])$
+    \IF {$\texttt{mean}(\mathbf{e}) > \text{mean\_error}$}
+	    \STATE{$v \gets \texttt{getJohansen}(\mathbf{Y},p)$}
+	    \STATE{$\mathbf{A} \gets
+        \texttt{vecMatrixUpdate}(\mathbf{A},\mathbf{Y},p,v)$}
+        \STATE \texttt{solver.init\_model($\mathbf{A},\mathbf{B}$)} 
+	    \STATE{$[\mathbf{a}_t \quad \mathbf{b}_t] \gets
+        \texttt{vecMatrixOnline}(\mathbf{Y},\mathbf{y}_t,p,v)$}
+        \STATE $[\mathbf{X} \quad \mathbf{y}_{\text{pred}}[t]] \gets \texttt{solver.regression}
+        (\mathbf{A},\mathbf{B},\mathbf{a}_t,\mathbf{b}_t)$
+    \ENDIF
+\STATE{$[\mathbf{A} \quad \mathbf{B}] \gets
+\texttt{matrixUpdate}(\mathbf{A},\mathbf{B},\mathbf{a}_t,\mathbf{b}t)$}
+\ENDFOR
+\end{algorithmic}
+\caption{OVECM: Online VECM}
+\label{alg:proposal}
+\end{algorithm}
+
+Our proposal considers the following:
+
+\begin{itemize}
+\item The function \texttt{getJohansen} returns cointegration vectors given by the
+Johansen method considering the trace statistic test at 95\% level of
+significance. This procedure is called at the first step of the algorithm and
+every time in-sample MAPE ($\mathbf{e}$) exceeds threshold error defined (mean\_error).
+\item The function \texttt{vecMatrix} returns VECM
+matrices $\mathbf{A},\mathbf{B}$ shown in equation~(\ref{eq:vecmatrix}). This
+method is only required at the first step.
+\item The function \texttt{vecMatrixOnline} returns new rows of $\mathbf{A}$ and
+$\mathbf{B}$ give new input data $\mathbf{y}_t$.
+\item solver (see algorithm \ref{alg:solver}) is an object which allows online
+calculation saving matrices $\mathbf{S}$ and $\mathbf{W}$ (see
+equations~\ref{eq:S} and \ref{eq:W}) for methods RR and AAR. This method is
+initialized in the first step and only if new cointegration are required. This
+method also returns in-sample MAPE saved in $\mathbf{e}$. 
+\item The function \texttt{vecMatrixUpdate} updates matrix $\mathbf{A}$ when new
+cointegration vectors are required (matrix $\mathbf{B}$ is not affected). Model
+parameters $\mathbf{X}$ is also updated later.
+\item Out-of-sample outputs are saved in the variables $\Delta
+\mathbf{Y}_{\text{true}}$ and $\Delta \mathbf{Y}_{\text{pred}}$ which are used
+to obtain Out-of-sample MAPE at the end.
+%\item The number of cointegration vectors is set as a parameter (aun no lo
+%coloco en el algoritmo).
+%\item In order to set VECM parameter: $L$ and $p$, we use the Akaike Information
+%Criterion (AIC). RR parameter $\lambda$ was done by cross-validation.
+%\item Obtain VECM parameter $\mathbf{X}(t)$ in equation \ref{eq:optsolSLAAR} requires
+%to calculate the inverse of matrix $\mathbf{S}(t+1)$ which is an expensive
+%routine.However, since we already know $\mathbf{S}(t)^{-1}$ we can use
+%Sherman-Morrison-Woodbury twice~\ref{eq:SMW}.
+\end{itemize}
+
+\begin{algorithm}[ht]
+\begin{algorithmic}[1]
+\REQUIRE $\,$ \\
+method: \{'OLS','RR','AAR'\} \\
+$\mathbf{A}$: VECM design matrix \\
+$\mathbf{B}$: VECM dependant variables \\
+$\mathbf{a}_t$: new row of $\mathbf{A}$ \\
+$\mathbf{b}_t$: new row of $\mathbf{B}$ \\
+$\gamma$: regularization parameter \\
+$n$: windows size to obtain in-sample MAPE \\
+\ENSURE  $\,$ \\
+$\mathbf{X}$: regression solution \\
+$\mathbf{e}$: in-sample MAPE \\
+%\quad \\
+%\texttt{init}($L,\lambda$)
+%\STATE self.L = L
+%\STATE self.lambda = $\lambda$
+\quad \\
+\texttt{init\_model}($\mathbf{A},\mathbf{B}$)\\
+\IF {method == 'RR' or method == 'AAR'}
+\STATE $ [m \quad n] = \text{size}(\mathbf{A}) $ 
+\STATE $\mathbf{S} = \displaystyle \sum_{i=1}^m \mathbf{a}_i \mathbf{a}_i^\top + \gamma \mathbb{I}$
+\STATE $\mathbf{W} = \displaystyle \sum_{i=1}^m \mathbf{a}_i \mathbf{b}_i$
+\ENDIF \\
+\quad \\
+\texttt{regression}($\mathbf{A},\mathbf{B},\mathbf{a_t},\mathbf{b_t},n$) \\
+\IF {method == 'OLS'}
+        \STATE $\mathbf{X}=(\mathbf{A}^\top \mathbf{A})^{-1}\mathbf{A}^\top
+        \mathbf{B}$
+\ELSIF {method == 'RR'}
+        \STATE $\mathbf{X} = \mathbf{S}^{-1} \mathbf{W} $
+        \STATE $\mathbf{S} = \mathbf{S}+
+        \mathbf{a}_t \mathbf{a}_t^\top-
+        \mathbf{a}_1 \mathbf{a}_1^\top$
+        \STATE $\mathbf{W} = \mathbf{W} + \mathbf{a}_t \mathbf{b}_t$
+\ELSIF {method == 'ARR'}
+        \STATE $\mathbf{S} = \mathbf{S}+
+        \mathbf{a}_t \mathbf{a}_t^\top-
+        \mathbf{a}_1 \mathbf{a}_1^\top$
+        \STATE $\mathbf{X} = \mathbf{S}^{-1} \mathbf{W} $
+        \STATE $\mathbf{W} = \mathbf{W} + \mathbf{a}_t \mathbf{b}_t$
+\ENDIF
+\STATE $\mathbf{y}_{\text{pred}} = \mathbf{X}^\top \mathbf{a}_t$
+\end{algorithmic}
+\caption{Solver Class for Regression Methods}
+\label{alg:solver}
+\end{algorithm}
+
+% Aqui se puede hablar de evaluation using competitive analysis. The idea
+%of competitiveness is to compare the output generated by an online algorithm to
+%the output produced by an optimal offline algorithm. An optimal online algorithm
+%is an omniscient algorithm that knows the entire input data in advance and can
+%compute an optimal output. The better an online algorithm approximates the
+%optimal solution, the more competitive this algorithm is.
+
+The proposal was compared against an optimal online algorithm which is a time varying
+VECM (OOVECM). Algorithm ~\ref{alg:OOVECM} shows this modification:
+
+\begin{algorithm}[ht]
+\begin{algorithmic}[1]
+\REQUIRE $\,$ \\
+$\mathbf{y}$: matrix with $N$ input vectors and $l$ time series\\
+$p$: number of past values \\
+$L$: sliding window size ($L<N$) \\
+\ENSURE  $\,$ \\
+$\{\Delta \mathbf{y}_{\text{pred}}[L+1],\dots,\Delta \mathbf{y}_{\text{pred}}[N]\}$: model predictions 
+\FOR { $i =0$ to $N-L$ }
+    \STATE $\mathbf{y}_i \gets \mathbf{y}[i:i+L]$
+	\STATE{$v \gets \texttt{getJohansen}(\mathbf{y}_i,p)$}
+	\STATE{$[\mathbf{A} \quad \mathbf{B}] \gets
+    \texttt{vecMatrix}(\mathbf{y}_i,p,v)$}
+    \STATE $\mathbf{X} \gets \text{OLS} (\mathbf{A},\mathbf{B})$%\mathbf{(A^\top A)^{-1}A^\top B}$
+    \STATE $\Delta \mathbf{Y}_{\text{true}}[i] \gets \mathbf{B}[-1,:]$
+    \STATE $\Delta \mathbf{Y}_{\text{pred}}[i] \gets \mathbf{A}[-1,:] \times \mathbf{X}$
+\ENDFOR
+    \STATE $\text{MAPE} \gets \texttt{mape}(\Delta \mathbf{Y}_{\text{true}}, \Delta
+    \mathbf{Y}_{\text{pred}})$
+\end{algorithmic}
+\caption{OOVECM: Optimal Online Vector Error Correction Model}
+\label{alg:OOVECM}
+\end{algorithm}
+
+%%Solo para tesis
+%\section{Online VEC model}
+%
+%If we want to update the VEC model with new input data
+%$\mathbf{y}_{T+1}$, we have to add a new row in matrices
+%$\mathbf{Y,X}$ and $\mathbf{E}$. 
+%
+%\begin{eqnarray}
+%\mathbf{Y} &=& 
+%                \left[ \begin{array}{ccc}
+%               \quad & \mathbf{\Delta y}_{p+1} & \quad \\ \hline
+%               \quad & \mathbf{\Delta y}_{p+2} & \quad \\ \hline
+%               \quad & \vdots & \quad \\ \hline 
+%               \quad & \mathbf{\Delta y}_T & \quad \\ 
+%               \quad & \color{red}\mathbf{\Delta y}_{T+1} & \quad 
+%               \end{array} \right]
+%             \\
+%\mathbf{\Phi^*} &=& 
+%                \left[ \begin{array}{ccc}
+%               \phi_{p-1}^* \\ 
+%               \vdots \\ 
+%               \phi_{1}^* \\
+%               \alpha \\
+%                c   
+%               \end{array} \right]
+%\\
+%\mathbf{X} &=& \begin{bmatrix} \label{offX}
+%   \mathbf{\Delta y}_2 & \dots & \mathbf{\Delta y}_{p-1} &
+%   \mathbf{\Delta y}_{p} & \beta'\mathbf{y}_{p} & 1\\
+%   \mathbf{\Delta y}_3 & \dots & \mathbf{\Delta y}_{p} &
+%   \mathbf{\Delta y}_{p+1} & \beta'\mathbf{y}_{p+1} &1\\
+%   \vdots &  \ddots & \vdots & \vdots & \vdots & \vdots\\
+%   \mathbf{\Delta y}_{T-p+1} & \dots & \mathbf{\Delta y}_{T-2} &
+%   \mathbf{\Delta y}_{T-1} & \beta'\mathbf{y}_{T-1} &1 \\
+%   \color{red} \mathbf{\Delta y}_{T-p+2} & \dots & \color{red}\mathbf{\Delta y}_{T-1} &
+%   \color{red}\mathbf{\Delta y}_{T} & \color{red}\beta'\mathbf{y}_{T}
+%   & \color{red} 1
+%   \end{bmatrix}
+%\\
+%\mathbf{E} &=& \begin{bmatrix}
+%              \mathbf{\epsilon}_{p+1} \\ \vdots \\ \mathbf{\epsilon}_T
+%              \\ \color{red}\mathbf{\epsilon}_{T+1}
+%             \end{bmatrix}
+%\end{eqnarray}

### 2015-04-19 src/5_3_Proposal.tex
ultimos cambios

@@
 
-In order to formulate this algorithm we need to define first the
-following matrices which considers only the last $L$ instances:
+VECM($p$) for $l$ cointegrated variables has the following form:
+
+\begin{equation}\label{eq:vecfull}
+\Delta\mathbf{y}_t 
+= \boldsymbol{\alpha\beta}^\top\mathbf{y}_{t-1} 
+  + \sum_{i=1}^{p-1}\boldsymbol{\Phi}_i^*\,\Delta\mathbf{y}_{t-i}
+  + \mathbf{c} + \boldsymbol{\epsilon}_t \qquad \forall t = p+1,\dots,N.
+\end{equation}
+
+Transposing each equation of the system (\ref{eq:vecfull}) we can write
+the VECM($p$) model in block-matrix form as:
+\begin{equation}\label{eq:vareq}
+\mathbf{B} = 
+\mathbf{A} \mathbf{X} + 
+\mathbf{E} \, , 
+\end{equation}
+%
+\noindent where $\mathbf{B}$ dimension is $((N-p)\times l)$, $\mathbf{A}$
+dimension is $((N-p)\times(r+(p-1)l +1))$, $\mathbf{X}$ dimension is $((r+(p-1)l
++1)\times l)$, $\mathbf{E}$ dimension is $((N-p)\times l)$ and $r$ is the number
+of cointegration vectors:
+%
+\begin{alignat}{3}
+\mathbf{B}
+&= \begin{bmatrix}
+   \Delta\mathbf{y}_{p+1}^\top \\
+   \Delta\mathbf{y}_{p+2}^\top \\
+   \vdots \\
+   \Delta\mathbf{y}_N^\top
+   \end{bmatrix}
+&\quad
+\mathbf{X}
+&= \begin{bmatrix}
+   \boldsymbol{\alpha}^\top \\
+   \boldsymbol{\Phi}_1^{*\top} \\
+   \boldsymbol{\Phi}_2^{*\top} \\
+   \vdots \\
+   \boldsymbol{\Phi}_{p-1}^{*\top} \\
+   \mathbf{c}^\top
+   \end{bmatrix}
+&\quad
+\mathbf{E}
+&= \begin{bmatrix}
+   \boldsymbol{\epsilon}_{p+1}^\top \\
+   \boldsymbol{\epsilon}_{p+2}^\top \\
+   \vdots \\
+   \boldsymbol{\epsilon}_N^\top \\
+   \end{bmatrix}
+\end{alignat}
+\noindent and 
+\begin{align}
+\mathbf{A} 
+&= \begin{pmat}[{....|}]
+   \mathbf{y}_p^\top \boldsymbol{\beta} & \Delta \mathbf{y}_p^\top & \Delta\mathbf{y}_{p-1}^\top & \dots 
+                    & \Delta\mathbf{y}_2^\top & 1 \cr
+   \mathbf{y}_{p+1}^\top  \boldsymbol{\beta} &\Delta\mathbf{y}_{p+1}^\top & \Delta\mathbf{y}_p^\top & \dots
+                       & \Delta\mathbf{y}_3^\top & 1 \cr
+   \vdots & \vdots & \vdots & \ddots & \vdots & \vdots \cr
+   \mathbf{y}_{N-1}^\top  \boldsymbol{\beta} &\Delta\mathbf{y}_{N-1}^\top & \Delta\mathbf{y}_{N-2}^\top & \dots 
+                       & \Delta\mathbf{y}_{N-p-1}^\top & 1 \cr
+   \end{pmat}\, .
+\end{align}
+%Taking into account the error term $\mathbf{E}$, equation~(\ref{eq:vareq}) 
+%can be solved with respect to $\mathbf{X}$ using the ordinary least
+%squares estimation.
+However, the number of rows of matrix $\mathbf{A}$ increases with the number of
+observations $\mathbf{y}_t$. Our proposal considers only the most recent $L$
+rows of matrices $\mathbf{A}$ and $\mathbf{B}$ defined as  $\mathbf{A}(t)$ and
+$\mathbf{B}(t)$:
 
@@
 
-The RR solution $\mathbf{X}(t)$ using the sliding window matrices defined above
+\noindent so that VECM parameters $\mathbf{X}(t)$ are found using the following
+equation:
+
+\begin{equation}
+\mathbf{B}(t) = \mathbf{A}(t) \mathbf{X}(t) + \mathbf{E}(t)\\
+\end{equation}
+
+The RR solution $\mathbf{\hat{X}}(t)$ using the sliding window matrices defined above
 is:
@@
 \begin{equation}
-\label{eq:optsolSLAAR}
-%\mathbf{X}(t)&=&()^{-1}{\bf A}(t)^\top{\bf B}(t) \\
-\mathbf{X}(t)=\mathbf{S}(t)^{-1} \mathbf{W}(t) \, ,
+\label{eq:oproblem}
+\mathbf{\hat{X}}(t)=\mathbf{S}(t)^{-1} \mathbf{W}(t) \, ,
 \end{equation}
@@
 \begin{eqnarray*}
-\mathbf{S}(t) =&{\bf A}(t)^\top{\bf A}(t)+ \gamma \mathbb{I} &=
-\sum_{i=0}^L \mathbf{a}_{t-i}\mathbf{a}_{t-i}^\top + \gamma \mathbb{I}
+\mathbf{S}(t) =&{\bf A}(t)^\top{\bf A}(t)+ \lambda \mathbb{I} &=
+\sum_{i=0}^L \mathbf{a}_{t-i}\mathbf{a}_{t-i}^\top + \lambda \mathbb{I}
 \label{eq:S} \\
@@
 
-Therefore Sherman-Morrison-Woodbury formula can be used to reduce inverse matrix
-calculations.
+%Agregar solo si se decide agregar Sherman-Morrison
+%Therefore the Sherman-Morrison-Woodbury formula can be used to reduce inverse matrix
+%calculations.
 
-The algorithm~\ref{alg:proposal} shows our online VECM using three different
-methods for geting model parameters: OLS, RR, AAR.
+The algorithm~\ref{alg:proposal} shows OVECM using three different
+methods for geting model parameters: OLS, RR, AAR which leads to three different
+algorithms OVECM-OLS, OVECM-RR, OVECM-AAR:
 
@@
 \item The function \texttt{vecMatrix} returns VECM
-matrices $\mathbf{A},\mathbf{B}$ shown in equation~(\ref{eq:vecmatrix}). This
+matrices $\mathbf{A}(t),\mathbf{B}(t)$ shown in equation~(\ref{eq:notation}). This
 method is only required at the first step.
-\item The function \texttt{vecMatrixOnline} returns new rows of $\mathbf{A}$ and
-$\mathbf{B}$ give new input data $\mathbf{y}_t$.
-\item solver (see algorithm \ref{alg:solver}) is an object which allows online
-calculation saving matrices $\mathbf{S}$ and $\mathbf{W}$ (see
-equations~\ref{eq:S} and \ref{eq:W}) for methods RR and AAR. This method is
-initialized in the first step and only if new cointegration are required. This
-method also returns in-sample MAPE saved in $\mathbf{e}$. 
+\item The function \texttt{vecMatrixOnline} returns new rows $\mathbf{a}_t^\top$ and
+$\mathbf{b}_t^\top$ given new input data $\mathbf{y}_t$.
+\item The Solver class (see algorithm \ref{alg:solver}) obtain VECM parameters  
+using RR, AAR or OLS methods.
 \item The function \texttt{vecMatrixUpdate} updates matrix $\mathbf{A}$ when new
@@
 parameters $\mathbf{X}$ is also updated later.
-\item Out-of-sample outputs are saved in the variables $\Delta
-\mathbf{Y}_{\text{true}}$ and $\Delta \mathbf{Y}_{\text{pred}}$ which are used
-to obtain Out-of-sample MAPE at the end.
 %\item The number of cointegration vectors is set as a parameter (aun no lo

### 2015-06-12 src/4_OnlineLearning.tex
pequeños detalles
demo variance and competitive analysis

@@
 	\begin{bmatrix} \quad &\mathbf{A}& \quad \\ \hdashline \quad
-& \lambda\mathbb{I}&\quad  
+& \sqrt\lambda\mathbb{I}&\quad  
 \end{bmatrix} 
@@
 \begin{eqnarray*}
-\mathbf{X}(\lambda) &=&( \mathbf{A}^\top \mathbf{A} + \lambda \mathbb{I})^{-1}\mathbf{Y} \\
+\mathbf{X}(\lambda) &=&( \mathbf{A}^\top \mathbf{A} + \lambda \mathbb{I})^{-1}\mathbf{A}^\top \mathbf{Y} \\
 &=& (\mathbb{I} + \lambda (\mathbf{A}^\top \mathbf{A})^{-1})^{-1} (\mathbf{A}^\top \mathbf{A})^{-1}\mathbf{A}^\top \mathbf{Y} \\
@@
 Var(\mathbf{X}(\lambda)) &=& Var(\mathbf{W}\hat{\mathbf{X}}) \\
+&=& E[(\mathbf{W}\hat{\mathbf{X}}-E[\mathbf{W}\hat{\mathbf{X}}])(\mathbf{W}\hat{\mathbf{X}}-E[\mathbf{W}\hat{\mathbf{X}}])^\top] \\
+&=& \mathbf{W}E[(\hat{\mathbf{X}}-E[\hat{\mathbf{X}}])(\hat{\mathbf{X}}-E[\hat{\mathbf{X}}])^\top] \mathbf{W}^\top \\
 &=& \mathbf{W}Var(\hat{\mathbf{X}})\mathbf{W}^\top \\
@@
 The figure~\ref{fig:biasvariance} shows the bias-variance tradeoff
-given by equation~(\ref{eq:prederror}). Ridge regression shows an
-increasing bias and a decreasing variance. Despite OLS has zero bias,
+given by equation~(\ref{eq:prederror}). Despite OLS has zero bias,
 its variance is greater than ridge for small values of $\lambda$.
 It can be shown that, in terms of prediction error, ridge (black line)
-is lower than OLS (dotted line)~\cite{hoerl1970}.
+is lower than OLS (dotted line)~\cite{hoerl1970}. Ridge regression shows an
+increasing squared bias and a decreasing variance.
+
+\textbf{Demo}\quad
+
+Since $Bias(\mathbf{X}(\lambda)) \neq 0$ this imply that
+
+\begin{equation*}
+Bias(\mathbf{X}(\lambda)) ^2 > 0 
+\end{equation*}
+
+\noindent we know that $\mathbf{A}^\top
+\mathbf{A}$ has an eigenvalue decomposition $\mathbf{A}^\top
+\mathbf{A} = \mathbf{V} \Sigma \mathbf{V}^{-1}$.
+
+\begin{eqnarray*}
+Var(\mathbf{X}(\lambda)) &=& \sigma^2 \mathbf{W}(\mathbf{A}^\top \mathbf{A} )^{-1}\mathbf{W}^\top\\
+Var(\mathbf{X}(\lambda)) &=& \sigma^2 (\mathbb{I} + \lambda (\mathbf{A}^\top
+\mathbf{A})^{-1})^{-1} (\mathbf{A}^\top \mathbf{A} )^{-1}((\mathbb{I} + \lambda (\mathbf{A}^\top
+\mathbf{A})^{-1})^{-1} )^\top \\
+&=& \sigma^2 (\mathbb{I} + \lambda (\mathbf{V} \Sigma \mathbf{V}^{-1})^{-1} (\mathbf{V} \Sigma \mathbf{V}^{-1})^{-1}((\mathbb{I} + \lambda (\mathbf{V} \Sigma \mathbf{V}^{-1})^{-1})^{-1} )^\top \\
+&=& \sigma^2 \mathbf{V} (\mathbb{I} + \lambda \Sigma^{-1})^{-1} \Sigma^{-1} (\mathbb{I} + \lambda \Sigma^{-1})^{-1}  \mathbf{V}^{-1}\\
+&=& \sigma^2 \mathbf{V} \mathbf{D}^{-1} \mathbf{V}^{-1}
+\end{eqnarray*}
+
+\noindent where $\mathbf{D}^{-1}  = (\mathbb{I} + \lambda \Sigma^{-1})^{-1} \Sigma^{-1} (\mathbb{I} + \lambda \Sigma^{-1})^{-1}$ is a diagonal matrix with coefficients:
+
+\begin{equation*}
+d_i = \frac{\sigma_i^{-1}}{(1+\lambda\sigma_i^{-1})^2}
+\end{equation*}
 
@@
 algorithm to the entire sequence of inputs received so far. It would be
-preferable compu- tationally if the existing hypothesis or other artifact of
+preferable computationally if the existing hypothesis or other artefact of
 learning could be updated in response to each newly received input observation.
@@
 data and only a summary needs to be retained. Due to this, incremental
-algorithms are often characterized as memoryless, because no memory of past data
+algorithms are often characterised as memoryless, because no memory of past data
 is required.  The algorithm is online but not incremental if it doesn't produce
@@
 \end{equation*}
+
+\subsection{Competitive analysis}
+
+Competitive analysis was designed for analysing the performance of an online algorithm compared with its optimal offline algorithm.
+An optimal offline algorithm can view the sequences of requests in advance.
+The effectiveness of an online algorithm \cite{} may be measured by its competitive ratio which is the worst-case ratio of the online algorithm and the the optimal offline algorithm.
 
@@
 
-The AAR, proposed by~\cite{vovk2001}, is an application of the aggregating
+The AAR, proposed by~\cite{vovk2001} (also known as Vovk-Azoury-Warmuth predictor ~\cite{azoury2001}), is an application of the aggregating
 algorithm to the problem of regression. The idea is introduce the new input

### 2015-06-17 src/5_3_Proposal.tex
reparando algo
pequeños detalles
Gráficos experimentos VECM vs OPT
plot de datos y distribución

@@
 
-\section{\uppercase{Methodology}}
-\label{sec:methodology}
+\section{Methodology}
 
@@
     \ENDIF
-	\STATE{$[\mathbf{a}_t \quad \mathbf{b}_t] \gets
+	\STATE{[$\mathbf{A} \quad \mathbf{B} \quad \mathbf{a}_t \quad \mathbf{a}_L \quad \mathbf{b}_t \quad
+    \mathbf{b}_L ] \gets
     \texttt{vecMatrixOnline}(\mathbf{Y},\mathbf{y}_t,p,v)$}
@@
 \quad \\
-\texttt{init\_model}($\mathbf{A},\mathbf{B}$)\\
-\IF {method == 'RR' or method == 'AAR'}
+\texttt{init\_model}($\mathbf{A},\mathbf{B}$)
 \STATE $ [m \quad n] = \text{size}(\mathbf{A}) $ 
@@
 \STATE $\mathbf{W} = \displaystyle \sum_{i=1}^m \mathbf{a}_i \mathbf{b}_i$
-\ENDIF \\
 \quad \\
-\texttt{regression}($\mathbf{A},\mathbf{B},\mathbf{a_t},\mathbf{b_t},n$) \\
+\texttt{regression}($\mathbf{A},\mathbf{B},\mathbf{a_t},\mathbf{b_t}$) \\
 \IF {method == 'OLS'}
@@
 %\end{eqnarray}
+
+\section{Experiments}
+
+\begin{figure}[!h]
+  %\vspace{-0.8cm}
+  \centering
+  \includegraphics[width=\textwidth]{img/forexdata}
+  \caption{10 seconds frequency forex data for EURUSD, GBPUSD and USDJPY}
+  \label{fig:forexdata}
+\end{figure}
+
+\begin{figure}[!h]
+  %\vspace{-0.8cm}
+  \centering
+  \includegraphics[width=0.8\textwidth]{img/distdata}
+  \caption{Distribution of EURUSD, GBPUSD and USDJPY data}
+  \label{fig:distdata}
+\end{figure}
+
+\begin{figure}[!h]
+  %\vspace{-0.8cm}
+  \centering
+  \includegraphics[width=0.9\textwidth]{img/onlinecomparison}
+  \caption{Cumulative loss of AAR, RR and OLS solutions for VECM against its optimal algorithm}
+  \label{fig:onlinecomparison}
+\end{figure}
+
+\begin{figure}[!h]
+  %\vspace{-0.8cm}
+  \centering
+  \includegraphics[width=0.9\textwidth]{img/RRcomparison}
+  \caption{Cumulative loss RR VECM against its optimal algorithm using different $\lambda$ values}
+  \label{fig:RRcomparison}
+\end{figure}
+
+\begin{figure}[!h]
+  %\vspace{-0.8cm}
+  \centering
+  \includegraphics[width=0.9\textwidth]{img/AARcomparison}
+  \caption{Cumulative loss AAR VECM against its optimal algorithm using different $\lambda$ values}
+  \label{fig:AARcomparison}
+\end{figure}

### 2015-06-19 manuscript/manuscript.tex
Primera version de articulo

--- 2015-06-17 src/5_3_Proposal.tex
@@
-\chapter{Real-time intraday Forex rates forecasting using a regularized Vector
-Error Correction model}
+\RequirePackage{fix-cm}
+%
+%\documentclass{svjour3}                     % onecolumn (standard format)
+%\documentclass[smallcondensed]{svjour3}     % onecolumn (ditto)
+%\documentclass[smallextended]{svjour3}       % onecolumn (second format)
+\documentclass[twocolumn]{svjour3}          % twocolumn
+%
+\smartqed  % flush right qed marks, e.g. at end of proof
+%
+\usepackage{graphicx}
+%
+% \usepackage{mathptmx}      % use Times fonts if available on your TeX system
+%
+% insert here the call for the packages your document requires
+%\usepackage{latexsym}
+% etc.
+%
+\usepackage{amsfonts,amssymb,amsmath,
+            amssymb,amsopn,amsxtra,amstext,amsbsy}
+\usepackage{epsfig}
+\usepackage[makeroom]{cancel}
+\usepackage{array,arydshln}
+\usepackage{array}
+\usepackage{url}
+%\usepackage{tikz}
+\usepackage[utf8x]{inputenc}
+\usepackage{fancyhdr}
+\usepackage{hyperref,wasysym}
+\usepackage{color}
+\usepackage{enumerate}
+\usepackage{algorithm}
+\usepackage{algorithmic}
+\renewcommand{\algorithmicrequire}{\textbf{Input:}}
+\renewcommand{\algorithmicensure}{\textbf{Output:}}
+\usepackage{multirow}
+%\usepackage[spanish]{babel}
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\usepackage{xcolor}%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\usepackage{hyperref}%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\usepackage{pmat}%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%\newtheorem*{myteo}{Teorema}
+%\newtheorem*{mydef}{Definici\'on}
+%% MYREMARK ENVIRONMENT %%%%%%%%%%%
+\newenvironment{mycolor}{\color{blue}}{}
+\newcounter{myremarkcount}
+\setcounter{myremarkcount}{0}
+\newsavebox{\myremarkbox}
+\newenvironment{myremark}
+{% This is the begin code
+\medskip
+\begin{flushright}
+\begin{lrbox}{\myremarkbox}%
+\begin{minipage}[t]{1.00\linewidth}
+\stepcounter{myremarkcount}
+\footnotesize
+\begin{mycolor}
+{{\bf Proofreading Remark.}\ }
+% \begin{it}
+}
+{% This is the end code
+% \end{it}
+\end{mycolor}
+\end{minipage}
+\end{lrbox}
+\fbox{\usebox{\myremarkbox}}
+\end{flushright}
+
+\bigskip}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\journalname{Applied Intelligence}
+
+\begin{document}
+
+\title{Intraday Forex rates forecasting using an online cointegration approach}
+%\subtitle{Do you have a subtitle?\\ If so, write it here}
+
+%\titlerunning{Short form of title}        % if too long for running head
+
+\author{Paola Arce \and Werner Kristjanpoller \and Luis Salinas }
+
+%\authorrunning{Short form of author list} % if too long for running head
+
+\institute{P. Arce  \at
+              Departamento de Inform\'atica, Universidad T\'ecnica Federico Santa
+              Mar\'ia\\
+              Tel.: +56-32-2654000\\
+              %Fax: +123-45-678910\\
+              \email{paola.arce@usm.cl}           %  \\
+%             \emph{Present address:} of F. Author  %  if needed
+           \and
+           W. Kristjanpoller \at
+              Departamento de Industrias, Universidad T\'ecnica Federico Santa
+              Mar\'ia\\
+              %Fax: +123-45-678910\\
+              \email{werner.kristjanpoller@usm.cl}             \\
+           \and
+           L. Salinas \at
+              Departamento de Inform\'atica, Universidad T\'ecnica Federico Santa
+              Mar\'ia\\
+              Centro Científico Tecnológico de Valparaíso (CCTVal) \\
+              \email{luis.salinas@usm.cl}           %  \\
+}
+
+\date{Received: date / Accepted: date}
+% The correct dates will be entered by the editor
+
+\maketitle
+
+\begin{abstract}
+
+%A Vector Error Correction Model (VECM) is an econometric model which
+%characterizes the joint dynamic behaviour of a set of cointegrated variables. It
+%is said that two non-stationary time series are cointegrated if a linear
+%combination of them exhibit a stationary behaviour. 
+%
+%Cointegration is commonly found in finance and it has been shown that
+%introducing cointegration restrictions often improves forecasting. However,
+%determining VECM parameters and how much of the historical data is useful for
+%making a forecast is a computational challenge.  
+%
+%In this study, we propose a new version of VECM whose parameters are obtained
+%adaptively using parallel computing taking into account the cointegrated
+%behaviour of the time series. The methodology is tested using data of the
+%Foreign Exchange (Forex) market. Numerical experiments show that our adaptive
+%and parallel approach significantly improves execution times with respect to the
+%common serial processing and forecasts are made before new data arrives.  
 Vector error correction model (VECM) parameters are commonly obtained using the
@@
 
-\section{The problem}
+\keywords{VECM \and Cointegration \and Forex \and MPI \and Parallel algorithm}
 
+\end{abstract}
+
+\section{Introduction}
+\label{sec:introduction}
 VECM introduces the long-run relationship among a set of cointegrated variables
@@
 currency rates from the foreign exchange market with a frequency of 10 seconds.
-Model efectiveness is focused on out-of-sample forecast rather than in-sample
+Model effectiveness is focused on out-of-sample forecast rather than in-sample
 fitting.  This criteria allows the OVECM prediction capability to be expressed
@@
 
-%The next sections are organized as follows: section~\ref{sec:background}
-%presents the VAR and VECM, the OVECM algorithm proposed is presented in
-%section~\ref{sec:methodology}. Section~\ref{sec:results} gives a description of
-%the data used and the tests carried on to show accuracy and time comparison of
-%our proposal against the traditional VECM and
-%section~\ref{sec:conclusions} includes conclusions and a proposal for future
-%tudy.
+\section{Background}
+\label{sec:background}
+\subsection{Integration and Cointegration}\label{sec:coint}\  
+Following Johansen \cite{johansen1995} we shall say that a stochastic process
+$Y_t$ which satisfies $Y_t-E(Y_t) = \sum_{i=0}^\infty C_i\,\varepsilon_{t-i}$ is
+called $I(0)$, and then we shall write $Y_t\sim I(0)$, whenever
+$\sum_{i=0}^\infty C_i \neq 0$ and $\sum_{i=0}^\infty C_i\,z^i$ converges for
+$z\in\mathbb{C}$ with $|z|<1$.  It is understood that the condition
+$\varepsilon_t\sim iid(0,\sigma^2)$ holds.
 
-\section{Methodology}
+A (vector) time series $\mathbf{y}_t$ is said to be {\em integrated of order\/}
+$d$, and then we shall write $\mathbf{y}_t\sim I(d)$, whenever after $d$ times
+(discrete) differentiation an stationary process is obtained~\cite{banerjee1993};
+more precisely, whenever
+$(1-L)^d\,\mathbf{y}_t\sim\text{I(0)}$, where $L$ is the usual lag operator:
+$(1-L)\,\mathbf{y}_t = \Delta\mathbf{y}_t = \mathbf{y}_t-\mathbf{y}_{t-1}$ for
+all $t$.
+
+Note that this definition includes the scalar case as time series of
+vectors of dimension 1; in this scalar case we will write the time series in
+nonbold format.
+
+Let $\mathbf{y}_t^\nu$, $\nu=1,\dots,p$, be a set of $p$ vector time series of
+order $I(1)$.  They are said to be {\em cointegrated\/} if a vector
+$\beta=[\beta(1),\dots,\beta(p)]^\top \in \mathbb{R}^p$ exists, such that the
+time series,
+\begin{equation}
+\mathbf{Z}_t:= 
+\sum_{\nu=1}^p \beta(\nu)\,\mathbf{y}_t^\nu\,\sim\,\text{I(0)}\,.
+\end{equation}
+In other words, a set of $I(1)$ time series is said to be cointegrated if a
+linear combination of them exists, which is I(0).
+
+\subsection{Vector Autorregresive Models}\label{sec:varvec}
+
+VECM is a special case of VAR model and both describe the joint behaviour
+of a set of variables.
+
+The VAR($p$) model is a general framework describing the behaviour of a
+set of $l$ endogenous variables as a linear combination of their last $p$
+values, where $l,p\in\mathbb{N}$. 
+In our case, each one of these $l$ variables is a scalar time series
+$y_{\lambda,t}$, $\lambda=1,\dots,l$, and we represent them all together
+at time $t$ by the the vector time series:
+\begin{equation}
+\label{eq:variables}
+\mathbf{y}_t = 
+\begin{bmatrix} y_{1,t} & y_{2,t} & \dots & y_{l,t} \end{bmatrix}^\top.
+\end{equation}
+\noindent
+Notice that the vectors $\mathbf{y}_t$ are assumed to be $l$-dimensional.
+
+The VAR($p$) model describes the behaviour of a dependent variable in terms of
+its own lagged values and the lags of the others variables in the system. The
+model with $p$ lags is formulated as the system of $N$:
+\begin{align}
+\label{eq:var}
+\mathbf{y}_t 
+= \boldsymbol{\Phi}_1 \mathbf{y}_{t-1} +
+  \boldsymbol{\Phi}_2 \mathbf{y}_{t-2} + \dots +
+  \boldsymbol{\Phi}_p\mathbf{y}_{t-p} +
+  \mathbf{c} + \boldsymbol{\epsilon}_t \nonumber \\
+t=p+1,\dots,N,
+\end{align}
+\noindent where 
+$\boldsymbol{\Phi}_1, \boldsymbol{\Phi}_2,\dots,\boldsymbol{\Phi}_p$
+are $l\times l$-matrices of real coefficients,
+$\boldsymbol{\epsilon}_{p+1},
+ \boldsymbol{\epsilon}_{p+2}, \dots, \boldsymbol{\epsilon}_N$ 
+are error terms, $\mathbf{c}$ is a constant vector and $N$ is the total
+number of samples.
+
+Notice that, regarding our notation of section (\ref{sec:coint}),
+we have here 
+$\mathbf{y}_t^0 = \mathbf{y}_t$,
+$\mathbf{y}_t^\nu = \mathbf{y}_{t-\nu}$ and
+the $\lambda$-th component of the vector time series $\mathbf{y}_t^\nu$
+is the scalar time series $y_{\lambda,t}^\nu$, where $\nu=1,\dots,p$ and
+$\lambda=1,\dots,l$.
+
+VECM is a special form of a VAR model for I(1) variables that are also
+cointegrated~\cite{banerjee1993}.
+
+It is obtained re-writing equation (\ref{eq:var}) in terms of the new
+variable $\Delta\mathbf{y}_t=\mathbf{y}_t-\mathbf{y}_{t-1}$.
+The VECM model, expressed in terms those differences, takes the form:
+\begin{equation}\label{eq:vec}
+\Delta \mathbf{y}_t 
+= \boldsymbol{\Omega}\,\mathbf{y}_{t-1}
+  + \sum_{i=1}^{p-1} \boldsymbol{\Phi}_i^*\,\Delta\mathbf{y}_{t-i}
+  + \mathbf{c} + \boldsymbol{\epsilon}_t\,,
+\end{equation}
+\noindent
+where the coefficients matrices $\boldsymbol{\Phi}_i^*$ and 
+$\boldsymbol{\Omega}$, expressed in terms of the matrices
+$\boldsymbol{\Phi}_i$ of (\ref{eq:var}), are:
+\begin{align*}
+\boldsymbol{\Phi}_i^* 
+&:= -\sum_{j=i+1}^{p}\boldsymbol{\Phi}_j\,, \\
+\boldsymbol{\Omega}
+&:= -\left( \mathbb{I} - \boldsymbol{\Phi}_1 - \dots 
+    - \boldsymbol{\phi}_p \right)\,. 
+\end{align*}
+The following well known properties of the matrix $\boldsymbol{\Omega}$
+\cite{johansen1995} will be useful in the sequel:
+\begin{itemize}
+\item
+If $\boldsymbol{\Omega} = \mathbf{0}$, there is no cointegration.
+\item 
+If $rank(\boldsymbol{\Omega})=l$, i.e., if $\boldsymbol{\Omega}$ has
+full rank, then the time series are not I(1) but stationary.
+\item
+If $rank(\boldsymbol{\Omega})=r$, $0<r<l$, then there is cointegration
+and the matrix $\boldsymbol{\Omega}$ can be expressed as
+$\boldsymbol{\Omega}=\boldsymbol{\alpha\beta}^\top$, where $\boldsymbol{\alpha}$
+and $\boldsymbol{\beta}$ are
+$l\times r$ matrices and
+$\text{rank}(\boldsymbol{\alpha})=\text{rank}(\boldsymbol{\beta})=r$.
+\item
+The columns of $\boldsymbol{\beta}$ contains the cointegration vectors and the rows of
+$\boldsymbol{\alpha}$ correspond with the adjusted vectors. 
+$\boldsymbol{\beta}$ is obtained by Johansen procedure~\cite{johansen1988},
+whereas $\boldsymbol{\alpha}$ has to be determined as a variable in the VECM.
+\end{itemize}
+It is worth noticing that the factorization of the matrix
+$\boldsymbol\Omega$ is not unique, since for any $r \times r$
+non-singular matrix $\mathbf{H}$, $\boldsymbol{\alpha}^*:=\boldsymbol{\alpha}\mathbf{H}$,
+and $\boldsymbol{\beta}^*=\boldsymbol{\beta}(\mathbf{H}^{-1})^\top$ we have
+$\boldsymbol{\alpha\beta}^\top=\boldsymbol{\alpha}^*(\boldsymbol{\beta}^*)^\top$.
+If cointegration exists, then equation (\ref{eq:vec}) can be written
+as follows:
+\begin{equation}\label{eq:vecfull}
+\Delta\mathbf{y}_t 
+= \boldsymbol{\alpha\beta}^\top\mathbf{y}_{t-1} 
+  + \sum_{i=1}^{p-1}\boldsymbol{\Phi}_i^*\,\Delta\mathbf{y}_{t-i}
+  + \mathbf{c} + \boldsymbol{\epsilon}_t\,,
+\end{equation}
+\noindent
+which is a VAR model but for time series differences.
+
+Transposing each equation of the system (\ref{eq:vecfull}) we can write
+the VECM($p$) model in block-matrix form as:
+\begin{equation}\label{eq:vareq}
+\mathbf{B} = 
+\mathbf{A} \mathbf{X} + 
+\mathbf{E} \, , 
+\end{equation}
+%
+\noindent where $\mathbf{B}$ dimension is $((N-p)\times l)$, $\mathbf{A}$
+dimension is $((N-p)\times(r+(p-1)l +1))$, $\mathbf{X}$ dimension is $((r+(p-1)l
++1)\times l)$ and $\mathbf{E}$ dimension is $((N-p)\times l)$:
+%
+\begin{alignat}{3}
+\mathbf{B}
+&= \begin{bmatrix}
+   \Delta\mathbf{y}_{p+1}^\top \\
+   \Delta\mathbf{y}_{p+2}^\top \\
+   \vdots \\
+   \Delta\mathbf{y}_N^\top
+   \end{bmatrix}
+&\quad
+\mathbf{X}
+&= \begin{bmatrix}
+   \boldsymbol{\alpha}^\top \\
+   \boldsymbol{\Phi}_1^{*\top} \\
+   \boldsymbol{\Phi}_2^{*\top} \\
+   \vdots \\
+   \boldsymbol{\Phi}_{p-1}^{*\top} \\
+   \mathbf{c}^\top
+   \end{bmatrix}
+&\quad
+\mathbf{E}
+&= \begin{bmatrix}
+   \boldsymbol{\epsilon}_{p+1}^\top \\
+   \boldsymbol{\epsilon}_{p+2}^\top \\
+   \vdots \\
+   \boldsymbol{\epsilon}_N^\top \\
+   \end{bmatrix}
+\end{alignat}
+\noindent and 
+\begin{align}
+\mathbf{A} 
+&= \begin{pmat}[{....|}]
+   \mathbf{y}_p^\top \boldsymbol{\beta} & \Delta \mathbf{y}_p^\top & \Delta\mathbf{y}_{p-1}^\top & \dots 
+                    & \Delta\mathbf{y}_2^\top & 1 \cr
+   \mathbf{y}_{p+1}^\top  \boldsymbol{\beta} &\Delta\mathbf{y}_{p+1}^\top & \Delta\mathbf{y}_p^\top & \dots
+                       & \Delta\mathbf{y}_3^\top & 1 \cr
+   \vdots & \vdots & \vdots & \ddots & \vdots & \vdots \cr
+   \mathbf{y}_{N-1}^\top  \boldsymbol{\beta} &\Delta\mathbf{y}_{N-1}^\top & \Delta\mathbf{y}_{N-2}^\top & \dots 
+                       & \Delta\mathbf{y}_{N-p-1}^\top & 1 \cr
+   \end{pmat}\, .
+\end{align}
+Taking into account the error term $\mathbf{E}$, equation~(\ref{eq:vareq}) 
+can be solved with respect to $\mathbf{X}$ using the ordinary least
+squares estimation.
+
+\subsection{Ordinary Least Squares method}
+
+The solution $\widehat{\mathbf{A}}$ to
+equation~(\ref{eq:vareq}) can be obtained by the ordinary least squares (OLS)
+method. $\widehat{\mathbf{X}}$ is the solution of the quadratic optimization problem
+\begin{equation*}
+\widehat{\mathbf{X}} = \underset{\mathbf{X}}{\text{Arg\;min}}
+\|\mathbf{A}\mathbf{X}-\mathbf{B}\|_2^2
+\end{equation*}
+\noindent for which the solution $\widehat{\mathbf{X}}$ is well-known:
+\begin{equation*}
+\label{eq:MP}
+\widehat{\mathbf{X}}=\mathbf{A}^{\!\!+}\,\mathbf{B}
+\end{equation*}
+\noindent where $\mathbf{A}^{\!\!+}$ is the Moore-Penrose pseudo-inverse
+which, when $\mathbf{A}$ is full rank, can be written as follows: 
+\begin{equation}
+\label{eq:pseudoinverse}
+\mathbf{A}^{\!\!+}= (\mathbf{A}^{\!\!\top} \mathbf{A})^{-1}\mathbf{A}^{\!\!\top} \, .
+\end{equation}
+when $\mathbf{A}$ is not full rank, i.e
+$rank(\mathbf{A})=k <  n \leq m$, $\mathbf{A}^\top \mathbf{A}$ is
+always singular and equation~(\ref{eq:pseudoinverse}) cannot be used.
+More generally, the pseudo-inverse is best computed using the compact
+singular value decomposition (SVD) of $\mathbf{A}$ which is:
+\begin{equation*}
+    \label{eq:compactsvd}
+    \mathbf{A}=
+    \mathbf{U_1}
+    \boldsymbol \Sigma_1
+    \mathbf{V}_1^{\top} \, .
+\end{equation*}
+Pseudo-inverse can then be written as follows:
+\begin{equation*}
+\label{eq:pseudoinversesvd}
+\mathbf{A}^{\!\!+} = \mathbf{V}_1 \boldsymbol \Sigma_1^{-1} \mathbf{U}_1^\top \, .
+\end{equation*}
+
+\section{Methodology} \label{sec:methodology}
 
@@
 
-\begin{equation}\label{eq:vecfull}
+\begin{align}
+\label{eq:vecfull}
 \Delta\mathbf{y}_t 
@@
   + \sum_{i=1}^{p-1}\boldsymbol{\Phi}_i^*\,\Delta\mathbf{y}_{t-i}
-  + \mathbf{c} + \boldsymbol{\epsilon}_t \qquad \forall t = p+1,\dots,N.
-\end{equation}
+  + \mathbf{c} + \boldsymbol{\epsilon}_t  \nonumber \\
+t=p+1,\dots,N,
+\end{align}
 
@@
 methods for geting model parameters: OLS, RR, AAR which leads to three different
-algorithms OVECM-OLS, OVECM-RR, OVECM-AAR:
+%algorithms OVECM-OLS, OVECM-RR, OVECM-AAR:
 
@@
 \end{algorithm}
-
 % Aqui se puede hablar de evaluation using competitive analysis. The idea
@@
 %optimal solution, the more competitive this algorithm is.
-
 The proposal was compared against an optimal online algorithm which is a time varying
-VECM (OOVECM). Algorithm ~\ref{alg:OOVECM} shows this modification:
-
+VECM. Algorithm \ref{alg:OOVECM} shows this modification:
 \begin{algorithm}[ht]
@@
 
+\section{Experimental results}
+\label{sec:results}
+\subsection{Data} \label{sec:unitroot}
+%AVECM tests were carried out using four foreign exchange rates all related to
+%USD: EURUSD, GBPUSD, USDCHF and USDJPY. This data was collected from the free
+%database Dukascopy which gives access to the Swiss Foreign Exchange marketplace
+%~\cite{Dukascopy2014}.
+%
+%The tests were done using 10-seconds frequency from ask prices which
+%corresponded to 8640 data points per day from the 11th to the 15th of August
+%2014. Since data datetimes were in GMT, opening market times for London, New
+%York, Sidney and Tokyo corresponded with data points 1440, 3240, 6120 and 6840
+%respectively. Our tests were made considering two hours after the opening time
+%of the 13th of August. Therefore, in order to obtain these times, we added 2
+%days and 2 hours to all the opening times ($2\times 8640 + 2 \times 360$) to obtain:
+%19440, 21240, 24120 and 24840. We called offset to these times in
+%table~\ref{tab:stats}, which allowed us to obtain performance at different
+%times of the day.
+
+\subsection{Unit root tests} \label{sec:unitroot}
+Before running the tests, we firstly checked whether the time series were
+I(1) using the Augmented Dickey Fuller (ADF) test at 95\% significance level.
+Table~\ref{tab:adf} shows that all currency rates cannot reject the unit root
+test but they rejected it with their first differences. This means that all of
+them are I(1) time series and we are allowed to use VECM and therefore AVECM.
+
+\begin{table}[h!]
+\label{tab:adf}
+\begin{center}
+\begin{tabular}{|l|c|c|c|c|c|}
+\hline
+& \textbf{Statistic} & \textbf{Critical value} & \textbf{Result}\\
+\hline
+EURUSD          &  -0.05398   & -1.94101 & True       \\
+$\Delta$ EURUSD & -89.12344   & -1.94101 & False       \\
+GBPUSD          &  -0.88208   & -1.94101 & True          \\
+$\Delta$ GBPUSD & -82.56501   & -1.94101 & False       \\
+CHFUSD          & -0.49941    & -1.94101 & True         \\
+$\Delta$ CHFUSD & -74.21940   & -1.94101 & False       \\
+JPYUSD          &  0.36721    & -1.94101 & True        \\
+$\Delta$ JPYUSD & -101.15589  & -1.94101 & False     \\ 
+\hline
+\end{tabular}
+\end{center}
+\caption{Unit roots tests for EURUSD, GBPUSD, USDCHF and USDJPY at 10-seconds
+frequency.}
+\end{table}
+
 \section{Experiments}
 
-\begin{figure}[!h]
+\begin{figure}[!ht]
   %\vspace{-0.8cm}
   \centering
-  \includegraphics[width=\textwidth]{img/forexdata}
+  \includegraphics[width=0.4\textwidth]{forexdata}
   \caption{10 seconds frequency forex data for EURUSD, GBPUSD and USDJPY}
@@
 
-\begin{figure}[!h]
+\begin{figure}[!ht]
   %\vspace{-0.8cm}
   \centering
-  \includegraphics[width=0.8\textwidth]{img/distdata}
+  \includegraphics[width=0.4\textwidth]{distdata}
   \caption{Distribution of EURUSD, GBPUSD and USDJPY data}
@@
 
-\begin{figure}[!h]
+\begin{figure}[!ht]
   %\vspace{-0.8cm}
   \centering
-  \includegraphics[width=0.9\textwidth]{img/onlinecomparison}
+  \includegraphics[width=0.4\textwidth]{onlinecomparison}
   \caption{Cumulative loss of AAR, RR and OLS solutions for VECM against its optimal algorithm}
@@
 
-\begin{figure}[!h]
+\begin{figure}[!ht]
   %\vspace{-0.8cm}
   \centering
-  \includegraphics[width=0.9\textwidth]{img/RRcomparison}
+  \includegraphics[width=0.4\textwidth]{RRcomparison}
   \caption{Cumulative loss RR VECM against its optimal algorithm using different $\lambda$ values}
@@
 
-\begin{figure}[!h]
+\begin{figure}[!ht]
   %\vspace{-0.8cm}
   \centering
-  \includegraphics[width=0.9\textwidth]{img/AARcomparison}
+  \includegraphics[width=0.4\textwidth]{AARcomparison}
   \caption{Cumulative loss AAR VECM against its optimal algorithm using different $\lambda$ values}
@@
 \end{figure}
+
+\section{Conclusions}
+\label{sec:conclusions}
+kk
+
+\bibliographystyle{plain}       % APS-like style for physics
+
+\bibliography{reference}   % name your BibTeX data base
+
+\end{document}
+% end of file template.tex

### 2015-06-19 manuscript/manuscript.tex
Avance articulo falta optimal algoritmo

@@
 
-\title{Intraday Forex rates forecasting using an online cointegration approach}
+\title{High frequency Intraday Forex rates forecasting using an online cointegration approach}
 %\subtitle{Do you have a subtitle?\\ If so, write it here}
@@
 
-%A Vector Error Correction Model (VECM) is an econometric model which
-%characterizes the joint dynamic behaviour of a set of cointegrated variables. It
-%is said that two non-stationary time series are cointegrated if a linear
-%combination of them exhibit a stationary behaviour. 
-%
-%Cointegration is commonly found in finance and it has been shown that
-%introducing cointegration restrictions often improves forecasting. However,
-%determining VECM parameters and how much of the historical data is useful for
-%making a forecast is a computational challenge.  
-%
-%In this study, we propose a new version of VECM whose parameters are obtained
-%adaptively using parallel computing taking into account the cointegrated
-%behaviour of the time series. The methodology is tested using data of the
-%Foreign Exchange (Forex) market. Numerical experiments show that our adaptive
-%and parallel approach significantly improves execution times with respect to the
-%common serial processing and forecasts are made before new data arrives.  
-Vector error correction model (VECM) parameters are commonly obtained using the
-ordinary least squares method. Since this parameters could be sensitive to the
-data, this proposal is to get these parameters using ridge regression which
-could lead to a better generalization capability. In order to give with a
-real-time response, an online version of VECM (OVECM) is proposed. OVECM solves
-VECM considering only a sliding window of historical data and therefore
-better response times could be achieved.
+This research examines the cointegration relation between three mayor currency pairs (MCP) of the foreign exchange markets:  EURUSD, GBPUSD, USDCHF.
+Specifically, we used a modified version of the Vector error correction model (VECM) which parameters are updated online using a limited amount of historical data.
+VECM parameters are commonly obtained using the
+ordinary least squares (OLS) method.
+Our proposal is to solve VECM using Ridge regression (RR) and the Aggregating algorithm for regression (AAR) which could lead to a better generalisation capability rather than OLS.
+Both approaches together with OLS are compared against an optimal offline algorithm which is able to see the sequences in advance.
+Our experiments were carried out using 10-seconds frequency Forex data and shows that RR and AAR approaches improve cumulative loss of OLS approach with respect to the optimal offline algorithm.
 
-\keywords{VECM \and Cointegration \and Forex \and MPI \and Parallel algorithm}
+\keywords{VECM \and Cointegration \and Forex \and High frequency data \and Online algorithm}
 
@@
 \label{sec:introduction}
-VECM introduces the long-run relationship among a set of cointegrated variables
-as an error correction term. VECM is a special case of the vector autorregresive
+
+The relationships between financial time series are commonly studied. In particular, in some of them
+it is possible to detect long-run relationships through to a cointegration analysis.
+Two or more non-stationary time series are said to be cointegrated if there exists a linear combination of them which is stationary.
+
+Vector error correction model (VECM) introduces the long-run relationship among a set of cointegrated variables
+as an error correction term. VECM is a special case of the vector autoregressive
 model (VAR) model. VAR model expresses future values as a linear combination of
@@
 adjusts coefficients to bring the variables back to equilibrium. In finance,
-many economic time series turn to be stationary when they are differentiated and
-cointegration restrictions often improves forecasting~\cite{duy1998}. Therefore,
+many economic time series (foreign exchange market rates for example) turns out to be stationary, when they are differentiated once ($I(1)$) and
+cointegration restrictions applied on the model often improves forecasting~\cite{duy1998}. Therefore,
 VECM has been widely adopted.
 
-Both VECM and VAR model parameters are obtained using ordinary least squares
-(OLS) method. OLS has two main problems: is sensitive to errors on input data
+VECM parameters are obtained using ordinary least squares
+(OLS) method \cite{golub1980}. OLS has two main problems: is sensitive to errors on input data
 and involves many calculations. The former problem is commonly solved using
-Ridge Regression (RR) \cite{hoerl1970} which introduces a regularization
-parameter that leads to an unbiased estimation with better generalization
+Ridge Regression (RR) \cite{hoerl1970} which introduces a regularisation
+parameter that leads to an unbiased estimation with better generalisation
 capability. The second problem of computational complexity depends on the number
@@
 their simplicity and their ability to update the model when new data is
-available.
+available. We use an online version of the linear ridge regression and also a modification introduced by 
+Vovk \cite{vovk2001} called the Aggregating algorithm regression (AAR), which is also known as Vovk-Azoury-Warmuth predictor ~\cite{azoury2001}.
 
 Our proposal is an online formulation of the VECM called Online VECM (OVECM)
-based on consideration of only a sliding window of the historical data.  OVECM
-introduces matrix optimizations in order to reduce the number of operations and
-also takes into account the fact that cointegration vector space doesn't
-experience large changes with small changes in the input data. Moreover, OVECM
-uses RR instead of OLS to obtain VECM parameters. Our method is later tested using four
-currency rates from the foreign exchange market with a frequency of 10 seconds.
-Model effectiveness is focused on out-of-sample forecast rather than in-sample
-fitting.  This criteria allows the OVECM prediction capability to be expressed
-rather than just explaining data history. Our method performance is compared
-with its optimal offline algorithm.
+based on consideration of only a sliding window of the historical data.  OVECM parameters are obtained
+via OLS, RR or AAR. A competitive analysis is done comparing its cumulative losses against the optimal offline algorithm.
+
+In finance, little academic research has been done using high-frequency Forex
+rates \cite{Bekiros2015}. Our proposal is tested using four
+three mayor currency pairs (MCP) from the foreign exchange market:  Euro (EUR) to United States Dollar (USD) (EURUSD), British Pound (GBP) to
+USD (GBPUSD) and USD to Swiss Franc (CHF) (USDCHF). We used samples with a frequency of 10 seconds.
+
+This paper is organised as follows: section~\ref{sec:background} presents concepts of integration and cointegration, the VECM model and an online learning introduction, the OVECM algorithm is presented in section~\ref{sec:methodology}.
+In section~\ref{sec:results} we describe the experiments carried out and their results. This section also includes a
+description of the test data.  Section~\ref{sec:conclusions} contains the
+conclusions and a discussion of future research.
 
@@
 
-\subsection{Vector Autorregresive Models}\label{sec:varvec}
+\subsection{Vector Error Correction Model}\label{sec:varvec}
 
@@
 \end{equation}
-when $\mathbf{A}$ is not full rank, i.e
-$rank(\mathbf{A})=k <  n \leq m$, $\mathbf{A}^\top \mathbf{A}$ is
-always singular and equation~(\ref{eq:pseudoinverse}) cannot be used.
-More generally, the pseudo-inverse is best computed using the compact
-singular value decomposition (SVD) of $\mathbf{A}$ which is:
+%when $\mathbf{A}$ is not full rank, i.e
+%$rank(\mathbf{A})=k <  n \leq m$, $\mathbf{A}^\top \mathbf{A}$ is
+%always singular and equation~(\ref{eq:pseudoinverse}) cannot be used.
+%More generally, the pseudo-inverse is best computed using the compact
+%singular value decomposition (SVD) of $\mathbf{A}$ which is:
+%\begin{equation*}
+%    \label{eq:compactsvd}
+%    \mathbf{A}=
+%    \mathbf{U_1}
+%    \boldsymbol \Sigma_1
+%    \mathbf{V}_1^{\top} \, .
+%\end{equation*}
+%Pseudo-inverse can then be written as follows:
+%\begin{equation*}
+%\label{eq:pseudoinversesvd}
+%\mathbf{A}^{\!\!+} = \mathbf{V}_1 \boldsymbol \Sigma_1^{-1} \mathbf{U}_1^\top \, .
+%\end{equation*}
+
+\subsection{Ridge regression method}
+
+In order to avoid the singularity of the matrix $\mathbf{A}^\top \mathbf{A}$, a regularisation term is introduced:
+
+\begin{equation}
+\label{eq:RRproblem} 
+J(\mathbf{\mathbf{X}}) =  \| \mathbf{A}\mathbf{\mathbf{X}} - \mathbf{Y} \|_2^2  + \lambda
+ \| \mathbf{\mathbf{X}}\| ^2
+\end{equation}
+
+\noindent which optimal solution $\mathbf{\mathbf{X}}_*$ is well known:
+
+\begin{equation}
+\label{eq:optsolRR}
+\mathbf{\mathbf{X}}_*=(\mathbf{A}^\top \mathbf{A}+\lambda \mathbb{I})^{-1}\mathbf{A}^\top y \, ,
+\end{equation}
+
+\subsection{Online learning}
+Online learning is a supervised machine learning framework that is useful when
+we have sequential access to a sample only once.  This differs from the
+classical batch learning, where there is an entire data set available and the
+learner can build the internal model without any limits in accessing the data.
+In batch learning, there is time enough to carefully analyse the dataset, build
+large predictive models and combine them in a sophisticated way. However, when
+execution time is critical and new data is required to be included in the model,
+online learning algorithms are a better option.
+
+In batch learning, the training phase is commonly a computationally expensive
+process. Therefore, when new data arrives it can't be included easily to the
+model. Moreover, it could happen that we won't have enough time to process the
+new data before more data arrives. In online learning algorithms there is no
+training phase, but the model is updated and evaluated at every time step. This
+model updating is computationally less expensive than a training phase.
+Online algorithms allow incremental learning by processing one instance at a
+time. This is done updating the current model instead of building the model from
+scratch.
+
+Online learning is also useful when some past data may be irrelevant
+or we want to improve computational efficiency. However, the use of less
+historical data could affect accuracy.
+
+The online learning framework was first introduced in the perceptron algorithm
+\cite{rosenblatt58}. There are several other widely used online methods such as
+passive-aggressive \cite{crammerETall2006}, stochastic gradient descent
+\cite{zhang2004}, aggregating algorithm \cite{vovk2001} and the second order
+perceptron \cite{cesa-bianchi2005}.  In \cite{cesa-bianchi2006} an in-depth
+analysis of online learning is provided.
+
+Algorithm~\ref{alg:onlinealg} shows the online learning algorithm structure:
+
+\begin{algorithm}[ht]
+\begin{algorithmic}[1]
+    \STATE Receives input $\mathbf{x}_t$
+    \STATE Makes prediction $\mathbf{\hat{y}}_t$
+    \STATE Receives response $\mathbf{y}_t$
+    \STATE Incurs loss $l_t(\mathbf{y}_t,\mathbf{\hat{y}}_t)$
+\end{algorithmic}
+\caption{Structure of a Learning System}
+\label{alg:onlinealg}
+\end{algorithm}
+
+\noindent where $l$ is some loss function. Performance is later measured after
+$T$ trials as:
+
 \begin{equation*}
-    \label{eq:compactsvd}
-    \mathbf{A}=
-    \mathbf{U_1}
-    \boldsymbol \Sigma_1
-    \mathbf{V}_1^{\top} \, .
+L_T = \sum_{t=1}^T l_t(\mathbf{y}_t,\mathbf{\hat{y}}_t)
 \end{equation*}
-Pseudo-inverse can then be written as follows:
-\begin{equation*}
-\label{eq:pseudoinversesvd}
-\mathbf{A}^{\!\!+} = \mathbf{V}_1 \boldsymbol \Sigma_1^{-1} \mathbf{U}_1^\top \, .
-\end{equation*}
+
+Competitive analysis was designed for analysing the performance of an online algorithm compared with its optimal offline algorithm.
+An optimal offline algorithm can view the sequences of requests in advance.
+The effectiveness of an online algorithm \cite{sleator1985} may be measured by its competitive ratio which is the worst-case ratio of the online algorithm and the the optimal offline algorithm.
+
+\subsection{Online Ridge Regression}
+
+Online RR is the online formulation of the regularised Least Squares method
+and is based on the following equivalent formulation of the RR optimal solution.
+
+%Since 
+%\begin{equation}
+%\label{eq:notation}
+%	\mathbf{A} = 
+%\left[
+%  \begin{tabular}{c>{$}c<{$}c}
+%    --- & \mathbf{a}^{\top}_{1} & ---\\
+%    --- & \mathbf{a}^{\top}_{2} & ---\\
+%    & \vdots & \\
+%    --- & \mathbf{a}^{\top}_{m} & ---
+%  \end{tabular}
+%\right]
+%\quad \text{and} \quad
+%\mathbf{B} =
+%\left[
+%  \begin{tabular}{c>{$}c<{$}c}
+%    --- & \mathbf{b}_{1} & ---\\
+%    --- & \mathbf{b}_{2} & ---\\
+%    & \vdots & \\
+%    --- & \mathbf{b}_{m} & ---
+%  \end{tabular}
+%\right]
+%\end{equation}
+
+Equation~(\ref{eq:optsolRR}) can also be written as:
+\begin{eqnarray*}
+\label{eq:RReapand}
+\mathbf{\mathbf{X}}_{\text{ridge}}&=&(\mathbf{A}^\top \mathbf{A}+ \lambda
+\mathbb{I})^{-1}\mathbf{A}^\top \mathbf{Y} \\
+&=& \displaystyle \big (\sum_{t=1}^m
+\mathbf{a}_t \mathbf{a}_t  ^\top + \lambda \mathbb{I}\big )^{-1}
+\sum_{t=1}^m \mathbf{a}_t \mathbf{y}_t \, .
+\end{eqnarray*}
+
+Lets define $\displaystyle\mathbf{S}= \sum_{t=1}^m \mathbf{a}_t
+\mathbf{a}_t  ^\top + \lambda \mathbb{I} $ and $\mathbf{W}=
+\displaystyle\sum_{t=1}^m \mathbf{a}_t \mathbf{y}_t$, so the
+algorithm~\ref{alg:RR} shows the iterative formulation:
+
+\begin{algorithm}[H]
+\begin{algorithmic}[1]
+\REQUIRE $\,$ \\
+$\{\mathbf{a}_1,\dots,\mathbf{a}_{m} \}$: $m$ input vectors \\
+$\{\mathbf{y}_1,\dots,\mathbf{y}_{m} \}$: $m$ target vectors \\
+$\lambda$: regularization parameter \\
+\ENSURE  $\,$ \\
+$\{f(\mathbf{a}_1),\dots,f(\mathbf{a}_{m}) \}$: model predictions \\
+\STATE Initialize $\mathbf{S}=\lambda \mathbb{I}$
+and $\mathbf{W}=0$
+\FOR { $t = 1$ to $m$ }
+	\STATE read new $\mathbf{a}_t$
+	\STATE $\mathbf{X}=\mathbf{S}^{-1}\mathbf{W}$
+	\STATE output prediction $f(\mathbf{a}_t) = \mathbf{X}^\top \mathbf{a}_t$
+   	\STATE $\mathbf{S} = \mathbf{S} + \mathbf{a}_t \mathbf{a}_t^\top$
+   	\STATE Read new $y_t$
+    	\STATE $\mathbf{W} = \mathbf{W} + \mathbf{a}_t \mathbf{y}_t$
+\ENDFOR
+\end{algorithmic}
+\caption{Online Ridge Regression}
+\label{alg:RR}
+\end{algorithm}
+
+\subsection{The Aggregating Algorithm for Regression}
+
+The AAR, proposed by~\cite{vovk2001} (also known as Vovk-Azoury-Warmuth predictor ~\cite{azoury2001}), is an application of the aggregating
+algorithm to the problem of regression. The idea is introduce the new input
+vector $\mathbf{x}_{m+1}$ to solve the model parameters:
+
+\begin{equation}
+\label{eq:AARexpand}
+\mathbf{X}_{aar} = \displaystyle \big (\sum_{t=1}^{m+1}
+\mathbf{a}_t \mathbf{a}_t  ^\intercal + \gamma \mathbb{I}\big )^{-1}
+\sum_{t=1}^m \mathbf{a}_t \mathbf{y}_t \, .
+\end{equation}
+
+If we define $\displaystyle\mathbf{S}= \sum_{t=1}^{m+1} \mathbf{a}_t
+\mathbf{a}_t  ^\intercal + \gamma \mathbb{I} $ and $\mathbf{W}=
+\displaystyle\sum_{t=1}^m \mathbf{a}_t \mathbf{y}_t$, the
+algorithm~\ref{alg:AAR} is slightly different to the algorithm~\ref{alg:RR}, 
+which updated matrix $\mathbf{S}$ before making the prediction:
+
+\begin{algorithm}[ht]
+\begin{algorithmic}[1]
+\REQUIRE $\,$ \\
+$\{\mathbf{a}_1,\dots,\mathbf{a}_{m} \}$: $m$ input vectors \\
+$\{\mathbf{y}_1,\dots,\mathbf{y}_{m} \}$: $m$ target vectors \\
+$\lambda$: regularisation parameter \\
+\ENSURE  $\,$ \\
+$\{f(\mathbf{a}_1),\dots,f(\mathbf{a}_{m}) \}$: model predictions \\
+\STATE Initialize $\mathbf{S}=\lambda \mathbb{I}$
+and $\mathbf{W}=0$
+\FOR { $t = 1$ to $m$ }
+	\STATE read new $\mathbf{a}_t$
+   	\STATE $\mathbf{S} = \mathbf{S} + \mathbf{a}_t \mathbf{a}_t^\intercal$
+	\STATE $\mathbf{X}=\mathbf{S}^{-1}\mathbf{W}$
+	\STATE output prediction $f(\mathbf{a}_t) = \mathbf{X}^\intercal \mathbf{a}_t$
+   	\STATE Read new $\mathbf{y}_t$
+    	\STATE $\mathbf{W} = \mathbf{W} + \mathbf{a}_t \mathbf{y}_t$
+\ENDFOR
+\end{algorithmic}
+\caption{{\em The aggregating algorithm for regression}}
+\label{alg:AAR}
+\end{algorithm}
 
@@
 \subsection{Data} \label{sec:unitroot}
+
+Experiments were run with data starting at 9:00 GMT of the 11th of August 2014,
+which corresponds to the opening of the New York financial market.
+
 %AVECM tests were carried out using four foreign exchange rates all related to
@@
   \centering
-  \includegraphics[width=0.4\textwidth]{forexdata}
+  \includegraphics[width=0.5\textwidth]{forexdata}
   \caption{10 seconds frequency forex data for EURUSD, GBPUSD and USDJPY}
@@
   \centering
-  \includegraphics[width=0.4\textwidth]{distdata}
+  \includegraphics[width=0.5\textwidth]{distdata}
   \caption{Distribution of EURUSD, GBPUSD and USDJPY data}
@@
   \centering
-  \includegraphics[width=0.4\textwidth]{onlinecomparison}
+  \includegraphics[width=0.5\textwidth]{onlinecomparison}
   \caption{Cumulative loss of AAR, RR and OLS solutions for VECM against its optimal algorithm}
@@
   \centering
-  \includegraphics[width=0.4\textwidth]{RRcomparison}
+  \includegraphics[width=0.5\textwidth]{RRcomparison}
   \caption{Cumulative loss RR VECM against its optimal algorithm using different $\lambda$ values}
@@
   \centering
-  \includegraphics[width=0.4\textwidth]{AARcomparison}
+  \includegraphics[width=0.5\textwidth]{AARcomparison}
   \caption{Cumulative loss AAR VECM against its optimal algorithm using different $\lambda$ values}

### 2015-06-19 src/4_OnlineLearning.tex
Avance articulo falta optimal algoritmo

@@
 process. Therefore, when new data arrives it can't be included easily to the
-model. Moreover, it could happen that we won't have enought time to process the
+model. Moreover, it could happen that we won't have enough time to process the
 new data before more data arrives. In online learning algorithms there is no
@@
 An optimal offline algorithm can view the sequences of requests in advance.
-The effectiveness of an online algorithm \cite{} may be measured by its competitive ratio which is the worst-case ratio of the online algorithm and the the optimal offline algorithm.
+The effectiveness of an online algorithm \cite{sleator1985} may be measured by its competitive ratio which is the worst-case ratio of the online algorithm and the the optimal offline algorithm.
 

### 2015-06-23 manuscript/manuscript.tex
saving my work

@@
 
-Vector error correction model (VECM) introduces the long-run relationship among a set of cointegrated variables
+The vector error correction model (VECM) introduces the long-run relationship among a set of cointegrated variables
 as an error correction term. VECM is a special case of the vector autoregressive
@@
 adjusts coefficients to bring the variables back to equilibrium. In finance,
-many economic time series (foreign exchange market rates for example) turns out to be stationary, when they are differentiated once ($I(1)$) and
-cointegration restrictions applied on the model often improves forecasting~\cite{duy1998}. Therefore,
-VECM has been widely adopted.
+many economic time series turns out to be stationary, when they are differentiated once (see section \ref{sec:coint} for more details).
+Additionally, cointegration restrictions applied on the model often improves forecasting~\cite{duy1998}.
+This is the reason why VECM has been widely adopted.
 
 VECM parameters are obtained using ordinary least squares
-(OLS) method \cite{golub1980}. OLS has two main problems: is sensitive to errors on input data
-and involves many calculations. The former problem is commonly solved using
+(OLS) method \cite{golub1980}. OLS has two main problems: it is sensitive to errors on input data
+and also involves many calculations. The former problem is commonly solved using
 Ridge Regression (RR) \cite{hoerl1970} which introduces a regularisation
@@
 Our proposal is an online formulation of the VECM called Online VECM (OVECM)
-based on consideration of only a sliding window of the historical data.  OVECM parameters are obtained
-via OLS, RR or AAR. A competitive analysis is done comparing its cumulative losses against the optimal offline algorithm.
+based on the consideration of only a sliding window of the historical data at every step.
+OVECM parameters are obtained via OLS, RR or AAR.
+In order to measure performance, a competitive analysis is done which compares the proposal cumulative loss against the optimal offline algorithm.
 
 In finance, little academic research has been done using high-frequency Forex
-rates \cite{Bekiros2015}. Our proposal is tested using four
+rates \cite{Bekiros2015}. Our proposal is tested using
 three mayor currency pairs (MCP) from the foreign exchange market:  Euro (EUR) to United States Dollar (USD) (EURUSD), British Pound (GBP) to
@@
 \widehat{\mathbf{X}} = \underset{\mathbf{X}}{\text{Arg\;min}}
-\|\mathbf{A}\mathbf{X}-\mathbf{B}\|_2^2
+\|\mathbf{A}\mathbf{X}-\mathbf{Y}\|_2^2
 \end{equation*}
@@
 \label{eq:MP}
-\widehat{\mathbf{X}}=\mathbf{A}^{\!\!+}\,\mathbf{B}
+\widehat{\mathbf{X}}=\mathbf{A}^{\!\!+}\,\mathbf{Y}
 \end{equation*}
@@
 \label{eq:RRproblem} 
-J(\mathbf{\mathbf{X}}) =  \| \mathbf{A}\mathbf{\mathbf{X}} - \mathbf{Y} \|_2^2  + \lambda
+\widehat{\mathbf{X}} (\lambda) = \underset{\mathbf{X}}{\text{Arg\;min}}  \| \mathbf{A}\mathbf{\mathbf{X}} - \mathbf{Y} \|_2^2  + \lambda
  \| \mathbf{\mathbf{X}}\| ^2
@@
 
-\noindent which optimal solution $\mathbf{\mathbf{X}}_*$ is well known:
+\noindent which optimal solution is well known:
 
@@
 \label{eq:optsolRR}
-\mathbf{\mathbf{X}}_*=(\mathbf{A}^\top \mathbf{A}+\lambda \mathbb{I})^{-1}\mathbf{A}^\top y \, ,
+\widehat{\mathbf{X}} (\lambda)=(\mathbf{A}^\top \mathbf{A}+\lambda \mathbb{I})^{-1}\mathbf{A}^\top \mathbf{Y} \, ,
 \end{equation}
@@
 process. Therefore, when new data arrives it can't be included easily to the
-model. Moreover, it could happen that we won't have enough time to process the
+model. Moreover, it could happen that there is  won't have enough time to process the
 new data before more data arrives. In online learning algorithms there is no

### 2015-06-23 manuscript/manuscript.tex
Primera version articulo

@@
 \documentclass[twocolumn]{svjour3}          % twocolumn
-%
 \smartqed  % flush right qed marks, e.g. at end of proof
-%
 \usepackage{graphicx}
-%
-% \usepackage{mathptmx}      % use Times fonts if available on your TeX system
-%
-% insert here the call for the packages your document requires
-%\usepackage{latexsym}
-% etc.
-%
 \usepackage{amsfonts,amssymb,amsmath,
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\usepackage{xcolor}%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\usepackage{hyperref}%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\usepackage{pmat}%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%\newtheorem*{myteo}{Teorema}
-%\newtheorem*{mydef}{Definici\'on}
-%% MYREMARK ENVIRONMENT %%%%%%%%%%%
-\newenvironment{mycolor}{\color{blue}}{}
-\newcounter{myremarkcount}
-\setcounter{myremarkcount}{0}
-\newsavebox{\myremarkbox}
-\newenvironment{myremark}
-{% This is the begin code
-\medskip
-\begin{flushright}
-\begin{lrbox}{\myremarkbox}%
-\begin{minipage}[t]{1.00\linewidth}
-\stepcounter{myremarkcount}
-\footnotesize
-\begin{mycolor}
-{{\bf Proofreading Remark.}\ }
-% \begin{it}
-}
-{% This is the end code
-% \end{it}
-\end{mycolor}
-\end{minipage}
-\end{lrbox}
-\fbox{\usebox{\myremarkbox}}
-\end{flushright}
-
-\bigskip}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\usepackage{xcolor}
+\usepackage{hyperref}
+\usepackage{pmat}
 
@@
 
-\title{High frequency Intraday Forex rates forecasting using an online cointegration approach}
+\title{High frequency intraday forex rates forecasting using an online cointegration approach}
 %\subtitle{Do you have a subtitle?\\ If so, write it here}
@@
 available. We use an online version of the linear ridge regression and also a modification introduced by 
-Vovk \cite{vovk2001} called the Aggregating algorithm regression (AAR), which is also known as Vovk-Azoury-Warmuth predictor ~\cite{azoury2001}.
-
+Vovk \cite{vovk2001} called the Aggregating algorithm regression (AAR).
 Our proposal is an online formulation of the VECM called Online VECM (OVECM)
@@
 \end{equation}
-%when $\mathbf{A}$ is not full rank, i.e
-%$rank(\mathbf{A})=k <  n \leq m$, $\mathbf{A}^\top \mathbf{A}$ is
-%always singular and equation~(\ref{eq:pseudoinverse}) cannot be used.
-%More generally, the pseudo-inverse is best computed using the compact
-%singular value decomposition (SVD) of $\mathbf{A}$ which is:
-%\begin{equation*}
-%    \label{eq:compactsvd}
-%    \mathbf{A}=
-%    \mathbf{U_1}
-%    \boldsymbol \Sigma_1
-%    \mathbf{V}_1^{\top} \, .
-%\end{equation*}
-%Pseudo-inverse can then be written as follows:
-%\begin{equation*}
-%\label{eq:pseudoinversesvd}
-%\mathbf{A}^{\!\!+} = \mathbf{V}_1 \boldsymbol \Sigma_1^{-1} \mathbf{U}_1^\top \, .
-%\end{equation*}
 
@@
 In batch learning, the training phase is commonly a computationally expensive
-process. Therefore, when new data arrives it can't be included easily to the
-model. Moreover, it could happen that there is  won't have enough time to process the
+process. Therefore, when new data arrives it can't be included easily into the
+model. Moreover, it could happen that it won't be enough time to process the
 new data before more data arrives. In online learning algorithms there is no
@@
 \noindent where $l$ is some loss function. Performance is later measured after
-$T$ trials as:
+$T$ trials using the cumulative loss function:
 
@@
 
-Competitive analysis was designed for analysing the performance of an online algorithm compared with its optimal offline algorithm.
-An optimal offline algorithm can view the sequences of requests in advance.
-The effectiveness of an online algorithm \cite{sleator1985} may be measured by its competitive ratio which is the worst-case ratio of the online algorithm and the the optimal offline algorithm.
+This cumulative loss function is compared against the cumulative loss function of its its optimal offline algorithm. An optimal offline algorithm can view the sequences of requests in advance. 
+This comparison is called competitive analysis  \cite{sleator1985}  which was designed for analysing the performance of an online algorithm. 
+The effectiveness of an online algorithm may be measured by its competitive ratio which is the worst-case ratio of the online algorithm and the the optimal offline algorithm.
 
@@
 
-Online RR is the online formulation of the regularised Least Squares method
+Online RR is the online formulation of the regularised OLS method
 and is based on the following equivalent formulation of the RR optimal solution.
-
-%Since 
-%\begin{equation}
-%\label{eq:notation}
-%	\mathbf{A} = 
-%\left[
-%  \begin{tabular}{c>{$}c<{$}c}
-%    --- & \mathbf{a}^{\top}_{1} & ---\\
-%    --- & \mathbf{a}^{\top}_{2} & ---\\
-%    & \vdots & \\
-%    --- & \mathbf{a}^{\top}_{m} & ---
-%  \end{tabular}
-%\right]
-%\quad \text{and} \quad
-%\mathbf{B} =
-%\left[
-%  \begin{tabular}{c>{$}c<{$}c}
-%    --- & \mathbf{b}_{1} & ---\\
-%    --- & \mathbf{b}_{2} & ---\\
-%    & \vdots & \\
-%    --- & \mathbf{b}_{m} & ---
-%  \end{tabular}
-%\right]
-%\end{equation}
 
@@
 \end{eqnarray*}
+
+\noindent where 
+\begin{equation}
+\label{eq:notation}
+	\mathbf{A} = 
+\left[
+  \begin{tabular}{c>{$}c<{$}c}
+    -- & \mathbf{a}^{\top}_{1} & --\\
+    -- & \mathbf{a}^{\top}_{2} & --\\
+    & \vdots & \\
+    -- & \mathbf{a}^{\top}_{m} & --
+  \end{tabular}
+\right]
+\quad \text{and} \quad
+\mathbf{B} =
+\left[
+  \begin{tabular}{c>{$}c<{$}c}
+    -- & \mathbf{b}_{1} & --\\
+    -- & \mathbf{b}_{2} & --\\
+    & \vdots & \\
+    -- & \mathbf{b}_{m} & --
+  \end{tabular}
+\right]
+\end{equation}
 
@@
 The AAR, proposed by~\cite{vovk2001} (also known as Vovk-Azoury-Warmuth predictor ~\cite{azoury2001}), is an application of the aggregating
-algorithm to the problem of regression. The idea is introduce the new input
+algorithm to the problem of regression. The idea is to introduce the new input
 vector $\mathbf{x}_{m+1}$ to solve the model parameters:
@@
 \mathbf{X}_{aar} = \displaystyle \big (\sum_{t=1}^{m+1}
-\mathbf{a}_t \mathbf{a}_t  ^\intercal + \gamma \mathbb{I}\big )^{-1}
+\mathbf{a}_t \mathbf{a}_t  ^\top+ \gamma \mathbb{I}\big )^{-1}
 \sum_{t=1}^m \mathbf{a}_t \mathbf{y}_t \, .
@@
 If we define $\displaystyle\mathbf{S}= \sum_{t=1}^{m+1} \mathbf{a}_t
-\mathbf{a}_t  ^\intercal + \gamma \mathbb{I} $ and $\mathbf{W}=
+\mathbf{a}_t  ^\top + \gamma \mathbb{I} $ and $\mathbf{W}=
 \displaystyle\sum_{t=1}^m \mathbf{a}_t \mathbf{y}_t$, the
@@
 VECM (OVECM) capable of updating the model with new arrival data and give
-responses in a short period of time. OVECM only considers a time varying window
+responses in a shorter period of time. OVECM only considers a time varying window
 of historical data and the using of RR or its variant AAR (Aggregating Algorithm
@@
 
-On the other hand, getting cointegration vectors using Johansen method is also
-an expensive procedure. However, since cointegration
-vectors represent the long-run relationship between the time series, they
-vary little in time. Our proposal obtains new cointegration vectors only when
-this long-term relationship changes. This change is detected by tracking 
-the Mean Absolute Percentage Error (MAPE) of the last $n$ in-sample forecasts.
-
-%Since VECM is a model based on time series differences, the MAPE is obtained
-%from $\Delta \mathbf{y}$ as following:
-% 
-%\begin{equation}\label{eq:MAPE}
-%\text{MAPE}[t] = \frac{1}{n} \sum_{i=1}^{n} \left| 
-%\frac{\Delta \mathbf{y}_{\text{true}}[t-i]-\Delta
-%\mathbf{y}_{\text{pred}}[t-i]}{\Delta \mathbf{y}_{\text{true}[t-i]}}
-%\right| \, , 
-%\end{equation}
-%
-%\noindent where $\Delta \mathbf{y}_{\text{true}}$ is the actual value of the
-%time series $\mathbf{y}$ differences and $\Delta \mathbf{y}_{\text{pred}}$ is
-%their forecast value.
-
-%The method proposed consists of
-%a modification of RR considering a sliding window which contains only
-%the last $L$ samples %and the new input $\mathbf{x}_t$, 
-%i.e. $\{\mathbf{x}_i\}_{i=t-L+1}^{t-1}$.
-
-VECM($p$) for $l$ cointegrated variables has the following form:
-
-\begin{align}
-\label{eq:vecfull}
-\Delta\mathbf{y}_t 
-= \boldsymbol{\alpha\beta}^\top\mathbf{y}_{t-1} 
-  + \sum_{i=1}^{p-1}\boldsymbol{\Phi}_i^*\,\Delta\mathbf{y}_{t-i}
-  + \mathbf{c} + \boldsymbol{\epsilon}_t  \nonumber \\
-t=p+1,\dots,N,
-\end{align}
-
-Transposing each equation of the system (\ref{eq:vecfull}) we can write
-the VECM($p$) model in block-matrix form as:
-\begin{equation}\label{eq:vareq}
-\mathbf{B} = 
-\mathbf{A} \mathbf{X} + 
-\mathbf{E} \, , 
-\end{equation}
-%
-\noindent where $\mathbf{B}$ dimension is $((N-p)\times l)$, $\mathbf{A}$
-dimension is $((N-p)\times(r+(p-1)l +1))$, $\mathbf{X}$ dimension is $((r+(p-1)l
-+1)\times l)$, $\mathbf{E}$ dimension is $((N-p)\times l)$ and $r$ is the number
-of cointegration vectors:
-%
-\begin{alignat}{3}
-\mathbf{B}
-&= \begin{bmatrix}
-   \Delta\mathbf{y}_{p+1}^\top \\
-   \Delta\mathbf{y}_{p+2}^\top \\
-   \vdots \\
-   \Delta\mathbf{y}_N^\top
-   \end{bmatrix}
-&\quad
-\mathbf{X}
-&= \begin{bmatrix}
-   \boldsymbol{\alpha}^\top \\
-   \boldsymbol{\Phi}_1^{*\top} \\
-   \boldsymbol{\Phi}_2^{*\top} \\
-   \vdots \\
-   \boldsymbol{\Phi}_{p-1}^{*\top} \\
-   \mathbf{c}^\top
-   \end{bmatrix}
-&\quad
-\mathbf{E}
-&= \begin{bmatrix}
-   \boldsymbol{\epsilon}_{p+1}^\top \\
-   \boldsymbol{\epsilon}_{p+2}^\top \\
-   \vdots \\
-   \boldsymbol{\epsilon}_N^\top \\
-   \end{bmatrix}
-\end{alignat}
-\noindent and 
-\begin{align}
-\mathbf{A} 
-&= \begin{pmat}[{....|}]
-   \mathbf{y}_p^\top \boldsymbol{\beta} & \Delta \mathbf{y}_p^\top & \Delta\mathbf{y}_{p-1}^\top & \dots 
-                    & \Delta\mathbf{y}_2^\top & 1 \cr
-   \mathbf{y}_{p+1}^\top  \boldsymbol{\beta} &\Delta\mathbf{y}_{p+1}^\top & \Delta\mathbf{y}_p^\top & \dots
-                       & \Delta\mathbf{y}_3^\top & 1 \cr
-   \vdots & \vdots & \vdots & \ddots & \vdots & \vdots \cr
-   \mathbf{y}_{N-1}^\top  \boldsymbol{\beta} &\Delta\mathbf{y}_{N-1}^\top & \Delta\mathbf{y}_{N-2}^\top & \dots 
-                       & \Delta\mathbf{y}_{N-p-1}^\top & 1 \cr
-   \end{pmat}\, .
-\end{align}
-%Taking into account the error term $\mathbf{E}$, equation~(\ref{eq:vareq}) 
-%can be solved with respect to $\mathbf{X}$ using the ordinary least
-%squares estimation.
-However, the number of rows of matrix $\mathbf{A}$ increases with the number of
-observations $\mathbf{y}_t$. Our proposal considers only the most recent $L$
-rows of matrices $\mathbf{A}$ and $\mathbf{B}$ defined as  $\mathbf{A}(t)$ and
-$\mathbf{B}(t)$:
+Our proposal considers only the most recent $L$
+rows of matrices $\mathbf{A}$ and $\mathbf{B}$ defined as  $\mathbf{A}_t$ and
+$\mathbf{B}_t$:
 
@@
 \label{eq:notation}
-	\mathbf{A}(t) = 
+	\mathbf{A}_t = 
 \left[
   \begin{tabular}{c>{$}c<{$}c}
-    --- & \mathbf{a}^{\top}_{t-L} & ---\\
-    --- & \mathbf{a}^{\top}_{t-L+1} & ---\\
+    -- & \mathbf{a}^{\top}_{t-L} & --\\
+    %-- & \mathbf{a}^{\top}_{t-L+1} & --\\
     & \vdots & \\
-    --- & \mathbf{a}^{\top}_{t} & ---
+    -- & \mathbf{a}^{\top}_{t} & --
   \end{tabular}
 \right]
-\quad \text{and} \quad
-\mathbf{B}(t) =
+,
+\mathbf{B}_t=
 \left[
   \begin{tabular}{c>{$}c<{$}c}
-    --- & \mathbf{b}_{t-L} & ---\\
-    --- & \mathbf{b}_{t-L+1} & ---\\
+    -- & \mathbf{b}_{t-L} & --\\
+    %-- & \mathbf{b}_{t-L+1} & --\\
     & \vdots & \\
-    --- & \mathbf{b}_{t} & ---
+    -- & \mathbf{b}_{t} & --
   \end{tabular}
@@
 
-\noindent so that VECM parameters $\mathbf{X}(t)$ are found using the following
+\noindent so that VECM parameters $\mathbf{X}_t$ are found using the following
 equation:
@@
 \begin{equation}
-\mathbf{B}(t) = \mathbf{A}(t) \mathbf{X}(t) + \mathbf{E}(t)\\
+\mathbf{Y}_t = \mathbf{A}_t \mathbf{X}_t + \mathbf{E}_t\\
 \end{equation}
 
-The RR solution $\mathbf{\hat{X}}(t)$ using the sliding window matrices defined above
+The RR solution $\mathbf{\hat{X}}_t(\lambda)$ using the sliding window matrices defined above
 is:
@@
 \label{eq:oproblem}
-\mathbf{\hat{X}}(t)=\mathbf{S}(t)^{-1} \mathbf{W}(t) \, ,
+\mathbf{\hat{X}}_t(\lambda)=\mathbf{S}_t^{-1} \mathbf{W}(t) \, ,
 \end{equation}
@@
 \begin{eqnarray*}
-\mathbf{S}(t) =&{\bf A}(t)^\top{\bf A}(t)+ \lambda \mathbb{I} &=
+\mathbf{S}_t =&{\bf A}_t^\top{\bf A}_t+ \lambda \mathbb{I} &=
 \sum_{i=0}^L \mathbf{a}_{t-i}\mathbf{a}_{t-i}^\top + \lambda \mathbb{I}
 \label{eq:S} \\
-\mathbf{W}(t) =& {\bf A}(t)^\top{\bf B}(t) &= 
+\mathbf{W}_t =& {\bf A}_t^\top{\bf B}_t &= 
 \sum_{i=0}^L \mathbf{a}_{t-i}\mathbf{b}_{t-i} 
@@
 
-It is worth noticing that, at the next time step, the matrices $\mathbf{S}(t+1)$
-and $\mathbf{W}(t+1)$ are slightly different to $\mathbf{S}(t)$ and
-$\mathbf{W}(t)$:
+It is worth noticing that, at the next time step, the matrices $\mathbf{S}_{t+1}$
+and $\mathbf{W}_{t+1}$ are slightly different to $\mathbf{S}_t$ and
+$\mathbf{W}_t$:
 
 \begin{eqnarray*}
-\mathbf{S}(t+1)&=&
-\mathbf{S}(t) +
+\mathbf{S}_{t+1}&=&
+\mathbf{S}_t +
 \mathbf{a}_{t+1}
@@
 \mathbf{a}_{t-L} \mathbf{a}_{t-L}^\top \\
-\mathbf{W}(t+1)&=&
-\mathbf{W}(t) +
+\mathbf{W}_{t+1}&=&
+\mathbf{W}_t +
 \mathbf{a}_{t+1}
@@
 
-%Agregar solo si se decide agregar Sherman-Morrison
-%Therefore the Sherman-Morrison-Woodbury formula can be used to reduce inverse matrix
-%calculations.
+Since we need to obtain the inverse of $\mathbf{S}_{t+1}$ at every step,
+we used the Sherman-Morrison-Woodbury formula in order to reduce the number of operations. 
+The Sherman-Morrison-Woodbury formula states if a matrix $\mathbf{A}$ is a positive
+definite matrix and its inverse matrix is known, then the inverse of the
+matrix $\mathbf{B}=\mathbf{A} + \mathbf{x}\mathbf{x}^\top$ can be
+obtained as:
+
+\begin{equation}
+\label{eq:SMW}
+\mathbf{B}^{-1} = \mathbf{A}^{-1} -
+\frac{(\mathbf{A}^{-1}\mathbf{x})(\mathbf{A}^{-1}\mathbf{x})^{\top}}
+{1 + \mathbf{x}^{\top} \mathbf{A}^{-1} \mathbf{x}} \, .
+\end{equation}
+
+This formula reduces the inverse matrix computation order from $O(n^3)$ to
+$O(n^2)$. In our case the inverse of $\mathbf{S}_{t+1}$ can be obtained applying twice the Sherman-Morrison-Woodbury formula.
 
@@
 methods for geting model parameters: OLS, RR, AAR which leads to three different
-%algorithms OVECM-OLS, OVECM-RR, OVECM-AAR:
+settings for OVECM:
 
@@
 	    \STATE{$v \gets \texttt{getJohansen}(\mathbf{Y},p)$}
-	    \STATE{$[\mathbf{A} \quad \mathbf{B}] \gets
+	    \STATE{$\mathbf{A} ,\quad \mathbf{B} \gets
         \texttt{vecMatrix}(\mathbf{Y},p,v)$}
@@
     \ENDIF
-	\STATE{[$\mathbf{A} \quad \mathbf{B} \quad \mathbf{a}_t \quad \mathbf{a}_L \quad \mathbf{b}_t \quad
-    \mathbf{b}_L ] \gets
+	\STATE{ $\mathbf{a}_t ,\quad \mathbf{b}_t \gets
     \texttt{vecMatrixOnline}(\mathbf{Y},\mathbf{y}_t,p,v)$}
-    \STATE $[\mathbf{X} \quad \mathbf{y}_{\text{pred}}[t]] \gets \texttt{solver.regression}
-    (\mathbf{A},\mathbf{B},\mathbf{a}_t,\mathbf{b}_t)$
-    \STATE $\mathbf{Y}_{\text{pred}} = \mathbf{AX}[-n:]+\mathbf{Y}[-n-1:-1]$
-    \STATE $\mathbf{e} = \texttt{mape}(\mathbf{Y}_{\text{pred}},\mathbf{Y}[-n:])$
-    \IF {$\texttt{mean}(\mathbf{e}) > \text{mean\_error}$}
+    \STATE $\mathbf{X} ,\quad \mathbf{Y}_{\text{pred}}[t] \gets \texttt{solver.regression}
+    (\mathbf{a}_t,\mathbf{b}_t)$
  	    \STATE{$v \gets \texttt{getJohansen}(\mathbf{Y},p)$}
-	    \STATE{$\mathbf{A} \gets
-        \texttt{vecMatrixUpdate}(\mathbf{A},\mathbf{Y},p,v)$}
-        \STATE \texttt{solver.init\_model($\mathbf{A},\mathbf{B}$)} 
-	    \STATE{$[\mathbf{a}_t \quad \mathbf{b}_t] \gets
-        \texttt{vecMatrixOnline}(\mathbf{Y},\mathbf{y}_t,p,v)$}
-        \STATE $[\mathbf{X} \quad \mathbf{y}_{\text{pred}}[t]] \gets \texttt{solver.regression}
-        (\mathbf{A},\mathbf{B},\mathbf{a}_t,\mathbf{b}_t)$
-    \ENDIF
-\STATE{$[\mathbf{A} \quad \mathbf{B}] \gets
-\texttt{matrixUpdate}(\mathbf{A},\mathbf{B},\mathbf{a}_t,\mathbf{b}t)$}
 \ENDFOR
@@
 Johansen method considering the trace statistic test at 95\% level of
-significance. This procedure is called at the first step of the algorithm and
-every time in-sample MAPE ($\mathbf{e}$) exceeds threshold error defined (mean\_error).
+significance. The first cointegration vector is only considered.
 \item The function \texttt{vecMatrix} returns VECM
-matrices $\mathbf{A}(t),\mathbf{B}(t)$ shown in equation~(\ref{eq:notation}). This
+matrices $\mathbf{A}_t,\mathbf{B}_t$ shown in equation~(\ref{eq:notation}). This
 method is only required at the first step.
@@
 \item The Solver class (see algorithm \ref{alg:solver}) obtain VECM parameters  
-using RR, AAR or OLS methods.
-\item The function \texttt{vecMatrixUpdate} updates matrix $\mathbf{A}$ when new
-cointegration vectors are required (matrix $\mathbf{B}$ is not affected). Model
-parameters $\mathbf{X}$ is also updated later.
-%\item The number of cointegration vectors is set as a parameter (aun no lo
-%coloco en el algoritmo).
-%\item In order to set VECM parameter: $L$ and $p$, we use the Akaike Information
-%Criterion (AIC). RR parameter $\lambda$ was done by cross-validation.
-%\item Obtain VECM parameter $\mathbf{X}(t)$ in equation \ref{eq:optsolSLAAR} requires
-%to calculate the inverse of matrix $\mathbf{S}(t+1)$ which is an expensive
-%routine.However, since we already know $\mathbf{S}(t)^{-1}$ we can use
-%Sherman-Morrison-Woodbury twice~\ref{eq:SMW}.
+using RR, AAR or OLS methods. This solver class obtains the inverse of matrix $\mathbf{S}_{t+1}$ using the  Sherman-Morrison-Woodbury twice (see equation \ref{eq:SMW}).
+\item In order to set VECM parameter: $L$ and $p$, we use the Akaike Information
+Criterion (AIC). RR parameter $\lambda$ was done by cross-validation.
 \end{itemize}
@@
 \STATE $\mathbf{W} = \displaystyle \sum_{i=1}^m \mathbf{a}_i \mathbf{b}_i$
+%\STATE{$ \mathbf{a}_L ,\quad  \mathbf{b}_L \gets \mathbf{A}[:,1]^\top,  \mathbf{B}[:,1] $} 
 \quad \\
 \texttt{regression}($\mathbf{A},\mathbf{B},\mathbf{a_t},\mathbf{b_t}$) \\
-\IF {method == 'OLS'}
-        \STATE $\mathbf{X}=(\mathbf{A}^\top \mathbf{A})^{-1}\mathbf{A}^\top
-        \mathbf{B}$
-\ELSIF {method == 'RR'}
+%\IF {method == 'OLS'}
+%        \STATE $\mathbf{X}=(\mathbf{A}^\top \mathbf{A})^{-1}\mathbf{A}^\top
+%        \mathbf{B}$
+\IF {method == 'RR' \OR method == 'OLS'}
         \STATE $\mathbf{X} = \mathbf{S}^{-1} \mathbf{W} $
@@
         \mathbf{a}_t \mathbf{a}_t^\top-
-        \mathbf{a}_1 \mathbf{a}_1^\top$
+        \mathbf{a}_{t-L} \mathbf{a}_{t-L}^\top$
         \STATE $\mathbf{W} = \mathbf{W} + \mathbf{a}_t \mathbf{b}_t$
@@
         \mathbf{a}_t \mathbf{a}_t^\top-
-        \mathbf{a}_1 \mathbf{a}_1^\top$
+        \mathbf{a}_{t-L} \mathbf{a}_{t-L}^\top$
         \STATE $\mathbf{X} = \mathbf{S}^{-1} \mathbf{W} $
@@
 \end{algorithm}
-% Aqui se puede hablar de evaluation using competitive analysis. The idea
-%of competitiveness is to compare the output generated by an online algorithm to
-%the output produced by an optimal offline algorithm. An optimal online algorithm
-%is an omniscient algorithm that knows the entire input data in advance and can
-%compute an optimal output. The better an online algorithm approximates the
-%optimal solution, the more competitive this algorithm is.
+
 The proposal was compared against an optimal online algorithm which is a time varying
@@
 
-%%Solo para tesis
-%\section{Online VEC model}
-%
-%If we want to update the VEC model with new input data
-%$\mathbf{y}_{T+1}$, we have to add a new row in matrices
-%$\mathbf{Y,X}$ and $\mathbf{E}$. 
-%
-%\begin{eqnarray}
-%\mathbf{Y} &=& 
-%                \left[ \begin{array}{ccc}
-%               \quad & \mathbf{\Delta y}_{p+1} & \quad \\ \hline
-%               \quad & \mathbf{\Delta y}_{p+2} & \quad \\ \hline
-%               \quad & \vdots & \quad \\ \hline 
-%               \quad & \mathbf{\Delta y}_T & \quad \\ 
-%               \quad & \color{red}\mathbf{\Delta y}_{T+1} & \quad 
-%               \end{array} \right]
-%             \\
-%\mathbf{\Phi^*} &=& 
-%                \left[ \begin{array}{ccc}
-%               \phi_{p-1}^* \\ 
-%               \vdots \\ 
-%               \phi_{1}^* \\
-%               \alpha \\
-%                c   
-%               \end{array} \right]
-%\\
-%\mathbf{X} &=& \begin{bmatrix} \label{offX}
-%   \mathbf{\Delta y}_2 & \dots & \mathbf{\Delta y}_{p-1} &
-%   \mathbf{\Delta y}_{p} & \beta'\mathbf{y}_{p} & 1\\
-%   \mathbf{\Delta y}_3 & \dots & \mathbf{\Delta y}_{p} &
-%   \mathbf{\Delta y}_{p+1} & \beta'\mathbf{y}_{p+1} &1\\
-%   \vdots &  \ddots & \vdots & \vdots & \vdots & \vdots\\
-%   \mathbf{\Delta y}_{T-p+1} & \dots & \mathbf{\Delta y}_{T-2} &
-%   \mathbf{\Delta y}_{T-1} & \beta'\mathbf{y}_{T-1} &1 \\
-%   \color{red} \mathbf{\Delta y}_{T-p+2} & \dots & \color{red}\mathbf{\Delta y}_{T-1} &
-%   \color{red}\mathbf{\Delta y}_{T} & \color{red}\beta'\mathbf{y}_{T}
-%   & \color{red} 1
-%   \end{bmatrix}
-%\\
-%\mathbf{E} &=& \begin{bmatrix}
-%              \mathbf{\epsilon}_{p+1} \\ \vdots \\ \mathbf{\epsilon}_T
-%              \\ \color{red}\mathbf{\epsilon}_{T+1}
-%             \end{bmatrix}
-%\end{eqnarray}
-
 \section{Experimental results}
@@
 
+OVECM tests were carried out using three foreign exchange rates all related to
+USD: EURUSD, GBPUSD and USDCHF. This data was collected from the free
+database Dukascopy which gives access to the Swiss Foreign Exchange marketplace
+~\cite{Dukascopy2014}.
+
+The tests were done using 10-seconds frequency from ask prices which
+corresponded to 8640 data points per day from the 11th to the 15th of August
+2014. 
 Experiments were run with data starting at 9:00 GMT of the 11th of August 2014,
-which corresponds to the opening of the New York financial market.
+which corresponds to the opening of the New York financial market. Figure \ref{fig:forexdata}  shows a sample of the data used and figure \ref{fig:distdata} shows distribution of the data.
 
-%AVECM tests were carried out using four foreign exchange rates all related to
-%USD: EURUSD, GBPUSD, USDCHF and USDJPY. This data was collected from the free
-%database Dukascopy which gives access to the Swiss Foreign Exchange marketplace
-%~\cite{Dukascopy2014}.
-%
-%The tests were done using 10-seconds frequency from ask prices which
-%corresponded to 8640 data points per day from the 11th to the 15th of August
-%2014. Since data datetimes were in GMT, opening market times for London, New
-%York, Sidney and Tokyo corresponded with data points 1440, 3240, 6120 and 6840
-%respectively. Our tests were made considering two hours after the opening time
-%of the 13th of August. Therefore, in order to obtain these times, we added 2
-%days and 2 hours to all the opening times ($2\times 8640 + 2 \times 360$) to obtain:
-%19440, 21240, 24120 and 24840. We called offset to these times in
-%table~\ref{tab:stats}, which allowed us to obtain performance at different
-%times of the day.
+\hspace*{-1.5in}
+\begin{figure}[!ht]
+  \vspace*{-1cm}
+  \hspace*{-0.4in}
+  \centering
+  \includegraphics[scale=0.34]{forexdata}
+  \caption{10 seconds frequency forex data for EURUSD, GBPUSD and USDJPY}
+  \label{fig:forexdata}
+\end{figure}
+
+\begin{figure}[!ht]
+  \vspace*{-1cm}
+  \hspace*{-0.3in}
+  \centering
+  \includegraphics[scale=0.3]{distdata}
+  \caption{Distribution of EURUSD, GBPUSD and USDJPY data}
+  \label{fig:distdata}
+\end{figure}
 
@@
 test but they rejected it with their first differences. This means that all of
-them are I(1) time series and we are allowed to use VECM and therefore AVECM.
+them are I(1) time series and we are allowed to use VECM and therefore OVECM.
 
@@
 $\Delta$ CHFUSD & -74.21940   & -1.94101 & False       \\
-JPYUSD          &  0.36721    & -1.94101 & True        \\
-$\Delta$ JPYUSD & -101.15589  & -1.94101 & False     \\ 
 \hline
@@
 \end{center}
-\caption{Unit roots tests for EURUSD, GBPUSD, USDCHF and USDJPY at 10-seconds
+\caption{Unit roots tests for EURUSD, GBPUSD and USDCHF at 10-seconds
 frequency.}
@@
 
-\begin{figure}[!ht]
-  %\vspace{-0.8cm}
-  \centering
-  \includegraphics[width=0.5\textwidth]{forexdata}
-  \caption{10 seconds frequency forex data for EURUSD, GBPUSD and USDJPY}
-  \label{fig:forexdata}
-\end{figure}
+Competitive analysis were done considering the optimal offline algorithm (OOVECM).
+Cumulative loss is shown for different settings of OVECM.
+Figure \ref{fig:RRcomparison}  shows a comparison between OVECM using RR with different values for $\lambda$.
+It should be notice that RR with $\lambda = 0$ setting is equivalent to OLS.
+Experiments show that it is possible to reduce cumulative loss using some $\lambda \neq 0$.
 
@@
   \centering
-  \includegraphics[width=0.5\textwidth]{distdata}
-  \caption{Distribution of EURUSD, GBPUSD and USDJPY data}
-  \label{fig:distdata}
+  \includegraphics[width=0.5\textwidth]{RRcomparison}
+  \caption{Cumulative loss RR VECM against its optimal algorithm using different $\lambda$ values}
+  \label{fig:RRcomparison}
 \end{figure}
+
+Same experiments were done using the AAR setting for different values for $\lambda$.
+The effectiveness of the regularization parameter is so clear than RR.
+However, AAR seems to be a good predictor compared with OOVECM.
+Results are shown in figure \ref{fig:AARcomparison}.
+
+\begin{figure}[!ht]
+  %\vspace{-0.8cm}
+  \centering
+  \includegraphics[width=0.5\textwidth]{AARcomparison}
+  \caption{Cumulative loss AAR VECM against its optimal algorithm using different $\lambda$ values}
+  \label{fig:AARcomparison}
+\end{figure}
+
+Finally, the three settings are compared with OOVECM. We can see that RR and AAR approaches are better than traditional OLS and also AAR seems to perform better than RR.
 
@@
 
-\begin{figure}[!ht]
-  %\vspace{-0.8cm}
-  \centering
-  \includegraphics[width=0.5\textwidth]{RRcomparison}
-  \caption{Cumulative loss RR VECM against its optimal algorithm using different $\lambda$ values}
-  \label{fig:RRcomparison}
-\end{figure}
-
-\begin{figure}[!ht]
-  %\vspace{-0.8cm}
-  \centering
-  \includegraphics[width=0.5\textwidth]{AARcomparison}
-  \caption{Cumulative loss AAR VECM against its optimal algorithm using different $\lambda$ values}
-  \label{fig:AARcomparison}
-\end{figure}
-
 \section{Conclusions}
 \label{sec:conclusions}
-kk
+
+A new online version of the VECM has been proposed in order to use it with high frequency time series such as foreign exchange market rates.
+VECM takes advantage of the cointegration evidence between rates related to the USD currency.
+In order to get an online version we used the online versions of RR and AAR which are commonly used with problems with sequential access to the data.
+RR and AAR are used instead of the traditional OLS approach which has several known problems.
+A competitive analysis was done and experiments have shown that a regularised version OVECM could lead to improve cumulative losses compared with its optimal offline algorithm.
+Moreover, we showed that AAR could be a better option to improve performance independently of  how we choose $\lambda$ parameter.
 

### 2015-06-29 manuscript/manuscript.tex
referencias Werner

@@
 VECM is a special form of a VAR model for I(1) variables that are also
-cointegrated~\cite{banerjee1993}.
+cointegrated~\cite{johansen1991,banerjee1993}.
 
@@
 \noindent
-which is a VAR model but for time series differences.
+which is a VAR model but for time series differences ~\cite{hansen1999}.
 
@@
 Before running the tests, we firstly checked whether the time series were
-I(1) using the Augmented Dickey Fuller (ADF) test at 95\% significance level.
+I(1) using the Augmented Dickey Fuller (ADF) \cite{dickey1979,dickey1981} test at 95\% significance level.
 Table~\ref{tab:adf} shows that all currency rates cannot reject the unit root

### 2015-06-30 manuscript/manuscript.tex
adding statistics table

@@
 
-Finally, the three settings are compared with OOVECM. We can see that RR and AAR approaches are better than traditional OLS and also AAR seems to perform better than RR.
+The three settings are compared with OOVECM. We can see that RR and AAR approaches are better than traditional OLS and also AAR seems to perform better than RR.
 
@@
 \end{figure}
+
+Additionally table \ref{tab:stats} shows the MAPE, MAE and RMSE statistics. This results also shows that AAR behaves better than RR and OLS in comparison to OOVECM.
+
+\begin{table*}[ht!]
+\caption{Model measures}
+\label{tab:stats}
+\begin{center}
+\begin{tabular}{|l|l|c|c|c|c|}
+\hline
+\hline
+Method & $Solver$  & $\lambda$ & 
+MAPE & MAE& RMSE \\
+\hline
+ OVECM  &   OLS   &  0.0&  0.00189&  1.169e-05&3.955e-05\\
+ OVECM  &   RR   & 0.00 &  0.00142& 9.167e-06 &1.713e-05\\
+ OVECM  &   AAR     &  1e-06 &  0.00115&  7.605e-06&1.197e-05\\
+ OOVECM  &   OLS    &  1e-06 &  0.00103&  6.880e-06&1.148e-05\\
+\hline
+\end{tabular}
+\end{center}
+\end{table*}
 

### 2015-12-28 src/2_Finance.tex
Avance capitulo 2
ordenando propuesta 2
previo rev LSalinas

@@
 
+\vspace{0.5cm}
+
 High frequency trading (HFT) strategies are based on buy and sell assets in
-short periods of time gaining a small profit in every transaction, the key is
-the amount of transactions that algorithms are capable to execute obtaining profit considering transaction costs. 
-In 2012 more than 50\% of US market share of exchange trades were HFT.  HFT has
+short periods of time gaining a small profit in every transaction despite the transaction costs. The key is
+the amount of transactions that HFT algorithms are capable to execute. 
+In 2012 more than 50\% of US market trades were HFT.  HFT has
 motivated computer-driven strategies capable of processing large amount of data
@@
 
-\section{High Frequency Trading}
+\section{High Frequency Trading Definition}
 
-HFT is not a strategy but a technology which allow to implement particular
+HFT is not a strategy but a technology that implements different
 trading strategies. The aim of HFT is to benefit from market short-term pricing
@@
 for modelling and executing orders efficiently. In order to make fast decisions,
-HFT firms require speed access to trading center servers and sometimes they are
-physically near minimising network latencies.
+HFT firms require speed access to trading servers and sometimes they are
+physically near so they can minimise network latencies.
 
@@
 day avoiding leaving opened positions to the next business day. HFT is
-frequently associate to algorithm trading strategies, but the former is focused
+frequently associated to algorithm trading strategies, but the former is focused
 in to reduce the market impact of large institutional positions and the later refers to trade execution strategies that are typically used by fund managers to buy or sell large amounts of assets.
@@
 of a position opened. Most of them agreed that HFT refers to positions between
-1-second and 10-minutes. Overnight positions are avoided since they are more
-risky and also have more expensive fees, HFT firms end the trading day closing all
+1-second and 10-minutes. Overnight positions are avoided since they are
+riskier and also have more expensive fees, HFT firms end the trading day closing all
 opened positions.
@@
 In 2014 HFT represented nearly 50\% of equity trades in the US and more than 20\%
-in Europe showing a consistent fall since 2009 which was its best year. US market Revenues
+in Europe showing a consistent fall since 2009 which was its best year. US market revenues
 have fallen dramatically.
@@
 One of the reasons of this fall is that exchange markets have adapted, having
-now faster and more transparent and efficient market structures than before.
+now faster, more transparent and efficient market structures than before.
 This has been possible due to its investment in technology enhancing
 reliability and stability of transactions. Even though this fall, HFT is still a major component of regulated markets and
-will probably remain a topic of interest for researchers in the near future.
+will probably remain as a topic of interest for researchers in the near future.
 
@@
 Capital markets include the stock or equity market and the bond or debt market
-and it their movements are the most widely followed.
+and their movements are the most widely followed.
 
@@
 \end{figure}
+
 Equity market and foreign exchange market (Forex) are the most popular markets
-for high frequency trading strategies. In equity markets can be traded stocks
+for high frequency trading strategies. In the Equity market, stocks
 such as futures and options, exchange-traded funds (ETFs) among others financial
-instruments.
-
-Forex market allows the trading of interest rates denominated pair currencies
-such as EURUSD (Euro to UD dollar) or USDJPY (US dollar to Japanese yen). Traders of this market are diverse, some of them are
+instruments can be traded. Additionally, in the Forex market, interest rates denominated pair currencies 
+such as EURUSD (Euro to UD dollar) or USDJPY (US dollar to Japanese yen) are traded. Traders of this market are diverse, some of them are
 high frequency traders, long-term investors and corporations. The Forex market
@@
   \centering
-  \includegraphics[width=0.8\textwidth]{img/forex-trading-hours}
+  \includegraphics[width=0.8\textwidth]{img/forex-trading-hours.png}
   \caption{Forex market trading hours (GMT).}
@@
 \end{figure}
+
+There are three periods of overlaps between different trading times:
+
+\begin{itemize}
+\item New York and London from 13:00 GMT until 17:00 GMT
+\item Tokyo and London from 8:00 GMT until 9:00 GMT
+\item Sydney and Tokyo from 23:00 GMT to 7:00 GMT 
+\end{itemize}
+
+These overlap hours are very important since the highest volume of trades are expected. Therefore, most HFT strategies are executed in these hours.
 
@@
 \begin{description}
-\item[Market order:] allows to buy or sell an asset at the current best price
+\item[Market order] allows to buy or sell an asset at the current best price
 whatever the price is. The transaction price is determined by the market
 considering previous executed orders and volume requested.
-\item[Limit order:] allow to specify the price you want to buy or sell an asset.
-Depending on the market conditions, this type of orders can never be executed.
-\item[Stop order:] are suitable for investors who are unable to monitor their
+\item[Limit order] allow to specify the price you want to buy or sell an asset.
+Depending on the market conditions, this type of orders could never been executed.
+\item[Stop order] are suitable for investors who are unable to monitor their
 investments for a period of time. Stop order allows to specify the price that a
 position should be closed called stop price. When the stop price is reached a
-market order is executed, this means that market price won't be exactly the same
+market order is executed, this means that market price could not be exactly the same
 specified in the stop order.
@@
 \begin{description}
-\item[Fill or Kill:] is an order that must be executed immediately or being
+\item[Fill or Kill] is an order that must be executed immediately or being
 cancelled, no partial fulfillments are allowed. 
-\item[Day:] the order is only valid during the day.
-\item[Good til canceled:] a order is active until the investor decides to cancel
+\item[Day] the order is only valid during the day.
+\item[Good til canceled] a order is active until the investor decides to cancel
 it or the trade is executed.
@@
 
-The strong form of the EMH additionally claims that prices instantly reflect even hidden, private or "insider" information.
+The strong form of the EMH additionally claims that prices instantly reflect even hidden, private or ``insider" information.
 
@@
 If the stock market efficiently digests all available information, there is little justification for seeking excess returns gains from investing.
-However, EMH doesn't lessen the the importance of investing only change its philosophy.
+However, EMH doesn't lessen the importance of investing only change its philosophy.
 The only way an investor can possibly obtain higher returns is by purchasing riskier investments.

### 2016-01-07 src/3_TimeSeries.tex
saving my work

@@
 explanatory variables (relationships with other variables, seasonal factors,
-etc.), signal processing, forecasting (predict future values).  There are two
-main approaches to study financial time series: to study directions of financial
-rates and to explain financial rate volatility.  This chapter reviews the more
+etc.), signal processing and forecasting (predict future values), which is the focus of this thesis. 
+There are two main approaches to study financial time series: to study directions of financial
+rates and to explain financial rate volatility.  This chapter reviews the most
 relevant techniques in the rich and rapidly growing field of time series
@@
 
-\section{Characteristics of Time Series}
+\section{Characteristics of Financial Time Series}
+%\section{Stylized facts of asset returns}
+\label{sec:stylizedfacts}
+
+There are several known features exhibited by financial instruments called
+stylized facts, which have been empirically studied and some of them have been
+documented only recently and accepted as truth. Stylized facts are usually
+formulated in terms of qualitative properties of asset returns and may not be
+precise enough to distinguish among different parametric models.
+\cite{cont2001}.
+
+Returns of a time series $y_t$ is defined as:
+\begin{equation}
+\label{eq:returndef}
+r_t = y_t - y_{t-1}  \, ,
+\end{equation}
+
+Some stylized facts which are common to a wide set of financial assets
+\cite{sewell2011} are:
+
+\begin{description}
+\item[Dependence] autocorrelation function (ACF) in returns is largely insignificant. The ACF measures the linear predictability of the series at time $t$ using only the value at time $s$:
+
+\begin{equation}
+\rho(s,t) = \frac{\gamma(s,t)}{\sqrt{\gamma(s,s)\gamma(t,t)}}
+\end{equation}
+\noindent where $\gamma(s,t)$ is the autocovariance function measures the linear dependence between two points on the same series observed at different times, it is defined as the second moment product:
+\begin{equation}
+\gamma_y(s,t) = E[(y_s-\mu_s)(y_t - \mu_t)]
+\end{equation}
+However, the autocorrelation in the absolute and squared returns is always positive,
+significant and decays slowly. In addition, the autocorrelation in the absolute
+returns is generally higher than the autocorrelation in the corresponding
+squared returns. The returns autocorrelation function (ACF) of SPY is shown in figure
+ \ref{fig:returnacf}. The figure shows how the correlations are significant even
+ for very long lags, this imply that a long-memory process because of its
+ long-range dependence. 
+ \begin{figure}[h]
+ \centering
+ \includegraphics[width=0.7\textwidth]{plots/spy_returns_acf.pdf}
+ \caption{SPY returns ACF}
+ \label{fig:returnacf}
+\end{figure}
+\item[Distribution] The distribution of returns is approximately symmetric and
+has high kurtosis (i.e fat tails and a peaked centre compared with the normal
+distribution). However, distribution of returns whose were obtained from higher
+frequencies looks more like a normal distribution.
+The returns distribution tails are larger than
+ what is hypothesised by common data generation process (generally normal
+ distribution assumption). In the markets, fat tails are an undesirable feature
+ because of the additional risk they imply.  In the figure \ref{fig:returndist}
+ is shown the SPY returns distribution based on daily dates from the period 1st
+ July 1998 to 4th April 2013. The distribution was compared against the normal
+ distribution which clearly doesn't fit the data.  
+ \begin{figure}[h]
+ \centering
+ \includegraphics[scale=0.5]{plots/spy_returns_dist.pdf}
+ % spy_returns_dist.pdf: 504x504 pixel, 72dpi, 17.78x17.78 cm, bb=0 0 504 504
+ \caption{SPY returns distribution}
+ \label{fig:returndist}
+\end{figure}
+\item[Heterogeneity] despite the fact that financial returns are non-stationary,
+stationary periods can be observed. Economist speaks in terms of a structural
+break \cite{stock1994}, in machine learning this is known as drift which means
+that the statistical properties of the target variable change over time
+\cite{widmer1996}, \cite{tsymbal2004}.
+\item[Non-Linearity] financial returns may be non-linear in mean and/or
+non-linear in variance.
+\item[Calendar effects] are also called seasonal effects and they are cyclical
+anomalies in returns where the cycle is based on the calendar. Some of known
+calendar effects are: intraday effect, weekend effect, Monday effect, intramonth
+effect, the January effect and Holiday effect. The most important calendar
+anomalies are the January effect and the weekend effect. Figure
+\ref{fig:mondayeffect} shows the Monday effect from 1928 to 2007. It shows that
+in average Monday has negative returns. However, this is changing in time,
+figure \ref{fig:mondayeffect2} shows returns only for year 2007 and its last
+three months and it shows that Mondays and Wednesdays has positive returns in
+average.
+
+\begin{figure}[h!]
+\centering
+\begin{subfigure}[b]{0.45\textwidth}
+ \includegraphics[width=\textwidth]{img/average_daily_change_1928_2007}
+ \caption{Average daily return: 1928-2007}
+ \label{fig:mondayeffect}
+\end{subfigure}
+\begin{subfigure}[b]{0.45\textwidth}
+ \includegraphics[width=\textwidth]{img/thumb-average_daily_change_2007}
+ \caption{Average daily return: 2007}
+ \label{fig:mondayeffect2}
+\end{subfigure}
+\end{figure}
+\end{description}
 
@@
 
-\section{Stylized facts of asset returns}
-\label{sec:stylizedfacts}
-
-There are several known features exhibited by financial instruments called
-stylized facts, which have been empirically studied and some of them have been
-documented only recently and accepted as truth. Stylized facts are usually
-formulated in terms of qualitative properties of asset returns and may not be
-precise enough to distinguish among different parametric models.
-\cite{cont2001}.
-
-Returns of a time series $y$ is defined as:
-\begin{equation}
-\label{eq:returndef}
-r_t = y_t - y_{t-1}  \, ,
-\end{equation}
-
-Some stylized facts which are common to a wide set of financial assets
-\cite{sewell2011} are:
-
-\begin{description}
-\item[Dependence] autocorrelation in returns is largely insignificant. However,
-autocorrelation in the absolute and squared returns is always positive and
-significant and decays slowly. In addition, the autocorrelation in the absolute
-returns is generally higher than the autocorrelation in the corresponding
-squared returns. The returns autocorrelation function (ACF) of SPY is shown in figure
- \ref{fig:returnacf}. The figure shows how the correlations are significant even
- for very long lags, this imply that a long-memory process because of its
- long-range dependence. 
- \begin{figure}[h]
- \centering
- \includegraphics[width=0.7\textwidth]{plots/spy_returns_acf.pdf}
- \caption{SPY returns ACF}
- \label{fig:returnacf}
-\end{figure}
-\item[Distribution] The distribution of returns is approximately symmetric and
-has high kurtosis (i.e fat tails and a peaked centre compared with the normal
-distribution). However, distribution of returns whose were obtained from higher
-frequencies looks more like a normal distribution.
-The returns distribution tails are larger than
- what is hypothesised by common data generation process (generally normal
- distribution assumption). In the markets, fat tails are an undesirable feature
- because of the additional risk they imply.  In the figure \ref{fig:returndist}
- is shown the SPY returns distribution based on daily dates from the period 1st
- July 1998 to 4th April 2013. The distribution was compared against the normal
- distribution which clearly doesn't fit the data.  
- \begin{figure}[h]
- \centering
- \includegraphics[scale=0.5]{plots/spy_returns_dist.pdf}
- % spy_returns_dist.pdf: 504x504 pixel, 72dpi, 17.78x17.78 cm, bb=0 0 504 504
- \caption{SPY returns distribution}
- \label{fig:returndist}
-\end{figure}
-\item[Heterogeneity] despite the fact that financial returns are non-stationary,
-stationary periods can be observed. Economist speaks in terms of a structural
-break \cite{stock1994}, in machine learning this is known as drift which means
-that the statistical properties of the target variable change over time
-\cite{widmer1996}, \cite{tsymbal2004}.
-\item[Non-Linearity] financial returns may be non-linear in mean and/or
-non-linear in variance.
-\item[Calendar effects] are also called seasonal effects and they are cyclical
-anomalies in returns where the cycle is based on the calendar. Some of known
-calendar effects are: intraday effect, weekend effect, Monday effect, intramonth
-effect, the January effect and Holiday effect. The most important calendar
-anomalies are the January effect and the weekend effect. Figure
-\ref{fig:mondayeffect} shows the Monday effect from 1928 to 2007. It shows that
-in average Monday has negative returns. However, this is changing in time,
-figure \ref{fig:mondayeffect2} shows returns only for year 2007 and its last
-three months and it shows that Mondays and Wednesdays has positive returns in
-average.
-
-\begin{figure}[h!]
-\centering
-\begin{subfigure}[b]{0.45\textwidth}
- \includegraphics[width=\textwidth]{img/average_daily_change_1928_2007}
- \caption{Average daily return: 1928-2007}
- \label{fig:mondayeffect}
-\end{subfigure}
-\begin{subfigure}[b]{0.45\textwidth}
- \includegraphics[width=\textwidth]{img/thumb-average_daily_change_2007}
- \caption{Average daily return: 2007}
- \label{fig:mondayeffect2}
-\end{subfigure}
-\end{figure}
-\end{description}
-
 \section{Volatility in the financial markets}
@@
 
-\section{Vector Autoregressive Models}\label{sec:varvec}
+%\section{Vector Autoregressive Models}\label{sec:varvec}
+\section{Multivariate Time Series modelling}
 

### 2016-01-26 src/4_OnlineLearning.tex
error in formulas OLS

@@
 \label{eq:proyectorsol}
-\mathbf{X \hat{\mathbf{X}} = PY} 
+\mathbf{A \hat{\mathbf{X}} = PY} 
 \end{equation*}
@@
 \begin{eqnarray*}
-\mathbf{X \hat{\mathbf{X}}} &=& \mathbf{PB} \\
-\mathbf{U_1 \Sigma_1 V_1^\top \hat{\mathbf{X}}} &=& \mathbf{U_1 U_1^\top B} \\
-\mathbf{ V_1^\top \hat{\mathbf{X}}} &=&  \mathbf{\Sigma_1^{-1} U_1^\top B} \\ 
+\mathbf{A \hat{\mathbf{X}}} &=& \mathbf{PY} \\
+\mathbf{U_1 \Sigma_1 V_1^\top \hat{\mathbf{X}}} &=& \mathbf{U_1 U_1^\top Y} \\
+\mathbf{ V_1^\top \hat{\mathbf{X}}} &=&  \mathbf{\Sigma_1^{-1} U_1^\top Y} \\ 
 \mathbf{ V_1^\top V_1 \mathbf{X}_1} &=& \mathbf{\Sigma_1^{-1}
-U_1^\top B} \\
-\mathbf{\mathbf{X}_1}&=& \mathbf{\Sigma_1^{-1} U_1^\top B}
+U_1^\top Y} \\
+\mathbf{\mathbf{X}_1}&=& \mathbf{\Sigma_1^{-1} U_1^\top Y}
 \end{eqnarray*}
@@
                 &=& \mathbf{V_1 \Sigma_1^{-1} U_1^\top Y} \\
-\mathbf{X^+} &=& \mathbf{V_1 \Sigma_1^{-1} U_1^\top} \, .
+\mathbf{A^+} &=& \mathbf{V_1 \Sigma_1^{-1} U_1^\top} \, .
 \end{eqnarray*}
@@
 \subsection{Efficient computation}
-
-In order to solve OLS in a efficient way, Coleman and
+In regression problems we always require getting a matrix inverse which is
+computationally expensive. However, in stream data problems the is a way to
+obtain a matrix inverse approximation using the Sherman-Morrison-Woodbury. This
+formula allows to get the inverse of a
+the matrix $\mathbf{A+uv^\top}$ if we already calculated the
+inverse of $\mathbf{A}$ as follows:
+
+\begin{equation}
+\label{eq:SMW}
+(\mathbf{A+uv^\top})^{-1}=\mathbf{A}^{-1}-
+\frac{\mathbf{A^{-1}uv^\top A^{-1}}}{\mathbf{1+v^\top A^{-1}u}}
+\end{equation}
+
+Alternatively, Coleman and
 Sun~\cite{coleman+sun2010} presented an iterative algorithm which

### 2016-03-22 articles-conferences/icpram2015/main.tex
agregando repo icpram

\documentclass[a4paper,twoside]{article}
\usepackage{graphicx} 
\usepackage{epstopdf}
\usepackage{epsfig}
\usepackage{subfigure}
\usepackage{calc}
\usepackage{amssymb}
\usepackage{amstext}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{lscape}
\usepackage{multicol}
\usepackage{pslatex}
\usepackage{apalike}
\usepackage{SCITEPRESS}
\usepackage[small]{caption}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{enumerate}
\renewcommand{\algorithmicrequire}{\textbf{Input:}}
\renewcommand{\algorithmicensure}{\textbf{Output:}}
\def\R{{\mathbb R }}
\def\N{{\mathbb N }}
\subfigtopskip=0pt
\subfigcapskip=0pt
\subfigbottomskip=0pt
\begin{document}

\title{An Online Vector Error Correction Model for Exchange Rates Forecasting}

\author{\authorname{Paola Arce\sup{1}, Jonathan Antognini\sup{1}, Werner
Kristjanpoller\sup{2} and Luis Salinas\sup{1,3}}
\affiliation{\sup{1}Departamento de Inform\'atica, Universidad T\'ecnica Federico
Santa Mar\'ia, Valpara\'iso, Chile}
\affiliation{\sup{2}Departamento de Industrias, Universidad T\'ecnica Federico
Santa Mar\'ia, Valpara\'iso, Chile}
\affiliation{\sup{3}CCTVal, Universidad T\'ecnica Federico
Santa Mar\'ia, Valpara\'iso, Chile}
\email{\{paola.arce,luis.salinas,werner.kristjanpoller\}@usm.cl, jantogni@csrg.inf.utfsm.cl}
}

\keywords{Vector Error Correction Model, Online Learning, Financial
Forecasting, Foreign Exchange Market}

\abstract{
Financial time series are known for their non-stationary behaviour. However,
sometimes they exhibit some stationary linear combinations. When this happens,
it is said that those time series are cointegrated.The Vector Error Correction
Model (VECM) is an econometric model which characterizes the joint dynamic
behaviour of a set of cointegrated variables in terms of forces pulling towards
equilibrium. In this study, we propose an Online VEC model (OVECM) which
optimizes how model parameters are obtained using a sliding window of the most
recent data. Our proposal also takes advantage of the long-run relationship
between the time series in order to obtain improved execution times. Our
proposed method is tested using four foreign exchange rates with a frequency of
1-minute, all related to the USD currency base. OVECM is compared with VECM and
ARIMA models in terms of forecasting accuracy and execution times. We show that
OVECM outperforms ARIMA forecasting and enables execution time to
be reduced considerably while maintaining good accuracy levels compared with
VECM.
}

\onecolumn \maketitle \normalsize \vfill

\input{src/intro}
\input{src/background}
\input{src/methodology}
\input{src/results}
\input{src/conclusions}

\vfill
\bibliographystyle{apalike}
{\small
\bibliography{reference}}

\vfill
\end{document}

### 2016-03-22 articles-conferences/icpram2015/slides/slides.tex
agregando repo icpram

\documentclass[xcolor=dvipsnames]{beamer}
\usepackage[spanish,activeacute]{babel}
\usepackage[utf8]{inputenc}
\usepackage{listings}
\usepackage{amsmath,amsfonts,amssymb}
\usepackage{adjustbox}
\usepackage{color}
\usepackage{tikz}
\usetikzlibrary{shapes,arrows}
\usepackage{multicol}
%\usepackage{slashbox,pict2e}
\usepackage{pgf}
\usepackage{algorithm} %REVISAR FORMATO DE ALGORITMO
\usepackage{algorithmic}
\renewcommand{\algorithmicrequire}{\textbf{Input:}}
\renewcommand{\algorithmicensure}{\textbf{Output:}}
\def\R{{\mathbb R }}
\def\N{{\mathbb N }}
\usetheme{Warsaw}
%\usecolortheme{lily}
%\usecolortheme{beaver}

\definecolor{myblue1}{RGB}{35,119,189}
\definecolor{myblue2}{RGB}{95,179,238}
\definecolor{myblue3}{RGB}{129,168,207}
\definecolor{myblue4}{RGB}{26,89,142}

\setbeamercolor*{structure}{fg=myblue1,bg=blue}
\setbeamercolor*{palette primary}{use=structure,fg=white,bg=structure.fg}
\setbeamercolor*{palette secondary}{use=structure,fg=white,bg=structure.fg!75!black}
\setbeamercolor*{palette tertiary}{use=structure,fg=white,bg=structure.fg!50!black}
\setbeamercolor*{palette quaternary}{fg=black,bg=white}
\setbeamercolor*{item projected}{fg=red,bg=myblue3!80}
\setbeamercolor*{block title example}{fg=white,bg=myblue4}
\setbeamercolor*{frametitle}{fg=black}

\setbeamertemplate{blocks}[rounded][shadow=true]

\makeatletter
\pgfdeclarehorizontalshading[frametitle.bg,frametitle right.bg]{beamer@frametitleshade}{\paperheight}{%
  color(0pt)=(myblue2);
  color(\paperwidth)=(white)}

\defbeamertemplate*{footline}{mysplit theme}
{%
  \leavevmode%
  \hbox{\begin{beamercolorbox}[wd=.5\paperwidth,ht=2.5ex,dp=1.125ex,leftskip=.3cm plus1fill,rightskip=.3cm]{author in head/foot}%
    \usebeamerfont{author in head/foot}\insertshortauthor
  \end{beamercolorbox}%
  \begin{beamercolorbox}[wd=.5\paperwidth,ht=2.5ex,dp=1.125ex,leftskip=.3cm,rightskip=.3cm plus1fil]{title in head/foot}%
    \usebeamerfont{title in head/foot}\insertshorttitle\hfill
    \insertframenumber/\inserttotalframenumber\hspace*{0.5em}
  \end{beamercolorbox}}%
  \vskip0pt%
}
\makeatother

\tikzstyle{startstop} = [rectangle, rounded corners, minimum width=3cm, minimum height=1cm,text centered, draw=black, fill=red!30]
\tikzstyle{io} = [trapezium, trapezium left angle=70, trapezium right angle=110, minimum width=3cm, minimum height=1cm, text centered, draw=black, fill=blue!30]
\tikzstyle{process} = [rectangle, minimum width=4cm, minimum height=1cm, text centered, text width=5cm, draw=black, fill=orange!30]
\tikzstyle{decision} = [diamond, minimum width=2cm, minimum height=1cm, text centered, draw=black, fill=green!30]
\tikzstyle{arrow} = [thick,->,>=stealth]

\author[P. Arce et al.]{Paola
Arce\inst{1}\and Jonathan Antognini\inst{1}\and Werner Kristjanpoller
\inst{2}\and Luis Salinas\inst{1,3}}

\institute[UTFSM]{
\inst{1} Departamento de Inform\'atica, UTFSM, Chile \\
\inst{2} Departamento de Industrias, UTFSM, Chile \\
\inst{3} CCTVal, UTFSM, Chile
}

\date[ICPRAM 2015]{4th International Conference on Pattern Recognition
Application and Methods 2015}
\title[Online VECM for FOREX rates forecasting]{\Large An Online Vector Error Correction Model for
Exchange Rates Forecasting}

\titlegraphic{
   \includegraphics[width=1cm]{img/logo_DI}\hspace*{2.5cm}~%
   \includegraphics[width=1.5cm]{img/logo_UTFSM}\hspace*{2.5cm}~%
   \includegraphics[width=1cm]{img/logo_IND}
}

\begin{document}
\begin{frame}[plain]
\titlepage
\end{frame}

\begin{frame}
\frametitle{Outline}
\begin{itemize}
\item Motivation: economic relations
\item Cointegration and Integration concepts
\item VECM model
\item Proposal
\item Experiments
\item Conclusions
\end{itemize}
\end{frame}

\section{Introduction}
\subsection{Motivation}
\begin{frame}
\frametitle{Motivation}
Financial time series are mainly non-stationary.  However, one or more linear
combinations of non-stationary time series can be stationary even though
individually they are not.

\begin{center}
\includegraphics[width=0.5\textwidth]{img/allcurrencies}
\end{center}
\end{frame}

\begin{frame}
When this happens it is said they have a long-run
equilibrium relationship and the variables are \textbf{cointegrated}.
\end{frame}

\section{Background}
\subsection{Definitions}
\begin{frame}
\frametitle{Integration }
\begin{block}{Integration of order $d$}
A time series $\mathbf{y}$ is said to be integrated of order $d$ if after
differentiating the variable $d$ times, we get an stationary process, more precisely:

\[
(1-L)^d \mathbf{y} \sim \text{I(0)} \, ,
\]

\noindent where I(0) is a stationary time series and $L$ is the lag operator:

\[
(1-L)\mathbf{y} = \Delta \mathbf{y}=\mathbf{y}_t  -\mathbf{y}_{t-1} \quad \forall t
\]
\end{block}
\end{frame}

\subsection{Integration example}
\begin{frame}
\frametitle{Financial Time Series}
\begin{columns}
\column[t]{5cm}
Financial time series are commonly a I(1) process 
\begin{center}
\includegraphics[width=0.85\textwidth]{img/EURUSD}
\end{center}
\column[t]{5cm}
Since their differences are stationary (I(0) process).
\begin{center}
\includegraphics[width=0.85\textwidth]{img/DEURUSD}
\end{center}
\end{columns}
\end{frame}

\subsection{Cointegration}
\begin{frame}
\frametitle{Cointegration}
\begin{block}{Cointegration definition}
Let $\mathbf{y} = \{\mathbf{y}^1, \dots, \mathbf{y}^l\}$ be a set of $l$
time series of order I(1) which are said to be cointegrated if a vector
$\beta=[\beta(1),\dots,\beta(l)]^\top \in \mathbb{R}^l$  exists such that the
time series,

\begin{equation*}
 \mathbf{Z}_t:= \beta^\top \mathbf{y} = \beta(1) \mathbf{y}^1 + \dots + \beta(l) \mathbf{y}^l \sim
 \text{I(0)}\, .
\end{equation*}
\end{block}
In other words, a set of I(1) variables is said to be cointegrated if
a linear combination of them exists which is I(0).

\end{frame}

\subsection{Vector Autorregresive models}
\begin{frame}
\frametitle{VECM: Vector Error Correction model}

VECM is a special form of a Vector Autorregresive model (VAR) for I(1)
variables $\mathbf{y}$ that are also cointegrated. Let

\begin{equation*}
\label{eq:variables}
\mathbf{y}_t = 
\begin{bmatrix} y_{1,t} & y_{2,t} & \dots & y_{l,t}
\end{bmatrix}^\top \, \qquad t=p+1, \dots N. 
\end{equation*}

\begin{block}{VAR}
VAR($p$) model describes the behaviour of a set of $l$
endogenous variables as a linear combination of their last $p$ values:
{\color{blue}
\begin{equation*}
\label{eq:var}
 \mathbf{y}_t = \phi_1 \mathbf{y}_{t-1}  + \dots +   \phi_p\mathbf{y}_{t-p}
+ \mathbf{c} + \mathbf{\epsilon}_t \, ,
\end{equation*}}
\noindent where  ${\phi_1,\dots,\phi_p}$ are $l \times l$
matrices of real coefficients,
$\mathbf{\epsilon}_{p+1},\dots,\mathbf{\epsilon}_N$ are error terms,
$\mathbf{c}$ is a constant vector and $N$ is the total number of samples.
\end{block}
\end{frame}

\begin{frame}
\frametitle{VECM: Vector Error Correction model}
\begin{block}{VECM}

It is obtained by replacing the form {\color{red} $\Delta
\mathbf{y}_t = \mathbf{y}_t - \mathbf{y}_{t-1}$} in the VAR form.
{\color{blue}
\begin{equation}
 \label{eq:vec}
 \Delta \mathbf{y}_t = 
 \underbrace{ \Omega\mathbf{y}_{t-1}}_\text{Error correction term} + 
 \sum_{i=1}^{p-1}
\phi_i^* \Delta \mathbf{y}_{t-i}  + \mathbf{c} + \mathbf{\epsilon}_t \quad ,
\end{equation}}
\end{block}

\noindent where 
%coefficients matrices $\Omega$ and $\phi_i^*$ are
%function of matrices $\phi_i$ as follows:

\begin{eqnarray*}
\phi_i^* &: =& -\sum_{j=i+1}^{p} \phi_j \\
\Omega &: =& -(\mathbb{I}-\phi_1-\dots-\phi_p) 
\end{eqnarray*}
\end{frame}

\begin{frame}
\frametitle{$\Omega$ matrix properties}
The matrix $\Omega$ has the following properties:
\begin{itemize}
\item If $\Omega = 0$ there is no cointegration
\item If $rank(\Omega)=l$ i.e full rank, then the time series are not
I(1) but stationary
\item If $rank(\Omega)=r,\quad 0 < r < l$ then, there is cointegration
and the matrix $\Omega$ can be expressed as $\Omega =
\alpha \beta^\top$, where $\alpha$ and $\beta$ are $(l \times r)$
matrices and $rank(\alpha)=rank(\beta)=r$.
\end{itemize}
The columns of $\beta$ contains the cointegration vectors and the rows of
$\alpha$ correspond with the adjusted vectors. $\beta$ is obtained by Johansen
procedure whereas $\alpha$ has to be determined as a
variable in the VECM.
\end{frame}

\begin{frame}
\frametitle{VECM matrix form}
\small
\begin{equation*}
\underbrace{
      \begin{bmatrix}
       \mathbf{\Delta y}_{p+1}  \\ 
       \mathbf{\Delta y}_{p+2}  \\ 
       \vdots                   \\ 
       \mathbf{\Delta y}_N      
      \end{bmatrix}}_{\mathbf{B} } =
\arraycolsep=1.4pt  
\underbrace{\begin{bmatrix}
   \alpha^\top & \phi^*_1 & \dots & \phi^*_{p-1} &\mathbf{c}   
  \end{bmatrix}}_{\mathbf{X}^\top}
\underbrace{\begin{bmatrix}
 \beta\mathbf{y}_p      & \beta\mathbf{y}_{p+1}   & \dots&   \beta\mathbf{y}_{N-1}   \\
 \mathbf{\Delta y}_{p}   & \mathbf{\Delta y}_{p+1}& \dots &  \mathbf{\Delta y}_{N-1} \\
 \mathbf{\Delta y}_{p-1} & \mathbf{\Delta y}_{p}  & \dots &  \mathbf{\Delta y}_{N-2}   \\
 \vdots                  & \vdots                 & \ddots&  \vdots                   \\
 \mathbf{\Delta y}_{2}   & \mathbf{\Delta y}_{3} & \dots &   \mathbf{\Delta y}_{N-p+1} \\
 1                      & 1                       & \dots     & 1   
 \end{bmatrix}}_{\mathbf{A}^\top}
+
\underbrace{\begin{bmatrix}
              \mathbf{\epsilon}_{p+1} \\ 
              \mathbf{\epsilon}_{p+2} \\ 
              \vdots \\ 
              \mathbf{\epsilon}_N
             \end{bmatrix}}_{\mathbf{E}^\top}
\end{equation*}
\end{frame}

\begin{frame}
\frametitle{VECM difficulties}
\begin{itemize}
\item VECM parameter estimation is done by the Johansen method and OLS which are
computationally expensive routines.
\item VECM parameters must be recalculated every time new data is available. 
\end{itemize}
\end{frame}

\section{Experiments}
\subsection{Proposal}
\begin{frame}
\frametitle{Proposal: Online VECM algorithm (OVECM)}
\begin{itemize}
\item Considers only a sliding window of the $L$ most recent data. 
\item Since cointegration vectors represent long-run relationships they vary
little in time. OVECM obtains these vectors only when they need to be updated. 
\item This updating is done when MAPEs of the last $n$ inputs is above a certain
error given by a threshold 
\item OVECM also implements matrix optimisations in order to reduce execution time,
such as updating matrices with new data, removing old data and introducing new
cointegration vectors.
\end{itemize}
\end{frame}

%\begin{frame}
%\begin{center}
%    \begin{adjustbox}{max size={.95\textwidth}{.8\textheight}} %height=0.8\textheight}
%    \begin{tikzpicture}[node distance=2cm]
%    %\node (start) [startstop] {Start};
%    %\node (in1) [io, below of=start] {Input};
%    \node (in1) [io] {Input: y, p, L, $\epsilon$, n};
%    \node (pro1) [process, below of=in1] {$y_i = y[i:i+L]$};
%    \node (dec1) [decision, below of=pro1, yshift=-0.5cm] {i = 0};
%    \node (pro2a) [process, below of=dec1, yshift=-0.5cm] {$[\mathbf{A} \quad \mathbf{B}] =\texttt{ovecm}(\mathbf{y}_i,p,v,\mathbf{A},\mathbf{B})$
%                                                        $\Delta \mathbf{Y}_{\text{true}}[i] = \mathbf{B}[-1,:]$
%                                                        $\Delta \mathbf{Y}_{\text{pred}}[i] = \mathbf{A}[-1,:] \times \mathbf{X}$};
%
%    \node (pro2b) [process, right of=dec1, xshift=4cm] {$v = \texttt{Johansen}(\mathbf{y}_i,p)$ $[\mathbf{A} \quad 
%                                                                \mathbf{B}] = \texttt{vecm}(\mathbf{y}_i,p,v)$};
%    \node (pro3) [process, below of=pro2a, yshift=-0.5cm] {$\mathbf{X} = \text{OLS} (\mathbf{A},\mathbf{B})$ \\
%                                                        $\mathbf{y}_{\text{true}}[i] = \mathbf{y}_i[-n,:]$ \\
%                                                        $\mathbf{y}_{\text{pred}}[i] = \mathbf{A}[-n,:] \times \mathbf{X} +$ \\$ \mathbf{y}_i[-n-1,:-1]$ \\
%                                                        $e = \texttt{mape}(\mathbf{y}_{\text{true}}, \mathbf{y}_{\text{pred}})$};
%
%    \node (dec2) [decision, below of=pro3, yshift=-0.5cm] { $e > \epsilon $ };
%
%    \node (pro4a) [process, right of=dec2, xshift=4cm] {$v = \texttt{Johansen}(\mathbf{y}_i,p)$ \\
%                                                        $\mathbf{A} = \texttt{uvecm}(\mathbf{y}_i,p,v,\mathbf{A})$ \\
%                                                        $\mathbf{X} = \texttt{OLS} (\mathbf{A},\mathbf{B})$};
%
%   
%    \node (pro5) [process, left of=pro2a, xshift=-4cm] {$ i++ $}; 
%   
%    \node (dec3) [decision, below of=dec2, yshift=-0.5cm] { $i > it $ };
%    
%    \node (out1) [io, below of=dec3] {Output: $\Delta \mathbf{y}_{true}, \Delta \mathbf{y}_{pred}$};
%    
%    %\node (stop) [startstop, below of=out1] {Stop};
%    
%    %\draw [arrow] (start) -- (in1);
%    \draw [arrow] (in1) -- (pro1);
%    \draw [arrow] (pro1) -- (dec1);
%    \draw [arrow] (dec1) -- node[anchor=east] {no} (pro2a);
%    \draw [arrow] (dec1) -- node[anchor=south] {si} (pro2b);
%    \draw [arrow] (pro2a) -- (pro3);
%    \draw [arrow] (pro2b) |- (pro3);
%    \draw [arrow] (pro3) -- (dec2);
%    \draw [arrow] (dec2) --  node[anchor=south] {si} (pro4a);
%    \draw [arrow] (pro4a) |- (dec3);
%    
%    \draw [arrow] (dec2) --  node[anchor=east] {no} (dec3);
%    \draw [arrow] (dec3.west) -| node[anchor=east] {no} (pro5);
%    \draw [arrow] (pro5) |- (pro1);
%    \draw [arrow] (dec3) --  node[anchor=east] {si} (out1);
%    %\draw [arrow] (out1) -- (stop);
%    \end{tikzpicture}
%    \end{adjustbox}
%\end{center}
%\end{frame}

\subsection{Data}
\begin{frame}
\frametitle{Data}
\begin{itemize}
\item We used data of four foreign exchange rates: EURUSD, GBPUSD, USDCHF and USDJPY. 
\item This data was collected from the free database Dukascopy which gives access to
the Swiss Foreign Exchange Market.
\item The reciprocal of the last two rates (CHFUSD, JPYUSD) were used in order to
obtain the same base currency for all rates. 
\item The tests were done using
1-minute frequency from ask prices which corresponded to 1.440 data points per
day from the 11th to the 15th of August 2014.
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Unit Root tests}
\begin{table}[h!]
\caption{Unit roots tests}
\label{tab:adf}
\begin{center}
\begin{tabular}{|l|c|c|c|c|c|}
\hline
& \textbf{Statistic} & \textbf{Critical value} & \textbf{Result}\\
\hline
EURUSD          & -0.64 & -1.94 & True       \\
$\Delta$EURUSD & -70.45   & -1.94 & False       \\
GBPUSD          & -0.63   & -1.94 & True          \\
$\Delta$GBPUSD & -54.53   & -1.94 & False       \\
CHFUSD          & -0.88   & -1.94 & True         \\
$\Delta$CHFUSD & -98.98   & -1.94 & False       \\
JPYUSD          & -0.65 & -1.94 & True        \\
$\Delta$JPYUSD & -85.78 & -1.94 & False     \\ 
\hline
\end{tabular}
\end{center}
\end{table}
\end{frame}

\subsection{Methodology}
\begin{frame}
\frametitle{Methodology}
OVECM were compared against VECM and ARIMA. VECM and ARIMA are called with a
sliding window of the most recent data, whereby the models are updated at every
time step. Therefore, we named them SLVECM and SLARIMA.
\end{frame}

\subsection{Parameters}
\begin{frame}
\frametitle{Parameters optimization}
VECM order and ARIMA parameters were selected using AIC.

\begin{table}[ht]
\label{tab:params}
\begin{center}
\begin{tabular}{|c|c|c|}
\hline
Windows size $L$ & VECM & ARIMA\\
 $L$ & order $(p)$ & order $(p,d,q)$ \\
\hline
100 & 2 & p=2,d=1,q=1\\
400 & 5 & p=1,d=1,q=1\\
700 & 3 &p=2,d=1,q=1\\
1000 &3 & p=2,d=1,q=1\\
\hline
\end{tabular}
\end{center}
\end{table}
\end{frame}

\begin{frame}
\frametitle{Execution times}
We ran 400 iterations of OVECM and SLVECM . SLARIMA was based on the ARIMA procedure
of the statsmodels library and therefore its execution time is excluded.
%because its is not comparable with OVECM and SLVECM. SLARIMA was created based
%on statsmodels library routine ARIMA.
\begin{center}
\adjustbox{max height=\dimexpr\textheight-5.5cm\relax,
           max width=\textwidth}{
\begin{tabular}{|l|c|c|c|c|c|}
\hline
& $L$ & order & e  & Time[s] \\
\hline
OVECM & 100 &p=2  & 0      & 2.492\\
OVECM & 100 &p=2  & 0.0026  & 1.606\\
SLVECM & 100 &p=2& -- & 2.100\\
\hline
OVECM & 400 & p=5  & 0      & 3.513\\
OVECM & 400 &p=5  & 0.0041  & 2.569\\
SLVECM & 400 & p=5 & -- & 3.222\\
\hline
OVECM & 700 &p=3  & 0      & 3.296\\
OVECM & 700 &p=3  & 0.0032  & 2.856\\
SLVECM & 700 &p=3 & -- & 3.581\\
\hline
OVECM & 1000 & p=3 & 0      & 4.387\\
OVECM & 1000 & p=3  & 0.0022  & 2.408\\
SLVECM & 1000 & p=3  & -- & 3.609\\
\hline
\end{tabular}
}\end{center}
\end{frame}

\begin{frame}
\frametitle{Performance accuracy}
\tiny
\begin{table}[H]
\caption{Model measures}
\label{tab:stats}
\begin{center}
\begin{tabular}{|l|l|c|c|c|c|c|c|c|}
\hline
\multicolumn{3}{|c|}{Model} & \multicolumn{3}{|c|}{In-sample} &
\multicolumn{3}{|c|}{Out-of-sample} \\ 
\hline
\hline
Method & $L$ & Parameters &  
MAPE & MAE& RMSE& 
MAPE & MAE& RMSE \\
\hline
 OVECM  &   100  &  P=2 e=0.0026 &  0.00263&  0.00085&  0.00114&  0.00309&  0.00094&  0.00131\\
 OVECM  &   400  &  P=5 e=0.0041&  0.00378&  0.00095&  0.00127&  0.00419&  0.00103&  0.00143\\
 OVECM  &   700  &  P=3 e=0.0032&  0.00323&  0.00099&  0.00130&  0.00322&  0.00097&  0.00132\\
 OVECM  &   1000 &  P=3 e=0.0022&  
 \textbf{0.00175}&  \textbf{0.00062}&  \textbf{0.00087} &  
 \textbf{0.00172}&  \textbf{0.00061}&  \textbf{0.00090}\\
\hline
 SLVECM  &   100 &  P=2 &  0.00262&  0.00085&  0.00113&  0.00310&  0.00095&  0.00132\\
 SLVECM  &   400 &  P=5 &  0.00375&  0.00095&  0.00126&  0.00419&  0.00103&  0.00143\\
 SLVECM  &   700 &  P=3 &  0.00324&  0.00099&  0.00130&  0.00322&  0.00098&  0.00132\\
 SLVECM  &   1000&  P=3 &  
 \textbf{0.00174}&  \textbf{0.00061}&  \textbf{0.00087}&  
 \textbf{0.00172}&  \textbf{0.00061}&  \textbf{0.00090}\\
\hline
SLARIMA & 100  &p=2 d=1 q=1 & 0.00285  &  0.00110  &  0.00308  &  0.00312  &0.00098  &  0.00144 \\
SLARIMA & 400  &p=1 d=1 q=1 & 0.00377  &  0.00101  &  0.00128  &  0.00418 & 0.00106 & 0.00145 \\
SLARIMA & 700  &p=2 d=1 q=1 & 0.00329  &  0.00102  &  0.00136  &  0.00324  & 0.00097  &  0.00133 \\
SLARIMA & 1000 &p=2 d=1 q=1 & \textbf{0.00281}  & \textbf{0.00074}  &  \textbf{0.00105}  &  \textbf{0.00177} & \textbf{0.00063}  &  \textbf{0.00092} \\
\hline
\end{tabular}
\end{center}
\end{table}
\end{frame}

%\begin{frame}
%\frametitle{Performance accuracy example}
%\begin{figure}[h]
%  \vspace{-0.2cm}
%  \centering
%   \includegraphics[width = 6cm]{img/accuracy} 
%  \caption{OVECM forecasting accuracy example for 50 minutes using $L = 1000$ and $p =3$}
%  \label{fig:accuracy}
% \end{figure}
%\end{frame}

\section{Conclusions}
\begin{frame}
\frametitle{Conclusions}
\begin{itemize}
\item  OVECM considerably reduces execution times
without compromising solution accuracy.  
\item OVECM was compared with VECM and ARIMA
with the same sliding window sizes and OVECM outperformed both in terms of
execution time. 
\item Traditional VECM slightly outperformed our proposal but the
OVECM execution time is lower.
\item We could see that our algorithm took much less than a minute at every
step. This means that it could also be used with higher frequency data and would
still provide responses before new data arrives.  
\end{itemize}
\end{frame}

\begin{frame}[plain,c]
%\frametitle{A first slide}

\begin{center}
\Huge Thank you for your attention
\Huge Questions?
\end{center}

\end{frame}
\end{document}

### 2016-03-22 articles-conferences/icpram2015/slides/src/slides.tex
agregando repo icpram

\section{HFT time series}
\subsection{Intro}

\begin{frame}
    \frametitle{Motivation: Financial Time Series Forecasting}
    \begin{itemize}
        \item High Frequency data
        \item Large number of factors affecting the market
    \end{itemize}
    \begin{center}
    \includegraphics[width=0.8\textwidth]{img/financial_ts}
    \end{center}
\end{frame}

\begin{frame}
    \frametitle{Learning Problems}
    \begin{itemize}
        \item Batch Learning
        \begin{itemize}
            \item Training phase then testing
            \item The model doesn't change in testing phase
        \end{itemize}
        \item Online Learning
        \begin{itemize}
            \item No separate phase
            \item "Learning while doing"
        \end{itemize}
    \end{itemize}
\end{frame}

\begin{frame}
    \frametitle{Batch methods}
    \framesubtitle{Ridge Regression}

    \begin{Definition}
    {\em Ridge regression} (RR) is a batch method which is a
        generalization of the least squares method (LS). 
    \end{Definition}

    \begin{equation*} \min_{\mathbf{w}} J(\mathbf{w}), \quad
    \text{where} \quad J(\mathbf{w}) = \sum_{t=1}^m
    (f(\mathbf{x}_t)-y_t)^2  + \gamma R(\mathbf{w}) \, ,
    \end{equation*}

    where $\gamma$ is a regularization parameter and
    $R(\mathbf{w}) = \| \mathbf{w}\| ^2$.
\end{frame}

\begin{frame}
    \frametitle{Batch methods}
    \framesubtitle{Ridge Regression}

    The RR solution $\mathbf{w}_*$ is well-known:

   \begin{equation*}
   {\color{red}\mathbf{w}_*}= (X^\intercal
   X+\gamma \mathbb{I})^{-1}X^\intercal y  
   \end{equation*}

   which is equivalent to:
   \begin{equation*}
   {\color{red}\mathbf{w}_*}= \Big (\sum_{t=1}^m \mathbf{x}_t \mathbf{x}_t
   ^\intercal + \gamma \mathbb{I})\Big )^{-1}
   \sum_{t=1}^m y_t \mathbf{x}_t
   \end{equation*}

\end{frame}

\begin{frame}
    \frametitle{Batch methods}
    \framesubtitle{Ridge Regression}
    The prediction for a new input $\color{blue}\mathbf{x}_{m+1}$ is given by:    
    \begin{equation*}
        f({\color{blue}\mathbf{x}_{m+1}})={\color{red}\mathbf{w}_*^\intercal}
        \color{blue}\mathbf{x}_{m+1}
    \end{equation*}

    \begin{equation*}
        f({\color{blue}\mathbf{x}_{m+1}}) = \displaystyle
        \overbrace{\underbrace{\Big( \sum_{t=1}^m y_t \mathbf{x}_t \Big
        )^\intercal}_{\color{ForestGreen}b} \underbrace{\Big (
        \sum_{t=1}^m \mathbf{x}_t \mathbf{x}_t  ^\intercal + \gamma
        \mathbb{I}}_{\color{ForestGreen}A}\Big)^{-1}}^{{\color{red}\mathbf{w}_*^\intercal}}  \color{blue}\mathbf{x}_{m+1} 
    \end{equation*}
\end{frame}

\begin{frame}
     \frametitle{RR Algorithm}

    \begin{columns}[T]
    \begin{column}{.6\textwidth}
    {\color{purple}\rule{\linewidth}{4pt}}
    \begin{algorithm}[H] \begin{algorithmic}[1] \REQUIRE $\,$ \\
    $\{\mathbf{x}_1,\dots,\mathbf{x}_m \}$: $m$ input vectors \\
    $\{y_1,\dots,y_m \}$: $m$ targets \\ \ENSURE  $\,$ \\
    $\{f(\mathbf{x}_1),\dots,f(\mathbf{x}_m) \}$: model predictions \\
    \STATE Initialize $\mathbf{A}=\gamma \mathbb{I}$ and $\mathbf{b}=0$
    \FOR { $t = 1$ to $m$ } \STATE read new $\mathbf{x}_t$ \STATE output
    prediction $f(\mathbf{x}_t) =  \mathbf{b}^\intercal
    \mathbf{A}^{-1}\mathbf{x}_t$ \STATE $\mathbf{A} = \mathbf{A} +
    \mathbf{x}_t \mathbf{x}_t^\intercal$ \STATE Read new $y_t$ \STATE
    $\mathbf{b} = \mathbf{b} + y_t \mathbf{x}_t$ \ENDFOR
    \end{algorithmic} \caption{Ridge Regression} \label{alg:RR}
    \end{algorithm}
    \end{column}
    \begin{column}{.3\textwidth}
    {\color{blue}\rule{\linewidth}{4pt}}
        \[ {\color{ForestGreen}\mathbf{b}}=  \sum_{t=1}^\intercal y_t \mathbf{x}_t  \]
    \[{\color{ForestGreen}\mathbf{A}}= \sum_{t=1}^\intercal \mathbf{x}_t \mathbf{x}_t
    ^\intercal + \gamma \mathbb{I} \]

    \end{column}
    \end{columns}
\end{frame}

\begin{frame}
    \frametitle{Online methods}
    \framesubtitle{Examples}
    There are several popular online methods such as:
    \begin{itemize}
    \item Perceptron
    \item Passive-aggressive
    \item Stochastic gradient descent
    \item {\color{red} Aggregating algorithm}
    \item The second order perceptron.
    \end{itemize}
\end{frame}

\begin{frame}
     \frametitle{AAR: The Aggregating Algorithm for Regression}

    The {\color{red}AAR} (proposed by Vovk in 2001) is an application of
    the AA to the problem of regression.

    The prediction formula is given by:

    \begin{equation}
    f({\color{blue}\mathbf{x}_{m+1}}) =
    \displaystyle {\Big( \underbrace{\sum_{t=1}^m y_t \mathbf{x}_t \Big
    )}_{\color{ForestGreen}b}}^\intercal \Big ( \underbrace{\sum_{t=1}^{{\color{red}m+1}} \mathbf{x}_t \mathbf{x}_t
    ^\intercal + \gamma \mathbb{I} \Big
    )}_{\color{ForestGreen}\mathbf{A}}^{-1}
    {\color{blue}\mathbf{x}_{m+1}} \, ,
    \end{equation}

    The only difference is the matriz $\color{ForestGreen}\mathbf{A}$ is updated before
    the prediction is made.
\end{frame}

\begin{frame}
     \frametitle{AAR: The Aggregating Algorithm for Regression}

%    Algorithm~\ref{alg:AAR} shows this procedure based on
%    equation~(\ref{eq:predAAR}). This algorithm can be obtained by
%    swapping lines 4 and 5 from algorithm~\ref{alg:RR}. 
    \begin{algorithm}[H] \begin{algorithmic}[1] \REQUIRE $\,$ \\
    $\{\mathbf{x}_1,\dots,\mathbf{x}_m \}$: $m$ input vectors \\
    $\{y_1,\dots,y_m \}$: $m$ targets \\ \ENSURE  $\,$ \\
    $\{f(\mathbf{x}_1),\dots,f(\mathbf{x}_m) \}$: model predictions \\
    \STATE Initialize $\mathbf{A}=\gamma \mathbb{I}$ and $\mathbf{b}=0$
    \FOR { $t = 1$ to $m$ } \STATE read new $\mathbf{x}_t$ \STATE
    $\color{red}\mathbf{A} = \mathbf{A} + \mathbf{x}_t \mathbf{x}_t^\intercal$ \STATE
    {\color{blue}output prediction $f(\mathbf{x}_t) =  \mathbf{b}^\intercal
    \mathbf{A}^{-1}\mathbf{x}_t$} 
    \STATE Read new $y_t$ \STATE $\mathbf{b}
    = \mathbf{b} + y_t \mathbf{x}_t$ \ENDFOR
    \end{algorithmic} \caption{{\em The aggregating algorithm for
    regression}} \label{alg:AAR} \end{algorithm}

\end{frame}

\section{Model} % (fold)
\label{sec:Method}

\begin{frame}
     \frametitle{Motivation}
    Despite of RR and AAR being very successful methods, they have
    some problems:
    \begin{itemize}
        \item They consider all the data for making predictions, but some time series show a strong
    dependence on the latest information instead of all the data.
        \item A matrix inverse is computed at every step of the
        algorithm without considering the matrices are slightly different.
    \end{itemize}
\end{frame}

\begin{frame}
     \frametitle{Proposal}

    The method proposed consists of a modification of AAR considering a
    {\color{red}sliding window} which contains only the last $L+1$ samples 
    $\{\mathbf{x}_m, \mathbf{x}_{m-1}, \dots, \mathbf{x}_{m-L} \}$. 
    If we define an input matrix with the sliding window data:

    \begin{equation*}
    \mathbf{X}(t) = 
    \begin{bmatrix}
    \mathbf{x}_{t-L}^{\intercal}\\
    \vdots \\ 
    \mathbf{x}_t^{\intercal} 
    \end{bmatrix} 
    , \qquad \text{and a target vector} \qquad 
    {\bf y}(t) =
    \begin{bmatrix}
    y_{t-L} \\
    \vdots \\ 
    y_t 
    \end{bmatrix} 
    \end{equation*}
    The optimal solution at each time $t$ is:

    \begin{equation*}
    {\bf w}(t)_*=({\bf
    X}(t)^\intercal{\bf X}(t)+\gamma \mathbb{I})^{-1}{\bf
    X}(t)^\intercal{\bf y}(t) \, .  
    \end{equation*}

\end{frame}

\begin{frame}
     \frametitle{Algorithm}

    \begin{algorithm}[H] \begin{algorithmic}[1] \REQUIRE $\,$ \\
    $\{\mathbf{x}_1,\dots,\mathbf{x}_m \}$: $m$ input vectors \\
    $\{y_1,\dots,y_m \}$: $m$ targets \\ $L$: sliding window size ($L<m$)
    \\ \ENSURE  $\,$ \\ $\{f(\mathbf{x}_{L+1}),\dots,f(\mathbf{x}_m) \}$:
    model predictions \\ \STATE {\color{blue}Initialize $\mathbf{A}=\displaystyle
    \sum_{t=1}^L \mathbf{x}_t \mathbf{x}_t^\intercal + \gamma \mathbb{I}$
    and $\mathbf{b}=\displaystyle \sum_{t=1}^Ly_t \mathbf{x}_t$} \FOR { $t
    = L+1$ to $m$ } \STATE read new $\mathbf{x}_t$

    \STATE $\color{red}\mathbf{A} =
    \mathbf{A} + \mathbf{x}_t \mathbf{x}_t^\intercal- \mathbf{x}_{t-L-1}
    \mathbf{x}_{t-L-1}^\intercal  $ \STATE output prediction
    $f(\mathbf{x}_t) =  \mathbf{b}^\intercal \mathbf{A}^{-1}\mathbf{x}_t$
    \STATE Read new $y_t$ \STATE $\mathbf{b} = \mathbf{b} + y_t
    \mathbf{x}_t$ \ENDFOR

    \end{algorithmic} \caption{Recursive Ridge Regression}
    \label{alg:SLAAR} \end{algorithm}
\end{frame}

\begin{frame}
     \frametitle{Matrix Inverse} 
   Additionally, it is worth noticing that the matrix $\mathbf{A}$ is
   only slightly different:

   \begin{equation*}
        \mathbf{A} = \mathbf{A} + {\color{red}\mathbf{x}_t \mathbf{x}_t^\intercal -\mathbf{x}_{t-L-1}
       \mathbf{x}_{t-L-1}^\intercal}
   \end{equation*}

   However the inverse matrix $\mathbf{A}^{-1}$ can be computed using
   the inverse matrix previously calculated.
\end{frame}

\begin{frame}
     \frametitle{Sherman Morrison formula}
     \begin{theorem}[Sherman Morrison]
     If $\mathbf{A}$ is a positive definite matrix and its inverse
     matrix is known, then the inverse of the matrix
     $\mathbf{B}=\mathbf{A} + \mathbf{x}\mathbf{x}^\intercal$ can be
     obtained as:
     \begin{equation*}
         \mathbf{B}^{-1} = \mathbf{A}^{-1} - 
         \frac{(\mathbf{A}^{-1}\mathbf{x})(\mathbf{A}^{-1}\mathbf{x})^{\intercal}}
         {1 + \mathbf{x}^{\intercal} \mathbf{A}^{-1} \mathbf{x}}
     \end{equation*}
     \end{theorem}   
     This reduces the inverse matrix computation order from $O(n^3)$ to $O(n^2)$.
\end{frame}

\section{Experiments} % (fold)
\label{sec:Experiments}

\begin{frame}
     \frametitle{Experiments}
     \framesubtitle{Data}

    We used daily stock returns of {\color{blue} 45 stocks} from the
    technology sector dated from the {\color{blue}1st of January 2000
    to the 1st of July 2012}.

    Returns $\{x_t\}_{t=1}^{m-1}$ were
    defined based on stock prices $\{{p_t}\}_{t=1}^{m}$ and they are
    related as:

    \begin{equation*} x_t = \frac{p_{t+1}-p_t}{p_t} \, .
    \end{equation*} 
    \begin{block}{Objective}
    The objective is to build a model of the returns of
    stock $k$ based on returns of other stocks of the same financial
    sector.
    \end{block}

\end{frame}

\begin{frame}
     \frametitle{Data}

    Therefore, every input vector $\mathbf{x}_t$ will have 44 stock
    returns at time $t$ without considering information of stock $k$. The
    target vector $\mathbf{y}$ will be the stocks returns of stock $k$.

\end{frame}

\begin{frame}
     \frametitle{Experiments}
     \begin{itemize}
         \item We compared RR, AAR and our proposal.
         \item Experiments were done using 45 stocks.
         \item The error was calculated considering the {\em mean
         squared error} (MSE) using the last 100 predictions made.
     \end{itemize}

\end{frame}

\begin{frame}
     \frametitle{Results} 
    \begin{figure}[H]
      \centering
        \includegraphics[width=0.78\textwidth]{img/data}
      \caption{SPY returns}
      \label{fig:data}
    \end{figure}
\end{frame}

\section{Conclusions} % (fold)
\label{sec:Conclusions}

\begin{frame}
     \frametitle{Conclusions}
    \begin{itemize}
        \item The experiments show that our method was better in 28 of
45 stocks.
        \item The smaller error in our method was for the SPY stock,
        which is an expected result considering that the SPY value is
        constructed based on the main stocks of the markets.
        \item For SPY, it is shown how AAR and
        our method fit better than RR to the target and also that our
        method is slightly better compared with AAR.
        \item It's also possible to notice that when returns change
        drastically none of the methods fit satisfactorily.
    \end{itemize}
\end{frame}

\begin{frame}
     \frametitle{Conclusions}
    \begin{itemize}
    \item The proposal shows an online version of RR which allows the
    number of calculations required to be reduced and it is therefore a better
    option for financial algorithms.
    \item The tests show that in some cases, a portion of data will
    provide a better model instead of using all data.
    \item The historical dependence should be studied in order to
    obtain more accurate algorithms.
    \end{itemize}
\end{frame}

### 2016-03-22 articles-conferences/icpram2015/slides/src/slides_example.tex
agregando repo icpram

\section{HFT time series}
\subsection{Intro}

\begin{frame}
\frametitle{Motivation: Financial Time Series Forecasting}
\begin{columns}
\begin{column}{.48\textwidth}
\color{purple}\rule{\linewidth}{4pt}
Characteristics
  \begin{itemize}
  \item High Frequency data
  \item Large number of factors affecting the market
  \end{itemize}
    \begin{center}
    \includegraphics[width=0.8\textwidth]{img/financial_ts}
    \end{center}
\end{column}%
\begin{column}{.48\textwidth}
\color{blue}\rule{\linewidth}{4pt}
Requires
  \begin{itemize}
  \item Fast implementations
  \item High Frequency Trading Algorithms 
  \end{itemize}
    \begin{center}
    \includegraphics[width=0.9\textwidth]{img/hft}
    \end{center}
\end{column}
\end{columns}

\end{frame}

\begin{frame}
\frametitle{FFN: Feed Forward Neural Network}

FFNs are widely-used to model problems for learning tasks and especially
used to forecast financial time series.

\begin{itemize}
\item FFNs have three layers: 
\begin{enumerate}[a)]
\item Input layer
\item Hidden layer
\item Output layer
\end{enumerate}
\item Each layer consists of a fixed set of neurons
\item All neurons in a layer are fully connected with the other neurons
of the following layer
\item Each connection has an associated weight related to its
dependence
\end{itemize}

\end{frame}

\begin{frame}
  \frametitle{Feed Forward Neural Network}
%\begin{center}
%\includegraphics[width=0.8\textwidth]{img/simple_feedforward_network}
%\end{center}

  %\input{src/ffn1}
\end{frame}

\begin{frame}
  \frametitle{Feed Forward Neural Network}
%\begin{center}
%\includegraphics[width=0.8\textwidth]{img/simple_feedforward_network}
%\end{center}

  %\input{src/ffn}
\end{frame}
\begin{frame}
    \frametitle{FFN: Training and Testing phases}

The learning process to set the weights and biases is done in a
{\color{red}training phase}, which is computationally expensive and
can be computed offline.

However, the output calculation or {\color{red}testing phase} can not
be done offline and needs to be as fast as possible.

\end{frame}

\section{Model}
\subsection{math}

\begin{frame}
\frametitle{Matrix representation}
\framesubtitle{Input Layer}

\begin{columns}
\begin{column}{.48\textwidth}
The input data can be represented as a pair $(\mathbf{x},y)$, where $\mathbf{x}$ is a vector:
\[
        \mathbf{x}=\begin{bmatrix}
        x_1 \\ x_2 \\ \vdots \\ x_n
        \end{bmatrix}
\]

and $y$ is the target.

\end{column}

\begin{column}{.48\textwidth}
In order to simplify the calculations, an extra row set to 1 is added:
\[
        x=\begin{bmatrix}
        x_1  \\ \vdots \\ x_n \\ 1
        \end{bmatrix}
\]

which is done to include the bias.

\end{column}
\end{columns}
\end{frame}

\begin{frame}
\frametitle{Input-Hidden weights}

The connection between the Input and the Hidden layer can be expressed
in a matricial form:
\[
    {\color{blue}
    \mathbf{W}_{h \times {n+1}}} =
    \left[
      \begin{array}{ccc|c}
        w_{1, 1} & \cdots & w_{1, n} & w_{1, n+1} \\
        \vdots   & \ddots & \vdots   & \vdots     \\
        w_{h, 1} & \cdots & w_{h, n} & w_{h, n+1}
      \end{array}
    \right],
  \]
where $w_{i,j}$ is the connection between the hidden neuron $i$ and
the input neuron $j$, $n$ is the length of the input vector and $h$ is
the number of neurons in the hidden layer.
The last column represents the bias for each hidden neuron.

\end{frame}

\begin{frame}
  \frametitle{Hidden-Output weights}

The weight matrix for the Hidden-Output weights is:

  \[
    {\color{blue}
    \mathbf{\Delta}_{m \times h+1}} =
    \left[
      \begin{array}{ccc|c}
        \delta_{1,1} & \cdots & \delta_{1,h} & \delta_{1,h+1}\\
        \vdots & \ddots &  \vdots & \vdots\\
        \delta_{m,1} & \cdots & \delta_{m,h} &\delta_{m,h+1}
      \end{array}
    \right],
  \]

where $\delta_{i,j}$ is the weight between the output neuron $i$ and
the hidden neuron $j$ and $m$ is the number of output neurons. The
bias is represented in the last column.

\end{frame}

\begin{frame}
  \frametitle{State of hidden neurons}

The input of a hidden neuron $i$ is

\[   
     {\color{blue}[\mathbf{Wx}]_{i}}=
     \sum_{j=1}^{n+1}
      w_{i,j}\,x_j 
  \]

and the output

\[
\eta_{i} = \phi \left(
{\color{blue}[\mathbf{Wx}]_{i}}
    \right)
\]

where $\phi$ is the logistic sigmoid function:
\[
\phi(x) = \frac{1}{1+e^{-x}}
\]

\end{frame}

\begin{frame}
  \frametitle{State of output neurons}

The input of an output neuron $i$ is
\[
    {\color{blue}[\mathbf{\Delta \eta}]_{i}} =
      \sum_{j=1}^{h+1}
      \delta_{i,j}\,\eta_i
  \]

and the output

\[
\gamma_{i} = \phi \left(
    {\color{blue}[\mathbf{\Delta \eta}]_{i}} 
    \right)
\]

which is the FFN output.

\end{frame}

\begin{frame}
  \frametitle{Algorithm}

The algorithm in matricial form for an input vector $\mathbf{x}$ is:

  \begin{align*}
    \mathbf{\eta} &= \mathbf{Wx} & \qquad \text{//Input for the
    Hidden layer }\\
    \mathbf{\eta} &= \mathbf{\phi}(\eta) & \qquad \text{//Applying activation
    function}\\
    \mathbf{\gamma} &= \mathbf{\Delta\eta} & \qquad \text{// Input for the
    Output layer}\\
    \mathbf{\gamma} &= \mathbf{\phi(\gamma)} & \qquad \text{//Applying
    activation function}
  \end{align*}

\end{frame}

%\begin{frame}
%  \frametitle{Interaction diagram}
%%  \includegraphics[width=0.7\textwidth]{img/interaction-diagram.eps}
%\begin{center}
%\includegraphics[width=0.9\textwidth]{img/interaction_diagram}
%\end{center}
%\end{frame}

%\begin{frame}
%  \frametitle{CUBLAS}
  %\lstinputlisting{src/cublas.cu}
%\end{frame}

%\begin{frame}
%  \frametitle{CUDA kernels}
  %\lstinputlisting[basicstyle=\small]{src/kernels.cu}
%\end{frame}

\section{Experiments}
\subsection{sol}
\begin{frame}
  \frametitle{Experiments}
  \begin{itemize}
    \item Problem: $n=h=m=1000,\dots,6000$
    \item Two implementations:
    \begin{itemize}
    \item GPU using CUBLAS (CUDA library for Basic Linear Algebra
    Subprograms)
    \item CPU using MKL (Math kernel library)    
    \end{itemize}
    \item Hardware
    \begin{itemize}
    \item Two 6-core Intel Xeon X5650 2.67GHz
    \item NVIDIA M2050 GPU,448 cores, 1.15Ghz    
    \end{itemize}
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Arquitecture of the solution}
  \begin{center}
  \includegraphics[width=0.7\textwidth]{img/arq1}    
  \end{center}
\end{frame}
\begin{frame}
\frametitle{Results}
    \begin{center}
     \includegraphics[width=0.9\textwidth]{img/data}    
    \end{center}
\end{frame}

\section{Conclusions}
\subsection{conclusion}

\begin{frame}
  \frametitle{Conclusions}
  \begin{itemize}
    \item Activation function time is negligible ($\sim$2\%).
    \item The GPU and CPU computation times are comparable when the
    FFN size is small.
    \item The GPU implementation is recommended for a large number of
    neurons (financial models).
    \item ZeroMQ library allows an efficient communication library for
    high frequency data.
  \end{itemize}
\end{frame}

%\section{Future work}
%\subsection{future}
%
\begin{frame}

    \begin{center}

    {\color{red}\Huge Questions?}
    \end{center}

\end{frame}

### 2016-03-22 articles-conferences/icpram2015/src/background.tex
agregando repo icpram

\section{\uppercase{Background}}
\label{sec:background}
\noindent
\subsection{Integration and Cointegration}
A time series $\mathbf{y}$ is said to be integrated of order $d$ if after
differentiating the variable $d$ times, we get an I(0) process, more precisely:

\[
(1-L)^d \mathbf{y} \sim \text{I(0)} \, ,
\]

\noindent where I(0) is a stationary time series and $L$ is the lag operator:

\[
(1-L)\mathbf{y} = \Delta \mathbf{y}=\mathbf{y}_t  -\mathbf{y}_{t-1} \quad \forall t
\]

Let $\mathbf{y}_t = \{\mathbf{y}^1, \dots, \mathbf{y}^l\}$ be a set of $l$
stationary time series I(1) which are said to be cointegrated if a vector
$\beta=[\beta(1),\dots,\beta(l)]^\top \in \mathbb{R}^l$  exists such that the
time series,

\begin{equation}
 \mathbf{Z}_t:= \beta^\top \mathbf{y}_t = \beta(1) \mathbf{y}^1 + \dots + \beta(l) \mathbf{y}^l \sim
 \text{I(0)}\, .
\end{equation}

In other words, a set of I(1) variables is said to be cointegrated if
a linear combination of them exists which is I(0).

\subsection{Vector Autorregresive Models}\label{sec:varvec}

VECM is a special case of VAR model and both describe the joint behaviour of a
set of variables.

VAR($p$) model is a general framework to describe the behaviour of a set of $l$
endogenous variables as a linear combination of their last $p$ values. These
$l$ variables at time $t$ are represented by the vector $\mathbf{y}_t$ as
follows:

\begin{equation*}
\label{eq:variables}
\mathbf{y}_t = 
\begin{bmatrix} y_{1,t} &
y_{2,t} &
\dots &
y_{l,t}
\end{bmatrix}^\top \, ,
\end{equation*}
\noindent where $y_{j,t}$ corresponds to the time series $j$ evaluated at
time $t$.

The VAR(p) model describes the behaviour of a dependent variable in terms
of its own lagged values and the lags of the others variables in the
system. The model with $p$ lags is formulated as the following:

\begin{equation}
\label{eq:var}
 \mathbf{y}_t = \phi_1 \mathbf{y}_{t-1}  + \dots +   \phi_p\mathbf{y}_{t-p}
+ \mathbf{c} + \mathbf{\epsilon}_t \, ,
\end{equation}

\noindent where $t=p+1, \dots N$,  ${\phi_1,\dots,\phi_p}$ are $l \times l$
matrices of real coefficients,
$\mathbf{\epsilon}_{p+1},\dots,\mathbf{\epsilon}_N$ are error terms,
$\mathbf{c}$ is a constant vector and $N$ is the total number of samples.

The VAR matrix form of equation~(\ref{eq:var}) is:

\begin{equation}\label{eq:vareq}
\mathbf{B} = \mathbf{A} \mathbf{X} + 
\mathbf{E} \, , 
\end{equation}

\noindent where:
\begin{eqnarray*}
\mathbf{A} &=&
\begin{bmatrix}
   \mathbf{y}_{p}     & \dots    & \mathbf{y}_{N-1}\\
   \mathbf{y}_{p-1}   & \dots    & \mathbf{y}_{N-2}\\
   \vdots             & \ddots   & \vdots\\
   \mathbf{y}_{1}     & \dots    & \mathbf{y}_{N-p}\\
   1                  & \dots    & 1 
   \end{bmatrix}^\top \, ,\\
\mathbf{B} &=&
\begin{bmatrix}
\quad \\
\mathbf{y}_{p+1} &
\dots & 
\mathbf{y}_N \\
\quad
\end{bmatrix}^\top  \, ,\\
\mathbf{X}&=&
\left[
\begin{array}{ccccc}
    \quad  & \quad & \quad & \quad \\
    \phi_1   & \cdots & \phi_p & \mathbf{c} \\  
    \quad & \quad & \quad & \quad
    \end{array}
\right]^\top \, ,\\
\mathbf{E} &=& 
\begin{bmatrix}
    \quad \\
    \mathbf{\epsilon}_{p+1}  & 
    \dots                & 
    \mathbf{\epsilon}_N \\
    \quad
\end{bmatrix}^\top \, . 
\end{eqnarray*}

Equation~(\ref{eq:vareq})  can be solved using ordinary least squares estimation.

\subsection{VECM}

VECM is a special form of a VAR model for I(1) variables that are also
cointegrated~\cite{banerjee1993}. It is obtained by replacing the form $\Delta
\mathbf{y}_t = \mathbf{y}_t - \mathbf{y}_{t-1}$ in equation (\ref{eq:var}).
VECM is expressed in terms of differences, has an error correction term and 
the following form:

\begin{equation}
 \label{eq:vec}
 \Delta \mathbf{y}_t = 
 \underbrace{ \Omega\mathbf{y}_{t-1}}_\text{Error correction term} + 
 \sum_{i=1}^{p-1}
\phi_i^* \Delta \mathbf{y}_{t-i}  + \mathbf{c} + \mathbf{\epsilon}_t \quad ,
\end{equation}

\noindent where coefficients matrices $\Omega$ and $\phi_i^*$ are
function of matrices $\phi_i$ (shown in equation (\ref{eq:var})) as follows:

\begin{eqnarray*}
\phi_i^* &: =& -\sum_{j=i+1}^{p} \phi_j \\
\Omega &: =& -(\mathbb{I}-\phi_1-\dots-\phi_p) 
\end{eqnarray*}

The matrix $\Omega$ has the following properties~\cite{johansen1995}:
\begin{itemize}
\item If $\Omega = 0$ there is no cointegration
\item If $rank(\Omega)=l$ i.e full rank, then the time series are not
I(1) but stationary
\item If $rank(\Omega)=r,\quad 0 < r < l$ then, there is cointegration
and the matrix $\Omega$ can be expressed as $\Omega =
\alpha \beta^\top$, where $\alpha$ and $\beta$ are $(l \times r)$
matrices and $rank(\alpha)=rank(\beta)=r$.

The columns of $\beta$ contains the cointegration vectors and the rows of
$\alpha$ correspond with the adjusted vectors. $\beta$ is obtained by Johansen
procedure~\cite{johansen1988} whereas $\alpha$ has to be determined as a
variable in the VECM.

It is worth noticing that the factorization of the matrix $\Omega$ is not
unique since for any $r \times r$ nonsingular matrix $H$ we have:

\begin{eqnarray*}
\alpha \beta^\top &=& \alpha \mathbf{HH^{-1}} \beta^\top\\
&=&(\alpha\mathbf{H})(\beta(\mathbf{H}^{-1})^\top)^\top \\
&=& \alpha^*(\beta^*)^\top
\end{eqnarray*}

\noindent with $\alpha^* = \alpha\mathbf{H}$ and $\beta^* =
\beta(\mathbf{H}^{-1})^\top$.

\end{itemize}

If cointegration exists, then equation (\ref{eq:vec}) can be written as follows:

\begin{equation*}
 \label{eq:vecfull}
 \Delta \mathbf{y}_t = \alpha \beta^\top\mathbf{y}_{t-1} 
 + \sum_{i=1}^{p-1} \phi_i^*\Delta
\mathbf{y}_{t-i}  + \mathbf{c} + \mathbf{\epsilon}_t \quad ,
\end{equation*}

\noindent which is a VAR model but for time series differences.

VECM has the same form shown in equation~(\ref{eq:vareq}) but with
different matrices:

\begin{eqnarray}\label{eq:vecmatrix}
 \mathbf{A}&=&
   \begin{bmatrix} 
   \beta^\top \mathbf{y}_{p} & 
   \cdots & \beta^\top \mathbf{y}_{N-1} \\
   \Delta \mathbf{y}_p  & \cdots 
   &\Delta\mathbf{y}_{N-1} \\ 
   \vdots  & \ddots & \vdots \\
   \Delta\mathbf{y}_2  & \cdots 
   & \Delta \mathbf{y}_{N-p+1} \\
   1 & \cdots & 1 
   \end{bmatrix}^\top \label{eq:vecA} \, ,\\
\mathbf{B} & = &
 \begin{bmatrix}
 \quad\\
  \Delta \mathbf{y}_{p+1} & 
  \dots &
  \Delta \mathbf{y}_N \label{eq:vecB}\\
  \quad
 \end{bmatrix}^\top \, ,\\
\mathbf{X}&=&
  \begin{bmatrix}
   \quad \\
   \alpha & \phi_1^* & \cdots & \phi_{p-1}^* & \mathbf{c} \\  
   \quad
   \end{bmatrix}^\top \, ,\\
\mathbf{E} &=&
\begin{bmatrix}
   \quad \\
   \mathbf{\epsilon}_{p+1} &
   \dots &
   \quad &\mathbf{\epsilon}_N \, \\ \quad
\end{bmatrix}^\top 
\end{eqnarray}

VAR and VECM parameters shown in equation (\ref{eq:vareq}) can be solved
using standard regression techniques, such as ordinary least squares (OLS).

\subsection{Ordinary Least Squares method}

When $\mathbf{A}$ is singular, solution to equation~(\ref{eq:vareq}) is given
by the ordinary least squares (OLS) method. OLS consists of minimizing the sum
of squared errors or equivalently minimizing the following expression:

\begin{equation*}
\label{eq:regressionproblem}
\underset{\mathbf{X}}{\text{min}} \quad \| \mathbf{A}\mathbf{\mathbf{X}} - \mathbf{B} \|_2^2
\end{equation*}

\noindent for which the solution $\hat{\mathbf{X}}$ is well-known:

\begin{equation*}
\label{eq:MP}
\hat{\mathbf{X}}=\mathbf{A}^{\!\!+}\,\mathbf{B}
\end{equation*}

\noindent where $\mathbf{A}^{\!\!+}$ is the Moore-Penrose pseudo-inverse
which can be written as follows:

\begin{equation}
\label{eq:pseudoinverse}
\mathbf{A}^{\!\!+}= (\mathbf{A}^{\!\!\top} \mathbf{A})^{-1}\mathbf{A}^{\!\!\top} \, .
\end{equation}

However, when $\mathbf{A}$ is not full rank, i.e
$rank(\mathbf{A})=k <  n \leq m$, $\mathbf{A}^\top \mathbf{A}$ is
always singular and equation~(\ref{eq:pseudoinverse}) cannot be used.
More generally, the pseudo-inverse is best computed using the compact
singular value decomposition (SVD) of $\mathbf{A}$:

\begin{equation*}
    \label{eq:compactsvd}
    \underset{m \times n}{\mathbf{A}}=
    \underset{m \times k}{\mathbf{U_1}} \enskip
    \underset{k \times k}{\Sigma_1} \enskip
    \underset{k \times n}{\mathbf{V}_1^{\top}} \, ,
\end{equation*}

\noindent as follows

\begin{equation*}
\label{eq:pseudoinversesvd}
\mathbf{A}^{\!\!+} = \mathbf{V}_1 \Sigma_1^{-1} \mathbf{U}_1^\top \, .
\end{equation*}

### 2016-03-22 articles-conferences/icpram2015/src/conclusions.tex
agregando repo icpram

--- 2014-11-27 src/5_Proposal.tex
@@
-\chapter{Online VECM proposal}
-\section{\uppercase{Methodology}}
-\label{sec:methodology}
-\noindent 
-\subsection{Online VECM} \label{sec:proposal}
+\section{\uppercase{Conclusions}}
+\label{sec:conclusions}
+\noindent A new online vector error correction method was presented. 
+We have shown that our proposed OVECM considerably reduces execution times
+without compromising solution accuracy.  OVECM was compared with VECM and ARIMA
+with the same sliding window sizes and OVECM outperformed both in terms of
+execution time. Traditional VECM slightly outperformed our proposal but the
+OVECM execution time is lower.
+This reduction of execution time is mainly because OVECM avoids the cointegration vector
+calculation using the Johansen method. The condition for getting new vectors is
+given by the mean\_error variable which controls how many times the Johansen method is
+called. Additionally, OVECM introduces matrix optimization in order to get the
+new model in an iterative way.
+We could see that our algorithm took much less than a minute at every
+step. This means that it could also be used with higher frequency data and would
+still provide responses before new data arrives.
 
-Since VECM parameter estimation is computationally expensive, we propose an
-online version of VECM (OVECM).  OVECM considers only a sliding window of the
-most recent data. Moreover, since cointegration vectors represent long-run
-relationships which vary little in time, OVECM determines firstly if they require calculation.
+For future study, it would be interesting to improve the out-of-sample forecast
+by considering more explicative variables, to increase window sizes or trying new
+conditions to obtain new cointegration vectors.
 
-OVECM also implements matrix optimizations in order to reduce execution time,
-such as updating matrices with new data, removing old data and introducing new
-cointegration vectors.
-
-Algorithm~\ref{alg:proposal} shows our OVECM proposal which considers the
-following:
-
-\begin{itemize}
-\item The function \texttt{getJohansen} returns cointegration vectors given by
-the Johansen method considering the trace statistic test at 95\%
-significance level.
-\item The function \texttt{vecMatrix} returns the matrices~(\ref{eq:vecA})
-and~(\ref{eq:vecB}) that allows VECM to be solved.
-\item The function \texttt{vecMatrixOnline} returns the
-matrices~(\ref{eq:vecA}) and~(\ref{eq:vecB}) aggregating new data and removing
-the old one, avoiding calculation of the matrix $\mathbf{A}$ from scratch.
-\item Out-of-sample outputs are saved in the variables 
-$\mathbf{Y}_{\text{true}}$ and $\mathbf{Y}_{\text{pred}}$.
-\item The model is solved using OLS.
-\item In-sample outputs are saved in the variables $\Delta
-\mathbf{y}_{\text{true}}$ and $\Delta \mathbf{y}_{\text{pred}}$.
-\item The function \texttt{mape} gets the in-sample MAPE for the $l$ time
-series.
-\item Cointegration vectors are obtained at the beginning and when they are required to be updated. This updating is decided based on the in-sample MAPE of the last $n$ inputs. The average of all
-MAPEs are stored in the variable $e$. If the average of MAPEs
-($\texttt{mean}(e)$) is above a certain error given by the mean\_error threshold, cointegration vectors are updated.
-\item If new cointegration vectors are required, the function
-\texttt{vecMatrixUpdate} only updates the corresponding columns of matrix
-$\mathbf{A}$ affected by those vectors (see equation~\ref{eq:vecA}).
-\end{itemize}
-
-\begin{algorithm}[ht]
-\begin{algorithmic}[1]
-\REQUIRE $\,$ \\
-$\mathbf{y}$: matrix with $N$ input vectors and $l$ time series\\
-$p$: number of past values \\
-$L$: sliding window size ($L<N$) \\
-$\text{mean\_error}$: MAPE threshold \\
-$n$: interpolation points to obtain MAPE \\
-\ENSURE  $\,$ \\
-$\{\mathbf{y}_{\text{pred}}[L+1],\dots,\mathbf{y}_{\text{pred}}[N]\}$: model predictions 
-\FOR { $i =0$ to $N-L$ }
-    \STATE $\mathbf{y}_i \gets \mathbf{y}[i:i+L]$
-	\IF {i = 0 }
-	    \STATE{$v \gets \texttt{getJohansen}(\mathbf{y}_i,p)$}
-	    \STATE{$[\mathbf{A} \quad \mathbf{B}] \gets
-        \texttt{vecMatrix}(\mathbf{y}_i,p,v)$}
-	\ELSE
-	    \STATE{$[\mathbf{A} \quad \mathbf{B}] \gets
-        \texttt{vecMatrixOnline}(\mathbf{y}_i,p,v,\mathbf{A},\mathbf{B})$}
-        %\STATE $\Delta \mathbf{Y}_{\text{true}}[i] \gets \mathbf{B}[-1,:]$
-        \STATE $\Delta \mathbf{Y}_{\text{pred}}[i] \gets \mathbf{A}[-1,:] \times \mathbf{X}$
-    \ENDIF
-    \STATE $\mathbf{X} \gets \texttt{OLS} (\mathbf{A},\mathbf{B})$%\mathbf{(A^\top A)^{-1}A^\top B}$
-    %\STATE $\Delta \mathbf{y}_{\text{true}} \gets \mathbf{B}[-n,:]$
-    %\STATE $\Delta \mathbf{y}_{\text{pred}} \gets \mathbf{A}[-n,:] \times \mathbf{X}$
-    \STATE $e \gets \texttt{mape}(\mathbf{B}[-n,:],\mathbf{A}[-n,:] \times \mathbf{X})$
-    %\STATE $e \gets \texttt{mape}(\Delta \mathbf{y}_{\text{true}}, \Delta
-    %\mathbf{y}_{\text{pred}})$
-    \IF {$\texttt{mean}(e) > \text{mean\_error}$}
-	    \STATE{$v \gets \texttt{getJohansen}(\mathbf{y}_i,p)$}
-	    \STATE{$\mathbf{A} \gets
-        \texttt{vecMatrixUpdate}(\mathbf{y}_i,p,v,\mathbf{A})$}
-        \STATE $\mathbf{X} \gets \texttt{OLS} (\mathbf{A},\mathbf{B})$%\mathbf{(A^\top A)^{-1}A^\top B}$
-        %\STATE $\mathbf{X} \gets \mathbf{(A^\top A)^{-1}A^\top B}$
-    \ENDIF
-\ENDFOR
-\STATE $\mathbf{Y}_{\text{true}} \gets \mathbf{Y}[L+1:N]$
-\STATE $\mathbf{Y}_{\text{pred}}\gets 
-\mathbf{Y}[L:N-1] +\Delta \mathbf{Y}_{\text{pred}}$
-%\STATE $\text{MAPE} \gets \texttt{mape}(\Delta \mathbf{Y}_{\text{true}}, \Delta
-%\mathbf{Y}_{\text{pred}})$
-\end{algorithmic}
-\caption{OVECM: Online VECM}
-\label{alg:proposal}
-\end{algorithm}
-
-Our proposal was compared against VECM and ARIMA. Both algorithms were adapted
-to an online context in order to get a reasonable comparison with our proposal
-(see algorithms \ref{alg:SLVECM} and \ref{alg:SLARIMA}). VECM and ARIMA are
-called with a sliding window of the most recent data, whereby the models are
-updated at every time step.
-
-\begin{algorithm}[ht]
-\begin{algorithmic}[1]
-\REQUIRE $\,$ \\
-$\mathbf{y}$: matrix with $N$ input vectors and $l$ time series\\
-$p$: number of past values \\
-$L$: sliding window size ($L<N$) \\
-\ENSURE  $\,$ \\
-$\{ \mathbf{y}_{\text{pred}}[L+1],\dots,\mathbf{y}_{\text{pred}}[N]\}$: model predictions 
-\FOR { $i =0$ to $N-L$ }
-    \STATE $\mathbf{y}_i \gets \mathbf{y}[i:i+L+1]$
-        \STATE $model = VECM(\mathbf{y}_i, p)$
-        \STATE $\mathbf{Y}_{\text{pred}}[i] = model.predict(\mathbf{y}[i+L])$
-\ENDFOR
-\STATE $\mathbf{Y}_{\text{true}} = \mathbf{y}[i+L+1:N] $
-\end{algorithmic}
-\caption{SLVECM: Sliding window VECM}
-\label{alg:SLVECM}
-\end{algorithm}
-
-Since we know out time series are I(1) SLARIMA is called with $d=1$. ARIMA is
-executed for every time series.
-
-\begin{algorithm}[ht]
-\begin{algorithmic}[1]
-\REQUIRE $\,$ \\
-$\mathbf{y}$: matrix with $N$ input vectors and $l$ time series\\
-$p$: autoregressive order \\
-$d$: integrated order\\
-$q$: moving average order\\
-$L$: sliding window size ($L<N$) \\
-\ENSURE  $\,$ \\
-$\{ \mathbf{y}_{\text{pred}}[L+1],\dots,\mathbf{y}_{\text{pred}}[N]\}$: model predictions 
-\FOR { $i =0$ to $N-L$ }
-\FOR { $j =0$ to $l-1$ }
-    \STATE $\mathbf{y}_i \gets \mathbf{y}[i:i+L+1,j]$
-        \STATE $model = ARIMA(\mathbf{y}_i, (p,d,q))$
-        \STATE $\mathbf{Y}_{\text{pred}}[i,j] = model.predict(\mathbf{y}[i+L,j])$
-\ENDFOR
-\ENDFOR
-\STATE $\mathbf{Y}_{\text{true}} = \mathbf{y}[i+L+1:N,:] $
-\end{algorithmic}
-\caption{SLARIMA: Sliding window ARIMA}
-\label{alg:SLARIMA}
-\end{algorithm}
-
-Both OVECM and SLVECM time complexity is dominated by Johansen method which is
-$O(n^3)$. Thus, both algorithms order is $O(Cn^3)$ where $C$ is the number of
-iterations.
-
-\subsection{Evaluation methods} \label{sec:evaluation}
-
-Forecast performance is evaluated using different methods. We have chosen three
-measures usually used:
-\begin{description}
-\item[MAPE] Mean Average Percent Error which presents forecast errors as a
-percentage.
-\begin{equation}\label{eq:MAPE}
-\text{MAPE} = \frac{1}{N} \sum_{t=1}^{N} 
-\frac{\left|\mathbf{y}_t-\hat{\mathbf{y}}_t\right|}{\left|\mathbf{y}_t\right|}
- \times 100 
-\end{equation}
-\item[MAE] Mean Average Error which measures the distance between forecasts to the
-true value.
-\begin{equation}\label{eq:MAE}
-\text{MAE} = \frac{1}{N} \sum_{t=1}^{N} 
-\left| 
-\mathbf{y}_t-\hat{\mathbf{y}}_t
-\right| 
-\end{equation}
-\item[RMSE] Root Mean Square Error also measures the distance between forecasts
-to the true values but, unlike MAE, large deviations from the true value have a
-large impact on RMSE due to squaring forecast error.
-\begin{equation}\label{eq:RMSE}
-\text{RMSE} = \sqrt{
-\frac{\displaystyle \sum_{t=1}^{N} (\mathbf{y}_t-\hat{\mathbf{y}}_t)^2}{N}}
-\end{equation}
-\end{description}
-
-\subsection{Model selection} \label{sec:pselection}
-Akaike Information Criterion (AIC) is often used in model selection where AIC
-with smaller values are preferred since they represent a trade-off between bias
-and variance.  AIC is obtained as follows:
-
-\begin{equation}
-\label{eq:aicformula}
-AIC = \underset{\text{bias}}{-\frac{2l}{N}} + 
-\underset{\text{variance}}{\frac{2k}{N}}
-\end{equation}
-
-\noindent where
-
-\begin{description}
-\item[l] is the loglikelihood function
-\item[k] number of estimated parameters
-\item[N] number of observations
-\end{description}
-
-Loglikelihood function is obtained from the Residual Sum of Squares (RSS):
-
-\begin{equation}
-\label{eq:ll}
-l = -\frac{N}{2} \left(1 + ln(2\pi) + ln\left(\frac{RSS}{N}\right)\right) 
-\end{equation}
-
-    \begin{tikzpicture}[node distance=2cm]
-    %\node (start) [startstop] {Start};
-    %\node (in1) [io, below of=start] {Input};
-    \node (in1) [io] {Input: y, p, L, $\epsilon$, n};
-    \node (pro1) [process, below of=in1] {$y_i = y[i:i+L]$};
-    \node (dec1) [decision, below of=pro1, yshift=-0.5cm] {i = 0};
-    \node (pro2a) [process, below of=dec1, yshift=-0.5cm] {$[\mathbf{A} \quad \mathbf{B}] =\texttt{ovecm}(\mathbf{y}_i,p,v,\mathbf{A},\mathbf{B})$
-                                                        $\Delta \mathbf{Y}_{\text{true}}[i] = \mathbf{B}[-1,:]$
-                                                        $\Delta \mathbf{Y}_{\text{pred}}[i] = \mathbf{A}[-1,:] \times \mathbf{X}$};
-
-    \node (pro2b) [process, right of=dec1, xshift=4cm] {$v = \texttt{Johansen}(\mathbf{y}_i,p)$ $[\mathbf{A} \quad 
-                                                                \mathbf{B}] = \texttt{vecm}(\mathbf{y}_i,p,v)$};
-    \node (pro3) [process, below of=pro2a, yshift=-0.5cm] {$\mathbf{X} = \text{OLS} (\mathbf{A},\mathbf{B})$ \\
-                                                        $\mathbf{y}_{\text{true}}[i] = \mathbf{y}_i[-n,:]$ \\
-                                                        $\mathbf{y}_{\text{pred}}[i] = \mathbf{A}[-n,:] \times \mathbf{X} +$ \\$ \mathbf{y}_i[-n-1,:-1]$ \\
-                                                        $e = \texttt{mape}(\mathbf{y}_{\text{true}}, \mathbf{y}_{\text{pred}})$};
-
-    \node (dec2) [decision, below of=pro3, yshift=-0.5cm] { $e > \epsilon $ };
-
-    \node (pro4a) [process, right of=dec2, xshift=4cm] {$v = \texttt{Johansen}(\mathbf{y}_i,p)$ \\
-                                                        $\mathbf{A} = \texttt{uvecm}(\mathbf{y}_i,p,v,\mathbf{A})$ \\
-                                                        $\mathbf{X} = \texttt{OLS} (\mathbf{A},\mathbf{B})$};
-
-    \node (pro5) [process, left of=pro2a, xshift=-4cm] {$ i++ $};
-
-    \node (dec3) [decision, below of=dec2, yshift=-0.5cm] { $i > it $ };
-
-    \node (out1) [io, below of=dec3] {Output: $\Delta \mathbf{y}_{true}, \Delta \mathbf{y}_{pred}$};
-
-    %\node (stop) [startstop, below of=out1] {Stop};
-
-    %\draw [arrow] (start) -- (in1);
-    \draw [arrow] (in1) -- (pro1);
-    \draw [arrow] (pro1) -- (dec1);
-    \draw [arrow] (dec1) -- node[anchor=east] {no} (pro2a);
-    \draw [arrow] (dec1) -- node[anchor=south] {si} (pro2b);
-    \draw [arrow] (pro2a) -- (pro3);
-    \draw [arrow] (pro2b) |- (pro3);
-    \draw [arrow] (pro3) -- (dec2);
-    \draw [arrow] (dec2) --  node[anchor=south] {si} (pro4a);
-    \draw [arrow] (pro4a) |- (dec3);
-
-    \draw [arrow] (dec2) --  node[anchor=east] {no} (dec3);
-    \draw [arrow] (dec3.west) -| node[anchor=east] {no} (pro5);
-    \draw [arrow] (pro5) |- (pro1);
-    \draw [arrow] (dec3) --  node[anchor=east] {si} (out1);
-    %\draw [arrow] (out1) -- (stop);
-    \end{tikzpicture}
+Since OVECM is an online algorithm which optimizes processing time, it could be
+used by investors as an input for strategy robots. Moreover, some technical
+analysis methods could be based on its output.

### 2016-03-22 articles-conferences/icpram2015/src/intro.tex
agregando repo icpram

\section{\uppercase{Introduction}}
\label{sec:introduction}
\noindent In finance, it is common to find variables with long-run equilibrium
relationships. This is called cointegration and it reflects the idea of that
some set of variables cannot wander too far from each other. Cointegration means
that one or more linear combinations of these variables are stationary even
though individually they are not~\cite{engle87}. Furthermore, the number of
cointegration vectors reflects how many of these linear combinations exist. Some
models, such as the Vector Error Correction (VECM), take advantage of this property
and describe the joint behaviour of several cointegrated variables.

VECM introduces this long-run relationship
among a set of cointegrated variables as an error correction term. VECM is a
special case of the vector autorregresive model (VAR) model. VAR model
expresses future values as a linear combination of variables past values.
However, VAR model cannot be used with non-stationary variables. VECM is a
linear model but in terms of variable differences. If cointegration exists,
variable differences are stationary and they introduce an error correction term
which adjusts coefficients to bring the variables back to equilibrium. In
finance, many economic time series are revealed to be stationary when they are
differentiated and cointegration restrictions often improves
forecasting~\cite{duy1998}. Therefore, VECM has been widely adopted.

In finance, pair trading is a very common example of cointegration
application~\cite{herlemont2003} but cointegration can also be extended to a
larger set of variables~\cite{mukherjee1995},\cite{engle2004}.

Both VECM and VAR model parameters are obtained using ordinary least squares
(OLS) method. Since OLS involves many calculations, the parameter estimation
method is computationally expensive when the number of past values and
observations increases. Moreover, obtaining cointegration vectors is also an
expensive routine.

Recently, online learning algorithms have been proposed to solve problems with
large data sets because of their simplicity and their ability to update the
model when new data is available. The study presented by ~\cite{arce+salinas2012}
applied this idea using ridge regression.

There are several popular online methods such as perceptron~\cite{rosenblatt58},
passive-aggressive~\cite{crammerETall2006}, stochastic gradient
descent~\cite{zhang2004}, aggregating algorithm~\cite{vovk2001} and the second
order perceptron~\cite{cesa-bianchi2005}.  In~\cite{cesa-bianchi2006}, an
in-deph analysis of online learning is provided.

In this paper, we propose an online formulation of the VECM called Online VECM
(OVECM). OVECM is a lighter version of VECM which considers only a sliding window
of the most recent data and introduces matrix optimizations in order to reduce
the number of operations and therefore execution times. OVECM also takes into
account the fact that cointegration vector space doesn't experience large
changes with small changes in the input data.

OVECM is later compared against VECM and ARIMA models using four currency rates
from the foreign exchange market with 1-minute frequency. VECM and ARIMA models
were used in an iterative way in order to allow fair comparison. Execution times
and forecast performance measures MAPE, MAE and RMSE were used to compare all
methods.

Model effectiveness is focused on out-of-sample forecast rather than in-sample
fitting. This criteria allows the OVECM prediction capability to be expressed
rather than just explaining data history.

The next sections are organized as follows: section~\ref{sec:background}
presents the VAR and VECM, the OVECM algorithm proposed is presented in
section~\ref{sec:methodology}. Section~\ref{sec:results} gives a description of
the data used and the tests carried on to show accuracy and time comparison of
our proposal against the traditional VECM and
section~\ref{sec:conclusions} includes conclusions and a proposal for future study.

### 2016-03-22 articles-conferences/icpram2015/src/results.tex
agregando repo icpram

--- 2015-03-11 src/5_1_Proposal.tex
@@
-\chapter{Online VECM proposal}
-Financial time series are known for their non-stationary behaviour. However,
-sometimes they exhibit some stationary linear combinations. When this happens,
-it is said that those time series are cointegrated. The Vector Error Correction
-Model (VECM) is an econometric model which characterises the joint dynamic
-behaviour of a set of cointegrated variables in terms of forces pulling towards
-equilibrium. In this study, we propose an Online VEC model (OVECM) which
-optimises how model parameters are obtained using a sliding window of the most
-recent data. Our proposal also takes advantage of the long-run relationship
-between the time series in order to obtain improved execution times. Our
-proposed method is tested using four foreign exchange rates with a frequency of
-1-minute, all related to the USD currency base. OVECM is compared with VECM and
-ARIMA models in terms of forecasting accuracy and execution times. We show that
-OVECM outperforms ARIMA forecasting and enables execution time to
-be reduced considerably while maintaining good accuracy levels compared with
-VECM.
-
-\vspace{0.5cm}
-
-\section{The problem}
-In finance, it is common to find variables with a long-run and/or a short-run
-equilibrium relationship. This relationship is called cointegration and it
-reflects the idea of that some set of variables cannot wander too far away from
-each other.
-
-Cointegration means that one or more combinations of these variables is
-stationary even though individually they are not.
-
-Some financial models, such as  vector error correction model (VECM), take advantage of
-this property and describe the joint behaviour of several cointegrated
-financial instruments.
-
-VECM is a special case of the vector autoregressive model (VAR). VAR is a linear
-model which express future values in terms of its own history and other
-variables past values. If cointegration exists, VECM allows to introduce an
-error correction term due to its cointegration estimation error.
-
-VECM as well as VAR model parameters are obtained using ordinary least squares
-method (OLS). Since OLS involves many calculations, the parameter estimation
-method is computationally expensive when the number of past values and
-observations considered increases. Moreover, OLS is an ill-posed problem which
-admits an infinite number of solutions.
-
-Ridge regression method (RR) tackles this ill-posed problem and is usually
-formulated instead of OLS. RR includes a regularisation parameter that leads to
-better generalisation capability. However, RR is still computationally
-expensive dealing with large datasets. Recently, online learning algorithms
-have been proposed to solve problems with large data sets because of their
-simplicity and their ability of updating the model when new data is available.
-
-We propose an online formulation of the VEC model based on the
-AAR method considering only a sliding window of the most recent data. The
-algorithm introduces matrix optimisations in order to reduce the number of
-operations. Our method is later tested with financial data from the foreign
-exchange market.
-
-\section{Methodology} \label{sec:methodology}
-
-Since VECM parameter estimation is computationally expensive, we propose an
-online version of VECM (OVECM).  OVECM considers only a sliding window of the
-most recent data. Moreover, since cointegration vectors represent long-run
-relationships which vary little in time, OVECM determines firstly if they require calculation.
-
-OVECM also implements matrix optimisations in order to reduce execution time,
-such as updating matrices with new data, removing old data and introducing new
-cointegration vectors.
-
-Algorithm~\ref{alg:proposal} shows our OVECM proposal which considers the
-following:
-
-\begin{itemize}
-\item The function \texttt{getJohansen} returns cointegration vectors given by
-the Johansen method considering the trace statistic test at 95\%
-significance level.
-\item The function \texttt{vecMatrix} returns the matrices~(\ref{eq:vecA})
-and~(\ref{eq:vecY}) that allows VECM to be solved.
-\item The function \texttt{vecMatrixOnline} returns the
-matrices~(\ref{eq:vecA}) and~(\ref{eq:vecY}) aggregating new data and removing
-the old one, avoiding calculation of the matrix $\mathbf{A}$ from scratch.
-\item Out-of-sample outputs are saved in the variables 
-$\mathbf{Y}_{\text{true}}$ and $\mathbf{Y}_{\text{pred}}$.
-\item The model is solved using OLS.
-\item In-sample outputs are saved in the variables $\Delta
-\mathbf{y}_{\text{true}}$ and $\Delta \mathbf{y}_{\text{pred}}$.
-\item The function \texttt{mape} gets the in-sample MAPE for the $l$ time
-series.
-\item Cointegration vectors are obtained at the beginning and when they are required to be updated. This updating is decided based on the in-sample MAPE of the last $n$ inputs. The average of all
-MAPEs are stored in the variable $e$. If the average of MAPEs
-($\texttt{mean}(e)$) is above a certain error given by the mean\_error threshold, cointegration vectors are updated.
-\item If new cointegration vectors are required, the function
-\texttt{vecMatrixUpdate} only updates the corresponding columns of matrix
-$\mathbf{A}$ affected by those vectors (see equation~\ref{eq:vecA}).
-\end{itemize}
-
-\begin{algorithm}[ht]
-\begin{algorithmic}[1]
-\REQUIRE $\,$ \\
-$\mathbf{y}$: matrix with $N$ input vectors and $l$ time series\\
-$p$: number of past values \\
-$L$: sliding window size ($L<N$) \\
-$\text{mean\_error}$: MAPE threshold \\
-$n$: interpolation points to obtain MAPE \\
-\ENSURE  $\,$ \\
-$\{\mathbf{y}_{\text{pred}}[L+1],\dots,\mathbf{y}_{\text{pred}}[N]\}$: model predictions 
-\FOR { $i =0$ to $N-L$ }
-    \STATE $\mathbf{y}_i \gets \mathbf{y}[i:i+L]$
-	\IF {i = 0 }
-	    \STATE{$v \gets \texttt{getJohansen}(\mathbf{y}_i,p)$}
-	    \STATE{$[\mathbf{A} \quad \mathbf{Y}] \gets
-        \texttt{vecMatrix}(\mathbf{y}_i,p,v)$}
-	\ELSE
-	    \STATE{$[\mathbf{A} \quad \mathbf{Y}] \gets
-        \texttt{vecMatrixOnline}(\mathbf{y}_i,p,v,\mathbf{A},\mathbf{Y})$}
-        \STATE $\Delta \mathbf{Y}_{\text{pred}}[i] \gets \mathbf{A}[-1,:] \times \mathbf{X}$
-    \ENDIF
-    \STATE $\mathbf{X} \gets \texttt{OLS} (\mathbf{A},\mathbf{Y})$
-    \STATE $e \gets \texttt{mape}(\mathbf{Y}[-n,:],\mathbf{A}[-n,:] \times \mathbf{X})$
-    \IF {$\texttt{mean}(e) > \text{mean\_error}$}
-	    \STATE{$v \gets \texttt{getJohansen}(\mathbf{y}_i,p)$}
-	    \STATE{$\mathbf{A} \gets
-        \texttt{vecMatrixUpdate}(\mathbf{y}_i,p,v,\mathbf{A})$}
-        \STATE $\mathbf{X} \gets \texttt{OLS} (\mathbf{A},\mathbf{Y})$
-    \ENDIF
-\ENDFOR
-\STATE $\mathbf{Y}_{\text{true}} \gets \mathbf{Y}[L+1:N]$
-\STATE $\mathbf{Y}_{\text{pred}}\gets 
-\mathbf{Y}[L:N-1] +\Delta \mathbf{Y}_{\text{pred}}$
-\end{algorithmic}
-\caption{OVECM: Online VECM}
-\label{alg:proposal}
-\end{algorithm}
-
-A pseudocode of the algorithm \ref{alg:proposal} is detailed in the appendix \ref{fig:pseudocodigo}.
-
-Our proposal was compared against VECM and ARIMA. Both algorithms were adapted
-to an online context in order to get a reasonable comparison with our proposal
-(see algorithms \ref{alg:SLVECM} and \ref{alg:SLARIMA}). VECM and ARIMA are
-called with a sliding window of the most recent data, whereby the models are
-updated at every time step.
-
-\begin{algorithm}[ht]
-\begin{algorithmic}[1]
-\REQUIRE $\,$ \\
-$\mathbf{y}$: matrix with $N$ input vectors and $l$ time series\\
-$p$: number of past values \\
-$L$: sliding window size ($L<N$) \\
-\ENSURE  $\,$ \\
-$\{ \mathbf{y}_{\text{pred}}[L+1],\dots,\mathbf{y}_{\text{pred}}[N]\}$: model predictions 
-\FOR { $i =0$ to $N-L$ }
-    \STATE $\mathbf{y}_i \gets \mathbf{y}[i:i+L+1]$
-        \STATE $model = VECM(\mathbf{y}_i, p)$
-        \STATE $\mathbf{Y}_{\text{pred}}[i] = model.predict(\mathbf{y}[i+L])$
-\ENDFOR
-\STATE $\mathbf{Y}_{\text{true}} = \mathbf{y}[i+L+1:N] $
-\end{algorithmic}
-\caption{SLVECM: Sliding window VECM}
-\label{alg:SLVECM}
-\end{algorithm}
-
-Since we know out time series are I(1) SLARIMA is called with $d=1$. ARIMA is
-executed for every time series.
-
-\begin{algorithm}[ht]
-\begin{algorithmic}[1]
-\REQUIRE $\,$ \\
-$\mathbf{y}$: matrix with $N$ input vectors and $l$ time series\\
-$p$: autoregressive order \\
-$d$: integrated order\\
-$q$: moving average order\\
-$L$: sliding window size ($L<N$) \\
-\ENSURE  $\,$ \\
-$\{ \mathbf{y}_{\text{pred}}[L+1],\dots,\mathbf{y}_{\text{pred}}[N]\}$: model predictions 
-\FOR { $i =0$ to $N-L$ }
-\FOR { $j =0$ to $l-1$ }
-    \STATE $\mathbf{y}_i \gets \mathbf{y}[i:i+L+1,j]$
-        \STATE $model = ARIMA(\mathbf{y}_i, (p,d,q))$
-        \STATE $\mathbf{Y}_{\text{pred}}[i,j] = model.predict(\mathbf{y}[i+L,j])$
-\ENDFOR
-\ENDFOR
-\STATE $\mathbf{Y}_{\text{true}} = \mathbf{y}[i+L+1:N,:] $
-\end{algorithmic}
-\caption{SLARIMA: Sliding window ARIMA}
-\label{alg:SLARIMA}
-\end{algorithm}
-
-Both OVECM and SLVECM time complexity is dominated by Johansen method which is
-$O(n^3)$. Thus, both algorithms order is $O(Cn^3)$ where $C$ is the number of
-iterations.
-
-\section{Results} \label{sec:results}
-
+\section{\uppercase{Results}}
+\label{sec:results}
 \subsection{Data} \label{sec:unitroot}
@@
 ARIMA parameters were also obtained using AIC.
-Parameters optimisation is presented in table~\ref{tab:params}:
+Parameters optimization is presented in table~\ref{tab:params}:
 
 \begin{table}[ht]
-\caption{Parameters optimisation. VECM order and ARIMA parameters were selected
+\caption{Parameters optimization. VECM order and ARIMA parameters were selected
 using AIC.}
@@
  \end{figure}
-
-\section{Conclusions} \label{sec:conclusions}
-
-A new online vector error correction method was presented. 
-We have shown that our proposed OVECM considerably reduces execution times
-without compromising solution accuracy.  OVECM was compared with VECM and ARIMA
-with the same sliding window sizes and OVECM outperformed both in terms of
-execution time. Traditional VECM slightly outperformed our proposal but the
-OVECM execution time is lower.
-This reduction of execution time is mainly because OVECM avoids the cointegration vector
-calculation using the Johansen method. The condition for getting new vectors is
-given by the mean\_error variable which controls how many times the Johansen method is
-called. Additionally, OVECM introduces matrix optimization in order to get the
-new model in an iterative way.
-We could see that our algorithm took much less than a minute at every
-step. This means that it could also be used with higher frequency data and would
-still provide responses before new data arrives.
-
-For future study, it would be interesting to improve the out-of-sample forecast
-by considering more explicative variables, to increase window sizes or trying new
-conditions to obtain new cointegration vectors.
-
-Since OVECM is an online algorithm which optimizes processing time, it could be
-used by investors as an input for strategy robots. Moreover, some technical
-analysis methods could be based on its output.

### 2016-03-22 src/5_3_Proposal.tex
agregando repo icpram

@@
-\chapter{Real-time intraday Forex rates forecasting using a regularized Vector
-Error Correction model}
+\chapter{Intraday Forex rates forecasting using an online cointegration approach}
 
+High frequency forex rates 
 Vector error correction model (VECM) parameters are commonly obtained using the

### 2016-03-22 articles-conferences/icpram2015/code/python/classes.py

@@ -1,0 +1,621 @@
+from __future__ import division
+
+import numpy as np
+import math as ma
+import pandas as pd
+import matplotlib.pyplot as plt
+
+from datetime import datetime, date, time
+
+import traceback
+import sys
+import random
+import time
+
+from statsmodels.tsa.johansen import coint_johansen
+from statsmodels.tsa.stattools import adfuller
+import glob
+
+BETA = 1.0 / 9.0
+
+ROWS = 3000
+COLS = 1600
+RANK = 632
+R = 1
+P = 3
+FILE = 'data.csv'
+size = 1
+
+
+class Reader:
+
+    def __init__(self, filename=FILE):
+        """ data Constructor
+
+        Parameters
+        ----------
+        filename: string
+                  Data path
+        """
+
+        self.filename = filename
+
+    def load_data(self, tickers, rtickers=[]):
+        """ Load data file setting time
+        column as the pandas index.
+
+        Parameters
+        ----------
+        filename: string
+                  Data path
+        tickers: string list
+                 List of currencies
+        rtickers: string list
+                 List of reciprocal currencies
+
+        Returns
+        -------
+        Pandas Dataframe object
+        """
+
+        data = pd.read_csv(self.filename)                 # Reading csv file
+        data.index = pd.to_datetime(data.pop('Time'))     # Indexed by time
+        # Selected currencies
+        data = data[tickers + rtickers]
+
+        for ticker in rtickers:
+            data[ticker] = 1 / data[ticker]
+            data = data.rename(columns={ticker: ticker[3:6] + ticker[0:3]})
+
+        # Getting all ticker names
+        self.alltickers = data.columns.values.tolist()
+
+        return data
+
+    def load_dukascopy(self):
+
+        ask_list = glob.glob('*Ask*.dat')
+        bid_list = glob.glob('*Bid*.dat')
+        ask_list.sort()
+        bid_list.sort()
+
+        file_lists = ask_list[1:]
+        prices = pd.read_csv(ask_list[0])
+        prices.columns = ['Time', ask_list[0][0:6]]
+
+        for filename in file_lists:
+            DF = pd.read_csv(filename)
+            DF.columns = ['Time', filename[0:6]]
+            prices = pd.merge(prices, DF, on='Time', how='outer')
+
+        prices.index = pd.to_datetime(prices.pop('Time'),
+                                      format='%d-%m-%Y %H:%M:%S')
+        prices.to_csv('data_ask_all.csv')
+
+        # The same has to be done using bid_list
+
+
[…]

### 2016-03-22 articles-conferences/icpram2015/code/python/main.py

@@ -1,0 +1,202 @@
+import classes as cl
+import numpy as np
+import time
+import sys
+from statsmodels.tsa.stattools import adfuller
+
+try:
+    from line_profiler import LineProfiler
+
+    def do_profile(follow=[]):
+        def inner(func):
+            def profiled_func(*args, **kwargs):
+                try:
+                    profiler = LineProfiler()
+                    profiler.add_function(func)
+                    for f in follow:
+                        profiler.add_function(f)
+                    profiler.enable_by_count()
+                    return func(*args, **kwargs)
+                finally:
+                    profiler.print_stats()
+            return profiled_func
+        return inner
+
+except ImportError:
+    def do_profile(follow=[]):
+        "Helpful if you accidentally leave in production!"
+        def inner(func):
+            def nothing(*args, **kwargs):
+                return func(*args, **kwargs)
+            return nothing
+        return inner
+
+
+def SLECM(y, L, P):
+    # SLECM Test
+    m = cl.Matrix()
+    it = y.shape[0] - L
+
+    for i in range(it):
+        y_i = y[i:i + L]
+        beta = m.get_johansen(y_i.as_matrix(), P)
+
+        A, B, beta = m.vec_matrix(y_i, P, beta.evecr)
+        x, residuals, rank, s = np.linalg.lstsq(A, B)
+
+
+def SLAAR_1(y, L, P, update):
+    # SLAAR Test
+    m = cl.Matrix()
+    #it = y.shape[0] - L
+    it = 100
+
+    for i in range(it):
+        sys.stdout.write("\r{0}%".format((float(i + 1) / it) * 100))
+        sys.stdout.flush()
+
+        y_i = y[i:i + L]
+
+        if(i % update == 0):
+            beta = m.get_johansen(y_i.as_matrix(), P)
+            A, B = m.vec_matrix(y_i, P, beta.evecr)
+            #lambda_svd = m.lambda_svd(A, np.linalg.matrix_rank(A))
+            lambda_qr = m.lambda_qr(A)
+        else:
+            A, B = m.vec_matrix(y_i, P, beta.evecr)
+            pass
+
+        #x = m.iteration(A, B, 3, 2, lambda_svd, 0.00025)
+        x, residuals, rank, s = np.linalg.lstsq(A, B)
+        x = m.iteration(A, B, 3, 2, lambda_qr)
+
+
+def SLAAR_2(y, L, P, it, avg_error, r, n):
[…]

### 2016-03-22 articles-conferences/iscef2016/ISCEF2016-slides/iscef2016-slides.tex
agregando repo iscef2016

%\documentclass[xcolor=dvipsnames]{beamer}
\documentclass[compress,red]{beamer}
\mode<presentation>
%\usepackage[spanish,activeacute]{babel}
\usepackage[utf8]{inputenc}
\usepackage{listings}
\usepackage{amsmath,amsfonts,amssymb}
\usepackage{adjustbox}
\usepackage{color}
\usepackage{tikz}
\usetikzlibrary{shapes,arrows}
\usepackage{multicol}
\usepackage{hhline}
\usepackage{pgf}
\usepackage{algorithm} 
\usepackage{algorithmic}
\renewcommand{\algorithmicrequire}{\textbf{Input:}}
\renewcommand{\algorithmicensure}{\textbf{Output:}}
\def\R{{\mathbb R }}
\def\N{{\mathbb N }}
\usetheme{Warsaw}
% other themes: AnnArbor, Antibes, Bergen, Berkeley, Berlin, Boadilla, boxes, CambridgeUS, Copenhagen, Darmstadt, default, Dresden, Frankfurt, Goettingen,
% Hannover, Ilmenau, JuanLesPins, Luebeck, Madrid, Maloe, Marburg, Montpellier, PaloAlto, Pittsburg, Rochester, Singapore, Szeged, classic
%\usecolortheme{lily}
% color themes: albatross, beaver, beetle, crane, default, dolphin, dov, fly, lily, orchid, rose, seagull, seahorse, sidebartab, structure, whale, wolverine

%define your own colours:
\definecolor{Red}{rgb}{1,0,0}
\definecolor{Blue}{rgb}{0,0,1}
\definecolor{Green}{rgb}{0,1,0}
\definecolor{magenta}{rgb}{1,0,.6}
\definecolor{lightblue}{rgb}{0,.5,1}
\definecolor{lightpurple}{rgb}{.6,.4,1}
\definecolor{gold}{rgb}{.6,.5,0}
\definecolor{orange}{rgb}{1,0.4,0}
\definecolor{hotpink}{rgb}{1,0,0.5}
\definecolor{newcolor2}{rgb}{.5,.3,.5}
\definecolor{newcolor}{rgb}{0,.3,1}
\definecolor{newcolor3}{rgb}{1,0,.35}
\definecolor{darkgreen1}{rgb}{0, .35, 0}
\definecolor{darkgreen}{rgb}{0, .6, 0}
\definecolor{darkred}{rgb}{.75,0,0}

\xdefinecolor{olive}{cmyk}{0.64,0,0.95,0.4}
\xdefinecolor{purpleish}{cmyk}{0.75,0.75,0,0}

\useoutertheme[subsection=false]{smoothbars}

\setbeamertemplate{blocks}[rounded][shadow=true]

\author[P. Arce et al.]{Paola
Arce\inst{1}\and Jonathan Antognini\inst{1}\and Werner Kristjanpoller
\inst{2}\and Luis Salinas\inst{1,3}}

\institute[UTFSM]{
\inst{1} Departamento de Inform\'atica, UTFSM, Chile \\
\inst{2} Departamento de Industrias, UTFSM, Chile \\
\inst{3} CCTVal, UTFSM, Chile
}

\date[ISCEF 2016]{4th International Symposium in Computational Economics and Finance 2016}
\title[Cointegration model
for forecasting HF financial time series]{\Large Fast and adaptive cointegration based model for forecasting high frequency financial time series}

\titlegraphic{
   \includegraphics[width=1cm]{img/logo_DI}\hspace*{2.5cm}~%
   \includegraphics[width=1.5cm]{img/logo_UTFSM}\hspace*{2.5cm}~%
   \includegraphics[width=1cm]{img/logo_IND}
}

\begin{document}
\begin{frame}[plain]
\titlepage
\end{frame}

\begin{frame}
\frametitle{Outline}
\begin{itemize}
\item Motivation: economic relations
\item Cointegration and Integration concepts
\item VECM model
\item Proposal
\item Experiments
\item Conclusions
\end{itemize}
\end{frame}

\section{Introduction}
\subsection{Motivation}

\begin{frame}
\frametitle{Motivation}
Financial time series are mainly non-stationary.  However, one or more linear
combinations of non-stationary time series can be stationary even though
individually they are not.

When this happens it is said they have a long-run
equilibrium relationship and the variables are \textbf{cointegrated}.
\end{frame}

\section{Background}
\subsection{Definitions}

\begin{frame}
\frametitle{Integration }
\begin{block}{Integration of order $d$}
A time series $\mathbf{y}$ is said to be integrated of order $d$ if after
differentiating the variable $d$ times, we get an stationary process, more precisely:
\[
(1-L)^d \mathbf{y} \sim \text{I(0)} \, ,
\]
\noindent where I(0) is a stationary time series and $L$ is the lag operator:
\[
(1-L)\mathbf{y} = \Delta \mathbf{y}=\mathbf{y}_t  -\mathbf{y}_{t-1} \quad \forall t
\]
\end{block}
\end{frame}

\subsection{Integration example}
\begin{frame}
\frametitle{Financial Time Series}
\begin{columns}
\column[t]{5cm}
Financial time series are commonly a I(1) process 
\begin{center}
\includegraphics[width=0.85\textwidth]{img/EURUSD}
\end{center}
\column[t]{5cm}
Since their differences are stationary (I(0) process).
\begin{center}
\includegraphics[width=0.85\textwidth]{img/DEURUSD}
\end{center}
\end{columns}
\end{frame}

\subsection{Cointegration}
\begin{frame}
\frametitle{Cointegration}
\begin{block}{Cointegration definition}
Let $\mathbf{y} = \{\mathbf{y}^1, \dots, \mathbf{y}^l\}$ be a set of $l$
time series of order I(1) which are said to be cointegrated if a vector
$\beta=[\beta(1),\dots,\beta(l)]^\top \in \mathbb{R}^l$  exists such that the
time series,

\begin{equation*}
 \mathbf{Z}_t:= \beta^\top \mathbf{y} = \beta(1) \mathbf{y}^1 + \dots + \beta(l) \mathbf{y}^l \sim
 \text{I(0)}\, .
\end{equation*}
\end{block}
In other words, a set of I(1) variables is said to be cointegrated if
a linear combination of them exists which is I(0).

\end{frame}

\subsection{Vector Autorregresive models}
\begin{frame}
\frametitle{VECM: Vector Error Correction model}

VECM is a special form of a Vector Autorregresive model (VAR) for I(1)
variables $\mathbf{y}$ that are also cointegrated. Let

\begin{equation*}
\label{eq:variables}
\mathbf{y}_t = 
\begin{bmatrix} y_{1,t} & y_{2,t} & \dots & y_{l,t}
\end{bmatrix}^\top \, \qquad t=p+1, \dots N. 
\end{equation*}

\begin{block}{VAR}
VAR($p$) model describes the behaviour of a set of $l$
endogenous variables as a linear combination of their last $p$ values:
{\color{blue}
\begin{equation*}
\label{eq:var}
 \mathbf{y}_t = \phi_1 \mathbf{y}_{t-1}  + \dots +   \phi_p\mathbf{y}_{t-p}
+ \mathbf{c} + \mathbf{\epsilon}_t \, ,
\end{equation*}}
\noindent where  ${\phi_1,\dots,\phi_p}$ are $l \times l$
matrices of real coefficients,
$\mathbf{\epsilon}_{p+1},\dots,\mathbf{\epsilon}_N$ are error terms,
$\mathbf{c}$ is a constant vector and $N$ is the total number of samples.
\end{block}
\end{frame}

\begin{frame}
\frametitle{VECM: Vector Error Correction model}
\begin{block}{VECM}

It is obtained by replacing the form {\color{red} $\Delta
\mathbf{y}_t = \mathbf{y}_t - \mathbf{y}_{t-1}$} in the VAR form.
{\color{blue}
\begin{equation}
 \label{eq:vec}
 \Delta \mathbf{y}_t = 
 \underbrace{ \Omega\mathbf{y}_{t-1}}_\text{Error correction term} + 
 \sum_{i=1}^{p-1}
\phi_i^* \Delta \mathbf{y}_{t-i}  + \mathbf{c} + \mathbf{\epsilon}_t \quad ,
\end{equation}}
\end{block}

\noindent where 
%coefficients matrices $\Omega$ and $\phi_i^*$ are
%function of matrices $\phi_i$ as follows:

\begin{eqnarray*}
\phi_i^* &: =& -\sum_{j=i+1}^{p} \phi_j \\
\Omega &: =& -(\mathbb{I}-\phi_1-\dots-\phi_p) 
\end{eqnarray*}
\end{frame}

\begin{frame}
\frametitle{$\Omega$ matrix properties}
The matrix $\Omega$ has the following properties:
\begin{itemize}
\item If $\Omega = 0$ there is no cointegration
\item If $rank(\Omega)=l$ i.e full rank, then the time series are not
I(1) but stationary
\item If $rank(\Omega)=r,\quad 0 < r < l$ then, there is cointegration
and the matrix $\Omega$ can be expressed as $\Omega =
\alpha \beta^\top$, where $\alpha$ and $\beta$ are $(l \times r)$
matrices and $rank(\alpha)=rank(\beta)=r$.
\end{itemize}
The columns of $\beta$ contains the cointegration vectors and the rows of
$\alpha$ correspond with the adjusted vectors. $\beta$ is obtained by Johansen
procedure whereas $\alpha$ has to be determined as a
variable in the VECM.
\end{frame}

\begin{frame}
\frametitle{VECM matrix form}
\small
\begin{equation*}
\underbrace{
      \begin{bmatrix}
       \mathbf{\Delta y}_{p+1}  \\ 
       \mathbf{\Delta y}_{p+2}  \\ 
       \vdots                   \\ 
       \mathbf{\Delta y}_N      
      \end{bmatrix}}_{\mathbf{B} } =
\arraycolsep=1.4pt  
\underbrace{\begin{bmatrix}
   \alpha^\top & \phi^*_1 & \dots & \phi^*_{p-1} &\mathbf{c}   
  \end{bmatrix}}_{\mathbf{X}^\top}
\underbrace{\begin{bmatrix}
 \beta\mathbf{y}_p      & \beta\mathbf{y}_{p+1}   & \dots&   \beta\mathbf{y}_{N-1}   \\
 \mathbf{\Delta y}_{p}   & \mathbf{\Delta y}_{p+1}& \dots &  \mathbf{\Delta y}_{N-1} \\
 \mathbf{\Delta y}_{p-1} & \mathbf{\Delta y}_{p}  & \dots &  \mathbf{\Delta y}_{N-2}   \\
 \vdots                  & \vdots                 & \ddots&  \vdots                   \\
 \mathbf{\Delta y}_{2}   & \mathbf{\Delta y}_{3} & \dots &   \mathbf{\Delta y}_{N-p+1} \\
 1                      & 1                       & \dots     & 1   
 \end{bmatrix}}_{\mathbf{A}^\top}
+
\underbrace{\begin{bmatrix}
              \mathbf{\epsilon}_{p+1} \\ 
              \mathbf{\epsilon}_{p+2} \\ 
              \vdots \\ 
              \mathbf{\epsilon}_N
             \end{bmatrix}}_{\mathbf{E}^\top}
\end{equation*}
\end{frame}

\section{Proposal}
\subsection{Motivation}
\begin{frame}
\frametitle{VECM difficulties}
\begin{itemize}
\item VECM assumes cointegration vectors do not change in time. However, these might change due to several economic factors.
\item To get updated cointegration vectors in time is computationally expensive because 
the Johansen method and OLS are used to obtain them.
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Cointegration vectors distribution}
\begin{figure}[!h]
  %\vspace{-0.8cm}
  \centering
  \includegraphics[width=0.75\textwidth]{img/histCointVectors-offset20520-p-1-freq-10s}
  \caption{Histogram of the number of cointegration vectors using $p=1$. Four
  possible values for $L$ are shown (100, 2000, 3000 and 4000).}
  \label{fig:hists}
\end{figure}
\end{frame}

\begin{frame}
\frametitle{Percentage of cointegration}
Since $r=0$ means no cointegration and $r=l=4$ reveals that no process is I(1) but stationary.
The interesting cases of cointegration are those where $r$ lies strictly
between $0$ and $l$, i.e. $0<r<l$.
\begin{block}{Percentage of cointegration (PC)}
{\color{blue}
\begin{equation*} \label{eq:pcoint}
PC = 
\frac{\#\{ it \mid \text{$it$ has $r$ c.v. with $0<r<l$}\}}
     {\#it}\times 100
\end{equation*}}
\end{block}
\end{frame}

\begin{frame}
\frametitle{PC vs MSE}
\begin{figure}[ht!]
  %\vspace{-0.8cm}
  \centering
  \includegraphics[width=0.7\textwidth]{img/MSE-offset20520-p-2-freq-10s}
  \caption{MSE versus the percentage of cointegration considering 1000
  iterations. }
  \label{fig:cointvsmse}
\end{figure}
\end{frame}
%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection{AVECM}
\begin{frame}
\frametitle{Proposal: Adaptive VECM algorithm (AVECM)}
\begin{itemize}
\item We proposed to choose $L$ and $p$ in order to maximise the percentage of
cointegration $PC$ in the near past. This process was done every time that new data was processed. 
\item To use a distributed environment to make this parameters search, since this is the most expensive routine.
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Proposal: Adaptive VECM algorithm (AVECM)}
\begin{itemize}
\item We proposed to choose $L$ and $p$ in order to maximise the percentage of
cointegration $PC$ in the near past. This process was done every time that new data was processed. 
\item To use a distributed environment to make this parameters search, since this is the most expensive routine.
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{AVECM Algorithm}
\begin{algorithmic}[1]
\REQUIRE $\,$ \\
$\mathbf{y}$: matrix with $N$ input vectors and $l$ time series\\
$j$: Starting point of testing \\
$it$: Ending point of testing \\
$ps$: list of $p$ values \\
$Ls$: list of $L$ values ($L<N$) \\
$m$: Iterations to determine parameters ($m < N-L$)\\
\ENSURE  $\,$ \\
$\{ \hat{\mathbf{y}}[1],\dots,\hat{\mathbf{y}}[it]\}$: prediction vectors \\
\FOR { $i =j$ to $it$ }
   \STATE $\mathbf{Y} \gets \mathbf{y}[:,i-1]$
    \STATE $L,p \gets
    \texttt{get\_best\_params}(Ls,ps,m,\mathbf{Y})$
        \STATE $model = \text{VECM}(\mathbf{Y},L, p)$
        \STATE $\hat{\mathbf{y}}[i-j] = model.predict()$
\ENDFOR
\end{algorithmic}
\end{frame}

\subsection{Model comparison} 
\begin{frame}
\frametitle{Random walk}
The random walk model is defined as:
\begin{equation}
\mathbf{y}_t = \mathbf{y}_{t-1} + \epsilon_{t}
\label{rwmodel}
\end{equation}
The naive forecast of the time series difference $\hat{\mathbf{y}}_{t+1}$ for the random walk model is defined as:
\begin{equation}
\hat{\mathbf{y}}_{t+1} = \mathbf{y}_t + \hat{\epsilon}_{t+1} 
\end{equation}
\noindent where  $\hat{\epsilon}_{t+1} = \epsilon_{t}$.
\end{frame}

\begin{frame}
\frametitle{ARIMA}
A process can be modelled as an ARIMA$(p,d,q)$ if $\mathbf{x}_t = \Delta^d \mathbf{y}_t $, is an ARMA$(p,q)$. An ARMA$(p,q)$ model is the following:
\begin{equation}
\mathbf{x}_t = \sum_{i=1}^p \phi_i \mathbf{x}_{t-i}  +  \sum_{j=1}^q \theta_j \epsilon_{t-j}  
\end{equation}
\noindent with coefficients $\phi_p \neq 0$, $\theta_q \neq 0$ and $\sigma_{\epsilon}^2 > 0$.
\end{frame}

\begin{frame}
\frametitle{Evaluation methods}
\begin{block}{MSE}
Measures the distance between forecasts
and the true values and large deviations from the true value have a
large impact due to squaring forecast error.
{\color{blue}
\begin{equation}\label{eq:MSE}
\text{MSE} = 
\frac{\displaystyle \sum_{t=1}^{N} (\mathbf{y}_t-\hat{\mathbf{y}}_t)^2}{N}
\end{equation}}
\end{block}
\begin{block}{\bf $U$-statistic}
Unit free measure obtained as the ratio between the root MSE (RMSE) of a model and the RMSE of the naive random walk model. 
\end{block}
\end{frame}

\section{Experiments}
\subsection{Uno}
\begin{frame}
\frametitle{Data}
AVECM tests were carried out using four foreign exchange rates all related to
USD: EURUSD, GBPUSD, USDCHF and USDJPY. This data was collected from Dukascopy, a free
database which gives access to the Swiss Foreign Exchange marketplace.
The tests were done using 10-seconds frequency from ask prices
from the 11th to the 15th of August 2014. Since one day corresponds to 8640 data points and we used 5 days of data, we have 43,200 data points in total.
\end{frame}

\begin{frame}
\frametitle{Unit root tests}
VECM can only be used if the time series are I(1). Therefore, 
before running the tests, we firstly checked whether the time series were
I(1) using the Augmented Dickey Fuller (ADF) test with lags $p=1,2,3,4,5$.
\end{frame}

\begin{frame}
\frametitle{Performance accuracy}
\small
\begin{tabular}{ccccccc}
& \multicolumn{3}{l}{MSE} & &
\multicolumn{2}{c}{$U$-statistic} \\ 
\hhline{~---~--} \\
 & AVECM & ARIMA & p-value & &
AVECM & ARIMA \\

\hline
 EURUSD & 1.0702 e-09 & 1.1481 e-09 &  9.250 e-12 & & 0.6863 & 0.7108\\
 GBPUSD & 1.6630 e-09 & 1.7408 e-09 &  6.951 e-02 & & 0.6866 & 0.7025\\
 USDCHF & 5.8503 e-10 & 6.3545 e-10 &  2.899 e-14 & & 0.6803 & 0.7091\\
 USDJPY  & 6.3483 e-06 & 6.5194 e-06 &  6.853 e-05 & & 0.6964 & 0.7057
 \end{tabular}
 \end{frame}
%For AVECM we considered different number of iterations (parameter $m$ in algorithm \ref{alg:AVECM}): 10, 50 and 100. We tried 12, 24 and 47 different pair of combinations for $L$ and $p$. Possible values for $L$ were in the interval $[100,4000]$ and $p$ can have values in between $[1,5]$. Best AVECM performance was compared against ARIMA and the random walk model.
%Table~\ref{tab:stats} shows the out-of-sample performance measures: MSE and $U$-statistic for AVECM and ARIMA. In terms of both measures we found that AVECM is superior to ARIMA and the naive random walk model. We also included the p-value that proves that the difference in the MSE is significant at the 99\% significance level in three of the four currency rates and at 90\% in the case of GBPUSD. The $U$-statistic shows that AVECM and ARIMA are superior to the naive random walk model and that our proposal is also superior to ARIMA.

\begin{frame}
\frametitle{Parallel implementation}
The function get\_best\_params($Ls,ps,m,\mathbf{Y}$) was implemented using MPI in Python. MPI allows large-scale parallel applications
with wide portability to be built, being able to run in large clusters
or on local computers.
We tested our proposal in a cluster with 2 servers Xeon E5-2667 (2.90GHz)
of 24 cores each (48 cores in total) and 24GB RAM.
\end{frame}

\begin{frame}
\frametitle{Parameters setting}
\begin{itemize}
\item Parameter $it=100$
\item The $L$ parameter was always chosen between 100 and 4000 and $p$ always took values
between 1 and 5. 
\item Parameter $nparams$ represents the number of pairs ($L$,$p$) used to maximise the percentage of cointegration. 
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Execution times}
\begin{figure}[ht]
  \centering
  \includegraphics[width=0.75\textwidth]{img/extimes}
  \caption{Computing time of sequential and parallel algorithm is shown in the
  upper figure. Speed-up is shown below.}
  \label{fig:extimes}
% The superior figure \ref{fig:extimes} shows that best performance accuracy measurements are
%achieved in times near or below 10 seconds in the parallel version. Contrarily,
%serial times are high, above 10 seconds in most cases. Figure~\ref{fig:extimes} also shows the speed-up in computing time for AVECM which is near 9X if we use more than 25 parameters ($nparams$).
%Execution times do not consider the loading data time, just that of finding
%best parameters and matrix operations. The time MPI spends
%transferring data and synchronising processes is about two seconds independently
%of the number of processes considered. Execution times were measured using the Time python library.
\end{figure}

\end{frame}

\section{Conclusions}
\subsection{Uno}
\begin{frame}
\frametitle{Conclusions}
\begin{itemize}
\item Cointegration relations change with time. We empirically showed that the Johansen method is sensitive to the number of lags but also to the amount of data considered.
\item We found that out-of-sample forecast performance MSE is related to the {\em percentage of cointegration\/} and in the last samples.  
\item VECM improves performance measures by finding
parameters of $L$ and $p$ maximising the percentage of cointegration.
\item Increasing the number of parameters will always lead to
better performance measures.
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Conclusions}
\begin{itemize}
\item The parallel implementation allowed the execution times to be reduced
more than 9 times and therefore a response time was obtain before 10
seconds. 
\item For future study, it would be interesting to explore the relationship between
cointegration and performance in order to propose new criteria for
improving VECM parameters. It would also be interesting to include more 
explaining variables such as bid-ask spread and change in volume.
\end{itemize}
\end{frame}

\begin{frame}[plain,c]
\begin{center}
\Huge Thank you for your attention
\Huge Questions?
\end{center}
\end{frame}

\end{document}

### 2016-03-22 articles-conferences/iscef2016/main.tex
agregando repo iscef2016

\RequirePackage{fix-cm}
%
%\documentclass{svjour3}                     % onecolumn (standard format)
%\documentclass[smallcondensed]{svjour3}     % onecolumn (ditto)
\documentclass[smallextended]{svjour3}       % onecolumn (second format)
%\documentclass[twocolumn]{svjour3}          % twocolumn
%
\smartqed  % flush right qed marks, e.g. at end of proof

\usepackage{graphicx}
%
% \usepackage{mathptmx}      % use Times fonts if available on your TeX system
%
% insert here the call for the packages your document requires
%\usepackage{latexsym}
% etc.
%
\usepackage{amsfonts,amssymb,amsmath,
            amssymb,amsopn,amsxtra,amstext,amsbsy}
\usepackage{epsfig}
\usepackage[makeroom]{cancel}
\usepackage{array,arydshln}
\usepackage{multirow,varwidth}
\usepackage{hhline}
\usepackage{url}
\usepackage[utf8x]{inputenc}
\usepackage{fancyhdr}
\usepackage{hyperref,wasysym}
\usepackage{color}
\usepackage{enumerate}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{setspace}
\usepackage{booktabs}
\let\Algorithm\algorithm
\renewcommand\algorithm[1][]{\Algorithm[#1]\setstretch{1.1}}

\renewcommand{\algorithmicrequire}{\textbf{Input:}}
\renewcommand{\algorithmicensure}{\textbf{Output:}}

\usepackage{xcolor}
\usepackage{hyperref}
\usepackage{pmat}

\journalname{Computational Economics}

\begin{document}

\title{Fast and adaptive cointegration based model
for forecasting high frequency financial time series}
%\subtitle{Do you have a subtitle?\\ If so, write it here}

\titlerunning{Cointegration based model
for forecasting high frequency financial time series}        % if too long for running head

%\author{Paola Arce  \and Jonathan Antognini \and Werner Kristjanpoller \and Luis Salinas }
\author{}
%\authorrunning{Short form of author list} % if too long for running head

%\institute{P. Arce  \at
%              Departamento de Inform\'atica, Universidad T\'ecnica Federico Santa
%              Mar\'ia\\
%              Tel.: +56-32-2654000\\
%              \email{paola.arce@usm.cl}           %  \\
%%             \emph{Present address:} of F. Author  %  if needed
%           \and
%           J. Antognini \at
%              Departamento de Inform\'atica, Universidad T\'ecnica Federico Santa
%              Mar\'ia\\
%              \email{jantognini@csrg.cl}             \\
%           \and
%           W. Kristjanpoller \at
%              Departamento de Industrias, Universidad T\'ecnica Federico Santa
%              Mar\'ia\\
%              %Fax: +123-45-678910\\
%              \email{werner.kristjanpoller@usm.cl}             \\
%           \and
%           L. Salinas \at
%              Departamento de Inform\'atica, Universidad T\'ecnica Federico Santa
%              Mar\'ia\\
%              Centro Cient\'ifico Tecnol\'ogico de Valpara\'iso (CCTVal) \\
%              \email{luis.salinas@usm.cl}           %  \\
%}

\date{Received: date / Accepted: date}
% The correct dates will be entered by the editor

\maketitle

\begin{abstract}

Cointegration is a long-run property of some non-stationary time series where a 
linear combination of those time series is stationary.
This behaviour has been studied in finance because cointegration restrictions often improve forecasting.
The Vector Error Correction Model (VECM) is a well-known econometric technique that
characterises short-run variations of a set of cointegrated time series incorporating long-run relationships as an error correction term. VECM has been broadly used with low frequency time series. We aimed to adapt VECM to be used in finance with high frequency stream data.
%

Cointegration relations change in time and therefore VECM parameters must be updated when new data is available. We studied how forecasting performance is affected when VECM parameters and the length of historical data used change in time. We observed that the number of cointegration relationships varies with the length of historical data used. Moreover, parameters that increased these relationships in time led to better forecasting performance. Our proposal, called an Adaptive VECM (AVECM) is 
to make a parameters grid search that maximises the number of cointegration relationships in the near past. To ensure the search can be executed fast enough, we used a distributed environment.

%
The methodology was tested using four 10-seconds frequency time series of the Foreign Exchange market. We compared our proposal with ARIMA and the naive forecast of the random walk model. Numerical experiments showed that on average AVECM performed better than ARIMA and random walk. Additionally, our parallel approach AVECM significantly improved execution times with respect to its serial version.

\keywords{VECM \and Cointegration \and Forex \and MPI \and Parallel algorithm}
% \PACS{PACS code1 \and PACS code2 \and more}
% \subclass{MSC code1 \and MSC code2 \and more}
\end{abstract}

\input{src/intro}
\input{src/background}
\input{src/methodology}
\input{src/results}
\input{src/conclusions}

\begin{acknowledgements}
L. Salinas acknowledges support from Basal Project
FB0821 CCTVal-Centro Científico Tecnológico de Valparaíso. P. Arce and W.
Kristjanpoller thank PIIC 2014 DGIP-UTFSM grant for supporting this research.
\end{acknowledgements}

% BibTeX users please use one of
%\bibliographystyle{spbasic}      % basic style, author-year citations
%\bibliographystyle{spmpsci}      % mathematics and physical sciences
%\bibliographystyle{spphys}       % APS-like style for physics

\bibliographystyle{plain}       % APS-like style for physics

\bibliography{reference}   % name your BibTeX data base

\end{document}
% end of file template.tex

### 2016-03-22 articles-conferences/iscef2016/src/background.tex
agregando repo iscef2016

--- 2015-06-23 manuscript/manuscript.tex
@@
-\RequirePackage{fix-cm}
-%
-%\documentclass{svjour3}                     % onecolumn (standard format)
-%\documentclass[smallcondensed]{svjour3}     % onecolumn (ditto)
-%\documentclass[smallextended]{svjour3}       % onecolumn (second format)
-\documentclass[twocolumn]{svjour3}          % twocolumn
-\smartqed  % flush right qed marks, e.g. at end of proof
-\usepackage{graphicx}
-\usepackage{amsfonts,amssymb,amsmath,
-            amssymb,amsopn,amsxtra,amstext,amsbsy}
-\usepackage{epsfig}
-\usepackage[makeroom]{cancel}
-\usepackage{array,arydshln}
-\usepackage{array}
-\usepackage{url}
-%\usepackage{tikz}
-\usepackage[utf8x]{inputenc}
-\usepackage{fancyhdr}
-\usepackage{hyperref,wasysym}
-\usepackage{color}
-\usepackage{enumerate}
-\usepackage{algorithm}
-\usepackage{algorithmic}
-\renewcommand{\algorithmicrequire}{\textbf{Input:}}
-\renewcommand{\algorithmicensure}{\textbf{Output:}}
-\usepackage{multirow}
-%\usepackage[spanish]{babel}
+\def\rot#1{{\color{red}{#1}}}
+\def\gruen#1{{\color{green}{#1}}}
+\def\blau#1{{\color{blue}{#1}}}
 
-\usepackage{xcolor}
-\usepackage{hyperref}
-\usepackage{pmat}
+\section{\uppercase{Background}}
+\label{sec:background}
 
-\journalname{Applied Intelligence}
-
-\begin{document}
-
-\title{High frequency intraday forex rates forecasting using an online cointegration approach}
-%\subtitle{Do you have a subtitle?\\ If so, write it here}
-
-%\titlerunning{Short form of title}        % if too long for running head
-
-\author{Paola Arce \and Werner Kristjanpoller \and Luis Salinas }
-
-%\authorrunning{Short form of author list} % if too long for running head
-
-\institute{P. Arce  \at
-              Departamento de Inform\'atica, Universidad T\'ecnica Federico Santa
-              Mar\'ia\\
-              Tel.: +56-32-2654000\\
-              %Fax: +123-45-678910\\
-              \email{paola.arce@usm.cl}           %  \\
-%             \emph{Present address:} of F. Author  %  if needed
-           \and
-           W. Kristjanpoller \at
-              Departamento de Industrias, Universidad T\'ecnica Federico Santa
-              Mar\'ia\\
-              %Fax: +123-45-678910\\
-              \email{werner.kristjanpoller@usm.cl}             \\
-           \and
-           L. Salinas \at
-              Departamento de Inform\'atica, Universidad T\'ecnica Federico Santa
-              Mar\'ia\\
-              Centro Científico Tecnológico de Valparaíso (CCTVal) \\
-              \email{luis.salinas@usm.cl}           %  \\
-}
-
-\date{Received: date / Accepted: date}
-% The correct dates will be entered by the editor
-
-\maketitle
-
-\begin{abstract}
-
-This research examines the cointegration relation between three mayor currency pairs (MCP) of the foreign exchange markets:  EURUSD, GBPUSD, USDCHF.
-Specifically, we used a modified version of the Vector error correction model (VECM) which parameters are updated online using a limited amount of historical data.
-VECM parameters are commonly obtained using the
-ordinary least squares (OLS) method.
-Our proposal is to solve VECM using Ridge regression (RR) and the Aggregating algorithm for regression (AAR) which could lead to a better generalisation capability rather than OLS.
-Both approaches together with OLS are compared against an optimal offline algorithm which is able to see the sequences in advance.
-Our experiments were carried out using 10-seconds frequency Forex data and shows that RR and AAR approaches improve cumulative loss of OLS approach with respect to the optimal offline algorithm.
-
-\keywords{VECM \and Cointegration \and Forex \and High frequency data \and Online algorithm}
-
-\end{abstract}
-
-\section{Introduction}
-\label{sec:introduction}
-
-The relationships between financial time series are commonly studied. In particular, in some of them
-it is possible to detect long-run relationships through to a cointegration analysis.
-Two or more non-stationary time series are said to be cointegrated if there exists a linear combination of them which is stationary.
-
-The vector error correction model (VECM) introduces the long-run relationship among a set of cointegrated variables
-as an error correction term. VECM is a special case of the vector autoregressive
-model (VAR) model. VAR model expresses future values as a linear combination of
-variables past values.  However, VAR model cannot be used with non-stationary
-variables, which is a common feature of financial assets. VECM is a linear model
-but in terms of variable differences. If cointegration exists, variable
-differences are stationary and they introduce an error correction term which
-adjusts coefficients to bring the variables back to equilibrium. In finance,
-many economic time series turns out to be stationary, when they are differentiated once (see section \ref{sec:coint} for more details).
-Additionally, cointegration restrictions applied on the model often improves forecasting~\cite{duy1998}.
-This is the reason why VECM has been widely adopted.
-
-VECM parameters are obtained using ordinary least squares
-(OLS) method \cite{golub1980}. OLS has two main problems: it is sensitive to errors on input data
-and also involves many calculations. The former problem is commonly solved using
-Ridge Regression (RR) \cite{hoerl1970} which introduces a regularisation
-parameter that leads to an unbiased estimation with better generalisation
-capability. The second problem of computational complexity depends on the number
-of past values and observations considered.  Recently, online learning
-algorithms have been proposed to solve problems with large data sets because of
-their simplicity and their ability to update the model when new data is
-available. We use an online version of the linear ridge regression and also a modification introduced by 
-Vovk \cite{vovk2001} called the Aggregating algorithm regression (AAR).
-Our proposal is an online formulation of the VECM called Online VECM (OVECM)
-based on the consideration of only a sliding window of the historical data at every step.
-OVECM parameters are obtained via OLS, RR or AAR.
-In order to measure performance, a competitive analysis is done which compares the proposal cumulative loss against the optimal offline algorithm.
-
-In finance, little academic research has been done using high-frequency Forex
-rates \cite{Bekiros2015}. Our proposal is tested using
-three mayor currency pairs (MCP) from the foreign exchange market:  Euro (EUR) to United States Dollar (USD) (EURUSD), British Pound (GBP) to
-USD (GBPUSD) and USD to Swiss Franc (CHF) (USDCHF). We used samples with a frequency of 10 seconds.
-
-This paper is organised as follows: section~\ref{sec:background} presents concepts of integration and cointegration, the VECM model and an online learning introduction, the OVECM algorithm is presented in section~\ref{sec:methodology}.
-In section~\ref{sec:results} we describe the experiments carried out and their results. This section also includes a
-description of the test data.  Section~\ref{sec:conclusions} contains the
-conclusions and a discussion of future research.
-
-\section{Background}
-\label{sec:background}
 \subsection{Integration and Cointegration}\label{sec:coint}\  
@@
 
-Let $\mathbf{y}_t^\nu$, $\nu=1,\dots,p$, be a set of $p$ vector time series of
+Let $\mathbf{y}_t^\nu$, $\nu=1,\dots,l$, be a set of $l$ vector time series of
 order $I(1)$.  They are said to be {\em cointegrated\/} if a vector
-$\beta=[\beta(1),\dots,\beta(p)]^\top \in \mathbb{R}^p$ exists, such that the
+$\beta=[\beta(1),\dots,\beta(l)]^\top \in \mathbb{R}^p$ exists, such that the
 time series,
@@
 \mathbf{Z}_t:= 
-\sum_{\nu=1}^p \beta(\nu)\,\mathbf{y}_t^\nu\,\sim\,\text{I(0)}\,.
+\sum_{\nu=1}^l \beta(\nu)\,\mathbf{y}_t^\nu\,\sim\,\text{I(0)}\,.
 \end{equation}
@@
 
-\subsection{Vector Error Correction Model}\label{sec:varvec}
+\subsection{Vector Autorregresive Models}\label{sec:varvec}
 
-VECM is a special case of VAR model and both describe the joint behaviour
+VECM is a special case of the VAR \cite{sims1980} model and both describe the joint behaviour
 of a set of variables.
@@
 
-VECM is a special form of a VAR model for I(1) variables that are also
-cointegrated~\cite{banerjee1993}.
+However, the VAR model cannot be used with non-stationary variables. VECM \cite{engle87} is also a linear model and a special form of a VAR model for I(1) variables that are also
+cointegrated~\cite{banerjee1993}. If cointegration exists, variable differences
+are stationary and they introduce an error correction term which adjusts
+coefficients to bring the variables back to equilibrium.
 
@@
 \begin{equation}\label{eq:vareq}
-\mathbf{Y} = 
+\mathbf{B} = 
 \mathbf{A} \mathbf{X} + 
@@
 %
-\noindent where $\mathbf{Y}$ dimension is $((N-p)\times l)$, $\mathbf{A}$
+\noindent where $\mathbf{B}$ dimension is $((N-p)\times l)$, $\mathbf{A}$
 dimension is $((N-p)\times(r+(p-1)l +1))$, $\mathbf{X}$ dimension is $((r+(p-1)l
@@
 \begin{alignat}{3}
-\mathbf{Y}
+\mathbf{B}
 &= \begin{bmatrix}
@@
    \end{pmat}\, .
+\label{eq:Amatrix}
 \end{align}
@@
 can be solved with respect to $\mathbf{X}$ using the ordinary least
-squares estimation.
+squares (OLS) estimation.
 
-\subsection{Ordinary Least Squares method}
-
-The solution $\widehat{\mathbf{A}}$ to
-equation~(\ref{eq:vareq}) can be obtained by the ordinary least squares (OLS)
-method. $\widehat{\mathbf{X}}$ is the solution of the quadratic optimization problem
-\begin{equation*}
-\widehat{\mathbf{X}} = \underset{\mathbf{X}}{\text{Arg\;min}}
-\|\mathbf{A}\mathbf{X}-\mathbf{Y}\|_2^2
-\end{equation*}
-\noindent for which the solution $\widehat{\mathbf{X}}$ is well-known:
-\begin{equation*}
-\label{eq:MP}
-\widehat{\mathbf{X}}=\mathbf{A}^{\!\!+}\,\mathbf{Y}
-\end{equation*}
-\noindent where $\mathbf{A}^{\!\!+}$ is the Moore-Penrose pseudo-inverse
-which, when $\mathbf{A}$ is full rank, can be written as follows: 
-\begin{equation}
-\label{eq:pseudoinverse}
-\mathbf{A}^{\!\!+}= (\mathbf{A}^{\!\!\top} \mathbf{A})^{-1}\mathbf{A}^{\!\!\top} \, .
-\end{equation}
-
-\subsection{Ridge regression method}
-
-In order to avoid the singularity of the matrix $\mathbf{A}^\top \mathbf{A}$, a regularisation term is introduced:
-
-\begin{equation}
-\label{eq:RRproblem} 
-\widehat{\mathbf{X}} (\lambda) = \underset{\mathbf{X}}{\text{Arg\;min}}  \| \mathbf{A}\mathbf{\mathbf{X}} - \mathbf{Y} \|_2^2  + \lambda
- \| \mathbf{\mathbf{X}}\| ^2
-\end{equation}
-
-\noindent which optimal solution is well known:
-
-\begin{equation}
-\label{eq:optsolRR}
-\widehat{\mathbf{X}} (\lambda)=(\mathbf{A}^\top \mathbf{A}+\lambda \mathbb{I})^{-1}\mathbf{A}^\top \mathbf{Y} \, ,
-\end{equation}
-
-\subsection{Online learning}
-Online learning is a supervised machine learning framework that is useful when
-we have sequential access to a sample only once.  This differs from the
-classical batch learning, where there is an entire data set available and the
-learner can build the internal model without any limits in accessing the data.
-In batch learning, there is time enough to carefully analyse the dataset, build
-large predictive models and combine them in a sophisticated way. However, when
-execution time is critical and new data is required to be included in the model,
-online learning algorithms are a better option.
-
-In batch learning, the training phase is commonly a computationally expensive
-process. Therefore, when new data arrives it can't be included easily into the
-model. Moreover, it could happen that it won't be enough time to process the
-new data before more data arrives. In online learning algorithms there is no
-training phase, but the model is updated and evaluated at every time step. This
-model updating is computationally less expensive than a training phase.
-Online algorithms allow incremental learning by processing one instance at a
-time. This is done updating the current model instead of building the model from
-scratch.
-
-Online learning is also useful when some past data may be irrelevant
-or we want to improve computational efficiency. However, the use of less
-historical data could affect accuracy.
-
-The online learning framework was first introduced in the perceptron algorithm
-\cite{rosenblatt58}. There are several other widely used online methods such as
-passive-aggressive \cite{crammerETall2006}, stochastic gradient descent
-\cite{zhang2004}, aggregating algorithm \cite{vovk2001} and the second order
-perceptron \cite{cesa-bianchi2005}.  In \cite{cesa-bianchi2006} an in-depth
-analysis of online learning is provided.
-
-Algorithm~\ref{alg:onlinealg} shows the online learning algorithm structure:
-
-\begin{algorithm}[ht]
-\begin{algorithmic}[1]
-    \STATE Receives input $\mathbf{x}_t$
-    \STATE Makes prediction $\mathbf{\hat{y}}_t$
-    \STATE Receives response $\mathbf{y}_t$
-    \STATE Incurs loss $l_t(\mathbf{y}_t,\mathbf{\hat{y}}_t)$
-\end{algorithmic}
-\caption{Structure of a Learning System}
-\label{alg:onlinealg}
-\end{algorithm}
-
-\noindent where $l$ is some loss function. Performance is later measured after
-$T$ trials using the cumulative loss function:
-
-\begin{equation*}
-L_T = \sum_{t=1}^T l_t(\mathbf{y}_t,\mathbf{\hat{y}}_t)
-\end{equation*}
-
-This cumulative loss function is compared against the cumulative loss function of its its optimal offline algorithm. An optimal offline algorithm can view the sequences of requests in advance. 
-This comparison is called competitive analysis  \cite{sleator1985}  which was designed for analysing the performance of an online algorithm. 
-The effectiveness of an online algorithm may be measured by its competitive ratio which is the worst-case ratio of the online algorithm and the the optimal offline algorithm.
-
-\subsection{Online Ridge Regression}
-
-Online RR is the online formulation of the regularised OLS method
-and is based on the following equivalent formulation of the RR optimal solution.
-
-Equation~(\ref{eq:optsolRR}) can also be written as:
-\begin{eqnarray*}
-\label{eq:RReapand}
-\mathbf{\mathbf{X}}_{\text{ridge}}&=&(\mathbf{A}^\top \mathbf{A}+ \lambda
-\mathbb{I})^{-1}\mathbf{A}^\top \mathbf{Y} \\
-&=& \displaystyle \big (\sum_{t=1}^m
-\mathbf{a}_t \mathbf{a}_t  ^\top + \lambda \mathbb{I}\big )^{-1}
-\sum_{t=1}^m \mathbf{a}_t \mathbf{y}_t \, .
-\end{eqnarray*}
-
-\noindent where 
-\begin{equation}
-\label{eq:notation}
-	\mathbf{A} = 
-\left[
-  \begin{tabular}{c>{$}c<{$}c}
-    -- & \mathbf{a}^{\top}_{1} & --\\
-    -- & \mathbf{a}^{\top}_{2} & --\\
-    & \vdots & \\
-    -- & \mathbf{a}^{\top}_{m} & --
-  \end{tabular}
-\right]
-\quad \text{and} \quad
-\mathbf{B} =
-\left[
-  \begin{tabular}{c>{$}c<{$}c}
-    -- & \mathbf{b}_{1} & --\\
-    -- & \mathbf{b}_{2} & --\\
-    & \vdots & \\
-    -- & \mathbf{b}_{m} & --
-  \end{tabular}
-\right]
-\end{equation}
-
-Lets define $\displaystyle\mathbf{S}= \sum_{t=1}^m \mathbf{a}_t
-\mathbf{a}_t  ^\top + \lambda \mathbb{I} $ and $\mathbf{W}=
-\displaystyle\sum_{t=1}^m \mathbf{a}_t \mathbf{y}_t$, so the
-algorithm~\ref{alg:RR} shows the iterative formulation:
-
-\begin{algorithm}[H]
-\begin{algorithmic}[1]
-\REQUIRE $\,$ \\
-$\{\mathbf{a}_1,\dots,\mathbf{a}_{m} \}$: $m$ input vectors \\
-$\{\mathbf{y}_1,\dots,\mathbf{y}_{m} \}$: $m$ target vectors \\
-$\lambda$: regularization parameter \\
-\ENSURE  $\,$ \\
-$\{f(\mathbf{a}_1),\dots,f(\mathbf{a}_{m}) \}$: model predictions \\
-\STATE Initialize $\mathbf{S}=\lambda \mathbb{I}$
-and $\mathbf{W}=0$
-\FOR { $t = 1$ to $m$ }
-	\STATE read new $\mathbf{a}_t$
-	\STATE $\mathbf{X}=\mathbf{S}^{-1}\mathbf{W}$
-	\STATE output prediction $f(\mathbf{a}_t) = \mathbf{X}^\top \mathbf{a}_t$
-   	\STATE $\mathbf{S} = \mathbf{S} + \mathbf{a}_t \mathbf{a}_t^\top$
-   	\STATE Read new $y_t$
-    	\STATE $\mathbf{W} = \mathbf{W} + \mathbf{a}_t \mathbf{y}_t$
-\ENDFOR
-\end{algorithmic}
-\caption{Online Ridge Regression}
-\label{alg:RR}
-\end{algorithm}
-
-\subsection{The Aggregating Algorithm for Regression}
-
-The AAR, proposed by~\cite{vovk2001} (also known as Vovk-Azoury-Warmuth predictor ~\cite{azoury2001}), is an application of the aggregating
-algorithm to the problem of regression. The idea is to introduce the new input
-vector $\mathbf{x}_{m+1}$ to solve the model parameters:
-
-\begin{equation}
-\label{eq:AARexpand}
-\mathbf{X}_{aar} = \displaystyle \big (\sum_{t=1}^{m+1}
-\mathbf{a}_t \mathbf{a}_t  ^\top+ \gamma \mathbb{I}\big )^{-1}
-\sum_{t=1}^m \mathbf{a}_t \mathbf{y}_t \, .
-\end{equation}
-
-If we define $\displaystyle\mathbf{S}= \sum_{t=1}^{m+1} \mathbf{a}_t
-\mathbf{a}_t  ^\top + \gamma \mathbb{I} $ and $\mathbf{W}=
-\displaystyle\sum_{t=1}^m \mathbf{a}_t \mathbf{y}_t$, the
-algorithm~\ref{alg:AAR} is slightly different to the algorithm~\ref{alg:RR}, 
-which updated matrix $\mathbf{S}$ before making the prediction:
-
-\begin{algorithm}[ht]
-\begin{algorithmic}[1]
-\REQUIRE $\,$ \\
-$\{\mathbf{a}_1,\dots,\mathbf{a}_{m} \}$: $m$ input vectors \\
-$\{\mathbf{y}_1,\dots,\mathbf{y}_{m} \}$: $m$ target vectors \\
-$\lambda$: regularisation parameter \\
-\ENSURE  $\,$ \\
-$\{f(\mathbf{a}_1),\dots,f(\mathbf{a}_{m}) \}$: model predictions \\
-\STATE Initialize $\mathbf{S}=\lambda \mathbb{I}$
-and $\mathbf{W}=0$
-\FOR { $t = 1$ to $m$ }
-	\STATE read new $\mathbf{a}_t$
-   	\STATE $\mathbf{S} = \mathbf{S} + \mathbf{a}_t \mathbf{a}_t^\intercal$
-	\STATE $\mathbf{X}=\mathbf{S}^{-1}\mathbf{W}$
-	\STATE output prediction $f(\mathbf{a}_t) = \mathbf{X}^\intercal \mathbf{a}_t$
-   	\STATE Read new $\mathbf{y}_t$
-    	\STATE $\mathbf{W} = \mathbf{W} + \mathbf{a}_t \mathbf{y}_t$
-\ENDFOR
-\end{algorithmic}
-\caption{{\em The aggregating algorithm for regression}}
-\label{alg:AAR}
-\end{algorithm}
-
-\section{Methodology} \label{sec:methodology}
-
-Since financial problems are stream data problems, it is unfeasible to include
-all input data in a VECM model. Our proposal consists on an online version of
-VECM (OVECM) capable of updating the model with new arrival data and give
-responses in a shorter period of time. OVECM only considers a time varying window
-of historical data and the using of RR or its variant AAR (Aggregating Algorithm
-for Regression) as an alternative of OLS to get model parameters.
-
-Our proposal considers only the most recent $L$
-rows of matrices $\mathbf{A}$ and $\mathbf{B}$ defined as  $\mathbf{A}_t$ and
-$\mathbf{B}_t$:
-
-\begin{equation}
-\label{eq:notation}
-	\mathbf{A}_t = 
-\left[
-  \begin{tabular}{c>{$}c<{$}c}
-    -- & \mathbf{a}^{\top}_{t-L} & --\\
-    %-- & \mathbf{a}^{\top}_{t-L+1} & --\\
-    & \vdots & \\
-    -- & \mathbf{a}^{\top}_{t} & --
-  \end{tabular}
-\right]
-,
-\mathbf{B}_t=
-\left[
-  \begin{tabular}{c>{$}c<{$}c}
-    -- & \mathbf{b}_{t-L} & --\\
-    %-- & \mathbf{b}_{t-L+1} & --\\
-    & \vdots & \\
-    -- & \mathbf{b}_{t} & --
-  \end{tabular}
-\right] \, ,
-\
-\end{equation}
-
-\noindent so that VECM parameters $\mathbf{X}_t$ are found using the following
-equation:
-
-\begin{equation}
-\mathbf{Y}_t = \mathbf{A}_t \mathbf{X}_t + \mathbf{E}_t\\
-\end{equation}
-
-The RR solution $\mathbf{\hat{X}}_t(\lambda)$ using the sliding window matrices defined above
-is:
-
-\begin{equation}
-\label{eq:oproblem}
-\mathbf{\hat{X}}_t(\lambda)=\mathbf{S}_t^{-1} \mathbf{W}(t) \, ,
-\end{equation}
-
-\noindent where $\mathbf{S}{t}$ and $\mathbf{W}{t}$ are define as:
-
-\begin{eqnarray*}
-\mathbf{S}_t =&{\bf A}_t^\top{\bf A}_t+ \lambda \mathbb{I} &=
-\sum_{i=0}^L \mathbf{a}_{t-i}\mathbf{a}_{t-i}^\top + \lambda \mathbb{I}
-\label{eq:S} \\
-\mathbf{W}_t =& {\bf A}_t^\top{\bf B}_t &= 
-\sum_{i=0}^L \mathbf{a}_{t-i}\mathbf{b}_{t-i} 
-\label{eq:W}
-\end{eqnarray*}
-
-It is worth noticing that, at the next time step, the matrices $\mathbf{S}_{t+1}$
-and $\mathbf{W}_{t+1}$ are slightly different to $\mathbf{S}_t$ and
-$\mathbf{W}_t$:
-
-\begin{eqnarray*}
-\mathbf{S}_{t+1}&=&
-\mathbf{S}_t +
-\mathbf{a}_{t+1}
-\mathbf{a}_{t+1}^\top -
-\mathbf{a}_{t-L} \mathbf{a}_{t-L}^\top \\
-\mathbf{W}_{t+1}&=&
-\mathbf{W}_t +
-\mathbf{a}_{t+1}
-\mathbf{b}_{t+1} -
-\mathbf{a}_{t-L} \mathbf{b}_{t-L} \, .
-\end{eqnarray*}
-
-Since we need to obtain the inverse of $\mathbf{S}_{t+1}$ at every step,
-we used the Sherman-Morrison-Woodbury formula in order to reduce the number of operations. 
-The Sherman-Morrison-Woodbury formula states if a matrix $\mathbf{A}$ is a positive
-definite matrix and its inverse matrix is known, then the inverse of the
-matrix $\mathbf{B}=\mathbf{A} + \mathbf{x}\mathbf{x}^\top$ can be
-obtained as:
-
-\begin{equation}
-\label{eq:SMW}
-\mathbf{B}^{-1} = \mathbf{A}^{-1} -
-\frac{(\mathbf{A}^{-1}\mathbf{x})(\mathbf{A}^{-1}\mathbf{x})^{\top}}
-{1 + \mathbf{x}^{\top} \mathbf{A}^{-1} \mathbf{x}} \, .
-\end{equation}
-
-This formula reduces the inverse matrix computation order from $O(n^3)$ to
-$O(n^2)$. In our case the inverse of $\mathbf{S}_{t+1}$ can be obtained applying twice the Sherman-Morrison-Woodbury formula.
-
-The algorithm~\ref{alg:proposal} shows OVECM using three different
-methods for geting model parameters: OLS, RR, AAR which leads to three different
-settings for OVECM:
-
-\begin{algorithm}[ht]
-\begin{algorithmic}[1]
-\REQUIRE $\,$ \\
-$\mathbf{y}$: matrix with $N$ input vectors and $l$ time series\\
-$p$: number of past values \\
-method: \{'OLS','RR','AAR'\} \\
-$\lambda$: regularization parameter (only requires for RR and AAR methods) \\
-$L$: window size of data($L<N$) \\
-$\text{mean\_error}$: MAPE threshold \\
-$n$: windows size to obtain in-sample MAPE \\
-\ENSURE  $\,$ \\
-$\{\mathbf{y}_{\text{pred}}[L+1],\dots, \mathbf{y}_{\text{pred}}[N]\}$: model predictions 
-\STATE solver = new \texttt{Solver}(method,$\lambda$) \\
-\FOR { $t =1$ to $N-L$ }
-    \STATE $\mathbf{Y} \gets \mathbf{y}[t:t+L-1]$
-    \STATE $\mathbf{y}_t \gets \mathbf{y}[t+L]$
-	\IF {$t = 1$}
-	    \STATE{$v \gets \texttt{getJohansen}(\mathbf{Y},p)$}
-	    \STATE{$\mathbf{A} ,\quad \mathbf{B} \gets
-        \texttt{vecMatrix}(\mathbf{Y},p,v)$}
-        \STATE \texttt{solver.init\_model($\mathbf{A},\mathbf{B}$)} 
-    \ENDIF
-	\STATE{ $\mathbf{a}_t ,\quad \mathbf{b}_t \gets
-    \texttt{vecMatrixOnline}(\mathbf{Y},\mathbf{y}_t,p,v)$}
-    \STATE $\mathbf{X} ,\quad \mathbf{Y}_{\text{pred}}[t] \gets \texttt{solver.regression}
-    (\mathbf{a}_t,\mathbf{b}_t)$
- 	    \STATE{$v \gets \texttt{getJohansen}(\mathbf{Y},p)$}
-\ENDFOR
-\end{algorithmic}
-\caption{OVECM: Online VECM}
-\label{alg:proposal}
-\end{algorithm}
-
-Our proposal considers the following:
-
-\begin{itemize}
-\item The function \texttt{getJohansen} returns cointegration vectors given by the
-Johansen method considering the trace statistic test at 95\% level of
-significance. The first cointegration vector is only considered.
-\item The function \texttt{vecMatrix} returns VECM
-matrices $\mathbf{A}_t,\mathbf{B}_t$ shown in equation~(\ref{eq:notation}). This
-method is only required at the first step.
-\item The function \texttt{vecMatrixOnline} returns new rows $\mathbf{a}_t^\top$ and
-$\mathbf{b}_t^\top$ given new input data $\mathbf{y}_t$.
-\item The Solver class (see algorithm \ref{alg:solver}) obtain VECM parameters  
-using RR, AAR or OLS methods. This solver class obtains the inverse of matrix $\mathbf{S}_{t+1}$ using the  Sherman-Morrison-Woodbury twice (see equation \ref{eq:SMW}).
-\item In order to set VECM parameter: $L$ and $p$, we use the Akaike Information
-Criterion (AIC). RR parameter $\lambda$ was done by cross-validation.
-\end{itemize}
-
-\begin{algorithm}[ht]
-\begin{algorithmic}[1]
-\REQUIRE $\,$ \\
-method: \{'OLS','RR','AAR'\} \\
-$\mathbf{A}$: VECM design matrix \\
-$\mathbf{B}$: VECM dependant variables \\
-$\mathbf{a}_t$: new row of $\mathbf{A}$ \\
-$\mathbf{b}_t$: new row of $\mathbf{B}$ \\
-$\gamma$: regularization parameter \\
-$n$: windows size to obtain in-sample MAPE \\
-\ENSURE  $\,$ \\
-$\mathbf{X}$: regression solution \\
-$\mathbf{e}$: in-sample MAPE \\
-%\quad \\
-%\texttt{init}($L,\lambda$)
-%\STATE self.L = L
-%\STATE self.lambda = $\lambda$
-\quad \\
-\texttt{init\_model}($\mathbf{A},\mathbf{B}$)
-\STATE $ [m \quad n] = \text{size}(\mathbf{A}) $ 
-\STATE $\mathbf{S} = \displaystyle \sum_{i=1}^m \mathbf{a}_i \mathbf{a}_i^\top + \gamma \mathbb{I}$
-\STATE $\mathbf{W} = \displaystyle \sum_{i=1}^m \mathbf{a}_i \mathbf{b}_i$
-%\STATE{$ \mathbf{a}_L ,\quad  \mathbf{b}_L \gets \mathbf{A}[:,1]^\top,  \mathbf{B}[:,1] $} 
-\quad \\
-\texttt{regression}($\mathbf{A},\mathbf{B},\mathbf{a_t},\mathbf{b_t}$) \\
-%\IF {method == 'OLS'}
-%        \STATE $\mathbf{X}=(\mathbf{A}^\top \mathbf{A})^{-1}\mathbf{A}^\top
-%        \mathbf{B}$
-\IF {method == 'RR' \OR method == 'OLS'}
-        \STATE $\mathbf{X} = \mathbf{S}^{-1} \mathbf{W} $
-        \STATE $\mathbf{S} = \mathbf{S}+
-        \mathbf{a}_t \mathbf{a}_t^\top-
-        \mathbf{a}_{t-L} \mathbf{a}_{t-L}^\top$
-        \STATE $\mathbf{W} = \mathbf{W} + \mathbf{a}_t \mathbf{b}_t$
-\ELSIF {method == 'ARR'}
-        \STATE $\mathbf{S} = \mathbf{S}+
-        \mathbf{a}_t \mathbf{a}_t^\top-
-        \mathbf{a}_{t-L} \mathbf{a}_{t-L}^\top$
-        \STATE $\mathbf{X} = \mathbf{S}^{-1} \mathbf{W} $
-        \STATE $\mathbf{W} = \mathbf{W} + \mathbf{a}_t \mathbf{b}_t$
-\ENDIF
-\STATE $\mathbf{y}_{\text{pred}} = \mathbf{X}^\top \mathbf{a}_t$
-\end{algorithmic}
-\caption{Solver Class for Regression Methods}
-\label{alg:solver}
-\end{algorithm}
-
-The proposal was compared against an optimal online algorithm which is a time varying
-VECM. Algorithm \ref{alg:OOVECM} shows this modification:
-\begin{algorithm}[ht]
-\begin{algorithmic}[1]
-\REQUIRE $\,$ \\
-$\mathbf{y}$: matrix with $N$ input vectors and $l$ time series\\
-$p$: number of past values \\
-$L$: sliding window size ($L<N$) \\
-\ENSURE  $\,$ \\
-$\{\Delta \mathbf{y}_{\text{pred}}[L+1],\dots,\Delta \mathbf{y}_{\text{pred}}[N]\}$: model predictions 
-\FOR { $i =0$ to $N-L$ }
-    \STATE $\mathbf{y}_i \gets \mathbf{y}[i:i+L]$
-	\STATE{$v \gets \texttt{getJohansen}(\mathbf{y}_i,p)$}
-	\STATE{$[\mathbf{A} \quad \mathbf{B}] \gets
-    \texttt{vecMatrix}(\mathbf{y}_i,p,v)$}
-    \STATE $\mathbf{X} \gets \text{OLS} (\mathbf{A},\mathbf{B})$%\mathbf{(A^\top A)^{-1}A^\top B}$
-    \STATE $\Delta \mathbf{Y}_{\text{true}}[i] \gets \mathbf{B}[-1,:]$
-    \STATE $\Delta \mathbf{Y}_{\text{pred}}[i] \gets \mathbf{A}[-1,:] \times \mathbf{X}$
-\ENDFOR
-    \STATE $\text{MAPE} \gets \texttt{mape}(\Delta \mathbf{Y}_{\text{true}}, \Delta
-    \mathbf{Y}_{\text{pred}})$
-\end{algorithmic}
-\caption{OOVECM: Optimal Online Vector Error Correction Model}
-\label{alg:OOVECM}
-\end{algorithm}
-
-\section{Experimental results}
-\label{sec:results}
-\subsection{Data} \label{sec:unitroot}
-
-OVECM tests were carried out using three foreign exchange rates all related to
-USD: EURUSD, GBPUSD and USDCHF. This data was collected from the free
-database Dukascopy which gives access to the Swiss Foreign Exchange marketplace
-~\cite{Dukascopy2014}.
-
-The tests were done using 10-seconds frequency from ask prices which
-corresponded to 8640 data points per day from the 11th to the 15th of August
-2014. 
-Experiments were run with data starting at 9:00 GMT of the 11th of August 2014,
-which corresponds to the opening of the New York financial market. Figure \ref{fig:forexdata}  shows a sample of the data used and figure \ref{fig:distdata} shows distribution of the data.
-
-\hspace*{-1.5in}
-\begin{figure}[!ht]
-  \vspace*{-1cm}
-  \hspace*{-0.4in}
-  \centering
-  \includegraphics[scale=0.34]{forexdata}
-  \caption{10 seconds frequency forex data for EURUSD, GBPUSD and USDJPY}
-  \label{fig:forexdata}
-\end{figure}
-
-\begin{figure}[!ht]
-  \vspace*{-1cm}
-  \hspace*{-0.3in}
-  \centering
-  \includegraphics[scale=0.3]{distdata}
-  \caption{Distribution of EURUSD, GBPUSD and USDJPY data}
-  \label{fig:distdata}
-\end{figure}
-
-\subsection{Unit root tests} \label{sec:unitroot}
-Before running the tests, we firstly checked whether the time series were
-I(1) using the Augmented Dickey Fuller (ADF) test at 95\% significance level.
-Table~\ref{tab:adf} shows that all currency rates cannot reject the unit root
-test but they rejected it with their first differences. This means that all of
-them are I(1) time series and we are allowed to use VECM and therefore OVECM.
-
-\begin{table}[h!]
-\label{tab:adf}
-\begin{center}
-\begin{tabular}{|l|c|c|c|c|c|}
-\hline
-& \textbf{Statistic} & \textbf{Critical value} & \textbf{Result}\\
-\hline
-EURUSD          &  -0.05398   & -1.94101 & True       \\
-$\Delta$ EURUSD & -89.12344   & -1.94101 & False       \\
-GBPUSD          &  -0.88208   & -1.94101 & True          \\
-$\Delta$ GBPUSD & -82.56501   & -1.94101 & False       \\
-CHFUSD          & -0.49941    & -1.94101 & True         \\
-$\Delta$ CHFUSD & -74.21940   & -1.94101 & False       \\
-\hline
-\end{tabular}
-\end{center}
-\caption{Unit roots tests for EURUSD, GBPUSD and USDCHF at 10-seconds
-frequency.}
-\end{table}
-
-\section{Experiments}
-
-Competitive analysis were done considering the optimal offline algorithm (OOVECM).
-Cumulative loss is shown for different settings of OVECM.
-Figure \ref{fig:RRcomparison}  shows a comparison between OVECM using RR with different values for $\lambda$.
-It should be notice that RR with $\lambda = 0$ setting is equivalent to OLS.
-Experiments show that it is possible to reduce cumulative loss using some $\lambda \neq 0$.
-
-\begin{figure}[!ht]
-  %\vspace{-0.8cm}
-  \centering
-  \includegraphics[width=0.5\textwidth]{RRcomparison}
-  \caption{Cumulative loss RR VECM against its optimal algorithm using different $\lambda$ values}
-  \label{fig:RRcomparison}
-\end{figure}
-
-Same experiments were done using the AAR setting for different values for $\lambda$.
-The effectiveness of the regularization parameter is so clear than RR.
-However, AAR seems to be a good predictor compared with OOVECM.
-Results are shown in figure \ref{fig:AARcomparison}.
-
-\begin{figure}[!ht]
-  %\vspace{-0.8cm}
-  \centering
-  \includegraphics[width=0.5\textwidth]{AARcomparison}
-  \caption{Cumulative loss AAR VECM against its optimal algorithm using different $\lambda$ values}
-  \label{fig:AARcomparison}
-\end{figure}
-
-Finally, the three settings are compared with OOVECM. We can see that RR and AAR approaches are better than traditional OLS and also AAR seems to perform better than RR.
-
-\begin{figure}[!ht]
-  %\vspace{-0.8cm}
-  \centering
-  \includegraphics[width=0.5\textwidth]{onlinecomparison}
-  \caption{Cumulative loss of AAR, RR and OLS solutions for VECM against its optimal algorithm}
-  \label{fig:onlinecomparison}
-\end{figure}
-
-\section{Conclusions}
-\label{sec:conclusions}
-
-A new online version of the VECM has been proposed in order to use it with high frequency time series such as foreign exchange market rates.
-VECM takes advantage of the cointegration evidence between rates related to the USD currency.
-In order to get an online version we used the online versions of RR and AAR which are commonly used with problems with sequential access to the data.
-RR and AAR are used instead of the traditional OLS approach which has several known problems.
-A competitive analysis was done and experiments have shown that a regularised version OVECM could lead to improve cumulative losses compared with its optimal offline algorithm.
-Moreover, we showed that AAR could be a better option to improve performance independently of  how we choose $\lambda$ parameter.
-
-\bibliographystyle{plain}       % APS-like style for physics
-
-\bibliography{reference}   % name your BibTeX data base
-
-\end{document}
-% end of file template.tex
+%\subsection{Ordinary Least Squares method}
+%
+%The solution $\widehat{\mathbf{A}}$ to
+%equation~(\ref{eq:vareq}) can be obtained by the ordinary least squares (OLS)
+%method. $\widehat{\mathbf{X}}$ is the solution of the quadratic optimization problem
+%\begin{equation*}
+%\widehat{\mathbf{X}} = \underset{\mathbf{X}}{\text{Arg\;min}}
+%\|\mathbf{A}\mathbf{X}-\mathbf{B}\|_2^2
+%\end{equation*}
+%\noindent for which the solution $\widehat{\mathbf{X}}$ is well-known:
+%\begin{equation*}
+%\label{eq:MP}
+%\widehat{\mathbf{X}}=\mathbf{A}^{\!\!+}\,\mathbf{B}
+%\end{equation*}
+%\noindent where $\mathbf{A}^{\!\!+}$ is the Moore-Penrose pseudo-inverse
+%which, when $\mathbf{A}$ is full rank, can be written as follows: 
+%\begin{equation}
+%\label{eq:pseudoinverse}
+%\mathbf{A}^{\!\!+}= (\mathbf{A}^{\!\!\top} \mathbf{A})^{-1}\mathbf{A}^{\!\!\top} \, .
+%\end{equation}
+%when $\mathbf{A}$ is not full rank, i.e
+%$rank(\mathbf{A})=k <  n \leq m$, $\mathbf{A}^\top \mathbf{A}$ is
+%always singular and equation~(\ref{eq:pseudoinverse}) cannot be used.
+%More generally, the pseudo-inverse is best computed using the compact
+%singular value decomposition (SVD) of $\mathbf{A}$ which is:
+%\begin{equation*}
+%    \label{eq:compactsvd}
+%    \mathbf{A}=
+%    \mathbf{U_1}
+%    \boldsymbol \Sigma_1
+%    \mathbf{V}_1^{\top} \, .
+%\end{equation*}
+%Pseudo-inverse can then be written as follows:
+%\begin{equation*}
+%\label{eq:pseudoinversesvd}
+%\mathbf{A}^{\!\!+} = \mathbf{V}_1 \boldsymbol \Sigma_1^{-1} \mathbf{U}_1^\top \, .
+%\end{equation*}

### 2016-03-22 articles-conferences/iscef2016/src/conclusions.tex
agregando repo iscef2016

\section{\uppercase{Conclusions}}
\label{sec:conclusions}
Cointegration in financial time series has been largely studied and the 
Johansen method is commonly used to obtain cointegration relationships. 
In practice, it has been found that cointegration relations change with time. 
However, model-based cointegration such as VECM assumes that cointegration
remains unchanged in time. 
We empirically showed that the Johansen method is sensitive to the number
of lags but also to the amount of data considered.

Moreover, we introduced the notion of {\em percentage of cointegration\/} and
found that out-of-sample forecast performance MSE is related to the value of
this figure in the last samples.  We used this information to set the model
parameters.  Our proposal AVECM consists of an adaptive algorithm to update VECM
parameters every time that new data is available. These parameters are found by
maximising the percentage of cointegration of the last samples or iterations.

Despite the fact that high frequency Forex data can be spurious, the model performance can be less reliable (and more spurious) relative to the lower frequencies (such as 1 minute or 5 minute intervals) adopted by some other studies. However, the deficiency is offset by gain in accuracy from parallel processing which is capable of searching or examining a much larger state space given the same computational time.

Determining VECM parameters was the most expensive routine and it was run
using parallel processes using MPI which allowed a grid search within a range of
values for $L$ and $p$ to be made.
Tests were done using real currency rates data. 
Our proposal was tested in several scenarios using different times of
the day to show its behaviour related to the opening times of
financial markets.

Results showed that our proposed AVECM improves performance measures by finding
parameters of $L$ and $p$ maximising the percentage of cointegration.
These facts were consistent for different times of the day. 
We found that increasing the number of parameters will always lead to
better performance measures.

The parallel implementation allowed the execution times to be reduced
more than 9 times and therefore a response time was obtain before 10
seconds. Since we used 10-second frequencies we can say that our proposal is suitable
for use in an online context for real applications because response times
were less than this frequency.

For future study, it would be interesting to explore the relationship between
cointegration and performance in order to propose new criteria for
improving VECM parameters. It would also be interesting to include more 
explaining variables such as bid-ask spread and change in volume.

### 2016-03-22 articles-conferences/iscef2016/src/intro.tex
agregando repo iscef2016

\section{Introduction}
\label{sec:introduction}
In finance, it is common to find variables with long-run equilibrium
relationships. This is termed cointegration and reflects the idea that
some sets of variables cannot wander too far from each other. Cointegration
means that one or more linear combinations of these variables are stationary
even though individually they are not. Some models,
such as the Vector Error Correction (VECM) \cite{engle87}, take advantage of
this property and describe the joint behaviour of several cointegrated
variables. VECM introduces this long-run relationship among a set of cointegrated
time series as an error correction term. These time series must be integrated of order 1, denoted I(1), i.e. they become stationary at their first differences. In finance, I(1) time series are very common and 
to introduce cointegration restrictions in models often improves forecasting \cite{duy1998}. Therefore, VECM has been widely adopted in
financial applications \cite{mukherjee1995,seong2013,maysami2000,arestis2001} such as pair trading \cite{herlemont2003} or models with more than two variables \cite{mukherjee1995,engle2004}.

VECM parameters are obtained using the ordinary least squares
(OLS) method~\cite{golub1980}. Since OLS involves many calculations, the parameter estimation is computationally expensive when the number of lagged values and
data increases. Moreover, obtaining cointegration vectors 
is also an expensive routine. This is a main limitation to use VECM with stream data with high frequency.
Chen and Lung's BVECM \cite{chen2003} addresses the advantage of distributed processing over conventional rolling window processing.

Our aims were to study if a parallel version of VECM can be used with high frequency stream data.
Our approach was to determine adaptively the number of observations and lags of VECM which
maximise cointegration relations in the past. 
 This search is computationally
expensive because the Johansen
method~\cite{johansen1995} is required to find cointegration vectors, which is of order $O(n^3)$. Therefore, our proposal, called Adaptively Vector Error Correction (AVECM), 
is to parallelise this search in order to get new parameters before new data arrives.
Model effectiveness is focused on out-of-sample forecast rather than in-sample
fitting. This criterion allows AVECM prediction capability to be
expressed rather than just explaining data history.
The forecast capability of our method was measured using MSE and the Theil's $U$-statistic \cite{theil1966} widely used in economic forecast. Tests were run using four currency rates: Euro (EUR) to
United States Dollar (USD) (EURUSD), British Pound (GBP) to USD (GBPUSD), USD to Swiss
Franc (CHF) (USDCHF) and USD to Japanese Yen (JPY) (USDJPY) with a 10-seconds frequency.

This paper is organised as follows: section~\ref{sec:background} presents the
VAR and VECM, the AVECM algorithm is presented in section~\ref{sec:methodology}.
In section~\ref{sec:results} we describe the tests carried on to assess the
accuracy and the execution time of AVECM.  This section also includes a
description of the test data.  Section~\ref{sec:conclusions} contains the
conclusions and a discussion of future research.

### 2016-03-22 articles-conferences/iscef2016/src/methodology.tex
agregando repo iscef2016

\section{\uppercase{Methodology}} \label{sec:methodology}
\subsection{Motivation} \label{sec:proposal}

Cointegration vectors can be found applying the Johansen method which uses a
sample of the last historical data. However, VECM assumes cointegration vectors do not change in time.
In fact, the long-run relationships between the time series might change due to
several economic factors that can lead to structural breaks in the cointegration
relationship \cite{gregoryETal1996}. 
In order to show the the number of cointegration vectors which depends on 
the amount $L$ of historical data and the number of lags $p$ in the VECM, we used a grid search. We defined a grid of possible values for $L$ and $p$. $L$ goes throughout $[100,5000]$ with a step size of 300 and $p$  throughout $[1,5]$ with step size of 1. The first experiment consisted in determining the number of cointegration vectors for all combinations of $L$ and $p$.
We used four forex rates: EURUSD, GBPUSD, USDCHF and USDJPY with 10-second frequency. Data started at 13:00 GMT of the 13th of August 2014,
when the New York and London financial markets opened.

Figure~\ref{fig:hists} shows the distribution of the number of cointegration vectors 
given by the Johansen method for different values of $L$ and $p=1$ . This procedure was carried out by a sliding window of historical data moving 1000 times.

\begin{figure}[!h]
  %\vspace{-0.8cm}
  \centering
  \includegraphics[width=0.9\textwidth]{img/histCointVectors-offset20520-p-1-freq-10s}
  \caption{Histogram of the number of cointegration vectors using $p=1$. Four
  possible values for $L$ are shown (100, 2000, 3000 and 4000).}
  \label{fig:hists}
\end{figure}

The same experiments, this time using $p=3$, were carried out in order to see the
effect on the number of cointegration vectors found. 
Figure~\ref{fig:histsp3} shows that $p$ affects the distribution of the number of cointegration vectors by small amounts.

\begin{figure}[!h]
  %\vspace{-0.8cm}
  \centering
  \includegraphics[width=0.9\textwidth]{img/histCointVectors-offset20520-p-3-freq-10s}
  \caption{Histogram of the number of cointegration vectors using $p=3$. The distribution is very similar to the one with $p=1$}
   \label{fig:histsp3}
\end{figure}

From section \ref{sec:background} we know that $r=0$ means no cointegration and $r=l$ (we are using four rates, so $l=4$) reveals that no process is I(1) but stationary.
The interesting cases of cointegration are those where $r$ lies strictly
between $0$ and $4$, i.e. $0<r<4$.

In order to measure the extent of cointegration, we introduce a
{\em percentage of cointegration\/} as following:
\begin{equation} \label{eq:pcoint}
PC = 
\frac{\#\{ it \mid \text{$it$ has $r$ c.v. with $0<r<l$}\}}
     {\#it}\times 100
\end{equation}
where c.v. stands for cointegration vectors and $it$ is the number of iterations.

The corresponding PC values for $L = [100, 2000, 3000, 4000]$ and $p=1$ (figure  \ref{fig:hists}) are: $PC = [21.3, 98.7, 79.8,  65.2]$. For the same values of $L$ and $p=3$ (figure  \ref{fig:histsp3}) the percentage of cointegration is $PC = [18.4, 92.5, 68,  52.5]$. From these results we observed that the number of cointegration vectors changed with different values for $L$ and $p$ and the percentage of cointegration depends mainly on $L$.

The goal of our next experiment was to find a relation between this ratio
$PC$ and the performance measure MSE (see equation 
\ref{eq:MSE}). $L$ was defined between $[100,4000]$ with step size 300 and $p$ takes values between $[1,5]$ with step size 1.

\begin{figure}[ht!]
  %\vspace{-0.8cm}
  \centering
  \includegraphics[width=0.8\textwidth]{img/MSE-offset20520-p-2-freq-10s}
  \caption{MSE versus the percentage of cointegration considering 1000
  iterations. Best $L$ found was 1600. Below MSE versus $L$ shows a rapidly decreasing behaviour, founding minimum at PC = 84.3\% }
  \label{fig:cointvsmse}
\end{figure}

Figure~\ref{fig:cointvsmse} shows the relation between MSE and $PC$ and $L$.
We found that better cointegration percentage leads to better
performance accuracy in terms of reducing MSE. We also found that MSE rapidly decreased with increasing $L$.

Therefore we proposed to choose $L$ and $p$ in order to maximise the percentage of
cointegration $PC$ in the near past. This process was done every time that new data was processed. However, this search can be slow if we try different values for $L$ and $p$ and therefore we distributed this calculation in order to reduce searching time.

Our proposal was then a modified version of VECM, with parameter $p$ and the amount of historical data used $L$ obtained at every step. Only the search of $L$ and $p$ was done in a distributed environment, since this is the most expensive routine. Our proposal is called Adaptive Vector Error Correction model (AVECM).
AVECM is detailed in the algorithm ~\ref{alg:AVECM} which summarises our proposal.

\begin{algorithm}[ht!]
\begin{algorithmic}[1]
\REQUIRE $\,$ \\
$\mathbf{y}$: matrix with $N$ input vectors and $l$ time series\\
$j$: Starting point of testing \\
$it$: Ending point of testing \\
$ps$: list of $p$ values \\
$Ls$: list of $L$ values ($L<N$) \\
$m$: Iterations to determine parameters ($m < N-L$)\\
\ENSURE  $\,$ \\
$\{ \hat{\mathbf{y}}[1],\dots,\hat{\mathbf{y}}[it]\}$: prediction vectors \\
\FOR { $i =j$ to $it$ }
   \STATE $\mathbf{Y} \gets \mathbf{y}[:,i-1]$
    \STATE $L,p \gets
    \texttt{get\_best\_params}(Ls,ps,m,\mathbf{Y})$
        \STATE $model = \text{VECM}(\mathbf{Y},L, p)$
        \STATE $\hat{\mathbf{y}}[i-j] = model.predict()$
\ENDFOR
\end{algorithmic}
\caption{AVECM: Adaptive VECM.}
\label{alg:AVECM}
\end{algorithm}

The input of AVECM is time series prices which are cointegrated. The starting point of testing is $j$ and the total number of iterations is $it$. We need to ensure that $j$ is at least the maximum value of $Ls$. $Ls$ and $ps$ are the possible values for $L$ and $p$.
The function {\bf get\_best\_params} makes this grid search on the two vector
lists $Ls$ and $ps$ and returns the parameters $L$ and $p$ which maximise
the percentage of cointegration $PC$ (see equation~\ref{eq:pcoint}) for a
pre-defined number of iterations $m$. This function is implemented in a distributed 
environment, thus ensuring a response
before new data is available. 
After $L$ and $p$ parameters are found, VECM is built and used
to forecast the next data point.

\subsection{Model comparison} \label{sec:random}
We compared our proposal, in terms of performance, with the naive forecast of the random walk model and ARIMA. It is still difficult to outperform the random walk model for standard econometric forecasting models \cite{lo2011} despite its simplicity. The random walk model is defined as:

\begin{equation}
\mathbf{y}_t = \mathbf{y}_{t-1} + \epsilon_{t}
\label{rwmodel}
\end{equation}

The naive forecast of the time series difference $\hat{\mathbf{y}}_{t+1}$ for the random walk model is defined as:
\begin{equation}
\hat{\mathbf{y}}_{t+1} = \mathbf{y}_t + \hat{\epsilon}_{t+1} 
\end{equation}
\noindent where  $\hat{\epsilon}_{t+1} = \epsilon_{t}$.

On the other hand, ARIMA is widely used to forecast returns in finance \cite{tsay2005}. A process can be modelled as an ARIMA$(p,d,q)$ model if $\mathbf{x}_t=\Delta^d \mathbf{y}_t $, i.e after differencing $d$ times the time series $\mathbf{y}_t$,  we get an ARMA$(p,q)$. Since we are modelling returns, we used $d=1$. An ARMA$(p,q)$ model is the following:

\begin{equation}
\mathbf{x}_t = \sum_{i=1}^p \phi_i \mathbf{x}_{t-i}  +  \sum_{j=1}^q \theta_j \epsilon_{t-j}  
\end{equation}
\noindent with coefficients $\phi_p \neq 0$, $\theta_q \neq 0$ and $\sigma_{\epsilon}^2 > 0$.

\subsection{Evaluation methods} \label{sec:evaluation}

Forecast performance was evaluated using two different methods which are frequently used in finance:
\begin{description}
\item
{\bf MSE},  Mean Square Error measures the distance between forecasts
and the true values and large deviations from the true value have a
large impact due to squaring forecast error.
\begin{equation}\label{eq:MSE}
\text{MSE} = 
\frac{\displaystyle \sum_{t=1}^{N} (\mathbf{y}_t-\hat{\mathbf{y}}_t)^2}{N}
\end{equation}
\item {\bf $U$-statistic}, the Theil's $U$-statistic \cite{theil1966} is a unit free measure obtained as the ratio between the root MSE (RMSE) of a model and the RMSE of the naive random walk model. A $U$-statistic less than 1 implies the performance is better than the naive model.
\end{description}

### 2016-03-22 articles-conferences/iscef2016/src/results.tex
agregando repo iscef2016

\section{\uppercase{Experimental results}}
\label{sec:results}
\subsection{Data} \label{sec:unitroot}
AVECM tests were carried out using four foreign exchange rates all related to
USD: EURUSD, GBPUSD, USDCHF and USDJPY. This data was collected from Dukascopy~\cite{Dukascopy2014}, a free
database which gives access to the Swiss Foreign Exchange marketplace.

The tests were done using 10-seconds frequency from ask prices
from the 11th to the 15th of August 2014. Since one day corresponds to 8640 data points and we used 5 days of data, we have 43,200 data points in total.

\subsection{Unit root tests} \label{sec:unitroot}

VECM can only be used if the time series are I(1). Therefore, 
before running the tests, we firstly checked whether the time series were
I(1) using the Augmented Dickey Fuller (ADF) test with lags $p=1,2,3,4,5$. 
MacKinnon \cite{mackinnon2010} critical values for rejection of hypothesis of a unit root are: -2.62 (1\%), 1.94 (5\%) and 1.62 (10\%).
\begin{table*}[ht]
\label{tab:adf}
\centering
\begin{tabular}{llllll}
\toprule
{Variable} & {ADF(1)} & {ADF(2)} & {ADF(3)} & {ADF(4)} & {ADF(5)}\\ 
\midrule
EURUSD &  -0.052   & -0.054  & -0.054  & -0.054  & -0.054  \\
GBPUSD &  -0.744  & -0.784  & -0.805  & -0.837  & -0.846  \\
USDCHF &  -0.476   & -0.493  & -0.493  & -0.495  & -0.502  \\
USDJPY &  0.357   & 0.360  & 0.360  & 0.367  & 0.367  \\
$\Delta$ EURUSD & -128.4*  & -128.4*  & -96.85* & -89.12*   & -89.12*\\
$\Delta$ GBPUSD & -131.4*  & -112.7*  & -102.5* & -92.86*   & -88.29*\\
$\Delta$ USDCHF & -127.8*  & -127.8*  & -96.94* & -88.82*   & -80.79*\\
$\Delta$ USDJPY & -135.1*  & -135.1*  & -101.2* & -101.2*   & -101.2*\\
\bottomrule
\addlinespace[1ex]
\multicolumn{6}{l}{ \textsuperscript{*} Indicates significance at 1\% level} \\
\multicolumn{6}{l}{ \textsuperscript{**} Indicates significance at 5\% level} \\
\multicolumn{6}{l}{ \textsuperscript{***} Indicates significance at 10\% level} \\
\multicolumn{6}{l}{MacKinnon critical values for rejection of hypothesis of a unit root are:}\\
\multicolumn{6}{l}{ -2.62 (1\%), -1.94 (5\%) and -1.62 (10\%)}\\
\multicolumn{6}{l}{ADF($d$) Augmented Dickey-Fuller test with lag $d$} 
\end{tabular}
\caption{Unit roots tests for EURUSD, GBPUSD, USDCHF and USDJPY at 10-seconds
frequency.}
\end{table*}
Table~\ref{tab:adf} shows that all currency rates cannot reject the unit root test in their level form considering different lags, but they rejected it with their first differences. This means that all of
them are I(1) time series and we are allowed to use VECM and therefore our proposed AVECM.

\subsection{Performance accuracy} \label{sec:performance}

Algorithms AVECM, ARIMA and the naive random walk were tested using four days of data (from the 12th to the 15th of August 2014).
For AVECM we considered different number of iterations (parameter $m$ in algorithm \ref{alg:AVECM}): 10, 50 and 100. We tried 12, 24 and 47 different pair of combinations for $L$ and $p$. Possible values for $L$ were in the interval $[100,4000]$ and $p$ can have values in between $[1,5]$. Best AVECM performance was compared against ARIMA and the random walk model.
Table~\ref{tab:stats} shows the out-of-sample performance measures: MSE and $U$-statistic for AVECM and ARIMA. In terms of both measures we found that AVECM is superior to ARIMA and the naive random walk model. We also included the p-value that proves that the difference in the MSE is significant at the 99\% significance level in three of the four currency rates and at 90\% in the case of GBPUSD. The $U$-statistic shows that AVECM and ARIMA are superior to the naive random walk model and that our proposal is also superior to ARIMA.

\begin{table}[!htbp]
\caption{AVECM performance}
\label{tab:stats}
\begin{center}
\begin{tabular}{ccccccc}

& \multicolumn{3}{l}{MSE} & &
\multicolumn{2}{c}{$U$-statistic} \\ 
\hhline{~---~--} \\
 &
AVECM & ARIMA & p-value & &
AVECM & ARIMA \\

\hline
 EURUSD & 1.0702 e-09 & 1.1481 e-09 &  9.2509 e-12 & & 0.6863 & 0.7108\\
 GBPUSD & 1.6630 e-09 & 1.7408 e-09 &  6.9519 e-02 & & 0.6866 & 0.7025\\
 USDCHF & 5.8503 e-10 & 6.3545 e-10 &  2.8999 e-14 & & 0.6803 & 0.7091\\
 USDJPY  & 6.3483 e-06 & 6.5194 e-06 &  6.8536 e-05 & & 0.6964 & 0.7057\\
 \end{tabular}
 \end{center}
\end{table}

\subsection{Parallel implementation} \label{sec:parallel}

To determine $L$ and $p$ based on maximising the percentage of cointegration requires use of the 
Johansen
method which is a computationally expensive routine. This procedure is done by the function get\_best\_params($Ls,ps,m,\mathbf{Y}$) shown in algorithm \ref{alg:AVECM}.
In order to improve the execution time of this search, our proposal included
a parallel search of VECM parameters using high performance computing.
The main objective was to obtain a response before a new data arrived in
the following10 seconds.

The Johansen method is already programmed in the Python Statsmodels
library~\cite{seabold2010} and the parallel implementation was developed using
MPI in Python. 
We chose MPI because it allows large-scale parallel applications
with wide portability to be built, being able to run in large clusters
or on local computers.
We tested our proposal in a cluster with 2 servers Xeon E5-2667 (2.90GHz)
of 24 cores each (48 cores in total) and 24GB RAM.
In order to compare serial and parallel execution times in AVECM, we set parameter $it=100$ in algorithm\ref{alg:AVECM}.

The $L$ parameter was always chosen between 100 and 4000 and $p$ always took values
between 1 and 5. Parameter $nparams$ represents the number of pairs ($L$,$p$) used to maximise the percentage of cointegration.

Execution time depends directly on $L$, $p$ and $nparams$ used, since they determine the size of
matrix $\mathbf{A}$ (see equation \ref{eq:Amatrix}) and therefore affect the OLS function execution time.
Therefore, if we try more combinations of $L$ and $p$ (increasing $nparams$) the
serial algorithm will take longer. For this financial time series we are interested in execution times below 10 seconds  (the time series frequency).

\begin{figure}[ht]
  \centering
  \includegraphics[width=0.75\textwidth]{img/extimes}
  \caption{Computing time of sequential and parallel algorithm is shown in the
  upper figure. Speed-up is shown below.}
  \label{fig:extimes}
\end{figure}

The superior figure \ref{fig:extimes} shows that best performance accuracy measurements are
achieved in times near or below 10 seconds in the parallel version. Contrarily,
serial times are high, above 10 seconds in most cases. Figure~\ref{fig:extimes} also shows the speed-up in computing time for AVECM which is near 9X if we use more than 25 parameters ($nparams$).

Execution times do not consider the loading data time, just that of finding
best parameters and matrix operations. The time MPI spends
transferring data and synchronising processes is about two seconds independently
of the number of processes considered. Execution times were measured using the Time python library.

### 2016-03-22 articles-conferences/elsevier-journal-2015/journal.tex
Agregar estos repos a la tesis

\documentclass[12pt,reqno]{amsart}
\usepackage{amsfonts,amssymb,amsmath,amsthm,
            amssymb,amsopn,amsxtra,amstext,amsbsy}
\usepackage{epsfig}
\usepackage[makeroom]{cancel}
\usepackage{array,arydshln}
\usepackage{array}
\usepackage{url}
%\usepackage{tikz}
\usepackage[spanish]{babel}
\usepackage[utf8x]{inputenc}
\usepackage{fancyhdr}
\usepackage{graphicx}
\usepackage{hyperref,wasysym}
\usepackage{color}
\usepackage{enumerate}
\usepackage{algorithm}
\usepackage{algorithmic}
\renewcommand{\algorithmicrequire}{\textbf{Input:}}
\renewcommand{\algorithmicensure}{\textbf{Output:}}

\newtheorem*{myteo}{Teorema}
\newtheorem*{mydef}{Definici\'on}

\voffset=-10mm
\oddsidemargin=0pt
\evensidemargin=0pt
\topmargin=-10mm
\headsep=10mm
\headheight=18pt
\footskip=30pt
\topskip=0mm
\textwidth 167mm
\textheight=230mm
\parindent=5mm
\marginparsep=3mm
\marginparwidth=19mm
\vfuzz2pt % Don't report over-full v-boxes if over-edge is small
\hfuzz6pt % Don't report over-full h-boxes if over-edge is small

%% MYREMARK ENVIRONMENT %%%%%%%%%%%
\newenvironment{mycolor}{\color{red}}{}
\newcounter{myremarkcount}
\setcounter{myremarkcount}{0}
\newsavebox{\myremarkbox}
\newenvironment{myremark}
{% This is the begin code
\medskip
\begin{flushright}
\begin{lrbox}{\myremarkbox}%
\begin{minipage}[t]{0.9\linewidth}
\stepcounter{myremarkcount}
\footnotesize
\begin{mycolor}
{\bf Remark \arabic{myremarkcount}.}
\begin{it}
}
{% This is the end code
\end{it}
\end{mycolor}
\end{minipage}
\end{lrbox}
\fbox{\usebox{\myremarkbox}}
\end{flushright}

\bigskip}

\title{A Time Varying Vector Error Correction Model Applied to Foreign Exchange
Market}
\author{Paola Arce \and Jonathan Antognini \and Luis Salinas \and Werner Kristjanpoller}

\begin{document}
\maketitle

\begin{abstract}
Financial time series are known for their non-stationary behaviour. However,
sometimes they exhibit some stationary linear combinations. When this happens,
it is said that those time series are cointegrated. Vector error correction
model (VECM) is an econometric model which characterizes the joint dynamic
behaviour of a set of cointegrated variables in terms of forces pulling towards
equilibrium.

VECM parameters are obtained using the ordinary least squares (OLS)
method.  Even though OLS is extensively used, it has over-fitting
issues and is computationally expensive. Ridge regression is commonly used
instead of OLS since they include a regularization parameter which could improve
prediction error.

In this study, we propose an online VECM which optimizes how model parameters
are obtained by using only relevant past data and RR instead of OLS.  Our
proposal also takes advantage of the long-run relationship between the time
series in order to obtain improved execution times. Our proposed method is
tested using synthetic data and foreign exchange rates.
Results show that our proposal enables execution time to be reduced by at least
50\% while maintaining good accuracy levels.
\end{abstract}

\input{src/intro}
\input{src/background}
\input{src/methodology}
\input{src/results}
\input{src/conclusions}
\bibliographystyle{apalike}
\bibliography{reference}
%\begin{thebibliography}{9}
%
%\bibitem{smola}
%Bernhard Scholkopf and Alexander J. Smola. 
%\emph{Learning with Kernels: Support Vector Machines, 
%Regularization, Optimization, and Beyond.}.
%MIT Press, Cambridge, MA, USA, 2001
%
%\bibitem{wiki}
%\url{http://en.wikipedia.org/wiki/Reproducing_kernel_Hilbert_space}
%
%\bibitem{wikibooksIsom}
%\url{http://en.wikibooks.org/wiki/Linear_Algebra/Definition_and_Examples_of_Isomorphisms}
%
%\bibitem{wolframREq}
%\url{http://mathworld.wolfram.com/EquivalenceRelation.html}
%
%\bibitem{wikiHomoM}
%\url{http://en.wikipedia.org/wiki/Homomorphism}
%
%
%\end{thebibliography}

\end{document}

### 2016-03-22 articles-conferences/elsevier-journal-2015/src/background.tex
Agregar estos repos a la tesis

--- 2016-01-26 src/4_OnlineLearning.tex
@@
-\chapter{Machine Learning Models}
-Machine learning is a scientific discipline focused on the development of algorithms focused on learning from examples.
-This idea has become central to the design of search engines, robots systems and forecasts applications which process large data sets.
-However, machines require an extended training period when developing algorithms to predict future behaviour such as financial time series forecasting where new data is available in a short period of time.
-Online machine learning techniques tackle this problem and allow to update the model with new data or to compute a new model using less data allowing to give a response in a short period of time.
-
-\vspace{0.5cm}
-
-\section{Introduction}
-Machine Learning (ML) studies computer algorithms for learning
-something such as to complete a task, to make accurate predictions or to behave intelligently.
-The learning is always based on samples and the objective is about to do better in the future based on the past experiences in an automatic way.
-ML is a subarea of artificial intelligence and broadly intersects with other fields such as statistics, mathematics, physics and computer science.
-There are many examples of machine learning problems: time series forecasting, image processing, face detection, spam filtering, weather prediction, search engines, among many others.
-
-ML models are often more accurate than what can be created through direct programming. The reason is that ML models are data driven and are able to examine large amounts of data.
-
-There are three types of ML classified depending on the nature of the learning input or output available to a learning system.
-\begin{description}
-\item[supervised learning]  the input data is a tuple which contains the example input and its desired outputs (also called labels).
-The list of tuples is called the training set.
-The concept of supervised learning comes from the supervisor, acting as a teacher in the learning process.
-The goal is to learn a general rule that maps inputs to outputs optimising a target function.
-There are two related problem types in supervised learning: classification and regression problems \cite{bishop2006}.
-Its two mainstream approaches are: support vector machines (SVMs) \cite{vapnik1998} and ensemble learning \cite{breiman1998}.
-Furthermore supervised learning can be categorised into offline or batch learning and online learning (see section \ref{sec:onoffline}).
-\item[unsupervised learning] also known as clustering \cite{ben2005}.
-In this type of learning no labels are given and the system has to find a structure on its own, discovering hidden patterns in data.
-\item[reinforcement learning] is the problem faced by an agent that must learn 
-through trial-and-error interactions with a dynamic environment. It is based on programming agents by reward and punishment without needing to specify how the task is to be achieved \cite{sutton1998}.
-\end{description}
-
-%%% Learning problem. Ac;a ma;ana, formulas y descripcion de tallada del problema de aprendizaje
-
-\section{Statistical learning theory} \label{sec:mltheory}
-
-All ML problems can be viewed as optimisation problems. The ML core task is to define a learning criterion, i.e the function to be optimised.
-
-In this thesis we will emphasised the supervised learning problem since it is the most common way of modelling financial problems.
-
-Supervised learning consists in to find a learning function $f: X \rightarrow Y$ which models a set of examples called training set $S$ which have been drawn randomly and independently according to a unknown joint distribution function $p(x,y)$ on $X \times Y$:
-
-$$
-S= \{(x_1,y_1),\,(x_2,y_2),\dots\,,\,(x_n,y_n)\} \in \mathcal{S}, (x_i,y_i) \in X\times Y \quad \forall i = 1\dots n
-$$
-
-\noindent where $X \subseteq \mathbb{R}^m$ and $Y = \mathbb{R}$ in case of regression problems and $Y = \{1,-1\}$  if it is a classification problem.
-
-A learning algorithm $\mathcal{A}$ takes as input a data set $S \in \mathcal{S}$ and output a function $f_S$:
-
-\begin{eqnarray*}
-\mathcal{A}: \mathcal{S} &\rightarrow & \mathcal{H} \\
-S &\rightarrow &\mathcal{A}(S) = f_S
-\end{eqnarray*}
-
-\noindent where $\mathcal{H}$ is called the hypothesis space, is the space of functions that the algorithm is allowed to search.
-The selection of $f_S$ is based on a loss function $V(f(x),y)$ which expected risk has to be minimised.
+\section{\uppercase{Background}}
+\label{sec:background}
+\noindent
+
+\subsection{Spurious Regression}
+
+Regression analysis is widely used to test economic theories. But sometimes
+regression is done between two variables, lets say $x_t$ and $y_t$, which are
+not causally related. This occurs at the presence of a third stochastic variable
+$z_t$ which is the true causal of the first two variables $x_t$ and $y_t$. The
+regression measure R-squared gives high values in presence of spurious
+relationships. This measure would means that the model explains all the
+variability of the response data around its mean which is not true.
+
+The R-squared measure is calculated as follows:
+
+\begin{equation}
+\label{eq:rsquared}
+R^2 = 1 - \frac{\sum_{t=1}^T \epsilon_t^2}{\sum_{t=1}^T(y_t - \bar{y})^2}
+\end{equation}
+
+In presence of spurious relationships the denominator of equation
+\ref{eq:rsquared} becomes very large because a large weight is placed on extreme
+observations on either side of the mean $\bar{y}$.
+In order to detect this, other statistic must be considered such as the
+Durbin-Watson statistic.
+
+% Acá agregar gráfico de spurious regression
+
+\subsection{Integration and Cointegration}
+A time series $\mathbf{y}$ is said to be integrated of order $d$ if after
+differentiating the variable $d$ times, we get an I(0) process, more precisely:
 
 \[
-E[V(f(x),y)] = \int V(f(x),y) dp(x,y)
+(1-L)^d \mathbf{y} \sim \text{I(0)} \, ,
 \]
 
-\noindent $V(f(x),y)$ denote the price paid for mistakes. Therefore, $V(f(x),y)=0$ if$f(x)=y$.
-
-For regression, the most common loss function is square loss or L2 loss function:
-\begin{equation}
-\label{eq:l2loss}
-V(f(x),y) = (f(x)-y)^2
-\end{equation}
-\noindent another option is the L1 loss
-\begin{equation}
-\label{eq:l1loss}
-V(f(x),y) = |f(x)-y| \, .
-\end{equation}
-\noindent the choice of loss function here gives rise to several well-known learning algorithms such as regularised least squares and support vector machines. %verificar esto
-
-Since the true distribution is unknown and only training samples are available, the objective is to estimate a function $\hat{f}$ through empirical risk (training error) minimisation (ERM):
-
-\begin{equation} 
-\label{eq:erm}
-R_{\text{emp}}[f] = \frac{1}{n} \sum_{i=1}^n V(f(x),y)
-\end{equation}
-
-\subsection{Learning algorithms}
-In all learning algorithms that are trained from example data, there is a tradeoff \cite{dietterich2003} between three factors:
-\begin{description}
-\item[Complexity of the hypothesis class] for best generalisation, we should adjust the complexity of the learner model $\hat{f}$ to the complexity of the data $f$.
-In polynomial regression, the complexity parameter is the order of the fitted polynomial, and therefore we need to find a way to choose the best order that minimises the generalisation error, that is, tune the complexity of the model to best fit the complexity of the function inherent in the data.
-\item[Generalisation accuracy on new examples] is the capability of the algorithm to generate the right output for an input instance outside the training set.
-\item[The amount of training data]  generally generalisation accuracy increases as the amount of training data increases.
-\end{description}
-
-Very complex models will fit the training data better (low bias) but it could overfit it and generalise poorly (high variance), this is also called the bias-variance tradeoff.
-Generalisation accuracy can sometimes be improved with more training data.
-The figure \ref{fig:tradeoff} illustrates how generalisation accuracy depends on the complexity of the model and the amount of training data.
-
-\begin{figure}[!h]
-  \centering
-  \includegraphics[width=0.8\textwidth]{img/3tradeoff}
-  \caption{Triple tradeoff in empirical learning}
-  \label{fig:tradeoff}
-\end{figure}
-
-\subsection{The bias-variance tradeoff} \label{sec:biasvar}
-
-Models learning error can be split into two main components: error due to bias and error due to variance.
-Bias measures how far off in general these models' predictions are from the correct value.
-Variance is taken as the variability of a model prediction for a given data point.
-
-If we have a learning model $\hat{f}$, its expected generalisation error on an unseen or testing sample $x$ can be decomposed as follows:
-
-\begin{align}
-\label{eq:bvtrade}
-\mathrm{E}\Big[\big(y - \hat{f}(x)\big)^2\Big]
- & = \mathrm{Bias}(\hat{f}(x))^2 + \mathrm{Var}(\hat{f}(x)) + \sigma^2
-\end{align}
-\noindent where:
-\begin{align*}
- \mathrm{Bias}(\hat{f}(x)) &= \mathrm{E}\big[\hat{f}(x)\big] - f(x) \\
- \mathrm{Var}(\hat{f}(x)) &= \mathrm{E}\Big[ \big( \hat{f}(x) - \mathrm{E}[\hat{f}(x)] \big)^2 \Big] 
-\end{align*}
-
-\textbf{Demo}\quad
-
-\begin{eqnarray*}
-E[(y-\hat{f}(\mathbf{x}))^2] & =&
-E[(y-f(\mathbf{x})+f(\mathbf{x})-\hat{f}(\mathbf{x}))^2] \\
-&=& E[(B-f(\mathbf{x}))^2] +
-E[(f(\mathbf{x})-\hat{f}(\mathbf{x}))^2] + \dots \\
-& &
-2\cancelto{0}{E[y-f(\mathbf{x})]}E[f(\mathbf{x})-\hat{f}(\mathbf{x})] \\
-&=& \sigma^2 + MSE(\hat{f}(\mathbf{x}))
-\end{eqnarray*}
-
-\noindent where
-
-\begin{eqnarray*}
-    MSE(\hat{f}(\mathbf{x})) &=&
-    E[f(\mathbf{x})-\hat{f}(\mathbf{x}))]^2 \\
-    &=& E[f(\mathbf{x})-E[\hat{f}(\mathbf{x})] +
-    E[\hat{f}(\mathbf{x})]-\hat{f}(\mathbf{x}))]^2 \\
-    &=&  E[(f(\mathbf{x})-E[\hat{f}(\mathbf{x})])^2] +
-      E[(\hat{f}(\mathbf{x})-E[\hat{f}(\mathbf{x})])^2] + \dots\\
-    & & E[(f(\mathbf{x})-E[\hat{f}(\mathbf{x})])] 
-    \cancelto{0}{E[(\hat{f}(\mathbf{x})-E[\hat{f}(\mathbf{x})])]} \\
-    &=& Bias(\hat{f}(\mathbf{x}))^2 + Var(\hat{f}(\mathbf{x}))
-\end{eqnarray*}
-
-$\blacksquare$
-
-Bias is introduced by the model selection.
-Therefore the model building process is repeated (through resampling) and substantially different averages of prediction values are obtained, bias will be high.
-The error due to variance is the amount by which the prediction, over one training set, differs from the expected predicted value, over all the training sets.
-Variance measures how inconsistent are the predictions from one another, over different training sets, not whether they are accurate or not.
-
-This generalisation error is unknown, but it can be estimated using a training sample, a testing sample or by estimating the model complexity.
-Figure \ref{fig:traintesterror} shows prediction error for training and testing sample.
-In general, in the training sample prediction error is decreasing with more complex models but it overfit the data and it is not a good model for the test data.
-Figure \ref{fig:bvtradeoff} shows how testing error can be decomposed into bias and variance components.
-The best model will be the one has a balance between bias and variance.
-
-\begin{figure}[!h]
-  \centering
-  \includegraphics[width=0.8\textwidth]{img/model_complexity}
-  \caption{Training and test error}
-  \label{fig:traintesterror}
-\end{figure}
-
-The objective is to simultaneously reduce bias and variance as much as possible in order to obtain as accurate model as is feasible.
-However, there is a tradeoff to be made when selecting models.
-Models that exhibit small variance and high bias underfit the truth target.
-Models that exhibit high variance and low bias overfit the truth target.
-
-This bias-variance tradeoff is the reason why in order to obtain the best model, data is usually broken in three subsets: training set, validation set and testing set.
-Training set is used to determine the model, validation set is used to estimate the generalisation error and finally testing set is used to estimate the accuracy of the model.
-Usually the partition is 50\% for training set, 25\% for validation and 25\% for testing purposes.
-This procedure can be extended to a Cross Validation (CV) procedure \cite{geisser1975}, usually used when the amount of data is limited.
-Various splitting strategies lead to various CV estimates.
-In K-fold cross-validation the training data is divided randomly into K distinct subsets, then the network is trained using K-1 subsets, and tested on the remaining subset.
-The process of training and testing is then repeated for each of the K possible choices of the subset omitted from the training.
-The average performance on the K omitted subsets is then our estimate of the generalisation performance.
-
-\begin{figure}[!h]
-  \centering
-  \includegraphics[width=0.8\textwidth]{img/biasvariancetradeoff}
-  \caption{Bias variance tradeoff}
-  \label{fig:bvtradeoff}
-\end{figure}
-
-Another way to control the complexity of the model is to include in the optimisation function not only the generalisation error but a penalisation of high complex models.
-
-\subsection{Regularization}
-Since ERM is an ill-posed problem, Tikhonov introduced a regularisation which ensures well-posedness and generalisation of ERM, i.e prevents overfit, by constraining the hypothesis space $\mathcal{H}$ usually called regularised ERM or Tikhonov regularisation.
-
-Tikhonov regularisation minimised over the hypothesis space $\mathcal{H}$ for a fixed positive parameter $\lambda$ the following:
-
-\begin{equation} 
-\label{eq:rerm}
-R_{\text{emp}}[f] = \frac{1}{n} \sum_{i=1}^n V(f(x),y) + \lambda \mathcal{R}(f)
-\end{equation}
-
-\noindent where $\mathcal{R}(f)$ is the regulariser, a penalisation on $f$.
-
-\subsection{Ridge Regression}
-
-One example of regularisation in linear models is Ridge Regression (RR), which is a regularised least squares method.
-The least squares (LS) method is a well known way to solve a regression
-problem. 
-LS method consists of minimising the sum of squared errors:
-
-\begin{eqnarray*}
-\label{eq:lsm}
- J(\mathbf{X}) =& \displaystyle \sum_{t=1}^N (f(\mathbf{x}_t)-y_t)^2 \\
- =&\displaystyle \sum_{t=1}^N (\mathbf{\mathbf{X}}^\top {\mathbf{x}}_t-y_t)^2 \\
- =& \| \mathbf{A}\mathbf{\mathbf{X}} - \mathbf{Y} \|_2^2 
-\end{eqnarray*}
-
-\noindent which is equivalent to find a solution to 
-\begin{equation}
-\label{eq:regproblem}
-\underset{m \times n}{\mathbf{A}} \enskip \underset{n \times
-l}{\mathbf{\mathbf{X}}} = \underset{m \times l}{\mathbf{Y}}
-\end{equation}
-
-The optimal solution for $\mathbf{\mathbf{X}}$ is:
+\noindent where I(0) is a stationary time series and $L$ is the lag operator:
+
+\[
+(1-L)\mathbf{y} = \Delta \mathbf{y}=\mathbf{y}_t  -\mathbf{y}_{t-1} \quad \forall t
+\]
+
+Let $\mathbf{y}_t = \{\mathbf{y}^1, \dots, \mathbf{y}^l\}$ be a set of $l$
+stationary time series I(1) which are said to be cointegrated if a vector
+$\beta=[\beta(1),\dots,\beta(l)]^\top \in \mathbb{R}^l$  exists such that the
+time series,
+
+\begin{equation}
+ \mathbf{Z}_t:= \beta^\top \mathbf{y}_t = \beta(1) \mathbf{y}^1 + \dots + \beta(l) \mathbf{y}^l \sim
+ \text{I(0)}\, .
+\end{equation}
+
+In other words, a set of I(1) variables is said to be cointegrated if
+a linear combination of them exists which is I(0).
+
+\subsection{Vector Autorregresive Models}\label{sec:varvec}
+
+VECM is a special case of VAR model and both describe the joint behaviour of a
+set of variables.
+
+VAR($p$) model is a general framework to describe the behaviour of a set of $l$
+endogenous variables as a linear combination of their last $p$ values. These
+$l$ variables at time $t$ are represented by the vector $\mathbf{y}_t$ as
+follows:
+
+\begin{equation}
+\label{eq:variables}
+\mathbf{y}_t = 
+\begin{bmatrix} y_{1,t} &
+y_{2,t} &
+\dots &
+y_{l,t}
+\end{bmatrix}^\top \, ,
+\end{equation}
+\noindent where $y_{j,t}$ corresponds to the time series $j$ evaluated at
+time $t$.
+
+The VAR(p) model describes the behaviour of a dependent variable in terms
+of its own lagged values and the lags of the others variables in the
+system. The model with $p$ lags is formulated as the following:
+
+\begin{equation}
+\label{eq:var}
+ \mathbf{y}_t = \phi_1 \mathbf{y}_{t-1}  + \dots +   \phi_p\mathbf{y}_{t-p}
++ \mathbf{c} + \mathbf{\epsilon}_t \, ,
+\end{equation}
+
+\noindent where $t=p+1, \dots N$,  ${\phi_1,\dots,\phi_p}$ are $l \times l$
+matrices of real coefficients,
+$\mathbf{\epsilon}_{p+1},\dots,\mathbf{\epsilon}_N$ are error terms,
+$\mathbf{c}$ is a constant vector and $N$ is the total number of samples.
+
+The VAR matrix form of equation~(\ref{eq:var}) is:
+
+%\begin{equation}\label{eq:vareq}
+%\mathbf{B} = \mathbf{A} \mathbf{X} + 
+%\mathbf{E} \, , 
+%\end{equation}
+%
+%\noindent where:
+%\begin{eqnarray}
+%\mathbf{A} &=&
+%\begin{bmatrix}
+%   \mathbf{y}_{p}     & \dots    & \mathbf{y}_{N-1}\\
+%   \mathbf{y}_{p-1}   & \dots    & \mathbf{y}_{N-2}\\
+%   \vdots             & \ddots   & \vdots\\
+%   \mathbf{y}_{1}     & \dots    & \mathbf{y}_{N-p}\\
+%   1                  & \dots    & 1 
+%   \end{bmatrix}^\top \, ,\\
+%\mathbf{B} &=&
+%\begin{bmatrix}
+%\quad \\
+%\mathbf{y}_{p+1} &
+%\dots & 
+%\mathbf{y}_N \\
+%\quad
+%\end{bmatrix}^\top  \, ,\\
+%\mathbf{X}&=&
+%\left[
+%\begin{array}{ccccc}
+%    \quad  & \quad & \quad & \quad \\
+%    \phi_1   & \cdots & \phi_p & \mathbf{c} \\  
+%    \quad & \quad & \quad & \quad
+%    \end{array}
+%\right]^\top \, ,\\
+%\mathbf{E} &=& 
+%\begin{bmatrix}
+%    \quad \\
+%    \mathbf{\epsilon}_{p+1}  & 
+%    \dots                & 
+%    \mathbf{\epsilon}_N \\
+%    \quad
+%\end{bmatrix}^\top \, . 
+%\end{eqnarray}
+
+\begin{equation}
+ \label{eq:vareq}
+               \underbrace{ \begin{bmatrix}
+               \quad \\
+               \mathbf{y}_{p+1} &
+               \mathbf{y}_{p+2} &
+               \dots & 
+               \mathbf{y}_N \\
+               \quad
+               \end{bmatrix}}_{\substack{ \mathbf{B}^\top\\l \times (N-p)}}   
+= 
+                \underbrace{\left[ 
+                \begin{array}{ccccc}
+                \quad & \quad & \quad & \quad & \quad \\
+                \phi_1  & \phi_2 & \cdots & \phi_p & \mathbf{c} \\  
+                \quad &\quad & \quad & \quad & \quad
+               \end{array} 
+               \right]}_{\substack{ \mathbf{X}^\top\\ l \times (l \times p + 1 )}}
+\underbrace{\begin{bmatrix}
+   \mathbf{y}_{p}  & \mathbf{y}_{p+1} & \dots    & \mathbf{y}_{N-1}\\
+   \mathbf{y}_{p-1}  & \mathbf{y}_{p} & \dots    & \mathbf{y}_{N-2}\\
+   \vdots        & \vdots   & \ddots   & \vdots\\
+   \mathbf{y}_{1} & \mathbf{y}_{2}   & \dots    & \mathbf{y}_{N-p}\\
+   1 & 1   & \dots    & 1 
+   \end{bmatrix}}_{\substack{ \mathbf{A}^\top\\ (l\times p +1 )\times (N-p)}}
++
+\underbrace{\begin{bmatrix}
+                \quad \\
+              \mathbf{\epsilon}_{p+1}  & 
+              \mathbf{\epsilon}_{p+2}  & 
+              \dots                & 
+              \mathbf{\epsilon}_N \\
+              \quad
+             \end{bmatrix}}_{\substack{\mathbf{E}^\top\\l \times (N-p) }} 
+\end{equation}
+Equation~(\ref{eq:vareq})  can be solved using ordinary least squares estimation.
+
+\subsection{Vector Error Correction}
+
+VECM is a special form of a VAR model for I(1) variables that are also
+cointegrated~\cite{banerjee1993}. It is obtained by replacing the form $\Delta
+\mathbf{y}_t = \mathbf{y}_t - \mathbf{y}_{t-1}$ in equation (\ref{eq:var}).
+VECM is expressed in terms of differences, has an error correction term and 
+the following form:
+
+\begin{equation}
+ \label{eq:vec}
+ \Delta \mathbf{y}_t = 
+ \underbrace{ \Omega\mathbf{y}_{t-1}}_\text{Error correction term} + 
+ \sum_{i=1}^{p-1}
+\phi_i^* \Delta \mathbf{y}_{t-i}  + \mathbf{c} + \mathbf{\epsilon}_t \quad ,
+\end{equation}
+
+\noindent where coefficients matrices $\Omega$ and $\phi_i^*$ are
+function of matrices $\phi_i$ (shown in equation (\ref{eq:var})) as follows:
+
+\begin{eqnarray*}
+\phi_i^* &: =& -\sum_{j=i+1}^{p} \phi_j \\
+\Omega &: =& -(\mathbb{I}-\phi_1-\dots-\phi_p) 
+\end{eqnarray*}
+
+The matrix $\Omega$ has the following properties~\cite{johansen1995}:
+\begin{itemize}
+\item If $\Omega = 0$ there is no cointegration
+\item If $rank(\Omega)=l$ i.e full rank, then the time series are not
+I(1) but stationary
+\item If $rank(\Omega)=r,\quad 0 < r < l$ then, there is cointegration
+and the matrix $\Omega$ can be expressed as $\Omega =
+\alpha \beta^\top$, where $\alpha$ and $\beta$ are $(l \times r)$
+matrices and $rank(\alpha)=rank(\beta)=r$.
+
+The columns of $\beta$ contains the cointegration vectors and the rows of
+$\alpha$ correspond with the adjusted vectors. $\beta$ is obtained by Johansen
+procedure~\cite{johansen1988} whereas $\alpha$ has to be determined as a
+variable in the VECM.
+
+It is worth noticing that the factorization of the matrix $\Omega$ is not
+unique since for any $r \times r$ nonsingular matrix $H$ we have:
+
+\begin{eqnarray*}
+\alpha \beta^\top &=& \alpha \mathbf{HH^{-1}} \beta^\top\\
+&=&(\alpha\mathbf{H})(\beta(\mathbf{H}^{-1})^\top)^\top \\
+&=& \alpha^*(\beta^*)^\top
+\end{eqnarray*}
+
+\noindent whith $\alpha^* = \alpha\mathbf{H}$ and $\beta^* =
+\beta(\mathbf{H}^{-1})^\top$.
+
+\end{itemize}
+
+If cointegration exists, then equation (\ref{eq:vec}) can be written as follows:
+
+\begin{equation}
+ \label{eq:vecfull}
+ \Delta \mathbf{y}_t = \alpha \beta^\top\mathbf{y}_{t-1} 
+ + \sum_{i=1}^{p-1} \phi_i^*\Delta
+\mathbf{y}_{t-i}  + \mathbf{c} + \mathbf{\epsilon}_t \quad ,
+\end{equation}
+
+\noindent which is a VAR model but for time series differences.
+
+VECM matricial form is the following:
+
+%\begin{eqnarray}\label{eq:vecmatrix}
+% \mathbf{A}&=&
+%   \begin{bmatrix} 
+%   \beta^\top \mathbf{y}_{p} & 
+%   \cdots & \beta^\top \mathbf{y}_{N-1} \\
+%   \Delta \mathbf{y}_p  & \cdots 
+%   &\Delta\mathbf{y}_{N-1} \\ 
+%   \vdots  & \ddots & \vdots \\
+%   \Delta\mathbf{y}_2  & \cdots 
+%   & \Delta \mathbf{y}_{N-p+1} \\
+%   1 & \cdots & 1 
+%   \end{bmatrix}^\top \label{eq:vecA} \, ,\\
+%\mathbf{B} & = &
+% \begin{bmatrix}
+% \quad\\
+%  \Delta \mathbf{y}_{p+1} & 
+%  \dots &
+%  \Delta \mathbf{y}_N \label{eq:vecB}\\
+%  \quad
+% \end{bmatrix}^\top \, ,\\
+%\mathbf{X}&=&
+%  \begin{bmatrix}
+%   \quad \\
+%   \alpha & \phi_1^* & \cdots & \phi_{p-1}^* & \mathbf{c} \\  
+%   \quad
+%   \end{bmatrix}^\top \, ,\\
+%\mathbf{E} &=&
+%\begin{bmatrix}
+%   \quad \\
+%   \mathbf{\epsilon}_{p+1} &
+%   \dots &
+%   \quad &\mathbf{\epsilon}_N \, \\ \quad
+%\end{bmatrix}^\top 
+%\end{eqnarray}
+%
+
+\begin{equation} \label{eq:vecmatrix}
+\underbrace{
+               \begin{bmatrix}
+               \quad\\
+                \mathbf{\Delta y}_{p+1} & 
+                \dots &
+                \mathbf{\Delta y}_N \\
+                \quad
+               \end{bmatrix}
+               }_{\substack{\mathbf{B}^\top\\ l \times (N-p) }} =
+   \underbrace{
+    \begin{bmatrix}
+     \quad \\
+     \alpha & \phi_1^* & \cdots & \phi_{p-1}^* & \mathbf{c} \\  
+     \quad
+     \end{bmatrix} 
+     }_{\substack{ \mathbf{X}^\top\\ l \times (l\times p +1)}}
+\underbrace{\begin{bmatrix} 
+   \beta^\top \mathbf{y}_{p} & 
+   \cdots & \beta^\top \mathbf{y}_{N-1} \\
+   \mathbf{\Delta y}_p  & \cdots 
+   & \mathbf{\Delta y}_{N-1} \\ 
+   \vdots  & \ddots & \vdots \\
+   \mathbf{\Delta y}_2  & \cdots 
+   & \mathbf{\Delta y}_{N-p+1} \\
+   1 & \cdots & 1 
+   \end{bmatrix}}_{\substack{\mathbf{A}^\top \\ (l \times p +1) \times (N-p) }}
++
+\underbrace{\begin{bmatrix}
+              \quad \\
+              \mathbf{\epsilon}_{p+1} &
+              \dots &
+              \quad &\mathbf{\epsilon}_N \\ \quad
+             \end{bmatrix}}_{\substack{\mathbf{E}^\top\\ l \times (N-p) }} 
+\end{equation}
+
+VAR and VECM parameters shown in equations (\ref{eq:vareq},\ref{eq:vecmatrix}) can be solved
+using standard regression techniques, such as ordinary least squares (OLS).
+
+\subsection{Unit Root Tests}
+In order to use VECM the time series are required to be I(1) which is detected
+using a unit root test. The Dickey-Fuller (ADF) test ~\cite{dickey1979} is a
+unit root test where the null hypothesis is that the
+variable has a unit root.
+
+If we have a variable with the form:
+\[
+\mathbf{x}_t = \alpha \mathbf{x}_{t-1} + \epsilon_t
+\]
+
+\noindent and if we substract $\mathbf{x}_{t-1}$ from both sides we get:
+
+\begin{equation}
+\label{eq:DF}
+\Delta \mathbf{x}_t = (\alpha-1) \mathbf{x}_{t-1} + \epsilon_t
+\end{equation}
+
+Dickey fuller test consist on determine if $\alpha$ equals 1. The alternative
+hypothesis is $\alpha < 1$ which means that $\mathbf{x}_t$ is a stationary
+process.
+
+In order to include more than 1 lagged variable the Augmented Dickey-Fuller
+test was developed and their tabulated critical values were obtained
+by~\cite{mackinnon2010}. The number of lags is often selected using Schwartz
+criterion.
+
+\subsection {Cointegration Tests}
+
+Johansen method~\cite{johansen1988} allows to find cointegration among several
+I(1) time series. This test is more general than the
+Engle-Granger~\cite{engle1987} test which only allow to find cointegration
+between two time series.  Johansen procedure takes the form of a
+likelihood-ratio test which determines $r$ cointegration relationships. The
+method provides two tests based on an eigenvector decomposition of the matrix
+$\Omega = \alpha \beta^\top$ on equation~\ref{eq:vec} called trace and maximal
+eigenvalue statistic.  Once the number of cointegration number is determined,
+$r$ more significative eigenvectors provided by Johansen procedure are used to
+get the cointegration vector matrix $\beta^\top$.  Critical values for both
+tests are limited to 12 and 5 variables in the trace and maximal eigenvalue
+statistic respectively.
+
+The trace test null hypothesis is $H_0: r = r_0$ against $H_1: r > r_0$ and the
+maximum eigenvalue test null hypothesis is $H_0: r = r_0$ against $H_1: r =
+r_0+1$ starting with $r_0 = 0$.  In the latest test, if we start with $r_0=0$
+and $H_0$ is rejected $H_1: r = 1$ will be accepted which is not true if there
+are more than one cointegration relationship, this is why this test is less
+powerful and the trace test is commonly used.
+
+\subsection{Ordinary Least Squares method}
+
+When $\mathbf{A}$ is singular, solution to equation~(\ref{eq:vecmatrix}) is given
+by the ordinary least squares (OLS) method. OLS consists of minimizing the sum
+of squared errors or equivalently minimizing the following expression:
+
+\begin{equation}
+\label{eq:regressionproblem}
+\underset{\mathbf{X}}{\text{min}} \quad \| \mathbf{A}\mathbf{\mathbf{X}} - \mathbf{B} \|_2^2
+\end{equation}
+
+\noindent for which the solution $\hat{\mathbf{X}}$ is well-known:
+
 \begin{equation}
 \label{eq:MP}
-\mathbf{\mathbf{X}}= \mathbf{A}^+ \mathbf{Y}
-\end{equation}
-
-\noindent where $\mathbf{A}^+$ is the Moore-Penrose pseudo-inverse
+\hat{\mathbf{X}}=\mathbf{A}^{\!\!+}\,\mathbf{B}
+\end{equation}
+
+\noindent where $\mathbf{A}^{\!\!+}$ is the Moore-Penrose pseudo-inverse
 which can be written as follows:
@@
 \label{eq:pseudoinverse}
-\mathbf{A}^+= (\mathbf{A}^\top \mathbf{A})^{-1}\mathbf{A}^\top \, .
-\end{equation}
-
-\textbf{Demo}\quad
-
-To solve equation~(\ref{eq:pseudoinverse}) is equivalent to solve the following
-optimisation problem:
-
-\begin{equation}
-\label{eq:RRproblem2}
-\underset{\mathbf{X}(\lambda)}{\text{min}} \quad \|
-\mathbf{C}\mathbf{\mathbf{X}} - \mathbf{F} \|_2^2
-\end{equation}
-
-\noindent where
-
-\begin{equation*}
-	\mathbf{C}=
-	\begin{bmatrix} \quad &\mathbf{A}& \quad \\ \hdashline \quad
-& \sqrt\lambda\mathbb{I}&\quad  
-\end{bmatrix} 
-\quad \text{and} \quad
-	\mathbf{F}=\begin{bmatrix} \quad &\mathbf{Y}& \quad \\ \hdashline \quad
-& \mathbf{0}&\quad  \end{bmatrix} \, .
-\end{equation*}
-
-Applying equation~(\ref{eq:MP}) and considering that $\mathbf{C}^\top
-\mathbf{C} = \mathbf{A}^\top \mathbf{A} + \lambda \mathbb{I}$ and 
-$\mathbf{C}^\top \mathbf{F}=\mathbf{A}^\top \mathbf{Y} $ we have:
-
-\begin{eqnarray*}
-\mathbf{X}(\lambda)&=&(\mathbf{C}^\top
-\mathbf{C})^{-1}\mathbf{C}^\top \mathbf{F} \\
-&=& (\mathbf{A}^\top \mathbf{A} + \lambda \mathbb{I})^{-1} \mathbf{A}^\top \mathbf{Y}
-\end{eqnarray*}
-
-$\blacksquare$
+\mathbf{A}^{\!\!+}= (\mathbf{A}^{\!\!\top} \mathbf{A})^{-1}\mathbf{A}^{\!\!\top} \, .
+\end{equation}
 
@@
     \underset{m \times k}{\mathbf{U_1}} \enskip
-    \underset{k \times k}{\mathbf{\Sigma_1}} \enskip
-    \underset{k \times n}{\mathbf{V_1}^\top}
+    \underset{k \times k}{\Sigma_1} \enskip
+    \underset{k \times n}{\mathbf{V}_1^{\top}} \, ,
 \end{equation}
@@
 \label{eq:pseudoinversesvd}
-\mathbf{A}^+ = \mathbf{V_1\Sigma_1^{-1}U_1^\top}
-\end{equation}
-
+\mathbf{A}^{\!\!+} = \mathbf{V}_1 \Sigma_1^{-1} \mathbf{U}_1^\top \, .
+\end{equation}
+
+%Descomentar en la tesis
 \textbf{Demo}\quad
 
-In the case matrix $\mathbf{A}$ is non singular, its solution is
-straight forward:
-
-\begin{equation}
-\label{OLSsolution}
-    \mathbf{\mathbf{X}}=\mathbf{A}^{-1}\mathbf{Y}
-\end{equation}
-
-Since the problem shown in equation~(\ref{eq:regproblem}) has not
-solution, the minimum norm given by equation~(\ref{OLSsolution}) is
-obtained by solving the equivalent problem:
+Since $\mathbf{A}$ is singular the problem shown in
+equation~(\ref{eq:regressionproblem}) has not solution, the minimum norm given
+by equation~(\ref{eq:MP}) is obtained by solving the equivalent problem:
 
@@
 \label{eq:proyectorsol}
-\mathbf{A \hat{\mathbf{X}} = PY} 
+\mathbf{A \hat{X} = PB} 
 \end{equation*}
@@
 \underset{(n \times k)}{\mathbf{V_2}}]$ and $\mathbf{V_1^\top V_2 =
-0}$ we can express $\mathbf{\hat{\mathbf{X}}} = \mathbf{V_1 \mathbf{X}_1 + V_2 \mathbf{X}_2}$
-with $\mathbf{\mathbf{X}_2=0}$ because $\mathbf{\hat{\mathbf{X}}}$ lives in the
+0}$ we can express $\mathbf{\hat{X}} = \mathbf{V_1 x_1 + V_2 x_2}$
+with $\mathbf{x_2=0}$ because $\mathbf{\hat{X}}$ lives in the
 $\text{Row}(\mathbf{A})$ given by $\mathbf{V_1}$, so we have:
@@
 \begin{eqnarray*}
-\mathbf{A \hat{\mathbf{X}}} &=& \mathbf{PY} \\
-\mathbf{U_1 \Sigma_1 V_1^\top \hat{\mathbf{X}}} &=& \mathbf{U_1 U_1^\top Y} \\
-\mathbf{ V_1^\top \hat{\mathbf{X}}} &=&  \mathbf{\Sigma_1^{-1} U_1^\top Y} \\ 
-\mathbf{ V_1^\top V_1 \mathbf{X}_1} &=& \mathbf{\Sigma_1^{-1}
-U_1^\top Y} \\
-\mathbf{\mathbf{X}_1}&=& \mathbf{\Sigma_1^{-1} U_1^\top Y}
-\end{eqnarray*}
-
-\noindent from this result we can obtain $\mathbf{\hat{\mathbf{X}}}$ and
+\mathbf{A \hat{X}} &=& \mathbf{PB} \\
+\mathbf{U_1 \Sigma_1 V_1^\top \hat{X}} &=& \mathbf{U_1 U_1^\top B} \\
+\mathbf{ V_1^\top \hat{X}} &=&  \mathbf{\Sigma_1^{-1} U_1^\top B} \\ 
+\mathbf{ V_1^\top V_1 x_1} &=& \mathbf{\Sigma_1^{-1}
+U_1^\top B} \\
+\mathbf{x_1}&=& \mathbf{\Sigma_1^{-1} U_1^\top B}
+\end{eqnarray*}
+
+\noindent from this result we can obtain $\mathbf{\hat{X}}$ and
 therefore the pseudo-inverse expression:
@@
 \begin{eqnarray*}
-\mathbf{\hat{\mathbf{X}}} &=& \mathbf{V_1 \mathbf{X}_1} \\
-                &=& \mathbf{V_1 \Sigma_1^{-1} U_1^\top Y} \\
+\mathbf{\hat{X}} &=& \mathbf{V_1 x_1} \\
+                &=& \mathbf{V_1 \Sigma_1^{-1} U_1^\top B} \\
 \mathbf{A^+} &=& \mathbf{V_1 \Sigma_1^{-1} U_1^\top} \, .
@@
 
-$\blacksquare$
-
-In order to avoid the singularity of the matrix $\mathbf{A}^\top \mathbf{A}$, a regularisation term is introduced:
+\subsection{Ridge Regression}\label{sec:RR}
+
+Ridge regression (RR) introduces a ridge estimation to avoid the
+singularity of the matrix $\mathbf{A}^\top \mathbf{A}$
+equation~(\ref{eq:pseudoinverse}). RR solves the following
+optimization problem:
 
@@
 \label{eq:RRproblem}
-J(\mathbf{\mathbf{X}}) =  \| \mathbf{A}\mathbf{\mathbf{X}} - \mathbf{Y} \|_2^2  + \lambda
- \| \mathbf{\mathbf{X}}\| ^2
-\end{equation}
-
-\noindent which optimal solution $\mathbf{\mathbf{X}}_*$ is well known:
+\underset{\mathbf{X}}{\text{min}} \quad \|
+\mathbf{A}\mathbf{\mathbf{X}} - \mathbf{B} \|_2^2 +\lambda \|
+\mathbf{\mathbf{X}}\|_2^2 
+\end{equation}
+
+\noindent where $\lambda$ is a regularization parameter.
+
+The RR optimal solution $\mathbf{X}(\lambda)$ is also well-known:
+
+\begin{equation}
+\label{eq:optsolRR}
+\mathbf{X}(\lambda)=(\mathbf{A}^\top \mathbf{A}+ \lambda
+\mathbb{I})^{-1}\mathbf{A}^\top \mathbf{B} \, . 
+\end{equation}
+
+\textbf{Demo}\quad
+
+To solve equation~(\ref{eq:optsolRR}) is equivalent to solve the following
+optimization problem:
+
+\begin{equation}
+\label{eq:RRproblem2}
+\underset{\mathbf{X}(\lambda)}{\text{min}} \quad \|
+\mathbf{C}\mathbf{\mathbf{X}} - \mathbf{F} \|_2^2
+\end{equation}
+
+\noindent where
 
 \begin{equation*}
-\label{eq:optsolRR}
-\mathbf{\mathbf{X}}_*=(\mathbf{A}^\top \mathbf{A}+\lambda \mathbb{I})^{-1}\mathbf{A}^\top y \, ,
+	\mathbf{C}=\begin{bmatrix} \quad &\mathbf{A}& \quad \\ \hdashline \quad
+& \lambda\mathbb{I}&\quad  \end{bmatrix} \quad \text{and} \quad
+	\mathbf{F}=\begin{bmatrix} \quad &\mathbf{B}& \quad \\ \hdashline \quad
+& \mathbf{0}&\quad  \end{bmatrix} \, .
 \end{equation*}
 
-\textbf{Demo}\quad
-
-\begin{eqnarray*}
-J(\mathbf{\mathbf{X}}) &=&  \| \mathbf{A}\mathbf{\mathbf{X}} - \mathbf{Y} \|_2^2  + \lambda
- \| \mathbf{\mathbf{X}}\| ^2 \\
- &=&  \sum_{t=1}^N (\mathbf{\mathbf{X}}^\top {\bf a}_t-y_t)^2 + \lambda \sum_{i=1}^p \mathbf{X}_i^2 \\
- &=& (\mathbf{X}^\top a_1-y_1)^2 + \dots + (\mathbf{X}^\top a_N-y_N)^2 + \lambda (\mathbf{X}_1^2+\dots+\mathbf{X}_p^2)
-\end{eqnarray*}
-\noindent taking derivatives
- \begin{eqnarray*}
- \frac{\partial J(\mathbf{\mathbf{X}})}{\partial \mathbf{X}_1}&=& 
- 2(\mathbf{X}^\top \mathbf{a}_1-y_1)\mathbf{a}_{11} + \dots + 2(\mathbf{X}^\top \mathbf{a}_N-y_N)\mathbf{a}_{N1} + 2\lambda \mathbf{X}_1 \\
- &=& 2\mathbf{a}_1^\top(\mathbf{A}\mathbf{X}-\mathbf{Y}) + 2\lambda\mathbf{X}_1\\
-& \vdots &\\
-  \frac{\partial J(\mathbf{\mathbf{X}})}{\partial \mathbf{X}_p}&=& 
- 2\mathbf{a}_p^\top(\mathbf{A}\mathbf{X}-\mathbf{Y}) + 2\lambda\mathbf{X}_p 
-\end{eqnarray*}
-
-Then we have that:
-\begin{eqnarray*}
-\frac{\partial J(\mathbf{\mathbf{X}})}{\partial \mathbf{X}}&=& 
- 2\mathbf{A}^\top(\mathbf{A}\mathbf{X}-\mathbf{Y}) + 2\lambda\mathbf{X}
-\end{eqnarray*}
-Since $\frac{\partial J(\mathbf{\mathbf{X}})}{\partial \mathbf{X}}=0$ we have:
-\begin{eqnarray*}
-2\mathbf{A}^\top(\mathbf{A}\mathbf{X}-\mathbf{Y}) + 2\lambda\mathbf{X}&=&0 \\
-\mathbf{A}^\top\mathbf{A}\mathbf{X} - \mathbf{A}^\top\mathbf{Y} + \lambda\mathbf{X} &=& 0\\
-(\mathbf{A}^\top\mathbf{A}+\lambda\mathbb{I})\mathbf{X} &=&  \mathbf{A}^\top\mathbf{Y} \\
-\mathbf{X} &=& (\mathbf{A}^\top\mathbf{A}+\lambda\mathbb{I})^{-1}  \mathbf{A}^\top\mathbf{Y}
-\end{eqnarray*}
-
-$\blacksquare$
-
-\subsection{The Lambda parameter}
-
-The additional term $\lambda \|\mathbf{\mathbf{X}}\|_2^2$ in the optimisation
-problem shown in equation~(\ref{eq:RRproblem}) has two effects on the solution:
-shrinks the coefficients towards zero and improves the conditioning of the
-problem.
-
-When $\mathbf{A}$ is orthonormal then $\mathbf{A}^\top \mathbf{A} =\mathbb{I}$ and there is a simple relation between the ridge estimator and the OLS estimator:
-\begin{eqnarray*}
-\mathbf{X}_* (\lambda) &=& (\mathbf{A}^\top \mathbf{A}+\lambda \mathbb{I})^{-1}\mathbf{A}^\top \mathbf{Y} \\
- &=& (\mathbb{I} + \lambda \mathbb{I})^{-1} \mathbf{A}^\top \mathbf{Y} \\
- &=&(1+\lambda)^{-1} \mathbb{I} \mathbf{A}^\top \mathbf{Y} \\
- &=&(1+\lambda)^{-1} (\mathbf{A}^\top \mathbf{A})^{-1}\mathbf{A}^\top \mathbf{Y} \\
- &=&(1+\lambda)^{-1} \mathbf{X}
-\end{eqnarray*}
-
-Figure~\ref{fig:shrinks} shows a visual example of the shrinking of the
-coefficients:
-
-\begin{figure}[h!]
-\includegraphics[width=0.5\linewidth]{img/shrinks}
-\caption{Shrink of regression coefficients}
-\label{fig:shrinks}
-\end{figure}
-
-On the other hand, the effect of adding the term $\lambda \mathbb{I}$
-to the matrix $\mathbf{A}^\top \mathbf{A}$
-(equation~(\ref{eq:optsolRR})) improves its condition number since it
-increases its diagonal values when $\lambda > 0 $.  The matrix
-$\mathbf{A}^\top \mathbf{A}$ is symmetrical ($(\mathbf{A}^\top
-\mathbf{A})^\top = \mathbf{A}^\top \mathbf{A}$) and therefore
-diagonalizable.  If we know the eigenvalue decomposition of $\mathbf{A}
-= \mathbf{U\Sigma V^\top}$, then:
-
-\begin{eqnarray*}
-\mathbf{A}^\top \mathbf{A}+\lambda \mathbb{I}&=&\mathbf{V\Sigma^2
-V}^\top + \lambda \mathbf{V} \mathbf{V}^\top\\ &=&\mathbf{V}
-(\mathbf{\Sigma}^2+\lambda\mathbb{I}) \mathbf{V}^\top \, ,
-\end{eqnarray*}
-
-\noindent where
-
-\begin{equation*}
-\mathbf{\Sigma}^2+\lambda\mathbb{I}=
-\begin{bmatrix}
-\sigma^2_1 + \lambda & \, & \, \\
-\, & \sigma^2_2 +\lambda & \, \\
-\, & \, & \ddots & \, \\
-\, & \, & \, & \sigma^2_n +\lambda \, .
-\end{bmatrix}
-\end{equation*}
-
-where $\sigma_1 \geq \sigma_2 \geq \dots \geq \sigma_n$.
-
-Since the condition number of a matrix $\mathbf{A}$ is defined as:
-
-\begin{equation*}
-	\kappa = \|\mathbf{A}\| \|\mathbf{A}^{-1}\|
-\end{equation*}
-
-If matrix $\mathbf{A}$ is non-singular, its condition number can be
-expressed in terms of its singular values. The effect of adding the
-regularization term affects the condition number as follows:
-
-\begin{eqnarray*}
-\kappa_{ols} &=& \|\mathbf{A}\| \|\mathbf{A}^{-1}\|=\frac{\sigma_1}{\sigma_n} \\
-\kappa_{ridge} &=& \|\mathbf{A}^\intercal \mathbf{A} + \lambda \mathbb{I}\| 
-\|(\mathbf{A}^\intercal \mathbf{A} + \lambda \mathbb{I})^{-1}\|=\frac{\sigma_1+\lambda}{\sigma_n + \lambda} \,
-\end{eqnarray*}
-
-It is easy to see that the term $\lambda$ improves the condition number:
-
-\begin{equation*}
-        \frac{\sigma_1+\lambda}{\sigma_n + \lambda} <
-        \frac{\sigma_1}{\sigma_n} \,  \qquad \forall \quad \lambda > 0
-\end{equation*}
-
-However, $\lambda$ cannot be too large. Tipically $\lambda$ is small and its
-magnitude depends on the matrix $\mathbf{A}$.
-
-For rank deficient matrices we know that $\text{det}(\mathbf{A A^\top})=0$, adding the
-term $\lambda \mathbb{I}$ we have that $\text{det}(\mathbf{A A^\top}+\lambda \mathbb{I}) =
-p(\lambda)$ where $p(\lambda)$ is a polynomial of degree $n$ ($\mathbf{A}$ is
-$m \times n$). The zeros of $p(\lambda)$ are discretes, so it can be represented
-as:
-
-\[
-p(\lambda) =
-\lambda(\lambda-\lambda_1)^{n_1}(\lambda-\lambda_2)^{n_2}\dots(\lambda-\lambda_s)^{n_s}
-\]
-
-\noindent where $n_1 + n_2 + \dots + n_s = n$.
-
-This means that $\lambda$ must be small in order to ensure that $p(\lambda)$
-does not vanish.
-
-\subsection{Selection of Lambda}
-
-One of the ways to determine parameter $\lambda$ is using the bias-variance tradeoff (see section \ref{sec:biasvar}). This parameter is crucial for ridge regression since it could
-reduce the expected prediction error by reducing variance, considering a biased
-estimator. 
-It is known that the prediction error can be express as a decomposition between bias and variance. The prediction error is
-obtained as:
-
-\begin{equation}
- \label{eq:prederror}
-E[(Y-\hat{f}(\mathbf{X}))^2] = \sigma^2 + Bias(\hat{f}(\mathbf{X}))^2 + Var(\hat{f}(\mathbf{X}))
-\end{equation}
-
-\noindent where $\hat{f}(\mathbf{X})=\mathbf{AX}(\lambda)$ for RR and
-$\hat{f}(\mathbf{X})=\mathbf{AX}$ in the case of OLS.
-
-For ridge regression we will use the following notation: $\hat{f}(\mathbf{X})=\mathbf{AX}(\lambda)$ and
-for OLS $f(\mathbf{X})=\mathbf{AX}$. The solution of OLS is well known $\hat{\mathbf{X}}=(\mathbf{A}^\top \mathbf{A})^{-1}\mathbf{A}^\top \mathbf{Y}$, and its bias is:
-
-\begin{eqnarray*}
-Bias(\hat{\mathbf{X}}) &=& E[\hat{\mathbf{X}}] - \mathbf{X} \\
-&=& E[ (\mathbf{A}^\top \mathbf{A})^{-1}\mathbf{A}^\top \mathbf{Y}] - \mathbf{X} \\
-&=& E[ (\mathbf{A}^\top \mathbf{A})^{-1}\mathbf{A}^\top (\mathbf{AX})] - \mathbf{X}  \\
-&=& \mathbf{X}  - \mathbf{X}  \\
-&=&  0
-\end{eqnarray*}
-
-The bias of ridge regression when $\mathbf{A A^\top}$ is non-singular
-can be obtained expressing ridge regression solution
-$\mathbf{\lambda}$ in terms of OLS solution $\hat{\mathbf{X}}$:
-
-\begin{eqnarray*}
-\mathbf{X}(\lambda) &=&( \mathbf{A}^\top \mathbf{A} + \lambda \mathbb{I})^{-1}\mathbf{A}^\top \mathbf{Y} \\
-&=& (\mathbb{I} + \lambda (\mathbf{A}^\top \mathbf{A})^{-1})^{-1} (\mathbf{A}^\top \mathbf{A})^{-1}\mathbf{A}^\top \mathbf{Y} \\
-&=&  (\mathbb{I} + \lambda (\mathbf{A}^\top \mathbf{A})^{-1})^{-1}  \hat{\mathbf{X}} \\
-&=& \mathbf{W} \hat{\mathbf{X}} 
-\end{eqnarray*}
-
-\noindent where $\mathbf{W}  = (\mathbb{I} + \lambda (\mathbf{A}^\top
-\mathbf{A})^{-1})^{-1}  $ it is defined for simplicity. Ridge
-regression bias is then obtained as:
-
-\begin{eqnarray*}
-Bias(\mathbf{X}(\lambda)) &=& E[\mathbf{X}(\lambda)] - \mathbf{X} \\
-&=& E[\mathbf{W}\hat{\mathbf{X}}] - \mathbf{X} \\
-&=&  \mathbf{W} \mathbf{X} - \mathbf{X} \neq 0 
-\end{eqnarray*}
-
-\begin{figure}[h!]
-\includegraphics[width=\linewidth]{img/biasvariance}
-\caption{Bias-variance tradeoff depending on lambda}
-\label{fig:biasvariance}
-\end{figure}
-
-The variance of OLS is:
-
-\begin{equation*}
-Var(\hat{\mathbf{X}}) = \sigma^2 (\mathbf{A}^\top \mathbf{A} )^{-1}
-\end{equation*}
-
-\noindent and the variance of ridge regression is:
-
-\begin{eqnarray*}
-Var(\mathbf{X}(\lambda)) &=& Var(\mathbf{W}\hat{\mathbf{X}}) \\
-&=& E[(\mathbf{W}\hat{\mathbf{X}}-E[\mathbf{W}\hat{\mathbf{X}}])(\mathbf{W}\hat{\mathbf{X}}-E[\mathbf{W}\hat{\mathbf{X}}])^\top] \\
-&=& \mathbf{W}E[(\hat{\mathbf{X}}-E[\hat{\mathbf{X}}])(\hat{\mathbf{X}}-E[\hat{\mathbf{X}}])^\top] \mathbf{W}^\top \\
-&=& \mathbf{W}Var(\hat{\mathbf{X}})\mathbf{W}^\top \\
-&=& \sigma^2 \mathbf{W}(\mathbf{A}^\top \mathbf{A} )^{-1}\mathbf{W}^\top
-\end{eqnarray*}
-
-The figure~\ref{fig:biasvariance} shows the bias-variance tradeoff
-given by equation~(\ref{eq:prederror}). Despite OLS has zero bias,
-its variance is greater than ridge for small values of $\lambda$.
-It can be shown that, in terms of prediction error, ridge (black line)
-is lower than OLS (dotted line)~\cite{hoerl1970}. Ridge regression shows an
-increasing squared bias and a decreasing variance.
-
-\textbf{Demo}\quad
-
-Since $Bias(\mathbf{X}(\lambda)) \neq 0$ this imply that
-
-\begin{equation*}
-Bias(\mathbf{X}(\lambda)) ^2 > 0 
-\end{equation*}
-
-\noindent we know that $\mathbf{A}^\top
-\mathbf{A}$ has an eigenvalue decomposition $\mathbf{A}^\top
-\mathbf{A} = \mathbf{V} \Sigma \mathbf{V}^{-1}$.
-
-\begin{eqnarray*}
-Var(\mathbf{X}(\lambda)) &=& \sigma^2 \mathbf{W}(\mathbf{A}^\top \mathbf{A} )^{-1}\mathbf{W}^\top\\
-Var(\mathbf{X}(\lambda)) &=& \sigma^2 (\mathbb{I} + \lambda (\mathbf{A}^\top
-\mathbf{A})^{-1})^{-1} (\mathbf{A}^\top \mathbf{A} )^{-1}((\mathbb{I} + \lambda (\mathbf{A}^\top
-\mathbf{A})^{-1})^{-1} )^\top \\
-&=& \sigma^2 (\mathbb{I} + \lambda (\mathbf{V} \Sigma \mathbf{V}^{-1})^{-1} (\mathbf{V} \Sigma \mathbf{V}^{-1})^{-1}((\mathbb{I} + \lambda (\mathbf{V} \Sigma \mathbf{V}^{-1})^{-1})^{-1} )^\top \\
-&=& \sigma^2 \mathbf{V} (\mathbb{I} + \lambda \Sigma^{-1})^{-1} \Sigma^{-1} (\mathbb{I} + \lambda \Sigma^{-1})^{-1}  \mathbf{V}^{-1}\\
-&=& \sigma^2 \mathbf{V} \mathbf{D}^{-1} \mathbf{V}^{-1}
-\end{eqnarray*}
-
-\noindent where $\mathbf{D}^{-1}  = (\mathbb{I} + \lambda \Sigma^{-1})^{-1} \Sigma^{-1} (\mathbb{I} + \lambda \Sigma^{-1})^{-1}$ is a diagonal matrix with coefficients:
-
-\begin{equation*}
-d_i = \frac{\sigma_i^{-1}}{(1+\lambda\sigma_i^{-1})^2}
-\end{equation*}
-
-\subsection{Machine learning algorithms}
-In recent years many successful machine learning applications have been developed.
-Artificial neural networks (ANN) and Support Vector Machines (SVM) have been some of the most popularly used machine learning algorithm \cite{haykin1998}.
-Historically, SVMs emerged after ANN.
-
-The main characteristics of neural networks are that they have the ability to learn
-complex nonlinear input-output relationships, they use sequential training procedures 
-they adapt themselves to the data.
-
-ANN is inspired by biological learning systems.
-ANN are organised into layers and have unidirectional connections between the layers, feed-forward networks  (FFN) are the most used.
-FFN consist of a series of layers.
-The first layer has a connection from the network input.
-Each subsequent layer has a connection from the previous layer.
-The final layer produces the network's output.
-Figure \ref{fig:ffn} shows a FFN with its different layers.
-
-\begin{figure}[!h]
-  \centering
-  \includegraphics[width=0.8\textwidth]{img/ffn}
-  \caption{Feed-forward neural network}
-  \label{fig:ffn}
-\end{figure}
-
-Support Vector Machine (SVM) was introduced in 1992 by Vapnik and his coworkers
-\cite{boser1992}.
-In its original form, SVM is a training algorithm for linear classification.
-Only later it was used for regression, principal component analysis, novelty detection and also for non-linear case.
-However, unlike ANN, it is very well
-based on theory in statistical learning \cite{cortes1995}.  SVM tunes the capacity of 
-maximising the margin between the training patterns and the decision boundary. The
-solution is expressed as a linear combination of supporting patterns, which are the 
-of training patterns close to the decision boundary, called the support vectors.
-
-The major difference between SVM and ANN is in the error optimisation. In ANN, the aim of learning is to obtain a set of weight values which minimise the training error 
-minimum while training adjust the capacity of the machine.
-
-For nonlinear case, SVM mapped the data sets of input space into a higher dimensional feature space, which is linear and the large margin learning algorithm is then applied.
-However, the mapping can be implicitly done by kernel functions.
-In the high dimensional feature space, simpler and linear hyper plane classifiers that have maximal margin between the classes can be obtained.
-
-%\subsection{Kernel methods}
-%Kernel functions may be used to introduce a nonlinear separation
-%in input space despite the linear classifier in decision space
-\newpage
-\section{Online learning} \label{sec:onoffline}
-
-Online learning is a supervised machine learning framework that is useful when
-we have sequential access to a sample only once.  This differs from the
-classical batch learning, where there is an entire data set available and the
-learner can build the internal model without any limits in accessing the data.
-In batch learning, there is time enough to carefully analyse the dataset, build
-large predictive models and combine them in a sophisticated way. However, when
-execution time is critical and new data is required to be included in the model,
-online learning algorithms are a better option.
-
-In batch learning, the training phase is commonly a computationally expensive
-process. Therefore, when new data arrives it can't be included easily to the
-model. Moreover, it could happen that we won't have enough time to process the
-new data before more data arrives. In online learning algorithms there is no
-training phase, but the model is updated and evaluated at every time step. This
-model updating is computationally less expensive than a training phase.
-Online algorithms allow incremental learning by processing one instance at a
-time. This is done updating the current model instead of building the model from
-scratch.
-
-Online learning is also useful when some past data may be irrelevant
-or we want to improve computational efficiency. However, the use of less
-historical data could affect accuracy.
-
-The goal of online learning is the same as batch learning, predicting targets as
-accurate as possible. For example, stock market prediction can be seen as online
-learning. The algorithm makes a prediction of the stock, little time after the
-real stock price is available, this information can be incorporated to the
-learner to further improve the prediction accuracy. In general, there is too
-much data available in an online learning setup, the data set grows
-continuously. Offline learning has equal or superior accuracy compared to online
-learning when the same amount of data is used.
-
-Classic statistical theory of sequential prediction enforces strong assumptions
-on the statistical properties of the input sequence (for example, stationary
-stochastic process). However, these assumptions can be unknown or change over
-time. In online learning there is no previous assumption about the data and the
-sequence is allowed to be deterministic, stochastic or even adaptive.
-
-Moreover, in case we receive data streams, ANN or SVM cannot introduce new
-information into the model without a re-training process, so we will have to use
-the same non-updated model until we decide to compute another one if it is
-possible.  Online learning algorithms allow one example at a time to be
-introduced into an existing model incrementally \cite{vovk2005}. This is
-extremely important when the problem has large data streams and real-time
-forecasting must be done.  This is the most common scenario when we want to
-forecast a wide range of data such as stock prices and volatilities, electricity
-power, intrusion detection, web-mining, server load, etc.  Besides, many
-problems of high interest in machine learning can be treated as online ones and
-they can also use these types of algorithms.
-
-The online learning framework was first introduced in the perceptron algorithm
-\cite{rosenblatt58}. There are several other widely used online methods such as
-passive-aggressive \cite{crammerETall2006}, stochastic gradient descent
-\cite{zhang2004}, aggregating algorithm \cite{vovk2001} and the second order
-perceptron \cite{cesa-bianchi2005}.  In \cite{cesa-bianchi2006} an in-depth
-analysis of online learning is provided.
-
-The motivation for online learning is to obtain computational efficiency and
-tackle the shifting problem, i.e. that the distribution of the data is unknown
-or changes over time. Online learning algorithms can deal with this problem
-because they have a tracking ability which is a strategy based on retaining weak
-dependence on past examples by using two types of models:
-
-\textit{a)} \textbf{memory boundedness:} consists of limiting the number of
-support vectors in order to improve computational efficiency. One example of
-this is the budget perceptron~\cite{crammeretal2004} which reduces the number of
-examples used for prediction. Alternatively, in the forgetron
-algorithm~\cite{dekeletal2008} the damage caused by removing old examples is
-discussed, which can be avoided by removing samples with small influences. Other
-examples are the sliding window kernel (RLS)~\cite{vanvaerenberghetal2006},
-which only considers a sliding window of the most recent data, and in
-\cite{arce+salinas2012} is shown a variant of aggregating algorithm for
-regression~\cite{vovk2001} considering only a sliding window of the most recent
-data, optimising also common matrix operations.
-
-\textit{b)} \textbf{weight decay:} one example of this is the shifting
-perceptron algorithm which implements an exponential decaying scheme for the
-examples~\cite{cavallantietal2007}.  Performance of an online learning algorithm
-is measured by the cumulative loss it suffers along its run on a sequence of
-examples. In order to minimise this loss, the learner may update the hypothesis
-after each round so as to be more accurate in later rounds.
-
-There is sometimes confusion about online and incremental learning concepts.
-Incremental learning refers to any online learning process that learns the same
-model as would be learnt by a batch learning algorithm.
-
-Incremental learning is useful when the input to a learning process is stream
-data, with the need or desire to be able to use the result of learning at any
-point in time, based on the input observations received so far.
-
-In principle, the stream of observations may be infinitely long, or the next
-observation long delayed, precluding any hope of waiting until all the
-observations have been received. One would rather not simply accumulate and
-store all the inputs and, upon receipt of each new one, apply a batch learning
-algorithm to the entire sequence of inputs received so far. It would be
-preferable computationally if the existing hypothesis or other artefact of
-learning could be updated in response to each newly received input observation.
-
-Incremental learning is very useful when there is no need to record fundamental
-data and only a summary needs to be retained. Due to this, incremental
-algorithms are often characterised as memoryless, because no memory of past data
-is required.  The algorithm is online but not incremental if it doesn't produce
-the same result for all observations that the corresponding batch algorithm
-would for these same observations.
-
-Algorithm~\ref{alg:onlinealg} shows the online learning algorithm structure:
-
-\begin{algorithm}[ht]
-\begin{algorithmic}[1]
-    \STATE Receives input $\mathbf{x}_t$
-    \STATE Makes prediction $\mathbf{\hat{y}}_t$
-    \STATE Receives response $\mathbf{y}_t$
-    \STATE Incurs loss $l_t(\mathbf{y}_t,\mathbf{\hat{y}}_t)$
-\end{algorithmic}
-\caption{Structure of a Learning System}
-\label{alg:onlinealg}
-\end{algorithm}
-
-\noindent where $l$ is some loss function. Performance is later measured after
-$T$ trials as:
-
-\begin{equation*}
-L_T = \sum_{t=1}^T l_t(\mathbf{y}_t,\mathbf{\hat{y}}_t)
-\end{equation*}
-
-The objective is minimize this loss function for all instances.
-The quality of online learning algorithms is measured by a quantity known as
-regret which is the difference between the performance of the online algorithm
-and its optimal predictor $E^* \in \Theta$ given by:
-
-\begin{equation*}
-L^*_T= \text{min}_{E \in \Theta} L_T^E \, ,
-\end{equation*}
-
-\noindent where $L_T^E = \sum_{t=1}^T
-l_t(\mathbf{y}_t,\mathbf{y}^E_t)$ and $\mathbf{y}^E_t$ is the expert estimation.
-
-Therefore regret is defined as:
-
-\begin{equation*}
-R_T = L_T - L^*_T
-\end{equation*}
-
-\subsection{Competitive analysis}
-
-Competitive analysis was designed for analysing the performance of an online algorithm compared with its optimal offline algorithm.
-An optimal offline algorithm can view the sequences of requests in advance.
-The effectiveness of an online algorithm \cite{sleator1985} may be measured by its competitive ratio which is the worst-case ratio of the online algorithm and the the optimal offline algorithm.
-
-\subsection{Online Ridge Regression}
-
-Online RR is the online formulation of the regularised Least Squares method
-and is based on the following equivalent formulation of the RR optimal solution.
-
-Since 
+Applying equation~(\ref{eq:MP}) and considering that $\mathbf{C}^\top
+\mathbf{C} = \mathbf{A}^\top \mathbf{A} + \lambda \mathbb{I}$ and 
+$\mathbf{C}^\top \mathbf{F}=\mathbf{A}^\top \mathbf{B} $ we have:
+
+\begin{eqnarray*}
+\mathbf{X}(\lambda)&=&(\mathbf{C}^\top
+\mathbf{C})^{-1}\mathbf{C}^\top \mathbf{F} \\
+&=& (\mathbf{A}^\top \mathbf{A} + \lambda \mathbb{I})^{-1} \mathbf{A}^\top \mathbf{B}
+\end{eqnarray*}
+
+\subsection{Online Ridge regression}
+
+If we represent matrices $\mathbf{A}$ and $\mathbf{B}$ such as:
+
 \begin{equation}
@@
   \end{tabular}
-\right]
-\end{equation}
-
-Equation~(\ref{eq:optsolRR}) can also be written as:
+\right] \, ,
+\
+\end{equation}
+
+\noindent equation~(\ref{eq:optsolRR}) can also be written as:
+
 \begin{eqnarray*}
@@
 \mathbf{\mathbf{X}}_{\text{ridge}}&=&(\mathbf{A}^\top \mathbf{A}+ \lambda
-\mathbb{I})^{-1}\mathbf{A}^\top \mathbf{Y} \\
+\mathbb{I})^{-1}\mathbf{A}^\top \mathbf{B} \\
 &=& \displaystyle \big (\sum_{t=1}^m
 \mathbf{a}_t \mathbf{a}_t  ^\top + \lambda \mathbb{I}\big )^{-1}
-\sum_{t=1}^m \mathbf{a}_t \mathbf{y}_t \, .
+\sum_{t=1}^m \mathbf{a}_t \mathbf{b}_t \, .
 \end{eqnarray*}
@@
 \mathbf{a}_t  ^\top + \lambda \mathbb{I} $ and $\mathbf{W}=
-\displaystyle\sum_{t=1}^m \mathbf{a}_t \mathbf{y}_t$, so the
+\displaystyle\sum_{t=1}^m \mathbf{a}_t \mathbf{b}_t$, so the
 algorithm~\ref{alg:RR} shows the iterative formulation:
@@
 $\{\mathbf{a}_1,\dots,\mathbf{a}_{m} \}$: $m$ input vectors \\
-$\{\mathbf{y}_1,\dots,\mathbf{y}_{m} \}$: $m$ target vectors \\
+$\{\mathbf{b}_1,\dots,\mathbf{b}_{m} \}$: $m$ target vectors \\
 $\lambda$: regularization parameter \\
@@
    	\STATE Read new $y_t$
-    	\STATE $\mathbf{W} = \mathbf{W} + \mathbf{a}_t \mathbf{y}_t$
+    	\STATE $\mathbf{W} = \mathbf{W} + \mathbf{a}_t \mathbf{b}_t$
 \ENDFOR
@@
 
-The AAR, proposed by~\cite{vovk2001} (also known as Vovk-Azoury-Warmuth predictor ~\cite{azoury2001}), is an application of the aggregating
-algorithm to the problem of regression. The idea is introduce the new input
+The AAR, proposed by~\cite{vovk2001}, is an application of the aggregating
+algorithm to the problem of regression. The idea is to introduce the new input
 vector $\mathbf{x}_{m+1}$ to solve the model parameters:
@@
 \mathbf{a}_t \mathbf{a}_t  ^\intercal + \gamma \mathbb{I}\big )^{-1}
-\sum_{t=1}^m \mathbf{a}_t \mathbf{y}_t \, .
+\sum_{t=1}^m \mathbf{a}_t \mathbf{b}_t \, .
 \end{equation}
@@
 \mathbf{a}_t  ^\intercal + \gamma \mathbb{I} $ and $\mathbf{W}=
-\displaystyle\sum_{t=1}^m \mathbf{a}_t \mathbf{y}_t$, the
+\displaystyle\sum_{t=1}^m \mathbf{a}_t \mathbf{b}_t$, the
 algorithm~\ref{alg:AAR} is slightly different to the algorithm~\ref{alg:RR}, 
@@
 $\{\mathbf{a}_1,\dots,\mathbf{a}_{m} \}$: $m$ input vectors \\
-$\{\mathbf{y}_1,\dots,\mathbf{y}_{m} \}$: $m$ target vectors \\
-$\lambda$: regularisation parameter \\
+$\{\mathbf{b}_1,\dots,\mathbf{b}_{m} \}$: $m$ target vectors \\
+$\lambda$: regularization parameter \\
 \ENSURE  $\,$ \\
@@
    	\STATE Read new $\mathbf{y}_t$
-    	\STATE $\mathbf{W} = \mathbf{W} + \mathbf{a}_t \mathbf{y}_t$
+    	\STATE $\mathbf{W} = \mathbf{W} + \mathbf{a}_t \mathbf{b}_t$
 \ENDFOR
@@
 
+\subsection{Selection of Lambda}
+
+The additional term $\lambda \|\mathbf{\mathbf{X}}\|_2^2$ in the optimization
+problem shown in equation~(\ref{eq:RRproblem}) has two effects on the solution:
+shrinks the coefficients towards zero and improves the conditioning of the
+problem.
+
+Figure~\ref{fig:shrinks} shows a visual example of the shrinking of the
+coefficients:
+
+\begin{figure}[h!]
+\includegraphics[width=0.5\linewidth]{img/shrinks}
+\caption{Shrink of regression coefficients}
+\label{fig:shrinks}
+\end{figure}
+
+On the other hand, the effect of adding the term $\lambda \mathbb{I}$
+to the matrix $\mathbf{A}^\top \mathbf{A}$
+(equation~(\ref{eq:optsolRR})) improves its condition number since it
+increases its diagonal values when $\lambda > 0 $.  The matrix
+$\mathbf{A}^\top \mathbf{A}$ is symmetrical ($(\mathbf{A}^\top
+\mathbf{A})^\top = \mathbf{A}^\top \mathbf{A}$) and therefore
+diagonalizable.  If we know the eigenvalue decomposition of $\mathbf{A}
+= \mathbf{U\Sigma V^\top}$, then:
+
+\begin{eqnarray*}
+\mathbf{A}^\top \mathbf{A}+\lambda \mathbb{I}&=&\mathbf{V\Sigma^2
+V}^\top + \lambda \mathbf{V} \mathbf{V}^\top\\ &=&\mathbf{V}
+(\mathbf{\Sigma}^2+\lambda\mathbb{I}) \mathbf{V}^\top \, ,
+\end{eqnarray*}
+
+\noindent where
+
+\begin{equation*}
+\mathbf{\Sigma}^2+\lambda\mathbb{I}=
+\begin{bmatrix}
+\sigma^2_1 + \lambda & \, & \, \\
+\, & \sigma^2_2 +\lambda & \, \\
+\, & \, & \ddots & \, \\
+\, & \, & \, & \sigma^2_n +\lambda \, .
+\end{bmatrix}
+\end{equation*}
+
+where $\sigma_1 \geq \sigma_2 \geq \dots \geq \sigma_n$.
+
+Since the condition number of a matrix $\mathbf{A}$ is defined as:
+
+\begin{equation*}
+	\kappa = \|\mathbf{A}\| \|\mathbf{A}^{-1}\|
+\end{equation*}
+
+If matrix $\mathbf{A}$ is non-singular, its condition number can be
+expressed in terms of its singular values. The effect of adding the
+regularization term affects the condition number as follows:
+
+\begin{eqnarray*}
+\kappa_{ols} &=& \|\mathbf{A}\| \|\mathbf{A}^{-1}\|=\frac{\sigma_1}{\sigma_n} \\
+\kappa_{ridge} &=& \|\mathbf{A}^\intercal \mathbf{A} + \lambda \mathbb{I}\| 
+\|(\mathbf{A}^\intercal \mathbf{A} + \lambda \mathbb{I})^{-1}\|=\frac{\sigma_1+\lambda}{\sigma_n + \lambda} \,
+\end{eqnarray*}
+
+It is easy to see that the term $\lambda$ improves the condition number:
+
+\begin{equation*}
+        \frac{\sigma_1+\lambda}{\sigma_n + \lambda} <
+        \frac{\sigma_1}{\sigma_n} \,  \qquad \forall \quad \lambda > 0
+\end{equation*}
+
+However, $\lambda$ cannot be too large. Tipically $\lambda$ is small and its
+magnitude depends on the matrix $\mathbf{A}$.
+
+For rank deficient matrices we know that $\text{det}(\mathbf{A A^\top})=0$, adding the
+term $\lambda \mathbb{I}$ we have that $\text{det}(\mathbf{A A^\top}+\lambda \mathbb{I}) =
+p(\lambda)$ where $p(\lambda)$ is a polynomial of degree $n$ ($\mathbf{A}$ is
+$m \times n$). The zeros of $p(\lambda)$ are discretes, so it can be represented
+as:
+
+\[
+p(\lambda) =
+\lambda(\lambda-\lambda_1)^{n_1}(\lambda-\lambda_2)^{n_2}\dots(\lambda-\lambda_s)^{n_s}
+\]
+
+\noindent where $n_1 + n_2 + \dots + n_s = n$.
+
+This means that $\lambda$ must be small in order to ensure that $p(\lambda)$
+does not vanish.
+
+\subsubsection{The bias-variance tradeoff}
+
+To determine parameter $\lambda$ is crucial for ridge regression since it could
+reduce the expected prediction error by reducing variance considering a biased
+estimator, this is called the bias-variance trafeoff. The prediction error is
+obtained as:
+
+\begin{equation}
+ \label{eq:prederror}
+E[(B-\hat{f}(\mathbf{X}))^2] = \sigma^2 + Bias(\hat{f}(\mathbf{X}))^2 + Var(\hat{f}(\mathbf{X}))
+\end{equation}
+
+\noindent where $\hat{f}(\mathbf{X})=\mathbf{AX}(\lambda)$ and
+$f(\mathbf{X})=\mathbf{AX}$.
+
+\textbf{Demo}\quad
+
+\begin{eqnarray*}
+E[(B-\hat{f}(\mathbf{X}))^2] & =&
+E[(B-f(\mathbf{X})+f(\mathbf{X})-\hat{f}(\mathbf{X}))^2] \\
+&=& E[(B-f(\mathbf{X}))^2] +
+E[(f(\mathbf{X})-\hat{f}(\mathbf{X}))^2] + \dots \\
+& &
+2\cancelto{0}{E[B-f(\mathbf{X})]}E[f(\mathbf{X})-\hat{f}(\mathbf{X})] \\
+&=& \sigma^2 + MSE(\hat{f}(\mathbf{X}))
+\end{eqnarray*}
+
+\noindent where
+
+\begin{eqnarray*}
+    MSE(\hat{f}(\mathbf{X})) &=&
+    E[f(\mathbf{X})-\hat{f}(\mathbf{X}))]^2 \\
+    &=& E[f(\mathbf{X})-E[\hat{f}(\mathbf{X})] +
+    E[\hat{f}(\mathbf{X})]-\hat{f}(\mathbf{X}))]^2 \\
+    &=&  E[(f(\mathbf{X})-E[\hat{f}(\mathbf{X})])^2] +
+      E[(\hat{f}(\mathbf{X})-E[\hat{f}(\mathbf{X})])^2] + \dots\\
+    & & E[(f(\mathbf{X})-E[\hat{f}(\mathbf{X})])] 
+    \cancelto{0}{E[(\hat{f}(\mathbf{X})-E[\hat{f}(\mathbf{X})])]} \\
+    &=& Bias(\hat{f}(\mathbf{X}))^2 + Var(\hat{f}(\mathbf{X}))
+\end{eqnarray*}
+
+The bias of OLS is:
+
+\begin{eqnarray*}
+Bias(\hat{\mathbf{X}}) &=& E[\hat{\mathbf{X}}] - \mathbf{X} \\
+&=& E[ (\mathbf{A}^\top \mathbf{A})^{-1}\mathbf{A}^\top \mathbf{B}] - \mathbf{X} \\
+&=& E[ (\mathbf{A}^\top \mathbf{A})^{-1}\mathbf{A}^\top (\mathbf{AX})] - \mathbf{X}  \\
+&=& \mathbf{X}  - \mathbf{X}  \\
+&=&  0
+\end{eqnarray*}
+
+The bias of ridge regression when $\mathbf{A A^\top}$ is non-singular
+can be obtained expressing ridge regression solution
+$\mathbf{\lambda}$ in terms of OLS solution $\hat{\mathbf{X}}$:
+
+\begin{eqnarray*}
+\mathbf{X}(\lambda) &=&( \mathbf{A}^\top \mathbf{A} + \lambda \mathbb{I})^{-1}\mathbf{B} \\
+&=& (\mathbb{I} + \lambda (\mathbf{A}^\top \mathbf{A})^{-1})^{-1} (\mathbf{A}^\top \mathbf{A})^{-1}\mathbf{A}^\top \mathbf{B} \\
+&=&  (\mathbb{I} + \lambda (\mathbf{A}^\top \mathbf{A})^{-1})^{-1}  \hat{\mathbf{X}} \\
+&=& \mathbf{W} \hat{\mathbf{X}} 
+\end{eqnarray*}
+
+\noindent where $\mathbf{W}  = (\mathbb{I} + \lambda (\mathbf{A}^\top
+\mathbf{A})^{-1})^{-1}  $ it is defined for simplicity. Ridge
+regression bias is then obtained as:
+
+\begin{eqnarray*}
+Bias(\mathbf{X}(\lambda)) &=& E[\mathbf{X}(\lambda)] - \mathbf{X} \\
+&=& E[\mathbf{W}\hat{\mathbf{X}}] - \mathbf{X} \\
+&=&  \mathbf{W} \mathbf{X} - \mathbf{X} \neq 0 
+\end{eqnarray*}
+
+The variance of OLS is:
+
+\begin{equation*}
+Var(\hat{\mathbf{X}}) = \sigma^2 (\mathbf{A}^\top \mathbf{A} )^{-1}
+\end{equation*}
+
+\noindent and the variance of ridge regression is:
+
+\begin{eqnarray*}
+Var(\mathbf{X}(\lambda)) &=& Var(\mathbf{W}\hat{\mathbf{X}}) \\ 
+&=& \mathbf{W}Var(\hat{\mathbf{X}})\mathbf{W}^\top \\
+&=& \sigma^2 \mathbf{W}(\mathbf{A}^\top \mathbf{A} )^{-1}\mathbf{W}^\top
+\end{eqnarray*}
+
+\begin{figure}[h!]
+\includegraphics[width=0.8\linewidth]{img/biasvariance}
+\caption{Bias-variance tradeoff}
+\label{fig:biasvariance}
+\end{figure}
+
+The figure~\ref{fig:biasvariance} shows the bias-variance tradeoff
+given by equation~(\ref{eq:prederror}). Ridge regression shows an
+increasing bias and a decreasing variance. Despite OLS has zero bias,
+its variance is greater than ridge for small values of $\lambda$.
+It can be shown that, in terms of prediction error, ridge (black line)
+is lower than OLS (dotted line)~\cite{hoerl1970}.
+
+\subsubsection{Methods for lambda selection}
+
+Despite the bias-variance tradeoff provides a conceptual framework for
+determining a good model, is not directly useful. Some popular methods
+for determine model selection are:
+
+\begin{itemize}
+    \item Akaike information criterion (AIC) (Akaike, 1974)
+    \item Bootstrap-based selection (Efron and Tibshirani, 1997)
+    \item Cross-validation (Stone, 1974)
+\end{itemize}
+
+On the other hand, $\lambda$ can also be used to reduce computational time as it
+is shown in equation~(\ref{eq:taylor}). Golub (~\cite{golub1965}) suggested that
+$\lambda$ had to be chosen so that:
+
+\[
+\frac{\lambda}{\lambda + \delta^2} < 0.1
+\]
+
+\noindent where $\delta$ is a lower bound of the smallest non-zero singular
+value $\sigma_k$. One approach is to choose the greatest lower bound for
+$\delta=\sigma_k$. Hence:
+
+\[
+\frac{\lambda}{\lambda + \sigma_k^2} < 0.1
+\]
+
+\noindent if we choose $\lambda=\beta \sigma_k^2$ then we get the following
+relation:
+
+\[
+\frac{\beta \sigma_k^2}{\beta \sigma_k^2 + \sigma_k^2} =
+\frac{\beta}{\beta+1} < 0.1 \Rightarrow \beta < \frac{1}{9}  
+\]
+
+Experiments done by ~\cite{coleman+sun2010} have shown that $\beta = 0.01$
+produces satisfactory results. This means that $\lambda = 0.01 \sigma_k^2$.
+
+However, get $\sigma_k$ imply obtaining first the SVD, which is
+computational expensive.
+
+Since the algorithm presented by ~\cite{coleman+sun2010} already compute the QR
+factorization of the matrix $\mathbf{A}$ they show a procedure for getting
+a $\lambda$ approximation using QR.
+
+\begin{enumerate}
+\item Compute QR factorization: $\mathbf{A} = \mathbf{Q_1R_1}$
+\item Let $\mathbf{W}$ denote the set of absolute values of the nonzero diagonal
+elements of $\mathbf{R_1}$. Let $w_{\text{min}}$ and $w_{\text{max}}$ denote the
+smallest and largest elements of $\mathbf{W}$ respectively. Both 
+$\lambda_1 = \hat{\beta} w_{\text{min}}^2$
+$\lambda_2 = \hat{\beta} \frac{w_{\text{min}}^2}{w_{\text{max}}^2}$ where
+$\hat{\beta}=0.00025$ produce satisfactory results.
+\item The $\lambda_{\text{QR}}$ approximation is obtained as:
+$\lambda_{\text{QR}} = \frac{\lambda_1 + \lambda_2}{2}$  
+\end{enumerate}
+
 \subsection{Efficient computation}
+
+%Revisar!!!
 In regression problems we always require getting a matrix inverse which is
@@
 
-Alternatively, Coleman and
-Sun~\cite{coleman+sun2010} presented an iterative algorithm which
+%Revisar!!
+
+Another way was presented by Coleman and
+Sun~\cite{coleman+sun2010} using an iterative algorithm which
 uses $\mathbf{X}(\lambda)$ to approximate $\mathbf{X}(0)$.
@@
 \mathbf{X}(\lambda) & = & (\mathbf{A}^\top \mathbf{A}+ \lambda
-\mathbb{I})^{-1}\mathbf{A}^\top \mathbf{Y} \nonumber \\
+\mathbb{I})^{-1}\mathbf{A}^\top \mathbf{B} \nonumber \\
 & = &\mathbf{V}_1(\Sigma_1^2+\lambda \mathbb{I})^{-1}\mathbf{\Sigma_1
-U_1^\top Y}
+U_1^\top B}
 \end{eqnarray}
@@
 \mathbf{\hat{X}}=\mathbf{V}_1 \mathbf{\Sigma}_1^{-1}\mathbf{U}_1^\top
-\mathbf{Y}$ as $\lambda \rightarrow 0$.
+\mathbf{B}$ as $\lambda \rightarrow 0$.
 
@@
 \begin{eqnarray*}
-(\mathbf{A}^\top \mathbf{A}+ \lambda\mathbb{I}) \mathbf{X}(\lambda) & = & \mathbf{A}^\top \mathbf{Y}\\
+(\mathbf{A}^\top \mathbf{A}+ \lambda\mathbb{I}) \mathbf{X}(\lambda) & = & \mathbf{A}^\top \mathbf{B}\\
 (\mathbf{A}^\top \mathbf{A}+ \lambda\mathbb{I}) \mathbf{X}(\lambda)^{(1)} + \mathbf{X}(\lambda)& = & 0 \\
@@
 $\mathbf{A}$: design matrix \\
-$\mathbf{Y}$: response matrix \\
+$\mathbf{B}$: response matrix \\
 $\lambda$: rank deficient parameter \\
@@
 \STATE Initialize $\mathbf{Q R}=\mathbf{M}$ \\
-\STATE $\mathbf{X} = \mathbf{R^{-1}Q^\top Y}$ \\
+\STATE $\mathbf{X} = \mathbf{R^{-1}Q^\top B}$ \\
 \STATE $\mathbf{s} = \mathbf{X}$ \\
@@
 The algorithm~\ref{alg:coleman} solves equation~(\ref{eq:taylor}). However, this
-version is unstable since typically $\|\mathbf{s}\|$ is very large and
+version is unstable since tipically $\|\mathbf{s}\|$ is very large and
 $\lambda^i$ is very small ($\lambda$ is small).
@@
 \STATE Initialize $\mathbf{Q R}=\mathbf{M}$ \\
-\STATE $\mathbf{X} = \mathbf{R^{-1}Q^\top Y}$ \\
+\STATE $\mathbf{X} = \mathbf{R^{-1}Q^\top B}$ \\
 \STATE $\mathbf{t} = \mathbf{X}$ \\
@@
 Both algorithms are equivalent, but algorithm~\ref{alg:colemanimproved} is more
-stable and converges in typically less than 10 steps.
-It is important to notice that the QR factorisation is computed only once and it
+stable and converges in tipically less than 10 steps.
+It is important to notice that the QR factorization is computed only once and it
 is computationally less expensive than the SVD.
 
-Despite the bias-variance tradeoff provides a conceptual framework for
-determining a good model, is not directly useful. Some popular methods
-for determine model selection are:
-
-\begin{itemize}
-    \item Akaike information criterion (AIC) (Akaike, 1974)
-    \item Bootstrap-based selection (Efron and Tibshirani, 1997)
-    \item Cross-validation (Stone, 1974)
-\end{itemize}
-
-On the other hand, $\lambda$ can also be used to reduce computational time as it
-is shown in equation~(\ref{eq:taylor}). Golub (~\cite{golub1965}) suggested that
-$\lambda$ be chosen so that:
-
-\[
-\frac{\lambda}{\lambda + \delta^2} < 0.1
-\]
-
-\noindent where $\delta$ is a lower bound of the smallest non-zero singular
-value $\sigma_k$. One approach is to choose the greatest lower bound for
-$\delta=\sigma_k$. Hence:
-
-\[
-\frac{\lambda}{\lambda + \sigma_k^2} < 0.1
-\]
-
-\noindent if we choose $\lambda=\beta \sigma_k^2$ then we get the following
-relation:
-
-\[
-\frac{\beta \sigma_k^2}{\beta \sigma_k^2 + \sigma_k^2} =
-\frac{\beta}{\beta+1} < 0.1 \Rightarrow \beta < \frac{1}{9}  
-\]
-
-Experiments done by ~\cite{coleman+sun2010} have shown that $\beta = 0.01$
-produces satisfactory results. This means that $\lambda = 0.01 \sigma_k^2$.
-
-However, get $\sigma_k$ imply obtaining first the SVD, which is
-computational expensive.
-
-Since the algorithm presented by ~\cite{coleman+sun2010} already compute the QR
-factorization of the matrix $\mathbf{A}$ they show a procedure for getting
-a $\lambda$ approximation using QR.
-
-\begin{enumerate}
-\item Compute QR factorization: $\mathbf{A} = \mathbf{Q_1R_1}$
-\item Let $\mathbf{W}$ denote the set of absolute values of the nonzero diagonal
-elements of $\mathbf{R_1}$. Let $w_{\text{min}}$ and $w_{\text{max}}$ denote the
-smallest and largest elements of $\mathbf{W}$ respectively. Both 
-$\lambda_1 = \hat{\beta} w_{\text{min}}^2$
-$\lambda_2 = \hat{\beta} \frac{w_{\text{min}}^2}{w_{\text{max}}^2}$ where
-$\hat{\beta}=0.00025$ produce satisfactory results.
-\item The $\lambda_{\text{QR}}$ approximation is obtained as:
-$\lambda_{\text{QR}} = \frac{\lambda_1 + \lambda_2}{2}$  
-\end{enumerate}
-
-\section{Evaluation methods} \label{sec:evaluation}
-
-Forecast performance is evaluated using different methods. We have chosen three
-measures usually used:
-\begin{description}
-\item[MAPE] Mean Average Percent Error which presents forecast errors as a
-percentage.
-\begin{equation}\label{eq:MAPE}
-\text{MAPE} = \frac{1}{N} \sum_{t=1}^{N} 
-\frac{\left|\mathbf{y}_t-\hat{\mathbf{y}}_t\right|}{\left|\mathbf{y}_t\right|}
- \times 100 
-\end{equation}
-\item[MAE] Mean Average Error which measures the distance between forecasts to the
-true value.
-\begin{equation}\label{eq:MAE}
-\text{MAE} = \frac{1}{N} \sum_{t=1}^{N} 
-\left| 
-\mathbf{y}_t-\hat{\mathbf{y}}_t
-\right| 
-\end{equation}
-\item[RMSE] Root Mean Square Error also measures the distance between forecasts
-to the true values but, unlike MAE, large deviations from the true value have a
-large impact on RMSE due to squaring forecast error.
-\begin{equation}\label{eq:RMSE}
-\text{RMSE} = \sqrt{
-\frac{\displaystyle \sum_{t=1}^{N} (\mathbf{y}_t-\hat{\mathbf{y}}_t)^2}{N}}
-\end{equation}
-\end{description}
-
-\section{Model selection} \label{sec:pselection}
-Akaike Information Criterion (AIC) is often used in model selection where AIC
-with smaller values are preferred since they represent a trade-off between bias
-and variance.  AIC is obtained as follows:
+\subsection{Online machine learning algorithms}
+
+An online algorithm allows incremental learning by processing one instance at a
+time. This is done updating the current model instead of building the model from
+scratch.
+
+Training phase is commonly a computationally expensive process. Therefore, when
+new data arrives it can't be included easily to the model. Moreover, it could
+happen that we won't have enought time to process the new data before more data
+arrives. In online learning algorithms there is no training phase, but the model
+is updated and evaluated at every time step. This model updating is
+computationally less expensive than a training phase.
+
+Online learning is specially useful for stream data problems and the model is
+sensitive to new data. It is also useful when some past data may be irrelevant
+or we want to improve computational efficiency. However, online algorithms could
+affect accuracy.
+
+There are several popular online methods such as
+perceptron~\cite{rosenblatt58},winnow~\cite{littlestone1988},
+passive-aggressive~\cite{crammerETall2006}, stochastic gradient
+descent~\cite{zhang2004}, aggregating algorithm~\cite{vovk2001} and the second
+order perceptron~\cite{cesa-bianchi2005}.  In \cite{blum1998} and
+\cite{cesa-bianchi2006} an in-deph analysis of online learning is provided.
+Applications in finance has been widely used: study presented
+by~\cite{arce+salinas2012} applied ridge regression in an online context and
+more recently, time series forecasting using online learning has been
+presented~\cite{anavaetAl2013}.
+
+Incremental learning refers to any online learning process that learns the same
+model as would be learnt by a batch learning algorithm.
+
+Incremental learning is useful when the input to a learning process is stream
+data, with the need or desire to be able to use the result of learning at any
+point in time, based on the input observations received so far.
+
+In principle, the stream of observations may be infinitely long, or the next
+observation long delayed, precluding any hope of waiting until all the
+observations have been received. One would rather not simply accumulate and
+store all the inputs and, upon receipt of each new one, apply a batch learning
+algorithm to the entire sequence of inputs received so far. It would be
+preferable compu- tationally if the existing hypothesis or other artifact of
+learning could be updated in response to each newly received input observation.
+
+Incremental learning is very useful when there is no need to record fundamental
+data and only a summary needs to be retained. Due to this, incremental
+algorithms are often characterized as memoryless, because no memory of past data
+is required.  The algorithm is online but not incremental if it doesn't produce
+the same result for all observations that the corresponding batch algorithm
+would for these same observations.
+
+Algorithm~\ref{alg:onlinealg} shows the online learning algorithm structure:
+
+\begin{algorithm}[ht]
+\begin{algorithmic}[1]
+    \STATE Receives input $\mathbf{x}_t$
+    \STATE Makes prediction $\mathbf{\hat{y}}_t$
+    \STATE Receives response $\mathbf{y}_t$
+    \STATE Incurs loss $l_t(\mathbf{y}_t,\mathbf{\hat{y}}_t)$
+\end{algorithmic}
+\caption{Structure of a Learning System}
+\label{alg:onlinealg}
+\end{algorithm}
+
+\noindent where $l$ is some loss function. Performance is later measured after
+$T$ trials as:
+
+\begin{equation*}
+L_T = \sum_{t=1}^T l_t(\mathbf{y}_t,\mathbf{\hat{y}}_t)
+\end{equation*}
+
+The objective is minimize this loss function for all instances.
+The quality of online learning algorithms is measured by a quantity known as
+regret which is the difference between the performance of the online algorithm
+and its optimal predictor $E^* \in \Theta$ given by:
+
+\begin{equation*}
+L^*_T= \text{min}_{E \in \Theta} L_T^E \, ,
+\end{equation*}
+
+\noindent where $L_T^E = \sum_{t=1}^T
+l_t(\mathbf{y}_t,\mathbf{y}^E_t)$ and $\mathbf{y}^E_t$ is the expert estimation.
+
+Therefore regret is defined as:
+
+\begin{equation*}
+R_T = L_T - L^*_T
+\end{equation*}
+
+%An online learning algorithm looks at every example exactly once.
+\subsection{Model selection}
+In order to set model parameters the Akaike Information Criterion (AIC) is
+used. AIC is often used in model selection where AIC whith smaller values are
+preferred.
+
+AIC is calculated as follows:
 
@@
 \item[l] is the loglikelihood function
-\item[k] number of estimated parameters
+\item[k] number of estimated parameters (including the variance)
 \item[N] number of observations
 \end{description}
-
-Loglikelihood function is obtained from the Residual Sum of Squares (RSS):
-
-\begin{equation}
-\label{eq:ll}
-l = -\frac{N}{2} \left(1 + ln(2\pi) + ln\left(\frac{RSS}{N}\right)\right) 
-\end{equation}

### 2016-03-22 articles-conferences/elsevier-journal-2015/src/intro.tex
Agregar estos repos a la tesis

\section{\uppercase{Introduction}}
\label{sec:introduction}
\noindent

Cointegration concept was introduced by~\cite{engle1987} and implies that one or
more linear combinations of non-stationary variables are stationary even though
individually they are not.  Moreover ~\cite{stock+watson1988} observed that
cointegration reflects the common stochastic trends providing a useful way to
understand cointegration relationships. These common stochastic trends can be
also interpreted as a long-run equilibrium relationships.

This idea was inmediatly adopted in finance since it could represent their
long-run relationship implied by economic theory~\cite{laietAl1991},
\cite{lence+falk2005}.  Economic theory suggest that economic time series are
mean-reverting process and therefore, it reflects the idea of that some set of
variables cannot wander too far from each other.

On the other hand, the efficient markets hypothesis, also known as the random
walk theory states that current stock prices fully reflect available
information related to its value and there is no way to earn excess
profits~\cite{fama1970}.  This means that if we have stock prices from a
jointly efficient market, they cannot be
cointegrated~\cite{granger1986},\cite{dwyer1992}. However, \cite{richards1995}
claims that cointegration is directly at odds with market efficiency, even
though, there is no evidende that cointegration among asset prices have
implications about market efficiency~\cite{lence+falk2005}.

Despite the fact that cointegration on closing daily rates of currency pairs has
not been found \cite{coleman1990}, \cite{copeland1991}, different time series
frequencies can have different behaviours~\cite{aldridge2009}. Pair trading is a
very common example of cointegration application~\cite{herlemont2003} 
but cointegration can also be extended to a larger set of
variables~\cite{mukherjee1995},\cite{engle2004}.

%So why do we care about cointegration? Someone else can probably give more
%econometric applications, but in quantitative finance, cointegration forms the
%basis of the pairs trading strategy: suppose we have two cointegrated stocks X
%and Y, with the particular (for concreteness) cointegrating relationship X - 2Y
%= Z, where Z is a stationary series of zero mean. For example, X could be
%McDonald's, Y could be Burger King, and the cointegration relationship would
%mean that X tends to be priced twice as high as Y, so that when X is more than
%twice the price of Y, we expect X to move down or Y to move up in the near
%future (and analogously, if X is less than twice the price of Y, we expect X to
%move up or Y to move down). This suggests the following trading strategy: if X -
%2Y > d, for some positive threshold d, then we should sell X and buy Y (since we
%expect X to decrease in price and Y to increase), and similarly, if X - 2Y < -d,
%then we should buy X and sell Y.

Vector error correction model (VECM) introduces this long-run relationship
among a set of cointegrated variables as an error correction term. VECM is a
special case of the vector autorregresive model (VAR) model. VAR model
expresses future values as a linear combination of variables past values.
However, VAR model cannot be used with non-stationary variables. VECM is a
linear model but in terms of variable differences. If cointegration exists,
variable differences are stationary and they introduce an error correction term
which adjusts coefficients to bring the variables back to equilibrium. In
finance, many economic time series turn to be stationary when they are
differentiated and cointegration restrictions often improves
forecasting~\cite{duy1998}. Therefore, VECM has been widely adopted.

Both VECM and VAR model parameters are obtained using ordinary least squares
(OLS) method. OLS has two main problems: is sensitive to errors on input data
and involves many calculations. The former problem is commonly solved using
Ridge Regression (RR) \cite{hoerl1970} which introduces a regularization
parameter that leads to an unbiased estimation with better generalization
capability. The second problem of computational complexity depends on the number
of past values and observations considered.  Recently, online learning
algorithms have been proposed to solve problems with large data sets because of
their simplicity and their ability to update the model when new data is
available.

In this paper, we propose an online formulation of the VECM called Onlive VECM
(OVECM) based on consideration of only a sliding window of the most recent data.
The algorithm introduces matrix optimizations in order to reduce the number of
operations and also takes into account the fact that cointegration vector space
doesn't experience large changes with small changes in the input data. Moreover,
RR instead of OLS is used to solve VECM. Our method is later tested using four
currency rates from the foreign exchange market with different frequencies.  Model
efectiveness is focused on out-of-sample forecast rather than in-sample fitting.
This criteria allows the OVECM prediction capability to be expressed rather than
just explaining data history. Our method performance is compared with its
optimal offline algorithm.

The next sections are organized as follows: section~\ref{sec:background}
presents the VAR and VECM, the OVECM algorithm proposed is presented in
section~\ref{sec:methodology}. Section~\ref{sec:results} gives a description of
the data used and the tests carried on to show accuracy and time comparison of
our proposal against the traditional VECM and
section~\ref{sec:conclusions} includes conclusions and a proposal for future
tudy.

### 2016-03-22 articles-conferences/elsevier-journal-2015/src/results.tex
Agregar estos repos a la tesis

\section{\uppercase{Results}}
\label{sec:results}
\noindent The forex market is a 24 hours market which main activity is
dominated by three major trading centers: Tokyo, London and New York. The seven
most important exchange rates are: the Australian Dollar (AUD), the Canadian
Dollar (CAD), the Swiss Franc (CHF), the British Pound Sterling (GBP),the Euro
(EUR), the Japanese Yen (JPY) and the Swedish Krona (SEK)~\cite{HaugetAl2000}.
Most HFT trades are done in those currencies, but increasingly New Zeland (NZD)
and Mexican Peso (MXN) have being automated traded.

Tests of OVECM were carried out using some of the main foreign exchange rates:
EURUSD, GBPUSD, USDCHF, USDJPY ~\cite{Dukascopy2014}. The reciprocal of the last
two rates (CHFUSD, JPYUSD) were used in order to obtain the same base currency
for all rates.  The tests were done using 1-minute frequency data from ask
prices which corresponded to 373.503 data points excluding flat weekends and
holidays from the 12th of August 2013 to the 12th of August 2014.

In estimating the VECM, we firstly checked for unit roots performing the
Augmented Dickey Fuller (ADF) test on the variables in levels and first
differences.

Table~\ref{tab:adf} shows that all currency rates cannot reject the unit root
test but they rejected it with their first differences. This means that all of
them are I(1) time series and we are allowed to use VECM and OVECM.

\begin{table}[h!]
\caption{Unit roots tests}
\label{tab:adf}
\begin{center}
\begin{tabular}{|l|c|c|c|c|c|}
\hline
& \textbf{Statistic} & \textbf{Critical value} & \textbf{Result}\\
\hline
EURUSD          & -0.64 & -1.94 & True       \\
$\Delta$ EURUSD & -70.45   & -1.94 & False       \\
GBPUSD          & -0.63   & -1.94 & True          \\
$\Delta$ GBPUSD & -54.53   & -1.94 & False       \\
CHFUSD          & -0.88   & -1.94 & True         \\
$\Delta$ CHFUSD & -98.98   & -1.94 & False       \\
JPYUSD          & -0.65 & -1.94 & True        \\
$\Delta$ JPYUSD & -85.78 & -1.94 & False     \\ 
\hline
\end{tabular}
\end{center}
\end{table}

\subsection{Setting parameters}

In order to set VECM parameter: $L$ and $p$, we use the Akaike Information
Criterion (AIC). RR parameter $\lambda$ was done by cross-validation.

\begin{figure}[h!]
\includegraphics[width=0.5\linewidth]{img/vecm-params}
\caption{Setting VECM parameters}
\label{fig:shrinks}
\end{figure}

We ran OVECM and SLVECM 10.000 iterations for different values of sliding
window size $L$, number of lags $p$ and mean\_error (the latest only appliesto
OVECM). The execution times are shown in the table~\ref{tab:extimes}.

\begin{table}[h!]
\caption{Execution times}
\label{tab:extimes}
\begin{center}
\begin{tabular}{|l|c|c|c|c|c|}
\hline
& \textbf{L} & \textbf{P} & \textbf{mean\_error}  & \textbf{Time} \\
\hline
SLVECM & 300 & 4  & -- & 144\\
OVECM & 300 & 4  & 0 & 98\\
OVECM & 300 & 4  & 120 & 78\\
\hline
SLVECM & 300 & 8  & -- & 298\\
OVECM & 300 & 8  & 0 & 196\\
OVECM & 300 & 8  & 120 & 177\\
\hline
SLVECM & 3000 & 4  & -- & 324\\
OVECM & 3000 & 4  & 0 & 241\\
OVECM & 3000 & 4  & 120 & 97\\
\hline
SLVECM & 3000 & 8  & -- & 860\\
OVECM & 3000 & 8  & 0 & 632\\
OVECM & 3000 & 8  & 120 & 303\\
\hline
\end{tabular}
\end{center}
\end{table}

It is clear that execution time depends directly on $L$ and $p$ since they
determine the size of matrix $\mathbf{A}$ and therefore affect the OLS function
execution time.
It is worthy of note that execution time also depends on mean\_error because it
determines how many times OVECM will recalculate cointegration vectors which is
an expensive routine. Also, our proposal reduces execution time more than  50\%
in some cases.

Figure~\ref{fig:mapes} shows an example of the
in-sample MAPE. In consequence, OVECM performance increases when
mean\_error increases, but it will affect accuracy. Table~\ref{tab:mapes} shows
differents MAPEs out-of-sample for SLVECM and OVECM and all of them are very
similar. This supports the theory that cointegration vectors little with time.

\begin{table*}[ht!]
\caption{Out-of-sample MAPEs}
\label{tab:mapes}
\begin{center}
\begin{tabular}{|l|c|c|c|c|c|c|c|}
\hline
Method & \textbf{L} & \textbf{P} & \textbf{mean}  & \textbf{MAPE} & \textbf{MAPE}& \textbf{MAPE}& \textbf{MAPE}\\
&  &  & \textbf{error}  & \textbf{EURUSD} & \textbf{GBPUSD}& \textbf{CHFUSD}& \textbf{JPYUSD}\\
\hline
 SLVECM &   L=300 &  P=4&    &  134.76&  136.50&  133.16&  137.20\\
 OVECM  &   L=300 &  P=4& 0  &  134.76&  136.50&  133.16&  137.20\\
 OVECM  &   L=300 &  P=4& 120&  134.30&  135.53&  132.88&  136.44\\
\hline
 SLVECM &   L=300 &  P=8&    &  157.76&  160.55&  155.24&  162.74\\
 OVECM  &   L=300 &  P=8& 0  &  157.76&  160.55&  155.24&  162.74\\
 OVECM  &   L=300 &  P=8& 120&  158.03&  160.89&  155.30&  162.61\\
\hline
 SLVECM &   L=3000&  P=4&    &  111.91&  110.60&  109.22&  105.69\\
 OVECM  &   L=3000&  P=4& 0  &  111.91&  110.60&  109.22&  105.69\\
 OVECM  &   L=3000&  P=4& 120&  112.49&  111.26&  109.25&  105.88\\
\hline
 SLVECM &   L=3000&  P=8&    &  118.96&  118.97&  114.99&  110.57\\
 OVECM  &   L=3000&  P=8& 0  &  118.96&  118.97&  114.99&  110.57\\
 OVECM  &   L=3000&  P=8& 120&  119.23&  119.05&  114.82&  110.58\\
\hline
\end{tabular}
\end{center}
\end{table*}

% SLVECM    L=300   P=4             [ 134.7630329035  136.5086625175  133.1658744763  137.2084992162]
% SLVECM    L=300   P=8             [ 157.761485001   160.5552885679  155.2402883372  162.7438222296]
% SLVECM    L=3000  P=4             [ 111.91007246    110.6026357988  109.2219354574  105.6932468374]
% SLVECM    L=3000  P=8             [ 118.9663817354  118.9770962212  114.9939599829  110.5712975915]
% OVECM     L=300   P=4 AVGE=0      [ 134.7630329035  136.5086625175  133.1658744763  137.2084992162]
% OVECM     L=300   P=8 AVGE=0      [ 157.761485001   160.5552885679  155.2402883372  162.7438222296]
% OVECM     L=3000  P=4 AVGE=0      [ 111.91007246    110.6026357988  109.2219354574  105.6932468374]
% OVECM     L=3000  P=8 AVGE=0      [ 118.9663817354  118.9770962212  114.9939599829  110.5712975915]
% OVECM     L=300   P=4 AVGE=120    [ 134.301881303   135.5315795893  132.8835616177  136.4493817757]
% OVECM     L=300   P=8 AVGE=120    [ 158.0383643503  160.8908753314  155.3006014391  162.6119716451]
% OVECM     L=3000  P=4 AVGE=120    [ 112.4957023005  111.2666209761  109.2525274819  105.8804205046]
% OVECM     L=3000  P=8 AVGE=120    [ 119.2383924719  119.0508650247  114.8288372278  110.5815700448]

%\begin{figure}[!h]
%  \vspace{-0.2cm}
%  \centering
%   {\epsfig{file = img/accuracy.eps, width = 7.5cm}}
%  \caption{Time series differences accuracy.}
%  \label{fig:accuracy}
% \end{figure}
%
%\begin{figure}[!h]
%  \vspace{-0.2cm}
%  \centering
%   {\epsfig{file = img/mapes.eps, width = 7.5cm}}
%  \caption{In-sample MAPEs example.}
%  \label{fig:mapes}
% \end{figure}

Despite the fact that MAPEs are high, figure ~\ref{fig:accuracy} shows some of
the out-of-sample forecasts made by our proposal OVECM which appears to follow
time series differences.

### 2016-03-22 articles-conferences/online-vec-article-2014/article.tex
Agregar estos repos a la tesis

--- 2016-03-22 articles-conferences/elsevier-journal-2015/journal.tex
@@
 \documentclass[12pt,reqno]{amsart}
+\usepackage{array,arydshln}
 \usepackage{amsfonts,amssymb,amsmath,amsthm,
             amssymb,amsopn,amsxtra,amstext,amsbsy}
+\usepackage{subfig} % To add subfigures
+\usepackage[makeroom]{cancel}
+%%\usepackage{graphicx}
 \usepackage{epsfig}
-\usepackage[makeroom]{cancel}
-\usepackage{array,arydshln}
 \usepackage{array}
 \usepackage{url}
-%\usepackage{tikz}
 \usepackage[spanish]{babel}
-\usepackage[utf8x]{inputenc}
-\usepackage{fancyhdr}
-\usepackage{graphicx}
-\usepackage{hyperref,wasysym}
 \usepackage{color}
-\usepackage{enumerate}
+%\usepackage{float}
+\usepackage{savesym}
+\savesymbol{AND}
 \usepackage{algorithm}
 \usepackage{algorithmic}
+\usepackage{enumerate}
+\newtheorem*{myteo}{Teorema}
+\newtheorem*{mydef}{Definici\'on}
 \renewcommand{\algorithmicrequire}{\textbf{Input:}}
 \renewcommand{\algorithmicensure}{\textbf{Output:}}
-
-\newtheorem*{myteo}{Teorema}
-\newtheorem*{mydef}{Definici\'on}
-
+\def\R{{\mathbb R }}
+\def\N{{\mathbb N }}
 \voffset=-10mm
@@
 \marginparwidth=19mm
+%\renewcommand{\baselinestretch}{2} %double spacing
 \vfuzz2pt % Don't report over-full v-boxes if over-edge is small
@@
 
-\title{A Time Varying Vector Error Correction Model Applied to Foreign Exchange
-Market}
-\author{Paola Arce \and Jonathan Antognini \and Luis Salinas \and Werner Kristjanpoller}
+\addtolength{\itemsep}{1.5\baselineskip}
+
+\title{An Online  Vector error correction model for forecasting exchange rates}
+\author[P. Arce]{Paola Arce}
+\address[Paola Arce]{Departamento de Inform\'atica, UTFSM}
+\email{paola.arce@usm.cl}
+\author[W. Kristjanpoller]{Werner Kristjanpoller} 
+\address[Werner Kristjanpoller]{Departamento de Industrias, UTFSM}
+\email{werner.kristjanpoller@usm.cl}
+\author[L. Salinas]{Luis Salinas}
+\address[Luis Salinas]{Departamento de Inform\'atica, UTFSM \and CCTVal, UTFSM}
+\email{luis.salinas@usm.cl}
 
 \begin{document}
+
 \maketitle
@@
 \begin{abstract}
-Financial time series are known for their non-stationary behaviour. However,
-sometimes they exhibit some stationary linear combinations. When this happens,
-it is said that those time series are cointegrated. Vector error correction
-model (VECM) is an econometric model which characterizes the joint dynamic
-behaviour of a set of cointegrated variables in terms of forces pulling towards
-equilibrium.
-
-VECM parameters are obtained using the ordinary least squares (OLS)
+Foreign exchange rates have been studied as a set of economic
+variables whose relationship is said to be cointegrated. When
+cointegration exists there are several models which attempt to predict
+their future behaviour using historical information. In particular,
+vector error correction (VEC) model is an econometric time series
+framework widely used to characterize the joint dynamic behaviour of a
+set of cointegrated variables. VEC is a particular case of the vector
+autoregression (VAR) model that models a set of any variables
+capturing linear interdependence among them.  As well as VAR and VEC
+model parameters are obtained using the ordinary least squares (OLS)
 method.  Even though OLS is extensively used, it has over-fitting
-issues and is computationally expensive. Ridge regression is commonly used
-instead of OLS since they include a regularization parameter which could improve
-prediction error.
-
-In this study, we propose an online VECM which optimizes how model parameters
-are obtained by using only relevant past data and RR instead of OLS.  Our
-proposal also takes advantage of the long-run relationship between the time
-series in order to obtain improved execution times. Our proposed method is
-tested using synthetic data and foreign exchange rates.
-Results show that our proposal enables execution time to be reduced by at least
-50\% while maintaining good accuracy levels.
+issues and is computationally expensive.  Due to the amount of new
+information available in foreign exchange markets, the using of VAR
+and VEC models is not suitable in online contexts.  In this article we
+present a new model able to obtain VEC parameters faster which are
+less adjusted to the data. Our model can be used in online contexts
+due to its reduced the response time and because it considers a
+portion of the historical data.
 \end{abstract}
@@
 \input{src/intro}
-\input{src/background}
-\input{src/methodology}
-\input{src/results}
-\input{src/conclusions}
-\bibliographystyle{apalike}
+\input{src/varvec}
+\input{src/OLS}
+\input{src/RR}
+\input{src/refinements}
+\input{src/lambda}
+\input{src/onlinemodels}
+\input{src/model}
+%\input{src/onlinevec}
+\input{src/data}
+%\input{src/experiments}
+%\input{src/conclusions}
+\input{src/appendix}
+
+\bibliographystyle{IEEEtran}
 \bibliography{reference}
-%\begin{thebibliography}{9}
-%
-%\bibitem{smola}
-%Bernhard Scholkopf and Alexander J. Smola. 
-%\emph{Learning with Kernels: Support Vector Machines, 
-%Regularization, Optimization, and Beyond.}.
-%MIT Press, Cambridge, MA, USA, 2001
-%
-%\bibitem{wiki}
-%\url{http://en.wikipedia.org/wiki/Reproducing_kernel_Hilbert_space}
-%
-%\bibitem{wikibooksIsom}
-%\url{http://en.wikibooks.org/wiki/Linear_Algebra/Definition_and_Examples_of_Isomorphisms}
-%
-%\bibitem{wolframREq}
-%\url{http://mathworld.wolfram.com/EquivalenceRelation.html}
-%
-%\bibitem{wikiHomoM}
-%\url{http://en.wikipedia.org/wiki/Homomorphism}
-%
-%
-%\end{thebibliography}
 

### 2016-03-22 articles-conferences/online-vec-article-2014/src/OLS.tex
Agregar estos repos a la tesis

\section{Regression problems} \label{sec:OLS}

VAR as also VEC model parameters requires solving the following
multiple regression problem:

\begin{equation}
\label{eq:regproblem}
\underset{m \times n}{\mathbf{A}} \enskip \underset{n \times
l}{\mathbf{\phi}} = \underset{m \times l}{\mathbf{B}}
\end{equation}

In the case matrix $\mathbf{X}$ is non singular, its solution is
straight forward:

\begin{equation}
\label{OLSsolution}
    \mathbf{X}=\mathbf{X}^{-1}\mathbf{B}
\end{equation}

\subsection{Ordinary least squares}
When $\mathbf{X}$ is singular, solution to
equation~(\ref{eq:regproblem}) is given by the ordinary least squares
(OLS) method. OLS consists of minimizing the sum of squared errors or
equivalently minimizing the following expression:

\begin{equation}
\label{eq:regressionproblem}
\underset{\mathbf{X}}{\text{min}} \quad \| \mathbf{X}\mathbf{\mathbf{X}} - \mathbf{B} \|_2^2
\end{equation}

\noindent which solution $\hat{\mathbf{X}}$ is well-known:

\begin{equation}
\label{eq:MP}
\hat{\mathbf{X}}=\mathbf{X}^+\mathbf{B}
\end{equation}

\noindent where $\mathbf{X}^+$ is the Moore-Penrose pseudo-inverse
which can be written as follows:

\begin{equation}
\label{eq:pseudoinverse}
\mathbf{X}^+= (\mathbf{X}^\top \mathbf{X})^{-1}\mathbf{X}^\top \, .
\end{equation}

However, when $\mathbf{X}$ is not full rank, i.e
$rank(\mathbf{X})=k <  n \leq m$, $\mathbf{X}^\top \mathbf{X}$ is
always singular and equation~(\ref{eq:pseudoinverse}) cannot be used.
More generally, the pseudo-inverse is best computed using the compact
singular value decomposition (SVD) of $\mathbf{X}$:

\begin{equation}
    \label{eq:compactsvd}
    \underset{m \times n}{\mathbf{X}}=
    \underset{m \times k}{\mathbf{U_1}} \enskip
    \underset{k \times k}{\mathbf{\Sigma_1}} \enskip
    \underset{k \times n}{\mathbf{V_1}^\top}
\end{equation}

\noindent as follows

\begin{equation}
\label{eq:pseudoinversesvd}
\mathbf{X}^+ = \mathbf{V_1\Sigma_1^{-1}U_1^\top}
\end{equation}

\textbf{Demo}\quad

Since the problem shown in equation~(\ref{eq:regproblem}) has not
solution, the minimum norm given by equation~(\ref{OLSsolution}) is
obtained by solving the equivalent problem:

\begin{equation*}
\label{eq:proyectorsol}
\mathbf{X \hat{\phi} = PB} 
\end{equation*}

\noindent where $\mathbf{P=U_1 U_1^\top}$ is the projection onto the
Col($\mathbf{X}$).

Since $\mathbf{V} = [\underset{(n \times k)}{\mathbf{V_1}} |
\underset{(n \times k)}{\mathbf{V_2}}]$ and $\mathbf{V_1^\top V_2 =
0}$ we can express $\mathbf{\hat{\phi}} = \mathbf{V_1 x_1 + V_2 x_2}$
with $\mathbf{x_2=0}$ because $\mathbf{\hat{\phi}}$ lives in the
$\text{Row}(\mathbf{X})$ given by $\mathbf{V_1}$, so we have:

\begin{eqnarray*}
\mathbf{X \hat{\phi}} &=& \mathbf{PB} \\
\mathbf{U_1 \Sigma_1 V_1^\top \hat{\phi}} &=& \mathbf{U_1 U_1^\top B} \\
\mathbf{ V_1^\top \hat{\phi}} &=&  \mathbf{\Sigma_1^{-1} U_1^\top B} \\ 
\mathbf{ V_1^\top V_1 x_1} &=& \mathbf{\Sigma_1^{-1}
U_1^\top B} \\
\mathbf{x_1}&=& \mathbf{\Sigma_1^{-1} U_1^\top B}
\end{eqnarray*}

\noindent from this result we can obtain $\mathbf{\hat{\phi}}$ and
therefore the pseudo-inverse expression:

\begin{eqnarray*}
\mathbf{\hat{\phi}} &=& \mathbf{V_1 x_1} \\
                &=& \mathbf{V_1 \Sigma_1^{-1} U_1^\top B} \\
\mathbf{X^+} &=& \mathbf{V_1 \Sigma_1^{-1} U_1^\top} \, .
\end{eqnarray*}

### 2016-03-22 articles-conferences/online-vec-article-2014/src/appendix.tex
Agregar estos repos a la tesis

\appendix

\subsection{Ridge vs. OLS estimator}
When $\mathbf{X}$ is orthonormal then $\mathbf{X}^\intercal \mathbf{X} =\mathbb{I}$ and there is a simple relation between the ridge estimator and the OLS estimator:

\begin{eqnarray*}
\phi_* (\lambda) &=& (\mathbf{X}^\intercal \mathbf{X}+\lambda \mathbb{I})^{-1}\mathbf{X}^\intercal \mathbf{Y} \\
 &=& (\mathbb{I} + \lambda \mathbb{I})^{-1} \mathbf{X}^\intercal \mathbf{Y} \\
 &=&(1+\lambda)^{-1} \mathbb{I} \mathbf{X}^\intercal \mathbf{Y} \\
 &=&(1+\lambda)^{-1} (\mathbf{X}^\intercal \mathbf{X})^{-1}\mathbf{X}^\intercal \mathbf{Y} \\
 &=&(1+\lambda)^{-1} \phi
\end{eqnarray*}

%\subsection{Geometric interpretation}
%\begin{figure}
%\includegraphics[width=0.9\linewidth]{img/geo1}
%\end{figure}
%\begin{figure}
%\includegraphics[width=0.9\linewidth]{img/geo2}
%\end{figure}

\section{Sherman Morrison formula} \label{sec:sherman}

Sherman Morrison formula states if a matrix $\mathbf{A}$ is a positive
definite matrix and its inverse matrix is known, then the inverse of the
matrix $\mathbf{B}=\mathbf{A} + \mathbf{x}\mathbf{x}^\intercal$ can be
obtained as:

\begin{equation*}
\mathbf{B}^{-1} = \mathbf{A}^{-1} -
\frac{(\mathbf{A}^{-1}\mathbf{x})(\mathbf{A}^{-1}\mathbf{x})^{\intercal}}
{1 + \mathbf{x}^{\intercal} \mathbf{A}^{-1} \mathbf{x}} \, .
\end{equation*}

This formula reduces the inverse matrix computation order from $O(n^3)$ to
$O(n^2)$.

### 2016-03-22 articles-conferences/online-vec-article-2014/src/data.tex
Agregar estos repos a la tesis

\section{Foreign Exchange Rate Data} \label{sec:data}

The data used in this article was obtained from Dukascopy historical
bid and ask prices~\cite{dukascopy}. We used currency prices every
minute from the 2nd of January 2012 to the 18th of November 2013,
which were 987454 inputs.

The available currencies pairs were: AUDCAD, AUDCHF, AUDJPY, AUDNZD,
AUDSGD, AUDUSD, CADCHF, CADHKD, CADJPY, CHFJPY, CHFSGD, EURAUD,
EURCAD, EURCHF, EURDKK, EURGBP, EURHKD, EURJPY, EURNOK, EURNZD,
EURPLN, EURSEK, EURSGD, EURTRY, EURUSD, GBPAUD, GBPCAD, GBPCHF,
GBPJPY, GBPNZD, GBPUSD, HKDJPY, NZDCAD, NZDCHF, NZDJPY, NZDUSD,
SGDJPY, USDCAD, USDCHF, USDDKK, USDHKD, USDJPY, USDMXN, USDNOK,
USDPLN, USDRUB, USDSEK, USDSGD, USDTRY, USDZAR, XAGUSD and XAUUSD.

### 2016-03-22 articles-conferences/online-vec-article-2014/src/experiments.tex
Agregar estos repos a la tesis

\section{Experiments} \label{sec:experiments}

In order to determine the parameter for sliding window size $L$  and
the number of lags, we define a set of base currencies:
EURUSD,GBPUSD,USDJPY,USDCHF and add an extra currency of the rest of
currencies available. The base currencies were chosen because they are
the most transacted currencies.

Later, we run VEC model with the 5 currencies considering a variable
window size data (from 100 to 4000) and number of lags (from 1 to 20)
and determine its akaike information criterion (AIC). The figure
~\ref{fig:lagoptimization} shows the zones with AIC levels varying in
color, were white represent the higher values and red represents the
lower values (best models have lower AIC).  We can see that more red
zones are related with high number of lags and windows size around
3000 in 4 examples. The situation is similar in

\begin{figure}[H]
\centering
\begin{tabular}{cc}

\subfloat[Base currencies + AUDSGD]{
    \includegraphics[width=0.4\linewidth]{img/AIC_AUDSGD}
    \label{fig:currency1}
}
&
\subfloat[Base currencies + NZDJPY]{
    \includegraphics[width=0.4\linewidth]{img/AIC_NZDJPY}
    \label{fig:currency2}
}
\\
\subfloat[Base currencies + AUDNZD]{
    \includegraphics[width=0.4\linewidth]{img/AIC_AUDNZD}
    \label{fig:currency3}
}
&
\subfloat[Base currencies + CHFJPY]{
    \includegraphics[width=0.4\linewidth]{img/AIC_CHFJPY}
    \label{fig:currency4}
}

\end{tabular}
\caption{AIC: Lags vs Sliding window size}
\label{fig:lagoptimization}
\end{figure}

### 2016-03-22 articles-conferences/online-vec-article-2014/src/intro.tex
Agregar estos repos a la tesis

--- 2014-10-29 src/5_Proposal.tex
@@
-\chapter{Online VECM proposal}
-\section{\uppercase{Methodology}}
-\label{sec:methodology}
-\noindent 
-\subsection{Online VECM} \label{sec:proposal}
+\section{Introduction}
+In finance, it is common to find variables with a long-run and/or a short-run
+equilibrium relationship. This relationship is called cointegration and it
+reflects the idea of that some set of variables cannot wander too far away from
+each other. 
+%and also high frequency relationship which avoid arbitrage opportunities
+%\footnote{Arbitrage is taking advantage of the price differences in different
+%markets}.  
+Cointegration means that one or more combinations of these variables is
+stationary even though individually they are not.
 
-Since VECM parameter estimation is computationally expensive, we propose an
-online version of VECM (OVECM).  OVECM considers only a sliding window of the
-most recent data. Moreover, since cointegration vectors represent long-run
-relationships which vary little in time, OVECM determines firstly if they require calculation.
+Some financial models, such as  vector error correction (VEC), take advantage of
+this property and describe the joint behaviour of several cointegrated
+financial instruments.
 
-OVECM also implements matrix optimizations in order to reduce execution time,
-such as updating matrices with new data, removing old data and introducing new
-cointegration vectors.
+VEC is a special case of the vector autorregresive model (VAR). VAR is a linear
+model which express future values in terms of its own history and other
+variables past values. If cointegration exists, VEC model allows to introduce an
+error correction term due to its cointegration estimation error.
 
-Algorithm~\ref{alg:proposal} shows our OVECM proposal which considers the
-following:
+VEC as well as VAR model parameters are obtained using ordinary least squares
+method (OLS).   Since OLS involves many calculations, the parameter estimation
+method is computationally expensive when the number of past values and
+observations considered increases. Moreover, OLS is an ill-posed problem which
+admits an infinite number of solutions.
 
-\begin{itemize}
-\item The function \texttt{getJohansen} returns cointegration vectors given by
-the Johansen method considering the trace statistic test at 95\%
-significance level.
-\item The function \texttt{vecMatrix} returns the matrices~(\ref{eq:vecA})
-and~(\ref{eq:vecB}) that allows VECM to be solved.
-\item The function \texttt{vecMatrixOnline} returns the
-matrices~(\ref{eq:vecA}) and~(\ref{eq:vecB}) aggregating new data and removing
-the old one, avoiding calculation of the matrix $\mathbf{A}$ from scratch.
-\item Out-of-sample outputs are saved in the variables 
-$\mathbf{Y}_{\text{true}}$ and $\mathbf{Y}_{\text{pred}}$.
-\item The model is solved using OLS.
-\item In-sample outputs are saved in the variables $\Delta
-\mathbf{y}_{\text{true}}$ and $\Delta \mathbf{y}_{\text{pred}}$.
-\item The function \texttt{mape} gets the in-sample MAPE for the $l$ time
-series.
-\item Cointegration vectors are obtained at the beginning and when they are required to be updated. This updating is decided based on the in-sample MAPE of the last $n$ inputs. The average of all
-MAPEs are stored in the variable $e$. If the average of MAPEs
-($\texttt{mean}(e)$) is above a certain error given by the mean\_error threshold, cointegration vectors are updated.
-\item If new cointegration vectors are required, the function
-\texttt{vecMatrixUpdate} only updates the corresponding columns of matrix
-$\mathbf{A}$ affected by those vectors (see equation~\ref{eq:vecA}).
-\end{itemize}
+Ridge regression method (RR) tackles this ill-posed problem and is usually
+formulated instead of OLS. RR includes a regularization parameter that leads to
+better generalization capability. However, RR is still computationally
+expensive dealing with large datasets. Recently, online learning algorithms
+have been proposed to solve problems with large data sets because of their
+simplicity and their ability of updating the model when new data is available.
 
-\begin{algorithm}[ht]
-\begin{algorithmic}[1]
-\REQUIRE $\,$ \\
-$\mathbf{y}$: matrix with $N$ input vectors and $l$ time series\\
-$p$: number of past values \\
-$L$: sliding window size ($L<N$) \\
-$\text{mean\_error}$: MAPE threshold \\
-$n$: interpolation points to obtain MAPE \\
-\ENSURE  $\,$ \\
-$\{\mathbf{y}_{\text{pred}}[L+1],\dots,\mathbf{y}_{\text{pred}}[N]\}$: model predictions 
-\FOR { $i =0$ to $N-L$ }
-    \STATE $\mathbf{y}_i \gets \mathbf{y}[i:i+L]$
-	\IF {i = 0 }
-	    \STATE{$v \gets \texttt{getJohansen}(\mathbf{y}_i,p)$}
-	    \STATE{$[\mathbf{A} \quad \mathbf{B}] \gets
-        \texttt{vecMatrix}(\mathbf{y}_i,p,v)$}
-	\ELSE
-	    \STATE{$[\mathbf{A} \quad \mathbf{B}] \gets
-        \texttt{vecMatrixOnline}(\mathbf{y}_i,p,v,\mathbf{A},\mathbf{B})$}
-        %\STATE $\Delta \mathbf{Y}_{\text{true}}[i] \gets \mathbf{B}[-1,:]$
-        \STATE $\Delta \mathbf{Y}_{\text{pred}}[i] \gets \mathbf{A}[-1,:] \times \mathbf{X}$
-    \ENDIF
-    \STATE $\mathbf{X} \gets \texttt{OLS} (\mathbf{A},\mathbf{B})$%\mathbf{(A^\top A)^{-1}A^\top B}$
-    %\STATE $\Delta \mathbf{y}_{\text{true}} \gets \mathbf{B}[-n,:]$
-    %\STATE $\Delta \mathbf{y}_{\text{pred}} \gets \mathbf{A}[-n,:] \times \mathbf{X}$
-    \STATE $e \gets \texttt{mape}(\mathbf{B}[-n,:],\mathbf{A}[-n,:] \times \mathbf{X})$
-    %\STATE $e \gets \texttt{mape}(\Delta \mathbf{y}_{\text{true}}, \Delta
-    %\mathbf{y}_{\text{pred}})$
-    \IF {$\texttt{mean}(e) > \text{mean\_error}$}
-	    \STATE{$v \gets \texttt{getJohansen}(\mathbf{y}_i,p)$}
-	    \STATE{$\mathbf{A} \gets
-        \texttt{vecMatrixUpdate}(\mathbf{y}_i,p,v,\mathbf{A})$}
-        \STATE $\mathbf{X} \gets \texttt{OLS} (\mathbf{A},\mathbf{B})$%\mathbf{(A^\top A)^{-1}A^\top B}$
-        %\STATE $\mathbf{X} \gets \mathbf{(A^\top A)^{-1}A^\top B}$
-    \ENDIF
-\ENDFOR
-\STATE $\mathbf{Y}_{\text{true}} \gets \mathbf{Y}[L+1:N]$
-\STATE $\mathbf{Y}_{\text{pred}}\gets 
-\mathbf{Y}[L:N-1] +\Delta \mathbf{Y}_{\text{pred}}$
-%\STATE $\text{MAPE} \gets \texttt{mape}(\Delta \mathbf{Y}_{\text{true}}, \Delta
-%\mathbf{Y}_{\text{pred}})$
-\end{algorithmic}
-\caption{OVECM: Online VECM}
-\label{alg:proposal}
-\end{algorithm}
+There are several popular online methods such as
+perceptron~\cite{rosenblatt58}, passive-aggressive~\cite{crammerETall2006},
+stochastic gradient descent~\cite{zhang2004}, aggregating
+algorithm~\cite{vovk2001} and the second order
+perceptron~\cite{cesa-bianchi2005}.  In~\cite{cesa-bianchi2006}, an in-deph
+analysis of online learning is provided.  In particular, the {\em aggregating
+algorithm for regression} (AAR) method is a recursive formulation of RR
+suitable for using in an online context.
 
-Our proposal was compared against VECM and ARIMA. Both algorithms were adapted
-to an online context in order to get a reasonable comparison with our proposal
-(see algorithms \ref{alg:SLVECM} and \ref{alg:SLARIMA}). VECM and ARIMA are
-called with a sliding window of the most recent data, whereby the models are
-updated at every time step.
+In this paper, we propose an online formulation of the VEC model based on the
+AAR method considering only a sliding window of the most recent data. The
+algorithm introduces matrix optimizations in order to reduce the number of
+operations. Our method is later tested with financial data from the foreign
+exchange market.
 
-\begin{algorithm}[ht]
-\begin{algorithmic}[1]
-\REQUIRE $\,$ \\
-$\mathbf{y}$: matrix with $N$ input vectors and $l$ time series\\
-$p$: number of past values \\
-$L$: sliding window size ($L<N$) \\
-\ENSURE  $\,$ \\
-$\{ \mathbf{y}_{\text{pred}}[L+1],\dots,\mathbf{y}_{\text{pred}}[N]\}$: model predictions 
-\FOR { $i =0$ to $N-L$ }
-    \STATE $\mathbf{y}_i \gets \mathbf{y}[i:i+L+1]$
-        \STATE $model = VECM(\mathbf{y}_i, p)$
-        \STATE $\mathbf{Y}_{\text{pred}}[i] = model.predict(\mathbf{y}[i+L])$
-\ENDFOR
-\STATE $\mathbf{Y}_{\text{true}} = \mathbf{y}[i+L+1:N] $
-\end{algorithmic}
-\caption{SLVECM: Sliding window VECM}
-\label{alg:SLVECM}
-\end{algorithm}
-
-Since we know out time series are I(1) SLARIMA is called with $d=1$. ARIMA is
-executed for every time series.
-
-\begin{algorithm}[ht]
-\begin{algorithmic}[1]
-\REQUIRE $\,$ \\
-$\mathbf{y}$: matrix with $N$ input vectors and $l$ time series\\
-$p$: autoregressive order \\
-$d$: integrated order\\
-$q$: moving average order\\
-$L$: sliding window size ($L<N$) \\
-\ENSURE  $\,$ \\
-$\{ \mathbf{y}_{\text{pred}}[L+1],\dots,\mathbf{y}_{\text{pred}}[N]\}$: model predictions 
-\FOR { $i =0$ to $N-L$ }
-\FOR { $j =0$ to $l-1$ }
-    \STATE $\mathbf{y}_i \gets \mathbf{y}[i:i+L+1,j]$
-        \STATE $model = ARIMA(\mathbf{y}_i, (p,d,q))$
-        \STATE $\mathbf{Y}_{\text{pred}}[i,j] = model.predict(\mathbf{y}[i+L,j])$
-\ENDFOR
-\ENDFOR
-\STATE $\mathbf{Y}_{\text{true}} = \mathbf{y}[i+L+1:N,:] $
-\end{algorithmic}
-\caption{SLARIMA: Sliding window ARIMA}
-\label{alg:SLARIMA}
-\end{algorithm}
-
-Both OVECM and SLVECM time complexity is dominated by Johansen method which is
-$O(n^3)$. Thus, both algorithms order is $O(Cn^3)$ where $C$ is the number of
-iterations.
-
-\subsection{Evaluation methods} \label{sec:evaluation}
-
-Forecast performance is evaluated using different methods. We have chosen three
-measures usually used:
-\begin{description}
-\item[MAPE] Mean Average Percent Error which presents forecast errors as a
-percentage.
-\begin{equation}\label{eq:MAPE}
-\text{MAPE} = \frac{1}{N} \sum_{t=1}^{N} 
-\frac{\left|\mathbf{y}_t-\hat{\mathbf{y}}_t\right|}{\left|\mathbf{y}_t\right|}
- \times 100 
-\end{equation}
-\item[MAE] Mean Average Error which measures the distance between forecasts to the
-true value.
-\begin{equation}\label{eq:MAE}
-\text{MAE} = \frac{1}{N} \sum_{t=1}^{N} 
-\left| 
-\mathbf{y}_t-\hat{\mathbf{y}}_t
-\right| 
-\end{equation}
-\item[RMSE] Root Mean Square Error also measures the distance between forecasts
-to the true values but, unlike MAE, large deviations from the true value have a
-large impact on RMSE due to squaring forecast error.
-\begin{equation}\label{eq:RMSE}
-\text{RMSE} = \sqrt{
-\frac{\displaystyle \sum_{t=1}^{N} (\mathbf{y}_t-\hat{\mathbf{y}}_t)^2}{N}}
-\end{equation}
-\end{description}
-
-\subsection{Model selection} \label{sec:pselection}
-Akaike Information Criterion (AIC) is often used in model selection where AIC
-with smaller values are preferred since they represent a trade-off between bias
-and variance.  AIC is obtained as follows:
-
-\begin{equation}
-\label{eq:aicformula}
-AIC = \underset{\text{bias}}{-\frac{2l}{N}} + 
-\underset{\text{variance}}{\frac{2k}{N}}
-\end{equation}
-
-\noindent where
-
-\begin{description}
-\item[l] is the loglikelihood function
-\item[k] number of estimated parameters
-\item[N] number of observations
-\end{description}
-
-Loglikelihood function is obtained from the Residual Sum of Squares (RSS):
-
-\begin{equation}
-\label{eq:ll}
-l = -\frac{N}{2} \left(1 + ln(2\pi) + ln\left(\frac{RSS}{N}\right)\right) 
-\end{equation}
+%The next sections are organized as follows: section~\ref{sec:varvec} presents
+%the VAR and VEC model, section~\ref{sec:RR} presents the RR formulation,
+%section~\ref{sec:ORR} shows the online version of RR, the online VEC algorithm
+%proposed is presented in section~\ref{sec:proposal}, section~\ref{sec:data}
+%gives a description of the data used, section~\ref{sec:experiments} shows the
+%tests carried on to show a comparision in accuracy and time of our proposal and
+%section~\label{sec:conclusion} concludes.

### 2016-03-22 articles-conferences/online-vec-article-2014/src/lambda.tex
Agregar estos repos a la tesis

--- 2016-03-22 articles-conferences/elsevier-journal-2015/src/background.tex
@@
-\section{\uppercase{Background}}
-\label{sec:background}
-\noindent
-
-\subsection{Spurious Regression}
-
-Regression analysis is widely used to test economic theories. But sometimes
-regression is done between two variables, lets say $x_t$ and $y_t$, which are
-not causally related. This occurs at the presence of a third stochastic variable
-$z_t$ which is the true causal of the first two variables $x_t$ and $y_t$. The
-regression measure R-squared gives high values in presence of spurious
-relationships. This measure would means that the model explains all the
-variability of the response data around its mean which is not true.
-
-The R-squared measure is calculated as follows:
-
-\begin{equation}
-\label{eq:rsquared}
-R^2 = 1 - \frac{\sum_{t=1}^T \epsilon_t^2}{\sum_{t=1}^T(y_t - \bar{y})^2}
-\end{equation}
-
-In presence of spurious relationships the denominator of equation
-\ref{eq:rsquared} becomes very large because a large weight is placed on extreme
-observations on either side of the mean $\bar{y}$.
-In order to detect this, other statistic must be considered such as the
-Durbin-Watson statistic.
-
-% Acá agregar gráfico de spurious regression
-
-\subsection{Integration and Cointegration}
-A time series $\mathbf{y}$ is said to be integrated of order $d$ if after
-differentiating the variable $d$ times, we get an I(0) process, more precisely:
-
-\[
-(1-L)^d \mathbf{y} \sim \text{I(0)} \, ,
-\]
-
-\noindent where I(0) is a stationary time series and $L$ is the lag operator:
-
-\[
-(1-L)\mathbf{y} = \Delta \mathbf{y}=\mathbf{y}_t  -\mathbf{y}_{t-1} \quad \forall t
-\]
-
-Let $\mathbf{y}_t = \{\mathbf{y}^1, \dots, \mathbf{y}^l\}$ be a set of $l$
-stationary time series I(1) which are said to be cointegrated if a vector
-$\beta=[\beta(1),\dots,\beta(l)]^\top \in \mathbb{R}^l$  exists such that the
-time series,
-
-\begin{equation}
- \mathbf{Z}_t:= \beta^\top \mathbf{y}_t = \beta(1) \mathbf{y}^1 + \dots + \beta(l) \mathbf{y}^l \sim
- \text{I(0)}\, .
-\end{equation}
-
-In other words, a set of I(1) variables is said to be cointegrated if
-a linear combination of them exists which is I(0).
-
-\subsection{Vector Autorregresive Models}\label{sec:varvec}
-
-VECM is a special case of VAR model and both describe the joint behaviour of a
-set of variables.
-
-VAR($p$) model is a general framework to describe the behaviour of a set of $l$
-endogenous variables as a linear combination of their last $p$ values. These
-$l$ variables at time $t$ are represented by the vector $\mathbf{y}_t$ as
-follows:
-
-\begin{equation}
-\label{eq:variables}
-\mathbf{y}_t = 
-\begin{bmatrix} y_{1,t} &
-y_{2,t} &
-\dots &
-y_{l,t}
-\end{bmatrix}^\top \, ,
-\end{equation}
-\noindent where $y_{j,t}$ corresponds to the time series $j$ evaluated at
-time $t$.
-
-The VAR(p) model describes the behaviour of a dependent variable in terms
-of its own lagged values and the lags of the others variables in the
-system. The model with $p$ lags is formulated as the following:
-
-\begin{equation}
-\label{eq:var}
- \mathbf{y}_t = \phi_1 \mathbf{y}_{t-1}  + \dots +   \phi_p\mathbf{y}_{t-p}
-+ \mathbf{c} + \mathbf{\epsilon}_t \, ,
-\end{equation}
-
-\noindent where $t=p+1, \dots N$,  ${\phi_1,\dots,\phi_p}$ are $l \times l$
-matrices of real coefficients,
-$\mathbf{\epsilon}_{p+1},\dots,\mathbf{\epsilon}_N$ are error terms,
-$\mathbf{c}$ is a constant vector and $N$ is the total number of samples.
-
-The VAR matrix form of equation~(\ref{eq:var}) is:
-
-%\begin{equation}\label{eq:vareq}
-%\mathbf{B} = \mathbf{A} \mathbf{X} + 
-%\mathbf{E} \, , 
-%\end{equation}
-%
-%\noindent where:
-%\begin{eqnarray}
-%\mathbf{A} &=&
-%\begin{bmatrix}
-%   \mathbf{y}_{p}     & \dots    & \mathbf{y}_{N-1}\\
-%   \mathbf{y}_{p-1}   & \dots    & \mathbf{y}_{N-2}\\
-%   \vdots             & \ddots   & \vdots\\
-%   \mathbf{y}_{1}     & \dots    & \mathbf{y}_{N-p}\\
-%   1                  & \dots    & 1 
-%   \end{bmatrix}^\top \, ,\\
-%\mathbf{B} &=&
-%\begin{bmatrix}
-%\quad \\
-%\mathbf{y}_{p+1} &
-%\dots & 
-%\mathbf{y}_N \\
-%\quad
-%\end{bmatrix}^\top  \, ,\\
-%\mathbf{X}&=&
-%\left[
-%\begin{array}{ccccc}
-%    \quad  & \quad & \quad & \quad \\
-%    \phi_1   & \cdots & \phi_p & \mathbf{c} \\  
-%    \quad & \quad & \quad & \quad
-%    \end{array}
-%\right]^\top \, ,\\
-%\mathbf{E} &=& 
-%\begin{bmatrix}
-%    \quad \\
-%    \mathbf{\epsilon}_{p+1}  & 
-%    \dots                & 
-%    \mathbf{\epsilon}_N \\
-%    \quad
-%\end{bmatrix}^\top \, . 
-%\end{eqnarray}
-
-\begin{equation}
- \label{eq:vareq}
-               \underbrace{ \begin{bmatrix}
-               \quad \\
-               \mathbf{y}_{p+1} &
-               \mathbf{y}_{p+2} &
-               \dots & 
-               \mathbf{y}_N \\
-               \quad
-               \end{bmatrix}}_{\substack{ \mathbf{B}^\top\\l \times (N-p)}}   
-= 
-                \underbrace{\left[ 
-                \begin{array}{ccccc}
-                \quad & \quad & \quad & \quad & \quad \\
-                \phi_1  & \phi_2 & \cdots & \phi_p & \mathbf{c} \\  
-                \quad &\quad & \quad & \quad & \quad
-               \end{array} 
-               \right]}_{\substack{ \mathbf{X}^\top\\ l \times (l \times p + 1 )}}
-\underbrace{\begin{bmatrix}
-   \mathbf{y}_{p}  & \mathbf{y}_{p+1} & \dots    & \mathbf{y}_{N-1}\\
-   \mathbf{y}_{p-1}  & \mathbf{y}_{p} & \dots    & \mathbf{y}_{N-2}\\
-   \vdots        & \vdots   & \ddots   & \vdots\\
-   \mathbf{y}_{1} & \mathbf{y}_{2}   & \dots    & \mathbf{y}_{N-p}\\
-   1 & 1   & \dots    & 1 
-   \end{bmatrix}}_{\substack{ \mathbf{A}^\top\\ (l\times p +1 )\times (N-p)}}
-+
-\underbrace{\begin{bmatrix}
-                \quad \\
-              \mathbf{\epsilon}_{p+1}  & 
-              \mathbf{\epsilon}_{p+2}  & 
-              \dots                & 
-              \mathbf{\epsilon}_N \\
-              \quad
-             \end{bmatrix}}_{\substack{\mathbf{E}^\top\\l \times (N-p) }} 
-\end{equation}
-Equation~(\ref{eq:vareq})  can be solved using ordinary least squares estimation.
-
-\subsection{Vector Error Correction}
-
-VECM is a special form of a VAR model for I(1) variables that are also
-cointegrated~\cite{banerjee1993}. It is obtained by replacing the form $\Delta
-\mathbf{y}_t = \mathbf{y}_t - \mathbf{y}_{t-1}$ in equation (\ref{eq:var}).
-VECM is expressed in terms of differences, has an error correction term and 
-the following form:
-
-\begin{equation}
- \label{eq:vec}
- \Delta \mathbf{y}_t = 
- \underbrace{ \Omega\mathbf{y}_{t-1}}_\text{Error correction term} + 
- \sum_{i=1}^{p-1}
-\phi_i^* \Delta \mathbf{y}_{t-i}  + \mathbf{c} + \mathbf{\epsilon}_t \quad ,
-\end{equation}
-
-\noindent where coefficients matrices $\Omega$ and $\phi_i^*$ are
-function of matrices $\phi_i$ (shown in equation (\ref{eq:var})) as follows:
-
-\begin{eqnarray*}
-\phi_i^* &: =& -\sum_{j=i+1}^{p} \phi_j \\
-\Omega &: =& -(\mathbb{I}-\phi_1-\dots-\phi_p) 
-\end{eqnarray*}
-
-The matrix $\Omega$ has the following properties~\cite{johansen1995}:
-\begin{itemize}
-\item If $\Omega = 0$ there is no cointegration
-\item If $rank(\Omega)=l$ i.e full rank, then the time series are not
-I(1) but stationary
-\item If $rank(\Omega)=r,\quad 0 < r < l$ then, there is cointegration
-and the matrix $\Omega$ can be expressed as $\Omega =
-\alpha \beta^\top$, where $\alpha$ and $\beta$ are $(l \times r)$
-matrices and $rank(\alpha)=rank(\beta)=r$.
-
-The columns of $\beta$ contains the cointegration vectors and the rows of
-$\alpha$ correspond with the adjusted vectors. $\beta$ is obtained by Johansen
-procedure~\cite{johansen1988} whereas $\alpha$ has to be determined as a
-variable in the VECM.
-
-It is worth noticing that the factorization of the matrix $\Omega$ is not
-unique since for any $r \times r$ nonsingular matrix $H$ we have:
-
-\begin{eqnarray*}
-\alpha \beta^\top &=& \alpha \mathbf{HH^{-1}} \beta^\top\\
-&=&(\alpha\mathbf{H})(\beta(\mathbf{H}^{-1})^\top)^\top \\
-&=& \alpha^*(\beta^*)^\top
-\end{eqnarray*}
-
-\noindent whith $\alpha^* = \alpha\mathbf{H}$ and $\beta^* =
-\beta(\mathbf{H}^{-1})^\top$.
-
-\end{itemize}
-
-If cointegration exists, then equation (\ref{eq:vec}) can be written as follows:
-
-\begin{equation}
- \label{eq:vecfull}
- \Delta \mathbf{y}_t = \alpha \beta^\top\mathbf{y}_{t-1} 
- + \sum_{i=1}^{p-1} \phi_i^*\Delta
-\mathbf{y}_{t-i}  + \mathbf{c} + \mathbf{\epsilon}_t \quad ,
-\end{equation}
-
-\noindent which is a VAR model but for time series differences.
-
-VECM matricial form is the following:
-
-%\begin{eqnarray}\label{eq:vecmatrix}
-% \mathbf{A}&=&
-%   \begin{bmatrix} 
-%   \beta^\top \mathbf{y}_{p} & 
-%   \cdots & \beta^\top \mathbf{y}_{N-1} \\
-%   \Delta \mathbf{y}_p  & \cdots 
-%   &\Delta\mathbf{y}_{N-1} \\ 
-%   \vdots  & \ddots & \vdots \\
-%   \Delta\mathbf{y}_2  & \cdots 
-%   & \Delta \mathbf{y}_{N-p+1} \\
-%   1 & \cdots & 1 
-%   \end{bmatrix}^\top \label{eq:vecA} \, ,\\
-%\mathbf{B} & = &
-% \begin{bmatrix}
-% \quad\\
-%  \Delta \mathbf{y}_{p+1} & 
-%  \dots &
-%  \Delta \mathbf{y}_N \label{eq:vecB}\\
-%  \quad
-% \end{bmatrix}^\top \, ,\\
-%\mathbf{X}&=&
-%  \begin{bmatrix}
-%   \quad \\
-%   \alpha & \phi_1^* & \cdots & \phi_{p-1}^* & \mathbf{c} \\  
-%   \quad
-%   \end{bmatrix}^\top \, ,\\
-%\mathbf{E} &=&
-%\begin{bmatrix}
-%   \quad \\
-%   \mathbf{\epsilon}_{p+1} &
-%   \dots &
-%   \quad &\mathbf{\epsilon}_N \, \\ \quad
-%\end{bmatrix}^\top 
-%\end{eqnarray}
-%
-
-\begin{equation} \label{eq:vecmatrix}
-\underbrace{
-               \begin{bmatrix}
-               \quad\\
-                \mathbf{\Delta y}_{p+1} & 
-                \dots &
-                \mathbf{\Delta y}_N \\
-                \quad
-               \end{bmatrix}
-               }_{\substack{\mathbf{B}^\top\\ l \times (N-p) }} =
-   \underbrace{
-    \begin{bmatrix}
-     \quad \\
-     \alpha & \phi_1^* & \cdots & \phi_{p-1}^* & \mathbf{c} \\  
-     \quad
-     \end{bmatrix} 
-     }_{\substack{ \mathbf{X}^\top\\ l \times (l\times p +1)}}
-\underbrace{\begin{bmatrix} 
-   \beta^\top \mathbf{y}_{p} & 
-   \cdots & \beta^\top \mathbf{y}_{N-1} \\
-   \mathbf{\Delta y}_p  & \cdots 
-   & \mathbf{\Delta y}_{N-1} \\ 
-   \vdots  & \ddots & \vdots \\
-   \mathbf{\Delta y}_2  & \cdots 
-   & \mathbf{\Delta y}_{N-p+1} \\
-   1 & \cdots & 1 
-   \end{bmatrix}}_{\substack{\mathbf{A}^\top \\ (l \times p +1) \times (N-p) }}
-+
-\underbrace{\begin{bmatrix}
-              \quad \\
-              \mathbf{\epsilon}_{p+1} &
-              \dots &
-              \quad &\mathbf{\epsilon}_N \\ \quad
-             \end{bmatrix}}_{\substack{\mathbf{E}^\top\\ l \times (N-p) }} 
-\end{equation}
-
-VAR and VECM parameters shown in equations (\ref{eq:vareq},\ref{eq:vecmatrix}) can be solved
-using standard regression techniques, such as ordinary least squares (OLS).
-
-\subsection{Unit Root Tests}
-In order to use VECM the time series are required to be I(1) which is detected
-using a unit root test. The Dickey-Fuller (ADF) test ~\cite{dickey1979} is a
-unit root test where the null hypothesis is that the
-variable has a unit root.
-
-If we have a variable with the form:
-\[
-\mathbf{x}_t = \alpha \mathbf{x}_{t-1} + \epsilon_t
-\]
-
-\noindent and if we substract $\mathbf{x}_{t-1}$ from both sides we get:
-
-\begin{equation}
-\label{eq:DF}
-\Delta \mathbf{x}_t = (\alpha-1) \mathbf{x}_{t-1} + \epsilon_t
-\end{equation}
-
-Dickey fuller test consist on determine if $\alpha$ equals 1. The alternative
-hypothesis is $\alpha < 1$ which means that $\mathbf{x}_t$ is a stationary
-process.
-
-In order to include more than 1 lagged variable the Augmented Dickey-Fuller
-test was developed and their tabulated critical values were obtained
-by~\cite{mackinnon2010}. The number of lags is often selected using Schwartz
-criterion.
-
-\subsection {Cointegration Tests}
-
-Johansen method~\cite{johansen1988} allows to find cointegration among several
-I(1) time series. This test is more general than the
-Engle-Granger~\cite{engle1987} test which only allow to find cointegration
-between two time series.  Johansen procedure takes the form of a
-likelihood-ratio test which determines $r$ cointegration relationships. The
-method provides two tests based on an eigenvector decomposition of the matrix
-$\Omega = \alpha \beta^\top$ on equation~\ref{eq:vec} called trace and maximal
-eigenvalue statistic.  Once the number of cointegration number is determined,
-$r$ more significative eigenvectors provided by Johansen procedure are used to
-get the cointegration vector matrix $\beta^\top$.  Critical values for both
-tests are limited to 12 and 5 variables in the trace and maximal eigenvalue
-statistic respectively.
-
-The trace test null hypothesis is $H_0: r = r_0$ against $H_1: r > r_0$ and the
-maximum eigenvalue test null hypothesis is $H_0: r = r_0$ against $H_1: r =
-r_0+1$ starting with $r_0 = 0$.  In the latest test, if we start with $r_0=0$
-and $H_0$ is rejected $H_1: r = 1$ will be accepted which is not true if there
-are more than one cointegration relationship, this is why this test is less
-powerful and the trace test is commonly used.
-
-\subsection{Ordinary Least Squares method}
-
-When $\mathbf{A}$ is singular, solution to equation~(\ref{eq:vecmatrix}) is given
-by the ordinary least squares (OLS) method. OLS consists of minimizing the sum
-of squared errors or equivalently minimizing the following expression:
-
-\begin{equation}
-\label{eq:regressionproblem}
-\underset{\mathbf{X}}{\text{min}} \quad \| \mathbf{A}\mathbf{\mathbf{X}} - \mathbf{B} \|_2^2
-\end{equation}
-
-\noindent for which the solution $\hat{\mathbf{X}}$ is well-known:
-
-\begin{equation}
-\label{eq:MP}
-\hat{\mathbf{X}}=\mathbf{A}^{\!\!+}\,\mathbf{B}
-\end{equation}
-
-\noindent where $\mathbf{A}^{\!\!+}$ is the Moore-Penrose pseudo-inverse
-which can be written as follows:
-
-\begin{equation}
-\label{eq:pseudoinverse}
-\mathbf{A}^{\!\!+}= (\mathbf{A}^{\!\!\top} \mathbf{A})^{-1}\mathbf{A}^{\!\!\top} \, .
-\end{equation}
-
-However, when $\mathbf{A}$ is not full rank, i.e
-$rank(\mathbf{A})=k <  n \leq m$, $\mathbf{A}^\top \mathbf{A}$ is
-always singular and equation~(\ref{eq:pseudoinverse}) cannot be used.
-More generally, the pseudo-inverse is best computed using the compact
-singular value decomposition (SVD) of $\mathbf{A}$:
-
-\begin{equation}
-    \label{eq:compactsvd}
-    \underset{m \times n}{\mathbf{A}}=
-    \underset{m \times k}{\mathbf{U_1}} \enskip
-    \underset{k \times k}{\Sigma_1} \enskip
-    \underset{k \times n}{\mathbf{V}_1^{\top}} \, ,
-\end{equation}
-
-\noindent as follows
-
-\begin{equation}
-\label{eq:pseudoinversesvd}
-\mathbf{A}^{\!\!+} = \mathbf{V}_1 \Sigma_1^{-1} \mathbf{U}_1^\top \, .
-\end{equation}
-
-%Descomentar en la tesis
-\textbf{Demo}\quad
-
-Since $\mathbf{A}$ is singular the problem shown in
-equation~(\ref{eq:regressionproblem}) has not solution, the minimum norm given
-by equation~(\ref{eq:MP}) is obtained by solving the equivalent problem:
-
-\begin{equation*}
-\label{eq:proyectorsol}
-\mathbf{A \hat{X} = PB} 
-\end{equation*}
-
-\noindent where $\mathbf{P=U_1 U_1^\top}$ is the projection onto the
-Col($\mathbf{A}$).
-
-Since $\mathbf{V} = [\underset{(n \times k)}{\mathbf{V_1}} |
-\underset{(n \times k)}{\mathbf{V_2}}]$ and $\mathbf{V_1^\top V_2 =
-0}$ we can express $\mathbf{\hat{X}} = \mathbf{V_1 x_1 + V_2 x_2}$
-with $\mathbf{x_2=0}$ because $\mathbf{\hat{X}}$ lives in the
-$\text{Row}(\mathbf{A})$ given by $\mathbf{V_1}$, so we have:
-
-\begin{eqnarray*}
-\mathbf{A \hat{X}} &=& \mathbf{PB} \\
-\mathbf{U_1 \Sigma_1 V_1^\top \hat{X}} &=& \mathbf{U_1 U_1^\top B} \\
-\mathbf{ V_1^\top \hat{X}} &=&  \mathbf{\Sigma_1^{-1} U_1^\top B} \\ 
-\mathbf{ V_1^\top V_1 x_1} &=& \mathbf{\Sigma_1^{-1}
-U_1^\top B} \\
-\mathbf{x_1}&=& \mathbf{\Sigma_1^{-1} U_1^\top B}
-\end{eqnarray*}
-
-\noindent from this result we can obtain $\mathbf{\hat{X}}$ and
-therefore the pseudo-inverse expression:
-
-\begin{eqnarray*}
-\mathbf{\hat{X}} &=& \mathbf{V_1 x_1} \\
-                &=& \mathbf{V_1 \Sigma_1^{-1} U_1^\top B} \\
-\mathbf{A^+} &=& \mathbf{V_1 \Sigma_1^{-1} U_1^\top} \, .
-\end{eqnarray*}
-
-\subsection{Ridge Regression}\label{sec:RR}
-
-Ridge regression (RR) introduces a ridge estimation to avoid the
-singularity of the matrix $\mathbf{A}^\top \mathbf{A}$
-equation~(\ref{eq:pseudoinverse}). RR solves the following
-optimization problem:
-
-\begin{equation}
-\label{eq:RRproblem}
-\underset{\mathbf{X}}{\text{min}} \quad \|
-\mathbf{A}\mathbf{\mathbf{X}} - \mathbf{B} \|_2^2 +\lambda \|
-\mathbf{\mathbf{X}}\|_2^2 
-\end{equation}
-
-\noindent where $\lambda$ is a regularization parameter.
-
-The RR optimal solution $\mathbf{X}(\lambda)$ is also well-known:
-
-\begin{equation}
-\label{eq:optsolRR}
-\mathbf{X}(\lambda)=(\mathbf{A}^\top \mathbf{A}+ \lambda
-\mathbb{I})^{-1}\mathbf{A}^\top \mathbf{B} \, . 
-\end{equation}
-
-\textbf{Demo}\quad
-
-To solve equation~(\ref{eq:optsolRR}) is equivalent to solve the following
-optimization problem:
-
-\begin{equation}
-\label{eq:RRproblem2}
-\underset{\mathbf{X}(\lambda)}{\text{min}} \quad \|
-\mathbf{C}\mathbf{\mathbf{X}} - \mathbf{F} \|_2^2
-\end{equation}
-
-\noindent where
-
-\begin{equation*}
-	\mathbf{C}=\begin{bmatrix} \quad &\mathbf{A}& \quad \\ \hdashline \quad
-& \lambda\mathbb{I}&\quad  \end{bmatrix} \quad \text{and} \quad
-	\mathbf{F}=\begin{bmatrix} \quad &\mathbf{B}& \quad \\ \hdashline \quad
-& \mathbf{0}&\quad  \end{bmatrix} \, .
-\end{equation*}
-
-Applying equation~(\ref{eq:MP}) and considering that $\mathbf{C}^\top
-\mathbf{C} = \mathbf{A}^\top \mathbf{A} + \lambda \mathbb{I}$ and 
-$\mathbf{C}^\top \mathbf{F}=\mathbf{A}^\top \mathbf{B} $ we have:
-
-\begin{eqnarray*}
-\mathbf{X}(\lambda)&=&(\mathbf{C}^\top
-\mathbf{C})^{-1}\mathbf{C}^\top \mathbf{F} \\
-&=& (\mathbf{A}^\top \mathbf{A} + \lambda \mathbb{I})^{-1} \mathbf{A}^\top \mathbf{B}
-\end{eqnarray*}
-
-\subsection{Online Ridge regression}
-
-If we represent matrices $\mathbf{A}$ and $\mathbf{B}$ such as:
-
-\begin{equation}
-\label{eq:notation}
-	\mathbf{A} = 
-\left[
-  \begin{tabular}{c>{$}c<{$}c}
-    --- & \mathbf{a}^{\top}_{1} & ---\\
-    --- & \mathbf{a}^{\top}_{2} & ---\\
-    & \vdots & \\
-    --- & \mathbf{a}^{\top}_{m} & ---
-  \end{tabular}
-\right]
-\quad \text{and} \quad
-\mathbf{B} =
-\left[
-  \begin{tabular}{c>{$}c<{$}c}
-    --- & \mathbf{b}_{1} & ---\\
-    --- & \mathbf{b}_{2} & ---\\
-    & \vdots & \\
-    --- & \mathbf{b}_{m} & ---
-  \end{tabular}
-\right] \, ,
-\
-\end{equation}
-
-\noindent equation~(\ref{eq:optsolRR}) can also be written as:
-
-\begin{eqnarray*}
-\label{eq:RReapand}
-\mathbf{\mathbf{X}}_{\text{ridge}}&=&(\mathbf{A}^\top \mathbf{A}+ \lambda
-\mathbb{I})^{-1}\mathbf{A}^\top \mathbf{B} \\
-&=& \displaystyle \big (\sum_{t=1}^m
-\mathbf{a}_t \mathbf{a}_t  ^\top + \lambda \mathbb{I}\big )^{-1}
-\sum_{t=1}^m \mathbf{a}_t \mathbf{b}_t \, .
-\end{eqnarray*}
-
-Lets define $\displaystyle\mathbf{S}= \sum_{t=1}^m \mathbf{a}_t
-\mathbf{a}_t  ^\top + \lambda \mathbb{I} $ and $\mathbf{W}=
-\displaystyle\sum_{t=1}^m \mathbf{a}_t \mathbf{b}_t$, so the
-algorithm~\ref{alg:RR} shows the iterative formulation:
-
-\begin{algorithm}[H]
-\begin{algorithmic}[1]
-\REQUIRE $\,$ \\
-$\{\mathbf{a}_1,\dots,\mathbf{a}_{m} \}$: $m$ input vectors \\
-$\{\mathbf{b}_1,\dots,\mathbf{b}_{m} \}$: $m$ target vectors \\
-$\lambda$: regularization parameter \\
-\ENSURE  $\,$ \\
-$\{f(\mathbf{a}_1),\dots,f(\mathbf{a}_{m}) \}$: model predictions \\
-\STATE Initialize $\mathbf{S}=\lambda \mathbb{I}$
-and $\mathbf{W}=0$
-\FOR { $t = 1$ to $m$ }
-	\STATE read new $\mathbf{a}_t$
-	\STATE $\mathbf{X}=\mathbf{S}^{-1}\mathbf{W}$
-	\STATE output prediction $f(\mathbf{a}_t) = \mathbf{X}^\top \mathbf{a}_t$
-   	\STATE $\mathbf{S} = \mathbf{S} + \mathbf{a}_t \mathbf{a}_t^\top$
-   	\STATE Read new $y_t$
-    	\STATE $\mathbf{W} = \mathbf{W} + \mathbf{a}_t \mathbf{b}_t$
-\ENDFOR
-\end{algorithmic}
-\caption{Online Ridge Regression}
-\label{alg:RR}
-\end{algorithm}
-
-\subsection{The Aggregating Algorithm for Regression}
-
-The AAR, proposed by~\cite{vovk2001}, is an application of the aggregating
-algorithm to the problem of regression. The idea is to introduce the new input
-vector $\mathbf{x}_{m+1}$ to solve the model parameters:
-
-\begin{equation}
-\label{eq:AARexpand}
-\mathbf{X}_{aar} = \displaystyle \big (\sum_{t=1}^{m+1}
-\mathbf{a}_t \mathbf{a}_t  ^\intercal + \gamma \mathbb{I}\big )^{-1}
-\sum_{t=1}^m \mathbf{a}_t \mathbf{b}_t \, .
-\end{equation}
-
-If we define $\displaystyle\mathbf{S}= \sum_{t=1}^{m+1} \mathbf{a}_t
-\mathbf{a}_t  ^\intercal + \gamma \mathbb{I} $ and $\mathbf{W}=
-\displaystyle\sum_{t=1}^m \mathbf{a}_t \mathbf{b}_t$, the
-algorithm~\ref{alg:AAR} is slightly different to the algorithm~\ref{alg:RR}, 
-which updated matrix $\mathbf{S}$ before making the prediction:
-
-\begin{algorithm}[ht]
-\begin{algorithmic}[1]
-\REQUIRE $\,$ \\
-$\{\mathbf{a}_1,\dots,\mathbf{a}_{m} \}$: $m$ input vectors \\
-$\{\mathbf{b}_1,\dots,\mathbf{b}_{m} \}$: $m$ target vectors \\
-$\lambda$: regularization parameter \\
-\ENSURE  $\,$ \\
-$\{f(\mathbf{a}_1),\dots,f(\mathbf{a}_{m}) \}$: model predictions \\
-\STATE Initialize $\mathbf{S}=\lambda \mathbb{I}$
-and $\mathbf{W}=0$
-\FOR { $t = 1$ to $m$ }
-	\STATE read new $\mathbf{a}_t$
-   	\STATE $\mathbf{S} = \mathbf{S} + \mathbf{a}_t \mathbf{a}_t^\intercal$
-	\STATE $\mathbf{X}=\mathbf{S}^{-1}\mathbf{W}$
-	\STATE output prediction $f(\mathbf{a}_t) = \mathbf{X}^\intercal \mathbf{a}_t$
-   	\STATE Read new $\mathbf{y}_t$
-    	\STATE $\mathbf{W} = \mathbf{W} + \mathbf{a}_t \mathbf{b}_t$
-\ENDFOR
-\end{algorithmic}
-\caption{{\em The aggregating algorithm for regression}}
-\label{alg:AAR}
-\end{algorithm}
-
-\subsection{Selection of Lambda}
+\section{Selection of Lambda}
 
@@
 \begin{figure}[h!]
-\includegraphics[width=0.5\linewidth]{img/shrinks}
+\includegraphics[width=0.5\linewidth]{plots/shrinks}
 \caption{Shrink of regression coefficients}
@@
 
-\subsubsection{The bias-variance tradeoff}
+\subsection{The bias-variance tradeoff}
 
@@
 \begin{figure}[h!]
-\includegraphics[width=0.8\linewidth]{img/biasvariance}
+\includegraphics[width=0.8\linewidth]{plots/biasvariance}
 \caption{Bias-variance tradeoff}
@@
 It can be shown that, in terms of prediction error, ridge (black line)
-is lower than OLS (dotted line)~\cite{hoerl1970}.
+is lower than OLS (dotted line)~\cite{hoerl+kennard1970}.
 
-\subsubsection{Methods for lambda selection}
+\subsection{Selection of lambda}
 
@@
 is shown in equation~(\ref{eq:taylor}). Golub (~\cite{golub1965}) suggested that
-$\lambda$ had to be chosen so that:
+$\lambda$ be chosen so that:
 
@@
 \end{enumerate}
-
-\subsection{Efficient computation}
-
-%Revisar!!!
-In regression problems we always require getting a matrix inverse which is
-computationally expensive. However, in stream data problems the is a way to
-obtain a matrix inverse approximation using the Sherman-Morrison-Woodbury. This
-formula allows to get the inverse of a
-the matrix $\mathbf{A+uv^\top}$ if we already calculated the
-inverse of $\mathbf{A}$ as follows:
-
-\begin{equation}
-\label{eq:SMW}
-(\mathbf{A+uv^\top})^{-1}=\mathbf{A}^{-1}-
-\frac{\mathbf{A^{-1}uv^\top A^{-1}}}{\mathbf{1+v^\top A^{-1}u}}
-\end{equation}
-
-%Revisar!!
-
-Another way was presented by Coleman and
-Sun~\cite{coleman+sun2010} using an iterative algorithm which
-uses $\mathbf{X}(\lambda)$ to approximate $\mathbf{X}(0)$.
-
-Using the compact SVD (shown in equation~(\ref{eq:compactsvd}))
-$\mathbf{X}(\lambda)$ is expressed as follows:
-
-\begin{eqnarray}
-\label{eq:optsolRRsvd}
-\mathbf{X}(\lambda) & = & (\mathbf{A}^\top \mathbf{A}+ \lambda
-\mathbb{I})^{-1}\mathbf{A}^\top \mathbf{B} \nonumber \\
-& = &\mathbf{V}_1(\Sigma_1^2+\lambda \mathbb{I})^{-1}\mathbf{\Sigma_1
-U_1^\top B}
-\end{eqnarray}
-
-\noindent where is easy to see that $\mathbf{X}(\lambda) \rightarrow
-\mathbf{\hat{X}}=\mathbf{V}_1 \mathbf{\Sigma}_1^{-1}\mathbf{U}_1^\top
-\mathbf{B}$ as $\lambda \rightarrow 0$.
-
-The method consists in obtaining $\mathbf{X}(\lambda)$ and then refine
-by adding more terms of its Taylor expansion to approximate
-$\mathbf{X}(0) = \mathbf{\hat{X}}$. The Taylor expansion about $\lambda_0$ is:
-
-\begin{equation}
-\label{eq:taylor}
-    \mathbf{X}(\lambda)=\mathbf{X}(\lambda_0) + \sum_{k=1}^\infty
-    \mathbf{s}_k(\lambda-\lambda_0)^{k}
-\end{equation}
-
-\noindent where $\mathbf{s}_k=\frac{1}{k!}\mathbf{X}(\lambda)^{(k)}$
-and $\mathbf{X}(\lambda)^{(k)}$ is obtained by taking differences 
-$\frac{\partial}{\partial \lambda}$ of:
-
-\begin{eqnarray*}
-(\mathbf{A}^\top \mathbf{A}+ \lambda\mathbb{I}) \mathbf{X}(\lambda) & = & \mathbf{A}^\top \mathbf{B}\\
-(\mathbf{A}^\top \mathbf{A}+ \lambda\mathbb{I}) \mathbf{X}(\lambda)^{(1)} + \mathbf{X}(\lambda)& = & 0 \\
-\mathbf{X}(\lambda)^{(1)}  &=& -(\mathbf{A}^\top \mathbf{A}+ \lambda\mathbb{I}) ^{-1} \mathbf{X}(\lambda) \\
-\mathbf{X}(\lambda)^{(2)}  &=& -2(\mathbf{A}^\top \mathbf{A}+ \lambda\mathbb{I}) ^{-1} \mathbf{X}(\lambda)^{(1)} \\
-& \vdots & \\
-\mathbf{X}(\lambda)^{(k)}  &=& -k(\mathbf{A}^\top \mathbf{A}+ \lambda\mathbb{I}) ^{-1} \mathbf{X}(\lambda)^{(k-1)} 
-\end{eqnarray*}
-
-\noindent for $\lambda=0$ we have:
-
-\begin{equation}
-\label{eq:taylor}
-    \mathbf{X}(0)=\mathbf{X}(\lambda_0) + \sum_{k=1}^\infty
-     (-1)^k \mathbf{s}_k \lambda_0^k
-\end{equation}
-
-\noindent therefore in order to ensure convergence, we can see that$\lambda_0$
-selection cannot be large.
-
-The algorithm for computing $\mathbf{X}(0)$ is the following:
-
-\begin{algorithm}[H]
-\begin{algorithmic}[1]
-\REQUIRE $\,$ \\
-$\mathbf{A}$: design matrix \\
-$\mathbf{B}$: response matrix \\
-$\lambda$: rank deficient parameter \\
-\ENSURE  $\,$ \\
-$\mathbf{X}$: parameters \\
-\STATE $\mathbf{M}=\mathbf{A^\top A}$ \\
-\STATE Initialize $\mathbf{Q R}=\mathbf{M}$ \\
-\STATE $\mathbf{X} = \mathbf{R^{-1}Q^\top B}$ \\
-\STATE $\mathbf{s} = \mathbf{X}$ \\
-\FOR { $i = 1,2,3,\dots$ }
-	\STATE $\mathbf{s} =
-        -(\mathbf{M}+\lambda\mathbb{I})^{-1}\mathbf{s}$\\
-	\STATE $\mathbf{X}=\mathbf{X} + (-1)^i {\color{red}\mathbf{s}
-        \lambda^i}$
-\ENDFOR
-\end{algorithmic}
-\caption{Algorithm for handling rank deficient matrices}
-\label{alg:coleman}
-\end{algorithm}
-
-The algorithm~\ref{alg:coleman} solves equation~(\ref{eq:taylor}). However, this
-version is unstable since tipically $\|\mathbf{s}\|$ is very large and
-$\lambda^i$ is very small ($\lambda$ is small).
-
-The following algorithm shows a more stable version of
-algorithm~\ref{alg:coleman}.
-
-\begin{algorithm}[H]
-\begin{algorithmic}[1]
-\REQUIRE $\,$ \\
-$\mathbf{A}$: design matrix \\
-$\mathbf{B}$: response matrix \\
-$\lambda$: rank deficient parameter \\
-\ENSURE  $\,$ \\
-$\mathbf{X}$: parameters \\
-\STATE $\mathbf{M}=\mathbf{A^\top A}$ \\
-\STATE Initialize $\mathbf{Q R}=\mathbf{M}$ \\
-\STATE $\mathbf{X} = \mathbf{R^{-1}Q^\top B}$ \\
-\STATE $\mathbf{t} = \mathbf{X}$ \\
-\FOR { $i = 1,2,3,\dots$ }
-        \STATE $\mathbf{t} =\lambda \mathbf{t}$  
-        \STATE $\mathbf{t} =  -(\mathbf{M}+\lambda\mathbb{I})^{-1}\mathbf{t}$
-	\STATE $\mathbf{X}=\mathbf{X} + \mathbf{t}$
-\ENDFOR
-\end{algorithmic}
-\caption{Algorithm for handling rank deficient matrices improved}
-\label{alg:colemanimproved}
-\end{algorithm}
-
-Both algorithms are equivalent, but algorithm~\ref{alg:colemanimproved} is more
-stable and converges in tipically less than 10 steps.
-It is important to notice that the QR factorization is computed only once and it
-is computationally less expensive than the SVD.
-
-\subsection{Online machine learning algorithms}
-
-An online algorithm allows incremental learning by processing one instance at a
-time. This is done updating the current model instead of building the model from
-scratch.
-
-Training phase is commonly a computationally expensive process. Therefore, when
-new data arrives it can't be included easily to the model. Moreover, it could
-happen that we won't have enought time to process the new data before more data
-arrives. In online learning algorithms there is no training phase, but the model
-is updated and evaluated at every time step. This model updating is
-computationally less expensive than a training phase.
-
-Online learning is specially useful for stream data problems and the model is
-sensitive to new data. It is also useful when some past data may be irrelevant
-or we want to improve computational efficiency. However, online algorithms could
-affect accuracy.
-
-There are several popular online methods such as
-perceptron~\cite{rosenblatt58},winnow~\cite{littlestone1988},
-passive-aggressive~\cite{crammerETall2006}, stochastic gradient
-descent~\cite{zhang2004}, aggregating algorithm~\cite{vovk2001} and the second
-order perceptron~\cite{cesa-bianchi2005}.  In \cite{blum1998} and
-\cite{cesa-bianchi2006} an in-deph analysis of online learning is provided.
-Applications in finance has been widely used: study presented
-by~\cite{arce+salinas2012} applied ridge regression in an online context and
-more recently, time series forecasting using online learning has been
-presented~\cite{anavaetAl2013}.
-
-Incremental learning refers to any online learning process that learns the same
-model as would be learnt by a batch learning algorithm.
-
-Incremental learning is useful when the input to a learning process is stream
-data, with the need or desire to be able to use the result of learning at any
-point in time, based on the input observations received so far.
-
-In principle, the stream of observations may be infinitely long, or the next
-observation long delayed, precluding any hope of waiting until all the
-observations have been received. One would rather not simply accumulate and
-store all the inputs and, upon receipt of each new one, apply a batch learning
-algorithm to the entire sequence of inputs received so far. It would be
-preferable compu- tationally if the existing hypothesis or other artifact of
-learning could be updated in response to each newly received input observation.
-
-Incremental learning is very useful when there is no need to record fundamental
-data and only a summary needs to be retained. Due to this, incremental
-algorithms are often characterized as memoryless, because no memory of past data
-is required.  The algorithm is online but not incremental if it doesn't produce
-the same result for all observations that the corresponding batch algorithm
-would for these same observations.
-
-Algorithm~\ref{alg:onlinealg} shows the online learning algorithm structure:
-
-\begin{algorithm}[ht]
-\begin{algorithmic}[1]
-    \STATE Receives input $\mathbf{x}_t$
-    \STATE Makes prediction $\mathbf{\hat{y}}_t$
-    \STATE Receives response $\mathbf{y}_t$
-    \STATE Incurs loss $l_t(\mathbf{y}_t,\mathbf{\hat{y}}_t)$
-\end{algorithmic}
-\caption{Structure of a Learning System}
-\label{alg:onlinealg}
-\end{algorithm}
-
-\noindent where $l$ is some loss function. Performance is later measured after
-$T$ trials as:
-
-\begin{equation*}
-L_T = \sum_{t=1}^T l_t(\mathbf{y}_t,\mathbf{\hat{y}}_t)
-\end{equation*}
-
-The objective is minimize this loss function for all instances.
-The quality of online learning algorithms is measured by a quantity known as
-regret which is the difference between the performance of the online algorithm
-and its optimal predictor $E^* \in \Theta$ given by:
-
-\begin{equation*}
-L^*_T= \text{min}_{E \in \Theta} L_T^E \, ,
-\end{equation*}
-
-\noindent where $L_T^E = \sum_{t=1}^T
-l_t(\mathbf{y}_t,\mathbf{y}^E_t)$ and $\mathbf{y}^E_t$ is the expert estimation.
-
-Therefore regret is defined as:
-
-\begin{equation*}
-R_T = L_T - L^*_T
-\end{equation*}
-
-%An online learning algorithm looks at every example exactly once.
-\subsection{Model selection}
-In order to set model parameters the Akaike Information Criterion (AIC) is
-used. AIC is often used in model selection where AIC whith smaller values are
-preferred.
-
-AIC is calculated as follows:
-
-\begin{equation}
-\label{eq:aicformula}
-AIC = \underset{\text{bias}}{-\frac{2l}{N}} + 
-\underset{\text{variance}}{\frac{2k}{N}}
-\end{equation}
-
-\noindent where
-
-\begin{description}
-\item[l] is the loglikelihood function
-\item[k] number of estimated parameters (including the variance)
-\item[N] number of observations
-\end{description}

### 2016-03-22 articles-conferences/online-vec-article-2014/src/refinements.tex
Agregar estos repos a la tesis

--- 2016-03-22 articles-conferences/elsevier-journal-2015/src/background.tex
@@
-\section{\uppercase{Background}}
-\label{sec:background}
-\noindent
+\section{Efficient computation}
 
-\subsection{Spurious Regression}
-
-Regression analysis is widely used to test economic theories. But sometimes
-regression is done between two variables, lets say $x_t$ and $y_t$, which are
-not causally related. This occurs at the presence of a third stochastic variable
-$z_t$ which is the true causal of the first two variables $x_t$ and $y_t$. The
-regression measure R-squared gives high values in presence of spurious
-relationships. This measure would means that the model explains all the
-variability of the response data around its mean which is not true.
-
-The R-squared measure is calculated as follows:
-
-\begin{equation}
-\label{eq:rsquared}
-R^2 = 1 - \frac{\sum_{t=1}^T \epsilon_t^2}{\sum_{t=1}^T(y_t - \bar{y})^2}
-\end{equation}
-
-In presence of spurious relationships the denominator of equation
-\ref{eq:rsquared} becomes very large because a large weight is placed on extreme
-observations on either side of the mean $\bar{y}$.
-In order to detect this, other statistic must be considered such as the
-Durbin-Watson statistic.
-
-% Acá agregar gráfico de spurious regression
-
-\subsection{Integration and Cointegration}
-A time series $\mathbf{y}$ is said to be integrated of order $d$ if after
-differentiating the variable $d$ times, we get an I(0) process, more precisely:
-
-\[
-(1-L)^d \mathbf{y} \sim \text{I(0)} \, ,
-\]
-
-\noindent where I(0) is a stationary time series and $L$ is the lag operator:
-
-\[
-(1-L)\mathbf{y} = \Delta \mathbf{y}=\mathbf{y}_t  -\mathbf{y}_{t-1} \quad \forall t
-\]
-
-Let $\mathbf{y}_t = \{\mathbf{y}^1, \dots, \mathbf{y}^l\}$ be a set of $l$
-stationary time series I(1) which are said to be cointegrated if a vector
-$\beta=[\beta(1),\dots,\beta(l)]^\top \in \mathbb{R}^l$  exists such that the
-time series,
-
-\begin{equation}
- \mathbf{Z}_t:= \beta^\top \mathbf{y}_t = \beta(1) \mathbf{y}^1 + \dots + \beta(l) \mathbf{y}^l \sim
- \text{I(0)}\, .
-\end{equation}
-
-In other words, a set of I(1) variables is said to be cointegrated if
-a linear combination of them exists which is I(0).
-
-\subsection{Vector Autorregresive Models}\label{sec:varvec}
-
-VECM is a special case of VAR model and both describe the joint behaviour of a
-set of variables.
-
-VAR($p$) model is a general framework to describe the behaviour of a set of $l$
-endogenous variables as a linear combination of their last $p$ values. These
-$l$ variables at time $t$ are represented by the vector $\mathbf{y}_t$ as
-follows:
-
-\begin{equation}
-\label{eq:variables}
-\mathbf{y}_t = 
-\begin{bmatrix} y_{1,t} &
-y_{2,t} &
-\dots &
-y_{l,t}
-\end{bmatrix}^\top \, ,
-\end{equation}
-\noindent where $y_{j,t}$ corresponds to the time series $j$ evaluated at
-time $t$.
-
-The VAR(p) model describes the behaviour of a dependent variable in terms
-of its own lagged values and the lags of the others variables in the
-system. The model with $p$ lags is formulated as the following:
-
-\begin{equation}
-\label{eq:var}
- \mathbf{y}_t = \phi_1 \mathbf{y}_{t-1}  + \dots +   \phi_p\mathbf{y}_{t-p}
-+ \mathbf{c} + \mathbf{\epsilon}_t \, ,
-\end{equation}
-
-\noindent where $t=p+1, \dots N$,  ${\phi_1,\dots,\phi_p}$ are $l \times l$
-matrices of real coefficients,
-$\mathbf{\epsilon}_{p+1},\dots,\mathbf{\epsilon}_N$ are error terms,
-$\mathbf{c}$ is a constant vector and $N$ is the total number of samples.
-
-The VAR matrix form of equation~(\ref{eq:var}) is:
-
-%\begin{equation}\label{eq:vareq}
-%\mathbf{B} = \mathbf{A} \mathbf{X} + 
-%\mathbf{E} \, , 
-%\end{equation}
-%
-%\noindent where:
-%\begin{eqnarray}
-%\mathbf{A} &=&
-%\begin{bmatrix}
-%   \mathbf{y}_{p}     & \dots    & \mathbf{y}_{N-1}\\
-%   \mathbf{y}_{p-1}   & \dots    & \mathbf{y}_{N-2}\\
-%   \vdots             & \ddots   & \vdots\\
-%   \mathbf{y}_{1}     & \dots    & \mathbf{y}_{N-p}\\
-%   1                  & \dots    & 1 
-%   \end{bmatrix}^\top \, ,\\
-%\mathbf{B} &=&
-%\begin{bmatrix}
-%\quad \\
-%\mathbf{y}_{p+1} &
-%\dots & 
-%\mathbf{y}_N \\
-%\quad
-%\end{bmatrix}^\top  \, ,\\
-%\mathbf{X}&=&
-%\left[
-%\begin{array}{ccccc}
-%    \quad  & \quad & \quad & \quad \\
-%    \phi_1   & \cdots & \phi_p & \mathbf{c} \\  
-%    \quad & \quad & \quad & \quad
-%    \end{array}
-%\right]^\top \, ,\\
-%\mathbf{E} &=& 
-%\begin{bmatrix}
-%    \quad \\
-%    \mathbf{\epsilon}_{p+1}  & 
-%    \dots                & 
-%    \mathbf{\epsilon}_N \\
-%    \quad
-%\end{bmatrix}^\top \, . 
-%\end{eqnarray}
-
-\begin{equation}
- \label{eq:vareq}
-               \underbrace{ \begin{bmatrix}
-               \quad \\
-               \mathbf{y}_{p+1} &
-               \mathbf{y}_{p+2} &
-               \dots & 
-               \mathbf{y}_N \\
-               \quad
-               \end{bmatrix}}_{\substack{ \mathbf{B}^\top\\l \times (N-p)}}   
-= 
-                \underbrace{\left[ 
-                \begin{array}{ccccc}
-                \quad & \quad & \quad & \quad & \quad \\
-                \phi_1  & \phi_2 & \cdots & \phi_p & \mathbf{c} \\  
-                \quad &\quad & \quad & \quad & \quad
-               \end{array} 
-               \right]}_{\substack{ \mathbf{X}^\top\\ l \times (l \times p + 1 )}}
-\underbrace{\begin{bmatrix}
-   \mathbf{y}_{p}  & \mathbf{y}_{p+1} & \dots    & \mathbf{y}_{N-1}\\
-   \mathbf{y}_{p-1}  & \mathbf{y}_{p} & \dots    & \mathbf{y}_{N-2}\\
-   \vdots        & \vdots   & \ddots   & \vdots\\
-   \mathbf{y}_{1} & \mathbf{y}_{2}   & \dots    & \mathbf{y}_{N-p}\\
-   1 & 1   & \dots    & 1 
-   \end{bmatrix}}_{\substack{ \mathbf{A}^\top\\ (l\times p +1 )\times (N-p)}}
-+
-\underbrace{\begin{bmatrix}
-                \quad \\
-              \mathbf{\epsilon}_{p+1}  & 
-              \mathbf{\epsilon}_{p+2}  & 
-              \dots                & 
-              \mathbf{\epsilon}_N \\
-              \quad
-             \end{bmatrix}}_{\substack{\mathbf{E}^\top\\l \times (N-p) }} 
-\end{equation}
-Equation~(\ref{eq:vareq})  can be solved using ordinary least squares estimation.
-
-\subsection{Vector Error Correction}
-
-VECM is a special form of a VAR model for I(1) variables that are also
-cointegrated~\cite{banerjee1993}. It is obtained by replacing the form $\Delta
-\mathbf{y}_t = \mathbf{y}_t - \mathbf{y}_{t-1}$ in equation (\ref{eq:var}).
-VECM is expressed in terms of differences, has an error correction term and 
-the following form:
-
-\begin{equation}
- \label{eq:vec}
- \Delta \mathbf{y}_t = 
- \underbrace{ \Omega\mathbf{y}_{t-1}}_\text{Error correction term} + 
- \sum_{i=1}^{p-1}
-\phi_i^* \Delta \mathbf{y}_{t-i}  + \mathbf{c} + \mathbf{\epsilon}_t \quad ,
-\end{equation}
-
-\noindent where coefficients matrices $\Omega$ and $\phi_i^*$ are
-function of matrices $\phi_i$ (shown in equation (\ref{eq:var})) as follows:
-
-\begin{eqnarray*}
-\phi_i^* &: =& -\sum_{j=i+1}^{p} \phi_j \\
-\Omega &: =& -(\mathbb{I}-\phi_1-\dots-\phi_p) 
-\end{eqnarray*}
-
-The matrix $\Omega$ has the following properties~\cite{johansen1995}:
-\begin{itemize}
-\item If $\Omega = 0$ there is no cointegration
-\item If $rank(\Omega)=l$ i.e full rank, then the time series are not
-I(1) but stationary
-\item If $rank(\Omega)=r,\quad 0 < r < l$ then, there is cointegration
-and the matrix $\Omega$ can be expressed as $\Omega =
-\alpha \beta^\top$, where $\alpha$ and $\beta$ are $(l \times r)$
-matrices and $rank(\alpha)=rank(\beta)=r$.
-
-The columns of $\beta$ contains the cointegration vectors and the rows of
-$\alpha$ correspond with the adjusted vectors. $\beta$ is obtained by Johansen
-procedure~\cite{johansen1988} whereas $\alpha$ has to be determined as a
-variable in the VECM.
-
-It is worth noticing that the factorization of the matrix $\Omega$ is not
-unique since for any $r \times r$ nonsingular matrix $H$ we have:
-
-\begin{eqnarray*}
-\alpha \beta^\top &=& \alpha \mathbf{HH^{-1}} \beta^\top\\
-&=&(\alpha\mathbf{H})(\beta(\mathbf{H}^{-1})^\top)^\top \\
-&=& \alpha^*(\beta^*)^\top
-\end{eqnarray*}
-
-\noindent whith $\alpha^* = \alpha\mathbf{H}$ and $\beta^* =
-\beta(\mathbf{H}^{-1})^\top$.
-
-\end{itemize}
-
-If cointegration exists, then equation (\ref{eq:vec}) can be written as follows:
-
-\begin{equation}
- \label{eq:vecfull}
- \Delta \mathbf{y}_t = \alpha \beta^\top\mathbf{y}_{t-1} 
- + \sum_{i=1}^{p-1} \phi_i^*\Delta
-\mathbf{y}_{t-i}  + \mathbf{c} + \mathbf{\epsilon}_t \quad ,
-\end{equation}
-
-\noindent which is a VAR model but for time series differences.
-
-VECM matricial form is the following:
-
-%\begin{eqnarray}\label{eq:vecmatrix}
-% \mathbf{A}&=&
-%   \begin{bmatrix} 
-%   \beta^\top \mathbf{y}_{p} & 
-%   \cdots & \beta^\top \mathbf{y}_{N-1} \\
-%   \Delta \mathbf{y}_p  & \cdots 
-%   &\Delta\mathbf{y}_{N-1} \\ 
-%   \vdots  & \ddots & \vdots \\
-%   \Delta\mathbf{y}_2  & \cdots 
-%   & \Delta \mathbf{y}_{N-p+1} \\
-%   1 & \cdots & 1 
-%   \end{bmatrix}^\top \label{eq:vecA} \, ,\\
-%\mathbf{B} & = &
-% \begin{bmatrix}
-% \quad\\
-%  \Delta \mathbf{y}_{p+1} & 
-%  \dots &
-%  \Delta \mathbf{y}_N \label{eq:vecB}\\
-%  \quad
-% \end{bmatrix}^\top \, ,\\
-%\mathbf{X}&=&
-%  \begin{bmatrix}
-%   \quad \\
-%   \alpha & \phi_1^* & \cdots & \phi_{p-1}^* & \mathbf{c} \\  
-%   \quad
-%   \end{bmatrix}^\top \, ,\\
-%\mathbf{E} &=&
-%\begin{bmatrix}
-%   \quad \\
-%   \mathbf{\epsilon}_{p+1} &
-%   \dots &
-%   \quad &\mathbf{\epsilon}_N \, \\ \quad
-%\end{bmatrix}^\top 
-%\end{eqnarray}
-%
-
-\begin{equation} \label{eq:vecmatrix}
-\underbrace{
-               \begin{bmatrix}
-               \quad\\
-                \mathbf{\Delta y}_{p+1} & 
-                \dots &
-                \mathbf{\Delta y}_N \\
-                \quad
-               \end{bmatrix}
-               }_{\substack{\mathbf{B}^\top\\ l \times (N-p) }} =
-   \underbrace{
-    \begin{bmatrix}
-     \quad \\
-     \alpha & \phi_1^* & \cdots & \phi_{p-1}^* & \mathbf{c} \\  
-     \quad
-     \end{bmatrix} 
-     }_{\substack{ \mathbf{X}^\top\\ l \times (l\times p +1)}}
-\underbrace{\begin{bmatrix} 
-   \beta^\top \mathbf{y}_{p} & 
-   \cdots & \beta^\top \mathbf{y}_{N-1} \\
-   \mathbf{\Delta y}_p  & \cdots 
-   & \mathbf{\Delta y}_{N-1} \\ 
-   \vdots  & \ddots & \vdots \\
-   \mathbf{\Delta y}_2  & \cdots 
-   & \mathbf{\Delta y}_{N-p+1} \\
-   1 & \cdots & 1 
-   \end{bmatrix}}_{\substack{\mathbf{A}^\top \\ (l \times p +1) \times (N-p) }}
-+
-\underbrace{\begin{bmatrix}
-              \quad \\
-              \mathbf{\epsilon}_{p+1} &
-              \dots &
-              \quad &\mathbf{\epsilon}_N \\ \quad
-             \end{bmatrix}}_{\substack{\mathbf{E}^\top\\ l \times (N-p) }} 
-\end{equation}
-
-VAR and VECM parameters shown in equations (\ref{eq:vareq},\ref{eq:vecmatrix}) can be solved
-using standard regression techniques, such as ordinary least squares (OLS).
-
-\subsection{Unit Root Tests}
-In order to use VECM the time series are required to be I(1) which is detected
-using a unit root test. The Dickey-Fuller (ADF) test ~\cite{dickey1979} is a
-unit root test where the null hypothesis is that the
-variable has a unit root.
-
-If we have a variable with the form:
-\[
-\mathbf{x}_t = \alpha \mathbf{x}_{t-1} + \epsilon_t
-\]
-
-\noindent and if we substract $\mathbf{x}_{t-1}$ from both sides we get:
-
-\begin{equation}
-\label{eq:DF}
-\Delta \mathbf{x}_t = (\alpha-1) \mathbf{x}_{t-1} + \epsilon_t
-\end{equation}
-
-Dickey fuller test consist on determine if $\alpha$ equals 1. The alternative
-hypothesis is $\alpha < 1$ which means that $\mathbf{x}_t$ is a stationary
-process.
-
-In order to include more than 1 lagged variable the Augmented Dickey-Fuller
-test was developed and their tabulated critical values were obtained
-by~\cite{mackinnon2010}. The number of lags is often selected using Schwartz
-criterion.
-
-\subsection {Cointegration Tests}
-
-Johansen method~\cite{johansen1988} allows to find cointegration among several
-I(1) time series. This test is more general than the
-Engle-Granger~\cite{engle1987} test which only allow to find cointegration
-between two time series.  Johansen procedure takes the form of a
-likelihood-ratio test which determines $r$ cointegration relationships. The
-method provides two tests based on an eigenvector decomposition of the matrix
-$\Omega = \alpha \beta^\top$ on equation~\ref{eq:vec} called trace and maximal
-eigenvalue statistic.  Once the number of cointegration number is determined,
-$r$ more significative eigenvectors provided by Johansen procedure are used to
-get the cointegration vector matrix $\beta^\top$.  Critical values for both
-tests are limited to 12 and 5 variables in the trace and maximal eigenvalue
-statistic respectively.
-
-The trace test null hypothesis is $H_0: r = r_0$ against $H_1: r > r_0$ and the
-maximum eigenvalue test null hypothesis is $H_0: r = r_0$ against $H_1: r =
-r_0+1$ starting with $r_0 = 0$.  In the latest test, if we start with $r_0=0$
-and $H_0$ is rejected $H_1: r = 1$ will be accepted which is not true if there
-are more than one cointegration relationship, this is why this test is less
-powerful and the trace test is commonly used.
-
-\subsection{Ordinary Least Squares method}
-
-When $\mathbf{A}$ is singular, solution to equation~(\ref{eq:vecmatrix}) is given
-by the ordinary least squares (OLS) method. OLS consists of minimizing the sum
-of squared errors or equivalently minimizing the following expression:
-
-\begin{equation}
-\label{eq:regressionproblem}
-\underset{\mathbf{X}}{\text{min}} \quad \| \mathbf{A}\mathbf{\mathbf{X}} - \mathbf{B} \|_2^2
-\end{equation}
-
-\noindent for which the solution $\hat{\mathbf{X}}$ is well-known:
-
-\begin{equation}
-\label{eq:MP}
-\hat{\mathbf{X}}=\mathbf{A}^{\!\!+}\,\mathbf{B}
-\end{equation}
-
-\noindent where $\mathbf{A}^{\!\!+}$ is the Moore-Penrose pseudo-inverse
-which can be written as follows:
-
-\begin{equation}
-\label{eq:pseudoinverse}
-\mathbf{A}^{\!\!+}= (\mathbf{A}^{\!\!\top} \mathbf{A})^{-1}\mathbf{A}^{\!\!\top} \, .
-\end{equation}
-
-However, when $\mathbf{A}$ is not full rank, i.e
-$rank(\mathbf{A})=k <  n \leq m$, $\mathbf{A}^\top \mathbf{A}$ is
-always singular and equation~(\ref{eq:pseudoinverse}) cannot be used.
-More generally, the pseudo-inverse is best computed using the compact
-singular value decomposition (SVD) of $\mathbf{A}$:
-
-\begin{equation}
-    \label{eq:compactsvd}
-    \underset{m \times n}{\mathbf{A}}=
-    \underset{m \times k}{\mathbf{U_1}} \enskip
-    \underset{k \times k}{\Sigma_1} \enskip
-    \underset{k \times n}{\mathbf{V}_1^{\top}} \, ,
-\end{equation}
-
-\noindent as follows
-
-\begin{equation}
-\label{eq:pseudoinversesvd}
-\mathbf{A}^{\!\!+} = \mathbf{V}_1 \Sigma_1^{-1} \mathbf{U}_1^\top \, .
-\end{equation}
-
-%Descomentar en la tesis
-\textbf{Demo}\quad
-
-Since $\mathbf{A}$ is singular the problem shown in
-equation~(\ref{eq:regressionproblem}) has not solution, the minimum norm given
-by equation~(\ref{eq:MP}) is obtained by solving the equivalent problem:
-
-\begin{equation*}
-\label{eq:proyectorsol}
-\mathbf{A \hat{X} = PB} 
-\end{equation*}
-
-\noindent where $\mathbf{P=U_1 U_1^\top}$ is the projection onto the
-Col($\mathbf{A}$).
-
-Since $\mathbf{V} = [\underset{(n \times k)}{\mathbf{V_1}} |
-\underset{(n \times k)}{\mathbf{V_2}}]$ and $\mathbf{V_1^\top V_2 =
-0}$ we can express $\mathbf{\hat{X}} = \mathbf{V_1 x_1 + V_2 x_2}$
-with $\mathbf{x_2=0}$ because $\mathbf{\hat{X}}$ lives in the
-$\text{Row}(\mathbf{A})$ given by $\mathbf{V_1}$, so we have:
-
-\begin{eqnarray*}
-\mathbf{A \hat{X}} &=& \mathbf{PB} \\
-\mathbf{U_1 \Sigma_1 V_1^\top \hat{X}} &=& \mathbf{U_1 U_1^\top B} \\
-\mathbf{ V_1^\top \hat{X}} &=&  \mathbf{\Sigma_1^{-1} U_1^\top B} \\ 
-\mathbf{ V_1^\top V_1 x_1} &=& \mathbf{\Sigma_1^{-1}
-U_1^\top B} \\
-\mathbf{x_1}&=& \mathbf{\Sigma_1^{-1} U_1^\top B}
-\end{eqnarray*}
-
-\noindent from this result we can obtain $\mathbf{\hat{X}}$ and
-therefore the pseudo-inverse expression:
-
-\begin{eqnarray*}
-\mathbf{\hat{X}} &=& \mathbf{V_1 x_1} \\
-                &=& \mathbf{V_1 \Sigma_1^{-1} U_1^\top B} \\
-\mathbf{A^+} &=& \mathbf{V_1 \Sigma_1^{-1} U_1^\top} \, .
-\end{eqnarray*}
-
-\subsection{Ridge Regression}\label{sec:RR}
-
-Ridge regression (RR) introduces a ridge estimation to avoid the
-singularity of the matrix $\mathbf{A}^\top \mathbf{A}$
-equation~(\ref{eq:pseudoinverse}). RR solves the following
-optimization problem:
-
-\begin{equation}
-\label{eq:RRproblem}
-\underset{\mathbf{X}}{\text{min}} \quad \|
-\mathbf{A}\mathbf{\mathbf{X}} - \mathbf{B} \|_2^2 +\lambda \|
-\mathbf{\mathbf{X}}\|_2^2 
-\end{equation}
-
-\noindent where $\lambda$ is a regularization parameter.
-
-The RR optimal solution $\mathbf{X}(\lambda)$ is also well-known:
-
-\begin{equation}
-\label{eq:optsolRR}
-\mathbf{X}(\lambda)=(\mathbf{A}^\top \mathbf{A}+ \lambda
-\mathbb{I})^{-1}\mathbf{A}^\top \mathbf{B} \, . 
-\end{equation}
-
-\textbf{Demo}\quad
-
-To solve equation~(\ref{eq:optsolRR}) is equivalent to solve the following
-optimization problem:
-
-\begin{equation}
-\label{eq:RRproblem2}
-\underset{\mathbf{X}(\lambda)}{\text{min}} \quad \|
-\mathbf{C}\mathbf{\mathbf{X}} - \mathbf{F} \|_2^2
-\end{equation}
-
-\noindent where
-
-\begin{equation*}
-	\mathbf{C}=\begin{bmatrix} \quad &\mathbf{A}& \quad \\ \hdashline \quad
-& \lambda\mathbb{I}&\quad  \end{bmatrix} \quad \text{and} \quad
-	\mathbf{F}=\begin{bmatrix} \quad &\mathbf{B}& \quad \\ \hdashline \quad
-& \mathbf{0}&\quad  \end{bmatrix} \, .
-\end{equation*}
-
-Applying equation~(\ref{eq:MP}) and considering that $\mathbf{C}^\top
-\mathbf{C} = \mathbf{A}^\top \mathbf{A} + \lambda \mathbb{I}$ and 
-$\mathbf{C}^\top \mathbf{F}=\mathbf{A}^\top \mathbf{B} $ we have:
-
-\begin{eqnarray*}
-\mathbf{X}(\lambda)&=&(\mathbf{C}^\top
-\mathbf{C})^{-1}\mathbf{C}^\top \mathbf{F} \\
-&=& (\mathbf{A}^\top \mathbf{A} + \lambda \mathbb{I})^{-1} \mathbf{A}^\top \mathbf{B}
-\end{eqnarray*}
-
-\subsection{Online Ridge regression}
-
-If we represent matrices $\mathbf{A}$ and $\mathbf{B}$ such as:
-
-\begin{equation}
-\label{eq:notation}
-	\mathbf{A} = 
-\left[
-  \begin{tabular}{c>{$}c<{$}c}
-    --- & \mathbf{a}^{\top}_{1} & ---\\
-    --- & \mathbf{a}^{\top}_{2} & ---\\
-    & \vdots & \\
-    --- & \mathbf{a}^{\top}_{m} & ---
-  \end{tabular}
-\right]
-\quad \text{and} \quad
-\mathbf{B} =
-\left[
-  \begin{tabular}{c>{$}c<{$}c}
-    --- & \mathbf{b}_{1} & ---\\
-    --- & \mathbf{b}_{2} & ---\\
-    & \vdots & \\
-    --- & \mathbf{b}_{m} & ---
-  \end{tabular}
-\right] \, ,
-\
-\end{equation}
-
-\noindent equation~(\ref{eq:optsolRR}) can also be written as:
-
-\begin{eqnarray*}
-\label{eq:RReapand}
-\mathbf{\mathbf{X}}_{\text{ridge}}&=&(\mathbf{A}^\top \mathbf{A}+ \lambda
-\mathbb{I})^{-1}\mathbf{A}^\top \mathbf{B} \\
-&=& \displaystyle \big (\sum_{t=1}^m
-\mathbf{a}_t \mathbf{a}_t  ^\top + \lambda \mathbb{I}\big )^{-1}
-\sum_{t=1}^m \mathbf{a}_t \mathbf{b}_t \, .
-\end{eqnarray*}
-
-Lets define $\displaystyle\mathbf{S}= \sum_{t=1}^m \mathbf{a}_t
-\mathbf{a}_t  ^\top + \lambda \mathbb{I} $ and $\mathbf{W}=
-\displaystyle\sum_{t=1}^m \mathbf{a}_t \mathbf{b}_t$, so the
-algorithm~\ref{alg:RR} shows the iterative formulation:
-
-\begin{algorithm}[H]
-\begin{algorithmic}[1]
-\REQUIRE $\,$ \\
-$\{\mathbf{a}_1,\dots,\mathbf{a}_{m} \}$: $m$ input vectors \\
-$\{\mathbf{b}_1,\dots,\mathbf{b}_{m} \}$: $m$ target vectors \\
-$\lambda$: regularization parameter \\
-\ENSURE  $\,$ \\
-$\{f(\mathbf{a}_1),\dots,f(\mathbf{a}_{m}) \}$: model predictions \\
-\STATE Initialize $\mathbf{S}=\lambda \mathbb{I}$
-and $\mathbf{W}=0$
-\FOR { $t = 1$ to $m$ }
-	\STATE read new $\mathbf{a}_t$
-	\STATE $\mathbf{X}=\mathbf{S}^{-1}\mathbf{W}$
-	\STATE output prediction $f(\mathbf{a}_t) = \mathbf{X}^\top \mathbf{a}_t$
-   	\STATE $\mathbf{S} = \mathbf{S} + \mathbf{a}_t \mathbf{a}_t^\top$
-   	\STATE Read new $y_t$
-    	\STATE $\mathbf{W} = \mathbf{W} + \mathbf{a}_t \mathbf{b}_t$
-\ENDFOR
-\end{algorithmic}
-\caption{Online Ridge Regression}
-\label{alg:RR}
-\end{algorithm}
-
-\subsection{The Aggregating Algorithm for Regression}
-
-The AAR, proposed by~\cite{vovk2001}, is an application of the aggregating
-algorithm to the problem of regression. The idea is to introduce the new input
-vector $\mathbf{x}_{m+1}$ to solve the model parameters:
-
-\begin{equation}
-\label{eq:AARexpand}
-\mathbf{X}_{aar} = \displaystyle \big (\sum_{t=1}^{m+1}
-\mathbf{a}_t \mathbf{a}_t  ^\intercal + \gamma \mathbb{I}\big )^{-1}
-\sum_{t=1}^m \mathbf{a}_t \mathbf{b}_t \, .
-\end{equation}
-
-If we define $\displaystyle\mathbf{S}= \sum_{t=1}^{m+1} \mathbf{a}_t
-\mathbf{a}_t  ^\intercal + \gamma \mathbb{I} $ and $\mathbf{W}=
-\displaystyle\sum_{t=1}^m \mathbf{a}_t \mathbf{b}_t$, the
-algorithm~\ref{alg:AAR} is slightly different to the algorithm~\ref{alg:RR}, 
-which updated matrix $\mathbf{S}$ before making the prediction:
-
-\begin{algorithm}[ht]
-\begin{algorithmic}[1]
-\REQUIRE $\,$ \\
-$\{\mathbf{a}_1,\dots,\mathbf{a}_{m} \}$: $m$ input vectors \\
-$\{\mathbf{b}_1,\dots,\mathbf{b}_{m} \}$: $m$ target vectors \\
-$\lambda$: regularization parameter \\
-\ENSURE  $\,$ \\
-$\{f(\mathbf{a}_1),\dots,f(\mathbf{a}_{m}) \}$: model predictions \\
-\STATE Initialize $\mathbf{S}=\lambda \mathbb{I}$
-and $\mathbf{W}=0$
-\FOR { $t = 1$ to $m$ }
-	\STATE read new $\mathbf{a}_t$
-   	\STATE $\mathbf{S} = \mathbf{S} + \mathbf{a}_t \mathbf{a}_t^\intercal$
-	\STATE $\mathbf{X}=\mathbf{S}^{-1}\mathbf{W}$
-	\STATE output prediction $f(\mathbf{a}_t) = \mathbf{X}^\intercal \mathbf{a}_t$
-   	\STATE Read new $\mathbf{y}_t$
-    	\STATE $\mathbf{W} = \mathbf{W} + \mathbf{a}_t \mathbf{b}_t$
-\ENDFOR
-\end{algorithmic}
-\caption{{\em The aggregating algorithm for regression}}
-\label{alg:AAR}
-\end{algorithm}
-
-\subsection{Selection of Lambda}
-
-The additional term $\lambda \|\mathbf{\mathbf{X}}\|_2^2$ in the optimization
-problem shown in equation~(\ref{eq:RRproblem}) has two effects on the solution:
-shrinks the coefficients towards zero and improves the conditioning of the
-problem.
-
-Figure~\ref{fig:shrinks} shows a visual example of the shrinking of the
-coefficients:
-
-\begin{figure}[h!]
-\includegraphics[width=0.5\linewidth]{img/shrinks}
-\caption{Shrink of regression coefficients}
-\label{fig:shrinks}
-\end{figure}
-
-On the other hand, the effect of adding the term $\lambda \mathbb{I}$
-to the matrix $\mathbf{A}^\top \mathbf{A}$
-(equation~(\ref{eq:optsolRR})) improves its condition number since it
-increases its diagonal values when $\lambda > 0 $.  The matrix
-$\mathbf{A}^\top \mathbf{A}$ is symmetrical ($(\mathbf{A}^\top
-\mathbf{A})^\top = \mathbf{A}^\top \mathbf{A}$) and therefore
-diagonalizable.  If we know the eigenvalue decomposition of $\mathbf{A}
-= \mathbf{U\Sigma V^\top}$, then:
-
-\begin{eqnarray*}
-\mathbf{A}^\top \mathbf{A}+\lambda \mathbb{I}&=&\mathbf{V\Sigma^2
-V}^\top + \lambda \mathbf{V} \mathbf{V}^\top\\ &=&\mathbf{V}
-(\mathbf{\Sigma}^2+\lambda\mathbb{I}) \mathbf{V}^\top \, ,
-\end{eqnarray*}
-
-\noindent where
-
-\begin{equation*}
-\mathbf{\Sigma}^2+\lambda\mathbb{I}=
-\begin{bmatrix}
-\sigma^2_1 + \lambda & \, & \, \\
-\, & \sigma^2_2 +\lambda & \, \\
-\, & \, & \ddots & \, \\
-\, & \, & \, & \sigma^2_n +\lambda \, .
-\end{bmatrix}
-\end{equation*}
-
-where $\sigma_1 \geq \sigma_2 \geq \dots \geq \sigma_n$.
-
-Since the condition number of a matrix $\mathbf{A}$ is defined as:
-
-\begin{equation*}
-	\kappa = \|\mathbf{A}\| \|\mathbf{A}^{-1}\|
-\end{equation*}
-
-If matrix $\mathbf{A}$ is non-singular, its condition number can be
-expressed in terms of its singular values. The effect of adding the
-regularization term affects the condition number as follows:
-
-\begin{eqnarray*}
-\kappa_{ols} &=& \|\mathbf{A}\| \|\mathbf{A}^{-1}\|=\frac{\sigma_1}{\sigma_n} \\
-\kappa_{ridge} &=& \|\mathbf{A}^\intercal \mathbf{A} + \lambda \mathbb{I}\| 
-\|(\mathbf{A}^\intercal \mathbf{A} + \lambda \mathbb{I})^{-1}\|=\frac{\sigma_1+\lambda}{\sigma_n + \lambda} \,
-\end{eqnarray*}
-
-It is easy to see that the term $\lambda$ improves the condition number:
-
-\begin{equation*}
-        \frac{\sigma_1+\lambda}{\sigma_n + \lambda} <
-        \frac{\sigma_1}{\sigma_n} \,  \qquad \forall \quad \lambda > 0
-\end{equation*}
-
-However, $\lambda$ cannot be too large. Tipically $\lambda$ is small and its
-magnitude depends on the matrix $\mathbf{A}$.
-
-For rank deficient matrices we know that $\text{det}(\mathbf{A A^\top})=0$, adding the
-term $\lambda \mathbb{I}$ we have that $\text{det}(\mathbf{A A^\top}+\lambda \mathbb{I}) =
-p(\lambda)$ where $p(\lambda)$ is a polynomial of degree $n$ ($\mathbf{A}$ is
-$m \times n$). The zeros of $p(\lambda)$ are discretes, so it can be represented
-as:
-
-\[
-p(\lambda) =
-\lambda(\lambda-\lambda_1)^{n_1}(\lambda-\lambda_2)^{n_2}\dots(\lambda-\lambda_s)^{n_s}
-\]
-
-\noindent where $n_1 + n_2 + \dots + n_s = n$.
-
-This means that $\lambda$ must be small in order to ensure that $p(\lambda)$
-does not vanish.
-
-\subsubsection{The bias-variance tradeoff}
-
-To determine parameter $\lambda$ is crucial for ridge regression since it could
-reduce the expected prediction error by reducing variance considering a biased
-estimator, this is called the bias-variance trafeoff. The prediction error is
-obtained as:
-
-\begin{equation}
- \label{eq:prederror}
-E[(B-\hat{f}(\mathbf{X}))^2] = \sigma^2 + Bias(\hat{f}(\mathbf{X}))^2 + Var(\hat{f}(\mathbf{X}))
-\end{equation}
-
-\noindent where $\hat{f}(\mathbf{X})=\mathbf{AX}(\lambda)$ and
-$f(\mathbf{X})=\mathbf{AX}$.
-
-\textbf{Demo}\quad
-
-\begin{eqnarray*}
-E[(B-\hat{f}(\mathbf{X}))^2] & =&
-E[(B-f(\mathbf{X})+f(\mathbf{X})-\hat{f}(\mathbf{X}))^2] \\
-&=& E[(B-f(\mathbf{X}))^2] +
-E[(f(\mathbf{X})-\hat{f}(\mathbf{X}))^2] + \dots \\
-& &
-2\cancelto{0}{E[B-f(\mathbf{X})]}E[f(\mathbf{X})-\hat{f}(\mathbf{X})] \\
-&=& \sigma^2 + MSE(\hat{f}(\mathbf{X}))
-\end{eqnarray*}
-
-\noindent where
-
-\begin{eqnarray*}
-    MSE(\hat{f}(\mathbf{X})) &=&
-    E[f(\mathbf{X})-\hat{f}(\mathbf{X}))]^2 \\
-    &=& E[f(\mathbf{X})-E[\hat{f}(\mathbf{X})] +
-    E[\hat{f}(\mathbf{X})]-\hat{f}(\mathbf{X}))]^2 \\
-    &=&  E[(f(\mathbf{X})-E[\hat{f}(\mathbf{X})])^2] +
-      E[(\hat{f}(\mathbf{X})-E[\hat{f}(\mathbf{X})])^2] + \dots\\
-    & & E[(f(\mathbf{X})-E[\hat{f}(\mathbf{X})])] 
-    \cancelto{0}{E[(\hat{f}(\mathbf{X})-E[\hat{f}(\mathbf{X})])]} \\
-    &=& Bias(\hat{f}(\mathbf{X}))^2 + Var(\hat{f}(\mathbf{X}))
-\end{eqnarray*}
-
-The bias of OLS is:
-
-\begin{eqnarray*}
-Bias(\hat{\mathbf{X}}) &=& E[\hat{\mathbf{X}}] - \mathbf{X} \\
-&=& E[ (\mathbf{A}^\top \mathbf{A})^{-1}\mathbf{A}^\top \mathbf{B}] - \mathbf{X} \\
-&=& E[ (\mathbf{A}^\top \mathbf{A})^{-1}\mathbf{A}^\top (\mathbf{AX})] - \mathbf{X}  \\
-&=& \mathbf{X}  - \mathbf{X}  \\
-&=&  0
-\end{eqnarray*}
-
-The bias of ridge regression when $\mathbf{A A^\top}$ is non-singular
-can be obtained expressing ridge regression solution
-$\mathbf{\lambda}$ in terms of OLS solution $\hat{\mathbf{X}}$:
-
-\begin{eqnarray*}
-\mathbf{X}(\lambda) &=&( \mathbf{A}^\top \mathbf{A} + \lambda \mathbb{I})^{-1}\mathbf{B} \\
-&=& (\mathbb{I} + \lambda (\mathbf{A}^\top \mathbf{A})^{-1})^{-1} (\mathbf{A}^\top \mathbf{A})^{-1}\mathbf{A}^\top \mathbf{B} \\
-&=&  (\mathbb{I} + \lambda (\mathbf{A}^\top \mathbf{A})^{-1})^{-1}  \hat{\mathbf{X}} \\
-&=& \mathbf{W} \hat{\mathbf{X}} 
-\end{eqnarray*}
-
-\noindent where $\mathbf{W}  = (\mathbb{I} + \lambda (\mathbf{A}^\top
-\mathbf{A})^{-1})^{-1}  $ it is defined for simplicity. Ridge
-regression bias is then obtained as:
-
-\begin{eqnarray*}
-Bias(\mathbf{X}(\lambda)) &=& E[\mathbf{X}(\lambda)] - \mathbf{X} \\
-&=& E[\mathbf{W}\hat{\mathbf{X}}] - \mathbf{X} \\
-&=&  \mathbf{W} \mathbf{X} - \mathbf{X} \neq 0 
-\end{eqnarray*}
-
-The variance of OLS is:
-
-\begin{equation*}
-Var(\hat{\mathbf{X}}) = \sigma^2 (\mathbf{A}^\top \mathbf{A} )^{-1}
-\end{equation*}
-
-\noindent and the variance of ridge regression is:
-
-\begin{eqnarray*}
-Var(\mathbf{X}(\lambda)) &=& Var(\mathbf{W}\hat{\mathbf{X}}) \\ 
-&=& \mathbf{W}Var(\hat{\mathbf{X}})\mathbf{W}^\top \\
-&=& \sigma^2 \mathbf{W}(\mathbf{A}^\top \mathbf{A} )^{-1}\mathbf{W}^\top
-\end{eqnarray*}
-
-\begin{figure}[h!]
-\includegraphics[width=0.8\linewidth]{img/biasvariance}
-\caption{Bias-variance tradeoff}
-\label{fig:biasvariance}
-\end{figure}
-
-The figure~\ref{fig:biasvariance} shows the bias-variance tradeoff
-given by equation~(\ref{eq:prederror}). Ridge regression shows an
-increasing bias and a decreasing variance. Despite OLS has zero bias,
-its variance is greater than ridge for small values of $\lambda$.
-It can be shown that, in terms of prediction error, ridge (black line)
-is lower than OLS (dotted line)~\cite{hoerl1970}.
-
-\subsubsection{Methods for lambda selection}
-
-Despite the bias-variance tradeoff provides a conceptual framework for
-determining a good model, is not directly useful. Some popular methods
-for determine model selection are:
-
-\begin{itemize}
-    \item Akaike information criterion (AIC) (Akaike, 1974)
-    \item Bootstrap-based selection (Efron and Tibshirani, 1997)
-    \item Cross-validation (Stone, 1974)
-\end{itemize}
-
-On the other hand, $\lambda$ can also be used to reduce computational time as it
-is shown in equation~(\ref{eq:taylor}). Golub (~\cite{golub1965}) suggested that
-$\lambda$ had to be chosen so that:
-
-\[
-\frac{\lambda}{\lambda + \delta^2} < 0.1
-\]
-
-\noindent where $\delta$ is a lower bound of the smallest non-zero singular
-value $\sigma_k$. One approach is to choose the greatest lower bound for
-$\delta=\sigma_k$. Hence:
-
-\[
-\frac{\lambda}{\lambda + \sigma_k^2} < 0.1
-\]
-
-\noindent if we choose $\lambda=\beta \sigma_k^2$ then we get the following
-relation:
-
-\[
-\frac{\beta \sigma_k^2}{\beta \sigma_k^2 + \sigma_k^2} =
-\frac{\beta}{\beta+1} < 0.1 \Rightarrow \beta < \frac{1}{9}  
-\]
-
-Experiments done by ~\cite{coleman+sun2010} have shown that $\beta = 0.01$
-produces satisfactory results. This means that $\lambda = 0.01 \sigma_k^2$.
-
-However, get $\sigma_k$ imply obtaining first the SVD, which is
-computational expensive.
-
-Since the algorithm presented by ~\cite{coleman+sun2010} already compute the QR
-factorization of the matrix $\mathbf{A}$ they show a procedure for getting
-a $\lambda$ approximation using QR.
-
-\begin{enumerate}
-\item Compute QR factorization: $\mathbf{A} = \mathbf{Q_1R_1}$
-\item Let $\mathbf{W}$ denote the set of absolute values of the nonzero diagonal
-elements of $\mathbf{R_1}$. Let $w_{\text{min}}$ and $w_{\text{max}}$ denote the
-smallest and largest elements of $\mathbf{W}$ respectively. Both 
-$\lambda_1 = \hat{\beta} w_{\text{min}}^2$
-$\lambda_2 = \hat{\beta} \frac{w_{\text{min}}^2}{w_{\text{max}}^2}$ where
-$\hat{\beta}=0.00025$ produce satisfactory results.
-\item The $\lambda_{\text{QR}}$ approximation is obtained as:
-$\lambda_{\text{QR}} = \frac{\lambda_1 + \lambda_2}{2}$  
-\end{enumerate}
-
-\subsection{Efficient computation}
-
-%Revisar!!!
-In regression problems we always require getting a matrix inverse which is
-computationally expensive. However, in stream data problems the is a way to
-obtain a matrix inverse approximation using the Sherman-Morrison-Woodbury. This
-formula allows to get the inverse of a
-the matrix $\mathbf{A+uv^\top}$ if we already calculated the
-inverse of $\mathbf{A}$ as follows:
-
-\begin{equation}
-\label{eq:SMW}
-(\mathbf{A+uv^\top})^{-1}=\mathbf{A}^{-1}-
-\frac{\mathbf{A^{-1}uv^\top A^{-1}}}{\mathbf{1+v^\top A^{-1}u}}
-\end{equation}
-
-%Revisar!!
-
-Another way was presented by Coleman and
-Sun~\cite{coleman+sun2010} using an iterative algorithm which
+In order to solve OLS in a efficient way, Coleman and
+Sun~\cite{coleman+sun2010} presented an iterative algorithm which
 uses $\mathbf{X}(\lambda)$ to approximate $\mathbf{X}(0)$.
@@
 is computationally less expensive than the SVD.
-
-\subsection{Online machine learning algorithms}
-
-An online algorithm allows incremental learning by processing one instance at a
-time. This is done updating the current model instead of building the model from
-scratch.
-
-Training phase is commonly a computationally expensive process. Therefore, when
-new data arrives it can't be included easily to the model. Moreover, it could
-happen that we won't have enought time to process the new data before more data
-arrives. In online learning algorithms there is no training phase, but the model
-is updated and evaluated at every time step. This model updating is
-computationally less expensive than a training phase.
-
-Online learning is specially useful for stream data problems and the model is
-sensitive to new data. It is also useful when some past data may be irrelevant
-or we want to improve computational efficiency. However, online algorithms could
-affect accuracy.
-
-There are several popular online methods such as
-perceptron~\cite{rosenblatt58},winnow~\cite{littlestone1988},
-passive-aggressive~\cite{crammerETall2006}, stochastic gradient
-descent~\cite{zhang2004}, aggregating algorithm~\cite{vovk2001} and the second
-order perceptron~\cite{cesa-bianchi2005}.  In \cite{blum1998} and
-\cite{cesa-bianchi2006} an in-deph analysis of online learning is provided.
-Applications in finance has been widely used: study presented
-by~\cite{arce+salinas2012} applied ridge regression in an online context and
-more recently, time series forecasting using online learning has been
-presented~\cite{anavaetAl2013}.
-
-Incremental learning refers to any online learning process that learns the same
-model as would be learnt by a batch learning algorithm.
-
-Incremental learning is useful when the input to a learning process is stream
-data, with the need or desire to be able to use the result of learning at any
-point in time, based on the input observations received so far.
-
-In principle, the stream of observations may be infinitely long, or the next
-observation long delayed, precluding any hope of waiting until all the
-observations have been received. One would rather not simply accumulate and
-store all the inputs and, upon receipt of each new one, apply a batch learning
-algorithm to the entire sequence of inputs received so far. It would be
-preferable compu- tationally if the existing hypothesis or other artifact of
-learning could be updated in response to each newly received input observation.
-
-Incremental learning is very useful when there is no need to record fundamental
-data and only a summary needs to be retained. Due to this, incremental
-algorithms are often characterized as memoryless, because no memory of past data
-is required.  The algorithm is online but not incremental if it doesn't produce
-the same result for all observations that the corresponding batch algorithm
-would for these same observations.
-
-Algorithm~\ref{alg:onlinealg} shows the online learning algorithm structure:
-
-\begin{algorithm}[ht]
-\begin{algorithmic}[1]
-    \STATE Receives input $\mathbf{x}_t$
-    \STATE Makes prediction $\mathbf{\hat{y}}_t$
-    \STATE Receives response $\mathbf{y}_t$
-    \STATE Incurs loss $l_t(\mathbf{y}_t,\mathbf{\hat{y}}_t)$
-\end{algorithmic}
-\caption{Structure of a Learning System}
-\label{alg:onlinealg}
-\end{algorithm}
-
-\noindent where $l$ is some loss function. Performance is later measured after
-$T$ trials as:
-
-\begin{equation*}
-L_T = \sum_{t=1}^T l_t(\mathbf{y}_t,\mathbf{\hat{y}}_t)
-\end{equation*}
-
-The objective is minimize this loss function for all instances.
-The quality of online learning algorithms is measured by a quantity known as
-regret which is the difference between the performance of the online algorithm
-and its optimal predictor $E^* \in \Theta$ given by:
-
-\begin{equation*}
-L^*_T= \text{min}_{E \in \Theta} L_T^E \, ,
-\end{equation*}
-
-\noindent where $L_T^E = \sum_{t=1}^T
-l_t(\mathbf{y}_t,\mathbf{y}^E_t)$ and $\mathbf{y}^E_t$ is the expert estimation.
-
-Therefore regret is defined as:
-
-\begin{equation*}
-R_T = L_T - L^*_T
-\end{equation*}
-
-%An online learning algorithm looks at every example exactly once.
-\subsection{Model selection}
-In order to set model parameters the Akaike Information Criterion (AIC) is
-used. AIC is often used in model selection where AIC whith smaller values are
-preferred.
-
-AIC is calculated as follows:
-
-\begin{equation}
-\label{eq:aicformula}
-AIC = \underset{\text{bias}}{-\frac{2l}{N}} + 
-\underset{\text{variance}}{\frac{2k}{N}}
-\end{equation}
-
-\noindent where
-
-\begin{description}
-\item[l] is the loglikelihood function
-\item[k] number of estimated parameters (including the variance)
-\item[N] number of observations
-\end{description}

### 2016-03-22 articles-conferences/online-vec-article-2014/src/varvec.tex
Agregar estos repos a la tesis

--- 2016-03-22 articles-conferences/elsevier-journal-2015/src/background.tex
@@
-\section{\uppercase{Background}}
-\label{sec:background}
-\noindent
+\section{Vector autorregresive models}\label{sec:varvec}
 
-\subsection{Spurious Regression}
+VAR as well as VEC are autorregresive models which describe the joint
+behaviour of a set variables. Since VEC is a special case of VAR
+model, we will formulate the VAR model first and then show how it is
+related to the VEC model.
 
-Regression analysis is widely used to test economic theories. But sometimes
-regression is done between two variables, lets say $x_t$ and $y_t$, which are
-not causally related. This occurs at the presence of a third stochastic variable
-$z_t$ which is the true causal of the first two variables $x_t$ and $y_t$. The
-regression measure R-squared gives high values in presence of spurious
-relationships. This measure would means that the model explains all the
-variability of the response data around its mean which is not true.
-
-The R-squared measure is calculated as follows:
-
-\begin{equation}
-\label{eq:rsquared}
-R^2 = 1 - \frac{\sum_{t=1}^T \epsilon_t^2}{\sum_{t=1}^T(y_t - \bar{y})^2}
-\end{equation}
-
-In presence of spurious relationships the denominator of equation
-\ref{eq:rsquared} becomes very large because a large weight is placed on extreme
-observations on either side of the mean $\bar{y}$.
-In order to detect this, other statistic must be considered such as the
-Durbin-Watson statistic.
-
-% Acá agregar gráfico de spurious regression
-
-\subsection{Integration and Cointegration}
-A time series $\mathbf{y}$ is said to be integrated of order $d$ if after
-differentiating the variable $d$ times, we get an I(0) process, more precisely:
-
-\[
-(1-L)^d \mathbf{y} \sim \text{I(0)} \, ,
-\]
-
-\noindent where I(0) is a stationary time series and $L$ is the lag operator:
-
-\[
-(1-L)\mathbf{y} = \Delta \mathbf{y}=\mathbf{y}_t  -\mathbf{y}_{t-1} \quad \forall t
-\]
-
-Let $\mathbf{y}_t = \{\mathbf{y}^1, \dots, \mathbf{y}^l\}$ be a set of $l$
-stationary time series I(1) which are said to be cointegrated if a vector
-$\beta=[\beta(1),\dots,\beta(l)]^\top \in \mathbb{R}^l$  exists such that the
-time series,
-
-\begin{equation}
- \mathbf{Z}_t:= \beta^\top \mathbf{y}_t = \beta(1) \mathbf{y}^1 + \dots + \beta(l) \mathbf{y}^l \sim
- \text{I(0)}\, .
-\end{equation}
-
-In other words, a set of I(1) variables is said to be cointegrated if
-a linear combination of them exists which is I(0).
-
-\subsection{Vector Autorregresive Models}\label{sec:varvec}
-
-VECM is a special case of VAR model and both describe the joint behaviour of a
-set of variables.
-
-VAR($p$) model is a general framework to describe the behaviour of a set of $l$
-endogenous variables as a linear combination of their last $p$ values. These
-$l$ variables at time $t$ are represented by the vector $\mathbf{y}_t$ as
-follows:
+VAR($p$) model is a general framework to describe
+the behaviour of a set of $l$ endogenous variables as a linear
+combination of their last $p$ values. These $l$ variables at time $t$
+are represented by the vector $\mathbf{y}_t$ as follows:
 
@@
 \mathbf{y}_t = 
-\begin{bmatrix} y_{1,t} &
-y_{2,t} &
-\dots &
+\begin{bmatrix} y_{1,t} \\
+y_{2,t} \\
+\vdots \\
 y_{l,t}
-\end{bmatrix}^\top \, ,
+\end{bmatrix}
 \end{equation}
@@
  \mathbf{y}_t = \phi_1 \mathbf{y}_{t-1}  + \dots +   \phi_p\mathbf{y}_{t-p}
-+ \mathbf{c} + \mathbf{\epsilon}_t \, ,
++ \mathbf{c} + \mathbf{\epsilon}_t, \qquad t=p+1, \dots, N
 \end{equation}
 
-\noindent where $t=p+1, \dots N$,  ${\phi_1,\dots,\phi_p}$ are $l \times l$
-matrices of real coefficients,
-$\mathbf{\epsilon}_{p+1},\dots,\mathbf{\epsilon}_N$ are error terms,
-$\mathbf{c}$ is a constant vector and $N$ is the total number of samples.
+\noindent where ${\phi_1,\dots,\phi_p}$ are $l \times l$ 
+matrices of real coefficients, $\mathbf{\epsilon}_{p+1},\dots,\mathbf{\epsilon}_N$ are
+error terms, $\mathbf{c}$ is a constant vector and $N$ is the total number of
+samples.
 
-The VAR matrix form of equation~(\ref{eq:var}) is:
-
-%\begin{equation}\label{eq:vareq}
-%\mathbf{B} = \mathbf{A} \mathbf{X} + 
-%\mathbf{E} \, , 
-%\end{equation}
-%
-%\noindent where:
-%\begin{eqnarray}
-%\mathbf{A} &=&
-%\begin{bmatrix}
-%   \mathbf{y}_{p}     & \dots    & \mathbf{y}_{N-1}\\
-%   \mathbf{y}_{p-1}   & \dots    & \mathbf{y}_{N-2}\\
-%   \vdots             & \ddots   & \vdots\\
-%   \mathbf{y}_{1}     & \dots    & \mathbf{y}_{N-p}\\
-%   1                  & \dots    & 1 
-%   \end{bmatrix}^\top \, ,\\
-%\mathbf{B} &=&
-%\begin{bmatrix}
-%\quad \\
-%\mathbf{y}_{p+1} &
-%\dots & 
-%\mathbf{y}_N \\
-%\quad
-%\end{bmatrix}^\top  \, ,\\
-%\mathbf{X}&=&
-%\left[
-%\begin{array}{ccccc}
-%    \quad  & \quad & \quad & \quad \\
-%    \phi_1   & \cdots & \phi_p & \mathbf{c} \\  
-%    \quad & \quad & \quad & \quad
-%    \end{array}
-%\right]^\top \, ,\\
-%\mathbf{E} &=& 
-%\begin{bmatrix}
-%    \quad \\
-%    \mathbf{\epsilon}_{p+1}  & 
-%    \dots                & 
-%    \mathbf{\epsilon}_N \\
-%    \quad
-%\end{bmatrix}^\top \, . 
-%\end{eqnarray}
+The VAR matrix form is the following:
 
 \begin{equation}
- \label{eq:vareq}
+ \label{eq:varmatrix}
                \underbrace{ \begin{bmatrix}
@@
 \end{equation}
-Equation~(\ref{eq:vareq})  can be solved using ordinary least squares estimation.
 
-\subsection{Vector Error Correction}
+\noindent which can be solve using ordinary least squares estimation.
 
-VECM is a special form of a VAR model for I(1) variables that are also
-cointegrated~\cite{banerjee1993}. It is obtained by replacing the form $\Delta
-\mathbf{y}_t = \mathbf{y}_t - \mathbf{y}_{t-1}$ in equation (\ref{eq:var}).
-VECM is expressed in terms of differences, has an error correction term and 
-the following form:
+\section{Integration and Cointegration}
+
+A time series $\mathbf{y}$ is called integrated of order $d$ if after
+differencing the variable $d$ times, we get a I(0) variable (stationary
+process), i.e:
+\[
+(1-L)^d \mathbf{y} \sim \text{I(0)}
+\]
+\noindent where I(0) is a stationary time series and $L$ is the lag operator, i.e,
+\[
+(1-L)\mathbf{y} = \Delta \mathbf{y}
+\]
+\noindent where $\Delta \mathbf{y}(t) = \mathbf{y}(t)  -\mathbf{y}(t-1) \quad \forall t $.
+
+Let $\mathbf{y}_t = \{\mathbf{y}^1, \dots, \mathbf{y}^l\}$ be a set of $l$ stationary time
+series I(1) which are said to be cointegrated if a vector
+$\beta=[\beta(1),\dots,\beta(l)]^\top \in \mathbb{R}^l$  exists such that the time series,
+
+\begin{equation}
+ \mathbf{Z}_t:= \beta^\top \mathbf{y}_t = \beta(1) \mathbf{y}^1 + \dots + \beta(l) \mathbf{y}^l \sim
+ \text{I(0)}
+\end{equation}
+
+In words, a set of I(1) variables is said to be cointegrated if
+exists a linear combination of them which is I(0).
+
+\subsection{Stationary Time Series}
+
+A strictly stationary times series is one for which the probabilistic behavior
+of every collection of values $\{y_{t_1},y_{t_2},\dots,y_{t_L}\}$ is identical
+to that of the time shifted set, more precisely: \[ P\{y_{t_1} \leq
+c_1,\dots,y_{t_L} \leq c_L\} = P\{y_{t_1+h} \leq c_1,\dots,y_{t_L+h}
+\leq c_L\}
+\quad \forall L \in \mathbb{N}, \forall h \in \mathbb{Z}\] \noindent where
+$c_1,\dots,c_L$ are constants.
+
+This definition is too strong and difficult to assess it from a single data
+set. The weak version of this definition imposes conditions only on the two
+first two moments.
+
+A weakly stationary time series is a process which mean, variance and auto
+covariance do not change over time:
+
+\begin{eqnarray*}
+E(Y_t) &=& \mu  \quad \forall t \in \mathbb{N} \\ E(Y^2_t) &=&
+\sigma^2  \quad \forall t \in \mathbb{N} \\
+\lambda(s,t)&=&\lambda(s+h,t+h) \quad \forall s,t \in \mathbb{N},
+\forall h \in \mathbb{Z}
+\end{eqnarray*}
+
+\noindent with $\lambda(s,t) = E[(y_s-\mu)(y_t - \mu)]$.
+
+The following example~\cite{johansen1995} helps to illustrate the meaning
+of $\beta$:
+
+\textbf{Example:}
+
+If we have two-dimensional process $\mathbf{X}_t$, $t=1,\dots,T$ by:
+
+\begin{eqnarray*}
+\mathbf{X}_{1t} &=& \sum_{i=1}^t \epsilon_{1i} + \epsilon_{2t} \\
+\mathbf{X}_{2t} &=& a \sum_{i=1}^t \epsilon_{1i} + \epsilon_{3t} 
+\end{eqnarray*}
+
+Since $\mathbf{X}_{1t}$ and $\mathbf{X}_{2t}$ are I(1) processes and there
+exist a vector $\beta = [a -1]$ such that:
+
+\[
+\beta^\top \mathbf{X}_t = a \mathbf{X}_{1t} -\mathbf{X}_{2t} = 
+a\epsilon_{2t} - \epsilon_{3t} \sim \text{I(0)}
+\]
+
+then, both processes are said to be cointegrated. If we add a I(0) process
+$\mathbf{X}_{3t} = \epsilon_{4t}$  we find that there exists two cointegration
+vectors now: $\begin{bmatrix}a &-1& 0\end{bmatrix}$ and $\begin{bmatrix}0
+&0&1\end{bmatrix}$ since:
+
+\[
+\beta^\top \mathbf{X}_t = 
+\begin{bmatrix}
+a & -1 & 0 \\
+0 & 0 & 1
+\end{bmatrix} 
+\begin{bmatrix} 
+\mathbf{X}_{1t} \\
+\mathbf{X}_{2t} \\
+\mathbf{X}_{3t}
+\end{bmatrix} = 
+\begin{bmatrix}
+a\epsilon_{2t} - \epsilon_{3t} \\
+\epsilon_{4t}
+\end{bmatrix}
+\]
+
+This example shows how cointegration vectors describes the stable relations
+between the processes by linear relations that are more stationary than the
+original process.
+
+\section{VEC model}
+
+A VEC model is a special form of a VAR model for I(1) variables that
+are also cointegrated. The VEC model is obtained by replacing the form
+$\Delta \mathbf{y}_t = \mathbf{y}_t - \mathbf{y}_{t-1}$ in equation
+(\ref{eq:var}). The VEC model is expressed in terms of differences,
+has an error correction term and it has the following form:
 
@@
 
-The matrix $\Omega$ has the following properties~\cite{johansen1995}:
+The matrix $\Omega$ has the following properties:
 \begin{itemize}
@@
 procedure~\cite{johansen1988} whereas $\alpha$ has to be determined as a
-variable in the VECM.
+variable in the VEC model.
 
-It is worth noticing that the factorization of the matrix $\Omega$ is not
-unique since for any $r \times r$ nonsingular matrix $H$ we have:
+It is worth noticing that the factorization of the matrix $\Omega$ is not unique since for any
+$r \times r$ nonsingular matrix $H$ we have:
 
@@
 
+Therefore, to obtain unique values, further restrictions on the model
+are required.
+
 \end{itemize}
+%Johansen  procedure also gives a set of
+%eigenvalues ordered in a descending way which do not correspond to
+%$\alpha$ but the number of eigenvalues larger than zero determine the rank
+%of $\beta$, i.e the number of cointegration relations.
 
@@
 
-VECM matricial form is the following:
-
-%\begin{eqnarray}\label{eq:vecmatrix}
-% \mathbf{A}&=&
-%   \begin{bmatrix} 
-%   \beta^\top \mathbf{y}_{p} & 
-%   \cdots & \beta^\top \mathbf{y}_{N-1} \\
-%   \Delta \mathbf{y}_p  & \cdots 
-%   &\Delta\mathbf{y}_{N-1} \\ 
-%   \vdots  & \ddots & \vdots \\
-%   \Delta\mathbf{y}_2  & \cdots 
-%   & \Delta \mathbf{y}_{N-p+1} \\
-%   1 & \cdots & 1 
-%   \end{bmatrix}^\top \label{eq:vecA} \, ,\\
-%\mathbf{B} & = &
-% \begin{bmatrix}
-% \quad\\
-%  \Delta \mathbf{y}_{p+1} & 
-%  \dots &
-%  \Delta \mathbf{y}_N \label{eq:vecB}\\
-%  \quad
-% \end{bmatrix}^\top \, ,\\
-%\mathbf{X}&=&
-%  \begin{bmatrix}
-%   \quad \\
-%   \alpha & \phi_1^* & \cdots & \phi_{p-1}^* & \mathbf{c} \\  
-%   \quad
-%   \end{bmatrix}^\top \, ,\\
-%\mathbf{E} &=&
-%\begin{bmatrix}
-%   \quad \\
-%   \mathbf{\epsilon}_{p+1} &
-%   \dots &
-%   \quad &\mathbf{\epsilon}_N \, \\ \quad
-%\end{bmatrix}^\top 
-%\end{eqnarray}
-%
+The matricial form of the VEC model is the following:
 
@@
 
-VAR and VECM parameters shown in equations (\ref{eq:vareq},\ref{eq:vecmatrix}) can be solved
-using standard regression techniques, such as ordinary least squares (OLS).
-
-\subsection{Unit Root Tests}
-In order to use VECM the time series are required to be I(1) which is detected
-using a unit root test. The Dickey-Fuller (ADF) test ~\cite{dickey1979} is a
-unit root test where the null hypothesis is that the
-variable has a unit root.
-
-If we have a variable with the form:
-\[
-\mathbf{x}_t = \alpha \mathbf{x}_{t-1} + \epsilon_t
-\]
-
-\noindent and if we substract $\mathbf{x}_{t-1}$ from both sides we get:
-
-\begin{equation}
-\label{eq:DF}
-\Delta \mathbf{x}_t = (\alpha-1) \mathbf{x}_{t-1} + \epsilon_t
-\end{equation}
-
-Dickey fuller test consist on determine if $\alpha$ equals 1. The alternative
-hypothesis is $\alpha < 1$ which means that $\mathbf{x}_t$ is a stationary
-process.
-
-In order to include more than 1 lagged variable the Augmented Dickey-Fuller
-test was developed and their tabulated critical values were obtained
-by~\cite{mackinnon2010}. The number of lags is often selected using Schwartz
-criterion.
-
-\subsection {Cointegration Tests}
-
-Johansen method~\cite{johansen1988} allows to find cointegration among several
-I(1) time series. This test is more general than the
-Engle-Granger~\cite{engle1987} test which only allow to find cointegration
-between two time series.  Johansen procedure takes the form of a
-likelihood-ratio test which determines $r$ cointegration relationships. The
-method provides two tests based on an eigenvector decomposition of the matrix
-$\Omega = \alpha \beta^\top$ on equation~\ref{eq:vec} called trace and maximal
-eigenvalue statistic.  Once the number of cointegration number is determined,
-$r$ more significative eigenvectors provided by Johansen procedure are used to
-get the cointegration vector matrix $\beta^\top$.  Critical values for both
-tests are limited to 12 and 5 variables in the trace and maximal eigenvalue
-statistic respectively.
-
-The trace test null hypothesis is $H_0: r = r_0$ against $H_1: r > r_0$ and the
-maximum eigenvalue test null hypothesis is $H_0: r = r_0$ against $H_1: r =
-r_0+1$ starting with $r_0 = 0$.  In the latest test, if we start with $r_0=0$
-and $H_0$ is rejected $H_1: r = 1$ will be accepted which is not true if there
-are more than one cointegration relationship, this is why this test is less
-powerful and the trace test is commonly used.
-
-\subsection{Ordinary Least Squares method}
-
-When $\mathbf{A}$ is singular, solution to equation~(\ref{eq:vecmatrix}) is given
-by the ordinary least squares (OLS) method. OLS consists of minimizing the sum
-of squared errors or equivalently minimizing the following expression:
-
-\begin{equation}
-\label{eq:regressionproblem}
-\underset{\mathbf{X}}{\text{min}} \quad \| \mathbf{A}\mathbf{\mathbf{X}} - \mathbf{B} \|_2^2
-\end{equation}
-
-\noindent for which the solution $\hat{\mathbf{X}}$ is well-known:
-
-\begin{equation}
-\label{eq:MP}
-\hat{\mathbf{X}}=\mathbf{A}^{\!\!+}\,\mathbf{B}
-\end{equation}
-
-\noindent where $\mathbf{A}^{\!\!+}$ is the Moore-Penrose pseudo-inverse
-which can be written as follows:
-
-\begin{equation}
-\label{eq:pseudoinverse}
-\mathbf{A}^{\!\!+}= (\mathbf{A}^{\!\!\top} \mathbf{A})^{-1}\mathbf{A}^{\!\!\top} \, .
-\end{equation}
-
-However, when $\mathbf{A}$ is not full rank, i.e
-$rank(\mathbf{A})=k <  n \leq m$, $\mathbf{A}^\top \mathbf{A}$ is
-always singular and equation~(\ref{eq:pseudoinverse}) cannot be used.
-More generally, the pseudo-inverse is best computed using the compact
-singular value decomposition (SVD) of $\mathbf{A}$:
-
-\begin{equation}
-    \label{eq:compactsvd}
-    \underset{m \times n}{\mathbf{A}}=
-    \underset{m \times k}{\mathbf{U_1}} \enskip
-    \underset{k \times k}{\Sigma_1} \enskip
-    \underset{k \times n}{\mathbf{V}_1^{\top}} \, ,
-\end{equation}
-
-\noindent as follows
-
-\begin{equation}
-\label{eq:pseudoinversesvd}
-\mathbf{A}^{\!\!+} = \mathbf{V}_1 \Sigma_1^{-1} \mathbf{U}_1^\top \, .
-\end{equation}
-
-%Descomentar en la tesis
-\textbf{Demo}\quad
-
-Since $\mathbf{A}$ is singular the problem shown in
-equation~(\ref{eq:regressionproblem}) has not solution, the minimum norm given
-by equation~(\ref{eq:MP}) is obtained by solving the equivalent problem:
-
-\begin{equation*}
-\label{eq:proyectorsol}
-\mathbf{A \hat{X} = PB} 
-\end{equation*}
-
-\noindent where $\mathbf{P=U_1 U_1^\top}$ is the projection onto the
-Col($\mathbf{A}$).
-
-Since $\mathbf{V} = [\underset{(n \times k)}{\mathbf{V_1}} |
-\underset{(n \times k)}{\mathbf{V_2}}]$ and $\mathbf{V_1^\top V_2 =
-0}$ we can express $\mathbf{\hat{X}} = \mathbf{V_1 x_1 + V_2 x_2}$
-with $\mathbf{x_2=0}$ because $\mathbf{\hat{X}}$ lives in the
-$\text{Row}(\mathbf{A})$ given by $\mathbf{V_1}$, so we have:
-
-\begin{eqnarray*}
-\mathbf{A \hat{X}} &=& \mathbf{PB} \\
-\mathbf{U_1 \Sigma_1 V_1^\top \hat{X}} &=& \mathbf{U_1 U_1^\top B} \\
-\mathbf{ V_1^\top \hat{X}} &=&  \mathbf{\Sigma_1^{-1} U_1^\top B} \\ 
-\mathbf{ V_1^\top V_1 x_1} &=& \mathbf{\Sigma_1^{-1}
-U_1^\top B} \\
-\mathbf{x_1}&=& \mathbf{\Sigma_1^{-1} U_1^\top B}
-\end{eqnarray*}
-
-\noindent from this result we can obtain $\mathbf{\hat{X}}$ and
-therefore the pseudo-inverse expression:
-
-\begin{eqnarray*}
-\mathbf{\hat{X}} &=& \mathbf{V_1 x_1} \\
-                &=& \mathbf{V_1 \Sigma_1^{-1} U_1^\top B} \\
-\mathbf{A^+} &=& \mathbf{V_1 \Sigma_1^{-1} U_1^\top} \, .
-\end{eqnarray*}
-
-\subsection{Ridge Regression}\label{sec:RR}
-
-Ridge regression (RR) introduces a ridge estimation to avoid the
-singularity of the matrix $\mathbf{A}^\top \mathbf{A}$
-equation~(\ref{eq:pseudoinverse}). RR solves the following
-optimization problem:
-
-\begin{equation}
-\label{eq:RRproblem}
-\underset{\mathbf{X}}{\text{min}} \quad \|
-\mathbf{A}\mathbf{\mathbf{X}} - \mathbf{B} \|_2^2 +\lambda \|
-\mathbf{\mathbf{X}}\|_2^2 
-\end{equation}
-
-\noindent where $\lambda$ is a regularization parameter.
-
-The RR optimal solution $\mathbf{X}(\lambda)$ is also well-known:
-
-\begin{equation}
-\label{eq:optsolRR}
-\mathbf{X}(\lambda)=(\mathbf{A}^\top \mathbf{A}+ \lambda
-\mathbb{I})^{-1}\mathbf{A}^\top \mathbf{B} \, . 
-\end{equation}
-
-\textbf{Demo}\quad
-
-To solve equation~(\ref{eq:optsolRR}) is equivalent to solve the following
-optimization problem:
-
-\begin{equation}
-\label{eq:RRproblem2}
-\underset{\mathbf{X}(\lambda)}{\text{min}} \quad \|
-\mathbf{C}\mathbf{\mathbf{X}} - \mathbf{F} \|_2^2
-\end{equation}
-
-\noindent where
-
-\begin{equation*}
-	\mathbf{C}=\begin{bmatrix} \quad &\mathbf{A}& \quad \\ \hdashline \quad
-& \lambda\mathbb{I}&\quad  \end{bmatrix} \quad \text{and} \quad
-	\mathbf{F}=\begin{bmatrix} \quad &\mathbf{B}& \quad \\ \hdashline \quad
-& \mathbf{0}&\quad  \end{bmatrix} \, .
-\end{equation*}
-
-Applying equation~(\ref{eq:MP}) and considering that $\mathbf{C}^\top
-\mathbf{C} = \mathbf{A}^\top \mathbf{A} + \lambda \mathbb{I}$ and 
-$\mathbf{C}^\top \mathbf{F}=\mathbf{A}^\top \mathbf{B} $ we have:
-
-\begin{eqnarray*}
-\mathbf{X}(\lambda)&=&(\mathbf{C}^\top
-\mathbf{C})^{-1}\mathbf{C}^\top \mathbf{F} \\
-&=& (\mathbf{A}^\top \mathbf{A} + \lambda \mathbb{I})^{-1} \mathbf{A}^\top \mathbf{B}
-\end{eqnarray*}
-
-\subsection{Online Ridge regression}
-
-If we represent matrices $\mathbf{A}$ and $\mathbf{B}$ such as:
-
-\begin{equation}
-\label{eq:notation}
-	\mathbf{A} = 
-\left[
-  \begin{tabular}{c>{$}c<{$}c}
-    --- & \mathbf{a}^{\top}_{1} & ---\\
-    --- & \mathbf{a}^{\top}_{2} & ---\\
-    & \vdots & \\
-    --- & \mathbf{a}^{\top}_{m} & ---
-  \end{tabular}
-\right]
-\quad \text{and} \quad
-\mathbf{B} =
-\left[
-  \begin{tabular}{c>{$}c<{$}c}
-    --- & \mathbf{b}_{1} & ---\\
-    --- & \mathbf{b}_{2} & ---\\
-    & \vdots & \\
-    --- & \mathbf{b}_{m} & ---
-  \end{tabular}
-\right] \, ,
-\
-\end{equation}
-
-\noindent equation~(\ref{eq:optsolRR}) can also be written as:
-
-\begin{eqnarray*}
-\label{eq:RReapand}
-\mathbf{\mathbf{X}}_{\text{ridge}}&=&(\mathbf{A}^\top \mathbf{A}+ \lambda
-\mathbb{I})^{-1}\mathbf{A}^\top \mathbf{B} \\
-&=& \displaystyle \big (\sum_{t=1}^m
-\mathbf{a}_t \mathbf{a}_t  ^\top + \lambda \mathbb{I}\big )^{-1}
-\sum_{t=1}^m \mathbf{a}_t \mathbf{b}_t \, .
-\end{eqnarray*}
-
-Lets define $\displaystyle\mathbf{S}= \sum_{t=1}^m \mathbf{a}_t
-\mathbf{a}_t  ^\top + \lambda \mathbb{I} $ and $\mathbf{W}=
-\displaystyle\sum_{t=1}^m \mathbf{a}_t \mathbf{b}_t$, so the
-algorithm~\ref{alg:RR} shows the iterative formulation:
-
-\begin{algorithm}[H]
-\begin{algorithmic}[1]
-\REQUIRE $\,$ \\
-$\{\mathbf{a}_1,\dots,\mathbf{a}_{m} \}$: $m$ input vectors \\
-$\{\mathbf{b}_1,\dots,\mathbf{b}_{m} \}$: $m$ target vectors \\
-$\lambda$: regularization parameter \\
-\ENSURE  $\,$ \\
-$\{f(\mathbf{a}_1),\dots,f(\mathbf{a}_{m}) \}$: model predictions \\
-\STATE Initialize $\mathbf{S}=\lambda \mathbb{I}$
-and $\mathbf{W}=0$
-\FOR { $t = 1$ to $m$ }
-	\STATE read new $\mathbf{a}_t$
-	\STATE $\mathbf{X}=\mathbf{S}^{-1}\mathbf{W}$
-	\STATE output prediction $f(\mathbf{a}_t) = \mathbf{X}^\top \mathbf{a}_t$
-   	\STATE $\mathbf{S} = \mathbf{S} + \mathbf{a}_t \mathbf{a}_t^\top$
-   	\STATE Read new $y_t$
-    	\STATE $\mathbf{W} = \mathbf{W} + \mathbf{a}_t \mathbf{b}_t$
-\ENDFOR
-\end{algorithmic}
-\caption{Online Ridge Regression}
-\label{alg:RR}
-\end{algorithm}
-
-\subsection{The Aggregating Algorithm for Regression}
-
-The AAR, proposed by~\cite{vovk2001}, is an application of the aggregating
-algorithm to the problem of regression. The idea is to introduce the new input
-vector $\mathbf{x}_{m+1}$ to solve the model parameters:
-
-\begin{equation}
-\label{eq:AARexpand}
-\mathbf{X}_{aar} = \displaystyle \big (\sum_{t=1}^{m+1}
-\mathbf{a}_t \mathbf{a}_t  ^\intercal + \gamma \mathbb{I}\big )^{-1}
-\sum_{t=1}^m \mathbf{a}_t \mathbf{b}_t \, .
-\end{equation}
-
-If we define $\displaystyle\mathbf{S}= \sum_{t=1}^{m+1} \mathbf{a}_t
-\mathbf{a}_t  ^\intercal + \gamma \mathbb{I} $ and $\mathbf{W}=
-\displaystyle\sum_{t=1}^m \mathbf{a}_t \mathbf{b}_t$, the
-algorithm~\ref{alg:AAR} is slightly different to the algorithm~\ref{alg:RR}, 
-which updated matrix $\mathbf{S}$ before making the prediction:
-
-\begin{algorithm}[ht]
-\begin{algorithmic}[1]
-\REQUIRE $\,$ \\
-$\{\mathbf{a}_1,\dots,\mathbf{a}_{m} \}$: $m$ input vectors \\
-$\{\mathbf{b}_1,\dots,\mathbf{b}_{m} \}$: $m$ target vectors \\
-$\lambda$: regularization parameter \\
-\ENSURE  $\,$ \\
-$\{f(\mathbf{a}_1),\dots,f(\mathbf{a}_{m}) \}$: model predictions \\
-\STATE Initialize $\mathbf{S}=\lambda \mathbb{I}$
-and $\mathbf{W}=0$
-\FOR { $t = 1$ to $m$ }
-	\STATE read new $\mathbf{a}_t$
-   	\STATE $\mathbf{S} = \mathbf{S} + \mathbf{a}_t \mathbf{a}_t^\intercal$
-	\STATE $\mathbf{X}=\mathbf{S}^{-1}\mathbf{W}$
-	\STATE output prediction $f(\mathbf{a}_t) = \mathbf{X}^\intercal \mathbf{a}_t$
-   	\STATE Read new $\mathbf{y}_t$
-    	\STATE $\mathbf{W} = \mathbf{W} + \mathbf{a}_t \mathbf{b}_t$
-\ENDFOR
-\end{algorithmic}
-\caption{{\em The aggregating algorithm for regression}}
-\label{alg:AAR}
-\end{algorithm}
-
-\subsection{Selection of Lambda}
-
-The additional term $\lambda \|\mathbf{\mathbf{X}}\|_2^2$ in the optimization
-problem shown in equation~(\ref{eq:RRproblem}) has two effects on the solution:
-shrinks the coefficients towards zero and improves the conditioning of the
-problem.
-
-Figure~\ref{fig:shrinks} shows a visual example of the shrinking of the
-coefficients:
-
-\begin{figure}[h!]
-\includegraphics[width=0.5\linewidth]{img/shrinks}
-\caption{Shrink of regression coefficients}
-\label{fig:shrinks}
-\end{figure}
-
-On the other hand, the effect of adding the term $\lambda \mathbb{I}$
-to the matrix $\mathbf{A}^\top \mathbf{A}$
-(equation~(\ref{eq:optsolRR})) improves its condition number since it
-increases its diagonal values when $\lambda > 0 $.  The matrix
-$\mathbf{A}^\top \mathbf{A}$ is symmetrical ($(\mathbf{A}^\top
-\mathbf{A})^\top = \mathbf{A}^\top \mathbf{A}$) and therefore
-diagonalizable.  If we know the eigenvalue decomposition of $\mathbf{A}
-= \mathbf{U\Sigma V^\top}$, then:
-
-\begin{eqnarray*}
-\mathbf{A}^\top \mathbf{A}+\lambda \mathbb{I}&=&\mathbf{V\Sigma^2
-V}^\top + \lambda \mathbf{V} \mathbf{V}^\top\\ &=&\mathbf{V}
-(\mathbf{\Sigma}^2+\lambda\mathbb{I}) \mathbf{V}^\top \, ,
-\end{eqnarray*}
-
-\noindent where
-
-\begin{equation*}
-\mathbf{\Sigma}^2+\lambda\mathbb{I}=
-\begin{bmatrix}
-\sigma^2_1 + \lambda & \, & \, \\
-\, & \sigma^2_2 +\lambda & \, \\
-\, & \, & \ddots & \, \\
-\, & \, & \, & \sigma^2_n +\lambda \, .
-\end{bmatrix}
-\end{equation*}
-
-where $\sigma_1 \geq \sigma_2 \geq \dots \geq \sigma_n$.
-
-Since the condition number of a matrix $\mathbf{A}$ is defined as:
-
-\begin{equation*}
-	\kappa = \|\mathbf{A}\| \|\mathbf{A}^{-1}\|
-\end{equation*}
-
-If matrix $\mathbf{A}$ is non-singular, its condition number can be
-expressed in terms of its singular values. The effect of adding the
-regularization term affects the condition number as follows:
-
-\begin{eqnarray*}
-\kappa_{ols} &=& \|\mathbf{A}\| \|\mathbf{A}^{-1}\|=\frac{\sigma_1}{\sigma_n} \\
-\kappa_{ridge} &=& \|\mathbf{A}^\intercal \mathbf{A} + \lambda \mathbb{I}\| 
-\|(\mathbf{A}^\intercal \mathbf{A} + \lambda \mathbb{I})^{-1}\|=\frac{\sigma_1+\lambda}{\sigma_n + \lambda} \,
-\end{eqnarray*}
-
-It is easy to see that the term $\lambda$ improves the condition number:
-
-\begin{equation*}
-        \frac{\sigma_1+\lambda}{\sigma_n + \lambda} <
-        \frac{\sigma_1}{\sigma_n} \,  \qquad \forall \quad \lambda > 0
-\end{equation*}
-
-However, $\lambda$ cannot be too large. Tipically $\lambda$ is small and its
-magnitude depends on the matrix $\mathbf{A}$.
-
-For rank deficient matrices we know that $\text{det}(\mathbf{A A^\top})=0$, adding the
-term $\lambda \mathbb{I}$ we have that $\text{det}(\mathbf{A A^\top}+\lambda \mathbb{I}) =
-p(\lambda)$ where $p(\lambda)$ is a polynomial of degree $n$ ($\mathbf{A}$ is
-$m \times n$). The zeros of $p(\lambda)$ are discretes, so it can be represented
-as:
-
-\[
-p(\lambda) =
-\lambda(\lambda-\lambda_1)^{n_1}(\lambda-\lambda_2)^{n_2}\dots(\lambda-\lambda_s)^{n_s}
-\]
-
-\noindent where $n_1 + n_2 + \dots + n_s = n$.
-
-This means that $\lambda$ must be small in order to ensure that $p(\lambda)$
-does not vanish.
-
-\subsubsection{The bias-variance tradeoff}
-
-To determine parameter $\lambda$ is crucial for ridge regression since it could
-reduce the expected prediction error by reducing variance considering a biased
-estimator, this is called the bias-variance trafeoff. The prediction error is
-obtained as:
-
-\begin{equation}
- \label{eq:prederror}
-E[(B-\hat{f}(\mathbf{X}))^2] = \sigma^2 + Bias(\hat{f}(\mathbf{X}))^2 + Var(\hat{f}(\mathbf{X}))
-\end{equation}
-
-\noindent where $\hat{f}(\mathbf{X})=\mathbf{AX}(\lambda)$ and
-$f(\mathbf{X})=\mathbf{AX}$.
-
-\textbf{Demo}\quad
-
-\begin{eqnarray*}
-E[(B-\hat{f}(\mathbf{X}))^2] & =&
-E[(B-f(\mathbf{X})+f(\mathbf{X})-\hat{f}(\mathbf{X}))^2] \\
-&=& E[(B-f(\mathbf{X}))^2] +
-E[(f(\mathbf{X})-\hat{f}(\mathbf{X}))^2] + \dots \\
-& &
-2\cancelto{0}{E[B-f(\mathbf{X})]}E[f(\mathbf{X})-\hat{f}(\mathbf{X})] \\
-&=& \sigma^2 + MSE(\hat{f}(\mathbf{X}))
-\end{eqnarray*}
-
-\noindent where
-
-\begin{eqnarray*}
-    MSE(\hat{f}(\mathbf{X})) &=&
-    E[f(\mathbf{X})-\hat{f}(\mathbf{X}))]^2 \\
-    &=& E[f(\mathbf{X})-E[\hat{f}(\mathbf{X})] +
-    E[\hat{f}(\mathbf{X})]-\hat{f}(\mathbf{X}))]^2 \\
-    &=&  E[(f(\mathbf{X})-E[\hat{f}(\mathbf{X})])^2] +
-      E[(\hat{f}(\mathbf{X})-E[\hat{f}(\mathbf{X})])^2] + \dots\\
-    & & E[(f(\mathbf{X})-E[\hat{f}(\mathbf{X})])] 
-    \cancelto{0}{E[(\hat{f}(\mathbf{X})-E[\hat{f}(\mathbf{X})])]} \\
-    &=& Bias(\hat{f}(\mathbf{X}))^2 + Var(\hat{f}(\mathbf{X}))
-\end{eqnarray*}
-
-The bias of OLS is:
-
-\begin{eqnarray*}
-Bias(\hat{\mathbf{X}}) &=& E[\hat{\mathbf{X}}] - \mathbf{X} \\
-&=& E[ (\mathbf{A}^\top \mathbf{A})^{-1}\mathbf{A}^\top \mathbf{B}] - \mathbf{X} \\
-&=& E[ (\mathbf{A}^\top \mathbf{A})^{-1}\mathbf{A}^\top (\mathbf{AX})] - \mathbf{X}  \\
-&=& \mathbf{X}  - \mathbf{X}  \\
-&=&  0
-\end{eqnarray*}
-
-The bias of ridge regression when $\mathbf{A A^\top}$ is non-singular
-can be obtained expressing ridge regression solution
-$\mathbf{\lambda}$ in terms of OLS solution $\hat{\mathbf{X}}$:
-
-\begin{eqnarray*}
-\mathbf{X}(\lambda) &=&( \mathbf{A}^\top \mathbf{A} + \lambda \mathbb{I})^{-1}\mathbf{B} \\
-&=& (\mathbb{I} + \lambda (\mathbf{A}^\top \mathbf{A})^{-1})^{-1} (\mathbf{A}^\top \mathbf{A})^{-1}\mathbf{A}^\top \mathbf{B} \\
-&=&  (\mathbb{I} + \lambda (\mathbf{A}^\top \mathbf{A})^{-1})^{-1}  \hat{\mathbf{X}} \\
-&=& \mathbf{W} \hat{\mathbf{X}} 
-\end{eqnarray*}
-
-\noindent where $\mathbf{W}  = (\mathbb{I} + \lambda (\mathbf{A}^\top
-\mathbf{A})^{-1})^{-1}  $ it is defined for simplicity. Ridge
-regression bias is then obtained as:
-
-\begin{eqnarray*}
-Bias(\mathbf{X}(\lambda)) &=& E[\mathbf{X}(\lambda)] - \mathbf{X} \\
-&=& E[\mathbf{W}\hat{\mathbf{X}}] - \mathbf{X} \\
-&=&  \mathbf{W} \mathbf{X} - \mathbf{X} \neq 0 
-\end{eqnarray*}
-
-The variance of OLS is:
-
-\begin{equation*}
-Var(\hat{\mathbf{X}}) = \sigma^2 (\mathbf{A}^\top \mathbf{A} )^{-1}
-\end{equation*}
-
-\noindent and the variance of ridge regression is:
-
-\begin{eqnarray*}
-Var(\mathbf{X}(\lambda)) &=& Var(\mathbf{W}\hat{\mathbf{X}}) \\ 
-&=& \mathbf{W}Var(\hat{\mathbf{X}})\mathbf{W}^\top \\
-&=& \sigma^2 \mathbf{W}(\mathbf{A}^\top \mathbf{A} )^{-1}\mathbf{W}^\top
-\end{eqnarray*}
-
-\begin{figure}[h!]
-\includegraphics[width=0.8\linewidth]{img/biasvariance}
-\caption{Bias-variance tradeoff}
-\label{fig:biasvariance}
-\end{figure}
-
-The figure~\ref{fig:biasvariance} shows the bias-variance tradeoff
-given by equation~(\ref{eq:prederror}). Ridge regression shows an
-increasing bias and a decreasing variance. Despite OLS has zero bias,
-its variance is greater than ridge for small values of $\lambda$.
-It can be shown that, in terms of prediction error, ridge (black line)
-is lower than OLS (dotted line)~\cite{hoerl1970}.
-
-\subsubsection{Methods for lambda selection}
-
-Despite the bias-variance tradeoff provides a conceptual framework for
-determining a good model, is not directly useful. Some popular methods
-for determine model selection are:
-
-\begin{itemize}
-    \item Akaike information criterion (AIC) (Akaike, 1974)
-    \item Bootstrap-based selection (Efron and Tibshirani, 1997)
-    \item Cross-validation (Stone, 1974)
-\end{itemize}
-
-On the other hand, $\lambda$ can also be used to reduce computational time as it
-is shown in equation~(\ref{eq:taylor}). Golub (~\cite{golub1965}) suggested that
-$\lambda$ had to be chosen so that:
-
-\[
-\frac{\lambda}{\lambda + \delta^2} < 0.1
-\]
-
-\noindent where $\delta$ is a lower bound of the smallest non-zero singular
-value $\sigma_k$. One approach is to choose the greatest lower bound for
-$\delta=\sigma_k$. Hence:
-
-\[
-\frac{\lambda}{\lambda + \sigma_k^2} < 0.1
-\]
-
-\noindent if we choose $\lambda=\beta \sigma_k^2$ then we get the following
-relation:
-
-\[
-\frac{\beta \sigma_k^2}{\beta \sigma_k^2 + \sigma_k^2} =
-\frac{\beta}{\beta+1} < 0.1 \Rightarrow \beta < \frac{1}{9}  
-\]
-
-Experiments done by ~\cite{coleman+sun2010} have shown that $\beta = 0.01$
-produces satisfactory results. This means that $\lambda = 0.01 \sigma_k^2$.
-
-However, get $\sigma_k$ imply obtaining first the SVD, which is
-computational expensive.
-
-Since the algorithm presented by ~\cite{coleman+sun2010} already compute the QR
-factorization of the matrix $\mathbf{A}$ they show a procedure for getting
-a $\lambda$ approximation using QR.
-
-\begin{enumerate}
-\item Compute QR factorization: $\mathbf{A} = \mathbf{Q_1R_1}$
-\item Let $\mathbf{W}$ denote the set of absolute values of the nonzero diagonal
-elements of $\mathbf{R_1}$. Let $w_{\text{min}}$ and $w_{\text{max}}$ denote the
-smallest and largest elements of $\mathbf{W}$ respectively. Both 
-$\lambda_1 = \hat{\beta} w_{\text{min}}^2$
-$\lambda_2 = \hat{\beta} \frac{w_{\text{min}}^2}{w_{\text{max}}^2}$ where
-$\hat{\beta}=0.00025$ produce satisfactory results.
-\item The $\lambda_{\text{QR}}$ approximation is obtained as:
-$\lambda_{\text{QR}} = \frac{\lambda_1 + \lambda_2}{2}$  
-\end{enumerate}
-
-\subsection{Efficient computation}
-
-%Revisar!!!
-In regression problems we always require getting a matrix inverse which is
-computationally expensive. However, in stream data problems the is a way to
-obtain a matrix inverse approximation using the Sherman-Morrison-Woodbury. This
-formula allows to get the inverse of a
-the matrix $\mathbf{A+uv^\top}$ if we already calculated the
-inverse of $\mathbf{A}$ as follows:
-
-\begin{equation}
-\label{eq:SMW}
-(\mathbf{A+uv^\top})^{-1}=\mathbf{A}^{-1}-
-\frac{\mathbf{A^{-1}uv^\top A^{-1}}}{\mathbf{1+v^\top A^{-1}u}}
-\end{equation}
-
-%Revisar!!
-
-Another way was presented by Coleman and
-Sun~\cite{coleman+sun2010} using an iterative algorithm which
-uses $\mathbf{X}(\lambda)$ to approximate $\mathbf{X}(0)$.
-
-Using the compact SVD (shown in equation~(\ref{eq:compactsvd}))
-$\mathbf{X}(\lambda)$ is expressed as follows:
-
-\begin{eqnarray}
-\label{eq:optsolRRsvd}
-\mathbf{X}(\lambda) & = & (\mathbf{A}^\top \mathbf{A}+ \lambda
-\mathbb{I})^{-1}\mathbf{A}^\top \mathbf{B} \nonumber \\
-& = &\mathbf{V}_1(\Sigma_1^2+\lambda \mathbb{I})^{-1}\mathbf{\Sigma_1
-U_1^\top B}
-\end{eqnarray}
-
-\noindent where is easy to see that $\mathbf{X}(\lambda) \rightarrow
-\mathbf{\hat{X}}=\mathbf{V}_1 \mathbf{\Sigma}_1^{-1}\mathbf{U}_1^\top
-\mathbf{B}$ as $\lambda \rightarrow 0$.
-
-The method consists in obtaining $\mathbf{X}(\lambda)$ and then refine
-by adding more terms of its Taylor expansion to approximate
-$\mathbf{X}(0) = \mathbf{\hat{X}}$. The Taylor expansion about $\lambda_0$ is:
-
-\begin{equation}
-\label{eq:taylor}
-    \mathbf{X}(\lambda)=\mathbf{X}(\lambda_0) + \sum_{k=1}^\infty
-    \mathbf{s}_k(\lambda-\lambda_0)^{k}
-\end{equation}
-
-\noindent where $\mathbf{s}_k=\frac{1}{k!}\mathbf{X}(\lambda)^{(k)}$
-and $\mathbf{X}(\lambda)^{(k)}$ is obtained by taking differences 
-$\frac{\partial}{\partial \lambda}$ of:
-
-\begin{eqnarray*}
-(\mathbf{A}^\top \mathbf{A}+ \lambda\mathbb{I}) \mathbf{X}(\lambda) & = & \mathbf{A}^\top \mathbf{B}\\
-(\mathbf{A}^\top \mathbf{A}+ \lambda\mathbb{I}) \mathbf{X}(\lambda)^{(1)} + \mathbf{X}(\lambda)& = & 0 \\
-\mathbf{X}(\lambda)^{(1)}  &=& -(\mathbf{A}^\top \mathbf{A}+ \lambda\mathbb{I}) ^{-1} \mathbf{X}(\lambda) \\
-\mathbf{X}(\lambda)^{(2)}  &=& -2(\mathbf{A}^\top \mathbf{A}+ \lambda\mathbb{I}) ^{-1} \mathbf{X}(\lambda)^{(1)} \\
-& \vdots & \\
-\mathbf{X}(\lambda)^{(k)}  &=& -k(\mathbf{A}^\top \mathbf{A}+ \lambda\mathbb{I}) ^{-1} \mathbf{X}(\lambda)^{(k-1)} 
-\end{eqnarray*}
-
-\noindent for $\lambda=0$ we have:
-
-\begin{equation}
-\label{eq:taylor}
-    \mathbf{X}(0)=\mathbf{X}(\lambda_0) + \sum_{k=1}^\infty
-     (-1)^k \mathbf{s}_k \lambda_0^k
-\end{equation}
-
-\noindent therefore in order to ensure convergence, we can see that$\lambda_0$
-selection cannot be large.
-
-The algorithm for computing $\mathbf{X}(0)$ is the following:
-
-\begin{algorithm}[H]
-\begin{algorithmic}[1]
-\REQUIRE $\,$ \\
-$\mathbf{A}$: design matrix \\
-$\mathbf{B}$: response matrix \\
-$\lambda$: rank deficient parameter \\
-\ENSURE  $\,$ \\
-$\mathbf{X}$: parameters \\
-\STATE $\mathbf{M}=\mathbf{A^\top A}$ \\
-\STATE Initialize $\mathbf{Q R}=\mathbf{M}$ \\
-\STATE $\mathbf{X} = \mathbf{R^{-1}Q^\top B}$ \\
-\STATE $\mathbf{s} = \mathbf{X}$ \\
-\FOR { $i = 1,2,3,\dots$ }
-	\STATE $\mathbf{s} =
-        -(\mathbf{M}+\lambda\mathbb{I})^{-1}\mathbf{s}$\\
-	\STATE $\mathbf{X}=\mathbf{X} + (-1)^i {\color{red}\mathbf{s}
-        \lambda^i}$
-\ENDFOR
-\end{algorithmic}
-\caption{Algorithm for handling rank deficient matrices}
-\label{alg:coleman}
-\end{algorithm}
-
-The algorithm~\ref{alg:coleman} solves equation~(\ref{eq:taylor}). However, this
-version is unstable since tipically $\|\mathbf{s}\|$ is very large and
-$\lambda^i$ is very small ($\lambda$ is small).
-
-The following algorithm shows a more stable version of
-algorithm~\ref{alg:coleman}.
-
-\begin{algorithm}[H]
-\begin{algorithmic}[1]
-\REQUIRE $\,$ \\
-$\mathbf{A}$: design matrix \\
-$\mathbf{B}$: response matrix \\
-$\lambda$: rank deficient parameter \\
-\ENSURE  $\,$ \\
-$\mathbf{X}$: parameters \\
-\STATE $\mathbf{M}=\mathbf{A^\top A}$ \\
-\STATE Initialize $\mathbf{Q R}=\mathbf{M}$ \\
-\STATE $\mathbf{X} = \mathbf{R^{-1}Q^\top B}$ \\
-\STATE $\mathbf{t} = \mathbf{X}$ \\
-\FOR { $i = 1,2,3,\dots$ }
-        \STATE $\mathbf{t} =\lambda \mathbf{t}$  
-        \STATE $\mathbf{t} =  -(\mathbf{M}+\lambda\mathbb{I})^{-1}\mathbf{t}$
-	\STATE $\mathbf{X}=\mathbf{X} + \mathbf{t}$
-\ENDFOR
-\end{algorithmic}
-\caption{Algorithm for handling rank deficient matrices improved}
-\label{alg:colemanimproved}
-\end{algorithm}
-
-Both algorithms are equivalent, but algorithm~\ref{alg:colemanimproved} is more
-stable and converges in tipically less than 10 steps.
-It is important to notice that the QR factorization is computed only once and it
-is computationally less expensive than the SVD.
-
-\subsection{Online machine learning algorithms}
-
-An online algorithm allows incremental learning by processing one instance at a
-time. This is done updating the current model instead of building the model from
-scratch.
-
-Training phase is commonly a computationally expensive process. Therefore, when
-new data arrives it can't be included easily to the model. Moreover, it could
-happen that we won't have enought time to process the new data before more data
-arrives. In online learning algorithms there is no training phase, but the model
-is updated and evaluated at every time step. This model updating is
-computationally less expensive than a training phase.
-
-Online learning is specially useful for stream data problems and the model is
-sensitive to new data. It is also useful when some past data may be irrelevant
-or we want to improve computational efficiency. However, online algorithms could
-affect accuracy.
-
-There are several popular online methods such as
-perceptron~\cite{rosenblatt58},winnow~\cite{littlestone1988},
-passive-aggressive~\cite{crammerETall2006}, stochastic gradient
-descent~\cite{zhang2004}, aggregating algorithm~\cite{vovk2001} and the second
-order perceptron~\cite{cesa-bianchi2005}.  In \cite{blum1998} and
-\cite{cesa-bianchi2006} an in-deph analysis of online learning is provided.
-Applications in finance has been widely used: study presented
-by~\cite{arce+salinas2012} applied ridge regression in an online context and
-more recently, time series forecasting using online learning has been
-presented~\cite{anavaetAl2013}.
-
-Incremental learning refers to any online learning process that learns the same
-model as would be learnt by a batch learning algorithm.
-
-Incremental learning is useful when the input to a learning process is stream
-data, with the need or desire to be able to use the result of learning at any
-point in time, based on the input observations received so far.
-
-In principle, the stream of observations may be infinitely long, or the next
-observation long delayed, precluding any hope of waiting until all the
-observations have been received. One would rather not simply accumulate and
-store all the inputs and, upon receipt of each new one, apply a batch learning
-algorithm to the entire sequence of inputs received so far. It would be
-preferable compu- tationally if the existing hypothesis or other artifact of
-learning could be updated in response to each newly received input observation.
-
-Incremental learning is very useful when there is no need to record fundamental
-data and only a summary needs to be retained. Due to this, incremental
-algorithms are often characterized as memoryless, because no memory of past data
-is required.  The algorithm is online but not incremental if it doesn't produce
-the same result for all observations that the corresponding batch algorithm
-would for these same observations.
-
-Algorithm~\ref{alg:onlinealg} shows the online learning algorithm structure:
-
-\begin{algorithm}[ht]
-\begin{algorithmic}[1]
-    \STATE Receives input $\mathbf{x}_t$
-    \STATE Makes prediction $\mathbf{\hat{y}}_t$
-    \STATE Receives response $\mathbf{y}_t$
-    \STATE Incurs loss $l_t(\mathbf{y}_t,\mathbf{\hat{y}}_t)$
-\end{algorithmic}
-\caption{Structure of a Learning System}
-\label{alg:onlinealg}
-\end{algorithm}
-
-\noindent where $l$ is some loss function. Performance is later measured after
-$T$ trials as:
-
-\begin{equation*}
-L_T = \sum_{t=1}^T l_t(\mathbf{y}_t,\mathbf{\hat{y}}_t)
-\end{equation*}
-
-The objective is minimize this loss function for all instances.
-The quality of online learning algorithms is measured by a quantity known as
-regret which is the difference between the performance of the online algorithm
-and its optimal predictor $E^* \in \Theta$ given by:
-
-\begin{equation*}
-L^*_T= \text{min}_{E \in \Theta} L_T^E \, ,
-\end{equation*}
-
-\noindent where $L_T^E = \sum_{t=1}^T
-l_t(\mathbf{y}_t,\mathbf{y}^E_t)$ and $\mathbf{y}^E_t$ is the expert estimation.
-
-Therefore regret is defined as:
-
-\begin{equation*}
-R_T = L_T - L^*_T
-\end{equation*}
-
-%An online learning algorithm looks at every example exactly once.
-\subsection{Model selection}
-In order to set model parameters the Akaike Information Criterion (AIC) is
-used. AIC is often used in model selection where AIC whith smaller values are
-preferred.
-
-AIC is calculated as follows:
-
-\begin{equation}
-\label{eq:aicformula}
-AIC = \underset{\text{bias}}{-\frac{2l}{N}} + 
-\underset{\text{variance}}{\frac{2k}{N}}
-\end{equation}
-
-\noindent where
-
-\begin{description}
-\item[l] is the loglikelihood function
-\item[k] number of estimated parameters (including the variance)
-\item[N] number of observations
-\end{description}
+VAR and VEC model parameters shown in equations (\ref{eq:varmatrix}) and
+(\ref{eq:vecmatrix}) can be solve using standard regression techniques, such as
+ordinary least squares (OLS). However, matrix $\mathbf{A}$ is usually
+rank-deficient and OLS solution could not be found. Ridge regression method (RR)
+is commonly used instead of OLS when matrices are ill-conditioned or
+rank-deficient since it has improved generalization capabilities.

### 2016-03-22 articles-conferences/elsevier-2016/journal.tex
Inicio articulo 2016

--- 2016-03-22 articles-conferences/elsevier-journal-2015/journal.tex
@@
 
-\title{A Time Varying Vector Error Correction Model Applied to Foreign Exchange
-Market}
+\title{A Regularized Time Varying Vector Error Correction Model}
 \author{Paola Arce \and Jonathan Antognini \and Luis Salinas \and Werner Kristjanpoller}
@@
 In this study, we propose an online VECM which optimizes how model parameters
-are obtained by using only relevant past data and RR instead of OLS.  Our
-proposal also takes advantage of the long-run relationship between the time
-series in order to obtain improved execution times. Our proposed method is
-tested using synthetic data and foreign exchange rates.
-Results show that our proposal enables execution time to be reduced by at least
-50\% while maintaining good accuracy levels.
+are obtained by using only relevant past data and RR instead of OLS.  
 \end{abstract}
 
-\input{src/intro}
+%\input{src/intro}
 \input{src/background}
-\input{src/methodology}
-\input{src/results}
-\input{src/conclusions}
+%\input{src/methodology}
+%\input{src/results}
+%\input{src/conclusions}
 \bibliographystyle{apalike}

### 2016-03-22 articles-conferences/elsevier-2016/src/background.tex
Inicio articulo 2016

--- 2016-03-22 articles-conferences/elsevier-journal-2015/src/background.tex
@@
 
-\subsection{Spurious Regression}
+\subsection{Integration and Cointegration}\label{sec:coint}\  
+Following Johansen \cite{johansen1995} we shall say that a stochastic process
+$Y_t$ which satisfies $Y_t-E(Y_t) = \sum_{i=0}^\infty C_i\,\varepsilon_{t-i}$ is
+called $I(0)$, and then we shall write $Y_t\sim I(0)$, whenever
+$\sum_{i=0}^\infty C_i \neq 0$ and $\sum_{i=0}^\infty C_i\,z^i$ converges for
+$z\in\mathbb{C}$ with $|z|<1$.  It is understood that the condition
+$\varepsilon_t\sim iid(0,\sigma^2)$ holds.
 
-Regression analysis is widely used to test economic theories. But sometimes
-regression is done between two variables, lets say $x_t$ and $y_t$, which are
-not causally related. This occurs at the presence of a third stochastic variable
-$z_t$ which is the true causal of the first two variables $x_t$ and $y_t$. The
-regression measure R-squared gives high values in presence of spurious
-relationships. This measure would means that the model explains all the
-variability of the response data around its mean which is not true.
+A (vector) time series $\mathbf{y}_t$ is said to be {\em integrated of order\/}
+$d$, and then we shall write $\mathbf{y}_t\sim I(d)$, whenever after $d$ times
+(discrete) differentiation an stationary process is obtained~\cite{banerjee1993};
+more precisely, whenever
+$(1-L)^d\,\mathbf{y}_t\sim\text{I(0)}$, where $L$ is the usual lag operator:
+$(1-L)\,\mathbf{y}_t = \Delta\mathbf{y}_t = \mathbf{y}_t-\mathbf{y}_{t-1}$ for
+all $t$.
 
-The R-squared measure is calculated as follows:
+Note that this definition includes the scalar case as time series of
+vectors of dimension 1; in this scalar case we will write the time series in
+nonbold format.
 
+Let $\mathbf{y}_t^\nu$, $\nu=1,\dots,l$, be a set of $l$ vector time series of
+order $I(1)$.  They are said to be {\em cointegrated\/} if a vector
+$\beta=[\beta(1),\dots,\beta(l)]^\top \in \mathbb{R}^p$ exists, such that the
+time series,
 \begin{equation}
-\label{eq:rsquared}
-R^2 = 1 - \frac{\sum_{t=1}^T \epsilon_t^2}{\sum_{t=1}^T(y_t - \bar{y})^2}
+\mathbf{Z}_t:= 
+\sum_{\nu=1}^l \beta(\nu)\,\mathbf{y}_t^\nu\,\sim\,\text{I(0)}\,.
 \end{equation}
-
-In presence of spurious relationships the denominator of equation
-\ref{eq:rsquared} becomes very large because a large weight is placed on extreme
-observations on either side of the mean $\bar{y}$.
-In order to detect this, other statistic must be considered such as the
-Durbin-Watson statistic.
-
-% Acá agregar gráfico de spurious regression
-
-\subsection{Integration and Cointegration}
-A time series $\mathbf{y}$ is said to be integrated of order $d$ if after
-differentiating the variable $d$ times, we get an I(0) process, more precisely:
-
-\[
-(1-L)^d \mathbf{y} \sim \text{I(0)} \, ,
-\]
-
-\noindent where I(0) is a stationary time series and $L$ is the lag operator:
-
-\[
-(1-L)\mathbf{y} = \Delta \mathbf{y}=\mathbf{y}_t  -\mathbf{y}_{t-1} \quad \forall t
-\]
-
-Let $\mathbf{y}_t = \{\mathbf{y}^1, \dots, \mathbf{y}^l\}$ be a set of $l$
-stationary time series I(1) which are said to be cointegrated if a vector
-$\beta=[\beta(1),\dots,\beta(l)]^\top \in \mathbb{R}^l$  exists such that the
-time series,
-
-\begin{equation}
- \mathbf{Z}_t:= \beta^\top \mathbf{y}_t = \beta(1) \mathbf{y}^1 + \dots + \beta(l) \mathbf{y}^l \sim
- \text{I(0)}\, .
-\end{equation}
-
-In other words, a set of I(1) variables is said to be cointegrated if
-a linear combination of them exists which is I(0).
+In other words, a set of $I(1)$ time series is said to be cointegrated if a
+linear combination of them exists, which is I(0).
 
@@
 
-\subsection{Unit Root Tests}
-In order to use VECM the time series are required to be I(1) which is detected
-using a unit root test. The Dickey-Fuller (ADF) test ~\cite{dickey1979} is a
-unit root test where the null hypothesis is that the
-variable has a unit root.
-
-If we have a variable with the form:
-\[
-\mathbf{x}_t = \alpha \mathbf{x}_{t-1} + \epsilon_t
-\]
-
-\noindent and if we substract $\mathbf{x}_{t-1}$ from both sides we get:
-
-\begin{equation}
-\label{eq:DF}
-\Delta \mathbf{x}_t = (\alpha-1) \mathbf{x}_{t-1} + \epsilon_t
-\end{equation}
-
-Dickey fuller test consist on determine if $\alpha$ equals 1. The alternative
-hypothesis is $\alpha < 1$ which means that $\mathbf{x}_t$ is a stationary
-process.
-
-In order to include more than 1 lagged variable the Augmented Dickey-Fuller
-test was developed and their tabulated critical values were obtained
-by~\cite{mackinnon2010}. The number of lags is often selected using Schwartz
-criterion.
-
-\subsection {Cointegration Tests}
-
-Johansen method~\cite{johansen1988} allows to find cointegration among several
-I(1) time series. This test is more general than the
-Engle-Granger~\cite{engle1987} test which only allow to find cointegration
-between two time series.  Johansen procedure takes the form of a
-likelihood-ratio test which determines $r$ cointegration relationships. The
-method provides two tests based on an eigenvector decomposition of the matrix
-$\Omega = \alpha \beta^\top$ on equation~\ref{eq:vec} called trace and maximal
-eigenvalue statistic.  Once the number of cointegration number is determined,
-$r$ more significative eigenvectors provided by Johansen procedure are used to
-get the cointegration vector matrix $\beta^\top$.  Critical values for both
-tests are limited to 12 and 5 variables in the trace and maximal eigenvalue
-statistic respectively.
-
-The trace test null hypothesis is $H_0: r = r_0$ against $H_1: r > r_0$ and the
-maximum eigenvalue test null hypothesis is $H_0: r = r_0$ against $H_1: r =
-r_0+1$ starting with $r_0 = 0$.  In the latest test, if we start with $r_0=0$
-and $H_0$ is rejected $H_1: r = 1$ will be accepted which is not true if there
-are more than one cointegration relationship, this is why this test is less
-powerful and the trace test is commonly used.
-
 \subsection{Ordinary Least Squares method}
@@
 
-%Descomentar en la tesis
-\textbf{Demo}\quad
-
-Since $\mathbf{A}$ is singular the problem shown in
-equation~(\ref{eq:regressionproblem}) has not solution, the minimum norm given
-by equation~(\ref{eq:MP}) is obtained by solving the equivalent problem:
-
-\begin{equation*}
-\label{eq:proyectorsol}
-\mathbf{A \hat{X} = PB} 
-\end{equation*}
-
-\noindent where $\mathbf{P=U_1 U_1^\top}$ is the projection onto the
-Col($\mathbf{A}$).
-
-Since $\mathbf{V} = [\underset{(n \times k)}{\mathbf{V_1}} |
-\underset{(n \times k)}{\mathbf{V_2}}]$ and $\mathbf{V_1^\top V_2 =
-0}$ we can express $\mathbf{\hat{X}} = \mathbf{V_1 x_1 + V_2 x_2}$
-with $\mathbf{x_2=0}$ because $\mathbf{\hat{X}}$ lives in the
-$\text{Row}(\mathbf{A})$ given by $\mathbf{V_1}$, so we have:
-
-\begin{eqnarray*}
-\mathbf{A \hat{X}} &=& \mathbf{PB} \\
-\mathbf{U_1 \Sigma_1 V_1^\top \hat{X}} &=& \mathbf{U_1 U_1^\top B} \\
-\mathbf{ V_1^\top \hat{X}} &=&  \mathbf{\Sigma_1^{-1} U_1^\top B} \\ 
-\mathbf{ V_1^\top V_1 x_1} &=& \mathbf{\Sigma_1^{-1}
-U_1^\top B} \\
-\mathbf{x_1}&=& \mathbf{\Sigma_1^{-1} U_1^\top B}
-\end{eqnarray*}
-
-\noindent from this result we can obtain $\mathbf{\hat{X}}$ and
-therefore the pseudo-inverse expression:
-
-\begin{eqnarray*}
-\mathbf{\hat{X}} &=& \mathbf{V_1 x_1} \\
-                &=& \mathbf{V_1 \Sigma_1^{-1} U_1^\top B} \\
-\mathbf{A^+} &=& \mathbf{V_1 \Sigma_1^{-1} U_1^\top} \, .
-\end{eqnarray*}
-
 \subsection{Ridge Regression}\label{sec:RR}
@@
 \end{equation}
-
-\textbf{Demo}\quad
-
-To solve equation~(\ref{eq:optsolRR}) is equivalent to solve the following
-optimization problem:
-
-\begin{equation}
-\label{eq:RRproblem2}
-\underset{\mathbf{X}(\lambda)}{\text{min}} \quad \|
-\mathbf{C}\mathbf{\mathbf{X}} - \mathbf{F} \|_2^2
-\end{equation}
-
-\noindent where
-
-\begin{equation*}
-	\mathbf{C}=\begin{bmatrix} \quad &\mathbf{A}& \quad \\ \hdashline \quad
-& \lambda\mathbb{I}&\quad  \end{bmatrix} \quad \text{and} \quad
-	\mathbf{F}=\begin{bmatrix} \quad &\mathbf{B}& \quad \\ \hdashline \quad
-& \mathbf{0}&\quad  \end{bmatrix} \, .
-\end{equation*}
-
-Applying equation~(\ref{eq:MP}) and considering that $\mathbf{C}^\top
-\mathbf{C} = \mathbf{A}^\top \mathbf{A} + \lambda \mathbb{I}$ and 
-$\mathbf{C}^\top \mathbf{F}=\mathbf{A}^\top \mathbf{B} $ we have:
-
-\begin{eqnarray*}
-\mathbf{X}(\lambda)&=&(\mathbf{C}^\top
-\mathbf{C})^{-1}\mathbf{C}^\top \mathbf{F} \\
-&=& (\mathbf{A}^\top \mathbf{A} + \lambda \mathbb{I})^{-1} \mathbf{A}^\top \mathbf{B}
-\end{eqnarray*}
 
@@
 
-Figure~\ref{fig:shrinks} shows a visual example of the shrinking of the
-coefficients:
-
-\begin{figure}[h!]
-\includegraphics[width=0.5\linewidth]{img/shrinks}
-\caption{Shrink of regression coefficients}
-\label{fig:shrinks}
-\end{figure}
-
 On the other hand, the effect of adding the term $\lambda \mathbb{I}$
@@
 
-\textbf{Demo}\quad
-
-\begin{eqnarray*}
-E[(B-\hat{f}(\mathbf{X}))^2] & =&
-E[(B-f(\mathbf{X})+f(\mathbf{X})-\hat{f}(\mathbf{X}))^2] \\
-&=& E[(B-f(\mathbf{X}))^2] +
-E[(f(\mathbf{X})-\hat{f}(\mathbf{X}))^2] + \dots \\
-& &
-2\cancelto{0}{E[B-f(\mathbf{X})]}E[f(\mathbf{X})-\hat{f}(\mathbf{X})] \\
-&=& \sigma^2 + MSE(\hat{f}(\mathbf{X}))
-\end{eqnarray*}
-
-\noindent where
-
-\begin{eqnarray*}
-    MSE(\hat{f}(\mathbf{X})) &=&
-    E[f(\mathbf{X})-\hat{f}(\mathbf{X}))]^2 \\
-    &=& E[f(\mathbf{X})-E[\hat{f}(\mathbf{X})] +
-    E[\hat{f}(\mathbf{X})]-\hat{f}(\mathbf{X}))]^2 \\
-    &=&  E[(f(\mathbf{X})-E[\hat{f}(\mathbf{X})])^2] +
-      E[(\hat{f}(\mathbf{X})-E[\hat{f}(\mathbf{X})])^2] + \dots\\
-    & & E[(f(\mathbf{X})-E[\hat{f}(\mathbf{X})])] 
-    \cancelto{0}{E[(\hat{f}(\mathbf{X})-E[\hat{f}(\mathbf{X})])]} \\
-    &=& Bias(\hat{f}(\mathbf{X}))^2 + Var(\hat{f}(\mathbf{X}))
-\end{eqnarray*}
-
-The bias of OLS is:
-
-\begin{eqnarray*}
-Bias(\hat{\mathbf{X}}) &=& E[\hat{\mathbf{X}}] - \mathbf{X} \\
-&=& E[ (\mathbf{A}^\top \mathbf{A})^{-1}\mathbf{A}^\top \mathbf{B}] - \mathbf{X} \\
-&=& E[ (\mathbf{A}^\top \mathbf{A})^{-1}\mathbf{A}^\top (\mathbf{AX})] - \mathbf{X}  \\
-&=& \mathbf{X}  - \mathbf{X}  \\
-&=&  0
-\end{eqnarray*}
-
-The bias of ridge regression when $\mathbf{A A^\top}$ is non-singular
-can be obtained expressing ridge regression solution
-$\mathbf{\lambda}$ in terms of OLS solution $\hat{\mathbf{X}}$:
-
-\begin{eqnarray*}
-\mathbf{X}(\lambda) &=&( \mathbf{A}^\top \mathbf{A} + \lambda \mathbb{I})^{-1}\mathbf{B} \\
-&=& (\mathbb{I} + \lambda (\mathbf{A}^\top \mathbf{A})^{-1})^{-1} (\mathbf{A}^\top \mathbf{A})^{-1}\mathbf{A}^\top \mathbf{B} \\
-&=&  (\mathbb{I} + \lambda (\mathbf{A}^\top \mathbf{A})^{-1})^{-1}  \hat{\mathbf{X}} \\
-&=& \mathbf{W} \hat{\mathbf{X}} 
-\end{eqnarray*}
-
-\noindent where $\mathbf{W}  = (\mathbb{I} + \lambda (\mathbf{A}^\top
-\mathbf{A})^{-1})^{-1}  $ it is defined for simplicity. Ridge
-regression bias is then obtained as:
-
-\begin{eqnarray*}
-Bias(\mathbf{X}(\lambda)) &=& E[\mathbf{X}(\lambda)] - \mathbf{X} \\
-&=& E[\mathbf{W}\hat{\mathbf{X}}] - \mathbf{X} \\
-&=&  \mathbf{W} \mathbf{X} - \mathbf{X} \neq 0 
-\end{eqnarray*}
-
-The variance of OLS is:
-
-\begin{equation*}
-Var(\hat{\mathbf{X}}) = \sigma^2 (\mathbf{A}^\top \mathbf{A} )^{-1}
-\end{equation*}
-
-\noindent and the variance of ridge regression is:
-
-\begin{eqnarray*}
-Var(\mathbf{X}(\lambda)) &=& Var(\mathbf{W}\hat{\mathbf{X}}) \\ 
-&=& \mathbf{W}Var(\hat{\mathbf{X}})\mathbf{W}^\top \\
-&=& \sigma^2 \mathbf{W}(\mathbf{A}^\top \mathbf{A} )^{-1}\mathbf{W}^\top
-\end{eqnarray*}
-
-\begin{figure}[h!]
-\includegraphics[width=0.8\linewidth]{img/biasvariance}
-\caption{Bias-variance tradeoff}
-\label{fig:biasvariance}
-\end{figure}
-
-The figure~\ref{fig:biasvariance} shows the bias-variance tradeoff
-given by equation~(\ref{eq:prederror}). Ridge regression shows an
+Ridge regression shows an
 increasing bias and a decreasing variance. Despite OLS has zero bias,

### 2016-03-22 articles-conferences/elsevier-2016/src/conclusions.tex
Inicio articulo 2016

--- 2016-03-22 articles-conferences/icpram2015/src/conclusions.tex
@@
 \label{sec:conclusions}
-\noindent A new online vector error correction method was presented. 
-We have shown that our proposed OVECM considerably reduces execution times
-without compromising solution accuracy.  OVECM was compared with VECM and ARIMA
-with the same sliding window sizes and OVECM outperformed both in terms of
-execution time. Traditional VECM slightly outperformed our proposal but the
-OVECM execution time is lower.
-This reduction of execution time is mainly because OVECM avoids the cointegration vector
-calculation using the Johansen method. The condition for getting new vectors is
-given by the mean\_error variable which controls how many times the Johansen method is
-called. Additionally, OVECM introduces matrix optimization in order to get the
-new model in an iterative way.
-We could see that our algorithm took much less than a minute at every
-step. This means that it could also be used with higher frequency data and would
-still provide responses before new data arrives.
-
-For future study, it would be interesting to improve the out-of-sample forecast
-by considering more explicative variables, to increase window sizes or trying new
-conditions to obtain new cointegration vectors.
-
+\noindent A new online vector error correction method was presented. The traditional VECM
+was modified in order to use it in an online context. We have shown that our
+proposed OVECM considerably reduces execution times without compromising
+solution accuracy.  However the execution time of OVECM is only better than
+SLVECM if we avoid the cointegration vector calculation using the Johansen
+method, this is given by the avg\_error variable which controls how many times
+the method is called.  We could see that our algorithm took much less than a
+minute at every step. This means that it could also be used with higher
+frequency data and would still provide responses before new data arrives. 
+In all the presented scenarios, MAPE is still very high and for future
+study, it would be interesting to improve the out-of-sample forecast by
+considering more explicative variables, optimizing the windows size, the
+number of lags or trying new condition to obtain new cointegration vectors.
 Since OVECM is an online algorithm which optimizes processing time, it could be

### 2016-03-29 articles-conferences/elsevier-2016/src/background.tex
avance martes

@@
 
-VECM is a special case of VAR model and both describe the joint behaviour of a
-set of variables.
+VECM is a special case of the VAR \cite{sims1980} model and both describe the joint behaviour
+of a set of variables.
 
-VAR($p$) model is a general framework to describe the behaviour of a set of $l$
-endogenous variables as a linear combination of their last $p$ values. These
-$l$ variables at time $t$ are represented by the vector $\mathbf{y}_t$ as
-follows:
-
+The VAR($p$) model is a general framework describing the behaviour of a
+set of $l$ endogenous variables as a linear combination of their last $p$
+values, where $l,p\in\mathbb{N}$. 
+In our case, each one of these $l$ variables is a scalar time series
+$y_{\lambda,t}$, $\lambda=1,\dots,l$, and we represent them all together
+at time $t$ by the the vector time series:
 \begin{equation}
@@
 \mathbf{y}_t = 
-\begin{bmatrix} y_{1,t} &
-y_{2,t} &
-\dots &
-y_{l,t}
-\end{bmatrix}^\top \, ,
+\begin{bmatrix} y_{1,t} & y_{2,t} & \dots & y_{l,t} \end{bmatrix}^\top.
 \end{equation}
-\noindent where $y_{j,t}$ corresponds to the time series $j$ evaluated at
-time $t$.
+\noindent
+Notice that the vectors $\mathbf{y}_t$ are assumed to be $l$-dimensional.
 
-The VAR(p) model describes the behaviour of a dependent variable in terms
-of its own lagged values and the lags of the others variables in the
-system. The model with $p$ lags is formulated as the following:
+The VAR($p$) model describes the behaviour of a dependent variable in terms of
+its own lagged values and the lags of the others variables in the system. The
+model with $p$ lags is formulated as the system of $N$:
+\begin{align}
+\label{eq:var}
+\mathbf{y}_t 
+= \boldsymbol{\Phi}_1 \mathbf{y}_{t-1} +
+  \boldsymbol{\Phi}_2 \mathbf{y}_{t-2} + \dots +
+  \boldsymbol{\Phi}_p\mathbf{y}_{t-p} +
+  \mathbf{c} + \boldsymbol{\epsilon}_t \nonumber \\
+t=p+1,\dots,N,
+\end{align}
+\noindent where 
+$\boldsymbol{\Phi}_1, \boldsymbol{\Phi}_2,\dots,\boldsymbol{\Phi}_p$
+are $l\times l$-matrices of real coefficients,
+$\boldsymbol{\epsilon}_{p+1},
+ \boldsymbol{\epsilon}_{p+2}, \dots, \boldsymbol{\epsilon}_N$ 
+are error terms, $\mathbf{c}$ is a constant vector and $N$ is the total
+number of samples.
 
-\begin{equation}
-\label{eq:var}
- \mathbf{y}_t = \phi_1 \mathbf{y}_{t-1}  + \dots +   \phi_p\mathbf{y}_{t-p}
-+ \mathbf{c} + \mathbf{\epsilon}_t \, ,
+Notice that, regarding our notation of section (\ref{sec:coint}),
+we have here 
+$\mathbf{y}_t^0 = \mathbf{y}_t$,
+$\mathbf{y}_t^\nu = \mathbf{y}_{t-\nu}$ and
+the $\lambda$-th component of the vector time series $\mathbf{y}_t^\nu$
+is the scalar time series $y_{\lambda,t}^\nu$, where $\nu=1,\dots,p$ and
+$\lambda=1,\dots,l$.
+
+However, the VAR model cannot be used with non-stationary variables. VECM \cite{engle87} is also a linear model and a special form of a VAR model for I(1) variables that are also
+cointegrated~\cite{banerjee1993}. If cointegration exists, variable differences
+are stationary and they introduce an error correction term which adjusts
+coefficients to bring the variables back to equilibrium.
+
+It is obtained re-writing equation (\ref{eq:var}) in terms of the new
+variable $\Delta\mathbf{y}_t=\mathbf{y}_t-\mathbf{y}_{t-1}$.
+The VECM model, expressed in terms those differences, takes the form:
+\begin{equation}\label{eq:vec}
+\Delta \mathbf{y}_t 
+= \boldsymbol{\Omega}\,\mathbf{y}_{t-1}
+  + \sum_{i=1}^{p-1} \boldsymbol{\Phi}_i^*\,\Delta\mathbf{y}_{t-i}
+  + \mathbf{c} + \boldsymbol{\epsilon}_t\,,
 \end{equation}
+\noindent
+where the coefficients matrices $\boldsymbol{\Phi}_i^*$ and 
+$\boldsymbol{\Omega}$, expressed in terms of the matrices
+$\boldsymbol{\Phi}_i$ of (\ref{eq:var}), are:
+\begin{align*}
+\boldsymbol{\Phi}_i^* 
+&:= -\sum_{j=i+1}^{p}\boldsymbol{\Phi}_j\,, \\
+\boldsymbol{\Omega}
+&:= -\left( \mathbb{I} - \boldsymbol{\Phi}_1 - \dots 
+    - \boldsymbol{\phi}_p \right)\,. 
+\end{align*}
+The following well known properties of the matrix $\boldsymbol{\Omega}$
+\cite{johansen1995} will be useful in the sequel:
+\begin{itemize}
+\item
+If $\boldsymbol{\Omega} = \mathbf{0}$, there is no cointegration.
+\item 
+If $rank(\boldsymbol{\Omega})=l$, i.e., if $\boldsymbol{\Omega}$ has
+full rank, then the time series are not I(1) but stationary.
+\item
+If $rank(\boldsymbol{\Omega})=r$, $0<r<l$, then there is cointegration
+and the matrix $\boldsymbol{\Omega}$ can be expressed as
+$\boldsymbol{\Omega}=\boldsymbol{\alpha\beta}^\top$, where $\boldsymbol{\alpha}$
+and $\boldsymbol{\beta}$ are
+$l\times r$ matrices and
+$\text{rank}(\boldsymbol{\alpha})=\text{rank}(\boldsymbol{\beta})=r$.
+\item
+The columns of $\boldsymbol{\beta}$ contains the cointegration vectors and the rows of
+$\boldsymbol{\alpha}$ correspond with the adjusted vectors. 
+$\boldsymbol{\beta}$ is obtained by Johansen procedure~\cite{johansen1988},
+whereas $\boldsymbol{\alpha}$ has to be determined as a variable in the VECM.
+\end{itemize}
+It is worth noticing that the factorization of the matrix
+$\boldsymbol\Omega$ is not unique, since for any $r \times r$
+non-singular matrix $\mathbf{H}$, $\boldsymbol{\alpha}^*:=\boldsymbol{\alpha}\mathbf{H}$,
+and $\boldsymbol{\beta}^*=\boldsymbol{\beta}(\mathbf{H}^{-1})^\top$ we have
+$\boldsymbol{\alpha\beta}^\top=\boldsymbol{\alpha}^*(\boldsymbol{\beta}^*)^\top$.
+If cointegration exists, then equation (\ref{eq:vec}) can be written
+as follows:
+\begin{equation}\label{eq:vecfull}
+\Delta\mathbf{y}_t 
+= \boldsymbol{\alpha\beta}^\top\mathbf{y}_{t-1} 
+  + \sum_{i=1}^{p-1}\boldsymbol{\Phi}_i^*\,\Delta\mathbf{y}_{t-i}
+  + \mathbf{c} + \boldsymbol{\epsilon}_t\,,
+\end{equation}
+\noindent
+which is a VAR model but for time series differences.
 
-\noindent where $t=p+1, \dots N$,  ${\phi_1,\dots,\phi_p}$ are $l \times l$
-matrices of real coefficients,
-$\mathbf{\epsilon}_{p+1},\dots,\mathbf{\epsilon}_N$ are error terms,
-$\mathbf{c}$ is a constant vector and $N$ is the total number of samples.
-
-The VAR matrix form of equation~(\ref{eq:var}) is:
-
-%\begin{equation}\label{eq:vareq}
-%\mathbf{B} = \mathbf{A} \mathbf{X} + 
-%\mathbf{E} \, , 
-%\end{equation}
+Transposing each equation of the system (\ref{eq:vecfull}) we can write
+the VECM($p$) model in block-matrix form as:
+\begin{equation}\label{eq:vareq}
+\mathbf{B} = 
+\mathbf{A} \mathbf{X} + 
+\mathbf{E} \, , 
+\end{equation}
 %
-%\noindent where:
-%\begin{eqnarray}
-%\mathbf{A} &=&
-%\begin{bmatrix}
-%   \mathbf{y}_{p}     & \dots    & \mathbf{y}_{N-1}\\
-%   \mathbf{y}_{p-1}   & \dots    & \mathbf{y}_{N-2}\\
-%   \vdots             & \ddots   & \vdots\\
-%   \mathbf{y}_{1}     & \dots    & \mathbf{y}_{N-p}\\
-%   1                  & \dots    & 1 
-%   \end{bmatrix}^\top \, ,\\
-%\mathbf{B} &=&
-%\begin{bmatrix}
-%\quad \\
-%\mathbf{y}_{p+1} &
-%\dots & 
-%\mathbf{y}_N \\
-%\quad
-%\end{bmatrix}^\top  \, ,\\
-%\mathbf{X}&=&
-%\left[
-%\begin{array}{ccccc}
-%    \quad  & \quad & \quad & \quad \\
-%    \phi_1   & \cdots & \phi_p & \mathbf{c} \\  
-%    \quad & \quad & \quad & \quad
-%    \end{array}
-%\right]^\top \, ,\\
-%\mathbf{E} &=& 
-%\begin{bmatrix}
-%    \quad \\
-%    \mathbf{\epsilon}_{p+1}  & 
-%    \dots                & 
-%    \mathbf{\epsilon}_N \\
-%    \quad
-%\end{bmatrix}^\top \, . 
-%\end{eqnarray}
-
-\begin{equation}
- \label{eq:vareq}
-               \underbrace{ \begin{bmatrix}
-               \quad \\
-               \mathbf{y}_{p+1} &
-               \mathbf{y}_{p+2} &
-               \dots & 
-               \mathbf{y}_N \\
-               \quad
-               \end{bmatrix}}_{\substack{ \mathbf{B}^\top\\l \times (N-p)}}   
-= 
-                \underbrace{\left[ 
-                \begin{array}{ccccc}
-                \quad & \quad & \quad & \quad & \quad \\
-                \phi_1  & \phi_2 & \cdots & \phi_p & \mathbf{c} \\  
-                \quad &\quad & \quad & \quad & \quad
-               \end{array} 
-               \right]}_{\substack{ \mathbf{X}^\top\\ l \times (l \times p + 1 )}}
-\underbrace{\begin{bmatrix}
-   \mathbf{y}_{p}  & \mathbf{y}_{p+1} & \dots    & \mathbf{y}_{N-1}\\
-   \mathbf{y}_{p-1}  & \mathbf{y}_{p} & \dots    & \mathbf{y}_{N-2}\\
-   \vdots        & \vdots   & \ddots   & \vdots\\
-   \mathbf{y}_{1} & \mathbf{y}_{2}   & \dots    & \mathbf{y}_{N-p}\\
-   1 & 1   & \dots    & 1 
-   \end{bmatrix}}_{\substack{ \mathbf{A}^\top\\ (l\times p +1 )\times (N-p)}}
-+
-\underbrace{\begin{bmatrix}
-                \quad \\
-              \mathbf{\epsilon}_{p+1}  & 
-              \mathbf{\epsilon}_{p+2}  & 
-              \dots                & 
-              \mathbf{\epsilon}_N \\
-              \quad
-             \end{bmatrix}}_{\substack{\mathbf{E}^\top\\l \times (N-p) }} 
-\end{equation}
-Equation~(\ref{eq:vareq})  can be solved using ordinary least squares estimation.
-
-\subsection{Vector Error Correction}
-
-VECM is a special form of a VAR model for I(1) variables that are also
-cointegrated~\cite{banerjee1993}. It is obtained by replacing the form $\Delta
-\mathbf{y}_t = \mathbf{y}_t - \mathbf{y}_{t-1}$ in equation (\ref{eq:var}).
-VECM is expressed in terms of differences, has an error correction term and 
-the following form:
-
-\begin{equation}
- \label{eq:vec}
- \Delta \mathbf{y}_t = 
- \underbrace{ \Omega\mathbf{y}_{t-1}}_\text{Error correction term} + 
- \sum_{i=1}^{p-1}
-\phi_i^* \Delta \mathbf{y}_{t-i}  + \mathbf{c} + \mathbf{\epsilon}_t \quad ,
-\end{equation}
-
-\noindent where coefficients matrices $\Omega$ and $\phi_i^*$ are
-function of matrices $\phi_i$ (shown in equation (\ref{eq:var})) as follows:
-
-\begin{eqnarray*}
-\phi_i^* &: =& -\sum_{j=i+1}^{p} \phi_j \\
-\Omega &: =& -(\mathbb{I}-\phi_1-\dots-\phi_p) 
-\end{eqnarray*}
-
-The matrix $\Omega$ has the following properties~\cite{johansen1995}:
-\begin{itemize}
-\item If $\Omega = 0$ there is no cointegration
-\item If $rank(\Omega)=l$ i.e full rank, then the time series are not
-I(1) but stationary
-\item If $rank(\Omega)=r,\quad 0 < r < l$ then, there is cointegration
-and the matrix $\Omega$ can be expressed as $\Omega =
-\alpha \beta^\top$, where $\alpha$ and $\beta$ are $(l \times r)$
-matrices and $rank(\alpha)=rank(\beta)=r$.
-
-The columns of $\beta$ contains the cointegration vectors and the rows of
-$\alpha$ correspond with the adjusted vectors. $\beta$ is obtained by Johansen
-procedure~\cite{johansen1988} whereas $\alpha$ has to be determined as a
-variable in the VECM.
-
-It is worth noticing that the factorization of the matrix $\Omega$ is not
-unique since for any $r \times r$ nonsingular matrix $H$ we have:
-
-\begin{eqnarray*}
-\alpha \beta^\top &=& \alpha \mathbf{HH^{-1}} \beta^\top\\
-&=&(\alpha\mathbf{H})(\beta(\mathbf{H}^{-1})^\top)^\top \\
-&=& \alpha^*(\beta^*)^\top
-\end{eqnarray*}
-
-\noindent whith $\alpha^* = \alpha\mathbf{H}$ and $\beta^* =
-\beta(\mathbf{H}^{-1})^\top$.
-
-\end{itemize}
-
-If cointegration exists, then equation (\ref{eq:vec}) can be written as follows:
-
-\begin{equation}
- \label{eq:vecfull}
- \Delta \mathbf{y}_t = \alpha \beta^\top\mathbf{y}_{t-1} 
- + \sum_{i=1}^{p-1} \phi_i^*\Delta
-\mathbf{y}_{t-i}  + \mathbf{c} + \mathbf{\epsilon}_t \quad ,
-\end{equation}
-
-\noindent which is a VAR model but for time series differences.
-
-VECM matricial form is the following:
-
-%\begin{eqnarray}\label{eq:vecmatrix}
-% \mathbf{A}&=&
-%   \begin{bmatrix} 
-%   \beta^\top \mathbf{y}_{p} & 
-%   \cdots & \beta^\top \mathbf{y}_{N-1} \\
-%   \Delta \mathbf{y}_p  & \cdots 
-%   &\Delta\mathbf{y}_{N-1} \\ 
-%   \vdots  & \ddots & \vdots \\
-%   \Delta\mathbf{y}_2  & \cdots 
-%   & \Delta \mathbf{y}_{N-p+1} \\
-%   1 & \cdots & 1 
-%   \end{bmatrix}^\top \label{eq:vecA} \, ,\\
-%\mathbf{B} & = &
-% \begin{bmatrix}
-% \quad\\
-%  \Delta \mathbf{y}_{p+1} & 
-%  \dots &
-%  \Delta \mathbf{y}_N \label{eq:vecB}\\
-%  \quad
-% \end{bmatrix}^\top \, ,\\
-%\mathbf{X}&=&
-%  \begin{bmatrix}
-%   \quad \\
-%   \alpha & \phi_1^* & \cdots & \phi_{p-1}^* & \mathbf{c} \\  
-%   \quad
-%   \end{bmatrix}^\top \, ,\\
-%\mathbf{E} &=&
-%\begin{bmatrix}
-%   \quad \\
-%   \mathbf{\epsilon}_{p+1} &
-%   \dots &
-%   \quad &\mathbf{\epsilon}_N \, \\ \quad
-%\end{bmatrix}^\top 
-%\end{eqnarray}
+\noindent where $\mathbf{B}$ dimension is $((N-p)\times l)$, $\mathbf{A}$
+dimension is $((N-p)\times(r+(p-1)l +1))$, $\mathbf{X}$ dimension is $((r+(p-1)l
++1)\times l)$ and $\mathbf{E}$ dimension is $((N-p)\times l)$:
 %
-
-\begin{equation} \label{eq:vecmatrix}
-\underbrace{
-               \begin{bmatrix}
-               \quad\\
-                \mathbf{\Delta y}_{p+1} & 
-                \dots &
-                \mathbf{\Delta y}_N \\
-                \quad
+\begin{alignat}{3}
+\mathbf{B}
+&= \begin{bmatrix}
+   \Delta\mathbf{y}_{p+1}^\top \\
+   \Delta\mathbf{y}_{p+2}^\top \\
+   \vdots \\
+   \Delta\mathbf{y}_N^\top
    \end{bmatrix}
-               }_{\substack{\mathbf{B}^\top\\ l \times (N-p) }} =
-   \underbrace{
-    \begin{bmatrix}
-     \quad \\
-     \alpha & \phi_1^* & \cdots & \phi_{p-1}^* & \mathbf{c} \\  
-     \quad
+&\quad
+\mathbf{X}
+&= \begin{bmatrix}
+   \boldsymbol{\alpha}^\top \\
+   \boldsymbol{\Phi}_1^{*\top} \\
+   \boldsymbol{\Phi}_2^{*\top} \\
+   \vdots \\
+   \boldsymbol{\Phi}_{p-1}^{*\top} \\
+   \mathbf{c}^\top
    \end{bmatrix}
-     }_{\substack{ \mathbf{X}^\top\\ l \times (l\times p +1)}}
-\underbrace{\begin{bmatrix} 
-   \beta^\top \mathbf{y}_{p} & 
-   \cdots & \beta^\top \mathbf{y}_{N-1} \\
-   \mathbf{\Delta y}_p  & \cdots 
-   & \mathbf{\Delta y}_{N-1} \\ 
-   \vdots  & \ddots & \vdots \\
-   \mathbf{\Delta y}_2  & \cdots 
-   & \mathbf{\Delta y}_{N-p+1} \\
-   1 & \cdots & 1 
-   \end{bmatrix}}_{\substack{\mathbf{A}^\top \\ (l \times p +1) \times (N-p) }}
-+
-\underbrace{\begin{bmatrix}
-              \quad \\
-              \mathbf{\epsilon}_{p+1} &
-              \dots &
-              \quad &\mathbf{\epsilon}_N \\ \quad
-             \end{bmatrix}}_{\substack{\mathbf{E}^\top\\ l \times (N-p) }} 
-\end{equation}
-
-VAR and VECM parameters shown in equations (\ref{eq:vareq},\ref{eq:vecmatrix}) can be solved
-using standard regression techniques, such as ordinary least squares (OLS).
+&\quad
+\mathbf{E}
+&= \begin{bmatrix}
+   \boldsymbol{\epsilon}_{p+1}^\top \\
+   \boldsymbol{\epsilon}_{p+2}^\top \\
+   \vdots \\
+   \boldsymbol{\epsilon}_N^\top \\
+   \end{bmatrix}
+\end{alignat}
+\noindent and 
+\begin{align}
+\mathbf{A} 
+&= \begin{pmat}[{....|}]
+   \mathbf{y}_p^\top \boldsymbol{\beta} & \Delta \mathbf{y}_p^\top & \Delta\mathbf{y}_{p-1}^\top & \dots 
+                    & \Delta\mathbf{y}_2^\top & 1 \cr
+   \mathbf{y}_{p+1}^\top  \boldsymbol{\beta} &\Delta\mathbf{y}_{p+1}^\top & \Delta\mathbf{y}_p^\top & \dots
+                       & \Delta\mathbf{y}_3^\top & 1 \cr
+   \vdots & \vdots & \vdots & \ddots & \vdots & \vdots \cr
+   \mathbf{y}_{N-1}^\top  \boldsymbol{\beta} &\Delta\mathbf{y}_{N-1}^\top & \Delta\mathbf{y}_{N-2}^\top & \dots 
+                       & \Delta\mathbf{y}_{N-p-1}^\top & 1 \cr
+   \end{pmat}\, .
+\label{eq:Amatrix}
+\end{align}
+Taking into account the error term $\mathbf{E}$, equation~(\ref{eq:vareq}) 
+can be solved with respect to $\mathbf{X}$ using the ordinary least
+squares (OLS) estimation.
 
@@
 
-When $\mathbf{A}$ is singular, solution to equation~(\ref{eq:vecmatrix}) is given
+When $\mathbf{A}$ is singular, solution to equation~(\ref{eq:vareq}) is given
 by the ordinary least squares (OLS) method. OLS consists of minimizing the sum
@@
 
-\subsection{Ridge Regression}\label{sec:RR}
+\subsection{Regularization}\label{sec:RR}
 
-Ridge regression (RR) introduces a ridge estimation to avoid the
+Regularization can be understood using two contexts: learning and inverse problems.
+In the context of learning, regularization refers to techniques allowing to avoid over-fitting, e.g.
+ridge regression (RR) or Tikhonov regularization \cite{ tikhonov1977}
+
+introduces a ridge estimation to avoid the
 singularity of the matrix $\mathbf{A}^\top \mathbf{A}$
@@
 optimization problem:
+
+which was independently by Tikhonov \cite{tikhonov1963} and Phillips \cite{phillips1962}.
 
@@
 
-\subsection{Online Ridge regression}
-
-If we represent matrices $\mathbf{A}$ and $\mathbf{B}$ such as:
-
-\begin{equation}
-\label{eq:notation}
-	\mathbf{A} = 
-\left[
-  \begin{tabular}{c>{$}c<{$}c}
-    --- & \mathbf{a}^{\top}_{1} & ---\\
-    --- & \mathbf{a}^{\top}_{2} & ---\\
-    & \vdots & \\
-    --- & \mathbf{a}^{\top}_{m} & ---
-  \end{tabular}
-\right]
-\quad \text{and} \quad
-\mathbf{B} =
-\left[
-  \begin{tabular}{c>{$}c<{$}c}
-    --- & \mathbf{b}_{1} & ---\\
-    --- & \mathbf{b}_{2} & ---\\
-    & \vdots & \\
-    --- & \mathbf{b}_{m} & ---
-  \end{tabular}
-\right] \, ,
-\
-\end{equation}
-
-\noindent equation~(\ref{eq:optsolRR}) can also be written as:
-
-\begin{eqnarray*}
-\label{eq:RReapand}
-\mathbf{\mathbf{X}}_{\text{ridge}}&=&(\mathbf{A}^\top \mathbf{A}+ \lambda
-\mathbb{I})^{-1}\mathbf{A}^\top \mathbf{B} \\
-&=& \displaystyle \big (\sum_{t=1}^m
-\mathbf{a}_t \mathbf{a}_t  ^\top + \lambda \mathbb{I}\big )^{-1}
-\sum_{t=1}^m \mathbf{a}_t \mathbf{b}_t \, .
-\end{eqnarray*}
-
-Lets define $\displaystyle\mathbf{S}= \sum_{t=1}^m \mathbf{a}_t
-\mathbf{a}_t  ^\top + \lambda \mathbb{I} $ and $\mathbf{W}=
-\displaystyle\sum_{t=1}^m \mathbf{a}_t \mathbf{b}_t$, so the
-algorithm~\ref{alg:RR} shows the iterative formulation:
-
-\begin{algorithm}[H]
-\begin{algorithmic}[1]
-\REQUIRE $\,$ \\
-$\{\mathbf{a}_1,\dots,\mathbf{a}_{m} \}$: $m$ input vectors \\
-$\{\mathbf{b}_1,\dots,\mathbf{b}_{m} \}$: $m$ target vectors \\
-$\lambda$: regularization parameter \\
-\ENSURE  $\,$ \\
-$\{f(\mathbf{a}_1),\dots,f(\mathbf{a}_{m}) \}$: model predictions \\
-\STATE Initialize $\mathbf{S}=\lambda \mathbb{I}$
-and $\mathbf{W}=0$
-\FOR { $t = 1$ to $m$ }
-	\STATE read new $\mathbf{a}_t$
-	\STATE $\mathbf{X}=\mathbf{S}^{-1}\mathbf{W}$
-	\STATE output prediction $f(\mathbf{a}_t) = \mathbf{X}^\top \mathbf{a}_t$
-   	\STATE $\mathbf{S} = \mathbf{S} + \mathbf{a}_t \mathbf{a}_t^\top$
-   	\STATE Read new $y_t$
-    	\STATE $\mathbf{W} = \mathbf{W} + \mathbf{a}_t \mathbf{b}_t$
-\ENDFOR
-\end{algorithmic}
-\caption{Online Ridge Regression}
-\label{alg:RR}
-\end{algorithm}
-
-\subsection{The Aggregating Algorithm for Regression}
-
-The AAR, proposed by~\cite{vovk2001}, is an application of the aggregating
-algorithm to the problem of regression. The idea is to introduce the new input
-vector $\mathbf{x}_{m+1}$ to solve the model parameters:
-
-\begin{equation}
-\label{eq:AARexpand}
-\mathbf{X}_{aar} = \displaystyle \big (\sum_{t=1}^{m+1}
-\mathbf{a}_t \mathbf{a}_t  ^\intercal + \gamma \mathbb{I}\big )^{-1}
-\sum_{t=1}^m \mathbf{a}_t \mathbf{b}_t \, .
-\end{equation}
-
-If we define $\displaystyle\mathbf{S}= \sum_{t=1}^{m+1} \mathbf{a}_t
-\mathbf{a}_t  ^\intercal + \gamma \mathbb{I} $ and $\mathbf{W}=
-\displaystyle\sum_{t=1}^m \mathbf{a}_t \mathbf{b}_t$, the
-algorithm~\ref{alg:AAR} is slightly different to the algorithm~\ref{alg:RR}, 
-which updated matrix $\mathbf{S}$ before making the prediction:
-
-\begin{algorithm}[ht]
-\begin{algorithmic}[1]
-\REQUIRE $\,$ \\
-$\{\mathbf{a}_1,\dots,\mathbf{a}_{m} \}$: $m$ input vectors \\
-$\{\mathbf{b}_1,\dots,\mathbf{b}_{m} \}$: $m$ target vectors \\
-$\lambda$: regularization parameter \\
-\ENSURE  $\,$ \\
-$\{f(\mathbf{a}_1),\dots,f(\mathbf{a}_{m}) \}$: model predictions \\
-\STATE Initialize $\mathbf{S}=\lambda \mathbb{I}$
-and $\mathbf{W}=0$
-\FOR { $t = 1$ to $m$ }
-	\STATE read new $\mathbf{a}_t$
-   	\STATE $\mathbf{S} = \mathbf{S} + \mathbf{a}_t \mathbf{a}_t^\intercal$
-	\STATE $\mathbf{X}=\mathbf{S}^{-1}\mathbf{W}$
-	\STATE output prediction $f(\mathbf{a}_t) = \mathbf{X}^\intercal \mathbf{a}_t$
-   	\STATE Read new $\mathbf{y}_t$
-    	\STATE $\mathbf{W} = \mathbf{W} + \mathbf{a}_t \mathbf{b}_t$
-\ENDFOR
-\end{algorithmic}
-\caption{{\em The aggregating algorithm for regression}}
-\label{alg:AAR}
-\end{algorithm}
-
-\subsection{Selection of Lambda}
+\subsection{Regularization parameter chose}
 
 The additional term $\lambda \|\mathbf{\mathbf{X}}\|_2^2$ in the optimization
-problem shown in equation~(\ref{eq:RRproblem}) has two effects on the solution:
-shrinks the coefficients towards zero and improves the conditioning of the
-problem.
-
-On the other hand, the effect of adding the term $\lambda \mathbb{I}$
-to the matrix $\mathbf{A}^\top \mathbf{A}$
-(equation~(\ref{eq:optsolRR})) improves its condition number since it
-increases its diagonal values when $\lambda > 0 $.  The matrix
-$\mathbf{A}^\top \mathbf{A}$ is symmetrical ($(\mathbf{A}^\top
-\mathbf{A})^\top = \mathbf{A}^\top \mathbf{A}$) and therefore
-diagonalizable.  If we know the eigenvalue decomposition of $\mathbf{A}
-= \mathbf{U\Sigma V^\top}$, then:
-
-\begin{eqnarray*}
-\mathbf{A}^\top \mathbf{A}+\lambda \mathbb{I}&=&\mathbf{V\Sigma^2
-V}^\top + \lambda \mathbf{V} \mathbf{V}^\top\\ &=&\mathbf{V}
-(\mathbf{\Sigma}^2+\lambda\mathbb{I}) \mathbf{V}^\top \, ,
-\end{eqnarray*}
-
-\noindent where
-
-\begin{equation*}
-\mathbf{\Sigma}^2+\lambda\mathbb{I}=
-\begin{bmatrix}
-\sigma^2_1 + \lambda & \, & \, \\
-\, & \sigma^2_2 +\lambda & \, \\
-\, & \, & \ddots & \, \\
-\, & \, & \, & \sigma^2_n +\lambda \, .
-\end{bmatrix}
-\end{equation*}
-
-where $\sigma_1 \geq \sigma_2 \geq \dots \geq \sigma_n$.
-
-Since the condition number of a matrix $\mathbf{A}$ is defined as:
-
-\begin{equation*}
-	\kappa = \|\mathbf{A}\| \|\mathbf{A}^{-1}\|
-\end{equation*}
-
-If matrix $\mathbf{A}$ is non-singular, its condition number can be
-expressed in terms of its singular values. The effect of adding the
-regularization term affects the condition number as follows:
-
-\begin{eqnarray*}
-\kappa_{ols} &=& \|\mathbf{A}\| \|\mathbf{A}^{-1}\|=\frac{\sigma_1}{\sigma_n} \\
-\kappa_{ridge} &=& \|\mathbf{A}^\intercal \mathbf{A} + \lambda \mathbb{I}\| 
-\|(\mathbf{A}^\intercal \mathbf{A} + \lambda \mathbb{I})^{-1}\|=\frac{\sigma_1+\lambda}{\sigma_n + \lambda} \,
-\end{eqnarray*}
-
-It is easy to see that the term $\lambda$ improves the condition number:
-
-\begin{equation*}
-        \frac{\sigma_1+\lambda}{\sigma_n + \lambda} <
-        \frac{\sigma_1}{\sigma_n} \,  \qquad \forall \quad \lambda > 0
-\end{equation*}
-
-However, $\lambda$ cannot be too large. Tipically $\lambda$ is small and its
-magnitude depends on the matrix $\mathbf{A}$.
-
-For rank deficient matrices we know that $\text{det}(\mathbf{A A^\top})=0$, adding the
-term $\lambda \mathbb{I}$ we have that $\text{det}(\mathbf{A A^\top}+\lambda \mathbb{I}) =
-p(\lambda)$ where $p(\lambda)$ is a polynomial of degree $n$ ($\mathbf{A}$ is
-$m \times n$). The zeros of $p(\lambda)$ are discretes, so it can be represented
-as:
-
-\[
-p(\lambda) =
-\lambda(\lambda-\lambda_1)^{n_1}(\lambda-\lambda_2)^{n_2}\dots(\lambda-\lambda_s)^{n_s}
-\]
-
-\noindent where $n_1 + n_2 + \dots + n_s = n$.
-
-This means that $\lambda$ must be small in order to ensure that $p(\lambda)$
-does not vanish.
-
-\subsubsection{The bias-variance tradeoff}
+problem shown in equation~(\ref{eq:RRproblem})
+improves the conditioning of the problem because is equivalent to add the term $\lambda \mathbb{I}$ to the matrix $\mathbf{A}^\top \mathbf{A}$
+(equation~(\ref{eq:optsolRR})) which effect is to improve its condition number.
 
@@
 reduce the expected prediction error by reducing variance considering a biased
-estimator, this is called the bias-variance trafeoff. The prediction error is
-obtained as:
-
-\begin{equation}
- \label{eq:prederror}
-E[(B-\hat{f}(\mathbf{X}))^2] = \sigma^2 + Bias(\hat{f}(\mathbf{X}))^2 + Var(\hat{f}(\mathbf{X}))
-\end{equation}
-
-\noindent where $\hat{f}(\mathbf{X})=\mathbf{AX}(\lambda)$ and
-$f(\mathbf{X})=\mathbf{AX}$.
-
-Ridge regression shows an
-increasing bias and a decreasing variance. Despite OLS has zero bias,
-its variance is greater than ridge for small values of $\lambda$.
-It can be shown that, in terms of prediction error, ridge (black line)
-is lower than OLS (dotted line)~\cite{hoerl1970}.
+estimator, this is called the bias-variance tradeoff. 
+For example, OLS has zero bias but its variance is greater than ridge regression for small values of $\lambda$.
+Additionally, in terms of prediction error, ridge regression is lower than OLS \cite{hoerl1970}.
 
@@
 Despite the bias-variance tradeoff provides a conceptual framework for
-determining a good model, is not directly useful. Some popular methods
+determining lambda is not directly useful.
+
+Some popular methods
 for determine model selection are:
@@
 On the other hand, $\lambda$ can also be used to reduce computational time as it
-is shown in equation~(\ref{eq:taylor}). Golub (~\cite{golub1965}) suggested that
-$\lambda$ had to be chosen so that:
-
-\[
-\frac{\lambda}{\lambda + \delta^2} < 0.1
-\]
-
-\noindent where $\delta$ is a lower bound of the smallest non-zero singular
-value $\sigma_k$. One approach is to choose the greatest lower bound for
-$\delta=\sigma_k$. Hence:
-
-\[
-\frac{\lambda}{\lambda + \sigma_k^2} < 0.1
-\]
-
-\noindent if we choose $\lambda=\beta \sigma_k^2$ then we get the following
-relation:
-
-\[
-\frac{\beta \sigma_k^2}{\beta \sigma_k^2 + \sigma_k^2} =
-\frac{\beta}{\beta+1} < 0.1 \Rightarrow \beta < \frac{1}{9}  
-\]
-
-Experiments done by ~\cite{coleman+sun2010} have shown that $\beta = 0.01$
-produces satisfactory results. This means that $\lambda = 0.01 \sigma_k^2$.
-
-However, get $\sigma_k$ imply obtaining first the SVD, which is
-computational expensive.
-
-Since the algorithm presented by ~\cite{coleman+sun2010} already compute the QR
-factorization of the matrix $\mathbf{A}$ they show a procedure for getting
-a $\lambda$ approximation using QR.
-
-\begin{enumerate}
-\item Compute QR factorization: $\mathbf{A} = \mathbf{Q_1R_1}$
-\item Let $\mathbf{W}$ denote the set of absolute values of the nonzero diagonal
-elements of $\mathbf{R_1}$. Let $w_{\text{min}}$ and $w_{\text{max}}$ denote the
-smallest and largest elements of $\mathbf{W}$ respectively. Both 
-$\lambda_1 = \hat{\beta} w_{\text{min}}^2$
-$\lambda_2 = \hat{\beta} \frac{w_{\text{min}}^2}{w_{\text{max}}^2}$ where
-$\hat{\beta}=0.00025$ produce satisfactory results.
-\item The $\lambda_{\text{QR}}$ approximation is obtained as:
-$\lambda_{\text{QR}} = \frac{\lambda_1 + \lambda_2}{2}$  
-\end{enumerate}
-
-\subsection{Efficient computation}
-
-%Revisar!!!
-In regression problems we always require getting a matrix inverse which is
-computationally expensive. However, in stream data problems the is a way to
-obtain a matrix inverse approximation using the Sherman-Morrison-Woodbury. This
-formula allows to get the inverse of a
-the matrix $\mathbf{A+uv^\top}$ if we already calculated the
-inverse of $\mathbf{A}$ as follows:
-
-\begin{equation}
-\label{eq:SMW}
-(\mathbf{A+uv^\top})^{-1}=\mathbf{A}^{-1}-
-\frac{\mathbf{A^{-1}uv^\top A^{-1}}}{\mathbf{1+v^\top A^{-1}u}}
-\end{equation}
-
-%Revisar!!
-
-Another way was presented by Coleman and
-Sun~\cite{coleman+sun2010} using an iterative algorithm which
-uses $\mathbf{X}(\lambda)$ to approximate $\mathbf{X}(0)$.
-
-Using the compact SVD (shown in equation~(\ref{eq:compactsvd}))
-$\mathbf{X}(\lambda)$ is expressed as follows:
-
-\begin{eqnarray}
-\label{eq:optsolRRsvd}
-\mathbf{X}(\lambda) & = & (\mathbf{A}^\top \mathbf{A}+ \lambda
-\mathbb{I})^{-1}\mathbf{A}^\top \mathbf{B} \nonumber \\
-& = &\mathbf{V}_1(\Sigma_1^2+\lambda \mathbb{I})^{-1}\mathbf{\Sigma_1
-U_1^\top B}
-\end{eqnarray}
-
-\noindent where is easy to see that $\mathbf{X}(\lambda) \rightarrow
-\mathbf{\hat{X}}=\mathbf{V}_1 \mathbf{\Sigma}_1^{-1}\mathbf{U}_1^\top
-\mathbf{B}$ as $\lambda \rightarrow 0$.
-
-The method consists in obtaining $\mathbf{X}(\lambda)$ and then refine
-by adding more terms of its Taylor expansion to approximate
-$\mathbf{X}(0) = \mathbf{\hat{X}}$. The Taylor expansion about $\lambda_0$ is:
-
-\begin{equation}
-\label{eq:taylor}
-    \mathbf{X}(\lambda)=\mathbf{X}(\lambda_0) + \sum_{k=1}^\infty
-    \mathbf{s}_k(\lambda-\lambda_0)^{k}
-\end{equation}
-
-\noindent where $\mathbf{s}_k=\frac{1}{k!}\mathbf{X}(\lambda)^{(k)}$
-and $\mathbf{X}(\lambda)^{(k)}$ is obtained by taking differences 
-$\frac{\partial}{\partial \lambda}$ of:
-
-\begin{eqnarray*}
-(\mathbf{A}^\top \mathbf{A}+ \lambda\mathbb{I}) \mathbf{X}(\lambda) & = & \mathbf{A}^\top \mathbf{B}\\
-(\mathbf{A}^\top \mathbf{A}+ \lambda\mathbb{I}) \mathbf{X}(\lambda)^{(1)} + \mathbf{X}(\lambda)& = & 0 \\
-\mathbf{X}(\lambda)^{(1)}  &=& -(\mathbf{A}^\top \mathbf{A}+ \lambda\mathbb{I}) ^{-1} \mathbf{X}(\lambda) \\
-\mathbf{X}(\lambda)^{(2)}  &=& -2(\mathbf{A}^\top \mathbf{A}+ \lambda\mathbb{I}) ^{-1} \mathbf{X}(\lambda)^{(1)} \\
-& \vdots & \\
-\mathbf{X}(\lambda)^{(k)}  &=& -k(\mathbf{A}^\top \mathbf{A}+ \lambda\mathbb{I}) ^{-1} \mathbf{X}(\lambda)^{(k-1)} 
-\end{eqnarray*}
-
-\noindent for $\lambda=0$ we have:
-
-\begin{equation}
-\label{eq:taylor}
-    \mathbf{X}(0)=\mathbf{X}(\lambda_0) + \sum_{k=1}^\infty
-     (-1)^k \mathbf{s}_k \lambda_0^k
-\end{equation}
-
-\noindent therefore in order to ensure convergence, we can see that$\lambda_0$
-selection cannot be large.
-
-The algorithm for computing $\mathbf{X}(0)$ is the following:
-
-\begin{algorithm}[H]
-\begin{algorithmic}[1]
-\REQUIRE $\,$ \\
-$\mathbf{A}$: design matrix \\
-$\mathbf{B}$: response matrix \\
-$\lambda$: rank deficient parameter \\
-\ENSURE  $\,$ \\
-$\mathbf{X}$: parameters \\
-\STATE $\mathbf{M}=\mathbf{A^\top A}$ \\
-\STATE Initialize $\mathbf{Q R}=\mathbf{M}$ \\
-\STATE $\mathbf{X} = \mathbf{R^{-1}Q^\top B}$ \\
-\STATE $\mathbf{s} = \mathbf{X}$ \\
-\FOR { $i = 1,2,3,\dots$ }
-	\STATE $\mathbf{s} =
-        -(\mathbf{M}+\lambda\mathbb{I})^{-1}\mathbf{s}$\\
-	\STATE $\mathbf{X}=\mathbf{X} + (-1)^i {\color{red}\mathbf{s}
-        \lambda^i}$
-\ENDFOR
-\end{algorithmic}
-\caption{Algorithm for handling rank deficient matrices}
-\label{alg:coleman}
-\end{algorithm}
-
-The algorithm~\ref{alg:coleman} solves equation~(\ref{eq:taylor}). However, this
-version is unstable since tipically $\|\mathbf{s}\|$ is very large and
-$\lambda^i$ is very small ($\lambda$ is small).
-
-The following algorithm shows a more stable version of
-algorithm~\ref{alg:coleman}.
-
-\begin{algorithm}[H]
-\begin{algorithmic}[1]
-\REQUIRE $\,$ \\
-$\mathbf{A}$: design matrix \\
-$\mathbf{B}$: response matrix \\
-$\lambda$: rank deficient parameter \\
-\ENSURE  $\,$ \\
-$\mathbf{X}$: parameters \\
-\STATE $\mathbf{M}=\mathbf{A^\top A}$ \\
-\STATE Initialize $\mathbf{Q R}=\mathbf{M}$ \\
-\STATE $\mathbf{X} = \mathbf{R^{-1}Q^\top B}$ \\
-\STATE $\mathbf{t} = \mathbf{X}$ \\
-\FOR { $i = 1,2,3,\dots$ }
-        \STATE $\mathbf{t} =\lambda \mathbf{t}$  
-        \STATE $\mathbf{t} =  -(\mathbf{M}+\lambda\mathbb{I})^{-1}\mathbf{t}$
-	\STATE $\mathbf{X}=\mathbf{X} + \mathbf{t}$
-\ENDFOR
-\end{algorithmic}
-\caption{Algorithm for handling rank deficient matrices improved}
-\label{alg:colemanimproved}
-\end{algorithm}
-
-Both algorithms are equivalent, but algorithm~\ref{alg:colemanimproved} is more
-stable and converges in tipically less than 10 steps.
-It is important to notice that the QR factorization is computed only once and it
-is computationally less expensive than the SVD.
+is shown in equation~(\ref{eq:taylor}). Golub (~\cite{golub1965})
 
@@
 \end{description}
+
+\subsection{Online Ridge regression}
+
+If we represent matrices $\mathbf{A}$ and $\mathbf{B}$ such as:
+
+\begin{equation}
+\label{eq:notation}
+	\mathbf{A} = 
+\left[
+  \begin{tabular}{c>{$}c<{$}c}
+    --- & \mathbf{a}^{\top}_{1} & ---\\
+    --- & \mathbf{a}^{\top}_{2} & ---\\
+    & \vdots & \\
+    --- & \mathbf{a}^{\top}_{m} & ---
+  \end{tabular}
+\right]
+\quad \text{and} \quad
+\mathbf{B} =
+\left[
+  \begin{tabular}{c>{$}c<{$}c}
+    --- & \mathbf{b}_{1} & ---\\
+    --- & \mathbf{b}_{2} & ---\\
+    & \vdots & \\
+    --- & \mathbf{b}_{m} & ---
+  \end{tabular}
+\right] \, ,
+\
+\end{equation}
+
+\noindent equation~(\ref{eq:optsolRR}) can also be written as:
+
+\begin{eqnarray*}
+\label{eq:RReapand}
+\mathbf{\mathbf{X}}_{\text{ridge}}&=&(\mathbf{A}^\top \mathbf{A}+ \lambda
+\mathbb{I})^{-1}\mathbf{A}^\top \mathbf{B} \\
+&=& \displaystyle \big (\sum_{t=1}^m
+\mathbf{a}_t \mathbf{a}_t  ^\top + \lambda \mathbb{I}\big )^{-1}
+\sum_{t=1}^m \mathbf{a}_t \mathbf{b}_t \, .
+\end{eqnarray*}
+
+Lets define $\displaystyle\mathbf{S}= \sum_{t=1}^m \mathbf{a}_t
+\mathbf{a}_t  ^\top + \lambda \mathbb{I} $ and $\mathbf{W}=
+\displaystyle\sum_{t=1}^m \mathbf{a}_t \mathbf{b}_t$, so the
+algorithm~\ref{alg:RR} shows the iterative formulation:
+
+\begin{algorithm}[H]
+\begin{algorithmic}[1]
+\REQUIRE $\,$ \\
+$\{\mathbf{a}_1,\dots,\mathbf{a}_{m} \}$: $m$ input vectors \\
+$\{\mathbf{b}_1,\dots,\mathbf{b}_{m} \}$: $m$ target vectors \\
+$\lambda$: regularization parameter \\
+\ENSURE  $\,$ \\
+$\{f(\mathbf{a}_1),\dots,f(\mathbf{a}_{m}) \}$: model predictions \\
+\STATE Initialize $\mathbf{S}=\lambda \mathbb{I}$
+and $\mathbf{W}=0$
+\FOR { $t = 1$ to $m$ }
+	\STATE read new $\mathbf{a}_t$
+	\STATE $\mathbf{X}=\mathbf{S}^{-1}\mathbf{W}$
+	\STATE output prediction $f(\mathbf{a}_t) = \mathbf{X}^\top \mathbf{a}_t$
+   	\STATE $\mathbf{S} = \mathbf{S} + \mathbf{a}_t \mathbf{a}_t^\top$
+   	\STATE Read new $y_t$
+    	\STATE $\mathbf{W} = \mathbf{W} + \mathbf{a}_t \mathbf{b}_t$
+\ENDFOR
+\end{algorithmic}
+\caption{Online Ridge Regression}
+\label{alg:RR}
+\end{algorithm}
+
+\subsection{The Aggregating Algorithm for Regression}
+
+The AAR, proposed by~\cite{vovk2001}, is an application of the aggregating
+algorithm to the problem of regression. The idea is to introduce the new input
+vector $\mathbf{x}_{m+1}$ to solve the model parameters:
+
+\begin{equation}
+\label{eq:AARexpand}
+\mathbf{X}_{aar} = \displaystyle \big (\sum_{t=1}^{m+1}
+\mathbf{a}_t \mathbf{a}_t  ^\intercal + \gamma \mathbb{I}\big )^{-1}
+\sum_{t=1}^m \mathbf{a}_t \mathbf{b}_t \, .
+\end{equation}
+
+If we define $\displaystyle\mathbf{S}= \sum_{t=1}^{m+1} \mathbf{a}_t
+\mathbf{a}_t  ^\intercal + \gamma \mathbb{I} $ and $\mathbf{W}=
+\displaystyle\sum_{t=1}^m \mathbf{a}_t \mathbf{b}_t$, the
+algorithm~\ref{alg:AAR} is slightly different to the algorithm~\ref{alg:RR}, 
+which updated matrix $\mathbf{S}$ before making the prediction:
+
+\begin{algorithm}[ht]
+\begin{algorithmic}[1]
+\REQUIRE $\,$ \\
+$\{\mathbf{a}_1,\dots,\mathbf{a}_{m} \}$: $m$ input vectors \\
+$\{\mathbf{b}_1,\dots,\mathbf{b}_{m} \}$: $m$ target vectors \\
+$\lambda$: regularization parameter \\
+\ENSURE  $\,$ \\
+$\{f(\mathbf{a}_1),\dots,f(\mathbf{a}_{m}) \}$: model predictions \\
+\STATE Initialize $\mathbf{S}=\lambda \mathbb{I}$
+and $\mathbf{W}=0$
+\FOR { $t = 1$ to $m$ }
+	\STATE read new $\mathbf{a}_t$
+   	\STATE $\mathbf{S} = \mathbf{S} + \mathbf{a}_t \mathbf{a}_t^\intercal$
+	\STATE $\mathbf{X}=\mathbf{S}^{-1}\mathbf{W}$
+	\STATE output prediction $f(\mathbf{a}_t) = \mathbf{X}^\intercal \mathbf{a}_t$
+   	\STATE Read new $\mathbf{y}_t$
+    	\STATE $\mathbf{W} = \mathbf{W} + \mathbf{a}_t \mathbf{b}_t$
+\ENDFOR
+\end{algorithmic}
+\caption{{\em The aggregating algorithm for regression}}
+\label{alg:AAR}
+\end{algorithm}

### 2016-04-04 articles-conferences/elsevier-2016/src/background.tex
saving

@@
 
-Regularization can be understood using two contexts: learning and inverse problems.
-In the context of learning, regularization refers to techniques allowing to avoid over-fitting, e.g.
-ridge regression (RR) or Tikhonov regularization \cite{ tikhonov1977}
+Regularization can be understood using two contexts: learning theory (probabilistic) and inverse problems (deterministic).
+In the context of learning, regularization refers to techniques allowing to avoid over-fitting and the desire property of the selected estimator is to perform well on new data (to generalize), e.g.
+regularized least squares.
+In the context of inverse problems, regularization objective is to obtain stable solutions with respect to noise e.g Tikhonov regularization or ridge regression (RR)\cite{ tikhonov1977}.
 
-introduces a ridge estimation to avoid the
-singularity of the matrix $\mathbf{A}^\top \mathbf{A}$
-equation~(\ref{eq:pseudoinverse}). RR solves the following
-optimization problem:
+In particular, it is well known that regularization schemes such as RR or Tikhonov regularization can be effectively used in the context of learning \cite{vito2005}.
 
-which was independently by Tikhonov \cite{tikhonov1963} and Phillips \cite{phillips1962}.
+Ridge regression was independently proposed by Tikhonov \cite{tikhonov1963} and Phillips \cite{phillips1962}.
+RR penalise the complexity of the  solution by adding a constraint with respect to OLS (see equation \ref{eq:regressionproblem}) in the optimisation problem:
 
@@
 
-\noindent where $\lambda$ is a regularization parameter.
-
-The RR optimal solution $\mathbf{X}(\lambda)$ is also well-known:
+\noindent where $\lambda>0$ is a regularization parameter. The RR optimal solution $\mathbf{X}(\lambda)$ is also well-known:
 
@@
 
-\subsection{Regularization parameter chose}
+The method is called ridge regression because the term $\lambda \mathbb{I}$ adds positive entries along the diagonal (ridge) to avoid the
+singularity of the covariance matrix $\mathbf{A}^\top \mathbf{A}$. This addition ensures that all of the covariance matrix eigenvalues will be strictly greater than 0, i.e the solution becomes unique.
 
-The additional term $\lambda \|\mathbf{\mathbf{X}}\|_2^2$ in the optimization
-problem shown in equation~(\ref{eq:RRproblem})
-improves the conditioning of the problem because is equivalent to add the term $\lambda \mathbb{I}$ to the matrix $\mathbf{A}^\top \mathbf{A}$
-(equation~(\ref{eq:optsolRR})) which effect is to improve its condition number.
+Despite the fact that RR is a biased estimator (OLS is unbiased) it could
+reduce the expected prediction error by reducing variance, this is called the bias-variance tradeoff \cite{hoerl1970}.
 
-To determine parameter $\lambda$ is crucial for ridge regression since it could
-reduce the expected prediction error by reducing variance considering a biased
-estimator, this is called the bias-variance tradeoff. 
-For example, OLS has zero bias but its variance is greater than ridge regression for small values of $\lambda$.
-Additionally, in terms of prediction error, ridge regression is lower than OLS \cite{hoerl1970}.
+To determine parameter $\lambda$ is crucial for ridge regression, but the bias-variance tradeoff is not directly useful to determine it, since que don't know the data we want to predict.
 
-\subsubsection{Methods for lambda selection}
-
-Despite the bias-variance tradeoff provides a conceptual framework for
-determining lambda is not directly useful.
-
-Some popular methods
-for determine model selection are:
-
+The two most  popular regularization methods are :
+%%%% Agregar metodos del paper spectral cut-off y cross valid
 \begin{itemize}
@@
 \end{itemize}
-
-On the other hand, $\lambda$ can also be used to reduce computational time as it
-is shown in equation~(\ref{eq:taylor}). Golub (~\cite{golub1965})
 

### 2016-04-08 articles-conferences/elsevier-2016/src/background.tex
saving my work

@@
 regularized least squares.
-In the context of inverse problems, regularization objective is to obtain stable solutions with respect to noise e.g Tikhonov regularization or ridge regression (RR)\cite{ tikhonov1977}.
+In the context of inverse problems, regularization objective is to stabilise, with respect to noise, a possibly ill-conditioned matrix inversion problem e.g spectral cut-off and Tikhonov regularization or ridge regression (RR)\cite{ tikhonov1977}.
 
@@
 To determine parameter $\lambda$ is crucial for ridge regression, but the bias-variance tradeoff is not directly useful to determine it, since que don't know the data we want to predict.
+Many techniques have been developed to determine suitable value for the regularization parameter.
+However, when no information is available on the error, one of the most popular techniques is the Generalized Cross Validation (GCV) \cite{bauer2011}.
+GCV originates from the older method leave-one-out Cross Validation \cite{stone1974}.
 
-The two most  popular regularization methods are :
-%%%% Agregar metodos del paper spectral cut-off y cross valid
-\begin{itemize}
-    \item Akaike information criterion (AIC) (Akaike, 1974)
-    \item Bootstrap-based selection (Efron and Tibshirani, 1997)
-    \item Cross-validation (Stone, 1974)
-\end{itemize}
+The leave-one-out cross validation method determine a regularization parameter that minimises the average of the squared predictions errors using each solution to predict the missing data value.
 
+%'---------------------------http://robjhyndman.com/hyndsight/crossvalidation/
+When the data are not independent cross-validation becomes more difficult as leaving out an observation does not remove all the associated information due to the correlations with other observations.
+For time series forecasting, a cross-validation statistic is obtained as follows
+
+Fit the model to the data y1,…,yt and let y^t+1 denote the forecast of the next observation. Then compute the error (et+1∗=yt+1−y^t+1) for the forecast observation.
+Repeat step 1 for t=m,…,n−1 where m is the minimum number of observations needed for fitting the model.
+Compute the MSE from em+1∗,…,en∗.
+References
+
+An excellent and comprehensive recent survey of cross-validation results is Arlot and Celisse (2010)
+
+%---------------------------------------------------
 \subsection{Online machine learning algorithms}

### 2016-04-18 articles-conferences/iscef2016/ISCEF2016-slides/iscef2016-slides.tex
Inicio articulo 2016
version final ISCEF 2016

@@
 
-\begin{frame}
-\frametitle{Outline}
-\begin{itemize}
-\item Motivation: economic relations
-\item Cointegration and Integration concepts
-\item VECM model
-\item Proposal
-\item Experiments
-\item Conclusions
-\end{itemize}
-\end{frame}
+%\begin{frame}
+%\frametitle{Outline}
+%\begin{itemize}
+%\item Motivation: economic relations
+%\item Cointegration and Integration concepts
+%\item VECM model
+%\item Proposal
+%\item Experiments
+%\item Conclusions
+%\end{itemize}
+%\end{frame}
 
@@
 \frametitle{Motivation}
-Financial time series are mainly non-stationary.  However, one or more linear
+\begin{itemize}
+\item Financial time series are mainly non-stationary.  However, one or more linear
 combinations of non-stationary time series can be stationary even though
-individually they are not.
+individually they are not. When this happens it is said they have a long-run
+equilibrium relationship and the variables are \textbf{cointegrated}.
+\item The Vector Error correction model (VECM) is used to model cointegrated time series but is used only with batch data and rarely used with high frequency data.
+\end{itemize}
+\end{frame}
 
-When this happens it is said they have a long-run
-equilibrium relationship and the variables are \textbf{cointegrated}.
+\subsection{Proposal}
+\begin{frame}
+\frametitle{Proposal}
+\begin{block}{}
+To use VECM with high frequency data where the best parameters are updated when new data is available and do this is done in a fast way.
+\end{block}
 \end{frame}
@@
 \section{Background}
-\subsection{Definitions}
-
-\begin{frame}
-\frametitle{Integration }
-\begin{block}{Integration of order $d$}
-A time series $\mathbf{y}$ is said to be integrated of order $d$ if after
-differentiating the variable $d$ times, we get an stationary process, more precisely:
-\[
-(1-L)^d \mathbf{y} \sim \text{I(0)} \, ,
-\]
-\noindent where I(0) is a stationary time series and $L$ is the lag operator:
-\[
-(1-L)\mathbf{y} = \Delta \mathbf{y}=\mathbf{y}_t  -\mathbf{y}_{t-1} \quad \forall t
-\]
-\end{block}
-\end{frame}
-
-\subsection{Integration example}
-\begin{frame}
-\frametitle{Financial Time Series}
-\begin{columns}
-\column[t]{5cm}
-Financial time series are commonly a I(1) process 
-\begin{center}
-\includegraphics[width=0.85\textwidth]{img/EURUSD}
-\end{center}
-\column[t]{5cm}
-Since their differences are stationary (I(0) process).
-\begin{center}
-\includegraphics[width=0.85\textwidth]{img/DEURUSD}
-\end{center}
-\end{columns}
-\end{frame}
+%
+%\begin{frame}
+%\frametitle{Integration }
+%\begin{block}{Integration of order $d$}
+%A time series $\mathbf{y}$ is said to be integrated of order $d$ if after
+%differentiating the variable $d$ times, we get an stationary process, more precisely:
+%\[
+%(1-L)^d \mathbf{y} \sim \text{I(0)} \, ,
+%\]
+%\noindent where I(0) is a stationary time series and $L$ is the lag operator:
+%\[
+%(1-L)\mathbf{y} = \Delta \mathbf{y}=\mathbf{y}_t  -\mathbf{y}_{t-1} \quad \forall t
+%\]
+%\end{block}
+%\end{frame}
+%
+%\subsection{Integration example}
+%\begin{frame}
+%\frametitle{Financial Time Series}
+%\begin{columns}
+%\column[t]{5cm}
+%High frequency financial time series are commonly an I(1) process 
+%\begin{center}
+%\includegraphics[width=0.85\textwidth]{img/EURUSD}
+%\end{center}
+%\column[t]{5cm}
+%Since their differences are stationary (I(0) process).
+%\begin{center}
+%\includegraphics[width=0.85\textwidth]{img/DEURUSD}
+%\end{center}
+%\end{columns}
+%\end{frame}
 
@@
 \begin{frame}
-\frametitle{VECM difficulties}
+\frametitle{VECM difficulties in an online context}
 \begin{itemize}
@@
 \item To get updated cointegration vectors in time is computationally expensive because 
-the Johansen method and OLS are used to obtain them.
+the Johansen method and OLS are used to obtain them. It would be good to know how the number of cointegration vectors are affected by the number of data used $L$ and the number of lags of VECM $p$.
+\end{itemize}
+\end{frame}
+
+\begin{frame}
+\frametitle{Data}
+\begin{itemize}
+\item Tests were carried out using four foreign exchange rates all related to
+USD: EURUSD, GBPUSD, USDCHF and USDJPY. 
+\item This data was collected from Dukascopy, a free
+database which gives access to the Swiss Foreign Exchange marketplace.
+\item The tests were done using 10-seconds frequency from ask prices
+from the 11th to the 15th of August 2014. 
+%\item Since one day corresponds to 8640 data points and we used 5 days of data, we have 43,200 data points in total.
 \end{itemize}
@@
 
-\begin{frame}
-\frametitle{Proposal: Adaptive VECM algorithm (AVECM)}
-\begin{itemize}
-\item We proposed to choose $L$ and $p$ in order to maximise the percentage of
-cointegration $PC$ in the near past. This process was done every time that new data was processed. 
-\item To use a distributed environment to make this parameters search, since this is the most expensive routine.
-\end{itemize}
-\end{frame}
-
-\begin{frame}
-\frametitle{AVECM Algorithm}
-\begin{algorithmic}[1]
-\REQUIRE $\,$ \\
-$\mathbf{y}$: matrix with $N$ input vectors and $l$ time series\\
-$j$: Starting point of testing \\
-$it$: Ending point of testing \\
-$ps$: list of $p$ values \\
-$Ls$: list of $L$ values ($L<N$) \\
-$m$: Iterations to determine parameters ($m < N-L$)\\
-\ENSURE  $\,$ \\
-$\{ \hat{\mathbf{y}}[1],\dots,\hat{\mathbf{y}}[it]\}$: prediction vectors \\
-\FOR { $i =j$ to $it$ }
-   \STATE $\mathbf{Y} \gets \mathbf{y}[:,i-1]$
-    \STATE $L,p \gets
-    \texttt{get\_best\_params}(Ls,ps,m,\mathbf{Y})$
-        \STATE $model = \text{VECM}(\mathbf{Y},L, p)$
-        \STATE $\hat{\mathbf{y}}[i-j] = model.predict()$
-\ENDFOR
-\end{algorithmic}
-\end{frame}
+%\begin{frame}
+%\frametitle{AVECM Algorithm}
+%\begin{algorithmic}[1]
+%\REQUIRE $\,$ \\
+%$\mathbf{y}$: matrix with $N$ input vectors and $l$ time series\\
+%$j$: Starting point of testing \\
+%$it$: Ending point of testing \\
+%$ps$: list of $p$ values \\
+%$Ls$: list of $L$ values ($L<N$) \\
+%$m$: Iterations to determine parameters ($m < N-L$)\\
+%\ENSURE  $\,$ \\
+%$\{ \hat{\mathbf{y}}[1],\dots,\hat{\mathbf{y}}[it]\}$: prediction vectors \\
+%\FOR { $i =j$ to $it$ }
+%   \STATE $\mathbf{Y} \gets \mathbf{y}[:,i-1]$
+%    \STATE $L,p \gets
+%    \texttt{get\_best\_params}(Ls,ps,m,\mathbf{Y})$
+%        \STATE $model = \text{VECM}(\mathbf{Y},L, p)$
+%        \STATE $\hat{\mathbf{y}}[i-j] = model.predict()$
+%\ENDFOR
+%\end{algorithmic}
+%\end{frame}
 
@@
 \begin{frame}
-\frametitle{Evaluation methods}
-\begin{block}{MSE}
-Measures the distance between forecasts
-and the true values and large deviations from the true value have a
-large impact due to squaring forecast error.
-{\color{blue}
-\begin{equation}\label{eq:MSE}
-\text{MSE} = 
-\frac{\displaystyle \sum_{t=1}^{N} (\mathbf{y}_t-\hat{\mathbf{y}}_t)^2}{N}
-\end{equation}}
-\end{block}
-\begin{block}{\bf $U$-statistic}
-Unit free measure obtained as the ratio between the root MSE (RMSE) of a model and the RMSE of the naive random walk model. 
-\end{block}
+\frametitle{Results}
+\begin{tabular}{ccccccc}
+
+& \multicolumn{3}{l}{MSE} & &
+\multicolumn{2}{c}{$U$-statistic} \\ 
+\hhline{~---~--} \\
+ &
+AVECM & ARIMA & p-value & &
+AVECM & ARIMA \\
+
+\hline
+ EURUSD & 1.07 e-09 & 1.14 e-09 &  9.25 e-12 & & 0.6863 & 0.7108\\
+ GBPUSD & 1.66 e-09 & 1.74 e-09 &  6.95 e-02 & & 0.6866 & 0.7025\\
+ USDCHF & 5.85 e-10 & 6.35 e-10 &  2.89 e-14 & & 0.6803 & 0.7091\\
+ USDJPY  & 6.34 e-06 & 6.51 e-06 &  6.85 e-05 & & 0.6964 & 0.7057\\
+ \end{tabular}
 \end{frame}
+
+%\begin{frame}
+%\frametitle{Evaluation methods}
+%\begin{block}{MSE}
+%Measures the distance between forecasts
+%and the true values and large deviations from the true value have a
+%large impact due to squaring forecast error.
+%{\color{blue}
+%\begin{equation}\label{eq:MSE}
+%\text{MSE} = 
+%\frac{\displaystyle \sum_{t=1}^{N} (\mathbf{y}_t-\hat{\mathbf{y}}_t)^2}{N}
+%\end{equation}}
+%\end{block}
+%\begin{block}{\bf $U$-statistic}
+%Unit free measure obtained as the ratio between the root MSE (RMSE) of a model and the RMSE of the naive random walk model. 
+%\end{block}
+%\end{frame}
 
@@
 \subsection{Uno}
-\begin{frame}
-\frametitle{Data}
-AVECM tests were carried out using four foreign exchange rates all related to
-USD: EURUSD, GBPUSD, USDCHF and USDJPY. This data was collected from Dukascopy, a free
-database which gives access to the Swiss Foreign Exchange marketplace.
-The tests were done using 10-seconds frequency from ask prices
-from the 11th to the 15th of August 2014. Since one day corresponds to 8640 data points and we used 5 days of data, we have 43,200 data points in total.
-\end{frame}
 
-\begin{frame}
-\frametitle{Unit root tests}
-VECM can only be used if the time series are I(1). Therefore, 
-before running the tests, we firstly checked whether the time series were
-I(1) using the Augmented Dickey Fuller (ADF) test with lags $p=1,2,3,4,5$.
-\end{frame}
-
-\begin{frame}
-\frametitle{Performance accuracy}
-\small
-\begin{tabular}{ccccccc}
-& \multicolumn{3}{l}{MSE} & &
-\multicolumn{2}{c}{$U$-statistic} \\ 
-\hhline{~---~--} \\
- & AVECM & ARIMA & p-value & &
-AVECM & ARIMA \\
-
-\hline
- EURUSD & 1.0702 e-09 & 1.1481 e-09 &  9.250 e-12 & & 0.6863 & 0.7108\\
- GBPUSD & 1.6630 e-09 & 1.7408 e-09 &  6.951 e-02 & & 0.6866 & 0.7025\\
- USDCHF & 5.8503 e-10 & 6.3545 e-10 &  2.899 e-14 & & 0.6803 & 0.7091\\
- USDJPY  & 6.3483 e-06 & 6.5194 e-06 &  6.853 e-05 & & 0.6964 & 0.7057
- \end{tabular}
- \end{frame}
+%\begin{frame}
+%\frametitle{Unit root tests}
+%VECM can only be used if the time series are I(1). Therefore, 
+%before running the tests, we firstly checked whether the time series were
+%I(1) using the Augmented Dickey Fuller (ADF) test with lags $p=1,2,3,4,5$.
+%\end{frame}
+%
+%\begin{frame}
+%\frametitle{Performance accuracy}
+%\small
+%\begin{tabular}{ccccccc}
+%& \multicolumn{3}{l}{MSE} & &
+%\multicolumn{2}{c}{$U$-statistic} \\ 
+%\hhline{~---~--} \\
+% & AVECM & ARIMA & p-value & &
+%AVECM & ARIMA \\ 
+%
+%\hline
+% EURUSD & 1.0702 e-09 & 1.1481 e-09 &  9.250 e-12 & & 0.6863 & 0.7108\\
+% GBPUSD & 1.6630 e-09 & 1.7408 e-09 &  6.951 e-02 & & 0.6866 & 0.7025\\
+% USDCHF & 5.8503 e-10 & 6.3545 e-10 &  2.899 e-14 & & 0.6803 & 0.7091\\
+% USDJPY  & 6.3483 e-06 & 6.5194 e-06 &  6.853 e-05 & & 0.6964 & 0.7057
+% \end{tabular}
+% \end{frame}
 %For AVECM we considered different number of iterations (parameter $m$ in algorithm \ref{alg:AVECM}): 10, 50 and 100.
@@
 \frametitle{Parallel implementation}
-The function get\_best\_params($Ls,ps,m,\mathbf{Y}$) was implemented using MPI in Python. MPI allows large-scale parallel applications
+The function that get best parameters $L$ and $p$ was implemented using MPI in Python. MPI allows large-scale parallel applications
 with wide portability to be built, being able to run in large clusters
@@
 \begin{itemize}
-\item Parameter $it=100$
+%\item Parameter $it=100$
 \item The $L$ parameter was always chosen between 100 and 4000 and $p$ always took values
@@
 \item Cointegration relations change with time. We empirically showed that the Johansen method is sensitive to the number of lags but also to the amount of data considered.
-\item We found that out-of-sample forecast performance MSE is related to the {\em percentage of cointegration\/} and in the last samples.  
+\item We found that out-of-sample forecast performance MSE is related to the {\em percentage of cointegration\/}.  
 \item VECM improves performance measures by finding

### 2016-04-19 articles-conferences/iscef2016/src/background.tex
avance martes
fixing definition of a VECM model (ISCEF comment)

@@
 
-VECM is a special case of the VAR \cite{sims1980} model and both describe the joint behaviour
-of a set of variables.
-
-The VAR($p$) model is a general framework describing the behaviour of a
-set of $l$ endogenous variables as a linear combination of their last $p$
+Vector error correction model (VECM) describe the joint behaviour
+of a set of variables and can be derived from the simple Vector Autoregressive model (VAR) \cite{sims1980} model.
+The VAR($p$) model is a framework describing the behaviour of a
+set of $l$ endogenous and stationary variables as a linear combination of their last $p$
 values, where $l,p\in\mathbb{N}$. 
@@
 squares (OLS) estimation.
-
-%\subsection{Ordinary Least Squares method}
-%
-%The solution $\widehat{\mathbf{A}}$ to
-%equation~(\ref{eq:vareq}) can be obtained by the ordinary least squares (OLS)
-%method. $\widehat{\mathbf{X}}$ is the solution of the quadratic optimization problem
-%\begin{equation*}
-%\widehat{\mathbf{X}} = \underset{\mathbf{X}}{\text{Arg\;min}}
-%\|\mathbf{A}\mathbf{X}-\mathbf{B}\|_2^2
-%\end{equation*}
-%\noindent for which the solution $\widehat{\mathbf{X}}$ is well-known:
-%\begin{equation*}
-%\label{eq:MP}
-%\widehat{\mathbf{X}}=\mathbf{A}^{\!\!+}\,\mathbf{B}
-%\end{equation*}
-%\noindent where $\mathbf{A}^{\!\!+}$ is the Moore-Penrose pseudo-inverse
-%which, when $\mathbf{A}$ is full rank, can be written as follows: 
-%\begin{equation}
-%\label{eq:pseudoinverse}
-%\mathbf{A}^{\!\!+}= (\mathbf{A}^{\!\!\top} \mathbf{A})^{-1}\mathbf{A}^{\!\!\top} \, .
-%\end{equation}
-%when $\mathbf{A}$ is not full rank, i.e
-%$rank(\mathbf{A})=k <  n \leq m$, $\mathbf{A}^\top \mathbf{A}$ is
-%always singular and equation~(\ref{eq:pseudoinverse}) cannot be used.
-%More generally, the pseudo-inverse is best computed using the compact
-%singular value decomposition (SVD) of $\mathbf{A}$ which is:
-%\begin{equation*}
-%    \label{eq:compactsvd}
-%    \mathbf{A}=
-%    \mathbf{U_1}
-%    \boldsymbol \Sigma_1
-%    \mathbf{V}_1^{\top} \, .
-%\end{equation*}
-%Pseudo-inverse can then be written as follows:
-%\begin{equation*}
-%\label{eq:pseudoinversesvd}
-%\mathbf{A}^{\!\!+} = \mathbf{V}_1 \boldsymbol \Sigma_1^{-1} \mathbf{U}_1^\top \, .
-%\end{equation*}

### 2016-04-19 src/3_TimeSeries.tex
avance capitulo 2. Non-stationary definition falta

@@
 \section{Characteristics of Financial Time Series}
-%\section{Stylized facts of asset returns}
+There are several characteristics and concepts of financial time series and here we will discuss the more important ones.
+\subsection{Stylized facts of asset returns}
 \label{sec:stylizedfacts}
@@
 
-Returns of a time series $y_t$ is defined as:
-\begin{equation}
-\label{eq:returndef}
-r_t = y_t - y_{t-1}  \, ,
-\end{equation}
-
 Some stylized facts which are common to a wide set of financial assets
@@
 \begin{description}
-\item[Dependence] autocorrelation function (ACF) in returns is largely insignificant. The ACF measures the linear predictability of the series at time $t$ using only the value at time $s$:
+\item[Dependence] autocorrelation function (ACF) in returns is largely insignificant.
+Returns of a time series $y_t$ is defined as $r_t = y_t - y_{t-1}$.
+The ACF measures the linear predictability of the time series $y_t$ at time $t$ using only the value at time $s$.
+ACF is defined as:
+\begin{equation*}
+\rho(s,t) = \frac{\gamma(s,t)}{\sqrt{\gamma(s,s)\gamma(t,t)}}
+\end{equation*}
+\noindent where $\gamma(s,t)$ is the auto-covariance function that measures the linear dependence between two points on the same series observed at different times.
+It is defined as the second moment product:
+\begin{equation*}
+\gamma_y(s,t) = E[(y_s-\mu_s)(y_t - \mu_t)]
+\end{equation*}
 
-\begin{equation}
-\rho(s,t) = \frac{\gamma(s,t)}{\sqrt{\gamma(s,s)\gamma(t,t)}}
-\end{equation}
-\noindent where $\gamma(s,t)$ is the autocovariance function measures the linear dependence between two points on the same series observed at different times, it is defined as the second moment product:
-\begin{equation}
-\gamma_y(s,t) = E[(y_s-\mu_s)(y_t - \mu_t)]
-\end{equation}
-However, the autocorrelation in the absolute and squared returns is always positive,
-significant and decays slowly. In addition, the autocorrelation in the absolute
-returns is generally higher than the autocorrelation in the corresponding
-squared returns. The returns autocorrelation function (ACF) of SPY is shown in figure
- \ref{fig:returnacf}. The figure shows how the correlations are significant even
- for very long lags, this imply that a long-memory process because of its
- long-range dependence. 
+The ACF in the absolute and squared returns is always positive,
+significant and decays slowly. In addition, the ACF in the absolute
+returns is generally higher than the ACF in the corresponding
+squared returns.
+
+Figure \ref{fig:returnacf} shows the ACF of the returns of SPY stock where the correlations are significant even
+ for very long lags, this implies a long-memory process.
+
  \begin{figure}[h]
@@
 effect, the January effect and Holiday effect. The most important calendar
-anomalies are the January effect and the weekend effect. Figure
-\ref{fig:mondayeffect} shows the Monday effect from 1928 to 2007. It shows that
-in average Monday has negative returns. However, this is changing in time,
-figure \ref{fig:mondayeffect2} shows returns only for year 2007 and its last
-three months and it shows that Mondays and Wednesdays has positive returns in
-average.
-
-\begin{figure}[h!]
-\centering
-\begin{subfigure}[b]{0.45\textwidth}
- \includegraphics[width=\textwidth]{img/average_daily_change_1928_2007}
- \caption{Average daily return: 1928-2007}
- \label{fig:mondayeffect}
-\end{subfigure}
-\begin{subfigure}[b]{0.45\textwidth}
- \includegraphics[width=\textwidth]{img/thumb-average_daily_change_2007}
- \caption{Average daily return: 2007}
- \label{fig:mondayeffect2}
-\end{subfigure}
-\end{figure}
+anomalies are the January effect and the weekend effect. 
 \end{description}
@@
 \subsection{Stationary }
-A strictly stationary times series $Y_t$ is one for which the probabilistic behaviour
-of every collection of values $\{Y_{t_1},Y_{t_2},\dots,Y_{t_L}\}$ is identical
-to that of the time shifted set, more precisely: \[ P\{Y_{t_1} \leq
-c_1,\dots,Y_{t_L} \leq c_L\} = P\{Y_{t_1+h} \leq c_1,\dots,Y_{t_L+h} \leq c_L\}
+A strictly stationary times series $y_t$ is one for which the probabilistic behaviour
+of every collection of values $\{y_{t_1},y_{t_2},\dots,y_{t_L}\}$ is identical
+to that of the time shifted set, more precisely: \[ P\{y_{t_1} \leq
+c_1,\dots,y_{t_L} \leq c_L\} = P\{y_{t_1+h} \leq c_1,\dots,y_{t_L+h} \leq c_L\}
 \quad \forall L \in \mathbb{N}, \forall h \in \mathbb{Z}\] \noindent where
@@
 A weakly stationary time series is a process which mean, variance and auto
-covariance do not change over time: \begin{eqnarray*} E(Y_t) &=& \mu  \quad
-\forall t \in \mathbb{N} \\ E(Y^2_t) &=& \sigma^2  \quad \forall t \in
+covariance do not change over time: \begin{eqnarray*} E(y_t) &=& \mu  \quad
+\forall t \in \mathbb{N} \\ E(y^2_t) &=& \sigma^2  \quad \forall t \in
 \mathbb{N} \\ \lambda(s,t)&=&\lambda(s+h,t+h) \quad \forall s,t \in \mathbb{N},
@@
 
-\noindent with $\lambda(s,t) = E[(Y_s-\mu)(Y_t - \mu)]$
+\noindent with $\lambda(s,t) = E[(y_s-\mu)(y_t - \mu)]$
 

### 2016-04-21 src/3_TimeSeries.tex
chapter 2 reviewed

@@
 
-\subsubsection{Deterministic trend}
-
-Deterministic trend or trend stationary processes have the following form:
+\begin{enumerate}[(a)]
+\item \textbf{Deterministic trend:} Deterministic trend or trend stationary processes have the following form:
 
 \[
-X_t = f(t) + \epsilon{t} \quad ,
+y_t = f(t) + \epsilon_{t} \quad ,
 \]
 
-\noindent where $t$ is the time trend and $\epsilon{t}$ represents a stationary error term (with mean 0 and variance $\sigma^2$) and $f(t)$ is a deterministic function of time:
+\noindent where $t$ is the time trend and $\epsilon_{t}$ represents a stationary error term (with mean 0 and variance $\sigma^2$) and $f(t)$ is a deterministic function of time:
 \begin{itemize}
@@
 \end{itemize}
-
-\subsubsection{Stochastic trend}
-Stochastic trend processes are also called unit root or difference stationarity processes and have the following form:
+\item \textbf{Stochastic trend:} Stochastic trend processes are also called unit root or difference stationarity processes and have the following form:
 
 \[
-X_t = \mu + X_{t-1} + \epsilon_t
+y_t = \mu + y_{t-1} + \epsilon_t
 \]
@@
 \[
-(1-L) X_t = \mu  + \epsilon_t
+(1-L) y_t = \mu  + \epsilon_t
 \]
@@
 Other less common forms of non-stationarity are structural break in mean and structural break in variance. 
+\end{enumerate}
 
-\newpage
 \subsection{Spurious regression} \label{sec:spurious}
-The use
-of non-stationary data can lead to spurious regressions. Spurious regression
-dates back to Yule in 1926 \cite{yule1926} If two stationary
+The use of non-stationary data can lead to spurious regressions. Spurious regression
+dates back to Yule in 1926 \cite{yule1926}. If two stationary
 variables are generated as independent random series and are trending over time,
@@
 \textbf{Example of spurious regression} \quad 
-An example is shown below. Two random walks series $x_1t$ and $x_2t$ with a small drift ($\alpha=0.09$) are generated and
+An example is shown below. Two random walks series $x_{1,t}$ and $x_{2,t}$ with a small drift ($\alpha=0.09$) are generated and
 then regressed one on the other:
  \begin{equation}
- x_i,t = \alpha + x_{i,t-1} + \epsilon_i,t  \qquad \text{where} \qquad \mathcal{N}(0,1), \quad i=1,2
+ x_{i,t} = \alpha + x_{i,t-1} + \epsilon_i,t  \qquad \text{where} \qquad \mathcal{N}(0,1), \quad i=1,2
  \end{equation}
@@
   \includegraphics[width=0.8\textwidth]{img/spurious1}
-  \caption{Two random walks time series $x_1,t$ and $x_2,t$}
+  \caption{Two random walks time series $x_{1,t}$ and $x_{2,t}$}
   \label{fig:spurious1}
@@
 
-%                            OLS Regression Results                            
-%==============================================================================
-%Dep. Variable:                      y   R-squared:                       0.883
-%Model:                            OLS   Adj. R-squared:                  0.882
-%Method:                 Least Squares   F-statistic:                     742.2
-%Date:                Fri, 06 Feb 2015   Prob (F-statistic):           1.61e-47
-%Time:                        06:07:47   Log-Likelihood:                -218.03
-%No. Observations:                 100   AIC:                             440.1
-%Df Residuals:                      98   BIC:                             445.3
-%Df Model:                           1                                         
-%==============================================================================
-%                 coef    std err          t      P>|t|      [95.0% Conf. Int.]
-%------------------------------------------------------------------------------
-%x1             0.8883      0.033     27.243      0.000         0.824     0.953
-%const          2.6726      0.405      6.606      0.000         1.870     3.475
-%==============================================================================
-%Omnibus:                        4.133   Durbin-Watson:                   0.363
-%Prob(Omnibus):                  0.127   Jarque-Bera (JB):                4.145
-%Skew:                          -0.469   Prob(JB):                        0.126
-%Kurtosis:                       2.662   Cond. No.                         23.3
-%==============================================================================
-$\blacksquare$
-
-\newpage
 \subsection{Integration}
@@
 $d$, and then we shall write $\mathbf{y}_t\sim I(d)$, whenever after $d$ times
-(discrete) differentiation an stationary process is
+(discrete) differentiation a stationary process is
 obtained~\cite{banerjee1993};
@@
 
-\subsection{Cointegration} \label{sec:coint}
+\subsection{Cointegration} \label{sec:cointegration}
 Cointegration concept was introduced by Engle in 1987 \cite{engle1987} and implies that one or
@@
 understand cointegration relationships. These common stochastic trends can be
-also interpreted as a long-run equilibrium relationships.
+also interpreted as a long-run equilibrium relationship.
 
@@
 
-There are two statistics for cointegration under Johansen approach: the trace (shown in equation \ref{eq:johansentrace}) and the maximum-eigenvalue statistic (shown in equation \ref{eq:johansenmax}).
+There are two statistics for cointegration under Johansen approach: the trace (shown in equation \ref{eq:johansentrace}) and the maximum-eigenvalue statistic (shown in equation \ref{eq:johansenmax}):
 
@@
 
-Where $r$ is the number of cointegration vectors and $\hat]{\lambda}_i$ is the estimated value for the ith ordered eigenvalue.
+\noindent where $r$ is the number of cointegration vectors and $\hat]{\lambda}_i$ is the estimated value for the ith ordered eigenvalue.
 $\lambda_{\text{trace}}$ null hypothesis is that the number of cointegration vectors is less than or equal to $r$ against that there are more than $r$.
@@
 
-If we have two-dimensional process $\mathbf{X}_t$, $t=1,\dots,T$ by:
+If we have two-dimensional process $\mathbf{y}_t$, $t=1,\dots,T$ by:
 
 \begin{eqnarray*}
-\mathbf{X}_{1t} &=& \sum_{i=1}^t \epsilon_{1i} + \epsilon_{2t} \\
-\mathbf{X}_{2t} &=& a \sum_{i=1}^t \epsilon_{1i} + \epsilon_{3t} 
+\mathbf{y}_{1t} &=& \sum_{i=1}^t \epsilon_{1i} + \epsilon_{2t} \\
+\mathbf{y}_{2t} &=& a \sum_{i=1}^t \epsilon_{1i} + \epsilon_{3t} 
 \end{eqnarray*}
 
-Since $\mathbf{X}_{1t}$ and $\mathbf{X}_{2t}$ are I(1) processes and there
+Since $\mathbf{y}_{1t}$ and $\mathbf{y}_{2t}$ are I(1) processes and there
 exist a vector $\beta = [a -1]$ such that:
@@
 \[
-\beta^\top \mathbf{X}_t = a \mathbf{X}_{1t} -\mathbf{X}_{2t} = 
+\beta^\top \mathbf{y}_t = a \mathbf{y}_{1t} -\mathbf{y}_{2t} = 
 a\epsilon_{2t} - \epsilon_{3t} \sim \text{I(0)}
@@
 If we add a I(0) process
-$\mathbf{X}_{3t} = \epsilon_{4t}$  we find that there exists two cointegration
+$\mathbf{y}_{3t} = \epsilon_{4t}$  we find that there exists two cointegration
 vectors now: $\begin{bmatrix}a &-1& 0\end{bmatrix}$ and $\begin{bmatrix}0
@@
 \[
-\beta^\top \mathbf{X}_t = 
+\beta^\top \mathbf{y}_t = 
 \begin{bmatrix}
@@
 \begin{bmatrix} 
-\mathbf{X}_{1t} \\
-\mathbf{X}_{2t} \\
-\mathbf{X}_{3t}
+\mathbf{y}_{1t} \\
+\mathbf{y}_{2t} \\
+\mathbf{y}_{3t}
 \end{bmatrix} = 
@@
   \includegraphics[width=\textwidth]{img/coint-example}
-  \caption{Cointegration example: two processes I(1)(blue and green) and one process I(0) (red)}
+  \caption{Cointegration example: two I(1) processes (blue and green) and one I(0) process (red).}
   \label{fig:cointex}
@@
 
-$\blacksquare$
-
 \section{Unit root tests}
@@
 In order to determine if a time series is an a I(1) or I(0) process unit root tests are used.
-Several tests have been developed but the most widely used is an extension of the Dickey and Fuller test \cite{dickey1979} called the Augmented Dickey-Fuller (ADF) test developed in 1984 by Said and Dickey in 1984\cite{said1984}.
+Several tests have been developed but the most widely used is an extension of the Dickey and Fuller test \cite{dickey1979} called the Augmented Dickey-Fuller (ADF) test developed by Said and Dickey in 1984\cite{said1984}.
 
@@
 trends, supply and demand. However, others such as monetary policies and news
-are not directly measurable and are included in other models out of the scope
-of this thesis.
+are not directly measurable.
 
@@
 
-The volatility of a stock is not directly observable~\cite{tsay2005,engle1993}.
+The volatility of a stock is not directly observable \cite{tsay2005,engle1993}.
 For example, daily volatility is not directly observable from only daily returns because there is only one observation in a trading day.
@@
 
-\subsubsection{realized volatility} is also known as historic volatility and
+\subsubsection{Realized volatility} is also known as historic volatility and
 it is the actual variance in the price of a stock over time.
@@
 inferior model.   
-\subsubsection{implied volatility volatility} not only can be extracted from returns
+\subsubsection{Implied volatility} not only can be extracted from returns
 but it can also be derived from option or future pricing models.  The
@@
 In our case, each one of these $l$ variables is a scalar time series
-$y_{\lambda,t}$, $\lambda=1,\dots,l$, and we represent them all together
+$y_{i,t}$, $i=1,\dots,l$, and we represent them all together
 at time $t$ by the the vector time series:
-\begin{equation}
+\begin{equation*}
 \label{eq:variables}
@@
 \begin{bmatrix} y_{1,t} & y_{2,t} & \dots & y_{l,t} \end{bmatrix}^\top.
-\end{equation}
+\end{equation*}
 \noindent
-Notice that the vectors $\mathbf{y}_t$ are assumed to be $l$-dimensional.
+Notice that the vector $\mathbf{y}_t$ is assumed to be $l$-dimensional.
 
@@
 
-Notice that, regarding our notation of section (\ref{sec:coint}),
+Notice that, regarding our notation of section (\ref{sec:cointegration}),
 we have here 
@@
 $\mathbf{y}_t^\nu = \mathbf{y}_{t-\nu}$ and
-the $\lambda$-th component of the vector time series $\mathbf{y}_t^\nu$
-is the scalar time series $y_{\lambda,t}^\nu$, where $\nu=1,\dots,p$ and
-$\lambda=1,\dots,l$.
+the $i$-th component of the vector time series $\mathbf{y}_t^\nu$
+is the scalar time series $y_{i,t}^\nu$, where $\nu=1,\dots,p$ and
+$i=1,\dots,l$.
 Transposing each equation of the system (\ref{eq:var}) we can write
 the VAR($p$) model in block-matrix form as:
-\begin{equation}\label{eq:vareq}
+\begin{equation}\label{eq:simplevareq}
 \mathbf{Y} = \mathbf{A} \mathbf{X} + \mathbf{E} \, , 
@@
 %%
-\begin{alignat}{2}
+\begin{alignat*}{2}
 \mathbf{Y}
@@
    \end{bmatrix}_{(N-p)\times l}
-\end{alignat}
-Taking into account the error term $\mathbf{E}$, equation~(\ref{eq:vareq}) 
+\end{alignat*}
+Taking into account the error term $\mathbf{E}$, equation \ref{eq:simplevareq}
 can be solved with respect to $\mathbf{X}$ using the ordinary least
-squares estimation.
+squares estimation (see section \ref{sec:OLS}).
 
 \section{Vector error correction model}
-Vector error correction model (VECM)  \cite{engle87} is a special form of a VAR model for I(1) variables that are also
-cointegrated~\cite{banerjee1993}.
+Vector error correction model (VECM)  \cite{engle87} describe the joint behaviour of a set of cointegrated I(1) time series \cite{banerjee1993} and can be derived from the simple VAR model.
 
-It is obtained re-writing equation (\ref{eq:var}) in terms of the new
+It is obtained re-writing equation \ref{eq:var} in terms of the new
 variable $\Delta\mathbf{y}_t=\mathbf{y}_t-\mathbf{y}_{t-1}$.
@@
 $\boldsymbol{\Omega}$, expressed in terms of the matrices
-$\boldsymbol{\Phi}_i$ of (\ref{eq:var}), are:
+$\boldsymbol{\Phi}_i$ of the model VAR shown in equation \ref{eq:var}, are:
 \begin{align*}
@@
 the VECM($p$) model in block-matrix form as:
-\begin{equation}\label{eq:vareq}
+\begin{equation}\label{eq:simpleveceq}
 \mathbf{Y} = 
@@
 \begin{align}
-\label{eq:vecX}
+\label{eq:vecE}
 \mathbf{E}
@@
 
-Taking into account the error term $\mathbf{E}$, equation~(\ref{eq:vareq}) 
+Taking into account the error term $\mathbf{E}$, equation~\ref{eq:simpleveceq}
 can be solved with respect to $\mathbf{X}$ using the ordinary least
@@
 
+\subsection{Ordinary Least Squares method} \label{sec:OLS}
+
+When $\mathbf{A}$ is singular, solution to equation~(\ref{eq:vareq}) is given
+by the ordinary least squares (OLS) method. OLS consists of minimizing the sum
+of squared errors or equivalently minimizing the following expression:
+
+\begin{equation}
+\label{eq:regressionproblem}
+\underset{\mathbf{X}}{\text{min}} \quad \| \mathbf{A}\mathbf{\mathbf{X}} - \mathbf{B} \|_2^2
+\end{equation}
+
+\noindent for which the solution $\hat{\mathbf{X}}$ is well-known:
+
+\begin{equation}
+\label{eq:MP}
+\hat{\mathbf{X}}=\mathbf{A}^{\!\!+}\,\mathbf{B}
+\end{equation}
+
+\noindent where $\mathbf{A}^{\!\!+}$ is the Moore-Penrose pseudo-inverse
+which can be written as follows:
+
+\begin{equation}
+\label{eq:pseudoinverse}
+\mathbf{A}^{\!\!+}= (\mathbf{A}^{\!\!\top} \mathbf{A})^{-1}\mathbf{A}^{\!\!\top} \, .
+\end{equation}
+
+However, when $\mathbf{A}$ is not full rank, i.e
+$rank(\mathbf{A})=k <  n \leq m$, $\mathbf{A}^\top \mathbf{A}$ is
+always singular and equation~(\ref{eq:pseudoinverse}) cannot be used.
+More generally, the pseudo-inverse is best computed using the compact
+singular value decomposition (SVD) of $\mathbf{A}$:
+
+\begin{equation}
+    \label{eq:compactsvd}
+    \underset{m \times n}{\mathbf{A}}=
+    \underset{m \times k}{\mathbf{U_1}} \enskip
+    \underset{k \times k}{\Sigma_1} \enskip
+    \underset{k \times n}{\mathbf{V}_1^{\top}} \, ,
+\end{equation}
+
+\noindent as follows
+
+\begin{equation}
+\label{eq:pseudoinversesvd}
+\mathbf{A}^{\!\!+} = \mathbf{V}_1 \Sigma_1^{-1} \mathbf{U}_1^\top \, .
+\end{equation}
+
+\newpage
 Conventional regression estimators, including VARs, have good properties when applied to covariance-stationary time series, but encounter difficulties when applied to non stationary or integrated processes.

### 2016-04-22 src/3_TimeSeries.tex
agregando def de RW and ARIMA

@@
 
-%\section{Vector Autoregressive Models}\label{sec:varvec}
+\section{Univariate Time Series modeling}
+
+The random walk model, despite its simplicity is still difficult to outperform for standard econometric forecasting models \cite{lo2011} despite its simplicity. The random walk model is defined as:
+
+\begin{equation}
+\mathbf{y}_t = \mathbf{y}_{t-1} + \epsilon_{t}
+\label{rwmodel}
+\end{equation}
+
+The naive forecast of the time series difference $\hat{\mathbf{y}}_{t+1}$ for the random walk model is defined as:
+\begin{equation}
+\hat{\mathbf{y}}_{t+1} = \mathbf{y}_t + \hat{\epsilon}_{t+1} 
+\end{equation}
+\noindent where  $\hat{\epsilon}_{t+1} = \epsilon_{t}$.
+
+On the other hand, ARIMA is widely used to forecast returns in finance \cite{tsay2005}.
+A process can be modelled as an ARIMA$(p,d,q)$ model if $\mathbf{x}_t=\Delta^d \mathbf{y}_t $, i.e after differencing $d$ times the time series $\mathbf{y}_t$,  we get an ARMA$(p,q)$.
+An ARMA$(p,q)$ model is the following:
+
+\begin{equation}
+\mathbf{x}_t = \sum_{i=1}^p \phi_i \mathbf{x}_{t-i}  +  \sum_{j=1}^q \theta_j \epsilon_{t-j}  
+\end{equation}
+\noindent with coefficients $\phi_p \neq 0$, $\theta_q \neq 0$ and $\sigma_{\epsilon}^2 > 0$.
+
 \section{Multivariate Time Series modelling}

### 2016-04-22 articles-conferences/iscef2016/src/methodology.tex
comentarios articulo Peretti

@@
 relationship \cite{gregoryETal1996}. 
-In order to show the the number of cointegration vectors which depends on 
+In order to show the number of cointegration vectors which depends on 
 the amount $L$ of historical data and the number of lags $p$ in the VECM, we used a grid search.
@@
 Since we are modelling returns, we used $d=1$.
-An ARMA$(p,q)$ model is the following:
-
-\begin{equation}
-\mathbf{x}_t = \sum_{i=1}^p \phi_i \mathbf{x}_{t-i}  +  \sum_{j=1}^q \theta_j \epsilon_{t-j}  
-\end{equation}
-\noindent with coefficients $\phi_p \neq 0$, $\theta_q \neq 0$ and $\sigma_{\epsilon}^2 > 0$.
 

### 2016-04-22 articles-conferences/iscef2016/src/results.tex
comentarios articulo Peretti

@@
 
-VECM can only be used if the time series are I(1). Therefore, 
-before running the tests, we firstly checked whether the time series were
+Before running the tests, we firstly checked whether the time series were
 I(1) using the Augmented Dickey Fuller (ADF) test with lags $p=1,2,3,4,5$. 

### 2016-04-25 articles-conferences/iscef2016/src/intro.tex
agregando imágenes nuevas

@@
 VECM parameters are obtained using the ordinary least squares
-(OLS) method~\cite{golub1980}. Since OLS involves many calculations, the parameter estimation is computationally expensive when the number of lagged values and
+(OLS) method \cite{golub1980}. Since OLS involves many calculations, the parameter estimation is computationally expensive when the number of lagged values and
 data increases. Moreover, obtaining cointegration vectors 
@@
 expensive because the Johansen
-method~\cite{johansen1995} is required to find cointegration vectors, which is of order $O(n^3)$. Therefore, our proposal, called Adaptively Vector Error Correction (AVECM), 
+method \cite{johansen1995} is required to find cointegration vectors, which is of order $O(n^3)$. Therefore, our proposal, called Adaptively Vector Error Correction (AVECM), 
 is to parallelise this search in order to get new parameters before new data arrives.
@@
 
-This paper is organised as follows: section~\ref{sec:background} presents the
-VAR and VECM, the AVECM algorithm is presented in section~\ref{sec:methodology}.
-In section~\ref{sec:results} we describe the tests carried on to assess the
+This paper is organised as follows: section \ref{sec:background} presents the
+VAR and VECM, the AVECM algorithm is presented in section \ref{sec:methodology}.
+In section \ref{sec:results} we describe the tests carried on to assess the
 accuracy and the execution time of AVECM.  This section also includes a
-description of the test data.  Section~\ref{sec:conclusions} contains the
+description of the test data.  Section \ref{sec:conclusions} contains the
 conclusions and a discussion of future research.

### 2016-04-25 articles-conferences/iscef2016/src/methodology.tex
agregando imágenes nuevas

@@
 
-The same experiments, this time using $p=3$, were carried out in order to see the
-effect on the number of cointegration vectors found. 
-Figure~\ref{fig:histsp3} shows that $p$ affects the distribution of the number of cointegration vectors by small amounts.
-
-\begin{figure}[!h]
-  %\vspace{-0.8cm}
-  \centering
-  \includegraphics[width=0.9\textwidth]{img/histCointVectors-offset20520-p-3-freq-10s}
-  \caption{Histogram of the number of cointegration vectors using $p=3$. The distribution is very similar to the one with $p=1$}
-   \label{fig:histsp3}
-\end{figure}
-
 From section \ref{sec:background} we know that $r=0$ means no cointegration and $r=l$ (we are using four rates, so $l=4$) reveals that no process is I(1) but stationary.
@@
   \centering
-  \includegraphics[width=0.8\textwidth]{img/MSE-offset20520-p-2-freq-10s}
+  \includegraphics[width=0.8\textwidth]{img/MSE-offset20520-p-1-freq-10s}
   \caption{MSE versus the percentage of cointegration considering 1000

### 2016-07-06 articles-conferences/iscef2016/manuscript-computational-economics/body.tex
Final manuscript

\section{Introduction}
\label{sec:introduction}
In finance, it is common to find variables with long-run equilibrium
relationships. This is termed cointegration and reflects the idea that some sets
of variables cannot wander too far from each other. Cointegration means that one
or more linear combinations of these variables are stationary even though
individually they are not. Some models, such as the Vector Error Correction
(VECM), see \cite{engle87}, take advantage of this property and describe the
joint behaviour of several cointegrated variables. VECM introduces this long-run
relationship among a set of cointegrated time series as an error correction
term. These time series must be integrated of order 1, denoted I(1), i.e. they
become stationary at their first differences. In finance, I(1) time series are
very common and to introduce cointegration restrictions in models often improves
forecasting, see \cite{duy1998}. Therefore, VECM has been widely adopted in
financial applications, among others: \cite{mukherjee1995}, \cite{seong2013},
\cite{maysami2000} and \cite{arestis2001}. VECM has also been used in pair
trading, see \cite{herlemont2003}, or models with more than two variables, see
for example \cite{mukherjee1995} and \cite{engle2004}).

VECM parameters are obtained using the ordinary least squares (OLS) method,
developed by \cite{golub1980}. Since OLS involves many calculations, the
parameter estimation is computationally expensive when the number of lagged
values and data increases. Moreover, obtaining cointegration vectors is also an
expensive routine. This is a main limitation to use VECM with stream data with
high frequency. \cite{chen2003} addresses the advantage of distributed
processing over conventional rolling window processing.

Our aims were to study if a parallel version of VECM can be used with high
frequency stream data.  Our approach was to determine adaptively the number of
observations and lags of VECM which maximise cointegration relations in the
past.  This search is computationally expensive because the Johansen method, see
\cite{johansen1995}, is required to find cointegration vectors, which is of
order $O(n^3)$. Therefore, our proposal, called Adaptively Vector Error
Correction (AVECM), is to parallelise this search in order to get new parameters
before new data arrives.  Model effectiveness is focused on out-of-sample
forecast rather than in-sample fitting. This criterion allows AVECM prediction
capability to be expressed rather than just explaining data history.  The
forecast capability of our method was measured using MSE and the Theil's
$U$-statistic, see \cite{theil1966}, widely used in economic forecast. Tests
were run using four currency rates: Euro (EUR) to United States Dollar (USD)
(EURUSD), British Pound (GBP) to USD (GBPUSD), USD to Swiss Franc (CHF) (USDCHF)
and USD to Japanese Yen (JPY) (USDJPY) with a 10-second frequency.

This paper is organised as follows: section \ref{sec:background} presents the
VAR and VECM, the AVECM algorithm is presented in section \ref{sec:methodology}.
In section \ref{sec:results} we describe the tests carried on to assess the
accuracy and the execution time of AVECM.  This section also includes a
description of the test data.  Section \ref{sec:conclusions} contains the
conclusions and a discussion of future research.
\def\rot#1{{\color{red}{#1}}}
\def\gruen#1{{\color{green}{#1}}}
\def\blau#1{{\color{blue}{#1}}}

\section{Background}
\label{sec:background}

\subsection{Integration and Cointegration}\label{sec:coint}\  
Following \cite{johansen1995} we shall say that a stochastic process
$Y_t$ which satisfies $Y_t-E(Y_t) = \sum_{i=0}^\infty C_i\,\varepsilon_{t-i}$ is
called $I(0)$, and then we shall write $Y_t\sim I(0)$, whenever
$\sum_{i=0}^\infty C_i \neq 0$ and $\sum_{i=0}^\infty C_i\,z^i$ converges for
$z\in\mathbb{C}$ with $|z|<1$.  It is understood that the condition
$\varepsilon_t\sim iid(0,\sigma^2)$ holds.

A (vector) time series $\mathbf{y}_t$ is said to be {\em integrated of order\/}
$d$, and then we shall write $\mathbf{y}_t\sim I(d)$, whenever after $d$ times
(discrete) differentiation an stationary process is
obtained, see \cite{banerjee1993}; more precisely, whenever
$(1-L)^d\,\mathbf{y}_t\sim\text{I(0)}$, where $L$ is the usual lag operator:
$(1-L)\,\mathbf{y}_t = \Delta\mathbf{y}_t = \mathbf{y}_t-\mathbf{y}_{t-1}$ for
all $t$.

Note that this definition includes the scalar case as time series of vectors of
dimension 1; in this scalar case we will write the time series in nonbold
format.

Let $\mathbf{y}_t^\nu$, $\nu=1,\dots,l$, be a set of $l$ vector time series of
order $I(1)$.  They are said to be {\em cointegrated\/} if a vector
$\beta=[\beta(1),\dots,\beta(l)]^\top \in \mathbb{R}^p$ exists, such that the
time series,
\begin{equation}
\mathbf{Z}_t:= 
\sum_{\nu=1}^l \beta(\nu)\,\mathbf{y}_t^\nu\,\sim\,\text{I(0)}\,.
\end{equation}
In other words, a set of $I(1)$ time series is said to be cointegrated if a
linear combination of them exists, which is I(0).

\subsection{Vector Autorregresive Models}\label{sec:varvec}

Vector error correction model (VECM) describe the joint behaviour of a set of
variables and can be derived from the simple Vector Autoregressive model (VAR)
presented by \cite{sims1980}.  The VAR($p$) model is a framework describing the
behaviour of a set of $l$ endogenous and stationary variables as a linear
combination of their last $p$ values, where $l,p\in\mathbb{N}$.  In our case,
each one of these $l$ variables is a scalar time series $y_{\lambda,t}$,
$\lambda=1,\dots,l$, and we represent them all together at time $t$ by the
vector time series:
\begin{equation}
\label{eq:variables}
\mathbf{y}_t = 
\begin{bmatrix} y_{1,t} & y_{2,t} & \dots & y_{l,t} \end{bmatrix}^\top.
\end{equation}
\noindent
Notice that the vectors $\mathbf{y}_t$ are assumed to be $l$-dimensional.

The VAR($p$) model describes the behaviour of a dependent variable in terms of
its own lagged values and the lags of the others variables in the system. The
model with $p$ lags is formulated as the system:
\begin{align}
\label{eq:var}
\mathbf{y}_t 
= \boldsymbol{\Phi}_1 \mathbf{y}_{t-1} +
  \boldsymbol{\Phi}_2 \mathbf{y}_{t-2} + \dots +
  \boldsymbol{\Phi}_p\mathbf{y}_{t-p} +
  \mathbf{c} + \boldsymbol{\epsilon}_t \nonumber \\
t=p+1,\dots,N,
\end{align}
\noindent where 
$\boldsymbol{\Phi}_1, \boldsymbol{\Phi}_2,\dots,\boldsymbol{\Phi}_p$
are $l\times l$-matrices of real coefficients,
$\boldsymbol{\epsilon}_{p+1},
 \boldsymbol{\epsilon}_{p+2}, \dots, \boldsymbol{\epsilon}_N$ 
are error terms, $\mathbf{c}$ is a constant vector and $N$ is the total
number of samples.

Notice that, regarding our notation of section (\ref{sec:coint}),
we have here 
$\mathbf{y}_t^0 = \mathbf{y}_t$,
$\mathbf{y}_t^\nu = \mathbf{y}_{t-\nu}$ and
the $\lambda$-th component of the vector time series $\mathbf{y}_t^\nu$
is the scalar time series $y_{\lambda,t}^\nu$, where $\nu=1,\dots,p$ and
$\lambda=1,\dots,l$.

However, the VAR model cannot be used with non-stationary variables. VECM ,
developed by \cite{engle87}, is also a linear model for I(1) variables that are
also cointegrated, see \cite{banerjee1993}. If cointegration exists, variable
differences are stationary and they introduce an error correction term which
adjusts coefficients to bring the variables back to equilibrium.

It is obtained re-writing equation (\ref{eq:var}) in terms of the new
variable $\Delta\mathbf{y}_t=\mathbf{y}_t-\mathbf{y}_{t-1}$.
The VECM model, expressed in terms those differences, takes the form:
\begin{equation}\label{eq:vec}
\Delta \mathbf{y}_t 
= \boldsymbol{\Omega}\,\mathbf{y}_{t-1}
  + \sum_{i=1}^{p-1} \boldsymbol{\Phi}_i^*\,\Delta\mathbf{y}_{t-i}
  + \mathbf{c} + \boldsymbol{\epsilon}_t\,,
\end{equation}
\noindent
where the coefficients matrices $\boldsymbol{\Phi}_i^*$ and 
$\boldsymbol{\Omega}$, expressed in terms of the matrices
$\boldsymbol{\Phi}_i$ of (\ref{eq:var}), are:
\begin{align*}
\boldsymbol{\Phi}_i^* 
&:= -\sum_{j=i+1}^{p}\boldsymbol{\Phi}_j\,, \\
\boldsymbol{\Omega}
&:= -\left( \mathbb{I} - \boldsymbol{\Phi}_1 - \dots 
    - \boldsymbol{\phi}_p \right)\,. 
\end{align*}
The following well known properties of the matrix $\boldsymbol{\Omega}$, see
\cite{johansen1995}, will be useful in the sequel:
\begin{itemize}
\item
If $\boldsymbol{\Omega} = \mathbf{0}$, there is no cointegration.
\item 
If $rank(\boldsymbol{\Omega})=l$, i.e., if $\boldsymbol{\Omega}$ has
full rank, then the time series are not I(1) but stationary.
\item
If $rank(\boldsymbol{\Omega})=r$, $0<r<l$, then there is cointegration
and the matrix $\boldsymbol{\Omega}$ can be expressed as
$\boldsymbol{\Omega}=\boldsymbol{\alpha\beta}^\top$, where $\boldsymbol{\alpha}$
and $\boldsymbol{\beta}$ are
$l\times r$ matrices and
$\text{rank}(\boldsymbol{\alpha})=\text{rank}(\boldsymbol{\beta})=r$.
\item
The columns of $\boldsymbol{\beta}$ contains the cointegration vectors and the rows of
$\boldsymbol{\alpha}$ correspond with the adjusted vectors. 
$\boldsymbol{\beta}$ is obtained by Johansen procedure,see \cite{johansen1988},
whereas $\boldsymbol{\alpha}$ has to be determined as a variable in the VECM.
\end{itemize}
It is worth noticing that the factorization of the matrix
$\boldsymbol\Omega$ is not unique, since for any $r \times r$
non-singular matrix $\mathbf{H}$, $\boldsymbol{\alpha}^*:=\boldsymbol{\alpha}\mathbf{H}$,
and $\boldsymbol{\beta}^*=\boldsymbol{\beta}(\mathbf{H}^{-1})^\top$ we have
$\boldsymbol{\alpha\beta}^\top=\boldsymbol{\alpha}^*(\boldsymbol{\beta}^*)^\top$.
If cointegration exists, then equation (\ref{eq:vec}) can be written
as follows:
\begin{equation}\label{eq:vecfull}
\Delta\mathbf{y}_t 
= \boldsymbol{\alpha\beta}^\top\mathbf{y}_{t-1} 
  + \sum_{i=1}^{p-1}\boldsymbol{\Phi}_i^*\,\Delta\mathbf{y}_{t-i}
  + \mathbf{c} + \boldsymbol{\epsilon}_t\,,
\end{equation}
\noindent
which is a VAR model but for time series differences.

Transposing each equation of the system (\ref{eq:vecfull}) we can write
the VECM($p$) model in block-matrix form as:
\begin{equation}\label{eq:vareq}
\mathbf{B} = 
\mathbf{A} \mathbf{X} + 
\mathbf{E} \, , 
\end{equation}
%
\noindent where $\mathbf{B}$ dimension is $((N-p)\times l)$, $\mathbf{A}$
dimension is $((N-p)\times(r+(p-1)l +1))$, $\mathbf{X}$ dimension is $((r+(p-1)l
+1)\times l)$ and $\mathbf{E}$ dimension is $((N-p)\times l)$:
%
\begin{alignat}{3}
\mathbf{B}
&= \begin{bmatrix}
   \Delta\mathbf{y}_{p+1}^\top \\
   \Delta\mathbf{y}_{p+2}^\top \\
   \vdots \\
   \Delta\mathbf{y}_N^\top
   \end{bmatrix}
&\quad
\mathbf{X}
&= \begin{bmatrix}
   \boldsymbol{\alpha}^\top \\
   \boldsymbol{\Phi}_1^{*\top} \\
   \boldsymbol{\Phi}_2^{*\top} \\
   \vdots \\
   \boldsymbol{\Phi}_{p-1}^{*\top} \\
   \mathbf{c}^\top
   \end{bmatrix}
&\quad
\mathbf{E}
&= \begin{bmatrix}
   \boldsymbol{\epsilon}_{p+1}^\top \\
   \boldsymbol{\epsilon}_{p+2}^\top \\
   \vdots \\
   \boldsymbol{\epsilon}_N^\top \\
   \end{bmatrix}
\end{alignat}
\noindent and 
\begin{align}
\mathbf{A} 
&= \begin{pmat}[{....|}]
   \mathbf{y}_p^\top \boldsymbol{\beta} & \Delta \mathbf{y}_p^\top & \Delta\mathbf{y}_{p-1}^\top & \dots 
                    & \Delta\mathbf{y}_2^\top & 1 \cr
   \mathbf{y}_{p+1}^\top  \boldsymbol{\beta} &\Delta\mathbf{y}_{p+1}^\top & \Delta\mathbf{y}_p^\top & \dots
                       & \Delta\mathbf{y}_3^\top & 1 \cr
   \vdots & \vdots & \vdots & \ddots & \vdots & \vdots \cr
   \mathbf{y}_{N-1}^\top  \boldsymbol{\beta} &\Delta\mathbf{y}_{N-1}^\top & \Delta\mathbf{y}_{N-2}^\top & \dots 
                       & \Delta\mathbf{y}_{N-p-1}^\top & 1 \cr
   \end{pmat}\, .
\label{eq:Amatrix}
\end{align}
Taking into account the error term $\mathbf{E}$, equation~(\ref{eq:vareq}) 
can be solved with respect to $\mathbf{X}$ using the OLS estimation.

\section{Methodology} \label{sec:methodology}
\subsection{Motivation} \label{sec:proposal}

Cointegration vectors can be found applying the Johansen method which uses a
sample of the last historical data. However, VECM assumes cointegration vectors
do not change in time.  In fact, \cite{gregoryETal1996} addresses that the
long-run relationships between the time series  might change due to several
economic factors that can lead to structural breaks in the cointegration
relationship.  In order to show that the number of cointegration vectors depends
on the amount $L$ of historical data and the number of lags $p$ in the VECM, we
used a grid search.  We defined a grid of possible values for $L$ and $p$. $L$
goes throughout $[2,14]$ hours (1 $hour = 360$ data points) with a step size of
4 hours and $p$ throughout $[1,5]$ with step size of 1. The first experiment
consisted in determining the number of cointegration vectors for all
combinations of $L$ and $p$.  We used four forex rates: EURUSD, GBPUSD, USDCHF
and USDJPY with 10-second frequency. Data started at 13:00 GMT of the 13th of
August 2014, when the New York and London financial markets opened.

Figure \ref{fig:hists} shows the distribution of the number of cointegration
vectors given by the Johansen method for different values of $L=[2,6,10,14]$
hours  and $p=1$ . This procedure was carried out by a sliding window of
historical data moving 1000 times.

\begin{figure}[!h]
  %\vspace{-0.8cm}
  \centering
  \includegraphics[width=0.9\textwidth]{Fig1}
  \caption{Distribution of the number of cointegration vectors using $p=1$ lags.
  Four possible values for windows size $L$ are shown (2, 6, 10 and 14) hours (1
  hour = 360 data points).}
  \label{fig:hists}
\end{figure}

From section \ref{sec:background} we know that $r=0$ means no cointegration and
$r=l$ (we are using four rates, so $l=4$) reveals that no process is I(1) but
stationary.  The interesting cases of cointegration are those where $r$ lies
strictly between $0$ and $4$, i.e. $0<r<4$.

In order to measure the extent of cointegration, we introduce a
{\em percentage of cointegration\/} as following:
\begin{equation} \label{eq:pcoint}
PC = 
\frac{\#\{ it \mid \text{$it$ has $r$ c.v. with $0<r<l$}\}}
     {\#it}\times 100
\end{equation}
where c.v. stands for cointegration vectors and $it$ is the number of iterations.

The goal of our next experiment was to find a relation between this ratio $PC$
and the performance measure MSE (see equation \ref{eq:MSE}). $L$ was defined
between $[700,1500]$ data points with step size 100 and $p$ took values between
$[1,5]$ with step size 1.

\begin{figure}[ht!]
  \centering
  \includegraphics[width=0.9\textwidth]{Fig2}
  \caption{MSE versus the percentage of cointegration considering 1000
  iterations. Best windows size $L$ found was 1060. Below MSE versus $L$ shows a
  rapidly decreasing behaviour, founding minimum at PC = 80\% }
  \label{fig:cointvsmse}
\end{figure}

Figure~\ref{fig:cointvsmse} shows the relation between MSE and $PC$ and $L$.  We
found that better cointegration percentage leads to better performance accuracy
in terms of reducing MSE.

Therefore we proposed to choose $L$ and $p$ in order to maximise the percentage
of cointegration $PC$ in the near past. This process was done every time that
new data was processed. However, this search can be slow if we try different
values for $L$ and $p$ and therefore we distributed this calculation in order to
reduce searching time.

Our proposal is then a modified version of VECM, with parameter $p$ and the
amount of historical data used $L$ obtained at every step. Only the search of
$L$ and $p$ was done in a distributed environment, since this is the most
expensive routine. Our proposal is called Adaptive Vector Error Correction model
(AVECM). AVECM is detailed in the algorithm \ref{alg:AVECM} which summarises
our proposal.

\begin{algorithm}[ht!]
\begin{algorithmic}[1]
\REQUIRE $\,$ \\
$\mathbf{y}$: matrix with $N$ input vectors and $l$ time series\\
$j$: Starting point of testing \\
$it$: Ending point of testing \\
$ps$: list of $p$ values \\
$Ls$: list of $L$ values ($L<N$) \\
$m$: Iterations to determine parameters ($m < N-L$)\\
\ENSURE  $\,$ \\
$\{ \hat{\mathbf{y}}[1],\dots,\hat{\mathbf{y}}[it]\}$: prediction vectors \\
\FOR { $i =j$ to $it$ }
   \STATE $\mathbf{Y} \gets \mathbf{y}[:,i-1]$
    \STATE $L,p \gets
    \texttt{get\_best\_params}(Ls,ps,m,\mathbf{Y})$
        \STATE $model = \text{VECM}(\mathbf{Y},L, p)$
        \STATE $\hat{\mathbf{y}}[i-j] = model.predict()$
\ENDFOR
\end{algorithmic}
\caption{AVECM: Adaptive VECM.}
\label{alg:AVECM}
\end{algorithm}

The input of AVECM is time series prices which are cointegrated. The starting
point of testing is $j$ and the total number of iterations is $it$. We need to
ensure that $j$ is at least the maximum value of $Ls$. $Ls$ and $ps$ are the
possible values for $L$ and $p$.  The function {\bf get\_best\_params} makes
this grid search on the two vector lists $Ls$ and $ps$ and returns the
parameters $L$ and $p$ which maximise the percentage of cointegration $PC$ (see
equation~\ref{eq:pcoint}) for a pre-defined number of iterations $m$. This
function is implemented in a distributed environment, thus ensuring a response
before new data is available.  After $L$ and $p$ parameters are found, VECM is
built and used to forecast the next data point.

\subsection{Model comparison} \label{sec:random}
We compared our proposal, in terms of performance, with the naive forecast of
the random walk model and ARIMA. It is still difficult to outperform the random
walk model for standard econometric forecasting models  despite its
simplicity, see \cite{lo2011}. The random walk model is defined as:

\begin{equation}
\mathbf{y}_t = \mathbf{y}_{t-1} + \epsilon_{t}
\label{rwmodel}
\end{equation}

The naive forecast of the time series difference $\hat{\mathbf{y}}_{t+1}$ for
the random walk model is defined as:
\begin{equation}
\hat{\mathbf{y}}_{t+1} = \mathbf{y}_t + \hat{\epsilon}_{t+1} 
\end{equation}
\noindent where  $\hat{\epsilon}_{t+1} = \epsilon_{t}$.

On the other hand, ARIMA is widely used to forecast returns in finance, see
\cite{tsay2005}. A process can be modelled as an ARIMA$(p,d,q)$ model if
$\mathbf{x}_t=\Delta^d \mathbf{y}_t $, i.e after differencing $d$ times the time
series $\mathbf{y}_t$,  we get an ARMA$(p,q)$. Since we are modelling returns,
we used $d=1$.

\subsection{Evaluation methods} \label{sec:evaluation}

Forecast performance was evaluated using two different methods which are
frequently used in finance:
\begin{description}
\item
{\bf MSE},  Mean Square Error measures the distance between forecasts
and the true values and large deviations from the true value have a
large impact due to squaring forecast error.
\begin{equation}\label{eq:MSE}
\text{MSE} = 
\frac{\displaystyle \sum_{t=1}^{N} (\mathbf{y}_t-\hat{\mathbf{y}}_t)^2}{N}
\end{equation}
\item {\bf $U$-statistic}, the Theil's $U$-statistic, presented by
\cite{theil1966}, is a unit free measure obtained as the ratio between the root
MSE (RMSE) of a model and the RMSE of the naive random walk model. A
$U$-statistic less than 1 implies the performance is better than the naive
model.
\end{description}

\section{Experimental results}
\label{sec:results}
\subsection{Data} \label{sec:unitroot}
AVECM tests were carried out using four foreign exchange rates all related to
USD: EURUSD, GBPUSD, USDCHF and USDJPY. This data was collected from
\cite{Dukascopy2014}, a free database which gives access to the Swiss Foreign
Exchange marketplace.

The tests were done using 10-second frequency from ask prices from the 11th to
the 15th of August 2014. Since one day corresponds to 8640 data points and we
used 5 days of data, we have 43,200 data points in total.

\subsection{Unit root tests} \label{sec:unitroottests}

Before running the tests, we firstly checked whether the time series were I(1)
using the Augmented Dickey Fuller (ADF) test with lags $p=1,2,3,4,5$.
\cite{mackinnon2010} presented critical values for rejection of hypothesis of a
unit root: -2.62 (1\%), 1.94 (5\%) and 1.62 (10\%).
\begin{table*}[ht]
\label{tab:adf}
\centering
\begin{tabular}{llllll}
\toprule
{Variable} & {ADF(1)} & {ADF(2)} & {ADF(3)} & {ADF(4)} & {ADF(5)}\\ 
\midrule
EURUSD &  -0.052   & -0.054  & -0.054  & -0.054  & -0.054  \\
GBPUSD &  -0.744  & -0.784  & -0.805  & -0.837  & -0.846  \\
USDCHF &  -0.476   & -0.493  & -0.493  & -0.495  & -0.502  \\
USDJPY &  0.357   & 0.360  & 0.360  & 0.367  & 0.367  \\
$\Delta$ EURUSD & -128.4*  & -128.4*  & -96.85* & -89.12*   & -89.12*\\
$\Delta$ GBPUSD & -131.4*  & -112.7*  & -102.5* & -92.86*   & -88.29*\\
$\Delta$ USDCHF & -127.8*  & -127.8*  & -96.94* & -88.82*   & -80.79*\\
$\Delta$ USDJPY & -135.1*  & -135.1*  & -101.2* & -101.2*   & -101.2*\\
\bottomrule
\addlinespace[1ex]
\multicolumn{6}{l}{ \textsuperscript{*} Indicates significance at 1\% level} \\
\multicolumn{6}{l}{ \textsuperscript{**} Indicates significance at 5\% level} \\
\multicolumn{6}{l}{ \textsuperscript{***} Indicates significance at 10\% level} \\
\multicolumn{6}{l}{MacKinnon critical values for rejection of hypothesis of a unit root are:}\\
\multicolumn{6}{l}{ -2.62 (1\%), -1.94 (5\%) and -1.62 (10\%)}\\
\multicolumn{6}{l}{ADF($d$) Augmented Dickey-Fuller test with lag $d$} 
\end{tabular}
\caption{Unit roots tests for EURUSD, GBPUSD, USDCHF and USDJPY at 10-second
frequency.}
\end{table*}
Table~\ref{tab:adf} shows that all currency rates cannot reject the unit root
test in their level form considering different lags, but they rejected it with
their first differences. This means that all of them are I(1) time series and we
are allowed to use VECM and therefore our proposed AVECM.

\subsection{Performance accuracy} \label{sec:performance}

Algorithms AVECM, ARIMA and the naive random walk were tested using four days of
data (from the 12th to the 15th of August 2014).  For AVECM we considered
different number of iterations (parameter $m$ in algorithm \ref{alg:AVECM}): 10,
50 and 100. We tried 12, 24 and 47 different pair of combinations for $L$ and
$p$. Possible values for $L$ were in the interval $[2,14]$ hours and $p$ can
have values in between $[1,5]$. Best AVECM performance was compared against
ARIMA and the random walk model.  Table~\ref{tab:stats} shows the out-of-sample
performance measures: MSE and $U$-statistic for AVECM and ARIMA. In terms of
both measures we found that AVECM is superior to ARIMA and the naive random walk
model. We also included the p-value that proves that the difference in the MSE
is significant at the 99\% significance level in three of the four currency
rates and at 90\% in the case of GBPUSD. The $U$-statistic shows that AVECM and
ARIMA are superior to the naive random walk model and that our proposal is also
superior to ARIMA.

\begin{table}[!htbp]
\caption{AVECM performance}
\label{tab:stats}
\begin{center}
\begin{tabular}{ccccccc}

& \multicolumn{3}{l}{MSE} & &
\multicolumn{2}{c}{$U$-statistic} \\ 
\hhline{~---~--} \\
 &
AVECM & ARIMA & p-value & &
AVECM & ARIMA \\

\hline
 EURUSD & 1.0702 e-09 & 1.1481 e-09 &  9.2509 e-12 & & 0.6863 & 0.7108\\
 GBPUSD & 1.6630 e-09 & 1.7408 e-09 &  6.9519 e-02 & & 0.6866 & 0.7025\\
 USDCHF & 5.8503 e-10 & 6.3545 e-10 &  2.8999 e-14 & & 0.6803 & 0.7091\\
 USDJPY  & 6.3483 e-06 & 6.5194 e-06 &  6.8536 e-05 & & 0.6964 & 0.7057\\
 \end{tabular}
 \end{center}
\end{table}

\subsection{Parallel implementation} \label{sec:parallel}

To determine $L$ and $p$ based on maximising the percentage of cointegration
requires use of the Johansen method which is a computationally expensive
routine. This procedure is done by the function
get\_best\_params($Ls,ps,m,\mathbf{Y}$) shown in algorithm \ref{alg:AVECM}.  In
order to improve the execution time of this search, our proposal included a
parallel search of VECM parameters using high performance computing.  The main
objective was to obtain a response before a new data arrived in the following 10
seconds.

The Johansen method is already programmed in the Python Statsmodels
library, see \cite{seabold2010}, and the parallel implementation was developed using
MPI in Python.  We chose MPI because it allows large-scale parallel applications
with wide portability to be built, being able to run in large clusters or on
local computers.  We tested our proposal in a cluster with 2 servers Xeon
E5-2667 (2.90GHz) of 24 cores each (48 cores in total) and 24GB RAM.  In order
to compare serial and parallel execution times in AVECM, we set parameter
$it=100$ in algorithm \ref{alg:AVECM}.

The $L$ parameter was always chosen between 2 and 14 hours and $p$ always took
values between 1 and 5. Parameter $nparams$ represents the number of pairs
($L$,$p$) used to maximise the percentage of cointegration.

Execution time depends directly on $L$, $p$ and $nparams$ used, since they
determine the size of matrix $\mathbf{A}$ (see equation \ref{eq:Amatrix}) and
therefore affect the OLS function execution time.  Therefore, if we try more
combinations of $L$ and $p$ (increasing $nparams$) the serial algorithm will
take longer. For this financial time series we are interested in execution times
below 10 seconds  (the time series frequency).

\begin{figure}[ht]
  \centering
  \includegraphics[width=0.9\textwidth]{Fig3}
  \caption{Computing time of sequential and parallel algorithm is shown in the
  upper figure. Speed-up is shown below.}
  \label{fig:extimes}
\end{figure}

The superior figure in \ref{fig:extimes} shows that best performance accuracy
measurements are achieved in times near or below 10 seconds in the parallel
version. Contrarily, serial times are higher, above 10 seconds in most cases.
Figure~\ref{fig:extimes} also shows the speed-up in computing time for AVECM
which is near 9X if we use more than 25 parameters ($nparams$).

Execution times do not consider the loading data time, just that of finding best
parameters and matrix operations. The time MPI spends transferring data and
synchronising processes is about two seconds independently of the number of
processes considered. Execution times were measured using the Time python
library.

\section{Conclusions}
\label{sec:conclusions}
Cointegration in financial time series has been largely studied and the Johansen
method is commonly used to obtain cointegration relationships.  In practice, it
has been found that cointegration relations change with time.  However,
model-based cointegration such as VECM assumes that cointegration remains
unchanged in time.  We empirically showed that the Johansen method is sensitive
to the number of lags but also to the amount of data considered.

Moreover, we introduced the notion of {\em percentage of cointegration\/} and
found that out-of-sample forecast performance MSE is related to the value of
this figure in the last samples.  We used this information to set the model
parameters.  Our proposal AVECM consists of an adaptive algorithm to update VECM
parameters every time that new data is available. These parameters are found by
maximising the percentage of cointegration of the last samples or iterations.

Despite the fact that high frequency Forex data can be spurious, the model
performance can be less reliable (and more spurious) relative to the lower
frequencies (such as 1 minute or 5 minute intervals) adopted by some other
studies. However, the deficiency is offset by gain in accuracy from parallel
processing which is capable of searching or examining a much larger state space
given the same computational time.

Determining VECM parameters was the most expensive routine and it was run using
parallel processes using MPI which allowed a grid search within a range of
values for $L$ and $p$ to be made.  Tests were done using real currency rates
data.

Results showed that our proposed AVECM improves performance measures by finding
parameters of $L$ and $p$ maximising the percentage of cointegration.

The parallel implementation allowed the execution times to be reduced more than
9 times and therefore a response time was obtain before 10 seconds. Since we
used 10-second frequencys we can say that our proposal is suitable for use in
an online context for real applications because response times were less than
this frequency.

For future study, it would be interesting to explore the relationship between
cointegration and performance in order to propose new criteria for improving
VECM parameters. It would also be interesting to include more explaining
variables such as bid-ask spread and change in volume.

### 2016-07-06 articles-conferences/iscef2016/manuscript-computational-economics/main.tex
Final manuscript

--- 2016-03-22 articles-conferences/iscef2016/main.tex
@@
 %
-%\documentclass{svjour3}                     % onecolumn (standard format)
+%\documentclass{svjour3}                     
 %\documentclass[smallcondensed]{svjour3}     % onecolumn (ditto)
-\documentclass[smallextended]{svjour3}       % onecolumn (second format)
+\documentclass[smallextended]{svjour3}       % onecolumn (second format) ******
 %\documentclass[twocolumn]{svjour3}          % twocolumn
-%
+
 \smartqed  % flush right qed marks, e.g. at end of proof
+\usepackage{natbib}
 
 \usepackage{graphicx}
-%
-% \usepackage{mathptmx}      % use Times fonts if available on your TeX system
-%
-% insert here the call for the packages your document requires
-%\usepackage{latexsym}
-% etc.
-%
+
 \usepackage{amsfonts,amssymb,amsmath,
@@
 for forecasting high frequency financial time series}
-%\subtitle{Do you have a subtitle?\\ If so, write it here}
 
@@
 
-%\author{Paola Arce  \and Jonathan Antognini \and Werner Kristjanpoller \and Luis Salinas }
-\author{}
+\author{Paola Arce  \and Jonathan Antognini \and Werner Kristjanpoller \and Luis Salinas }
 %\authorrunning{Short form of author list} % if too long for running head
 
-%\institute{P. Arce  \at
-%              Departamento de Inform\'atica, Universidad T\'ecnica Federico Santa
-%              Mar\'ia\\
-%              Tel.: +56-32-2654000\\
-%              \email{paola.arce@usm.cl}           %  \\
-%%             \emph{Present address:} of F. Author  %  if needed
-%           \and
-%           J. Antognini \at
-%              Departamento de Inform\'atica, Universidad T\'ecnica Federico Santa
-%              Mar\'ia\\
-%              \email{jantognini@csrg.cl}             \\
-%           \and
-%           W. Kristjanpoller \at
-%              Departamento de Industrias, Universidad T\'ecnica Federico Santa
-%              Mar\'ia\\
-%              %Fax: +123-45-678910\\
-%              \email{werner.kristjanpoller@usm.cl}             \\
-%           \and
-%           L. Salinas \at
-%              Departamento de Inform\'atica, Universidad T\'ecnica Federico Santa
-%              Mar\'ia\\
-%              Centro Cient\'ifico Tecnol\'ogico de Valpara\'iso (CCTVal) \\
-%              \email{luis.salinas@usm.cl}           %  \\
-%}
-
+\institute{P. Arce  \at
+              Departamento de Inform\'atica, Universidad T\'ecnica Federico Santa
+              Mar\'ia\\
+              Tel.: +56-32-2654000\\
+              \email{paola.arce@usm.cl}           %  \\
+%             \emph{Present address:} of F. Author  %  if needed
+           \and
+           J. Antognini \at
+              Departamento de Inform\'atica, Universidad T\'ecnica Federico Santa
+              Mar\'ia\\
+              \email{jantognini@csrg.cl}             \\
+           \and
+           W. Kristjanpoller \at
+              Departamento de Industrias, Universidad T\'ecnica Federico Santa
+              Mar\'ia\\
+              %Fax: +123-45-678910\\
+              \email{werner.kristjanpoller@usm.cl}             \\
+           \and
+           L. Salinas \at
+              Departamento de Inform\'atica, Universidad T\'ecnica Federico Santa
+              Mar\'ia\\
+              Centro Cient\'ifico Tecnol\'ogico de Valpara\'iso (CCTVal) \\
+              \email{luis.salinas@usm.cl}           %  \\
+}
 \date{Received: date / Accepted: date}
-% The correct dates will be entered by the editor
 
@@
 Cointegration is a long-run property of some non-stationary time series where a
-linear combination of those time series is stationary.
-This behaviour has been studied in finance because cointegration restrictions often improve forecasting.
-The Vector Error Correction Model (VECM) is a well-known econometric technique that
-characterises short-run variations of a set of cointegrated time series incorporating long-run relationships as an error correction term.
-VECM has been broadly used with low frequency time series.
-We aimed to adapt VECM to be used in finance with high frequency stream data.
-%
+linear combination of those time series is stationary.  This behaviour has been
+studied in finance because cointegration restrictions often improve forecasting.
+The Vector Error Correction Model (VECM) is a well-known econometric technique
+that characterises short-run variations of a set of cointegrated time series
+incorporating long-run relationships as an error correction term. VECM has been
+broadly used with low frequency time series. We aimed to adapt VECM to be used
+in finance with high frequency stream data.
 
-Cointegration relations change in time and therefore VECM parameters must be updated when new data is available.
-We studied how forecasting performance is affected when VECM parameters and the length of historical data used change in time.
-We observed that the number of cointegration relationships varies with the length of historical data used.
-Moreover, parameters that increased these relationships in time led to better forecasting performance.
-Our proposal, called an Adaptive VECM (AVECM) is
-to make a parameters grid search that maximises the number of cointegration relationships in the near past. To ensure the search can be executed fast enough, we used a distributed environment.
+Cointegration relations change in time and therefore VECM parameters must be
+updated when new data is available. We studied how forecasting performance is
+affected when VECM parameters and the length of historical data used change in
+time. We observed that the number of cointegration relationships varies with the
+length of historical data used. Moreover, parameters that increased these
+relationships in time led to better forecasting performance. Our proposal,
+called an Adaptive VECM (AVECM) is to make a parameters grid search that
+maximises the number of cointegration relationships in the near past. To ensure
+the search can be executed fast enough, we used a distributed environment.
 
-%
-The methodology was tested using four 10-seconds frequency time series of the Foreign Exchange market.
-We compared our proposal with ARIMA and the naive forecast of the random walk model.
-Numerical experiments showed that on average AVECM performed better than ARIMA and random walk.
-Additionally, our parallel approach AVECM significantly improved execution times with respect to its serial version.
+The methodology was tested using four 10-second frequency time series of the
+Foreign Exchange market. We compared our proposal with ARIMA and the naive
+forecast of the random walk model. Numerical experiments showed that on average
+AVECM performed better than ARIMA and random walk. Additionally, AVECM
+significantly improved execution times with respect to its serial version.
 
 \keywords{VECM \and Cointegration \and Forex \and MPI \and Parallel algorithm}
-% \PACS{PACS code1 \and PACS code2 \and more}
-% \subclass{MSC code1 \and MSC code2 \and more}
+
 \end{abstract}
 
-\input{src/intro}
-\input{src/background}
-\input{src/methodology}
-\input{src/results}
-\input{src/conclusions}
-
-\begin{acknowledgements}
-L. Salinas acknowledges support from Basal Project
-FB0821 CCTVal-Centro Científico Tecnológico de Valparaíso. P. Arce and W.
-Kristjanpoller thank PIIC 2014 DGIP-UTFSM grant for supporting this research.
-\end{acknowledgements}
+\input{body}
 
 % BibTeX users please use one of
-%\bibliographystyle{spbasic}      % basic style, author-year citations
+\bibliographystyle{spbasic}      % basic style, author-year citations
 %\bibliographystyle{spmpsci}      % mathematics and physical sciences
@@
 
-\bibliographystyle{plain}       % APS-like style for physics
+%\bibliographystyle{plain}       % APS-like style for physics
 
@@
 \end{document}
-% end of file template.tex

### 2016-07-06 articles-conferences/iscef2016/old-manuscript/manuscript.tex
Final manuscript

--- 2015-04-09 src/5_2_Proposal.tex
@@
-\chapter{Adaptive parallel algorithm for VECM}
+\RequirePackage{fix-cm}
+%
+%\documentclass{svjour3}                     % onecolumn (standard format)
+%\documentclass[smallcondensed]{svjour3}     % onecolumn (ditto)
+%\documentclass[smallextended]{svjour3}       % onecolumn (second format)
+\documentclass[twocolumn]{svjour3}          % twocolumn
+%
+\smartqed  % flush right qed marks, e.g. at end of proof
+%
+\usepackage{graphicx}
+%
+% \usepackage{mathptmx}      % use Times fonts if available on your TeX system
+%
+% insert here the call for the packages your document requires
+%\usepackage{latexsym}
+% etc.
+%
+\usepackage{amsfonts,amssymb,amsmath,
+            amssymb,amsopn,amsxtra,amstext,amsbsy}
+\usepackage{epsfig}
+\usepackage[makeroom]{cancel}
+\usepackage{array,arydshln}
+\usepackage{array}
+\usepackage{url}
+%\usepackage{tikz}
+\usepackage[utf8x]{inputenc}
+\usepackage{fancyhdr}
+\usepackage{hyperref,wasysym}
+\usepackage{color}
+\usepackage{enumerate}
+\usepackage{algorithm}
+\usepackage{algorithmic}
+\renewcommand{\algorithmicrequire}{\textbf{Input:}}
+\renewcommand{\algorithmicensure}{\textbf{Output:}}
+\usepackage{multirow}
+%\usepackage[spanish]{babel}
 
-\vspace{0.5cm}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\usepackage{xcolor}%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\usepackage{hyperref}%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\usepackage{pmat}%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-VECM parameters are obtained using ordinary least squares (OLS) method
-\cite{golub1980}, which is computationally expensive when the number of past
-values $L$ and observations $p$ increases. Moreover, obtaining cointegration
-vectors is also an expensive routine because of the use of the Johansen method.
+%\newtheorem*{myteo}{Teorema}
+%\newtheorem*{mydef}{Definici\'on}
+%% MYREMARK ENVIRONMENT %%%%%%%%%%%
+\newenvironment{mycolor}{\color{blue}}{}
+\newcounter{myremarkcount}
+\setcounter{myremarkcount}{0}
+\newsavebox{\myremarkbox}
+\newenvironment{myremark}
+{% This is the begin code
+\medskip
+\begin{flushright}
+\begin{lrbox}{\myremarkbox}%
+\begin{minipage}[t]{1.00\linewidth}
+\stepcounter{myremarkcount}
+\footnotesize
+\begin{mycolor}
+{{\bf Proofreading Remark.}\ }
+% \begin{it}
+}
+{% This is the end code
+% \end{it}
+\end{mycolor}
+\end{minipage}
+\end{lrbox}
+\fbox{\usebox{\myremarkbox}}
+\end{flushright}
 
-In this study we found that for some combinations of $L$ and $p$, the number of
-cointegration relations is increased and this leads to better performance. We
-propose to parallelise this search in order to achive faster execution times.
+\bigskip}
 
-\section{The problem} \label{sec:proposal}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-Cointegration vectors can be found applying the Johansen method which uses a
-sample of the most recent historical data. However, one of the main limitations of this
-methodology is that it assumes cointegration vectors do not change in time.  In
-fact, the long-run relationships between the time series might change due to
-several economic factors that can lead to structural breaks in the cointegrating
-relation \cite{gregoryETal1996}.
+\journalname{Cluster Computing}
 
-In this study, we show that cointegration relations depend on the amount $L$ of
-historical data considered, and how many lags $p$ were used. This search is
-computationally expensive, considering that the Johansen
-method~\cite{johansen1995} (which is of order $O(n^3)$) is used and these
-parameters change at every iteration.
+\begin{document}
 
-Our proposal consists on to parallelise this search in order to get new
-parameters at every iteration.
+\title{An adaptive parallel algorithm for Vector Error Correction models applied
+to the Foreign Exchange
+Market
+}
+%\subtitle{Do you have a subtitle?\\ If so, write it here}
+
+%\titlerunning{Short form of title}        % if too long for running head
+
+\author{Paola Arce \and Jonathan Antognini \and Werner Kristjanpoller \and Luis Salinas }
+
+%\authorrunning{Short form of author list} % if too long for running head
+
+\institute{P. Arce  \at
+              Departamento de Inform\'atica, Universidad T\'ecnica Federico Santa
+              Mar\'ia\\
+              Tel.: +56-32-2654000\\
+              %Fax: +123-45-678910\\
+              \email{paola.arce@usm.cl}           %  \\
+%             \emph{Present address:} of F. Author  %  if needed
+           \and
+           J. Antognini \at
+              Departamento de Inform\'atica, Universidad T\'ecnica Federico Santa
+              Mar\'ia\\
+              %Fax: +123-45-678910\\
+              \email{jantognini@csrg.cl}             \\
+           \and
+           W. Kristjanpoller \at
+              Departamento de Industrias, Universidad T\'ecnica Federico Santa
+              Mar\'ia\\
+              %Fax: +123-45-678910\\
+              \email{werner.kristjanpoller@usm.cl}             \\
+           \and
+           L. Salinas \at
+              Departamento de Inform\'atica, Universidad T\'ecnica Federico Santa
+              Mar\'ia\\
+              Centro Científico Tecnológico de Valparaíso (CCTVal) \\
+              \email{luis.salinas@usm.cl}           %  \\
+}
+
+\date{Received: date / Accepted: date}
+% The correct dates will be entered by the editor
+
+\maketitle
+
+\begin{abstract}
+
+A Vector Error Correction Model (VECM) is an econometric model which
+characterizes the joint dynamic behaviour of a set of cointegrated variables. It
+is said that two non-stationary time series are cointegrated if a linear
+combination of them exhibit a stationary behaviour.
+
+Cointegration is commonly found in finance and it has been shown that
+introducing cointegration restrictions often improves forecasting. However,
+determining VECM parameters and how much of the historical data is useful for
+making a forecast is a computational challenge.
+
+In this study, we propose a new version of VECM whose parameters are obtained
+adaptively using parallel computing taking into account the cointegrated
+behaviour of the time series. The methodology is tested using data of the
+Foreign Exchange (Forex) market. Numerical experiments show that our adaptive
+and parallel approach significantly improves execution times with respect to the
+common serial processing and forecasts are made before new data arrives.
+
+\keywords{VECM \and Cointegration \and Forex \and MPI \and Parallel algorithm}
+
+\end{abstract}
+
+\section{Introduction}
+\label{sec:introduction}
+In finance, it is common to find variables with long-run equilibrium
+relationships. This is called cointegration and it reflects the idea of that
+some sets of variables cannot wander too far from each other. Cointegration
+means that one or more linear combinations of these variables are stationary
+even though individually they are not. Furthermore, the number of cointegration
+vectors reflects how many of these linear combinations exist.  Some models,
+such as the Vector Error Correction (VECM) \cite{engle87}, take advantage of
+this property and describe the joint behaviour of several cointegrated
+variables.
+
+VECM introduces this long-run relationship among a set of cointegrated
+variables as an error correction term. VECM is a special case of the vector
+autoregressive model (VAR) model \cite{sims1980}. VAR model expresses future
+values as a linear combination of variables past values.  However, the VAR
+model cannot be used with non-stationary variables. VECM is a linear model but
+in terms of variable differences. If cointegration exists, variable differences
+are stationary and they introduce an error correction term which adjusts
+coefficients to bring the variables back to equilibrium.
+
+Both VECM and VAR model parameters are obtained using ordinary least squares
+(OLS) method~\cite{golub1980}. Since OLS involves many calculations, the parameter estimation
+method is computationally expensive when the number of past values and
+observations increases. Moreover, obtaining cointegration vectors 
+is also an expensive routine.
+
+In finance, many economic time series are revealed to be stationary when they
+are differentiated and moreover cointegration restrictions often improve
+forecasting \cite{duy1998}. Therefore, VECM has been widely adopted in
+finance: \cite{mukherjee1995}, \cite{maysami2000}, \cite{arestis2001} and
+\cite{seong2013} to name a few. Pair
+trading is a very common example of a cointegration
+application \cite{herlemont2003} but also can also be extended to a larger set
+of variables \cite{mukherjee1995}, \cite{engle2004}.
+
+Based on this fact, we proposed in this paper a new forecast method which
+determines adaptively the number of observations and lags of VECM based on
+cointegration which we called Adaptively Vector Error Correction (AVECM).  Since
+cointegration often improves forecasting, model parameters will be those which
+maximize cointegration relations in the past.  This search is computationally
+expensive because to find cointegration vectors we used the Johansen
+method~\cite{johansen1995}, which is of order $O(n^3)$. Therefore, our proposal
+is to parallelize this search in order to get new parameters at every iteration.
+
+Model effectiveness is focused on out-of-sample forecast rather than in-sample
+fitting. This criterion allows our proposal AVECM prediction capability to be
+expressed rather than just explaining data history.
 
 The forecast capability of our method was measured using MAPE, MAE and
-RMSE \cite{armstrong1992}. Tests were run using four currency rates: Euro (EUR) to
+RMSE~\cite{armstrong1992}. Tests were run using four currency rates: Euro (EUR) to
 United States Dollar (USD) (EURUSD), British Pound (GBP) to USD (GBPUSD), USD to Swiss
@@
 
-\section{Methodology}
+This paper is organized as follows: section~\ref{sec:background} presents the
+VAR and VECM, the AVECM algorithm is presented in section~\ref{sec:methodology}.
+In section~\ref{sec:results} we describe the tests carried on to assess the
+accuracy and the execution time of AVECM.  This section also includes a
+description of the test data.  Section~\ref{sec:conclusions} contains the
+conclusions and a discussion of future research.
 
-VECM is obtained re-writing equation (\ref{eq:var}) in terms of the new
+\section{Background}
+\label{sec:background}
+\subsection{Integration and Cointegration}\label{sec:coint}\  
+Following Johansen \cite{johansen1995} we shall say that a stochastic process
+$Y_t$ which satisfies $Y_t-E(Y_t) = \sum_{i=0}^\infty C_i\,\varepsilon_{t-i}$ is
+called $I(0)$, and then we shall write $Y_t\sim I(0)$, whenever
+$\sum_{i=0}^\infty C_i \neq 0$ and $\sum_{i=0}^\infty C_i\,z^i$ converges for
+$z\in\mathbb{C}$ with $|z|<1$.  It is understood that the condition
+$\varepsilon_t\sim iid(0,\sigma^2)$ holds.
+
+A (vector) time series $\mathbf{y}_t$ is said to be {\em integrated of order\/}
+$d$, and then we shall write $\mathbf{y}_t\sim I(d)$, whenever after $d$ times
+(discrete) differentiation an stationary process is obtained~\cite{banerjee1993};
+more precisely, whenever
+$(1-L)^d\,\mathbf{y}_t\sim\text{I(0)}$, where $L$ is the usual lag operator:
+$(1-L)\,\mathbf{y}_t = \Delta\mathbf{y}_t = \mathbf{y}_t-\mathbf{y}_{t-1}$ for
+all $t$.
+
+Note that this definition includes the scalar case as time series of
+vectors of dimension 1; in this scalar case we will write the time series in
+nonbold format.
+
+Let $\mathbf{y}_t^\nu$, $\nu=1,\dots,p$, be a set of $p$ vector time series of
+order $I(1)$.  They are said to be {\em cointegrated\/} if a vector
+$\beta=[\beta(1),\dots,\beta(p)]^\top \in \mathbb{R}^p$ exists, such that the
+time series,
+\begin{equation}
+\mathbf{Z}_t:= 
+\sum_{\nu=1}^p \beta(\nu)\,\mathbf{y}_t^\nu\,\sim\,\text{I(0)}\,.
+\end{equation}
+In other words, a set of $I(1)$ time series is said to be cointegrated if a
+linear combination of them exists, which is I(0).
+
+\subsection{Vector Autorregresive Models}\label{sec:varvec}
+
+VECM is a special case of VAR model and both describe the joint behaviour
+of a set of variables.
+
+The VAR($p$) model is a general framework describing the behaviour of a
+set of $l$ endogenous variables as a linear combination of their last $p$
+values, where $l,p\in\mathbb{N}$. 
+In our case, each one of these $l$ variables is a scalar time series
+$y_{\lambda,t}$, $\lambda=1,\dots,l$, and we represent them all together
+at time $t$ by the the vector time series:
+\begin{equation}
+\label{eq:variables}
+\mathbf{y}_t = 
+\begin{bmatrix} y_{1,t} & y_{2,t} & \dots & y_{l,t} \end{bmatrix}^\top.
+\end{equation}
+\noindent
+Notice that the vectors $\mathbf{y}_t$ are assumed to be $l$-dimensional.
+
+The VAR($p$) model describes the behaviour of a dependent variable in terms of
+its own lagged values and the lags of the others variables in the system. The
+model with $p$ lags is formulated as the system of $N$:
+\begin{align}
+\label{eq:var}
+\mathbf{y}_t 
+= \boldsymbol{\Phi}_1 \mathbf{y}_{t-1} +
+  \boldsymbol{\Phi}_2 \mathbf{y}_{t-2} + \dots +
+  \boldsymbol{\Phi}_p\mathbf{y}_{t-p} +
+  \mathbf{c} + \boldsymbol{\epsilon}_t \nonumber \\
+t=p+1,\dots,N,
+\end{align}
+\noindent where 
+$\boldsymbol{\Phi}_1, \boldsymbol{\Phi}_2,\dots,\boldsymbol{\Phi}_p$
+are $l\times l$-matrices of real coefficients,
+$\boldsymbol{\epsilon}_{p+1},
+ \boldsymbol{\epsilon}_{p+2}, \dots, \boldsymbol{\epsilon}_N$ 
+are error terms, $\mathbf{c}$ is a constant vector and $N$ is the total
+number of samples.
+
+Notice that, regarding our notation of section (\ref{sec:coint}),
+we have here 
+$\mathbf{y}_t^0 = \mathbf{y}_t$,
+$\mathbf{y}_t^\nu = \mathbf{y}_{t-\nu}$ and
+the $\lambda$-th component of the vector time series $\mathbf{y}_t^\nu$
+is the scalar time series $y_{\lambda,t}^\nu$, where $\nu=1,\dots,p$ and
+$\lambda=1,\dots,l$.
+
+VECM is a special form of a VAR model for I(1) variables that are also
+cointegrated~\cite{banerjee1993}.
+
+It is obtained re-writing equation (\ref{eq:var}) in terms of the new
 variable $\Delta\mathbf{y}_t=\mathbf{y}_t-\mathbf{y}_{t-1}$.
@@
 \end{itemize}
-
+It is worth noticing that the factorization of the matrix
+$\boldsymbol\Omega$ is not unique, since for any $r \times r$
+non-singular matrix $\mathbf{H}$, $\boldsymbol{\alpha}^*:=\boldsymbol{\alpha}\mathbf{H}$,
+and $\boldsymbol{\beta}^*=\boldsymbol{\beta}(\mathbf{H}^{-1})^\top$ we have
+$\boldsymbol{\alpha\beta}^\top=\boldsymbol{\alpha}^*(\boldsymbol{\beta}^*)^\top$.
 If cointegration exists, then equation (\ref{eq:vec}) can be written
@@
 
-\subsection{Motivation}
-The relation between the number of cointegration relations
-and parameters $L$ and $p$ is studied and then a relation between cointegration
-and performance is stated.
+\subsection{Ordinary Least Squares method}
+
+The solution $\widehat{\mathbf{A}}$ to
+equation~(\ref{eq:vareq}) can be obtained by the ordinary least squares (OLS)
+method. $\widehat{\mathbf{X}}$ is the solution of the quadratic optimization problem
+\begin{equation*}
+\widehat{\mathbf{X}} = \underset{\mathbf{X}}{\text{Arg\;min}}
+\|\mathbf{A}\mathbf{X}-\mathbf{B}\|_2^2
+\end{equation*}
+\noindent for which the solution $\widehat{\mathbf{X}}$ is well-known:
+\begin{equation*}
+\label{eq:MP}
+\widehat{\mathbf{X}}=\mathbf{A}^{\!\!+}\,\mathbf{B}
+\end{equation*}
+\noindent where $\mathbf{A}^{\!\!+}$ is the Moore-Penrose pseudo-inverse
+which, when $\mathbf{A}$ is full rank, can be written as follows: 
+\begin{equation}
+\label{eq:pseudoinverse}
+\mathbf{A}^{\!\!+}= (\mathbf{A}^{\!\!\top} \mathbf{A})^{-1}\mathbf{A}^{\!\!\top} \, .
+\end{equation}
+when $\mathbf{A}$ is not full rank, i.e
+$rank(\mathbf{A})=k <  n \leq m$, $\mathbf{A}^\top \mathbf{A}$ is
+always singular and equation~(\ref{eq:pseudoinverse}) cannot be used.
+More generally, the pseudo-inverse is best computed using the compact
+singular value decomposition (SVD) of $\mathbf{A}$ which is:
+\begin{equation*}
+    \label{eq:compactsvd}
+    \mathbf{A}=
+    \mathbf{U_1}
+    \boldsymbol \Sigma_1
+    \mathbf{V}_1^{\top} \, .
+\end{equation*}
+Pseudo-inverse can then be written as follows:
+\begin{equation*}
+\label{eq:pseudoinversesvd}
+\mathbf{A}^{\!\!+} = \mathbf{V}_1 \boldsymbol \Sigma_1^{-1} \mathbf{U}_1^\top \, .
+\end{equation*}
+
+\section{Methodology} \label{sec:methodology}
+\subsection{Motivation} \label{sec:proposal}
+
+Cointegration vectors can be found applying the Johansen method which uses a
+sample of the last historical data. However, one of the main limitations of this
+methodology is that it assumes cointegration vectors do not change in time.
+In fact, the long-run relationships between the time series might change due to
+several economic factors that can lead to structural breaks in the cointegrating
+relation \cite{gregoryETal1996}. 
+Experiments show that cointegration depends on the amount $L$ of historical data
+considered, and how many lags $p$ we use.
 
@@
 the Johansen method for different values of the parameters $L$ and $p$, using
-four currencies data, i.e., $l=4$, and 1000 considering iterations ($it=1000$).
-\begin{figure}[!h]
-  %\vspace{-0.8cm}
-  \centering
-  \includegraphics[width=\textwidth]{img/histCointVectors}
-  \caption{Histogram of the number of cointegration vectors using $p=1$. Four
-  possible values for $L$ are shown (100, 1900, 6100 and 10700).
-  In the cases of $L=100$ and $L=6100$, most of the time there is little
-  cointegration (0 or 4 cointegration vectors). However, for $L=1900$ and
-  $L=10700$ significant cointegration is found.}
-  \label{fig:hists}
-\end{figure}
-We know that when there is no cointegration vectors, i.e $r=0$, 
-there is no cointegration. In the same way, when $r=l$ reveals that no process is I(1) but,
+four currencies data, i.e., $l=4$, and considering 1000 iterations $it$.
+From the previous section \ref{sec:background} we know that when $r=0$, 
+there is no cointegration and $r=l$ reveals that no process is I(1) but,
 instead, they are all stationary.
@@
   \centering
-  \includegraphics[width=\textwidth]{img/histCointVectorsp5}
+  \includegraphics[width=0.45\textwidth]{histCointVectors}
+  \caption{Histogram of the number of cointegration vectors using $p=1$. Four
+  possible values for $L$ are shown (100, 1900, 6100 and 10700).
+  In the cases of $L=100$ and $L=6100$, most of the time there is little
+  cointegration (0 or 4 cointegration vectors). However, for $L=1900$ and
+  $L=10700$ significant cointegration is found.}
+  \label{fig:hists}
+\end{figure}
+
+\begin{figure}[!h]
+  %\vspace{-0.8cm}
+  \centering
+  \includegraphics[width=0.45\textwidth]{histCointVectorsp5}
   \caption{Histogram of the number of cointegration vectors using $p=5$. 
-   Despite the change in the distribution of cointegration vectors, 
+   Despite the chande in the distribution of cointegration vectors, 
    the percentage of cointegration is nearly the same as in the case $p=1$. 
@@
   \centering
-  \includegraphics[width=0.7\textwidth]{img/MAPEvsCoint-offset21600-p-2-freq-10s}
+  \includegraphics[width=0.48\textwidth]{MAPEvsCoint-offset21600-p-2-freq-10s}
   \caption{MAPE versus percentage of cointegration. MAPE was obtained from 1000
@@
 for $L$ and $p$ and also that better cointegration percentage leads to better
-accuracy performance.  Therefore we propose to choose best $L$ and $p$
-parameters determined at every step.  The criterion for choosing $L$ and $p$ is
-to maximise the percentage of cointegration $PC$ considering a number of past
-iterations $it$.
+accuracy performance.
+Therefore we propose to choose best $L$ and $p$ parameters determined at every
+step. 
+The criterion for choosing $L$ and $p$ is to maximize the percentage of
+cointegration $PC$ considering a number of past iterations $it$.
 
-Algorithm ~\ref{alg:AVECM} solves equation \ref{eq:vareq} defining first
-parameters $L$ and $p$ using the function {\bf get\_best\_params} which makes a
-grid search on the two vector lists $Ls$ and $ps$ and returns the parameters $L$
-and $p$ which maximise the percentage of cointegration $PC$ (see
-equation~\ref{eq:pcoint}) for a pre-defined number of iterations.  Since this
-search is computational expensive, the number of cointegration vectors at each
-iteration is computed in parallel, thus ensuring a response before new data is
-available.  This parallel routine also allows to refine our grid search leading
-to obtain improved forecasting performance.  After that optimum $L$ and $p$
-parameters are found, VECM is built and used to forecast next data.
+Algorithm ~\ref{alg:AVECM} summarizes our proposal. 
+The function {\bf get\_best\_params} makes a grid search on the two vector
+lists $Ls$ and $ps$ and returns the parameters $L$ and $p$ which maximize
+the percentage of cointegration $PC$ (see equation~\ref{eq:pcoint}) for a
+pre-defined number of iterations.
+Since this search is computational expensive, the number of cointegration
+vectors at each iteration is computed in parallel, thus ensuring a response
+before new data is available. 
+This parallel routine also allows to refine our grid search leading to obtain
+improved forecasting performance.
+After that optimum $L$ and $p$ parameters are found, VECM is built and used
+to forecast next data.
 
@@
 \end{algorithm}
-%
-%
-%\subsection{Evaluation methods} \label{sec:evaluation}
-%
-%Forecast performance was evaluated using different methods. We chose three
-%measures that are frequently used:
-%\begin{description}
-%\item
-%{\bf MAPE}, Mean Average Percent Error, which presents forecast errors as a
-%percentage:
-%\begin{equation}\label{eq:MAPE}
-%\text{MAPE} = \frac{1}{N} \sum_{t=1}^{N} 
-%\frac{\left|\mathbf{y}_t-\hat{\mathbf{y}}_t\right|}{\left|\mathbf{y}_t\right|}
-% \times 100 
-%\end{equation}
-%\item
-%{\bf MAE}, Mean Average Error, which measures the distance between forecasts and the
-%true value.
-%\begin{equation}\label{eq:MAE}
-%\text{MAE} = \frac{1}{N} \sum_{t=1}^{N} 
-%\left| 
-%\mathbf{y}_t-\hat{\mathbf{y}}_t
-%\right| 
-%\end{equation}
-%\item
-%{\bf RMSE}, Root Mean Square Error, also measures the distance between forecasts
-%and the true values but, unlike MAE, large deviations from the true value have a
-%large impact on RMSE due to squaring forecast error.
-%\begin{equation}\label{eq:RMSE}
-%\text{RMSE} = \sqrt{
-%\frac{\displaystyle \sum_{t=1}^{N} (\mathbf{y}_t-\hat{\mathbf{y}}_t)^2}{N}}
-%\end{equation}
-%\end{description}
-%
-%
-%
+
+\subsection{Evaluation methods} \label{sec:evaluation}
+
+Forecast performance was evaluated using different methods. We chose three
+measures that are frequently used:
+\begin{description}
+\item
+{\bf MAPE}, Mean Average Percent Error, which presents forecast errors as a
+percentage:
+\begin{equation}\label{eq:MAPE}
+\text{MAPE} = \frac{1}{N} \sum_{t=1}^{N} 
+\frac{\left|\mathbf{y}_t-\hat{\mathbf{y}}_t\right|}{\left|\mathbf{y}_t\right|}
+ \times 100 
+\end{equation}
+\item
+{\bf MAE}, Mean Average Error, which measures the distance between forecasts and the
+true value.
+\begin{equation}\label{eq:MAE}
+\text{MAE} = \frac{1}{N} \sum_{t=1}^{N} 
+\left| 
+\mathbf{y}_t-\hat{\mathbf{y}}_t
+\right| 
+\end{equation}
+\item
+{\bf RMSE}, Root Mean Square Error, also measures the distance between forecasts
+and the true values but, unlike MAE, large deviations from the true value have a
+large impact on RMSE due to squaring forecast error.
+\begin{equation}\label{eq:RMSE}
+\text{RMSE} = \sqrt{
+\frac{\displaystyle \sum_{t=1}^{N} (\mathbf{y}_t-\hat{\mathbf{y}}_t)^2}{N}}
+\end{equation}
+\end{description}
+
 \section{Experimental results}
@@
 \begin{table}[h!]
+\label{tab:adf}
 \begin{center}
@@
 frequency.}
-\label{tab:adf}
 \end{table}
 
-\section{Parallel implementation} \label{sec:paralell}
+\subsection{Parallel implementation} \label{sec:paralell}
 
@@
   \centering
-  \includegraphics[width=\textwidth]{img/accuracy}
+  \includegraphics[width=0.48\textwidth]{accuracy}
   \caption{AVECM out-of-sample forecasting for four currency rates.}
@@
 \begin{table*}[ht!]
-\tiny
 \caption{Performance and execution times for min MAPE criteria}
@@
   \centering
-  \includegraphics[width=0.7\textwidth]{img/extimes}
+  \includegraphics[width=0.5\textwidth]{extimes}
   \caption{Computing time of sequential and parallel algorithm is shown in the
@@
 best parameters and the building the VECM model. The time MPI spends
-transferring data and synchronising process is about two seconds independently
+transferring data and synchronizing process is about two seconds independently
 of the number of process considered.
@@
 parameters every time that new data is available. These parameters are found by
-maximising the percentage of cointegration of the last samples or iterations.
+maximizing the percentage of cointegration of the last samples or iterations.
 
@@
 Results showed that our proposed AVECM improves performance measures by finding
-parameters of $L$ and $p$ maximising the percentage of cointegration.
+parameters of $L$ and $p$ maximizing the percentage of cointegration.
 These facts were consistent for different times of the day. 
@@
 improving VECM parameters.
+
+\begin{acknowledgements}
+L. Salinas acknowledges support from Basal Project
+FB0821 CCTVal-Centro Científico Tecnológico de Valparaíso. P. Arce and W.
+Kristjanpoller thank PIIC 2014 DGIP-UTFSM grant for supporting this research.
+\end{acknowledgements}
+
+% BibTeX users please use one of
+%\bibliographystyle{spbasic}      % basic style, author-year citations
+%\bibliographystyle{spmpsci}      % mathematics and physical sciences
+%\bibliographystyle{spphys}       % APS-like style for physics
+
+\bibliographystyle{plain}       % APS-like style for physics
+
+%\bibliography{reference}   % name your BibTeX data base
+
+\begin{thebibliography}{10}
+
+\bibitem{arestis2001}
+Philip Arestis, Panicos~O Demetriades, and Kul~B Luintel.
+\newblock Financial development and economic growth: the role of stock markets.
+\newblock {\em Journal of Money, Credit and Banking}, pages 16--41, 2001.
+
+\bibitem{armstrong1992}
+J~Scott Armstrong and Fred Collopy.
+\newblock Error measures for generalizing about forecasting methods: Empirical
+  comparisons.
+\newblock {\em International journal of forecasting}, 8:69--80, 1992.
+
+\bibitem{banerjee1993}
+A.~Banerjee.
+\newblock {\em Co-integration, Error Correction, and the Econometric Analysis
+  of Non-stationary Data}.
+\newblock Advanced texts in econometrics. Oxford University Press, 1993.
+
+\bibitem{Dukascopy2014}
+Dukascopy.
+\newblock Dukascopy historical data feed, 2014.
+
+\bibitem{duy1998}
+Timothy~A Duy and Mark~A Thoma.
+\newblock Modeling and forecasting cointegrated variables: some practical
+  experience.
+\newblock {\em Journal of Economics and Business}, 50(3):291--307, 1998.
+
+\bibitem{engle87}
+Robert~F. Engle and C.~W.~J. Granger.
+\newblock {Co-Integration and Error Correction: Representation, Estimation, and
+  Testing}.
+\newblock {\em Econometrica}, 55(2):251--276, March 1987.
+
+\bibitem{engle2004}
+Robert~F Engle and Andrew~J Patton.
+\newblock Impacts of trades in an error-correction model of quote prices.
+\newblock {\em Journal of Financial Markets}, 7(1):1--25, 2004.
+
+\bibitem{golub1980}
+Gene~H Golub and Charles~F Van~Loan.
+\newblock An analysis of the total least squares problem.
+\newblock {\em SIAM Journal on Numerical Analysis}, 17(6):883--893, 1980.
+
+\bibitem{gregoryETal1996}
+Allan~W Gregory, James~M Nason, and David~G Watt.
+\newblock Testing for structural breaks in cointegrated relationships.
+\newblock {\em Journal of Econometrics}, 71(1):321--341, 1996.
+
+\bibitem{herlemont2003}
+Daniel Herlemont.
+\newblock Pairs trading, convergence trading, cointegration.
+\newblock {\em YATS Finances \& Techonologies}, 5, 2003.
+
+\bibitem{johansen1988}
+Soren Johansen.
+\newblock {Statistical analysis of cointegration vectors}.
+\newblock {\em Journal of Economic Dynamics and Control}, 12(2-3):231--254,
+  1988.
+
+\bibitem{johansen1995}
+Soren Johansen.
+\newblock {\em Likelihood-Based Inference in Cointegrated Vector Autoregressive
+  Models}.
+\newblock Oxford University Press, 1995.
+
+\bibitem{maysami2000}
+Ramin~Cooper Maysami and Tiong~Sim Koh.
+\newblock A vector error correction model of the singapore stock market.
+\newblock {\em International Review of Economics \& Finance}, 9(1):79--96,
+  2000.
+
+\bibitem{mukherjee1995}
+Tarun~K Mukherjee and Atsuyuki Naka.
+\newblock Dynamic relations between macroeconomic variables and the japanese
+  stock market: an application of a vector error correction model.
+\newblock {\em Journal of Financial Research}, 18(2):223--37, 1995.
+
+\bibitem{seabold2010}
+Skipper Seabold and Josef Perktold.
+\newblock Statsmodels: Econometric and statistical modeling with python.
+\newblock In {\em Proceedings of the 9th Python in Science Conference}, pages
+  57--61, 2010.
+
+\bibitem{seong2013}
+Byeongchan Seong, Sung~K Ahn, and Peter~A Zadrozny.
+\newblock Estimation of vector error correction models with mixed-frequency
+  data.
+\newblock {\em Journal of Time Series Analysis}, 34(2):194--205, 2013.
+
+\bibitem{sims1980}
+Christopher~A Sims.
+\newblock Macroeconomics and reality.
+\newblock {\em Econometrica: Journal of the Econometric Society}, pages 1--48,
+  1980.
+
+\end{thebibliography}
+\end{document}
+% end of file template.tex

### 2016-07-06 articles-conferences/iscef2016/src/background.tex
Final manuscript

@@
 
-\section{\uppercase{Background}}
+\section{Background}
 \label{sec:background}
@@
 \subsection{Integration and Cointegration}\label{sec:coint}\  
-Following Johansen \cite{johansen1995} we shall say that a stochastic process
+Following Johansen \citep{johansen1995} we shall say that a stochastic process
 $Y_t$ which satisfies $Y_t-E(Y_t) = \sum_{i=0}^\infty C_i\,\varepsilon_{t-i}$ is
@@
 $d$, and then we shall write $\mathbf{y}_t\sim I(d)$, whenever after $d$ times
-(discrete) differentiation an stationary process is obtained~\cite{banerjee1993};
-more precisely, whenever
+(discrete) differentiation an stationary process is
+obtained \citep{banerjee1993}; more precisely, whenever
 $(1-L)^d\,\mathbf{y}_t\sim\text{I(0)}$, where $L$ is the usual lag operator:
@@
 
-Note that this definition includes the scalar case as time series of
-vectors of dimension 1; in this scalar case we will write the time series in
-nonbold format.
+Note that this definition includes the scalar case as time series of vectors of
+dimension 1; in this scalar case we will write the time series in nonbold
+format.
 
@@
 
-Vector error correction model (VECM) describe the joint behaviour
-of a set of variables and can be derived from the simple Vector Autoregressive model (VAR) \cite{sims1980} model.
-The VAR($p$) model is a framework describing the behaviour of a
-set of $l$ endogenous and stationary variables as a linear combination of their last $p$
-values, where $l,p\in\mathbb{N}$. 
-In our case, each one of these $l$ variables is a scalar time series
-$y_{\lambda,t}$, $\lambda=1,\dots,l$, and we represent them all together
-at time $t$ by the the vector time series:
+Vector error correction model (VECM) describe the joint behaviour of a set of
+variables and can be derived from the simple Vector Autoregressive model (VAR)
+\citep{sims1980} model.  The VAR($p$) model is a framework describing the
+behaviour of a set of $l$ endogenous and stationary variables as a linear
+combination of their last $p$ values, where $l,p\in\mathbb{N}$.  In our case,
+each one of these $l$ variables is a scalar time series $y_{\lambda,t}$,
+$\lambda=1,\dots,l$, and we represent them all together at time $t$ by the
+vector time series:
 \begin{equation}
@@
 its own lagged values and the lags of the others variables in the system. The
-model with $p$ lags is formulated as the system of $N$:
+model with $p$ lags is formulated as the system:
 \begin{align}
@@
 
-However, the VAR model cannot be used with non-stationary variables. VECM \cite{engle87} is also a linear model and a special form of a VAR model for I(1) variables that are also
-cointegrated~\cite{banerjee1993}. If cointegration exists, variable differences
-are stationary and they introduce an error correction term which adjusts
-coefficients to bring the variables back to equilibrium.
+However, the VAR model cannot be used with non-stationary variables. VECM
+\citep{engle87} is also a linear model and a special form of a VAR model for I(1)
+variables that are also cointegrated \citep{banerjee1993}. If cointegration
+exists, variable differences are stationary and they introduce an error
+correction term which adjusts coefficients to bring the variables back to
+equilibrium.
 
@@
 The following well known properties of the matrix $\boldsymbol{\Omega}$
-\cite{johansen1995} will be useful in the sequel:
+\citep{johansen1995} will be useful in the sequel:
 \begin{itemize}
@@
 $\boldsymbol{\alpha}$ correspond with the adjusted vectors. 
-$\boldsymbol{\beta}$ is obtained by Johansen procedure~\cite{johansen1988},
+$\boldsymbol{\beta}$ is obtained by Johansen procedure~\citep{johansen1988},
 whereas $\boldsymbol{\alpha}$ has to be determined as a variable in the VECM.

### 2016-07-06 articles-conferences/iscef2016/src/intro.tex
Final manuscript

@@
 In finance, it is common to find variables with long-run equilibrium
-relationships. This is termed cointegration and reflects the idea that
-some sets of variables cannot wander too far from each other. Cointegration
-means that one or more linear combinations of these variables are stationary
-even though individually they are not. Some models,
-such as the Vector Error Correction (VECM) \cite{engle87}, take advantage of
-this property and describe the joint behaviour of several cointegrated
-variables. VECM introduces this long-run relationship among a set of cointegrated
-time series as an error correction term.
-These time series must be integrated of order 1, denoted I(1), i.e.
-they become stationary at their first differences.
-In finance, I(1) time series are very common and
-to introduce cointegration restrictions in models often improves forecasting \cite{duy1998}. Therefore, VECM has been widely adopted in
-financial applications \cite{mukherjee1995,seong2013,maysami2000,arestis2001} such as pair trading \cite{herlemont2003} or models with more than two variables \cite{mukherjee1995,engle2004}.
+relationships. This is termed cointegration and reflects the idea that some sets
+of variables cannot wander too far from each other. Cointegration means that one
+or more linear combinations of these variables are stationary even though
+individually they are not. Some models, such as the Vector Error Correction
+(VECM) \citep{engle87}, take advantage of this property and describe the joint
+behaviour of several cointegrated variables. VECM introduces this long-run
+relationship among a set of cointegrated time series as an error correction
+term. These time series must be integrated of order 1, denoted I(1), i.e. they
+become stationary at their first differences. In finance, I(1) time series are
+very common and to introduce cointegration restrictions in models often improves
+forecasting \citep{duy1998}. Therefore, VECM has been widely adopted in financial
+applications (\cite{mukherjee1995}, \cite{seong2013}, \cite{maysami2000} and
+\cite{arestis2001}). VECM has also been used in pair trading \citep{herlemont2003} or
+models with more than two variables (\cite{mukherjee1995} and \cite{engle2004}).
 
-VECM parameters are obtained using the ordinary least squares
-(OLS) method \cite{golub1980}. Since OLS involves many calculations, the parameter estimation is computationally expensive when the number of lagged values and
-data increases. Moreover, obtaining cointegration vectors 
-is also an expensive routine. This is a main limitation to use VECM with stream data with high frequency.
-Chen and Lung's BVECM \cite{chen2003} addresses the advantage of distributed processing over conventional rolling window processing.
+VECM parameters are obtained using the ordinary least squares (OLS) method
+\citep{golub1980}. Since OLS involves many calculations, the parameter estimation
+is computationally expensive when the number of lagged values and data
+increases. Moreover, obtaining cointegration vectors is also an expensive
+routine. This is a main limitation to use VECM with stream data with high
+frequency.  Chen and Lung's BVECM \citep{chen2003} addresses the advantage of
+distributed processing over conventional rolling window processing.
 
-Our aims were to study if a parallel version of VECM can be used with high frequency stream data.
-Our approach was to determine adaptively the number of observations and lags of VECM which
-maximise cointegration relations in the past. 
- This search is computationally
-expensive because the Johansen
-method \cite{johansen1995} is required to find cointegration vectors, which is of order $O(n^3)$. Therefore, our proposal, called Adaptively Vector Error Correction (AVECM), 
-is to parallelise this search in order to get new parameters before new data arrives.
-Model effectiveness is focused on out-of-sample forecast rather than in-sample
-fitting. This criterion allows AVECM prediction capability to be
-expressed rather than just explaining data history.
-The forecast capability of our method was measured using MSE and the Theil's $U$-statistic \cite{theil1966} widely used in economic forecast. Tests were run using four currency rates: Euro (EUR) to
-United States Dollar (USD) (EURUSD), British Pound (GBP) to USD (GBPUSD), USD to Swiss
-Franc (CHF) (USDCHF) and USD to Japanese Yen (JPY) (USDJPY) with a 10-seconds frequency.
+Our aims were to study if a parallel version of VECM can be used with high
+frequency stream data.  Our approach was to determine adaptively the number of
+observations and lags of VECM which maximise cointegration relations in the
+past.  This search is computationally expensive because the Johansen method
+\citep{johansen1995} is required to find cointegration vectors, which is of order
+$O(n^3)$. Therefore, our proposal, called Adaptively Vector Error Correction
+(AVECM), is to parallelise this search in order to get new parameters before new
+data arrives.  Model effectiveness is focused on out-of-sample forecast rather
+than in-sample fitting. This criterion allows AVECM prediction capability to be
+expressed rather than just explaining data history.  The forecast capability of
+our method was measured using MSE and the Theil's $U$-statistic \citep{theil1966}
+widely used in economic forecast. Tests were run using four currency rates: Euro
+(EUR) to United States Dollar (USD) (EURUSD), British Pound (GBP) to USD
+(GBPUSD), USD to Swiss Franc (CHF) (USDCHF) and USD to Japanese Yen (JPY)
+(USDJPY) with a 10-seconds frequency.
 

### 2016-07-06 articles-conferences/iscef2016/src/methodology.tex
back to normal
Final manuscript

@@
 Cointegration vectors can be found applying the Johansen method which uses a
-sample of the last historical data. However, VECM assumes cointegration vectors do not change in time.
-In fact, the long-run relationships between the time series might change due to
-several economic factors that can lead to structural breaks in the cointegration
-relationship \cite{gregoryETal1996}. 
-In order to show the number of cointegration vectors which depends on 
-the amount $L$ of historical data and the number of lags $p$ in the VECM, we used a grid search.
-We defined a grid of possible values for $L$ and $p$.
-$L$ goes throughout $[100,5000]$ with a step size of 300 and $p$  throughout $[1,5]$ with step size of 1.
-The first experiment consisted in determining the number of cointegration vectors for all combinations of $L$ and $p$.
-We used four forex rates: EURUSD, GBPUSD, USDCHF and USDJPY with 10-second frequency. Data started at 13:00 GMT of the 13th of August 2014,
-when the New York and London financial markets opened.
+sample of the last historical data. However, VECM assumes cointegration vectors
+do not change in time.  In fact, the long-run relationships between the time
+series might change due to several economic factors that can lead to structural
+breaks in the cointegration relationship \citep{gregoryETal1996}.  In order to
+show the number of cointegration vectors which depends on the amount $L$ of
+historical data and the number of lags $p$ in the VECM, we used a grid search.
+We defined a grid of possible values for $L$ and $p$. $L$ goes throughout
+$[2,14]$ hours with a step size of 4 hours and $p$  throughout $[1,5]$ with step
+size of 1. The first experiment consisted in determining the number of
+cointegration vectors for all combinations of $L$ and $p$.  We used four forex
+rates: EURUSD, GBPUSD, USDCHF and USDJPY with 10-second frequency. Data started
+at 13:00 GMT of the 13th of August 2014, when the New York and London financial
+markets opened.
 
-Figure~\ref{fig:hists} shows the distribution of the number of cointegration vectors 
-given by the Johansen method for different values of $L$ and $p=1$ . This procedure was carried out by a sliding window of historical data moving 1000 times.
+Figure~\ref{fig:hists} shows the distribution of the number of cointegration
+vectors given by the Johansen method for different values of $L=[2,6,10,14]$ and
+$p=1$ . This procedure was carried out by a sliding window of historical data
+moving 1000 times.
 
@@
   \centering
-  \includegraphics[width=0.9\textwidth]{img/histCointVectors-offset20520-p-1-freq-10s}
+  \includegraphics[width=0.9\textwidth]{img/Fig1}
   \caption{Histogram of the number of cointegration vectors using $p=1$. Four
-  possible values for $L$ are shown (100, 2000, 3000 and 4000).}
+  possible values for $L$ are shown (2, 6, 10 and 14) hours (1 hour = 360 data
+  points).}
   \label{fig:hists}
@@
 
-From section \ref{sec:background} we know that $r=0$ means no cointegration and $r=l$ (we are using four rates, so $l=4$) reveals that no process is I(1) but stationary.
-The interesting cases of cointegration are those where $r$ lies strictly
-between $0$ and $4$, i.e. $0<r<4$.
+From section \ref{sec:background} we know that $r=0$ means no cointegration and
+$r=l$ (we are using four rates, so $l=4$) reveals that no process is I(1) but
+stationary.  The interesting cases of cointegration are those where $r$ lies
+strictly between $0$ and $4$, i.e. $0<r<4$.
 
@@
 
-The corresponding PC values for $L = [100, 2000, 3000, 4000]$ and $p=1$ (figure  \ref{fig:hists}) are: $PC = [21.3, 98.7, 79.8,  65.2]$.
-For the same values of $L$ and $p=3$ (figure  \ref{fig:histsp3}) the percentage of cointegration is $PC = [18.4, 92.5, 68,  52.5]$.
-From these results we observed that the number of cointegration vectors changed with different values for $L$ and $p$ and the percentage of cointegration depends mainly on $L$.
-
-The goal of our next experiment was to find a relation between this ratio
-$PC$ and the performance measure MSE (see equation 
-\ref{eq:MSE}). $L$ was defined between $[100,4000]$ with step size 300 and $p$ takes values between $[1,5]$ with step size 1.
+The goal of our next experiment was to find a relation between this ratio $PC$
+and the performance measure MSE (see equation \ref{eq:MSE}). $L$ was defined
+between $[700,1500]$ data points with step size 100 and $p$ takes values between
+$[1,5]$ with step size 1.
 
 \begin{figure}[ht!]
-  %\vspace{-0.8cm}
   \centering
-  \includegraphics[width=0.8\textwidth]{img/MSE-offset20520-p-1-freq-10s}
+  \includegraphics[width=0.8\textwidth]{img/Fig2}
   \caption{MSE versus the percentage of cointegration considering 1000
-  iterations. Best $L$ found was 1600. Below MSE versus $L$ shows a rapidly decreasing behaviour, founding minimum at PC = 84.3\% }
+  iterations. Best $L$ found was 1060. Below MSE versus $L$ shows a rapidly
+  decreasing behaviour, founding minimum at PC = 80\% }
   \label{fig:cointvsmse}
@@
 
-Figure~\ref{fig:cointvsmse} shows the relation between MSE and $PC$ and $L$.
-We found that better cointegration percentage leads to better
-performance accuracy in terms of reducing MSE. We also found that MSE rapidly decreased with increasing $L$.
+Figure~\ref{fig:cointvsmse} shows the relation between MSE and $PC$ and $L$.  We
+found that better cointegration percentage leads to better performance accuracy
+in terms of reducing MSE.
 
-Therefore we proposed to choose $L$ and $p$ in order to maximise the percentage of
-cointegration $PC$ in the near past.
-This process was done every time that new data was processed.
-However, this search can be slow if we try different values for $L$ and $p$ and therefore we distributed this calculation in order to reduce searching time.
+Therefore we proposed to choose $L$ and $p$ in order to maximise the percentage
+of cointegration $PC$ in the near past. This process was done every time that
+new data was processed. However, this search can be slow if we try different
+values for $L$ and $p$ and therefore we distributed this calculation in order to
+reduce searching time.
 
-Our proposal was then a modified version of VECM, with parameter $p$ and the amount of historical data used $L$ obtained at every step.
-Only the search of $L$ and $p$ was done in a distributed environment, since this is the most expensive routine.
-Our proposal is called Adaptive Vector Error Correction model (AVECM).
-AVECM is detailed in the algorithm ~\ref{alg:AVECM} which summarises our proposal.
+Our proposal was then a modified version of VECM, with parameter $p$ and the
+amount of historical data used $L$ obtained at every step. Only the search of
+$L$ and $p$ was done in a distributed environment, since this is the most
+expensive routine. Our proposal is called Adaptive Vector Error Correction model
+(AVECM). AVECM is detailed in the algorithm \ref{alg:AVECM} which summarises
+our proposal.
 
@@
 
-The input of AVECM is time series prices which are cointegrated.
-The starting point of testing is $j$ and the total number of iterations is $it$.
-We need to ensure that $j$ is at least the maximum value of $Ls$.
-$Ls$ and $ps$ are the possible values for $L$ and $p$.
-The function {\bf get\_best\_params} makes this grid search on the two vector
-lists $Ls$ and $ps$ and returns the parameters $L$ and $p$ which maximise
-the percentage of cointegration $PC$ (see equation~\ref{eq:pcoint}) for a
-pre-defined number of iterations $m$. This function is implemented in a distributed 
-environment, thus ensuring a response
-before new data is available. 
-After $L$ and $p$ parameters are found, VECM is built and used
-to forecast the next data point.
+The input of AVECM is time series prices which are cointegrated. The starting
+point of testing is $j$ and the total number of iterations is $it$. We need to
+ensure that $j$ is at least the maximum value of $Ls$. $Ls$ and $ps$ are the
+possible values for $L$ and $p$.  The function {\bf get\_best\_params} makes
+this grid search on the two vector lists $Ls$ and $ps$ and returns the
+parameters $L$ and $p$ which maximise the percentage of cointegration $PC$ (see
+equation~\ref{eq:pcoint}) for a pre-defined number of iterations $m$. This
+function is implemented in a distributed environment, thus ensuring a response
+before new data is available.  After $L$ and $p$ parameters are found, VECM is
+built and used to forecast the next data point.
 
 \subsection{Model comparison} \label{sec:random}
-We compared our proposal, in terms of performance, with the naive forecast of the random walk model and ARIMA.
-It is still difficult to outperform the random walk model for standard econometric forecasting models \cite{lo2011} despite its simplicity.
-The random walk model is defined as:
+We compared our proposal, in terms of performance, with the naive forecast of
+the random walk model and ARIMA. It is still difficult to outperform the random
+walk model for standard econometric forecasting models \citep{lo2011} despite its
+simplicity. The random walk model is defined as:
 
@@
 
-The naive forecast of the time series difference $\hat{\mathbf{y}}_{t+1}$ for the random walk model is defined as:
+The naive forecast of the time series difference $\hat{\mathbf{y}}_{t+1}$ for
+the random walk model is defined as:
 \begin{equation}
@@
 
-On the other hand, ARIMA is widely used to forecast returns in finance \cite{tsay2005}.
-A process can be modelled as an ARIMA$(p,d,q)$ model if $\mathbf{x}_t=\Delta^d \mathbf{y}_t $, i.e after differencing $d$ times the time series $\mathbf{y}_t$,  we get an ARMA$(p,q)$.
-Since we are modelling returns, we used $d=1$.
+On the other hand, ARIMA is widely used to forecast returns in finance
+\citep{tsay2005}. A process can be modelled as an ARIMA$(p,d,q)$ model if
+$\mathbf{x}_t=\Delta^d \mathbf{y}_t $, i.e after differencing $d$ times the time
+series $\mathbf{y}_t$,  we get an ARMA$(p,q)$. Since we are modelling returns,
+we used $d=1$.
 
@@
 
-Forecast performance was evaluated using two different methods which are frequently used in finance:
+Forecast performance was evaluated using two different methods which are
+frequently used in finance:
 \begin{description}
@@
 \end{equation}
-\item {\bf $U$-statistic}, the Theil's $U$-statistic \cite{theil1966} is a unit free measure obtained as the ratio between the root MSE (RMSE) of a model and the RMSE of the naive random walk model.
-A $U$-statistic less than 1 implies the performance is better than the naive model.
+\item {\bf $U$-statistic}, the Theil's $U$-statistic \citep{theil1966} is a unit
+free measure obtained as the ratio between the root MSE (RMSE) of a model and
+the RMSE of the naive random walk model. A $U$-statistic less than 1 implies the
+performance is better than the naive model.
 \end{description}

### 2016-07-06 articles-conferences/iscef2016/src/results.tex
Final manuscript

@@
-\section{\uppercase{Experimental results}}
+\section{Experimental results}
 \label{sec:results}
@@
 AVECM tests were carried out using four foreign exchange rates all related to
-USD: EURUSD, GBPUSD, USDCHF and USDJPY. This data was collected from Dukascopy~\cite{Dukascopy2014}, a free
-database which gives access to the Swiss Foreign Exchange marketplace.
+USD: EURUSD, GBPUSD, USDCHF and USDJPY. This data was collected from
+Dukascopy \citep{Dukascopy2014}, a free database which gives access to the Swiss
+Foreign Exchange marketplace.
 
-The tests were done using 10-seconds frequency from ask prices
-from the 11th to the 15th of August 2014. Since one day corresponds to 8640 data points and we used 5 days of data, we have 43,200 data points in total.
+The tests were done using 10-seconds frequency from ask prices from the 11th to
+the 15th of August 2014. Since one day corresponds to 8640 data points and we
+used 5 days of data, we have 43,200 data points in total.
 
@@
 
-Before running the tests, we firstly checked whether the time series were
-I(1) using the Augmented Dickey Fuller (ADF) test with lags $p=1,2,3,4,5$. 
-MacKinnon \cite{mackinnon2010} critical values for rejection of hypothesis of a unit root are: -2.62 (1\%), 1.94 (5\%) and 1.62 (10\%).
+Before running the tests, we firstly checked whether the time series were I(1)
+using the Augmented Dickey Fuller (ADF) test with lags $p=1,2,3,4,5$.  MacKinnon
+\citep{mackinnon2010} critical values for rejection of hypothesis of a unit root
+are: -2.62 (1\%), 1.94 (5\%) and 1.62 (10\%).
 \begin{table*}[ht]
@@
 \end{table*}
-Table~\ref{tab:adf} shows that all currency rates cannot reject the unit root test in their level form considering different lags, but they rejected it with their first differences.
-This means that all of
-them are I(1) time series and we are allowed to use VECM and therefore our proposed AVECM.
+Table~\ref{tab:adf} shows that all currency rates cannot reject the unit root
+test in their level form considering different lags, but they rejected it with
+their first differences. This means that all of them are I(1) time series and we
+are allowed to use VECM and therefore our proposed AVECM.
 
@@
 
-Algorithms AVECM, ARIMA and the naive random walk were tested using four days of data (from the 12th to the 15th of August 2014).
-For AVECM we considered different number of iterations (parameter $m$ in algorithm \ref{alg:AVECM}): 10, 50 and 100.
-We tried 12, 24 and 47 different pair of combinations for $L$ and $p$.
-Possible values for $L$ were in the interval $[100,4000]$ and $p$ can have values in between $[1,5]$.
-Best AVECM performance was compared against ARIMA and the random walk model.
-Table~\ref{tab:stats} shows the out-of-sample performance measures: MSE and $U$-statistic for AVECM and ARIMA.
-In terms of both measures we found that AVECM is superior to ARIMA and the naive random walk model.
-We also included the p-value that proves that the difference in the MSE is significant at the 99\% significance level in three of the four currency rates and at 90\% in the case of GBPUSD.
-The $U$-statistic shows that AVECM and ARIMA are superior to the naive random walk model and that our proposal is also superior to ARIMA.
+Algorithms AVECM, ARIMA and the naive random walk were tested using four days of
+data (from the 12th to the 15th of August 2014).  For AVECM we considered
+different number of iterations (parameter $m$ in algorithm \ref{alg:AVECM}): 10,
+50 and 100. We tried 12, 24 and 47 different pair of combinations for $L$ and
+$p$. Possible values for $L$ were in the interval $[2,14]$ hours and $p$ can
+have values in between $[1,5]$. Best AVECM performance was compared against
+ARIMA and the random walk model.  Table~\ref{tab:stats} shows the out-of-sample
+performance measures: MSE and $U$-statistic for AVECM and ARIMA. In terms of
+both measures we found that AVECM is superior to ARIMA and the naive random walk
+model. We also included the p-value that proves that the difference in the MSE
+is significant at the 99\% significance level in three of the four currency
+rates and at 90\% in the case of GBPUSD. The $U$-statistic shows that AVECM and
+ARIMA are superior to the naive random walk model and that our proposal is also
+superior to ARIMA.
 
@@
 
-To determine $L$ and $p$ based on maximising the percentage of cointegration requires use of the 
-Johansen
-method which is a computationally expensive routine. This procedure is done by the function get\_best\_params($Ls,ps,m,\mathbf{Y}$) shown in algorithm \ref{alg:AVECM}.
-In order to improve the execution time of this search, our proposal included
-a parallel search of VECM parameters using high performance computing.
-The main objective was to obtain a response before a new data arrived in
-the following10 seconds.
+To determine $L$ and $p$ based on maximising the percentage of cointegration
+requires use of the Johansen method which is a computationally expensive
+routine. This procedure is done by the function
+get\_best\_params($Ls,ps,m,\mathbf{Y}$) shown in algorithm \ref{alg:AVECM}.  In
+order to improve the execution time of this search, our proposal included a
+parallel search of VECM parameters using high performance computing.  The main
+objective was to obtain a response before a new data arrived in the following10
+seconds.
 
 The Johansen method is already programmed in the Python Statsmodels
-library~\cite{seabold2010} and the parallel implementation was developed using
-MPI in Python. 
-We chose MPI because it allows large-scale parallel applications
-with wide portability to be built, being able to run in large clusters
-or on local computers.
-We tested our proposal in a cluster with 2 servers Xeon E5-2667 (2.90GHz)
-of 24 cores each (48 cores in total) and 24GB RAM.
-In order to compare serial and parallel execution times in AVECM, we set parameter $it=100$ in algorithm\ref{alg:AVECM}.
+library \citep{seabold2010} and the parallel implementation was developed using
+MPI in Python.  We chose MPI because it allows large-scale parallel applications
+with wide portability to be built, being able to run in large clusters or on
+local computers.  We tested our proposal in a cluster with 2 servers Xeon
+E5-2667 (2.90GHz) of 24 cores each (48 cores in total) and 24GB RAM.  In order
+to compare serial and parallel execution times in AVECM, we set parameter
+$it=100$ in algorithm\ref{alg:AVECM}.
 
-The $L$ parameter was always chosen between 100 and 4000 and $p$ always took values
-between 1 and 5. Parameter $nparams$ represents the number of pairs ($L$,$p$) used to maximise the percentage of cointegration.
+The $L$ parameter was always chosen between 2 and 14 hours and $p$ always took
+values between 1 and 5. Parameter $nparams$ represents the number of pairs
+($L$,$p$) used to maximise the percentage of cointegration.
 
-Execution time depends directly on $L$, $p$ and $nparams$ used, since they determine the size of
-matrix $\mathbf{A}$ (see equation \ref{eq:Amatrix}) and therefore affect the OLS function execution time.
-Therefore, if we try more combinations of $L$ and $p$ (increasing $nparams$) the
-serial algorithm will take longer. For this financial time series we are interested in execution times below 10 seconds  (the time series frequency).
+Execution time depends directly on $L$, $p$ and $nparams$ used, since they
+determine the size of matrix $\mathbf{A}$ (see equation \ref{eq:Amatrix}) and
+therefore affect the OLS function execution time.  Therefore, if we try more
+combinations of $L$ and $p$ (increasing $nparams$) the serial algorithm will
+take longer. For this financial time series we are interested in execution times
+below 10 seconds  (the time series frequency).
 
@@
   \centering
-  \includegraphics[width=0.75\textwidth]{img/extimes}
+  \includegraphics[width=0.75\textwidth]{img/Fig3}
   \caption{Computing time of sequential and parallel algorithm is shown in the
@@
 
-The superior figure \ref{fig:extimes} shows that best performance accuracy measurements are
-achieved in times near or below 10 seconds in the parallel version. Contrarily,
-serial times are high, above 10 seconds in most cases. Figure~\ref{fig:extimes} also shows the speed-up in computing time for AVECM which is near 9X if we use more than 25 parameters ($nparams$).
+The superior figure \ref{fig:extimes} shows that best performance accuracy
+measurements are achieved in times near or below 10 seconds in the parallel
+version. Contrarily, serial times are high, above 10 seconds in most cases.
+Figure~\ref{fig:extimes} also shows the speed-up in computing time for AVECM
+which is near 9X if we use more than 25 parameters ($nparams$).
 
-Execution times do not consider the loading data time, just that of finding
-best parameters and matrix operations. The time MPI spends
-transferring data and synchronising processes is about two seconds independently
-of the number of processes considered. Execution times were measured using the Time python library.
+Execution times do not consider the loading data time, just that of finding best
+parameters and matrix operations. The time MPI spends transferring data and
+synchronising processes is about two seconds independently of the number of
+processes considered. Execution times were measured using the Time python
+library.

### 2016-07-06 articles-conferences/iscef2016/src/background.tex
manuscript updates

@@
-\def\rot#1{{\color{red}{#1}}}
-\def\gruen#1{{\color{green}{#1}}}
-\def\blau#1{{\color{blue}{#1}}}
-
 \section{Background}
@@
 \subsection{Integration and Cointegration}\label{sec:coint}\  
-Following Johansen \citep{johansen1995} we shall say that a stochastic process
+Following \cite{johansen1995} we shall say that a stochastic process
 $Y_t$ which satisfies $Y_t-E(Y_t) = \sum_{i=0}^\infty C_i\,\varepsilon_{t-i}$ is
@@
 (discrete) differentiation an stationary process is
-obtained \citep{banerjee1993}; more precisely, whenever
+obtained, see \cite{banerjee1993}; more precisely, whenever
 $(1-L)^d\,\mathbf{y}_t\sim\text{I(0)}$, where $L$ is the usual lag operator:
@@
 variables and can be derived from the simple Vector Autoregressive model (VAR)
-\citep{sims1980} model.  The VAR($p$) model is a framework describing the
+presented by \cite{sims1980}.  The VAR($p$) model is a framework describing the
 behaviour of a set of $l$ endogenous and stationary variables as a linear
@@
 
-However, the VAR model cannot be used with non-stationary variables. VECM
-\citep{engle87} is also a linear model and a special form of a VAR model for I(1)
-variables that are also cointegrated \citep{banerjee1993}. If cointegration
-exists, variable differences are stationary and they introduce an error
-correction term which adjusts coefficients to bring the variables back to
-equilibrium.
+However, the VAR model cannot be used with non-stationary variables. VECM ,
+developed by \cite{engle87}, is also a linear model for I(1) variables that are
+also cointegrated, see \cite{banerjee1993}. If cointegration exists, variable
+differences are stationary and they introduce an error correction term which
+adjusts coefficients to bring the variables back to equilibrium.
 
@@
 \end{align*}
-The following well known properties of the matrix $\boldsymbol{\Omega}$
-\citep{johansen1995} will be useful in the sequel:
+The following well known properties of the matrix $\boldsymbol{\Omega}$, see
+\cite{johansen1995}, will be useful in the sequel:
 \begin{itemize}
@@
 $\boldsymbol{\alpha}$ correspond with the adjusted vectors. 
-$\boldsymbol{\beta}$ is obtained by Johansen procedure~\citep{johansen1988},
+$\boldsymbol{\beta}$ is obtained by Johansen procedure,see \cite{johansen1988},
 whereas $\boldsymbol{\alpha}$ has to be determined as a variable in the VECM.
@@
 Taking into account the error term $\mathbf{E}$, equation~(\ref{eq:vareq}) 
-can be solved with respect to $\mathbf{X}$ using the ordinary least
-squares (OLS) estimation.
+can be solved with respect to $\mathbf{X}$ using the OLS estimation.

### 2016-07-06 articles-conferences/iscef2016/src/conclusions.tex
Final manuscript
manuscript updates

@@
-\section{\uppercase{Conclusions}}
+\section{Conclusions}
 \label{sec:conclusions}
-Cointegration in financial time series has been largely studied and the 
-Johansen method is commonly used to obtain cointegration relationships. 
-In practice, it has been found that cointegration relations change with time. 
-However, model-based cointegration such as VECM assumes that cointegration
-remains unchanged in time. 
-We empirically showed that the Johansen method is sensitive to the number
-of lags but also to the amount of data considered.
+Cointegration in financial time series has been largely studied and the Johansen
+method is commonly used to obtain cointegration relationships.  In practice, it
+has been found that cointegration relations change with time.  However,
+model-based cointegration such as VECM assumes that cointegration remains
+unchanged in time.  We empirically showed that the Johansen method is sensitive
+to the number of lags but also to the amount of data considered.
 
@@
 
-Despite the fact that high frequency Forex data can be spurious, the model performance can be less reliable (and more spurious) relative to the lower frequencies (such as 1 minute or 5 minute intervals) adopted by some other studies.
-However, the deficiency is offset by gain in accuracy from parallel processing which is capable of searching or examining a much larger state space given the same computational time.
+Despite the fact that high frequency Forex data can be spurious, the model
+performance can be less reliable (and more spurious) relative to the lower
+frequencies (such as 1 minute or 5 minute intervals) adopted by some other
+studies. However, the deficiency is offset by gain in accuracy from parallel
+processing which is capable of searching or examining a much larger state space
+given the same computational time.
 
-Determining VECM parameters was the most expensive routine and it was run
-using parallel processes using MPI which allowed a grid search within a range of
-values for $L$ and $p$ to be made.
-Tests were done using real currency rates data. 
-Our proposal was tested in several scenarios using different times of
-the day to show its behaviour related to the opening times of
-financial markets.
+Determining VECM parameters was the most expensive routine and it was run using
+parallel processes using MPI which allowed a grid search within a range of
+values for $L$ and $p$ to be made.  Tests were done using real currency rates
+data.
 
@@
 parameters of $L$ and $p$ maximising the percentage of cointegration.
-These facts were consistent for different times of the day. 
-We found that increasing the number of parameters will always lead to
-better performance measures.
 
-The parallel implementation allowed the execution times to be reduced
-more than 9 times and therefore a response time was obtain before 10
-seconds. Since we used 10-second frequencies we can say that our proposal is suitable
-for use in an online context for real applications because response times
-were less than this frequency.
+The parallel implementation allowed the execution times to be reduced more than
+9 times and therefore a response time was obtain before 10 seconds. Since we
+used 10-second frequencys we can say that our proposal is suitable for use in
+an online context for real applications because response times were less than
+this frequency.
 
 For future study, it would be interesting to explore the relationship between
-cointegration and performance in order to propose new criteria for
-improving VECM parameters. It would also be interesting to include more 
-explaining variables such as bid-ask spread and change in volume.
+cointegration and performance in order to propose new criteria for improving
+VECM parameters. It would also be interesting to include more explaining
+variables such as bid-ask spread and change in volume.

### 2016-07-06 articles-conferences/iscef2016/src/intro.tex
manuscript updates

--- 2016-04-25 articles-conferences/iscef2016/src/intro.tex
@@
 In finance, it is common to find variables with long-run equilibrium
-relationships. This is termed cointegration and reflects the idea that
-some sets of variables cannot wander too far from each other. Cointegration
-means that one or more linear combinations of these variables are stationary
-even though individually they are not. Some models,
-such as the Vector Error Correction (VECM) \cite{engle87}, take advantage of
-this property and describe the joint behaviour of several cointegrated
-variables. VECM introduces this long-run relationship among a set of cointegrated
-time series as an error correction term.
-These time series must be integrated of order 1, denoted I(1), i.e.
-they become stationary at their first differences.
-In finance, I(1) time series are very common and
-to introduce cointegration restrictions in models often improves forecasting \cite{duy1998}. Therefore, VECM has been widely adopted in
-financial applications \cite{mukherjee1995,seong2013,maysami2000,arestis2001} such as pair trading \cite{herlemont2003} or models with more than two variables \cite{mukherjee1995,engle2004}.
+relationships. This is termed cointegration and reflects the idea that some sets
+of variables cannot wander too far from each other. Cointegration means that one
+or more linear combinations of these variables are stationary even though
+individually they are not. Some models, such as the Vector Error Correction
+(VECM), see \cite{engle87}, take advantage of this property and describe the
+joint behaviour of several cointegrated variables. VECM introduces this long-run
+relationship among a set of cointegrated time series as an error correction
+term. These time series must be integrated of order 1, denoted I(1), i.e. they
+become stationary at their first differences. In finance, I(1) time series are
+very common and to introduce cointegration restrictions in models often improves
+forecasting, see \cite{duy1998}. Therefore, VECM has been widely adopted in
+financial applications, among others: \cite{mukherjee1995}, \cite{seong2013},
+\cite{maysami2000} and \cite{arestis2001}. VECM has also been used in pair
+trading, see \cite{herlemont2003}, or models with more than two variables, see
+for example \cite{mukherjee1995} and \cite{engle2004}).
 
-VECM parameters are obtained using the ordinary least squares
-(OLS) method \cite{golub1980}. Since OLS involves many calculations, the parameter estimation is computationally expensive when the number of lagged values and
-data increases. Moreover, obtaining cointegration vectors 
-is also an expensive routine. This is a main limitation to use VECM with stream data with high frequency.
-Chen and Lung's BVECM \cite{chen2003} addresses the advantage of distributed processing over conventional rolling window processing.
+VECM parameters are obtained using the ordinary least squares (OLS) method,
+developed by \cite{golub1980}. Since OLS involves many calculations, the
+parameter estimation is computationally expensive when the number of lagged
+values and data increases. Moreover, obtaining cointegration vectors is also an
+expensive routine. This is a main limitation to use VECM with stream data with
+high frequency. \cite{chen2003} addresses the advantage of distributed
+processing over conventional rolling window processing.
 
-Our aims were to study if a parallel version of VECM can be used with high frequency stream data.
-Our approach was to determine adaptively the number of observations and lags of VECM which
-maximise cointegration relations in the past. 
- This search is computationally
-expensive because the Johansen
-method \cite{johansen1995} is required to find cointegration vectors, which is of order $O(n^3)$. Therefore, our proposal, called Adaptively Vector Error Correction (AVECM), 
-is to parallelise this search in order to get new parameters before new data arrives.
-Model effectiveness is focused on out-of-sample forecast rather than in-sample
-fitting. This criterion allows AVECM prediction capability to be
-expressed rather than just explaining data history.
-The forecast capability of our method was measured using MSE and the Theil's $U$-statistic \cite{theil1966} widely used in economic forecast. Tests were run using four currency rates: Euro (EUR) to
-United States Dollar (USD) (EURUSD), British Pound (GBP) to USD (GBPUSD), USD to Swiss
-Franc (CHF) (USDCHF) and USD to Japanese Yen (JPY) (USDJPY) with a 10-seconds frequency.
+Our aims were to study if a parallel version of VECM can be used with high
+frequency stream data.  Our approach was to determine adaptively the number of
+observations and lags of VECM which maximise cointegration relations in the
+past.  This search is computationally expensive because the Johansen method, see
+\cite{johansen1995}, is required to find cointegration vectors, which is of
+order $O(n^3)$. Therefore, our proposal, called Adaptively Vector Error
+Correction (AVECM), is to parallelise this search in order to get new parameters
+before new data arrives.  Model effectiveness is focused on out-of-sample
+forecast rather than in-sample fitting. This criterion allows AVECM prediction
+capability to be expressed rather than just explaining data history.  The
+forecast capability of our method was measured using MSE and the Theil's
+$U$-statistic, see \cite{theil1966}, widely used in economic forecast. Tests
+were run using four currency rates: Euro (EUR) to United States Dollar (USD)
+(EURUSD), British Pound (GBP) to USD (GBPUSD), USD to Swiss Franc (CHF) (USDCHF)
+and USD to Japanese Yen (JPY) (USDJPY) with a 10-second frequency.
 

### 2016-07-06 articles-conferences/iscef2016/src/methodology.tex
manuscript updates

@@
-\section{\uppercase{Methodology}} \label{sec:methodology}
+\section{Methodology} \label{sec:methodology}
 \subsection{Motivation} \label{sec:proposal}
@@
 sample of the last historical data. However, VECM assumes cointegration vectors
-do not change in time.  In fact, the long-run relationships between the time
-series might change due to several economic factors that can lead to structural
-breaks in the cointegration relationship \citep{gregoryETal1996}.  In order to
-show the number of cointegration vectors which depends on the amount $L$ of
-historical data and the number of lags $p$ in the VECM, we used a grid search.
-We defined a grid of possible values for $L$ and $p$. $L$ goes throughout
-$[2,14]$ hours with a step size of 4 hours and $p$  throughout $[1,5]$ with step
-size of 1. The first experiment consisted in determining the number of
-cointegration vectors for all combinations of $L$ and $p$.  We used four forex
-rates: EURUSD, GBPUSD, USDCHF and USDJPY with 10-second frequency. Data started
-at 13:00 GMT of the 13th of August 2014, when the New York and London financial
-markets opened.
+do not change in time.  In fact, \cite{gregoryETal1996} addresses that the
+long-run relationships between the time series  might change due to several
+economic factors that can lead to structural breaks in the cointegration
+relationship.  In order to show that the number of cointegration vectors depends
+on the amount $L$ of historical data and the number of lags $p$ in the VECM, we
+used a grid search.  We defined a grid of possible values for $L$ and $p$. $L$
+goes throughout $[2,14]$ hours (1 $hour = 360$ data points) with a step size of
+4 hours and $p$ throughout $[1,5]$ with step size of 1. The first experiment
+consisted in determining the number of cointegration vectors for all
+combinations of $L$ and $p$.  We used four forex rates: EURUSD, GBPUSD, USDCHF
+and USDJPY with 10-second frequency. Data started at 13:00 GMT of the 13th of
+August 2014, when the New York and London financial markets opened.
 
-Figure~\ref{fig:hists} shows the distribution of the number of cointegration
-vectors given by the Johansen method for different values of $L=[2,6,10,14]$ and
-$p=1$ . This procedure was carried out by a sliding window of historical data
-moving 1000 times.
+Figure \ref{fig:hists} shows the distribution of the number of cointegration
+vectors given by the Johansen method for different values of $L=[2,6,10,14]$
+hours  and $p=1$ . This procedure was carried out by a sliding window of
+historical data moving 1000 times.
 
@@
   \centering
-  \includegraphics[width=0.9\textwidth]{img/Fig1}
-  \caption{Histogram of the number of cointegration vectors using $p=1$. Four
-  possible values for $L$ are shown (2, 6, 10 and 14) hours (1 hour = 360 data
-  points).}
+  \includegraphics[width=0.9\textwidth]{Fig1}
+  \caption{Distribution of the number of cointegration vectors using $p=1$ lags.
+  Four possible values for windows size $L$ are shown (2, 6, 10 and 14) hours (1
+  hour = 360 data points).}
   \label{fig:hists}
@@
 and the performance measure MSE (see equation \ref{eq:MSE}). $L$ was defined
-between $[700,1500]$ data points with step size 100 and $p$ takes values between
+between $[700,1500]$ data points with step size 100 and $p$ took values between
 $[1,5]$ with step size 1.
@@
   \centering
-  \includegraphics[width=0.8\textwidth]{img/Fig2}
+  \includegraphics[width=0.9\textwidth]{Fig2}
   \caption{MSE versus the percentage of cointegration considering 1000
-  iterations. Best $L$ found was 1060. Below MSE versus $L$ shows a rapidly
-  decreasing behaviour, founding minimum at PC = 80\% }
+  iterations. Best windows size $L$ found was 1060. Below MSE versus $L$ shows a
+  rapidly decreasing behaviour, founding minimum at PC = 80\% }
   \label{fig:cointvsmse}
@@
 
-Our proposal was then a modified version of VECM, with parameter $p$ and the
+Our proposal is then a modified version of VECM, with parameter $p$ and the
 amount of historical data used $L$ obtained at every step. Only the search of
@@
 the random walk model and ARIMA. It is still difficult to outperform the random
-walk model for standard econometric forecasting models \citep{lo2011} despite its
-simplicity. The random walk model is defined as:
+walk model for standard econometric forecasting models  despite its
+simplicity, see \cite{lo2011}. The random walk model is defined as:
 
@@
 
-On the other hand, ARIMA is widely used to forecast returns in finance
-\citep{tsay2005}. A process can be modelled as an ARIMA$(p,d,q)$ model if
+On the other hand, ARIMA is widely used to forecast returns in finance, see
+\cite{tsay2005}. A process can be modelled as an ARIMA$(p,d,q)$ model if
 $\mathbf{x}_t=\Delta^d \mathbf{y}_t $, i.e after differencing $d$ times the time
@@
 \end{equation}
-\item {\bf $U$-statistic}, the Theil's $U$-statistic \citep{theil1966} is a unit
-free measure obtained as the ratio between the root MSE (RMSE) of a model and
-the RMSE of the naive random walk model. A $U$-statistic less than 1 implies the
-performance is better than the naive model.
+\item {\bf $U$-statistic}, the Theil's $U$-statistic, presented by
+\cite{theil1966}, is a unit free measure obtained as the ratio between the root
+MSE (RMSE) of a model and the RMSE of the naive random walk model. A
+$U$-statistic less than 1 implies the performance is better than the naive
+model.
 \end{description}

### 2016-07-06 articles-conferences/iscef2016/src/results.tex
manuscript updates

@@
 USD: EURUSD, GBPUSD, USDCHF and USDJPY. This data was collected from
-Dukascopy \citep{Dukascopy2014}, a free database which gives access to the Swiss
-Foreign Exchange marketplace.
+\cite{Dukascopy2014}, a free database which gives access to the Swiss Foreign
+Exchange marketplace.
 
-The tests were done using 10-seconds frequency from ask prices from the 11th to
+The tests were done using 10-second frequency from ask prices from the 11th to
 the 15th of August 2014. Since one day corresponds to 8640 data points and we
@@
 
-\subsection{Unit root tests} \label{sec:unitroot}
+\subsection{Unit root tests} \label{sec:unitroottests}
 
 Before running the tests, we firstly checked whether the time series were I(1)
-using the Augmented Dickey Fuller (ADF) test with lags $p=1,2,3,4,5$.  MacKinnon
-\citep{mackinnon2010} critical values for rejection of hypothesis of a unit root
-are: -2.62 (1\%), 1.94 (5\%) and 1.62 (10\%).
+using the Augmented Dickey Fuller (ADF) test with lags $p=1,2,3,4,5$.
+\cite{mackinnon2010} presented critical values for rejection of hypothesis of a
+unit root: -2.62 (1\%), 1.94 (5\%) and 1.62 (10\%).
 \begin{table*}[ht]
@@
 \end{tabular}
-\caption{Unit roots tests for EURUSD, GBPUSD, USDCHF and USDJPY at 10-seconds
+\caption{Unit roots tests for EURUSD, GBPUSD, USDCHF and USDJPY at 10-second
 frequency.}
@@
 parallel search of VECM parameters using high performance computing.  The main
-objective was to obtain a response before a new data arrived in the following10
+objective was to obtain a response before a new data arrived in the following 10
 seconds.
@@
 The Johansen method is already programmed in the Python Statsmodels
-library \citep{seabold2010} and the parallel implementation was developed using
+library, see \cite{seabold2010}, and the parallel implementation was developed using
 MPI in Python.  We chose MPI because it allows large-scale parallel applications
@@
 to compare serial and parallel execution times in AVECM, we set parameter
-$it=100$ in algorithm\ref{alg:AVECM}.
+$it=100$ in algorithm \ref{alg:AVECM}.
 
@@
   \centering
-  \includegraphics[width=0.75\textwidth]{img/Fig3}
+  \includegraphics[width=0.9\textwidth]{Fig3}
   \caption{Computing time of sequential and parallel algorithm is shown in the
@@
 
-The superior figure \ref{fig:extimes} shows that best performance accuracy
+The superior figure in \ref{fig:extimes} shows that best performance accuracy
 measurements are achieved in times near or below 10 seconds in the parallel
-version. Contrarily, serial times are high, above 10 seconds in most cases.
+version. Contrarily, serial times are higher, above 10 seconds in most cases.
 Figure~\ref{fig:extimes} also shows the speed-up in computing time for AVECM

### 2016-07-06 src/4_OnlineLearning.tex
chapter 3 ready for revision

@@
 \chapter{Machine Learning Models}
-Machine learning is a scientific discipline focused on the development of algorithms focused on learning from examples.
-This idea has become central to the design of search engines, robots systems and forecasts applications which process large data sets.
-However, machines require an extended training period when developing algorithms to predict future behaviour such as financial time series forecasting where new data is available in a short period of time.
-Online machine learning techniques tackle this problem and allow to update the model with new data or to compute a new model using less data allowing to give a response in a short period of time.
+Machine learning is a scientific discipline focused on the development of algorithms based on learning from examples.
+This idea has become central to the design of search engines, robots systems and forecasts applications that process large data sets.
+Usually machines require long training period when developing algorithms to predict future behaviour such as financial time series forecasting where new data is available in a short period of time.
+Online machine learning techniques tackle this problem by updating the model with new data or computing a new model using less data so they  can give a response in a short period of time.
 
@@
 The learning is always based on samples and the objective is about to do better in the future based on the past experiences in an automatic way.
-ML is a subarea of artificial intelligence and broadly intersects with other fields such as statistics, mathematics, physics and computer science.
+ML is a sub-area of artificial intelligence and broadly intersects with other fields such as statistics, mathematics, physics and computer science.
 There are many examples of machine learning problems: time series forecasting, image processing, face detection, spam filtering, weather prediction, search engines, among many others.
@@
 
-There are three types of ML classified depending on the nature of the learning input or output available to a learning system.
+There are three types of ML classified depending on the nature of the learning input or output available to a learning system:
+
 \begin{description}
-\item[supervised learning]  the input data is a tuple which contains the example input and its desired outputs (also called labels).
+\item[Supervised learning]  the input data is a tuple which contains the example input and its desired outputs (also called labels).
 The list of tuples is called the training set.
@@
 Its two mainstream approaches are: support vector machines (SVMs) \cite{vapnik1998} and ensemble learning \cite{breiman1998}.
-Furthermore supervised learning can be categorised into offline or batch learning and online learning (see section \ref{sec:onoffline}).
-\item[unsupervised learning] also known as clustering \cite{ben2005}.
+Furthermore, supervised learning can be categorised into offline or batch learning and online learning (see section \ref{sec:onoffline}).
+\item[Unsupervised learning] also known as clustering \cite{ben2005}.
 In this type of learning no labels are given and the system has to find a structure on its own, discovering hidden patterns in data.
-\item[reinforcement learning] is the problem faced by an agent that must learn 
-through trial-and-error interactions with a dynamic environment. It is based on programming agents by reward and punishment without needing to specify how the task is to be achieved \cite{sutton1998}.
+\item[Reinforcement learning] is the problem faced by an agent that must learn 
+through trial-and-error interactions with a dynamic environment.
+It is based on programming agents by reward and punishment without the need to specify how the task is to be achieved \cite{sutton1998}.
 \end{description}
-
-%%% Learning problem. Ac;a ma;ana, formulas y descripcion de tallada del problema de aprendizaje
 
@@
   \includegraphics[width=0.8\textwidth]{img/3tradeoff}
-  \caption{Triple tradeoff in empirical learning}
+  \caption{Tradeoff in empirical learning.
+The generalization accuracy is not affected by the amount of examples for low complex classifiers.
+However, for some more complex classifiers, the number of examples improves the generalization accuracy.
+The key is to find the best generalization accuracy and the lowest complex classifier as possible.
+}
   \label{fig:tradeoff}
@@
 estimator. 
-It is known that the prediction error can be express as a decomposition between bias and variance. The prediction error is
-obtained as:
-
-\begin{equation}
- \label{eq:prederror}
-E[(Y-\hat{f}(\mathbf{X}))^2] = \sigma^2 + Bias(\hat{f}(\mathbf{X}))^2 + Var(\hat{f}(\mathbf{X}))
-\end{equation}
-
-\noindent where $\hat{f}(\mathbf{X})=\mathbf{AX}(\lambda)$ for RR and
-$\hat{f}(\mathbf{X})=\mathbf{AX}$ in the case of OLS.
-
-For ridge regression we will use the following notation: $\hat{f}(\mathbf{X})=\mathbf{AX}(\lambda)$ and
-for OLS $f(\mathbf{X})=\mathbf{AX}$. The solution of OLS is well known $\hat{\mathbf{X}}=(\mathbf{A}^\top \mathbf{A})^{-1}\mathbf{A}^\top \mathbf{Y}$, and its bias is:
-
-\begin{eqnarray*}
-Bias(\hat{\mathbf{X}}) &=& E[\hat{\mathbf{X}}] - \mathbf{X} \\
+It is known that the prediction error can be express as a decomposition between bias and variance.
+
+The solution of OLS is well known $\hat{\mathbf{X}}=(\mathbf{A}^\top \mathbf{A})^{-1}\mathbf{A}^\top \mathbf{Y}$, and its bias is:
+
+\begin{eqnarray*}
+Bias(\hat{f}(\hat{\mathbf{X}}) &=& E[\hat{\mathbf{X}}] - \mathbf{X} \\
 &=& E[ (\mathbf{A}^\top \mathbf{A})^{-1}\mathbf{A}^\top \mathbf{Y}] - \mathbf{X} \\
@@
 \end{eqnarray*}
-
-\begin{figure}[h!]
-\includegraphics[width=\linewidth]{img/biasvariance}
-\caption{Bias-variance tradeoff depending on lambda}
-\label{fig:biasvariance}
-\end{figure}
 
@@
 
+\begin{figure}[h!]
+\includegraphics[width=\linewidth]{img/biasvariance}
+\caption{Bias-variance tradeoff depending on lambda.
+Dotted line corresponds to OLS prediction error (invariant with lambda).
+Black line corresponds to RR prediction error which can be decomposed in $Bias^2$ (in red) + Variance (in blue).}
+\label{fig:biasvariance}
+\end{figure}
+
 The figure~\ref{fig:biasvariance} shows the bias-variance tradeoff
-given by equation~(\ref{eq:prederror}). Despite OLS has zero bias,
+given by equation~(\ref{eq:bvtrade}).
+
+Despite OLS has zero bias,
 its variance is greater than ridge for small values of $\lambda$.
@@
 
-ANN is inspired by biological learning systems.
-ANN are organised into layers and have unidirectional connections between the layers, feed-forward networks  (FFN) are the most used.
+ANN is inspired by biological learning systems organised into layers and have unidirectional connections between the layers, feed-forward networks  (FFN) are the most used.
 FFN consist of a series of layers.
@@
   \includegraphics[width=0.8\textwidth]{img/ffn}
-  \caption{Feed-forward neural network}
+  \caption{Feed-forward neural network (FFN). FNN consists of an input layer, an output layer and at least one hidden layer between the input and output layer. }
   \label{fig:ffn}
@@
 maximising the margin between the training patterns and the decision boundary. The
-solution is expressed as a linear combination of supporting patterns, which are the 
-of training patterns close to the decision boundary, called the support vectors.
+solution is expressed as a linear combination of supporting patterns, which are the training patterns close to the decision boundary, called the support vectors.
 
@@
 
-In principle, the stream of observations may be infinitely long, or the next
-observation long delayed, precluding any hope of waiting until all the
-observations have been received. One would rather not simply accumulate and
-store all the inputs and, upon receipt of each new one, apply a batch learning
-algorithm to the entire sequence of inputs received so far. It would be
-preferable computationally if the existing hypothesis or other artefact of
-learning could be updated in response to each newly received input observation.
-
 Incremental learning is very useful when there is no need to record fundamental
@@
 An optimal offline algorithm can view the sequences of requests in advance.
-The effectiveness of an online algorithm \cite{sleator1985} may be measured by its competitive ratio which is the worst-case ratio of the online algorithm and the the optimal offline algorithm.
+The effectiveness of an online algorithm \cite{sleator1985} may be measured by its competitive ratio which is the worst-case ratio of the online algorithm and the optimal offline algorithm.
 
@@
 In regression problems we always require getting a matrix inverse which is
-computationally expensive. However, in stream data problems the is a way to
+computationally expensive. However, in stream data problems there is a way to
 obtain a matrix inverse approximation using the Sherman-Morrison-Woodbury. This
-formula allows to get the inverse of a
-the matrix $\mathbf{A+uv^\top}$ if we already calculated the
+formula allows to get the inverse of a matrix $\mathbf{A+uv^\top}$ if we previously calculated the
 inverse of $\mathbf{A}$ as follows:
@@
 
+\newpage
+
 \section{Evaluation methods} \label{sec:evaluation}

### 2016-07-06 articles-conferences/iscef2016/main.tex
agregando imágenes nuevas
change order chapters

@@
 %
-%\documentclass{svjour3}                     % onecolumn (standard format)
+%\documentclass{svjour3}                     
 %\documentclass[smallcondensed]{svjour3}     % onecolumn (ditto)
-\documentclass[smallextended]{svjour3}       % onecolumn (second format)
+\documentclass[smallextended]{svjour3}       % onecolumn (second format) ******
+%\usepackage[doublespacing]{setspace}   
 %\documentclass[twocolumn]{svjour3}          % twocolumn
@@
 
+%\usepackage{natbib}
 \usepackage{graphicx}
+
 %
@@
 Cointegration is a long-run property of some non-stationary time series where a
-linear combination of those time series is stationary.
-This behaviour has been studied in finance because cointegration restrictions often improve forecasting.
-The Vector Error Correction Model (VECM) is a well-known econometric technique that
-characterises short-run variations of a set of cointegrated time series incorporating long-run relationships as an error correction term.
-VECM has been broadly used with low frequency time series.
-We aimed to adapt VECM to be used in finance with high frequency stream data.
+linear combination of those time series is stationary.  This behaviour has been
+studied in finance because cointegration restrictions often improve forecasting.
+The Vector Error Correction Model (VECM) is a well-known econometric technique
+that characterises short-run variations of a set of cointegrated time series
+incorporating long-run relationships as an error correction term. VECM has been
+broadly used with low frequency time series. We aimed to adapt VECM to be used
+in finance with high frequency stream data.
 %
-
-Cointegration relations change in time and therefore VECM parameters must be updated when new data is available.
-We studied how forecasting performance is affected when VECM parameters and the length of historical data used change in time.
-We observed that the number of cointegration relationships varies with the length of historical data used.
-Moreover, parameters that increased these relationships in time led to better forecasting performance.
-Our proposal, called an Adaptive VECM (AVECM) is
-to make a parameters grid search that maximises the number of cointegration relationships in the near past. To ensure the search can be executed fast enough, we used a distributed environment.
-
+Cointegration relations change in time and therefore VECM parameters must be
+updated when new data is available. We studied how forecasting performance is
+affected when VECM parameters and the length of historical data used change in
+time. We observed that the number of cointegration relationships varies with the
+length of historical data used. Moreover, parameters that increased these
+relationships in time led to better forecasting performance. Our proposal,
+called an Adaptive VECM (AVECM) is to make a parameters grid search that
+maximises the number of cointegration relationships in the near past. To ensure
+the search can be executed fast enough, we used a distributed environment.
 %
-The methodology was tested using four 10-seconds frequency time series of the Foreign Exchange market.
-We compared our proposal with ARIMA and the naive forecast of the random walk model.
-Numerical experiments showed that on average AVECM performed better than ARIMA and random walk.
-Additionally, our parallel approach AVECM significantly improved execution times with respect to its serial version.
+The methodology was tested using four 10-seconds frequency time series of the
+Foreign Exchange market. We compared our proposal with ARIMA and the naive
+forecast of the random walk model. Numerical experiments showed that on average
+AVECM performed better than ARIMA and random walk. Additionally, AVECM
+significantly improved execution times with respect to its serial version.
 
@@
 \end{abstract}
+
+\begin{acknowledgements}
+L. Salinas acknowledges support from Basal Project
+FB0821 CCTVal-Centro Científico Tecnológico de Valparaíso. P. Arce and W.
+Kristjanpoller thank PIIC 2014 DGIP-UTFSM grant for supporting this research.
+We also acknowledge to the Fourth International Symposium in Computational Economics and Finance (ISCEF) 2016 for valuable comments and suggestions.
+\end{acknowledgements}
 
@@
 
-\begin{acknowledgements}
-L. Salinas acknowledges support from Basal Project
-FB0821 CCTVal-Centro Científico Tecnológico de Valparaíso. P. Arce and W.
-Kristjanpoller thank PIIC 2014 DGIP-UTFSM grant for supporting this research.
-\end{acknowledgements}
-
 % BibTeX users please use one of
-%\bibliographystyle{spbasic}      % basic style, author-year citations
+\bibliographystyle{spbasic}      % basic style, author-year citations
 %\bibliographystyle{spmpsci}      % mathematics and physical sciences

### 2016-07-06 src/5_2_Proposal.tex
change order chapters

@@
-\chapter{Adaptive parallel algorithm for Vector error correction model}
+\chapter{Online VECM proposal}
+Financial time series are known for their non-stationary behaviour. However,
+sometimes they exhibit some stationary linear combinations. When this happens,
+it is said that those time series are cointegrated. The Vector Error Correction
+Model (VECM) is an econometric model which characterises the joint dynamic
+behaviour of a set of cointegrated variables in terms of forces pulling towards
+equilibrium. In this study, we propose an Online VEC model (OVECM) which
+optimises how model parameters are obtained using a sliding window of the most
+recent data. Our proposal also takes advantage of the long-run relationship
+between the time series in order to obtain improved execution times. Our
+proposed method is tested using four foreign exchange rates with a frequency of
+1-minute, all related to the USD currency base. OVECM is compared with VECM and
+ARIMA models in terms of forecasting accuracy and execution times. We show that
+OVECM outperforms ARIMA forecasting and enables execution time to
+be reduced considerably while maintaining good accuracy levels compared with
+VECM.
 
@@
 
-Vector error correction model (VECM) parameters are obtained using ordinary
-least squares (OLS) method \cite{golub1980}, which is computationally expensive
-when the number of past values $L$ and observations $p$ increases. Moreover,
-obtaining cointegration vectors is also an expensive routine because of the use
-of the Johansen method.
+\section{The problem}
+In finance, it is common to find variables with a long-run and/or a short-run
+equilibrium relationship. This relationship is called cointegration and it
+reflects the idea of that some set of variables cannot wander too far away from
+each other.
 
-In this study we found that for some combinations of $L$ and $p$, the number of
-cointegration relations is increased and this leads to better performance. We
-propose to parallelise this search in order to achive faster execution times.
+Cointegration means that one or more combinations of these variables is
+stationary even though individually they are not.
 
-\section{The problem} \label{sec:proposal}
+Some financial models, such as  vector error correction model (VECM), take advantage of
+this property and describe the joint behaviour of several cointegrated
+financial instruments.
 
-Cointegration vectors can be found applying the Johansen method which uses a
-sample of the most recent historical data. However, one of the main limitations of this
-methodology is that it assumes cointegration vectors do not change in time.  In
-fact, the long-run relationships between the time series might change due to
-several economic factors that can lead to structural breaks in the cointegrating
-relation \cite{gregoryETal1996}.
+VECM is a special case of the vector autoregressive model (VAR). VAR is a linear
+model which express future values in terms of its own history and other
+variables past values. If cointegration exists, VECM allows to introduce an
+error correction term due to its cointegration estimation error.
 
-In this study, we show that cointegration relations depend on the amount $L$ of
-historical data considered, and how many lags $p$ were used. This search is
-computationally expensive, considering that the Johansen
-method~\cite{johansen1995} (which is of order $O(n^3)$) is used and these
-parameters change at every iteration.
+VECM as well as VAR model parameters are obtained using ordinary least squares
+method (OLS). Since OLS involves many calculations, the parameter estimation
+method is computationally expensive when the number of past values and
+observations considered increases. Moreover, OLS is an ill-posed problem which
+admits an infinite number of solutions.
 
-Our proposal consists on to parallelise this search in order to get new
-parameters at every iteration.
+Ridge regression method (RR) tackles this ill-posed problem and is usually
+formulated instead of OLS. RR includes a regularisation parameter that leads to
+better generalisation capability. However, RR is still computationally
+expensive dealing with large datasets. Recently, online learning algorithms
+have been proposed to solve problems with large data sets because of their
+simplicity and their ability of updating the model when new data is available.
 
-In finance, little academic reasearch has been done using high-frequency Forex
-rates \cite{Bekiros2015}. In our study, tests were run using four currency
-rates: Euro (EUR) to United States Dollar (USD) (EURUSD), British Pound (GBP) to
-USD (GBPUSD), USD to Swiss Franc (CHF) (USDCHF) and USD to Japanese Yen (JPY)
-(USDJPY) with a 10-seconds frequency.  The forecast capability of our method was
-later measured using MAPE, MAE and RMSE \cite{armstrong1992}.
+We propose an online formulation of the VEC model based on the
+AAR method considering only a sliding window of the most recent data. The
+algorithm introduces matrix optimisations in order to reduce the number of
+operations. Our method is later tested with financial data from the foreign
+exchange market.
 
-\section{Methodology}
+\section{Methodology} \label{sec:methodology}
 
-VECM is obtained re-writing the VAR model, shown in equation (\ref{eq:var}), in
-terms of differences defines as: $\Delta\mathbf{y}_t=\mathbf{y}_t-\mathbf{y}_{t-1}$.
-The VECM model, expressed in terms those differences, takes the form:
-\begin{equation}\label{eq:vec}
-\Delta \mathbf{y}_t 
-= \boldsymbol{\Omega}\,\mathbf{y}_{t-1}
-  + \sum_{i=1}^{p-1} \boldsymbol{\Phi}_i^*\,\Delta\mathbf{y}_{t-i}
-  + \mathbf{c} + \boldsymbol{\epsilon}_t\,,
-\end{equation}
-\noindent
-where the coefficients matrices $\boldsymbol{\Phi}_i^*$ and 
-$\boldsymbol{\Omega}$, expressed in terms of the matrices
-$\boldsymbol{\Phi}_i$ of (\ref{eq:var}), are:
-\begin{align*}
-\boldsymbol{\Phi}_i^* 
-&:= -\sum_{j=i+1}^{p}\boldsymbol{\Phi}_j\,, \\
-\boldsymbol{\Omega}
-&:= -\left( \mathbb{I} - \boldsymbol{\Phi}_1 - \dots 
-    - \boldsymbol{\phi}_p \right)\,. 
-\end{align*}
-The following well known properties of the matrix $\boldsymbol{\Omega}$
-\cite{johansen1995} will be useful in the sequel:
+Since VECM parameter estimation is computationally expensive, we propose an
+online version of VECM (OVECM).  OVECM considers only a sliding window of the
+most recent data. Moreover, since cointegration vectors represent long-run
+relationships which vary little in time, OVECM determines firstly if they require calculation.
+
+OVECM also implements matrix optimisations in order to reduce execution time,
+such as updating matrices with new data, removing old data and introducing new
+cointegration vectors.
+
+Algorithm~\ref{alg:proposal} shows our OVECM proposal which considers the
+following:
+
 \begin{itemize}
-\item
-If $\boldsymbol{\Omega} = \mathbf{0}$, there is no cointegration.
-\item 
-If $rank(\boldsymbol{\Omega})=l$, i.e., if $\boldsymbol{\Omega}$ has
-full rank, then the time series are not I(1) but stationary.
-\item
-If $rank(\boldsymbol{\Omega})=r$, $0<r<l$, then there is cointegration
-and the matrix $\boldsymbol{\Omega}$ can be expressed as
-$\boldsymbol{\Omega}=\boldsymbol{\alpha\beta}^\top$, where $\boldsymbol{\alpha}$
-and $\boldsymbol{\beta}$ are
-$l\times r$ matrices and
-$\text{rank}(\boldsymbol{\alpha})=\text{rank}(\boldsymbol{\beta})=r$.
-\item
-The columns of $\boldsymbol{\beta}$ contains the cointegration vectors and the rows of
-$\boldsymbol{\alpha}$ correspond with the adjusted vectors. 
-$\boldsymbol{\beta}$ is obtained by Johansen procedure~\cite{johansen1988},
-whereas $\boldsymbol{\alpha}$ has to be determined as a variable in the VECM.
+\item The function \texttt{getJohansen} returns cointegration vectors given by
+the Johansen method considering the trace statistic test at 95\%
+significance level.
+\item The function \texttt{vecMatrix} returns the matrices~(\ref{eq:vecA})
+and~(\ref{eq:vecY}) that allows VECM to be solved.
+\item The function \texttt{vecMatrixOnline} returns the
+matrices~(\ref{eq:vecA}) and~(\ref{eq:vecY}) aggregating new data and removing
+the old one, avoiding calculation of the matrix $\mathbf{A}$ from scratch.
+\item Out-of-sample outputs are saved in the variables 
+$\mathbf{Y}_{\text{true}}$ and $\mathbf{Y}_{\text{pred}}$.
+\item The model is solved using OLS.
+\item In-sample outputs are saved in the variables $\Delta
+\mathbf{y}_{\text{true}}$ and $\Delta \mathbf{y}_{\text{pred}}$.
+\item The function \texttt{mape} gets the in-sample MAPE for the $l$ time
+series.
+\item Cointegration vectors are obtained at the beginning and when they are required to be updated. This updating is decided based on the in-sample MAPE of the last $n$ inputs. The average of all
+MAPEs are stored in the variable $e$. If the average of MAPEs
+($\texttt{mean}(e)$) is above a certain error given by the mean\_error threshold, cointegration vectors are updated.
+\item If new cointegration vectors are required, the function
+\texttt{vecMatrixUpdate} only updates the corresponding columns of matrix
+$\mathbf{A}$ affected by those vectors (see equation~\ref{eq:vecA}).
 \end{itemize}
-
-If cointegration exists, then equation (\ref{eq:vec}) can be written
-as follows:
-\begin{equation}\label{eq:vecfull}
-\Delta\mathbf{y}_t 
-= \boldsymbol{\alpha\beta}^\top\mathbf{y}_{t-1} 
-  + \sum_{i=1}^{p-1}\boldsymbol{\Phi}_i^*\,\Delta\mathbf{y}_{t-i}
-  + \mathbf{c} + \boldsymbol{\epsilon}_t\,,
-\end{equation}
-\noindent
-which is a VAR model but for time series differences.
-
-Transposing each equation of the system (\ref{eq:vecfull}) we can write
-the VECM($p$) model in block-matrix form as:
-\begin{equation}\label{eq:vareq}
-\mathbf{B} = 
-\mathbf{A} \mathbf{X} + 
-\mathbf{E} \, , 
-\end{equation}
-%
-\noindent where $\mathbf{B}$ dimension is $((N-p)\times l)$, $\mathbf{A}$
-dimension is $((N-p)\times(r+(p-1)l +1))$, $\mathbf{X}$ dimension is $((r+(p-1)l
-+1)\times l)$ and $\mathbf{E}$ dimension is $((N-p)\times l)$:
-%
-\begin{alignat}{3}
-\mathbf{B}
-&= \begin{bmatrix}
-   \Delta\mathbf{y}_{p+1}^\top \\
-   \Delta\mathbf{y}_{p+2}^\top \\
-   \vdots \\
-   \Delta\mathbf{y}_N^\top
-   \end{bmatrix}
-&\quad
-\mathbf{X}
-&= \begin{bmatrix}
-   \boldsymbol{\alpha}^\top \\
-   \boldsymbol{\Phi}_1^{*\top} \\
-   \boldsymbol{\Phi}_2^{*\top} \\
-   \vdots \\
-   \boldsymbol{\Phi}_{p-1}^{*\top} \\
-   \mathbf{c}^\top
-   \end{bmatrix}
-&\quad
-\mathbf{E}
-&= \begin{bmatrix}
-   \boldsymbol{\epsilon}_{p+1}^\top \\
-   \boldsymbol{\epsilon}_{p+2}^\top \\
-   \vdots \\
-   \boldsymbol{\epsilon}_N^\top \\
-   \end{bmatrix}
-\end{alignat}
-\noindent and 
-\begin{align}
-\mathbf{A} 
-&= \begin{pmat}[{....|}]
-   \mathbf{y}_p^\top \boldsymbol{\beta} & \Delta \mathbf{y}_p^\top & \Delta\mathbf{y}_{p-1}^\top & \dots 
-                    & \Delta\mathbf{y}_2^\top & 1 \cr
-   \mathbf{y}_{p+1}^\top  \boldsymbol{\beta} &\Delta\mathbf{y}_{p+1}^\top & \Delta\mathbf{y}_p^\top & \dots
-                       & \Delta\mathbf{y}_3^\top & 1 \cr
-   \vdots & \vdots & \vdots & \ddots & \vdots & \vdots \cr
-   \mathbf{y}_{N-1}^\top  \boldsymbol{\beta} &\Delta\mathbf{y}_{N-1}^\top & \Delta\mathbf{y}_{N-2}^\top & \dots 
-                       & \Delta\mathbf{y}_{N-p-1}^\top & 1 \cr
-   \end{pmat}\, .
-\end{align}
-Taking into account the error term $\mathbf{E}$, equation~(\ref{eq:vareq}) 
-can be solved with respect to $\mathbf{X}$ using the ordinary least
-squares estimation.
-
-\subsection{Motivation}
-The relation between the number of cointegration relations
-and parameters $L$ and $p$ is studied and then a relation between cointegration
-and performance is stated.
-
-Figure~\ref{fig:hists} shows the number $r$ of cointegration vectors given by
-the Johansen method for different values of the parameters $L$ leaving $p$ set
-as $1$, using four currencies data, i.e., $l=4$, and 1000 considering iterations
-($it=1000$).
-\begin{figure}[!h]
-  %\vspace{-0.8cm}
-  \centering
-  \includegraphics[width=\textwidth]{img/histCointVectors}
-  \caption{Histogram of the number of cointegration vectors using $p=1$. Four
-  possible values for $L$ are shown (100, 1900, 6100 and 10700).
-  In the cases of $L=100$ and $L=6100$, most of the time there is little
-  cointegration (0 or 4 cointegration vectors). However, for $L=1900$ and
-  $L=10700$ significant cointegration is found.}
-  \label{fig:hists}
-\end{figure}
-We know that when no cointegration vectors can be found, i.e $r=0$, implies that
-the variables are not cointegrated. In the same way, when $r=l$ reveals that
-none of the processes is I(1) but, they are all stationary.  The interesting cases of
-cointegration are those, where $r$ lies strictly between $0$ and $l$, i.e.
-$0<r<l$,
-
-Experiments were run with data starting at 9:00 GMT of the 11th of August 2014,
-which corresponds to the opening of the New York financial market.
-
-\begin{figure}[!h]
-  %\vspace{-0.8cm}
-  \centering
-  \includegraphics[width=\textwidth]{img/histCointVectorsp5}
-  \caption{Histogram of the number of cointegration vectors using $p=5$. 
-   Despite the change in the distribution of cointegration vectors, 
-   the percentage of cointegration is nearly the same as in the case $p=1$. 
-   For example, for $L=10700$ the sum of 1 and 2 number of cointegration
-   vectors is very similar to 1 cointegration vector for the case $L=10700$
-   in figure \ref{fig:hists}.}
-   \label{fig:histsp5}
-\end{figure}
-
-The same experiments, setting $p=5$, were carried out in order to see the
-effect on the number of cointegration vectors found. 
-Figure~\ref{fig:histsp5} shows that, despite the change in the distribution
-of the number of cointegration vectors, the sum of the cases where $0<r<4$,
-remains nearly the same. 
-Therefore, we can say that the number of lags doesn't significantly affect
-the extent of cointegration found.
-
-In order to measure the extent of cointegration, we introduce a
-{\em percentage of cointegration\/} as following:
-\begin{equation} \label{eq:pcoint}
-PC = 
-\frac{\#\{ it \mid \text{$it$ has $r$ c.v. with $0<r<l$}\}}
-     {\#it}\times 100
-\end{equation}
-where c.v. stands for cointegration vectors and $it$ is the number of iterations.
-
-The goal of our next experiment was to find a relation between this ratio
-$PC$ and the performance of the accuracy measure MAPE (see equation 
-\ref{eq:MAPE}). 
-The experiments suggest that there is indeed a relation between the $PC$ ratio
-and MAPE.
-Figure~\ref{fig:cointvsmape} shows that MAPE rapidly decreases with increasing
-$PC$.
-
-\begin{figure}[!h]
-  %\vspace{-0.8cm}
-  \centering
-  \includegraphics[width=0.7\textwidth]{img/MAPEvsCoint-offset21600-p-2-freq-10s}
-  \caption{MAPE versus percentage of cointegration. MAPE was obtained from 1000
-  iterations considering several possible $L=[100,10000,400]$ and $p=[1,5]$.
-  Best $L$ found was 8900.}
-  \label{fig:cointvsmape}
-\end{figure}
-
-From these results we observe that cointegration changes with different values
-for $L$ and $p$ and also that better cointegration percentage leads to better
-accuracy performance.  Therefore we propose to choose best $L$ and $p$
-parameters determined at every step.  The criterion for choosing $L$ and $p$ is
-to maximise the percentage of cointegration $PC$ considering a number of past
-iterations $it$.
-
-Algorithm ~\ref{alg:AVECM} solves equation \ref{eq:vareq} defining first
-parameters $L$ and $p$ using the function {\bf get\_best\_params} which makes a
-grid search on the two vector lists $Ls$ and $ps$ and returns the parameters $L$
-and $p$ which maximise the percentage of cointegration $PC$ (see
-equation~\ref{eq:pcoint}) for a pre-defined number of iterations.  Since this
-search is computational expensive, the number of cointegration vectors at each
-iteration is computed in parallel, thus ensuring a response before new data is
-available.  This parallel routine also allows to refine our grid search leading
-to obtain improved forecasting performance.  After that optimum $L$ and $p$
-parameters are found, VECM is built and used to forecast next data.
 
@@
 $\mathbf{y}$: matrix with $N$ input vectors and $l$ time series\\
-$ps$: vector with the number of past values range \\
-$Ls$: vector with window sizes range ($L<N$) \\
-$outit$: Number of out-of-sample iterations($outit < N-L$)
-$it$: Number of iterations of testing (in-sample)\\
-$of$: Starting point of Testing \\
+$p$: number of past values \\
+$L$: sliding window size ($L<N$) \\
+$\text{mean\_error}$: MAPE threshold \\
+$n$: interpolation points to obtain MAPE \\
 \ENSURE  $\,$ \\
-$\{ \mathbf{y}_{\text{pred}}[1],\dots,\mathbf{y}_{\text{pred}}[N-L]\}$: model predictions 
-\FOR { $i =of$ to $of + outit$ }
-   \STATE $\mathbf{Y} \gets \mathbf{y}[i-\texttt{max}(Ls):i]$
-    \STATE $L,p \gets
-    \texttt{get\_best\_params}(Ls,ps,it,Y)$
-    \STATE $\mathbf{y}_i \gets \mathbf{y}[i-L:i]$
-        \STATE $model = VECM(\mathbf{y}_i, p)$
-        \STATE $\mathbf{y}_{\text{pred}}[i] = model.predict()$
+$\{\mathbf{y}_{\text{pred}}[L+1],\dots,\mathbf{y}_{\text{pred}}[N]\}$: model predictions 
+\FOR { $i =0$ to $N-L$ }
+    \STATE $\mathbf{y}_i \gets \mathbf{y}[i:i+L]$
+	\IF {i = 0 }
+	    \STATE{$v \gets \texttt{getJohansen}(\mathbf{y}_i,p)$}
+	    \STATE{$[\mathbf{A} \quad \mathbf{Y}] \gets
+        \texttt{vecMatrix}(\mathbf{y}_i,p,v)$}
+	\ELSE
+	    \STATE{$[\mathbf{A} \quad \mathbf{Y}] \gets
+        \texttt{vecMatrixOnline}(\mathbf{y}_i,p,v,\mathbf{A},\mathbf{Y})$}
+        \STATE $\Delta \mathbf{Y}_{\text{pred}}[i] \gets \mathbf{A}[-1,:] \times \mathbf{X}$
+    \ENDIF
+    \STATE $\mathbf{X} \gets \texttt{OLS} (\mathbf{A},\mathbf{Y})$
+    \STATE $e \gets \texttt{mape}(\mathbf{Y}[-n,:],\mathbf{A}[-n,:] \times \mathbf{X})$
+    \IF {$\texttt{mean}(e) > \text{mean\_error}$}
+	    \STATE{$v \gets \texttt{getJohansen}(\mathbf{y}_i,p)$}
+	    \STATE{$\mathbf{A} \gets
+        \texttt{vecMatrixUpdate}(\mathbf{y}_i,p,v,\mathbf{A})$}
+        \STATE $\mathbf{X} \gets \texttt{OLS} (\mathbf{A},\mathbf{Y})$
+    \ENDIF
 \ENDFOR
+\STATE $\mathbf{Y}_{\text{true}} \gets \mathbf{Y}[L+1:N]$
+\STATE $\mathbf{Y}_{\text{pred}}\gets 
+\mathbf{Y}[L:N-1] +\Delta \mathbf{Y}_{\text{pred}}$
 \end{algorithmic}
-\caption{AVECM: Adaptive VECM.}
-\label{alg:AVECM}
+\caption{OVECM: Online VECM}
+\label{alg:proposal}
 \end{algorithm}
 
-\section{Experimental results}
-\label{sec:results}
+A pseudocode of the algorithm \ref{alg:proposal} is detailed in the appendix \ref{fig:pseudocodigo}.
+
+\resizebox{\textwidth}{!}{
+   \begin{tikzpicture}[node distance=2cm]
+    \label{fig:pseudocodigo}
+    %\node (start) [startstop] {Start};
+    %\node (in1) [io, below of=start] {Input};
+    \node (in1) [io] {Input: y, p, L, $\epsilon$, n};
+    \node (pro1) [process, below of=in1] {$y_i = y[i:i+L]$};
+    \node (dec1) [decision, below of=pro1, yshift=-0.5cm] {i = 0};
+    \node (pro2a) [process, below of=dec1, yshift=-0.5cm] {$[\mathbf{A} \quad \mathbf{B}] =\texttt{ovecm}(\mathbf{y}_i,p,v,\mathbf{A},\mathbf{B})$
+                                                        $\Delta \mathbf{Y}_{\text{true}}[i] = \mathbf{B}[-1,:]$
+                                                        $\Delta \mathbf{Y}_{\text{pred}}[i] = \mathbf{A}[-1,:] \times \mathbf{X}$};
+
+    \node (pro2b) [process, right of=dec1, xshift=4cm] {$v = \texttt{Johansen}(\mathbf{y}_i,p)$ $[\mathbf{A} \quad 
+                                                                \mathbf{B}] = \texttt{vecm}(\mathbf{y}_i,p,v)$};
+    \node (pro3) [process, below of=pro2a, yshift=-0.5cm] {$\mathbf{X} = \text{OLS} (\mathbf{A},\mathbf{B})$ \\
+                                                        $\mathbf{y}_{\text{true}}[i] = \mathbf{y}_i[-n,:]$ \\
+                                                        $\mathbf{y}_{\text{pred}}[i] = \mathbf{A}[-n,:] \times \mathbf{X} +$ \\$ \mathbf{y}_i[-n-1,:-1]$ \\
+                                                        $e = \texttt{mape}(\mathbf{y}_{\text{true}}, \mathbf{y}_{\text{pred}})$};
+
+    \node (dec2) [decision, below of=pro3, yshift=-0.5cm] { $e > \epsilon $ };
+
+    \node (pro4a) [process, right of=dec2, xshift=4cm] {$v = \texttt{Johansen}(\mathbf{y}_i,p)$ \\
+                                                        $\mathbf{A} = \texttt{uvecm}(\mathbf{y}_i,p,v,\mathbf{A})$ \\
+                                                        $\mathbf{X} = \texttt{OLS} (\mathbf{A},\mathbf{B})$};
+
+    \node (pro5) [process, left of=pro2a, xshift=-4cm] {$ i++ $};
+
+    \node (dec3) [decision, below of=dec2, yshift=-0.5cm] { $i > it $ };
+
+    \node (out1) [io, below of=dec3] {Output: $\Delta \mathbf{y}_{true}, \Delta \mathbf{y}_{pred}$};
+
+    %\node (stop) [startstop, below of=out1] {Stop};
+
+    %\draw [arrow] (start) -- (in1);
+    \draw [arrow] (in1) -- (pro1);
+    \draw [arrow] (pro1) -- (dec1);
+    \draw [arrow] (dec1) -- node[anchor=east] {no} (pro2a);
+    \draw [arrow] (dec1) -- node[anchor=south] {si} (pro2b);
+    \draw [arrow] (pro2a) -- (pro3);
+    \draw [arrow] (pro2b) |- (pro3);
+    \draw [arrow] (pro3) -- (dec2);
+    \draw [arrow] (dec2) --  node[anchor=south] {si} (pro4a);
+    \draw [arrow] (pro4a) |- (dec3);
+
+    \draw [arrow] (dec2) --  node[anchor=east] {no} (dec3);
+    \draw [arrow] (dec3.west) -| node[anchor=east] {no} (pro5);
+    \draw [arrow] (pro5) |- (pro1);
+    \draw [arrow] (dec3) --  node[anchor=east] {si} (out1);
+    %\draw [arrow] (out1) -- (stop);
+    \end{tikzpicture}
+}
+
+Our proposal was compared against VECM and ARIMA. Both algorithms were adapted
+to an online context in order to get a reasonable comparison with our proposal
+(see algorithms \ref{alg:SLVECM} and \ref{alg:SLARIMA}). VECM and ARIMA are
+called with a sliding window of the most recent data, whereby the models are
+updated at every time step.
+
+\begin{algorithm}[ht]
+\begin{algorithmic}[1]
+\REQUIRE $\,$ \\
+$\mathbf{y}$: matrix with $N$ input vectors and $l$ time series\\
+$p$: number of past values \\
+$L$: sliding window size ($L<N$) \\
+\ENSURE  $\,$ \\
+$\{ \mathbf{y}_{\text{pred}}[L+1],\dots,\mathbf{y}_{\text{pred}}[N]\}$: model predictions 
+\FOR { $i =0$ to $N-L$ }
+    \STATE $\mathbf{y}_i \gets \mathbf{y}[i:i+L+1]$
+        \STATE $model = VECM(\mathbf{y}_i, p)$
+        \STATE $\mathbf{Y}_{\text{pred}}[i] = model.predict(\mathbf{y}[i+L])$
+\ENDFOR
+\STATE $\mathbf{Y}_{\text{true}} = \mathbf{y}[i+L+1:N] $
+\end{algorithmic}
+\caption{SLVECM: Sliding window VECM}
+\label{alg:SLVECM}
+\end{algorithm}
+
+Since we know out time series are I(1) SLARIMA is called with $d=1$. ARIMA is
+executed for every time series.
+
+\begin{algorithm}[ht]
+\begin{algorithmic}[1]
+\REQUIRE $\,$ \\
+$\mathbf{y}$: matrix with $N$ input vectors and $l$ time series\\
+$p$: autoregressive order \\
+$d$: integrated order\\
+$q$: moving average order\\
+$L$: sliding window size ($L<N$) \\
+\ENSURE  $\,$ \\
+$\{ \mathbf{y}_{\text{pred}}[L+1],\dots,\mathbf{y}_{\text{pred}}[N]\}$: model predictions 
+\FOR { $i =0$ to $N-L$ }
+\FOR { $j =0$ to $l-1$ }
+    \STATE $\mathbf{y}_i \gets \mathbf{y}[i:i+L+1,j]$
+        \STATE $model = ARIMA(\mathbf{y}_i, (p,d,q))$
+        \STATE $\mathbf{Y}_{\text{pred}}[i,j] = model.predict(\mathbf{y}[i+L,j])$
+\ENDFOR
+\ENDFOR
+\STATE $\mathbf{Y}_{\text{true}} = \mathbf{y}[i+L+1:N,:] $
+\end{algorithmic}
+\caption{SLARIMA: Sliding window ARIMA}
+\label{alg:SLARIMA}
+\end{algorithm}
+
+Both OVECM and SLVECM time complexity is dominated by Johansen method which is
+$O(n^3)$. Thus, both algorithms order is $O(Cn^3)$ where $C$ is the number of
+iterations.
+
+\section{Results} \label{sec:results}
+
 \subsection{Data} \label{sec:unitroot}
-AVECM tests were carried out using four foreign exchange rates all related to
-USD: EURUSD, GBPUSD, USDCHF and USDJPY. This data was collected from the free
-database Dukascopy which gives access to the Swiss Foreign Exchange marketplace
-~\cite{Dukascopy2014}.
+Tests of SLVECM, SLARIMA and our proposal OVECM were carried out using foreign
+four exchange data rates: EURUSD, GBPUSD, USDCHF and USDJPY. This data was
+collected from the free database Dukascopy which gives access to the Swiss
+Foreign Exchange Marketplace ~\cite{Dukascopy2014}.
 
-The tests were done using 10-seconds frequency from ask prices which
-corresponded to 8640 data points per day from the 11th to the 15th of August
-2014. Since data datetimes were in GMT, opening market times for London, New
-York, Sidney and Tokyo corresponded with data points 1440, 3240, 6120 and 6840
-respectively. Our tests were made considering two hours after the opening time
-of the 13th of August. Therefore, in order to obtain these times, we added 2
-days and 2 hours to all the opening times ($2\times 8640 + 2 \times 360$) to obtain:
-19440, 21240, 24120 and 24840. We called offset to these times in
-table~\ref{tab:stats}, which allowed us to obtain performance at different
-times of the day.
+The reciprocal of the last two rates (CHFUSD, JPYUSD) were used in order to
+obtain the same base currency for all rates.  The tests were done using
+1-minute frequency from ask prices which corresponded to 1.440 data points per
+day from the 11th to the 15th of August 2014.
 
 \subsection{Unit root tests} \label{sec:unitroot}
-Before running the tests, we firstly checked whether the time series were
-I(1) using the Augmented Dickey Fuller (ADF) test at 95\% significance level.
-Table~\ref{tab:adf} shows that all currency rates cannot reject the unit root
-test but they rejected it with their first differences. This means that all of
-them are I(1) time series and we are allowed to use VECM and therefore AVECM.
+Before running the tests, we firstly checked if they were I(1) time series
+using the Augmented Dickey Fuller (ADF) test.
 
 \begin{table}[h!]
+\caption{Unit roots tests}
+\label{tab:adf}
 \begin{center}
@@
 \hline
-EURUSD          &  -0.05398   & -1.94101 & True       \\
-$\Delta$ EURUSD & -89.12344   & -1.94101 & False       \\
-GBPUSD          &  -0.88208   & -1.94101 & True          \\
-$\Delta$ GBPUSD & -82.56501   & -1.94101 & False       \\
-CHFUSD          & -0.49941    & -1.94101 & True         \\
-$\Delta$ CHFUSD & -74.21940   & -1.94101 & False       \\
-JPYUSD          &  0.36721    & -1.94101 & True        \\
-$\Delta$ JPYUSD & -101.15589  & -1.94101 & False     \\ 
+EURUSD          & -0.64 & -1.94 & True       \\
+$\Delta$EURUSD & -70.45   & -1.94 & False       \\
+GBPUSD          & -0.63   & -1.94 & True          \\
+$\Delta$GBPUSD & -54.53   & -1.94 & False       \\
+CHFUSD          & -0.88   & -1.94 & True         \\
+$\Delta$CHFUSD & -98.98   & -1.94 & False       \\
+JPYUSD          & -0.65 & -1.94 & True        \\
+$\Delta$JPYUSD & -85.78 & -1.94 & False     \\ 
 \hline
@@
 \end{center}
-\caption{Unit roots tests for EURUSD, GBPUSD, USDCHF and USDJPY at 10-seconds
-frequency.}
-\label{tab:adf}
 \end{table}
 
-\section{Parallel implementation} \label{sec:paralell}
+Table~\ref{tab:adf} shows that all currency rates cannot reject the unit root
+test but they rejected it with their first differences. This means that all of
+them are I(1) time series and we are allowed to use VECM and therefore OVECM.
 
-To determine optimal $L$ and $p$ parameters require the using of the Johansen
-method which is a computationally expensive routine. 
-In order to improve the execution time of this search, our proposal included
-a parallel search of VECM parameters using high performance computing.
-The main objective is to obtain a response before a new data arrives in
-the next 10 seconds.
+\subsection{Parameter selection} \label{sec:paramselection}
 
-The Johansen method is implemented in the Python Statsmodels
-library~\cite{seabold2010} and the parallel implementation was done using
-MPI in Python. 
-MPI was chosen because it allows large-scale parallel applications
-with wide portability to be built, being able to run in large clusters
-or on local computers.
-Our tests were done in a cluster with 2 servers Xeon E5-2667 (2.90GHz)
-of 24 cores each (48 cores in total) and 24GB RAM.
+In order to set OVECM parameters: windows size $L$ and lag order $p$,
+we propose to use several window sizes: $L = 100, 400, 700, 1000$. For every window size $L$ we chose lag order with minimum AIC.
 
-Since financial market are constantly changing, we ran our
-algorithm using different times of the day. We chose to use the algorithm
-starting on the third day and two hours after the market opening times. These
-times corresponded to data points 19440 (5 GMT London), 21240 (10 GMT New
-York), 24120 (19 GMT Sydney) and 24840 (9 GMT Tokyo).
+ARIMA parameters were also obtained using AIC.
+Parameters optimisation is presented in table~\ref{tab:params}:
 
-In order to determine the percentage of cointegration vectors, we considered
-different number of iterations: 10, 50 and 100. The number of possible
-combinations ($nparam$) of $L$ and $p$ is shown for 12, 24 and 47.
+\begin{table}[ht]
+\caption{Parameters optimisation. VECM order and ARIMA parameters were selected
+using AIC.}
+\label{tab:params}
+\begin{center}
+\begin{tabular}{|c|c|c|}
+\hline
+Windows size $L$ & VECM & ARIMA\\
+ $L$ & order $(p)$ & order $(p,d,q)$ \\
+\hline
+100 & 2 & p=2,d=1,q=1\\
+400 & 5 & p=1,d=1,q=1\\
+700 & 3 &p=2,d=1,q=1\\
+1000 &3 & p=2,d=1,q=1\\
+\hline
+\end{tabular}
+\end{center}
+\end{table}
+
+In OVECM we also define a mean\_error variable, which was defined based on the
+in-sample MAPEs. Figure \ref{fig:mapes} shows how MAPE moves and how mean\_error
+variable help us to decide whether new cointegration vectors are needed.
+
+\begin{figure}[!h]
+  \centering
+   \includegraphics[width = 10.5cm]{img/mapes} 
+  \caption{In-sample MAPEs example for 50 minutes. The average of them is
+  considered to obtain new cointegration vectors.}
+  \label{fig:mapes}
+ \end{figure}
+
+\subsection{Execution times} \label{sec:exectimes}
+We ran OVECM and SLVECM 400 iterations. SLARIMA execution time is excluded
+because its is not comparable with OVECM and SLVECM. SLARIMA was created based
+on statsmodels library routine ARIMA.
+
+The execution times are shown in the table~\ref{tab:extimes}.
+
+\begin{table}[h!]
+\caption{Execution times}
+\label{tab:extimes}
+\begin{center}
+\begin{tabular}{|l|c|c|c|c|c|}
+\hline
+& $L$ & order & e  & Time[s] \\
+\hline
+OVECM & 100 &p=2  & 0      & 2.492\\
+OVECM & 100 &p=2  & 0.0026  & 1.606\\
+SLVECM & 100 &p=2& -- & 2.100\\
+\hline
+OVECM & 400 & p=5  & 0      & 3.513\\
+OVECM & 400 &p=5  & 0.0041  & 2.569\\
+SLVECM & 400 & p=5 & -- & 3.222\\
+\hline
+OVECM & 700 &p=3  & 0      & 3.296\\
+OVECM & 700 &p=3  & 0.0032  & 2.856\\
+SLVECM & 700 &p=3 & -- & 3.581\\
+\hline
+OVECM & 1000 & p=3 & 0      & 4.387\\
+OVECM & 1000 & p=3  & 0.0022  & 2.408\\
+SLVECM & 1000 & p=3  & -- & 3.609\\
+\hline
+\end{tabular}
+\end{center}
+\end{table}
+
+It is clear that execution time depends directly on $L$ and $p$ since they
+determine the size of matrix $\mathbf{A}$ and therefore affect the OLS function
+execution time.
+It is worthy of note that execution time also depends on mean\_error because it
+determines how many times OVECM will recalculate cointegration vectors which is
+an expensive routine.
+
+Figure~\ref{fig:mapes} shows an example of the in-sample MAPE for 50 iterations.
+When the average of the in-sample MAPEs is above mean\_error new cointegration
+vectors are obtained. In consequence, OVECM performance increases when
+mean\_error increases. However, this could affect accuracy, but
+table~\ref{tab:stats} shows that using an appropriate mean\_error doesn't affect
+accuracy considerable.
 
@@
 
-Table~\ref{tab:stats} shows the out-of-sample performance measures: MAPE, MAE
-and RMSE for AVECM for different offsets, iterations and number of parameters.
-We observed that performance measures differ for different offsets having even
-in some cases different orders. London and New York offsets are order $10^{-3}$
-while Sidney and Tokyo vary between $10^{-3}$ and $10^{-4}$. Despite this
-difference in the order, they all achieve their minimum MAPE, MAE and RMSE
-measurements at 100 iterations. In the case of Tokyo, it also reduces its MAPE
-order. In all cases, we can see that performance measures are improving with
-the number of parameters.  This is because there are more possible options to find
-optimal cointegration parameters.
-
-Figure ~\ref{fig:accuracy} shows part of the out-of-sample forecasts made by
-our proposed AVECM using 100 iterations and offset New York.
-
-\begin{figure}[!h]
-  \centering
-  \includegraphics[width=\textwidth]{img/accuracy}
-  \caption{AVECM out-of-sample forecasting for four currency rates.}
-  \label{fig:accuracy}
-\end{figure}
+Table~\ref{tab:stats} shows in-sample and out-of-sample performance measures:
+MAPE, MAE and RMSE for OVECM, SLVECM and SLARIMA. Test were done using the
+parameters defined in table \ref{tab:params}.  We can see that OVECM has very
+similar performance than SLVECM and this support the theory that cointegration
+vectors vary little in time. Moreover, OVECM also out performed SLARIMA based on
+these three performance measures.
 
 \begin{table*}[ht!]
-\tiny
-\caption{Performance and execution times for min MAPE criteria}
+\caption{Model measures}
 \label{tab:stats}
 \begin{center}
-\begin{tabular}{|l|l|l|c|c|c|c|c|}
+\tiny
+\begin{tabular}{|l|l|c|c|c|c|c|c|c|c|}
 \hline
-\multicolumn{3}{|c}{Parameters} & 
-\multicolumn{3}{|c}{Out-of-sample Performance} &
-\multicolumn{2}{|c|}{Time[s]} \\ 
-\hline
-
-\hline
-Offset & Iterations & $nparam$ & 
-MAPE & MAE& RMSE& 
-Sequencial & Parallel \\
-\hline
- \multirow{12}{*}{19440 (London)} &
- \multirow{3}{*}{10} 
-  &  12 & 1.8481E-03 & 5.1974E-04 & 7.9714E-04 & 2.64 & 2.61    \\
- &&  24 & 1.8793E-03 & 5.2935E-04 & 8.0128E-04 & 7.38 & 3.03 \\ 
- &&  47 & 1.8804E-03 & 5.2477E-04 & 7.9274E-04 & 18.86 & 3.07 \\ 
- \cline{2-8} 
- & \multirow{3}{*}{50} 
- &  12&  1.5760E-03 & 3.8559E-04 & 6.5579E-04 & 12.49 & 3.99 \\
- && 24 & 1.5645E-03 & 3.7379E-04 & 6.3128E-04 & 36.69 & 6.45 \\
- && 47 & 1.5757E-03 & 3.7336E-04 & \textbf{6.2828E-04} & 92.30 & 5.75 \\
- \cline{2-8} 
- & \multirow{3}{*}{100} 
- &  12&  1.5760E-03 & 3.8559E-04 & 6.5579E-04 & 25.00 & 5.90 \\ 
- && 24 & \textbf{1.5501E-03} & \textbf{3.6575E-04} & 6.4058E-04 & 70.58 & 9.74 \\ 
- && 47 & 1.5812E-03 & 3.8773E-04 & 6.4740E-04 & 185.05 & 11.09 \\
- \cline{2-8} 
- & \multirow{3}{*}{150} 
- &  12&  1.5760E-03 & 3.8559E-04 & 6.5579E-04 & 37.09 & 7.55 \\ 
- && 24 & 1.5501E-03 & 3.6575E-04 & 6.4058E-04 & 104.52 & 14.33 \\
- && 47 & 1.5812E-03 & 3.8773E-04 & 6.4740E-04 & 292.84 & 15.44 \\
+\multicolumn{4}{|c|}{Model} & \multicolumn{3}{|c|}{In-sample} &
+\multicolumn{3}{|c|}{Out-of-sample} \\ 
 \hline
 \hline
- \multirow{12}{*}{21240 (New York)} &
- \multirow{3}{*}{10} 
-  &  12 & 1.8358E-03 & 3.4906E-04 & 4.9314E-04 & 2.64 & 2.45 \\
- &&  24 & 1.8287E-03 & 3.4661E-04 & 4.9140E-04 & 7.38 & 3.03 \\
- &&  47 & 1.8402E-03 & 3.4937E-04 & 4.9298E-04 & 18.86 & 3.00 \\
- \cline{2-8} 
- & \multirow{3}{*}{50} 
- &  12&  1.6911E-03 & 3.3614E-04 & 4.8197E-04 & 12.49 & 4.00 \\
- && 24 & 1.6871E-03 & 3.3546E-04 & 4.8198E-04 & 36.69 & 6.15 \\
- && 47 & 1.6991E-03 & 3.3790E-04 & \textbf{4.8141E-04} & 92.30 & 6.17 \\
- \cline{2-8} 
- & \multirow{3}{*}{100} 
- &  12&  1.6911E-03 & 3.3614E-04 & 4.8197E-04 & 25.00 & 6.03 \\ 
- && 24 & \textbf{1.6857E-03} & \textbf{3.3510E-04} & 4.8308E-04 & 70.58 & 9.34 \\ 
- && 47 & 1.6978E-03 & 3.3766E-04 & 4.8285E-04 & 185.05 & 10.36 \\
- \cline{2-8} 
- & \multirow{3}{*}{150} 
- &  12& 1.6911E-03 & 3.3614E-04 & 4.8197E-04 & 37.09 & 7.77 \\   
- && 24 &1.6857E-03 & 3.3510E-04 & 4.8308E-04 & 104.52 & 13.84 \\
- && 47 &1.6978E-03 & 3.3766E-04 & 4.8285E-04 & 292.84 & 13.06 \\
+Method & $L$ & Parameters & e & 
+MAPE & MAE& RMSE& 
+MAPE & MAE& RMSE \\
 \hline
+ OVECM  &   100  &  P=2& 0.0026  &  0.00263&  0.00085&  0.00114&  0.00309&  0.00094&  0.00131\\
+ OVECM  &   400  &  P=5& 0.0041  &  0.00378&  0.00095&  0.00127&  0.00419&  0.00103&  0.00143\\
+ OVECM  &   700  &  P=3& 0.0032  &  0.00323&  0.00099&  0.00130&  0.00322&  0.00097&  0.00132\\
+ OVECM  &   1000 &  P=3& 0.0022  &  
+ \textbf{0.00175}&  \textbf{0.00062}&  \textbf{0.00087} &  
+ \textbf{0.00172}&  \textbf{0.00061}&  \textbf{0.00090}\\
 \hline
- \multirow{12}{*}{24120 (Sidney)} &
- \multirow{3}{*}{10} 
-  &  12 & 8.5847E-04 & 2.2284E-04 & 3.1202E-04 & 2.64 & 2.41 \\ 
- &&  24 & 7.9127E-04 & 1.9526E-04 & 2.9580E-04 & 7.38 & 2.86 \\ 
- &&  47 & 8.0491E-04 & 1.9578E-04 & 2.9669E-04 & 18.86 & 3.13 \\ 
- \cline{2-8} 
- & \multirow{3}{*}{50} 
- &  12& 9.1471E-04 & 2.2555E-04 & 3.1371E-04 & 12.49 & 4.02 \\ 
- && 24 &7.3172E-04 & 1.8017E-04 & 2.8054E-04 & 36.69 & 5.45 \\
- && 47 &7.3172E-04 & 1.8017E-04 & 2.8054E-04 & 92.30 & 6.48 \\
- \cline{2-8} 
- & \multirow{3}{*}{100} 
- &  12& 8.0458E-04 & 2.0706E-04 & 2.9542E-04 & 25.00 & 5.90 \\ 
- && 24 &\textbf{6.2547E-04} & 1.6451E-04 & 2.5945E-04 & 70.58 & 9.23 \\ 
- && 47 &6.3424E-04 & \textbf{1.6400E-04} & \textbf{2.5849E-04} & 185.05 & 11.13 \\
- \cline{2-8} 
- & \multirow{3}{*}{150} 
- &  12&  7.2437E-04 & 1.8782E-04 & 2.8016E-04 & 37.09 & 7.83 \\ 
- && 24 & 6.2547E-04 & 1.6451E-04 & 2.5945E-04 & 104.52 & 12.42 \\
- && 47 & 6.2547E-04 & 1.6451E-04 & 2.5945E-04 & 292.84 & 13.95 \\
+ SLVECM  &   100 &  P=2& -  &  0.00262&  0.00085&  0.00113&  0.00310&  0.00095&  0.00132\\
+ SLVECM  &   400 &  P=5& -  &  0.00375&  0.00095&  0.00126&  0.00419&  0.00103&  0.00143\\
+ SLVECM  &   700 &  P=3& -  &  0.00324&  0.00099&  0.00130&  0.00322&  0.00098&  0.00132\\
+ SLVECM  &   1000 &  P=3& -  &  
+ \textbf{0.00174}&  \textbf{0.00061}&  \textbf{0.00087}&  
+ \textbf{0.00172}&  \textbf{0.00061}&  \textbf{0.00090}\\
 \hline
-\hline
- \multirow{12}{*}{24840 (Tokyo)} &
- \multirow{3}{*}{10} 
-  &  12 & 1.0787E-03 & 3.3948E-04 & 5.5145E-04 & 2.64 & 2.45 \\ 
- &&  24 & 1.0823E-03 & 3.3992E-04 & 5.4697E-04 & 7.38 & 2.96 \\ 
- &&  47 & 1.0885E-03 & 3.4462E-04 & 5.5231E-04 & 18.86 & 3.05 \\
- \cline{2-8} 
- & \multirow{3}{*}{50} 
- &  12& 1.0753E-03 & 3.4955E-04 & 5.7100E-04 & 12.49 & 3.97 \\ 
- && 24 &1.1042E-03 & 3.5395E-04 & 5.6747E-04 & 36.69 & 6.08 \\ 
- && 47 &1.1238E-03 & 3.5794E-04 & 5.7330E-04 & 92.30 & 6.67 \\ 
- \cline{2-8} 
- & \multirow{3}{*}{100} 
- &  12& \textbf{8.4691E-04} & \textbf{2.8979E-04} & \textbf{5.2553E-04} & 25.00 & 5.74 \\ 
- && 24 &8.6649E-04 & 2.9405E-04 & 5.2874E-04 & 70.58 & 10.29 \\
- && 47 &8.8877E-04 & 2.9821E-04 & 5.3796E-04 & 185.05 & 9.00 \\
- \cline{2-8} 
- & \multirow{3}{*}{150} 
- &  12& 8.5021E-04 & 2.9579E-04 & 5.2707E-04 & 37.09 & 7.54 \\ 
- && 24 &8.6350E-04 & 2.9183E-04 & 5.2837E-04 & 104.52 & 13.05 \\
- && 47 &8.8597E-04 & 2.9594E-04 & 5.3759E-04 & 292.84 & 13.82 \\
+SLARIMA & 100  &p,d,q=2,1,1 & - & 0.00285  &  0.00110  &  0.00308  &  0.00312  &0.00098  &  0.00144 \\
+SLARIMA & 400  &p,d,q=1,1,1 & - & 0.00377  &  0.00101  &  0.00128  &  0.00418 & 0.00106 & 0.00145 \\
+SLARIMA & 700  &p,d,q=2,1,1 & - & 0.00329  &  0.00102  &  0.00136  &  0.00324  & 0.00097  &  0.00133 \\
+SLARIMA & 1000 &p,d,q=2,1,1 & - & \textbf{0.00281}  & \textbf{0.00074}  &  \textbf{0.00105}  &  \textbf{0.00177} & \textbf{0.00063}  &  \textbf{0.00092} \\
 \hline
@@
 
-\subsection{Execution times} \label{sec:exectimes}
-We ran AVECM 100 ($outit$ in algorithm~\ref{alg:AVECM}) iterations.
-Table~\ref{tab:stats} shows execution times and performance measures for
-different times of the market, in-sample iterations and number of parameters used.
-The number of parameters represents the possible combinations of values for $L$ and
-$P$. The $L$ parameter was always chosen between 100 and 4000 and $nparams$
-determines how many values in this range were considered. $p$ always took values
-between 1 and 5. Execution times were measured using the Time python library.
+We can also notice that in-sample performance in OVECM and SLVECM is related
+with the out-of-sample performance.  This differs with SLARIMA which models with
+good in-sample performance are not necessarily good out-of-sample models.
+Moreover OVECM outperformed SLARIMA using the same window size.
 
-Execution time depends directly on $L$ and $p$ since they determine the size of
-matrix $\mathbf{A}$ and therefore affects the OLS function execution time.
-Therefore, if we try more combinations of $L$ and $p$ (increasing $nparam$) the
-algorithm will take longer. However, the idea is to improve accuracy measures
-ensuring execution times below 10 seconds (the time series frequency).
+Figure ~\ref{fig:accuracy} shows the out-of-sample forecasts made by our
+proposal OVECM with the best parameters found based on table~\ref{tab:stats}
+which follows the time series very well.
 
-Table ~\ref{tab:stats} shows that best performance accuracy measures are
-achieved in times near or below 10 seconds in the parallel version. Contrarily,
-sequential times are high, above 10 seconds in most of the cases.
-
-Despite there being some parallel execution times above 10 seconds, the
-performance measures didn't improve and so they can be dismissed.
-
-Figure~\ref{fig:extimes} shows sequential and parallel time and speed-up
-computing time for VECM with 100 iterations.
-
-\begin{figure}[!h]
+\begin{figure}[h]
+  \vspace{-0.2cm}
   \centering
-  \includegraphics[width=0.7\textwidth]{img/extimes}
-  \caption{Computing time of sequential and parallel algorithm is shown in the
-  upper figure. Speed-up is shown below.}
-  \label{fig:extimes}
+   \includegraphics[width = 10cm]{img/accuracy} 
+  \caption{OVECM forecasting accuracy example for 50 minutes using $L = 1000$ and $p =3$}
+  \label{fig:accuracy}
  \end{figure}
 
-Execution times do not consider the loading data time, just the finding of
-best parameters and the building the VECM model. The time MPI spends
-transferring data and synchronising process is about two seconds independently
-of the number of process considered.
+\section{Conclusions} \label{sec:conclusions}
 
-\section{Conclusions}
-\label{sec:conclusions}
-Cointegration in financial time series has been largely studied and the 
-Johansen method is commonly used to obtain it. 
-In practice it has been found that cointegration relations change with time. 
-However, cointegration model-based such as VECM assumes that cointegration
-remains unchanged in time. 
-We empirically showed that the Johansen method is sensitive to the number
-of lags but also to the amount of data considered.
+A new online vector error correction method was presented. 
+We have shown that our proposed OVECM considerably reduces execution times
+without compromising solution accuracy.  OVECM was compared with VECM and ARIMA
+with the same sliding window sizes and OVECM outperformed both in terms of
+execution time. Traditional VECM slightly outperformed our proposal but the
+OVECM execution time is lower.
+This reduction of execution time is mainly because OVECM avoids the cointegration vector
+calculation using the Johansen method. The condition for getting new vectors is
+given by the mean\_error variable which controls how many times the Johansen method is
+called. Additionally, OVECM introduces matrix optimisation in order to get the
+new model in an iterative way.
+We could see that our algorithm took much less than a minute at every
+step. This means that it could also be used with higher frequency data and would
+still provide responses before new data arrives.
 
-Moreover, we introduced the notion of {\em percentage of cointegration\/} and
-have found that out-of-sample forecast performance is related to the value of
-this figure in the last samples.  We used this information to set the model
-parameters.  Our proposal AVECM consists then on an adaptive algorithm to update VECM
-parameters every time that new data is available. These parameters are found by
-maximising the percentage of cointegration of the last samples or iterations.
+For future study, it would be interesting to improve the out-of-sample forecast
+by considering more explicative variables, to increase window sizes or trying new
+conditions to obtain new cointegration vectors.
 
-Determining VECM parameters was the most expensive routine and it was run
-using parallel processes using MPI which allowed a grid search within a range of
-values for $L$ and $p$ to be made.
-Tests were done using real currency rates data. 
-Our proposal was tested in several scenarios using different times of
-the day to show its behaviour related to the opening times of
-financial markets.
-
-Results showed that our proposed AVECM improves performance measures by finding
-parameters of $L$ and $p$ maximising the percentage of cointegration.
-These facts were consistent for different times of the day. 
-We found that increasing the number of parameters will always lead to
-better performance measures. Furthermore, in the case of Tokyo 
-financial market opening time, MAPE reduced an order from $10^{-3}$ to $10^{-4}$.
-
-The parallel implementation allowed the execution times to be reduced
-more than ten times and therefore a response time was obtain before 10
-seconds. Since we used 10-second frequency we can say that our proposal is suitable
-for use in an online context for real applications because response times
-were less than this frequency.
-
-For future study, it would be interesting to explore the relation between
-cointegration and performance in order to propose new criteria for
-improving VECM parameters.
+Since OVECM is an online algorithm which optimises processing time, it could be
+used by investors as an input for strategy robots. Moreover, some technical
+analysis methods could be based on its output.

### 2016-07-06 src/3_TimeSeries.tex
Adding first proposal

@@
 
-\section{Vector error correction model}
+\section{Vector error correction model}\label{sec:vecm}
 Vector error correction model (VECM)  \cite{engle87} describe the joint behaviour of a set of cointegrated I(1) time series \cite{banerjee1993} and can be derived from the simple VAR model.

### 2016-07-06 src/5_1_Proposal.tex
Adding first proposal

--- 2014-11-27 src/5_Proposal.tex
@@
-\chapter{Online VECM proposal}
-\section{\uppercase{Methodology}}
-\label{sec:methodology}
-\noindent 
-\subsection{Online VECM} \label{sec:proposal}
+\chapter{Adaptive parallel algorithm for Vector error correction model}
+VECM (presented in section \ref{sec:vecm}) has been
+broadly used with low frequency time series. We aimed to adapt VECM to be used
+in finance with high frequency stream data.
 
-Since VECM parameter estimation is computationally expensive, we propose an
-online version of VECM (OVECM).  OVECM considers only a sliding window of the
-most recent data. Moreover, since cointegration vectors represent long-run
-relationships which vary little in time, OVECM determines firstly if they require calculation.
+We observed that the number of cointegration relationships varies with the
+length of historical data used. Moreover, parameters that increased these
+relationships in time led to better forecasting performance. Our proposal,
+called an Adaptive VECM (AVECM) is to make a parameters grid search that
+maximises the number of cointegration relationships in the near past. To ensure
+the search can be executed fast enough, we used a distributed environment.
 
-OVECM also implements matrix optimizations in order to reduce execution time,
-such as updating matrices with new data, removing old data and introducing new
-cointegration vectors.
+The methodology was tested using four 10-second frequency time series of the
+Foreign Exchange market. We compared our proposal with ARIMA and the naive
+forecast of the random walk model. Numerical experiments showed that on average
+AVECM performed better than ARIMA and random walk. Additionally, AVECM
+significantly improved execution times with respect to its serial version.
 
-Algorithm~\ref{alg:proposal} shows our OVECM proposal which considers the
-following:
+\section{Introduction}
+\label{sec:introduction}
+In finance, it is common to find variables with long-run equilibrium
+relationships. This is termed cointegration and reflects the idea that some sets
+of variables cannot wander too far from each other. Cointegration means that one
+or more linear combinations of these variables are stationary even though
+individually they are not. Some models, such as the Vector Error Correction
+(VECM), see \cite{engle87}, take advantage of this property and describe the
+joint behaviour of several cointegrated variables. VECM introduces this long-run
+relationship among a set of cointegrated time series as an error correction
+term. These time series must be integrated of order 1, denoted I(1), i.e. they
+become stationary at their first differences. In finance, I(1) time series are
+very common and to introduce cointegration restrictions in models often improves
+forecasting, see \cite{duy1998}. Therefore, VECM has been widely adopted in
+financial applications, among others: \cite{mukherjee1995}, \cite{seong2013},
+\cite{maysami2000} and \cite{arestis2001}. VECM has also been used in pair
+trading, see \cite{herlemont2003}, or models with more than two variables, see
+for example \cite{mukherjee1995} and \cite{engle2004}).
 
-\begin{itemize}
-\item The function \texttt{getJohansen} returns cointegration vectors given by
-the Johansen method considering the trace statistic test at 95\%
-significance level.
-\item The function \texttt{vecMatrix} returns the matrices~(\ref{eq:vecA})
-and~(\ref{eq:vecB}) that allows VECM to be solved.
-\item The function \texttt{vecMatrixOnline} returns the
-matrices~(\ref{eq:vecA}) and~(\ref{eq:vecB}) aggregating new data and removing
-the old one, avoiding calculation of the matrix $\mathbf{A}$ from scratch.
-\item Out-of-sample outputs are saved in the variables 
-$\mathbf{Y}_{\text{true}}$ and $\mathbf{Y}_{\text{pred}}$.
-\item The model is solved using OLS.
-\item In-sample outputs are saved in the variables $\Delta
-\mathbf{y}_{\text{true}}$ and $\Delta \mathbf{y}_{\text{pred}}$.
-\item The function \texttt{mape} gets the in-sample MAPE for the $l$ time
-series.
-\item Cointegration vectors are obtained at the beginning and when they are required to be updated. This updating is decided based on the in-sample MAPE of the last $n$ inputs. The average of all
-MAPEs are stored in the variable $e$. If the average of MAPEs
-($\texttt{mean}(e)$) is above a certain error given by the mean\_error threshold, cointegration vectors are updated.
-\item If new cointegration vectors are required, the function
-\texttt{vecMatrixUpdate} only updates the corresponding columns of matrix
-$\mathbf{A}$ affected by those vectors (see equation~\ref{eq:vecA}).
-\end{itemize}
+VECM parameters are obtained using the ordinary least squares (OLS) method,
+developed by \cite{golub1980}. Since OLS involves many calculations, the
+parameter estimation is computationally expensive when the number of lagged
+values and data increases. Moreover, obtaining cointegration vectors is also an
+expensive routine. This is a main limitation to use VECM with stream data with
+high frequency. \cite{chen2003} addresses the advantage of distributed
+processing over conventional rolling window processing.
 
-\begin{algorithm}[ht]
+Our aims were to study if a parallel version of VECM can be used with high
+frequency stream data.  Our approach was to determine adaptively the number of
+observations and lags of VECM which maximise cointegration relations in the
+past.  This search is computationally expensive because the Johansen method, see
+\cite{johansen1995}, is required to find cointegration vectors, which is of
+order $O(n^3)$. Therefore, our proposal, called Adaptively Vector Error
+Correction (AVECM), is to parallelise this search in order to get new parameters
+before new data arrives.  Model effectiveness is focused on out-of-sample
+forecast rather than in-sample fitting. This criterion allows AVECM prediction
+capability to be expressed rather than just explaining data history.  The
+forecast capability of our method was measured using MSE and the Theil's
+$U$-statistic, see \cite{theil1966}, widely used in economic forecast. Tests
+were run using four currency rates: Euro (EUR) to United States Dollar (USD)
+(EURUSD), British Pound (GBP) to USD (GBPUSD), USD to Swiss Franc (CHF) (USDCHF)
+and USD to Japanese Yen (JPY) (USDJPY) with a 10-second frequency.
+
+\section{Methodology} \label{sec:methodology}
+\subsection{Motivation} \label{sec:proposal}
+
+Cointegration vectors can be found applying the Johansen method which uses a
+sample of the last historical data. However, VECM assumes cointegration vectors
+do not change in time.  In fact, \cite{gregoryETal1996} addresses that the
+long-run relationships between the time series  might change due to several
+economic factors that can lead to structural breaks in the cointegration
+relationship.  In order to show that the number of cointegration vectors depends
+on the amount $L$ of historical data and the number of lags $p$ in the VECM, we
+used a grid search.  We defined a grid of possible values for $L$ and $p$. $L$
+goes throughout $[2,14]$ hours (1 $hour = 360$ data points) with a step size of
+4 hours and $p$ throughout $[1,5]$ with step size of 1. The first experiment
+consisted in determining the number of cointegration vectors for all
+combinations of $L$ and $p$.  We used four forex rates: EURUSD, GBPUSD, USDCHF
+and USDJPY with 10-second frequency. Data started at 13:00 GMT of the 13th of
+August 2014, when the New York and London financial markets opened.
+
+Figure \ref{fig:hists} shows the distribution of the number of cointegration
+vectors given by the Johansen method for different values of $L=[2,6,10,14]$
+hours  and $p=1$ . This procedure was carried out by a sliding window of
+historical data moving 1000 times.
+
+\begin{figure}[!h]
+  %\vspace{-0.8cm}
+  \centering
+  \includegraphics[width=0.9\textwidth]{img/Fig1}
+  \caption{Distribution of the number of cointegration vectors using $p=1$ lags.
+  Four possible values for windows size $L$ are shown (2, 6, 10 and 14) hours (1
+  hour = 360 data points).}
+  \label{fig:hists}
+\end{figure}
+
+From section \ref{sec:background} we know that $r=0$ means no cointegration and
+$r=l$ (we are using four rates, so $l=4$) reveals that no process is I(1) but
+stationary.  The interesting cases of cointegration are those where $r$ lies
+strictly between $0$ and $4$, i.e. $0<r<4$.
+
+In order to measure the extent of cointegration, we introduce a
+{\em percentage of cointegration\/} as following:
+\begin{equation} \label{eq:pcoint}
+PC = 
+\frac{\#\{ it \mid \text{$it$ has $r$ c.v. with $0<r<l$}\}}
+     {\#it}\times 100
+\end{equation}
+where c.v. stands for cointegration vectors and $it$ is the number of iterations.
+
+The goal of our next experiment was to find a relation between this ratio $PC$
+and the performance measure MSE (see equation \ref{eq:MSE}). $L$ was defined
+between $[700,1500]$ data points with step size 100 and $p$ took values between
+$[1,5]$ with step size 1.
+
+\begin{figure}[ht!]
+  \centering
+  \includegraphics[width=0.9\textwidth]{img/Fig2}
+  \caption{MSE versus the percentage of cointegration considering 1000
+  iterations. Best windows size $L$ found was 1060. Below MSE versus $L$ shows a
+  rapidly decreasing behaviour, founding minimum at PC = 80\% }
+  \label{fig:cointvsmse}
+\end{figure}
+
+Figure~\ref{fig:cointvsmse} shows the relation between MSE and $PC$ and $L$.  We
+found that better cointegration percentage leads to better performance accuracy
+in terms of reducing MSE.
+
+Therefore we proposed to choose $L$ and $p$ in order to maximise the percentage
+of cointegration $PC$ in the near past. This process was done every time that
+new data was processed. However, this search can be slow if we try different
+values for $L$ and $p$ and therefore we distributed this calculation in order to
+reduce searching time.
+
+Our proposal is then a modified version of VECM, with parameter $p$ and the
+amount of historical data used $L$ obtained at every step. Only the search of
+$L$ and $p$ was done in a distributed environment, since this is the most
+expensive routine. Our proposal is called Adaptive Vector Error Correction model
+(AVECM). AVECM is detailed in the algorithm \ref{alg:AVECM} which summarises
+our proposal.
+
+\begin{algorithm}[ht!]
 \begin{algorithmic}[1]
@@
 $\mathbf{y}$: matrix with $N$ input vectors and $l$ time series\\
-$p$: number of past values \\
-$L$: sliding window size ($L<N$) \\
-$\text{mean\_error}$: MAPE threshold \\
-$n$: interpolation points to obtain MAPE \\
+$j$: Starting point of testing \\
+$it$: Ending point of testing \\
+$ps$: list of $p$ values \\
+$Ls$: list of $L$ values ($L<N$) \\
+$m$: Iterations to determine parameters ($m < N-L$)\\
 \ENSURE  $\,$ \\
-$\{\mathbf{y}_{\text{pred}}[L+1],\dots,\mathbf{y}_{\text{pred}}[N]\}$: model predictions 
-\FOR { $i =0$ to $N-L$ }
-    \STATE $\mathbf{y}_i \gets \mathbf{y}[i:i+L]$
-	\IF {i = 0 }
-	    \STATE{$v \gets \texttt{getJohansen}(\mathbf{y}_i,p)$}
-	    \STATE{$[\mathbf{A} \quad \mathbf{B}] \gets
-        \texttt{vecMatrix}(\mathbf{y}_i,p,v)$}
-	\ELSE
-	    \STATE{$[\mathbf{A} \quad \mathbf{B}] \gets
-        \texttt{vecMatrixOnline}(\mathbf{y}_i,p,v,\mathbf{A},\mathbf{B})$}
-        %\STATE $\Delta \mathbf{Y}_{\text{true}}[i] \gets \mathbf{B}[-1,:]$
-        \STATE $\Delta \mathbf{Y}_{\text{pred}}[i] \gets \mathbf{A}[-1,:] \times \mathbf{X}$
-    \ENDIF
-    \STATE $\mathbf{X} \gets \texttt{OLS} (\mathbf{A},\mathbf{B})$%\mathbf{(A^\top A)^{-1}A^\top B}$
-    %\STATE $\Delta \mathbf{y}_{\text{true}} \gets \mathbf{B}[-n,:]$
-    %\STATE $\Delta \mathbf{y}_{\text{pred}} \gets \mathbf{A}[-n,:] \times \mathbf{X}$
-    \STATE $e \gets \texttt{mape}(\mathbf{B}[-n,:],\mathbf{A}[-n,:] \times \mathbf{X})$
-    %\STATE $e \gets \texttt{mape}(\Delta \mathbf{y}_{\text{true}}, \Delta
-    %\mathbf{y}_{\text{pred}})$
-    \IF {$\texttt{mean}(e) > \text{mean\_error}$}
-	    \STATE{$v \gets \texttt{getJohansen}(\mathbf{y}_i,p)$}
-	    \STATE{$\mathbf{A} \gets
-        \texttt{vecMatrixUpdate}(\mathbf{y}_i,p,v,\mathbf{A})$}
-        \STATE $\mathbf{X} \gets \texttt{OLS} (\mathbf{A},\mathbf{B})$%\mathbf{(A^\top A)^{-1}A^\top B}$
-        %\STATE $\mathbf{X} \gets \mathbf{(A^\top A)^{-1}A^\top B}$
-    \ENDIF
+$\{ \hat{\mathbf{y}}[1],\dots,\hat{\mathbf{y}}[it]\}$: prediction vectors \\
+\FOR { $i =j$ to $it$ }
+   \STATE $\mathbf{Y} \gets \mathbf{y}[:,i-1]$
+    \STATE $L,p \gets
+    \texttt{get\_best\_params}(Ls,ps,m,\mathbf{Y})$
+        \STATE $model = \text{VECM}(\mathbf{Y},L, p)$
+        \STATE $\hat{\mathbf{y}}[i-j] = model.predict()$
 \ENDFOR
-\STATE $\mathbf{Y}_{\text{true}} \gets \mathbf{Y}[L+1:N]$
-\STATE $\mathbf{Y}_{\text{pred}}\gets 
-\mathbf{Y}[L:N-1] +\Delta \mathbf{Y}_{\text{pred}}$
-%\STATE $\text{MAPE} \gets \texttt{mape}(\Delta \mathbf{Y}_{\text{true}}, \Delta
-%\mathbf{Y}_{\text{pred}})$
 \end{algorithmic}
-\caption{OVECM: Online VECM}
-\label{alg:proposal}
+\caption{AVECM: Adaptive VECM.}
+\label{alg:AVECM}
 \end{algorithm}
 
-Our proposal was compared against VECM and ARIMA. Both algorithms were adapted
-to an online context in order to get a reasonable comparison with our proposal
-(see algorithms \ref{alg:SLVECM} and \ref{alg:SLARIMA}). VECM and ARIMA are
-called with a sliding window of the most recent data, whereby the models are
-updated at every time step.
+The input of AVECM is time series prices which are cointegrated. The starting
+point of testing is $j$ and the total number of iterations is $it$. We need to
+ensure that $j$ is at least the maximum value of $Ls$. $Ls$ and $ps$ are the
+possible values for $L$ and $p$.  The function {\bf get\_best\_params} makes
+this grid search on the two vector lists $Ls$ and $ps$ and returns the
+parameters $L$ and $p$ which maximise the percentage of cointegration $PC$ (see
+equation~\ref{eq:pcoint}) for a pre-defined number of iterations $m$. This
+function is implemented in a distributed environment, thus ensuring a response
+before new data is available.  After $L$ and $p$ parameters are found, VECM is
+built and used to forecast the next data point.
 
-\begin{algorithm}[ht]
-\begin{algorithmic}[1]
-\REQUIRE $\,$ \\
-$\mathbf{y}$: matrix with $N$ input vectors and $l$ time series\\
-$p$: number of past values \\
-$L$: sliding window size ($L<N$) \\
-\ENSURE  $\,$ \\
-$\{ \mathbf{y}_{\text{pred}}[L+1],\dots,\mathbf{y}_{\text{pred}}[N]\}$: model predictions 
-\FOR { $i =0$ to $N-L$ }
-    \STATE $\mathbf{y}_i \gets \mathbf{y}[i:i+L+1]$
-        \STATE $model = VECM(\mathbf{y}_i, p)$
-        \STATE $\mathbf{Y}_{\text{pred}}[i] = model.predict(\mathbf{y}[i+L])$
-\ENDFOR
-\STATE $\mathbf{Y}_{\text{true}} = \mathbf{y}[i+L+1:N] $
-\end{algorithmic}
-\caption{SLVECM: Sliding window VECM}
-\label{alg:SLVECM}
-\end{algorithm}
+\subsection{Model comparison} \label{sec:random}
+We compared our proposal, in terms of performance, with the naive forecast of
+the random walk model and ARIMA. It is still difficult to outperform the random
+walk model for standard econometric forecasting models  despite its
+simplicity, see \cite{lo2011}. The random walk model is defined as:
 
-Since we know out time series are I(1) SLARIMA is called with $d=1$. ARIMA is
-executed for every time series.
+\begin{equation}
+\mathbf{y}_t = \mathbf{y}_{t-1} + \epsilon_{t}
+\label{rwmodel}
+\end{equation}
 
-\begin{algorithm}[ht]
-\begin{algorithmic}[1]
-\REQUIRE $\,$ \\
-$\mathbf{y}$: matrix with $N$ input vectors and $l$ time series\\
-$p$: autoregressive order \\
-$d$: integrated order\\
-$q$: moving average order\\
-$L$: sliding window size ($L<N$) \\
-\ENSURE  $\,$ \\
-$\{ \mathbf{y}_{\text{pred}}[L+1],\dots,\mathbf{y}_{\text{pred}}[N]\}$: model predictions 
-\FOR { $i =0$ to $N-L$ }
-\FOR { $j =0$ to $l-1$ }
-    \STATE $\mathbf{y}_i \gets \mathbf{y}[i:i+L+1,j]$
-        \STATE $model = ARIMA(\mathbf{y}_i, (p,d,q))$
-        \STATE $\mathbf{Y}_{\text{pred}}[i,j] = model.predict(\mathbf{y}[i+L,j])$
-\ENDFOR
-\ENDFOR
-\STATE $\mathbf{Y}_{\text{true}} = \mathbf{y}[i+L+1:N,:] $
-\end{algorithmic}
-\caption{SLARIMA: Sliding window ARIMA}
-\label{alg:SLARIMA}
-\end{algorithm}
+The naive forecast of the time series difference $\hat{\mathbf{y}}_{t+1}$ for
+the random walk model is defined as:
+\begin{equation}
+\hat{\mathbf{y}}_{t+1} = \mathbf{y}_t + \hat{\epsilon}_{t+1} 
+\end{equation}
+\noindent where  $\hat{\epsilon}_{t+1} = \epsilon_{t}$.
 
-Both OVECM and SLVECM time complexity is dominated by Johansen method which is
-$O(n^3)$. Thus, both algorithms order is $O(Cn^3)$ where $C$ is the number of
-iterations.
+On the other hand, ARIMA is widely used to forecast returns in finance, see
+\cite{tsay2005}. A process can be modelled as an ARIMA$(p,d,q)$ model if
+$\mathbf{x}_t=\Delta^d \mathbf{y}_t $, i.e after differencing $d$ times the time
+series $\mathbf{y}_t$,  we get an ARMA$(p,q)$. Since we are modelling returns,
+we used $d=1$.
 
@@
 
-Forecast performance is evaluated using different methods. We have chosen three
-measures usually used:
+Forecast performance was evaluated using two different methods which are
+frequently used in finance:
 \begin{description}
-\item[MAPE] Mean Average Percent Error which presents forecast errors as a
-percentage.
-\begin{equation}\label{eq:MAPE}
-\text{MAPE} = \frac{1}{N} \sum_{t=1}^{N} 
-\frac{\left|\mathbf{y}_t-\hat{\mathbf{y}}_t\right|}{\left|\mathbf{y}_t\right|}
- \times 100 
+\item
+{\bf MSE},  Mean Square Error measures the distance between forecasts
+and the true values and large deviations from the true value have a
+large impact due to squaring forecast error.
+\begin{equation}\label{eq:MSE}
+\text{MSE} = 
+\frac{\displaystyle \sum_{t=1}^{N} (\mathbf{y}_t-\hat{\mathbf{y}}_t)^2}{N}
 \end{equation}
-\item[MAE] Mean Average Error which measures the distance between forecasts to the
-true value.
-\begin{equation}\label{eq:MAE}
-\text{MAE} = \frac{1}{N} \sum_{t=1}^{N} 
-\left| 
-\mathbf{y}_t-\hat{\mathbf{y}}_t
-\right| 
-\end{equation}
-\item[RMSE] Root Mean Square Error also measures the distance between forecasts
-to the true values but, unlike MAE, large deviations from the true value have a
-large impact on RMSE due to squaring forecast error.
-\begin{equation}\label{eq:RMSE}
-\text{RMSE} = \sqrt{
-\frac{\displaystyle \sum_{t=1}^{N} (\mathbf{y}_t-\hat{\mathbf{y}}_t)^2}{N}}
-\end{equation}
+\item {\bf $U$-statistic}, the Theil's $U$-statistic, presented by
+\cite{theil1966}, is a unit free measure obtained as the ratio between the root
+MSE (RMSE) of a model and the RMSE of the naive random walk model. A
+$U$-statistic less than 1 implies the performance is better than the naive
+model.
 \end{description}
 
-\subsection{Model selection} \label{sec:pselection}
-Akaike Information Criterion (AIC) is often used in model selection where AIC
-with smaller values are preferred since they represent a trade-off between bias
-and variance.  AIC is obtained as follows:
+\section{Experimental results}
+\label{sec:results}
+\subsection{Data} \label{sec:unitroot}
+AVECM tests were carried out using four foreign exchange rates all related to
+USD: EURUSD, GBPUSD, USDCHF and USDJPY. This data was collected from
+\cite{Dukascopy2014}, a free database which gives access to the Swiss Foreign
+Exchange marketplace.
 
-\begin{equation}
-\label{eq:aicformula}
-AIC = \underset{\text{bias}}{-\frac{2l}{N}} + 
-\underset{\text{variance}}{\frac{2k}{N}}
-\end{equation}
+The tests were done using 10-second frequency from ask prices from the 11th to
+the 15th of August 2014. Since one day corresponds to 8640 data points and we
+used 5 days of data, we have 43,200 data points in total.
 
-\noindent where
+\subsection{Unit root tests} \label{sec:unitroottests}
 
-\begin{description}
-\item[l] is the loglikelihood function
-\item[k] number of estimated parameters
-\item[N] number of observations
-\end{description}
+Before running the tests, we firstly checked whether the time series were I(1)
+using the Augmented Dickey Fuller (ADF) test with lags $p=1,2,3,4,5$.
+\cite{mackinnon2010} presented critical values for rejection of hypothesis of a
+unit root: -2.62 (1\%), 1.94 (5\%) and 1.62 (10\%).
+\begin{table*}[ht]
+\label{tab:adf}
+\centering
+\begin{tabular}{llllll}
+\toprule
+{Variable} & {ADF(1)} & {ADF(2)} & {ADF(3)} & {ADF(4)} & {ADF(5)}\\ 
+\midrule
+EURUSD &  -0.052   & -0.054  & -0.054  & -0.054  & -0.054  \\
+GBPUSD &  -0.744  & -0.784  & -0.805  & -0.837  & -0.846  \\
+USDCHF &  -0.476   & -0.493  & -0.493  & -0.495  & -0.502  \\
+USDJPY &  0.357   & 0.360  & 0.360  & 0.367  & 0.367  \\
+$\Delta$ EURUSD & -128.4*  & -128.4*  & -96.85* & -89.12*   & -89.12*\\
+$\Delta$ GBPUSD & -131.4*  & -112.7*  & -102.5* & -92.86*   & -88.29*\\
+$\Delta$ USDCHF & -127.8*  & -127.8*  & -96.94* & -88.82*   & -80.79*\\
+$\Delta$ USDJPY & -135.1*  & -135.1*  & -101.2* & -101.2*   & -101.2*\\
+\bottomrule
+\addlinespace[1ex]
+\multicolumn{6}{l}{ \textsuperscript{*} Indicates significance at 1\% level} \\
+\multicolumn{6}{l}{ \textsuperscript{**} Indicates significance at 5\% level} \\
+\multicolumn{6}{l}{ \textsuperscript{***} Indicates significance at 10\% level} \\
+\multicolumn{6}{l}{MacKinnon critical values for rejection of hypothesis of a unit root are:}\\
+\multicolumn{6}{l}{ -2.62 (1\%), -1.94 (5\%) and -1.62 (10\%)}\\
+\multicolumn{6}{l}{ADF($d$) Augmented Dickey-Fuller test with lag $d$} 
+\end{tabular}
+\caption{Unit roots tests for EURUSD, GBPUSD, USDCHF and USDJPY at 10-second
+frequency.}
+\end{table*}
+Table~\ref{tab:adf} shows that all currency rates cannot reject the unit root
+test in their level form considering different lags, but they rejected it with
+their first differences. This means that all of them are I(1) time series and we
+are allowed to use VECM and therefore our proposed AVECM.
 
-Loglikelihood function is obtained from the Residual Sum of Squares (RSS):
+\subsection{Performance accuracy} \label{sec:performance}
 
-\begin{equation}
-\label{eq:ll}
-l = -\frac{N}{2} \left(1 + ln(2\pi) + ln\left(\frac{RSS}{N}\right)\right) 
-\end{equation}
+Algorithms AVECM, ARIMA and the naive random walk were tested using four days of
+data (from the 12th to the 15th of August 2014).  For AVECM we considered
+different number of iterations (parameter $m$ in algorithm \ref{alg:AVECM}): 10,
+50 and 100. We tried 12, 24 and 47 different pair of combinations for $L$ and
+$p$. Possible values for $L$ were in the interval $[2,14]$ hours and $p$ can
+have values in between $[1,5]$. Best AVECM performance was compared against
+ARIMA and the random walk model.  Table~\ref{tab:stats} shows the out-of-sample
+performance measures: MSE and $U$-statistic for AVECM and ARIMA. In terms of
+both measures we found that AVECM is superior to ARIMA and the naive random walk
+model. We also included the p-value that proves that the difference in the MSE
+is significant at the 99\% significance level in three of the four currency
+rates and at 90\% in the case of GBPUSD. The $U$-statistic shows that AVECM and
+ARIMA are superior to the naive random walk model and that our proposal is also
+superior to ARIMA.
 
-    \begin{tikzpicture}[node distance=2cm]
-    %\node (start) [startstop] {Start};
-    %\node (in1) [io, below of=start] {Input};
-    \node (in1) [io] {Input: y, p, L, $\epsilon$, n};
-    \node (pro1) [process, below of=in1] {$y_i = y[i:i+L]$};
-    \node (dec1) [decision, below of=pro1, yshift=-0.5cm] {i = 0};
-    \node (pro2a) [process, below of=dec1, yshift=-0.5cm] {$[\mathbf{A} \quad \mathbf{B}] =\texttt{ovecm}(\mathbf{y}_i,p,v,\mathbf{A},\mathbf{B})$
-                                                        $\Delta \mathbf{Y}_{\text{true}}[i] = \mathbf{B}[-1,:]$
-                                                        $\Delta \mathbf{Y}_{\text{pred}}[i] = \mathbf{A}[-1,:] \times \mathbf{X}$};
+\begin{table}[!htbp]
+\caption{AVECM performance}
+\label{tab:stats}
+\begin{center}
+\begin{tabular}{ccccccc}
 
-    \node (pro2b) [process, right of=dec1, xshift=4cm] {$v = \texttt{Johansen}(\mathbf{y}_i,p)$ $[\mathbf{A} \quad 
-                                                                \mathbf{B}] = \texttt{vecm}(\mathbf{y}_i,p,v)$};
-    \node (pro3) [process, below of=pro2a, yshift=-0.5cm] {$\mathbf{X} = \text{OLS} (\mathbf{A},\mathbf{B})$ \\
-                                                        $\mathbf{y}_{\text{true}}[i] = \mathbf{y}_i[-n,:]$ \\
-                                                        $\mathbf{y}_{\text{pred}}[i] = \mathbf{A}[-n,:] \times \mathbf{X} +$ \\$ \mathbf{y}_i[-n-1,:-1]$ \\
-                                                        $e = \texttt{mape}(\mathbf{y}_{\text{true}}, \mathbf{y}_{\text{pred}})$};
+& \multicolumn{3}{l}{MSE} & &
+\multicolumn{2}{c}{$U$-statistic} \\ 
+\hhline{~---~--} \\
+ &
+AVECM & ARIMA & p-value & &
+AVECM & ARIMA \\
 
-    \node (dec2) [decision, below of=pro3, yshift=-0.5cm] { $e > \epsilon $ };
+\hline
+ EURUSD & 1.0702 e-09 & 1.1481 e-09 &  9.2509 e-12 & & 0.6863 & 0.7108\\
+ GBPUSD & 1.6630 e-09 & 1.7408 e-09 &  6.9519 e-02 & & 0.6866 & 0.7025\\
+ USDCHF & 5.8503 e-10 & 6.3545 e-10 &  2.8999 e-14 & & 0.6803 & 0.7091\\
+ USDJPY  & 6.3483 e-06 & 6.5194 e-06 &  6.8536 e-05 & & 0.6964 & 0.7057\\
+ \end{tabular}
+ \end{center}
+\end{table}
 
-    \node (pro4a) [process, right of=dec2, xshift=4cm] {$v = \texttt{Johansen}(\mathbf{y}_i,p)$ \\
-                                                        $\mathbf{A} = \texttt{uvecm}(\mathbf{y}_i,p,v,\mathbf{A})$ \\
-                                                        $\mathbf{X} = \texttt{OLS} (\mathbf{A},\mathbf{B})$};
+\subsection{Parallel implementation} \label{sec:parallel}
 
-    \node (pro5) [process, left of=pro2a, xshift=-4cm] {$ i++ $};
+To determine $L$ and $p$ based on maximising the percentage of cointegration
+requires use of the Johansen method which is a computationally expensive
+routine. This procedure is done by the function
+get\_best\_params($Ls,ps,m,\mathbf{Y}$) shown in algorithm \ref{alg:AVECM}.  In
+order to improve the execution time of this search, our proposal included a
+parallel search of VECM parameters using high performance computing.  The main
+objective was to obtain a response before a new data arrived in the following 10
+seconds.
 
-    \node (dec3) [decision, below of=dec2, yshift=-0.5cm] { $i > it $ };
+The Johansen method is already programmed in the Python Statsmodels
+library, see \cite{seabold2010}, and the parallel implementation was developed using
+MPI in Python.  We chose MPI because it allows large-scale parallel applications
+with wide portability to be built, being able to run in large clusters or on
+local computers.  We tested our proposal in a cluster with 2 servers Xeon
+E5-2667 (2.90GHz) of 24 cores each (48 cores in total) and 24GB RAM.  In order
+to compare serial and parallel execution times in AVECM, we set parameter
+$it=100$ in algorithm \ref{alg:AVECM}.
 
-    \node (out1) [io, below of=dec3] {Output: $\Delta \mathbf{y}_{true}, \Delta \mathbf{y}_{pred}$};
+The $L$ parameter was always chosen between 2 and 14 hours and $p$ always took
+values between 1 and 5. Parameter $nparams$ represents the number of pairs
+($L$,$p$) used to maximise the percentage of cointegration.
 
-    %\node (stop) [startstop, below of=out1] {Stop};
+Execution time depends directly on $L$, $p$ and $nparams$ used, since they
+determine the size of matrix $\mathbf{A}$ (see equation \ref{eq:Amatrix}) and
+therefore affect the OLS function execution time.  Therefore, if we try more
+combinations of $L$ and $p$ (increasing $nparams$) the serial algorithm will
+take longer. For this financial time series we are interested in execution times
+below 10 seconds  (the time series frequency).
 
-    %\draw [arrow] (start) -- (in1);
-    \draw [arrow] (in1) -- (pro1);
-    \draw [arrow] (pro1) -- (dec1);
-    \draw [arrow] (dec1) -- node[anchor=east] {no} (pro2a);
-    \draw [arrow] (dec1) -- node[anchor=south] {si} (pro2b);
-    \draw [arrow] (pro2a) -- (pro3);
-    \draw [arrow] (pro2b) |- (pro3);
-    \draw [arrow] (pro3) -- (dec2);
-    \draw [arrow] (dec2) --  node[anchor=south] {si} (pro4a);
-    \draw [arrow] (pro4a) |- (dec3);
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width=0.9\textwidth]{img/Fig3}
+  \caption{Computing time of sequential and parallel algorithm is shown in the
+  upper figure. Speed-up is shown below.}
+  \label{fig:extimes}
+\end{figure}
 
-    \draw [arrow] (dec2) --  node[anchor=east] {no} (dec3);
-    \draw [arrow] (dec3.west) -| node[anchor=east] {no} (pro5);
-    \draw [arrow] (pro5) |- (pro1);
-    \draw [arrow] (dec3) --  node[anchor=east] {si} (out1);
-    %\draw [arrow] (out1) -- (stop);
-    \end{tikzpicture}
+The superior figure in \ref{fig:extimes} shows that best performance accuracy
+measurements are achieved in times near or below 10 seconds in the parallel
+version. Contrarily, serial times are higher, above 10 seconds in most cases.
+Figure~\ref{fig:extimes} also shows the speed-up in computing time for AVECM
+which is near 9X if we use more than 25 parameters ($nparams$).
+
+Execution times do not consider the loading data time, just that of finding best
+parameters and matrix operations. The time MPI spends transferring data and
+synchronising processes is about two seconds independently of the number of
+processes considered. Execution times were measured using the Time python
+library.
+
+\section{Conclusions}
+\label{sec:conclusions}
+Cointegration in financial time series has been largely studied and the Johansen
+method is commonly used to obtain cointegration relationships.  In practice, it
+has been found that cointegration relations change with time.  However,
+model-based cointegration such as VECM assumes that cointegration remains
+unchanged in time.  We empirically showed that the Johansen method is sensitive
+to the number of lags but also to the amount of data considered.
+
+Moreover, we introduced the notion of {\em percentage of cointegration\/} and
+found that out-of-sample forecast performance MSE is related to the value of
+this figure in the last samples.  We used this information to set the model
+parameters.  Our proposal AVECM consists of an adaptive algorithm to update VECM
+parameters every time that new data is available. These parameters are found by
+maximising the percentage of cointegration of the last samples or iterations.
+
+Despite the fact that high frequency Forex data can be spurious, the model
+performance can be less reliable (and more spurious) relative to the lower
+frequencies (such as 1 minute or 5 minute intervals) adopted by some other
+studies. However, the deficiency is offset by gain in accuracy from parallel
+processing which is capable of searching or examining a much larger state space
+given the same computational time.
+
+Determining VECM parameters was the most expensive routine and it was run using
+parallel processes using MPI which allowed a grid search within a range of
+values for $L$ and $p$ to be made.  Tests were done using real currency rates
+data.
+
+Results showed that our proposed AVECM improves performance measures by finding
+parameters of $L$ and $p$ maximising the percentage of cointegration.
+
+The parallel implementation allowed the execution times to be reduced more than
+9 times and therefore a response time was obtain before 10 seconds. Since we
+used 10-second frequencys we can say that our proposal is suitable for use in
+an online context for real applications because response times were less than
+this frequency.
+
+For future study, it would be interesting to explore the relationship between
+cointegration and performance in order to propose new criteria for improving
+VECM parameters. It would also be interesting to include more explaining
+variables such as bid-ask spread and change in volume.

### 2016-08-19 src/4_OnlineLearning.tex
avance jueves

@@
 \chapter{Machine Learning Models}
-Machine learning is a scientific discipline focused on the development of algorithms based on learning from examples.
-This idea has become central to the design of search engines, robots systems and forecasts applications that process large data sets.
-Usually machines require long training period when developing algorithms to predict future behaviour such as financial time series forecasting where new data is available in a short period of time.
-Online machine learning techniques tackle this problem by updating the model with new data or computing a new model using less data so they  can give a response in a short period of time.
+Machine learning is a scientific discipline focused on the development of
+algorithms based on learning from examples. This idea has become central to the
+design of search engines, robots systems and forecasts applications that
+process large data sets. Usually machines designed to forecast financial time
+series requires long training period and therefore they are not suitable to be updated
+with stream data. Online machine learning techniques tackle this problem by updating the model with new
+data or by computing a new model using less data so they can give a response in a
+short period of time.
 
@@
 
-In this thesis we will emphasised the supervised learning problem since it is the most common way of modelling financial problems.
-
-Supervised learning consists in to find a learning function $f: X \rightarrow Y$ which models a set of examples called training set $S$ which have been drawn randomly and independently according to a unknown joint distribution function $p(x,y)$ on $X \times Y$:
+In this thesis we will emphasise the supervised learning problem since it is the most common way of modelling financial problems.
+
+Supervised learning consists in finding a learning function $f: X \rightarrow Y$ which models a set of examples, called training set $S$, which have been drawn randomly and independently according to a unknown joint distribution function $p(x,y)$ on $X \times Y$:
 
 $$
-S= \{(x_1,y_1),\,(x_2,y_2),\dots\,,\,(x_n,y_n)\} \in \mathcal{S}, (x_i,y_i) \in X\times Y \quad \forall i = 1\dots n
+S= \{(x_1,y_1),\,(x_2,y_2),\dots\,,\,(x_n,y_n)\} \in \mathcal{S}, (x_i,y_i) \in X\times Y \quad \forall i = 1..n
 $$
 
-\noindent where $X \subseteq \mathbb{R}^m$ and $Y = \mathbb{R}$ in case of regression problems and $Y = \{1,-1\}$  if it is a classification problem.
+\noindent where $\mathcal{S}$ is the space of instances, $X \subseteq \mathbb{R}^m$ and $Y = \mathbb{R}$ in case of regression problems and $Y = \{1,-1\}$  if it is a classification problem.
 
@@
 
-\noindent where $\mathcal{H}$ is called the hypothesis space, is the space of functions that the algorithm is allowed to search.
-The selection of $f_S$ is based on a loss function $V(f(x),y)$ which expected risk has to be minimised.
-
-\[
-E[V(f(x),y)] = \int V(f(x),y) dp(x,y)
-\]
-
-\noindent $V(f(x),y)$ denote the price paid for mistakes. Therefore, $V(f(x),y)=0$ if$f(x)=y$.
+\noindent where $\mathcal{H}$ is called the {\em hypothesis space}, is the space of functions that the algorithm is allowed to search.
+The selection of $f_S$ is based on a loss function $V(f(x),y)$ the expected risk, which has to be minimised.
+
+\begin{equation}
+\label{eq:expetedrisk}
+E[V(f(x),y)] = \int_{X \times Y} V(f(x),y) \, \, dp(x,y)
+\end{equation}
+
+\noindent $V(f(x),y)$ denote the price paid for mistakes. Therefore, $V(f(x),y)=0$ if $f(x)=y$.
 
@@
 \item[Complexity of the hypothesis class] for best generalisation, we should adjust the complexity of the learner model $\hat{f}$ to the complexity of the data $f$.
+The complexity of a learning problem mostly depends on the number of training examples and the size of the searched hypothesis space.
 In polynomial regression, the complexity parameter is the order of the fitted polynomial, and therefore we need to find a way to choose the best order that minimises the generalisation error, that is, tune the complexity of the model to best fit the complexity of the function inherent in the data.
 \item[Generalisation accuracy on new examples] is the capability of the algorithm to generate the right output for an input instance outside the training set.
-\item[The amount of training data]  generally generalisation accuracy increases as the amount of training data increases.
+It is measured by the expected risk (equation \ref{eq:expetedrisk}).
+\item[The amount of training data]  in most cases, generalisation accuracy increases as the amount of training data increases.
 \end{description}
@@
   \includegraphics[width=0.8\textwidth]{img/3tradeoff}
-  \caption{Tradeoff in empirical learning.
+  \caption{Tradeoff in empirical learning from \cite{dietterich2003}.
 The generalization accuracy is not affected by the amount of examples for low complex classifiers.

### 2016-10-20 comentarios-tesis/20161009_Wuerzburg_Paolas Thesis Reviewing.tex
comentarios profe

% 20161009_Wuwerzburg_Paolas Thesis Reviewing.tex --- 09.10.2016

\documentclass[12pt,reqno]{amsart}
\usepackage{amsfonts,amssymb,amsmath,amsthm,mathrsfs,
            amsopn,amsxtra,amstext,amsbsy}
\usepackage{epsfig}
\usepackage{array}
\usepackage{url}
\usepackage{tikz}
%% \usepackage[spanish]{babel}
\usepackage[utf8x]{inputenc}
\usepackage{fancyhdr}
\usepackage{graphicx}
\usepackage{hyperref,wasysym}
\usepackage{xcolor}
\usepackage{lipsum}
\usepackage{enumerate}
\usepackage{enumitem}% http://ctan.org/pkg/enumitem
\usepackage{verbatim}

\voffset=-10mm
\oddsidemargin=0pt
\evensidemargin=0pt
\topmargin=-10mm
\headsep=18pt
\headheight=18pt
\footskip=30pt
\topskip=0mm
\textwidth 167mm
\textheight=230mm
\parindent=0em
\marginparsep=0mm
\marginparwidth=0mm
\vfuzz2pt % Don't report over-full v-boxes if over-edge is small
\hfuzz6pt % Don't report over-full h-boxes if over-edge is small

% \newtheorem{conjecture}[theorem]{Conjecture}
\newtheorem{definicion}{Definición}
% \newtheorem{theorem}{Theorem}%[section]
% \newtheorem{lemma}[theorem]{Lemma}
% \newtheorem{corollary}[theorem]{Corollary}
% \newtheorem{conjecture}[theorem]{Conjecture}
% \theoremstyle{definition}
% \newtheorem{definition}[theorem]{Definition}
% \newtheorem{example}[theorem]{Example}
% \newtheorem{xca}[theorem]{Exercise}
% \theoremstyle{remark}
% \newtheorem{remark}[theorem]{Remark}
% \numberwithin{equation}{section}

% \def\phi{\varphi}
% \def\epsilon{\varepsilon}
\def\RE{\mbox{\rm Re}\,}
\def\IM{\mbox{\rm Im}\,}
% \def\AZ{|z|}
\newcommand{\rectangle}{\fboxsep0pt\fbox{\rule{0.3em}{0pt}\rule{0pt}{1.5ex}}}

%-------------------------mathcal

\def\A{{\mathcal A }}
\def\CC{{\mathcal C }}
\def\CH{{\mathcal H }}
\def\CK{{\mathcal K }}
\def\CM{{\mathcal M }}
\def\CP{{\mathcal P }}
\def\CS{{\mathcal S }}
\def\CR{{\mathcal R}}
\def\CRu{{\mathcal R}^{\rm u}}
\def\CT{{\mathcal T}}
\def\CTT{\widetilde{{\mathcal T}}}
\def\CW{{\mathcal W}}
\def\Li{{\mbox{\rm Li}}}
\def\WO{{\mathcal W}_{\Omega }}

%-------------------------------Bbb
\def\K{{\mathbb K }}
\def\C{{\mathbb C }}
\def\N{{\mathbb N }}
\def\NO{{\mathbb{N}_0 }}
\def\Z{{\mathbb Z }}
\def\R{{\mathbb R }}
\def\U{{\mathbb U }}
\def\D{{\mathbb D }}
%----------------------------Zusammengesetztes

\def\DB{\overline{\D}}
\def\Db#1{\overline{\D_{#1}}}
\def\DBR{\overline{\D_\rho}}
\def\DV{de la Vall\'ee Poussin }
\def\d#1{\D_{#1}}
\def\H{{\mathcal H}(\D)}
\def\Hb{{\mathcal H}(\DB)}
\def\HO{{\mathcal H}_0(\D )}
\def\Hoo#1{{\mathcal H}(\overline{\D_{#1}})}
\def\HB#1{{\mathcal H}(\D_{#1})}
\def\Ho#1{{\mathcal H}({#1})}
\def\Hyp{{}_2F_1}

\def\achtung#1{\alert{#1}}
\def\blue#1{{\color{blue} #1}}
\def\green#1{{\color{green} #1}}
\def\red#1{{\color{red} #1}}

\def\pick{{\mathscr P}}
\def\logpick{{\text{log}\,{\mathscr P}}}
\def\stammpick{\pick_{\!\!\!\raisebox{-2pt}{\scalebox{0.5}{$\int$}}}}
\def\stammpickpick{\stammpick\cap\pick}
\def\stammpicklogpick{\stammpick\cap\logpick}
\def\expstammpicklogpick{e^{\stammpicklogpick}}

%-------------------------Bakan-Ruscheweyh-Salinas 2015
\def\mid{\;:\ }
\def\hol#1{\text{Hol}(#1)}
\def\normhol#1#2{\text{Hol}_{#1}(#2)}

\DeclareMathOperator*{\cosig}{cosig}
\DeclareMathOperator*{\supp}{supp}

%graficos
%\pgfdeclareimage[width=5cm, height=3cm ]{recursiva}{./Png/recursiva.png}
%\pgfdeclareimage[width=5cm, height=5cm ]{alan}{./Png/Turing.jpeg}
%\pgfdeclareimage[width=5cm, height=5cm ]{programa}{./Png/maquina.png}
%\pgfuseimage{lala}

\renewcommand{\restriction}{\mathord{|}}

\newcommand\blfootnote[1]{%
  \begingroup
  \renewcommand\thefootnote{}\footnote{#1}%
  \addtocounter{footnote}{-1}%
  \endgroup }

\title[Paola's Thesis Comments]{Paola's Thesis Comments}

\author[L. Salinas]{Luis Salinas}

\address{CCTVal and Departamento de Inform\'atica, UTFSM, Valpara\'\i so;
         \newline
         Mathematisches Institut, Universit\"at W\"urzburg;
         \newline
         \today} % October 06, 2016}

\email{luis.salinas@usm.cl}

\begin{document}

\maketitle%

\section{Chapter 3: Machine Learning Models.}

\def\myitem#1{ \item[\blue{\bf #1}] }

\begin{description}[style=unboxed,leftmargin=0cm,itemsep=3ex]

\myitem{P 33, Abstract}
It seems that some commas are missing:
"\dots training period when \dots" 
\quad$\longrightarrow$\quad
"\dots training period, when\dots" \\
"\dots forecasting where\dots"
\quad$\longrightarrow$\quad
"\dots forecasting, where\dots" \\
But the whole sentence "Usually machines \dots period of time"
does not make too much sense to me. Please review.

\myitem{P 33, L -4}
Insert \red{for this} in the text:
"The reason \red{for this} is that ML models are data driven and are
able to examine large amounts of data."

\myitem{P 34, section 3.2}\mbox{}\\
\blue{L 3:}\quad
"In this thesis we will emphasise\red{d} the supervised learning
problem since it is the most common way of modelling financial problems."\\
\blue{COMMENT:}\quad
(a) \red{d} in ``emphasize\red{d}'' is wrong. \\
(b) That almost everybody uses this approach is not a good reason for you
to use it as well!
Think, for instance, of "almost everybody eats junk food; thus I will
eat junk food as well"!
I suggest instead the following phrasing: \\
\green{\em "Supervised learning is most popular and most commonly used
in modelling financial problems and the assessments of this method and
their results in praxis are fairly good. Therefore, in this thesis we
will adhere to this trend and we will use (an improved version of)
supervised learning."}

\vspace{1ex}\blue{L 5:}\quad
"\dots consists in to find\dots"
\quad$\longrightarrow$\quad
"\dots consists in finding\dots"\quad (?)  CHECK!

\vspace{1ex}\blue{L 6:}
"\dots set of examples called training set S which have\dots"
\quad$\longrightarrow$\quad
"\dots set of examples, called training set S, which have\dots"

\vspace{1ex}\blue{First formula:}\quad
There are two "S": a slanted one and a calligraphic one; 
slanted-S belongs (in set theory sense) to calligraphic-S; 
slanted S is defined, but what is calligraphic-S?

\vspace{1ex}\blue{First formula notation: }\quad
Written "$\forall\ i=1...n$"; but it seems to me that the standard
form (in this format) is "$\forall\ i=1..n$" (two points only) CHECK!

\vspace{1ex}\blue{GENERAL:}\quad Put new names with 
\verb!\em (\emphasized)!.
Example: {\em hypothesis space}; 
What kind of functions are allowed in calligraphic H? 
Polynomials in several variables? 
Continuous functions? etc.

\vspace{1ex}\blue{Last sentence:}\quad 
"\dots $V(f(x),y)$ which expected risk has to be minimised."
\quad$\longrightarrow$\quad
"\dots $V(f(x),y)$, THE expected risk, WHICH has to be minimised." CHECK!

\vspace{1ex}
\red{I sent you some of these last remarks in a mail from August 14.
I will repeat them in this LaTeX file.
These last remarks are discussed in a more expanded form below,
because there are more serious mathematical problems here. 
There are repetitions among the remarks.}

\myitem{P 34, LL -16 to -1 (section 3.2)}
``A learning algorithm $\mathcal{A}$ takes as input a data set
$S\in\mathcal{S}$ and output a function $f_S$:''

It should say ``...and output\red{s} a function...''

\red{HOWEVER, there are more serious problems here and in the
following paragraphs.}

It is really a bad idea to work with a set $S$ and something
called $\mathcal{S}$, which seems to be a class of sets like $S$.

To clearly distinguish them, let us call $S$ the \red{set $S$} and
$\mathcal{S}$, \red{the class $\mathcal{S}$ of sets $S$} or simply
\red{the class $\mathcal{S}$} for short.

You see, the set $S$ and the class $\mathcal{S}$, as fonts, are almost
indistinguishable.
It seems that the set $S$ is a subset of $X\times Y$ and the class
$\mathcal{S}$ would be the subset of the power set of $X\times Y$
containing all training sets $S$.

To avoid confusions I suggest to use \verb!\mathscr! fonts and
call $\mathscr{T}$ your set $\mathcal{S}$.
The $\mathscr{T}$ stands for "training".

Achtung!: \verb!\mathscr! needs \verb!\usepackage{mathrsfs}!

However, is it really necessary to use $\mathcal{S}$ or $\mathscr{T}$?

Have all sets $S$ in $\mathscr{T}$ the same cardinality?

If not, I think one can safely use the power set $\mathscr{P}(X\times Y)$
as the set $\mathscr{T}$ and forget $\mathscr{T}$.

But you have to judge whether this would be appropriate or not.

If your $\mathcal{S}$ ( = my $\mathscr{T}$ ) is really a subset of the
power set of $X\times Y$, then I would suggest to define the set $S$
simply as
$$
S = \{ (x_i,y_i)\in X\times Y : k=1,...,n \}
$$
The more serious problem is related to the functions $f_S$.

What is a \red{``learning function''} $f$?

What are the domain and codomain of these functions?
If your answer is ``$f:X\to Y$'', where $X\subseteq\R^m$ and $Y=\R$ or
$Y=\{1,-1\}$, then the problems is transferred to the ``loss function''
$V(f(x),y)$ because then both arguments of $V(\cdot,\cdot)$ lie in $Y=\R$.
How do you define then the function $V$?
Suppose that $Y=\R$ as allowed above.
Then it seems that the loss function $V$ is defined on
$X\times Y=\R^m\times\R$, i.e. $V:\R^m\times\R\to\R$ assuming that the
codomain of $V$ is $\R$ (you must declare what the codomain of $V$ is).
But then, when you write $V(f(x),y)$, then $f(x)$ is in $Y=\R$ and
hence $V:\R\times\R\to\R$.
Thus, what is true: $V:\R^m\times\R\to\R$ or $V:\R\times\R\to\R$?

How is the functions $f_S$ related to the sets $S$?

\myitem{P 34, PARAGRAPH STARTING AT L -6}
Something is wrong with this definition. 
It seems that a learning algorithm $\mathcal{A}$ assigns an object
$\mathcal{A}(S)$ to each set $S\in\mathscr{T}$.

What is exactly this object $\mathcal{A}(S)$?

You say that the object $\mathcal{A}(S)$ is an element of a so called
"hypothesis space" $\mathcal{H}$.
But what is this "hypothesis space"?
You do not really describe the members of this "hypothesis space"
$\mathcal{H}$.
You just say that the members of $\mathcal{H}$ are functions.
But what kind of functions? Domain? Range? What do they do these
functions?

And furthermore you say that the algorithm $\mathcal{A}$ \red{searches}
these functions in $\mathcal{H}$.
How is this search exactly performed?

The whole PARAGRAPH is rather unintelligible.
Before proceeding any further, a CONCRETE EXAMPLE is needed. 
We must come back to this definition after we discuss your example.

In the last 2 lines of this paragraph and in the first 2 lines of page
35 you write:

"The selection of $f_S$ is based on a loss function $V(f(x),y)$
\red{WHICH} expected risk has to be minimised."
Instead of \red{WHICH} you should use \blue{WHOSE}, I guess.
Then comes this formula:
$$
E[V(f(x),y)] = \int V(f(x),y)\;dp(x,y)
$$
"$V(f(x),y)$ denote the price paid for mistakes. Therefore,
$V(f(x),y) = 0$ if $f(x) = y$."

There are many comments in order here:\\[1ex]

1. By integrals always put a small space \verb!"\,"! or \verb!"\;"!
between integrand and differential ($dp(x,y)$ here ).\\[1ex]

2. What is the domain of integration for the integral?
Judging by the differential, it seems that this domain is $X\times Y$
and hence the function $f$ is a function from $X$ into (onto?) $Y$.\\[1ex]

3. Then another function appears: $V(f(x),y)$, so that $V$ should be
defined on $Y\times Y$. Is this true? \\[1ex]

4. Whenever you define functions you must first clearly define its
domain and its range (codomain).
The same applies to algorithms.
If there is not yet a concrete pseudo-code description of an algorithm,
it is necessary at least to say something about the class of algorithms
that you are going to consider, and what should your algorithms do.
An algoritmh is just a kind of function, which operate on a set or
space of some objects and produce an output, which a subset (consisting
may be of only one member or none at all!) of objects taken fron some
other set or space. \\[1ex]

5. It is important to define the units that will be used to measure
the different variables appearing in your proble.
But this might be optative. For instance it seems that the value
$V(f(x),y)$ is measured in USD or EUR. \\[1ex]

6. Your formula has a major problem in the following sense: both sides
depend on $x$ and $y$, but on the right hand side they are integration
variables (i.e., running presumably on the whole of $X\times Y$) and
on the left hand side they seem to denote that the variable representing
the first argument of $V$ is $x$ running on $X$, and the variable
representing the second argument of $V$ is $y$ running on $Y$.
This is cumbersome.
I suggest:
$$
E[V(f(\cdot),\cdot)] = \int_{X\times Y} V(f(x),y)\;dp(x,y)\,,
$$
but this must be critically reviewed.

Recall that some lines above you wrote $X=\R^m$, i.e. dimension $m$,
and $Y=\R$, i.e. dimension $1$.
Note that in this case $f:X\to Y$ means $f:\R^m\to\R$.
If you write $V(f(x),y)$, then we are in troubles because the first
argument of $V$ is now $f(x)$, which is in $\R$.
Thus, in this case, $V$ should be defined as $V:\R\times\R\to\R$,
but you wrote before $V:X\times Y\to\R$, i.e., $V:\R^m\times\R\to\R$.

Must then $V(\xi,\eta)$ be defined --for a given $f$-- only on the graph 
$$
\{ \big( f(\xi),\eta \big)\mid \xi\in X, \eta\in Y \}\subset\R\times\R
$$
and not on the whole of $X\times Y=\R^m\times\R$?

This feeling is re-inforce after looking to equations (3.1) and (3.2).
But then all would depend on $f$ and I think this is not very rational.
I would dare to say that that the function $V$ --as well as the differential
$dp(x,y)$-- is defined (somehow; examples?) on the whole of $X\times Y$:
$$
\begin{array}{rccl}
V: & X\times Y  & \longrightarrow & \R     \\
   & (\xi,\eta) & \longmapsto     & V(\xi,\eta)
\end{array}
\qquad\text{and}\qquad
\begin{array}{rccl}
dp: & X\times Y  & \longrightarrow & \R     \\
    & (\xi,\eta) & \longmapsto     & dp(\xi,\eta)
\end{array}
$$
Under this notation, when $X=\R^m$ and $Y=\R$ there are big problems
with your equations (3.1) and (3.2):
$$
V(\red{\mathbf{x}},y)=(\red{\mathbf{x}}-y)^2 
\qquad\text{or}\qquad 
V(\red{\mathbf{x}},y)=|\red{\mathbf{x}}-y|\,,
\qquad\text{where}\quad\red{\mathbf{x}\in\R^m}\,,\ y\in\R\,.
$$
There is no way to assign a sound meaning to these equations.

However, if you now consider a particular function 
$f:X=\R^m\to Y$, $y=f(x)\in\R$, then these equations \red{can} have a
sound meaning, namely:
$$
V(f(x),y)=(f(x)-y)^2 \qquad\text{or}\qquad V(f(x),y)=|f(x)-y)|
$$
because $f(\red{\mathbf{x}})$, for $\red{\mathbf{x}\in\R^m}$,
and $y$ are now both in $\R$.
But then $V:\R\times\R\to\R$!

All this is very different from \red{your} equations (3.1), (3.2).
In these equations (3.1), (3.2) the domain of the function $f$
would be $X$, but its codomain cannot be $Y$, because if this were the
case, then $V$ should be defined on $Y\times Y$ which seems to be nonsense
at this point.

Note that for $X=\R^m$, $Y=\R$ and $f:X=\R^m\longrightarrow Y=\R$
we have:
$$
\begin{array}{rccl}
V(f(\cdot),\cdot): & \R^m\times\R  & \longrightarrow & \R \\
   & (x,y) & \longmapsto     & V(f(x),y)
\end{array}
\qquad\text{but}\qquad
\begin{array}{rccl}
V(\cdot,\cdot): & \R\times\R  & \longrightarrow & \R     \\
    & (\xi,\eta) & \longmapsto     & V(\xi,\eta)
\end{array}
$$
which is not consistent with your definitions.

Summarizing: you must carefully review this section. In particular,
you must clarify what is the domain of $V$: $X\times Y$? $Y\times Y$?

\myitem{P.35, LL 9-12, Formula (3.3)}
The same objections as before must be applied here.
The main problem is with the definition of $V(x,y)$ and $V(f(x),y)$,
as before.
We already had $f$ and $f_S$ and now, without any warning, a $\widehat{f}$
appears.
How are all these functions (I suppose!) related?

You say ``The objective is to estimate a function $\widehat{f}$ through
empirical risk (training error) minimisation (ERM).''

To estimate from what set or from where?
From the class $\mathscr{T}$ of sets $S$?
This means that in your formula (3.3) you have (at least) to declare
from where you select the $f$'s and what they are. 
Discrete functions?
Are they related to the sets $S$ in $\mathscr{T}$?

Formulas (3.1) and (3.2) would be OK assuming $x\in X=\R^m$ and
$f:X=\R^m\to\R$.
But again: what is the relation to the sets $S$?
In formula (3.3) I suppose that $x$ should be $x_i$ and $y$ should be
$y_i$. Is this so?

With this interpretation in mind, the right hand side of formula (3.3)
would be a Riemann sum approximation of the integral in the first line
of P 35, \red{where $p(x,y)$ is the uniform distribution and hence
$dp(x,y) \approx 1/n$}.
Thus, you have:
$$
E[V(f(x),y)] \approx R_{\text{emp}}[f]\,.
$$
But what is this $R_{\text{emp}}[f]$ conceptually?
So, this is a well known joint (?) distribution.
But in the previous page you said that the joint distribution $p(x,y)$
was \red{unknown}.
Note that you can arbitrarily change the value of $R_{\text{emp}}[f]$ just
by arbitrarily changing the unknown joint distribution $p(x,y)$.
So what is going on here?
What do you want to estimate and how exactly do you want to do it?

\myitem{P 35, several lines}

\blue{L 1:}quad 
First formula: What is the domain of integration for the integral? 
Also, separate "$V(f(x),y)$" from "$dp(x, y)$" a little!

\vspace{1ex}
\blue{L 2:}\quad Separate "V(f(x),y)= 0 if" from "f(x) = y" a little!

\vspace{1ex}
\blue{L  3:}\quad
Are you writing "$L2$" for "$L^2$" or "$L_2$"?
This would be OK if you systematically use "$L2$". Idem for $L1$.

\vspace{1ex}
\blue{L 15:}\quad
Complexity of the hypothesis class. 
What notion of complexity are you using here?
There are two complexities: one of the learner model, and one of the data.
An explanation is needed.
You give a kind of explanation in the case of polynomial regression; 
is this the one you will be using? 
Anyway, an explanation is needed.

It says: ``complexity of the learner model $\widehat{f}$ to the
complexity of the data $f$''

Do you call ``learner model $\widehat{f}$'' the one $f$ you get by
maximizing (3.3) among nobody knows what?

What is it and how do you compute the ``\red{complexity} of the
learner model $\widehat{f}$''?

What is the data $f$?
Has it something to do with the sets $S$?

What is it (definition) and how do you compute the 
``complexity of the data $f$''?

You say: ``In polynomial regression, the complexity parameter is
the order of the fitted polynomial''.

Why this is so?
The word ``complexity'' always suggest how difficult is to perform
some computation.
Is this ``difficulty'' related to the something called ``order''
of a polynomial?

What is the \red{order} of a polynomial?
Do you mean the \red{degree}?

\vspace{1ex}
\blue{L -13:}\quad
"\dots data: generally generalisation accuracy\dots"

Too many "general's" together!

\vspace{1ex}
\blue{L -11:}\quad
"Very complex models will\dots"
\quad$\longrightarrow$\quad
"a very complex model will..."

(so that the grammar number of the subject coincides with "it could...".

\myitem{P 36, some remarks}

\blue{Figure 3.1.}\quad
How do you define "Generalization Accuracy"?
Idem, ``classifier complexity'' and hence "low complexity classifier''?
\red{How do you measure them?}
Without these definitions the diagram makes little sense.

\vspace{1ex}
\blue{Ad (3.4):}\quad
\red{{\bf NEW (as for 10.10.2016):}\quad
The presentation here has some problems.
Formula (3.4) and its proof appears very suddenly.
How is it connected to the last parapraph 3.2.2. in page 35?
Why are you interested in dealing with formula (3.4) here? \\[0.6ex]
More exactly, something like a roadmap is missing here. \\[0.6ex]
I think somewhere before or in an Appendix you should  define (very well) 
a minimal set of concepts from Statistics in order to understand the
various formulas appearing in your work. \\[0.6ex]
Good modern references should be given for those concepts. \\[0.6ex]
You use variables like $x$ and $y$ which are usually assumed to be
continuous variables and hence the base theory would be integration
theory. \\[0.6ex]
But it seems to me that in the whole problem you are dealing with,
samples or discrete values of these variables is what you are using. \\[0.6ex]
If this is so, I would suggest that you use discrete variables, like
$x_k$, $y_k$, samples, etc.
Then you would have simple formulas for defining bias, expected values,
sigma, etc.
You should clearly state, for instance:
\begin{itemize}
\item
what are you assuming to be known;
\item
what is not known;
\item
what quantitities you want to determine;
\item
how you intend to determine those quantities;
\item
etc.
\end{itemize}
}

Is this (3.4) a theorem? a remark?
It seems a standard formula to me;
I think there are textbooks discussing this formula.
But what is $\sigma$ here?
If it is indeed a formula from the mathematical folklore,
you could omit the proof and simply put a reference where this proof
can be found.

By the way, instead of "Demo" you should write "Proof".
In the "Demo"-Proof, the ellipsis "\dots" is misleading, since in formulas
"\dots" is used to show that some terms have been left out, but this not
the case here \red{since the only term missing in this line is the term
written in the line immediately below}

\red{Why is it $E[y-f(x)]=0$ here?}

\red{In the second line of the "Demo" appears a "$B$". What is this "$B$"?
It seems to me, that it should be "$y$".}

\myitem{P 37, LL 1-5 (formula).}

\blue{L 1:}\quad
"$E[f(x)-\widehat{f}(x))]^2$" has two mistakes:

(1) A parenthesis "(" is missing between "[" and "$f$".

(2) The "${}^2$" should go between ")" and "]", i.e. the "${}^2$" should
go inside the brackets "$[\dots]$" and not outside.

A general recommendation: use \verb!\widehat! instead of \verb!\hat!;
\verb!\hat! produces a too tiny wedge.

\blue{L 2:}\quad
The parenthesis "(" is still missing between "[" and "$f$", and the
"${}^2$" is still outside the brackets "[...]".

\blue{LL 3-4:}
Again the problem of the misleading ellipsis "$+\dots$";
it must be omitted

\red{{\bf GENERAL REMARK:}\quad
In a scientific paper or in a thesis, you should NOT present ALL
algebraic details, but just the minimum necessary for the reader
to understand what is going on.
Details must always be left to the reader.
NEVER, but really NEVER, use weird school signs to denote cancelation
of terms!}

\myitem{P 37, L 6.}
``Bias is introduced by the model selection.'' \\
How does model selection introduces bias?
This should be explained.

\myitem{P 37, LL 6-17}
"Therefore the model building process is repeated (through resampling)
and substantially different averages of prediction values are obtained,
bias will be high."

Is this sentence correct?
It seems to me that a better versiob of this sentence would be:

"Therefore\red{, if} the model building process is repeated (through
resampling) and substantially different averages of prediction values
are obtained, \red{then} bias will be high."

Is this your idea? Please discuss!

But I still do not see \red{why this would be so}, i.e., why is it so
that resampling could produce substantially different averages of
prediction values.
When or under what circumstances this occurs?
You have to prove this implication, or cite some reference where this
is proved.

\red{You have not yet described the model building process.
You now should describe it.}

In the next sentences in this paragraph the notion of "model complexity"
appears again.
Of course everybody has an intuitive meaning of this concept, but here
we need a mathematically consistent definition and not only intuitions.

Grammar is again to be reviewed: in the sentence

\myitem{P 37, L 10}
"Variance measures how inconsistent are the predictions from one another,
over different training sets, not whether they are accurate or not."

Inconsistence? Accuracy? Definition and/or references are again needed here.

\blue{QUESTION:}\ 
If this is as claimed, how this would affect the results of your predictions
in the sequel? 
What would be the effect on the confidence of your predictions?

The effects on your results could be devastating.
Thus, how are you planning to avoid these bad effects?

\myitem{P 37, L 17}
"The best model will be the one has a balance between bias and variance".

Why this is so? 
Something like "which" or "that" between "the one" and "has a balance".

\myitem{P37, Figure 3.2}
Somewhere near Figure 3.2 (before, after, or in the caption) clear
definitions of \red{Prediction error} and \red{Model complexity}
should be given, together with an explanation about how they are
measured in order to produce the diagram.

Explain the reasons of the behaviour of the red (test sample) and the
blue (training samples) curves.

Why the blue one is decreasing?

Intuitively one can think that the more complex the model (whatever
this could mean), the more over-fit results and hence less prediction
error, but precisely this over-fit makes that the generalisation
capacity of the model will be bad, and hence the error when the model
is applied to the testing samples will increase (red curve). 
Is this correct?
Please discuss, providing the right definitions.

\myitem{P 38, L 2}
``\dots in order to obtain as accurate model as is feasible \dots''\\
$\rightarrow$
``\dots in order to obtain as \red{an} accurate model as is feasible \dots''
You must discuss before (somewhere) the concepts of \red{accuracy} and
\red{feasibility} you are using.

\myitem{P 38, L 4} 
"truth target" Is it correct? Or is it "true target"?
Two times in this line.

\myitem{GENERAL REMARK}
Pay attention to USA and UK English dialects!

Generalization $\quad\longleftrightarrow\quad$ Generalisation

Optimization  $\quad\longleftrightarrow\quad$ Optimisation

Are "penalisation", "regularisation", etc., English words? Etc.

\myitem{P 38, L 5} 
``\dots is the reason why in order to obtain the best model, data\dots'' \\
$\rightarrow$
``\dots is the reason why, in order to obtain the best model, data\dots'' \\
Note the comma between ``why'' and ``in''.

\myitem{P 38, L 6} 
``Training set is used to determine the model, validation set is used 
to estimate the generalisation error and finally testing set is used
to estimate the accuracy of the model.'' \\
$\rightarrow$
``\red{The} training set is used to \red{construct} the model, 
\red{the} validation set to estimate the generalisation error and finally
\red{the} testing set to estimate the accuracy of the model.'' \\
\red{Note that the suggested sentence has only one ``is used to''.
We are using here the \blue{ellipsis} feauture of indoeuropean languages.}

Where did you rigorously define \red{generalisation error} and
\red{model accuracy}?

\myitem{P 38, L 10} 
``\dots procedure [Gei75], \red{usually used} when\dots''
Too many \red{``use''} of \red{``uses''}!

\myitem{P 38, L 12}
``In K-fold cross-validation''\quad References? This is most important
because there should be an statistical theory as a basis for this
procedure.

Must the $K$ subsets of data be disjoint?

\myitem{P 38, LL -4, -3; and P 39, LL 1,2}
This must be expanded.
You now talk about an optimisation function.
Which is the optimisation function?
How is it related to the ``complexity of the model'' (whatever it might be)?
How is the generalisation error included in this optimisation function?
Linearly?
What are the ``high complex models''?

\myitem{PP 38/39, paragraph stating on L -2, and finishing on L 2}
Did you discussed already what ``ill/well-posed problems'' are?
Did you include Jean Jacques Hadamard as the source of these concepts?
References for this circle of ideas?
How do you prove that ERM is an ill-posed problem?
Where is this proof?

What is a \red{regularisation} here?
This concept is usually represented by a functional over a class of
functions.
The functional have many different forms and some additional terms
are introduced as penalisations.
These additional terms are usually additive, like Lagrange multipliers.
But not always.

\red{How does it looks like the Tikhonov regularisation? References?}

\red{Why and how does exactly the Tikhonov regularisation prevents
overfitting?}

You should present clearly your function class (which seems to be your
$\mathcal H$ space), your optimisation functional, your penalisation
terms, your regularisation parameters, etc.
Of course, references are necessary.

\red{Achtung!}
In L -2 you write ``Regulari\red{z}ation'' but in L -1 you you write 
``Regulari\red{s}ation''.

\myitem{P 39, LL 3-6}
It says: ``Tikhonov regularisation minimised over the hypothesis space
$\mathcal H$ for a fixed positive parameter $\lambda$ the following:''

The sentence is too cumbersome and has mistakes (``minimise'' should be
``minimises'').
I suggest:

``Tikhonov regularisation considers a functional of the form
$$
( FORMULA (3.5) ) \rule{3cm}{0pt} R_{\text{emp}}[f] 
= \frac{1}{n} \sum_{i=1}^n V(f(x),y) + \lambda \mathcal{R}(f)
$$
where $\lambda$ is a real parameter and $\mathcal{R}(f)$ is the
regulariser, which is a penalisation on $f$.
The goal is to minimise the functional $R_{\text{emp}}[f]$ over the
hypothesis space $\mathcal H$ for a fixed positive parameter $\lambda$.''

There are many things unclear here:

\begin{enumerate}
\item
In formula (3.5) the variables $x,y$ on the right hand side have no
subindex. I assume they are $x_i,y_i$. Is this correct?
\item
Where do they run $x,y$, resp $x_i,y_i$?
\item
The same objection discussed in a previous item applies here:
What is known? What is unknown? etc.
\item
Is $V$ known? Is it one of those $L^p$-norms? If yes, which one in
your specific case?
\item
What form has your penalisation $\mathcal{R}$?
\item
See the problems related to $V(f(x),y)$ discussed above.
\item
What exactly is the space of the $f$ functions?
$\mathcal{H}$?
How will you represent your functions $f$?
\item
How exactly do you plan to perfom the minimisation?
At least a pseudocode of your minimisation algorithm would be necesary
at this point.
\end{enumerate}

\myitem{P 39, L 9}
``...to solve a regression problem.'' \\
\quad$\rightarrow$\quad
``...to solve a \red{linear} regression problem. 
In this case $f(x)$ has the form $f(x_t)=X^T.\,x_t$, where $X^T$ is an
appropriate matrix and $x_t\in X=\R^m$.''
Is it correct that $X=\R^m$ for some $m\in\N$?

\myitem{P 39, equation in lines 11 - 13}
What are $\mathbf{A}$, $\mathbf{X}$, $\mathbf{x}_t$, $\mathbf{Y}$\,?

What is the relationship between $\mathbf{A}$ and $\mathbf{X}^T$?

What is the relationship between $\mathbf{x}_t$ and $\mathbf{X}^T$?

A random reader of your thesis could think that:
$$
\mathbf{X} 
= \begin{bmatrix} x_1 \\ \vdots \\ x_t \\ \vdots \\ x_N \end{bmatrix}\,,
\qquad
\mathbf{Y} 
= \begin{bmatrix} y_1 \\ \vdots \\ y_t \\ \vdots \\ y_N \end{bmatrix}\,.
$$
But these are vectors of dimension $N$.
In equation (3.6) you say that $\mathbf{X}$ and $\mathbf{Y}$ are matrices
of dimensiosn $n\times l$ and $m\times l$, respectively.

But then, what could $\mathbf{X}^T$ be?
Is it a matrix built from sequentially shifted copies of the 
{\em row vector\/} $\mathbf{X}^T$?
If yes, you must explain this.

In seems that the equation, as it is, makes no sense!
Clear definitions are needed.

I suggest you review and present this equation using elementary matrix
notation. This would make the incoherences evident.

\myitem{From P 39, L -7, until P 40, L 3}
You put this paragraph as someting called ``Demo''.
What is it? A proof? A proof of what? 
What is your claim before starting with ``Demo''?

Then you say ``to solve equation (3.8)''.
But (3.8) is just the definition of More-Penrose!

Are you trying to solve (3.6)?
If this is true, what is known and what is not known?

You should prove the equivalence you claim.
At least write $\mathbf{C}_\lambda$ to make explicit the dependence of
$\mathbf{C}$ from the parameter $\lambda$.

I think you shoud better write something like:
$$
(3.9)\rule{50mm}{0pt} \min_{\mathbf X \in \text{???}}
\|\mathbf{C}_\lambda\,\mathbf{X}-\mathbf{F}\|
$$
You should make clear in which space $\mathbf{X}$ is running for the
optimisation.

The outcome of this minimisation problem depends on $\lambda$, but
equation (3.6) does not depend on $\lambda$.

Thus, why is (3.6) equivalent to the minimization problem (3.9)?
A proof is needed.

Which value of $\lambda$ will you consider for a solution of (3.6) and why?

What is $\mathbf{A}$ and what is the rationale for introducing $\mathbf{C}$
and $\mathbf{F}$.

The equation for $\mathbf{X}(\lambda)$ in LL 1, 2 in P 40, seems to be a
More-Penrose solution of $\mathbf{C}_\lambda\,\mathbf{X}=\mathbf{F}$.
But why this solution coincides with the solution --if it exists-- of the
minimisation problem?

By the way, is it sure that the minimisation Problem has a solution?
Is it unique? etc.

\end{description}

Reviewing work in progress/DATE: 19.10.2016

\begin{thebibliography}{99}
\bibitem{Arce-2016}
Paola Arce, 
{\em Online Learning Method for Financial Time Series Forecasting\/}.
Dissertation, UTFSM, June 2016.

\end{thebibliography}

\end{document}

### 2016-11-08 src/4_OnlineLearning.tex
inicio revision profe
revision page 2

@@
 
-ML models are often more accurate than what can be created through direct programming. The reason is that ML models are data driven and are able to examine large amounts of data.
+ML models are often more accurate than what can be created through direct programming. The reason for this is that ML models are data driven and are able to examine large amounts of data.
 
@@
 
-In this thesis we will emphasise the supervised learning problem since it is the most common way of modelling financial problems.
+Supervised learning is most popular and most commonly used in modelling financial problems and the assessments of this method and their results in practice are fairly good.
+Therefore, in this thesis we will adhere to this trend and we will use supervised learning.
 
@@
 $$
-S= \{(x_1,y_1),\,(x_2,y_2),\dots\,,\,(x_n,y_n)\} \in \mathcal{S}, (x_i,y_i) \in X\times Y \quad \forall i = 1..n
+S = \{ (x_k,y_k)\in X\times Y : k=1,...,n \}
 $$
 
-\noindent where $\mathcal{S}$ is the space of instances, $X \subseteq \mathbb{R}^m$ and $Y = \mathbb{R}$ in case of regression problems and $Y = \{1,-1\}$  if it is a classification problem.
-
-A learning algorithm $\mathcal{A}$ takes as input a data set $S \in \mathcal{S}$ and output a function $f_S$:
+\noindent where $X \subseteq \mathbb{R}^m$ and $Y = \mathbb{R}$ in case of regression problems and $Y = \{1,-1\}$  if it is a classification problem.
+
+A learning algorithm $\mathcal{A}$ takes as input a data set $S \in \mathcal{S}$ and outputs a function $f$:
 
@@
 \mathcal{A}: \mathcal{S} &\rightarrow & \mathcal{H} \\
-S &\rightarrow &\mathcal{A}(S) = f_S
+S &\rightarrow &\mathcal{A}(S) = f
 \end{eqnarray*}
@@
 \noindent where $\mathcal{H}$ is called the {\em hypothesis space}, is the space of functions that the algorithm is allowed to search.
-The selection of $f_S$ is based on a loss function $V(f(x),y)$ the expected risk, which has to be minimised.
+The selection of $f_S$ is based on the minimization if the expected risk of the loss function $V(f(x),y): \R\times\R\to\R$.
 

### 2016-11-08 src/4_OnlineLearning.tex
fixing learning alg definition

@@
 
-A learning algorithm $\mathcal{A}$ takes as input a data set $S \in \mathcal{S}$ and outputs a function $f$:
+A learning algorithm $\mathcal{A}$ takes as input a data set $S \in \mathscr{T}$ and selects from $\mathcal{H}$ a function $f$ such that $f(x)\approx y $ in a predictive way:
 
@@
 \noindent where $\mathcal{H}$ is called the {\em hypothesis space}, is the space of functions that the algorithm is allowed to search.
-The selection of $f_S$ is based on the minimization if the expected risk of the loss function $V(f(x),y): \R\times\R\to\R$.
+The selection of $f$ is based on the minimization of the expected error of the loss function $V(f(x),y)$ defined as:
+
+\begin{eqnarray*}
+ V: Y\times Y &\rightarrow & Y \\
+f(x), y &\rightarrow &V(f(x),y)
+\end{eqnarray*}
+
+Therefore, given a function $f$ a loss function $V$ and a probability distribution $p$ over $X \times Y$, the expected error of $f$ is:
 
@@
 \label{eq:expetedrisk}
-E[V(f(x),y)] = \int_{X \times Y} V(f(x),y) \, \, dp(x,y)
+I[f] = E[V(f(x),y)] = \int_{X \times Y} V(f(x),y) \; dp(x,y)
 \end{equation}

### 2016-11-23 src/4_OnlineLearning.tex
reunion 20 nov

--- 2016-11-08 src/4_OnlineLearning.tex
@@
 
-A learning algorithm $\mathcal{A}$ takes as input a data set $S \in \mathcal{S}$ and outputs a function $f$:
-
-\begin{eqnarray*}
-\mathcal{A}: \mathcal{S} &\rightarrow & \mathcal{H} \\
+A learning algorithm $\mathcal{A}$ takes as input a data set $S \in \mathscr{T}=\mathscr{P}(X\times Y)$ and selects from $\mathcal{H}$ a function $f$ such that $f(x)\approx y $ in a predictive way:
+
+\begin{eqnarray*}
+\mathcal{A}: \mathscr{T} &\rightarrow & \mathcal{H} \\
 S &\rightarrow &\mathcal{A}(S) = f
@@
 \noindent where $\mathcal{H}$ is called the {\em hypothesis space}, is the space of functions that the algorithm is allowed to search.
-The selection of $f_S$ is based on the minimization if the expected risk of the loss function $V(f(x),y): \R\times\R\to\R$.
+The selection of $f$ is based on the minimization of the expected error of the loss function $V(f(x),y)$ defined as:
+
+\begin{eqnarray*}
+ V(f(\cdot),\cdot): X\times Y &\rightarrow & Y \\
+  (x, y) &\rightarrow &V(f(x),y)
+\end{eqnarray*}
+
+\noindent $V(f(x),y)$ denote the price paid for mistakes. Therefore, $V(f(x),y)=0$ if $f(x)=y$.
+
+Given a function $f$ a loss function $V$ and a probability distribution $p$ over $X \times Y$, the expected error of $f$ is:
 
@@
 \label{eq:expetedrisk}
-E[V(f(x),y)] = \int_{X \times Y} V(f(x),y) \, \, dp(x,y)
-\end{equation}
-
-\noindent $V(f(x),y)$ denote the price paid for mistakes. Therefore, $V(f(x),y)=0$ if $f(x)=y$.
+E[V(f(\cdot),\cdot)] = \int_{X\times Y} V(f(x),y)\;dp(x,y)\,,
+\end{equation}
 
@@
 \label{eq:erm}
-R_{\text{emp}}[f] = \frac{1}{n} \sum_{i=1}^n V(f(x),y)
-\end{equation}
+R_{\text{emp}}[f] = \frac{1}{n} \sum_{i=1}^n V(f(x_i),y_i)
+\end{equation}
+
+The ERM is a Riemann sum approximation of the expected error of $f$ shown in equation \ref{eq:expetedrisk}, where $dp(x,y) \approx 1/n$.
 

### 2017-01-13 articles-conferences/iscef2016/body.tex
Review Intro
General comments 3 ready

--- 2016-07-06 articles-conferences/iscef2016/manuscript-computational-economics/body.tex
@@
 
-Our aims were to study if a parallel version of VECM can be used with high
-frequency stream data.  Our approach was to determine adaptively the number of
-observations and lags of VECM which maximise cointegration relations in the
-past.  This search is computationally expensive because the Johansen method, see
-\cite{johansen1995}, is required to find cointegration vectors, which is of
-order $O(n^3)$. Therefore, our proposal, called Adaptively Vector Error
-Correction (AVECM), is to parallelise this search in order to get new parameters
-before new data arrives.  Model effectiveness is focused on out-of-sample
-forecast rather than in-sample fitting. This criterion allows AVECM prediction
-capability to be expressed rather than just explaining data history.  The
-forecast capability of our method was measured using MSE and the Theil's
-$U$-statistic, see \cite{theil1966}, widely used in economic forecast. Tests
-were run using four currency rates: Euro (EUR) to United States Dollar (USD)
-(EURUSD), British Pound (GBP) to USD (GBPUSD), USD to Swiss Franc (CHF) (USDCHF)
-and USD to Japanese Yen (JPY) (USDJPY) with a 10-second frequency.
+Cointegration relationship can be found in low and high frequency data of two
+or more assets. While cointegration in low frequency data is motivated by a
+long-run equilibrium relationship between economic forces, cointegration in
+high frequency data has its foundation in statistical arbitrage which is very
+helpful to detect mean-reverting trades. Cointegration at different data
+frequencies may differ. \cite{rittler2012} shows how cointegration appears when
+the data frequency is increased. Our aims were to study if a parallel
+version of VECM can be used with high frequency stream data.
+
+Our approach was to determine adaptively the number of observations and lags of
+VECM which maximise cointegration relations in the past.  This search is
+computationally expensive because the Johansen method, see \cite{johansen1995},
+is required to find cointegration vectors, which is of order $O(n^3)$.
+Therefore, our proposal, called Adaptively Vector Error Correction (AVECM), is
+to parallelise this search in order to get new parameters before new data
+arrives.  Model effectiveness is focused on out-of-sample forecast rather than
+in-sample fitting. This criterion allows AVECM prediction capability to be
+expressed rather than just explaining data history.  The forecast capability of
+our method was measured using MSE and the Theil's $U$-statistic, see
+\cite{theil1966}, widely used in economic forecast. Tests were run using four
+currency rates: Euro (EUR) to United States Dollar (USD) (EURUSD), British
+Pound (GBP) to USD (GBPUSD), USD to Swiss Franc (CHF) (USDCHF) and USD to
+Japanese Yen (JPY) (USDJPY) with a 10-second frequency.
 
@@
 conclusions and a discussion of future research.
-\def\rot#1{{\color{red}{#1}}}
-\def\gruen#1{{\color{green}{#1}}}
-\def\blau#1{{\color{blue}{#1}}}
 

### 2017-01-16 articles-conferences/iscef2016/body.tex
intro re estructurado

@@
 
-VECM parameters are obtained using the ordinary least squares (OLS) method,
-developed by \cite{golub1980}. Since OLS involves many calculations, the
-parameter estimation is computationally expensive when the number of lagged
-values and data increases. Moreover, obtaining cointegration vectors is also an
-expensive routine. This is a main limitation to use VECM with stream data with
-high frequency. \cite{chen2003} addresses the advantage of distributed
-processing over conventional rolling window processing.
-
 Cointegration relationship can be found in low and high frequency data of two
@@
 frequencies may differ. \cite{rittler2012} shows how cointegration appears when
-the data frequency is increased. Our aims were to study if a parallel
-version of VECM can be used with high frequency stream data.
+the data frequency is increased.
+
+The use of VECM with high frequency data is mainly limited by computational 
+expensive routines. Firstly, VECM parameters are obtained using the ordinary least squares (OLS) method, developed by \cite{golub1980}. Since OLS involves many calculations, the
+parameter estimation is computationally expensive when the number of lagged
+values and data increases. Secondly, obtaining cointegration vectors is also an
+expensive routine because the Johansen method, see \cite{johansen1995}, is required, which is of order $O(n^3)$. \cite{chen2003} addresses the advantage of distributed
+processing over conventional rolling window processing. Therefore, our aims were to study if a parallel version of VECM can be used with high frequency stream data.
 
 Our approach was to determine adaptively the number of observations and lags of
-VECM which maximise cointegration relations in the past.  This search is
-computationally expensive because the Johansen method, see \cite{johansen1995},
-is required to find cointegration vectors, which is of order $O(n^3)$.
-Therefore, our proposal, called Adaptively Vector Error Correction (AVECM), is
-to parallelise this search in order to get new parameters before new data
+VECM which maximise cointegration relations in the past in a distributed environment. 
+We called Adaptively Vector Error Correction (AVECM) to our proposal. AVECM
+parallelise this search of parameters in order to update them before new data
 arrives. Model effectiveness is focused on out-of-sample forecast rather than
@@
   \centering
-  \includegraphics[width=0.9\textwidth]{Fig1}
+  %\includegraphics[width=0.9\textwidth]{Fig1}
+   \includegraphics[width=0.9\textwidth]{img/Fig1}
   \caption{Distribution of the number of cointegration vectors using $p=1$ lags.
@@
   \centering
-  \includegraphics[width=0.9\textwidth]{Fig2}
+  %\includegraphics[width=0.9\textwidth]{Fig2}
+   \includegraphics[width=0.9\textwidth]{img/Fig2}
   \caption{MSE versus the percentage of cointegration considering 1000
@@
   \centering
-  \includegraphics[width=0.9\textwidth]{Fig3}
+  %\includegraphics[width=0.9\textwidth]{Fig3}
+  \includegraphics[width=0.9\textwidth]{img/Fig3}
   \caption{Computing time of sequential and parallel algorithm is shown in the

### 2017-01-16 articles-conferences/iscef2016/body.tex
specific comments 4,5

@@
 
-Cointegration relationship can be found in low and high frequency data of two
-or more assets. While cointegration in low frequency data is motivated by a
-long-run equilibrium relationship between economic forces, cointegration in
-high frequency data has its foundation in statistical arbitrage which is very
-helpful to detect mean-reverting trades. Cointegration at different data
-frequencies may differ. \cite{rittler2012} shows how cointegration appears when
-the data frequency is increased.
+Cointegration relationship can be found in low and high frequency data of two or
+more assets. While cointegration in low frequency data is motivated by a
+long-run equilibrium relationship between economic forces, cointegration in high
+frequency data has its foundation in statistical arbitrage which is very helpful
+to detect mean-reverting trades. Cointegration at different data frequencies may
+differ. \cite{rittler2012} shows how cointegration appears when the data
+frequency is increased.
 
 The use of VECM with high frequency data is mainly limited by computational
-expensive routines. Firstly, VECM parameters are obtained using the ordinary least squares (OLS) method, developed by \cite{golub1980}. Since OLS involves many calculations, the
-parameter estimation is computationally expensive when the number of lagged
-values and data increases. Secondly, obtaining cointegration vectors is also an
-expensive routine because the Johansen method, see \cite{johansen1995}, is required, which is of order $O(n^3)$. \cite{chen2003} addresses the advantage of distributed
-processing over conventional rolling window processing. Therefore, our aims were to study if a parallel version of VECM can be used with high frequency stream data.
+expensive routines. Firstly, VECM parameters are obtained using the ordinary
+least squares (OLS) method, developed by \cite{golub1980}. Since OLS involves
+many calculations, the parameter estimation is computationally expensive when
+the number of lagged values and data increases. Secondly, obtaining
+cointegration vectors is also an expensive routine because the Johansen method,
+see \cite{johansen1995}, is required, which is of order $O(n^3)$.
+\cite{chen2003} addresses the advantage of distributed processing over
+conventional rolling window processing. Therefore, our aims were to study if a
+parallel version of VECM can be used with high frequency stream data.
 
 Our approach was to determine adaptively the number of observations and lags of
-VECM which maximise cointegration relations in the past in a distributed environment. 
-We called Adaptively Vector Error Correction (AVECM) to our proposal. AVECM
-parallelise this search of parameters in order to update them before new data
-arrives. Model effectiveness is focused on out-of-sample forecast rather than
-in-sample fitting. This criterion allows AVECM prediction capability to be
-expressed rather than just explaining data history.  The forecast capability of
-our method was measured using MSE and the Theil's $U$-statistic, see
-\cite{theil1966}, widely used in economic forecast. Tests were run using four
-currency rates: Euro (EUR) to United States Dollar (USD) (EURUSD), British
-Pound (GBP) to USD (GBPUSD), USD to Swiss Franc (CHF) (USDCHF) and USD to
-Japanese Yen (JPY) (USDJPY) with a 10-second frequency.
+VECM which maximise cointegration relations in the past in a distributed
+environment.  We called Adaptively Vector Error Correction (AVECM) to our
+proposal. AVECM parallelise this search of parameters in order to update them
+before new data arrives. Model effectiveness is focused on out-of-sample
+forecast rather than in-sample fitting. This criterion allows AVECM prediction
+capability to be expressed rather than just explaining data history.  The
+forecast capability of our method was measured using MSE and the Theil's
+$U$-statistic, see \cite{theil1966}, widely used in economic forecast. Tests
+were run using four currency rates: Euro (EUR) to United States Dollar (USD)
+(EURUSD), British Pound (GBP) to USD (GBPUSD), USD to Swiss Franc (CHF) (USDCHF)
+and USD to Japanese Yen (JPY) (USDJPY) with a 10-second frequency.
 
@@
 on the amount $L$ of historical data and the number of lags $p$ in the VECM, we
-used a grid search.  We defined a grid of possible values for $L$ and $p$. $L$
-goes throughout $[2,14]$ hours (1 $hour = 360$ data points) with a step size of
-4 hours and $p$ throughout $[1,5]$ with step size of 1. The first experiment
-consisted in determining the number of cointegration vectors for all
-combinations of $L$ and $p$.  We used four forex rates: EURUSD, GBPUSD, USDCHF
-and USDJPY with 10-second frequency. Data started at 13:00 GMT of the 13th of
-August 2014, when the New York and London financial markets opened.
-
-Figure \ref{fig:hists} shows the distribution of the number of cointegration
-vectors given by the Johansen method for different values of $L=[2,6,10,14]$
-hours  and $p=1$ . This procedure was carried out by a sliding window of
-historical data moving 1000 times.
+used a grid search.  We defined arbitrarily a grid of possible values for $L$
+and $p$. $L$ goes throughout $[2,14]$ hours (1 $hour = 360$ data points) with a
+step size of 4 hours and $p$ throughout $[1,5]$ with step size of 1. The idea is
+to show the variability of the number of cointegration vectors when we changed
+these two parameters. We used four forex rates: EURUSD, GBPUSD, USDCHF and
+USDJPY with 10-second frequency. Data started at 13:00 GMT of the 13th of August
+2014, when the New York and London financial markets opened.
 
@@
 \end{figure}
+
+Figure \ref{fig:hists} shows the distribution of the number of cointegration
+vectors given by the Johansen method for different values of $L=[2,6,10,14]$
+hours and $p=1$ . This procedure was carried out by a sliding window of
+historical data moving 1000 times. We can observe that the distribution of
+cointegration vectors is always changing with different values of $L$. When
+$L=2$ hours, there were no cointegration in more than 60\% of the iterations.
+Cointegration increases when $L=6$ hours and is maximum when $L=10$ hours where
+one cointegration vectors was found for all 1000 iterations. The ocurrence of
+cointegration starts to decrease with $L=14$ hours.
 

### 2017-01-17 articles-conferences/iscef2016/body.tex
comments 6,7,9

@@
 frequency data has its foundation in statistical arbitrage which is very helpful
-to detect mean-reverting trades. Cointegration at different data frequencies may
-differ. \cite{rittler2012} shows how cointegration appears when the data
-frequency is increased.
+to detect mean-reverting trades. Information, about cointegrated assets in high
+frequency data could be used as an input for high frequency trading strategies,
+using it as signal to capture small profits in short term trades, see
+\cite{miao2014}. \cite{zhou2001} addresses the benefits of using higher
+frequency data to analize cointegration. \cite{rittler2012} also found that
+cointegration may differ with different data frequencies and it could appear
+when the data frequency is increased.
 
@@
 and the performance measure MSE (see equation \ref{eq:MSE}). $L$ was defined
-between $[700,1500]$ data points with step size 100 and $p$ took values between
-$[1,5]$ with step size 1.
+between $[2,14]$ hours, that corresponds to $[720, 5040]$ data points, and $p$
+took values between $[1,5]$.
 
@@
 found that better cointegration percentage leads to better performance accuracy
-in terms of reducing MSE.
+in terms of reducing MSE. Also, increasing the size of the sliding window $L$
+doesn't necessarily help to reduce MSE.
 
@@
 \subsection{Data} \label{sec:unitroot}
-AVECM tests were carried out using four foreign exchange rates all related to
-USD: EURUSD, GBPUSD, USDCHF and USDJPY. This data was collected from
+All the experiments and AVECM tests were carried out using four foreign exchange rates all related to
+USD: EURUSD, GBPUSD, USDCHF and USDJPY. We chose the most traded rates related
+to USD so they are likely to be cointegrated. 
+This data was collected from
 \cite{Dukascopy2014}, a free database which gives access to the Swiss Foreign

### 2017-01-20 articles-conferences/iscef2016/body.tex
adding conclusions
correccion Jon

--- 2017-01-16 articles-conferences/iscef2016/body.tex
@@
 trading, see \cite{herlemont2003}, or models with more than two variables, see
-for example \cite{mukherjee1995} and \cite{engle2004}).
+for example \cite{mukherjee1995} and \cite{engle2004}.
 
-Cointegration relationship can be found in low and high frequency data of two
-or more assets. While cointegration in low frequency data is motivated by a
-long-run equilibrium relationship between economic forces, cointegration in
-high frequency data has its foundation in statistical arbitrage which is very
-helpful to detect mean-reverting trades. Cointegration at different data
-frequencies may differ. \cite{rittler2012} shows how cointegration appears when
-the data frequency is increased.
+Cointegration relationships can be found in low and high frequency data of two or
+more assets. While cointegration in low frequency data is motivated by a
+long-run equilibrium relationship between economic forces, cointegration in high
+frequency data has its foundation in statistical arbitrage which is very helpful
+to detect mean-reverting trades. Information about cointegrated assets in high
+frequency data could be used as an input for high frequency trading strategies,
+using it as a signal to capture small profits in short term trades, see
+\cite{miao2014}. \cite{zhou2001} addressed the benefits of using higher
+frequency data to analize cointegration. \cite{rittler2012} also found that
+cointegration may differ with different data frequencies and could appear
+whith increased data frequency.
 
-The use of VECM with high frequency data is mainly limited by computational 
-expensive routines. Firstly, VECM parameters are obtained using the ordinary least squares (OLS) method, developed by \cite{golub1980}. Since OLS involves many calculations, the
-parameter estimation is computationally expensive when the number of lagged
-values and data increases. Secondly, obtaining cointegration vectors is also an
-expensive routine because the Johansen method, see \cite{johansen1995}, is required, which is of order $O(n^3)$. \cite{chen2003} addresses the advantage of distributed
-processing over conventional rolling window processing. Therefore, our aims were to study if a parallel version of VECM can be used with high frequency stream data.
+The use of VECM with high frequency data is mainly limited by computationally
+expensive routines. Firstly, VECM parameters are obtained using the ordinary
+least squares (OLS) method, developed by \cite{golub1980}. Since OLS involves
+many calculations, the parameter estimation is computationally expensive when
+the number of lagged values and data increases. Secondly, obtaining
+cointegration vectors is also an expensive routine because the Johansen method 
+is required, which is of order $O(n^3)$ (\cite{johansen1995}).
+\cite{chen2003} addressed the advantage of distributed processing over
+conventional rolling window processing. Therefore, our aim was to study if a
+parallel version of VECM can be used with high frequency stream data.
 
-Our approach was to determine adaptively the number of observations and lags of
-VECM which maximise cointegration relations in the past in a distributed environment. 
-We called Adaptively Vector Error Correction (AVECM) to our proposal. AVECM
-parallelise this search of parameters in order to update them before new data
-arrives. Model effectiveness is focused on out-of-sample forecast rather than
-in-sample fitting. This criterion allows AVECM prediction capability to be
-expressed rather than just explaining data history.  The forecast capability of
-our method was measured using MSE and the Theil's $U$-statistic, see
-\cite{theil1966}, widely used in economic forecast. Tests were run using four
-currency rates: Euro (EUR) to United States Dollar (USD) (EURUSD), British
-Pound (GBP) to USD (GBPUSD), USD to Swiss Franc (CHF) (USDCHF) and USD to
-Japanese Yen (JPY) (USDJPY) with a 10-second frequency.
+Our approach was to determine, adaptively, the number of observations and lags of
+VECM which maximise cointegration relations in the past in a distributed
+environment. We called our proposal Adaptive Vector Error Correction (AVECM). 
+AVECM parallelises this search of parameters in order to update them
+before new data arrives. Model effectiveness is focused on out-of-sample
+forecast rather than in-sample fitting. This criterion allows AVECM prediction
+capability to be expressed rather than just explaining data history.  The
+forecast capability of our method was measured using MSE and the Theil's
+$U$-statistic, see \cite{theil1966}, widely used in economic forecast. Tests
+were run using four currency rates: Euro (EUR) to United States Dollar (USD)
+(EURUSD), British Pound (GBP) to USD (GBPUSD), USD to Swiss Franc (CHF) (USDCHF)
+and USD to Japanese Yen (JPY) (USDJPY) with a 10-second frequency.
 
@@
 on the amount $L$ of historical data and the number of lags $p$ in the VECM, we
-used a grid search.  We defined a grid of possible values for $L$ and $p$. $L$
-goes throughout $[2,14]$ hours (1 $hour = 360$ data points) with a step size of
-4 hours and $p$ throughout $[1,5]$ with step size of 1. The first experiment
-consisted in determining the number of cointegration vectors for all
-combinations of $L$ and $p$.  We used four forex rates: EURUSD, GBPUSD, USDCHF
-and USDJPY with 10-second frequency. Data started at 13:00 GMT of the 13th of
-August 2014, when the New York and London financial markets opened.
-
-Figure \ref{fig:hists} shows the distribution of the number of cointegration
-vectors given by the Johansen method for different values of $L=[2,6,10,14]$
-hours  and $p=1$ . This procedure was carried out by a sliding window of
-historical data moving 1000 times.
+used a grid search.  We arbitrarily defined a grid of possible values for $L$
+and $p$. $L$ goes throughout $[2,14]$ hours (1 $hour = 360$ data points) with a
+step size of 4 hours and $p$ throughout $[1,5]$ with a step size of 1. The idea
+was to show the variability of the number of cointegration vectors when we changed
+these two parameters. We used four forex rates: EURUSD, GBPUSD, USDCHF and
+USDJPY with 10-second frequency. Data started at 13:00 GMT of the 13th of August
+2014, when the New York and London financial markets opened.
 
@@
 \end{figure}
+
+Figure \ref{fig:hists} shows the distribution of the number of cointegration
+vectors given by the Johansen method for different values of $L=[2,6,10,14]$
+hours and $p=1$ . This procedure was carried out by a sliding window of
+historical data moving 1000 times. We observed that the distribution of
+cointegration vectors changed with different values of $L$. When
+$L=2$ hours, there was no cointegration in more than 60\% of the iterations.
+Cointegration increased when $L=6$ hours and was maximum when $L=10$ hours where
+one cointegration vectors was found for all 1000 iterations. The ocurrence of
+cointegration started to decrease at $L=14$ hours.
 
@@
 
-The goal of our next experiment was to find a relation between this ratio $PC$
+The goal of our next experiment was to find a relationship between the ratio $PC$
 and the performance measure MSE (see equation \ref{eq:MSE}). $L$ was defined
-between $[700,1500]$ data points with step size 100 and $p$ took values between
-$[1,5]$ with step size 1.
+between $[2,14]$ hours, that corresponded to $[720, 5040]$ data points, and $p$
+took values between $[1,5]$.
 
@@
   \caption{MSE versus the percentage of cointegration considering 1000
-  iterations. Best windows size $L$ found was 1060. Below MSE versus $L$ shows a
-  rapidly decreasing behaviour, founding minimum at PC = 80\% }
+  iterations. Optimum windows size $L$ found was 1060. Below MSE versus $L$ shows a
+  rapidly decreasing behaviour, founding minimum at PC = 73\% }
   \label{fig:cointvsmse}
@@
 
-Figure~\ref{fig:cointvsmse} shows the relation between MSE and $PC$ and $L$.  We
-found that better cointegration percentage leads to better performance accuracy
-in terms of reducing MSE.
+Figure~\ref{fig:cointvsmse} shows the relationship between MSE and $PC$ and $L$. We
+found that higher cointegration percentage leads to improved performance accuracy
+in terms of lower MSE. Also, increasing the size of the sliding window $L$
+doesn't necessarily help to reduce MSE.
 
@@
 \subsection{Data} \label{sec:unitroot}
-AVECM tests were carried out using four foreign exchange rates all related to
-USD: EURUSD, GBPUSD, USDCHF and USDJPY. This data was collected from
+All the experiments and AVECM tests were carried out using four foreign exchange rates all related to
+USD: EURUSD, GBPUSD, USDCHF and USDJPY. We chose the most traded rates related
+to USD so they were likely to be cointegrated. 
+This data was collected from
 \cite{Dukascopy2014}, a free database which gives access to the Swiss Foreign
@@
 9 times and therefore a response time was obtain before 10 seconds. Since we
-used 10-second frequencys we can say that our proposal is suitable for use in
-an online context for real applications because response times were less than
-this frequency.
+used 10-second frequencys we can say that our proposal is suitable for use in an
+online context for real applications because response times were less than this
+frequency. Cointegration information can now easily be used as an integration tool
+to detect arbitrage opportunities or risk control.
 

### 2017-02-01 articles-conferences/iscef2016/body.tex
first version of editor letter

@@
 long-run equilibrium relationship between economic forces, cointegration in high
-frequency data has its foundation in statistical arbitrage which is very helpful
+frequency data has its foundation in statistical arbitrage theory which is very helpful
 to detect mean-reverting trades. Information about cointegrated assets in high

### 2017-03-31 articles-conferences/elsevier-2016/src/background.tex
adding comment characters in crossvalidation section

@@
 %'---------------------------http://robjhyndman.com/hyndsight/crossvalidation/
-When the data are not independent cross-validation becomes more difficult as leaving out an observation does not remove all the associated information due to the correlations with other observations.
+%When the data are not independent cross-validation becomes more difficult as leaving out an observation does not remove all the associated information due to the correlations with other observations.
 For time series forecasting, a cross-validation statistic is obtained as follows
 
-Fit the model to the data y1,…,yt and let y^t+1 denote the forecast of the next observation. Then compute the error (et+1∗=yt+1−y^t+1) for the forecast observation.
-Repeat step 1 for t=m,…,n−1 where m is the minimum number of observations needed for fitting the model.
-Compute the MSE from em+1∗,…,en∗.
-References
+%Fit the model to the data y1,…,yt and let y^t+1 denote the forecast of the next observation. Then compute the error (et+1∗=yt+1−y^t+1) for the forecast observation.
+%Repeat step 1 for t=m,…,n−1 where m is the minimum number of observations needed for fitting the model.
+%Compute the MSE from em+1∗,…,en∗.
+%References
 
-An excellent and comprehensive recent survey of cross-validation results is Arlot and Celisse (2010)
+%An excellent and comprehensive recent survey of cross-validation results is Arlot and Celisse (2010)
 

### 2017-03-31 articles-conferences/elsevier-2016/src/background.tex
deleting matrix demonstrations

@@
 Following Johansen \cite{johansen1995} we shall say that a stochastic process
-$Y_t$ which satisfies $Y_t-E(Y_t) = \sum_{i=0}^\infty C_i\,\varepsilon_{t-i}$ is
-called $I(0)$, and then we shall write $Y_t\sim I(0)$, whenever
+$Y_t$ which satisfies $Y_t-E(Y_t) = \sum_{i=0}^\infty C_i\,\varepsilon_{t-i}$
+is called $I(0)$, and then we shall write $Y_t\sim I(0)$, whenever
 $\sum_{i=0}^\infty C_i \neq 0$ and $\sum_{i=0}^\infty C_i\,z^i$ converges for
@@
 $d$, and then we shall write $\mathbf{y}_t\sim I(d)$, whenever after $d$ times
-(discrete) differentiation an stationary process is obtained~\cite{banerjee1993};
-more precisely, whenever
+(discrete) differentiation an stationary process is
+obtained~\cite{banerjee1993}; more precisely, whenever
 $(1-L)^d\,\mathbf{y}_t\sim\text{I(0)}$, where $L$ is the usual lag operator:
@@
 $\beta=[\beta(1),\dots,\beta(l)]^\top \in \mathbb{R}^p$ exists, such that the
-time series,
-\begin{equation}
-\mathbf{Z}_t:= 
-\sum_{\nu=1}^l \beta(\nu)\,\mathbf{y}_t^\nu\,\sim\,\text{I(0)}\,.
-\end{equation}
-In other words, a set of $I(1)$ time series is said to be cointegrated if a
-linear combination of them exists, which is I(0).
+time series, \begin{equation} \mathbf{Z}_t:= \sum_{\nu=1}^l
+\beta(\nu)\,\mathbf{y}_t^\nu\,\sim\,\text{I(0)}\,.  \end{equation} In other
+words, a set of $I(1)$ time series is said to be cointegrated if a linear
+combination of them exists, which is I(0).
 
@@
 
-VECM is a special case of the VAR \cite{sims1980} model and both describe the joint behaviour
-of a set of variables.
+VECM is a special case of the VAR \cite{sims1980} model and both describe the
+joint behaviour of a set of variables.
 
-The VAR($p$) model is a general framework describing the behaviour of a
-set of $l$ endogenous variables as a linear combination of their last $p$
-values, where $l,p\in\mathbb{N}$. 
-In our case, each one of these $l$ variables is a scalar time series
-$y_{\lambda,t}$, $\lambda=1,\dots,l$, and we represent them all together
-at time $t$ by the the vector time series:
-\begin{equation}
-\label{eq:variables}
-\mathbf{y}_t = 
-\begin{bmatrix} y_{1,t} & y_{2,t} & \dots & y_{l,t} \end{bmatrix}^\top.
-\end{equation}
-\noindent
-Notice that the vectors $\mathbf{y}_t$ are assumed to be $l$-dimensional.
+The VAR($p$) model is a general framework describing the behaviour of a set of
+$l$ endogenous variables as a linear combination of their last $p$ values,
+where $l,p\in\mathbb{N}$.  In our case, each one of these $l$ variables is a
+scalar time series $y_{\lambda,t}$, $\lambda=1,\dots,l$, and we represent them
+all together at time $t$ by the the vector time series: \begin{equation}
+\label{eq:variables} \mathbf{y}_t = \begin{bmatrix} y_{1,t} & y_{2,t} & \dots &
+y_{l,t} \end{bmatrix}^\top.  \end{equation} \noindent Notice that the vectors
+$\mathbf{y}_t$ are assumed to be $l$-dimensional.
 
@@
 
-However, the VAR model cannot be used with non-stationary variables. VECM \cite{engle87} is also a linear model and a special form of a VAR model for I(1) variables that are also
-cointegrated~\cite{banerjee1993}. If cointegration exists, variable differences
-are stationary and they introduce an error correction term which adjusts
-coefficients to bring the variables back to equilibrium.
+However, the VAR model cannot be used with non-stationary variables. VECM
+\cite{engle87} is also a linear model and a special form of a VAR model for
+I(1) variables that are also cointegrated~\cite{banerjee1993}. If cointegration
+exists, variable differences are stationary and they introduce an error
+correction term which adjusts coefficients to bring the variables back to
+equilibrium.
 
@@
 %
-\noindent where $\mathbf{B}$ dimension is $((N-p)\times l)$, $\mathbf{A}$
-dimension is $((N-p)\times(r+(p-1)l +1))$, $\mathbf{X}$ dimension is $((r+(p-1)l
-+1)\times l)$ and $\mathbf{E}$ dimension is $((N-p)\times l)$:
-%
-\begin{alignat}{3}
-\mathbf{B}
-&= \begin{bmatrix}
-   \Delta\mathbf{y}_{p+1}^\top \\
-   \Delta\mathbf{y}_{p+2}^\top \\
-   \vdots \\
-   \Delta\mathbf{y}_N^\top
-   \end{bmatrix}
-&\quad
-\mathbf{X}
-&= \begin{bmatrix}
-   \boldsymbol{\alpha}^\top \\
-   \boldsymbol{\Phi}_1^{*\top} \\
-   \boldsymbol{\Phi}_2^{*\top} \\
-   \vdots \\
-   \boldsymbol{\Phi}_{p-1}^{*\top} \\
-   \mathbf{c}^\top
-   \end{bmatrix}
-&\quad
-\mathbf{E}
-&= \begin{bmatrix}
-   \boldsymbol{\epsilon}_{p+1}^\top \\
-   \boldsymbol{\epsilon}_{p+2}^\top \\
-   \vdots \\
-   \boldsymbol{\epsilon}_N^\top \\
-   \end{bmatrix}
-\end{alignat}
-\noindent and 
-\begin{align}
-\mathbf{A} 
-&= \begin{pmat}[{....|}]
-   \mathbf{y}_p^\top \boldsymbol{\beta} & \Delta \mathbf{y}_p^\top & \Delta\mathbf{y}_{p-1}^\top & \dots 
-                    & \Delta\mathbf{y}_2^\top & 1 \cr
-   \mathbf{y}_{p+1}^\top  \boldsymbol{\beta} &\Delta\mathbf{y}_{p+1}^\top & \Delta\mathbf{y}_p^\top & \dots
-                       & \Delta\mathbf{y}_3^\top & 1 \cr
-   \vdots & \vdots & \vdots & \ddots & \vdots & \vdots \cr
-   \mathbf{y}_{N-1}^\top  \boldsymbol{\beta} &\Delta\mathbf{y}_{N-1}^\top & \Delta\mathbf{y}_{N-2}^\top & \dots 
-                       & \Delta\mathbf{y}_{N-p-1}^\top & 1 \cr
-   \end{pmat}\, .
-\label{eq:Amatrix}
-\end{align}
 Taking into account the error term $\mathbf{E}$, equation~(\ref{eq:vareq}) 
@@
 squares (OLS) estimation.
-
-\subsection{Ordinary Least Squares method}
-
-When $\mathbf{A}$ is singular, solution to equation~(\ref{eq:vareq}) is given
-by the ordinary least squares (OLS) method. OLS consists of minimizing the sum
-of squared errors or equivalently minimizing the following expression:
-
-\begin{equation}
-\label{eq:regressionproblem}
-\underset{\mathbf{X}}{\text{min}} \quad \| \mathbf{A}\mathbf{\mathbf{X}} - \mathbf{B} \|_2^2
-\end{equation}
-
-\noindent for which the solution $\hat{\mathbf{X}}$ is well-known:
-
-\begin{equation}
-\label{eq:MP}
-\hat{\mathbf{X}}=\mathbf{A}^{\!\!+}\,\mathbf{B}
-\end{equation}
-
-\noindent where $\mathbf{A}^{\!\!+}$ is the Moore-Penrose pseudo-inverse
-which can be written as follows:
-
-\begin{equation}
-\label{eq:pseudoinverse}
-\mathbf{A}^{\!\!+}= (\mathbf{A}^{\!\!\top} \mathbf{A})^{-1}\mathbf{A}^{\!\!\top} \, .
-\end{equation}
-
-However, when $\mathbf{A}$ is not full rank, i.e
-$rank(\mathbf{A})=k <  n \leq m$, $\mathbf{A}^\top \mathbf{A}$ is
-always singular and equation~(\ref{eq:pseudoinverse}) cannot be used.
-More generally, the pseudo-inverse is best computed using the compact
-singular value decomposition (SVD) of $\mathbf{A}$:
-
-\begin{equation}
-    \label{eq:compactsvd}
-    \underset{m \times n}{\mathbf{A}}=
-    \underset{m \times k}{\mathbf{U_1}} \enskip
-    \underset{k \times k}{\Sigma_1} \enskip
-    \underset{k \times n}{\mathbf{V}_1^{\top}} \, ,
-\end{equation}
-
-\noindent as follows
-
-\begin{equation}
-\label{eq:pseudoinversesvd}
-\mathbf{A}^{\!\!+} = \mathbf{V}_1 \Sigma_1^{-1} \mathbf{U}_1^\top \, .
-\end{equation}
 

### 2017-05-10 articles-conferences/elsevier-2016/src/background.tex
elsevier

--- 2016-04-08 articles-conferences/elsevier-2016/src/background.tex
@@
 
-Fit the model to the data y1,…,yt and let y^t+1 denote the forecast of the next observation. Then compute the error (et+1∗=yt+1−y^t+1) for the forecast observation.
+Fit the model to the data $y1,…,yt$ and let $y^t+1$ denote the forecast of the next observation. Then compute the error ($et+1∗=yt+1−y^t+1$) for the forecast observation.
 Repeat step 1 for t=m,…,n−1 where m is the minimum number of observations needed for fitting the model.
-Compute the MSE from em+1∗,…,en∗.
+Compute the MSE from $em+1∗,…,en∗$.
 References

### 2017-05-10 src/3_TimeSeries.tex

@@
 \subsection{Spurious regression} \label{sec:spurious}
-The use of non-stationary data can lead to spurious regressions. Spurious regression
-dates back to Yule in 1926 \cite{yule1926}. If two stationary
-variables are generated as independent random series and are trending over time,
-when one of those variables is regressed on the other, they could have a high
-$R^2$ even if the two are totally unrelated. So, if standard regression
-techniques are applied to non-stationary data, it could look good under standard
-measures but valueless \cite{brooks2002}.
+The use of non-stationary data can lead to spurious regressions. Spurious
+regression dates back to Yule in 1926 \cite{yule1926}. If two stationary
+variables are generated as independent random series and are trending over
+time, when one of those variables is regressed on the other, they could have a
+high $R^2$ even if the two are totally unrelated. So, if standard regression
+techniques are applied to non-stationary data, it could look good under
+standard measures but valueless \cite{brooks2002}.
+The
+regression measure $R^2$ gives high values in presence of spurious
+relationships. This measure would means that the model explains all the
+variability of the response data around its mean which is not true.
+
+The $R^2$ measure is calculated as follows:
+
+\begin{equation}
+\label{eq:rsquared}
+R^2 = 1 - \frac{\sum_{t=1}^T \epsilon_t^2}{\sum_{t=1}^T(y_t - \bar{y})^2}
+\end{equation}
+
+In presence of spurious relationships the denominator of equation
+\ref{eq:rsquared} becomes very large because a large weight is placed on
+extreme observations on either side of the mean $\bar{y}$.
 
 \textbf{Example of spurious regression} \quad 
-An example is shown below. Two random walks series $x_{1,t}$ and $x_{2,t}$ with a small drift ($\alpha=0.09$) are generated and
-then regressed one on the other:
+An example is shown below. Two random walks series $x_{1,t}$ and $x_{2,t}$ with
+a small drift ($\alpha=0.09$) are generated and then regressed one on the
+other:
  \begin{equation}
- x_{i,t} = \alpha + x_{i,t-1} + \epsilon_i,t  \qquad \text{where} \qquad \mathcal{N}(0,1), \quad i=1,2
+ x_{i,t} = \alpha + x_{i,t-1} + \epsilon_i,t  \qquad \text{where} \qquad
+\mathcal{N}(0,1), \quad i=1,2
  \end{equation}
@@
 
-These two random walks, which are independent by construction, appear to be related with a $R^2=0.883$.
-However, what it is happening is that they are drifting in the same direction.
-Drift can be removed using returns or first differences.
-If after taking returns or first differences the regression fit is still good we will say that the variables are cointegrated, this means that both series may drift but their residuals will not.
-A more formal definition of cointegration is given in the section \ref{sec:cointegration}.
+These two random walks, which are independent by construction, appear to be
+related with a $R^2=0.883$. However, what it is happening is that they are
+drifting in the same direction. Drift can be removed using returns or first
+differences.  If after taking returns or first differences the regression fit
+is still good we will say that the variables are cointegrated, this means that
+both series may drift but their residuals will not. A more formal definition of
+cointegration is given in the section \ref{sec:cointegration}.
 
@@
 \newpage
-Conventional regression estimators, including VARs, have good properties when applied to covariance-stationary time series, but encounter difficulties when applied to non stationary or integrated processes.
+Conventional regression estimators, including VARs, have good properties when
+applied to covariance-stationary time series, but encounter difficulties when
+applied to non stationary or integrated processes.
 
-These difficulties were illustrated by Granger and Newbold in 1974 \cite{granger1974} when they introduced the concept of spurious regressions (see section \ref{sec:spurious}).
-In 1982, Nelson and Plosser  \cite{nelson1982} showed that unit roots might be present in a wide variety of macroeconomic series
-in levels or logarithms. This finding gave rise to the industry of unit root testing, and
-the implication that variables should be rendered stationary by
-differencing before they are included in an econometric model.
-Further theoretical developments by Granger and Engle in 1987 \cite{engle87} raised the possibility that two or more integrated, non stationary time series might be cointegrated, so that some linear combination of these series could be stationary even though each series were not.
+These difficulties were illustrated by Granger and Newbold in 1974
+\cite{granger1974} when they introduced the concept of spurious regressions
+(see section \ref{sec:spurious}). In 1982, Nelson and Plosser \cite{nelson1982}
+showed that unit roots might be present in a wide variety of macroeconomic
+series in levels or logarithms. This finding gave rise to the industry of unit
+root testing, and the implication that variables should be rendered stationary
+by differencing before they are included in an econometric model.  Further
+theoretical developments by Granger and Engle in 1987 \cite{engle87} raised the
+possibility that two or more integrated, non stationary time series might be
+cointegrated, so that some linear combination of these series could be
+stationary even though each series were not.
 
-If two series are both integrated (of order one, or I(1)) we could model
-their interrelationship by taking first differences of each series and
-including the differences in a VAR or a structural model.
-However, this approach would be suboptimal if it was determined that
-these series are indeed cointegrated. In that case, the VAR only
-express the short-run responses of these series to innovations in each
-series. This implies that the simple regression in first differences is
-misspecified.
-If the series are cointegrated, they move together in the long-run. A
-VAR in first differences will not capture those long-run tendencies.
-VECM fit to the first differences of the non stationary variables, but a lagged error-correction term is added to the relationship representing the long-run relationship.
+If two series are both integrated (of order one, or I(1)) we could model their
+interrelationship by taking first differences of each series and including the
+differences in a VAR or a structural model.  However, this approach would be
+suboptimal if it was determined that these series are indeed cointegrated. In
+that case, the VAR only express the short-run responses of these series to
+innovations in each series. This implies that the simple regression in first
+differences is misspecified.  If the series are cointegrated, they move
+together in the long-run. A VAR in first differences will not capture those
+long-run tendencies.  VECM fit to the first differences of the non stationary
+variables, but a lagged error-correction term is added to the relationship
+representing the long-run relationship.
 
@@
 are differentiated and moreover cointegration restrictions often improve
-forecasting \cite{duy1998}. Therefore, VECM has been widely adopted in
-finance: \cite{mukherjee1995}, \cite{maysami2000}, \cite{arestis2001} and
-\cite{seong2013} to name a few. Pair
-trading is a very common example of a cointegration
-application \cite{herlemont2003} but also can also be extended to a larger set
-of variables \cite{mukherjee1995}, \cite{engle2004}.
+forecasting \cite{duy1998}. Therefore, VECM has been widely adopted in finance:
+\cite{mukherjee1995}, \cite{maysami2000}, \cite{arestis2001} and
+\cite{seong2013} to name a few. Pair trading is a very common example of a
+cointegration application \cite{herlemont2003} but also can also be extended to
+a larger set of variables \cite{mukherjee1995}, \cite{engle2004}.

### 2017-05-29 src/5_1_Proposal.tex
adding last changes from Springer journal

@@
-\chapter{Adaptive parallel algorithm for Vector error correction model}
-VECM (presented in section \ref{sec:vecm}) has been
+\chapter{Fast and adaptive cointegration based model
+for forecasting high frequency financial time series}
+
+Cointegration is a long-run property of some non-stationary time series where a
+linear combination of those time series is stationary.  This behaviour has been
+studied in finance because cointegration restrictions often improve forecasting.
+The Vector Error Correction Model (VECM) is a well-known econometric technique
+that characterises short-run variations of a set of cointegrated time series
+incorporating long-run relationships as an error correction term. VECM has been
 broadly used with low frequency time series. We aimed to adapt VECM to be used
@@
 
-We observed that the number of cointegration relationships varies with the
+Cointegration relations change in time and therefore VECM parameters must be
+updated when new data is available. We studied how forecasting performance is
+affected when VECM parameters and the length of historical data used change in
+time. We observed that the number of cointegration relationships varies with the
 length of historical data used. Moreover, parameters that increased these
@@
 trading, see \cite{herlemont2003}, or models with more than two variables, see
-for example \cite{mukherjee1995} and \cite{engle2004}).
+for example \cite{mukherjee1995} and \cite{engle2004}.
 
-VECM parameters are obtained using the ordinary least squares (OLS) method,
-developed by \cite{golub1980}. Since OLS involves many calculations, the
-parameter estimation is computationally expensive when the number of lagged
-values and data increases. Moreover, obtaining cointegration vectors is also an
-expensive routine. This is a main limitation to use VECM with stream data with
-high frequency. \cite{chen2003} addresses the advantage of distributed
-processing over conventional rolling window processing.
+Cointegration relationships can be found in low and high frequency data of two or
+more assets. While cointegration in low frequency data is motivated by a
+long-run equilibrium relationship between economic forces, cointegration in high
+frequency data has its foundation in statistical arbitrage theory which is very helpful
+to detect mean-reverting trades. Information about cointegrated assets in high
+frequency data could be used as an input for high frequency trading strategies,
+using it as a signal to capture small profits in short term trades, see
+\cite{miao2014}. \cite{zhou2001} addressed the benefits of using higher
+frequency data to analize cointegration. \cite{rittler2012} also found that
+cointegration may differ with different data frequencies and could appear
+whith increased data frequency.
 
-Our aims were to study if a parallel version of VECM can be used with high
-frequency stream data.  Our approach was to determine adaptively the number of
-observations and lags of VECM which maximise cointegration relations in the
-past.  This search is computationally expensive because the Johansen method, see
-\cite{johansen1995}, is required to find cointegration vectors, which is of
-order $O(n^3)$. Therefore, our proposal, called Adaptively Vector Error
-Correction (AVECM), is to parallelise this search in order to get new parameters
+The use of VECM with high frequency data is mainly limited by computationally
+expensive routines. Firstly, VECM parameters are obtained using the ordinary
+least squares (OLS) method, developed by \cite{golub1980}. Since OLS involves
+many calculations, the parameter estimation is computationally expensive when
+the number of lagged values and data increases. Secondly, obtaining
+cointegration vectors is also an expensive routine because the Johansen method 
+is required, which is of order $O(n^3)$ (\cite{johansen1995}).
+\cite{chen2003} addressed the advantage of distributed processing over
+conventional rolling window processing. Therefore, our aim was to study if a
+parallel version of VECM can be used with high frequency stream data.
+
+Our approach was to determine, adaptively, the number of observations and lags of
+VECM which maximise cointegration relations in the past in a distributed
+environment. We called our proposal Adaptive Vector Error Correction (AVECM). 
+AVECM parallelises this search of parameters in order to update them
 before new data arrives. Model effectiveness is focused on out-of-sample
@@
 and USD to Japanese Yen (JPY) (USDJPY) with a 10-second frequency.
+
+This paper is organised as follows: section \ref{sec:background} presents the
+VAR and VECM, the AVECM algorithm is presented in section \ref{sec:methodology}.
+In section \ref{sec:results} we describe the tests carried on to assess the
+accuracy and the execution time of AVECM.  This section also includes a
+description of the test data.  Section \ref{sec:conclusions} contains the
+conclusions and a discussion of future research.
+
+\section{Background}
+\label{sec:background}
+
+\subsection{Integration and Cointegration}\label{sec:coint}\  
+Following \cite{johansen1995} we shall say that a stochastic process
+$Y_t$ which satisfies $Y_t-E(Y_t) = \sum_{i=0}^\infty C_i\,\varepsilon_{t-i}$ is
+called $I(0)$, and then we shall write $Y_t\sim I(0)$, whenever
+$\sum_{i=0}^\infty C_i \neq 0$ and $\sum_{i=0}^\infty C_i\,z^i$ converges for
+$z\in\mathbb{C}$ with $|z|<1$.  It is understood that the condition
+$\varepsilon_t\sim iid(0,\sigma^2)$ holds.
+
+A (vector) time series $\mathbf{y}_t$ is said to be {\em integrated of order\/}
+$d$, and then we shall write $\mathbf{y}_t\sim I(d)$, whenever after $d$ times
+(discrete) differentiation an stationary process is
+obtained, see \cite{banerjee1993}; more precisely, whenever
+$(1-L)^d\,\mathbf{y}_t\sim\text{I(0)}$, where $L$ is the usual lag operator:
+$(1-L)\,\mathbf{y}_t = \Delta\mathbf{y}_t = \mathbf{y}_t-\mathbf{y}_{t-1}$ for
+all $t$.
+
+Note that this definition includes the scalar case as time series of vectors of
+dimension 1; in this scalar case we will write the time series in nonbold
+format.
+
+Let $\mathbf{y}_t^\nu$, $\nu=1,\dots,l$, be a set of $l$ vector time series of
+order $I(1)$.  They are said to be {\em cointegrated\/} if a vector
+$\beta=[\beta(1),\dots,\beta(l)]^\top \in \mathbb{R}^p$ exists, such that the
+time series,
+\begin{equation}
+\mathbf{Z}_t:= 
+\sum_{\nu=1}^l \beta(\nu)\,\mathbf{y}_t^\nu\,\sim\,\text{I(0)}\,.
+\end{equation}
+In other words, a set of $I(1)$ time series is said to be cointegrated if a
+linear combination of them exists, which is I(0).
+
+\subsection{Vector Autorregresive Models}\label{sec:varvec}
+
+Vector error correction model (VECM) describe the joint behaviour of a set of
+variables and can be derived from the simple Vector Autoregressive model (VAR)
+presented by \cite{sims1980}.  The VAR($p$) model is a framework describing the
+behaviour of a set of $l$ endogenous and stationary variables as a linear
+combination of their last $p$ values, where $l,p\in\mathbb{N}$.  In our case,
+each one of these $l$ variables is a scalar time series $y_{\lambda,t}$,
+$\lambda=1,\dots,l$, and we represent them all together at time $t$ by the
+vector time series:
+\begin{equation}
+\label{eq:variables}
+\mathbf{y}_t = 
+\begin{bmatrix} y_{1,t} & y_{2,t} & \dots & y_{l,t} \end{bmatrix}^\top.
+\end{equation}
+\noindent
+Notice that the vectors $\mathbf{y}_t$ are assumed to be $l$-dimensional.
+
+The VAR($p$) model describes the behaviour of a dependent variable in terms of
+its own lagged values and the lags of the others variables in the system. The
+model with $p$ lags is formulated as the system:
+\begin{align}
+\label{eq:var}
+\mathbf{y}_t 
+= \boldsymbol{\Phi}_1 \mathbf{y}_{t-1} +
+  \boldsymbol{\Phi}_2 \mathbf{y}_{t-2} + \dots +
+  \boldsymbol{\Phi}_p\mathbf{y}_{t-p} +
+  \mathbf{c} + \boldsymbol{\epsilon}_t \nonumber \\
+t=p+1,\dots,N,
+\end{align}
+\noindent where 
+$\boldsymbol{\Phi}_1, \boldsymbol{\Phi}_2,\dots,\boldsymbol{\Phi}_p$
+are $l\times l$-matrices of real coefficients,
+$\boldsymbol{\epsilon}_{p+1},
+ \boldsymbol{\epsilon}_{p+2}, \dots, \boldsymbol{\epsilon}_N$ 
+are error terms, $\mathbf{c}$ is a constant vector and $N$ is the total
+number of samples.
+
+Notice that, regarding our notation of section (\ref{sec:coint}),
+we have here 
+$\mathbf{y}_t^0 = \mathbf{y}_t$,
+$\mathbf{y}_t^\nu = \mathbf{y}_{t-\nu}$ and
+the $\lambda$-th component of the vector time series $\mathbf{y}_t^\nu$
+is the scalar time series $y_{\lambda,t}^\nu$, where $\nu=1,\dots,p$ and
+$\lambda=1,\dots,l$.
+
+However, the VAR model cannot be used with non-stationary variables. VECM ,
+developed by \cite{engle87}, is also a linear model for I(1) variables that are
+also cointegrated, see \cite{banerjee1993}. If cointegration exists, variable
+differences are stationary and they introduce an error correction term which
+adjusts coefficients to bring the variables back to equilibrium.
+
+It is obtained re-writing equation (\ref{eq:var}) in terms of the new
+variable $\Delta\mathbf{y}_t=\mathbf{y}_t-\mathbf{y}_{t-1}$.
+The VECM model, expressed in terms those differences, takes the form:
+\begin{equation}\label{eq:vec}
+\Delta \mathbf{y}_t 
+= \boldsymbol{\Omega}\,\mathbf{y}_{t-1}
+  + \sum_{i=1}^{p-1} \boldsymbol{\Phi}_i^*\,\Delta\mathbf{y}_{t-i}
+  + \mathbf{c} + \boldsymbol{\epsilon}_t\,,
+\end{equation}
+\noindent
+where the coefficients matrices $\boldsymbol{\Phi}_i^*$ and 
+$\boldsymbol{\Omega}$, expressed in terms of the matrices
+$\boldsymbol{\Phi}_i$ of (\ref{eq:var}), are:
+\begin{align*}
+\boldsymbol{\Phi}_i^* 
+&:= -\sum_{j=i+1}^{p}\boldsymbol{\Phi}_j\,, \\
+\boldsymbol{\Omega}
+&:= -\left( \mathbb{I} - \boldsymbol{\Phi}_1 - \dots 
+    - \boldsymbol{\phi}_p \right)\,. 
+\end{align*}
+The following well known properties of the matrix $\boldsymbol{\Omega}$, see
+\cite{johansen1995}, will be useful in the sequel:
+\begin{itemize}
+\item
+If $\boldsymbol{\Omega} = \mathbf{0}$, there is no cointegration.
+\item 
+If $rank(\boldsymbol{\Omega})=l$, i.e., if $\boldsymbol{\Omega}$ has
+full rank, then the time series are not I(1) but stationary.
+\item
+If $rank(\boldsymbol{\Omega})=r$, $0<r<l$, then there is cointegration
+and the matrix $\boldsymbol{\Omega}$ can be expressed as
+$\boldsymbol{\Omega}=\boldsymbol{\alpha\beta}^\top$, where $\boldsymbol{\alpha}$
+and $\boldsymbol{\beta}$ are
+$l\times r$ matrices and
+$\text{rank}(\boldsymbol{\alpha})=\text{rank}(\boldsymbol{\beta})=r$.
+\item
+The columns of $\boldsymbol{\beta}$ contains the cointegration vectors and the rows of
+$\boldsymbol{\alpha}$ correspond with the adjusted vectors. 
+$\boldsymbol{\beta}$ is obtained by Johansen procedure,see \cite{johansen1988},
+whereas $\boldsymbol{\alpha}$ has to be determined as a variable in the VECM.
+\end{itemize}
+It is worth noticing that the factorization of the matrix
+$\boldsymbol\Omega$ is not unique, since for any $r \times r$
+non-singular matrix $\mathbf{H}$, $\boldsymbol{\alpha}^*:=\boldsymbol{\alpha}\mathbf{H}$,
+and $\boldsymbol{\beta}^*=\boldsymbol{\beta}(\mathbf{H}^{-1})^\top$ we have
+$\boldsymbol{\alpha\beta}^\top=\boldsymbol{\alpha}^*(\boldsymbol{\beta}^*)^\top$.
+If cointegration exists, then equation (\ref{eq:vec}) can be written
+as follows:
+\begin{equation}\label{eq:vecfull}
+\Delta\mathbf{y}_t 
+= \boldsymbol{\alpha\beta}^\top\mathbf{y}_{t-1} 
+  + \sum_{i=1}^{p-1}\boldsymbol{\Phi}_i^*\,\Delta\mathbf{y}_{t-i}
+  + \mathbf{c} + \boldsymbol{\epsilon}_t\,,
+\end{equation}
+\noindent
+which is a VAR model but for time series differences.
+
+Transposing each equation of the system (\ref{eq:vecfull}) we can write
+the VECM($p$) model in block-matrix form as:
+\begin{equation}\label{eq:vareq}
+\mathbf{B} = 
+\mathbf{A} \mathbf{X} + 
+\mathbf{E} \, , 
+\end{equation}
+%
+\noindent where $\mathbf{B}$ dimension is $((N-p)\times l)$, $\mathbf{A}$
+dimension is $((N-p)\times(r+(p-1)l +1))$, $\mathbf{X}$ dimension is $((r+(p-1)l
++1)\times l)$ and $\mathbf{E}$ dimension is $((N-p)\times l)$:
+%
+\begin{alignat}{3}
+\mathbf{B}
+&= \begin{bmatrix}
+   \Delta\mathbf{y}_{p+1}^\top \\
+   \Delta\mathbf{y}_{p+2}^\top \\
+   \vdots \\
+   \Delta\mathbf{y}_N^\top
+   \end{bmatrix}
+&\quad
+\mathbf{X}
+&= \begin{bmatrix}
+   \boldsymbol{\alpha}^\top \\
+   \boldsymbol{\Phi}_1^{*\top} \\
+   \boldsymbol{\Phi}_2^{*\top} \\
+   \vdots \\
+   \boldsymbol{\Phi}_{p-1}^{*\top} \\
+   \mathbf{c}^\top
+   \end{bmatrix}
+&\quad
+\mathbf{E}
+&= \begin{bmatrix}
+   \boldsymbol{\epsilon}_{p+1}^\top \\
+   \boldsymbol{\epsilon}_{p+2}^\top \\
+   \vdots \\
+   \boldsymbol{\epsilon}_N^\top \\
+   \end{bmatrix}
+\end{alignat}
+\noindent and 
+\begin{align}
+\mathbf{A} 
+&= \begin{pmat}[{....|}]
+   \mathbf{y}_p^\top \boldsymbol{\beta} & \Delta \mathbf{y}_p^\top & \Delta\mathbf{y}_{p-1}^\top & \dots 
+                    & \Delta\mathbf{y}_2^\top & 1 \cr
+   \mathbf{y}_{p+1}^\top  \boldsymbol{\beta} &\Delta\mathbf{y}_{p+1}^\top & \Delta\mathbf{y}_p^\top & \dots
+                       & \Delta\mathbf{y}_3^\top & 1 \cr
+   \vdots & \vdots & \vdots & \ddots & \vdots & \vdots \cr
+   \mathbf{y}_{N-1}^\top  \boldsymbol{\beta} &\Delta\mathbf{y}_{N-1}^\top & \Delta\mathbf{y}_{N-2}^\top & \dots 
+                       & \Delta\mathbf{y}_{N-p-1}^\top & 1 \cr
+   \end{pmat}\, .
+\label{eq:Amatrix}
+\end{align}
+Taking into account the error term $\mathbf{E}$, equation~(\ref{eq:vareq}) 
+can be solved with respect to $\mathbf{X}$ using the OLS estimation.
 
@@
   \centering
-  \includegraphics[width=0.9\textwidth]{img/Fig1}
+   \includegraphics[width=0.9\textwidth]{img/51_Fig1}
   \caption{Distribution of the number of cointegration vectors using $p=1$ lags.
@@
   \centering
-  \includegraphics[width=0.9\textwidth]{img/Fig2}
+   \includegraphics[width=0.9\textwidth]{img/51_Fig2}
   \caption{MSE versus the percentage of cointegration considering 1000
-  iterations. Best windows size $L$ found was 1060. Below MSE versus $L$ shows a
-  rapidly decreasing behaviour, founding minimum at PC = 80\% }
+  iterations. Optimum windows size $L$ found was 1060. Below MSE versus $L$ shows a
+  rapidly decreasing behaviour, founding minimum at PC = 73\% }
   \label{fig:cointvsmse}
@@
   \centering
-  \includegraphics[width=0.9\textwidth]{img/Fig3}
+  \includegraphics[width=0.9\textwidth]{img/51_Fig3}
   \caption{Computing time of sequential and parallel algorithm is shown in the

### 2017-05-29 src/5_1_Proposal.tex
check formulas references before finishing

--- 2016-07-06 src/5_1_Proposal.tex
@@
 maximises the number of cointegration relationships in the near past. To ensure
-the search can be executed fast enough, we used a distributed environment.
+the search can be executed fast enough, we used a distributed environment 
+\cite{Arce2017}.
 
@@
 \section{Introduction}
-\label{sec:introduction}
 In finance, it is common to find variables with long-run equilibrium
@@
 
-\section{Methodology} \label{sec:methodology}
-\subsection{Motivation} \label{sec:proposal}
+The AVECM algorithm is presented in section \ref{sec:51methodology}.
+In section \ref{sec:51results} we describe the tests carried on to assess the
+accuracy and the execution time of AVECM.  This section also includes a
+description of the test data.  Section \ref{sec:51conclusions} contains the
+conclusions and a discussion of future research.
+
+%\section{Background}
+%\label{sec:background}
+%
+%\subsection{Integration and Cointegration}\label{sec:coint}\  
+%Following \cite{johansen1995} we shall say that a stochastic process
+%$Y_t$ which satisfies $Y_t-E(Y_t) = \sum_{i=0}^\infty C_i\,\varepsilon_{t-i}$ is
+%called $I(0)$, and then we shall write $Y_t\sim I(0)$, whenever
+%$\sum_{i=0}^\infty C_i \neq 0$ and $\sum_{i=0}^\infty C_i\,z^i$ converges for
+%$z\in\mathbb{C}$ with $|z|<1$.  It is understood that the condition
+%$\varepsilon_t\sim iid(0,\sigma^2)$ holds.
+%
+%A (vector) time series $\mathbf{y}_t$ is said to be {\em integrated of order\/}
+%$d$, and then we shall write $\mathbf{y}_t\sim I(d)$, whenever after $d$ times
+%(discrete) differentiation an stationary process is
+%obtained, see \cite{banerjee1993}; more precisely, whenever
+%$(1-L)^d\,\mathbf{y}_t\sim\text{I(0)}$, where $L$ is the usual lag operator:
+%$(1-L)\,\mathbf{y}_t = \Delta\mathbf{y}_t = \mathbf{y}_t-\mathbf{y}_{t-1}$ for
+%all $t$.  
+%
+%Note that this definition includes the scalar case as time series of vectors of
+%dimension 1; in this scalar case we will write the time series in nonbold
+%format.
+%
+%Let $\mathbf{y}_t^\nu$, $\nu=1,\dots,l$, be a set of $l$ vector time series of
+%order $I(1)$.  They are said to be {\em cointegrated\/} if a vector
+%$\beta=[\beta(1),\dots,\beta(l)]^\top \in \mathbb{R}^p$ exists, such that the
+%time series,
+%\begin{equation}
+%\mathbf{Z}_t:= 
+%\sum_{\nu=1}^l \beta(\nu)\,\mathbf{y}_t^\nu\,\sim\,\text{I(0)}\,.
+%\end{equation}
+%In other words, a set of $I(1)$ time series is said to be cointegrated if a
+%linear combination of them exists, which is I(0).
+%
+%
+%\subsection{Vector Autorregresive Models}\label{sec:varvec}
+%
+%Vector error correction model (VECM) describe the joint behaviour of a set of
+%variables and can be derived from the simple Vector Autoregressive model (VAR)
+%presented by \cite{sims1980}.  The VAR($p$) model is a framework describing the
+%behaviour of a set of $l$ endogenous and stationary variables as a linear
+%combination of their last $p$ values, where $l,p\in\mathbb{N}$.  In our case,
+%each one of these $l$ variables is a scalar time series $y_{\lambda,t}$,
+%$\lambda=1,\dots,l$, and we represent them all together at time $t$ by the
+%vector time series:
+%\begin{equation}
+%\label{eq:variables}
+%\mathbf{y}_t = 
+%\begin{bmatrix} y_{1,t} & y_{2,t} & \dots & y_{l,t} \end{bmatrix}^\top.
+%\end{equation}
+%\noindent
+%Notice that the vectors $\mathbf{y}_t$ are assumed to be $l$-dimensional.
+%
+%The VAR($p$) model describes the behaviour of a dependent variable in terms of
+%its own lagged values and the lags of the others variables in the system. The
+%model with $p$ lags is formulated as the system:
+%\begin{align}
+%\label{eq:var}
+%\mathbf{y}_t 
+%= \boldsymbol{\Phi}_1 \mathbf{y}_{t-1} +
+%  \boldsymbol{\Phi}_2 \mathbf{y}_{t-2} + \dots +
+%  \boldsymbol{\Phi}_p\mathbf{y}_{t-p} +
+%  \mathbf{c} + \boldsymbol{\epsilon}_t \nonumber \\
+%t=p+1,\dots,N,
+%\end{align}
+%\noindent where 
+%$\boldsymbol{\Phi}_1, \boldsymbol{\Phi}_2,\dots,\boldsymbol{\Phi}_p$
+%are $l\times l$-matrices of real coefficients,
+%$\boldsymbol{\epsilon}_{p+1},
+% \boldsymbol{\epsilon}_{p+2}, \dots, \boldsymbol{\epsilon}_N$ 
+%are error terms, $\mathbf{c}$ is a constant vector and $N$ is the total
+%number of samples.
+%
+%Notice that, regarding our notation of section (\ref{sec:coint}),
+%we have here 
+%$\mathbf{y}_t^0 = \mathbf{y}_t$,
+%$\mathbf{y}_t^\nu = \mathbf{y}_{t-\nu}$ and
+%the $\lambda$-th component of the vector time series $\mathbf{y}_t^\nu$
+%is the scalar time series $y_{\lambda,t}^\nu$, where $\nu=1,\dots,p$ and
+%$\lambda=1,\dots,l$.
+%
+%However, the VAR model cannot be used with non-stationary variables. VECM ,
+%developed by \cite{engle87}, is also a linear model for I(1) variables that are
+%also cointegrated, see \cite{banerjee1993}. If cointegration exists, variable
+%differences are stationary and they introduce an error correction term which
+%adjusts coefficients to bring the variables back to equilibrium. 
+%
+%
+%It is obtained re-writing equation (\ref{eq:var}) in terms of the new
+%variable $\Delta\mathbf{y}_t=\mathbf{y}_t-\mathbf{y}_{t-1}$.
+%The VECM model, expressed in terms those differences, takes the form:
+%\begin{equation}\label{eq:vec}
+%\Delta \mathbf{y}_t 
+%= \boldsymbol{\Omega}\,\mathbf{y}_{t-1}
+%  + \sum_{i=1}^{p-1} \boldsymbol{\Phi}_i^*\,\Delta\mathbf{y}_{t-i}
+%  + \mathbf{c} + \boldsymbol{\epsilon}_t\,,
+%\end{equation}
+%\noindent
+%where the coefficients matrices $\boldsymbol{\Phi}_i^*$ and 
+%$\boldsymbol{\Omega}$, expressed in terms of the matrices
+%$\boldsymbol{\Phi}_i$ of (\ref{eq:var}), are:
+%\begin{align*}
+%\boldsymbol{\Phi}_i^* 
+%&:= -\sum_{j=i+1}^{p}\boldsymbol{\Phi}_j\,, \\
+%\boldsymbol{\Omega}
+%&:= -\left( \mathbb{I} - \boldsymbol{\Phi}_1 - \dots 
+%    - \boldsymbol{\phi}_p \right)\,. 
+%\end{align*}
+%The following well known properties of the matrix $\boldsymbol{\Omega}$, see
+%\cite{johansen1995}, will be useful in the sequel:
+%\begin{itemize}
+%\item
+%If $\boldsymbol{\Omega} = \mathbf{0}$, there is no cointegration.
+%\item 
+%If $rank(\boldsymbol{\Omega})=l$, i.e., if $\boldsymbol{\Omega}$ has
+%full rank, then the time series are not I(1) but stationary.
+%\item
+%If $rank(\boldsymbol{\Omega})=r$, $0<r<l$, then there is cointegration
+%and the matrix $\boldsymbol{\Omega}$ can be expressed as
+%$\boldsymbol{\Omega}=\boldsymbol{\alpha\beta}^\top$, where $\boldsymbol{\alpha}$
+%and $\boldsymbol{\beta}$ are
+%$l\times r$ matrices and
+%$\text{rank}(\boldsymbol{\alpha})=\text{rank}(\boldsymbol{\beta})=r$.
+%\item
+%The columns of $\boldsymbol{\beta}$ contains the cointegration vectors and the rows of
+%$\boldsymbol{\alpha}$ correspond with the adjusted vectors. 
+%$\boldsymbol{\beta}$ is obtained by Johansen procedure,see \cite{johansen1988},
+%whereas $\boldsymbol{\alpha}$ has to be determined as a variable in the VECM.
+%\end{itemize}
+%It is worth noticing that the factorization of the matrix
+%$\boldsymbol\Omega$ is not unique, since for any $r \times r$
+%non-singular matrix $\mathbf{H}$, $\boldsymbol{\alpha}^*:=\boldsymbol{\alpha}\mathbf{H}$,
+%and $\boldsymbol{\beta}^*=\boldsymbol{\beta}(\mathbf{H}^{-1})^\top$ we have
+%$\boldsymbol{\alpha\beta}^\top=\boldsymbol{\alpha}^*(\boldsymbol{\beta}^*)^\top$.
+%If cointegration exists, then equation (\ref{eq:vec}) can be written
+%as follows:
+%\begin{equation}\label{eq:vecfull}
+%\Delta\mathbf{y}_t 
+%= \boldsymbol{\alpha\beta}^\top\mathbf{y}_{t-1} 
+%  + \sum_{i=1}^{p-1}\boldsymbol{\Phi}_i^*\,\Delta\mathbf{y}_{t-i}
+%  + \mathbf{c} + \boldsymbol{\epsilon}_t\,,
+%\end{equation}
+%\noindent
+%which is a VAR model but for time series differences.
+%
+%
+%Transposing each equation of the system (\ref{eq:vecfull}) we can write
+%the VECM($p$) model in block-matrix form as:
+%\begin{equation}\label{eq:vareq}
+%\mathbf{B} = 
+%\mathbf{A} \mathbf{X} + 
+%\mathbf{E} \, , 
+%\end{equation}
+%%
+%\noindent where $\mathbf{B}$ dimension is $((N-p)\times l)$, $\mathbf{A}$
+%dimension is $((N-p)\times(r+(p-1)l +1))$, $\mathbf{X}$ dimension is $((r+(p-1)l
+%+1)\times l)$ and $\mathbf{E}$ dimension is $((N-p)\times l)$:
+%%
+%\begin{alignat}{3}
+%\mathbf{B}
+%&= \begin{bmatrix}
+%   \Delta\mathbf{y}_{p+1}^\top \\
+%   \Delta\mathbf{y}_{p+2}^\top \\
+%   \vdots \\
+%   \Delta\mathbf{y}_N^\top
+%   \end{bmatrix}
+%&\quad
+%\mathbf{X}
+%&= \begin{bmatrix}
+%   \boldsymbol{\alpha}^\top \\
+%   \boldsymbol{\Phi}_1^{*\top} \\
+%   \boldsymbol{\Phi}_2^{*\top} \\
+%   \vdots \\
+%   \boldsymbol{\Phi}_{p-1}^{*\top} \\
+%   \mathbf{c}^\top
+%   \end{bmatrix}
+%&\quad
+%\mathbf{E}
+%&= \begin{bmatrix}
+%   \boldsymbol{\epsilon}_{p+1}^\top \\
+%   \boldsymbol{\epsilon}_{p+2}^\top \\
+%   \vdots \\
+%   \boldsymbol{\epsilon}_N^\top \\
+%   \end{bmatrix}
+%\end{alignat}
+%\noindent and 
+%\begin{align}
+%\mathbf{A} 
+%&= \begin{pmat}[{....|}]
+%   \mathbf{y}_p^\top \boldsymbol{\beta} & \Delta \mathbf{y}_p^\top & \Delta\mathbf{y}_{p-1}^\top & \dots 
+%                    & \Delta\mathbf{y}_2^\top & 1 \cr
+%   \mathbf{y}_{p+1}^\top  \boldsymbol{\beta} &\Delta\mathbf{y}_{p+1}^\top & \Delta\mathbf{y}_p^\top & \dots
+%                       & \Delta\mathbf{y}_3^\top & 1 \cr
+%   \vdots & \vdots & \vdots & \ddots & \vdots & \vdots \cr
+%   \mathbf{y}_{N-1}^\top  \boldsymbol{\beta} &\Delta\mathbf{y}_{N-1}^\top & \Delta\mathbf{y}_{N-2}^\top & \dots 
+%                       & \Delta\mathbf{y}_{N-p-1}^\top & 1 \cr
+%   \end{pmat}\, .
+%\label{eq:Amatrix}
+%\end{align}
+%Taking into account the error term $\mathbf{E}$, equation~(\ref{eq:vareq}) 
+%can be solved with respect to $\mathbf{X}$ using the OLS estimation.
+
+\section{Methodology} \label{sec:51methodology}
+\subsection{Motivation}
 
@@
 
-From section \ref{sec:background} we know that $r=0$ means no cointegration and
+Figure \ref{fig:hists} shows the distribution of the number of cointegration
+vectors given by the Johansen method for different values of $L=[2,6,10,14]$
+hours and $p=1$ . This procedure was carried out by a sliding window of
+historical data moving 1000 times. We observed that the distribution of
+cointegration vectors changed with different values of $L$. When
+$L=2$ hours, there was no cointegration in more than 60\% of the iterations.
+Cointegration increased when $L=6$ hours and was maximum when $L=10$ hours where
+one cointegration vectors was found for all 1000 iterations. The ocurrence of
+cointegration started to decrease at $L=14$ hours.
+
+From section \ref{sec:vecm} we know that $r=0$ means no cointegration and
 $r=l$ (we are using four rates, so $l=4$) reveals that no process is I(1) but
@@
 
-\subsection{Model comparison} \label{sec:random}
+\subsection{Model comparison}
 We compared our proposal, in terms of performance, with the naive forecast of
@@
 
-\subsection{Evaluation methods} \label{sec:evaluation}
+\subsection{Evaluation methods}
 
@@
 \begin{description}
-\item
-{\bf MSE},  Mean Square Error measures the distance between forecasts
+\item[MSE]  Mean Square Error measures the distance between forecasts
 and the true values and large deviations from the true value have a
@@
 \end{equation}
-\item {\bf $U$-statistic}, the Theil's $U$-statistic, presented by
+\item[$U$-statistic] the Theil's $U$-statistic, presented by
 \cite{theil1966}, is a unit free measure obtained as the ratio between the root
@@
 \section{Experimental results}
-\label{sec:results}
-\subsection{Data} \label{sec:unitroot}
-AVECM tests were carried out using four foreign exchange rates all related to
-USD: EURUSD, GBPUSD, USDCHF and USDJPY. This data was collected from
+\label{sec:51results}
+\subsection{Data}
+All the experiments and AVECM tests were carried out using four foreign exchange rates all related to
+USD: EURUSD, GBPUSD, USDCHF and USDJPY. We chose the most traded rates related
+to USD so they were likely to be cointegrated. 
+This data was collected from
 \cite{Dukascopy2014}, a free database which gives access to the Swiss Foreign
@@
 
-\subsection{Unit root tests} \label{sec:unitroottests}
+\subsection{Unit root tests}
 
@@
 
-\subsection{Performance accuracy} \label{sec:performance}
+\subsection{Performance accuracy}
 
@@
 
-\subsection{Parallel implementation} \label{sec:parallel}
+\subsection{Parallel implementation}
 
@@
 \section{Conclusions}
-\label{sec:conclusions}
+\label{sec:51conclusions}
 Cointegration in financial time series has been largely studied and the Johansen

### 2017-06-07 src/3_TimeSeries.tex
Chapter 4: Springer paper finished

@@
 \section{Vector error correction model}\label{sec:vecm}
-Vector error correction model (VECM)  \cite{engle87} describe the joint behaviour of a set of cointegrated I(1) time series \cite{banerjee1993} and can be derived from the simple VAR model.
+The VAR model cannot be used with non-stationary variables. VECM ,
+developed by \cite{engle87}, is also a linear model for I(1) variables that are
+also cointegrated, see \cite{banerjee1993}. If cointegration exists, variable
+differences are stationary and they introduce an error correction term which
+adjusts coefficients to bring the variables back to equilibrium.
 
@@
 the VECM($p$) model in block-matrix form as:
-\begin{equation}\label{eq:simpleveceq}
-\mathbf{Y} = 
+\begin{equation}\label{eq:simpleveceq} %vareq}
+\mathbf{B} = 
 \mathbf{A} \mathbf{X} + 
@@
 \label{eq:vecY}
-\mathbf{Y}
+\mathbf{B}
 = \begin{bmatrix}

### 2017-06-07 src/4_OnlineLearning.tex
Chapter 4: Springer paper finished

@@
 \end{equation}
+\item[MSE]  Mean Square Error measures the distance between forecasts
+and the true values and large deviations from the true value have a
+large impact due to squaring forecast error.
+\begin{equation}\label{eq:MSE}
+\text{MSE} = 
+\frac{\displaystyle \sum_{t=1}^{N} (\mathbf{y}_t-\hat{\mathbf{y}}_t)^2}{N}
+\end{equation}
 \item[RMSE] Root Mean Square Error also measures the distance between forecasts
@@
 \end{equation}
+\item[$U$-statistic] the Theil's $U$-statistic, presented by
+\cite{theil1966}, is a unit free measure obtained as the ratio between the root
+MSE (RMSE) of a model and the RMSE of the naive random walk model. A
+$U$-statistic less than 1 implies the performance is better than the naive
+model.
 \end{description}

### 2017-06-07 src/5_1_Proposal.tex
my citation
Chapter 4: Springer paper finished

@@
-\chapter{Fast and adaptive cointegration based model
+\chapter[Adaptive cointegration model for time series forecasting]{Fast and adaptive cointegration based model
 for forecasting high frequency financial time series}
@@
 maximises the number of cointegration relationships in the near past. To ensure
-the search can be executed fast enough, we used a distributed environment 
-\cite{Arce2017}.
+the search can be executed fast enough, we used a distributed environment .
 
@@
 significantly improved execution times with respect to its serial version.
+
+This work is published in \cite{Arce2017}.
 
@@
 
-%\section{Background}
-%\label{sec:background}
-%
-%\subsection{Integration and Cointegration}\label{sec:coint}\  
-%Following \cite{johansen1995} we shall say that a stochastic process
-%$Y_t$ which satisfies $Y_t-E(Y_t) = \sum_{i=0}^\infty C_i\,\varepsilon_{t-i}$ is
-%called $I(0)$, and then we shall write $Y_t\sim I(0)$, whenever
-%$\sum_{i=0}^\infty C_i \neq 0$ and $\sum_{i=0}^\infty C_i\,z^i$ converges for
-%$z\in\mathbb{C}$ with $|z|<1$.  It is understood that the condition
-%$\varepsilon_t\sim iid(0,\sigma^2)$ holds.
-%
-%A (vector) time series $\mathbf{y}_t$ is said to be {\em integrated of order\/}
-%$d$, and then we shall write $\mathbf{y}_t\sim I(d)$, whenever after $d$ times
-%(discrete) differentiation an stationary process is
-%obtained, see \cite{banerjee1993}; more precisely, whenever
-%$(1-L)^d\,\mathbf{y}_t\sim\text{I(0)}$, where $L$ is the usual lag operator:
-%$(1-L)\,\mathbf{y}_t = \Delta\mathbf{y}_t = \mathbf{y}_t-\mathbf{y}_{t-1}$ for
-%all $t$.  
-%
-%Note that this definition includes the scalar case as time series of vectors of
-%dimension 1; in this scalar case we will write the time series in nonbold
-%format.
-%
-%Let $\mathbf{y}_t^\nu$, $\nu=1,\dots,l$, be a set of $l$ vector time series of
-%order $I(1)$.  They are said to be {\em cointegrated\/} if a vector
-%$\beta=[\beta(1),\dots,\beta(l)]^\top \in \mathbb{R}^p$ exists, such that the
-%time series,
-%\begin{equation}
-%\mathbf{Z}_t:= 
-%\sum_{\nu=1}^l \beta(\nu)\,\mathbf{y}_t^\nu\,\sim\,\text{I(0)}\,.
-%\end{equation}
-%In other words, a set of $I(1)$ time series is said to be cointegrated if a
-%linear combination of them exists, which is I(0).
-%
-%
-%\subsection{Vector Autorregresive Models}\label{sec:varvec}
-%
-%Vector error correction model (VECM) describe the joint behaviour of a set of
-%variables and can be derived from the simple Vector Autoregressive model (VAR)
-%presented by \cite{sims1980}.  The VAR($p$) model is a framework describing the
-%behaviour of a set of $l$ endogenous and stationary variables as a linear
-%combination of their last $p$ values, where $l,p\in\mathbb{N}$.  In our case,
-%each one of these $l$ variables is a scalar time series $y_{\lambda,t}$,
-%$\lambda=1,\dots,l$, and we represent them all together at time $t$ by the
-%vector time series:
-%\begin{equation}
-%\label{eq:variables}
-%\mathbf{y}_t = 
-%\begin{bmatrix} y_{1,t} & y_{2,t} & \dots & y_{l,t} \end{bmatrix}^\top.
-%\end{equation}
-%\noindent
-%Notice that the vectors $\mathbf{y}_t$ are assumed to be $l$-dimensional.
-%
-%The VAR($p$) model describes the behaviour of a dependent variable in terms of
-%its own lagged values and the lags of the others variables in the system. The
-%model with $p$ lags is formulated as the system:
-%\begin{align}
-%\label{eq:var}
-%\mathbf{y}_t 
-%= \boldsymbol{\Phi}_1 \mathbf{y}_{t-1} +
-%  \boldsymbol{\Phi}_2 \mathbf{y}_{t-2} + \dots +
-%  \boldsymbol{\Phi}_p\mathbf{y}_{t-p} +
-%  \mathbf{c} + \boldsymbol{\epsilon}_t \nonumber \\
-%t=p+1,\dots,N,
-%\end{align}
-%\noindent where 
-%$\boldsymbol{\Phi}_1, \boldsymbol{\Phi}_2,\dots,\boldsymbol{\Phi}_p$
-%are $l\times l$-matrices of real coefficients,
-%$\boldsymbol{\epsilon}_{p+1},
-% \boldsymbol{\epsilon}_{p+2}, \dots, \boldsymbol{\epsilon}_N$ 
-%are error terms, $\mathbf{c}$ is a constant vector and $N$ is the total
-%number of samples.
-%
-%Notice that, regarding our notation of section (\ref{sec:coint}),
-%we have here 
-%$\mathbf{y}_t^0 = \mathbf{y}_t$,
-%$\mathbf{y}_t^\nu = \mathbf{y}_{t-\nu}$ and
-%the $\lambda$-th component of the vector time series $\mathbf{y}_t^\nu$
-%is the scalar time series $y_{\lambda,t}^\nu$, where $\nu=1,\dots,p$ and
-%$\lambda=1,\dots,l$.
-%
-%However, the VAR model cannot be used with non-stationary variables. VECM ,
-%developed by \cite{engle87}, is also a linear model for I(1) variables that are
-%also cointegrated, see \cite{banerjee1993}. If cointegration exists, variable
-%differences are stationary and they introduce an error correction term which
-%adjusts coefficients to bring the variables back to equilibrium. 
-%
-%
-%It is obtained re-writing equation (\ref{eq:var}) in terms of the new
-%variable $\Delta\mathbf{y}_t=\mathbf{y}_t-\mathbf{y}_{t-1}$.
-%The VECM model, expressed in terms those differences, takes the form:
-%\begin{equation}\label{eq:vec}
-%\Delta \mathbf{y}_t 
-%= \boldsymbol{\Omega}\,\mathbf{y}_{t-1}
-%  + \sum_{i=1}^{p-1} \boldsymbol{\Phi}_i^*\,\Delta\mathbf{y}_{t-i}
-%  + \mathbf{c} + \boldsymbol{\epsilon}_t\,,
-%\end{equation}
-%\noindent
-%where the coefficients matrices $\boldsymbol{\Phi}_i^*$ and 
-%$\boldsymbol{\Omega}$, expressed in terms of the matrices
-%$\boldsymbol{\Phi}_i$ of (\ref{eq:var}), are:
-%\begin{align*}
-%\boldsymbol{\Phi}_i^* 
-%&:= -\sum_{j=i+1}^{p}\boldsymbol{\Phi}_j\,, \\
-%\boldsymbol{\Omega}
-%&:= -\left( \mathbb{I} - \boldsymbol{\Phi}_1 - \dots 
-%    - \boldsymbol{\phi}_p \right)\,. 
-%\end{align*}
-%The following well known properties of the matrix $\boldsymbol{\Omega}$, see
-%\cite{johansen1995}, will be useful in the sequel:
-%\begin{itemize}
-%\item
-%If $\boldsymbol{\Omega} = \mathbf{0}$, there is no cointegration.
-%\item 
-%If $rank(\boldsymbol{\Omega})=l$, i.e., if $\boldsymbol{\Omega}$ has
-%full rank, then the time series are not I(1) but stationary.
-%\item
-%If $rank(\boldsymbol{\Omega})=r$, $0<r<l$, then there is cointegration
-%and the matrix $\boldsymbol{\Omega}$ can be expressed as
-%$\boldsymbol{\Omega}=\boldsymbol{\alpha\beta}^\top$, where $\boldsymbol{\alpha}$
-%and $\boldsymbol{\beta}$ are
-%$l\times r$ matrices and
-%$\text{rank}(\boldsymbol{\alpha})=\text{rank}(\boldsymbol{\beta})=r$.
-%\item
-%The columns of $\boldsymbol{\beta}$ contains the cointegration vectors and the rows of
-%$\boldsymbol{\alpha}$ correspond with the adjusted vectors. 
-%$\boldsymbol{\beta}$ is obtained by Johansen procedure,see \cite{johansen1988},
-%whereas $\boldsymbol{\alpha}$ has to be determined as a variable in the VECM.
-%\end{itemize}
-%It is worth noticing that the factorization of the matrix
-%$\boldsymbol\Omega$ is not unique, since for any $r \times r$
-%non-singular matrix $\mathbf{H}$, $\boldsymbol{\alpha}^*:=\boldsymbol{\alpha}\mathbf{H}$,
-%and $\boldsymbol{\beta}^*=\boldsymbol{\beta}(\mathbf{H}^{-1})^\top$ we have
-%$\boldsymbol{\alpha\beta}^\top=\boldsymbol{\alpha}^*(\boldsymbol{\beta}^*)^\top$.
-%If cointegration exists, then equation (\ref{eq:vec}) can be written
-%as follows:
-%\begin{equation}\label{eq:vecfull}
-%\Delta\mathbf{y}_t 
-%= \boldsymbol{\alpha\beta}^\top\mathbf{y}_{t-1} 
-%  + \sum_{i=1}^{p-1}\boldsymbol{\Phi}_i^*\,\Delta\mathbf{y}_{t-i}
-%  + \mathbf{c} + \boldsymbol{\epsilon}_t\,,
-%\end{equation}
-%\noindent
-%which is a VAR model but for time series differences.
-%
-%
-%Transposing each equation of the system (\ref{eq:vecfull}) we can write
-%the VECM($p$) model in block-matrix form as:
-%\begin{equation}\label{eq:vareq}
-%\mathbf{B} = 
-%\mathbf{A} \mathbf{X} + 
-%\mathbf{E} \, , 
-%\end{equation}
-%%
-%\noindent where $\mathbf{B}$ dimension is $((N-p)\times l)$, $\mathbf{A}$
-%dimension is $((N-p)\times(r+(p-1)l +1))$, $\mathbf{X}$ dimension is $((r+(p-1)l
-%+1)\times l)$ and $\mathbf{E}$ dimension is $((N-p)\times l)$:
-%%
-%\begin{alignat}{3}
-%\mathbf{B}
-%&= \begin{bmatrix}
-%   \Delta\mathbf{y}_{p+1}^\top \\
-%   \Delta\mathbf{y}_{p+2}^\top \\
-%   \vdots \\
-%   \Delta\mathbf{y}_N^\top
-%   \end{bmatrix}
-%&\quad
-%\mathbf{X}
-%&= \begin{bmatrix}
-%   \boldsymbol{\alpha}^\top \\
-%   \boldsymbol{\Phi}_1^{*\top} \\
-%   \boldsymbol{\Phi}_2^{*\top} \\
-%   \vdots \\
-%   \boldsymbol{\Phi}_{p-1}^{*\top} \\
-%   \mathbf{c}^\top
-%   \end{bmatrix}
-%&\quad
-%\mathbf{E}
-%&= \begin{bmatrix}
-%   \boldsymbol{\epsilon}_{p+1}^\top \\
-%   \boldsymbol{\epsilon}_{p+2}^\top \\
-%   \vdots \\
-%   \boldsymbol{\epsilon}_N^\top \\
-%   \end{bmatrix}
-%\end{alignat}
-%\noindent and 
-%\begin{align}
-%\mathbf{A} 
-%&= \begin{pmat}[{....|}]
-%   \mathbf{y}_p^\top \boldsymbol{\beta} & \Delta \mathbf{y}_p^\top & \Delta\mathbf{y}_{p-1}^\top & \dots 
-%                    & \Delta\mathbf{y}_2^\top & 1 \cr
-%   \mathbf{y}_{p+1}^\top  \boldsymbol{\beta} &\Delta\mathbf{y}_{p+1}^\top & \Delta\mathbf{y}_p^\top & \dots
-%                       & \Delta\mathbf{y}_3^\top & 1 \cr
-%   \vdots & \vdots & \vdots & \ddots & \vdots & \vdots \cr
-%   \mathbf{y}_{N-1}^\top  \boldsymbol{\beta} &\Delta\mathbf{y}_{N-1}^\top & \Delta\mathbf{y}_{N-2}^\top & \dots 
-%                       & \Delta\mathbf{y}_{N-p-1}^\top & 1 \cr
-%   \end{pmat}\, .
-%\label{eq:Amatrix}
-%\end{align}
-%Taking into account the error term $\mathbf{E}$, equation~(\ref{eq:vareq}) 
-%can be solved with respect to $\mathbf{X}$ using the OLS estimation.
-
 \section{Methodology} \label{sec:51methodology}
-\subsection{Motivation}
 
@@
 
-\subsection{Evaluation methods}
-
-Forecast performance was evaluated using two different methods which are
-frequently used in finance:
-\begin{description}
-\item[MSE]  Mean Square Error measures the distance between forecasts
-and the true values and large deviations from the true value have a
-large impact due to squaring forecast error.
-\begin{equation}\label{eq:MSE}
-\text{MSE} = 
-\frac{\displaystyle \sum_{t=1}^{N} (\mathbf{y}_t-\hat{\mathbf{y}}_t)^2}{N}
-\end{equation}
-\item[$U$-statistic] the Theil's $U$-statistic, presented by
-\cite{theil1966}, is a unit free measure obtained as the ratio between the root
-MSE (RMSE) of a model and the RMSE of the naive random walk model. A
-$U$-statistic less than 1 implies the performance is better than the naive
-model.
-\end{description}
-
 \section{Experimental results}
@@
 have values in between $[1,5]$. Best AVECM performance was compared against
-ARIMA and the random walk model.  Table~\ref{tab:stats} shows the out-of-sample
+ARIMA and the random walk model.  Table~\ref{tab:51stats} shows the out-of-sample
 performance measures: MSE and $U$-statistic for AVECM and ARIMA. In terms of
@@
 \caption{AVECM performance}
-\label{tab:stats}
+\label{tab:51stats}
 \begin{center}
@@
 routine. This procedure is done by the function
-get\_best\_params($Ls,ps,m,\mathbf{Y}$) shown in algorithm \ref{alg:AVECM}.  In
+get\_best\_params shown in algorithm \ref{alg:AVECM}.  In
 order to improve the execution time of this search, our proposal included a
@@
 Execution time depends directly on $L$, $p$ and $nparams$ used, since they
-determine the size of matrix $\mathbf{A}$ (see equation \ref{eq:Amatrix}) and
+determine the size of matrix $\mathbf{A}$ (see equation \ref{eq:vecA}) and
 therefore affect the OLS function execution time.  Therefore, if we try more

### 2017-06-07 src/5_2_Proposal.tex
change title 5-2 and date
chapter 5 src/5_2 Icpram 2015 ready

@@
-\chapter{Online VECM proposal}
-Financial time series are known for their non-stationary behaviour. However,
-sometimes they exhibit some stationary linear combinations. When this happens,
-it is said that those time series are cointegrated. The Vector Error Correction
-Model (VECM) is an econometric model which characterises the joint dynamic
-behaviour of a set of cointegrated variables in terms of forces pulling towards
-equilibrium. In this study, we propose an Online VEC model (OVECM) which
-optimises how model parameters are obtained using a sliding window of the most
-recent data. Our proposal also takes advantage of the long-run relationship
-between the time series in order to obtain improved execution times. Our
-proposed method is tested using four foreign exchange rates with a frequency of
-1-minute, all related to the USD currency base. OVECM is compared with VECM and
-ARIMA models in terms of forecasting accuracy and execution times. We show that
-OVECM outperforms ARIMA forecasting and enables execution time to
-be reduced considerably while maintaining good accuracy levels compared with
-VECM.
+\chapter{An Online Vector Error Correction Model for Exchange Rates Forecasting}
+
+The Vector Error Correction Model (VECM) is an econometric model which
+characterises the joint dynamic behaviour of a set of cointegrated variables in
+terms of forces pulling towards equilibrium. In the previous proposal, the
+Adaptive VECM (AVECM) showed that to be able to use VECM at every time step can
+be used for forecasting purposes. However, to avoid extensive calculations, a
+parallel approach was implemented.
+
+In this study, an Online VEC model (OVECM) is proposed, which optimises how
+model parameters are obtained using a sliding window of the most recent data.
+OVECM, unlile AVECM, updates the parameters at each step, instead of obtaining
+new ones. This proposal also takes advantage of the long-run relationship
+between the time series in order to obtain improved execution times.
+
+Our proposed method is tested using four foreign exchange rates with a frequency
+of 1-minute, all related to the USD currency base.
+
+OVECM is compared with VECM and ARIMA models in terms of forecasting accuracy
+and execution times. We show that OVECM outperforms ARIMA forecasting and
+enables execution time to be reduced considerably while maintaining good
+accuracy levels compared with VECM.
+
+This work is published in \cite{icpram15}.
 
@@
 
+\newpage
 \section{The problem}
-In finance, it is common to find variables with a long-run and/or a short-run
-equilibrium relationship. This relationship is called cointegration and it
-reflects the idea of that some set of variables cannot wander too far away from
-each other.
 
-Cointegration means that one or more combinations of these variables is
-stationary even though individually they are not.
+Both VECM and VAR model parameters are obtained using ordinary least squares
+(OLS) method. Since OLS involves many calculations, the parameter estimation
+method is computationally expensive when the number of past values and
+observations increases. Moreover, obtaining cointegration vectors is also an
+expensive routine.
 
-Some financial models, such as  vector error correction model (VECM), take advantage of
-this property and describe the joint behaviour of several cointegrated
-financial instruments.
+Recently, online learning algorithms have been proposed to solve problems with
+large data sets because of their simplicity and their ability to update the
+model when new data is available. The study presented by ~\cite{arce+salinas2012}
+applied this idea using ridge regression.
 
-VECM is a special case of the vector autoregressive model (VAR). VAR is a linear
-model which express future values in terms of its own history and other
-variables past values. If cointegration exists, VECM allows to introduce an
-error correction term due to its cointegration estimation error.
+There are several popular online methods such as perceptron \cite{rosenblatt58},
+passive-aggressive \cite{crammerETall2006}, stochastic gradient
+descent \cite{zhang2004}, aggregating algorithm \cite{vovk2001} and the second
+order perceptron \cite{cesa-bianchi2005}.  In \cite{cesa-bianchi2006}, an
+in-deph analysis of online learning is provided.
 
-VECM as well as VAR model parameters are obtained using ordinary least squares
-method (OLS). Since OLS involves many calculations, the parameter estimation
-method is computationally expensive when the number of past values and
-observations considered increases. Moreover, OLS is an ill-posed problem which
-admits an infinite number of solutions.
+In this proposal, an online formulation of the VECM called Online VECM (OVECM)
+is presented. OVECM is a lighter version of VECM which considers only a sliding
+window of the most recent data and introduces matrix optimizations in order to
+reduce the number of operations and therefore execution times. OVECM also takes
+into account the fact that cointegration vector space doesn't experience large
+changes with small changes in the input data.
 
-Ridge regression method (RR) tackles this ill-posed problem and is usually
-formulated instead of OLS. RR includes a regularisation parameter that leads to
-better generalisation capability. However, RR is still computationally
-expensive dealing with large datasets. Recently, online learning algorithms
-have been proposed to solve problems with large data sets because of their
-simplicity and their ability of updating the model when new data is available.
+OVECM is later compared against VECM and ARIMA models using four currency rates
+from the foreign exchange market with 1-minute frequency. VECM and ARIMA models
+were used in an iterative way in order to allow fair comparison. Execution times
+and forecast performance measures MAPE, MAE and RMSE were used to compare all
+methods.
 
-We propose an online formulation of the VEC model based on the
-AAR method considering only a sliding window of the most recent data. The
-algorithm introduces matrix optimisations in order to reduce the number of
-operations. Our method is later tested with financial data from the foreign
-exchange market.
+Model effectiveness is focused on out-of-sample forecast rather than in-sample
+fitting. This criteria allows the OVECM prediction capability to be expressed
+rather than just explaining data history.
 
-\section{Methodology} \label{sec:methodology}
+The next sections are organized as follows: the OVECM algorithm proposed is
+presented in section~\ref{sec:52methodology}. Section~\ref{sec:52results} gives a
+description of the data used and the tests carried on to show accuracy and time
+comparison of our proposal against the traditional VECM and
+section~\ref{sec:52conclusions} includes conclusions and a proposal for future
+study.
+
+\section{Methodology} \label{sec:52methodology}
 
@@
 most recent data. Moreover, since cointegration vectors represent long-run
-relationships which vary little in time, OVECM determines firstly if they require calculation.
+relationships which vary little in time, OVECM determines firstly if they
+require calculation.
 
@@
 
-Algorithm~\ref{alg:proposal} shows our OVECM proposal which considers the
-following:
-
-\begin{itemize}
-\item The function \texttt{getJohansen} returns cointegration vectors given by
-the Johansen method considering the trace statistic test at 95\%
-significance level.
-\item The function \texttt{vecMatrix} returns the matrices~(\ref{eq:vecA})
-and~(\ref{eq:vecY}) that allows VECM to be solved.
-\item The function \texttt{vecMatrixOnline} returns the
-matrices~(\ref{eq:vecA}) and~(\ref{eq:vecY}) aggregating new data and removing
-the old one, avoiding calculation of the matrix $\mathbf{A}$ from scratch.
-\item Out-of-sample outputs are saved in the variables 
-$\mathbf{Y}_{\text{true}}$ and $\mathbf{Y}_{\text{pred}}$.
-\item The model is solved using OLS.
-\item In-sample outputs are saved in the variables $\Delta
-\mathbf{y}_{\text{true}}$ and $\Delta \mathbf{y}_{\text{pred}}$.
-\item The function \texttt{mape} gets the in-sample MAPE for the $l$ time
-series.
-\item Cointegration vectors are obtained at the beginning and when they are required to be updated. This updating is decided based on the in-sample MAPE of the last $n$ inputs. The average of all
-MAPEs are stored in the variable $e$. If the average of MAPEs
-($\texttt{mean}(e)$) is above a certain error given by the mean\_error threshold, cointegration vectors are updated.
-\item If new cointegration vectors are required, the function
-\texttt{vecMatrixUpdate} only updates the corresponding columns of matrix
-$\mathbf{A}$ affected by those vectors (see equation~\ref{eq:vecA}).
-\end{itemize}
-
-\begin{algorithm}[ht]
+Algorithm~\ref{alg:52proposal} shows our OVECM proposal.
+\begin{algorithm}[!htp]
 \begin{algorithmic}[1]
@@
 \caption{OVECM: Online VECM}
-\label{alg:proposal}
+\label{alg:52proposal}
 \end{algorithm}
 
-A pseudocode of the algorithm \ref{alg:proposal} is detailed in the appendix \ref{fig:pseudocodigo}.
+\begin{itemize}
+\item The function \texttt{getJohansen} returns cointegration vectors given by
+the Johansen method considering the trace statistic test at 95\%
+significance level.
+\item The function \texttt{vecMatrix} returns the matrices~(\ref{eq:vecA})
+and~(\ref{eq:vecY}) that allows VECM to be solved.
+\item The function \texttt{vecMatrixOnline} returns the
+matrices~(\ref{eq:vecA}) and~(\ref{eq:vecY}) aggregating new data and removing
+the old one, avoiding calculation of the matrix $\mathbf{A}$ from scratch.
+\item Out-of-sample outputs are saved in the variables 
+$\mathbf{Y}_{\text{true}}$ and $\mathbf{Y}_{\text{pred}}$.
+\item The model is solved using OLS.
+\item In-sample outputs are saved in the variables $\Delta
+\mathbf{y}_{\text{true}}$ and $\Delta \mathbf{y}_{\text{pred}}$.
+\item The function \texttt{mape} gets the in-sample MAPE for the $l$ time
+series.
+\item Cointegration vectors are obtained at the beginning and when they are required to be updated. This updating is decided based on the in-sample MAPE of the last $n$ inputs. The average of all
+MAPEs are stored in the variable $e$. If the average of MAPEs
+($\texttt{mean}(e)$) is above a certain error given by the mean\_error threshold, cointegration vectors are updated.
+\item If new cointegration vectors are required, the function
+\texttt{vecMatrixUpdate} only updates the corresponding columns of matrix
+$\mathbf{A}$ affected by those vectors (see equation~\ref{eq:vecA}).
+\end{itemize}
+\newpage
+
+A pseudocode of the algorithm \ref{alg:52proposal} is detailed in the figure\ref{fig:pseudocodigo}.
+\newline
 
@@
 
-\section{Results} \label{sec:results}
+\section{Results} \label{sec:52results}
 
@@
 \caption{Unit roots tests}
-\label{tab:adf}
+\label{tab:52adf}
 \begin{center}
@@
 
-Table~\ref{tab:adf} shows that all currency rates cannot reject the unit root
+Table~\ref{tab:52adf} shows that all currency rates cannot reject the unit root
 test but they rejected it with their first differences. This means that all of
@@
 ARIMA parameters were also obtained using AIC.
-Parameters optimisation is presented in table~\ref{tab:params}:
+Parameters optimisation is presented in table~\ref{tab:52params}:
 
@@
 using AIC.}
-\label{tab:params}
+\label{tab:52params}
 \begin{center}
@@
 MAPE, MAE and RMSE for OVECM, SLVECM and SLARIMA. Test were done using the
-parameters defined in table \ref{tab:params}.  We can see that OVECM has very
+parameters defined in table \ref{tab:52params}.  We can see that OVECM has very
 similar performance than SLVECM and this support the theory that cointegration
@@
 
-\section{Conclusions} \label{sec:conclusions}
+\section{Conclusions} \label{sec:52conclusions}
 

### 2017-06-15 src/5_3_Proposal.tex
change tittle
Intro 5-3 ready

--- 2015-06-17 src/5_3_Proposal.tex
@@
-\chapter{Real-time intraday Forex rates forecasting using a regularized Vector
-Error Correction model}
+\chapter{A Regularized Time Varying Vector Error Correction Model}
 
-Vector error correction model (VECM) parameters are commonly obtained using the
-ordinary least squares method. Since this parameters could be sensitive to the
-data, this proposal is to get these parameters using ridge regression which
-could lead to a better generalization capability. In order to give with a
-real-time response, an online version of VECM (OVECM) is proposed. OVECM solves
-VECM considering only a sliding window of historical data and therefore
-better response times could be achieved.
+In the last chapter, the Online Vector Error Correction Model (OVECM) was proposed.
+OVECM parameters are obtained using the Ordinary Least Squares (OLS) method.
+Even though OLS is extensively used, it has over-fitting issues and is computationally expensive.
+Therefore, OVECM parameters could be very sensitive to the data.
+Ridge Regression (RR)  is commonly used
+instead of OLS to tackle this problem, introducing a regularization parameter which could improve
+prediction error.
+In this chapter, the use of Ridge Regression (RR) and its variant called Aggregating Algorithm for Regression (AAR) is explored in order to get better generalisation capability.
 

### 2017-06-15 src/5_3_Proposal.tex
saving my work: adding background sections to thesis

@@
 
-VECM introduces the long-run relationship among a set of cointegrated variables
-as an error correction term. VECM is a special case of the vector autorregresive
-model (VAR) model. VAR model expresses future values as a linear combination of
-variables past values.  However, VAR model cannot be used with non-stationary
-variables, which is a common feature of financial assets. VECM is a linear model
-but in terms of variable differences. If cointegration exists, variable
-differences are stationary and they introduce an error correction term which
-adjusts coefficients to bring the variables back to equilibrium. In finance,
-many economic time series turn to be stationary when they are differentiated and
-cointegration restrictions often improves forecasting~\cite{duy1998}. Therefore,
-VECM has been widely adopted.
+%So why do we care about cointegration? Someone else can probably give more
+%econometric applications, but in quantitative finance, cointegration forms the
+%basis of the pairs trading strategy: suppose we have two cointegrated stocks X
+%and Y, with the particular (for concreteness) cointegrating relationship X - 2Y
+%= Z, where Z is a stationary series of zero mean. For example, X could be
+%McDonald's, Y could be Burger King, and the cointegration relationship would
+%mean that X tends to be priced twice as high as Y, so that when X is more than
+%twice the price of Y, we expect X to move down or Y to move up in the near
+%future (and analogously, if X is less than twice the price of Y, we expect X to
+%move up or Y to move down). This suggests the following trading strategy: if X -
+%2Y > d, for some positive threshold d, then we should sell X and buy Y (since we
+%expect X to decrease in price and Y to increase), and similarly, if X - 2Y < -d,
+%then we should buy X and sell Y.
 
@@
 
-Our proposal is an online formulation of the VECM called Online VECM (OVECM)
-based on consideration of only a sliding window of the historical data.  OVECM
-introduces matrix optimizations in order to reduce the number of operations and
-also takes into account the fact that cointegration vector space doesn't
-experience large changes with small changes in the input data. Moreover, OVECM
-uses RR instead of OLS to obtain VECM parameters. Our method is later tested using four
-currency rates from the foreign exchange market with a frequency of 10 seconds.
-Model efectiveness is focused on out-of-sample forecast rather than in-sample
-fitting.  This criteria allows the OVECM prediction capability to be expressed
-rather than just explaining data history. Our method performance is compared
-with its optimal offline algorithm.
+In this paper, we propose an online formulation of the VECM called Onlive VECM
+(OVECM) based on consideration of only a sliding window of the most recent data.
+The algorithm introduces matrix optimizations in order to reduce the number of
+operations and also takes into account the fact that cointegration vector space
+doesn't experience large changes with small changes in the input data. Moreover,
+RR instead of OLS is used to solve VECM. Our method is later tested using four
+currency rates from the foreign exchange market with different frequencies.  Model
+efectiveness is focused on out-of-sample forecast rather than in-sample fitting.
+This criteria allows the OVECM prediction capability to be expressed rather than
+just explaining data history. Our method performance is compared with its
+optimal offline algorithm.
 
-%The next sections are organized as follows: section~\ref{sec:background}
-%presents the VAR and VECM, the OVECM algorithm proposed is presented in
-%section~\ref{sec:methodology}. Section~\ref{sec:results} gives a description of
-%the data used and the tests carried on to show accuracy and time comparison of
-%our proposal against the traditional VECM and
-%section~\ref{sec:conclusions} includes conclusions and a proposal for future
-%tudy.
+The next sections are organized as follows: section~\ref{sec:background}
+presents the VAR and VECM, the OVECM algorithm proposed is presented in
+section~\ref{sec:methodology}. Section~\ref{sec:results} gives a description of
+the data used and the tests carried on to show accuracy and time comparison of
+our proposal against the traditional VECM and
+section~\ref{sec:conclusions} includes conclusions and a proposal for future
+tudy.
 

### 2017-06-23 src/4_OnlineLearning.tex
Adding Regularisation

@@
 \subsection{Regularization}
-Since ERM is an ill-posed problem, Tikhonov introduced a regularisation which ensures well-posedness and generalisation of ERM, i.e prevents overfit, by constraining the hypothesis space $\mathcal{H}$ usually called regularised ERM or Tikhonov regularisation.
+
+Regularization can be understood using two contexts: learning theory (probabilistic) and inverse problems (deterministic).
+In the context of learning, regularization refers to techniques allowing to avoid over-fitting and the desire property of the selected estimator is to perform well on new data (to generalize), e.g.
+regularized least squares.
+In the context of inverse problems, regularization objective is to stabilise, with respect to noise, a possibly ill-conditioned matrix inversion problem e.g spectral cut-off and Tikhonov regularization or ridge regression (RR)\cite{ tikhonov1977}.
+
+In particular, it is well known that regularization schemes such as RR or Tikhonov regularization can be effectively used in the context of learning \cite{vito2005}.
+
+Tikhonov introduced a regularisation which ensures well-posedness and generalisation of ERM, i.e prevents overfit, by constraining the hypothesis space $\mathcal{H}$ usually called regularised ERM or Tikhonov regularisation.
 

### 2017-06-23 src/4_OnlineLearning.tex
Avance hasta Competitive analysis

@@
 
+Ridge regression was independently proposed by Tikhonov \cite{tikhonov1963} and Phillips \cite{phillips1962}. 
 One example of regularisation in linear models is Ridge Regression (RR), which is a regularised least squares method.
@@
 
+The method is called ridge regression because the term $\lambda \mathbb{I}$ adds positive entries along the diagonal (ridge) to avoid the
+singularity of the covariance matrix $\mathbf{A}^\top \mathbf{A}$. This addition ensures that all of the covariance matrix eigenvalues will be strictly greater than 0, i.e the solution becomes unique.
+
 \subsection{The Lambda parameter}
@@
 
+$\blacksquare$
+
+\subsection{Generalized Cross Validation}
+Since we don't know the data we want to predict, the bias-variance tradeoff is not directly useful to determine $\lambda$.
+Many techniques have been developed to determine a suitable value for the regularization parameter.
+However, when no information is available on the error, one of the most popular techniques is the Generalized Cross Validation (GCV) \cite{bauer2011}.
+GCV originates from the older method leave-one-out Cross Validation \cite{stone1974}.
+
+The leave-one-out cross validation method determine a regularization parameter that minimises the average of the squared predictions errors using each solution to predict the missing data value.
+
+%'---------------------------http://robjhyndman.com/hyndsight/crossvalidation/
+%When the data are not independent cross-validation becomes more difficult as leaving out an observation does not remove all the associated information due to the correlations with other observations.
+
+For time series forecasting, a cross-validation statistic is obtained as follows: fit the model to the data $y1,…,yt$ and let $y^t+1$ denote the forecast of the next observation.
+Then compute the error ($et+1∗=yt+1−y^t+1$) for the forecast observation.
+Repeat step 1 for t=m,…,n−1 where m is the minimum number of observations needed for fitting the model.
+Compute the MSE from $em+1∗,…,en∗$.
+
+%An excellent and comprehensive recent survey of cross-validation results is Arlot and Celisse (2010)
+
+%---------------------------------------------------
+
 \subsection{Machine learning algorithms}
@@
 model updating is computationally less expensive than a training phase.
+
 Online algorithms allow incremental learning by processing one instance at a
@@
 analysis of online learning is provided.
+Applications in finance has been widely used: study presented by~\cite{arce+salinas2012} applied ridge regression in an online context and more recently, time series forecasting using online learning has been presented~\cite{anavaetAl2013}.
 
@@
 \end{equation*}
+
+%REVISADO HASTA ACA
 
@@
 
-Despite the bias-variance tradeoff provides a conceptual framework for
-determining a good model, is not directly useful. Some popular methods
-for determine model selection are:
-
-\begin{itemize}
-    \item Akaike information criterion (AIC) (Akaike, 1974)
-    \item Bootstrap-based selection (Efron and Tibshirani, 1997)
-    \item Cross-validation (Stone, 1974)
-\end{itemize}
-
-On the other hand, $\lambda$ can also be used to reduce computational time as it
-is shown in equation~(\ref{eq:taylor}). Golub (~\cite{golub1965}) suggested that
-$\lambda$ be chosen so that:
-
-\[
-\frac{\lambda}{\lambda + \delta^2} < 0.1
-\]
-
-\noindent where $\delta$ is a lower bound of the smallest non-zero singular
-value $\sigma_k$. One approach is to choose the greatest lower bound for
-$\delta=\sigma_k$. Hence:
-
-\[
-\frac{\lambda}{\lambda + \sigma_k^2} < 0.1
-\]
-
-\noindent if we choose $\lambda=\beta \sigma_k^2$ then we get the following
-relation:
-
-\[
-\frac{\beta \sigma_k^2}{\beta \sigma_k^2 + \sigma_k^2} =
-\frac{\beta}{\beta+1} < 0.1 \Rightarrow \beta < \frac{1}{9}  
-\]
-
-Experiments done by ~\cite{coleman+sun2010} have shown that $\beta = 0.01$
-produces satisfactory results. This means that $\lambda = 0.01 \sigma_k^2$.
-
-However, get $\sigma_k$ imply obtaining first the SVD, which is
-computational expensive.
-
-Since the algorithm presented by ~\cite{coleman+sun2010} already compute the QR
-factorization of the matrix $\mathbf{A}$ they show a procedure for getting
-a $\lambda$ approximation using QR.
-
-\begin{enumerate}
-\item Compute QR factorization: $\mathbf{A} = \mathbf{Q_1R_1}$
-\item Let $\mathbf{W}$ denote the set of absolute values of the nonzero diagonal
-elements of $\mathbf{R_1}$. Let $w_{\text{min}}$ and $w_{\text{max}}$ denote the
-smallest and largest elements of $\mathbf{W}$ respectively. Both 
-$\lambda_1 = \hat{\beta} w_{\text{min}}^2$
-$\lambda_2 = \hat{\beta} \frac{w_{\text{min}}^2}{w_{\text{max}}^2}$ where
-$\hat{\beta}=0.00025$ produce satisfactory results.
-\item The $\lambda_{\text{QR}}$ approximation is obtained as:
-$\lambda_{\text{QR}} = \frac{\lambda_1 + \lambda_2}{2}$  
-\end{enumerate}
+%Despite the bias-variance tradeoff provides a conceptual framework for
+%determining a good model, is not directly useful. Some popular methods
+%for determine model selection are:
+%
+%\begin{itemize}
+%    \item Akaike information criterion (AIC) (Akaike, 1974)
+%    \item Bootstrap-based selection (Efron and Tibshirani, 1997)
+%    \item Cross-validation (Stone, 1974)
+%\end{itemize}
+%
+%On the other hand, $\lambda$ can also be used to reduce computational time as it
+%is shown in equation~(\ref{eq:taylor}). Golub (~\cite{golub1965}) suggested that
+%$\lambda$ be chosen so that:
+%
+%\[
+%\frac{\lambda}{\lambda + \delta^2} < 0.1
+%\]
+%
+%\noindent where $\delta$ is a lower bound of the smallest non-zero singular
+%value $\sigma_k$. One approach is to choose the greatest lower bound for
+%$\delta=\sigma_k$. Hence:
+%
+%\[
+%\frac{\lambda}{\lambda + \sigma_k^2} < 0.1
+%\]
+%
+%\noindent if we choose $\lambda=\beta \sigma_k^2$ then we get the following
+%relation:
+%
+%
+%\[
+%\frac{\beta \sigma_k^2}{\beta \sigma_k^2 + \sigma_k^2} =
+%\frac{\beta}{\beta+1} < 0.1 \Rightarrow \beta < \frac{1}{9}  
+%\]
+%
+%Experiments done by ~\cite{coleman+sun2010} have shown that $\beta = 0.01$
+%produces satisfactory results. This means that $\lambda = 0.01 \sigma_k^2$.
+%
+%However, get $\sigma_k$ imply obtaining first the SVD, which is
+%computational expensive.
+%
+%Since the algorithm presented by ~\cite{coleman+sun2010} already compute the QR
+%factorization of the matrix $\mathbf{A}$ they show a procedure for getting
+%a $\lambda$ approximation using QR.
+%
+%\begin{enumerate}
+%\item Compute QR factorization: $\mathbf{A} = \mathbf{Q_1R_1}$
+%\item Let $\mathbf{W}$ denote the set of absolute values of the nonzero diagonal
+%elements of $\mathbf{R_1}$. Let $w_{\text{min}}$ and $w_{\text{max}}$ denote the
+%smallest and largest elements of $\mathbf{W}$ respectively. Both 
+%$\lambda_1 = \hat{\beta} w_{\text{min}}^2$
+%$\lambda_2 = \hat{\beta} \frac{w_{\text{min}}^2}{w_{\text{max}}^2}$ where
+%$\hat{\beta}=0.00025$ produce satisfactory results.
+%\item The $\lambda_{\text{QR}}$ approximation is obtained as:
+%$\lambda_{\text{QR}} = \frac{\lambda_1 + \lambda_2}{2}$  
+%\end{enumerate}
 

### 2017-06-24 src/4_OnlineLearning.tex
background 2016  include finished
Background revision 2015 2016 elsevier ready

@@
 
-In order to avoid the singularity of the matrix $\mathbf{A}^\top \mathbf{A}$, a regularisation term is introduced:
+In order to avoid the singularity of the matrix $\mathbf{A}^\top \mathbf{A}$, a regularisation term $\lambda$ is introduced:
 
@@
 \end{equation*}
-
-%REVISADO HASTA ACA
 
@@
 
-Equation~(\ref{eq:optsolRR}) can also be written as:
+\noindent equation~(\ref{eq:optsolRR}) can also be written as:
+
 \begin{eqnarray*}

### 2017-07-13 src/5_3_Proposal.tex
elsevier 2016 journal.tex and intro.tex revised

--- 2016-03-22 src/5_3_Proposal.tex
@@
-\chapter{Intraday Forex rates forecasting using an online cointegration approach}
-
-High frequency forex rates 
+\chapter{A Regularized Time Varying Vector Error Correction Model}
 Vector error correction model (VECM) parameters are commonly obtained using the
-ordinary least squares method. Since this parameters could be sensitive to the
-data, this proposal is to get these parameters using ridge regression which
-could lead to a better generalization capability. In order to give with a
-real-time response, an online version of VECM (OVECM) is proposed. OVECM solves
-VECM considering only a sliding window of historical data and therefore
-better response times could be achieved.
+ordinary least squares method (OLS). Since these parameters could be very
+sensitive to the data, this proposal explores the use of ridge regression (RR)
+and the agregating algorithm for regression (AAR) which could lead to a better
+generalization capability. In the previous chapter OVECM was presented, which is
+an online version of the traditional VECM (Vector Error Correction) that uses
+OLS to get the model parameters. In this chapter a variation of OVECM is
+presented. RR and AAR will be used instead of OLS to get the model parameters in
+an online way.
 
@@
 
-VECM introduces the long-run relationship among a set of cointegrated variables
-as an error correction term. VECM is a special case of the vector autorregresive
-model (VAR) model. VAR model expresses future values as a linear combination of
-variables past values.  However, VAR model cannot be used with non-stationary
-variables, which is a common feature of financial assets. VECM is a linear model
-but in terms of variable differences. If cointegration exists, variable
-differences are stationary and they introduce an error correction term which
-adjusts coefficients to bring the variables back to equilibrium. In finance,
-many economic time series turn to be stationary when they are differentiated and
-cointegration restrictions often improves forecasting~\cite{duy1998}. Therefore,
-VECM has been widely adopted.
-
-Both VECM and VAR model parameters are obtained using ordinary least squares
+Traditionally VECM parameters are obtained using ordinary least squares
 (OLS) method. OLS has two main problems: is sensitive to errors on input data
@@
 with its optimal offline algorithm.
-
-%The next sections are organized as follows: section~\ref{sec:background}
-%presents the VAR and VECM, the OVECM algorithm proposed is presented in
-%section~\ref{sec:methodology}. Section~\ref{sec:results} gives a description of
-%the data used and the tests carried on to show accuracy and time comparison of
-%our proposal against the traditional VECM and
-%section~\ref{sec:conclusions} includes conclusions and a proposal for future
-%tudy.
 

### 2017-07-13 src/3_TimeSeries.tex

@@
 In order to determine if a time series is an a I(1) or I(0) process unit root tests are used.
-Several tests have been developed but the most widely used is an extension of the Dickey and Fuller test \cite{dickey1979} called the Augmented Dickey-Fuller (ADF) test developed by Said and Dickey in 1984\cite{said1984}.
+Several tests have been developed but the most widely used is an extension of the Dickey and Fuller test \cite{dickey1979} called the Augmented Dickey-Fuller (ADF) test developed by Said and Dickey in 1984\cite{said1984} which tabulated critical values were obtained by~\cite{mackinnon2010}.
 
@@
 The $\text{ADF}_t$ statistic is based on the least squares estimates of equation \ref{eq:adftest}.
+
+\section {Cointegration Tests}
+
+Johansen method~\cite{johansen1988} allows to find cointegration among several
+I(1) time series. This test is more general than the
+Engle-Granger~\cite{engle1987} test which only allow to find cointegration
+between two time series.  Johansen procedure takes the form of a
+likelihood-ratio test which determines $r$ cointegration relationships. The
+method provides two tests based on an eigenvector decomposition of the matrix
+$\Omega = \alpha \beta^\top$ on equation~\ref{eq:vec} called trace and maximal
+eigenvalue statistic.  Once the number of cointegration number is determined,
+$r$ more significative eigenvectors provided by Johansen procedure are used to
+get the cointegration vector matrix $\beta^\top$.  Critical values for both
+tests are limited to 12 and 5 variables in the trace and maximal eigenvalue
+statistic respectively.
+
+The trace test null hypothesis is $H_0: r = r_0$ against $H_1: r > r_0$ and the
+maximum eigenvalue test null hypothesis is $H_0: r = r_0$ against $H_1: r =
+r_0+1$ starting with $r_0 = 0$.  In the latest test, if we start with $r_0=0$
+and $H_0$ is rejected $H_1: r = 1$ will be accepted which is not true if there
+are more than one cointegration relationship, this is why this test is less
+powerful and the trace test is commonly used.
 
@@
 \begin{equation}
-\label{eq:pseudoinverse}
+\label{eq:pseudoinv}
 \mathbf{A}^{\!\!+}= (\mathbf{A}^{\!\!\top} \mathbf{A})^{-1}\mathbf{A}^{\!\!\top} \, .
@@
 $rank(\mathbf{A})=k <  n \leq m$, $\mathbf{A}^\top \mathbf{A}$ is
-always singular and equation~(\ref{eq:pseudoinverse}) cannot be used.
+always singular and equation~(\ref{eq:pseudoinv}) cannot be used.
 More generally, the pseudo-inverse is best computed using the compact
@@
 \end{equation}
+
+\textbf{Demo}\quad
+
+Since $\mathbf{A}$ is singular the problem shown in
+equation~(\ref{eq:regressionproblem}) has not solution, the minimum norm given
+by equation~(\ref{eq:MP}) is obtained by solving the equivalent problem:
+
+\begin{equation*}
+\label{eq:proyectorsol}
+\mathbf{A \hat{X} = PB} 
+\end{equation*}
+
+\noindent where $\mathbf{P=U_1 U_1^\top}$ is the projection onto the
+Col($\mathbf{A}$).
+
+Since $\mathbf{V} = [\underset{(n \times k)}{\mathbf{V_1}} |
+\underset{(n \times k)}{\mathbf{V_2}}]$ and $\mathbf{V_1^\top V_2 =
+0}$ we can express $\mathbf{\hat{X}} = \mathbf{V_1 x_1 + V_2 x_2}$
+with $\mathbf{x_2=0}$ because $\mathbf{\hat{X}}$ lives in the
+$\text{Row}(\mathbf{A})$ given by $\mathbf{V_1}$, so we have:
+
+\begin{eqnarray*}
+\mathbf{A \hat{X}} &=& \mathbf{PB} \\
+\mathbf{U_1 \Sigma_1 V_1^\top \hat{X}} &=& \mathbf{U_1 U_1^\top B} \\
+\mathbf{ V_1^\top \hat{X}} &=&  \mathbf{\Sigma_1^{-1} U_1^\top B} \\ 
+\mathbf{ V_1^\top V_1 x_1} &=& \mathbf{\Sigma_1^{-1}
+U_1^\top B} \\
+\mathbf{x_1}&=& \mathbf{\Sigma_1^{-1} U_1^\top B}
+\end{eqnarray*}
+
+\noindent from this result we can obtain $\mathbf{\hat{X}}$ and
+therefore the pseudo-inverse expression:
+
+\begin{eqnarray*}
+\mathbf{\hat{X}} &=& \mathbf{V_1 x_1} \\
+                &=& \mathbf{V_1 \Sigma_1^{-1} U_1^\top B} \\
+\mathbf{A^+} &=& \mathbf{V_1 \Sigma_1^{-1} U_1^\top} \, .
+\end{eqnarray*}
+
+$\blacksquare$
 

### 2017-07-13 src/5_3_Proposal.tex

--- 2017-06-15 src/5_3_Proposal.tex
@@
 
-In the last chapter, the Online Vector Error Correction Model (OVECM) was proposed.
-OVECM parameters are obtained using the Ordinary Least Squares (OLS) method.
-Even though OLS is extensively used, it has over-fitting issues and is computationally expensive.
-Therefore, OVECM parameters could be very sensitive to the data.
-Ridge Regression (RR)  is commonly used
-instead of OLS to tackle this problem, introducing a regularization parameter which could improve
-prediction error.
-In this chapter, the use of Ridge Regression (RR) and its variant called Aggregating Algorithm for Regression (AAR) is explored in order to get better generalisation capability.
+In the last chapter, the Online Vector Error Correction Model (OVECM) was
+proposed. OVECM parameters are obtained using the Ordinary Least Squares (OLS)
+method. Even though OLS is extensively used, it has over-fitting issues and is
+computationally expensive. Therefore, OVECM parameters could be very sensitive
+to the data. Ridge Regression (RR) is commonly used instead of OLS to tackle
+this problem, introducing a regularization parameter which could improve
+prediction error.  In this chapter, the use of Ridge Regression (RR) and its
+variant called Aggregating Algorithm for Regression (AAR) is explored in order
+to get better generalisation capability.
 
 \section{The problem}
-
-VECM introduces the long-run relationship among a set of cointegrated variables
-as an error correction term. VECM is a special case of the vector autorregresive
-model (VAR) model. VAR model expresses future values as a linear combination of
-variables past values.  However, VAR model cannot be used with non-stationary
-variables, which is a common feature of financial assets. VECM is a linear model
-but in terms of variable differences. If cointegration exists, variable
-differences are stationary and they introduce an error correction term which
-adjusts coefficients to bring the variables back to equilibrium. In finance,
-many economic time series turn to be stationary when they are differentiated and
-cointegration restrictions often improves forecasting~\cite{duy1998}. Therefore,
-VECM has been widely adopted.
 

### 2017-07-13 src/4_OnlineLearning.tex
comment section of Cross Validation

@@
 
-For time series forecasting, a cross-validation statistic is obtained as follows: fit the model to the data $y1,…,yt$ and let $y^t+1$ denote the forecast of the next observation.
-Then compute the error ($et+1∗=yt+1−y^t+1$) for the forecast observation.
-Repeat step 1 for t=m,…,n−1 where m is the minimum number of observations needed for fitting the model.
-Compute the MSE from $em+1∗,…,en∗$.
+%For time series forecasting, a cross-validation statistic is obtained as
+%follows: fit the model to the data $y1,\dots,yt$ and let $y^t+1$ denote the forecast
+%of the next observation. Then compute the error $et+1=yt+1-y^t+1$ for the
+%forecast observation.  Repeat step 1 for $t=m,\dots,n−1$ where m is the minimum number
+%of observations needed for fitting the model.
+%Compute the MSE from $em+1∗,…,en∗$.
 

### 2017-07-13 src/5_3_Proposal.tex
falta revisar algoritmos

@@
 
-Both VECM and VAR model parameters are obtained using ordinary least squares
+VECM model parameters are traditionally obtained using ordinary least squares
 (OLS) method. OLS has two main problems: is sensitive to errors on input data
@@
 
-Our proposal is an online formulation of the VECM called Online VECM (OVECM)
-based on consideration of only a sliding window of the historical data.  OVECM
-introduces matrix optimizations in order to reduce the number of operations and
-also takes into account the fact that cointegration vector space doesn't
-experience large changes with small changes in the input data. Moreover, OVECM
-uses RR instead of OLS to obtain VECM parameters. Our method is later tested using four
-currency rates from the foreign exchange market with a frequency of 10 seconds.
-Model efectiveness is focused on out-of-sample forecast rather than in-sample
-fitting.  This criteria allows the OVECM prediction capability to be expressed
-rather than just explaining data history. Our method performance is compared
-with its optimal offline algorithm.
+The proposal is to modify OVECM so it wan be an incremental algorithm and also
+to use RR and its variant AAR instead of OLS. Therefore, three different
+variations of OVECM using as a solver: OLS, RR and AAR will be compared.
 
@@
 
-On the other hand, getting cointegration vectors using Johansen method is also
-an expensive procedure. However, since cointegration
-vectors represent the long-run relationship between the time series, they
-vary little in time. Our proposal obtains new cointegration vectors only when
-this long-term relationship changes. This change is detected by tracking 
-the Mean Absolute Percentage Error (MAPE) of the last $n$ in-sample forecasts.
+On the other hand, getting cointegration vectors using Johansen method is also a
+computationally expensive procedure. However, since cointegration vectors
+represent the long-run relationship between the time series, they vary little in
+time. This proposal also include to obtain new cointegration vectors only when
+this long-term relationship changes. This change is detected by tracking the
+Mean Absolute Percentage Error (MAPE) of the last $n$ in-sample forecasts.
 
@@
 
-VECM($p$) for $l$ cointegrated variables has the following form:
+VECM($p$) was defined in equation \ref{eq:vecfull} and its matrix form is:
 
-\begin{equation}\label{eq:vecfull}
-\Delta\mathbf{y}_t 
-= \boldsymbol{\alpha\beta}^\top\mathbf{y}_{t-1} 
-  + \sum_{i=1}^{p-1}\boldsymbol{\Phi}_i^*\,\Delta\mathbf{y}_{t-i}
-  + \mathbf{c} + \boldsymbol{\epsilon}_t \qquad \forall t = p+1,\dots,N.
-\end{equation}
-
-Transposing each equation of the system (\ref{eq:vecfull}) we can write
-the VECM($p$) model in block-matrix form as:
 \begin{equation}\label{eq:vareq}
@@
 \end{equation}
-%
-\noindent where $\mathbf{B}$ dimension is $((N-p)\times l)$, $\mathbf{A}$
-dimension is $((N-p)\times(r+(p-1)l +1))$, $\mathbf{X}$ dimension is $((r+(p-1)l
-+1)\times l)$, $\mathbf{E}$ dimension is $((N-p)\times l)$ and $r$ is the number
-of cointegration vectors:
-%
-\begin{alignat}{3}
-\mathbf{B}
-&= \begin{bmatrix}
-   \Delta\mathbf{y}_{p+1}^\top \\
-   \Delta\mathbf{y}_{p+2}^\top \\
-   \vdots \\
-   \Delta\mathbf{y}_N^\top
-   \end{bmatrix}
-&\quad
-\mathbf{X}
-&= \begin{bmatrix}
-   \boldsymbol{\alpha}^\top \\
-   \boldsymbol{\Phi}_1^{*\top} \\
-   \boldsymbol{\Phi}_2^{*\top} \\
-   \vdots \\
-   \boldsymbol{\Phi}_{p-1}^{*\top} \\
-   \mathbf{c}^\top
-   \end{bmatrix}
-&\quad
-\mathbf{E}
-&= \begin{bmatrix}
-   \boldsymbol{\epsilon}_{p+1}^\top \\
-   \boldsymbol{\epsilon}_{p+2}^\top \\
-   \vdots \\
-   \boldsymbol{\epsilon}_N^\top \\
-   \end{bmatrix}
-\end{alignat}
-\noindent and 
-\begin{align}
-\mathbf{A} 
-&= \begin{pmat}[{....|}]
-   \mathbf{y}_p^\top \boldsymbol{\beta} & \Delta \mathbf{y}_p^\top & \Delta\mathbf{y}_{p-1}^\top & \dots 
-                    & \Delta\mathbf{y}_2^\top & 1 \cr
-   \mathbf{y}_{p+1}^\top  \boldsymbol{\beta} &\Delta\mathbf{y}_{p+1}^\top & \Delta\mathbf{y}_p^\top & \dots
-                       & \Delta\mathbf{y}_3^\top & 1 \cr
-   \vdots & \vdots & \vdots & \ddots & \vdots & \vdots \cr
-   \mathbf{y}_{N-1}^\top  \boldsymbol{\beta} &\Delta\mathbf{y}_{N-1}^\top & \Delta\mathbf{y}_{N-2}^\top & \dots 
-                       & \Delta\mathbf{y}_{N-p-1}^\top & 1 \cr
-   \end{pmat}\, .
-\end{align}
-%Taking into account the error term $\mathbf{E}$, equation~(\ref{eq:vareq}) 
-%can be solved with respect to $\mathbf{X}$ using the ordinary least
-%squares estimation.
-However, the number of rows of matrix $\mathbf{A}$ increases with the number of
-observations $\mathbf{y}_t$. Our proposal considers only the most recent $L$
+
+\noindent where matrix $\mathbf{B}$ is defined in equation
+\ref{eq:vecY}, matrix $\mathbf{X}$ is detailed in equation \ref{eq:vecX}, matrix
+$\mathbf{E}$ in equation \ref{eq:vecE} and matrix $\mathbf{A}$ in equation \ref{eq:vecA}.
+
+In the traditional VECM the number of rows of matrix $\mathbf{A}$ increases with the number of
+observations $\mathbf{y}_t$. This proposal considers only the most recent $L$
 rows of matrices $\mathbf{A}$ and $\mathbf{B}$ defined as  $\mathbf{A}(t)$ and
@@
 
-Our proposal considers the following:
+The proposal considers the following:
 

### 2017-07-20 src/5_2_Proposal.tex
details only

@@
     \node (pro4a) [process, right of=dec2, xshift=4cm] {$v = \texttt{Johansen}(\mathbf{y}_i,p)$ \\
-                                                        $\mathbf{A} = \texttt{uvecm}(\mathbf{y}_i,p,v,\mathbf{A})$ \\
+                                                        $\mathbf{A} = \texttt{ovecm}(\mathbf{y}_i,p,v,\mathbf{A})$ \\
                                                         $\mathbf{X} = \texttt{OLS} (\mathbf{A},\mathbf{B})$};

### 2017-07-20 src/5_3_Proposal.tex
details only

@@
 VECM model parameters are traditionally obtained using ordinary least squares
-(OLS) method. OLS has two main problems: is sensitive to errors on input data
+(OLS) method. OLS has two main problems: is sensitive to input data errors
 and involves many calculations. The former problem is commonly solved using
@@
 
-The proposal is to modify OVECM so it wan be an incremental algorithm and also
-to use RR and its variant AAR instead of OLS. Therefore, three different
-variations of OVECM using as a solver: OLS, RR and AAR will be compared.
+The proposal is to create an online algorithm based on the idea of OVECM
+using either RR or its variant AAR instead of OLS 
+.
 
@@
 
-Since financial problems are stream data problems, it is unfeasible to include
-all input data in a VECM model. Our proposal consists on an online version of
-VECM (OVECM) capable of updating the model with new arrival data and give
-responses in a short period of time. OVECM only considers a time varying window
-of historical data and the using of RR or its variant AAR (Aggregating Algorithm
-for Regression) as an alternative of OLS to get model parameters.
+OVECM only considers a time varying window of historical data and uses OLS to
+get solve model parameters. This proposal is about using RR or its variant AAR
+as an alternative of OLS to get model parameters.
 
-On the other hand, getting cointegration vectors using Johansen method is also a
-computationally expensive procedure. However, since cointegration vectors
-represent the long-run relationship between the time series, they vary little in
-time. This proposal also include to obtain new cointegration vectors only when
-this long-term relationship changes. This change is detected by tracking the
-Mean Absolute Percentage Error (MAPE) of the last $n$ in-sample forecasts.
+Since cointegration vectors represent a long-run relationship between the time
+series, they vary little in time. This proposal also includes to obtain new
+cointegration vectors only when this long-term relationship changes. This change
+is detected by tracking the Mean Absolute Percentage Error (MAPE) of the last
+$n$ in-sample forecasts.
 
-%Since VECM is a model based on time series differences, the MAPE is obtained
-%from $\Delta \mathbf{y}$ as following:
-% 
-%\begin{equation}\label{eq:MAPE}
-%\text{MAPE}[t] = \frac{1}{n} \sum_{i=1}^{n} \left| 
-%\frac{\Delta \mathbf{y}_{\text{true}}[t-i]-\Delta
-%\mathbf{y}_{\text{pred}}[t-i]}{\Delta \mathbf{y}_{\text{true}[t-i]}}
-%\right| \, , 
-%\end{equation}
-%
-%\noindent where $\Delta \mathbf{y}_{\text{true}}$ is the actual value of the
-%time series $\mathbf{y}$ differences and $\Delta \mathbf{y}_{\text{pred}}$ is
-%their forecast value.
-
-%The method proposed consists of
-%a modification of RR considering a sliding window which contains only
-%the last $L$ samples %and the new input $\mathbf{x}_t$, 
-%i.e. $\{\mathbf{x}_i\}_{i=t-L+1}^{t-1}$.
-
-VECM($p$) was defined in equation \ref{eq:vecfull} and its matrix form is:
+VECM($p$) matrix form is:
 
@@
 
-\noindent where matrix $\mathbf{B}$ is defined in equation
-\ref{eq:vecY}, matrix $\mathbf{X}$ is detailed in equation \ref{eq:vecX}, matrix
-$\mathbf{E}$ in equation \ref{eq:vecE} and matrix $\mathbf{A}$ in equation \ref{eq:vecA}.
+\noindent where matrix $\mathbf{B},\mathbf{X},\mathbf{E}$ and $\mathbf{A}$ were
+previously defined in equations \ref{eq:vecY} , \ref{eq:vecX} , \ref{eq:vecE} and
+\ref{eq:vecA} respectively.
 

### 2017-07-20 src/5_3_Proposal.tex
model updates before pseudoalgorithms

@@
 \begin{equation}
+\mathbf{B}(t) = \mathbf{A}(t) \mathbf{X}(t) + \mathbf{E}(t)\\
+\label{eq:vareqt}
+\end{equation}
+
+\noindent where,
+
+\begin{equation}
 \label{eq:notation}
@@
 
-\noindent so that VECM parameters $\mathbf{X}(t)$ are found using the following
-equation:
-
-\begin{equation}
-\mathbf{B}(t) = \mathbf{A}(t) \mathbf{X}(t) + \mathbf{E}(t)\\
-\end{equation}
-
-The RR solution $\mathbf{\hat{X}}(t)$ using the sliding window matrices defined above
-is:
+The RR solution $\mathbf{\hat{X}}(t)$ of equation \ref{eq:vareqt} using the
+sliding window matrices defined above is:
 
@@
 \end{eqnarray*}
+
+\noindent here $\lambda$ corresponds to the regularization parameter (see section
+\ref{sec:regularization}). When $\lambda=0$, the solution is equivalent to OLS.
+
+The AAR solution only differs in the calculation $\mathbf{S}(t)$ which is
+updated with the last input received $\mathbf{a}_{L+1}$:
+
+\begin{equation}
+\mathbf{S}(t) =
+\sum_{i=0}^{L+1} \mathbf{a}_{t-i}\mathbf{a}_{t-i}^\top + \lambda \mathbb{I}
+\label{eq:SAAR}
+\end{equation}
 

### 2017-07-23 src/5_3_Proposal.tex
changing minor errors

@@
 and involves many calculations. The former problem is commonly solved using
-Ridge Regression (RR) \cite{hoerl1970} which introduces a regularization
+Ridge Regression (RR) \cite{hoerl1970} that introduces a regularization
 parameter that leads to an unbiased estimation with better generalization
@@
 The proposal is to create an online algorithm based on the idea of OVECM
-using either RR or its variant AAR instead of OLS 
-.
+using either RR or its variant AAR instead of OLS.
 
@@
 OVECM only considers a time varying window of historical data and uses OLS to
-get solve model parameters. This proposal is about using RR or its variant AAR
-as an alternative of OLS to get model parameters.
+get model parameters. This proposal is about using RR or its variant AAR
+as an alternative of OLS to estimate model parameters.
 
@@
 cointegration vectors only when this long-term relationship changes. This change
-is detected by tracking the Mean Absolute Percentage Error (MAPE) of the last
-$n$ in-sample forecasts.
+is detected by tracking the Mean Absolute Percentage Error (MAPE) defined in the
+equation \ref{eq:MAPE}, of the last $n$ in-sample forecasts.
 
@@
 
-\begin{equation}\label{eq:vareq}
+\begin{equation*}
 \mathbf{B} = 
@@
 \mathbf{E} \, , 
-\end{equation}
+\end{equation*}
 
@@
 \mathbf{B}(t) = \mathbf{A}(t) \mathbf{X}(t) + \mathbf{E}(t)\\
-\label{eq:vareqt}
+\label{eq:vareqt1}
 \end{equation}
@@
   \end{tabular}
-\right] \, ,
+\right] \, .
 \
@@
 
-The RR solution $\mathbf{\hat{X}}(t)$ of equation \ref{eq:vareqt} using the
+The RR solution $\mathbf{\hat{X}}(t)$ of equation \ref{eq:vareqt1} using the
 sliding window matrices defined above is:
@@
 \noindent here $\lambda$ corresponds to the regularization parameter (see section
-\ref{sec:regularization}). When $\lambda=0$, the solution is equivalent to OLS.
+\ref{sec:regularization}). When $\lambda=0$, the solution is equivalent to the
+one given by OLS.
 

### 2017-07-23 src/5_3_Proposal.tex
algoritmos revisados falta actualizar graficos

@@
 
-\noindent where $\mathbf{S}{t}$ and $\mathbf{W}{t}$ are define as:
+\noindent where $\mathbf{S}{t}$ and $\mathbf{W}{t}$ are defined as:
 
@@
 methods for geting model parameters: OLS, RR, AAR which leads to three different
-algorithms OVECM-OLS, OVECM-RR, OVECM-AAR:
+algorithms OVECM-OLS, OVECM-RR, OVECM-AAR.
+The proposal considers the following:
 
+\begin{itemize}
+\item The function \texttt{getJohansen} returns cointegration vectors given by the
+Johansen method considering the trace statistic test at 95\% level of
+significance. This procedure is called at the first step of the algorithm and
+every time in-sample MAPE ($\mathbf{e}$) exceeds threshold error defined (mean\_error).
+\item The function \texttt{vecMatrix} returns VECM
+matrices $\mathbf{A}(t),\mathbf{B}(t)$ shown in equation~(\ref{eq:notation}). This
+method is only required at the first step.
+\item The function \texttt{vecMatrixOnline} returns new rows $\mathbf{a}_t^\top$ and
+$\mathbf{b}_t^\top$ given new input data $\mathbf{y}_t$.
+\item The Solver class (see algorithm \ref{alg:solver}) obtain VECM parameters  
+using RR, AAR or OLS methods.
+\item The function \texttt{vecMatrixUpdate} updates matrix $\mathbf{A}$ when new
+cointegration vectors are required (matrix $\mathbf{B}$ is not affected). Model
+parameters matrix $\mathbf{X}$ is also updated later.
+\item The number of cointegration vectors is set as a parameter % agregar a alg
+\item RR parameter $\lambda$ was done by cross-validation. %agregar cap timeseries
+\end{itemize}
+
+\newpage
 \begin{algorithm}[ht]
@@
 
-The proposal considers the following:
-
-\begin{itemize}
-\item The function \texttt{getJohansen} returns cointegration vectors given by the
-Johansen method considering the trace statistic test at 95\% level of
-significance. This procedure is called at the first step of the algorithm and
-every time in-sample MAPE ($\mathbf{e}$) exceeds threshold error defined (mean\_error).
-\item The function \texttt{vecMatrix} returns VECM
-matrices $\mathbf{A}(t),\mathbf{B}(t)$ shown in equation~(\ref{eq:notation}). This
-method is only required at the first step.
-\item The function \texttt{vecMatrixOnline} returns new rows $\mathbf{a}_t^\top$ and
-$\mathbf{b}_t^\top$ given new input data $\mathbf{y}_t$.
-\item The Solver class (see algorithm \ref{alg:solver}) obtain VECM parameters  
-using RR, AAR or OLS methods.
-\item The function \texttt{vecMatrixUpdate} updates matrix $\mathbf{A}$ when new
-cointegration vectors are required (matrix $\mathbf{B}$ is not affected). Model
-parameters $\mathbf{X}$ is also updated later.
-%\item The number of cointegration vectors is set as a parameter (aun no lo
-%coloco en el algoritmo).
-%\item In order to set VECM parameter: $L$ and $p$, we use the Akaike Information
-%Criterion (AIC). RR parameter $\lambda$ was done by cross-validation.
-%\item Obtain VECM parameter $\mathbf{X}(t)$ in equation \ref{eq:optsolSLAAR} requires
-%to calculate the inverse of matrix $\mathbf{S}(t+1)$ which is an expensive
-%routine.However, since we already know $\mathbf{S}(t)^{-1}$ we can use
-%Sherman-Morrison-Woodbury twice~\ref{eq:SMW}.
-\end{itemize}
-
-\begin{algorithm}[ht]
+\newpage
+\begin{algorithm}[ht!]
 \begin{algorithmic}[1]
@@
 
+\newpage
 The proposal was compared against an optimal online algorithm which is a time varying
-VECM (OOVECM). Algorithm ~\ref{alg:OOVECM} shows this modification:
+VECM (OVECM-OPT). Algorithm ~\ref{alg:OOVECM} shows this modification:
 
-\begin{algorithm}[ht]
+\begin{algorithm}[ht!]
 \begin{algorithmic}[1]
@@
 \end{algorithmic}
-\caption{OOVECM: Optimal Online Vector Error Correction Model}
+\caption{OVECM-OPT: Optimal Online Vector Error Correction Model}
 \label{alg:OOVECM}

### 2017-07-23 src/4_OnlineLearning.tex
model updates before pseudoalgorithms
inicio revision capitulo 3x
last pdf
two references

@@
 algorithms based on learning from examples. This idea has become central to the
-design of search engines, robots systems and forecasts applications that
-process large data sets. Usually machines designed to forecast financial time
-series requires long training period and therefore they are not suitable to be updated
-with stream data. Online machine learning techniques tackle this problem by updating the model with new
-data or by computing a new model using less data so they can give a response in a
-short period of time.
+design of search engines, robots systems and forecasts applications that process
+large data sets. Usually machines designed to forecast financial time series
+requires long training period and therefore they are not suitable to be updated
+with stream data. Online machine learning techniques tackle this problem by
+updating the model with new data or by computing a new model using less data so
+they can give a response in a short period of time.
 
@@
 \section{Introduction}
-Machine Learning (ML) studies computer algorithms for learning
-something such as to complete a task, to make accurate predictions or to behave intelligently.
-The learning is always based on samples and the objective is about to do better in the future based on the past experiences in an automatic way.
-ML is a sub-area of artificial intelligence and broadly intersects with other fields such as statistics, mathematics, physics and computer science.
-There are many examples of machine learning problems: time series forecasting, image processing, face detection, spam filtering, weather prediction, search engines, among many others.
-
-ML models are often more accurate than what can be created through direct programming. The reason for this is that ML models are data driven and are able to examine large amounts of data.
-
-There are three types of ML classified depending on the nature of the learning input or output available to a learning system:
+Machine Learning (ML) studies computer algorithms for learning something such as
+to complete a task, to make accurate predictions or to behave intelligently. The
+learning is always based on samples and the objective is about to do better in
+the future based on the past experiences in an automatic way. ML is a sub-area
+of artificial intelligence and broadly intersects with other fields such as
+statistics, mathematics, physics and computer science.  There are many examples
+of machine learning problems: time series forecasting, image processing, face
+detection, spam filtering, weather prediction, search engines, among many
+others.
+
+ML models are often more accurate than what can be created through direct
+programming. The reason for this is that ML models are data driven and are able
+to examine large amounts of data.
+
+There are three types of ML classified depending on the nature of the learning
+input or output available to a learning system:
 
 \begin{description}
-\item[Supervised learning]  the input data is a tuple which contains the example input and its desired outputs (also called labels).
-The list of tuples is called the training set.
-The concept of supervised learning comes from the supervisor, acting as a teacher in the learning process.
-The goal is to learn a general rule that maps inputs to outputs optimising a target function.
-There are two related problem types in supervised learning: classification and regression problems \cite{bishop2006}.
-Its two mainstream approaches are: support vector machines (SVMs) \cite{vapnik1998} and ensemble learning \cite{breiman1998}.
-Furthermore, supervised learning can be categorised into offline or batch learning and online learning (see section \ref{sec:onoffline}).
-\item[Unsupervised learning] also known as clustering \cite{ben2005}.
-In this type of learning no labels are given and the system has to find a structure on its own, discovering hidden patterns in data.
+\item[Supervised learning]  the input data is a tuple which contains the example
+input and its desired outputs (also called labels). The list of tuples is called
+the training set. The concept of supervised learning comes from the supervisor,
+acting as a teacher in the learning process. The goal is to learn a general rule
+that maps inputs to outputs optimising a target function. There are two related
+problem types in supervised learning: classification and regression problems
+\cite{bishop2006}. Its two mainstream approaches are: support vector machines
+(SVMs) \cite{vapnik1998} and ensemble learning \cite{breiman1998}. Furthermore,
+supervised learning can be categorised into offline or batch learning and online
+learning (see section \ref{sec:onoffline}). 
+\item[Unsupervised learning] also known as clustering \cite{ben2005}. In this
+type of learning no labels are given and the system has to find a structure on
+its own, discovering hidden patterns in data.
 \item[Reinforcement learning] is the problem faced by an agent that must learn
-through trial-and-error interactions with a dynamic environment.
-It is based on programming agents by reward and punishment without the need to specify how the task is to be achieved \cite{sutton1998}.
+through trial-and-error interactions with a dynamic environment. It is based on
+programming agents by reward and punishment without the need to specify how the
+task is to be achieved \cite{sutton1998}.
 \end{description}
@@
 
-All ML problems can be viewed as optimisation problems. The ML core task is to define a learning criterion, i.e the function to be optimised.
-
-Supervised learning is most popular and most commonly used in modelling financial problems and the assessments of this method and their results in practice are fairly good.
-Therefore, in this thesis we will adhere to this trend and we will use supervised learning.
-
-Supervised learning consists in finding a learning function $f: X \rightarrow Y$ which models a set of examples, called training set $S$, which have been drawn randomly and independently according to a unknown joint distribution function $p(x,y)$ on $X \times Y$:
+All ML problems can be viewed as optimisation problems. The ML core task is to
+define a learning criterion, i.e the function to be optimised.
+
+Supervised learning is most popular and most commonly used in modelling
+financial problems and the assessments of this method and their results in
+practice are fairly good. Therefore, in this thesis we will adhere to this trend
+and we will use supervised learning.
+
+Supervised learning consists in finding a learning function $f: X \rightarrow Y$
+which models a set of examples, called training set $S$, which have been drawn
+randomly and independently according to a unknown joint distribution function
+$p(x,y)$ on $X \times Y$:
 
@@
 
-\noindent where $X \subseteq \mathbb{R}^m$ and $Y = \mathbb{R}$ in case of regression problems and $Y = \{1,-1\}$  if it is a classification problem.
-
+\noindent where $X \subseteq \mathbb{R}^m$ and $Y = \mathbb{R}$ in case of
+regression problems and $Y = \{1,-1\}$  if it is a classification problem.
+
+%=======================HASTA ACA REVISADO PROFE =========================
 A learning algorithm $\mathcal{A}$ takes as input a data set $S \in \mathscr{T}=\mathscr{P}(X\times Y)$ and selects from $\mathcal{H}$ a function $f$ such that $f(x)\approx y $ in a predictive way:
@@
 
-\subsection{Regularization}
-
-Regularization can be understood using two contexts: learning theory (probabilistic) and inverse problems (deterministic).
-In the context of learning, regularization refers to techniques allowing to avoid over-fitting and the desire property of the selected estimator is to perform well on new data (to generalize), e.g.
-regularized least squares.
-In the context of inverse problems, regularization objective is to stabilise, with respect to noise, a possibly ill-conditioned matrix inversion problem e.g spectral cut-off and Tikhonov regularization or ridge regression (RR)\cite{ tikhonov1977}.
-
-In particular, it is well known that regularization schemes such as RR or Tikhonov regularization can be effectively used in the context of learning \cite{vito2005}.
-
-Tikhonov introduced a regularisation which ensures well-posedness and generalisation of ERM, i.e prevents overfit, by constraining the hypothesis space $\mathcal{H}$ usually called regularised ERM or Tikhonov regularisation.
-
-Tikhonov regularisation minimised over the hypothesis space $\mathcal{H}$ for a fixed positive parameter $\lambda$ the following:
+\subsection{Regularization}\label{sec:regularization}
+
+Regularization can be understood using two contexts: learning theory
+(probabilistic) and inverse problems (deterministic). In the context of
+learning, regularization refers to techniques allowing to avoid over-fitting and
+the desire property of the selected estimator is to perform well on new data (to
+generalize), e.g. regularized least squares. In the context of inverse problems,
+regularization objective is to stabilise, with respect to noise, a possibly
+ill-conditioned matrix inversion problem e.g spectral cut-off and Tikhonov
+regularization or ridge regression (RR)\cite{ tikhonov1977}.
+
+In particular, it is well known that regularization schemes such as RR or
+Tikhonov regularization can be effectively used in the context of learning
+\cite{vito2005}.
+
+Tikhonov introduced a regularisation which ensures well-posedness and
+generalisation of ERM, i.e prevents overfit, by constraining the hypothesis
+space $\mathcal{H}$ usually called regularised ERM or Tikhonov regularisation.
+
+Tikhonov regularisation minimised over the hypothesis space $\mathcal{H}$ for a
+fixed positive parameter $\lambda$ the following:
 
@@
 
-Ridge regression was independently proposed by Tikhonov \cite{tikhonov1963} and Phillips \cite{phillips1962}. 
+Ridge regression was independently proposed by Tikhonov \cite{tikhonov1963} and
+Phillips \cite{phillips1962}. 
 One example of regularisation in linear models is Ridge Regression (RR), which is a regularised least squares method.
@@
 \begin{eqnarray*}
-Bias(\hat{f}(\hat{\mathbf{X}}) &=& E[\hat{\mathbf{X}}] - \mathbf{X} \\
+Bias(\hat{f}(\hat{\mathbf{X}})) &=& E[\hat{\mathbf{X}}] - \mathbf{X} \\
 &=& E[ (\mathbf{A}^\top \mathbf{A})^{-1}\mathbf{A}^\top \mathbf{Y}] - \mathbf{X} \\

### 2017-07-24 src/4_OnlineLearning.tex
avance correcciones cap 3

@@
 %=======================HASTA ACA REVISADO PROFE =========================
-A learning algorithm $\mathcal{A}$ takes as input a data set $S \in \mathscr{T}=\mathscr{P}(X\times Y)$ and selects from $\mathcal{H}$ a function $f$ such that $f(x)\approx y $ in a predictive way:
+A learning algorithm $\mathcal{A}$ takes as input a data set $S \in
+\mathscr{T}=\mathscr{P}(X\times Y)$ and selects from $\mathcal{H}$ a function
+$f$ such that $f(x)\approx y $ in a predictive way:
 
@@
 
-\noindent where $\mathcal{H}$ is called the {\em hypothesis space}, is the space of functions that the algorithm is allowed to search.
-The selection of $f$ is based on the minimization of the expected error of the loss function $V(f(x),y)$ defined as:
-
-\begin{eqnarray*}
- V(f(\cdot),\cdot): X\times Y &\rightarrow & Y \\
-  (x, y) &\rightarrow &V(f(x),y)
-\end{eqnarray*}
-
-\noindent $V(f(x),y)$ denote the price paid for mistakes. Therefore, $V(f(x),y)=0$ if $f(x)=y$.
+\noindent where $\mathcal{H}$ is called the {\em hypothesis space}, i.e. the space
+of functions that the algorithm is allowed to search. The selection of
+$f$ is based on the minimization of the expected error of the loss
+function $V(f(x),y): Y \times Y$ defined as:
+
+\begin{eqnarray*}
+ V & \rightarrow & \mathbb{R}^+ \\
+  (f(x), y) &\rightarrow & V(f(x),y)
+\end{eqnarray*}
+
+\noindent where $V(f(x),y)$ denote the price paid for mistakes, $f:X=\R^m\to Y$ and
+$y=f(x)\in\R$. Therefore, $V(f(x),y)=0$ if $f(x)=y$.
 
@@
 \label{eq:expetedrisk}
-E[V(f(\cdot),\cdot)] = \int_{X\times Y} V(f(x),y)\;dp(x,y)\,,
+E[V(f(x),y)] = \int_{X\times Y} V(f(x),y)\;dp(x,y)\,,
 \end{equation}
@@
 
+\begin{equation*}
+\hat{f} = \argmin_{f \in \mathcal{H}} R_{\text{emp}}[f] 
+\end{equation*}
+
+\noindent where,
+
 \begin{equation} 
@@
 
-The ERM is a Riemann sum approximation of the expected error of $f$ shown in equation \ref{eq:expetedrisk}, where $dp(x,y) \approx 1/n$.
+The ERM is a Riemann sum approximation of the expected error of $f$ shown in
+equation \ref{eq:expetedrisk}. By the law of large numbers it is known that:
+
+\begin{equation}
+\frac{1}{n} \sum_{i=1}^n V(f(x_i),y_i) \xrightarrow{n \rightarrow \infty}  E[V(f(x),y)]
+\end{equation}
 
 \subsection{Learning algorithms}
-In all learning algorithms that are trained from example data, there is a tradeoff \cite{dietterich2003} between three factors:
+In all learning algorithms that are trained from example data, there is a
+tradeoff \cite{dietterich2003} between three factors:
 \begin{description}
-\item[Complexity of the hypothesis class] for best generalisation, we should adjust the complexity of the learner model $\hat{f}$ to the complexity of the data $f$.
-The complexity of a learning problem mostly depends on the number of training examples and the size of the searched hypothesis space.
-In polynomial regression, the complexity parameter is the order of the fitted polynomial, and therefore we need to find a way to choose the best order that minimises the generalisation error, that is, tune the complexity of the model to best fit the complexity of the function inherent in the data.
-\item[Generalisation accuracy on new examples] is the capability of the algorithm to generate the right output for an input instance outside the training set.
-It is measured by the expected risk (equation \ref{eq:expetedrisk}).
-\item[The amount of training data]  in most cases, generalisation accuracy increases as the amount of training data increases.
+\item[Complexity of the hypothesis class] for best generalisation, we should
+adjust the complexity of the learner model $\hat{f}$ to the complexity of the
+data $f$. The complexity of a learning problem mostly depends on the number of
+training examples and the size of the searched hypothesis space. In polynomial
+regression, the complexity parameter is the order of the fitted polynomial, and
+therefore we need to find a way to choose the best order that minimises the
+generalisation error, that is, tune the complexity of the model to best fit the
+complexity of the function inherent in the data.
+\item[Generalisation accuracy on new examples] is the capability of the
+algorithm to generate the right output for an input instance outside the
+training set. It is measured by the expected risk (equation \ref{eq:expetedrisk}).
+\item[The amount of training data] in most cases, generalisation accuracy
+increases as the amount of training data increases.
 \end{description}

### 2017-07-25 src/4_OnlineLearning.tex
avance correcciones profe

@@
 \noindent where $V(f(x),y)$ denote the price paid for mistakes, $f:X=\R^m\to Y$ and
-$y=f(x)\in\R$. Therefore, $V(f(x),y)=0$ if $f(x)=y$.
+$y=f(x)\in\R$. Therefore, $V(f(x),y)=0$ if $\;f(x)=y$.
 
@@
 training examples and the size of the searched hypothesis space. In polynomial
-regression, the complexity parameter is the order of the fitted polynomial, and
+regression, the complexity parameter is the degree of the fitted polynomial, and
 therefore we need to find a way to choose the best order that minimises the
@@
 
-Very complex models will fit the training data better (low bias) but it could overfit it and generalise poorly (high variance), this is also called the bias-variance tradeoff.
+A very complex models will fit the training data better (low bias) but it could overfit it and generalise poorly (high variance), this is also called the bias-variance tradeoff.
 Generalisation accuracy can sometimes be improved with more training data.

### 2017-07-28 src/4_OnlineLearning.tex
complexity of hypothesis

--- 2017-07-24 src/4_OnlineLearning.tex
@@
 
-%=======================HASTA ACA REVISADO PROFE =========================
 A learning algorithm $\mathcal{A}$ takes as input a data set $S \in
@@
 
+%=======================HASTA ACA REVISADO PROFE =========================
 \subsection{Learning algorithms}
 In all learning algorithms that are trained from example data, there is a
-tradeoff \cite{dietterich2003} between three factors:
+tradeoff between three factors \cite{dietterich2003}:
 \begin{description}
-\item[Complexity of the hypothesis class] for best generalisation, we should
-adjust the complexity of the learner model $\hat{f}$ to the complexity of the
-data $f$. The complexity of a learning problem mostly depends on the number of
-training examples and the size of the searched hypothesis space. In polynomial
-regression, the complexity parameter is the order of the fitted polynomial, and
+\item[Complexity or capacity of the hypothesis class]
+
+and can be measured by the Vapnik\-Chervonenkis (VC) dimension
+
+for best generalisation, we should
+adjust the complexity of the learner model $\hat{f}$ o the complexity of the
+data.
+
+In polynomial regression, the complexity parameter is the order of the fitted polynomial, and
 therefore we need to find a way to choose the best order that minimises the
@@
 complexity of the function inherent in the data.
+
+The complexity of a learning problem mostly depends on the number of
+training examples and the size of the searched hypothesis space.
+
 \item[Generalisation accuracy on new examples] is the capability of the

### 2017-07-28 src/4_OnlineLearning.tex

@@
 \item[Complexity or capacity of the hypothesis class] 
-
-and can be measured by the Vapnik\-Chervonenkis (VC) dimension
+ can be measured by the Vapnik\-Chervonenkis (VC) dimension
 

### 2017-08-02 src/4_OnlineLearning.tex
Complexity of the hypothesis space

@@
 \begin{description}
-\item[Complexity or capacity of the hypothesis class] 
- can be measured by the Vapnik\-Chervonenkis (VC) dimension
-
-for best generalisation, we should
-adjust the complexity of the learner model $\hat{f}$ o the complexity of the
-data.
-
-In polynomial regression, the complexity parameter is the order of the fitted polynomial, and
-therefore we need to find a way to choose the best order that minimises the
-generalisation error, that is, tune the complexity of the model to best fit the
-complexity of the function inherent in the data.
-
-The complexity of a learning problem mostly depends on the number of
-training examples and the size of the searched hypothesis space.
+\item[Complexity or capacity of the hypothesis class] is defined in terms of the
+Vapnik-Chervonenkis (VC) dimension, which corresponds to the number of training
+points that can be classified exactly by the hypothesis space. More formally,
+the VC dimension of the hypothesis space $\mathcal{H}$ defined over instance
+space $X$ is the size of the largest finite subset of $X$ shattered by
+$\mathcal{H}$. The dataset $S$ is shattered by hypothesis space $\mathcal{H}$ if
+and only if for every dichotomy of $S$ there exists some hypothesis in
+$\mathcal{H}$ consistent with this dichotomy. A dichotomy of a set $S$ is a
+partition of $S$ into two disjoint subsets \cite{vapnik2013nature}.
+
+For example, the VC dimension of the lines in the plane is explained in figure
+\ref{fig:vcdimension}:
+
+\begin{figure}[!h]
+  \centering
+  \includegraphics[width=0.8\textwidth]{img/VCdimension}
+  \caption{VC dimension example from \cite{vapnik2013nature}. The left figure
+  shows that three vectors can be shattered by a line in a plane. However, if we
+  add an extra vector $z_4$, it is impossible to shatter them with a
+  plane.Therefore, linear classifiers has VC dimension equals to $d+1$ where $d$
+  is the dimension of the instance space $X$.}
+  \label{fig:vcdimension}
+\end{figure}
 
@@
 
-A very complex models will fit the training data better (low bias) but it could overfit it and generalise poorly (high variance), this is also called the bias-variance tradeoff.
-Generalisation accuracy can sometimes be improved with more training data.
-The figure \ref{fig:tradeoff} illustrates how generalisation accuracy depends on the complexity of the model and the amount of training data.
+A very complex models will fit the training data better (low bias) but it could
+overfit it and generalise poorly (high variance), this is also called the
+bias-variance tradeoff.  Generalisation accuracy can sometimes be improved with
+more training data. The figure \ref{fig:tradeoff} illustrates how generalisation
+accuracy depends on the complexity of the model and the amount of training data.
 

### 2017-08-06 src/4_OnlineLearning.tex
Siguiente comentario P37 L17

@@
 
-%=======================HASTA ACA REVISADO PROFE =========================
 \subsection{Learning algorithms}
@@
 Models learning error can be split into two main components: error due to bias and error due to variance.
-Bias measures how far off in general these models' predictions are from the correct value.
-Variance is taken as the variability of a model prediction for a given data point.
-
-If we have a learning model $\hat{f}$, its expected generalisation error on an unseen or testing sample $x$ can be decomposed as follows:
+Bias measures the prediction error of the model $\hat{f}$ from the real value, see equation \ref{eq:biastrade}.
+Variance is taken as the variability of a model prediction for a given data point or its sensitivity to small fluctuations in the input data, see equation \ref{eq:variancetrade}.
+
+If we have a learning model $\hat{f}$, its expected generalisation error on a testing sample $x$ can be decomposed as shown in equation \ref{eq:bvtrade} (demo in \cite{geman1992neural}):
 
@@
 \noindent where:
-\begin{align*}
- \mathrm{Bias}(\hat{f}(x)) &= \mathrm{E}\big[\hat{f}(x)\big] - f(x) \\
- \mathrm{Var}(\hat{f}(x)) &= \mathrm{E}\Big[ \big( \hat{f}(x) - \mathrm{E}[\hat{f}(x)] \big)^2 \Big] 
-\end{align*}
-
-\textbf{Demo}\quad
-
-\begin{eqnarray*}
-E[(y-\hat{f}(\mathbf{x}))^2] & =&
-E[(y-f(\mathbf{x})+f(\mathbf{x})-\hat{f}(\mathbf{x}))^2] \\
-&=& E[(B-f(\mathbf{x}))^2] +
-E[(f(\mathbf{x})-\hat{f}(\mathbf{x}))^2] + \dots \\
-& &
-2\cancelto{0}{E[y-f(\mathbf{x})]}E[f(\mathbf{x})-\hat{f}(\mathbf{x})] \\
-&=& \sigma^2 + MSE(\hat{f}(\mathbf{x}))
-\end{eqnarray*}
-
-\noindent where
-
-\begin{eqnarray*}
-    MSE(\hat{f}(\mathbf{x})) &=&
-    E[f(\mathbf{x})-\hat{f}(\mathbf{x}))]^2 \\
-    &=& E[f(\mathbf{x})-E[\hat{f}(\mathbf{x})] +
-    E[\hat{f}(\mathbf{x})]-\hat{f}(\mathbf{x}))]^2 \\
-    &=&  E[(f(\mathbf{x})-E[\hat{f}(\mathbf{x})])^2] +
-      E[(\hat{f}(\mathbf{x})-E[\hat{f}(\mathbf{x})])^2] + \dots\\
-    & & E[(f(\mathbf{x})-E[\hat{f}(\mathbf{x})])] 
-    \cancelto{0}{E[(\hat{f}(\mathbf{x})-E[\hat{f}(\mathbf{x})])]} \\
-    &=& Bias(\hat{f}(\mathbf{x}))^2 + Var(\hat{f}(\mathbf{x}))
-\end{eqnarray*}
-
-$\blacksquare$
-
-Bias is introduced by the model selection.
-Therefore the model building process is repeated (through resampling) and substantially different averages of prediction values are obtained, bias will be high.
+\begin{align}
+ \mathrm{Bias}(\hat{f}(x)) &= \mathrm{E}\big[\hat{f}(x)\big] - f(x) \label{eq:biastrade}\\
+ \mathrm{Var}(\hat{f}(x)) &= \mathrm{E}\Big[ \big( \hat{f}(x) - \mathrm{E}[\hat{f}(x)] \big)^2 \Big]  \label{eq:variancetrade}
+\end{align}
+
+The bias shown in equation \ref{eq:biastrade} can be obtain through resampling of the input data $x$, therefore an average of the predicted values can be calculated.
+If these prediction values are substantially different to the true value, the bias will be high.
+
 The error due to variance is the amount by which the prediction, over one training set, differs from the expected predicted value, over all the training sets.
-Variance measures how inconsistent are the predictions from one another, over different training sets, not whether they are accurate or not.
-
-This generalisation error is unknown, but it can be estimated using a training sample, a testing sample or by estimating the model complexity.
+Variance measures how inconsistent are the predictions from one another, over different training sets, not whether they are accurate or not \cite{gutierrez2015machine}.
+%=======================HASTA ACA REVISADO PROFE =========================
+
+The generalisation error is unknown, but it can be estimated using a training sample, a testing sample or by estimating the model complexity.
 Figure \ref{fig:traintesterror} shows prediction error for training and testing sample.
@@
   \includegraphics[width=0.8\textwidth]{img/model_complexity}
-  \caption{Training and test error}
+  \caption{Training and test error. Image source is \cite{friedman2001elements}}
   \label{fig:traintesterror}

### 2017-08-07 src/4_OnlineLearning.tex
REVISAR DESDE P38 L6

@@
 of functions that the algorithm is allowed to search. The selection of
-$f$ is based on the minimization of the expected error of the loss
+$f$ is based on the minimisation of the expected error of the loss
 function $V(f(x),y): Y \times Y$ defined as:
@@
 \end{equation}
-\noindent the choice of loss function here gives rise to several well-known learning algorithms such as regularised least squares and support vector machines. %verificar esto
-
-Since the true distribution is unknown and only training samples are available, the objective is to estimate a function $\hat{f}$ through empirical risk (training error) minimisation (ERM):
+\noindent the choice of loss function here gives rise to several well-known
+learning algorithms such as regularised least squares and support vector
+machines. %verificar esto
+
+Since the true distribution is unknown and only training samples are available,
+the objective is to estimate a function $\hat{f}$ through empirical risk
+(training error) minimisation (ERM):
 
@@
 
-A very complex models will fit the training data better (low bias) but it could
-overfit it and generalise poorly (high variance), this is also called the
-bias-variance tradeoff.  Generalisation accuracy can sometimes be improved with
-more training data. The figure \ref{fig:tradeoff} illustrates how generalisation
+The relationship between generalisation error and the amount of training or
+input data is intuitive: the more data is given to the learning model $f$, the
+more evidence the algorithm has about the problem. In the limit, the input data
+will contain every possible example, so the algorithm will generalize perfectly.
+However, for some $f \in H$, the best hypothesis $\hat{f}$ might be far away to
+the best hypothesis class but it is the best that all the learners can hope.
+Therefore, to choose the right hypothesis class $H$ is crucial. The error of
+this best hypothesis is formalised as the {\em approximation or generalisation
+error} \cite{niyogi1996relationship}.
+
+Additionally, we do not have infinite data but only some finite random sample
+set to find an hypothesis, the additional error caused by finiteness of the
+data, is called {\em estimation error}. The amount of data needed to ensure a
+small {\em estimation error} is referred to as {\em sample complexity} of the
+problem.
+
+The relationship between the generalisation error and the hypothesis complexity
+is less intuitive. If the class $H$ complexity (VC dimension) increases, then to
+mantain an {\em estimation error}, the {\em sample complexity} increases.
+Therefore, if we have a low complex model we will reduce the {\em sample complexity}
+but it will increase {\em generalisation error}. The hypothesis has to be
+sufficiently complex to capture the characteristics of the data.
+
+The figure \ref{fig:tradeoff} illustrates how generalisation
 accuracy depends on the complexity of the model and the amount of training data.
@@
   \includegraphics[width=0.8\textwidth]{img/3tradeoff}
-  \caption{Tradeoff in empirical learning from \cite{dietterich2003}.
-The generalization accuracy is not affected by the amount of examples for low complex classifiers.
-However, for some more complex classifiers, the number of examples improves the generalization accuracy.
-The key is to find the best generalization accuracy and the lowest complex classifier as possible.
-}
+  \caption{Tradeoff in empirical learning from \cite{dietterich2003}. The
+generalization accuracy is not affected by the amount of examples for low
+complex classifiers. However, for some more complex classifiers, the number of
+examples improves the generalization accuracy. The key is to find the best
+generalization accuracy and the lowest complex classifier with the available
+training data. }
   \label{fig:tradeoff}
@@
 
-Models learning error can be split into two main components: error due to bias and error due to variance.
-Bias measures the prediction error of the model $\hat{f}$ from the real value, see equation \ref{eq:biastrade}.
-Variance is taken as the variability of a model prediction for a given data point or its sensitivity to small fluctuations in the input data, see equation \ref{eq:variancetrade}.
-
-If we have a learning model $\hat{f}$, its expected generalisation error on a testing sample $x$ can be decomposed as shown in equation \ref{eq:bvtrade} (demo in \cite{geman1992neural}):
+A very complex models will fit the training data better (low bias) but it could
+overfit it and generalise poorly (high variance), this is also called the
+bias-variance tradeoff.
+
+Figure \ref{fig:traintesterror} shows {\em prediction or estimation error} for
+training and testing sample.  In general, in the training sample prediction
+error is decreasing with more complex models but it overfit the data and it is
+not a good model for the test data.
+
+\begin{figure}[!h]
+  \centering
+  \includegraphics[width=0.8\textwidth]{img/model_complexity}
+  \caption{Training and test error. Image source is
+\cite{friedman2001elements}.The blue curve represents the training error which
+decreases with more complex models. However, the model starts to overfit the
+data reducing the generalisation capacity of the model,this occurs when the
+prediction error in the test sample starts to increase (red curve).}
+  \label{fig:traintesterror}
+\end{figure}
+
+Models learning error can be split into two main components: error due to bias
+and error due to variance. Bias measures the prediction error of the model
+$\hat{f}$ from the real value, see equation \ref{eq:biastrade}. Variance is
+taken as the variability of a model prediction for a given data point or its
+sensitivity to small fluctuations in the input data, see equation
+\ref{eq:variancetrade}.
+
+If we have a learning model $\hat{f}$, its expected generalisation error on a
+testing sample $x$ can be decomposed as shown in equation \ref{eq:bvtrade} (demo
+in \cite{geman1992neural}):
 
@@
 
-The bias shown in equation \ref{eq:biastrade} can be obtain through resampling of the input data $x$, therefore an average of the predicted values can be calculated.
-If these prediction values are substantially different to the true value, the bias will be high.
-
-The error due to variance is the amount by which the prediction, over one training set, differs from the expected predicted value, over all the training sets.
-Variance measures how inconsistent are the predictions from one another, over different training sets, not whether they are accurate or not \cite{gutierrez2015machine}.
-%=======================HASTA ACA REVISADO PROFE =========================
-
-The generalisation error is unknown, but it can be estimated using a training sample, a testing sample or by estimating the model complexity.
-Figure \ref{fig:traintesterror} shows prediction error for training and testing sample.
-In general, in the training sample prediction error is decreasing with more complex models but it overfit the data and it is not a good model for the test data.
-Figure \ref{fig:bvtradeoff} shows how testing error can be decomposed into bias and variance components.
-The best model will be the one has a balance between bias and variance.
-
-\begin{figure}[!h]
-  \centering
-  \includegraphics[width=0.8\textwidth]{img/model_complexity}
-  \caption{Training and test error. Image source is \cite{friedman2001elements}}
-  \label{fig:traintesterror}
-\end{figure}
-
-The objective is to simultaneously reduce bias and variance as much as possible in order to obtain as accurate model as is feasible.
-However, there is a tradeoff to be made when selecting models.
-Models that exhibit small variance and high bias underfit the truth target.
-Models that exhibit high variance and low bias overfit the truth target.
-
-This bias-variance tradeoff is the reason why in order to obtain the best model, data is usually broken in three subsets: training set, validation set and testing set.
-Training set is used to determine the model, validation set is used to estimate the generalisation error and finally testing set is used to estimate the accuracy of the model.
-Usually the partition is 50\% for training set, 25\% for validation and 25\% for testing purposes.
-This procedure can be extended to a Cross Validation (CV) procedure \cite{geisser1975}, usually used when the amount of data is limited.
-Various splitting strategies lead to various CV estimates.
-In K-fold cross-validation the training data is divided randomly into K distinct subsets, then the network is trained using K-1 subsets, and tested on the remaining subset.
-The process of training and testing is then repeated for each of the K possible choices of the subset omitted from the training.
-The average performance on the K omitted subsets is then our estimate of the generalisation performance.
+Figure \ref{fig:bvtradeoff} shows how
+testing error can be decomposed into bias and variance components. 
+The bias shown in equation \ref{eq:biastrade} can be obtain through resampling
+of the input data $x$, therefore an average of the predicted values can be
+calculated. If these prediction values are substantially different to the true
+value, the bias will be high.
 
@@
 
-Another way to control the complexity of the model is to include in the optimisation function not only the generalisation error but a penalisation of high complex models.
-
+The error due to variance is the amount by which the prediction, over one
+training set, differs from the expected predicted value, over all the training
+sets. Variance measures how inconsistent are the predictions from one another,
+over different training sets, not whether they are accurate or not
+\cite{gutierrez2015machine}.
+
+This bias-variance tradeoff is the reason why, in order to obtain the best model,
+data is usually broken in three subsets: training set, validation set and
+testing set. Training set is used to determine the model, validation set is used
+to estimate the generalisation error and finally testing set is used to estimate
+the accuracy of the model. Usually the partition is 50\% for training set, 25\%
+for validation and 25\% for testing purposes. This procedure can be extended to
+a Cross Validation (CV) procedure \cite{geisser1975}, usually used when the
+amount of data is limited. Various splitting strategies lead to various CV
+estimates. In K-fold cross-validation the training data is divided randomly into
+K distinct subsets, then the network is trained using K-1 subsets, and tested on
+the remaining subset. The process of training and testing is then repeated for
+each of the K possible choices of the subset omitted from the training. The
+average performance on the K omitted subsets is then our estimate of the
+generalisation performance.
+
+Another way to control the complexity of the model is to include in the
+optimisation function not only the generalisation error but a penalisation of
+high complex models.
+
+%=======================HASTA ACA REVISADO PROFE =========================
 \subsection{Regularization}\label{sec:regularization}

### 2017-08-12 src/4_OnlineLearning.tex
comentarios cap 3 finalizados

@@
 
-\begin{eqnarray*}
+\begin{eqnarray}
+\label{eq:lossfunction}
  V & \rightarrow & \mathbb{R}^+ \\
   (f(x), y) &\rightarrow & V(f(x),y)
-\end{eqnarray*}
+\end{eqnarray}
 
@@
 If we have a learning model $\hat{f}$, its expected generalisation error on a
-testing sample $x$ can be decomposed as shown in equation \ref{eq:bvtrade} (demo
+testing sample $x$ can be decomposed as shown in equation \ref{eq:bvtrade} (proof
 in \cite{geman1992neural}):
@@
 
-This bias-variance tradeoff is the reason why, in order to obtain the best model,
-data is usually broken in three subsets: training set, validation set and
-testing set. Training set is used to determine the model, validation set is used
-to estimate the generalisation error and finally testing set is used to estimate
-the accuracy of the model. Usually the partition is 50\% for training set, 25\%
-for validation and 25\% for testing purposes. This procedure can be extended to
-a Cross Validation (CV) procedure \cite{geisser1975}, usually used when the
-amount of data is limited. Various splitting strategies lead to various CV
-estimates. In K-fold cross-validation the training data is divided randomly into
-K distinct subsets, then the network is trained using K-1 subsets, and tested on
+\subsection{Generalised Cross Validation}
+This bias-variance tradeoff is the reason why the 
+data is usually divided in three subsets: training, validation and
+testing set. The training set is used to construct the model, the validation set
+to estimate the generalisation error and finally the testing set to estimate
+the accuracy of the model. Usually the partition is 50\% for training, 25\%
+for validation and 25\% for testing purposes. When the
+amount of data is limited, this procedure can be extended to
+a Cross Validation (CV) approach \cite{geisser1975} . Various splitting strategies lead to various CV
+estimates. In K-fold cross-validation  \cite{stone1974} the training data is divided randomly into
+K distinct subsets, then the model is trained using K-1 subsets, and tested on
 the remaining subset. The process of training and testing is then repeated for
 each of the K possible choices of the subset omitted from the training. The
-average performance on the K omitted subsets is then our estimate of the
+average performance on the K omitted subsets is the estimate of the
 generalisation performance.
 
-Another way to control the complexity of the model is to include in the
-optimisation function not only the generalisation error but a penalisation of
-high complex models.
-
-%=======================HASTA ACA REVISADO PROFE =========================
+%Another way to control the complexity of the model is to include in the
+%optimisation function not only the generalisation error but a penalisation of
+%high complex models.
+
+%Since we don't know the data we want to predict, the bias-variance tradeoff is not directly useful to determine $\lambda$.
+Many techniques have been developed to determine a suitable value for the regularization parameter.
+However, when no information is available on the error, one of the most popular techniques is the Generalized Cross Validation (GCV) \cite{bauer2011}.
+GCV originates from the older method leave-one-out Cross Validation.
+
+%The leave-one-out cross validation method determine a regularization parameter that minimises the average of the squared predictions errors using each solution to predict the missing data value.
+
+%'---------------------------http://robjhyndman.com/hyndsight/crossvalidation/
+%When the data are not independent cross-validation becomes more difficult as leaving out an observation does not remove all the associated information due to the correlations with other observations.
+
+%For time series forecasting, a cross-validation statistic is obtained as
+%follows: fit the model to the data $y1,\dots,yt$ and let $y^t+1$ denote the forecast
+%of the next observation. Then compute the error $et+1=yt+1-y^t+1$ for the
+%forecast observation.  Repeat step 1 for $t=m,\dots,n−1$ where m is the minimum number
+%of observations needed for fitting the model.
+%Compute the MSE from $em+1∗,…,en∗$.
+
+%An excellent and comprehensive recent survey of cross-validation results is Arlot and Celisse (2010)
+
+%---------------------------------------------------
+
 \subsection{Regularization}\label{sec:regularization}
@@
 ill-conditioned matrix inversion problem e.g spectral cut-off and Tikhonov
-regularization or ridge regression (RR)\cite{ tikhonov1977}.
+regularization or ridge regression (RR) \cite{ tikhonov1977}.
 
@@
 
-Tikhonov introduced a regularisation which ensures well-posedness and
+Tikhonov introduced a regularization which ensures well-posedness and
 generalisation of ERM, i.e prevents overfit, by constraining the hypothesis
@@
 
-Tikhonov regularisation minimised over the hypothesis space $\mathcal{H}$ for a
-fixed positive parameter $\lambda$ the following:
+%Tikhonov regularisation minimised over the hypothesis space $\mathcal{H}$ for a
+%fixed positive parameter $\lambda$ the following:
+Tikhonov regularisation considers a functional of the form:
 
@@
 \label{eq:rerm}
-R_{\text{emp}}[f] = \frac{1}{n} \sum_{i=1}^n V(f(x),y) + \lambda \mathcal{R}(f)
-\end{equation}
-
-\noindent where $\mathcal{R}(f)$ is the regulariser, a penalisation on $f$.
+R_{\text{emp}}[f] = \frac{1}{n} \sum_{i=1}^n V(f(x_i),y_i) + \lambda \mathcal{R}(f)
+\end{equation}
+
+\noindent where $X \subseteq \mathbb{R}^m$ and $Y = \mathbb{R}$ , $V$ is the loss function defined in the equation \ref{eq:lossfunction} and $\mathcal{R}(f)$ is the regulariser, a penalisation on $f$.
+One example of regularisation in linear models is Ridge Regression (RR), which is a regularised least squares method.
 
@@
 
-Ridge regression was independently proposed by Tikhonov \cite{tikhonov1963} and
-Phillips \cite{phillips1962}. 
-One example of regularisation in linear models is Ridge Regression (RR), which is a regularised least squares method.
-The least squares (LS) method is a well known way to solve a regression
-problem. 
-LS method consists of minimising the sum of squared errors:
+Ridge regression (RR) was independently proposed by Tikhonov \cite{tikhonov1963} and
+Phillips \cite{phillips1962}. RR corresponds to a regularised least squares method. The least squares (LS) method is a well known way to solve a regression
+problem. LS method consists of minimising the sum of squared errors:
 
@@
  J(\mathbf{X}) =& \displaystyle \sum_{t=1}^N (f(\mathbf{x}_t)-y_t)^2 \\
- =&\displaystyle \sum_{t=1}^N (\mathbf{\mathbf{X}}^\top {\mathbf{x}}_t-y_t)^2 \\
+ =&\displaystyle \sum_{t=1}^N (\mathbf{\mathbf{A}}^\top {\mathbf{x}}_t-y_t)^2 \\
  =& \| \mathbf{A}\mathbf{\mathbf{X}} - \mathbf{Y} \|_2^2 
@@
 
-\noindent which is equivalent to find a solution to 
+\noindent where $f(\mathbf{x}_t)={\mathbf{A}}^\top {\mathbf{x}}_t$. To solve equation \ref{eq:lsm} is equivalent to find a solution of:
 \begin{equation}
@@
 
-\textbf{Demo}\quad
-
-To solve equation~(\ref{eq:pseudoinverse}) is equivalent to solve the following
-optimisation problem:
-
-\begin{equation}
-\label{eq:RRproblem2}
-\underset{\mathbf{X}(\lambda)}{\text{min}} \quad \|
-\mathbf{C}\mathbf{\mathbf{X}} - \mathbf{F} \|_2^2
-\end{equation}
-
-\noindent where
-
-\begin{equation*}
-	\mathbf{C}=
-	\begin{bmatrix} \quad &\mathbf{A}& \quad \\ \hdashline \quad
-& \sqrt\lambda\mathbb{I}&\quad  
-\end{bmatrix} 
-\quad \text{and} \quad
-	\mathbf{F}=\begin{bmatrix} \quad &\mathbf{Y}& \quad \\ \hdashline \quad
-& \mathbf{0}&\quad  \end{bmatrix} \, .
-\end{equation*}
-
-Applying equation~(\ref{eq:MP}) and considering that $\mathbf{C}^\top
-\mathbf{C} = \mathbf{A}^\top \mathbf{A} + \lambda \mathbb{I}$ and 
-$\mathbf{C}^\top \mathbf{F}=\mathbf{A}^\top \mathbf{Y} $ we have:
-
-\begin{eqnarray*}
-\mathbf{X}(\lambda)&=&(\mathbf{C}^\top
-\mathbf{C})^{-1}\mathbf{C}^\top \mathbf{F} \\
-&=& (\mathbf{A}^\top \mathbf{A} + \lambda \mathbb{I})^{-1} \mathbf{A}^\top \mathbf{Y}
-\end{eqnarray*}
-
-$\blacksquare$
+%\textbf{Demo}\quad
+%
+%To solve equation~(\ref{eq:pseudoinverse}) is equivalent to solve the following
+%optimisation problem:
+%
+%
+%\begin{equation}
+%\label{eq:RRproblem2}
+%\underset{\mathbf{X}(\lambda)}{\text{min}} \quad \|
+%\mathbf{C}\mathbf{\mathbf{X}} - \mathbf{F} \|_2^2
+%\end{equation}
+%
+%\noindent where
+%
+%\begin{equation*}
+%	\mathbf{C}=
+%	\begin{bmatrix} \quad &\mathbf{A}& \quad \\ \hdashline \quad
+%& \sqrt\lambda\mathbb{I}&\quad  
+%\end{bmatrix} 
+%\quad \text{and} \quad
+%	\mathbf{F}=\begin{bmatrix} \quad &\mathbf{Y}& \quad \\ \hdashline \quad
+%& \mathbf{0}&\quad  \end{bmatrix} \, .
+%\end{equation*}
+%
+%Applying equation~(\ref{eq:MP}) and considering that $\mathbf{C}^\top
+%\mathbf{C} = \mathbf{A}^\top \mathbf{A} + \lambda \mathbb{I}$ and 
+%$\mathbf{C}^\top \mathbf{F}=\mathbf{A}^\top \mathbf{Y} $ we have:
+%
+%\begin{eqnarray*}
+%\mathbf{X}(\lambda)&=&(\mathbf{C}^\top
+%\mathbf{C})^{-1}\mathbf{C}^\top \mathbf{F} \\
+%&=& (\mathbf{A}^\top \mathbf{A} + \lambda \mathbb{I})^{-1} \mathbf{A}^\top \mathbf{Y}
+%\end{eqnarray*}
+%
+%$\blacksquare$
 
@@
 
-\textbf{Demo}\quad
+\textbf{Proof}\quad
 
@@
 singularity of the covariance matrix $\mathbf{A}^\top \mathbf{A}$. This addition ensures that all of the covariance matrix eigenvalues will be strictly greater than 0, i.e the solution becomes unique.
+%=======================HASTA ACA REVISADO PROFE =========================
 
@@
 $\blacksquare$
-
-\subsection{Generalized Cross Validation}
-Since we don't know the data we want to predict, the bias-variance tradeoff is not directly useful to determine $\lambda$.
-Many techniques have been developed to determine a suitable value for the regularization parameter.
-However, when no information is available on the error, one of the most popular techniques is the Generalized Cross Validation (GCV) \cite{bauer2011}.
-GCV originates from the older method leave-one-out Cross Validation \cite{stone1974}.
-
-The leave-one-out cross validation method determine a regularization parameter that minimises the average of the squared predictions errors using each solution to predict the missing data value.
-
-%'---------------------------http://robjhyndman.com/hyndsight/crossvalidation/
-%When the data are not independent cross-validation becomes more difficult as leaving out an observation does not remove all the associated information due to the correlations with other observations.
-
-%For time series forecasting, a cross-validation statistic is obtained as
-%follows: fit the model to the data $y1,\dots,yt$ and let $y^t+1$ denote the forecast
-%of the next observation. Then compute the error $et+1=yt+1-y^t+1$ for the
-%forecast observation.  Repeat step 1 for $t=m,\dots,n−1$ where m is the minimum number
-%of observations needed for fitting the model.
-%Compute the MSE from $em+1∗,…,en∗$.
-
-%An excellent and comprehensive recent survey of cross-validation results is Arlot and Celisse (2010)
-
-%---------------------------------------------------
 
@@
 
-\newpage
-
 \section{Evaluation methods} \label{sec:evaluation}

### 2017-08-14 src/2_Finance.tex
avance cap 1

@@
 High frequency trading (HFT) strategies are based on buy and sell assets in
-short periods of time gaining a small profit in every transaction despite the transaction costs. The key is
-the amount of transactions that HFT algorithms are capable to execute. 
-In 2012 more than 50\% of US market trades were HFT.  HFT has
-motivated computer-driven strategies capable of processing large amount of data
-in short periods of time.
+short periods of time gaining a small profit in every transaction despite the
+transaction costs. The key is the amount of transactions that HFT algorithms are
+capable to execute.  In 2012 more than 50\% of US market trades were HFT.  HFT
+has motivated computer-driven strategies capable of processing large amount of
+data in short periods of time.
 
@@
 
-HFT is not a strategy but a technology that implements different
-trading strategies. The aim of HFT is to benefit from market short-term pricing
+HFT is not a strategy but a technology that implements different trading
+strategies. The aim of HFT is to benefit from market short-term pricing
 inefficiencies~\cite{chlistalla2011}. HFT is characterised for the use of
@@
 
-High frequency trades are short-term positions that commonly end the trading
-day avoiding leaving opened positions to the next business day. HFT is
-frequently associated to algorithm trading strategies, but the former is focused
-in to reduce the market impact of large institutional positions and the later refers to trade execution strategies that are typically used by fund managers to buy or sell large amounts of assets.
-The duration of the positions in HFT it is not well-defined.
-Figure~\ref{fig:HFTtimes} shows a survey done to traders about the holding time
-of a position opened. Most of them agreed that HFT refers to positions between
-1-second and 10-minutes. Overnight positions are avoided since they are
-riskier and also have more expensive fees, HFT firms end the trading day closing all
-opened positions.
+High frequency trades are short-term positions that commonly end the trading day
+avoiding leaving opened positions to the next business day. HFT is frequently
+associated to algorithm trading strategies, but the former is focused into
+reduce the market impact of large institutional positions and the later refers
+to trade execution strategies that are typically used by fund managers to buy or
+sell large amounts of assets. The duration of the positions in HFT it is not
+well-defined. Figure~\ref{fig:HFTtimes} shows a survey done to traders about the
+holding time of a position opened. Most of them agreed that HFT refers to
+positions between 1-second and 10-minutes. Overnight positions are avoided since
+they are riskier and also have more expensive fees, HFT firms end the trading
+day closing all opened positions.
 
@@
   \includegraphics[width=\textwidth]{img/HFTtradestime}
-  \caption{Holding time of an opened position of a high frequency trade.}
+  \caption{Holding time of an opened position of a high frequency trade.Source
+  \cite{aldridge2009}}
   \label{fig:HFTtimes}
 \end{figure}
-In 2014 HFT represented nearly 50\% of equity trades in the US and more than 20\%
-in Europe showing a consistent fall since 2009 which was its best year. US market revenues
-have fallen dramatically.
-Figure~\ref{fig:HFTmarket} on the left shows HFT trades percentage of US and Europe equity shares. The right figure shows how revenues in US stocks have fallen the last years.
+In 2014 HFT represented nearly 50\% of equity trades in the US and more than
+20\% in Europe showing a consistent fall since 2009 which was its best year. US
+market revenues have fallen dramatically.  Figure~\ref{fig:HFTmarket} on the
+left shows HFT trades percentage of US and Europe equity shares. The right
+figure shows how revenues in US stocks have fallen the last years.
 

### 2017-08-18 src/2_Finance.tex
Fin revision cap 1

@@
   \caption{Holding time of an opened position of a high frequency trade.Source
-  \cite{aldridge2009}}
+  \cite{aldridge2009}.}
   \label{fig:HFTtimes}
@@
   \caption{Left figure shows HFT market share in the US and Europe. Right figure
-  shows revenue in the US.}
+  shows revenue in the US. Source: TABB Group.}
   \label{fig:HFTmarket}
@@
 Capital markets include the stock or equity market and the bond or debt market
-and their movements are the most widely followed.
+and their movements are the most widely followed \cite{aldridge2009}.
 
@@
   \includegraphics[width=0.7\textwidth]{img/capitalmarkets}
-  \caption{Old structure of capital markets.}
+  \caption{Old structure of capital markets. Source \cite{aldridge2009}.}
   \label{fig:capitalmarket}
@@
 Broker-dealers clients are institutional investors, corporate clients,
-commercial clients, and high-net-worth individuals.
+commercial clients, and high-net-worth individuals \cite{aldridge2009}.
 
@@
   \includegraphics[width=0.8\textwidth]{img/capitalmarketsnow}
-  \caption{Actual structure of capital markets.}
+  \caption{Actual structure of capital markets. Source \cite{aldridge2009}.}
   \label{fig:capitalmarketnow}
@@
 Equity market and foreign exchange market (Forex) are the most popular markets
-for high frequency trading strategies. In the Equity market, stocks
-such as futures and options, exchange-traded funds (ETFs) among others financial
-instruments can be traded. Additionally, in the Forex market, interest rates denominated pair currencies 
-such as EURUSD (Euro to UD dollar) or USDJPY (US dollar to Japanese yen) are traded. Traders of this market are diverse, some of them are
-high frequency traders, long-term investors and corporations. The Forex market
-used to be centralised, only commercial banks had exclusive access to
-inter-dealer networks. Today, Forex market is decentralised and has become the
-biggest market in terms of volume of trading. The main participants are
-international banks geographically dispersed with continuous 24 hours operations
-excepting weekends. Figure~\ref{fig:Forextimes} shows different market trading
-hours in GMT for London, New York, Sydney and Tokyo.
+for high frequency trading strategies \cite{genccay2001introduction}. In the
+Equity market, stocks such as futures and options, exchange-traded funds (ETFs)
+among others financial instruments can be traded. Additionally, in the Forex
+market, interest rates denominated pair currencies such as EURUSD (Euro to UD
+dollar) or USDJPY (US dollar to Japanese yen) are traded. Traders of this
+market are diverse, some of them are high frequency traders, long-term
+investors and corporations. The Forex market used to be centralised, only
+commercial banks had exclusive access to inter-dealer networks. Today, Forex
+market is decentralised and has become the biggest market in terms of volume of
+trading. The main participants are international banks geographically dispersed
+with continuous 24 hours operations excepting weekends.
+Figure~\ref{fig:Forextimes} shows different market trading hours in GMT for
+London, New York, Sydney and Tokyo.
 
@@
   \includegraphics[width=0.8\textwidth]{img/forex-trading-hours.png}
-  \caption{Forex market trading hours (GMT).}
+  \caption{Forex market trading hours (GMT). Source \cite{dailyprice}.}
   \label{fig:Forextimes}
@@
 
-These overlap hours are very important since the highest volume of trades are expected. Therefore, most HFT strategies are executed in these hours.
+These overlap hours are very important since the highest volume of trades are
+expected. Therefore, most HFT strategies are executed in these hours.
 
@@
 investments for a period of time. Stop order allows to specify the price that a
-position should be closed called stop price. When the stop price is reached a
-market order is executed, this means that market price could not be exactly the same
-specified in the stop order.
+position should be closed; this price is called stop price. When the stop price
+is reached a market order is executed, this means that market price could not be
+exactly the same specified in the stop order.
 \end{description}
@@
 willing to pay for a good. Orders with the same price are prioritised by
-arrival time and placed on top of the book.  The difference between current ask and bid
-price is called spread and their average is called mid-price.
+arrival time and placed on top of the book.  The difference between current ask
+and bid price is called spread and their average is called mid-price
+\cite{bouchaud2002statistical}.
 
@@
   \caption{Order book. Buyers and sellers are ordered according to the bid or
-  ask price and the market determines the mid-price and transacted volume.}
+ask price and the market determines the mid-price and transacted volume. Source
+\cite{gould2013limit}.}
   \label{fig:orderbook}
@@
 
-Efficient Market Hypothesis (EMH) was developed independently by  Paul Samuelson and Eugene Fama in the 1960s.
-EMH also known as the random walk theory states that current stock prices fully reflect available information related to its value and there is no way to earn excess profits
-\cite{fama1970}.
-EMH requires that all the participants have rational expectations, and investors reactions be random and follow a normal distribution pattern.
-Thus, any one can be wrong about the market, but the market is always right as a whole.
-There are three common forms of EMH: weak-form efficiency, semi-strong efficiency and strong-form efficiency.
+Efficient Market Hypothesis (EMH) was developed independently by  Paul
+Samuelson and Eugene Fama in the 1960s.  EMH also known as the random walk
+theory states that current stock prices fully reflect available information
+related to its value and there is no way to earn excess profits
+\cite{fama1970}. EMH requires that all the participants have rational
+expectations, and investors reactions be random and follow a normal
+distribution pattern. Thus, any one can be wrong about the market, but the
+market is always right as a whole.  There are three common forms of EMH:
+weak-form efficiency, semi-strong efficiency and strong-form efficiency.
 
-The weak form claims that prices already reflect all past publicly available information.
-Therefore, future prices cannot be predicted based on analysis of historical data.
-This implies that future prices movements are determined entirely by information not contained in the past prices and participants are unable to systematically profit from market inefficiencies.
-However, many studies have shown a marked tendency for the stock markets to trend over time periods.
-Various explanations for such large and apparently non-random price movements have been promulgated.
-Even Fama has accepted price anomalies which do not follow the weak-form efficiency hypothesis \cite{fama+french2008}.
+The weak form claims that prices already reflect all past publicly available
+information. Therefore, future prices cannot be predicted based on analysis of
+historical data. This implies that future prices movements are determined
+entirely by information not contained in the past prices and participants are
+unable to systematically profit from market inefficiencies.  However, many
+studies have shown a marked tendency for the stock markets to trend over time
+periods. Various explanations for such large and apparently non-random price
+movements have been promulgated. Even Fama has accepted price anomalies which
+do not follow the weak-form efficiency hypothesis \cite{fama+french2008}.
 
-The semi-strong form of the EMH claims that prices instantly change to reflect new public (not private) information such that no excess return can be obtained.
+The semi-strong form of the EMH claims that prices instantly change to reflect
+new public (not private) information such that no excess return can be
+obtained.
 
-The strong form of the EMH additionally claims that prices instantly reflect even hidden, private or ``insider" information.
+The strong form of the EMH additionally claims that prices instantly reflect
+even hidden, private or ``insider" information.
 
-EMH is related with two approaches to investment analysis: fundamental and technical analysis.
-Fundamental analysts base their predictions of stock price behaviour on fundamental factors such as internal information of a company, its industry or the economy.
-Technical analysts, by contrast, consider that all this financial information is already included in the prices and believe that future stock prices can be predicted studying the historical market behaviour.
-A market technician bases its predictions on historical patterns of prices of volume changes.
+EMH is related with two approaches to investment analysis: fundamental and
+technical analysis. Fundamental analysts base their predictions of stock price
+behaviour on fundamental factors such as internal information of a company, its
+industry or the economy. Technical analysts, by contrast, consider that all
+this financial information is already included in the prices and believe that
+future stock prices can be predicted studying the historical market behaviour.
+A market technician bases his predictions on historical patterns of prices of
+volume changes.
 
-Under the weak form of EMH, strategies based on technical analysis will not be able to produce excess returns.
-However, the weak form accepts that some forms of fundamental analysis may still provide excess return.
-Similarly, the semi-strong form of EMH  neither technical or fundamental analysis can produce excess return.
+Under the weak form of EMH, strategies based on technical analysis will not be
+able to produce excess returns. However, the weak form accepts that some forms
+of fundamental analysis may still provide excess return. Similarly, the
+semi-strong form of EMH  neither technical or fundamental analysis can produce
+excess return.
 
-If the stock market efficiently digests all available information, there is little justification for seeking excess returns gains from investing.
-However, EMH doesn't lessen the importance of investing only change its philosophy.
-The only way an investor can possibly obtain higher returns is by purchasing riskier investments.
+If the stock market efficiently digests all available information, there is
+little justification for seeking excess returns gains from investing. However,
+EMH doesn't lessen the importance of investing only change its philosophy. The
+only way an investor can possibly obtain higher returns is by purchasing
+riskier investments.
 
- Researches can determine now how efficient a financial market is, i.e how efficiently information is processed.
+ Researches can determine now how efficient a financial market is, i.e how
+efficiently information is processed.
 
-EMH is based on rational human behaviour and its validity has been criticised by
-psychologists and behavioural economists who argue that the EMH is based
-on counterfactual assumptions regarding human behaviour, that is,
-rationality. Recent advances in evolutionary psychology and the cognitive
-neurosciences may be able to reconcile the EMH with behavioural
-anomalies.
-On the other hand, if the EMH is true, the market really walks randomly and therefore there shouldn't be any difference between experienced and novice traders.
-Kim Man Lui proved the contrary in a controlled experiment \cite{man2013}.
+EMH is based on rational human behaviour and its validity has been criticised
+by psychologists and behavioural economists who argue that the EMH is based on
+counterfactual assumptions regarding human behaviour, that is, rationality.
+Recent advances in evolutionary psychology and the cognitive neurosciences may
+be able to reconcile the EMH with behavioural anomalies. On the other hand, if
+the EMH is true, the market really walks randomly and therefore there shouldn't
+be any difference between experienced and novice traders. Kim Man Lui proved
+the contrary in a controlled experiment \cite{man2013}.

### 2017-08-20 src/4_OnlineLearning.tex
ultimos comentarios

@@
 
-Supervised learning consists in finding a learning function $f: X \rightarrow Y$
-which models a set of examples, called training set $S$, which have been drawn
-randomly and independently according to a unknown joint distribution function
-$p(x,y)$ on $X \times Y$:
-
+%Supervised learning consists in finding a learning function $f: X \rightarrow Y$
+%which models a set of examples, called training set $S$, which have been drawn
+%randomly and independently according to a unknown joint distribution function
+%$p(x,y)$ on $X \times Y$:
+%
+%
+%$$
+%S = \{ (x_k,y_k)\in X\times Y : k=1,...,n \}
+%$$
+%
+%\noindent where $X \subseteq \mathbb{R}^m$ and $Y = \mathbb{R}$ in case of
+%regression problems and $Y = \{1,-1\}$  if it is a classification problem. 
+%
+%
+%A learning algorithm $\mathcal{A}$ takes as input a data set $S \in
+%\mathscr{T}=\mathscr{P}(X\times Y)$ and selects from $\mathscr{H}$ a function
+%$f$ such that $f(\mathbf{x})\approx y $ in a predictive way:
+%
+%\begin{eqnarray*}
+%\mathcal{A}: \mathscr{T} &\rightarrow & \mathscr{H} \\
+%S &\rightarrow &\mathcal{A}(S) = f
+%\end{eqnarray*}
+%
+%\noindent where $\mathscr{H}$ is called the {\em hypothesis space}, i.e. the space
+%of functions that the algorithm is allowed to search.
+
+%------------------------DRAFT PROFE -------------------------------
+The basic setting for supervised learning is a {\em data set\/} $X$, 
+and an  {\em outcome set\/} $Y$.
+For our purposes $X$ will be $\R^m$, $m\in\N$, and $Y$ either $\R$
+in the case of regression problems, or $\{-1,1\}$ in the case of
+classification problems.
+In addition we have a plurality of {\em training sets\/}
 $$
-S = \{ (x_k,y_k)\in X\times Y : k=1,...,n \}
+S=\bigg\{ (\mathbf{x}_k,y_k)\in X\times Y \mid k=1,\dots,n \bigg\}
+\subseteq X\times Y\,,
 $$
-
-\noindent where $X \subseteq \mathbb{R}^m$ and $Y = \mathbb{R}$ in case of
-regression problems and $Y = \{1,-1\}$  if it is a classification problem.
-
-A learning algorithm $\mathcal{A}$ takes as input a data set $S \in
-\mathscr{T}=\mathscr{P}(X\times Y)$ and selects from $\mathcal{H}$ a function
-$f$ such that $f(x)\approx y $ in a predictive way:
-
-\begin{eqnarray*}
-\mathcal{A}: \mathscr{T} &\rightarrow & \mathcal{H} \\
-S &\rightarrow &\mathcal{A}(S) = f
-\end{eqnarray*}
-
-\noindent where $\mathcal{H}$ is called the {\em hypothesis space}, i.e. the space
-of functions that the algorithm is allowed to search. The selection of
-$f$ is based on the minimisation of the expected error of the loss
-function $V(f(x),y): Y \times Y$ defined as:
-
-\begin{eqnarray}
+where the members of $S$ have been drawn randomly and independently
+from $X\times Y$ according to an {\em unknown\/} joint distribution
+function $p(\mathbf{x},y)$ on $X\times Y$.
+We gather all these training sets in a subset $\mathscr{T}$ of
+$\mathscr{P}(X\times Y)$.
+
+In addition, there is a {\em learning algorithm\/} $\mathcal{A}$ that
+associates to every data set one and only one learning function
+$f\in\mathscr{H}$.
+Thus $\mathcal{A}$ can be considered as a function:
+$$
+\mathcal{A}:\mathscr{T}\to\mathscr{H}\,,\qquad
+S\mapsto\mathcal{A}(S)=f\quad\forall S\in\mathscr{T}\,.
+$$
+
+%------------------------FIN DRAFT PROFE -------------------------------
+{\em Supervised learning\/} consists then in finding a {\em learning function\/}
+$f:X\to Y$ that minimisr the expected error of the loss function $V(f(\mathbf{x}),y): Y
+\times Y$ defined as:
+
+\begin{eqnarray*}
 \label{eq:lossfunction}
  V & \rightarrow & \mathbb{R}^+ \\
-  (f(x), y) &\rightarrow & V(f(x),y)
-\end{eqnarray}
-
-\noindent where $V(f(x),y)$ denote the price paid for mistakes, $f:X=\R^m\to Y$ and
-$y=f(x)\in\R$. Therefore, $V(f(x),y)=0$ if $\;f(x)=y$.
-
-Given a function $f$ a loss function $V$ and a probability distribution $p$ over $X \times Y$, the expected error of $f$ is:
+  (f(\mathbf{x}), y) &\rightarrow & V(f(\mathbf{x}),y)
+\end{eqnarray*}
+
+\noindent where $V(f(\mathbf{x}),y)$ denote the price paid for mistakes, $f:X=\R^m\to Y$
+and $y=f(\mathbf{x})\in\R$. Therefore, $V(f(\mathbf{x}),y)=0$ if $\;f(\mathbf{x})=y$.
+
+Given a function $f$ a loss function $V$ and a probability distribution $p$ over
+$X \times Y$, the expected error of $f$ is:
 
@@
 \label{eq:expetedrisk}
-E[V(f(x),y)] = \int_{X\times Y} V(f(x),y)\;dp(x,y)\,,
+%E[V(f(\mathbf{x}),y)] = \int_{X\times Y} V(f(\mathbf{x}),y)\;dp(x,y)\,,
+E\big[ V( f(\cdot),\cdot) \big]
+= \int_{X\times Y} V\big( f(\mathbf{x},y \big)\,dp(\mathbf{x},y)\,.
 \end{equation}
@@
 \label{eq:l2loss}
-V(f(x),y) = (f(x)-y)^2
+V(f(\mathbf{x}),y) = (f(\mathbf{x})-y)^2
 \end{equation}
@@
 \label{eq:l1loss}
-V(f(x),y) = |f(x)-y| \, .
+V(f(\mathbf{x}),y) = |f(\mathbf{x})-y| \, .
 \end{equation}
@@
 \begin{equation*}
-\hat{f} = \argmin_{f \in \mathcal{H}} R_{\text{emp}}[f] 
+\hat{f} = \argmin_{f \in \mathscr{H}} R_{\text{emp}}[f] 
 \end{equation*}
@@
 \begin{equation}
-\frac{1}{n} \sum_{i=1}^n V(f(x_i),y_i) \xrightarrow{n \rightarrow \infty}  E[V(f(x),y)]
+\frac{1}{n} \sum_{i=1}^n V(f(x_i),y_i) \xrightarrow{n \rightarrow \infty}  E[V(f(\mathbf{x}),y)]
 \end{equation}
@@
 points that can be classified exactly by the hypothesis space. More formally,
-the VC dimension of the hypothesis space $\mathcal{H}$ defined over instance
+the VC dimension of the hypothesis space $\mathscr{H}$ defined over instance
 space $X$ is the size of the largest finite subset of $X$ shattered by
-$\mathcal{H}$. The dataset $S$ is shattered by hypothesis space $\mathcal{H}$ if
+$\mathscr{H}$. The dataset $S$ is shattered by hypothesis space $\mathscr{H}$ if
 and only if for every dichotomy of $S$ there exists some hypothesis in
-$\mathcal{H}$ consistent with this dichotomy. A dichotomy of a set $S$ is a
+$\mathscr{H}$ consistent with this dichotomy. A dichotomy of a set $S$ is a
 partition of $S$ into two disjoint subsets \cite{vapnik2013nature}.
@@
 \label{eq:bvtrade}
-\mathrm{E}\Big[\big(y - \hat{f}(x)\big)^2\Big]
- & = \mathrm{Bias}(\hat{f}(x))^2 + \mathrm{Var}(\hat{f}(x)) + \sigma^2
+\mathrm{E}\Big[\big(y - \hat{f}(\mathbf{x})\big)^2\Big]
+ & = \mathrm{Bias}(\hat{f}(\mathbf{x}))^2 + \mathrm{Var}(\hat{f}(\mathbf{x})) + \sigma^2
 \end{align}
@@
 \begin{align}
- \mathrm{Bias}(\hat{f}(x)) &= \mathrm{E}\big[\hat{f}(x)\big] - f(x) \label{eq:biastrade}\\
- \mathrm{Var}(\hat{f}(x)) &= \mathrm{E}\Big[ \big( \hat{f}(x) - \mathrm{E}[\hat{f}(x)] \big)^2 \Big]  \label{eq:variancetrade}
+ \mathrm{Bias}(\hat{f}(\mathbf{x})) &= \mathrm{E}\big[\hat{f}(\mathbf{x})\big] - f(\mathbf{x}) \label{eq:biastrade}\\
+ \mathrm{Var}(\hat{f}(\mathbf{x})) &= \mathrm{E}\Big[ \big( \hat{f}(\mathbf{x}) - \mathrm{E}[\hat{f}(\mathbf{x})] \big)^2 \Big]  \label{eq:variancetrade}
 \end{align}
@@
 generalisation of ERM, i.e prevents overfit, by constraining the hypothesis
-space $\mathcal{H}$ usually called regularised ERM or Tikhonov regularisation.
-
-%Tikhonov regularisation minimised over the hypothesis space $\mathcal{H}$ for a
+space $\mathscr{H}$ usually called regularised ERM or Tikhonov regularisation.
+
+%Tikhonov regularisation minimised over the hypothesis space $\mathscr{H}$ for a
 %fixed positive parameter $\lambda$ the following:
@@
 
-\noindent where $X \subseteq \mathbb{R}^m$ and $Y = \mathbb{R}$ , $V$ is the loss function defined in the equation \ref{eq:lossfunction} and $\mathcal{R}(f)$ is the regulariser, a penalisation on $f$.
-One example of regularisation in linear models is Ridge Regression (RR), which is a regularised least squares method.
+\noindent where $X \subseteq \mathbb{R}^m$ and $Y = \mathbb{R}$ , $V$ is the
+loss function defined in the equation \ref{eq:lossfunction} and $\mathcal{R}(f)$
+is the regulariser, a penalisation on $f$. One example of regularisation in
+linear models is Ridge Regression (RR), which is a regularised least squares
+method.
 
@@
 
-Ridge regression (RR) was independently proposed by Tikhonov \cite{tikhonov1963} and
-Phillips \cite{phillips1962}. RR corresponds to a regularised least squares method. The least squares (LS) method is a well known way to solve a regression
+Ridge regression (RR) was independently proposed by Tikhonov \cite{tikhonov1963}
+and Phillips \cite{phillips1962}. RR corresponds to a regularised least squares
+method. The least squares (LS) method is a well known way to solve a regression
 problem. LS method consists of minimising the sum of squared errors:

### 2017-08-27 src/4_OnlineLearning.tex
explicacion de ridge regression
last comments chapter 3

@@
 {\em Supervised learning\/} consists then in finding a {\em learning function\/}
-$f:X\to Y$ that minimisr the expected error of the loss function $V(f(\mathbf{x}),y): Y
+$f:X\to Y$ that minimise the expected error of the loss function $V(f(\mathbf{x}),y): Y
 \times Y$ defined as:
 
-\begin{eqnarray*}
-\label{eq:lossfunction}
- V & \rightarrow & \mathbb{R}^+ \\
-  (f(\mathbf{x}), y) &\rightarrow & V(f(\mathbf{x}),y)
-\end{eqnarray*}
+%\begin{eqnarray*}
+%\label{eq:lossfunction}
+% V & \rightarrow & \mathbb{R}^+ \\
+%  (f(\mathbf{x}), y) &\rightarrow & V(f(\mathbf{x}),y)
+%\end{eqnarray*}
+
+\begin{equation}\label{eq:lossfunction}
+V(f(\cdot),\cdot):X\times Y\to\R_0^+\,,\qquad
+(\bfx,y)\mapsto V\big(f(\bfx),y\big)
+\end{equation}
 
@@
 Since the true distribution is unknown and only training samples are available,
-the objective is to estimate a function $\hat{f}$ through empirical risk
+the objective is to estimate a function $\widehat{f}$ through empirical risk
 (training error) minimisation (ERM):
@@
 \begin{equation*}
-\hat{f} = \argmin_{f \in \mathscr{H}} R_{\text{emp}}[f] 
+\widehat{f} = \argmin_{f \in \mathscr{H}} R_{\text{emp}}[f] 
 \end{equation*}
@@
 \begin{equation}
-\frac{1}{n} \sum_{i=1}^n V(f(x_i),y_i) \xrightarrow{n \rightarrow \infty}  E[V(f(\mathbf{x}),y)]
+%\frac{1}{n} \sum_{i=1}^n V(f(x_i),y_i) \xrightarrow{n \rightarrow \infty}  E[V(f(\mathbf{x}),y)]
+\frac1n\,\sum_{i=1}^n V\big( f(\bfx_i,y_i \big)
+\overset{n\to\infty}{\longrightarrow} E\big[ V(f(\cdot),\cdot)\big]\,.
 \end{equation}
@@
   \caption{VC dimension example from \cite{vapnik2013nature}. The left figure
-  shows that three vectors can be shattered by a line in a plane. However, if we
+  shows that the 3 dichotomies can be shattered by a line in a plane. However, if we
   add an extra vector $z_4$, it is impossible to shatter them with a
-  plane.Therefore, linear classifiers has VC dimension equals to $d+1$ where $d$
+  line. Therefore, linear classifiers has VC dimension equals to $d+1$ where $d$
   is the dimension of the instance space $X$.}
@@
 \end{figure}
+
+A key point is Radon's theorem in discrete geometry \cite{peterson1972geometry}
+($\text{co}(A)$ denotes the convex hull of $A$):
+
+\begin{theorem}[J. Radon]
+Let $S\subset\R^d$ be a subset containing at least $d+2$ elements. 
+Then there exist {\bf disjoint} subsets $A\subset S$ and $B\subset S$
+such that
+$$
+\text{co}(A)\cap\text{co}(B)\neq\emptyset
+$$
+\end{theorem}
+
+Radon's theorem implies that in $\R^d$ any set of $d+2$ or more vectors has at least one dichotomy which cannot be
+separated (shattered) by any plane in $\R^d$. Thus the VC-dimension in $\R^d$ cannot be greater than $d+1$.
 
@@
 Ridge regression (RR) was independently proposed by Tikhonov \cite{tikhonov1963}
-and Phillips \cite{phillips1962}. RR corresponds to a regularised least squares
+and Phillips \cite{phillips1962}.
+
+RR corresponds to a regularised least squares
 method. The least squares (LS) method is a well known way to solve a regression
-problem. LS method consists of minimising the sum of squared errors:
-
+problem.
+
+\begin{equation}
+\label{eq:RRdefinitions}
+\mathbf{X} = 
+\left[
+  \begin{array}{cccc}
+    \vertbar & \vertbar &        & \vertbar \\
+    \mathbf{x}_{1}    & \mathbf{x}_{2}    & \ldots & \mathbf{x}_{n}    \\
+    \vertbar & \vertbar &        & \vertbar 
+  \end{array}
+\right]
+\quad , \quad
+\mathbf{A} = 
+\left[
+  \begin{tabular}{c>{$}c<{$}c}
+    --- & \mathbf{a}^{\top}_{1} & ---\\
+    --- & \mathbf{a}^{\top}_{2} & ---\\
+    & \vdots & \\
+    --- & \mathbf{a}^{\top}_{m} & ---
+  \end{tabular}
+\right]
+\quad \text{and} \quad
+\mathbf{Y} = 
+\left[
+  \begin{array}{cccc}
+    \vertbar & \vertbar &        & \vertbar \\
+    \mathbf{y}_{1}    & \mathbf{y}_{2}    & \ldots & \mathbf{y}_{n}    \\
+    \vertbar & \vertbar &        & \vertbar 
+  \end{array}
+\right]
+\end{equation}
+
+LS method consists of minimising the sum of squared errors:
 \begin{eqnarray*}
 \label{eq:lsm}
- J(\mathbf{X}) =& \displaystyle \sum_{t=1}^N (f(\mathbf{x}_t)-y_t)^2 \\
- =&\displaystyle \sum_{t=1}^N (\mathbf{\mathbf{A}}^\top {\mathbf{x}}_t-y_t)^2 \\
+ J(\mathbf{X}) =& \displaystyle \sum_{t=1}^n (f(\mathbf{x}_t)-y_t)^2 \\
+ =&\displaystyle \sum_{t=1}^n (\mathbf{\mathbf{A}}^\top {\mathbf{x}}_t-y_t)^2 \\
  =& \| \mathbf{A}\mathbf{\mathbf{X}} - \mathbf{Y} \|_2^2 

### 2017-08-27 comentarios-tesis/chapter3ok/20170814_Paola_Chapter_3_again.tex
last changes to chapter 3

%% 20170814_Paola_Chapter_3_again.tex

\red{\section{\Large AUGUST 14, 2017 -- Chapter 3 once again!}}

\begin{description}[style=unboxed,leftmargin=0cm,itemsep=3ex]

\myitem{P 36, section 3.2, L 6}
It says:\quad ``...in finding a learning function $f:X\to Y$
which models... etc.''
It is not said here what are $X$ and $Y$; 
further on it is said. 
This is not usual in a scientific text.
Thus, I would suggest a text like this:

\red{``...in finding a learning function $f:X\to Y$, where $X\subseteq\R^m$
and $Y\subset\R$, which models a set of examples $S\subseteq X\times Y$,
called {\em training sets}, which have been drawn randomly and
independently from $X\times Y$ according to \blue{an} unknown joint
distribution function $p(x,y)$ on $X\times Y$:
$$
S=\big\{ (\mathbf{x}_k,y_k)\in X\times Y \mid k=1,\dots,n \big\}
$$
Here, $Y=\R$ in case of regression problems and $Y=\{-1,1\}$ if it is
a classification problem.''}

\blue{However, below there is a deeper discussion of these facts and
I propose you a more comprehensive composition.}

%% \myitem{P 36, section 3.2, L -4}
%% The expression $S\in\mathscr{T}=\mathscr{P}(X\times Y)$ seems
%% unnecessarily complicated.

%% Isn't it the same (and much simpler) to say that the training sets
%% fulfill $S\subseteq X\times Y$ ?

%% The introduction of $\mathscr{T}$ suggests that you want to have the
%% possibility of choosing \red{several} training sets $S$ taken from
%% $\mathscr{T}=\mathscr{P}(X\times Y)$, but this is unnecessary since
%% if you indeed had several training sets, say, $S_\alpha,\;S_\beta,\;\dots$, 
%% you coud always consider a super-training set
%% $S=S_\alpha \cup S_\beta \cup \dots$ and you would achieve the same generality.

%% Thus in the sequel I shall abandone the set $\mathscr{T}$ and shall assume
%% that we have just \red{ONE} training set $S$ and that this is a \red{subset}
%% of $X\times Y$: $S\subseteq X\times Y$.

\myitem{P 36, section 3.2, L -3}
Suddenly the symbol \red{$\mathscr{H}$} appears on the stage!
But it is not defined before or in the same line!
Only after the formula for $\mathcal{A}$ there is a hint about
$\mathscr{H}$.
This writing style is highly non-standard in math!

\myitem{P 36-37, section 3.2. Some remarks trying to make sense of all these!}

\mbox{}\newline
After reading your definitions several times, it seems to me that:

\begin{enumerate}

\item
Which are the references for the many definitions in these pages ?

\item
The most basic data sets you have are $X\subseteq\R^m$ and $Y\subseteq\R$.
You distinguish two cases: $Y=\R$, the case of regression problems;
and $Y=\{-1,1\}$, the case of classification problems.

\item
Then you have a training set $S\subseteq X\times Y$ of the form:
$$
S=\big\{ (\mathbf{x}_k,y_k)\in X\times Y \mid k=1,\dots,n \big\}\,,
$$
where the members of $S$ have been drawn randomly and independently
according to a unknown joint distribution function
$p(\mathbf{x},y)$ on $X\times Y$.

\item
The problem is to find a function $f:X\to Y$ such that:
\begin{itemize}
\item
it \red{models} the training set $S$ in the sense that
$f(\mathbf{x}_k)=y_k$ for all $k=1,\dots,n$; \\
\red{Is this true? If not, then in what sense $f$ models $S$ ?}

\item
it has a \red{predictive capacity} in the sense that $f(\mathbf{x})=y$
for a newly arrived data $(\mathbf{x},y)\in X\times Y\setminus S$.
\end{itemize}
These conditions seem to be too restrictive.
We can weaken them a little by accepting that the function $f:X\to Y$ should:
\begin{itemize}
\item
just \red{closely models} the training set $S$ in the sense that
$f(\mathbf{x}_k)\approx y_k$ with an error $|f(x_k)-y_k|<\varepsilon$ for
a given $\varepsilon>0$ and all $k=1,\dots,n$; \\
\red{Is this true? If not, then in what sense $f$ models $S$ ?}
\item
have an \red{approximate predictive capacity} in the sense that
$f(\mathbf{x})\approx y$ with error $|f(\mathbf{x})-y|<\varepsilon$
for a newly arrived data $(\mathbf{x},y)\in X\times Y\setminus S$.
\end{itemize}
\item 
This function $f:X\to Y$ should be searched after (and hopefully found)
among a set of functions $\mathscr{H}$ from $X$ to $Y$.

The set of functions $\mathscr{H}$ is called \red{hypothesis space}.

Technically speaking, $\mathscr{H}$ is a subset of the set $Y^X$
of \red{all} functions $\varphi:X\to Y$, but this is mathematically too
technical for our purposes in this work.

For the search process it would be convenient to introduce a topology
on $\mathscr{H}$, preferrable a norm topology (like an $L^p$-norm for
instance) and to require topological completeness.
It would be even better if we would require that $\mathscr{H}$ is Banach
space, but this would be perhaps too technical for our goals here.

\item
There is also a learning algorithm $\mathcal{A}$ which allows $f$
to be found in the \red{hypothesis space} $\mathscr{H}$.

We discussed this subject in October last year.
Please see my remarks from that month.

You now interpret $\mathcal{A}$ as an application (function) from
$\mathscr{T}\subseteq\mathscr{P}(X\times Y)$ to $\mathscr{H}$, 
$\mathcal{A}:S\mapsto\mathcal{A}(S)=f$.

\red{Note that you write $\mathscr{T}=\mathscr{P}(X\times Y)$, but
$\mathscr{T}\subseteq\mathscr{P}(X\times Y)$ gives you more freedom.}

I think there is still a conceptual ``darkness'' associated with 
``$S\in\mathscr{T}\subseteq\mathscr{P}(X\times Y)$''.

The problem is \red{to find a learning function $f$}, with the
properties stated above, for some particular (financial) experiment,
but it is one and only one well defined experiment, for instance,
the analysis of a vector valued (financial) time series (this would
be the $\mathbf{x}_k=\left[x_k^1,\dots,x_k^m\right]^T$ for the
(time-) instants $k=1,\dots,n$, but what would represent the 
``$y_k\in\R$''?
But never mind, it will be a real number (possibly $\pm1$) for each $k$.

Now comes the delicate point: for this one and only one experiment
you have:
\begin{itemize}
\item[1.]
\red{Either} one and only one data set 
$S\in\mathscr{T}\subseteq\mathscr{P}(X\times Y)$
and hence your algorithm $\mathcal{A}$ acting on $S$
produces one and only one learning function $f$, $\mathcal{A}(S)=f$,
for the whole experiment.

In this case it doesn't make sense (you could, but it doesn't help too
much!) to consider the algorithm $\mathcal{A}$
as a mapping $\mathcal{A}:\mathscr{T}\to\mathscr{H}$ since $\mathscr{T}$
has only one member, namely, $\mathscr{T}=\big\{S\big\}$.
\item[2.]
\red{Or} several different data sets
$S_\alpha,S_\beta,\dots\in\mathscr{T}\subseteq\mathscr{P}(X\times Y)$
of the same one and only one experiment, and hence your algorithm
$\mathcal{A}$ acting on $S_\alpha$, $S_\beta$, $\dots$ should produce
learning functions $\mathcal{A}(S_\alpha)=f_\alpha$, 
$\mathcal{A}(S_\beta)=f_\beta$, $\dots$

In this case you could have 
$\mathscr{T}=\big\{ S_\alpha,S_\beta,\dots\big\}$
and it make sense to consider the algorithm $\mathcal{A}$ as a mapping
$\mathcal{A}:\mathscr{T}\to\mathscr{H}$:
$$
S_\alpha\mapsto \mathcal{A}(S_\alpha)=f_\alpha\in\mathscr{H}\,,\quad
S_\beta \mapsto \mathcal{A}(S_\beta )=f_\beta \in\mathscr{H}\,,\quad\dots
$$
\end{itemize}

\item
\blue{\bf An example of Case 2. as above, to fix the ideas!}
This would be the case if you consider the data sets 
$S_\alpha,S_\beta,\dots\in\mathscr{T}\subseteq\mathscr{P}(X\times Y)$
as \red{segments} or \red{windows} of a time series
$\big\{(\mathbf{x}_t,y_t)\big\}_{t\in\Z}$ of vectors 
$(\mathbf{x}_t,y_t)\in\R^m\times\R$, say
\begin{align*}
S_\alpha &= \bigg\{ 
\left( \mathbf{x}_{t_\alpha}, y_{t_\alpha} \right), 
\left( \mathbf{x}_{t_\alpha+1}, y_{t_\alpha+1} \right), \dots, 
\left( \mathbf{x}_{t_\alpha+n-1}, y_{t_\alpha+n-1} \right) \bigg\}\,, \\
S_\beta &= \bigg\{ 
\left( \mathbf{x}_{t_\beta}, y_{t_\beta} \right), 
\left( \mathbf{x}_{t_\beta+1}, y_{t_\beta+1} \right),\dots, 
\left( \mathbf{x}_{t_\beta+n-1}, y_{t_\beta+n-1} \right) \bigg\}\,, \\
\vdots & \rule{10em}{0ex}\vdots
\end{align*}
for $t_\alpha,t_\beta,\dots\in\Z$, which without restriction we can assume
to be increasing: $t_\alpha<t_\beta<\dots$.

Note that here all data sets $S_\alpha,S_\beta,\dots$ have the same cardinality
(windows of the same length), namely $n$.
But this is not obligatory: windows could have different length, but
computations would then be more cumbersome.

Also note that nothing is said about the convergence of the sequence
$f_\alpha,f_\beta,\dots$
In fact this sequence may very well not converge in $\mathscr{H}$.
The terms of this sequence are functions that model the behaviour of the
time series within the particular window associated to each data set
$S_\alpha,S_\beta,\dots$ and what is expected is that the corresponding
learning functions $f_\alpha,f_\beta,\dots$ have good predictive capacity,
namely that: 
\begin{align*}
\widehat{y}_{t_\beta}
&:= \mathcal{A}(S_\alpha)(\mathbf{x}_{t_\beta})
\equiv f_\alpha(\mathbf{x}_{t_\beta}) \approx y_{t_\beta} \quad
\text{with an error $\big|\widehat{y}_{t_\beta}-y_{t_\beta}\big|<\varepsilon$}\,, \\
\widehat{y}_{t_\gamma}
&:= \mathcal{A}(S_\beta)(\mathbf{x}_{t_\gamma})
\equiv f_\beta(\mathbf{x}_{t_\gamma}) \approx y_{t_\gamma} \quad
\text{with an error $\big|\widehat{y}_{t_\gamma}-y_{t_\gamma}\big|<\varepsilon$}\,, \\
\vdots & \rule{10em}{0ex}\vdots
\end{align*}
Note that in these equations $f_\alpha$ is evaluated in $\mathbf{x}_{t_\beta}=$
the $X$-coordinate of the first member of the next ``window'' $S_\beta$.

Similarly, $f_\beta$ is evaluated in $\mathbf{x}_{t_\gamma}=$ the $X$-coordinate
of the first member of the next ``window'' $S_\gamma$.
Etc.

Note, however, that each data set in Case 2. is just another version of
the one and only one data set in Case 1.
Thus, Case 2. is \red{not} a generalisation of Case 1.
In fact, Case 2. can be reduced to Case 1. just by taking the union of
all data sets of Case 2.

\red{\bf Thus, it seems to me that you are considering Case 2.
I will assume this in the sequel!}

\item\label{referenceFunction}
Note that there is always an {\em observed\/} or {\em reference function\/}
$$
\mathscr{Y}:X\to Y\,,\qquad
\mathscr{Y}: \mathbf{x}_k \mapsto \mathscr{Y}(\mathbf{x}_k):=y_k\,,
\quad k=1,\dots,n\,.
$$
We will always assume that $\mathscr{Y}\in\mathscr{H}$.
In fact, $\mathscr{Y}$ is from the very beginning the most natural member
of $\mathscr{H}$!

Besides this reference function $\mathscr{Y}$ there are the learning
functions
$f\in\mathscr{H}$:
$$
f:X\to Y\,,\qquad
f: \mathbf{x}_k \mapsto f(\mathbf{x}_k)=:\widehat{y}_k\,,
\quad k=1,\dots,n\,.
$$
The {\em error function\/} associated to any $f\in\mathscr{H}$
is defined by:
$$
\texttt{Error}_f(\mathbf{x}_k)
:=|f(\mathbf{x_k})-\mathscr{Y}(\mathbf{x}_k)|=|\widehat{y}_k-y_k|\,,
\quad k=1,\dots,n\,,
$$
where the error function \texttt{Error} should ideally be small, say, 
$\texttt{Error}(\mathbf{x}_k)<\varepsilon$ for all $k=1,\dots,n$, for
a given $\varepsilon>0$.

Note that $f,\mathscr{Y}:X\to Y$ but $\texttt{Error}_f:X\to\R_0^+$.

In function-analytic notation we observe that, for $f\in\mathscr{H}$,
$\texttt{Error}_f$ is a function on $X$:
$$
\texttt{Error}_f:X\to\R_0^+\,,\qquad
\texttt{Error}_f(\mathbf{x}) 
= |f(\mathbf{x})-\mathscr{Y}(\mathbf{x})| 
= |\widehat{y} - y| \quad\forall x\in X\,.
$$
and since the functions $f$, $\mathscr{Y}$, and hence $\texttt{Error}_f$,
are defined on the whole of $X$ we can also write for any
$f\in\mathscr{H}$:
$$
\texttt{Error}_f(\mathbf{x})
= |f(\mathbf{x})-\mathscr{Y}(\mathbf{x})| 
= |f(\mathbf{x}) - y| = |\widehat{y} - y|\,,\quad \mathbf{x}\in X,\ 
\text{provided that}\ (\mathbf{x},y)\in S\in\mathscr{T}.
$$

\item\label{lossFunction}
For a given $S\in\mathscr{T}$, the algorithm $\mathcal{A}$ selects
one (and only one) learning function $f\in\mathscr{H}$ by minimising
the expected error of a {\em loss function\/}:
$$
V:Y\times Y\to\R_0^+\,,\qquad V: (y_1,y_2)\mapsto V(y_1,y_2)\,.
$$
Note that both arguments of $V$ are points in $Y$.
The prototype of this loss function in the case $Y=\R$ is the {\em metric\/}
\begin{equation}\label{BasicLossFunction}
V(y_1,y_2):=|y_1-y_2|\qquad \forall y_1,y_2\in Y=\R\,. 
\end{equation} 
Two simple loss functions for regression problemas derived from
(\ref{BasicLossFunction}) for $f\in\mathscr{H}$ are:
$$
V\big( f(\mathbf{x}),y \big) 
= |f(\mathbf{x})-y| = |\widehat{y}-y|\,,\quad 
\mathbf{x}\in X\ \text{and}\ y\in Y\ 
\text{with}\ (\mathbf{x},y)\in S\in\mathscr{T}\,,
$$
which is the so called $L^1$-{\em loss function\/} and:
$$
V\big( f(\mathbf{x}),y \big) 
= \big(f(\mathbf{x})-y\big)^2 = \big( \widehat{y}-y\big)^2\,,\quad 
\mathbf{x}\in X\ \text{and}\ y\in Y\ 
\text{with}\ (\mathbf{x},y)\in S\in\mathscr{T}\,,
$$
which is the so called {\em square\/} or $L^2$-{\em loss function\/}.

\item
Note that $V\big( f(\mathbf{x}),y \big)$, $\mathbf{x}\in X$ and $y\in Y$ 
with $(\mathbf{x},y)\in S\in\mathscr{T}$, is actually a {\em function\/}
on $X\times Y$:
$$
V\big(f(\mathbf{\cdot}),\cdot):X\times Y\to\R_0^+\,,\quad
\big( \mathbf{x},y \big) \mapsto V\big( f(\mathbf{x}),y \big)\,,\ 
\text{where}\ (\mathbf{x},y)\in S\in\mathscr{T}\,.
$$
Condition ``$(\mathbf{x},y)\in S\in\mathscr{T}$'' is a small nuisance here,
since the function ``$V\big(f(\mathbf{\cdot}),\cdot)$''  \red{is not defined
{\em for all\/} $\mathbf{x}\in X$ and {\em all\/} $y\in Y$.}
But this is inevitable, since the members of the data sets which we are
dealing with are of the form $(\mathbf{x},y)$.
It could be even the case that the variable $\mathbf{x}$ does not run on
the whole of $X$, and this is most surely the case for $y$ regarding $Y$.
When we come to compute the expected value of the function
$V\big(f(\mathbf{\cdot}),\cdot)$, we will see an easy solution to
this ``beauty defect'' by means of a clever probability distribution on
$X\times Y$ (see (\ref{expectedError}) below ).

There is another function closely related to $V\big(f(\mathbf{\cdot}),\cdot)$
but different fro that one:
$$
V\big(f(\cdot), \mathscr{Y}(\cdot)\big):X\to Y\,,\qquad
\mathbf{x}\mapsto V\big(f(\mathbf{x}), \mathscr{Y}(\mathbf{x})\big)
= \big( \widehat{y}, y \big)\,,\qquad f\in\mathscr{H}\,.
$$
Note that here we do not need the ``ugly'' condition
``$(\mathbf{x},y)\in S\in\mathscr{T}$'' because this is built-in in the
definition of the reference function $\mathscr{Y}:X\to Y$
(see (\ref{referenceFunction}) above).

\item\label{expectedError}
Consider a {\em learning function\/} $f\in\mathscr{H}$,
$f(\mathbf{x})=\widehat{y}$.
Also consider the {\em reference function\/} $\mathscr{Y}:X\to Y$, 
$\mathscr{Y}(\mathbf{x})=y$, where $(\mathbf{x},y)\in S\in\mathscr{T}$,
$S\subseteq X\times Y$.

Note that $S$ may be considered as the {\em graph\/} of $\mathscr{Y}$:
$$
\text{Graph}(\mathscr{Y})
:= \big\{ (\mathbf{x}, y)\in X\times Y \mid \mathscr{Y}(\mathbf{x})=y\big\}
= S\subseteq X\times Y\quad !
$$
There are clearly (many) probability distributions $p$ on $X\times Y$
such that:
\begin{equation}\label{probdist}
p\ \text{assigns probability 0 to all ``events''}\  
(\mathbf{x},y)\in X\times Y\setminus\text{Graph}(\mathscr{Y})
= X\times Y\setminus S\,.
\end{equation}
Consider, in addition, a given {\em loss function\/} $V:Y\times Y\to\R_0^+$.
(see (\ref{lossFunction}) above) and a probability distribution $p$ over
$X\times Y$ with property (\ref{probdist}).

The {\em expected error\/} of $V(f(\cdot),\cdot)$ may then be defined as:
$$
E\big[ V( f(\cdot),\cdot) \big]
= \int_{X\times Y} V\big( f(\mathbf{x},y \big)\,dp(\mathbf{x},y)\,.
$$
\color{red}{
Compare this last formula with your formula (3.3).
Note that in your formula it doesn't make sense to have the variables
$\mathbf{x}$ and $y$ on the left-hand-side since this variebles are
used in the integration on the righ-hand-side, hence they are 
``destroyed'' by the integration and they are not available any more:
the whole formul IS NOT a function of $\mathbf{x}$ and $y$.

Also, in these formulas adopt the convention of writing the variable
$\mathbf{x}$ in {\bf boldface} to recall that this variable is a
vector in $X=\R^m$.

}

%%%%%%% DRAFT PROPOSAL %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\vspace{4ex}
\item
{\color{blue}
{\Large\bf A draft proposal.} \\[1ex]
The basic setting for supervised learning is a {\em data set\/} $X$, 
and an  {\em outcome set\/} $Y$.
For our purposes $X$ will be $\R^m$, $m\in\N$, and $Y$ either $\R$
in the case of regression problems, or $\{-1,1\}$ in the case of
classification problems.
In addition we have a plurality of {\em training sets\/}
$$
S=\bigg\{ (\mathbf{x}_k,y_k)\in X\times Y \mid k=1,\dots,n \bigg\}
\subseteq X\times Y\,,
$$
where the members of $S$ have been drawn randomly and independently
from $X\times Y$ according to an {\em unknown\/} joint distribution
function $p(\mathbf{x},y)$ on $X\times Y$.
We gather all these training sets in a subset $\mathscr{T}$ of
$\mathscr{P}(X\times Y)$.

{\em Supervised learning\/} consists then in finding a
{\em learning function\/} $f:X\to Y$ such that:

%% \begin{itemize}
%% \item
%% it \red{models} the training set $S$ in the sense that $f(x_k)=y_k$
%% for all $k=1,\dots,n$ and
%% \item
%% it has a \red{predictive capacity} in the sense that $f(x)=y$ for a
%% newly arrived data $(x,y)\not\in S$.
%% \end{itemize}
%% These conditions seem to be too restrictive.
%% We can weaken them a little by accepting that the function $f:X\to Y$ should:

\begin{itemize}
\item
it closely models the training set $S$ in the sense that
$f(\mathbf{x}_k)\approx y_k$ with an error
$|f(\mathbf{x}_k)-y_k|<\varepsilon$ for a given $\varepsilon>0$
and all $k=1,\dots,n$;
\red{Is this true? If not, then in what sense $f$ models $S$ ?}
\item
it has an approximate predictive capacity in the sense that
$f(\mathbf{x})\approx y$ with error $|f(\mathbf{x})-y|<\varepsilon$
for a newly arrived data $(\mathbf{x},y)\in X\times Y\setminus S$.
\end{itemize}
The learning function $f:X\to Y$ is to be searched after among a set
of functions $\mathscr{H}$ from $X$ to $Y$.
The set of functions $\mathscr{H}$ is called {\em hypothesis space\/}.

%% For the search process it is convenient to consider a vector space
%% structure and a topology on the hypothesis space $\mathscr{H}$.
%% A popular topology in the field is the $L^p$-norm topology for some
%% $p\in\R$ with $1\leq p\leq \infty$.
%% Topological completeness of $\mathscr{H}$ is convenient as well.
%% A convenient comprehensive requirement would be that $\mathscr{H}$ be
%% a Banach space.

In addition, there is a {\em learning algorithm\/} $\mathcal{A}$ that
associates to every data set one and only one learning function
$f\in\mathscr{H}$.
Thus $\mathcal{A}$ can be considered as a function:
$$
\mathcal{A}:\mathscr{T}\to\mathscr{H}\,,\qquad
S\mapsto\mathcal{A}(S)=f\quad\forall S\in\mathscr{T}\,.
$$

PLEASE CONTINUE HERE WITH THE PERTINENT REMARKS WRITTEN ABOVE !

}

\end{enumerate}

\end{description}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\red{\section{\Large AUGUST 19, 2017 -- Chapter 3 once again!}}

\def\bfx{\mathbf{x}}

\begin{description}[style=unboxed,leftmargin=0cm,itemsep=3ex]

\myitem{P 37, L 3}
It says: ``\dots selection of $f$ is based\dots''.
It should say: ``\dots selection of $f\in\mathscr{H}$ is based\dots''.

\myitem{P 37, Eqs. (3.1) and (3.2)}
\mbox{}\newline
It is NOT advisable to put 2 equation numbers for this only one function!

I guess you now know that the $V$ in (3.1) is wrong.

Moreover, when you write ``$\big(f(\bfx),y\big)\mapsto V\big(f(\bfx),y\big)$''
in (3.2), you are saying that the {\em domain\/} of your function is
$f(X)\times Y$ and not $X\times Y$, where $f(X)\subseteq Y$ but it may
be very well possible that $f(X)\neq Y$ ($f$ could be not {\em onto\/}).

Thus, in $(3.2)$ you are tacitly assuming that the function is defined on
$f(X)\times Y$, where $f(x)\subseteq Y$.

For equation (3.2) to be correct you should write:
\begin{equation}\label{20170819:1}
V:f(X)\times Y\to\R_0^+\,,\qquad
\big(f(\bfx),y\big)\mapsto V\big(f(\bfx),y\big)\,.
\end{equation}

Note the difference between \verb|\to|: $\to$\,, 
and \verb|\mapsto|: $\mapsto$\,.

Note that $\R^+:=\big\{ x\in\R\mid x>0 \big\}$ and
$\R_0^+:=\big\{ x\in\R\mid x\geq0 \big\}$.
$V(f(x),y)$ can be 0!
Hence, instead of $\R^+$ you should write $\R_0^+$.

Here is my suggestion for (3.1), (3.2):
\begin{equation}\label{20170819:2}
V(f(\cdot),\cdot):X\times Y\to\R_0^+\,,\qquad
(\bfx,y)\mapsto V\big(f(\bfx),y\big)
\end{equation}
Note that, although the {\em values\/} of the functions (\ref{20170819:1})
and (\ref{20170819:2}) are the same, namely $V\big(f(\bfx),y\big)$, the
{\em functions\/} $V$ and $V(f(\cdot),\cdot)$ are {\em very\/} different.

\myitem{P 37, L -6}
Formula\quad
$\displaystyle\widehat{f}
=\underset{f\in\mathscr{H}}{\text{arg min\ }} R_{\text{emp}}[f]$\,.

Don't use \verb|\hat| because the hat is too small: $\hat{f}$ !\quad
Use \verb|\widehat| instead: $\widehat{f}$.

You are using \verb|\mathcal{H}|: $\mathcal{H}$.
I like more \verb|\mathscr{H}|: $\mathscr{H}$.
But this is a matter of taste!

\myitem{P 37, formulae (3.3) and (3.6)}
Formula (3.6) tacitly assumes that the distribution $p$ is actually constant
although in L. -9 you wrote: 
``Since the true distribution is {\em unknown\/}...''

There is a contradiction here! Please clarify!

The Riemann sum approximmation of the integral in (3.3) is:
$$
\sum_{i=1}^n V\big( f(\bfx_i,y_i \big)\,p(\bfx_i,y_i)\,\frac1n\,,
$$
which is different fron your formula (3.6).
This formula would coincides with your formula (3.6) in the case
$$
p(\bfx_i,y_i)=1\quad \text{for all}\ i=1,\dots,n\,.
$$
You are tacitly assuming that your data set here is
$S=\big\{ (\bfx_i,y_i)\mid i=1,\dots,n\big\}\in\mathscr{T}$ and
that $p(\bfx,y)=0$ for all $(\bfx,y)\in X\times Y\setminus S$.
Thus everything would be OK for this particular distribution $p$.

\myitem{P 37, formula (3.7)}
The right hand side of formula (3.7) is wrong:
it cannot contain variables (like $\bfx\in X$ or $y\in Y$) because these
would be ``destroyed'' by the operator $E$, producing numerical values.
This operator $E$ acts on {\em functions\/} like $f$ or $V(f(\cdot),\cdot)$.

This formula should be as follows:
$$
\frac1n\,\sum_{i=1}^n V\big( f(\bfx_i,y_i \big)
\overset{n\to\infty}{\longrightarrow} E\big[ V(f(\cdot),\cdot)\big]\,.
$$

\myitem{P 38, Complexity \dots of the hypothesis class + Fig. 3.1}
\mbox{}\newline
It seems that in the left diagram in Fig. 3.1 the data set is
$S=\{z_1,z_2,z_3\}$.
I guess that the possible ``dichotomies'' are the following $3$: 
$D_1=\big\{ \{z_1\}, \{z_2,z_3\} \big\}$,
$D_2=\big\{ \{z_2\}, \{z_1,z_3\} \big\}$,
$D_3=\big\{ \{z_3\}, \{z_1,z_2\} \big\}$.
There are no other.
The two components of each one of the 3 dichotomies $D_1, D_2,D_3$ can
be separated (i.e., shattered) by one line in the {\em plane\/}.

On the right diagram of Fig. 3.1 the data set woul be $S=\{z_1,z_2,z_3,z_4\}$.
The possible dichotomies are then the following $7$:
$D_1=\big\{ \{z_1\}, \{z_2,z_3,z_4\} \big\}$,
$D_2=\big\{ \{z_2\}, \{z_1,z_3,z_4\} \big\}$,
$D_3=\big\{ \{z_3\}, \{z_1,z_2,z_4\} \big\}$.
$D_4=\big\{ \{z_4\}, \{z_1,z_2,z_4\} \big\}$.
$D_5=\big\{ \{z_1,z_2\}, \{z_3,z_4\} \big\}$,
$D_6=\big\{ \{z_1,z_3\}, \{z_2,z_4\} \big\}$,
$D_7=\big\{ \{z_1,z_4\}, \{z_2,z_3\} \big\}$,
If the 4 data $z_1,z_2,z_3,z_4$ are arranged as the vertices of a square
with sides $z_1z_2$, $z_2z_3$, $z_3z_4$, $z_4z_1$ on a plane, then it is
impossible to separate the dichothomy
$D_6=\big\{ \{z_1,z_3\}, \{z_2,z_4\} \big\}$
by means of a straight line on the plane.

In the left side of Fig. 3.1 you say that you shatter {\em the three vectors\/}
by {\em lines\/} in a plane.
You should say that what you separate are the 3 {\em dichotomies\/}.
Dichotomy is the crucial concept here!

In the right side of Fig. 3.1 yo say that is is impossible to shatter appart
the 4 vectors by means of a {\em plane\/}.
This is misleading.
You should talk of {\em straight lines\/}

Both diagrams, left and right, are planar, 2-dimensional diagrams.
So you cannot talk about planes on a plane diagram.

The names $z^1$ and $z^2$ of the coordinate axis, are misleading.
It is better to change them by $x$ and $y$, so that the vectors are
2-dimensional: $z_k=(x_k,y_k)$,  $k=1,2,3,4$.
2-dimensional vectors are enough to explain the concept.

Your definition of the VC-dimension is very obscure.
Compare with the following transcription from Wikipedia
(\texttt{https://en.wikipedia.org/wiki/VC\_dimension}),
where the classification model $f$ is a straight line:

\vspace{2ex}
\red{{\em
``$f$ is a straight line as a classification model on points in a
two-dimensional plane (this is the model used by a perceptron). 
The line should separate positive data points from negative data points. 
There exist sets of 3 points that can indeed be shattered using this
model (any 3 points that are not collinear can be shattered). 
However, no set of 4 points can be shattered: by Radon's theorem,
any four points can be partitioned into two subsets with intersecting
convex hulls, so it is not possible to separate one of these two subsets
from the other. 
Thus, the VC-dimension of this particular classifier is 3.''
}}

\vspace{3ex}
This is a clear description in order to understand the concept of
VC-dimension.
Don't forget the references!
But, of course, in a PhD thesis (or in a paper) it is NOT vey wise to
quote Wikipedia.
You must go to the original sources.

A key point is Radon's theorem in discrete geometry
($\text{co}(A)$ denotes the convex hull of $A$):

\begin{theorem}[J. Radon]
Let $S\subset\R^d$ be a subset containing at least $d+2$ elements. 
Then there exist {\bf disjoint} subsets $A\subset S$ and $B\subset S$
such that
$$
\text{co}(A)\cap\text{co}(B)\neq\emptyset
$$
\end{theorem}
\noindent
Ref.: B.B. Peterson, {\em ``The Geometry of Radon's Theorem''\/}, The
American Mathematical Monthly, Vol. 9, No. 9 (Nov., 1972), 949-963.

\vspace{2ex}
Radon's theorem implies that in $\R^d$ any set of $d+2$ or more points
(vectors!) has at least one dichotomy which cannot be separated (shattered)
by any plane in $\R^d$.

Thus the VC-dimension in $\R^d$ cannot be greater than $d+1$.

\end{description}

%% END OF FILE

### 2017-08-27 comentarios-tesis/lecturadetesis18-09/20161019_Paola_Chapter_3_Part_A.tex
last changes to chapter 3

--- 2016-10-20 comentarios-tesis/20161009_Wuerzburg_Paolas Thesis Reviewing.tex
@@
 
-\documentclass[12pt,reqno]{amsart}
-\usepackage{amsfonts,amssymb,amsmath,amsthm,mathrsfs,
-            amsopn,amsxtra,amstext,amsbsy}
-\usepackage{epsfig}
-\usepackage{array}
-\usepackage{url}
-\usepackage{tikz}
-%% \usepackage[spanish]{babel}
-\usepackage[utf8x]{inputenc}
-\usepackage{fancyhdr}
-\usepackage{graphicx}
-\usepackage{hyperref,wasysym}
-\usepackage{xcolor}
-\usepackage{lipsum}
-\usepackage{enumerate}
-\usepackage{enumitem}% http://ctan.org/pkg/enumitem
-\usepackage{verbatim}
+%% \documentclass[12pt,reqno]{amsart}
+%% \usepackage{amsfonts,amssymb,amsmath,amsthm,mathrsfs,
+%%             amsopn,amsxtra,amstext,amsbsy}
+%% \usepackage{epsfig}
+%% \usepackage{array}
+%% \usepackage{url}
+%% \usepackage{tikz}
+%% %% \usepackage[spanish]{babel}
+%% \usepackage[utf8x]{inputenc}
+%% \usepackage{fancyhdr}
+%% \usepackage{graphicx}
+%% \usepackage{hyperref,wasysym}
+%% \usepackage{xcolor}
+%% \usepackage{lipsum}
+%% \usepackage{enumerate}
+%% \usepackage{enumitem}% http://ctan.org/pkg/enumitem
+%% \usepackage{verbatim}
 
-\voffset=-10mm
-\oddsidemargin=0pt
-\evensidemargin=0pt
-\topmargin=-10mm
-\headsep=18pt
-\headheight=18pt
-\footskip=30pt
-\topskip=0mm
-\textwidth 167mm
-\textheight=230mm
-\parindent=0em
-\marginparsep=0mm
-\marginparwidth=0mm
-\vfuzz2pt % Don't report over-full v-boxes if over-edge is small
-\hfuzz6pt % Don't report over-full h-boxes if over-edge is small
+%% \voffset=-10mm
+%% \oddsidemargin=0pt
+%% \evensidemargin=0pt
+%% \topmargin=-10mm
+%% \headsep=18pt
+%% \headheight=18pt
+%% \footskip=30pt
+%% \topskip=0mm
+%% \textwidth 167mm
+%% \textheight=230mm
+%% \parindent=0em
+%% \marginparsep=0mm
+%% \marginparwidth=0mm
+%% \vfuzz2pt % Don't report over-full v-boxes if over-edge is small
+%% \hfuzz6pt % Don't report over-full h-boxes if over-edge is small
 
-% \newtheorem{conjecture}[theorem]{Conjecture}
-\newtheorem{definicion}{Definición}
-% \newtheorem{theorem}{Theorem}%[section]
-% \newtheorem{lemma}[theorem]{Lemma}
-% \newtheorem{corollary}[theorem]{Corollary}
-% \newtheorem{conjecture}[theorem]{Conjecture}
-% \theoremstyle{definition}
-% \newtheorem{definition}[theorem]{Definition}
-% \newtheorem{example}[theorem]{Example}
-% \newtheorem{xca}[theorem]{Exercise}
-% \theoremstyle{remark}
-% \newtheorem{remark}[theorem]{Remark}
-% \numberwithin{equation}{section}
+%% % \newtheorem{conjecture}[theorem]{Conjecture}
+%% \newtheorem{definicion}{Definición}
+%% % \newtheorem{theorem}{Theorem}%[section]
+%% % \newtheorem{lemma}[theorem]{Lemma}
+%% % \newtheorem{corollary}[theorem]{Corollary}
+%% % \newtheorem{conjecture}[theorem]{Conjecture}
+%% % \theoremstyle{definition}
+%% % \newtheorem{definition}[theorem]{Definition}
+%% % \newtheorem{example}[theorem]{Example}
+%% % \newtheorem{xca}[theorem]{Exercise}
+%% % \theoremstyle{remark}
+%% % \newtheorem{remark}[theorem]{Remark}
+%% % \numberwithin{equation}{section}
 
-% \def\phi{\varphi}
-% \def\epsilon{\varepsilon}
-\def\RE{\mbox{\rm Re}\,}
-\def\IM{\mbox{\rm Im}\,}
-% \def\AZ{|z|}
-\newcommand{\rectangle}{\fboxsep0pt\fbox{\rule{0.3em}{0pt}\rule{0pt}{1.5ex}}}
+%% % \def\phi{\varphi}
+%% % \def\epsilon{\varepsilon}
+%% \def\RE{\mbox{\rm Re}\,}
+%% \def\IM{\mbox{\rm Im}\,}
+%% % \def\AZ{|z|}
+%% \newcommand{\rectangle}{\fboxsep0pt\fbox{\rule{0.3em}{0pt}\rule{0pt}{1.5ex}}}
 
-%-------------------------mathcal
+%% %-------------------------mathcal
 
-\def\A{{\mathcal A }}
-\def\CC{{\mathcal C }}
-\def\CH{{\mathcal H }}
-\def\CK{{\mathcal K }}
-\def\CM{{\mathcal M }}
-\def\CP{{\mathcal P }}
-\def\CS{{\mathcal S }}
-\def\CR{{\mathcal R}}
-\def\CRu{{\mathcal R}^{\rm u}}
-\def\CT{{\mathcal T}}
-\def\CTT{\widetilde{{\mathcal T}}}
-\def\CW{{\mathcal W}}
-\def\Li{{\mbox{\rm Li}}}
-\def\WO{{\mathcal W}_{\Omega }}
+%% \def\A{{\mathcal A }}
+%% \def\CC{{\mathcal C }}
+%% \def\CH{{\mathcal H }}
+%% \def\CK{{\mathcal K }}
+%% \def\CM{{\mathcal M }}
+%% \def\CP{{\mathcal P }}
+%% \def\CS{{\mathcal S }}
+%% \def\CR{{\mathcal R}}
+%% \def\CRu{{\mathcal R}^{\rm u}}
+%% \def\CT{{\mathcal T}}
+%% \def\CTT{\widetilde{{\mathcal T}}}
+%% \def\CW{{\mathcal W}}
+%% \def\Li{{\mbox{\rm Li}}}
+%% \def\WO{{\mathcal W}_{\Omega }}
 
-%-------------------------------Bbb
-\def\K{{\mathbb K }}
-\def\C{{\mathbb C }}
-\def\N{{\mathbb N }}
-\def\NO{{\mathbb{N}_0 }}
-\def\Z{{\mathbb Z }}
-\def\R{{\mathbb R }}
-\def\U{{\mathbb U }}
-\def\D{{\mathbb D }}
-%----------------------------Zusammengesetztes
+%% %-------------------------------Bbb
+%% \def\K{{\mathbb K }}
+%% \def\C{{\mathbb C }}
+%% \def\N{{\mathbb N }}
+%% \def\NO{{\mathbb{N}_0 }}
+%% \def\Z{{\mathbb Z }}
+%% \def\R{{\mathbb R }}
+%% \def\U{{\mathbb U }}
+%% \def\D{{\mathbb D }}
+%% %----------------------------Zusammengesetztes
 
-\def\DB{\overline{\D}}
-\def\Db#1{\overline{\D_{#1}}}
-\def\DBR{\overline{\D_\rho}}
-\def\DV{de la Vall\'ee Poussin }
-\def\d#1{\D_{#1}}
-\def\H{{\mathcal H}(\D)}
-\def\Hb{{\mathcal H}(\DB)}
-\def\HO{{\mathcal H}_0(\D )}
-\def\Hoo#1{{\mathcal H}(\overline{\D_{#1}})}
-\def\HB#1{{\mathcal H}(\D_{#1})}
-\def\Ho#1{{\mathcal H}({#1})}
-\def\Hyp{{}_2F_1}
+%% \def\DB{\overline{\D}}
+%% \def\Db#1{\overline{\D_{#1}}}
+%% \def\DBR{\overline{\D_\rho}}
+%% \def\DV{de la Vall\'ee Poussin }
+%% \def\d#1{\D_{#1}}
+%% \def\H{{\mathcal H}(\D)}
+%% \def\Hb{{\mathcal H}(\DB)}
+%% \def\HO{{\mathcal H}_0(\D )}
+%% \def\Hoo#1{{\mathcal H}(\overline{\D_{#1}})}
+%% \def\HB#1{{\mathcal H}(\D_{#1})}
+%% \def\Ho#1{{\mathcal H}({#1})}
+%% \def\Hyp{{}_2F_1}
 
-\def\achtung#1{\alert{#1}}
-\def\blue#1{{\color{blue} #1}}
-\def\green#1{{\color{green} #1}}
-\def\red#1{{\color{red} #1}}
+%% \def\achtung#1{\alert{#1}}
+%% \def\blue#1{{\color{blue} #1}}
+%% \def\green#1{{\color{green} #1}}
+%% \def\red#1{{\color{red} #1}}
 
-\def\pick{{\mathscr P}}
-\def\logpick{{\text{log}\,{\mathscr P}}}
-\def\stammpick{\pick_{\!\!\!\raisebox{-2pt}{\scalebox{0.5}{$\int$}}}}
-\def\stammpickpick{\stammpick\cap\pick}
-\def\stammpicklogpick{\stammpick\cap\logpick}
-\def\expstammpicklogpick{e^{\stammpicklogpick}}
+%% \def\pick{{\mathscr P}}
+%% \def\logpick{{\text{log}\,{\mathscr P}}}
+%% \def\stammpick{\pick_{\!\!\!\raisebox{-2pt}{\scalebox{0.5}{$\int$}}}}
+%% \def\stammpickpick{\stammpick\cap\pick}
+%% \def\stammpicklogpick{\stammpick\cap\logpick}
+%% \def\expstammpicklogpick{e^{\stammpicklogpick}}
 
-%-------------------------Bakan-Ruscheweyh-Salinas 2015
-\def\mid{\;:\ }
-\def\hol#1{\text{Hol}(#1)}
-\def\normhol#1#2{\text{Hol}_{#1}(#2)}
+%% %-------------------------Bakan-Ruscheweyh-Salinas 2015
+%% \def\mid{\;:\ }
+%% \def\hol#1{\text{Hol}(#1)}
+%% \def\normhol#1#2{\text{Hol}_{#1}(#2)}
 
-\DeclareMathOperator*{\cosig}{cosig}
-\DeclareMathOperator*{\supp}{supp}
+%% \DeclareMathOperator*{\cosig}{cosig}
+%% \DeclareMathOperator*{\supp}{supp}
 
-%graficos
-%\pgfdeclareimage[width=5cm, height=3cm ]{recursiva}{./Png/recursiva.png}
-%\pgfdeclareimage[width=5cm, height=5cm ]{alan}{./Png/Turing.jpeg}
-%\pgfdeclareimage[width=5cm, height=5cm ]{programa}{./Png/maquina.png}
-%\pgfuseimage{lala}
+%% %graficos
+%% %\pgfdeclareimage[width=5cm, height=3cm ]{recursiva}{./Png/recursiva.png}
+%% %\pgfdeclareimage[width=5cm, height=5cm ]{alan}{./Png/Turing.jpeg}
+%% %\pgfdeclareimage[width=5cm, height=5cm ]{programa}{./Png/maquina.png}
+%% %\pgfuseimage{lala}
 
-\renewcommand{\restriction}{\mathord{|}}
+%% \renewcommand{\restriction}{\mathord{|}}
 
-\newcommand\blfootnote[1]{%
-  \begingroup
-  \renewcommand\thefootnote{}\footnote{#1}%
-  \addtocounter{footnote}{-1}%
-  \endgroup }
+%% \newcommand\blfootnote[1]{%
+%%   \begingroup
+%%   \renewcommand\thefootnote{}\footnote{#1}%
+%%   \addtocounter{footnote}{-1}%
+%%   \endgroup }
 
-\title[Paola's Thesis Comments]{Paola's Thesis Comments}
+%% \title[Paola's Thesis Comments]{Paola's Thesis Comments}
 
-\author[L. Salinas]{Luis Salinas}
+%% \author[L. Salinas]{Luis Salinas}
 
-\address{CCTVal and Departamento de Inform\'atica, UTFSM, Valpara\'\i so;
-         \newline
-         Mathematisches Institut, Universit\"at W\"urzburg;
-         \newline
-         \today} % October 06, 2016}
+%% \address{CCTVal and Departamento de Inform\'atica, UTFSM, Valpara\'\i so;
+%%          \newline
+%%          Mathematisches Institut, Universit\"at W\"urzburg;
+%%          \newline
+%%          \today} % October 06, 2016}
 
-\email{luis.salinas@usm.cl}
+%% \email{luis.salinas@usm.cl}
 
-\begin{document}
+%% \begin{document}
 
-\maketitle%
+%% \maketitle%
 
-\section{Chapter 3: Machine Learning Models.}
+%% \section{Chapter 3: Machine Learning Models.}
 
-\def\myitem#1{ \item[\blue{\bf #1}] }
+%% \def\myitem#1{ \item[\blue{\bf #1}] }
+
+\section{Chapter 3: REMARKS FROM 20161019}
 
@@
 
-\begin{thebibliography}{99}
-\bibitem{Arce-2016}
-Paola Arce, 
-{\em Online Learning Method for Financial Time Series Forecasting\/}.
-Dissertation, UTFSM, June 2016.
-
-\end{thebibliography}
-
-\end{document}
+%% \end{document}

### 2017-08-27 comentarios-tesis/lecturadetesis18-09/20161031_Paola_Chapter_3_Part_B.tex
last changes to chapter 3

% 20161009_Paola_Chapter_3_Part_B.tex --- 31.10.2016

\section{Chapter 3: Remarks from 20161031}

\begin{description}[style=unboxed,leftmargin=0cm,itemsep=3ex]

\myitem{P 39, LL 11-13, \red{3.2.4 Ridge Regression. 
        Discussion of the equation for $J(\mathbf{X})$}}
I discuss here the equation for $J(\mathbf{X})$ starting from equation (3.6).
For any matrix $\mathbf{A}$ let 
$\mathbf{A}_i$ be the $i$-th row,
$\mathbf{A}^j$ be the $j$-th column, and
$\mathbf{A}_i^j$ be the $(i,j)$-element of matrix the $\mathbf{A}$.
Then:
\begin{alignat*}{2}
&\text{$(i,j)$-element of the \red{left} hand side of (3.6)}
= \mathbf{A}_i\;\mathbf{X}^j\,, &\quad& i=1:n\,,\ j=1:l\,, \\
&\text{$(i,j)$-element of the \red{right} hand side of (3.6)}
= \mathbf{Y}_i^j\,, &\quad& i=1:n\,,\ j=1:l\,.
\end{alignat*}
If $\|\cdot\|_2$ denotes the Frobenius matrix norm, we would have:
\begin{equation}\label{20161031:1}
J(\mathbf{X})
= \|\mathbf{A}\,\mathbf{X}-\mathbf{Y}\|_2^2
= \sum_{j=1}^l \left\| \mathbf{A}\,\mathbf{X}^j - \mathbf{Y}^j \right\|_2^2
= \sum_{i=1}^m \sum_{j=1}^l 
  \left( \mathbf{A}_i\,\mathbf{X}^j - \mathbf{Y}_i^j \right)^2
\end{equation}
Now the first problem is the meaning that you assign to the equality
\begin{equation}\label{20161031:2}
J(\mathbf{X})=\sum_{t=1}^N \left(f(\mathbf{x}_t)-y_t\right)^2\,.
\end{equation}
Everything is undefined here: $t,\,N,\,f,\,\mathbf{x}_t,y_t$.
Thus, the equality does not make sense.
We have to find a plausible interpretation.
It might be that the term with the sum in the right hand side of
(\ref{20161031:2}) concides with the last term in (\ref{20161031:1}).
In this interpretation, 
\begin{equation}\label{20161031:3}
f(\mathbf{x}_t) \quad\text{would be}\quad 
f_i(\mathbf{X}^j):=\mathbf{A}_i\,\mathbf{X}^j\,,
\end{equation}
but you would still need two subindices: $i$ and $j$.

One could, of course, introduce a numbering $t=j+(i-1)\,l$ of the
coefficients $(i,j)$ involved, and hence $N=i\times j$, but then you
would need expresions of $i$ and $j$ in terms of $t$, which are easy
to obtain but makes everything too cumbersome.

Another problem is the matrix $\mathbf{A}$ which suddenly appears here,
without definition.
One might think that $\mathbf{A}$ is just $\mathbf{X}^T$,
but then there are problems with the dimension of the matrices in (3.6).

Summarizing, you need to give a clear presentation for the various
expressions of the functional $J(\mathbf{X})$.

\myitem{P 39, LL 15-19, Eqs. (3.6), (3.7), (3.8)}
Here you start from equation (3.6), $\mathbf{A}\,\mathbf{X}=\mathbf{Y}$,
but you have not defined these matrices, except for their dimensions.

Then you say something about the \red{optimal} solution of (3.6).
What do you mean by \red{optimal} here?

An equation, alone, like (3.6), cannot have an \red{optimal} solution.
If (3.6) has many solutions, then some of them might be optimal from
some point of view, regarding --for instance-- some particular mimisation
or maximisation problems, but they could be not optimal for other
minimax problems.
If (3.6) has no solution, then still one can find some $\mathbf{X}$'s
such that $\mathbf{A}\,\mathbf{X}$ is as close as possible to $\mathbf{Y}$
in some appropriate sense.
These are dubbed \red{pseudo-solutions}.
One of these \red{pseudo-solutions} is obtained via More-Penrose.
This is clear when $\mathbf{X}$ and $\mathbf{Y}$ are vectors.
But what is the situation when $\mathbf{X}$ and $\mathbf{Y}$ are matrices
like here?

Then in (3.7) and (3.8) you talk about $\mathbf{A}^+$ and give the usual
formula when $\mathbf{X}$ and $\mathbf{Y}$ are vectors.
Holds this formula when $\mathbf{X}$ and $\mathbf{Y}$ are matrices

Is all this related to the minimisation problem for $J(\mathbf{X})$?
If yes, you should give proofs!

In an \red{Appendix} you should provide proofs of all claims involving
(3.6), (3.7), (3.8).

\myitem{From P 39, L -7, until P 40, L 3. 
\red{3.2.4 Ridge Regression. THIS IS AN EXPANDED VERSION OF THE LAST
ITEM FROM PART A}}
You present this paragraph as someting called ``Demo''.
What is it? A proof? A proof of what? 
What is your claim before starting with ``Demo''?

\red{It seems that your ``Demo'' has to do with the case when
$\mathbf{A}^T\mathbf{A}$ is not invertible and then you introduce a
regularisation by means of a parameter $\lambda$.
But before your ``Demo'' there is no mention of this fact.
Please clarify!}

You say ``to solve equation (3.8)''.
But (3.8) is just the definition of More-Penrose!

Are you trying to solve (3.6)?
If this is true, what is known and what is not known?
Moreover, what do you mean by \red{solving (3.6)}, specially in the
case (as in your case) when the equation is \red{overdetermined}?

You should prove the equivalence you claim.
At least write $\mathbf{C}_\lambda$ to make explicit the dependence of
$\mathbf{C}$ from the parameter $\lambda$.

I think you should better write something like:
$$
(3.9)\rule{50mm}{0pt} \min_{\mathbf X \in \text{\red{???}}}
\|\mathbf{C}_\lambda\,\mathbf{X}-\mathbf{F}\|
$$
You should make clear in which space $\mathbf{X}$ is running for the
optimisation.

The outcome of this minimisation problem depends on $\lambda$, but
equation (3.6) does not depend on $\lambda$.

Thus, why is (3.6) equivalent to the minimization problem (3.9)?
A proof is needed.

Which value of $\lambda$ will you consider for a solution of (3.6) and why?

What is $\mathbf{A}$ and what is the rationale for introducing $\mathbf{C}$
and $\mathbf{F}$.

The equation for $\mathbf{X}(\lambda)$ in LL 1, 2 in P 40, seems to be a
More-Penrose solution of $\mathbf{C}_\lambda\,\mathbf{X}=\mathbf{F}$.
But why this solution coincides with the solution --if it exists-- of the
minimisation problem?

By the way, is it sure that the minimisation Problem has a solution?
Is it unique? etc.

\vspace{3ex}
We are given
an $m\times n$ matrix of known real coefficients $\mathbf{A}$,
an $n\times l$ matrix of real unknowns $\mathbf{X}$,
an $m\times l$ matrix of real known data $\mathbf{Y}$.

Our problem is to find a matrix $\mathbf{X}$ that minimises the functional
\begin{equation}
J(\mathbf{X}):=\|\mathbf{A}\,\mathbf{X}-\mathbf{Y}\|_2^2\,,
\end{equation}
where $\|\cdot\|_2$ is the Frobenius matrix norm.
Any matrix $\mathbf{X}$ that minimizes $J(\mathbf{X})$ will be called a
{\em minimum\/} of $J(\mathbf{X})$.

\vspace{3ex}
{\bf Claim}. 
1. A minimum $\mathbf{X}_0$ of $J(\mathbf{X})$ always exist. \\
1. If $\mathbf{X}_0$ is a minimum of $J(\mathbf{X})$, then
$\mathbf{X}_0+\mathbf{X}_1$ is also a minimum of $J(\mathbf{X})$
if and only if: 
\begin{equation}
\mathbf{X}_1\in \mathbf{A}^{\perp}:=
\left\{ \mathbf{X}\in\R^n\times\R^l \mid
        \mathbf{A}\,\mathbf{X}=\mathbf{0} \right\}\,.
\end{equation}
3. If the $n\times n$ matrix $\mathbf{A}^{\mathsf{T}}\,\mathbf{A}$ is
invertible and $\mathbf{X}$ and $\mathbf{Y}$ are vectors, then 
the minimum $\mathbf{X}_0$ is given by
$\mathbf{A}^{\mathsf{T}}\,\mathbf{A}$

\end{description}

\vspace{2ex}
Reviewing work in progress/DATE: 31.10.2016

%% \end{document}

### 2017-08-27 comentarios-tesis/lecturadetesis18-09/20170814_Paola_Chapter_3_again.tex
last changes to chapter 3

--- 2017-08-27 comentarios-tesis/chapter3ok/20170814_Paola_Chapter_3_again.tex
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\red{\section{\Large AUGUST 19, 2017 -- Chapter 3 once again!}}
-
-\def\bfx{\mathbf{x}}
-
-\begin{description}[style=unboxed,leftmargin=0cm,itemsep=3ex]
-
-\myitem{P 37, L 3}
-It says: ``\dots selection of $f$ is based\dots''.
-It should say: ``\dots selection of $f\in\mathscr{H}$ is based\dots''.
-
-\myitem{P 37, Eqs. (3.1) and (3.2)}
-\mbox{}\newline
-It is NOT advisable to put 2 equation numbers for this only one function!
-
-I guess you now know that the $V$ in (3.1) is wrong.
-
-Moreover, when you write ``$\big(f(\bfx),y\big)\mapsto V\big(f(\bfx),y\big)$''
-in (3.2), you are saying that the {\em domain\/} of your function is
-$f(X)\times Y$ and not $X\times Y$, where $f(X)\subseteq Y$ but it may
-be very well possible that $f(X)\neq Y$ ($f$ could be not {\em onto\/}).
-
-Thus, in $(3.2)$ you are tacitly assuming that the function is defined on
-$f(X)\times Y$, where $f(x)\subseteq Y$.
-
-For equation (3.2) to be correct you should write:
-\begin{equation}\label{20170819:1}
-V:f(X)\times Y\to\R_0^+\,,\qquad
-\big(f(\bfx),y\big)\mapsto V\big(f(\bfx),y\big)\,.
-\end{equation}
-
-Note the difference between \verb|\to|: $\to$\,, 
-and \verb|\mapsto|: $\mapsto$\,.
-
-Note that $\R^+:=\big\{ x\in\R\mid x>0 \big\}$ and
-$\R_0^+:=\big\{ x\in\R\mid x\geq0 \big\}$.
-$V(f(x),y)$ can be 0!
-Hence, instead of $\R^+$ you should write $\R_0^+$.
-
-Here is my suggestion for (3.1), (3.2):
-\begin{equation}\label{20170819:2}
-V(f(\cdot),\cdot):X\times Y\to\R_0^+\,,\qquad
-(\bfx,y)\mapsto V\big(f(\bfx),y\big)
-\end{equation}
-Note that, although the {\em values\/} of the functions (\ref{20170819:1})
-and (\ref{20170819:2}) are the same, namely $V\big(f(\bfx),y\big)$, the
-{\em functions\/} $V$ and $V(f(\cdot),\cdot)$ are {\em very\/} different.
-
-\myitem{P 37, L -6}
-Formula\quad
-$\displaystyle\widehat{f}
-=\underset{f\in\mathscr{H}}{\text{arg min\ }} R_{\text{emp}}[f]$\,.
-
-Don't use \verb|\hat| because the hat is too small: $\hat{f}$ !\quad
-Use \verb|\widehat| instead: $\widehat{f}$.
-
-You are using \verb|\mathcal{H}|: $\mathcal{H}$.
-I like more \verb|\mathscr{H}|: $\mathscr{H}$.
-But this is a matter of taste!
-
-\myitem{P 37, formulae (3.3) and (3.6)}
-Formula (3.6) tacitly assumes that the distribution $p$ is actually constant
-although in L. -9 you wrote: 
-``Since the true distribution is {\em unknown\/}...''
-
-There is a contradiction here! Please clarify!
-
-The Riemann sum approximmation of the integral in (3.3) is:
-$$
-\sum_{i=1}^n V\big( f(\bfx_i,y_i \big)\,p(\bfx_i,y_i)\,\frac1n\,,
-$$
-which is different fron your formula (3.6).
-This formula would coincides with your formula (3.6) in the case
-$$
-p(\bfx_i,y_i)=1\quad \text{for all}\ i=1,\dots,n\,.
-$$
-You are tacitly assuming that your data set here is
-$S=\big\{ (\bfx_i,y_i)\mid i=1,\dots,n\big\}\in\mathscr{T}$ and
-that $p(\bfx,y)=0$ for all $(\bfx,y)\in X\times Y\setminus S$.
-Thus everything would be OK for this particular distribution $p$.
-
-\myitem{P 37, formula (3.7)}
-The right hand side of formula (3.7) is wrong:
-it cannot contain variables (like $\bfx\in X$ or $y\in Y$) because these
-would be ``destroyed'' by the operator $E$, producing numerical values.
-This operator $E$ acts on {\em functions\/} like $f$ or $V(f(\cdot),\cdot)$.
-
-This formula should be as follows:
-$$
-\frac1n\,\sum_{i=1}^n V\big( f(\bfx_i,y_i \big)
-\overset{n\to\infty}{\longrightarrow} E\big[ V(f(\cdot),\cdot)\big]\,.
-$$
-
-\myitem{P 38, Complexity \dots of the hypothesis class + Fig. 3.1}
-\mbox{}\newline
-It seems that in the left diagram in Fig. 3.1 the data set is
-$S=\{z_1,z_2,z_3\}$.
-I guess that the possible ``dichotomies'' are the following $3$: 
-$D_1=\big\{ \{z_1\}, \{z_2,z_3\} \big\}$,
-$D_2=\big\{ \{z_2\}, \{z_1,z_3\} \big\}$,
-$D_3=\big\{ \{z_3\}, \{z_1,z_2\} \big\}$.
-There are no other.
-The two components of each one of the 3 dichotomies $D_1, D_2,D_3$ can
-be separated (i.e., shattered) by one line in the {\em plane\/}.
-
-On the right diagram of Fig. 3.1 the data set woul be $S=\{z_1,z_2,z_3,z_4\}$.
-The possible dichotomies are then the following $7$:
-$D_1=\big\{ \{z_1\}, \{z_2,z_3,z_4\} \big\}$,
-$D_2=\big\{ \{z_2\}, \{z_1,z_3,z_4\} \big\}$,
-$D_3=\big\{ \{z_3\}, \{z_1,z_2,z_4\} \big\}$.
-$D_4=\big\{ \{z_4\}, \{z_1,z_2,z_4\} \big\}$.
-$D_5=\big\{ \{z_1,z_2\}, \{z_3,z_4\} \big\}$,
-$D_6=\big\{ \{z_1,z_3\}, \{z_2,z_4\} \big\}$,
-$D_7=\big\{ \{z_1,z_4\}, \{z_2,z_3\} \big\}$,
-If the 4 data $z_1,z_2,z_3,z_4$ are arranged as the vertices of a square
-with sides $z_1z_2$, $z_2z_3$, $z_3z_4$, $z_4z_1$ on a plane, then it is
-impossible to separate the dichothomy
-$D_6=\big\{ \{z_1,z_3\}, \{z_2,z_4\} \big\}$
-by means of a straight line on the plane.
-
-In the left side of Fig. 3.1 you say that you shatter {\em the three vectors\/}
-by {\em lines\/} in a plane.
-You should say that what you separate are the 3 {\em dichotomies\/}.
-Dichotomy is the crucial concept here!
-
-In the right side of Fig. 3.1 yo say that is is impossible to shatter appart
-the 4 vectors by means of a {\em plane\/}.
-This is misleading.
-You should talk of {\em straight lines\/}
-
-Both diagrams, left and right, are planar, 2-dimensional diagrams.
-So you cannot talk about planes on a plane diagram.
-
-The names $z^1$ and $z^2$ of the coordinate axis, are misleading.
-It is better to change them by $x$ and $y$, so that the vectors are
-2-dimensional: $z_k=(x_k,y_k)$,  $k=1,2,3,4$.
-2-dimensional vectors are enough to explain the concept.
-
-Your definition of the VC-dimension is very obscure.
-Compare with the following transcription from Wikipedia
-(\texttt{https://en.wikipedia.org/wiki/VC\_dimension}),
-where the classification model $f$ is a straight line:
-
-\vspace{2ex}
-\red{{\em
-``$f$ is a straight line as a classification model on points in a
-two-dimensional plane (this is the model used by a perceptron). 
-The line should separate positive data points from negative data points. 
-There exist sets of 3 points that can indeed be shattered using this
-model (any 3 points that are not collinear can be shattered). 
-However, no set of 4 points can be shattered: by Radon's theorem,
-any four points can be partitioned into two subsets with intersecting
-convex hulls, so it is not possible to separate one of these two subsets
-from the other. 
-Thus, the VC-dimension of this particular classifier is 3.''
-}}
-
-\vspace{3ex}
-This is a clear description in order to understand the concept of
-VC-dimension.
-Don't forget the references!
-But, of course, in a PhD thesis (or in a paper) it is NOT vey wise to
-quote Wikipedia.
-You must go to the original sources.
-
-A key point is Radon's theorem in discrete geometry
-($\text{co}(A)$ denotes the convex hull of $A$):
-
-\begin{theorem}[J. Radon]
-Let $S\subset\R^d$ be a subset containing at least $d+2$ elements. 
-Then there exist {\bf disjoint} subsets $A\subset S$ and $B\subset S$
-such that
-$$
-\text{co}(A)\cap\text{co}(B)\neq\emptyset
-$$
-\end{theorem}
-\noindent
-Ref.: B.B. Peterson, {\em ``The Geometry of Radon's Theorem''\/}, The
-American Mathematical Monthly, Vol. 9, No. 9 (Nov., 1972), 949-963.
-
-\vspace{2ex}
-Radon's theorem implies that in $\R^d$ any set of $d+2$ or more points
-(vectors!) has at least one dichotomy which cannot be separated (shattered)
-by any plane in $\R^d$.
-
-Thus the VC-dimension in $\R^d$ cannot be greater than $d+1$.
-
-\end{description}
-
 %% END OF FILE

### 2017-08-27 src/4_OnlineLearning.tex
last changes to chapter 3

@@
   \includegraphics[width=0.8\textwidth]{img/VCdimension}
-  \caption{VC dimension example from \cite{vapnik2013nature}. The left figure
+  \caption{VC dimension example from \cite{vapnik2013nature}. The example shows 2-dimensional vectors $z_k=(x_k,y_k)$,  $k=1,2,3,4$. The left figure
   shows that the 3 dichotomies can be shattered by a line in a plane. However, if we
@@
 will contain every possible example, so the algorithm will generalize perfectly.
-However, for some $f \in H$, the best hypothesis $\hat{f}$ might be far away to
+However, for some $f \in \mathscr{H}$, the best hypothesis $\hat{f}$ might be far away to
 the best hypothesis class but it is the best that all the learners can hope.
-Therefore, to choose the right hypothesis class $H$ is crucial. The error of
+Therefore, to choose the right hypothesis class $\mathscr{H}$ is crucial. The error of
 this best hypothesis is formalised as the {\em approximation or generalisation
@@
 The relationship between the generalisation error and the hypothesis complexity
-is less intuitive. If the class $H$ complexity (VC dimension) increases, then to
+is less intuitive. If the class $\mathscr{H}$ complexity (VC dimension) increases, then to
 mantain an {\em estimation error}, the {\em sample complexity} increases.

### 2017-09-23 src/0_Abstract.tex
intro fix
Conclusions first part

@@
 
-\emph{VECM parameters are obtained using the ordinary least squares (OLS)
-method.  Even though OLS is extensively used, it has over-fitting
-issues and is computationally expensive. Ridge regression is commonly used
-instead of OLS since they include a regularisation parameter which could improve
-prediction error.}
+\emph{Cointegration relationships are found on different time series
+frequencies. Cointegration in low frequency time series is motivated by a
+long-run equilibrium relationship between economic forces, whereas cointegration
+in high frequency data has its foundation in statistical arbitrage theory.
+However, the use of cointegration models such as VECM is mainly limited by
+computationally expensive routines.}
 
 \emph{In this thesis, financial time series features are studied and different
-algorithms were developed in order to optimise parameters and increase performance.
-In particular, I propose an online algorithm based on VECM which optimises how model parameters are obtained and reduces execution times.
-This is achieved considering only a sliding window of the last historical data and using machine learning techniques to solve the model.
-Moreover, the long-run relationship between the time series is used in order to make
-optimisations and obtain improved execution times. 
-Due to the large amount of financial data available and the need of quick response, 
-the algorithm presented
-was optimised using high performance computing.
-Our experiments were tested using synthetic data and foreign exchange rates.
-Results show that cointegration and high performance computing allow to obtain
-models with better performance accuracy and reduced execution times.}
+algorithms were developed in order to optimise parameters and increase
+performance.  In particular, an online algorithm based on VECM which optimises
+how model parameters are obtained and reduces execution times is proposed. This
+is achieved considering only a sliding window of the last historical data and
+using machine learning techniques to solve the model. Moreover, the long-run
+relationship between the time series is used in order to make optimisations and
+obtain improved execution times.  Due to the large amount of financial data
+available and the need of quick response, the algorithm presented was optimised
+using high performance computing. The experiments were tested using foreign
+exchange rates.  Results show that cointegration and high performance computing
+allow to obtain models with better performance accuracy and reduced execution
+times.}
 
@@
 \emph{\textbf{Keywords:} Computational finance  - Cointegration  - Time Series - 
-Online algorithms - Resampling  - Regression}
+Online algorithms  - Regression}

### 2017-09-23 src/5_1_Proposal.tex
chapter 5 src/5_2 Icpram 2015 ready
Conclusions first part

@@
-\chapter[Adaptive cointegration model for time series forecasting]{Fast and adaptive cointegration based model
-for forecasting high frequency financial time series}
+\chapter[Adaptive cointegration model for time series forecasting]{Fast and
+adaptive cointegration based model for forecasting high frequency financial time
+series}\label{chapter:proposal1}
 
@@
 This work is published in \cite{Arce2017}.
+
+\vspace{0.5cm}
 

### 2017-09-23 src/6_Conclusion.tex
ordenando propuesta 2
Conclusions first part

@@
-\chapter{Conclusion}
+\chapter{Conclusions and Future Work}
+
+Main results and contributions are presented in this chapter along with some
+suggested future work.
+
+\vspace{0.5cm}
 
@@
 
+This thesis is a multidisciplinary work that involves knowledge of Finance
+and Economics, Time Series, Machine Learning, Parallel Computing and Scientific
+Computing. We specifically addressed the joint dynamic behaviour of financial time
+series that said to be cointegrated. A broadly used approach, called VECM (Vector
+Error Correction Model), characterises this behaviour in terms of forces pulling
+towards equilibrium called cointegration. However, the use of VECM has been
+limited to low frequency time series processed in batch mode. In this thesis we
+explored the use of VECM with high frequency data and we found that the main
+limitation was computational. VECM parameter estimation uses two computationally
+expensive routines: the Johansen method, to obtain cointegration vectors, and
+the ordinary least squares method to solve the system. In this thesis, different
+ways to explore cointegration in high frequency time series were proposed,
+always considering the computational limitations of the VECM.
+
+The study of cointegration in high frequency data was done using two different
+approaches presented in Chapter \ref{chapter:proposal1} and
+\ref{chapter:proposal2}. First, we used a parallel version of VECM, called AVECM
+(Adaptive VECM), which includes a new method to choose VECM parameters based on
+the maximisation of the percentage of cointegration. AVECM was implemented using
+MPI to search on a grid of possible values. Results showed that AVECM improves
+performance measures by finding parameters of L and p maximising the percentage
+of cointegration and also execution times were reduced more than 9 times
+ensuring a response time before the processing of the next data point.
+Secondly, we proposed and online version of VECM (OVECM). OVECM optimises how
+model parameters are obtained using a sliding window of the most recent data.
+OVECM, unlike AVECM, updates the parameters at each step, instead of obtaining
+new ones. This proposal also takes advantage of the long-run relationship
+between the time series in order to obtain improved execution times. OVECM also
+introduces matrix optimisation in order to get the new model in an iterative
+way.  As a result, OVECM took much less than a minute at every
+step making it possible to use with higher frequency data.
+
 \section{Contributions of this thesis}
 
+Based on the observation that cointegration
+relationships are sensitive to the choose of the amount of data and VECM
+parameters,
+
 \section{Future Work}
+
+For future study, it would be interesting to explore the relationship between
+cointegration and performance for more assets, including not only forex rates
+but also stocks. It would also be interesting to include more explaining
+variables such as bid-ask spread and change in volume.
+
+The online approach for other econometrics models it will be also worth of
+study, many of the could be adapted and be used with higher frequency data. In
+this thesis, the online version of OLS was studied, but the online version of
+ridge regression could also be applied to obtain a solution with better
+generalisation capabilities.
+
+Finally, it will be also interesting to improve the out-of-sample forecast
+by considering more explicative variables, to increase window sizes or trying new
+conditions to obtain new cointegration vectors.

### 2017-09-23 src/1_Introduction.tex
cambio intro
Solo falta relacion con los objetivos en capitulo de conclusiones

@@
 approaches. More recently, machine learning models have been extensively used in
-forecasting. However, their main disadvantage is that getting model parameters is a computational challenge.
-The computational complexity of machine learning algorithms has become a
-limiting factor for problems that require processing large volumes of data and
-where response time is crucial.
+forecasting. However, their main disadvantage is that getting model parameters
+is a computational challenge.  The computational complexity of machine learning
+algorithms has become a limiting factor for problems that require processing
+large volumes of data and where response time is crucial.
 
@@
 %%
-In this thesis, an online formulation of the VECM called Online VECM
-(OVECM) is proposed. OVECM is based on the consideration of a sliding window of the most recent data.
-The algorithm introduces matrix optimisations in order to reduce the number of
-operations and also takes into account the fact that cointegration vector space
-doesn't experience large changes with small changes in the input data. Moreover, VECM parameters are obtained using machine learning methods. Our method is later tested using four
-currency rates from the foreign exchange market with different frequencies.  
-On the other hand, in order to improve VECM parameters, an adaptive VECM algorithm is presented called AVECM.
-AVECM allows VECM parameters to be found by maximising the number of cointegration relations for a given set of parameters on a grid search.
-This grid search is done in parallel.
+In this thesis, an online formulation of the VECM called Online VECM (OVECM) is
+proposed. OVECM is based on the consideration of a sliding window of the most
+recent data.  The algorithm introduces matrix optimisations in order to reduce
+the number of operations and also takes into account the fact that cointegration
+vector space doesn't experience large changes with small changes in the input
+data. Moreover, VECM parameters are obtained using machine learning methods. Our
+method is later tested using four currency rates from the foreign exchange
+market with different frequencies.  On the other hand, in order to improve VECM
+parameters, an adaptive VECM algorithm is presented called AVECM. AVECM allows
+VECM parameters to be found by maximising the number of cointegration relations
+for a given set of parameters on a grid search. This grid search is done in
+parallel.
 
-Models effectiveness were focused on the out-of-sample forecast rather than on the in-sample fitting.
-This criteria allows the OVECM and AVECM prediction capability to be expressed rather than just explaining data history.
-Our method performance is compared with its optimal offline algorithm.
+Models effectiveness were focused on the out-of-sample forecast rather than on
+the in-sample fitting. This criteria allows the OVECM and AVECM prediction
+capability to be expressed rather than just explaining data history. Our method
+performance is compared with the naive forecast of the random walk model and
+ARIMA which are the most widely used algorithms for modelling a multivariate
+time series.
 
 \section{Research Objectives}
-The main motivation for this research is the development of efficient methods for
-forecasting financial time series.
+The main motivation for this research is the development of efficient methods
+for forecasting financial time series.
 
@@
 
-Cointegration concept was introduced by Engle and Granger in 1987 \cite{engle1987} and implies that one or
-more linear combinations of non-stationary variables are stationary even though
-individually they are not.  Moreover Stock and Watson in 1988 \cite{stock+watson1988} observed that
+Cointegration concept was introduced by Engle and Granger in 1987
+\cite{engle1987} and implies that one or more linear combinations of
+non-stationary variables are stationary even though individually they are not.
+Moreover Stock and Watson in 1988 \cite{stock+watson1988} observed that
 cointegration reflects the common stochastic trends providing a useful way to
@@
 Vector error correction model (VECM) introduces this long-run relationship
-among a set of cointegrated variables as an error correction term. VECM is a
-special case of the vector autoregresive model (VAR) model. VAR model
+among a set of cointegrated variables as an error correction term. VAR model
 expresses future values as a linear combination of variables past values.
@@
 
-Chapter 1 contains relevant finance concepts required to understand financial
+Chapter \ref{chapter:HFT} contains relevant finance concepts required to understand financial
 time series models such as market hypothesis, frequency, order book generation,
 market microstructure.
-Chapter 2 discuss main time series models including classic ones such as ARMA,
-GARCH and more recent ones such as VECM, VAR and volatility models. %mejorar
-Chapter 3 gives a brief introduction to machine learning and its variant online
-learning. 
-Chapter 4 presents different algorithms propose to forecast financial time
-series using machine learning techniques.
-In Chapter 5 a deep study of foreign exchange market features is presented and
-experiments of cointegration are carried out in order to show its dynamic.
-Chapter 6 presents a discussion of found results and summarise the main
+Chapter \ref{chapter:FTT} discuss main characteristics and concepts of financial
+time series including classic models such as ARMA, ARIMA,
+GARCH and more recent ones such as VECM, VAR and volatility models.
+Chapter \ref{chapter:MLM} gives an introduction to machine learning and its
+variant online learning. 
+Chapter \ref{chapter:proposal1} presents the first proposal called AVECM which
+includes a new parallel method to choose VECM parameters based on the maximisation of the
+percentage of cointegration.
+In Chapter \ref{chapter:proposal2} a second approach is presented called OVECM,
+which is an online version of OVECM for high frequency data.
+Chapter \ref{chapter:conclusions} presents a discussion of found results and summarise the main
 conclusions of this thesis. It also presents some research directions for future
-studied.
+study.

### 2017-09-23 src/5_1_Proposal.tex
Solo falta relacion con los objetivos en capitulo de conclusiones

@@
 unit root: -2.62 (1\%), 1.94 (5\%) and 1.62 (10\%).
-\begin{table*}[ht]
-\label{tab:adf}
+Table \ref{tab:adf} shows that all currency rates cannot reject the unit root
+test in their level form considering different lags, but they rejected it with
+their first differences. This means that all of them are I(1) time series and we
+are allowed to use VECM and therefore our proposed AVECM.
+\begin{table}[ht]
 \centering
@@
 frequency.}
-\end{table*}
-Table~\ref{tab:adf} shows that all currency rates cannot reject the unit root
-test in their level form considering different lags, but they rejected it with
-their first differences. This means that all of them are I(1) time series and we
-are allowed to use VECM and therefore our proposed AVECM.
+\label{tab:adf}
+\end{table}
 

### 2017-10-08 CCTValproject/CCTVal-Project.tex
CCTVal project

\documentclass[12pt,reqno]{amsart}
\usepackage{amsfonts,amssymb,amsmath,amsthm,mathrsfs,
            amsopn,amsxtra,amstext,amsbsy}
\usepackage{epsfig}
\usepackage{array}
\usepackage{url}
\usepackage{tikz}
%% \usepackage[spanish]{babel}
\usepackage[utf8x]{inputenc}
\usepackage{fancyhdr}
\usepackage{graphicx}
\usepackage{hyperref,wasysym}
\usepackage{xcolor}
\usepackage{lipsum}
\usepackage{enumerate}
\usepackage{enumitem}% http://ctan.org/pkg/enumitem
\usepackage{verbatim}
%% \usepackage[strings]{underscore}

\voffset=-10mm
\oddsidemargin=0pt
\evensidemargin=0pt
\topmargin=-10mm
\headsep=18pt
\headheight=18pt
\footskip=30pt
\topskip=0mm
\textwidth 167mm
\textheight=230mm
\parindent=0em
\marginparsep=0mm
\marginparwidth=0mm
\vfuzz2pt % Don't report over-full v-boxes if over-edge is small
\hfuzz6pt % Don't report over-full h-boxes if over-edge is small

% \newtheorem{conjecture}[theorem]{Conjecture}
\newtheorem{definicion}{Definición}
\newtheorem{theorem}{Theorem}%[section]
% \newtheorem{lemma}[theorem]{Lemma}
% \newtheorem{corollary}[theorem]{Corollary}
% \newtheorem{conjecture}[theorem]{Conjecture}
% \theoremstyle{definition}
% \newtheorem{definition}[theorem]{Definition}
% \newtheorem{example}[theorem]{Example}
% \newtheorem{xca}[theorem]{Exercise}
% \theoremstyle{remark}
% \newtheorem{remark}[theorem]{Remark}
% \numberwithin{equation}{section}

% \def\phi{\varphi}
% \def\epsilon{\varepsilon}
\def\RE{\mbox{\rm Re}\,}
\def\IM{\mbox{\rm Im}\,}
% \def\AZ{|z|}
\newcommand{\rectangle}{\fboxsep0pt\fbox{\rule{0.3em}{0pt}\rule{0pt}{1.5ex}}}

%-------------------------mathcal

\def\A{{\mathcal A }}
\def\CC{{\mathcal C }}
\def\CH{{\mathcal H }}
\def\CK{{\mathcal K }}
\def\CM{{\mathcal M }}
\def\CP{{\mathcal P }}
\def\CS{{\mathcal S }}
\def\CR{{\mathcal R}}
\def\CRu{{\mathcal R}^{\rm u}}
\def\CT{{\mathcal T}}
\def\CTT{\widetilde{{\mathcal T}}}
\def\CW{{\mathcal W}}
\def\Li{{\mbox{\rm Li}}}
\def\WO{{\mathcal W}_{\Omega }}

%-------------------------------Bbb
\def\K{{\mathbb K }}
\def\C{{\mathbb C }}
\def\N{{\mathbb N }}
\def\NO{{\mathbb{N}_0 }}
\def\Z{{\mathbb Z }}
\def\R{{\mathbb R }}
\def\U{{\mathbb U }}
\def\D{{\mathbb D }}
%----------------------------Zusammengesetztes

\def\DB{\overline{\D}}
\def\Db#1{\overline{\D_{#1}}}
\def\DBR{\overline{\D_\rho}}
\def\DV{de la Vall\'ee Poussin }
\def\d#1{\D_{#1}}
\def\H{{\mathcal H}(\D)}
\def\Hb{{\mathcal H}(\DB)}
\def\HO{{\mathcal H}_0(\D )}
\def\Hoo#1{{\mathcal H}(\overline{\D_{#1}})}
\def\HB#1{{\mathcal H}(\D_{#1})}
\def\Ho#1{{\mathcal H}({#1})}
\def\Hyp{{}_2F_1}

\def\achtung#1{\alert{#1}}
\def\blue#1{{\color{blue} #1}}
\def\green#1{{\color{green} #1}}
\def\red#1{{\color{red} #1}}

\def\pick{{\mathscr P}}
\def\logpick{{\text{log}\,{\mathscr P}}}
\def\stammpick{\pick_{\!\!\!\raisebox{-2pt}{\scalebox{0.5}{$\int$}}}}
\def\stammpickpick{\stammpick\cap\pick}
\def\stammpicklogpick{\stammpick\cap\logpick}
\def\expstammpicklogpick{e^{\stammpicklogpick}}

\title[CCTVal Research project]{CCTVal Research project}

\author{Paola Arce}
\author{Werner Kristjanpoller}
\author{Luis Salinas}

\address{CCTVal and Departamento de Inform\'atica, UTFSM, Valpara\'\i so;
         %% \newline
         %% Mathematisches Institut, Universit\"at W\"urzburg;
         \newline
         \today} % October 06, 2016}

%\email{luis.salinas@usm.cl}

\begin{document}

\maketitle%

\def\myitem#1{ \item[\blue{\bf #1}] }

\section{Introduction}
Cointegration is a long-run property of some non-stationary time series where a
linear combination of those time series is stationary.  This behaviour has been
studied in finance because cointegration restrictions often improve forecasting.
The Vector Error Correction Model (VECM)  \cite{engle1987}, is a well-known econometric technique
that characterises short-run variations of a set of cointegrated time series
incorporating long-run relationships as an error correction term. However, the use of VECM has been limited to low frequency time series processed in batch mode mainly due to computational limitations.

The first computational limitation to address is that VECM parameters are obtained using the ordinary least squares (OLS)
method.  Even though OLS is extensively used, it has over-fitting
issues and it gets computationally expensive when more data is added. In this project we explore the use of Ridge regression which is commonly used
instead of OLS since they include a regularisation parameter which could improve
prediction error.

The second limitation is the calculation of the cointegration vectors. Our proposal is to avoid their calculation at every step since they represent a long-run relationship between the time series.

For this research project proposal, we will design and evaluate an online version of VECM which optimises how model parameters are obtained so the model can be used with high frequency data. The proposal is an improvement of the online VECM presented in \cite{icpram15} by adding the following modifications:
\begin{itemize}
\item The use of RR instead of OLS to get the model parameters. This will imply the finding of the best regularisation parameter at every time step.
\item The evaluation of the Aggregating Algorithm for Regression (AAR) which is a variation of the online version of Ridge regression.
\item Addition of rules to define when the cointegration vectors must be recalculated.
\item Analysis of different time series frequencies to determine where the model works best.
\item The selection of other relevant data to the model such as bid, ask and volume.
\end{itemize}

The findings of this project could be easily used as an integration tool
to detect arbitrage opportunities or risk control in financial time series. Also, the methodology could be extended to other econometric methods which could be benefited of machine learning techniques.

\section{Background}

VECM introduces the long-run
relationship among a set of cointegrated time series as an error correction
term. These time series must be integrated of order 1, denoted I(1), i.e. they
become stationary at their first differences. In finance, I(1) time series are
very common and to introduce cointegration restrictions in models often improves
forecasting, see \cite{duy1998}. Therefore, VECM has been widely adopted in
financial applications \cite{mukherjee1995}, \cite{seong2013},
\cite{maysami2000} and \cite{arestis2001}. VECM has also been used in pair
trading, see \cite{herlemont2003}, or models with more than two variables, see
for example \cite{mukherjee1995} and \cite{engle2004}.

Cointegration relationships can be found in low and high frequency data of two or
more assets. While cointegration in low frequency data is motivated by a
long-run equilibrium relationship between economic forces, cointegration in high
frequency data has its foundation in statistical arbitrage theory which is very helpful
to detect mean-reverting trades. Information about cointegrated assets in high
frequency data could be used as an input for high frequency trading strategies,
using it as a signal to capture small profits in short term trades, see
\cite{miao2014}. \cite{zhou2001} addressed the benefits of using higher
frequency data to analize cointegration. \cite{rittler2012} also found that
cointegration may differ with different data frequencies and could appear
with increased data frequency.

The use of VECM with high frequency data is mainly limited by computationally
expensive routines. Firstly, VECM parameters are obtained using the ordinary
least squares (OLS) it has over-fitting
issues and is computationally expensive when
the number of lagged values and data increases. This problem is commonly adressed using
Ridge Regression (RR) \cite{hoerl1970} which introduces a regularisation
parameter that leads to an unbiased estimation with better generalisation
capability. Secondly, obtaining
cointegration vectors is also an expensive routine because the Johansen method 
is of order $O(n^3)$ (\cite{johansen1995}).
\cite{chen2003}, \cite{Arce2017} addressed the advantage of distributed processing over
conventional rolling window processing. Therefore, our aim is to design
an online method based on cointegration using RR instead of OLS and that works on a distributed environment so it can be used with high frequency stream data.

\section{Research Design and Methods}
We intend to extend the approach presented in \cite{icpram15}, where the proposed method called OVECM only considered a time varying window of historical data using OLS to
get model parameters. This proposal is about using RR or its variant  The  Aggregating Algorithm for Regression (AAR)  \cite{vovk2001} as an alternative of OLS to estimate model parameters.

Since cointegration vectors represent a long-run relationship between the time
series, they vary little in time. This proposal also includes to obtain new
cointegration vectors only when this long-term relationship changes.
We will estimate the performance of our method using different frequencies of financial time series, so we can measure the effect of cointegration in high frequency data.

\section{Hypothesis}

We hypothesise that a regularisation solution could lead to better forecasting performance than OVECM. RR is usually introduced to get better generalisation capabilities but the choice of the regularisation parameter is a challenge to accomplish this goal.

We also hypothesise that, using high frequency data, cointegration vectors will remain the same for some period of time, since they represent long-run relationship.

\bibliographystyle{amscls/amsalpha}
\bibliography{reference}

\end{document}

### 2017-10-14 CCTValproject/CCTVal-Project.tex
Last approved version

@@
 
-\title[CCTVal Research project]{CCTVal Research project}
+\title[CCTVal Research project proposal]{CCTVal Research project proposal}
 
@@
 The findings of this project could be easily used as an integration tool
-to detect arbitrage opportunities or risk control in financial time series.
-Also, the methodology could be extended to other econometric methods which could be benefited of machine learning techniques.
+to detect arbitrage opportunities or risk control in financial time series. Also, the methodology could be extended to other econometric methods which could be enhanced by machine learning techniques.
 
@@
 
-VECM introduces the long-run
-relationship among a set of cointegrated time series as an error correction
-term. These time series must be integrated of order 1, denoted I(1), i.e. they
+In order to use the VECM, the time series must be integrated of order 1, denoted I(1), i.e. they
 become stationary at their first differences. In finance, I(1) time series are
-very common and to introduce cointegration restrictions in models often improves
-forecasting, see \cite{duy1998}. Therefore, VECM has been widely adopted in
-financial applications \cite{mukherjee1995}, \cite{seong2013},
-\cite{maysami2000} and \cite{arestis2001}. VECM has also been used in pair
-trading, see \cite{herlemont2003}, or models with more than two variables, see
+very common. VECM introduces the cointegration relationship as an error correction
+term which often improves forecasting, see \cite{duy1998}. Therefore, VECM has been widely adopted in
+financial applications \cite{mukherjee1995,seong2013,maysami2000,arestis2001}. VECM has also been used in pair trading, see \cite{herlemont2003}, or models with more than two variables, see
 for example \cite{mukherjee1995} and \cite{engle2004}.
@@
 \cite{miao2014}. \cite{zhou2001} addressed the benefits of using higher
-frequency data to analize cointegration. \cite{rittler2012} also found that
+frequency data to analyse cointegration. \cite{rittler2012} also found that
 cointegration may differ with different data frequencies and could appear
@@
 The use of VECM with high frequency data is mainly limited by computationally
-expensive routines. Firstly, VECM parameters are obtained using the ordinary
-least squares (OLS) it has over-fitting
+expensive routines:
+\begin{itemize}
+\item VECM parameters are obtained using the ordinary
+least squares (OLS) which has over-fitting
 issues and is computationally expensive when
-the number of lagged values and data increases. This problem is commonly adressed using
-Ridge Regression (RR) \cite{hoerl1970} which introduces a regularisation
+the number of lagged values and data increases. This problem is commonly addressed using
+Ridge Regression (RR) \cite{hoerl1970}. RR introduces a regularisation
 parameter that leads to an unbiased estimation with better generalisation
-capability. Secondly, obtaining
-cointegration vectors is also an expensive routine because the Johansen method 
-is of order $O(n^3)$ (\cite{johansen1995}).
-\cite{chen2003}, \cite{Arce2017} addressed the advantage of distributed processing over
-conventional rolling window processing. Therefore, our aim is to design
-an online method based on cointegration using RR instead of OLS and that works on a distributed environment so it can be used with high frequency stream data.
+capability. 
+\item The Johansen method is used to obtain 
+cointegration vectors and it is also an expensive routine  of order $O(n^3)$  \cite{johansen1995}.
+A distributed processing approach has been used recently successfully reducing execution times, see \cite{chen2003}, \cite{Arce2017}.
+Specially in \cite{Arce2017}we showed the advantage of distributed processing over conventional rolling window processing.
+\end{itemize}
+Therefore, our aim is to design an online method based on cointegration using RR instead of OLS and that works on a distributed environment so it can be used with high frequency stream data.
 
 \section{Research Design and Methods}
-We intend to extend the approach presented in \cite{icpram15}, where the proposed method called OVECM only considered a time varying window of historical data using OLS to
-get model parameters. This proposal is about using RR or its variant  The  Aggregating Algorithm for Regression (AAR)  \cite{vovk2001} as an alternative of OLS to estimate model parameters.
 
-Since cointegration vectors represent a long-run relationship between the time
-series, they vary little in time. This proposal also includes to obtain new
-cointegration vectors only when this long-term relationship changes.
+In this proposal, we will extend the approach presented in \cite{icpram15}.
+In this article we proposed method called OVECM that only considered a time varying window of historical data using OLS to get model parameters.
+This proposal will explore the use of RR or its variant The  Aggregating Algorithm for Regression (AAR)  \cite{vovk2001} as an alternative of OLS to estimate model parameters.
+
+Additionally, since cointegration vectors represent a long-run relationship between the time
+series, they vary little in time. Therefore, we plan to obtain new
+cointegration vectors only when this long-term relationship changes, this can be tracked using a threshold of forecast error for example.
+
 We will estimate the performance of our method using different frequencies of financial time series, so we can measure the effect of cointegration in high frequency data.
@@
 
-We also hypothesise that, using high frequency data, cointegration vectors will remain the same for some period of time, since they represent long-run relationship.
+We also hypothesise that, using high frequency data, cointegration vectors will remain the same for some period of time, since they represent long-run relationships.
 

### 2017-10-14 src/1_Introduction.tex
Last approved version

@@
 
-"An individual economic variable, views as a time series, can wander extensively
+"An individual economic variable, viewed as a time series, can wander extensively
 and yet some pairs of series may be expected to move so that they do not drift
@@
 \item $\Large \mathcal{O}_2$: \emph{Development of a set of known features of the
-studied time series and applied them to improve forecasting.}
+studied time series and the application to improve forecasting.}
 \item $\Large \mathcal{O}_3$: \emph{Development of parallel and efficient
@@
 \item $\Large \mathcal{O}_5$: \emph{Design and implement representative set of
-experiments in order to show when and why proposal performs better.}
+experiments in order to show when and why the proposal performs better.}
 \end{itemize}

### 2017-10-14 src/5_1_Proposal.tex
Last approved version

@@
 9 times and therefore a response time was obtain before 10 seconds. Since we
-used 10-second frequencys we can say that our proposal is suitable for use in an
+used 10-second frequencies we can say that our proposal is suitable for use in an
 online context for real applications because response times were less than this

### 2017-10-14 src/6_Conclusion.tex
Solo falta relacion con los objetivos en capitulo de conclusiones
Last approved version

--- 2015-04-01 src/6_Conclusion.tex
@@
-\chapter{Conclusion}
+\chapter{Conclusions and Future Work}
+\label{chapter:conclusions}
+Main results and contributions are presented in this chapter along with some
+suggested future work.
+
+\vspace{0.5cm}
 
@@
 
+This thesis is a multidisciplinary work that involves knowledge of Finance
+and Economics, Time Series, Machine Learning, Parallel Computing and Scientific
+Computing. We specifically addressed the joint dynamic behaviour of financial time
+series that are said to be cointegrated. A broadly used approach, called VECM (Vector
+Error Correction Model), characterises this behaviour in terms of forces pulling
+towards equilibrium called cointegration. However, the use of VECM has been
+limited to low frequency time series processed in batch mode. In this thesis we
+explored the use of VECM with high frequency data and we found that the main
+limitation was computational.
+The focus of our work was developing a model to improve the performance forecasting of financial time series maintaining good execution times in order to use it with high frequency data.
+VECM parameter estimation uses two computationally
+expensive routines: the Johansen method, to obtain cointegration vectors, and
+the ordinary least squares method to solve the system. In this thesis, different
+ways to explore cointegration in high frequency time series were proposed,
+always considering the computational limitations of the VECM.
+
+The study of cointegration in high frequency data was done using two different
+approaches presented in Chapters \ref{chapter:proposal1} and
+\ref{chapter:proposal2}, called AVECM and OVECM:
+\begin{description}
+\item[AVECM] AVECM is an adaptive version of VECM which includes a new method to choose VECM parameters based on
+the maximisation of the percentage of cointegration. AVECM was implemented using
+MPI to search on a grid of possible values. We evaluated our results in terms of performance and execution times:
+\begin{description}
+\item[Performance] Results showed that AVECM improves
+performance measures by finding parameters of L and p maximising the percentage
+of cointegration.
+AVECM performance was compared against ARIMA and the random walk model.
+We showed that the out-of-sample performance for AVECM is superior to ARIMA and the naive random walk model in terms of the MSE and $U$-statistic.
+This result is very important since it is still difficult to outperform ARIMA and the random walk model for standard econometric forecasting models  despite their simplicity.
+\item[Execution times] execution times were reduced more than 9 times
+ensuring a response time before the processing of the next data point. Despite the fact that high frequency Forex data can be spurious, the model
+performance can be less reliable (and more spurious) relative to the lower
+frequencies (such as 1 minute or 5 minute intervals) adopted by some other
+studies. However, the deficiency is offset by gain in accuracy from parallel
+processing which is capable of searching or examining a much larger state space
+given the same computational time.
+\end{description}
+\item[OVECM] We proposed an online version of VECM (OVECM). OVECM optimises how
+model parameters are obtained using a sliding window of the most recent data.
+OVECM, unlike AVECM, updates the parameters at each step, instead of obtaining
+new ones. This proposal takes advantage of the long-run relationship
+between the time series in order to obtain improved execution times. OVECM also
+introduces matrix optimisation in order to obtain the new model in an iterative
+way.  
+\begin{description}
+\item[Performance] 
+OVECM was compared with the traditional VECM and ARIMA.
+Despite the fact that traditional VECM slightly outperformed our proposal, the OVECM execution time is lower.
+This result can be explained because OVECM avoids the calculation of cointegration vectors at each time step which improves execution times but can affect forecast performance.
+\item[Execution times] OVECM was compared with the traditional VECM and ARIMA
+with the same sliding window sizes. As a result, OVECM outperformed both in terms of
+execution time. This reduction of execution time is mainly because OVECM avoids the cointegration vector calculation using the Johansen method. Aditionally, OVECM took much less than a minute at every
+step making it possible to use with higher frequency data.  
+\end{description}
+\end{description}
+Finally, cointegration information can now easily be used as an integration tool
+to detect arbitrage opportunities or risk control in financial time series.
+
 \section{Contributions of this thesis}
 
+This research helped to integrate machine learning techniques and parallel computing with econometric models such as VECM so they can be used with high frequency data.
+We also observed that cointegration relationships are sensitive to the choice of the amount of data and VECM parameters and this can be used to improve forecasting performance.
+We proposed two variations of VECM, OVECM and AVECM which could be extended to other econometric models.
+
+Regarding the initial objectives defined at the beginning of this thesis we can say:
+
+\begin{itemize}
+\item $\Large \mathcal{O}_1$: \emph{A review of the literature on time series
+analysis models including machine learning techniques.}\\
+The literature review was focused in finance, time series concepts and models and machine learning including theory and applications.
+\item $\Large \mathcal{O}_2$: \emph{Development of a set of known features of the
+studied time series and the application to improve forecasting.} \\
+We included specific knowledge about forex rates, which were those used in our experiments, such as trade best trading times, stylised facts, percentage of cointegration, time series frequency, among others.
+\item $\Large \mathcal{O}_3$: \emph{Development of parallel and efficient
+algorithms to ensure quick response times .} \\
+We showed that high performance computing will allow the use of computationally expensive methods to forecast high frequency data ensuring quick response times.
+\item $\Large \mathcal{O}_4$: \emph{Deep mathematical analysis of the proposal
+and financial concepts involved.} \\
+The two proposals were presented with a strong mathematical foundation, we added definitions and proofs of all the concepts involved.
+\item $\Large \mathcal{O}_5$: \emph{Design and implement representative set of
+experiments in order to show when and why the proposal performs better.}
+All the experiments were designed to give a general view of the method performance. We clearly showed the use and limitations of our proposals.
+\end{itemize}
+
 \section{Future Work}
+
+For future study, it would be interesting to explore the relationship between
+cointegration and performance for more assets, including not only forex rates
+but also stocks. It would also be interesting to include more explaining
+variables such as bid-ask spread and change in volume.
+
+The online approach for other econometrics models it will be also worthy of 
+study. Many of them could be adapted and used with higher frequency data. In
+this thesis, the online version of OLS was studied, but the online version of
+ridge regression could also be applied to obtain a solution with better
+generalisation capabilities.
+
+Finally, it will be also interesting to improve the out-of-sample forecast
+by considering more explicative variables, to increase window sizes or try different
+conditions to obtain new cointegration vectors.

### 2017-10-15 src/2_Finance.tex
Solo falta relacion con los objetivos en capitulo de conclusiones
fixing table captions

@@
 \chapter{High Frequency Trading}
+\label{chapter:HFT}
 
@@
   \includegraphics[width=\textwidth]{img/HFTtradestime}
-  \caption{Holding time of an opened position of a high frequency trade.Source
+  \caption[Holding time of an opened position of a high frequency trade]{Holding time of an opened position of a high frequency trade.Source
   \cite{aldridge2009}.}
@@
   \includegraphics[width=\textwidth]{img/HFTmarket}
-  \caption{Left figure shows HFT market share in the US and Europe. Right figure
+  \caption[HFT market share in the US and Europe]{Left figure shows HFT market share in the US and Europe. Right figure
   shows revenue in the US. Source: TABB Group.}
@@
   \includegraphics[width=0.7\textwidth]{img/capitalmarkets}
-  \caption{Old structure of capital markets. Source \cite{aldridge2009}.}
+  \caption[Old structure of capital markets]{Old structure of capital markets. Source \cite{aldridge2009}.}
   \label{fig:capitalmarket}
@@
   \includegraphics[width=0.8\textwidth]{img/capitalmarketsnow}
-  \caption{Actual structure of capital markets. Source \cite{aldridge2009}.}
+  \caption[Actual structure of capital markets]{Actual structure of capital markets. Source \cite{aldridge2009}.}
   \label{fig:capitalmarketnow}
@@
   \includegraphics[width=0.8\textwidth]{img/forex-trading-hours.png}
-  \caption{Forex market trading hours (GMT). Source \cite{dailyprice}.}
+  \caption[Forex market trading hours (GMT)]{Forex market trading hours (GMT). Source \cite{dailyprice}.}
   \label{fig:Forextimes}
@@
   \includegraphics[width=0.5\textwidth]{img/orderbook}
-  \caption{Order book. Buyers and sellers are ordered according to the bid or
+  \caption[Order book]{Order book. Buyers and sellers are ordered according to the bid or
 ask price and the market determines the mid-price and transacted volume. Source

### 2017-10-15 src/3_TimeSeries.tex
changing minor errors
Solo falta relacion con los objetivos en capitulo de conclusiones
fixing table captions

@@
 \chapter{Financial Time Series}
+\label{chapter:FTT}
 
@@
   \includegraphics[width=0.8\textwidth]{img/spurious1}
-  \caption{Two random walks time series $x_{1,t}$ and $x_{2,t}$}
+  \caption[Random walks time series]{Two random walks time series $x_{1,t}$ and $x_{2,t}$}
   \label{fig:spurious1}
@@
   \includegraphics[width=\textwidth]{img/coint-example}
-  \caption{Cointegration example: two I(1) processes (blue and green) and one I(0) process (red).}
+  \caption[Cointegration example]{Cointegration example: two I(1) processes (blue and green) and one I(0) process (red).}
   \label{fig:cointex}
@@
 the VECM($p$) model in block-matrix form as:
-\begin{equation}\label{eq:simpleveceq} %vareq}
+\begin{equation}\label{eq:simpleveceq}
 \mathbf{B} = 
@@
 
-When $\mathbf{A}$ is singular, solution to equation~(\ref{eq:vareq}) is given
+When $\mathbf{A}$ is singular, solution to equation~(\ref{eq:simpleveceq}) is given
 by the ordinary least squares (OLS) method. OLS consists of minimizing the sum

### 2017-10-15 src/4_OnlineLearning.tex
Solo falta relacion con los objetivos en capitulo de conclusiones
fixing table captions

@@
 \chapter{Machine Learning Models}
+\label{chapter:MLM}
 Machine learning is a scientific discipline focused on the development of
@@
   \includegraphics[width=0.8\textwidth]{img/VCdimension}
-  \caption{VC dimension example from \cite{vapnik2013nature}. The example shows 2-dimensional vectors $z_k=(x_k,y_k)$,  $k=1,2,3,4$. The left figure
+  \caption[VC dimension example]{VC dimension example from \cite{vapnik2013nature}. The example shows 2-dimensional vectors $z_k=(x_k,y_k)$,  $k=1,2,3,4$. The left figure
   shows that the 3 dichotomies can be shattered by a line in a plane. However, if we
@@
   \includegraphics[width=0.8\textwidth]{img/3tradeoff}
-  \caption{Tradeoff in empirical learning from \cite{dietterich2003}. The
+  \caption[Tradeoff in empirical learning ]{Tradeoff in empirical learning from \cite{dietterich2003}. The
 generalization accuracy is not affected by the amount of examples for low
@@
   \includegraphics[width=0.8\textwidth]{img/model_complexity}
-  \caption{Training and test error. Image source is
+  \caption[Training and test error]{Training and test error. Image source is
 \cite{friedman2001elements}.The blue curve represents the training error which
@@
 \includegraphics[width=\linewidth]{img/biasvariance}
-\caption{Bias-variance tradeoff depending on lambda.
+\caption[Bias-variance tradeoff depending on lambda]{Bias-variance tradeoff depending on lambda.
 Dotted line corresponds to OLS prediction error (invariant with lambda).
@@
   \includegraphics[width=0.8\textwidth]{img/ffn}
-  \caption{Feed-forward neural network (FFN). FNN consists of an input layer, an output layer and at least one hidden layer between the input and output layer. }
+  \caption[Feed-forward neural network (FFN)]{Feed-forward neural network (FFN). FNN consists of an input layer, an output layer and at least one hidden layer between the input and output layer. }
   \label{fig:ffn}

### 2017-10-15 src/5_1_Proposal.tex
fixing table captions

@@
    \includegraphics[width=0.9\textwidth]{img/51_Fig1}
-  \caption{Distribution of the number of cointegration vectors using $p=1$ lags.
+  \caption[Distribution of the number of cointegration vectors using $p=1$ lags]{Distribution of the number of cointegration vectors using $p=1$ lags.
   Four possible values for windows size $L$ are shown (2, 6, 10 and 14) hours (1
@@
    \includegraphics[width=0.9\textwidth]{img/51_Fig2}
-  \caption{MSE versus the percentage of cointegration considering 1000
+  \caption[MSE versus the percentage of cointegration considering 1000
+  iterations]{MSE versus the percentage of cointegration considering 1000
   iterations. Optimum windows size $L$ found was 1060. Below MSE versus $L$ shows a
@@
   \includegraphics[width=0.9\textwidth]{img/51_Fig3}
-  \caption{Computing time of sequential and parallel algorithm is shown in the
+  \caption[Computing time and Speed-up]{Computing time of sequential and parallel algorithm is shown in the
   upper figure. Speed-up is shown below.}

### 2017-10-15 src/5_2_Proposal.tex
Conclusions first part
fixing table captions

@@
 \chapter{An Online Vector Error Correction Model for Exchange Rates Forecasting}
+\label{chapter:proposal2}
 
@@
 model parameters are obtained using a sliding window of the most recent data.
-OVECM, unlile AVECM, updates the parameters at each step, instead of obtaining
+OVECM, unlike AVECM, updates the parameters at each step, instead of obtaining
 new ones. This proposal also takes advantage of the long-run relationship
@@
 
-\section{Results} \label{sec:52results}
+\section{Experimental results} \label{sec:52results}
 
@@
    \includegraphics[width = 10.5cm]{img/mapes} 
-  \caption{In-sample MAPEs example for 50 minutes. The average of them is
+  \caption[In-sample MAPEs example for 50 minutes]{In-sample MAPEs example for 50 minutes. The average of them is
   considered to obtain new cointegration vectors.}
@@
    \includegraphics[width = 10cm]{img/accuracy} 
-  \caption{OVECM forecasting accuracy example for 50 minutes using $L = 1000$ and $p =3$}
+  \caption[OVECM forecasting accuracy example]{OVECM forecasting accuracy example for 50 minutes using $L = 1000$ and $p =3$}
   \label{fig:accuracy}

### 2017-10-15 src/firmas.tex
fixing table captions

@@
 {\textsc Prof. Dr. Luis Salinas} & \begin{picture}(150,5) \line(1,0){150} \end{picture}\\
-{Universidad Técnica Federico Santa María} & {Advisor}\\
+{Universidad T\'ecnica Federico Santa Mar\'ia} & {Advisor}\\
 Chile & \\
@@
 & \\
-{\textsc Prof. Dr. } &  \begin{picture}(150,5) \line(1,0){150} \end{picture}\\
-{Universidad Técnica Federico Santa María} & {Committee Member} \\
+{\textsc Dr. Claudio Torres} &  \begin{picture}(150,5) \line(1,0){150} \end{picture}\\
+{Universidad T\'ecnica Federico Santa Mar\'ia} & {Co-advisor} \\
+Chile & \\
+
+& \\
+& \\
+{\textsc Dr. Werner Kristjanpoller} &  \begin{picture}(150,5) \line(1,0){150} \end{picture}\\
+{Universidad T\'ecnica Federico Santa Mar\'ia} & {Co-advisor} \\
 Chile & \\
@@
 & \\
-{Prof. Dr. } & \begin{picture}(150,5) \line(1,0){150} \end{picture}\\
-{University} & {External Peer Reviewer}\\
+{Prof. Dr. Christian de Peretti} & \begin{picture}(150,5) \line(1,0){150} \end{picture}\\
+{Ecole Centrale de Lyon} & {External Peer Reviewer}\\
+{France} & \\
+& \\
+& \\
+{Prof. Dr. Wilfredo Palma} & \begin{picture}(150,5) \line(1,0){150} \end{picture}\\
+{Pontificia Universidad Cat\'olica de Chile} & {External Peer Reviewer}\\
 {Chile} & \\
+& \\
+& \\
+{\textsc Prof. Dr. Hern\'an Astudillo} &  \begin{picture}(150,5) \line(1,0){150} \end{picture}\\
+{Universidad T\'ecnica Federico Santa Mar\'ia} & { PhD program Director} \\
+Chile & \\
+%& \\
+%& \\
+%{\textsc Prof. Dr. Carlos Castro} &  \begin{picture}(150,5) \line(1,0){150} \end{picture}\\
+%{Universidad T\'ecnica Federico Santa Mar\'ia} & {Committee Member} \\
+%Chile & \\
 \end{tabular}

### 2017-10-15 src/1_Introduction.tex
updating Financial info

@@
 amount of data available and the need of performing accurate forecasting in
-short periods of time has motivated researchers to create efficient and fast
+short periods of time has motivated researchers to create efficient and fast forecast
 algorithms. This study involves interdisciplinary knowledge such as: finance,
@@
 
-\textit{An online learning algorithms based on cointegration and high
+\textit{An online learning algorithm based on cointegration and high
 performance computing will allow faster forecasting

### 2017-10-15 src/2_Finance.tex
updating Financial info

@@
 transaction costs. The key is the amount of transactions that HFT algorithms are
-capable to execute.  In 2012 more than 50\% of US market trades were HFT.  HFT
+capable to execute.  In 2016 nearly 50\% of US market trades were HFT.  HFT
 has motivated computer-driven strategies capable of processing large amount of
@@
   \includegraphics[width=\textwidth]{img/HFTtradestime}
-  \caption[Holding time of an opened position of a high frequency trade]{Holding time of an opened position of a high frequency trade.Source
+  \caption[Holding time of an opened position of a high frequency trade]{Holding time of an opened position of a high frequency trade. Source
   \cite{aldridge2009}.}
@@
 \end{figure}
-In 2014 HFT represented nearly 50\% of equity trades in the US and more than
-20\% in Europe showing a consistent fall since 2009 which was its best year. US
-market revenues have fallen dramatically.  Figure~\ref{fig:HFTmarket} on the
-left shows HFT trades percentage of US and Europe equity shares. The right
-figure shows how revenues in US stocks have fallen the last years.
+In 2016 HFT represented nearly 50\% of equity trades in the US showing a consistent fall since 2009, which was its best year, until 2013 where this rate remains the same until 2016.
+Figure~\ref{fig:HFTmarket1} shows HFT trades percentage of US equity shares.
 
@@
   \centering
-  \includegraphics[width=\textwidth]{img/HFTmarket}
-  \caption[HFT market share in the US and Europe]{Left figure shows HFT market share in the US and Europe. Right figure
-  shows revenue in the US. Source: TABB Group.}
-  \label{fig:HFTmarket}
+  \includegraphics[width=\textwidth]{img/HFTmarket1.png}
+  \caption[HFT market share in the US and Europe]{HFT market share in the US. Source: TABB Group.}
+  \label{fig:HFTmarket1}
 \end{figure}
+
+\begin{figure}[!h]
+  %\vspace{-0.8cm}
+  \centering
+  \includegraphics[width=\textwidth]{img/HFTmarket2.png}
+  \caption[Revenue in the US]{Revenue in the US. Source: TABB Group.}
+  \label{fig:HFTmarket2}
+\end{figure}
+US market revenues have fallen dramatically.
+Figure~\ref{fig:HFTmarket2} shows how revenues in US stocks have fallen the last years.
 One of the reasons of this fall is that exchange markets have adapted, having

### 2017-10-16 src/2_Finance.tex
new bid ask spread plot

@@
   \centering
-  \includegraphics[width=\textwidth]{img/HFTmarket1.png}
+  \includegraphics[width=0.9\textwidth]{img/HFTmarket1.png}
   \caption[HFT market share in the US and Europe]{HFT market share in the US. Source: TABB Group.}
@@
   \centering
-  \includegraphics[width=\textwidth]{img/HFTmarket2.png}
+  \includegraphics[width=0.9\textwidth]{img/HFTmarket2.png}
   \caption[Revenue in the US]{Revenue in the US. Source: TABB Group.}
@@
 On the other hand, HFT has been criticised on qualitative issues concerning
-fairness and systemic risk. However, empirical research shows that HFT has lead to
-beneficial impacts such as reducing spreads (difference between buyers and
-sellers prices), increased liquidity, allowing more efficient price formation,
-reduced transaction costs and lower market volatility. HFT makes the stock market more efficient and helps small investors who trade at random times over the day.
+fairness and systemic risk. However, HFT has lead to
+beneficial impacts such as reducing spreads (difference between buyers (bid) and
+sellers (ask) prices), increased liquidity, allowing more efficient price formation,
+reduced transaction costs and lower market volatility.
+
+\begin{figure}[!ht]
+  %\vspace{-0.8cm}
+  \centering
+  \includegraphics[width=0.7\textwidth]{img/bid-ask-spreads.jpg}
+  \caption[Bid-ask spreads]{Bid-ask spreads for large and small caps stock US. Source: Credit Suisse.}
+  \label{fig:Spreads}
+\end{figure}
+In figure \ref{fig:Spreads}  shows the Bid-ask spread for large and small caps. A large cap refers to an asset with market capitalisation value of more than \$10 billion dollars.
+
+In theory, Bid-ask spreads for large caps and small caps move on the same direction, meaning they both are affected by volatility.
+This scenario is only true until 2009, where the large caps (in red) gets tight spreads and small caps have wider  spreads, suggesting a concentrating of trading in the most liquid, biggest stocks.
 

### 2017-10-21 src/2_Finance.tex
minor changes

@@
   \includegraphics[width=\textwidth]{img/HFTtradestime}
-  \caption[Holding time of an opened position of a high frequency trade]{Holding time of an opened position of a high frequency trade. Source
+  \caption[Holding time of an opened position of a high frequency trade]{Holding
+  time of an opened position of a high frequency trade. Source
   \cite{aldridge2009}.}
@@
 \end{figure}
-In 2016 HFT represented nearly 50\% of equity trades in the US showing a consistent fall since 2009, which was its best year, until 2013 where this rate remains the same until 2016.
-Figure~\ref{fig:HFTmarket1} shows HFT trades percentage of US equity shares.
+In 2016 HFT represented nearly 50\% of equity trades in the US showing a
+consistent fall since 2009, which was its best year, until 2013 where this rate
+remains the same until 2016.  Figure~\ref{fig:HFTmarket1} shows HFT trades
+percentage of US equity shares.
 
@@
   \includegraphics[width=0.9\textwidth]{img/HFTmarket1.png}
-  \caption[HFT market share in the US and Europe]{HFT market share in the US. Source: TABB Group.}
+  \caption[HFT market share in the US and Europe]{HFT market share in the US.
+  Source: TABB Group.}
   \label{fig:HFTmarket1}
@@
 \end{figure}
-US market revenues have fallen dramatically.
-Figure~\ref{fig:HFTmarket2} shows how revenues in US stocks have fallen the last years.
-One of the reasons of this fall is that exchange markets have adapted, having
-now faster, more transparent and efficient market structures than before.
-This has been possible due to its investment in technology enhancing
-reliability and stability of transactions. Even though this fall, HFT is still a major component of regulated markets and
+US market revenues have fallen dramatically. Figure~\ref{fig:HFTmarket2} shows
+how revenues in US stocks have fallen the last years. One of the reasons of this
+fall is that exchange markets have adapted, having now faster, more transparent
+and efficient market structures than before.  This has been possible due to its
+investment in technology enhancing reliability and stability of transactions.
+Even though this fall, HFT is still a major component of regulated markets and
 will probably remain as a topic of interest for researchers in the near future.
@@
 On the other hand, HFT has been criticised on qualitative issues concerning
-fairness and systemic risk. However, HFT has lead to
-beneficial impacts such as reducing spreads (difference between buyers (bid) and
-sellers (ask) prices), increased liquidity, allowing more efficient price formation,
-reduced transaction costs and lower market volatility.
+fairness and systemic risk. However, HFT has lead to beneficial impacts such as
+reducing spreads (difference between buyers (bid) and sellers (ask) prices),
+increased liquidity, allowing more efficient price formation, reduced
+transaction costs and lower market volatility.
 
@@
   \includegraphics[width=0.7\textwidth]{img/bid-ask-spreads.jpg}
-  \caption[Bid-ask spreads]{Bid-ask spreads for large and small caps stock US. Source: Credit Suisse.}
+  \caption[Bid-ask spreads]{Bid-ask spreads for large and small caps stock US.
+  Source: Credit Suisse.}
   \label{fig:Spreads}
 \end{figure}
-In figure \ref{fig:Spreads}  shows the Bid-ask spread for large and small caps. A large cap refers to an asset with market capitalisation value of more than \$10 billion dollars.
+In figure \ref{fig:Spreads}  shows the Bid-ask spread for large and small caps.
+A large cap refers to an asset with market capitalisation value of more than
+\$10 billion dollars.
 
-In theory, Bid-ask spreads for large caps and small caps move on the same direction, meaning they both are affected by volatility.
-This scenario is only true until 2009, where the large caps (in red) gets tight spreads and small caps have wider  spreads, suggesting a concentrating of trading in the most liquid, biggest stocks.
+In theory, Bid-ask spreads for large caps and small caps move on the same
+direction, meaning they both are affected by volatility. This scenario is only
+true until 2009, where the large caps (in red) gets tight spreads and small caps
+have wider  spreads, suggesting a concentrating of trading in the most liquid,
+biggest stocks.
 
@@
 markets and capital markets. Money markets are used for short-term basis,
-usually for assets up to one year, for greater periods capital markets are used.
+usually for assets up to one year, for greater periods, capital markets are used.
 Capital markets include the stock or equity market and the bond or debt market
@@
   \includegraphics[width=0.7\textwidth]{img/capitalmarkets}
-  \caption[Old structure of capital markets]{Old structure of capital markets. Source \cite{aldridge2009}.}
+  \caption[Old structure of capital markets]{Old structure of capital markets.
+  Source \cite{aldridge2009}.}
   \label{fig:capitalmarket}
@@
 liquidity pools or Electronic Communication Networks (ECNs) which are able to
-transmit order quickly matching buyers and sellers optimally. There are also
+transmit and order quickly matching buyers and sellers optimally. There are also
 dark liquidity pools where trader identity and orders remain anonymous.

### 2017-10-21 src/2_Finance.tex
correcting chapter 2

@@
 \item[Fill or Kill] is an order that must be executed immediately or being
-cancelled, no partial fulfillments are allowed. 
+cancelled, no partial fulfilments are allowed. 
 \item[Day] the order is only valid during the day.
-\item[Good til canceled] a order is active until the investor decides to cancel
+\item[Good till cancelled] a order is active until the investor decides to cancel
 it or the trade is executed.
@@
 counterfactual assumptions regarding human behaviour, that is, rationality.
-Recent advances in evolutionary psychology and the cognitive neurosciences may
+Recent advances in evolutionary psychology and the cognitive neuroscience may
 be able to reconcile the EMH with behavioural anomalies. On the other hand, if

### 2017-10-21 src/3_TimeSeries.tex
correcting chapter 2

@@
 There are several characteristics and concepts of financial time series and here we will discuss the more important ones.
-\subsection{Stylized facts of asset returns}
-\label{sec:stylizedfacts}
+\subsection{Stylised facts of asset returns}
+\label{sec:stylisedfacts}
 
 There are several known features exhibited by financial instruments called
-stylized facts, which have been empirically studied and some of them have been
-documented only recently and accepted as truth. Stylized facts are usually
+stylised facts, which have been empirically studied and some of them have been
+documented only recently and accepted as truth. Stylised facts are usually
 formulated in terms of qualitative properties of asset returns and may not be
@@
 
-Some stylized facts which are common to a wide set of financial assets
+Some stylised facts which are common to a wide set of financial assets
 \cite{sewell2011} are:
@@
 has high kurtosis (i.e fat tails and a peaked centre compared with the normal
-distribution). However, distribution of returns whose were obtained from higher
-frequencies looks more like a normal distribution.
+distribution). 
 The returns distribution tails are larger than
@@
  distribution assumption). In the markets, fat tails are an undesirable feature
- because of the additional risk they imply.  In the figure \ref{fig:returndist}
+ because of the additional risk they imply.  However, distribution of returns whose were obtained from higher
+frequencies looks more like a normal distribution.
+In the figure \ref{fig:returndist}
  is shown the SPY returns distribution based on daily dates from the period 1st
@@
 $c_1,\dots,c_L$ are constants.  This definition is too strong and difficult to
-assess it from a single data set. The weak version of this definition imposes
-conditions only on the two first two moments.
+assess from a single data set. The weak version of this definition imposes
+conditions only on the first two moments.
 
@@
 
-In finance, a shock represents an unexpected change in a variable or a in its
-error term in a particular time period. For stationary time series, shocks to
+One of the consequences of stationary processes is how they recover from a shock. A shock represents an unexpected change in a variable or a in its
+error term in a particular period of time. For stationary time series, shocks to
 the system will gradually die away. That is, a shock during time $t$ will have a
 smaller effect in time $t+1$, a smaller effect on $t+2$ and so on. For
-non-stationary data, the persistence of shocks will always be infinite.
+non-stationary data, the persistence of shocks will always have permanent effects because is not a mean-reverting process.
 
@@
 time, when one of those variables is regressed on the other, they could have a
-high $R^2$ even if the two are totally unrelated. So, if standard regression
+high coefficient of determination ($R^2$) even if the two are totally unrelated. So, if standard regression
 techniques are applied to non-stationary data, it could look good under
@@
 
-\subsubsection{Realized volatility} is also known as historic volatility and
+\subsubsection{Realised volatility} is also known as historic volatility and
 it is the actual variance in the price of a stock over time.
-Realized volatility is measured in terms of the standard deviation
+Realised volatility is measured in terms of the standard deviation
 using the historical stock prices. It is commonly calculated based on
@@
 \noindent where $p_{t,n}$ is the price observed at day $t=1,\dots,T$ and
-intraday sample $n=2,\dots,N$. Realized volatility is defined as:
+intraday sample $n=2,\dots,N$. Realised volatility is defined as:
 \begin{equation}
@@
 Lunde~\cite{hansen+lunde2005} introduced a scaling version of
-realized volatility using the following definitions:
+realised volatility using the following definitions:
 \begin{eqnarray}
@@
 equation~(\ref{eq: ret 2}), where $p_{t,N}$ is the last intraday 
-sample at day $t$. The scaled realized volatility $\rho(t)$ is 
+sample at day $t$. The scaled realised volatility $\rho(t)$ is 
 defined as:
@@
 \end{eqnarray}
-Realized volatility has also been defined as the absolute value return or as
+Realised volatility has also been defined as the absolute value return or as
 the mean of the sum of intraday squared returns at short intervals of time. The
@@
 and Lunde~\cite{hansen+lunde2006} stated that the use of a noisy proxy could
-result in an inferior model being chosen as the best one. The realized
+result in an inferior model being chosen as the best one. The realised
 volatility, as calculated by the cumulative sum of squared intraday returns and
@@
 
-For trading strategies, the interest is centred in forecasting realized
+For trading strategies, the interest is centred in forecasting realised
 volatility over the life of an option and to take advantage when this
@@
 arbitrage. For example, a trader will buy an option and hedge the underlying
-asset if the implied volatility is under the realized volatility.
+asset if the implied volatility is under the realised volatility.
 
@@
 heteroskedasticity (GARCH) models, the stochastic volatility (SV)
-models, the realized volatility models and the machine learning based
+models, the realised volatility models and the machine learning based
 models. A comparison of the first three models can be found
@@
 
-The realized volatility constructed from high frequency intraday returns gave
-rise to the realized volatility models mainly because the realized volatility
+The realised volatility constructed from high frequency intraday returns gave
+rise to the realised volatility models mainly because the realised volatility
 series is much more homoskedastic and seems to be a long memory
-process~\cite{andersonetal2003}. For realized volatility, the autoregressive
+process~\cite{andersonetal2003}. For realised volatility, the autoregressive
 fractionally integrated moving average (ARFIMA) process emerged as a standard

### 2017-10-21 src/3_TimeSeries.tex
cap 1 y 2 revisados

@@
 explanatory variables (relationships with other variables, seasonal factors,
-etc.), signal processing and forecasting (predict future values), which is the focus of this thesis. 
-There are two main approaches to study financial time series: to study directions of financial
-rates and to explain financial rate volatility.  This chapter reviews the most
-relevant techniques in the rich and rapidly growing field of time series
-analysis considering these two approaches.
+etc.), signal processing and forecasting (predict future values), which is the
+focus of this thesis.  There are two main approaches to study financial time
+series: to study directions of financial rates and to explain financial rate
+volatility.  This chapter reviews the most relevant techniques in the rich and
+rapidly growing field of time series analysis considering these two approaches.
 
 \section{Characteristics of Financial Time Series}
-There are several characteristics and concepts of financial time series and here we will discuss the more important ones.
+There are several characteristics and concepts of financial time series and here
+we will discuss the more important ones.
 \subsection{Stylised facts of asset returns}
@@
 \begin{description}
-\item[Dependence] autocorrelation function (ACF) in returns is largely insignificant.
-Returns of a time series $y_t$ is defined as $r_t = y_t - y_{t-1}$.
-The ACF measures the linear predictability of the time series $y_t$ at time $t$ using only the value at time $s$.
-ACF is defined as:
+\item[Dependence] autocorrelation function (ACF) in returns is largely
+insignificant. Returns of a time series $y_t$ is defined as $r_t = y_t -
+y_{t-1}$. The ACF measures the linear predictability of the time series $y_t$ at
+time $t$ using only the value at time $s$. ACF is defined as:
 \begin{equation*}
@@
 \end{equation*}
-\noindent where $\gamma(s,t)$ is the auto-covariance function that measures the linear dependence between two points on the same series observed at different times.
-It is defined as the second moment product:
+\noindent where $\gamma(s,t)$ is the auto-covariance function that measures the
+linear dependence between two points on the same series observed at different
+times. It is defined as the second moment product:
 \begin{equation*}
@@
 
-Figure \ref{fig:returnacf} shows the ACF of the returns of SPY stock where the correlations are significant even
- for very long lags, this implies a long-memory process.
+Figure \ref{fig:returnacf} shows the ACF of the returns of SPY stock where the
+correlations are significant even for very long lags, this implies a long-memory
+process.
 
@@
  \centering
- \includegraphics[width=0.7\textwidth]{plots/spy_returns_acf.pdf}
+ \includegraphics[width=0.9\textwidth]{plots/spy_returns_acf.pdf}
  \caption{SPY returns ACF}
@@
 distribution). 
-The returns distribution tails are larger than
- what is hypothesised by common data generation process (generally normal
- distribution assumption). In the markets, fat tails are an undesirable feature
- because of the additional risk they imply.  However, distribution of returns whose were obtained from higher
+The returns distribution tails are larger than what is hypothesised by common
+data generation process (generally normal distribution assumption). In the
+markets, fat tails are an undesirable feature because of the additional risk
+they imply.  However, distribution of returns whose were obtained from higher
 frequencies looks more like a normal distribution.
-In the figure \ref{fig:returndist}
- is shown the SPY returns distribution based on daily dates from the period 1st
- July 1998 to 4th April 2013. The distribution was compared against the normal
- distribution which clearly doesn't fit the data.  
+In the figure \ref{fig:returndist} is shown the SPY returns distribution based
+on daily dates from the period 1st July 1998 to 4th April 2013. The distribution
+was compared against the normal distribution which clearly doesn't fit the data.  
  \begin{figure}[h]
@@
 
+\vspace{15mm}
+
 \subsection{Stationary }
@@
 
-One of the consequences of stationary processes is how they recover from a shock. A shock represents an unexpected change in a variable or a in its
-error term in a particular period of time. For stationary time series, shocks to
-the system will gradually die away. That is, a shock during time $t$ will have a
+One of the consequences of stationary processes is how they recover from a
+shock. A shock represents an unexpected change in a variable or a in its error
+term in a particular period of time. For stationary time series, shocks to the
+system will gradually die away. That is, a shock during time $t$ will have a
 smaller effect in time $t+1$, a smaller effect on $t+2$ and so on. For
-non-stationary data, the persistence of shocks will always have permanent effects because is not a mean-reverting process.
+non-stationary data, the persistence of shocks will always have permanent
+effects because is not a mean-reverting process.
 
@@
 \begin{enumerate}[(a)]
-\item \textbf{Deterministic trend:} Deterministic trend or trend stationary processes have the following form:
+\item \textbf{Deterministic trend:} Deterministic trend or trend stationary
+processes have the following form:
 
@@
 
-\noindent where $t$ is the time trend and $\epsilon_{t}$ represents a stationary error term (with mean 0 and variance $\sigma^2$) and $f(t)$ is a deterministic function of time:
+\noindent where $t$ is the time trend and $\epsilon_{t}$ represents a stationary
+error term (with mean 0 and variance $\sigma^2$) and $f(t)$ is a deterministic
+function of time:
 \begin{itemize}
@@
 \end{itemize}
-\item \textbf{Stochastic trend:} Stochastic trend processes are also called unit root or difference stationarity processes and have the following form:
+\item \textbf{Stochastic trend:} Stochastic trend processes are also called unit
+root or difference stationarity processes and have the following form:
 
@@
 \]
-\noindent where $\epsilon_t$ is a stationary process. When $\mu = 0 $ the process is called pure random walk and when $\mu \neq 0$  the process is called random walk with drift.
+\noindent where $\epsilon_t$ is a stationary process. When $\mu = 0 $ the
+process is called pure random walk and when $\mu \neq 0$  the process is called
+random walk with drift.
 
@@
 
-This process is also called unit root because the root of the characteristic equation ($1-z = 0$) is the unity.
-
-A random walk with or without drift can be transformed to a stationary process by differencing the time serie once.
-The disadvantage of differencing is that the process loses one observation each time the time series is differentiated.
-
-Apart from a stochastic trend, many economic financial time series seem to involve an exponential trend, this is the reason why researchers often take the logarithmic transformation before doing analysis.
-
-Other less common forms of non-stationarity are structural break in mean and structural break in variance. 
+This process is also called unit root because the root of the characteristic
+equation ($1-z = 0$) is the unity.
+
+A random walk with or without drift can be transformed to a stationary process
+by differencing the time serie once. The disadvantage of differencing is that
+the process loses one observation each time the time series is differentiated.
+
+Apart from a stochastic trend, many economic financial time series seem to
+involve an exponential trend, this is the reason why researchers often take the
+logarithmic transformation before doing analysis.
+
+Other less common forms of non-stationarity are structural break in mean and
+structural break in variance. 
 \end{enumerate}
@@
 regression dates back to Yule in 1926 \cite{yule1926}. If two stationary
-variables are generated as independent random series and are trending over
-time, when one of those variables is regressed on the other, they could have a
-high coefficient of determination ($R^2$) even if the two are totally unrelated. So, if standard regression
-techniques are applied to non-stationary data, it could look good under
-standard measures but valueless \cite{brooks2002}.
-The
-regression measure $R^2$ gives high values in presence of spurious
+variables are generated as independent random series and are trending over time,
+when one of those variables is regressed on the other, they could have a high
+coefficient of determination ($R^2$) even if the two are totally unrelated. So,
+if standard regression techniques are applied to non-stationary data, it could
+look good under standard measures but valueless \cite{brooks2002}.
+The regression measure $R^2$ gives high values in presence of spurious
 relationships. This measure would means that the model explains all the
@@
 
+\vspace{15mm}
 \textbf{Example of spurious regression} \quad 
@@
 
+Figure \ref{fig:spurious1} shows the time series $x_{1,t}$ and $x_{2,t}$ and
+figure \ref{fig:spurious2} shows the regression between them and how they seem
+to be related.
 \begin{figure}[!h]
@@
 
-A (vector) time series $\mathbf{y}_t$ is said to be {\em integrated of order\/}
+A vector time series $\mathbf{y}_t$ is said to be {\em integrated of order\/}
 $d$, and then we shall write $\mathbf{y}_t\sim I(d)$, whenever after $d$ times
@@
 \subsection{Cointegration} \label{sec:cointegration}
-Cointegration concept was introduced by Engle in 1987 \cite{engle1987} and implies that one or
-more linear combinations of non-stationary variables are stationary even though
-individually they are not.
+Cointegration concept was introduced by Engle in 1987 \cite{engle1987} and
+implies that one or more linear combinations of non-stationary variables are
+stationary even though individually they are not.
 
@@
 \subsection{Johansen method}
-Johansen in 1988 \cite{johansen1988} suggests a method for determining cointegration vectors.
-
-There are two statistics for cointegration under Johansen approach: the trace (shown in equation \ref{eq:johansentrace}) and the maximum-eigenvalue statistic (shown in equation \ref{eq:johansenmax}):
+Johansen in 1988 \cite{johansen1988} suggests a method for determining
+cointegration vectors.
+
+There are two statistics for cointegration under Johansen approach: the trace
+(shown in equation \ref{eq:johansentrace}) and the maximum-eigenvalue statistic
+(shown in equation \ref{eq:johansenmax}):
 
@@
 
-\noindent where $r$ is the number of cointegration vectors and $\hat]{\lambda}_i$ is the estimated value for the ith ordered eigenvalue.
-$\lambda_{\text{trace}}$ null hypothesis is that the number of cointegration vectors is less than or equal to $r$ against that there are more than $r$.
-$\lambda_{\text{max}}$ has the null hypothesis that the number of cointegration vectors is $r$ against an alternative of $r
-+1$.
-
-\newpage
+\noindent where $r$ is the number of cointegration vectors and
+$\hat]{\lambda}_i$ is the estimated value for the ith ordered eigenvalue.
+$\lambda_{\text{trace}}$ null hypothesis is that the number of cointegration
+vectors is less than or equal to $r$ against that there are more than $r$.
+$\lambda_{\text{max}}$ has the null hypothesis that the number of cointegration
+vectors is $r$ against an alternative of $r +1$.
+
+\vspace{15mm}
+
 \textbf{Cointegration example} \quad
@@
 \end{eqnarray*}
+
+\begin{figure}[!h]
+  \centering
+  \includegraphics[width=\textwidth]{img/coint-example}
+  \caption[Cointegration example]{Cointegration example: two I(1) processes (blue and green) and one I(0) process (red).}
+  \label{fig:cointex}
+\end{figure}
 
@@
 
-\begin{figure}[!h]
-  \centering
-  \includegraphics[width=\textwidth]{img/coint-example}
-  \caption[Cointegration example]{Cointegration example: two I(1) processes (blue and green) and one I(0) process (red).}
-  \label{fig:cointex}
-\end{figure}
+Figure \ref{fig:cointex} shows the three time series examples.
 
@@
 between the processes by linear relations that are more stationary than the
-original process. Cointegration vectors can be obtained using the Johansen method which gives with a certain probability the number of significant vectors. Figures    
-\ref{fig:cointvectors3}, \ref{fig:cointvectors4}  and \ref{fig:cointvectors5}  shows the linear combinations obtained using cointegration vectors given by Johansen method.
-The method confirm that only two vectors are found and the third one given doesn't result in a I(0) process.
+original process. Cointegration vectors can be obtained using the Johansen
+method which gives with a certain probability the number of significant vectors.
+Figures \ref{fig:cointvectors3}, \ref{fig:cointvectors4}  and
+\ref{fig:cointvectors5}  shows the linear combinations obtained using
+cointegration vectors given by Johansen method. The method confirm that only two
+vectors are found and the third one given doesn't result in a I(0) process.
+\vfill
 
 \section{Unit root tests}
-Many economic and financial time series exhibit trending behaviour or non-stationary behaviour in the mean or variance.
-An important econometric task is to determine the type of non-stationary.
-For trend stationary I(0) time series a time-trend regression is appropriate.
-For I(1) time series taking first differences transforms it into a stationary one.
-For cointegration to determine if time series are I(1) is mandatory to model their long-run relationship.
-In order to determine if a time series is an a I(1) or I(0) process unit root tests are used.
-Several tests have been developed but the most widely used is an extension of the Dickey and Fuller test \cite{dickey1979} called the Augmented Dickey-Fuller (ADF) test developed by Said and Dickey in 1984\cite{said1984} which tabulated critical values were obtained by~\cite{mackinnon2010}.
-
-The ADF test tests the null hypothesis that a time series $y_t$ is I(1) against the alternative that it is I(0), assuming that the dynamics in the data have an ARMA structure.
-The ADF test is based on estimating the test regression:
+Many economic and financial time series exhibit trending behaviour or
+non-stationary behaviour in the mean or variance. An important econometric task
+is to determine the type of non-stationarity. For trend stationary I(0) time
+series a time-trend regression is appropriate. For I(1) time series taking first
+differences transforms it into a stationary one.  For cointegration to determine
+if time series are I(1) is mandatory to model their long-run relationship. In
+order to determine if a time series is an a I(1) or I(0) process unit root tests
+are used. Several tests have been developed but the most widely used is an
+extension of the Dickey and Fuller test \cite{dickey1979} called the Augmented
+Dickey-Fuller (ADF) test developed by Said and Dickey in 1984 \cite{said1984}
+which tabulated critical values were obtained by~\cite{mackinnon2010}.
+
+The ADF test tests the null hypothesis that a time series $y_t$ is I(1) against
+the alternative that it is I(0), assuming that the dynamics in the data have an
+ARMA structure. The ADF test is based on estimating the test regression:
 
@@
 
-\noindent where $\mathbb{D}_t$ is a vector of deterministic terms. Lag length $p$ has to be determined when applying the test, Akaike Information Criteria (AIC) is commonly used to determine it.
-The p-lagged difference terms $\Delta y_{t-j}$ are used to approximate an ARMA process and the value of $p$ is set so that the error $\epsilon_t$ is serially uncorrelated. 
-Under the null hypothesis, $y_t$ is I(1) which implies that $\phi=1$. 
-The test statistic is the following:
+\noindent where $\mathbb{D}_t$ is a vector of deterministic terms. Lag length
+$p$ has to be determined when applying the test, Akaike Information Criteria
+(AIC) is commonly used to determine it.  The p-lagged difference terms $\Delta
+y_{t-j}$ are used to approximate an ARMA process and the value of $p$ is set so
+that the error $\epsilon_t$ is serially uncorrelated. Under the null hypothesis,
+$y_t$ is I(1) which implies that $\phi=1$. The statistic test is the following:
 
@@
 
-\section {Cointegration Tests}
-
-Johansen method~\cite{johansen1988} allows to find cointegration among several
-I(1) time series. This test is more general than the
-Engle-Granger~\cite{engle1987} test which only allow to find cointegration
-between two time series.  Johansen procedure takes the form of a
-likelihood-ratio test which determines $r$ cointegration relationships. The
-method provides two tests based on an eigenvector decomposition of the matrix
-$\Omega = \alpha \beta^\top$ on equation~\ref{eq:vec} called trace and maximal
-eigenvalue statistic.  Once the number of cointegration number is determined,
-$r$ more significative eigenvectors provided by Johansen procedure are used to
-get the cointegration vector matrix $\beta^\top$.  Critical values for both
-tests are limited to 12 and 5 variables in the trace and maximal eigenvalue
-statistic respectively.
-
-The trace test null hypothesis is $H_0: r = r_0$ against $H_1: r > r_0$ and the
-maximum eigenvalue test null hypothesis is $H_0: r = r_0$ against $H_1: r =
-r_0+1$ starting with $r_0 = 0$.  In the latest test, if we start with $r_0=0$
-and $H_0$ is rejected $H_1: r = 1$ will be accepted which is not true if there
-are more than one cointegration relationship, this is why this test is less
-powerful and the trace test is commonly used.
+\vfill
 
@@
 Besides, volatility provides important information for studying asset returns
-because the latter are largely uncorrelated and nonlinearly dependent. However,
+because they largely uncorrelated and nonlinearly dependent. However,
 it has been found that volatility exhibits significant autocorrelation and
@@
 
-Despite the fact variance and volatility are related, they are not the same concept. Variance is a measure of distribution of returns and is not necessarily
+Despite the fact variance and volatility are related, they are not the same
+concept. Variance is a measure of distribution of returns and is not necessarily
 bound by any time period.  Volatility is a measure of the standard deviation
-(square root of the variance) over a certain time interval. In finance,
-variance and volatility both gives you a sense of an asset's risk. Variance
-gives you a sense of the risk in the asset over its lifetime, while volatility
-gives you a sense of the movement of the asset in, for example, the past month or the
-past year. The main underlying difference is in their definition. Variance has a
+(square root of the variance) over a certain time interval. In finance, variance
+and volatility both gives you a sense of an asset's risk. Variance gives you a
+sense of the risk in the asset over its lifetime, while volatility gives you a
+sense of the movement of the asset in, for example, the past month or the past
+year. The main underlying difference is in their definition. Variance has a
 fixed mathematical definition, however volatility does not as such. Volatility
@@
 The volatility of a stock is not directly observable \cite{tsay2005,engle1993}.
-For example, daily volatility is not directly observable from only daily returns because there is only one observation in a trading day.
-If intraday data is available, then volatility could be estimated.
-However, intraday returns are not the only explanatory variables for volatility and several estimators have been proposed.
-These estimators are observable variables that are related to the latent variable of interest called volatility proxies~\cite{devilderetal2007}.
+For example, daily volatility is not directly observable from only daily returns
+because there is only one observation in a trading day.  If intraday data is
+available, then volatility could be estimated. However, intraday returns are not
+the only explanatory variables for volatility and several estimators have been
+proposed. These estimators are observable variables that are related to the
+latent variable of interest called volatility proxies~\cite{devilderetal2007}.
 Examples of volatility proxies are the following:
@@
 \begin{description}
-\item [$\sigma$:] underlying asset price volatility which measures the standard
+\item [$\sigma$] underlying asset price volatility which measures the standard
 deviation of the returns
-\item [$\mu$:] underlying asset drift which is a measure of the average rate of
+\item [$\mu$] underlying asset drift which is a measure of the average rate of
 growth of the stock
-\item[$E$:] option strike or excersice price
-\item[$T$:] option date of expiry
-\item[$t$:] current time
-\item[$r$:] risk-free interest rate
+\item[$E$] option strike or excersice price
+\item[$T$] option date of expiry
+\item[$t$] current time
+\item[$r$] risk-free interest rate
 \end{description}
@@
 has two components: a deterministic term given by $\mu S dt$ and a random term
-given by $\sigma S dX$.
+given by $\sigma S dB$.
 
 A brownian motion $B$ (also called a Wiener process) is a stochastic process
-characterized by the three following properties:
+characterised by the three following properties:
 
 \begin{description}
-\item[Continuity:] $B(t)$ is a continuos function
-\item[Normal increments:]  $B(t)-B(s)$ has a normal distribution with mean $0$
+\item[Continuity] $B(t)$ is a continuos function
+\item[Normal increments]  $B(t)-B(s)$ has a normal distribution with mean $0$
 and variance $t-s$.
-\item[Independence of increments:] for every choice of nonnegative real numbers
+\item[Independence of increments] for every choice of nonnegative real numbers
 $0 \leq s_1 <  t_1 \leq \cdots \leq s_n < t_n < \infty$, the increment random
@@
 be  differentiable function of $S$ and once differentiable function of $t$.
-
-These are known or can be easily obtained from the market, excepting by
-volatility which must be estimated. However, rather than assuming a volatility
-a priori and computing option prices from it, the model can be used to estimate
-volatility at given prices, time to expiration and strike price. This obtained
-volatility is called the implied volatility of an option. Additionally, some
-models obtain implied volatility from futures (other derivative from prices).
 
@@
 
-In the existing literature, there are four main classes of asset
-return volatility models: the general autoregressive conditional
-heteroskedasticity (GARCH) models, the stochastic volatility (SV)
-models, the realised volatility models and the machine learning based
-models. A comparison of the first three models can be found
-in~\cite{wei2012}.
-
-For many years the most popular methods for estimating financial
-volatility were the autoregressive conditional heteroskedasticity
-(ARCH) models~\cite{engle1982} and the general ARCH (GARCH)
-models~\cite{bollerslev1986}. For instance, the GARCH(1,1) defines
-returns $y_t$ and volatility $\sigma_t$ as:
+In the existing literature, there are four main classes of asset return
+volatility models: the general autoregressive conditional heteroskedasticity
+(GARCH) models, the stochastic volatility (SV) models, the realised volatility
+models and the machine learning based models. A comparison of the first three
+models can be found in~\cite{wei2012}.
+
+For many years the most popular methods for estimating financial volatility were
+the autoregressive conditional heteroskedasticity (ARCH) models~\cite{engle1982}
+and the general ARCH (GARCH) models~\cite{bollerslev1986}. For instance, the
+GARCH(1,1) defines returns $y_t$ and volatility $\sigma_t$ as:
 
@@
 
-The random walk model, despite its simplicity is still difficult to outperform for standard econometric forecasting models \cite{lo2011} despite its simplicity. The random walk model is defined as:
+The random walk model, despite its simplicity, is still difficult to outperform
+for standard econometric forecasting models \cite{lo2011}. The random walk model
+is defined as:
 
@@
 
-On the other hand, ARIMA is widely used to forecast returns in finance \cite{tsay2005}.
-A process can be modelled as an ARIMA$(p,d,q)$ model if $\mathbf{x}_t=\Delta^d \mathbf{y}_t $, i.e after differencing $d$ times the time series $\mathbf{y}_t$,  we get an ARMA$(p,q)$.
-An ARMA$(p,q)$ model is the following:
+On the other hand, ARIMA is widely used to forecast returns in finance
+\cite{tsay2005}. A process can be modelled as an ARIMA$(p,d,q)$ model if
+$\mathbf{x}_t=\Delta^d \mathbf{y}_t $, i.e after differencing $d$ times the time
+series $\mathbf{y}_t$,  we get an ARMA$(p,q)$. An ARMA$(p,q)$ model is the
+following:
 
@@
 
-The vector autoregressive VAR($p$) model \cite{sims1980} is one of the most easy to use, successful and flexible models for the analysis of multivariate time series.
-It is a natural extension of the univariate autoregressive (AR) model.
-The VAR model has proven to be useful for describing the dynamic behaviour of economic and financial time series.
-VAR is a general framework describing the behaviour of a
-set of $l$ endogenous variables as a linear combination of their last $p$
-values, where $l,p\in\mathbb{N}$. 
-In our case, each one of these $l$ variables is a scalar time series
-$y_{i,t}$, $i=1,\dots,l$, and we represent them all together
-at time $t$ by the the vector time series:
+The vector autoregressive VAR($p$) model \cite{sims1980} is one of the most easy
+to use, successful and flexible models for the analysis of multivariate time
+series. It is a natural extension of the univariate autoregressive (AR) model.
+The VAR model has proven to be useful for describing the dynamic behaviour of
+economic and financial time series.  VAR is a general framework describing the
+behaviour of a set of $l$ endogenous variables as a linear combination of their
+last $p$ values, where $l,p\in\mathbb{N}$.  In our case, each one of these $l$
+variables is a scalar time series $y_{i,t}$, $i=1,\dots,l$, and we represent
+them all together at time $t$ by the the vector time series:
 \begin{equation*}
@@
 
-Notice that, regarding our notation of section (\ref{sec:cointegration}),
-we have here 
+Notice that, regarding our notation of the cointegration section
+(\ref{sec:cointegration}), we have here 
 $\mathbf{y}_t^0 = \mathbf{y}_t$,
@@
 
+Conventional regression estimators, including VARs, have good properties when
+applied to covariance-stationary time series, but encounter difficulties when
+applied to non stationary or integrated processes.
+
+These difficulties were illustrated by Granger and Newbold in 1974
+\cite{granger1974} when they introduced the concept of spurious regressions (see
+section \ref{sec:spurious}). In 1982, Nelson and Plosser \cite{nelson1982}
+showed that unit roots might be present in a wide variety of macroeconomic
+series in levels or logarithms. This finding made the unit root testing very
+popular. Aditionally, the implication that variables should be rendered
+stationary by differencing before they are included in an econometric model was
+generalised.  Further theoretical developments by Granger and Engle in 1987
+\cite{engle87} raised the possibility that two or more integrated, non
+stationary time series might be cointegrated, so that some linear combination of
+these series could be stationary even though each series were not.
+
 \section{Vector error correction model}\label{sec:vecm}
-The VAR model cannot be used with non-stationary variables. VECM ,
-developed by \cite{engle87}, is also a linear model for I(1) variables that are
-also cointegrated, see \cite{banerjee1993}. If cointegration exists, variable
+
+VECM, developed by \cite{engle87}, is a linear model for I(1) variables that are
+cointegrated, see \cite{banerjee1993}. If cointegration exists, variable
 differences are stationary and they introduce an error correction term which
@@
 
-It is obtained re-writing equation \ref{eq:var} in terms of the new
+In finance, many economic time series are revealed to be stationary when they
+are differentiated and moreover cointegration restrictions often improve
+forecasting \cite{duy1998}. Therefore, VECM has been widely adopted in finance:
+\cite{mukherjee1995}, \cite{maysami2000}, \cite{arestis2001} and
+\cite{seong2013} to name a few. Pair trading is a very common example of a
+cointegration application \cite{herlemont2003} but also can also be extended to
+a larger set of variables \cite{mukherjee1995}, \cite{engle2004}.
+
+VECM  is obtained re-writing equation \ref{eq:var} in terms of the new
 variable $\Delta\mathbf{y}_t=\mathbf{y}_t-\mathbf{y}_{t-1}$.
@@
 \end{itemize}
-It is worth noticing that the factorization of the matrix
+It is worth noticing that the factorisation of the matrix
 $\boldsymbol\Omega$ is not unique, since for any $r \times r$
@@
 squares estimation (see section \ref{sec:OLS}).
-
-\subsection{Ordinary Least Squares method} \label{sec:OLS}
+\section{Cointegration Tests}
+
+Johansen method~\cite{johansen1988} allows to find cointegration among several
+I(1) time series. This test is more general than the
+Engle-Granger~\cite{engle1987} test which only allow to find cointegration
+between two time series.  Johansen procedure takes the form of a
+likelihood-ratio test which determines $r$ cointegration relationships. The
+method provides two tests based on an eigenvector decomposition of the matrix
+$\Omega = \alpha \beta^\top$ on equation~\ref{eq:vec} called trace and maximal
+eigenvalue statistic.  Once the number of cointegration number is determined,
+$r$ more significative eigenvectors provided by Johansen procedure are used to
+get the cointegration vector matrix $\beta^\top$.  Critical values for both
+tests are limited to 12 and 5 variables in the trace and maximal eigenvalue
+statistic respectively.
+
+The trace test null hypothesis is $H_0: r = r_0$ against $H_1: r > r_0$ and the
+maximum eigenvalue test null hypothesis is $H_0: r = r_0$ against $H_1: r =
+r_0+1$ starting with $r_0 = 0$.  In the latest test, if we start with $r_0=0$
+and $H_0$ is rejected $H_1: r = 1$ will be accepted which is not true if there
+are more than one cointegration relationship, this is why this test is less
+powerful and the trace test is commonly used.
+
+\section{Ordinary Least Squares method} \label{sec:OLS}
 
@@
 \newpage
-Conventional regression estimators, including VARs, have good properties when
-applied to covariance-stationary time series, but encounter difficulties when
-applied to non stationary or integrated processes.
-
-These difficulties were illustrated by Granger and Newbold in 1974
-\cite{granger1974} when they introduced the concept of spurious regressions
-(see section \ref{sec:spurious}). In 1982, Nelson and Plosser \cite{nelson1982}
-showed that unit roots might be present in a wide variety of macroeconomic
-series in levels or logarithms. This finding gave rise to the industry of unit
-root testing, and the implication that variables should be rendered stationary
-by differencing before they are included in an econometric model.  Further
-theoretical developments by Granger and Engle in 1987 \cite{engle87} raised the
-possibility that two or more integrated, non stationary time series might be
-cointegrated, so that some linear combination of these series could be
-stationary even though each series were not.
-
-If two series are both integrated (of order one, or I(1)) we could model their
-interrelationship by taking first differences of each series and including the
-differences in a VAR or a structural model.  However, this approach would be
-suboptimal if it was determined that these series are indeed cointegrated. In
-that case, the VAR only express the short-run responses of these series to
-innovations in each series. This implies that the simple regression in first
-differences is misspecified.  If the series are cointegrated, they move
-together in the long-run. A VAR in first differences will not capture those
-long-run tendencies.  VECM fit to the first differences of the non stationary
-variables, but a lagged error-correction term is added to the relationship
-representing the long-run relationship.
-
-In finance, many economic time series are revealed to be stationary when they
-are differentiated and moreover cointegration restrictions often improve
-forecasting \cite{duy1998}. Therefore, VECM has been widely adopted in finance:
-\cite{mukherjee1995}, \cite{maysami2000}, \cite{arestis2001} and
-\cite{seong2013} to name a few. Pair trading is a very common example of a
-cointegration application \cite{herlemont2003} but also can also be extended to
-a larger set of variables \cite{mukherjee1995}, \cite{engle2004}.

### 2017-10-26 src/4_OnlineLearning.tex
Adding appendix chapter. page 49

@@
 with stream data. Online machine learning techniques tackle this problem by
-updating the model with new data or by computing a new model using less data so
+updating the model with new data or by computing a new model considering less data so
 they can give a response in a short period of time.
@@
 the desire property of the selected estimator is to perform well on new data (to
-generalize), e.g. regularized least squares. In the context of inverse problems,
+generalise), e.g. regularized least squares. In the context of inverse problems,
 regularization objective is to stabilise, with respect to noise, a possibly
@@
 
+LS method consists of minimising the sum of squared errors:
+\begin{eqnarray*}
+\label{eq:lsm}
+ J(\mathbf{X}) =& \displaystyle \sum_{t=1}^n (f(\mathbf{x}_t)-y_t)^2 \\
+ =&\displaystyle \sum_{t=1}^n (\mathbf{\mathbf{A}}^\top {\mathbf{x}}_t-y_t)^2 \\
+ =& \| \mathbf{A}\mathbf{\mathbf{X}} - \mathbf{Y} \|_2^2 
+\end{eqnarray*}
+
+\noindent where 
 \begin{equation}
@@
 
-LS method consists of minimising the sum of squared errors:
-\begin{eqnarray*}
-\label{eq:lsm}
- J(\mathbf{X}) =& \displaystyle \sum_{t=1}^n (f(\mathbf{x}_t)-y_t)^2 \\
- =&\displaystyle \sum_{t=1}^n (\mathbf{\mathbf{A}}^\top {\mathbf{x}}_t-y_t)^2 \\
- =& \| \mathbf{A}\mathbf{\mathbf{X}} - \mathbf{Y} \|_2^2 
-\end{eqnarray*}
-
-\noindent where $f(\mathbf{x}_t)={\mathbf{A}}^\top {\mathbf{x}}_t$. To solve equation \ref{eq:lsm} is equivalent to find a solution of:
+\noindent and $f(\mathbf{x}_t)={\mathbf{A}}^\top {\mathbf{x}}_t$.
+
+To solve equation \ref{eq:lsm} is equivalent to find a solution of:
 \begin{equation}
@@
 
-\textbf{Proof}\quad
-
-In the case matrix $\mathbf{A}$ is non singular, its solution is
-straight forward:
-
-\begin{equation}
-\label{OLSsolution}
-    \mathbf{\mathbf{X}}=\mathbf{A}^{-1}\mathbf{Y}
-\end{equation}
-
-Since the problem shown in equation~(\ref{eq:regproblem}) has not
-solution, the minimum norm given by equation~(\ref{OLSsolution}) is
-obtained by solving the equivalent problem:
-
-\begin{equation*}
-\label{eq:proyectorsol}
-\mathbf{A \hat{\mathbf{X}} = PY} 
-\end{equation*}
-
-\noindent where $\mathbf{P=U_1 U_1^\top}$ is the projection onto the
-Col($\mathbf{A}$).
-
-Since $\mathbf{V} = [\underset{(n \times k)}{\mathbf{V_1}} |
-\underset{(n \times k)}{\mathbf{V_2}}]$ and $\mathbf{V_1^\top V_2 =
-0}$ we can express $\mathbf{\hat{\mathbf{X}}} = \mathbf{V_1 \mathbf{X}_1 + V_2 \mathbf{X}_2}$
-with $\mathbf{\mathbf{X}_2=0}$ because $\mathbf{\hat{\mathbf{X}}}$ lives in the
-$\text{Row}(\mathbf{A})$ given by $\mathbf{V_1}$, so we have:
-
-\begin{eqnarray*}
-\mathbf{A \hat{\mathbf{X}}} &=& \mathbf{PY} \\
-\mathbf{U_1 \Sigma_1 V_1^\top \hat{\mathbf{X}}} &=& \mathbf{U_1 U_1^\top Y} \\
-\mathbf{ V_1^\top \hat{\mathbf{X}}} &=&  \mathbf{\Sigma_1^{-1} U_1^\top Y} \\ 
-\mathbf{ V_1^\top V_1 \mathbf{X}_1} &=& \mathbf{\Sigma_1^{-1}
-U_1^\top Y} \\
-\mathbf{\mathbf{X}_1}&=& \mathbf{\Sigma_1^{-1} U_1^\top Y}
-\end{eqnarray*}
-
-\noindent from this result we can obtain $\mathbf{\hat{\mathbf{X}}}$ and
-therefore the pseudo-inverse expression:
-
-\begin{eqnarray*}
-\mathbf{\hat{\mathbf{X}}} &=& \mathbf{V_1 \mathbf{X}_1} \\
-                &=& \mathbf{V_1 \Sigma_1^{-1} U_1^\top Y} \\
-\mathbf{A^+} &=& \mathbf{V_1 \Sigma_1^{-1} U_1^\top} \, .
-\end{eqnarray*}
-
-$\blacksquare$
+Proof can be found in the appendix section \ref{app:pseudoproof}.
 
@@
 
-\textbf{Proof}\quad
-
-\begin{eqnarray*}
-J(\mathbf{\mathbf{X}}) &=&  \| \mathbf{A}\mathbf{\mathbf{X}} - \mathbf{Y} \|_2^2  + \lambda
- \| \mathbf{\mathbf{X}}\| ^2 \\
- &=&  \sum_{t=1}^N (\mathbf{\mathbf{X}}^\top {\bf a}_t-y_t)^2 + \lambda \sum_{i=1}^p \mathbf{X}_i^2 \\
- &=& (\mathbf{X}^\top a_1-y_1)^2 + \dots + (\mathbf{X}^\top a_N-y_N)^2 + \lambda (\mathbf{X}_1^2+\dots+\mathbf{X}_p^2)
-\end{eqnarray*}
-\noindent taking derivatives
- \begin{eqnarray*}
- \frac{\partial J(\mathbf{\mathbf{X}})}{\partial \mathbf{X}_1}&=& 
- 2(\mathbf{X}^\top \mathbf{a}_1-y_1)\mathbf{a}_{11} + \dots + 2(\mathbf{X}^\top \mathbf{a}_N-y_N)\mathbf{a}_{N1} + 2\lambda \mathbf{X}_1 \\
- &=& 2\mathbf{a}_1^\top(\mathbf{A}\mathbf{X}-\mathbf{Y}) + 2\lambda\mathbf{X}_1\\
-& \vdots &\\
-  \frac{\partial J(\mathbf{\mathbf{X}})}{\partial \mathbf{X}_p}&=& 
- 2\mathbf{a}_p^\top(\mathbf{A}\mathbf{X}-\mathbf{Y}) + 2\lambda\mathbf{X}_p 
-\end{eqnarray*}
-
-Then we have that:
-\begin{eqnarray*}
-\frac{\partial J(\mathbf{\mathbf{X}})}{\partial \mathbf{X}}&=& 
- 2\mathbf{A}^\top(\mathbf{A}\mathbf{X}-\mathbf{Y}) + 2\lambda\mathbf{X}
-\end{eqnarray*}
-Since $\frac{\partial J(\mathbf{\mathbf{X}})}{\partial \mathbf{X}}=0$ we have:
-\begin{eqnarray*}
-2\mathbf{A}^\top(\mathbf{A}\mathbf{X}-\mathbf{Y}) + 2\lambda\mathbf{X}&=&0 \\
-\mathbf{A}^\top\mathbf{A}\mathbf{X} - \mathbf{A}^\top\mathbf{Y} + \lambda\mathbf{X} &=& 0\\
-(\mathbf{A}^\top\mathbf{A}+\lambda\mathbb{I})\mathbf{X} &=&  \mathbf{A}^\top\mathbf{Y} \\
-\mathbf{X} &=& (\mathbf{A}^\top\mathbf{A}+\lambda\mathbb{I})^{-1}  \mathbf{A}^\top\mathbf{Y}
-\end{eqnarray*}
-
-$\blacksquare$
+Proof can be found in section \ref{app:rroptsection}.
 
@@
 singularity of the covariance matrix $\mathbf{A}^\top \mathbf{A}$. This addition ensures that all of the covariance matrix eigenvalues will be strictly greater than 0, i.e the solution becomes unique.
-%=======================HASTA ACA REVISADO PROFE =========================
 
@@
 
-The solution of OLS is well known $\hat{\mathbf{X}}=(\mathbf{A}^\top \mathbf{A})^{-1}\mathbf{A}^\top \mathbf{Y}$, and its bias is:
-
-\begin{eqnarray*}
-Bias(\hat{f}(\hat{\mathbf{X}})) &=& E[\hat{\mathbf{X}}] - \mathbf{X} \\
-&=& E[ (\mathbf{A}^\top \mathbf{A})^{-1}\mathbf{A}^\top \mathbf{Y}] - \mathbf{X} \\
-&=& E[ (\mathbf{A}^\top \mathbf{A})^{-1}\mathbf{A}^\top (\mathbf{AX})] - \mathbf{X}  \\
-&=& \mathbf{X}  - \mathbf{X}  \\
-&=&  0
-\end{eqnarray*}
+The solution of OLS is well known $\hat{\mathbf{X}}=(\mathbf{A}^\top \mathbf{A})^{-1}\mathbf{A}^\top \mathbf{Y}$, and its bias ($Bias(\hat{f}(\hat{\mathbf{X}})$)  is 0 (Proof in section \ref{app:biasandvariance}).
 
@@
 
-\begin{eqnarray*}
-\mathbf{X}(\lambda) &=&( \mathbf{A}^\top \mathbf{A} + \lambda \mathbb{I})^{-1}\mathbf{A}^\top \mathbf{Y} \\
-&=& (\mathbb{I} + \lambda (\mathbf{A}^\top \mathbf{A})^{-1})^{-1} (\mathbf{A}^\top \mathbf{A})^{-1}\mathbf{A}^\top \mathbf{Y} \\
-&=&  (\mathbb{I} + \lambda (\mathbf{A}^\top \mathbf{A})^{-1})^{-1}  \hat{\mathbf{X}} \\
-&=& \mathbf{W} \hat{\mathbf{X}} 
-\end{eqnarray*}
+\begin{equation}
+ Bias(\mathbf{X}(\lambda))  = \mathbf{W} \mathbf{X} - \mathbf{X} \neq 0 
+ \end{equation}
 
  \noindent where $\mathbf{W}  = (\mathbb{I} + \lambda (\mathbf{A}^\top
-\mathbf{A})^{-1})^{-1}  $ it is defined for simplicity. Ridge
-regression bias is then obtained as:
-
-\begin{eqnarray*}
-Bias(\mathbf{X}(\lambda)) &=& E[\mathbf{X}(\lambda)] - \mathbf{X} \\
-&=& E[\mathbf{W}\hat{\mathbf{X}}] - \mathbf{X} \\
-&=&  \mathbf{W} \mathbf{X} - \mathbf{X} \neq 0 
-\end{eqnarray*}
+\mathbf{A})^{-1})^{-1}  $
 
@@
 
-\begin{eqnarray*}
-Var(\mathbf{X}(\lambda)) &=& Var(\mathbf{W}\hat{\mathbf{X}}) \\
-&=& E[(\mathbf{W}\hat{\mathbf{X}}-E[\mathbf{W}\hat{\mathbf{X}}])(\mathbf{W}\hat{\mathbf{X}}-E[\mathbf{W}\hat{\mathbf{X}}])^\top] \\
-&=& \mathbf{W}E[(\hat{\mathbf{X}}-E[\hat{\mathbf{X}}])(\hat{\mathbf{X}}-E[\hat{\mathbf{X}}])^\top] \mathbf{W}^\top \\
-&=& \mathbf{W}Var(\hat{\mathbf{X}})\mathbf{W}^\top \\
-&=& \sigma^2 \mathbf{W}(\mathbf{A}^\top \mathbf{A} )^{-1}\mathbf{W}^\top
-\end{eqnarray*}
+\begin{equation*}
+Var(\mathbf{X}(\lambda)) = \sigma^2 \mathbf{W}(\mathbf{A}^\top \mathbf{A} )^{-1}\mathbf{W}^\top
+\end{equation*}
+
+Proof in section \ref{app:biasandvariance}.
 
@@
 is lower than OLS (dotted line)~\cite{hoerl1970}. Ridge regression shows an
-increasing squared bias and a decreasing variance.
-
-\textbf{Proof}\quad
-
-Since $Bias(\mathbf{X}(\lambda)) \neq 0$ this imply that
-
-\begin{equation*}
-Bias(\mathbf{X}(\lambda)) ^2 > 0 
-\end{equation*}
-
-\noindent we know that $\mathbf{A}^\top
-\mathbf{A}$ has an eigenvalue decomposition $\mathbf{A}^\top
-\mathbf{A} = \mathbf{V} \Sigma \mathbf{V}^{-1}$.
-
-\begin{eqnarray*}
-Var(\mathbf{X}(\lambda)) &=& \sigma^2 \mathbf{W}(\mathbf{A}^\top \mathbf{A} )^{-1}\mathbf{W}^\top\\
-Var(\mathbf{X}(\lambda)) &=& \sigma^2 (\mathbb{I} + \lambda (\mathbf{A}^\top
-\mathbf{A})^{-1})^{-1} (\mathbf{A}^\top \mathbf{A} )^{-1}((\mathbb{I} + \lambda (\mathbf{A}^\top
-\mathbf{A})^{-1})^{-1} )^\top \\
-&=& \sigma^2 (\mathbb{I} + \lambda (\mathbf{V} \Sigma \mathbf{V}^{-1})^{-1} (\mathbf{V} \Sigma \mathbf{V}^{-1})^{-1}((\mathbb{I} + \lambda (\mathbf{V} \Sigma \mathbf{V}^{-1})^{-1})^{-1} )^\top \\
-&=& \sigma^2 \mathbf{V} (\mathbb{I} + \lambda \Sigma^{-1})^{-1} \Sigma^{-1} (\mathbb{I} + \lambda \Sigma^{-1})^{-1}  \mathbf{V}^{-1}\\
-&=& \sigma^2 \mathbf{V} \mathbf{D}^{-1} \mathbf{V}^{-1}
-\end{eqnarray*}
-
-\noindent where $\mathbf{D}^{-1}  = (\mathbb{I} + \lambda \Sigma^{-1})^{-1} \Sigma^{-1} (\mathbb{I} + \lambda \Sigma^{-1})^{-1}$ is a diagonal matrix with coefficients:
-
-\begin{equation*}
-d_i = \frac{\sigma_i^{-1}}{(1+\lambda\sigma_i^{-1})^2}
-\end{equation*}
-
-$\blacksquare$
+increasing squared bias and a decreasing variance. See proof in section \ref{app:rrbiasvar}.
 

### 2017-10-27 src/4_OnlineLearning.tex
Chapter 3 reviewed

@@
 ill-conditioned matrix inversion problem e.g spectral cut-off and Tikhonov
-regularization or ridge regression (RR) \cite{ tikhonov1977}.
+regularization or ridge regression (RR) \cite{ tikhonov1977}. More details in section \ref{app:effcomp}.
 
@@
 
-\begin{equation*}
+\begin{equation}
 \label{eq:optsolRR}
 \mathbf{\mathbf{X}}_*=(\mathbf{A}^\top \mathbf{A}+\lambda \mathbb{I})^{-1}\mathbf{A}^\top y \, ,
-\end{equation*}
+\end{equation}
 
@@
 
-The major difference between SVM and ANN is in the error optimisation. In ANN, the aim of learning is to obtain a set of weight values which minimise the training error 
-minimum while training adjust the capacity of the machine.
+The major difference between SVM and ANN is the error optimisation.
+In ANN, the aim of learning is to obtain a set of weight values which minimise the training error while training adjust the capacity of the machine.
 
@@
 In batch learning, the training phase is commonly a computationally expensive
-process. Therefore, when new data arrives it can't be included easily to the
+process. Therefore, when new data arrives it can't be included easily into the
 model. Moreover, it could happen that we won't have enough time to process the
 new data before more data arrives. In online learning algorithms there is no
-training phase, but the model is updated and evaluated at every time step. This
+training phase, the model is updated and evaluated at every time step. This
 model updating is computationally less expensive than a training phase.
@@
 
-The objective is minimize this loss function for all instances.
+The objective is to minimise this loss function for all instances.
 The quality of online learning algorithms is measured by a quantity known as
@@
 \end{equation*}
-
-\subsection{Competitive analysis}
-
-Competitive analysis was designed for analysing the performance of an online algorithm compared with its optimal offline algorithm.
-An optimal offline algorithm can view the sequences of requests in advance.
-The effectiveness of an online algorithm \cite{sleator1985} may be measured by its competitive ratio which is the worst-case ratio of the online algorithm and the optimal offline algorithm.
 
@@
 
-If we define $\displaystyle\mathbf{S}= \sum_{t=1}^{m+1} \mathbf{a}_t
-\mathbf{a}_t  ^\intercal + \gamma \mathbb{I} $ and $\mathbf{W}=
-\displaystyle\sum_{t=1}^m \mathbf{a}_t \mathbf{y}_t$, the
-algorithm~\ref{alg:AAR} is slightly different to the algorithm~\ref{alg:RR}, 
-which updated matrix $\mathbf{S}$ before making the prediction:
-
 \begin{algorithm}[ht]
@@
 
-\subsection{Efficient computation}
-In regression problems we always require getting a matrix inverse which is
-computationally expensive. However, in stream data problems there is a way to
-obtain a matrix inverse approximation using the Sherman-Morrison-Woodbury. This
-formula allows to get the inverse of a matrix $\mathbf{A+uv^\top}$ if we previously calculated the
-inverse of $\mathbf{A}$ as follows:
-
-\begin{equation}
-\label{eq:SMW}
-(\mathbf{A+uv^\top})^{-1}=\mathbf{A}^{-1}-
-\frac{\mathbf{A^{-1}uv^\top A^{-1}}}{\mathbf{1+v^\top A^{-1}u}}
-\end{equation}
-
-Alternatively, Coleman and
-Sun~\cite{coleman+sun2010} presented an iterative algorithm which
-uses $\mathbf{X}(\lambda)$ to approximate $\mathbf{X}(0)$.
-
-Using the compact SVD (shown in equation~(\ref{eq:compactsvd}))
-$\mathbf{X}(\lambda)$ is expressed as follows:
-
-\begin{eqnarray}
-\label{eq:optsolRRsvd}
-\mathbf{X}(\lambda) & = & (\mathbf{A}^\top \mathbf{A}+ \lambda
-\mathbb{I})^{-1}\mathbf{A}^\top \mathbf{Y} \nonumber \\
-& = &\mathbf{V}_1(\Sigma_1^2+\lambda \mathbb{I})^{-1}\mathbf{\Sigma_1
-U_1^\top Y}
-\end{eqnarray}
-
-\noindent where is easy to see that $\mathbf{X}(\lambda) \rightarrow
-\mathbf{\hat{X}}=\mathbf{V}_1 \mathbf{\Sigma}_1^{-1}\mathbf{U}_1^\top
-\mathbf{Y}$ as $\lambda \rightarrow 0$.
-
-The method consists in obtaining $\mathbf{X}(\lambda)$ and then refine
-by adding more terms of its Taylor expansion to approximate
-$\mathbf{X}(0) = \mathbf{\hat{X}}$. The Taylor expansion about $\lambda_0$ is:
-
-\begin{equation}
-\label{eq:taylor}
-    \mathbf{X}(\lambda)=\mathbf{X}(\lambda_0) + \sum_{k=1}^\infty
-    \mathbf{s}_k(\lambda-\lambda_0)^{k}
-\end{equation}
-
-\noindent where $\mathbf{s}_k=\frac{1}{k!}\mathbf{X}(\lambda)^{(k)}$
-and $\mathbf{X}(\lambda)^{(k)}$ is obtained by taking differences 
-$\frac{\partial}{\partial \lambda}$ of:
-
-\begin{eqnarray*}
-(\mathbf{A}^\top \mathbf{A}+ \lambda\mathbb{I}) \mathbf{X}(\lambda) & = & \mathbf{A}^\top \mathbf{Y}\\
-(\mathbf{A}^\top \mathbf{A}+ \lambda\mathbb{I}) \mathbf{X}(\lambda)^{(1)} + \mathbf{X}(\lambda)& = & 0 \\
-\mathbf{X}(\lambda)^{(1)}  &=& -(\mathbf{A}^\top \mathbf{A}+ \lambda\mathbb{I}) ^{-1} \mathbf{X}(\lambda) \\
-\mathbf{X}(\lambda)^{(2)}  &=& -2(\mathbf{A}^\top \mathbf{A}+ \lambda\mathbb{I}) ^{-1} \mathbf{X}(\lambda)^{(1)} \\
-& \vdots & \\
-\mathbf{X}(\lambda)^{(k)}  &=& -k(\mathbf{A}^\top \mathbf{A}+ \lambda\mathbb{I}) ^{-1} \mathbf{X}(\lambda)^{(k-1)} 
-\end{eqnarray*}
-
-\noindent for $\lambda=0$ we have:
-
-\begin{equation}
-\label{eq:taylor}
-    \mathbf{X}(0)=\mathbf{X}(\lambda_0) + \sum_{k=1}^\infty
-     (-1)^k \mathbf{s}_k \lambda_0^k
-\end{equation}
-
-\noindent therefore in order to ensure convergence, we can see that$\lambda_0$
-selection cannot be large.
-
-The algorithm for computing $\mathbf{X}(0)$ is the following:
-
-\begin{algorithm}[H]
-\begin{algorithmic}[1]
-\REQUIRE $\,$ \\
-$\mathbf{A}$: design matrix \\
-$\mathbf{Y}$: response matrix \\
-$\lambda$: rank deficient parameter \\
-\ENSURE  $\,$ \\
-$\mathbf{X}$: parameters \\
-\STATE $\mathbf{M}=\mathbf{A^\top A}$ \\
-\STATE Initialize $\mathbf{Q R}=\mathbf{M}$ \\
-\STATE $\mathbf{X} = \mathbf{R^{-1}Q^\top Y}$ \\
-\STATE $\mathbf{s} = \mathbf{X}$ \\
-\FOR { $i = 1,2,3,\dots$ }
-	\STATE $\mathbf{s} =
-        -(\mathbf{M}+\lambda\mathbb{I})^{-1}\mathbf{s}$\\
-	\STATE $\mathbf{X}=\mathbf{X} + (-1)^i {\color{red}\mathbf{s}
-        \lambda^i}$
-\ENDFOR
-\end{algorithmic}
-\caption{Algorithm for handling rank deficient matrices}
-\label{alg:coleman}
-\end{algorithm}
-
-The algorithm~\ref{alg:coleman} solves equation~(\ref{eq:taylor}). However, this
-version is unstable since typically $\|\mathbf{s}\|$ is very large and
-$\lambda^i$ is very small ($\lambda$ is small).
-
-The following algorithm shows a more stable version of
-algorithm~\ref{alg:coleman}.
-
-\begin{algorithm}[H]
-\begin{algorithmic}[1]
-\REQUIRE $\,$ \\
-$\mathbf{A}$: design matrix \\
-$\mathbf{Y}$: response matrix \\
-$\lambda$: rank deficient parameter \\
-\ENSURE  $\,$ \\
-$\mathbf{X}$: parameters \\
-\STATE $\mathbf{M}=\mathbf{A^\top A}$ \\
-\STATE Initialize $\mathbf{Q R}=\mathbf{M}$ \\
-\STATE $\mathbf{X} = \mathbf{R^{-1}Q^\top Y}$ \\
-\STATE $\mathbf{t} = \mathbf{X}$ \\
-\FOR { $i = 1,2,3,\dots$ }
-        \STATE $\mathbf{t} =\lambda \mathbf{t}$  
-        \STATE $\mathbf{t} =  -(\mathbf{M}+\lambda\mathbb{I})^{-1}\mathbf{t}$
-	\STATE $\mathbf{X}=\mathbf{X} + \mathbf{t}$
-\ENDFOR
-\end{algorithmic}
-\caption{Algorithm for handling rank deficient matrices improved}
-\label{alg:colemanimproved}
-\end{algorithm}
-
-Both algorithms are equivalent, but algorithm~\ref{alg:colemanimproved} is more
-stable and converges in typically less than 10 steps.
-It is important to notice that the QR factorisation is computed only once and it
-is computationally less expensive than the SVD.
+If we define $\displaystyle\mathbf{S}= \sum_{t=1}^{m+1} \mathbf{a}_t
+\mathbf{a}_t  ^\intercal + \gamma \mathbb{I} $ and $\mathbf{W}=
+\displaystyle\sum_{t=1}^m \mathbf{a}_t \mathbf{y}_t$, the
+algorithm~\ref{alg:AAR} is slightly different to the algorithm~\ref{alg:RR}, 
+which updated matrix $\mathbf{S}$ before making the prediction.
+
+\subsection{Competitive analysis}
+
+Competitive analysis was designed for analysing the performance of an online algorithm compared with its optimal offline algorithm.
+An optimal offline algorithm can view the sequences of requests in advance.
+The effectiveness of an online algorithm \cite{sleator1985} may be measured by its competitive ratio which is the worst-case ratio of the online algorithm and the optimal offline algorithm.
 
@@
 %\end{enumerate}
+\newpage
 
@@
 \cite{theil1966}, is a unit free measure obtained as the ratio between the root
-MSE (RMSE) of a model and the RMSE of the naive random walk model. A
-$U$-statistic less than 1 implies the performance is better than the naive
-model.
+MSE (RMSE) of a model and the RMSE of the naive random walk model. 
 \end{description}

### 2017-10-27 src/3_TimeSeries.tex
first draft ready to print

@@
 
-\textbf{Proof}\quad
-
-Since $\mathbf{A}$ is singular the problem shown in
-equation~(\ref{eq:regressionproblem}) has not solution, the minimum norm given
-by equation~(\ref{eq:MP}) is obtained by solving the equivalent problem:
-
-\begin{equation*}
-\label{eq:proyectorsol}
-\mathbf{A \hat{X} = PB} 
-\end{equation*}
-
-\noindent where $\mathbf{P=U_1 U_1^\top}$ is the projection onto the
-Col($\mathbf{A}$).
-
-Since $\mathbf{V} = [\underset{(n \times k)}{\mathbf{V_1}} |
-\underset{(n \times k)}{\mathbf{V_2}}]$ and $\mathbf{V_1^\top V_2 =
-0}$ we can express $\mathbf{\hat{X}} = \mathbf{V_1 x_1 + V_2 x_2}$
-with $\mathbf{x_2=0}$ because $\mathbf{\hat{X}}$ lives in the
-$\text{Row}(\mathbf{A})$ given by $\mathbf{V_1}$, so we have:
-
-\begin{eqnarray*}
-\mathbf{A \hat{X}} &=& \mathbf{PB} \\
-\mathbf{U_1 \Sigma_1 V_1^\top \hat{X}} &=& \mathbf{U_1 U_1^\top B} \\
-\mathbf{ V_1^\top \hat{X}} &=&  \mathbf{\Sigma_1^{-1} U_1^\top B} \\ 
-\mathbf{ V_1^\top V_1 x_1} &=& \mathbf{\Sigma_1^{-1}
-U_1^\top B} \\
-\mathbf{x_1}&=& \mathbf{\Sigma_1^{-1} U_1^\top B}
-\end{eqnarray*}
-
-\noindent from this result we can obtain $\mathbf{\hat{X}}$ and
-therefore the pseudo-inverse expression:
-
-\begin{eqnarray*}
-\mathbf{\hat{X}} &=& \mathbf{V_1 x_1} \\
-                &=& \mathbf{V_1 \Sigma_1^{-1} U_1^\top B} \\
-\mathbf{A^+} &=& \mathbf{V_1 \Sigma_1^{-1} U_1^\top} \, .
-\end{eqnarray*}
-
-$\blacksquare$
-
-\newpage
+More details can be found in section \ref{app:pscsvd}.

### 2017-11-12 slides/defense-slides.tex
First phd defense slide template

%\documentclass[xcolor=dvipsnames]{beamer}
\documentclass[compress,red]{beamer}
\mode<presentation>
%\usepackage[spanish,activeacute]{babel}
\usepackage[utf8]{inputenc}
\usepackage{listings}
\usepackage{amsmath,amsfonts,amssymb}
\usepackage{adjustbox}
\usepackage{color}
\usepackage{tikz}
\usetikzlibrary{shapes,arrows}
\usepackage{multicol}
\usepackage{hhline}
\usepackage{pgf}
\usepackage{algorithm} 
\usepackage{algorithmic}
\renewcommand{\algorithmicrequire}{\textbf{Input:}}
\renewcommand{\algorithmicensure}{\textbf{Output:}}
\def\R{{\mathbb R }}
\def\N{{\mathbb N }}
\usetheme{Warsaw}
% other themes: AnnArbor, Antibes, Bergen, Berkeley, Berlin, Boadilla, boxes, CambridgeUS, Copenhagen, Darmstadt, default, Dresden, Frankfurt, Goettingen,
% Hannover, Ilmenau, JuanLesPins, Luebeck, Madrid, Maloe, Marburg, Montpellier, PaloAlto, Pittsburg, Rochester, Singapore, Szeged, classic
%\usecolortheme{lily}
% color themes: albatross, beaver, beetle, crane, default, dolphin, dov, fly, lily, orchid, rose, seagull, seahorse, sidebartab, structure, whale, wolverine

%define your own colours:
\definecolor{Red}{rgb}{1,0,0}
\definecolor{Blue}{rgb}{0,0,1}
\definecolor{Green}{rgb}{0,1,0}
\definecolor{magenta}{rgb}{1,0,.6}
\definecolor{lightblue}{rgb}{0,.5,1}
\definecolor{lightpurple}{rgb}{.6,.4,1}
\definecolor{gold}{rgb}{.6,.5,0}
\definecolor{orange}{rgb}{1,0.4,0}
\definecolor{hotpink}{rgb}{1,0,0.5}
\definecolor{newcolor2}{rgb}{.5,.3,.5}
\definecolor{newcolor}{rgb}{0,.3,1}
\definecolor{newcolor3}{rgb}{1,0,.35}
\definecolor{darkgreen1}{rgb}{0, .35, 0}
\definecolor{darkgreen}{rgb}{0, .6, 0}
\definecolor{darkred}{rgb}{.75,0,0}

\xdefinecolor{olive}{cmyk}{0.64,0,0.95,0.4}
\xdefinecolor{purpleish}{cmyk}{0.75,0.75,0,0}

\useoutertheme[subsection=false]{smoothbars}

\setbeamertemplate{blocks}[rounded][shadow=true]

\author[P. Arce et al.]{Paola
Arce\inst{1}\and Jonathan Antognini\inst{1}\and Werner Kristjanpoller
\inst{2}\and Luis Salinas\inst{1,3}}

\institute[UTFSM]{
\inst{1} Departamento de Inform\'atica, UTFSM, Chile \\
\inst{2} Departamento de Industrias, UTFSM, Chile \\
\inst{3} CCTVal, UTFSM, Chile
}

\date[ISCEF 2016]{4th International Symposium in Computational Economics and Finance 2016}
\title[Cointegration model
for forecasting HF financial time series]{\Large Fast and adaptive cointegration based model for forecasting high frequency financial time series}

%\titlegraphic{
%   \includegraphics[width=1cm]{img/logo_DI}\hspace*{2.5cm}~%
%   \includegraphics[width=1.5cm]{img/logo_UTFSM}\hspace*{2.5cm}~%
%   \includegraphics[width=1cm]{img/logo_IND}
%}

\begin{document}
\begin{frame}[plain]
\titlepage
\end{frame}

%\begin{frame}
%\frametitle{Outline}
%\begin{itemize}
%\item Motivation: economic relations
%\item Cointegration and Integration concepts
%\item VECM model
%\item Proposal
%\item Experiments
%\item Conclusions
%\end{itemize}
%\end{frame}

\section{Introduction}
\subsection{Motivation}

\begin{frame}
\frametitle{Motivation}
Most financial time series are non-stationary in nature. This implies that the statistical properties, such as mean and variance of the data change over time.
\end{frame}

\begin{frame}
\hspace*{-4mm}
\includegraphics[width=0.9\paperwidth]{img/motivation.png}
\end{frame}

\begin{frame}
\frametitle{Motivation}
\begin{itemize}
\item Financial time series are mainly non-stationary.  However, one or more linear
combinations of non-stationary time series can be stationary even though
individually they are not. When this happens it is said they have a long-run
equilibrium relationship and the variables are \textbf{cointegrated}.
\item The Vector Error correction model (VECM) is used to model cointegrated time series but is used only with batch data and rarely used with high frequency data.
\end{itemize}
\end{frame}

%\subsection{Proposal}
%\begin{frame}
%\frametitle{Proposal}
%\begin{block}{}
%To use VECM with high frequency data where the best parameters are updated when new data is available and do this is done in a fast way.
%\end{block}
%\end{frame}
%
%\section{Background}
%%
%%\begin{frame}
%%\frametitle{Integration }
%%\begin{block}{Integration of order $d$}
%%A time series $\mathbf{y}$ is said to be integrated of order $d$ if after
%%differentiating the variable $d$ times, we get an stationary process, more precisely:
%%\[
%%(1-L)^d \mathbf{y} \sim \text{I(0)} \, ,
%%\]
%%\noindent where I(0) is a stationary time series and $L$ is the lag operator:
%%\[
%%(1-L)\mathbf{y} = \Delta \mathbf{y}=\mathbf{y}_t  -\mathbf{y}_{t-1} \quad \forall t
%%\]
%%\end{block}
%%\end{frame}
%%
%%\subsection{Integration example}
%%\begin{frame}
%%\frametitle{Financial Time Series}
%%\begin{columns}
%%\column[t]{5cm}
%%High frequency financial time series are commonly an I(1) process 
%%\begin{center}
%%\includegraphics[width=0.85\textwidth]{img/EURUSD}
%%\end{center}
%%\column[t]{5cm}
%%Since their differences are stationary (I(0) process).
%%\begin{center}
%%\includegraphics[width=0.85\textwidth]{img/DEURUSD}
%%\end{center}
%%\end{columns}
%%\end{frame}
%
%\subsection{Cointegration}
%\begin{frame}
%\frametitle{Cointegration}
%\begin{block}{Cointegration definition}
%Let $\mathbf{y} = \{\mathbf{y}^1, \dots, \mathbf{y}^l\}$ be a set of $l$
%time series of order I(1) which are said to be cointegrated if a vector
%$\beta=[\beta(1),\dots,\beta(l)]^\top \in \mathbb{R}^l$  exists such that the
%time series,
%
%\begin{equation*}
% \mathbf{Z}_t:= \beta^\top \mathbf{y} = \beta(1) \mathbf{y}^1 + \dots + \beta(l) \mathbf{y}^l \sim
% \text{I(0)}\, .
%\end{equation*}
%\end{block}
%In other words, a set of I(1) variables is said to be cointegrated if
%a linear combination of them exists which is I(0).
%
%\end{frame}
%
%\subsection{Vector Autorregresive models}
%\begin{frame}
%\frametitle{VECM: Vector Error Correction model}
%
%VECM is a special form of a Vector Autorregresive model (VAR) for I(1)
%variables $\mathbf{y}$ that are also cointegrated. Let 
%
%\begin{equation*}
%\label{eq:variables}
%\mathbf{y}_t = 
%\begin{bmatrix} y_{1,t} & y_{2,t} & \dots & y_{l,t}
%\end{bmatrix}^\top \, \qquad t=p+1, \dots N. 
%\end{equation*}
%
%\begin{block}{VAR}
%VAR($p$) model describes the behaviour of a set of $l$
%endogenous variables as a linear combination of their last $p$ values:
%{\color{blue}
%\begin{equation*}
%\label{eq:var}
% \mathbf{y}_t = \phi_1 \mathbf{y}_{t-1}  + \dots +   \phi_p\mathbf{y}_{t-p}
%+ \mathbf{c} + \mathbf{\epsilon}_t \, ,
%\end{equation*}}
%\noindent where  ${\phi_1,\dots,\phi_p}$ are $l \times l$
%matrices of real coefficients,
%$\mathbf{\epsilon}_{p+1},\dots,\mathbf{\epsilon}_N$ are error terms,
%$\mathbf{c}$ is a constant vector and $N$ is the total number of samples.
%\end{block}
%\end{frame}
%
%
%
%\begin{frame}
%\frametitle{VECM: Vector Error Correction model}
%\begin{block}{VECM}
%
%It is obtained by replacing the form {\color{red} $\Delta
%\mathbf{y}_t = \mathbf{y}_t - \mathbf{y}_{t-1}$} in the VAR form.
%{\color{blue}
%\begin{equation}
% \label{eq:vec}
% \Delta \mathbf{y}_t = 
% \underbrace{ \Omega\mathbf{y}_{t-1}}_\text{Error correction term} + 
% \sum_{i=1}^{p-1}
%\phi_i^* \Delta \mathbf{y}_{t-i}  + \mathbf{c} + \mathbf{\epsilon}_t \quad ,
%\end{equation}}
%\end{block}
%
%\noindent where 
%%coefficients matrices $\Omega$ and $\phi_i^*$ are
%%function of matrices $\phi_i$ as follows:
%
%\begin{eqnarray*}
%\phi_i^* &: =& -\sum_{j=i+1}^{p} \phi_j \\
%\Omega &: =& -(\mathbb{I}-\phi_1-\dots-\phi_p) 
%\end{eqnarray*}
%\end{frame}
%
%\begin{frame}
%\frametitle{$\Omega$ matrix properties}
%The matrix $\Omega$ has the following properties:
%\begin{itemize}
%\item If $\Omega = 0$ there is no cointegration
%\item If $rank(\Omega)=l$ i.e full rank, then the time series are not
%I(1) but stationary
%\item If $rank(\Omega)=r,\quad 0 < r < l$ then, there is cointegration
%and the matrix $\Omega$ can be expressed as $\Omega =
%\alpha \beta^\top$, where $\alpha$ and $\beta$ are $(l \times r)$
%matrices and $rank(\alpha)=rank(\beta)=r$.
%\end{itemize}
%The columns of $\beta$ contains the cointegration vectors and the rows of
%$\alpha$ correspond with the adjusted vectors. $\beta$ is obtained by Johansen
%procedure whereas $\alpha$ has to be determined as a
%variable in the VECM.
%\end{frame}
%
%
%\begin{frame}
%\frametitle{VECM matrix form}
%\small
%\begin{equation*}
%\underbrace{
%      \begin{bmatrix}
%       \mathbf{\Delta y}_{p+1}  \\ 
%       \mathbf{\Delta y}_{p+2}  \\ 
%       \vdots                   \\ 
%       \mathbf{\Delta y}_N      
%      \end{bmatrix}}_{\mathbf{B} } =
%\arraycolsep=1.4pt  
%\underbrace{\begin{bmatrix}
%   \alpha^\top & \phi^*_1 & \dots & \phi^*_{p-1} &\mathbf{c}   
%  \end{bmatrix}}_{\mathbf{X}^\top}
%\underbrace{\begin{bmatrix}
% \beta\mathbf{y}_p      & \beta\mathbf{y}_{p+1}   & \dots&   \beta\mathbf{y}_{N-1}   \\
% \mathbf{\Delta y}_{p}   & \mathbf{\Delta y}_{p+1}& \dots &  \mathbf{\Delta y}_{N-1} \\
% \mathbf{\Delta y}_{p-1} & \mathbf{\Delta y}_{p}  & \dots &  \mathbf{\Delta y}_{N-2}   \\
% \vdots                  & \vdots                 & \ddots&  \vdots                   \\
% \mathbf{\Delta y}_{2}   & \mathbf{\Delta y}_{3} & \dots &   \mathbf{\Delta y}_{N-p+1} \\
% 1                      & 1                       & \dots     & 1   
% \end{bmatrix}}_{\mathbf{A}^\top}
%+
%\underbrace{\begin{bmatrix}
%              \mathbf{\epsilon}_{p+1} \\ 
%              \mathbf{\epsilon}_{p+2} \\ 
%              \vdots \\ 
%              \mathbf{\epsilon}_N
%             \end{bmatrix}}_{\mathbf{E}^\top}
%\end{equation*}
%\end{frame}
%
%\section{Proposal}
%\subsection{Motivation}
%\begin{frame}
%\frametitle{VECM difficulties in an online context}
%\begin{itemize}
%\item VECM assumes cointegration vectors do not change in time. However, these might change due to several economic factors.
%\item To get updated cointegration vectors in time is computationally expensive because 
%the Johansen method and OLS are used to obtain them. It would be good to know how the number of cointegration vectors are affected by the number of data used $L$ and the number of lags of VECM $p$.
%\end{itemize}
%\end{frame}
%
%
%\begin{frame}
%\frametitle{Data}
%\begin{itemize}
%\item Tests were carried out using four foreign exchange rates all related to
%USD: EURUSD, GBPUSD, USDCHF and USDJPY. 
%\item This data was collected from Dukascopy, a free
%database which gives access to the Swiss Foreign Exchange marketplace.
%\item The tests were done using 10-seconds frequency from ask prices
%from the 11th to the 15th of August 2014. 
%%\item Since one day corresponds to 8640 data points and we used 5 days of data, we have 43,200 data points in total.
%\end{itemize}
%\end{frame}
%
%
%\begin{frame}
%\frametitle{Cointegration vectors distribution}
%\begin{figure}[!h]
%  %\vspace{-0.8cm}
%  \centering
%  \includegraphics[width=0.75\textwidth]{img/histCointVectors-offset20520-p-1-freq-10s}
%  \caption{Histogram of the number of cointegration vectors using $p=1$. Four
%  possible values for $L$ are shown (100, 2000, 3000 and 4000).}
%  \label{fig:hists}
%\end{figure}
%\end{frame}
%
%\begin{frame}
%\frametitle{Percentage of cointegration}
%Since $r=0$ means no cointegration and $r=l=4$ reveals that no process is I(1) but stationary.
%The interesting cases of cointegration are those where $r$ lies strictly
%between $0$ and $l$, i.e. $0<r<l$.
%\begin{block}{Percentage of cointegration (PC)}
%{\color{blue}
%\begin{equation*} \label{eq:pcoint}
%PC = 
%\frac{\#\{ it \mid \text{$it$ has $r$ c.v. with $0<r<l$}\}}
%     {\#it}\times 100
%\end{equation*}}
%\end{block}
%\end{frame}
%
%
%\begin{frame}
%\frametitle{PC vs MSE}
%\begin{figure}[ht!]
%  %\vspace{-0.8cm}
%  \centering
%  \includegraphics[width=0.7\textwidth]{img/MSE-offset20520-p-2-freq-10s}
%  \caption{MSE versus the percentage of cointegration considering 1000
%  iterations. }
%  \label{fig:cointvsmse}
%\end{figure}
%\end{frame}
%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%\subsection{AVECM}
%\begin{frame}
%\frametitle{Proposal: Adaptive VECM algorithm (AVECM)}
%\begin{itemize}
%\item We proposed to choose $L$ and $p$ in order to maximise the percentage of
%cointegration $PC$ in the near past. This process was done every time that new data was processed. 
%\item To use a distributed environment to make this parameters search, since this is the most expensive routine.
%\end{itemize}
%\end{frame}
%
%
%%\begin{frame}
%%\frametitle{AVECM Algorithm}
%%\begin{algorithmic}[1]
%%\REQUIRE $\,$ \\
%%$\mathbf{y}$: matrix with $N$ input vectors and $l$ time series\\
%%$j$: Starting point of testing \\
%%$it$: Ending point of testing \\
%%$ps$: list of $p$ values \\
%%$Ls$: list of $L$ values ($L<N$) \\
%%$m$: Iterations to determine parameters ($m < N-L$)\\
%%\ENSURE  $\,$ \\
%%$\{ \hat{\mathbf{y}}[1],\dots,\hat{\mathbf{y}}[it]\}$: prediction vectors \\
%%\FOR { $i =j$ to $it$ }
%%   \STATE $\mathbf{Y} \gets \mathbf{y}[:,i-1]$
%%    \STATE $L,p \gets
%%    \texttt{get\_best\_params}(Ls,ps,m,\mathbf{Y})$
%%        \STATE $model = \text{VECM}(\mathbf{Y},L, p)$
%%        \STATE $\hat{\mathbf{y}}[i-j] = model.predict()$
%%\ENDFOR
%%\end{algorithmic}
%%\end{frame}
%
%\subsection{Model comparison} 
%\begin{frame}
%\frametitle{Random walk}
%The random walk model is defined as:
%\begin{equation}
%\mathbf{y}_t = \mathbf{y}_{t-1} + \epsilon_{t}
%\label{rwmodel}
%\end{equation}
%The naive forecast of the time series difference $\hat{\mathbf{y}}_{t+1}$ for the random walk model is defined as:
%\begin{equation}
%\hat{\mathbf{y}}_{t+1} = \mathbf{y}_t + \hat{\epsilon}_{t+1} 
%\end{equation}
%\noindent where  $\hat{\epsilon}_{t+1} = \epsilon_{t}$.
%\end{frame}
%
%
%\begin{frame}
%\frametitle{ARIMA}
%A process can be modelled as an ARIMA$(p,d,q)$ if $\mathbf{x}_t = \Delta^d \mathbf{y}_t $, is an ARMA$(p,q)$. An ARMA$(p,q)$ model is the following:
%\begin{equation}
%\mathbf{x}_t = \sum_{i=1}^p \phi_i \mathbf{x}_{t-i}  +  \sum_{j=1}^q \theta_j \epsilon_{t-j}  
%\end{equation}
%\noindent with coefficients $\phi_p \neq 0$, $\theta_q \neq 0$ and $\sigma_{\epsilon}^2 > 0$.
%\end{frame}
%
%\begin{frame}
%\frametitle{Results}
%\begin{tabular}{ccccccc}
%
%& \multicolumn{3}{l}{MSE} & &
%\multicolumn{2}{c}{$U$-statistic} \\ 
%\hhline{~---~--} \\
% &
%AVECM & ARIMA & p-value & &
%AVECM & ARIMA \\ 
%
%\hline
% EURUSD & 1.07 e-09 & 1.14 e-09 &  9.25 e-12 & & 0.6863 & 0.7108\\
% GBPUSD & 1.66 e-09 & 1.74 e-09 &  6.95 e-02 & & 0.6866 & 0.7025\\
% USDCHF & 5.85 e-10 & 6.35 e-10 &  2.89 e-14 & & 0.6803 & 0.7091\\
% USDJPY  & 6.34 e-06 & 6.51 e-06 &  6.85 e-05 & & 0.6964 & 0.7057\\
% \end{tabular}
%\end{frame}
%
%%\begin{frame}
%%\frametitle{Evaluation methods}
%%\begin{block}{MSE}
%%Measures the distance between forecasts
%%and the true values and large deviations from the true value have a
%%large impact due to squaring forecast error.
%%{\color{blue}
%%\begin{equation}\label{eq:MSE}
%%\text{MSE} = 
%%\frac{\displaystyle \sum_{t=1}^{N} (\mathbf{y}_t-\hat{\mathbf{y}}_t)^2}{N}
%%\end{equation}}
%%\end{block}
%%\begin{block}{\bf $U$-statistic}
%%Unit free measure obtained as the ratio between the root MSE (RMSE) of a model and the RMSE of the naive random walk model. 
%%\end{block}
%%\end{frame}
%
%\section{Experiments}
%\subsection{Uno}
%
%%\begin{frame}
%%\frametitle{Unit root tests}
%%VECM can only be used if the time series are I(1). Therefore, 
%%before running the tests, we firstly checked whether the time series were
%%I(1) using the Augmented Dickey Fuller (ADF) test with lags $p=1,2,3,4,5$.
%%\end{frame}
%%
%%\begin{frame}
%%\frametitle{Performance accuracy}
%%\small
%%\begin{tabular}{ccccccc}
%%& \multicolumn{3}{l}{MSE} & &
%%\multicolumn{2}{c}{$U$-statistic} \\ 
%%\hhline{~---~--} \\
%% & AVECM & ARIMA & p-value & &
%%AVECM & ARIMA \\ 
%%
%%\hline
%% EURUSD & 1.0702 e-09 & 1.1481 e-09 &  9.250 e-12 & & 0.6863 & 0.7108\\
%% GBPUSD & 1.6630 e-09 & 1.7408 e-09 &  6.951 e-02 & & 0.6866 & 0.7025\\
%% USDCHF & 5.8503 e-10 & 6.3545 e-10 &  2.899 e-14 & & 0.6803 & 0.7091\\
%% USDJPY  & 6.3483 e-06 & 6.5194 e-06 &  6.853 e-05 & & 0.6964 & 0.7057
%% \end{tabular}
%% \end{frame}
%%For AVECM we considered different number of iterations (parameter $m$ in algorithm \ref{alg:AVECM}): 10, 50 and 100. We tried 12, 24 and 47 different pair of combinations for $L$ and $p$. Possible values for $L$ were in the interval $[100,4000]$ and $p$ can have values in between $[1,5]$. Best AVECM performance was compared against ARIMA and the random walk model.
%%Table~\ref{tab:stats} shows the out-of-sample performance measures: MSE and $U$-statistic for AVECM and ARIMA. In terms of both measures we found that AVECM is superior to ARIMA and the naive random walk model. We also included the p-value that proves that the difference in the MSE is significant at the 99\% significance level in three of the four currency rates and at 90\% in the case of GBPUSD. The $U$-statistic shows that AVECM and ARIMA are superior to the naive random walk model and that our proposal is also superior to ARIMA.
%
%
%
%\begin{frame}
%\frametitle{Parallel implementation}
%The function that get best parameters $L$ and $p$ was implemented using MPI in Python. MPI allows large-scale parallel applications
%with wide portability to be built, being able to run in large clusters
%or on local computers.
%We tested our proposal in a cluster with 2 servers Xeon E5-2667 (2.90GHz)
%of 24 cores each (48 cores in total) and 24GB RAM.
%\end{frame}
%
%\begin{frame}
%\frametitle{Parameters setting}
%\begin{itemize}
%%\item Parameter $it=100$
%\item The $L$ parameter was always chosen between 100 and 4000 and $p$ always took values
%between 1 and 5. 
%\item Parameter $nparams$ represents the number of pairs ($L$,$p$) used to maximise the percentage of cointegration. 
%\end{itemize}
%\end{frame}
%
%\begin{frame}
%\frametitle{Execution times}
%\begin{figure}[ht]
%  \centering
%  \includegraphics[width=0.75\textwidth]{img/extimes}
%  \caption{Computing time of sequential and parallel algorithm is shown in the
%  upper figure. Speed-up is shown below.}
%  \label{fig:extimes}
%% The superior figure \ref{fig:extimes} shows that best performance accuracy measurements are
%%achieved in times near or below 10 seconds in the parallel version. Contrarily,
%%serial times are high, above 10 seconds in most cases. Figure~\ref{fig:extimes} also shows the speed-up in computing time for AVECM which is near 9X if we use more than 25 parameters ($nparams$).
%%Execution times do not consider the loading data time, just that of finding
%%best parameters and matrix operations. The time MPI spends
%%transferring data and synchronising processes is about two seconds independently
%%of the number of processes considered. Execution times were measured using the Time python library.
%\end{figure}
%
%\end{frame}
%
%\section{Conclusions}
%\subsection{Uno}
%\begin{frame}
%\frametitle{Conclusions}
%\begin{itemize}
%\item Cointegration relations change with time. We empirically showed that the Johansen method is sensitive to the number of lags but also to the amount of data considered.
%\item We found that out-of-sample forecast performance MSE is related to the {\em percentage of cointegration\/}.  
%\item VECM improves performance measures by finding
%parameters of $L$ and $p$ maximising the percentage of cointegration.
%\item Increasing the number of parameters will always lead to
%better performance measures.
%\end{itemize}
%\end{frame}
%
%\begin{frame}
%\frametitle{Conclusions}
%\begin{itemize}
%\item The parallel implementation allowed the execution times to be reduced
%more than 9 times and therefore a response time was obtain before 10
%seconds. 
%\item For future study, it would be interesting to explore the relationship between
%cointegration and performance in order to propose new criteria for
%improving VECM parameters. It would also be interesting to include more 
%explaining variables such as bid-ask spread and change in volume.
%\end{itemize}
%\end{frame}

\begin{frame}[plain,c]
\begin{center}
\Huge Thank you for your attention
\Huge Questions?
\end{center}
\end{frame}
%Have a slide for each of the following
%9 Problem statement or Hypothesis
%9 “Why is this a Hard Problem?”
%9 Approach or Methodology
%9 “Why is this Innovative?”
%9 Assumptions and Constraints
%9 Initial Results (Promise of Great Things to Come)
%9 Validation Plan
%9 Limitations and Applicability
%9 Expected Contributions
%9 “What’s beyond this thesis?”
%9 Roadmap of Thesis

% Financial time series are not the only ones with this behaviour, this study can be extended to other non-stationary time series such as: weather, earthquakes, energy demand and sales forecasting.
\end{document}

### 2017-11-12 slides/defense-slides.tex
New template and motivation slides

@@
-%\documentclass[xcolor=dvipsnames]{beamer}
-\documentclass[compress,red]{beamer}
-\mode<presentation>
+\documentclass{beamer}
+\usetheme{metropolis}
 %\usepackage[spanish,activeacute]{babel}
@@
 \def\N{{\mathbb N }}
-\usetheme{Warsaw}
-% other themes: AnnArbor, Antibes, Bergen, Berkeley, Berlin, Boadilla, boxes, CambridgeUS, Copenhagen, Darmstadt, default, Dresden, Frankfurt, Goettingen,
-% Hannover, Ilmenau, JuanLesPins, Luebeck, Madrid, Maloe, Marburg, Montpellier, PaloAlto, Pittsburg, Rochester, Singapore, Szeged, classic
-%\usecolortheme{lily}
-% color themes: albatross, beaver, beetle, crane, default, dolphin, dov, fly, lily, orchid, rose, seagull, seahorse, sidebartab, structure, whale, wolverine
 
@@
 
-\useoutertheme[subsection=false]{smoothbars}
+\author{Paola Arce \\ Departamento de Inform\'atica, UTFSM \\ Thesis advisor: Luis Salinas}
+\date{December 15, 2017}
+\title{Online learning methods for financial time series forecasting}
+\subtitle{PhD Thesis Defense}
 
-\setbeamertemplate{blocks}[rounded][shadow=true]
-
-\author[P. Arce et al.]{Paola
-Arce\inst{1}\and Jonathan Antognini\inst{1}\and Werner Kristjanpoller
-\inst{2}\and Luis Salinas\inst{1,3}}
-
-\institute[UTFSM]{
-\inst{1} Departamento de Inform\'atica, UTFSM, Chile \\
-\inst{2} Departamento de Industrias, UTFSM, Chile \\
-\inst{3} CCTVal, UTFSM, Chile
+\titlegraphic{
+    \tikz[overlay,remember picture]
+        \node[at=(current page.south east), anchor=south east] {
+            \includegraphics[height=1.5cm]{img/DIT}
+        };
 }
-
-\date[ISCEF 2016]{4th International Symposium in Computational Economics and Finance 2016}
-\title[Cointegration model
-for forecasting HF financial time series]{\Large Fast and adaptive cointegration based model for forecasting high frequency financial time series}
-
-%\titlegraphic{
-%   \includegraphics[width=1cm]{img/logo_DI}\hspace*{2.5cm}~%
-%   \includegraphics[width=1.5cm]{img/logo_UTFSM}\hspace*{2.5cm}~%
-%   \includegraphics[width=1cm]{img/logo_IND}
-%}
 
@@
 \section{Introduction}
-\subsection{Motivation}
 
@@
 \frametitle{Motivation}
-Most financial time series are non-stationary in nature. This implies that the statistical properties, such as mean and variance of the data change over time.
+Most financial time series are non-stationary in nature.
+This implies that the statistical properties, such as mean and variance of the data change over time.
+Therefore, financial time series forecasting is still a challenge.
+\begin{figure}
+\includegraphics[width=0.4\paperwidth]{img/forecast}
+\end{figure}
+
 \end{frame}

### 2017-11-12 slides/defense-slides.tex
new theme

@@
 \documentclass{beamer}
-\usetheme{metropolis}
+\usetheme{Warsaw}
+%\usetheme{metropolis}
 %\usepackage[spanish,activeacute]{babel}
@@
 
-%define your own colours:
-\definecolor{Red}{rgb}{1,0,0}
-\definecolor{Blue}{rgb}{0,0,1}
-\definecolor{Green}{rgb}{0,1,0}
-\definecolor{magenta}{rgb}{1,0,.6}
-\definecolor{lightblue}{rgb}{0,.5,1}
-\definecolor{lightpurple}{rgb}{.6,.4,1}
-\definecolor{gold}{rgb}{.6,.5,0}
-\definecolor{orange}{rgb}{1,0.4,0}
-\definecolor{hotpink}{rgb}{1,0,0.5}
-\definecolor{newcolor2}{rgb}{.5,.3,.5}
-\definecolor{newcolor}{rgb}{0,.3,1}
-\definecolor{newcolor3}{rgb}{1,0,.35}
-\definecolor{darkgreen1}{rgb}{0, .35, 0}
-\definecolor{darkgreen}{rgb}{0, .6, 0}
-\definecolor{darkred}{rgb}{.75,0,0}
+   \definecolor{myblue1}{RGB}{35,119,189}
+    \definecolor{myblue2}{RGB}{95,179,238}
+    \definecolor{myblue3}{RGB}{129,168,207}
+    \definecolor{myblue4}{RGB}{26,89,142}
 
-\xdefinecolor{olive}{cmyk}{0.64,0,0.95,0.4}
-\xdefinecolor{purpleish}{cmyk}{0.75,0.75,0,0}
+    \setbeamercolor*{structure}{fg=myblue1,bg=blue}
+    \setbeamercolor*{palette primary}{use=structure,fg=white,bg=structure.fg}
+    \setbeamercolor*{palette secondary}{use=structure,fg=white,bg=structure.fg!75!black}
+    \setbeamercolor*{palette tertiary}{use=structure,fg=white,bg=structure.fg!50!black}
+    \setbeamercolor*{palette quaternary}{fg=black,bg=white}
 
-\author{Paola Arce \\ Departamento de Inform\'atica, UTFSM \\ Thesis advisor: Luis Salinas}
+    \setbeamercolor*{item projected}{fg=red,bg=myblue3!80}
+    \setbeamercolor*{block title example}{fg=white,bg=myblue4}
+    \setbeamercolor*{frametitle}{fg=black}
+
+    \setbeamertemplate{blocks}[rounded][shadow=true]
+
+    \makeatletter
+    \pgfdeclarehorizontalshading[frametitle.bg,frametitle right.bg]{beamer@frametitleshade}{\paperheight}{%
+      color(0pt)=(myblue2);
+      color(\paperwidth)=(white)}
+
+    \defbeamertemplate*{footline}{mysplit theme}
+    {%
+      \leavevmode%
+      \hbox{\begin{beamercolorbox}[wd=.5\paperwidth,ht=2.5ex,dp=1.125ex,leftskip=.3cm plus1fill,rightskip=.3cm]{author in head/foot}%
+        \usebeamerfont{author in head/foot}\insertshortauthor
+      \end{beamercolorbox}%
+      \begin{beamercolorbox}[wd=.5\paperwidth,ht=2.5ex,dp=1.125ex,leftskip=.3cm,rightskip=.3cm plus1fil]{title in head/foot}%
+        \usebeamerfont{title in head/foot}\insertshorttitle\hfill
+        \insertframenumber/\inserttotalframenumber\hspace*{0.5em}
+      \end{beamercolorbox}}%
+      \vskip0pt%
+    }
+    \makeatother
+
+\author[Paola Arce]{Paola Arce \\ Departamento de Inform\'atica, UTFSM \\ Thesis advisor: Luis Salinas}
 \date{December 15, 2017}
-\title{Online learning methods for financial time series forecasting}
+\title[Thesis defense]{Online learning methods for financial time series forecasting}
 \subtitle{PhD Thesis Defense}

### 2017-11-18 slides/defense-slides.tex
Drunk and the dog example

@@
     \tikz[overlay,remember picture]
-        \node[at=(current page.south east), anchor=south east] {
+        \node[at=(current page.south), anchor=south] {
             \includegraphics[height=1.5cm]{img/DIT}
@@
 
-%\begin{frame}
-%\frametitle{Outline}
-%\begin{itemize}
-%\item Motivation: economic relations
-%\item Cointegration and Integration concepts
-%\item VECM model
-%\item Proposal
-%\item Experiments
-%\item Conclusions
-%\end{itemize}
-%\end{frame}
+\begin{frame}
+\frametitle{Outline}
+\begin{itemize}
+\item Motivation: economic relations
+\item Cointegration and Integration concepts
+\item VECM model
+\item Proposal
+\item Experiments
+\item Conclusions
+\end{itemize}
+\end{frame}
 
@@
 
+\begin{frame}[t,fragile]{Random walk and Cointegration Illustration}
+ \begin{itemize}
+  \item<1-> The drunkard's walk
+  \item<2-> The dog's walk
+  %\item<3-> The drunk and dog's walk
+ \end{itemize}
+
+\begin{columns}
+\begin{column}{.4\linewidth}<1->
+\begin{center}
+        \begin{tikzpicture}[overlay]
+            \node[yshift=-50] (img2) {\includegraphics[scale=1.4]{img/drunk}};
+        \end{tikzpicture}
+\end{center}
+\end{column}
+\begin{column}{.08\linewidth}<2->
+        \begin{tikzpicture}[overlay]
+            \node[yshift=-50] (img2) {\includegraphics[scale=0.06]{img/dog}};
+        \end{tikzpicture}
+\end{column}
+%\begin{column}{.5\linewidth}<3->
+%    \begin{center}
+%        \begin{tikzpicture}[overlay]
+%            \node[yshift=-50] (img2) {\includegraphics[scale=0.1]{img/mananddog}};
+%        \end{tikzpicture}
+%    \end{center}
+%\end{column}
+\end{columns}
+\end{frame}
+
+\begin{frame}[t,fragile]{Random walk and Cointegration Illustration}
+ \begin{itemize}
+  \item The drunk and dog's walk
+ \end{itemize}
+    \begin{center}
+        \begin{tikzpicture}[overlay]
+            \node[yshift=-60] (img2) {\includegraphics[scale=0.1]{img/mananddog}};
+        \end{tikzpicture}
+    \end{center}
+\end{frame}
+
 \begin{frame}
 \frametitle{Motivation}
-\begin{itemize}
-\item Financial time series are mainly non-stationary.  However, one or more linear
-combinations of non-stationary time series can be stationary even though
-individually they are not. When this happens it is said they have a long-run
-equilibrium relationship and the variables are \textbf{cointegrated}.
-\item The Vector Error correction model (VECM) is used to model cointegrated time series but is used only with batch data and rarely used with high frequency data.
-\end{itemize}
+
+The relationship between non-stationary time series can be modelled if exists a linear combination of them that turns out to be stationary.
+%Financial time series are mainly non-stationary.  However, one or more linear
+%combinations of non-stationary time series can be stationary even though
+%individually they are not. 
+When this happens it is said they have a long-run
+equilibrium relationship and the variables are {\color{red} cointegrated}.
+\
+% The Vector Error correction model (VECM) is used to model cointegrated time series but is used only with batch data and rarely used with high frequency data.
+
 \end{frame}

### 2017-11-19 slides/defense-slides.tex
SATURDAY

@@
 
-The relationship between non-stationary time series can be modelled if exists a linear combination of them that turns out to be stationary.
+The relationship between non-stationary time series can be modelled if a stationary linear combination is shown to exist.
 %Financial time series are mainly non-stationary.  However, one or more linear
@@
 \
-% The Vector Error correction model (VECM) is used to model cointegrated time series but is used only with batch data and rarely used with high frequency data.
 
@@
 
-%\subsection{Proposal}
-%\begin{frame}
-%\frametitle{Proposal}
-%\begin{block}{}
-%To use VECM with high frequency data where the best parameters are updated when new data is available and do this is done in a fast way.
-%\end{block}
-%\end{frame}
-%
+\tikzstyle{every picture}+=[remember picture]
+\everymath{\displaystyle}
+\begin{frame}
+\frametitle{The Vector Error correction}
+\tikzstyle{na} = [baseline=-.5ex]
+
+The Vector Error correction model (VECM) is used to model cointegrated time series.
+
+\Large
+\begin{equation*}
+ \Delta \mathbf{y}_t = \underbrace{
+        \tikz[baseline]{
+            \node[fill=red!20,ellipse,anchor=base] (t1)
+            {$ \Omega\mathbf{y}_{t-1}$};
+        }}_\text{Error correction term} +
+        \underbrace{
+        \tikz[baseline]{
+            \node[fill=blue!20,anchor=base] (t2)
+            {$\sum_{i=1}^{p-1} \phi_i^* \Delta \mathbf{y}_{t-i} $};
+        }}_\text{Autorregresive term} +
+        \underbrace{
+        \tikz[baseline]{
+            \node[fill=green!20,anchor=base] (t3)
+            {$c$};
+        }}_\text{Constant}
+        +
+        \underbrace{
+        \tikz[baseline]{
+            \node[fill=yellow!20,anchor=base] (t3)
+            {$\mathbf{\epsilon}_t $};
+        }}_\text{i.i.e}
+\end{equation*}
+
+\end{frame}
+
+\begin{frame}
+The Vector Error correction model (VECM) is used to model cointegrated time series but is used only with batch data and rarely used with high frequency data.
+\end{frame}
+
+\subsection{Proposal}
+\begin{frame}
+\frametitle{Proposal}
+\begin{block}{}
+To use VECM with high frequency data where the best parameters are updated when new data is available and do this is done in a fast way.
+\end{block}
+\end{frame}
+
 %\section{Background}
@@
 %
-%\begin{frame}
-%\frametitle{VECM: Vector Error Correction model}
-%\begin{block}{VECM}
-%
-%It is obtained by replacing the form {\color{red} $\Delta
-%\mathbf{y}_t = \mathbf{y}_t - \mathbf{y}_{t-1}$} in the VAR form.
-%{\color{blue}
-%\begin{equation}
-% \label{eq:vec}
-% \Delta \mathbf{y}_t = 
-% \underbrace{ \Omega\mathbf{y}_{t-1}}_\text{Error correction term} + 
-% \sum_{i=1}^{p-1}
-%\phi_i^* \Delta \mathbf{y}_{t-i}  + \mathbf{c} + \mathbf{\epsilon}_t \quad ,
-%\end{equation}}
-%\end{block}
+
 %

### 2017-11-19 slides/defense-slides.tex
Motivation nearly finished

@@
 \begin{frame}
-\hspace*{-4mm}
-\includegraphics[width=0.9\paperwidth]{img/motivation.png}
+\frametitle{Stationarity}
+\begin{block}{Strong definition}
+
+The hole distribution of the variable does not depend on time, more precisely:
+\small
+\begin{equation*}
+P\{y_{t_1} \leq
+c_1,..,y_{t_L} \leq c_L\} = P\{y_{t_1+h} \leq c_1,..,y_{t_L+h} \leq c_L\}\enspace\forall L \in \mathbb{N}, \forall h \in \mathbb{Z}
+\end{equation*}
+\end{block}
 \end{frame}
 
-\begin{frame}[t,fragile]{Random walk and Cointegration Illustration}
+\begin{frame}
+\frametitle{Stationarity}
+\begin{block}{Weak definition}
+
+A weakly stationary time series is a process which mean, variance and auto covariance do not change over time:
+\small
+{\color{blue}
+ \begin{eqnarray*} E(y_t) &=& \mu  \quad
+\forall t \in \mathbb{N} \\ E(y^2_t) &=& \sigma^2  \quad \forall t \in
+\mathbb{N} \\ \lambda(s,t)&=&\lambda(s+h,t+h) \quad \forall s,t \in \mathbb{N},
+\forall h \in \mathbb{Z} \end{eqnarray*}}
+\end{block}
+One of the consequences of stationary processes is how they recover from a shock.
+A shock represents an unexpected change in a variable or a in its error term in a particular period of time.
+For stationary time series, shocks to the system will gradually die away.
+\end{frame}
+
+\begin{frame}
+\frametitle{Non-stationary processes}
+There are different types of non-stationary time series models often found in economics:
+\begin{block}{Deterministic trend}
+ Deterministic trend or trend stationary
+processes have the following form:
+{\color{blue}
+\[
+y_t = f(t) + \epsilon_{t}
+\]}
+\noindent where $\epsilon_{t}$ represents a stationary error term (with mean 0 and variance $\sigma^2$) and $f(t)$ is a deterministic function of time.
+\end{block}
+\end{frame}
+
+\begin{frame}
+\frametitle{Non-stationary processes}
+
+\begin{block}{Stochastic trend}
+Stochastic trend processes are also called {\color{red}unit root or I(1)} and have the following form:
+{\color{blue}
+\[
+y_t = \mu + y_{t-1} + \epsilon_t
+\]}
+\noindent where $\epsilon_t$ is a stationary process. When $\mu = 0 $ the process is called pure random walk and when $\mu \neq 0$  the process is called random walk with drift.
+\end{block}
+\end{frame}
+
+\begin{frame}
+\frametitle{Integration }
+\begin{block}{Integration of order $d$}
+A time series $\mathbf{y}$ is said to be integrated of order $d$ if after
+differentiating the variable $d$ times, we get an stationary process, more precisely:
+\[
+(1-L)^d \mathbf{y} \sim \text{I(0)} \, ,
+\]
+\noindent where I(0) is a stationary time series and $L$ is the lag operator:
+\[
+(1-L)\mathbf{y} = \Delta \mathbf{y}=\mathbf{y}_t  -\mathbf{y}_{t-1} \quad \forall t
+\]
+\end{block}
+\end{frame}
+
+\begin{frame}
+\frametitle{Integration example}
+\begin{columns}
+\column[t]{5cm}
+High frequency financial time series are commonly an I(1) process 
+\begin{center}
+\includegraphics[width=0.85\textwidth]{img/EURUSD}
+\end{center}
+\column[t]{5cm}
+Since their differences are stationary (I(0) process).
+\newline
+\begin{center}
+\includegraphics[width=0.85\textwidth]{img/DEURUSD}
+\end{center}
+\end{columns}
+\end{frame}
+
+\begin{frame}
+\frametitle{Spurious regression}
+Standard regression techniques are applied to non-stationary data.
+The use of non-stationary data can lead to spurious regressions.
+
+%\begin{block}{Spurious regression}
+%If two stationary
+%variables are generated as independent random series and are trending over time, when one of those variables is regressed on the other, they could have a high coefficient of determination ($R^2$) even if the two are totally unrelated.
+%\end{block}
+
+\begin{columns}
+\begin{column}{.45\linewidth}<1->
+\begin{figure}
+\includegraphics[width=0.4\paperwidth]{img/spurious1}
+\end{figure}
+\end{column}
+\begin{column}{.45\linewidth}<2->
+\begin{figure}
+\includegraphics[width=0.4\paperwidth]{img/spurious2}
+\end{figure}
+\end{column}
+\end{columns}
+\end{frame}
+
+\begin{frame}[t,fragile]{Random walk Illustration}
  \begin{itemize}
@@
 \begin{frame}
+\hspace*{-4mm}
+\includegraphics[width=0.9\paperwidth]{img/motivation.png}
+\end{frame}
+
+\begin{frame}
 \frametitle{Motivation}
@@
 equilibrium relationship and the variables are {\color{red} cointegrated}.
-\
+\end{frame}
+
+\begin{frame}
+\frametitle{Cointegration}
+\begin{block}{Cointegration definition}
+Let {\color{blue}$\mathbf{y} = \{\mathbf{y}^1, \dots, \mathbf{y}^l\}$} be a set of $l$
+time series of order I(1) which are said to be cointegrated if a vector
+$\beta=[\beta(1),\dots,\beta(l)]^\top \in \mathbb{R}^l$  exists such that the
+time series,
+
+\begin{equation*}
+ \mathbf{Z}_t:= \beta^\top \mathbf{y} = \beta(1) \mathbf{y}^1 + \dots + \beta(l) \mathbf{y}^l \sim
+ \text{I(0)}\, .
+\end{equation*}
+\end{block}
+In other words, a set of I(1) variables is said to be cointegrated if
+a linear combination of them exists which is I(0).
 
@@
             {$c$};
-        }}_\text{Constant}
+        }}_\text{Intercept}
         +
@@
             {$\mathbf{\epsilon}_t $};
-        }}_\text{i.i.e}
+        }}_\text{i.i.d}
 \end{equation*}
@@
 \begin{frame}
-The Vector Error correction model (VECM) is used to model cointegrated time series but is used only with batch data and rarely used with high frequency data.
+\begin{alertblock}{VECM limitation}
+The Vector Error correction model (VECM) is used to model cointegrated time series but only with batch data and rarely used with high frequency data.
+\end{alertblock}
 \end{frame}
@@
 \frametitle{Proposal}
-\begin{block}{}
+\begin{block}{Proposal}
 To use VECM with high frequency data where the best parameters are updated when new data is available and do this is done in a fast way.
@@
 %\section{Background}
-%%
-%%\begin{frame}
-%%\frametitle{Integration }
-%%\begin{block}{Integration of order $d$}
-%%A time series $\mathbf{y}$ is said to be integrated of order $d$ if after
-%%differentiating the variable $d$ times, we get an stationary process, more precisely:
-%%\[
-%%(1-L)^d \mathbf{y} \sim \text{I(0)} \, ,
-%%\]
-%%\noindent where I(0) is a stationary time series and $L$ is the lag operator:
-%%\[
-%%(1-L)\mathbf{y} = \Delta \mathbf{y}=\mathbf{y}_t  -\mathbf{y}_{t-1} \quad \forall t
-%%\]
-%%\end{block}
-%%\end{frame}
-%%
-%%\subsection{Integration example}
-%%\begin{frame}
-%%\frametitle{Financial Time Series}
-%%\begin{columns}
-%%\column[t]{5cm}
-%%High frequency financial time series are commonly an I(1) process 
-%%\begin{center}
-%%\includegraphics[width=0.85\textwidth]{img/EURUSD}
-%%\end{center}
-%%\column[t]{5cm}
-%%Since their differences are stationary (I(0) process).
-%%\begin{center}
-%%\includegraphics[width=0.85\textwidth]{img/DEURUSD}
-%%\end{center}
-%%\end{columns}
-%%\end{frame}
+%
 %
 %\subsection{Cointegration}
-%\begin{frame}
-%\frametitle{Cointegration}
-%\begin{block}{Cointegration definition}
-%Let $\mathbf{y} = \{\mathbf{y}^1, \dots, \mathbf{y}^l\}$ be a set of $l$
-%time series of order I(1) which are said to be cointegrated if a vector
-%$\beta=[\beta(1),\dots,\beta(l)]^\top \in \mathbb{R}^l$  exists such that the
-%time series,
-%
-%\begin{equation*}
-% \mathbf{Z}_t:= \beta^\top \mathbf{y} = \beta(1) \mathbf{y}^1 + \dots + \beta(l) \mathbf{y}^l \sim
-% \text{I(0)}\, .
-%\end{equation*}
-%\end{block}
-%In other words, a set of I(1) variables is said to be cointegrated if
-%a linear combination of them exists which is I(0).
-%
-%\end{frame}
+
 %

### 2017-11-20 slides/defense-slides.tex
adding Online Learning slide

@@
 \end{figure}
-
 \end{frame}
@@
 \begin{frame}
+\frametitle{$\Omega$ matrix properties}
+The matrix $\Omega$ has the following properties:
+\begin{itemize}
+\item If $\Omega = 0$ there is no cointegration
+\item If $rank(\Omega)=l$ i.e full rank, then the time series are not
+I(1) but stationary
+\item If $rank(\Omega)=r,\quad 0 < r < l$ then, there is cointegration
+and the matrix $\Omega$ can be expressed as $\Omega =
+\alpha \beta^\top$, where $\alpha$ and $\beta$ are $(l \times r)$
+matrices and $rank(\alpha)=rank(\beta)=r$.
+\end{itemize}
+The columns of $\beta$ contains the cointegration vectors and the rows of
+$\alpha$ correspond with the adjusted vectors. $\beta$ is obtained by Johansen
+procedure whereas $\alpha$ has to be determined as a
+variable in the VECM.
+\end{frame}
+
+\begin{frame}
 \begin{alertblock}{VECM limitation}
@@
 \end{alertblock}
+\end{frame}
+
+\begin{frame}
+\frametitle{Online learning}
+\begin{figure}
+\includegraphics[width=0.8\paperwidth]{img/onlinelearning}
+\end{figure}
 \end{frame}
@@
 %
-%\begin{frame}
-%\frametitle{$\Omega$ matrix properties}
-%The matrix $\Omega$ has the following properties:
-%\begin{itemize}
-%\item If $\Omega = 0$ there is no cointegration
-%\item If $rank(\Omega)=l$ i.e full rank, then the time series are not
-%I(1) but stationary
-%\item If $rank(\Omega)=r,\quad 0 < r < l$ then, there is cointegration
-%and the matrix $\Omega$ can be expressed as $\Omega =
-%\alpha \beta^\top$, where $\alpha$ and $\beta$ are $(l \times r)$
-%matrices and $rank(\alpha)=rank(\beta)=r$.
-%\end{itemize}
-%The columns of $\beta$ contains the cointegration vectors and the rows of
-%$\alpha$ correspond with the adjusted vectors. $\beta$ is obtained by Johansen
-%procedure whereas $\alpha$ has to be determined as a
-%variable in the VECM.
-%\end{frame}
 %

### 2017-11-25 slides/defense-slides.tex
Drunk example ready
partial final draft

@@
 \usepackage{tikz}
+\usepackage{array}
+\usepackage{booktabs}
 \usetikzlibrary{shapes,arrows}
@@
 
-A weakly stationary time series is a process which mean, variance and auto covariance do not change over time:
+A weak stationary time series is a process which mean, variance and auto covariance do not change over time:
 \small
@@
         \begin{tikzpicture}[overlay]
-            \node[yshift=-50] (img2) {\includegraphics[scale=0.06]{img/dog}};
+            \node[yshift=-50] (img2) {\includegraphics[scale=0.1]{img/dog}};
         \end{tikzpicture}
@@
         \begin{tikzpicture}[overlay]
-            \node[yshift=-60] (img2) {\includegraphics[scale=0.1]{img/mananddog}};
+            \node[yshift=-60] (img2) {\includegraphics[scale=0.7]{img/mananddog}};
         \end{tikzpicture}
@@
 \begin{frame}
+\frametitle{VECM matrix form}
+\small
+\begin{equation*}
+\underbrace{
+      \begin{bmatrix}
+       \mathbf{\Delta y}_{p+1}  \\ 
+       \mathbf{\Delta y}_{p+2}  \\ 
+       \vdots                   \\ 
+       \mathbf{\Delta y}_N      
+      \end{bmatrix}}_{\mathbf{B} } =
+\arraycolsep=1.4pt  
+\underbrace{\begin{bmatrix}
+   \alpha^\top & \phi^*_1 & \dots & \phi^*_{p-1} &\mathbf{c}   
+  \end{bmatrix}}_{\mathbf{X}^\top}
+\underbrace{\begin{bmatrix}
+ \beta\mathbf{y}_p      & \beta\mathbf{y}_{p+1}   & \dots&   \beta\mathbf{y}_{N-1}   \\
+ \mathbf{\Delta y}_{p}   & \mathbf{\Delta y}_{p+1}& \dots &  \mathbf{\Delta y}_{N-1} \\
+ \mathbf{\Delta y}_{p-1} & \mathbf{\Delta y}_{p}  & \dots &  \mathbf{\Delta y}_{N-2}   \\
+ \vdots                  & \vdots                 & \ddots&  \vdots                   \\
+ \mathbf{\Delta y}_{2}   & \mathbf{\Delta y}_{3} & \dots &   \mathbf{\Delta y}_{N-p+1} \\
+ 1                      & 1                       & \dots     & 1   
+ \end{bmatrix}}_{\mathbf{A}^\top}
++
+\underbrace{\begin{bmatrix}
+              \mathbf{\epsilon}_{p+1} \\ 
+              \mathbf{\epsilon}_{p+2} \\ 
+              \vdots \\ 
+              \mathbf{\epsilon}_N
+             \end{bmatrix}}_{\mathbf{E}^\top}
+\end{equation*}
+\end{frame}
+
+\begin{frame}
 \frametitle{$\Omega$ matrix properties}
@@
 \begin{frame}
-\begin{alertblock}{VECM limitation}
-The Vector Error correction model (VECM) is used to model cointegrated time series but only with batch data and rarely used with high frequency data.
-\end{alertblock}
+\frametitle{VECM limitations}
+
+The Vector Error correction model (VECM) is used to model cointegrated time series but only with batch data and rarely used with high frequency data  mainly due to computational limitations:
+
+\begin{itemize}
+\item Calculation of the cointegration vectors
+\item VECM parameters are obtained using the ordinary least squares (OLS) method.
+\end{itemize}
+
+\end{frame}
+
+\section{Proposal}
+\begin{frame}
+\frametitle{Proposal}
+\begin{block}{Proposal}
+To adapt VECM to be used with high frequency data, where the best parameters are updated when new data.
+\end{block}
+
+\begin{block}{Hypothesis}
+An algorithm based on cointegration and high
+performance computing will allow faster forecasting
+algorithms for financial time series to be obtained while maintaining good accuracy levels.
+\end{block}
+\end{frame}
+
+\section{First Contribution}
+\begin{frame}
+\frametitle{First approach}
+To use VECM with high frequency data, where optimised parameters are obtained when new data. The optimisation was made in a distributed environment.
+\end{frame}
+
+\subsection{Data}
+\begin{frame}
+\frametitle{Data}
+\begin{itemize}
+\item Tests were carried out using four foreign exchange rates all related to
+USD: EURUSD, GBPUSD, USDCHF and USDJPY. 
+\item This data was collected from Dukascopy, a free
+database which gives access to the Swiss Foreign Exchange marketplace.
+\item The tests were done using 10-seconds frequency from ask prices
+from the 11th to the 15th of August 2014. 
+%\item Since one day corresponds to 8640 data points and we used 5 days of data, we have 43,200 data points in total.
+\end{itemize}
 \end{frame}
@@
 \begin{frame}
-\frametitle{Online learning}
-\begin{figure}
-\includegraphics[width=0.8\paperwidth]{img/onlinelearning}
+\frametitle{Cointegration vectors distribution}
+  \begin{figure}[!h]
+  %\vspace{-0.8cm}
+  \centering
+   \includegraphics[width=0.65\textwidth]{img/51_Fig1}
+  \caption[Distribution of the number of cointegration vectors using $p=1$ lags]{Distribution of the number of cointegration vectors using $p=1$ lags.
+  Four possible values for windows size $L$ are shown (2, 6, 10 and 14) hours (1
+  hour = 360 data points).}
+  \label{fig:hists}
 \end{figure}
@@
 
-\subsection{Proposal}
 \begin{frame}
-\frametitle{Proposal}
-\begin{block}{Proposal}
-To use VECM with high frequency data where the best parameters are updated when new data is available and do this is done in a fast way.
+\frametitle{Percentage of cointegration}
+Since $r=0$ means no cointegration and $r=l=4$ reveals that no process is I(1) but stationary.
+The interesting cases of cointegration are those where $r$ lies strictly
+between $0$ and $l$, i.e. $0<r<l$.
+\begin{block}{Percentage of cointegration (PC)}
+{\color{blue}
+\begin{equation*} \label{eq:pcoint}
+PC = 
+\frac{\#\{ it \mid \text{$it$ has $r$ c.v. with $0<r<l$}\}}
+     {\#it}\times 100
+\end{equation*}}
 \end{block}
@@
 
-%\section{Background}
-%
-%
-%\subsection{Cointegration}
-
-%
-%\subsection{Vector Autorregresive models}
-%\begin{frame}
-%\frametitle{VECM: Vector Error Correction model}
-%
-%VECM is a special form of a Vector Autorregresive model (VAR) for I(1)
-%variables $\mathbf{y}$ that are also cointegrated. Let 
-%
-%\begin{equation*}
-%\label{eq:variables}
-%\mathbf{y}_t = 
-%\begin{bmatrix} y_{1,t} & y_{2,t} & \dots & y_{l,t}
-%\end{bmatrix}^\top \, \qquad t=p+1, \dots N. 
-%\end{equation*}
-%
-%\begin{block}{VAR}
-%VAR($p$) model describes the behaviour of a set of $l$
-%endogenous variables as a linear combination of their last $p$ values:
-%{\color{blue}
-%\begin{equation*}
-%\label{eq:var}
-% \mathbf{y}_t = \phi_1 \mathbf{y}_{t-1}  + \dots +   \phi_p\mathbf{y}_{t-p}
-%+ \mathbf{c} + \mathbf{\epsilon}_t \, ,
-%\end{equation*}}
-%\noindent where  ${\phi_1,\dots,\phi_p}$ are $l \times l$
-%matrices of real coefficients,
-%$\mathbf{\epsilon}_{p+1},\dots,\mathbf{\epsilon}_N$ are error terms,
-%$\mathbf{c}$ is a constant vector and $N$ is the total number of samples.
-%\end{block}
-%\end{frame}
-%
-%
-%
-
-%
-%\noindent where 
-%%coefficients matrices $\Omega$ and $\phi_i^*$ are
-%%function of matrices $\phi_i$ as follows:
-%
-%\begin{eqnarray*}
-%\phi_i^* &: =& -\sum_{j=i+1}^{p} \phi_j \\
-%\Omega &: =& -(\mathbb{I}-\phi_1-\dots-\phi_p) 
-%\end{eqnarray*}
-%\end{frame}
-%
-%
-%
-%\begin{frame}
-%\frametitle{VECM matrix form}
-%\small
-%\begin{equation*}
-%\underbrace{
-%      \begin{bmatrix}
-%       \mathbf{\Delta y}_{p+1}  \\ 
-%       \mathbf{\Delta y}_{p+2}  \\ 
-%       \vdots                   \\ 
-%       \mathbf{\Delta y}_N      
-%      \end{bmatrix}}_{\mathbf{B} } =
-%\arraycolsep=1.4pt  
-%\underbrace{\begin{bmatrix}
-%   \alpha^\top & \phi^*_1 & \dots & \phi^*_{p-1} &\mathbf{c}   
-%  \end{bmatrix}}_{\mathbf{X}^\top}
-%\underbrace{\begin{bmatrix}
-% \beta\mathbf{y}_p      & \beta\mathbf{y}_{p+1}   & \dots&   \beta\mathbf{y}_{N-1}   \\
-% \mathbf{\Delta y}_{p}   & \mathbf{\Delta y}_{p+1}& \dots &  \mathbf{\Delta y}_{N-1} \\
-% \mathbf{\Delta y}_{p-1} & \mathbf{\Delta y}_{p}  & \dots &  \mathbf{\Delta y}_{N-2}   \\
-% \vdots                  & \vdots                 & \ddots&  \vdots                   \\
-% \mathbf{\Delta y}_{2}   & \mathbf{\Delta y}_{3} & \dots &   \mathbf{\Delta y}_{N-p+1} \\
-% 1                      & 1                       & \dots     & 1   
-% \end{bmatrix}}_{\mathbf{A}^\top}
-%+
-%\underbrace{\begin{bmatrix}
-%              \mathbf{\epsilon}_{p+1} \\ 
-%              \mathbf{\epsilon}_{p+2} \\ 
-%              \vdots \\ 
-%              \mathbf{\epsilon}_N
-%             \end{bmatrix}}_{\mathbf{E}^\top}
-%\end{equation*}
-%\end{frame}
-%
-%\section{Proposal}
-%\subsection{Motivation}
-%\begin{frame}
-%\frametitle{VECM difficulties in an online context}
-%\begin{itemize}
-%\item VECM assumes cointegration vectors do not change in time. However, these might change due to several economic factors.
-%\item To get updated cointegration vectors in time is computationally expensive because 
-%the Johansen method and OLS are used to obtain them. It would be good to know how the number of cointegration vectors are affected by the number of data used $L$ and the number of lags of VECM $p$.
-%\end{itemize}
-%\end{frame}
-%
-%
-%\begin{frame}
-%\frametitle{Data}
-%\begin{itemize}
-%\item Tests were carried out using four foreign exchange rates all related to
-%USD: EURUSD, GBPUSD, USDCHF and USDJPY. 
-%\item This data was collected from Dukascopy, a free
-%database which gives access to the Swiss Foreign Exchange marketplace.
-%\item The tests were done using 10-seconds frequency from ask prices
-%from the 11th to the 15th of August 2014. 
-%%\item Since one day corresponds to 8640 data points and we used 5 days of data, we have 43,200 data points in total.
-%\end{itemize}
-%\end{frame}
-%
-%
-%\begin{frame}
-%\frametitle{Cointegration vectors distribution}
-%\begin{figure}[!h]
-%  %\vspace{-0.8cm}
-%  \centering
-%  \includegraphics[width=0.75\textwidth]{img/histCointVectors-offset20520-p-1-freq-10s}
-%  \caption{Histogram of the number of cointegration vectors using $p=1$. Four
-%  possible values for $L$ are shown (100, 2000, 3000 and 4000).}
-%  \label{fig:hists}
-%\end{figure}
-%\end{frame}
-%
-%\begin{frame}
-%\frametitle{Percentage of cointegration}
-%Since $r=0$ means no cointegration and $r=l=4$ reveals that no process is I(1) but stationary.
-%The interesting cases of cointegration are those where $r$ lies strictly
-%between $0$ and $l$, i.e. $0<r<l$.
-%\begin{block}{Percentage of cointegration (PC)}
-%{\color{blue}
-%\begin{equation*} \label{eq:pcoint}
-%PC = 
-%\frac{\#\{ it \mid \text{$it$ has $r$ c.v. with $0<r<l$}\}}
-%     {\#it}\times 100
-%\end{equation*}}
-%\end{block}
-%\end{frame}
-%
-%
 %\begin{frame}
@@
 %
-%\subsection{AVECM}
-%\begin{frame}
-%\frametitle{Proposal: Adaptive VECM algorithm (AVECM)}
-%\begin{itemize}
-%\item We proposed to choose $L$ and $p$ in order to maximise the percentage of
-%cointegration $PC$ in the near past. This process was done every time that new data was processed. 
-%\item To use a distributed environment to make this parameters search, since this is the most expensive routine.
-%\end{itemize}
-%\end{frame}
-%
-%
-%%\begin{frame}
-%%\frametitle{AVECM Algorithm}
-%%\begin{algorithmic}[1]
-%%\REQUIRE $\,$ \\
-%%$\mathbf{y}$: matrix with $N$ input vectors and $l$ time series\\
-%%$j$: Starting point of testing \\
-%%$it$: Ending point of testing \\
-%%$ps$: list of $p$ values \\
-%%$Ls$: list of $L$ values ($L<N$) \\
-%%$m$: Iterations to determine parameters ($m < N-L$)\\
-%%\ENSURE  $\,$ \\
-%%$\{ \hat{\mathbf{y}}[1],\dots,\hat{\mathbf{y}}[it]\}$: prediction vectors \\
-%%\FOR { $i =j$ to $it$ }
-%%   \STATE $\mathbf{Y} \gets \mathbf{y}[:,i-1]$
-%%    \STATE $L,p \gets
-%%    \texttt{get\_best\_params}(Ls,ps,m,\mathbf{Y})$
-%%        \STATE $model = \text{VECM}(\mathbf{Y},L, p)$
-%%        \STATE $\hat{\mathbf{y}}[i-j] = model.predict()$
-%%\ENDFOR
-%%\end{algorithmic}
-%%\end{frame}
-%
+\subsection{AVECM}
+\begin{frame}
+\frametitle{Adaptive VECM algorithm (AVECM)}
+\begin{itemize}
+\item We proposed to choose $L$ and $p$ in order to maximise the percentage of
+cointegration $PC$ in the near past. This process was done every time that new data was processed. 
+\item To use a distributed environment to make this parameters search, since this is the most expensive routine.
+\end{itemize}
+\end{frame}
+
+\begin{frame}
+\frametitle{AVECM Algorithm}
+\small
+\begin{algorithmic}[1]
+\REQUIRE $\,$ \\
+$\mathbf{y}$: matrix with $N$ input vectors and $l$ time series\\
+$j$: Starting point of testing \\
+$it$: Ending point of testing \\
+$ps$: list of $p$ values \\
+$Ls$: list of $L$ values ($L<N$) \\
+$m$: Iterations to determine parameters ($m < N-L$)\\
+\ENSURE  $\,$ \\
+$\{ \hat{\mathbf{y}}[1],\dots,\hat{\mathbf{y}}[it]\}$: prediction vectors \\
+\FOR { $i =j$ to $it$ }
+   \STATE $\mathbf{Y} \gets \mathbf{y}[:,i-1]$
+    \STATE $L,p \gets
+    \texttt{get\_best\_params}(Ls,ps,m,\mathbf{Y})$
+        \STATE $model = \text{VECM}(\mathbf{Y},L, p)$
+        \STATE $\hat{\mathbf{y}}[i-j] = model.predict()$
+\ENDFOR
+\end{algorithmic}
+\end{frame}
+
 %\subsection{Model comparison} 
@@
 %\end{frame}
-%
-%\begin{frame}
-%\frametitle{Results}
-%\begin{tabular}{ccccccc}
-%
-%& \multicolumn{3}{l}{MSE} & &
-%\multicolumn{2}{c}{$U$-statistic} \\ 
-%\hhline{~---~--} \\
-% &
-%AVECM & ARIMA & p-value & &
-%AVECM & ARIMA \\ 
-%
-%\hline
-% EURUSD & 1.07 e-09 & 1.14 e-09 &  9.25 e-12 & & 0.6863 & 0.7108\\
-% GBPUSD & 1.66 e-09 & 1.74 e-09 &  6.95 e-02 & & 0.6866 & 0.7025\\
-% USDCHF & 5.85 e-10 & 6.35 e-10 &  2.89 e-14 & & 0.6803 & 0.7091\\
-% USDJPY  & 6.34 e-06 & 6.51 e-06 &  6.85 e-05 & & 0.6964 & 0.7057\\
-% \end{tabular}
-%\end{frame}
-%
+
 %%\begin{frame}
@@
 %
-%%\begin{frame}
-%%\frametitle{Unit root tests}
-%%VECM can only be used if the time series are I(1). Therefore, 
-%%before running the tests, we firstly checked whether the time series were
-%%I(1) using the Augmented Dickey Fuller (ADF) test with lags $p=1,2,3,4,5$.
-%%\end{frame}
-%%
-%%\begin{frame}
-%%\frametitle{Performance accuracy}
-%%\small
-%%\begin{tabular}{ccccccc}
-%%& \multicolumn{3}{l}{MSE} & &
-%%\multicolumn{2}{c}{$U$-statistic} \\ 
-%%\hhline{~---~--} \\
-%% & AVECM & ARIMA & p-value & &
-%%AVECM & ARIMA \\ 
-%%
-%%\hline
-%% EURUSD & 1.0702 e-09 & 1.1481 e-09 &  9.250 e-12 & & 0.6863 & 0.7108\\
-%% GBPUSD & 1.6630 e-09 & 1.7408 e-09 &  6.951 e-02 & & 0.6866 & 0.7025\\
-%% USDCHF & 5.8503 e-10 & 6.3545 e-10 &  2.899 e-14 & & 0.6803 & 0.7091\\
-%% USDJPY  & 6.3483 e-06 & 6.5194 e-06 &  6.853 e-05 & & 0.6964 & 0.7057
-%% \end{tabular}
-%% \end{frame}
+\begin{frame}
+\frametitle{Unit root tests}
+VECM can only be used if the time series are I(1). Therefore, 
+before running the tests, we firstly checked whether the time series were
+I(1) using the Augmented Dickey Fuller (ADF) test with lags $p=1,2,3,4,5$.
+\begin{table}[ht]
+\centering
+\tiny
+\begin{tabular}{llllll}
+\toprule
+{Variable} & {ADF(1)} & {ADF(2)} & {ADF(3)} & {ADF(4)} & {ADF(5)}\\ 
+\midrule
+EURUSD &  -0.052   & -0.054  & -0.054  & -0.054  & -0.054  \\
+GBPUSD &  -0.744  & -0.784  & -0.805  & -0.837  & -0.846  \\
+USDCHF &  -0.476   & -0.493  & -0.493  & -0.495  & -0.502  \\
+USDJPY &  0.357   & 0.360  & 0.360  & 0.367  & 0.367  \\
+$\Delta$ EURUSD & -128.4*  & -128.4*  & -96.85* & -89.12*   & -89.12*\\
+$\Delta$ GBPUSD & -131.4*  & -112.7*  & -102.5* & -92.86*   & -88.29*\\
+$\Delta$ USDCHF & -127.8*  & -127.8*  & -96.94* & -88.82*   & -80.79*\\
+$\Delta$ USDJPY & -135.1*  & -135.1*  & -101.2* & -101.2*   & -101.2*\\
+\bottomrule
+\addlinespace[1ex]
+\multicolumn{6}{l}{ \textsuperscript{*} Indicates significance at 1\% level}
+\end{tabular}
+\caption{Unit roots tests for EURUSD, GBPUSD, USDCHF and USDJPY at 10-second
+frequency.}
+\label{tab:adf}
+\end{table}
+\end{frame}
+
+\begin{frame}
+\frametitle{Performance accuracy}
+\small
+\begin{tabular}{ccccccc}
+& \multicolumn{3}{l}{MSE} & &
+\multicolumn{2}{c}{$U$-statistic} \\ 
+\hhline{~---~--} \\
+ & AVECM & ARIMA & p-value & &
+AVECM & ARIMA \\
+
+\hline
+ EURUSD & 1.0702 e-09 & 1.1481 e-09 &  9.250 e-12 & & 0.6863 & 0.7108\\
+ GBPUSD & 1.6630 e-09 & 1.7408 e-09 &  6.951 e-02 & & 0.6866 & 0.7025\\
+ USDCHF & 5.8503 e-10 & 6.3545 e-10 &  2.899 e-14 & & 0.6803 & 0.7091\\
+ USDJPY  & 6.3483 e-06 & 6.5194 e-06 &  6.853 e-05 & & 0.6964 & 0.7057
+ \end{tabular}
+ \end{frame}
 %%For AVECM we considered different number of iterations (parameter $m$ in algorithm \ref{alg:AVECM}): 10, 50 and 100.
@@
 %
-%\begin{frame}
-%\frametitle{Parallel implementation}
-%The function that get best parameters $L$ and $p$ was implemented using MPI in Python. MPI allows large-scale parallel applications
-%with wide portability to be built, being able to run in large clusters
-%or on local computers.
-%We tested our proposal in a cluster with 2 servers Xeon E5-2667 (2.90GHz)
-%of 24 cores each (48 cores in total) and 24GB RAM.
-%\end{frame}
+\begin{frame}
+\frametitle{Parallel implementation}
+The function that get best parameters $L$ and $p$ was implemented using MPI in Python. MPI allows large-scale parallel applications
+with wide portability to be built, being able to run in large clusters
+or on local computers.
+Tests ran in a cluster with 2 servers Xeon E5-2667 (2.90GHz)
+of 24 cores each (48 cores in total) and 24GB RAM.
+Parameter settings:
+\begin{itemize}
+%\item Parameter $it=100$
+\item The $L$ parameter was always chosen between 100 and 4000 and $p$ always took values
+between 1 and 5. 
+\item Parameter $nparams$ represents the number of pairs ($L$,$p$) used to maximise the percentage of cointegration. 
+\end{itemize}
+\end{frame}
 %
-%\begin{frame}
-%\frametitle{Parameters setting}
-%\begin{itemize}
-%%\item Parameter $it=100$
-%\item The $L$ parameter was always chosen between 100 and 4000 and $p$ always took values
-%between 1 and 5. 
-%\item Parameter $nparams$ represents the number of pairs ($L$,$p$) used to maximise the percentage of cointegration. 
-%\end{itemize}
-%\end{frame}
-%
-%\begin{frame}
-%\frametitle{Execution times}
-%\begin{figure}[ht]
-%  \centering
-%  \includegraphics[width=0.75\textwidth]{img/extimes}
-%  \caption{Computing time of sequential and parallel algorithm is shown in the
-%  upper figure. Speed-up is shown below.}
-%  \label{fig:extimes}
-%% The superior figure \ref{fig:extimes} shows that best performance accuracy measurements are
-%%achieved in times near or below 10 seconds in the parallel version. Contrarily,
-%%serial times are high, above 10 seconds in most cases. Figure~\ref{fig:extimes} also shows the speed-up in computing time for AVECM which is near 9X if we use more than 25 parameters ($nparams$).
-%%Execution times do not consider the loading data time, just that of finding
-%%best parameters and matrix operations. The time MPI spends
-%%transferring data and synchronising processes is about two seconds independently
-%%of the number of processes considered. Execution times were measured using the Time python library.
-%\end{figure}
-%
-%\end{frame}
-%
-%\section{Conclusions}
-%\subsection{Uno}
-%\begin{frame}
-%\frametitle{Conclusions}
-%\begin{itemize}
-%\item Cointegration relations change with time. We empirically showed that the Johansen method is sensitive to the number of lags but also to the amount of data considered.
-%\item We found that out-of-sample forecast performance MSE is related to the {\em percentage of cointegration\/}.  
-%\item VECM improves performance measures by finding
-%parameters of $L$ and $p$ maximising the percentage of cointegration.
-%\item Increasing the number of parameters will always lead to
-%better performance measures.
-%\end{itemize}
-%\end{frame}
-%
-%\begin{frame}
-%\frametitle{Conclusions}
-%\begin{itemize}
-%\item The parallel implementation allowed the execution times to be reduced
-%more than 9 times and therefore a response time was obtain before 10
-%seconds. 
-%\item For future study, it would be interesting to explore the relationship between
-%cointegration and performance in order to propose new criteria for
-%improving VECM parameters. It would also be interesting to include more 
-%explaining variables such as bid-ask spread and change in volume.
-%\end{itemize}
-%\end{frame}
+\begin{frame}
+\frametitle{Execution times}
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width=0.7\textwidth]{img/51_Fig3}
+  \caption[Computing time and Speed-up]{Computing time of sequential and parallel algorithm is shown in the
+  upper figure. Speed-up is shown below.}
+  \label{fig:extimes}
+\end{figure}
+\end{frame}
+
+\section{Conclusions}
+\subsection{Uno}
+\begin{frame}
+\frametitle{Conclusions}
+\begin{itemize}
+\item Cointegration relations change with time. We empirically showed that the Johansen method is sensitive to the number of lags but also to the amount of data considered.
+\item We found that out-of-sample forecast performance MSE is related to the {\em percentage of cointegration\/}.  
+\item VECM improves performance measures by finding
+parameters of $L$ and $p$ maximising the percentage of cointegration.
+\item Increasing the number of parameters will always lead to
+better performance measures.
+\item The parallel implementation allowed the execution times to be reduced
+more than 9 times and therefore a response time was obtain before 10
+seconds. 
+\end{itemize}
+\end{frame}
+
+\section{Second Contribution}
+
+\begin{frame}
+\frametitle{Second approach}
+ Recently, online learning algorithms have been proposed to solve problems with large data sets because of their simplicity and their ability to update the model when new data is
+available.
+\begin{figure}
+\includegraphics[width=0.6\paperwidth]{img/onlinelearning}
+\end{figure} 
+\end{frame}
+
+\begin{frame}
+\frametitle{Proposal: Online VECM algorithm (OVECM)}
+\begin{itemize}
+\item Considers only a sliding window of the $L$ most recent data. 
+\item Since cointegration vectors represent long-run relationships they vary
+little in time. OVECM obtains these vectors only when they need to be updated. 
+\item This updating is done when MAPEs of the last $n$ inputs is above a certain
+error given by a threshold 
+\item OVECM also implements matrix optimisations in order to reduce execution time,
+such as updating matrices with new data, removing old data and introducing new
+cointegration vectors.
+\end{itemize}
+\end{frame}
+
+\begin{frame}
+\frametitle{Methodology}
+OVECM were compared against VECM and ARIMA. VECM and ARIMA are called with a
+sliding window of the most recent data, whereby the models are updated at every
+time step. Therefore, we named them SLVECM and SLARIMA.
+\end{frame}
+
+\subsection{Parameters}
+\begin{frame}
+\frametitle{Parameters optimization}
+VECM order and ARIMA parameters were selected using AIC.
+
+\begin{table}[ht]
+\label{tab:params}
+\begin{center}
+\begin{tabular}{|c|c|c|}
+\hline
+Windows size $L$ & VECM & ARIMA\\
+ $L$ & order $(p)$ & order $(p,d,q)$ \\
+\hline
+100 & 2 & p=2,d=1,q=1\\
+400 & 5 & p=1,d=1,q=1\\
+700 & 3 &p=2,d=1,q=1\\
+1000 &3 & p=2,d=1,q=1\\
+\hline
+\end{tabular}
+\end{center}
+\end{table}
+\end{frame}
+
+\begin{frame}
+\frametitle{Execution times}
+We ran 400 iterations of OVECM and SLVECM . SLARIMA was based on the ARIMA procedure
+of the statsmodels library and therefore its execution time is excluded.
+%because its is not comparable with OVECM and SLVECM. SLARIMA was created based
+%on statsmodels library routine ARIMA.
+\begin{center}
+\adjustbox{max height=\dimexpr\textheight-5.5cm\relax,
+           max width=\textwidth}{
+\begin{tabular}{|l|c|c|c|c|c|}
+\hline
+& $L$ & order & e  & Time[s] \\
+\hline
+OVECM & 100 &p=2  & 0      & 2.492\\
+OVECM & 100 &p=2  & 0.0026  & 1.606\\
+SLVECM & 100 &p=2& -- & 2.100\\
+\hline
+OVECM & 400 & p=5  & 0      & 3.513\\
+OVECM & 400 &p=5  & 0.0041  & 2.569\\
+SLVECM & 400 & p=5 & -- & 3.222\\
+\hline
+OVECM & 700 &p=3  & 0      & 3.296\\
+OVECM & 700 &p=3  & 0.0032  & 2.856\\
+SLVECM & 700 &p=3 & -- & 3.581\\
+\hline
+OVECM & 1000 & p=3 & 0      & 4.387\\
+OVECM & 1000 & p=3  & 0.0022  & 2.408\\
+SLVECM & 1000 & p=3  & -- & 3.609\\
+\hline
+\end{tabular}
+}\end{center}
+\end{frame}
+
+\begin{frame}
+\frametitle{Performance accuracy}
+\tiny
+\begin{table}[H]
+\caption{Model measures}
+\label{tab:stats}
+\begin{center}
+\begin{tabular}{|l|l|c|c|c|c|c|c|c|}
+\hline
+\multicolumn{3}{|c|}{Model} & \multicolumn{3}{|c|}{In-sample} &
+\multicolumn{3}{|c|}{Out-of-sample} \\ 
+\hline
+\hline
+Method & $L$ & Parameters &  
+MAPE & MAE& RMSE& 
+MAPE & MAE& RMSE \\
+\hline
+ OVECM  &   100  &  P=2 e=0.0026 &  0.00263&  0.00085&  0.00114&  0.00309&  0.00094&  0.00131\\
+ OVECM  &   400  &  P=5 e=0.0041&  0.00378&  0.00095&  0.00127&  0.00419&  0.00103&  0.00143\\
+ OVECM  &   700  &  P=3 e=0.0032&  0.00323&  0.00099&  0.00130&  0.00322&  0.00097&  0.00132\\
+ OVECM  &   1000 &  P=3 e=0.0022&  
+ \textbf{0.00175}&  \textbf{0.00062}&  \textbf{0.00087} &  
+ \textbf{0.00172}&  \textbf{0.00061}&  \textbf{0.00090}\\
+\hline
+ SLVECM  &   100 &  P=2 &  0.00262&  0.00085&  0.00113&  0.00310&  0.00095&  0.00132\\
+ SLVECM  &   400 &  P=5 &  0.00375&  0.00095&  0.00126&  0.00419&  0.00103&  0.00143\\
+ SLVECM  &   700 &  P=3 &  0.00324&  0.00099&  0.00130&  0.00322&  0.00098&  0.00132\\
+ SLVECM  &   1000&  P=3 &  
+ \textbf{0.00174}&  \textbf{0.00061}&  \textbf{0.00087}&  
+ \textbf{0.00172}&  \textbf{0.00061}&  \textbf{0.00090}\\
+\hline
+SLARIMA & 100  &p=2 d=1 q=1 & 0.00285  &  0.00110  &  0.00308  &  0.00312  &0.00098  &  0.00144 \\
+SLARIMA & 400  &p=1 d=1 q=1 & 0.00377  &  0.00101  &  0.00128  &  0.00418 & 0.00106 & 0.00145 \\
+SLARIMA & 700  &p=2 d=1 q=1 & 0.00329  &  0.00102  &  0.00136  &  0.00324  & 0.00097  &  0.00133 \\
+SLARIMA & 1000 &p=2 d=1 q=1 & \textbf{0.00281}  & \textbf{0.00074}  &  \textbf{0.00105}  &  \textbf{0.00177} & \textbf{0.00063}  &  \textbf{0.00092} \\
+\hline
+\end{tabular}
+\end{center}
+\end{table}
+\end{frame}
+
+\begin{frame}
+\frametitle{Conclusions}
+\begin{itemize}
+\item  OVECM considerably reduces execution times
+without compromising solution accuracy.  
+\item OVECM was compared with VECM and ARIMA
+with the same sliding window sizes and OVECM outperformed both in terms of
+execution time. 
+\item Traditional VECM slightly outperformed our proposal but the
+OVECM execution time is lower.
+\item We could see that our algorithm took much less than a minute at every
+step. This means that it could also be used with higher frequency data and would
+still provide responses before new data arrives.  
+\end{itemize}
+\end{frame}
+
+\begin{frame}
+\frametitle{Summary of Contributions}
+\begin{itemize}
+\item Two different approaches based on cointegration were presented: an adaptive version of VECM and an online approach.
+\item The concept of cointegration can now be used with high frequency time series.
+\item VECM computational limitations can either be tackled using a distribution computations or considering an online approach.
+\item An interesting connection between the percentage cointegration and forecasting performance.
+\item Cointegration information can be used as an integration tool to detect arbitrage opportunities.
+\end{itemize}
+\end{frame}
+
+\begin{frame}
+\frametitle{Author Contributions}
+\begin{itemize}
+\item 1
+\item 2
+\end{itemize}
+\end{frame}
+
+\begin{frame}
+\frametitle{Future research}
+\begin{itemize}
+\item 1
+\item 2
+\end{itemize}
+\end{frame}
 
@@
 \end{frame}
+
 %Have a slide for each of the following

### 2017-11-26 slides/defense-slides.tex
more images

@@
 \usetheme{Warsaw}
+%\usecolortheme{lily}
 %\usetheme{metropolis}
@@
 \usepackage{amsmath,amsfonts,amssymb}
-\usepackage{adjustbox}
 \usepackage{color}
@@
 \usepackage{array}
+%\usepackage{graphicx}
+\usepackage{adjustbox}
 \usepackage{booktabs}
 \usetikzlibrary{shapes,arrows}
-\usepackage{multicol}
+%\usepackage{multicol}
 \usepackage{hhline}
@@
 
-   \definecolor{myblue1}{RGB}{35,119,189}
-    \definecolor{myblue2}{RGB}{95,179,238}
-    \definecolor{myblue3}{RGB}{129,168,207}
-    \definecolor{myblue4}{RGB}{26,89,142}
+%   \definecolor{myblue1}{RGB}{35,119,189}
+%    \definecolor{myblue2}{RGB}{95,179,238}
+%    \definecolor{myblue3}{RGB}{129,168,207}
+%    \definecolor{myblue4}{RGB}{26,89,142}
+%
+%    \setbeamercolor*{structure}{fg=myblue1,bg=blue}
+%    \setbeamercolor*{palette primary}{use=structure,fg=white,bg=structure.fg}
+%    \setbeamercolor*{palette secondary}{use=structure,fg=white,bg=structure.fg!75!black}
+%    \setbeamercolor*{palette tertiary}{use=structure,fg=white,bg=structure.fg!50!black}
+%    \setbeamercolor*{palette quaternary}{fg=black,bg=white}
+%
+%    \setbeamercolor*{item projected}{fg=red,bg=myblue3!80}
+%    \setbeamercolor*{block title example}{fg=white,bg=myblue4}
+%    \setbeamercolor*{frametitle}{fg=black}
+%
+%    \setbeamertemplate{blocks}[rounded][shadow=true]
+%
+%    \makeatletter
+%    \pgfdeclarehorizontalshading[frametitle.bg,frametitle right.bg]{beamer@frametitleshade}{\paperheight}{%
+%      color(0pt)=(myblue2);
+%      color(\paperwidth)=(white)}
+%
+%    \defbeamertemplate*{footline}{mysplit theme}
+%    {%
+%      \leavevmode%
+%      \hbox{\begin{beamercolorbox}[wd=.5\paperwidth,ht=2.5ex,dp=1.125ex,leftskip=.3cm plus1fill,rightskip=.3cm]{author in head/foot}%
+%        \usebeamerfont{author in head/foot}\insertshortauthor
+%      \end{beamercolorbox}%
+%      \begin{beamercolorbox}[wd=.5\paperwidth,ht=2.5ex,dp=1.125ex,leftskip=.3cm,rightskip=.3cm plus1fil]{title in head/foot}%
+%        \usebeamerfont{title in head/foot}\insertshorttitle\hfill
+%        \insertframenumber/\inserttotalframenumber\hspace*{0.5em}
+%      \end{beamercolorbox}}%
+%      \vskip0pt%
+%    }
+%    \makeatother
 
-    \setbeamercolor*{structure}{fg=myblue1,bg=blue}
-    \setbeamercolor*{palette primary}{use=structure,fg=white,bg=structure.fg}
-    \setbeamercolor*{palette secondary}{use=structure,fg=white,bg=structure.fg!75!black}
-    \setbeamercolor*{palette tertiary}{use=structure,fg=white,bg=structure.fg!50!black}
-    \setbeamercolor*{palette quaternary}{fg=black,bg=white}
-
-    \setbeamercolor*{item projected}{fg=red,bg=myblue3!80}
-    \setbeamercolor*{block title example}{fg=white,bg=myblue4}
-    \setbeamercolor*{frametitle}{fg=black}
-
-    \setbeamertemplate{blocks}[rounded][shadow=true]
-
-    \makeatletter
-    \pgfdeclarehorizontalshading[frametitle.bg,frametitle right.bg]{beamer@frametitleshade}{\paperheight}{%
-      color(0pt)=(myblue2);
-      color(\paperwidth)=(white)}
-
-    \defbeamertemplate*{footline}{mysplit theme}
-    {%
-      \leavevmode%
-      \hbox{\begin{beamercolorbox}[wd=.5\paperwidth,ht=2.5ex,dp=1.125ex,leftskip=.3cm plus1fill,rightskip=.3cm]{author in head/foot}%
-        \usebeamerfont{author in head/foot}\insertshortauthor
-      \end{beamercolorbox}%
-      \begin{beamercolorbox}[wd=.5\paperwidth,ht=2.5ex,dp=1.125ex,leftskip=.3cm,rightskip=.3cm plus1fil]{title in head/foot}%
-        \usebeamerfont{title in head/foot}\insertshorttitle\hfill
-        \insertframenumber/\inserttotalframenumber\hspace*{0.5em}
-      \end{beamercolorbox}}%
-      \vskip0pt%
+\usebackgroundtemplate{%             declare it
+\tikz[overlay,remember picture] \node[opacity=0.2, at=(current page.center)] {
+   \includegraphics[height=\paperheight,width=\paperwidth]{img/background}};
 }
-    \makeatother
 
@@
     \tikz[overlay,remember picture]
-        \node[at=(current page.south), anchor=south] {
+        \node[at=(current page.south east), anchor=south east] {
             \includegraphics[height=1.5cm]{img/DIT}
@@
 \begin{document}
+
 \begin{frame}[plain]
@@
 \end{frame}
+\usebackgroundtemplate{ }
 
@@
 
-A weak stationary time series is a process which mean, variance and auto covariance do not change over time:
+A weakly stationary time series is a process which mean, variance and auto covariance do not change over time:
 \small
@@
 
-\Large
+{\Large
 \begin{equation*}
@@
 \end{equation*}
+}
+The columns of $\beta$ contains the cointegration vectors and the rows of
+$\alpha$ correspond with the adjusted vectors. $\beta$ is obtained by Johansen
+procedure whereas $\alpha$ has to be determined as a
+variable in the VECM.
 
@@
 \end{itemize}
-The columns of $\beta$ contains the cointegration vectors and the rows of
-$\alpha$ correspond with the adjusted vectors. $\beta$ is obtained by Johansen
-procedure whereas $\alpha$ has to be determined as a
-variable in the VECM.
 \end{frame}
@@
 \frametitle{VECM limitations}
-
 The Vector Error correction model (VECM) is used to model cointegrated time series but only with batch data and rarely used with high frequency data  mainly due to computational limitations:
-
 \begin{itemize}
@@
 \end{itemize}
-
 \end{frame}
@@
 \begin{block}{Proposal}
-To adapt VECM to be used with high frequency data, where the best parameters are updated when new data.
+To adapt VECM to be used with high frequency data, where the best parameters are updated when new data arrives.
 \end{block}
-
 \begin{block}{Hypothesis}
-An algorithm based on cointegration and high
-performance computing will allow faster forecasting
+An algorithm based on cointegration and high performance computing will allow faster forecasting
 algorithms for financial time series to be obtained while maintaining good accuracy levels.
@@
 \frametitle{First approach}
-To use VECM with high frequency data, where optimised parameters are obtained when new data. The optimisation was made in a distributed environment.
+An adaptive Vector Error Correction algorithm that:
+\begin{itemize}
+\item considers the last most recent data (windows size L)
+\item finds optimal parameters of VECM in that window.
+\item the optimisation is made in a distributed environment.
+\end{itemize}
+\end{frame}
+
+\begin{frame}
+\frametitle{First approach idea}
+\begin{figure}
+\includegraphics[width=0.65\textwidth]{img/slidingwindow}
+\end{figure}
+\end{frame}
+
+\begin{frame}
+\frametitle{How to choose L?}
+  \begin{figure}[!h]
+  %\vspace{-0.8cm}
+  \centering
+   \includegraphics[width=0.65\textwidth]{img/51_Fig1}
+  \caption[Distribution of the number of cointegration vectors using $p=1$ lags]{Distribution of the number of cointegration vectors using $p=1$ lags.
+  Four possible values for windows size $L$ are shown (2, 6, 10 and 14) hours (1
+  hour = 360 data points).}
+  \label{fig:hists}
+\end{figure}
 \end{frame}
@@
 \end{itemize}
-\end{frame}
-
-\begin{frame}
-\frametitle{Cointegration vectors distribution}
  \begin{figure}[!h]
-  %\vspace{-0.8cm}
-  \centering
-   \includegraphics[width=0.65\textwidth]{img/51_Fig1}
-  \caption[Distribution of the number of cointegration vectors using $p=1$ lags]{Distribution of the number of cointegration vectors using $p=1$ lags.
-  Four possible values for windows size $L$ are shown (2, 6, 10 and 14) hours (1
-  hour = 360 data points).}
-  \label{fig:hists}
+    \includegraphics[width=0.2\textwidth]{img/dukascopy}
  \end{figure}
@@
 VECM order and ARIMA parameters were selected using AIC.
-
 \begin{table}[ht]

### 2017-12-07 src/firmas.tex
Firmas definitivo

--- 2014-10-29 src/firmas.tex
@@
 \begin{tabular}[c]{lc}
-{\textsc Prof. Dr. Luis Salinas} & \begin{picture}(150,5) \line(1,0){150} \end{picture}\\
-{Universidad Técnica Federico Santa María} & {Advisor}\\
+{\textsc Dr. Luis Salinas} & \begin{picture}(150,5) \line(1,0){150} \end{picture}\\
+{Universidad T\'ecnica Federico Santa Mar\'ia} & {Advisor}\\
 Chile & \\
@@
 & \\
-{\textsc Prof. Dr. } &  \begin{picture}(150,5) \line(1,0){150} \end{picture}\\
-{Universidad Técnica Federico Santa María} & {Committee Member} \\
+{\textsc Dr. Werner Kristjanpoller} &  \begin{picture}(150,5) \line(1,0){150} \end{picture}\\
+{Universidad T\'ecnica Federico Santa Mar\'ia} & {Co-advisor} \\
 Chile & \\
@@
 & \\
-{Prof. Dr. } & \begin{picture}(150,5) \line(1,0){150} \end{picture}\\
-{University} & {External Peer Reviewer}\\
+{\textsc Dr. Claudio Torres} &  \begin{picture}(150,5) \line(1,0){150} \end{picture}\\
+{Universidad T\'ecnica Federico Santa Mar\'ia} & {Special Reviewer} \\
+Chile & \\
+& \\
+& \\
+{\textsc Dr. Marcelo Mendoza} &  \begin{picture}(150,5) \line(1,0){150} \end{picture}\\
+{Universidad T\'ecnica Federico Santa Mar\'ia} & {Internal Reviewer} \\
+Chile & \\
+& \\
+& \\
+{Dr. Christian de Peretti} & \begin{picture}(150,5) \line(1,0){150} \end{picture}\\
+{Ecole Centrale de Lyon} & {External Peer Reviewer}\\
+{France} & \\
+& \\
+& \\
+{Dr. Wilfredo Palma} & \begin{picture}(150,5) \line(1,0){150} \end{picture}\\
+{Pontificia Universidad Cat\'olica de Chile} & {External Peer Reviewer}\\
 {Chile} & \\
+%{\textsc Dr. Hern\'an Astudillo} &  \begin{picture}(150,5) \line(1,0){150} \end{picture}\\
+%{Universidad T\'ecnica Federico Santa Mar\'ia} & { PhD program Director} \\
+%Chile & \\
 \end{tabular}

### 2017-12-07 slides/defense-slides.tex
Preparing first changes

@@
 \documentclass{beamer}
-\usetheme{Warsaw}
+\usetheme{CambridgeUS}
 %\usecolortheme{lily}
@@
 
-\usebackgroundtemplate{%             declare it
-\tikz[overlay,remember picture] \node[opacity=0.2, at=(current page.center)] {
-   \includegraphics[height=\paperheight,width=\paperwidth]{img/background}};
-}
+%\usebackgroundtemplate{%             declare it
+%\tikz[overlay,remember picture] \node[opacity=0.2, at=(current page.center)] {
+%   \includegraphics[height=\paperheight,width=\paperwidth]{img/background}};
+%}
 
@@
 \end{frame}
-\usebackgroundtemplate{ }
+%\usebackgroundtemplate{ }
 
@@
 \begin{itemize}
-\item Motivation: economic relations
+\item Motivation: Financial time series and HFT
 \item Cointegration and Integration concepts
@@
 \begin{frame}
-\frametitle{Motivation}
-Most financial time series are non-stationary in nature.
-This implies that the statistical properties, such as mean and variance of the data change over time.
-Therefore, financial time series forecasting is still a challenge.
+\frametitle{HFT: High Frequency Trading}
+High frequency trading has motivated computer-driven strategies capable of processing large amount of data in short periods of time.
 \begin{figure}
-\includegraphics[width=0.4\paperwidth]{img/forecast}
+\includegraphics[width=0.7\paperwidth]{img/HFTmarket}
+\caption{Percentage of HFT trades of total US Equity Trading. Source: TABB group}
 \end{figure}
+\end{frame}
+
+\begin{frame}
+\frametitle{HFT Revenue}
+However, the revenues have fallen dramatically mainly because of more regulation, increasing cost of HFT infrastructure and competition.
+\begin{figure}
+\includegraphics[width=0.7\paperwidth]{img/HFTmarket2}
+\caption{Revenue in the US. Source: TABB group}
+\end{figure}
+\end{frame}
+
+\begin{frame}
+\frametitle{High Frequency data}
+High frequency time series are commonly:
+\begin{columns}
+\begin{column}{0.5\textwidth}
+\begin{itemize}
+\item Non-stationary
+\item Irregularly spaced over time.
+%\item Heterogeneity
+\item Heteroscedastic
+\end{itemize}
+\end{column}
+\begin{column}{0.5\textwidth}
+\begin{figure}
+\includegraphics[width=0.35\paperwidth]{img/forecast}
+\end{figure}
+\end{column}
+\end{columns}
+\vspace{5mm}
+This imply that classical statistical models can't be used. Therefore, financial time series forecasting is still a challenge.
 \end{frame}
@@
 \begin{block}{Strong definition}
+A strictly stationary time series $y_t \in \mathbb{R}$, $\forall t \in \mathbb{Z}$ is one for which the probabilistic behaviour of every collection of values $\{y_{t_1},y_{t_2},\dots,y_{t_L}\}$ is identical
+to that of the time shifted set, more precisely:
+\small
 
-The hole distribution of the variable does not depend on time, more precisely:
-\small
 \begin{equation*}
+{\color{blue}
 P\{y_{t_1} \leq
-c_1,..,y_{t_L} \leq c_L\} = P\{y_{t_1+h} \leq c_1,..,y_{t_L+h} \leq c_L\}\enspace\forall L \in \mathbb{N}, \forall h \in \mathbb{Z}
+c_1,..,y_{t_L} \leq c_L\} = P\{y_{t_1+h} \leq c_1,..,y_{t_L+h} \leq c_L\}\enspace }\forall L \in \mathbb{N}, \forall h \in \mathbb{Z}
 \end{equation*}
+\noindent where $c_1,\dots,c_L$ are constants.
 \end{block}
@@
 \end{block}
+\noindent with $\lambda(s,t) = E[(y_s-\mu)(y_t - \mu)]$ 
+\end{frame}
+
+\begin{frame}
+\frametitle{Stationarity: consequences}
 One of the consequences of stationary processes is how they recover from a shock.
@@
 There are different types of non-stationary time series models often found in economics:
-\begin{block}{Deterministic trend}
- Deterministic trend or trend stationary
-processes have the following form:
-{\color{blue}
-\[
-y_t = f(t) + \epsilon_{t}
-\]}
-\noindent where $\epsilon_{t}$ represents a stationary error term (with mean 0 and variance $\sigma^2$) and $f(t)$ is a deterministic function of time.
-\end{block}
-\end{frame}
-
-\begin{frame}
-\frametitle{Non-stationary processes}
-
 \begin{block}{Stochastic trend}
-Stochastic trend processes are also called {\color{red}unit root or I(1)} and have the following form:
+Stochastic trend processes are also called {\color{red}unit root or I(1)} if they have the following form:
 {\color{blue}
@@
 \end{block}
+%\begin{block}{Deterministic trend}
+% Deterministic trend or trend stationary
+%processes have the following form:
+%{\color{blue}
+%\[
+%y_t = f(t) + \epsilon_{t}
+%\]}
+%\noindent where $\epsilon_{t}$ represents a stationary error term (with mean 0 and variance $\sigma^2$) and $f(t)$ is a deterministic function of time.
+%\end{block}
+\end{frame}
+
+%\begin{frame}[t,fragile]{Random walk Illustration}
+% \begin{itemize}
+%  \item<1-> The drunkard's walk
+%  \item<2-> The dog's walk
+%  %\item<3-> The drunk and dog's walk
+% \end{itemize}
+% 
+%\begin{columns}
+%\begin{column}{.4\linewidth}<1->
+%\begin{center}
+%        \begin{tikzpicture}[overlay]
+%            \node[yshift=-50] (img2) {\includegraphics[scale=1.4]{img/drunk}};
+%        \end{tikzpicture}
+%\end{center}
+%\end{column}
+%\begin{column}{.08\linewidth}<2->
+%        \begin{tikzpicture}[overlay]
+%            \node[yshift=-50] (img2) {\includegraphics[scale=0.1]{img/dog}};
+%        \end{tikzpicture}
+%\end{column}
+%%\begin{column}{.5\linewidth}<3->
+%%    \begin{center}
+%%        \begin{tikzpicture}[overlay]
+%%            \node[yshift=-50] (img2) {\includegraphics[scale=0.1]{img/mananddog}};
+%%        \end{tikzpicture}
+%%    \end{center}
+%%\end{column}
+%\end{columns}
+%\end{frame}
+%
+%\begin{frame}[t,fragile]{Random walk and Cointegration Illustration}
+% \begin{itemize}
+%  \item The drunk and dog's walk
+% \end{itemize}
+%    \begin{center}
+%        \begin{tikzpicture}[overlay]
+%            \node[yshift=-60] (img2) {\includegraphics[scale=0.7]{img/mananddog}};
+%        \end{tikzpicture}
+%    \end{center}
+%\end{frame}
+
+\begin{frame}
+\frametitle{Irregularly spaced over time}
+\begin{figure}
+\includegraphics[width=\paperwidth]{img/hft-ticks-zoom.png}
+\end{figure}
 \end{frame}
@@
 \begin{frame}
-\frametitle{Integration }
-\begin{block}{Integration of order $d$}
-A time series $\mathbf{y}$ is said to be integrated of order $d$ if after
-differentiating the variable $d$ times, we get an stationary process, more precisely:
-\[
-(1-L)^d \mathbf{y} \sim \text{I(0)} \, ,
-\]
-\noindent where I(0) is a stationary time series and $L$ is the lag operator:
-\[
-(1-L)\mathbf{y} = \Delta \mathbf{y}=\mathbf{y}_t  -\mathbf{y}_{t-1} \quad \forall t
-\]
-\end{block}
-\end{frame}
-
-\begin{frame}
-\frametitle{Integration example}
-\begin{columns}
-\column[t]{5cm}
-High frequency financial time series are commonly an I(1) process 
-\begin{center}
-\includegraphics[width=0.85\textwidth]{img/EURUSD}
-\end{center}
-\column[t]{5cm}
-Since their differences are stationary (I(0) process).
-\newline
-\begin{center}
-\includegraphics[width=0.85\textwidth]{img/DEURUSD}
-\end{center}
-\end{columns}
-\end{frame}
-
-\begin{frame}
-\frametitle{Spurious regression}
-Standard regression techniques are applied to non-stationary data.
-The use of non-stationary data can lead to spurious regressions.
-
-%\begin{block}{Spurious regression}
-%If two stationary
-%variables are generated as independent random series and are trending over time, when one of those variables is regressed on the other, they could have a high coefficient of determination ($R^2$) even if the two are totally unrelated.
-%\end{block}
-
-\begin{columns}
-\begin{column}{.45\linewidth}<1->
-\begin{figure}
-\includegraphics[width=0.4\paperwidth]{img/spurious1}
-\end{figure}
-\end{column}
-\begin{column}{.45\linewidth}<2->
-\begin{figure}
-\includegraphics[width=0.4\paperwidth]{img/spurious2}
-\end{figure}
-\end{column}
-\end{columns}
-\end{frame}
-
-\begin{frame}[t,fragile]{Random walk Illustration}
- \begin{itemize}
-  \item<1-> The drunkard's walk
-  \item<2-> The dog's walk
-  %\item<3-> The drunk and dog's walk
- \end{itemize}
-
-\begin{columns}
-\begin{column}{.4\linewidth}<1->
-\begin{center}
-        \begin{tikzpicture}[overlay]
-            \node[yshift=-50] (img2) {\includegraphics[scale=1.4]{img/drunk}};
-        \end{tikzpicture}
-\end{center}
-\end{column}
-\begin{column}{.08\linewidth}<2->
-        \begin{tikzpicture}[overlay]
-            \node[yshift=-50] (img2) {\includegraphics[scale=0.1]{img/dog}};
-        \end{tikzpicture}
-\end{column}
-%\begin{column}{.5\linewidth}<3->
-%    \begin{center}
-%        \begin{tikzpicture}[overlay]
-%            \node[yshift=-50] (img2) {\includegraphics[scale=0.1]{img/mananddog}};
-%        \end{tikzpicture}
-%    \end{center}
-%\end{column}
-\end{columns}
-\end{frame}
-
-\begin{frame}[t,fragile]{Random walk and Cointegration Illustration}
- \begin{itemize}
-  \item The drunk and dog's walk
- \end{itemize}
-    \begin{center}
-        \begin{tikzpicture}[overlay]
-            \node[yshift=-60] (img2) {\includegraphics[scale=0.7]{img/mananddog}};
-        \end{tikzpicture}
-    \end{center}
-\end{frame}
-
-\begin{frame}
-\hspace*{-4mm}
-\includegraphics[width=0.9\paperwidth]{img/motivation.png}
+\frametitle{Heteroscedastic}
+%HASTA AQUI
 \end{frame}
@@
 
+%\begin{frame}
+%\hspace*{-4mm}
+%\includegraphics[width=\paperwidth]{img/hft-ticks.png}
+%\end{frame}
+%
+
 %Have a slide for each of the following
@@
 
+\begin{frame}
+\frametitle{Integration }
+\begin{block}{Integration of order $d$}
+A time series $\mathbf{y}$ is said to be integrated of order $d$ if after
+differentiating the variable $d$ times, we get an stationary process, more precisely:
+\[
+(1-L)^d \mathbf{y} \sim \text{I(0)} \, ,
+\]
+\noindent where I(0) is a stationary time series and $L$ is the lag operator:
+\[
+(1-L)\mathbf{y} = \Delta \mathbf{y}=\mathbf{y}_t  -\mathbf{y}_{t-1} \quad \forall t
+\]
+\end{block}
+\end{frame}
+
+\begin{frame}
+\frametitle{Integration example}
+\begin{columns}
+\column[t]{5cm}
+High frequency financial time series are commonly an I(1) process 
+\begin{center}
+\includegraphics[width=0.85\textwidth]{img/EURUSD}
+\end{center}
+\column[t]{5cm}
+Since their differences are stationary (I(0) process).
+\newline
+\begin{center}
+\includegraphics[width=0.85\textwidth]{img/DEURUSD}
+\end{center}
+\end{columns}
+\end{frame}
+
+\begin{frame}
+\frametitle{Spurious regression}
+Standard regression techniques are applied to non-stationary data.
+The use of non-stationary data can lead to spurious regressions.
+
+%\begin{block}{Spurious regression}
+%If two stationary
+%variables are generated as independent random series and are trending over time, when one of those variables is regressed on the other, they could have a high coefficient of determination ($R^2$) even if the two are totally unrelated.
+%\end{block}
+
+\begin{columns}
+\begin{column}{.45\linewidth}<1->
+\begin{figure}
+\includegraphics[width=0.4\paperwidth]{img/spurious1}
+\end{figure}
+\end{column}
+\begin{column}{.45\linewidth}<2->
+\begin{figure}
+\includegraphics[width=0.4\paperwidth]{img/spurious2}
+\end{figure}
+\end{column}
+\end{columns}
+\end{frame}
+
 % Financial time series are not the only ones with this behaviour, this study can be extended to other non-stationary time series such as: weather, earthquakes, energy demand and sales forecasting.
+
+\bibliographystyle{amscls/amsalpha}
+\bibliography{reference}
 \end{document}

### 2017-12-08 slides/defense-slides.tex
New theme and heteroscedasticity def

@@
 \documentclass{beamer}
-\usetheme{CambridgeUS}
-%\usecolortheme{lily}
-%\usetheme{metropolis}
 %\usepackage[spanish,activeacute]{babel}
@@
 
-%   \definecolor{myblue1}{RGB}{35,119,189}
-%    \definecolor{myblue2}{RGB}{95,179,238}
-%    \definecolor{myblue3}{RGB}{129,168,207}
-%    \definecolor{myblue4}{RGB}{26,89,142}
+\usetheme{metropolis}
+
+\definecolor{BlueTOL}{HTML}{222255}
+\definecolor{BrownTOL}{HTML}{666633}
+\definecolor{GreenTOL}{HTML}{225522}
+\setbeamercolor{normal text}{fg=BlueTOL,bg=white}
+\setbeamercolor{alerted text}{fg=BrownTOL}
+\setbeamercolor{example text}{fg=GreenTOL}
+
+\metroset{block=fill}
+
+\setbeamercolor{block title alerted}{use=alerted text,
+    fg=alerted text.fg,
+    bg=alerted text.bg!80!alerted text.fg}
+\setbeamercolor{block body alerted}{use={block title alerted, alerted text},
+    fg=alerted text.fg,
+    bg=block title alerted.bg!50!alerted text.bg}
+\setbeamercolor{block title example}{use=example text,
+    fg=example text.fg,
+    bg=example text.bg!80!example text.fg}
+\setbeamercolor{block body example}{use={block title example, example text},
+    fg=example text.fg,
+    bg=block title example.bg!50!example text.bg}
+
+%\author[Paola Arce]{Paola Arce \\ Departamento de Inform\'atica, UTFSM \\ Thesis advisor: Luis Salinas}
+%\date{December 15, 2017}
+%\title[Thesis defense]{Online learning methods for financial time series forecasting}
+%\subtitle{PhD Thesis Defense}
 %
-%    \setbeamercolor*{structure}{fg=myblue1,bg=blue}
-%    \setbeamercolor*{palette primary}{use=structure,fg=white,bg=structure.fg}
-%    \setbeamercolor*{palette secondary}{use=structure,fg=white,bg=structure.fg!75!black}
-%    \setbeamercolor*{palette tertiary}{use=structure,fg=white,bg=structure.fg!50!black}
-%    \setbeamercolor*{palette quaternary}{fg=black,bg=white}
-%
-%    \setbeamercolor*{item projected}{fg=red,bg=myblue3!80}
-%    \setbeamercolor*{block title example}{fg=white,bg=myblue4}
-%    \setbeamercolor*{frametitle}{fg=black}
-%
-%    \setbeamertemplate{blocks}[rounded][shadow=true]
-%
-%    \makeatletter
-%    \pgfdeclarehorizontalshading[frametitle.bg,frametitle right.bg]{beamer@frametitleshade}{\paperheight}{%
-%      color(0pt)=(myblue2);
-%      color(\paperwidth)=(white)}
-%
-%    \defbeamertemplate*{footline}{mysplit theme}
-%    {%
-%      \leavevmode%
-%      \hbox{\begin{beamercolorbox}[wd=.5\paperwidth,ht=2.5ex,dp=1.125ex,leftskip=.3cm plus1fill,rightskip=.3cm]{author in head/foot}%
-%        \usebeamerfont{author in head/foot}\insertshortauthor
-%      \end{beamercolorbox}%
-%      \begin{beamercolorbox}[wd=.5\paperwidth,ht=2.5ex,dp=1.125ex,leftskip=.3cm,rightskip=.3cm plus1fil]{title in head/foot}%
-%        \usebeamerfont{title in head/foot}\insertshorttitle\hfill
-%        \insertframenumber/\inserttotalframenumber\hspace*{0.5em}
-%      \end{beamercolorbox}}%
-%      \vskip0pt%
-%    }
-%    \makeatother
-
-%\usebackgroundtemplate{%             declare it
-%\tikz[overlay,remember picture] \node[opacity=0.2, at=(current page.center)] {
-%   \includegraphics[height=\paperheight,width=\paperwidth]{img/background}};
+%\titlegraphic{
+%    \tikz[overlay,remember picture]
+%        \node[at=(current page.south east), anchor=south east] {
+%            \includegraphics[height=1.5cm]{img/DIT}
+%        };
 %}
@@
 \author[Paola Arce]{Paola Arce \\ Departamento de Inform\'atica, UTFSM \\ Thesis advisor: Luis Salinas}
-\date{December 15, 2017}
-\title[Thesis defense]{Online learning methods for financial time series forecasting}
-\subtitle{PhD Thesis Defense}
+\date{December 11, 2017}
+\title[Thesis defense]{Fast and adaptive cointegration based model for forecasting financial time series}
+\subtitle{Seminario Inform\'atica CCTVal}
 
@@
 \frametitle{HFT Revenue}
-However, the revenues have fallen dramatically mainly because of more regulation, increasing cost of HFT infrastructure and competition.
+However, the revenues have fallen dramatically mainly because of more regulation, increasing cost of HFT infrastructure and more competitive algorithms.
 \begin{figure}
@@
 \item Heteroscedastic
+%\item Volatility
 \end{itemize}
@@
 \begin{block}{Strong definition}
-A strictly stationary time series $y_t \in \mathbb{R}$, $\forall t \in \mathbb{Z}$ is one for which the probabilistic behaviour of every collection of values $\{y_{t_1},y_{t_2},\dots,y_{t_L}\}$ is identical
+A {\color{red}strictly stationary time series} $y_t \in \mathbb{R}$, $\forall t \in \mathbb{Z}$ is one for which the probabilistic behaviour of every collection of values $\{y_{t_1},y_{t_2},\dots,y_{t_L}\}$ is identical
 to that of the time shifted set, more precisely:
-\small
-
 \begin{equation*}
@@
 P\{y_{t_1} \leq
-c_1,..,y_{t_L} \leq c_L\} = P\{y_{t_1+h} \leq c_1,..,y_{t_L+h} \leq c_L\}\enspace }\forall L \in \mathbb{N}, \forall h \in \mathbb{Z}
+c_1,..,y_{t_L} \leq c_L\} = P\{y_{t_1+h} \leq c_1,..,y_{t_L+h} \leq c_L\}\enspace }
 \end{equation*}
-\noindent where $c_1,\dots,c_L$ are constants.
+\noindent $\forall L \in \mathbb{N}, \forall h \in \mathbb{Z}$ where $c_1,\dots,c_L$ are constants.
 \end{block}
@@
 \frametitle{Heteroscedastic}
-%HASTA AQUI
+Conditional heteroskedasticity identifies non-constant volatility.
+Volatility gives a measure of uncertainty about future returns. Many factors affect volatility such as: historical prices, trends, monetary policies and news.
+\begin{figure}
+\includegraphics[width=0.8\paperwidth]{img/vix}
+\end{figure}
 \end{frame}

### 2017-12-09 slides/defense-slides.tex
primera contribucion lista

@@
 %\usepackage[spanish,activeacute]{babel}
+\usetheme{metropolis}
 \usepackage[utf8]{inputenc}
@@
 \usepackage{tikz}
+\usetikzlibrary{positioning}
 \usepackage{array}
@@
 
-\usetheme{metropolis}
-
 \definecolor{BlueTOL}{HTML}{222255}
-\definecolor{BrownTOL}{HTML}{666633}
+\definecolor{BrownTOL}{HTML}{f43d00}%666633}
 \definecolor{GreenTOL}{HTML}{225522}
@@
 \setbeamercolor{block body alerted}{use={block title alerted, alerted text},
-    fg=alerted text.fg,
+    fg=normal text.fg,
     bg=block title alerted.bg!50!alerted text.bg}
@@
 
-\author[Paola Arce]{Paola Arce \\ Departamento de Inform\'atica, UTFSM \\ Thesis advisor: Luis Salinas}
+\author[Paola Arce]{Paola Arce \\ Thesis advisor: Luis Salinas}
 \date{December 11, 2017}
@@
 \subtitle{Seminario Inform\'atica CCTVal}
-
-\titlegraphic{
-    \tikz[overlay,remember picture]
-        \node[at=(current page.south east), anchor=south east] {
-            \includegraphics[height=1.5cm]{img/DIT}
-        };
-}
 
@@
 \titlepage
+\begin{tikzpicture}[overlay, remember picture]
+\node[above right=0.4cm and 1.2cm of current page.south west] {\includegraphics[width=2.5cm]{img/UTFSM}};
+%\node[above =0.5cm of current page.south] {\includegraphics[width=2cm]{img/DIT}};
+\node[above left=0.5cm and .8cm of current page.south east] {\includegraphics[width=5cm]{img/logo-di}};
+\end{tikzpicture}
+
 \end{frame}
@@
 \begin{itemize}
-\item Motivation: Financial time series and HFT
-\item Cointegration and Integration concepts
-\item VECM model
-\item Proposal
+\item Motivation: High Frequency trading
+\item Cointegration and the VECM model
+\item Challenges and Hypothesis
+\item Contributions
 \item Experiments
 \item Conclusions
+\item Future work
 \end{itemize}
@@
 
-\section{Introduction}
+\section{Motivation}
 
@@
 \item Heteroscedastic
-%\item Volatility
+\item Potentially cointegrated
 \end{itemize}
@@
 One of the consequences of stationary processes is how they recover from a shock.
-A shock represents an unexpected change in a variable or a in its error term in a particular period of time.
+A shock represents an unexpected change in a variable in a particular period of time.
 For stationary time series, shocks to the system will gradually die away.
+\begin{figure}
+\includegraphics[width=0.6\paperwidth]{img/shock}
+\end{figure}
 \end{frame}
@@
 \frametitle{Non-stationary processes}
-There are different types of non-stationary time series models often found in economics:
-\begin{block}{Stochastic trend}
-Stochastic trend processes are also called {\color{red}unit root or I(1)} if they have the following form:
+There are different types of non-stationary time series models often found in economics. Unit root process is one of them commonly used to model prices:
+\begin{block}{Unit root process}
+{\color{red}Unit root or I(1)} processes are also called stochastic trend if they have the following form:
 {\color{blue}
@@
 \]}
-\noindent where $\epsilon_t$ is a stationary process. When $\mu = 0 $ the process is called pure random walk and when $\mu \neq 0$  the process is called random walk with drift.
+\noindent where $\epsilon_t$ is a stationary process. When $\mu = 0 $ the process is called {\bf pure random walk} and when $\mu \neq 0$  the process is called {\bf random walk with drift}.
 \end{block}
-%\begin{block}{Deterministic trend}
-% Deterministic trend or trend stationary
-%processes have the following form:
-%{\color{blue}
-%\[
-%y_t = f(t) + \epsilon_{t}
-%\]}
-%\noindent where $\epsilon_{t}$ represents a stationary error term (with mean 0 and variance $\sigma^2$) and $f(t)$ is a deterministic function of time.
-%\end{block}
 \end{frame}
 
-%\begin{frame}[t,fragile]{Random walk Illustration}
-% \begin{itemize}
-%  \item<1-> The drunkard's walk
-%  \item<2-> The dog's walk
-%  %\item<3-> The drunk and dog's walk
-% \end{itemize}
-% 
-%\begin{columns}
-%\begin{column}{.4\linewidth}<1->
-%\begin{center}
-%        \begin{tikzpicture}[overlay]
-%            \node[yshift=-50] (img2) {\includegraphics[scale=1.4]{img/drunk}};
-%        \end{tikzpicture}
-%\end{center}
-%\end{column}
-%\begin{column}{.08\linewidth}<2->
-%        \begin{tikzpicture}[overlay]
-%            \node[yshift=-50] (img2) {\includegraphics[scale=0.1]{img/dog}};
-%        \end{tikzpicture}
-%\end{column}
-%%\begin{column}{.5\linewidth}<3->
-%%    \begin{center}
-%%        \begin{tikzpicture}[overlay]
-%%            \node[yshift=-50] (img2) {\includegraphics[scale=0.1]{img/mananddog}};
-%%        \end{tikzpicture}
-%%    \end{center}
-%%\end{column}
-%\end{columns}
-%\end{frame}
-%
-%\begin{frame}[t,fragile]{Random walk and Cointegration Illustration}
-% \begin{itemize}
-%  \item The drunk and dog's walk
-% \end{itemize}
-%    \begin{center}
-%        \begin{tikzpicture}[overlay]
-%            \node[yshift=-60] (img2) {\includegraphics[scale=0.7]{img/mananddog}};
-%        \end{tikzpicture}
-%    \end{center}
-%\end{frame}
+\begin{frame}
+\frametitle{Integration }
+\begin{block}{Integration of order $d$}
+A time series $\mathbf{y}$ is said to be I(d), or integrated of order $d$, if after
+differentiating the variable $d$ times, we get an stationary process, more precisely:
+\[
+{\color{blue}
+(1-L)^d \mathbf{y} \sim \text{I(0)}} \, ,
+\]
+\noindent where I(0) is a stationary time series and $L$ is the lag operator:
+\[
+(1-L)\mathbf{y} = \Delta \mathbf{y}=\mathbf{y}_t  -\mathbf{y}_{t-1} \quad \forall t
+\]
+\end{block}
+\end{frame}
 
@@
 \begin{figure}
-\includegraphics[width=\paperwidth]{img/hft-ticks-zoom.png}
+\includegraphics[width=0.8\paperwidth]{img/hft-ticks-zoom.png}
 \end{figure}
@@
 
+\section{Cointegration and the VECM model}
 \begin{frame}
-\frametitle{Motivation}
-
-The relationship between non-stationary time series can be modelled if a stationary linear combination is shown to exist.
-%Financial time series are mainly non-stationary.  However, one or more linear
-%combinations of non-stationary time series can be stationary even though
-%individually they are not. 
-When this happens it is said they have a long-run
-equilibrium relationship and the variables are {\color{red} cointegrated}.
+\frametitle{Cointegration}
+\begin{figure}
+\includegraphics[width=0.8\paperwidth]{img/motivation}
+\end{figure}
 \end{frame}
@@
 \begin{frame}
-\frametitle{Cointegration}
+\frametitle{Cointegration definition}
+
+\begin{block}{Cointegration}
+The relationship between non-stationary time series can be modelled if a {\bf stationary linear combination} is shown to exist.
+When this happens it is said they have a {\bf long-run
+equilibrium} relationship and the variables are {\color{red} cointegrated}.
+\end{block}
+
+\end{frame}
+
+\begin{frame}
+\frametitle{Cointegration: formal definition}
 \begin{block}{Cointegration definition}
@@
 time series of order I(1) which are said to be cointegrated if a vector
-$\beta=[\beta(1),\dots,\beta(l)]^\top \in \mathbb{R}^l$  exists such that the
+${\color{blue}\boldsymbol\beta=[\beta(1),\dots,\beta(l)]^\top} \in \mathbb{R}^l$  exists such that the
 time series,
@@
 \begin{equation*}
- \mathbf{Z}_t:= \beta^\top \mathbf{y} = \beta(1) \mathbf{y}^1 + \dots + \beta(l) \mathbf{y}^l \sim
+ \mathbf{Z}_t:= \boldsymbol \beta^\top \mathbf{y} = \beta(1) \mathbf{y}^1 + \dots + \beta(l) \mathbf{y}^l \sim
  \text{I(0)}\, .
@@
 \end{block}
-In other words, a set of I(1) variables is said to be cointegrated if
-a linear combination of them exists which is I(0).
+%In other words, a set of I(1) variables is said to be cointegrated if
+%a linear combination of them exists which is I(0).
 
@@
 
-The Vector Error correction model (VECM) is used to model cointegrated time series.
+The Vector Error correction model (VECM) is used to model cointegrated time series:
 
@@
 }
-The columns of $\beta$ contains the cointegration vectors and the rows of
-$\alpha$ correspond with the adjusted vectors. $\beta$ is obtained by Johansen
-procedure whereas $\alpha$ has to be determined as a
-variable in the VECM.
-
+\noindent where $p$ is the number of lags and $\boldsymbol \Omega = \boldsymbol \alpha \boldsymbol \beta^\top$.
+The columns of $\boldsymbol\beta$ contains the cointegration vectors and the rows of $\boldsymbol\alpha$ correspond with the adjusted vectors.
 \end{frame}
@@
 \begin{frame}
-\frametitle{VECM matrix form}
-\small
-\begin{equation*}
-\underbrace{
-      \begin{bmatrix}
-       \mathbf{\Delta y}_{p+1}  \\ 
-       \mathbf{\Delta y}_{p+2}  \\ 
-       \vdots                   \\ 
-       \mathbf{\Delta y}_N      
-      \end{bmatrix}}_{\mathbf{B} } =
-\arraycolsep=1.4pt  
-\underbrace{\begin{bmatrix}
-   \alpha^\top & \phi^*_1 & \dots & \phi^*_{p-1} &\mathbf{c}   
-  \end{bmatrix}}_{\mathbf{X}^\top}
-\underbrace{\begin{bmatrix}
- \beta\mathbf{y}_p      & \beta\mathbf{y}_{p+1}   & \dots&   \beta\mathbf{y}_{N-1}   \\
- \mathbf{\Delta y}_{p}   & \mathbf{\Delta y}_{p+1}& \dots &  \mathbf{\Delta y}_{N-1} \\
- \mathbf{\Delta y}_{p-1} & \mathbf{\Delta y}_{p}  & \dots &  \mathbf{\Delta y}_{N-2}   \\
- \vdots                  & \vdots                 & \ddots&  \vdots                   \\
- \mathbf{\Delta y}_{2}   & \mathbf{\Delta y}_{3} & \dots &   \mathbf{\Delta y}_{N-p+1} \\
- 1                      & 1                       & \dots     & 1   
- \end{bmatrix}}_{\mathbf{A}^\top}
-+
-\underbrace{\begin{bmatrix}
-              \mathbf{\epsilon}_{p+1} \\ 
-              \mathbf{\epsilon}_{p+2} \\ 
-              \vdots \\ 
-              \mathbf{\epsilon}_N
-             \end{bmatrix}}_{\mathbf{E}^\top}
-\end{equation*}
-\end{frame}
-
-\begin{frame}
-\frametitle{$\Omega$ matrix properties}
-The matrix $\Omega$ has the following properties:
+\frametitle{$\boldsymbol \Omega$ matrix properties}
+The matrix $\boldsymbol \Omega$ has the following properties:
 \begin{itemize}
-\item If $\Omega = 0$ there is no cointegration
-\item If $rank(\Omega)=l$ i.e full rank, then the time series are not
+\item If ${\color{blue}\boldsymbol \Omega = 0}$ there is no cointegration
+\item If ${\color{blue}rank(\boldsymbol\Omega)=l}$ i.e full rank, then the time series are not
 I(1) but stationary
-\item If $rank(\Omega)=r,\quad 0 < r < l$ then, there is cointegration
-and the matrix $\Omega$ can be expressed as $\Omega =
-\alpha \beta^\top$, where $\alpha$ and $\beta$ are $(l \times r)$
-matrices and $rank(\alpha)=rank(\beta)=r$.
+\item If ${\color{blue}rank(\boldsymbol\Omega)=r,\quad 0 < r < l}$ then, there is cointegration.
+%where $\boldsymbol\alpha$ and $\boldsymbol\beta$ are $(l \times r)$
+%matrices and $rank(\boldsymbol\alpha)=rank(\boldsymbol\beta)=r$.
 \end{itemize}
@@
 \frametitle{VECM limitations}
-The Vector Error correction model (VECM) is used to model cointegrated time series but only with batch data and rarely used with high frequency data  mainly due to computational limitations:
+
+\begin{alertblock}{VECM limitations}
+The Vector Error correction model is used to model cointegrated time series but only with {\bf batch data and rarely used with high frequency data} mainly due to {\bf computational limitations}:
 \begin{itemize}
-\item Calculation of the cointegration vectors
-\item VECM parameters are obtained using the ordinary least squares (OLS) method.
+\item Calculation of the cointegration vectors $\boldsymbol \beta$ is obtained by the Johansen procedure which is of order $O(n^3)$.
+%whereas $\alpha$ has to be determined as a variable in the VECM.
+\item VECM parameters are obtained using the ordinary least squares ({\bf OLS}) method.
 \end{itemize}
+\end{alertblock}
 \end{frame}
 
-\section{Proposal}
+\section{Challenges and Hypothesis}
 \begin{frame}
-\frametitle{Proposal}
+\frametitle{Challenge}
 \begin{block}{Proposal}
@@
 \end{block}
-\begin{block}{Hypothesis}
+\begin{exampleblock}{Hypothesis}
 An algorithm based on cointegration and high performance computing will allow faster forecasting
 algorithms for financial time series to be obtained while maintaining good accuracy levels.
-\end{block}
+\end{exampleblock}
 \end{frame}
 
-\section{First Contribution}
+\section{Contributions}
 \begin{frame}
-\frametitle{First approach}
-An adaptive Vector Error Correction algorithm that:
+\frametitle{First contribution: Adaptive VECM}
+\begin{figure}
+\includegraphics[width=\textwidth]{img/slidingwindow}
+\end{figure}
+\end{frame}
+
+\begin{frame}
+\frametitle{First approach idea}
+An adaptive Vector Error Correction (AVECM) algorithm that:
 \begin{itemize}
@@
 \begin{frame}
-\frametitle{First approach idea}
-\begin{figure}
-\includegraphics[width=0.65\textwidth]{img/slidingwindow}
-\end{figure}
-\end{frame}
-
-\begin{frame}
 \frametitle{How to choose L?}
@@
   \centering
-   \includegraphics[width=0.65\textwidth]{img/51_Fig1}
-  \caption[Distribution of the number of cointegration vectors using $p=1$ lags]{Distribution of the number of cointegration vectors using $p=1$ lags.
-  Four possible values for windows size $L$ are shown (2, 6, 10 and 14) hours (1
-  hour = 360 data points).}
+   \includegraphics[width=0.85\textwidth]{img/51_Fig1}
+  \caption[Distribution of the number of cointegration vectors using $p=1$ lags]{Distribution of the number of cointegration vectors using $p=1$ lags.}
+  %Four possible values for windows size $L$ are shown (2, 6, 10 and 14) hours (1 hour = 360 data points).}
   \label{fig:hists}
-\end{figure}
-\end{frame}
-
-\subsection{Data}
-\begin{frame}
-\frametitle{Data}
-\begin{itemize}
-\item Tests were carried out using four foreign exchange rates all related to
-USD: EURUSD, GBPUSD, USDCHF and USDJPY. 
-\item This data was collected from Dukascopy, a free
-database which gives access to the Swiss Foreign Exchange marketplace.
-\item The tests were done using 10-seconds frequency from ask prices
-from the 11th to the 15th of August 2014. 
-%\item Since one day corresponds to 8640 data points and we used 5 days of data, we have 43,200 data points in total.
-\end{itemize}
- \begin{figure}[!h]
-    \includegraphics[width=0.2\textwidth]{img/dukascopy}
 \end{figure}
@@
 
-%\begin{frame}
-%\frametitle{PC vs MSE}
-%\begin{figure}[ht!]
-%  %\vspace{-0.8cm}
-%  \centering
-%  \includegraphics[width=0.7\textwidth]{img/MSE-offset20520-p-2-freq-10s}
-%  \caption{MSE versus the percentage of cointegration considering 1000
-%  iterations. }
-%  \label{fig:cointvsmse}
-%\end{figure}
-%\end{frame}
+\begin{frame}
+\frametitle{Data}
+\begin{columns}
+\begin{column}{0.7\textwidth}
+\begin{itemize}
+\item This data was collected from {\bf Dukascopy}, a free
+database which gives access to the {\bf Swiss Foreign Exchange marketplace}.
+\item Tests were carried out using four foreign exchange rates all related to
+USD: {\bf EURUSD, GBPUSD, USDCHF} and {\bf USDJPY}. 
+\item The tests were done using 10-seconds frequency from ask prices
+from the 11th to the 15th of August 2014. 
+%\item Since one day corresponds to 8640 data points and we used 5 days of data, we have 43,200 data points in total.
+\end{itemize}
+\end{column}
+\begin{column}{0.3\textwidth}
+ \begin{figure}[!h]
+    \includegraphics[width=\textwidth]{img/dukascopy}
+ \end{figure}
+\end{column}
+\end{columns}
+\end{frame}
+
+\begin{frame}
+\frametitle{PC vs MSE}
+\begin{figure}[ht!]
+  %\vspace{-0.8cm}
+  \centering
+  \includegraphics[width=0.7\textwidth]{img/51_Fig2}
+  \caption{MSE versus the percentage of cointegration considering 1000
+  iterations. }
+  \label{fig:cointvsmse}
+\end{figure}
+\end{frame}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%
 %
-\subsection{AVECM}
 \begin{frame}
@@
 
-\section{Conclusions}
-\subsection{Uno}
 \begin{frame}
@@
 \begin{frame}
-\frametitle{Integration }
-\begin{block}{Integration of order $d$}
-A time series $\mathbf{y}$ is said to be integrated of order $d$ if after
-differentiating the variable $d$ times, we get an stationary process, more precisely:
-\[
-(1-L)^d \mathbf{y} \sim \text{I(0)} \, ,
-\]
-\noindent where I(0) is a stationary time series and $L$ is the lag operator:
-\[
-(1-L)\mathbf{y} = \Delta \mathbf{y}=\mathbf{y}_t  -\mathbf{y}_{t-1} \quad \forall t
-\]
-\end{block}
+\frametitle{VECM(p) matrix form}
+If we have $N$ data points, with $N>p$, we can estimate the model parameters $\mathbf{X}$ using the following matrix form:
+\small
+\begin{equation*}
+\underbrace{
+      \begin{bmatrix}
+       \mathbf{\Delta y}_{p+1}  \\ 
+       \mathbf{\Delta y}_{p+2}  \\ 
+       \vdots                   \\ 
+       \mathbf{\Delta y}_N      
+      \end{bmatrix}}_{\mathbf{B} } =
+\arraycolsep=1.4pt  
+\underbrace{\begin{bmatrix}
+   \alpha^\top & \phi^*_1 & \dots & \phi^*_{p-1} &\mathbf{c}   
+  \end{bmatrix}}_{\mathbf{X}^\top}
+\underbrace{\begin{bmatrix}
+ \beta\mathbf{y}_p      & \beta\mathbf{y}_{p+1}   & \dots&   \beta\mathbf{y}_{N-1}   \\
+ \mathbf{\Delta y}_{p}   & \mathbf{\Delta y}_{p+1}& \dots &  \mathbf{\Delta y}_{N-1} \\
+ \mathbf{\Delta y}_{p-1} & \mathbf{\Delta y}_{p}  & \dots &  \mathbf{\Delta y}_{N-2}   \\
+ \vdots                  & \vdots                 & \ddots&  \vdots                   \\
+ \mathbf{\Delta y}_{2}   & \mathbf{\Delta y}_{3} & \dots &   \mathbf{\Delta y}_{N-p+1} \\
+ 1                      & 1                       & \dots     & 1   
+ \end{bmatrix}}_{\mathbf{A}^\top}
++
+\underbrace{\begin{bmatrix}
+              \mathbf{\epsilon}_{p+1} \\ 
+              \mathbf{\epsilon}_{p+2} \\ 
+              \vdots \\ 
+              \mathbf{\epsilon}_N
+             \end{bmatrix}}_{\mathbf{E}^\top}
+\end{equation*}
 \end{frame}

### 2017-12-10 slides/defense-slides.tex
nocite references, first contribution ready

@@
 \frametitle{First contribution: Adaptive VECM}
-\begin{figure}
-\includegraphics[width=\textwidth]{img/slidingwindow}
-\end{figure}
-\end{frame}
-
-\begin{frame}
-\frametitle{First approach idea}
-An adaptive Vector Error Correction (AVECM) algorithm that:
-\begin{itemize}
-\item considers the last most recent data (windows size L)
-\item finds optimal parameters of VECM in that window.
-\item the optimisation is made in a distributed environment.
-\end{itemize}
+\hspace*{-9mm}
+\includegraphics[width=1.15\textwidth]{img/slidingwindow}
 \end{frame}
@@
 between $0$ and $l$, i.e. $0<r<l$.
+\vspace{5mm}
 \begin{block}{Percentage of cointegration (PC)}
@@
 \end{frame}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%
+
 \begin{frame}
 \frametitle{Adaptive VECM algorithm (AVECM)}
-\begin{itemize}
-\item We proposed to choose $L$ and $p$ in order to maximise the percentage of
-cointegration $PC$ in the near past. This process was done every time that new data was processed. 
+AVECM has the following features: 
+\begin{enumerate}
+\item Uses VECM only considering a sliding windows of size $L$ of data.
+\item We proposed to choose $L$ and the number of lags $p$ in order to maximise the percentage of cointegration $PC$ in the near past. This process was done every time that new data was processed. 
 \item To use a distributed environment to make this parameters search, since this is the most expensive routine.
-\end{itemize}
+\end{enumerate}
 \end{frame}
@@
 
-%\subsection{Model comparison} 
-%\begin{frame}
-%\frametitle{Random walk}
-%The random walk model is defined as:
-%\begin{equation}
-%\mathbf{y}_t = \mathbf{y}_{t-1} + \epsilon_{t}
-%\label{rwmodel}
-%\end{equation}
-%The naive forecast of the time series difference $\hat{\mathbf{y}}_{t+1}$ for the random walk model is defined as:
-%\begin{equation}
-%\hat{\mathbf{y}}_{t+1} = \mathbf{y}_t + \hat{\epsilon}_{t+1} 
-%\end{equation}
-%\noindent where  $\hat{\epsilon}_{t+1} = \epsilon_{t}$.
-%\end{frame}
-%
-%
-%\begin{frame}
-%\frametitle{ARIMA}
-%A process can be modelled as an ARIMA$(p,d,q)$ if $\mathbf{x}_t = \Delta^d \mathbf{y}_t $, is an ARMA$(p,q)$. An ARMA$(p,q)$ model is the following:
-%\begin{equation}
-%\mathbf{x}_t = \sum_{i=1}^p \phi_i \mathbf{x}_{t-i}  +  \sum_{j=1}^q \theta_j \epsilon_{t-j}  
-%\end{equation}
-%\noindent with coefficients $\phi_p \neq 0$, $\theta_q \neq 0$ and $\sigma_{\epsilon}^2 > 0$.
-%\end{frame}
-
 %%\begin{frame}
@@
 \frametitle{Unit root tests}
-VECM can only be used if the time series are I(1). Therefore, 
-before running the tests, we firstly checked whether the time series were
-I(1) using the Augmented Dickey Fuller (ADF) test with lags $p=1,2,3,4,5$.
+Augmented Dickey Fuller (ADF) test with lags $p=1,2,3,4,5$.
 \begin{table}[ht]
 \centering
-\tiny
+\small
 \begin{tabular}{llllll}
@@
 \addlinespace[1ex]
-\multicolumn{6}{l}{ \textsuperscript{*} Indicates significance at 1\% level}
 \end{tabular}
@@
 \frametitle{Parallel implementation}
-The function that get best parameters $L$ and $p$ was implemented using MPI in Python. MPI allows large-scale parallel applications
-with wide portability to be built, being able to run in large clusters
-or on local computers.
-Tests ran in a cluster with 2 servers Xeon E5-2667 (2.90GHz)
-of 24 cores each (48 cores in total) and 24GB RAM.
+The function that get optimal parameters $L$ and $p$ was implemented using MPI in Python.
+Tests ran in a cluster with 2 servers Xeon E5-2667 (2.90GHz) of 24 cores each (48 cores in total) and 24GB RAM.
+
 Parameter settings:
@@
 %\item Parameter $it=100$
-\item The $L$ parameter was always chosen between 100 and 4000 and $p$ always took values
-between 1 and 5. 
+\item The $L$ parameter was always chosen between 100 and 4000 and $p$ always took values between 1 and 5. 
 \item Parameter $nparams$ represents the number of pairs ($L$,$p$) used to maximise the percentage of cointegration. 
@@
   \centering
-  \includegraphics[width=0.7\textwidth]{img/51_Fig3}
+  \includegraphics[width=0.9\textwidth]{img/51_Fig3}
   \caption[Computing time and Speed-up]{Computing time of sequential and parallel algorithm is shown in the
@@
 \item Cointegration relations change with time. We empirically showed that the Johansen method is sensitive to the number of lags but also to the amount of data considered.
-\item We found that out-of-sample forecast performance MSE is related to the {\em percentage of cointegration\/}.  
+%\item We found that out-of-sample forecast performance MSE is related to the {\em percentage of cointegration\/}.  
 \item VECM improves performance measures by finding
 parameters of $L$ and $p$ maximising the percentage of cointegration.
-\item Increasing the number of parameters will always lead to
-better performance measures.
+%\item Increasing the number of parameters will always lead to
+%better performance measures.
 \item The parallel implementation allowed the execution times to be reduced
@@
 
+\nocite{arce+salinas2012,icpram15,Arce2017}
+\bibliographystyle{apalike}
+\bibliography{reference}
+
 %\begin{frame}
@@
 %9 Roadmap of Thesis
+
+\appendix
 
@@
 
+%\subsection{Model comparison} 
+\begin{frame}
+\frametitle{Random walk}
+The random walk model is defined as:
+\begin{equation}
+\mathbf{y}_t = \mathbf{y}_{t-1} + \epsilon_{t}
+\label{rwmodel}
+\end{equation}
+The naive forecast of the time series difference $\hat{\mathbf{y}}_{t+1}$ for the random walk model is defined as:
+\begin{equation}
+\hat{\mathbf{y}}_{t+1} = \mathbf{y}_t + \hat{\epsilon}_{t+1} 
+\end{equation}
+\noindent where  $\hat{\epsilon}_{t+1} = \epsilon_{t}$.
+\end{frame}
+%
+%
+\begin{frame}
+\frametitle{ARIMA}
+A process can be modelled as an ARIMA$(p,d,q)$ if $\mathbf{x}_t = \Delta^d \mathbf{y}_t $, is an ARMA$(p,q)$. An ARMA$(p,q)$ model is the following:
+\begin{equation}
+\mathbf{x}_t = \sum_{i=1}^p \phi_i \mathbf{x}_{t-i}  +  \sum_{j=1}^q \theta_j \epsilon_{t-j}  
+\end{equation}
+\noindent with coefficients $\phi_p \neq 0$, $\theta_q \neq 0$ and $\sigma_{\epsilon}^2 > 0$.
+\end{frame}
+
 % Financial time series are not the only ones with this behaviour, this study can be extended to other non-stationary time series such as: weather, earthquakes, energy demand and sales forecasting.
 
-\bibliographystyle{amscls/amsalpha}
-\bibliography{reference}
 \end{document}

### 2017-12-10 slides/defense-slides.tex
before adding second contribution

@@
 
+\begin{frame}
+\frametitle{Online VECM}
+Recently, online learning algorithms have been proposed to solve problems with large data sets main because: 
+\begin{itemize}
+\item Their simplicity
+\item Process one instance at a time
+\item The hypothesis is updated every new data arrives
+\end{itemize} 
+\end{frame}
+
+\begin{frame}
+\frametitle{Online learning algorithms}
+\begin{algorithm}[H]
+\begin{algorithmic}[1]
+    \STATE Receives input $\mathbf{x}_t$
+    \STATE Makes prediction $\mathbf{\hat{y}}_t$
+    \STATE Receives response $\mathbf{y}_t$
+    \STATE Incurs loss $l_t(\mathbf{y}_t,\mathbf{\hat{y}}_t)$
+\end{algorithmic}
+\caption{Structure of an Online Learning System}
+\end{algorithm}
+Performance is later measured after $T$ trials as:
+\begin{equation*}
+L_T = \sum_{t=1}^T l_t(\mathbf{y}_t,\mathbf{\hat{y}}_t)
+\end{equation*}
+\end{frame}
+
 \section{Challenges and Hypothesis}
@@
 \begin{frame}
-\frametitle{First contribution: Adaptive VECM}
+\frametitle{Adaptive VECM}
 \hspace*{-9mm}
@@
 
-\section{Second Contribution}
-
-\begin{frame}
-\frametitle{Second approach}
- Recently, online learning algorithms have been proposed to solve problems with large data sets because of their simplicity and their ability to update the model when new data is
-available.
-\begin{figure}
-\includegraphics[width=0.6\paperwidth]{img/onlinelearning}
-\end{figure} 
-\end{frame}
-
-\begin{frame}
-\frametitle{Proposal: Online VECM algorithm (OVECM)}
-\begin{itemize}
-\item Considers only a sliding window of the $L$ most recent data. 
-\item Since cointegration vectors represent long-run relationships they vary
-little in time. OVECM obtains these vectors only when they need to be updated. 
-\item This updating is done when MAPEs of the last $n$ inputs is above a certain
-error given by a threshold 
-\item OVECM also implements matrix optimisations in order to reduce execution time,
-such as updating matrices with new data, removing old data and introducing new
-cointegration vectors.
-\end{itemize}
-\end{frame}
-
-\begin{frame}
-\frametitle{Methodology}
-OVECM were compared against VECM and ARIMA. VECM and ARIMA are called with a
-sliding window of the most recent data, whereby the models are updated at every
-time step. Therefore, we named them SLVECM and SLARIMA.
-\end{frame}
-
-\subsection{Parameters}
-\begin{frame}
-\frametitle{Parameters optimization}
-VECM order and ARIMA parameters were selected using AIC.
-\begin{table}[ht]
-\label{tab:params}
-\begin{center}
-\begin{tabular}{|c|c|c|}
-\hline
-Windows size $L$ & VECM & ARIMA\\
- $L$ & order $(p)$ & order $(p,d,q)$ \\
-\hline
-100 & 2 & p=2,d=1,q=1\\
-400 & 5 & p=1,d=1,q=1\\
-700 & 3 &p=2,d=1,q=1\\
-1000 &3 & p=2,d=1,q=1\\
-\hline
-\end{tabular}
-\end{center}
-\end{table}
-\end{frame}
-
-\begin{frame}
-\frametitle{Execution times}
-We ran 400 iterations of OVECM and SLVECM . SLARIMA was based on the ARIMA procedure
-of the statsmodels library and therefore its execution time is excluded.
-%because its is not comparable with OVECM and SLVECM. SLARIMA was created based
-%on statsmodels library routine ARIMA.
-\begin{center}
-\adjustbox{max height=\dimexpr\textheight-5.5cm\relax,
-           max width=\textwidth}{
-\begin{tabular}{|l|c|c|c|c|c|}
-\hline
-& $L$ & order & e  & Time[s] \\
-\hline
-OVECM & 100 &p=2  & 0      & 2.492\\
-OVECM & 100 &p=2  & 0.0026  & 1.606\\
-SLVECM & 100 &p=2& -- & 2.100\\
-\hline
-OVECM & 400 & p=5  & 0      & 3.513\\
-OVECM & 400 &p=5  & 0.0041  & 2.569\\
-SLVECM & 400 & p=5 & -- & 3.222\\
-\hline
-OVECM & 700 &p=3  & 0      & 3.296\\
-OVECM & 700 &p=3  & 0.0032  & 2.856\\
-SLVECM & 700 &p=3 & -- & 3.581\\
-\hline
-OVECM & 1000 & p=3 & 0      & 4.387\\
-OVECM & 1000 & p=3  & 0.0022  & 2.408\\
-SLVECM & 1000 & p=3  & -- & 3.609\\
-\hline
-\end{tabular}
-}\end{center}
-\end{frame}
-
-\begin{frame}
-\frametitle{Performance accuracy}
-\tiny
-\begin{table}[H]
-\caption{Model measures}
-\label{tab:stats}
-\begin{center}
-\begin{tabular}{|l|l|c|c|c|c|c|c|c|}
-\hline
-\multicolumn{3}{|c|}{Model} & \multicolumn{3}{|c|}{In-sample} &
-\multicolumn{3}{|c|}{Out-of-sample} \\ 
-\hline
-\hline
-Method & $L$ & Parameters &  
-MAPE & MAE& RMSE& 
-MAPE & MAE& RMSE \\
-\hline
- OVECM  &   100  &  P=2 e=0.0026 &  0.00263&  0.00085&  0.00114&  0.00309&  0.00094&  0.00131\\
- OVECM  &   400  &  P=5 e=0.0041&  0.00378&  0.00095&  0.00127&  0.00419&  0.00103&  0.00143\\
- OVECM  &   700  &  P=3 e=0.0032&  0.00323&  0.00099&  0.00130&  0.00322&  0.00097&  0.00132\\
- OVECM  &   1000 &  P=3 e=0.0022&  
- \textbf{0.00175}&  \textbf{0.00062}&  \textbf{0.00087} &  
- \textbf{0.00172}&  \textbf{0.00061}&  \textbf{0.00090}\\
-\hline
- SLVECM  &   100 &  P=2 &  0.00262&  0.00085&  0.00113&  0.00310&  0.00095&  0.00132\\
- SLVECM  &   400 &  P=5 &  0.00375&  0.00095&  0.00126&  0.00419&  0.00103&  0.00143\\
- SLVECM  &   700 &  P=3 &  0.00324&  0.00099&  0.00130&  0.00322&  0.00098&  0.00132\\
- SLVECM  &   1000&  P=3 &  
- \textbf{0.00174}&  \textbf{0.00061}&  \textbf{0.00087}&  
- \textbf{0.00172}&  \textbf{0.00061}&  \textbf{0.00090}\\
-\hline
-SLARIMA & 100  &p=2 d=1 q=1 & 0.00285  &  0.00110  &  0.00308  &  0.00312  &0.00098  &  0.00144 \\
-SLARIMA & 400  &p=1 d=1 q=1 & 0.00377  &  0.00101  &  0.00128  &  0.00418 & 0.00106 & 0.00145 \\
-SLARIMA & 700  &p=2 d=1 q=1 & 0.00329  &  0.00102  &  0.00136  &  0.00324  & 0.00097  &  0.00133 \\
-SLARIMA & 1000 &p=2 d=1 q=1 & \textbf{0.00281}  & \textbf{0.00074}  &  \textbf{0.00105}  &  \textbf{0.00177} & \textbf{0.00063}  &  \textbf{0.00092} \\
-\hline
-\end{tabular}
-\end{center}
-\end{table}
-\end{frame}
+%\begin{frame}
+%\frametitle{Proposal: Online VECM algorithm (OVECM)}
+%\begin{itemize}
+%\item Considers only a sliding window of the $L$ most recent data. 
+%\item Since cointegration vectors represent long-run relationships they vary
+%little in time. OVECM obtains these vectors only when they need to be updated. 
+%\item This updating is done when MAPEs of the last $n$ inputs is above a certain
+%error given by a threshold 
+%\item OVECM also implements matrix optimisations in order to reduce execution time,
+%such as updating matrices with new data, removing old data and introducing new
+%cointegration vectors.
+%\end{itemize}
+%\end{frame}
+%
+%\begin{frame}
+%\frametitle{Methodology}
+%OVECM were compared against VECM and ARIMA. VECM and ARIMA are called with a
+%sliding window of the most recent data, whereby the models are updated at every
+%time step. Therefore, we named them SLVECM and SLARIMA.
+%\end{frame}
+%
+%\subsection{Parameters}
+%\begin{frame}
+%\frametitle{Parameters optimization}
+%VECM order and ARIMA parameters were selected using AIC.
+%\begin{table}[ht]
+%\label{tab:params}
+%\begin{center}
+%\begin{tabular}{|c|c|c|}
+%\hline
+%Windows size $L$ & VECM & ARIMA\\
+% $L$ & order $(p)$ & order $(p,d,q)$ \\
+%\hline
+%100 & 2 & p=2,d=1,q=1\\
+%400 & 5 & p=1,d=1,q=1\\
+%700 & 3 &p=2,d=1,q=1\\
+%1000 &3 & p=2,d=1,q=1\\
+%\hline
+%\end{tabular}
+%\end{center}
+%\end{table}
+%\end{frame}
+%
+%\begin{frame}
+%\frametitle{Execution times}
+%We ran 400 iterations of OVECM and SLVECM . SLARIMA was based on the ARIMA procedure
+%of the statsmodels library and therefore its execution time is excluded.
+%%because its is not comparable with OVECM and SLVECM. SLARIMA was created based
+%%on statsmodels library routine ARIMA.
+%\begin{center}
+%\adjustbox{max height=\dimexpr\textheight-5.5cm\relax,
+%           max width=\textwidth}{
+%\begin{tabular}{|l|c|c|c|c|c|}
+%\hline
+%& $L$ & order & e  & Time[s] \\
+%\hline
+%OVECM & 100 &p=2  & 0      & 2.492\\
+%OVECM & 100 &p=2  & 0.0026  & 1.606\\
+%SLVECM & 100 &p=2& -- & 2.100\\
+%\hline
+%OVECM & 400 & p=5  & 0      & 3.513\\
+%OVECM & 400 &p=5  & 0.0041  & 2.569\\
+%SLVECM & 400 & p=5 & -- & 3.222\\
+%\hline
+%OVECM & 700 &p=3  & 0      & 3.296\\
+%OVECM & 700 &p=3  & 0.0032  & 2.856\\
+%SLVECM & 700 &p=3 & -- & 3.581\\
+%\hline
+%OVECM & 1000 & p=3 & 0      & 4.387\\
+%OVECM & 1000 & p=3  & 0.0022  & 2.408\\
+%SLVECM & 1000 & p=3  & -- & 3.609\\
+%\hline
+%\end{tabular}
+%}\end{center}
+%\end{frame}
+%
+%\begin{frame}
+%\frametitle{Performance accuracy}
+%\tiny
+%\begin{table}[H]
+%\caption{Model measures}
+%\label{tab:stats}
+%\begin{center}
+%\begin{tabular}{|l|l|c|c|c|c|c|c|c|}
+%\hline
+%\multicolumn{3}{|c|}{Model} & \multicolumn{3}{|c|}{In-sample} &
+%\multicolumn{3}{|c|}{Out-of-sample} \\ 
+%\hline
+%\hline
+%Method & $L$ & Parameters &  
+%MAPE & MAE& RMSE& 
+%MAPE & MAE& RMSE \\
+%\hline
+% OVECM  &   100  &  P=2 e=0.0026 &  0.00263&  0.00085&  0.00114&  0.00309&  0.00094&  0.00131\\
+% OVECM  &   400  &  P=5 e=0.0041&  0.00378&  0.00095&  0.00127&  0.00419&  0.00103&  0.00143\\
+% OVECM  &   700  &  P=3 e=0.0032&  0.00323&  0.00099&  0.00130&  0.00322&  0.00097&  0.00132\\
+% OVECM  &   1000 &  P=3 e=0.0022&  
+% \textbf{0.00175}&  \textbf{0.00062}&  \textbf{0.00087} &  
+% \textbf{0.00172}&  \textbf{0.00061}&  \textbf{0.00090}\\
+%\hline
+% SLVECM  &   100 &  P=2 &  0.00262&  0.00085&  0.00113&  0.00310&  0.00095&  0.00132\\
+% SLVECM  &   400 &  P=5 &  0.00375&  0.00095&  0.00126&  0.00419&  0.00103&  0.00143\\
+% SLVECM  &   700 &  P=3 &  0.00324&  0.00099&  0.00130&  0.00322&  0.00098&  0.00132\\
+% SLVECM  &   1000&  P=3 &  
+% \textbf{0.00174}&  \textbf{0.00061}&  \textbf{0.00087}&  
+% \textbf{0.00172}&  \textbf{0.00061}&  \textbf{0.00090}\\
+%\hline
+%SLARIMA & 100  &p=2 d=1 q=1 & 0.00285  &  0.00110  &  0.00308  &  0.00312  &0.00098  &  0.00144 \\
+%SLARIMA & 400  &p=1 d=1 q=1 & 0.00377  &  0.00101  &  0.00128  &  0.00418 & 0.00106 & 0.00145 \\
+%SLARIMA & 700  &p=2 d=1 q=1 & 0.00329  &  0.00102  &  0.00136  &  0.00324  & 0.00097  &  0.00133 \\
+%SLARIMA & 1000 &p=2 d=1 q=1 & \textbf{0.00281}  & \textbf{0.00074}  &  \textbf{0.00105}  &  \textbf{0.00177} & \textbf{0.00063}  &  \textbf{0.00092} \\
+%\hline
+%\end{tabular}
+%\end{center}
+%\end{table}
+%\end{frame}
 

### 2017-12-10 slides/defense-slides.tex
Adding online component

@@
 \includegraphics[width=1.15\textwidth]{img/slidingwindow}
+\end{frame}
+
+\begin{frame}
+\frametitle{VECM(p) matrix form}
+If we have $N$ data points, with $N>p$, we can estimate the model parameters $\mathbf{X}$ using the following matrix form:
+\small
+\begin{equation*}
+\underbrace{
+      \begin{bmatrix}
+       \mathbf{\Delta y}_{p+1}  \\ 
+       \mathbf{\Delta y}_{p+2}  \\ 
+       \vdots                   \\ 
+       \mathbf{\Delta y}_N      
+      \end{bmatrix}}_{\mathbf{B} } =
+\arraycolsep=1.4pt  
+\underbrace{\begin{bmatrix}
+   \alpha^\top & \phi^*_1 & \dots & \phi^*_{p-1} &\mathbf{c}   
+  \end{bmatrix}}_{\mathbf{X}^\top}
+\underbrace{\begin{bmatrix}
+ \beta\mathbf{y}_p      & \beta\mathbf{y}_{p+1}   & \dots&   \beta\mathbf{y}_{N-1}   \\
+ \mathbf{\Delta y}_{p}   & \mathbf{\Delta y}_{p+1}& \dots &  \mathbf{\Delta y}_{N-1} \\
+ \mathbf{\Delta y}_{p-1} & \mathbf{\Delta y}_{p}  & \dots &  \mathbf{\Delta y}_{N-2}   \\
+ \vdots                  & \vdots                 & \ddots&  \vdots                   \\
+ \mathbf{\Delta y}_{2}   & \mathbf{\Delta y}_{3} & \dots &   \mathbf{\Delta y}_{N-p+1} \\
+ 1                      & 1                       & \dots     & 1   
+ \end{bmatrix}}_{\mathbf{A}^\top}
++
+\underbrace{\begin{bmatrix}
+              \mathbf{\epsilon}_{p+1} \\ 
+              \mathbf{\epsilon}_{p+2} \\ 
+              \vdots \\ 
+              \mathbf{\epsilon}_N
+             \end{bmatrix}}_{\mathbf{E}^\top}
+\end{equation*}
+\end{frame}
+
+\begin{frame}
+VECM can be solved using OLS. Using a sliding windows of data, the problem can be expressed as follows:
+
+\[
+\mathbf{X}(t) = 
+\begin{bmatrix} 
+\mathbf{x}_{t-L}^\intercal \\ \vdots \\ \mathbf{x}_t^\intercal
+\end{bmatrix} \; , 
+{\bf y}(t) = \begin{bmatrix} y_{t-L} \\ \vdots \\ y_t \end{bmatrix} \; .
+\]
+
+The optimal solution using a sliding windows is then:
+
+\begin{equation}
+{\bf w}(t)_*=({\bf X}(t)^\intercal{\bf X}(t))^{-1}{\bf X}(t)^\intercal{\bf y}(t) \, .
+=\displaystyle \sum_{t=1}^L \mathbf{x}_t
+\mathbf{x}_t^\intercal \displaystyle \sum_{t=1}^Ly_t \mathbf{x}_t
+\end{equation}
+\end{frame}
+
+\begin{frame}
+\begin{algorithm}[H]
+\begin{algorithmic}[1]
+\REQUIRE $\,$ \\
+$\{\mathbf{x}_1,\dots,\mathbf{x}_m \}$: $m$ input vectors \\
+$\{y_1,\dots,y_m \}$: $m$ targets \\
+$L$: sliding window size ($L<m$) \\
+\ENSURE  $\,$ \\
+$\{f(\mathbf{x}_{L+1}),\dots,f(\mathbf{x}_m) \}$: model predictions \\
+\STATE Initialize $\mathbf{A}=\displaystyle \sum_{t=1}^L \mathbf{x}_t
+\mathbf{x}_t^\intercal + \gamma \mathbb{I}$
+and $\mathbf{b}=\displaystyle \sum_{t=1}^Ly_t \mathbf{x}_t$
+\FOR { $t = L+1$ to $m$ }
+        \STATE read new $\mathbf{x}_t$
+        \STATE $\mathbf{A} = \mathbf{A} + \mathbf{x}_t \mathbf{x}_t^\intercal-
+\mathbf{x}_{t-L-1} \mathbf{x}_{t-L-1}^\intercal  $
+        \STATE output prediction $f(\mathbf{x}_t) =  \mathbf{b}^\intercal
+\mathbf{A}^{-1}\mathbf{x}_t$
+        \STATE Read new $y_t$
+        \STATE $\mathbf{b} = \mathbf{b} + y_t \mathbf{x}_t$
+\ENDFOR
+
+\end{algorithmic}
+\caption{Recursive Ridge Regression}
+\label{alg:SLAAR}
+\end{algorithm}
 \end{frame}
@@
 \begin{frame}
-\frametitle{VECM(p) matrix form}
-If we have $N$ data points, with $N>p$, we can estimate the model parameters $\mathbf{X}$ using the following matrix form:
-\small
-\begin{equation*}
-\underbrace{
-      \begin{bmatrix}
-       \mathbf{\Delta y}_{p+1}  \\ 
-       \mathbf{\Delta y}_{p+2}  \\ 
-       \vdots                   \\ 
-       \mathbf{\Delta y}_N      
-      \end{bmatrix}}_{\mathbf{B} } =
-\arraycolsep=1.4pt  
-\underbrace{\begin{bmatrix}
-   \alpha^\top & \phi^*_1 & \dots & \phi^*_{p-1} &\mathbf{c}   
-  \end{bmatrix}}_{\mathbf{X}^\top}
-\underbrace{\begin{bmatrix}
- \beta\mathbf{y}_p      & \beta\mathbf{y}_{p+1}   & \dots&   \beta\mathbf{y}_{N-1}   \\
- \mathbf{\Delta y}_{p}   & \mathbf{\Delta y}_{p+1}& \dots &  \mathbf{\Delta y}_{N-1} \\
- \mathbf{\Delta y}_{p-1} & \mathbf{\Delta y}_{p}  & \dots &  \mathbf{\Delta y}_{N-2}   \\
- \vdots                  & \vdots                 & \ddots&  \vdots                   \\
- \mathbf{\Delta y}_{2}   & \mathbf{\Delta y}_{3} & \dots &   \mathbf{\Delta y}_{N-p+1} \\
- 1                      & 1                       & \dots     & 1   
- \end{bmatrix}}_{\mathbf{A}^\top}
-+
-\underbrace{\begin{bmatrix}
-              \mathbf{\epsilon}_{p+1} \\ 
-              \mathbf{\epsilon}_{p+2} \\ 
-              \vdots \\ 
-              \mathbf{\epsilon}_N
-             \end{bmatrix}}_{\mathbf{E}^\top}
-\end{equation*}
-\end{frame}
-
-\begin{frame}
 \frametitle{Integration example}

### 2017-12-11 slides/defense-slides.tex
First complete version for CCTVal

--- 2017-12-10 slides/defense-slides.tex
@@
 \usepackage{amsmath,amsfonts,amssymb}
+\usepackage{pmat}
 \usepackage{color}
@@
 time series of order I(1) which are said to be cointegrated if a vector
-${\color{blue}\boldsymbol\beta=[\beta(1),\dots,\beta(l)]^\top} \in \mathbb{R}^l$  exists such that the
+${\color{blue}\boldsymbol\beta=[\beta(1),\dots,\beta(l)]^\intercal} \in \mathbb{R}^l$  exists such that the
 time series,
@@
 \begin{equation*}
- \mathbf{Z}_t:= \boldsymbol \beta^\top \mathbf{y} = \beta(1) \mathbf{y}^1 + \dots + \beta(l) \mathbf{y}^l \sim
+ \mathbf{Z}_t:= \boldsymbol \beta^\intercal \mathbf{y} = \beta(1) \mathbf{y}^1 + \dots + \beta(l) \mathbf{y}^l \sim
  \text{I(0)}\, .
@@
 }
-\noindent where $p$ is the number of lags and $\boldsymbol \Omega = \boldsymbol \alpha \boldsymbol \beta^\top$.
+\noindent where $p$ is the number of lags and $\boldsymbol \Omega = \boldsymbol \alpha \boldsymbol \beta^\intercal$.
 The columns of $\boldsymbol\beta$ contains the cointegration vectors and the rows of $\boldsymbol\alpha$ correspond with the adjusted vectors.
@@
 \begin{frame}
+\frametitle{VECM(p) matrix form}
+If we have $N$ data points, with $N>p$, the VECM matrix form is as follows:
+\small
+{\color{blue}
+\begin{equation*}
+\underbrace{
+      \begin{bmatrix}
+   \Delta\mathbf{y}_{p+1}^\intercal \\
+   \Delta\mathbf{y}_{p+2}^\intercal \\
+   \vdots \\
+   \Delta\mathbf{y}_N^\intercal
+   \end{bmatrix}}_{\mathbf{Y} } =
+\arraycolsep=1.4pt
+\underbrace{
+\begin{pmat}[{....|}]
+   \mathbf{y}_p^\intercal \boldsymbol{\beta} & \Delta \mathbf{y}_p^\intercal & \Delta\mathbf{y}_{p-1}^\intercal & \dots 
+                    & \Delta\mathbf{y}_2^\intercal & 1 \cr
+   \mathbf{y}_{p+1}^\intercal  \boldsymbol{\beta} &\Delta\mathbf{y}_{p+1}^\intercal & \Delta\mathbf{y}_p^\intercal & \dots
+                       & \Delta\mathbf{y}_3^\intercal & 1 \cr
+   \vdots & \vdots & \vdots & \ddots & \vdots & \vdots \cr
+   \mathbf{y}_{N-1}^\intercal  \boldsymbol{\beta} &\Delta\mathbf{y}_{N-1}^\intercal & \Delta\mathbf{y}_{N-2}^\intercal & \dots 
+                       & \Delta\mathbf{y}_{N-p-1}^\intercal & 1 \cr
+   \end{pmat}}_{\mathbf{X}}
+\underbrace{
+   \begin{bmatrix}
+   \boldsymbol{\alpha}^\intercal \\
+   \boldsymbol{\Phi}_1^{*\top} \\
+   \boldsymbol{\Phi}_2^{*\top} \\
+   \vdots \\
+   \boldsymbol{\Phi}_{p-1}^{*\top} \\
+   \mathbf{c}^\intercal
+   \end{bmatrix}
+}_{\mathbf{W}}
++
+\underbrace{
+\begin{bmatrix}
+   \boldsymbol{\epsilon}_{p+1}^\intercal \\
+   \boldsymbol{\epsilon}_{p+2}^\intercal \\
+   \vdots \\
+   \boldsymbol{\epsilon}_N^\intercal \\
+   \end{bmatrix}
+}_{\mathbf{E}}
+\end{equation*}}
+
+We can estimate the model parameters $\mathbf{W}$ using OLS:
+\begin{equation}
+{\color{blue}
+\mathbf{Y} = \mathbf{X}\mathbf{W}+\mathbf{E}
+}
+\end{equation}
+
+\end{frame}
+
+\begin{frame}
+\frametitle{Online version of VECM}
+VECM can be solved using a sliding windows of data, the problem can be expressed as follows:
+
+{\color{blue}
+\[
+\mathbf{X}(t) = 
+\begin{bmatrix} 
+\mathbf{x}_{t-L}^\intercal \\ \vdots \\ \mathbf{x}_t^\intercal
+\end{bmatrix} \; , 
+{\bf Y}(t) = \begin{bmatrix} \mathbf{y}^\intercal_{t-L} \\ \vdots \\ \mathbf{y}^\intercal_t \end{bmatrix} \; .
+\]
+}
+
+The optimal solution using a OLS is then:
+
+{\color{blue}
+\begin{eqnarray}
+{\bf W}(t)_*&=&({\bf X}(t)^\intercal{\bf X}(t))^{-1}{\bf X}(t)^\intercal{\bf Y}(t) \\
+&=&\Big( \displaystyle \sum_{i=0}^L \mathbf{x}_{t-i}
+\mathbf{x}_{t-i}^\intercal \Big)^{-1} \displaystyle \sum_{i=0}^L \mathbf{x}_{t-i} \mathbf{y}^\intercal_{t-i}
+\end{eqnarray}
+}
+\end{frame}
+
+\begin{frame}
+\begin{algorithm}[H]
+\begin{algorithmic}[1]
+\REQUIRE $\,$ \\
+$\{\mathbf{x}_1,\dots,\mathbf{x}_N \}$: $N$ input vectors \\
+$\{\mathbf{y}_1,\dots,\mathbf{y}_N \}$: $N$ targets \\
+$L$: sliding window size ($L<N$) \\
+\ENSURE  $\,$ \\
+$\{f(\mathbf{x}_{L+1}),\dots,f(\mathbf{x}_N) \}$: model predictions \\
+\STATE Initialize $\mathbf{A}=\displaystyle \sum_{t=1}^L \mathbf{x}_t
+\mathbf{x}_t^\intercal$
+and $\mathbf{b}=\displaystyle \sum_{t=1}^L \mathbf{x}_t \mathbf{y}^\intercal_{t}$
+\FOR { $t = L+1$ to $N$ }
+        \STATE read new $\mathbf{x}_t$
+        \STATE $\mathbf{A} = \mathbf{A} + \mathbf{x}_t \mathbf{x}_t^\intercal-
+\mathbf{x}_{t-L-1} \mathbf{x}_{t-L-1}^\intercal  $
+        \STATE output prediction $f(\mathbf{x}_t) =  \mathbf{b}^\intercal
+\mathbf{A}^{-1}\mathbf{x}_t$
+        \STATE Read new $y_t$
+        \STATE $\mathbf{b} = \mathbf{b} + y_t \mathbf{x}_t$
+\ENDFOR
+
+\end{algorithmic}
+\caption{Online Ordinary Least Squares}
+\label{alg:SLAAR}
+\end{algorithm}
+\end{frame}
+
+\begin{frame}
 \frametitle{How to choose L?}
@@
 USD: {\bf EURUSD, GBPUSD, USDCHF} and {\bf USDJPY}. 
-\item The tests were done using 10-seconds frequency from ask prices
+\item The tests were done using {\bf 10-seconds frequency} from ask prices
 from the 11th to the 15th of August 2014. 
@@
   \centering
-  \includegraphics[width=0.7\textwidth]{img/51_Fig2}
+  \includegraphics[width=0.8\textwidth]{img/51_Fig2}
   \caption{MSE versus the percentage of cointegration considering 1000
@@
 \end{enumerate}
-\end{frame}
-
-\begin{frame}
-\frametitle{AVECM Algorithm}
-\small
-\begin{algorithmic}[1]
-\REQUIRE $\,$ \\
-$\mathbf{y}$: matrix with $N$ input vectors and $l$ time series\\
-$j$: Starting point of testing \\
-$it$: Ending point of testing \\
-$ps$: list of $p$ values \\
-$Ls$: list of $L$ values ($L<N$) \\
-$m$: Iterations to determine parameters ($m < N-L$)\\
-\ENSURE  $\,$ \\
-$\{ \hat{\mathbf{y}}[1],\dots,\hat{\mathbf{y}}[it]\}$: prediction vectors \\
-\FOR { $i =j$ to $it$ }
-   \STATE $\mathbf{Y} \gets \mathbf{y}[:,i-1]$
-    \STATE $L,p \gets
-    \texttt{get\_best\_params}(Ls,ps,m,\mathbf{Y})$
-        \STATE $model = \text{VECM}(\mathbf{Y},L, p)$
-        \STATE $\hat{\mathbf{y}}[i-j] = model.predict()$
-\ENDFOR
-\end{algorithmic}
 \end{frame}
@@
 \begin{frame}
-\frametitle{Conclusions}
+\frametitle{Conclusions I}
 \begin{itemize}
@@
 %\item We found that out-of-sample forecast performance MSE is related to the {\em percentage of cointegration\/}.  
-\item VECM improves performance measures by finding
-parameters of $L$ and $p$ maximising the percentage of cointegration.
-%\item Increasing the number of parameters will always lead to
-%better performance measures.
-\item The parallel implementation allowed the execution times to be reduced
-more than 9 times and therefore a response time was obtain before 10
-seconds. 
-\end{itemize}
-\end{frame}
-
-%\begin{frame}
-%\frametitle{Proposal: Online VECM algorithm (OVECM)}
-%\begin{itemize}
-%\item Considers only a sliding window of the $L$ most recent data. 
-%\item Since cointegration vectors represent long-run relationships they vary
-%little in time. OVECM obtains these vectors only when they need to be updated. 
-%\item This updating is done when MAPEs of the last $n$ inputs is above a certain
-%error given by a threshold 
-%\item OVECM also implements matrix optimisations in order to reduce execution time,
-%such as updating matrices with new data, removing old data and introducing new
-%cointegration vectors.
-%\end{itemize}
-%\end{frame}
-%
-%\begin{frame}
-%\frametitle{Methodology}
-%OVECM were compared against VECM and ARIMA. VECM and ARIMA are called with a
-%sliding window of the most recent data, whereby the models are updated at every
-%time step. Therefore, we named them SLVECM and SLARIMA.
-%\end{frame}
-%
-%\subsection{Parameters}
-%\begin{frame}
-%\frametitle{Parameters optimization}
-%VECM order and ARIMA parameters were selected using AIC.
-%\begin{table}[ht]
-%\label{tab:params}
-%\begin{center}
-%\begin{tabular}{|c|c|c|}
-%\hline
-%Windows size $L$ & VECM & ARIMA\\
-% $L$ & order $(p)$ & order $(p,d,q)$ \\
-%\hline
-%100 & 2 & p=2,d=1,q=1\\
-%400 & 5 & p=1,d=1,q=1\\
-%700 & 3 &p=2,d=1,q=1\\
-%1000 &3 & p=2,d=1,q=1\\
-%\hline
-%\end{tabular}
-%\end{center}
-%\end{table}
-%\end{frame}
-%
-%\begin{frame}
-%\frametitle{Execution times}
-%We ran 400 iterations of OVECM and SLVECM . SLARIMA was based on the ARIMA procedure
-%of the statsmodels library and therefore its execution time is excluded.
-%%because its is not comparable with OVECM and SLVECM. SLARIMA was created based
-%%on statsmodels library routine ARIMA.
-%\begin{center}
-%\adjustbox{max height=\dimexpr\textheight-5.5cm\relax,
-%           max width=\textwidth}{
-%\begin{tabular}{|l|c|c|c|c|c|}
-%\hline
-%& $L$ & order & e  & Time[s] \\
-%\hline
-%OVECM & 100 &p=2  & 0      & 2.492\\
-%OVECM & 100 &p=2  & 0.0026  & 1.606\\
-%SLVECM & 100 &p=2& -- & 2.100\\
-%\hline
-%OVECM & 400 & p=5  & 0      & 3.513\\
-%OVECM & 400 &p=5  & 0.0041  & 2.569\\
-%SLVECM & 400 & p=5 & -- & 3.222\\
-%\hline
-%OVECM & 700 &p=3  & 0      & 3.296\\
-%OVECM & 700 &p=3  & 0.0032  & 2.856\\
-%SLVECM & 700 &p=3 & -- & 3.581\\
-%\hline
-%OVECM & 1000 & p=3 & 0      & 4.387\\
-%OVECM & 1000 & p=3  & 0.0022  & 2.408\\
-%SLVECM & 1000 & p=3  & -- & 3.609\\
-%\hline
-%\end{tabular}
-%}\end{center}
-%\end{frame}
-%
-%\begin{frame}
-%\frametitle{Performance accuracy}
-%\tiny
-%\begin{table}[H]
-%\caption{Model measures}
-%\label{tab:stats}
-%\begin{center}
-%\begin{tabular}{|l|l|c|c|c|c|c|c|c|}
-%\hline
-%\multicolumn{3}{|c|}{Model} & \multicolumn{3}{|c|}{In-sample} &
-%\multicolumn{3}{|c|}{Out-of-sample} \\ 
-%\hline
-%\hline
-%Method & $L$ & Parameters &  
-%MAPE & MAE& RMSE& 
-%MAPE & MAE& RMSE \\
-%\hline
-% OVECM  &   100  &  P=2 e=0.0026 &  0.00263&  0.00085&  0.00114&  0.00309&  0.00094&  0.00131\\
-% OVECM  &   400  &  P=5 e=0.0041&  0.00378&  0.00095&  0.00127&  0.00419&  0.00103&  0.00143\\
-% OVECM  &   700  &  P=3 e=0.0032&  0.00323&  0.00099&  0.00130&  0.00322&  0.00097&  0.00132\\
-% OVECM  &   1000 &  P=3 e=0.0022&  
-% \textbf{0.00175}&  \textbf{0.00062}&  \textbf{0.00087} &  
-% \textbf{0.00172}&  \textbf{0.00061}&  \textbf{0.00090}\\
-%\hline
-% SLVECM  &   100 &  P=2 &  0.00262&  0.00085&  0.00113&  0.00310&  0.00095&  0.00132\\
-% SLVECM  &   400 &  P=5 &  0.00375&  0.00095&  0.00126&  0.00419&  0.00103&  0.00143\\
-% SLVECM  &   700 &  P=3 &  0.00324&  0.00099&  0.00130&  0.00322&  0.00098&  0.00132\\
-% SLVECM  &   1000&  P=3 &  
-% \textbf{0.00174}&  \textbf{0.00061}&  \textbf{0.00087}&  
-% \textbf{0.00172}&  \textbf{0.00061}&  \textbf{0.00090}\\
-%\hline
-%SLARIMA & 100  &p=2 d=1 q=1 & 0.00285  &  0.00110  &  0.00308  &  0.00312  &0.00098  &  0.00144 \\
-%SLARIMA & 400  &p=1 d=1 q=1 & 0.00377  &  0.00101  &  0.00128  &  0.00418 & 0.00106 & 0.00145 \\
-%SLARIMA & 700  &p=2 d=1 q=1 & 0.00329  &  0.00102  &  0.00136  &  0.00324  & 0.00097  &  0.00133 \\
-%SLARIMA & 1000 &p=2 d=1 q=1 & \textbf{0.00281}  & \textbf{0.00074}  &  \textbf{0.00105}  &  \textbf{0.00177} & \textbf{0.00063}  &  \textbf{0.00092} \\
-%\hline
-%\end{tabular}
-%\end{center}
-%\end{table}
-%\end{frame}
-
-\begin{frame}
-\frametitle{Conclusions}
-\begin{itemize}
-\item  OVECM considerably reduces execution times
-without compromising solution accuracy.  
-\item OVECM was compared with VECM and ARIMA
-with the same sliding window sizes and OVECM outperformed both in terms of
-execution time. 
-\item Traditional VECM slightly outperformed our proposal but the
-OVECM execution time is lower.
-\item We could see that our algorithm took much less than a minute at every
-step. This means that it could also be used with higher frequency data and would
-still provide responses before new data arrives.  
+\item AVECM improves performance measures by finding parameters of $L$ and $p$ maximising the percentage of cointegration.
+\item AVECM performance is superior to ARIMA and the naive random walk model in terms of MSE and $U$-statistic. 
+%This result is very important since it is still difficult to outperform ARIMA and the random walk model for standard econometric forecasting models  despite their simplicity.
 \end{itemize}
@@
 \begin{frame}
-\frametitle{Summary of Contributions}
+\frametitle{Conclusions II}
 \begin{itemize}
-\item Two different approaches based on cointegration were presented: an adaptive version of VECM and an online approach.
-\item The concept of cointegration can now be used with high frequency time series.
-\item VECM computational limitations can either be tackled using a distribution computations or considering an online approach.
-\item An interesting connection between the percentage cointegration and forecasting performance.
+\item The parallel implementation allowed the execution times to be reduced
+more than 9 times and therefore a response time was obtain before 10
+seconds.
+\item We showed that VECM computational limitations can be tackled using a an online approach in a distributed environment.
 \item Cointegration information can be used as an integration tool to detect arbitrage opportunities.
-\end{itemize}
-\end{frame}
-
-\begin{frame}
-\frametitle{Author Contributions}
-\begin{itemize}
-\item 1
-\item 2
 \end{itemize}
@@
 \begin{itemize}
-\item 1
-\item 2
+\item To explore the relationship between cointegration and performance for different assets (stocks, other forex rates).
+\item To include more explicative variables such as bid-ask spread and change in volume.
+\item The online approach for other econometrics models it will be also worthy of  study.% Many of them could be adapted and used with higher frequency data.
+\item  In this thesis, the online version of OLS was studied, but the online version of
+ridge regression could also be applied to obtain a solution with better
+generalisation capabilities.
 \end{itemize}
@@
 \appendix
-
-\begin{frame}
-\frametitle{VECM(p) matrix form}
-If we have $N$ data points, with $N>p$, we can estimate the model parameters $\mathbf{X}$ using the following matrix form:
-\small
-\begin{equation*}
-\underbrace{
-      \begin{bmatrix}
-       \mathbf{\Delta y}_{p+1}  \\ 
-       \mathbf{\Delta y}_{p+2}  \\ 
-       \vdots                   \\ 
-       \mathbf{\Delta y}_N      
-      \end{bmatrix}}_{\mathbf{B} } =
-\arraycolsep=1.4pt  
-\underbrace{\begin{bmatrix}
-   \alpha^\top & \phi^*_1 & \dots & \phi^*_{p-1} &\mathbf{c}   
-  \end{bmatrix}}_{\mathbf{X}^\top}
-\underbrace{\begin{bmatrix}
- \beta\mathbf{y}_p      & \beta\mathbf{y}_{p+1}   & \dots&   \beta\mathbf{y}_{N-1}   \\
- \mathbf{\Delta y}_{p}   & \mathbf{\Delta y}_{p+1}& \dots &  \mathbf{\Delta y}_{N-1} \\
- \mathbf{\Delta y}_{p-1} & \mathbf{\Delta y}_{p}  & \dots &  \mathbf{\Delta y}_{N-2}   \\
- \vdots                  & \vdots                 & \ddots&  \vdots                   \\
- \mathbf{\Delta y}_{2}   & \mathbf{\Delta y}_{3} & \dots &   \mathbf{\Delta y}_{N-p+1} \\
- 1                      & 1                       & \dots     & 1   
- \end{bmatrix}}_{\mathbf{A}^\top}
-+
-\underbrace{\begin{bmatrix}
-              \mathbf{\epsilon}_{p+1} \\ 
-              \mathbf{\epsilon}_{p+2} \\ 
-              \vdots \\ 
-              \mathbf{\epsilon}_N
-             \end{bmatrix}}_{\mathbf{E}^\top}
-\end{equation*}
-\end{frame}
 
@@
 
+\begin{frame}
+\frametitle{AVECM Algorithm}
+\small
+\begin{algorithmic}[1]
+\REQUIRE $\,$ \\
+$\mathbf{y}$: matrix with $N$ input vectors and $l$ time series\\
+$j$: Starting point of testing \\
+$it$: Ending point of testing \\
+$ps$: list of $p$ values \\
+$Ls$: list of $L$ values ($L<N$) \\
+$m$: Iterations to determine parameters ($m < N-L$)\\
+\ENSURE  $\,$ \\
+$\{ \hat{\mathbf{y}}[1],\dots,\hat{\mathbf{y}}[it]\}$: prediction vectors \\
+\FOR { $i =j$ to $it$ }
+   \STATE $\mathbf{Y} \gets \mathbf{y}[:,i-1]$
+    \STATE $L,p \gets
+    \texttt{get\_best\_params}(Ls,ps,m,\mathbf{Y})$
+        \STATE $model = \text{VECM}(\mathbf{Y},L, p)$
+        \STATE $\hat{\mathbf{y}}[i-j] = model.predict()$
+\ENDFOR
+\end{algorithmic}
+\end{frame}
+
 % Financial time series are not the only ones with this behaviour, this study can be extended to other non-stationary time series such as: weather, earthquakes, energy demand and sales forecasting.

### 2017-12-11 slides/defense-slides.tex
Version para el seminario CCTVal

@@
 
-\author[Paola Arce]{Paola Arce \\ Thesis advisor: Luis Salinas}
+\author[Paola Arce]{Paola Arce}% \\ Thesis advisor: Luis Salinas}
 \date{December 11, 2017}
@@
 \begin{itemize}
-\item Motivation: High Frequency trading
+\item Motivation: High Frequency Trading
 \item Cointegration and the VECM model
@@
 \item Contributions
-\item Experiments
-\item Conclusions
-\item Future work
+\item Limitations and Applicability
+\item Conclusions and Future Research
 \end{itemize}
@@
 \frametitle{HFT: High Frequency Trading}
-High frequency trading has motivated computer-driven strategies capable of processing large amount of data in short periods of time.
+High Frequency Trading has motivated computer-driven strategies capable of processing large amounts of data in short periods of time.
 \begin{figure}
@@
 \frametitle{HFT Revenue}
-However, the revenues have fallen dramatically mainly because of more regulation, increasing cost of HFT infrastructure and more competitive algorithms.
+However, the revenues have fallen dramatically mainly because of more regulation, increasing the cost of HFT infrastructure and more competitive algorithms.
 \begin{figure}
@@
 %\item Heterogeneity
-\item Heteroscedastic
+%\item Heteroscedastic
 \item Potentially cointegrated
@@
 \vspace{5mm}
-This imply that classical statistical models can't be used. Therefore, financial time series forecasting is still a challenge.
+This implies that classical statistical models as such can't be used. Therefore, financial time series forecasting is still a challenge.
 \end{frame}
@@
 \begin{block}{Strong definition}
-A {\color{red}strictly stationary time series} $y_t \in \mathbb{R}$, $\forall t \in \mathbb{Z}$ is one for which the probabilistic behaviour of every collection of values $\{y_{t_1},y_{t_2},\dots,y_{t_L}\}$ is identical
+A {\color{red}strictly stationary time series} $y_t \in \mathbb{R}$, $\forall t \in \mathbb{Z}$, is one for which the probabilistic behaviour of every collection of values $\{y_{t_1},y_{t_2},\dots,y_{t_L}\}$ is identical
 to that of the time shifted set, more precisely:
@@
 \end{equation*}
-\noindent $\forall L \in \mathbb{N}, \forall h \in \mathbb{Z}$ where $c_1,\dots,c_L$ are constants.
+\noindent $\forall L \in \mathbb{N}, \forall h \in \mathbb{Z}$, where $c_1,\dots,c_L$ are constants.
 \end{block}
@@
 
-A weakly stationary time series is a process which mean, variance and auto covariance do not change over time:
+A weakly stationary time series is a process where the  mean, variance and auto covariance do not change over time:
 \small
@@
 \frametitle{Irregularly spaced over time}
-\begin{figure}
-\includegraphics[width=0.8\paperwidth]{img/hft-ticks-zoom.png}
-\end{figure}
+\centerline{
+\includegraphics[width=\paperwidth,height=0.7\paperheight]{img/hft-ticks-zoom.png}
+}
 \end{frame}
 
-\begin{frame}
-\frametitle{Heteroscedastic}
-Conditional heteroskedasticity identifies non-constant volatility.
-Volatility gives a measure of uncertainty about future returns. Many factors affect volatility such as: historical prices, trends, monetary policies and news.
-\begin{figure}
-\includegraphics[width=0.8\paperwidth]{img/vix}
-\end{figure}
-\end{frame}
+%\begin{frame}
+%\frametitle{Heteroscedastic}
+%Conditional Heteroscedasticity identifies non-constant volatility.
+%Volatility gives a measure of uncertainty about future returns. Many factors affect volatility such as: historical prices, trends, monetary policies and news.
+%\begin{figure}
+%\includegraphics[width=0.8\paperwidth]{img/vix}
+%\end{figure}
+%\end{frame}
 
 \section{Cointegration and the VECM model}
+
 \begin{frame}
 \frametitle{Cointegration}
+Cointegration reflects the idea of that some set of non-stationary variables cannot wander too far from each other.
 \begin{figure}
@@
 \frametitle{Cointegration definition}
-
+%The idea of cointegration is that there is an I(1) process, underlying two or more processes.
 \begin{block}{Cointegration}
@@
 \end{block}
-
 \end{frame}
@@
 \frametitle{Online VECM}
-Recently, online learning algorithms have been proposed to solve problems with large data sets main because: 
+Recently, online learning algorithms have been proposed to solve problems with large data sets mainly because of: 
 \begin{itemize}
 \item Their simplicity
-\item Process one instance at a time
-\item The hypothesis is updated every new data arrives
+\item They process one instance at a time
+\item The hypothesis is updated every time new data arrives
 \end{itemize} 
@@
 \frametitle{VECM(p) matrix form}
+\begin{equation*}
+ \Delta \mathbf{y}_t = \underbrace{
+        \tikz[baseline]{
+            \node[fill=red!20,ellipse,anchor=base] (t1)
+            {$ \Omega\mathbf{y}_{t-1}$};
+        }}_\text{Error correction term} +
+        \underbrace{
+        \tikz[baseline]{
+            \node[fill=blue!20,anchor=base] (t2)
+            {$\sum_{i=1}^{p-1} \phi_i^* \Delta \mathbf{y}_{t-i} $};
+        }}_\text{Autorregresive term} +
+        \underbrace{
+        \tikz[baseline]{
+            \node[fill=green!20,anchor=base] (t3)
+            {$c$};
+        }}_\text{Intercept}
+        +
+        \underbrace{
+        \tikz[baseline]{
+            \node[fill=yellow!20,anchor=base] (t3)
+            {$\mathbf{\epsilon}_t $};
+        }}_\text{i.i.d}
+\end{equation*}
 If we have $N$ data points, with $N>p$, the VECM matrix form is as follows:
@@
 
-We can estimate the model parameters $\mathbf{W}$ using OLS:
-\begin{equation}
-{\color{blue}
-\mathbf{Y} = \mathbf{X}\mathbf{W}+\mathbf{E}
-}
-\end{equation}
+%We can estimate the model parameters $\mathbf{W}$ using OLS:
+%\begin{equation}
+%{\color{blue}
+%\mathbf{Y} = \mathbf{X}\mathbf{W}+\mathbf{E}
+%}
+%\end{equation}
 
@@
 \mathbf{A}^{-1}\mathbf{x}_t$
-        \STATE Read new $y_t$
-        \STATE $\mathbf{b} = \mathbf{b} + y_t \mathbf{x}_t$
+        \STATE Read new $\mathbf{y}_t$
+        \STATE $\mathbf{b} = \mathbf{b} + \mathbf{x}_t \mathbf{y}^\intercal_{t}$
 \ENDFOR
-
 \end{algorithmic}
@@
 \begin{enumerate}
-\item Uses VECM only considering a sliding windows of size $L$ of data.
+\item It uses VECM only considering a sliding windows of size $L$ of data.
 \item We proposed to choose $L$ and the number of lags $p$ in order to maximise the percentage of cointegration $PC$ in the near past. This process was done every time that new data was processed. 
-\item To use a distributed environment to make this parameters search, since this is the most expensive routine.
+\item And to use a distributed environment to make this parameters search, since this is the most expensive routine.
 \end{enumerate}
@@
 
+\section{Limitations and Applicability}
+
+\begin{frame}
+\frametitle{Limitations and Applicability}
+\begin{itemize}
+\item Cointegration information can be used as an integration tool to detect arbitrage opportunities.
+\item Use with stocks, forex rates and crypto-currency?
+\item Applicable to other non-stationary but cointegrated time series
+\end{itemize}
+\end{frame}
+
+\section{Conclusions and Future Research}
+
 \begin{frame}
@@
 \item We showed that VECM computational limitations can be tackled using a an online approach in a distributed environment.
-\item Cointegration information can be used as an integration tool to detect arbitrage opportunities.
 \end{itemize}
@@
 \begin{itemize}
-\item To explore the relationship between cointegration and performance for different assets (stocks, other forex rates).
-\item To include more explicative variables such as bid-ask spread and change in volume.
-\item The online approach for other econometrics models it will be also worthy of  study.% Many of them could be adapted and used with higher frequency data.
-\item  In this thesis, the online version of OLS was studied, but the online version of
-ridge regression could also be applied to obtain a solution with better
-generalisation capabilities.
+\item To add {\bf matrix optimizations} to obtain VECM parameters.
+\item To add a regularization parameter to get better generalisation capabilities using {\bf Ridge Regression} and the {\bf Aggregating Algorithm for Regression}.
+\item To include {\bf more explicative variables} such as bid-ask spread and change in volume.
+\item The online approach for {\bf other econometrics models} it is also worthy of  study.% Many of them could be adapted and used with higher frequency data.
 \end{itemize}
@@
 %9 Validation Plan
-%9 Limitations and Applicability
+
 %9 Expected Contributions
@@
 \appendix
+
+\begin{frame}
+\frametitle{The order book}
+\begin{figure}
+\includegraphics[width=0.7\textwidth]{img/orderbook}
+\end{figure}
+\end{frame}
 
@@
 
-%\subsection{Model comparison} 
 \begin{frame}

### 2017-12-12 slides/defense-slides.tex
mejoras dePeretti

@@
 %\usepackage[spanish,activeacute]{babel}
-\usetheme{metropolis}
+\usetheme[sectionpage=none,progressbar=frametitle]{metropolis}
 \usepackage[utf8]{inputenc}
@@
 \def\N{{\mathbb N }}
+\def\Z{{\mathbb Z }}
+
+\makeatletter
+%\setlength{\metropolis@titleseparator@linewidth}{2pt}
+\setlength{\metropolis@progressonsectionpage@linewidth}{2pt}
+\setlength{\metropolis@progressinheadfoot@linewidth}{2pt}
+\makeatother
 
@@
 
-%\author[Paola Arce]{Paola Arce \\ Departamento de Inform\'atica, UTFSM \\ Thesis advisor: Luis Salinas}
-%\date{December 15, 2017}
-%\title[Thesis defense]{Online learning methods for financial time series forecasting}
-%\subtitle{PhD Thesis Defense}
+\author[Paola Arce]{Paola Arce \\ Thesis advisor: Luis Salinas}
+\date{December 15, 2017}
+\title[Thesis defense]{Online learning methods for financial time series forecasting}
+\subtitle{PhD Thesis Defense}
 %
@@
 %}
-
-\author[Paola Arce]{Paola Arce}% \\ Thesis advisor: Luis Salinas}
-\date{December 11, 2017}
-\title[Thesis defense]{Fast and adaptive cointegration based model for forecasting financial time series}
-\subtitle{Seminario Inform\'atica CCTVal}
+%
+%\author[Paola Arce]{Paola Arce}% \\ Thesis advisor: Luis Salinas}
+%\date{December 11, 2017}
+%\title[Thesis defense]{Fast and adaptive cointegration based model for forecasting financial time series}
+%\subtitle{Seminario Inform\'atica CCTVal}
 
@@
 %\item Heterogeneity
-%\item Heteroscedastic
+\item Heteroscedastic
 \item Potentially cointegrated
@@
 \begin{block}{Strong definition}
-A {\color{red}strictly stationary time series} $y_t \in \mathbb{R}$, $\forall t \in \mathbb{Z}$, is one for which the probabilistic behaviour of every collection of values $\{y_{t_1},y_{t_2},\dots,y_{t_L}\}$ is identical
-to that of the time shifted set, more precisely:
+A {\color{red}strictly stationary time series} $y_t: r.v$ with values in $\mathbb{R}$ defined for $t \in \mathbb{Z}$, is one for which the probabilistic behaviour of every collection of values $\{y_{t_1},y_{t_2},\dots,y_{t_L}\}$ is identical to that of the time shifted set, more precisely:
 \begin{equation*}
@@
 \end{frame}
-
+%
 %\begin{frame}
@@
             \node[fill=blue!20,anchor=base] (t2)
-            {$\sum_{i=1}^{p-1} \phi_i^* \Delta \mathbf{y}_{t-i} $};
+            {$\sum_{i=1}^{p-1} {\boldsymbol\phi}_i^* \Delta \mathbf{y}_{t-i} $};
         }}_\text{Autorregresive term} +
@@
             \node[fill=yellow!20,anchor=base] (t3)
-            {$\mathbf{\epsilon}_t $};
+            {$\boldsymbol{\epsilon}_t $};
         }}_\text{i.i.d}
@@
 }
-\noindent where $p$ is the number of lags and $\boldsymbol \Omega = \boldsymbol \alpha \boldsymbol \beta^\intercal$.
+\noindent where $\mathbf{y}_t = [y^1_t,\dots,y^l_t]^\intercal$, $p$ is the number of lags and $\boldsymbol \Omega = \boldsymbol \alpha \boldsymbol \beta^\intercal$.
 The columns of $\boldsymbol\beta$ contains the cointegration vectors and the rows of $\boldsymbol\alpha$ correspond with the adjusted vectors.
@@
 \begin{frame}
+\frametitle{Articles and conferences}
+\nocite{arce+salinas2012,icpram15,Arce2017}
+\bibliographystyle{apalike}
+\bibliography{reference}
+\end{frame}
+
+\begin{frame}
 \frametitle{Future research}
@@
 \end{frame}
-
-\nocite{arce+salinas2012,icpram15,Arce2017}
-\bibliographystyle{apalike}
-\bibliography{reference}
 
@@
 \begin{frame}
+\frametitle{Stationarity}
+A time series $y_t: \Z \to \R $ is said to be strictly stationary if, for a finite set of $L$ indices $\{t_{1},\dots,t_{L}\}  \, \forall L \in \Z_{\ge 0}$, the joint probability distribution of the random variables $\{y_{t_{1}},y_{t_{2}},\dots,y_{t_{L}}\}$ remains unchanged for any time shifted $h$.
+The probability distribution for the random variable $y_{t_1}$ is:
+
+\begin{equation}
+P\{y_{t} \leq c_1 \} = P\{t \in \Z:  y_t \in  ]-\infty, c_1] \} 
+\end{equation}
+
+\end{frame}
+
+\begin{frame}
 \frametitle{The order book}
@@
 \end{figure}
+\end{frame}
+
+\begin{frame}
+\frametitle{Unit root process}
+
 \end{frame}

### 2017-12-13 slides/defense-slides.tex
Contribucion esquema y mejor motivacion

@@
 \begin{itemize}
-\item Motivation: High Frequency Trading
-\item Cointegration and the VECM model
+\item Motivation: High Frequency Trading and Cointegration
 \item Challenges and Hypothesis
@@
 \section{Motivation}
+
+\begin{frame}
+\frametitle{Motivation: HFT}
+\begin{block}{High Frequency Trading}
+High Frequency Trading (HFT) is a technology that implements different trading strategies to buy or sell an stock using sophisticated infrastructure and high speed algorithms.
+\end{block}
+
+Today, the volume of data has rapidly increased:
+
+\begin{table}
+\begin{tabular}{ c | c }
+Period & Data precision \\
+\hline \hline
+1993-2003 & seconds \\
+2003-2015 & milliseconds \\
+27th July 2015-onwards & microseconds \\
+24th October 2016-onwards & nanoseconds
+\end{tabular}
+\caption{High Frequency data precision}
+\end{table}
+
+\end{frame}
 
@@
 \caption{Revenue in the US. Source: TABB group}
+\end{figure}
+\end{frame}
+
+\begin{frame}
+\frametitle{Motivation: HFT}
+\begin{alertblock}{Arbitrage}
+The practice of taking advantage of market inefficiency or discrepancies in the market price.
+\end{alertblock}
+\begin{columns}
+
+\begin{column}{0.4\textwidth}
+\begin{exampleblock}{Advantages}
+\begin{itemize}
+\item Increases liquidity
+\item Allows more efficient price formation
+\item Reduce the difference between buyers and sellers prices.
+\end{itemize}
+\end{exampleblock}
+\end{column}
+\begin{column}{0.4\textwidth}
+\begin{exampleblock}{Risks}
+\begin{itemize}
+\item Removal of human decision making could lead to failures in untested or incorrect strategies
+\end{itemize}
+\end{exampleblock}
+\end{column}
+\end{columns}
+\end{frame}
+
+\begin{frame}
+\frametitle{Motivation: Cointegration}
+Cointegration reflects the idea of that some set of non-stationary variables cannot wander too far from each other.
+\begin{figure}
+\includegraphics[width=0.8\paperwidth]{img/motivation}
+\end{figure}
+\end{frame}
+
+\begin{frame}
+\frametitle{Contribution}
+This thesis contribution has several components:
+\begin{figure}
+\includegraphics[width=\textwidth]{img/contribution}
 \end{figure}
@@
 \begin{frame}
-\frametitle{Cointegration}
-Cointegration reflects the idea of that some set of non-stationary variables cannot wander too far from each other.
-\begin{figure}
-\includegraphics[width=0.8\paperwidth]{img/motivation}
-\end{figure}
-\end{frame}
-
-\begin{frame}
 \frametitle{Cointegration definition}
@@
 When this happens it is said they have a {\bf long-run
-equilibrium} relationship and the variables are {\color{red} cointegrated}.
+equilibrium} relationship and the variables are {\color{red} {\bf cointegrated}}.
 \end{block}
+\begin{figure}
+\includegraphics[width=0.8\paperwidth, height=0.4\paperheight]{img/motivation}
+\end{figure}
 \end{frame}
@@
 
-The Vector Error correction model (VECM) is used to model cointegrated time series:
+The Vector Error correction model (VECM), considering $p$ lags, is used to model cointegrated time series:
 
@@
 }
-\noindent where $\mathbf{y}_t = [y^1_t,\dots,y^l_t]^\intercal$, $p$ is the number of lags and $\boldsymbol \Omega = \boldsymbol \alpha \boldsymbol \beta^\intercal$.
+\noindent where 
+\begin{itemize}
+\item $\mathbf{y}_t = [y^1_t,\dots,y^l_t]^\intercal$,
+\item the errors $\mathbf{e}_t$ are assumed to follow i.i.d l-dimensional multivariate normal distribution $\mathcal{N}({\bf O,\Sigma})$, 
+\end{itemize}
+\end{frame}
+
+\begin{frame}
+\frametitle{The Vector Error correction}
+\tikzstyle{na} = [baseline=-.5ex]
+
+The Vector Error correction model (VECM), considering $p$ lags, is used to model cointegrated time series:
+
+{\Large
+\begin{equation*}
+ \Delta \mathbf{y}_t = \underbrace{
+        \tikz[baseline]{
+            \node[fill=red!20,ellipse,anchor=base] (t1)
+            {$ \Omega\mathbf{y}_{t-1}$};
+        }}_\text{Error correction term} +
+        \underbrace{
+        \tikz[baseline]{
+            \node[fill=blue!20,anchor=base] (t2)
+            {$\sum_{i=1}^{p-1} {\boldsymbol\phi}_i^* \Delta \mathbf{y}_{t-i} $};
+        }}_\text{Autorregresive term} +
+        \underbrace{
+        \tikz[baseline]{
+            \node[fill=green!20,anchor=base] (t3)
+            {$c$};
+        }}_\text{Intercept}
+        +
+        \underbrace{
+        \tikz[baseline]{
+            \node[fill=yellow!20,anchor=base] (t3)
+            {$\boldsymbol{\epsilon}_t $};
+        }}_\text{i.i.d}
+\end{equation*}
+}
+\noindent where 
+\begin{itemize}
+\item $\boldsymbol \Omega = \boldsymbol \alpha \boldsymbol \beta^\intercal$.
 The columns of $\boldsymbol\beta$ contains the cointegration vectors and the rows of $\boldsymbol\alpha$ correspond with the adjusted vectors.
+\end{itemize}
 \end{frame}
@@
 \frametitle{$\boldsymbol \Omega$ matrix properties}
+
+Since $\boldsymbol \Omega = \boldsymbol \alpha \boldsymbol \beta^\intercal$ we can also say that:
+
+$$\boldsymbol \Omega \mathbf{y} = \boldsymbol \alpha \boldsymbol \beta^\intercal \mathbf{y} = \boldsymbol \alpha \mathbf{Z}_t \sim \text{I(0)}$$
+
 The matrix $\boldsymbol \Omega$ has the following properties:
 \begin{itemize}
-\item If ${\color{blue}\boldsymbol \Omega = 0}$ there is no cointegration
-\item If ${\color{blue}rank(\boldsymbol\Omega)=l}$ i.e full rank, then the time series are not
-I(1) but stationary
-\item If ${\color{blue}rank(\boldsymbol\Omega)=r,\quad 0 < r < l}$ then, there is cointegration.
+\item If ${\color{blue}rank(\boldsymbol \Omega) = 0}$ there is no cointegration
+\item If ${\color{blue}rank(\boldsymbol\Omega)=l}$ i.e full rank, $\boldsymbol\Omega$ is non-singular or invertible, therefore we can express $\mathbf{y}$ as follows:
+$$\mathbf{y} = \boldsymbol \Omega ^{-1} \boldsymbol \alpha \mathbf{Z}_t \sim \text{I(0)} \, ,$$
+then the time series are not I(1) but stationary.
+\item If ${\color{blue}rank(\boldsymbol\Omega)=r,\quad 0 < r < l}$ then, r cointegration vectors exist such that: 
+$$\boldsymbol \Omega \mathbf{y} = \boldsymbol \alpha \mathbf{Z}_t \sim \text{I(0)} \, .$$
 %where $\boldsymbol\alpha$ and $\boldsymbol\beta$ are $(l \times r)$
@@
 \section{Contributions}
+
 \begin{frame}

### 2017-12-14 slides/defense-slides.tex
presentacion antes de reunion con el profe

@@
 \usepackage{pgf}
+\usepackage{pgfplots}
+\pgfplotsset{compat=1.12}
 \usepackage{algorithm} 
@@
 \frametitle{Motivation: HFT}
-\begin{block}{High Frequency Trading}
+\begin{alertblock}{High Frequency Trading}
 High Frequency Trading (HFT) is a technology that implements different trading strategies to buy or sell an stock using sophisticated infrastructure and high speed algorithms.
-\end{block}
-
-Today, the volume of data has rapidly increased:
-
+\end{alertblock}
+HFT challenges are related to {\bf time series forecasting} and also {\bf computational} (hardware and software). Today, the volume of data has rapidly increased:
 \begin{table}
-\begin{tabular}{ c | c }
+\small
+\begin{tabular}{ l | c }
 Period & Data precision \\
@@
 \end{table}
-
 \end{frame}
@@
 \begin{alertblock}{Arbitrage}
-The practice of taking advantage of market inefficiency or discrepancies in the market price.
+The practice of taking advantage of market inefficiency or discrepancies in the market price. Arbitrage contradicts the Efficient Market Hypothesis. 
+%that states, the current stock price fully reflects available information related to its value and there is no way to earn excess profit.
 \end{alertblock}
@@
 
-\begin{column}{0.4\textwidth}
+\begin{column}{0.5\textwidth}
 \begin{exampleblock}{Advantages}
@@
 \item Allows more efficient price formation
-\item Reduce the difference between buyers and sellers prices.
+\item Reduce the difference between buyers (bid) and sellers (ask) prices.
 \end{itemize}
@@
 \end{column}
-\begin{column}{0.4\textwidth}
-\begin{exampleblock}{Risks}
+\begin{column}{0.5\textwidth}
+\begin{exampleblock}{Disadvantages}
 \begin{itemize}
 \item Removal of human decision making could lead to failures in untested or incorrect strategies
+\item fairness
 \end{itemize}
@@
 \end{frame}
-%
-%\begin{frame}
-%\frametitle{Heteroscedastic}
-%Conditional Heteroscedasticity identifies non-constant volatility.
-%Volatility gives a measure of uncertainty about future returns. Many factors affect volatility such as: historical prices, trends, monetary policies and news.
-%\begin{figure}
-%\includegraphics[width=0.8\paperwidth]{img/vix}
-%\end{figure}
-%\end{frame}
+
+\begin{frame}
+\frametitle{Heteroscedastic}
+
+\begin{block}{Heteroscedasticity}
+Heteroscedasticity identifies {\bf non-constant volatility}.
+Volatility gives a measure of uncertainty about future returns. It can either be obtained from historical returns or implied from derivatives such as options. 
+\end{block}
+
+\begin{figure}
+\begin{tikzpicture}[ % Define Normal Probability Function
+declare function={
+            normal(\x,\m,\s) = 1/(2*\s*sqrt(pi))*exp(-(\x-\m)^2/(2*\s^2));
+        }
+       ]
+\begin{axis}[
+    no markers,
+    domain=0:12,
+    zmin=0, zmax=1,
+    xmin=0, xmax=3,
+    samples=200,
+   samples y=0,
+    axis lines=middle,
+    xtick={0.5,1.5,2.5},
+    xmajorgrids,
+    xticklabels={},
+    ytick=\empty,
+    xticklabels={$\mathbf{y}_{t_1}$, $\mathbf{y}_{t_2}$, $\mathbf{y}_{t_3}$},
+    ztick=\empty,
+    xlabel=$t$, xlabel style={at={(rel axis cs:1,0,0)}, anchor=west},
+    ylabel=$\mathbf{y}$, ylabel style={at={(rel axis cs:0,1,0)}, anchor=south west},
+    zlabel=Probability density, zlabel style={at={(rel axis cs:0,0,0.5)}, rotate=90, anchor=south},
+    set layers
+  ]
+
+\addplot3 [samples=2, samples y=0, domain=0:3] (x, {1.5*(x-0.5)+3}, 0);
+\addplot3 [cyan!50!black, thick] (0.5, x, {normal(x, 3, 0.5)});
+\addplot3 [cyan!50!black, thick] (1.5, x, {normal(x, 4.5, 1)});
+\addplot3 [cyan!50!black, thick] (2.5, x, {normal(x, 6, 1.5)});
+
+\pgfplotsextra{
+\begin{pgfonlayer}{axis background}
+\draw [on layer=axis background] (0.5, 3, 0) -- (0.5, 3, {normal(0,0,0.5)}) (0.5,0,0) -- (0.5,12,0);
+\draw (1.5, 4.5, 0) -- (1.5, 4.5, {normal(0,0,1)}) (1.5,0,0) -- (1.5,12,0);
+\draw (2.5, 6, 0) -- (2.5, 6, {normal(0,0,1.5)}) (2.5,0,0) -- (2.5,12,0);
+\end{pgfonlayer}
+}
+\end{axis}
+
+\end{tikzpicture}
+\end{figure}
+% An option is a contract which gives the owner the right but not the obligation to buy or sell an underlying asset at a given price called strike price.
+
+\end{frame}
 
@@
 The columns of $\boldsymbol\beta$ contains the cointegration vectors and the rows of $\boldsymbol\alpha$ correspond with the adjusted vectors.
-\end{itemize}
-\end{frame}
-
-\begin{frame}
-\frametitle{$\boldsymbol \Omega$ matrix properties}
-
-Since $\boldsymbol \Omega = \boldsymbol \alpha \boldsymbol \beta^\intercal$ we can also say that:
-
-$$\boldsymbol \Omega \mathbf{y} = \boldsymbol \alpha \boldsymbol \beta^\intercal \mathbf{y} = \boldsymbol \alpha \mathbf{Z}_t \sim \text{I(0)}$$
-
-The matrix $\boldsymbol \Omega$ has the following properties:
-\begin{itemize}
-\item If ${\color{blue}rank(\boldsymbol \Omega) = 0}$ there is no cointegration
-\item If ${\color{blue}rank(\boldsymbol\Omega)=l}$ i.e full rank, $\boldsymbol\Omega$ is non-singular or invertible, therefore we can express $\mathbf{y}$ as follows:
-$$\mathbf{y} = \boldsymbol \Omega ^{-1} \boldsymbol \alpha \mathbf{Z}_t \sim \text{I(0)} \, ,$$
-then the time series are not I(1) but stationary.
-\item If ${\color{blue}rank(\boldsymbol\Omega)=r,\quad 0 < r < l}$ then, r cointegration vectors exist such that: 
-$$\boldsymbol \Omega \mathbf{y} = \boldsymbol \alpha \mathbf{Z}_t \sim \text{I(0)} \, .$$
-%where $\boldsymbol\alpha$ and $\boldsymbol\beta$ are $(l \times r)$
-%matrices and $rank(\boldsymbol\alpha)=rank(\boldsymbol\beta)=r$.
 \end{itemize}
@@
 \end{equation*}}
-
 %We can estimate the model parameters $\mathbf{W}$ using OLS:
@@
 %\end{equation}
-
 \end{frame}
@@
 \end{algorithm}
+\end{frame}
+
+\begin{frame}
+\frametitle{How to choose $L$?: $\boldsymbol \Omega$ matrix properties}
+Since $\boldsymbol \Omega = \boldsymbol \alpha \boldsymbol \beta^\intercal$ we can also say that:
+$$\boldsymbol \Omega \mathbf{y} = \boldsymbol \alpha \boldsymbol \beta^\intercal \mathbf{y} = \boldsymbol \alpha \mathbf{Z}_t \sim \text{I(0)}$$
+The matrix $\boldsymbol \Omega$ has the following properties:
+\begin{itemize}
+\item If ${\color{blue}rank(\boldsymbol \Omega) = 0}$ there is no cointegration
+\item If ${\color{blue}rank(\boldsymbol\Omega)=l}$ i.e full rank, $\boldsymbol\Omega$ is non-singular or invertible, therefore we can express $\mathbf{y}$ as follows:
+$$\mathbf{y} = \boldsymbol \Omega ^{-1} \boldsymbol \alpha \mathbf{Z}_t \sim \text{I(0)} \, ,$$
+then the time series are not I(1) but stationary.
+\item If ${\color{blue}rank(\boldsymbol\Omega)=r,\quad 0 < r < l}$ then, r cointegration vectors exist such that: 
+$$\boldsymbol \Omega \mathbf{y} = \boldsymbol \alpha \mathbf{Z}_t \sim \text{I(0)} \, .$$
+%where $\boldsymbol\alpha$ and $\boldsymbol\beta$ are $(l \times r)$
+%matrices and $rank(\boldsymbol\alpha)=rank(\boldsymbol\beta)=r$.
+\end{itemize}
 \end{frame}
@@
 \frametitle{Adaptive VECM algorithm (AVECM)}
-AVECM has the following features: 
 \begin{enumerate}
@@
 \begin{tabular}{ccccccc}
-& \multicolumn{3}{l}{MSE} & &
-\multicolumn{2}{c}{$U$-statistic} \\ 
+& \multicolumn{3}{l}{{\bf MSE}} & &
+\multicolumn{2}{c}{{\bf $\mathbf{U}$-statistic}} \\ 
 \hhline{~---~--} \\
@@
 \hline
- EURUSD & 1.0702 e-09 & 1.1481 e-09 &  9.250 e-12 & & 0.6863 & 0.7108\\
- GBPUSD & 1.6630 e-09 & 1.7408 e-09 &  6.951 e-02 & & 0.6866 & 0.7025\\
- USDCHF & 5.8503 e-10 & 6.3545 e-10 &  2.899 e-14 & & 0.6803 & 0.7091\\
- USDJPY  & 6.3483 e-06 & 6.5194 e-06 &  6.853 e-05 & & 0.6964 & 0.7057
+ EURUSD & 1.0702 e-09 & 1.1481 e-09 &  \bf{9.250 e-12} & & {\color{GreenTOL}\bf{0.6863}} & {\color{red}\bf{0.7108}}\\
+ GBPUSD & 1.6630 e-09 & 1.7408 e-09 &  6.951 e-02 & & {\color{GreenTOL}\bf{0.6866}} & {\color{red}\bf{0.7025}}\\
+ USDCHF & 5.8503 e-10 & 6.3545 e-10 &  \bf{2.899 e-14} & & {\color{GreenTOL}\bf{0.6803}} & {\color{red}\bf{0.7091}}\\
+ USDJPY  & 6.3483 e-06 & 6.5194 e-06 &  \bf{6.853 e-05} & & {\color{GreenTOL}\bf{0.6964}} & {\color{red}\bf{0.7057}}
  \end{tabular}
@@
 The function that get optimal parameters $L$ and $p$ was implemented using MPI in Python.
-Tests ran in a cluster with 2 servers Xeon E5-2667 (2.90GHz) of 24 cores each (48 cores in total) and 24GB RAM.
+Tests ran in our cluster {\bf hpc.utfsm.cl} considering 2 servers Xeon E5-2667 (2.90GHz) of 24 cores each (48 cores in total) and 24GB RAM.
 
@@
 %\item Parameter $it=100$
-\item The $L$ parameter was always chosen between 100 and 4000 and $p$ always took values between 1 and 5. 
+\item The $L$ parameter was always chosen between 100 and 4000 (1 hour=360 data points) and $p$ always took values between 1 and 5. 
 \item Parameter $nparams$ represents the number of pairs ($L$,$p$) used to maximise the percentage of cointegration. 
@@
 \end{figure}
+\end{frame}
+
+\begin{frame}
+\frametitle{Contributions}
+%\begin{figure}
+%\includegraphics[width=0.7\textwidth]{img/contribution}
+%\end{figure}
+\begin{description}
+\item[Cointegration]Cointegration relations change with time.
+We empirically showed that the number of coitegration vectors is sensitive to the number of lags but also to the amount of data considered.
+The percentage of cointegration is used to find VECM parameters.
+\item[Online learning] The adaptation of the VECM to be used in an online context expressing the OLS method with matrix actualization.
+\item[Distributed environment] The use of a distributed environment to calculate $L$ and $p$ and obtain a response below 10 seconds.
+\end{description}
 \end{frame}
@@
 \begin{frame}
-\frametitle{Conclusions I}
+\frametitle{Conclusions}
 \begin{itemize}
-\item Cointegration relations change with time. We empirically showed that the Johansen method is sensitive to the number of lags but also to the amount of data considered.
 %\item We found that out-of-sample forecast performance MSE is related to the {\em percentage of cointegration\/}.  
@@
 %This result is very important since it is still difficult to outperform ARIMA and the random walk model for standard econometric forecasting models  despite their simplicity.
-\end{itemize}
-\end{frame}
-
-\begin{frame}
-\frametitle{Conclusions II}
-\begin{itemize}
-\item The parallel implementation allowed the execution times to be reduced
-more than 9 times and therefore a response time was obtain before 10
-seconds.
 \item We showed that VECM computational limitations can be tackled using a an online approach in a distributed environment.

### 2017-12-15 slides/defense-slides.tex
reducing slides

@@
 \begin{frame}
-\frametitle{HFT: High Frequency Trading}
-High Frequency Trading has motivated computer-driven strategies capable of processing large amounts of data in short periods of time.
-\begin{figure}
-\includegraphics[width=0.7\paperwidth]{img/HFTmarket}
-\caption{Percentage of HFT trades of total US Equity Trading. Source: TABB group}
-\end{figure}
-\end{frame}
-
-\begin{frame}
-\frametitle{HFT Revenue}
-However, the revenues have fallen dramatically mainly because of more regulation, increasing the cost of HFT infrastructure and more competitive algorithms.
-\begin{figure}
-\includegraphics[width=0.7\paperwidth]{img/HFTmarket2}
-\caption{Revenue in the US. Source: TABB group}
-\end{figure}
-\end{frame}
-
-\begin{frame}
 \frametitle{Motivation: HFT}
@@
 {\color{blue}
- \begin{eqnarray*} E(y_t) &=& \mu  \quad
-\forall t \in \mathbb{N} \\ E(y^2_t) &=& \sigma^2  \quad \forall t \in
+ \begin{eqnarray*} E[y_t] &=& \mu  \quad
+\forall t \in \mathbb{N} \\ E[y^2_t] &=& \sigma^2  \quad \forall t \in
 \mathbb{N} \\ \lambda(s,t)&=&\lambda(s+h,t+h) \quad \forall s,t \in \mathbb{N},
@@
 \end{block}
-\noindent with $\lambda(s,t) = E[(y_s-\mu)(y_t - \mu)]$ 
-\end{frame}
-
-\begin{frame}
-\frametitle{Stationarity: consequences}
-One of the consequences of stationary processes is how they recover from a shock.
-A shock represents an unexpected change in a variable in a particular period of time.
-For stationary time series, shocks to the system will gradually die away.
-\begin{figure}
-\includegraphics[width=0.6\paperwidth]{img/shock}
-\end{figure}
+\noindent with $\lambda(s,t) = E[(y_s-\mu)(y_t - \mu)]$ and $E[y_t]= \displaystyle \int_0^\infty t p(t)dt$. 
 \end{frame}
@@
 \noindent where $\epsilon_t$ is a stationary process. When $\mu = 0 $ the process is called {\bf pure random walk} and when $\mu \neq 0$  the process is called {\bf random walk with drift}.
-\end{block}
-\end{frame}
-
-\begin{frame}
-\frametitle{Integration }
-\begin{block}{Integration of order $d$}
-A time series $\mathbf{y}$ is said to be I(d), or integrated of order $d$, if after
-differentiating the variable $d$ times, we get an stationary process, more precisely:
-\[
-{\color{blue}
-(1-L)^d \mathbf{y} \sim \text{I(0)}} \, ,
-\]
-\noindent where I(0) is a stationary time series and $L$ is the lag operator:
-\[
-(1-L)\mathbf{y} = \Delta \mathbf{y}=\mathbf{y}_t  -\mathbf{y}_{t-1} \quad \forall t
-\]
 \end{block}
@@
 
-\begin{frame}
-\frametitle{Online learning algorithms}
-\begin{algorithm}[H]
-\begin{algorithmic}[1]
-    \STATE Receives input $\mathbf{x}_t$
-    \STATE Makes prediction $\mathbf{\hat{y}}_t$
-    \STATE Receives response $\mathbf{y}_t$
-    \STATE Incurs loss $l_t(\mathbf{y}_t,\mathbf{\hat{y}}_t)$
-\end{algorithmic}
-\caption{Structure of an Online Learning System}
-\end{algorithm}
-Performance is later measured after $T$ trials as:
-\begin{equation*}
-L_T = \sum_{t=1}^T l_t(\mathbf{y}_t,\mathbf{\hat{y}}_t)
-\end{equation*}
-\end{frame}
-
 \section{Challenges and Hypothesis}
@@
         \STATE read new $\mathbf{x}_t$
+        \STATE output prediction $f(\mathbf{x}_t) =  \mathbf{b}^\intercal
+\mathbf{A}^{-1}\mathbf{x}_t$
 \STATE $\mathbf{A} = \mathbf{A} + \mathbf{x}_t \mathbf{x}_t^\intercal-
 \mathbf{x}_{t-L-1} \mathbf{x}_{t-L-1}^\intercal  $
-        \STATE output prediction $f(\mathbf{x}_t) =  \mathbf{b}^\intercal
-\mathbf{A}^{-1}\mathbf{x}_t$
         \STATE Read new $\mathbf{y}_t$
@@
 \begin{enumerate}
-\item It uses VECM only considering a sliding windows of size $L$ of data.
-\item We proposed to choose $L$ and the number of lags $p$ in order to maximise the percentage of cointegration $PC$ in the near past. This process was done every time that new data was processed. 
-\item And to use a distributed environment to make this parameters search, since this is the most expensive routine.
+\item It uses VECM only considering a {\bf sliding windows of size $L$} of data.
+\item We proposed to choose $L$ and the number of lags $p$ in order to {\bf maximise the percentage of cointegration $PC$} in the near past.
+This process was done every time that new data was processed.
+\item And to use a {\bf distributed environment to make this parameters search}, since this is the most expensive routine.
 \end{enumerate}
@@
 \centering
-\small
+\tiny
 \begin{tabular}{llllll}
@@
 \caption{Unit roots tests for EURUSD, GBPUSD, USDCHF and USDJPY at 10-second
-frequency.}
+frequency. MacKinnon critical values for rejection of hypothesis of a unit root are: -2.62 (1\%), -1.94 (5\%) and -1.62 (10\%)}
 \label{tab:adf}
@@
 \item[Cointegration]Cointegration relations change with time.
-We empirically showed that the number of coitegration vectors is sensitive to the number of lags but also to the amount of data considered.
+We empirically showed that the number of cointegration vectors is sensitive to the number of lags but also to the amount of data considered.
 The percentage of cointegration is used to find VECM parameters.
@@
 
-\begin{frame}
-\frametitle{Stationarity}
-A time series $y_t: \Z \to \R $ is said to be strictly stationary if, for a finite set of $L$ indices $\{t_{1},\dots,t_{L}\}  \, \forall L \in \Z_{\ge 0}$, the joint probability distribution of the random variables $\{y_{t_{1}},y_{t_{2}},\dots,y_{t_{L}}\}$ remains unchanged for any time shifted $h$.
-The probability distribution for the random variable $y_{t_1}$ is:
-
-\begin{equation}
-P\{y_{t} \leq c_1 \} = P\{t \in \Z:  y_t \in  ]-\infty, c_1] \} 
-\end{equation}
-
+\section{HFT}
+\begin{frame}
+\frametitle{HFT: High Frequency Trading}
+High Frequency Trading has motivated computer-driven strategies capable of processing large amounts of data in short periods of time.
+\begin{figure}
+\includegraphics[width=0.7\paperwidth]{img/HFTmarket}
+\caption{Percentage of HFT trades of total US Equity Trading. Source: TABB group}
+\end{figure}
+\end{frame}
+
+\begin{frame}
+\frametitle{HFT Revenue}
+However, the revenues have fallen dramatically mainly because of more regulation, increasing the cost of HFT infrastructure and more competitive algorithms.
+\begin{figure}
+\includegraphics[width=0.7\paperwidth]{img/HFTmarket2}
+\caption{Revenue in the US. Source: TABB group}
+\end{figure}
 \end{frame}
@@
 
-\begin{frame}
-\frametitle{Unit root process}
-
+\section{Shock}
+
+\begin{frame}
+\frametitle{Stationarity: consequences}
+One of the consequences of stationary processes is how they recover from a shock.
+A shock represents an unexpected change in a variable in a particular period of time.
+For stationary time series, shocks to the system will gradually die away.
+\begin{figure}
+\includegraphics[width=0.6\paperwidth]{img/shock}
+\end{figure}
+\end{frame}
+
+\section{Order of Integration}
+\begin{frame}
+\frametitle{Integration }
+\begin{block}{Integration of order $d$}
+A time series $\mathbf{y}$ is said to be I(d), or integrated of order $d$, if after
+differentiating the variable $d$ times, we get an stationary process, more precisely:
+\[
+{\color{blue}
+(1-L)^d \mathbf{y} \sim \text{I(0)}} \, ,
+\]
+\noindent where I(0) is a stationary time series and $L$ is the lag operator:
+\[
+(1-L)\mathbf{y} = \Delta \mathbf{y}=\mathbf{y}_t  -\mathbf{y}_{t-1} \quad \forall t
+\]
+\end{block}
 \end{frame}
@@
 
+\section{Spurious regression}
 \begin{frame}
@@
 \begin{columns}
-\begin{column}{.45\linewidth}<1->
+\begin{column}{.45\linewidth}
 \begin{figure}
@@
 \end{column}
-\begin{column}{.45\linewidth}<2->
+\begin{column}{.45\linewidth}
 \begin{figure}
@@
 
+\section{Random walk forecast}
 \begin{frame}
@@
 %
+\section{ARIMA}
 \begin{frame}
@@
 \noindent with coefficients $\phi_p \neq 0$, $\theta_q \neq 0$ and $\sigma_{\epsilon}^2 > 0$.
+\end{frame}
+
+\section{Online Learning System}
+\begin{frame}
+\frametitle{Online learning algorithms}
+\begin{algorithm}[H]
+\begin{algorithmic}[1]
+    \STATE Receives input $\mathbf{x}_t$
+    \STATE Makes prediction $\mathbf{\hat{y}}_t$
+    \STATE Receives response $\mathbf{y}_t$
+    \STATE Incurs loss $l_t(\mathbf{y}_t,\mathbf{\hat{y}}_t)$
+\end{algorithmic}
+\caption{Structure of an Online Learning System}
+\end{algorithm}
+Performance is later measured after $T$ trials as:
+\begin{equation*}
+L_T = \sum_{t=1}^T l_t(\mathbf{y}_t,\mathbf{\hat{y}}_t)
+\end{equation*}
 \end{frame}
@@
 
+\section{Online Ridge Regression}
+
+\begin{frame}
+\frametitle{Online Ridge Regression}
+\begin{eqnarray*}
+\label{eq:RReapand}
+\mathbf{\mathbf{W}}_{\text{ridge}}&=&(\mathbf{X}^\top \mathbf{X}+ \lambda
+\mathbb{I})^{-1}\mathbf{X}^\top \mathbf{Y} \\
+&=& \displaystyle \big (\sum_{t=1}^m
+\mathbf{x}_t \mathbf{x}_t  ^\top + \lambda \mathbb{I}\big )^{-1}
+\sum_{t=1}^m \mathbf{x}_t \mathbf{y}_t \, .
+\end{eqnarray*}
+\end{frame}
+
+\begin{frame}
+%\lambda \mathbb{I}
+\begin{algorithm}[H]
+\begin{algorithmic}[1]
+\REQUIRE $\,$ \\
+$\{\mathbf{x}_1,\dots,\mathbf{x}_N \}$: $N$ input vectors \\
+$\{\mathbf{y}_1,\dots,\mathbf{y}_N \}$: $N$ targets \\
+$L$: sliding window size ($L<N$) \\
+\ENSURE  $\,$ \\
+$\{f(\mathbf{x}_{L+1}),\dots,f(\mathbf{x}_N) \}$: model predictions \\
+\STATE Initialize $\mathbf{A}=\displaystyle \sum_{t=1}^L \mathbf{x}_t
+\mathbf{x}_t^\intercal + {\color{red}\lambda \mathbb{I}}$
+and $\mathbf{b}=\displaystyle \sum_{t=1}^L \mathbf{x}_t \mathbf{y}^\intercal_{t}$
+\FOR { $t = L+1$ to $N$ }
+        \STATE read new $\mathbf{x}_t$
+        \STATE output prediction $f(\mathbf{x}_t) =  \mathbf{b}^\intercal
+\mathbf{A}^{-1}\mathbf{x}_t$
+\STATE $\mathbf{A} = \mathbf{A} + \mathbf{x}_t \mathbf{x}_t^\intercal-
+\mathbf{x}_{t-L-1} \mathbf{x}_{t-L-1}^\intercal  $
+        \STATE Read new $\mathbf{y}_t$
+        \STATE $\mathbf{b} = \mathbf{b} + \mathbf{x}_t \mathbf{y}^\intercal_{t}$
+\ENDFOR
+\end{algorithmic}
+\caption{Online Ridge Regression}
+\label{alg:SLAAR}
+\end{algorithm}
+\end{frame}
+
+\begin{frame}
+\begin{algorithm}[H]
+\begin{algorithmic}[1]
+\REQUIRE $\,$ \\
+$\{\mathbf{x}_1,\dots,\mathbf{x}_N \}$: $N$ input vectors \\
+$\{\mathbf{y}_1,\dots,\mathbf{y}_N \}$: $N$ targets \\
+$L$: sliding window size ($L<N$) \\
+\ENSURE  $\,$ \\
+$\{f(\mathbf{x}_{L+1}),\dots,f(\mathbf{x}_N) \}$: model predictions \\
+\STATE Initialize $\mathbf{A}=\displaystyle \sum_{t=1}^L \mathbf{x}_t
+\mathbf{x}_t^\intercal + {\color{red}\lambda \mathbb{I}}$
+and $\mathbf{b}=\displaystyle \sum_{t=1}^L \mathbf{x}_t \mathbf{y}^\intercal_{t}$
+\FOR { $t = L+1$ to $N$ }
+        \STATE read new $\mathbf{x}_t$
+        \STATE {\color{red} $\mathbf{A} = \mathbf{A} + \mathbf{x}_t \mathbf{x}_t^\intercal-
+\mathbf{x}_{t-L-1} \mathbf{x}_{t-L-1}^\intercal$}
+        \STATE output prediction $f(\mathbf{x}_t) =  \mathbf{b}^\intercal
+\mathbf{A}^{-1}\mathbf{x}_t$
+        \STATE Read new $\mathbf{y}_t$
+        \STATE $\mathbf{b} = \mathbf{b} + \mathbf{x}_t \mathbf{y}^\intercal_{t}$
+\ENDFOR
+\end{algorithmic}
+\caption{The Aggregating Algorithm for Regression}
+\label{alg:SLAAR}
+\end{algorithm}
+\end{frame}
 % Financial time series are not the only ones with this behaviour, this study can be extended to other non-stationary time series such as: weather, earthquakes, energy demand and sales forecasting.
 
+\section{The Johansen method}
+
+\begin{frame}
+\frametitle{The Johansen method}
+Johansen in 1988 suggests a method for determining
+cointegration vectors. 
+There are two statistics for cointegration under Johansen approach: the trace
+(shown in equation \ref{eq:johansentrace}) and the maximum-eigenvalue statistic
+(shown in equation \ref{eq:johansenmax}):
+\begin{equation}
+\label{eq:johansentrace}
+\lambda_{\text{trace}} (r) = -T \sum_{i=r+1}^g \ln(1-\hat{\lambda_i})
+\end{equation}
+\begin{equation}
+\label{eq:johansenmax}
+\lambda_{\text{max}} (r,r+1) = -T \ln(1-\hat{\lambda_{r+1}})
+\end{equation}
+\noindent where $r$ is the number of cointegration vectors and
+$\hat{\lambda}_i$ is the estimated value for the ith ordered eigenvalue.
+$\lambda_{\text{trace}}$ null hypothesis is that the number of cointegration
+vectors is less than or equal to $r$ against that there are more than $r$.
+$\lambda_{\text{max}}$ has the null hypothesis that the number of cointegration
+vectors is $r$ against an alternative of $r +1$.
+\end{frame}
+
+\section{Evaluation methods}
+\begin{frame}
+\frametitle{Evaluation methods}
+\begin{description}
+\item[MSE]  Mean Square Error measures the distance between forecasts
+and the true values and large deviations from the true value have a
+large impact due to squaring forecast error.
+\begin{equation}\label{eq:MSE}
+\text{MSE} = 
+\frac{\displaystyle \sum_{t=1}^{N} (\mathbf{y}_t-\hat{\mathbf{y}}_t)^2}{N}
+\end{equation}
+\item[$U$-statistic] the Theil's $U$-statistic, presented by, is a unit free measure obtained as the ratio between the root
+MSE (RMSE) of a model and the RMSE of the naive random walk model. 
+\end{description}
+\end{frame}
+
+\section{Inverse calculation}
+\begin{frame}
+\frametitle{Sherman-Morrison-Woodbury}
+The Sherman-Morrison-Woodbury formula can be used to reduce inverse matrix
+This
+formula allows to get the inverse of a
+the matrix $\mathbf{A+uv^\top}$ if we already calculated the
+inverse of $\mathbf{A}$ as follows:
+
+\begin{equation}
+\label{eq:SMW}
+(\mathbf{A+uv^\top})^{-1}=\mathbf{A}^{-1}-
+\frac{\mathbf{A^{-1}uv^\top A^{-1}}}{\mathbf{1+v^\top A^{-1}u}}
+\end{equation}
+\end{frame}
+
 \end{document}

### 2017-12-15 slides/defense-slides.tex
Ready to present

@@
 
+\section{Volatility models}
+
+\begin{frame}
+\frametitle{Volatility models}
+For many years the most popular methods for estimating financial volatility were
+the autoregressive conditional heteroskedasticity (ARCH) models
+and the general ARCH (GARCH) models. For instance, the
+GARCH(1,1) defines returns $y_t$ and volatility $\sigma_t$ as:
+
+\begin{eqnarray*}
+    y_t &=& \sigma_t \epsilon_t \\
+     \sigma_t^2 &=& \alpha_0 + \alpha_1 \epsilon_{t-1}^2 + \beta_1
+     \sigma_{t-1}^2
+\end{eqnarray*}
+
+\noindent where $\epsilon_t$ is standard Gaussian white noise,
+$\alpha_0,\alpha_1,\beta_1 \geq 0$ are required to ensure that the
+variance will never be negative and $\alpha_1+\beta_1 <1$ is needed to
+guarantee a weakly stationary process.
+\end{frame}
+
 \end{document}

### 2017-12-18 articles-conferences/jcc2012/JCC2012/src/AAR.tex
Adding all before the exam

\section{The Aggregating Algorithm for Regression}

The AAR, proposed by~\cite{vovk2001}, is an application of the
aggregating algorithm to the problem of regression.

The prediction formula for AAR is given by
equation~(\ref{eq:predAAR}):

\begin{equation}
\label{eq:predAAR}
f(\mathbf{x}_{m+1}) = \displaystyle \big( \sum_{t=1}^m y_t \mathbf{x}_t \big )^\intercal 
  \big ( \sum_{t=1}^{m+1} \mathbf{x}_t \mathbf{x}_t  ^\intercal + \gamma
\mathbb{I})\big )^{-1}  \mathbf{x}_{m+1} \, ,
\end{equation}

\noindent which is very similar to the RR method except because AAR includes
information of the new input $\mathbf{x}_{m+1}$.  This means that AAR
updates matrix $A$ with the new input $\mathbf{x}_{m+1}$ before the
prediction $f(\mathbf{x}_{m+1})$ is made.

Algorithm~\ref{alg:AAR} shows this procedure based on
equation~(\ref{eq:predAAR}). This algorithm can be obtained by swapping lines 4 and 5 from
algorithm~\ref{alg:RR}.

\begin{algorithm}[ht]
\begin{algorithmic}[1]
\REQUIRE $\,$ \\
$\{\mathbf{x}_1,\dots,\mathbf{x}_m \}$: $m$ input vectors \\
$\{y_1,\dots,y_m \}$: $m$ targets \\
\ENSURE  $\,$ \\
$\{f(\mathbf{x}_1),\dots,f(\mathbf{x}_m) \}$: model predictions \\
\STATE Initialize $\mathbf{A}=\gamma \mathbb{I}$
and $\mathbf{b}=0$
\FOR { $t = 1$ to $m$ }
   	\STATE read new $\mathbf{x}_t$
	\STATE $\mathbf{A} = \mathbf{A} + \mathbf{x}_t \mathbf{x}_t^\intercal$
	\STATE output prediction $f(\mathbf{x}_t) =  \mathbf{b}^\intercal \mathbf{A}^{-1}\mathbf{x}_t$
   	\STATE Read new $y_t$
    	\STATE $\mathbf{b} = \mathbf{b} + y_t \mathbf{x}_t$
\ENDFOR

\end{algorithmic}
\caption{{\em The aggregating algorithm for regression}}
\label{alg:AAR}
\end{algorithm}

### 2017-12-18 articles-conferences/jcc2012/JCC2012/src/experiments.tex
Adding all before the exam

\section{Experimental Results}

In order to measure performance of our algorithm, we used daily stock
returns of 45 stocks from the technology sector dated from the 1st of
January 2000 to the 1st of July 2012.

\begin{figure}[h]
  \centering
  \subfigure[AAR v/s the target for SPY]{
  \includegraphics[width=0.4\textwidth]{img/AAR}
  %\caption{AAR v/s the target for SPY}
  \label{fig:data}
  }
  \subfigure[RR v/s the target for SPY]{
  \includegraphics[width=0.4\textwidth]{img/RR}
  %\caption{AAR v/s the target for SPY}
  \label{fig:data2}
  }
  \subfigure[SLAAR v/s the target for SPY]{
  \includegraphics[width=0.4\textwidth]{img/SLAAR}
  %\caption{AAR v/s the target for SPY}
  \label{fig:data3}
  }
\end{figure}

Returns $\{x_t\}_{t=1}^{m-1}$ were defined in based on stock prices
$\{{p_t}\}_{t=1}^m$ and are related as:

\begin{equation*}
x_t = \frac{p_{t+1}-p_t}{p_t} \, .
\end{equation*}

The objective is to build a model of the returns of stock $k$ based on
returns of other stocks of the same financial sector.

Therefore, every input vector $\mathbf{x}_t$ will have 44 stock
returns at time $t$ without considering information of stock $k$. The
target vector $\mathbf{y}$ will be the stocks returns of stock $k$.

We compared RR, AAR and our proposal called sliding windows AAR
(SLAAR) and the results are shown in table~\ref{tab:results}. The MSE
was shown in bold when our method was better than RR and AAR. The
table shows that our method outperforms in 28 of 45 stocks.

\begin{table}[h!]
    \centering
    \footnotesize
    \begin{tabular}{*{5}{r}}
        \toprule
          & AAR & RR & SLAAR& Best L \\
         \midrule
Stock  &  AAR error  &   RR error  &   SLAAR error  &   Best L  \\
IBM  &  0.230307  &  0.231317  &  \textbf{0.228963}  &  70   \\
MMM  &  0.199237  &  0.197172  &  \textbf{ 0.195550} &  990\\
CVX  &  0.155454  &  0.152070  &  0.155150  &  750\\
UTX  &  0.260980  &  0.268351  &  0.261140  &  340\\
CAT  &  0.223338  &  0.222240  &  \textbf{ 0.210500} &  360\\
MCD  &  0.268211  &  0.266997  &  0.267182  &  990\\
BA  &  0.292699  &  0.295623  &  \textbf{ 0.292180} &  150\\
XOM  &  0.160376  &  0.157125  &  0.159074  &  780\\
JNJ  &  0.213276  &  0.214747  &  \textbf{ 0.212601}  &  680\\
PG  &  0.494937  &  0.493036  &  0.495395  &  950\\
KO  &  0.272203  &  0.274337  &  \textbf{ 0.270888} &  870\\
WMT  &  0.272945  &  0.269190  &  \textbf{ 0.254820}  &  630\\
HPQ  &  0.276595  &  0.275459  &  0.280236  &  980\\
AXP  &  0.204162  &  0.211216  &  \textbf{ 0.195888}  &  230\\
DD  &  0.204220  &  0.202591  &  0.203177  &  330\\
JPM  &  0.238405  &  0.255824  &  \textbf{ 0.204777}  &  290\\
MRK  &  0.214294  &  0.211708  &  \textbf{ 0.196368}  &  850\\
DIS  &  0.324324  &  0.326105  &  \textbf{ 0.297895}  &  290\\
VZ  &  0.203936  &  0.206390  &  \textbf{ 0.196078}  &  490\\
HD  &  0.236339  &  0.237688  &  0.242606  &  990\\
T  &  0.235940  &  0.234002  &  \textbf{ 0.230230}  &  570\\
MSFT  &  0.332521  &  0.331694  &  0.334283  &  710\\
CSCO  &  0.227675  &  0.228196  &  0.233057  &  990\\
INTC  &  0.262545  &  0.264391  &  0.271204  &  1000\\
GE  &  0.168935  &  0.168258  &  0.168924  &  1000\\
PFE  &  0.200335  &  0.199163  &  0.204552  &  1000\\
BAC  &  0.299956  &  0.333396  &  \textbf{ 0.200134}  &  530\\
AA  &  0.249913  &  0.250263  &  \textbf{ 0.246067}  &  1000\\
XLF  &  0.300486  &  0.299604  &  0.301322  &  830\\
EWH  &  0.186584  &  0.188475  &  \textbf{ 0.180674}  &  260\\
EWG  &  0.156459  &  0.158946  &  \textbf{ 0.152092}  &  470\\
EWA  &  0.231424  &  0.233347  &  \textbf{ 0.224692}  &  960\\
XLV  &  0.140344  &  0.139772  &  \textbf{ 0.118340}  &  610\\
XLI  &  0.089607  &  0.087342  &  \textbf{ 0.086304}  &  450\\
XLU  &  0.107657  &  0.106929  &  \textbf{ 0.091362}  &  550\\
XLY  &  0.103068  &  0.102359  &  \textbf{ 0.095318}  &  590\\
XLB  &  0.105090  &  0.102791  &  \textbf{ 0.099280}  &  940\\
XLE  &  0.118998  &  0.117319  &  \textbf{ 0.111371} &  1000\\
SPY  &  0.073800  &  0.071908  &  \textbf{ 0.067655} &  230\\
EWS  &  0.223738  &  0.223721  &  0.228166  &  990\\
EWC  &  0.172954  &  0.174107  &  \textbf{ 0.164855} &  330\\
EWU  &  0.146232  &  0.145031  &  0.146491  &  640\\
EWW  &  0.205101  &  0.203012  &  \textbf{ 0.199228}  &  80\\
EWM  &  0.249645  &  0.248694  &  0.257393  &  80\\
XLK  &  0.098086  &  0.098599  &  \textbf{ 0.097944}  &  440  \\

        \bottomrule
    \end{tabular}
    \caption{MSE and best L}
    \label{tab:results}
\end{table}

%\begin{figure}[h!]
%  \centering
%    \includegraphics[width=0.55\textwidth]{img/results}
%  \caption{Comparison AAR, RR and SLAAR for SPY}
%  \label{fig:data}
%\end{figure}

%\begin{figure}[h!]
%  \centering
%    \includegraphics[width=0.45\textwidth]{img/RR}
%  \caption{RR v/s the target for SPY}
%  \label{fig:data2}
%\end{figure}
%
%
%\begin{figure}[h!]
%  \centering
%    \includegraphics[width=0.45\textwidth]{img/SLAAR}
%  \caption{SLAAR v/s the target for SPY}
%  \label{fig:data3}
%\end{figure}

The error was calculated considering the {\em mean squared error}
(MSE) using the last 70 predictions made.

Table~\ref{tab:results} also shows that the smaller error in our
method was for the SPY stock, which is an expected result considering
that the SPY value is constructed based on the main stocks of the
markets.

A graphical comparison between AAR, RR and SLAAR with the target for
stock SPY is shown in figures~\ref{fig:data},\ref{fig:data2} and
\ref{fig:data3}. It shows how AAR and our method fit better than RR to
the target and also that our method is slightly better compared with
AAR. It's also possible to notice that when returns change drastically
none of the methods fit satisfactorily.

\section{Discussion and Conclusion}

The existing approaches to solve regression problems are well known,
however they are not used in an online context because they involve
many calculations and in order to make a prediction, they consider all
the data available. However, in some cases, only a portion of data is
needed or available to make a prediction and sometimes the accuracy
could be improved considering less data.

In this article, this historical dependence of data is studied and the
results show that in some cases, if we use a sliding window of data
instead of all the data, the MSE of the predictions is minimized.
It's possible to check that there are some stocks which depend more on
the historical data than others and this is valuable information for
future algorithms.

On the other hand, RR and AAR calculates an inverse matrix at every
step, which is computationally very expensive. In our proposal, this
inverse matrix calculation is avoided and therefore it's more suitable
to be used in an online context.

### 2017-12-18 articles-conferences/jcc2012/JCC2012/src/intro.tex
Adding all before the exam

\section{Introduction}
The learning from examples problem is an ill-posed problem which
admits an infinite number of solutions.  In order to restrict the
space of admissible solutions, the regression problem is usually
formulated in terms of regularization theory~\cite{girosiETal1995} as
an optimization problem, which minimizes the functional:

\begin{equation}
\label{eq:problem} 
J(\mathbf{w}) = \sum_{t=1}^m (y_t - f(\mathbf{x}_t))^2 + \gamma R(\mathbf{w})
, \qquad f \in \mathcal{H} \, ,
\end{equation}

\noindent where $m$ is the number of samples $(\mathbf{x}_t ,y_t)$ with
$\mathbf{x}_t \in \R^l$, $l$ correspond to the number of features, $y_t
\in \R$ is the target, $R(\mathbf{w}) = ||w||^2$ where $||.||^2_K$ is a norm in a
{\em reproducing kernel hilbert space} $\mathcal{H}$ defined by the
positive definite form $K$, and $\gamma$ is a regularization
parameter~\cite{evgeniouETal2000}.
%Qué norma debe usarse? falta definir K
When the hypothesis space is reduced, the risk of overfitting the training data
decreases and therefore leading to better generalization capability. 
%aclarar porqué se reduce el espacio de hipótesis
{\em Ridge regression} (RR) is a batch method generally used to solve this
problem, which is a generalization of least squares method (LS).

%RR is a batch method and it is not used in an online context mainly because
%it uses large amount of data and the number of operations increases
%with data size.

However, online algorithms are more attractive than batch algorithms because
their simplicity and ability to manage large data sets, this is why
they are popular in financial applications.

There are several popular online methods such as
perceptron~\cite{rosenblatt58},
passive-aggressive~\cite{crammerETall2006}, stochastic gradient
descent~\cite{zhang2004}, aggregating algorithm~\cite{vovk2001} and
the second order perceptron~\cite{cesa-bianchi2005}.
In~\cite{cesa-bianchi2006} it is provided an in-deph analysis of
online learning.

The {\em aggregating algorithm for regression} (AAR) method formulates
a recursive formulation of RR in an online way. AAR
consider all data to make a prediction, but in highly variant
scenarios the old data could be useless.

In this paper, we propose an online method based on the idea presented
by~\cite{vovk2001} considering in our case a single sliding window of
the most recent data. This proposal also reduces the number of
operations at every step of the algorithm by expressing the inverse
matrix in an iterative form. Our algorithm is later tested with
financial data from stock market.

### 2017-12-18 articles-conferences/jcc2012/JCC2012/src/model-lsr.tex
Adding all before the exam

\section{Least Squares with Regularization} \label{sec:proposal}

The given data is a sequence $\left\{ \left(d(k),\mathbf{u}(k) \right) \right\}_{k\in\N}$ where $
d(k)\in\R $ and $\mathbf{u}(k) = \big[ u_1(k), u_2(k), \dots\,, u_N(k) \big]^T\in\R^N\,$.

For each fixed $k\in\N$, $k\geq L$, we want to find scalars $w_j\in\R$,
$j\in\N$, $1\leq j\leq N$,
which minimize the functional:

\begin{eqnarray}
J_k(\mathbf{w}) &=\sum_{i=k-L+1}^{k}\left( d(i) - \mathbf{w}^T \mathbf{u}(i)\right)^2 + \gamma \sum_{i=1}^{L} w_i^2 \nonumber\\
&=\sum_{i=k-L+1}^{k}\left( d(i) - \sum_{j=0}^{N} w_j u_j(i)\right)^2 + \gamma \sum_{i=1}^{L} w_i^2
\label{eq:f1}
\end{eqnarray}

Note that $w_j$ is the factor associated with the input data
$u_j(k),u_j(k-1),u_j(k-2),\dots,u_j(k-L+1)$.

The equivalent matricial representation of equation (\ref{eq:f1}) is the following:

\begin{equation}
J_k(\mathbf{w})=\bigg\| 
\begin{bmatrix}
d(k) \\ d(k-1) \\ d(k-2) \\ \vdots \\ d(k-L+1)
\end{bmatrix}
- \begin{bmatrix}
\, & \mathbf{u}(k)^T & \, \\
\, & \mathbf{u}(k-1)^T & \, \\
\, & \mathbf{u}(k-2)^T & \, \\
\,  & \vdots & \, \\
\, & \mathbf{u}(k-L+1)^T & \,
\end{bmatrix}
\begin{bmatrix}
w_1 \\ w_2 \\ w_3 \\ \vdots \\ w_N
\end{bmatrix}
\bigg\|_{L^2}^2
+
\gamma \bigg\|
\begin{bmatrix}
w_1 \\ w_2 \\ w_3 \\ \vdots \\ w_N
\end{bmatrix}
\bigg\|_{L^2}^2
\label{eq:f2}
\end{equation}

A reduced expression of (\ref{eq:f2}) is:

\begin{equation}
J_k(\mathbf{w})
= \big\|\mathbf{d}(k)-\mathbf{U}(k)\mathbf{w} \big\|_{L^2}^2 + \gamma \big\| \mathbf{w} \big\|_{L^2}^2 \,,
\end{equation}

where $\mathbf{d}(k)\in\R^{L}\,,\;
\mathbf{U}(k)=\left[ u_j(k) \right]_{(L)\times N}\in\R^{(L)\times N}\,,\;
\mathbf{w}\in\R^{N}$.

The solution of this optimization problem is:

\begin{equation}\label{20120421:1}
\mathbf{w} = \big(\mathbf{U}(k)^T\mathbf{U}(k) + \gamma \mathbb{I} \big)^{-1}\mathbf{U}(k)^T\mathbf{d}(k)\,.
\end{equation}
Equation (\ref{20120421:1}) gives the best value of the unknown
$\mathbf{w}\in\R^N$ {\em at the instant\/} $k$, in de sense that it
minimizes the value of $J_k(\mathbf{w})$.

\subsection{Recursive formulation}

If a new pair of data arrives$(d(k+1),\mathbf{w}(k+1))$ the matrix $\mathbf{U}(k+1)$ will be:

\begin{equation}
\mathbf{U}(k+1)=
\begin{bmatrix}
\, & \mathbf{u}(k+1)^T & \, \\
\, & \mathbf{u}(k)^T & \, \\
\, & \mathbf{u}(k-1)^T & \, \\
\,  & \vdots & \, \\
\, & \mathbf{u}(k-L+2)^T & \,
\end{bmatrix}\, .
\end{equation}

Using the new information the best value of the
unknown $\mathbf{w}\in\R^N$ at instant $k+1$ will be:

\begin{equation}\label{eq:f3}
\mathbf{w} = \big(\mathbf{U}(k+1)^T\mathbf{U}(k+1) + 
\gamma \mathbb{I} \big)^{-1}\mathbf{U}(k+1)^T\mathbf{d}(k+1)\,.
\end{equation}

The matrix $\mathbf{U}(k+1)$ differs from the previous matrix 
$\mathbf{U}(k)$ only because it was added the vector 
$\mathbf{u}(k+1)$ and removed the last row ($\mathbf{u}(k-L+1)$).
Using this fact, we will formulate a recursive form of the equation 
(\ref{eq:f3}) based on that:

\begin{equation}\label{eq:f4}
\mathbf{U}(k+1)^T\mathbf{U}(k+1)= 
\mathbf{U}(k)^T\mathbf{U}(k) + A(k)\, ,
\end{equation}

where, 
\begin{equation}
A(k) =\mathbf{u}(k+1) \mathbf{u}(k+1)^T - \mathbf{u}(k-L+1) \mathbf{u}(k-L+1)^T \; .
\label{eq:f41}
\end{equation}

Considering equations (\ref{eq:f4}) and (\ref{eq:f41}) we can express the inverse term of equation (\ref{eq:f3}) as:

\begin{align}
\big(\mathbf{U}(k+1)^T\mathbf{U}(k+1) + 
\gamma \mathbb{I} \big)^{-1}= &
 \nonumber\\
= \big(\mathbf{U}(k)^T\mathbf{U}(k) + A(&k) +
\gamma \mathbb{I} \big)^{-1} \nonumber\\
= \big( ( \mathbb{I} + A(k) \big(\mathbf{U}(k)^T&\mathbf{U}(k) +  \gamma \mathbb{I} \big)^{-1} ) 
\big(\mathbf{U}(k)^T\mathbf{U}(k) + \gamma \mathbb{I} \big) \big)^{-1} \,
\label{eq:f5}
\end{align}

and for simplicity we define:

\begin{equation}
	\Delta(k) = A(k) \big(\mathbf{U}(k)^T\mathbf{U}(k) +  \gamma \mathbb{I} \big)^{-1},
\end{equation}

so we can reduce the equation (\ref{eq:f5}) as the following:

\begin{eqnarray}
	\big(\mathbf{U}(k+1)^T\mathbf{U}(k+1) + 
\gamma \mathbb{I} \big)^{-1}=&\big( ( \mathbb{I} + \Delta(k)) 
\big(\mathbf{U}(k)^T\mathbf{U}(k) + \gamma \mathbb{I} \big) \big)^{-1} \nonumber \\
=& \big(\mathbf{U}(k)^T\mathbf{U}(k) + \gamma \mathbb{I} \big) ^{-1} ( \mathbb{I} + \Delta(k)) ^{-1}.
\label{eq:f6}
\end{eqnarray}

Equation (\ref{eq:f6}) can be approximated using the Taylor expansion for matrices:

\begin{equation}
\begin{split}
	\big(\mathbf{U}(k+1)^T\mathbf{U}(k+1) + 
\gamma \mathbb{I} \big)^{-1} & \approx \\
\big(\mathbf{U}(k)^T\mathbf{U}(k) &+ \gamma \mathbb{I} \big) ^{-1} 
( \mathbb{I} - \Delta(k) + \Delta(k)^2 \pm \dots),
\end{split}
\end{equation}

where the term $\big(\mathbf{U}(k)^T\mathbf{U}(k) + 
\gamma \mathbb{I} \big) ^{-1}$ has been previously calculated at time $k$.
By this way is avoided the inverse calculation, which is the most expensive
procedure in the algorithm.

\subsection{Matrix Inversion} \label{sec:matrixinv}

The matrix inverse can be improved using the approximation used in \cite{obrien1982}. If we want to obtain the matrix inverse of $A \in Mat(n,n)$ by using some numerical approximation and we get $B$ then:

\begin{equation} \label{eq:m1}
A^{-1} = B + \epsilon \, ,
\end{equation}

where $\epsilon$ is a $n \times n$ matrix with the error approximations values. This $\epsilon$ matrix can be approximated by premultiplying by $BA$ in the equation (\ref{eq:m1}):

\begin{align} \label{eq:m2}
& A^{-1} = B + \epsilon  \nonumber \\
& BAA^{-1} =  BAB + BA\epsilon \, ,
\end{align}

but $BAA^{-1}=B$ and considering that $BA\epsilon \approx \epsilon$ we can obtain an the following expression for $\epsilon$ using equation (\ref{eq:m2}):

\begin{align} \label{eq:m3}
B =  BAB + BA\epsilon \nonumber \\
\epsilon \approx B-BAB \, .
\end{align}

Replacing expression (\ref{eq:m3}) in equation (\ref{eq:m1}) we have:

\begin{align} \label{eq:m4}
A^{-1} \approx B + B-BAB \nonumber \\
A^{-1} \approx 2B - BAB
\end{align}

The equation (\ref{eq:m4}) allows us to improve the matrix inverse approximation.

The algorithm (\ref{alg:rlms}) includes the recursive model formulation and the matrix inverse optimization. Every row of $X$ has an input vector and $Y$ has the corresponding input target. 
In order to initialize the algorithm we need to get first the exact inverse calculation $\Gamma$ considering the first $L$ rows of $X$. For the next rows of $X$, the inverse calculation will be approximated by our proposal shown in section (\ref{sec:proposal}). We need to get first the matrices $A$ and $\Delta$ in lines 3 and 4 in order to express the new matrix inverse approximation using the Taylor expansion in line 5. In line 6 the improvement of the matrix inverse approximation is done as it was shown in section (\ref{sec:matrixinv}). Finally the model parameters $w$ can be updated using the matrix inverse approximation and the new information of the pair of data $(X_{i},d_{i})$.

\begin{algorithm}[ht]
\begin{algorithmic}[1]
\REQUIRE $\,$ \\
$L$: windows size \\
$X$: matrix data with $n$ rows as input vectors \\
$Y$: target with size $n$ \\ 
\ENSURE  $\,$ \\
$w$: model parameters vector\\
\STATE Initialize $\Gamma=(X_{L:1}' X_{L:1} + \gamma \mathbb{I})^{-1}$

\FOR { $i=L+1$ to $n$ }

    \STATE $A=X_i X_i' - X_{i-L} X_{i-L}'$
    \STATE $\Delta = A \Gamma$
    \STATE $\Gamma=\Gamma (\mathbb{I} - \Delta + \Delta^2) \quad$  
    \STATE $\Gamma=2\Gamma-\Gamma(X_{i:i-L+1}' X_{i:i-L+1})\Gamma$ 
     \STATE $w_{i-L} = \Gamma X_{i:i-L+1}'d_{i:i-L+1}$
\ENDFOR

\end{algorithmic}
\caption{Recursive Least Mean Square}
\label{alg:rlms}
\end{algorithm}

### 2017-12-18 slides/defense-slides.tex
Adding all before the exam

@@
 \begin{alertblock}{High Frequency Trading}
-High Frequency Trading (HFT) is a technology that implements different trading strategies to buy or sell an stock using sophisticated infrastructure and high speed algorithms.
+High Frequency Trading (HFT) is a technology that implements different trading strategies to buy or sell an asset using sophisticated infrastructure and high speed algorithms.
 \end{alertblock}
@@
 \item Allows more efficient price formation
-\item Reduce the difference between buyers (bid) and sellers (ask) prices.
+\item Reduces the difference between buyers (bid) and sellers (ask) prices.
 \end{itemize}
@@
 \item Removal of human decision making could lead to failures in untested or incorrect strategies
-\item fairness
+\item Fairness
 \end{itemize}

### 2017-12-18 articles-conferences/jcc2012/JCC2012/code-paper/AAR.m

@@ -1,0 +1,14 @@
+
+function f = AAR(X,Y,gamma,TD,N)
+
+f=zeros(N,1);
+b=zeros(TD,1);
+A = gamma*eye(TD);
+
+for i=1:N
+    A = A + X(:,i)*X(:,i)';
+    f(i)=b'*(A\X(:,i));
+    b = b + Y(i)*X(:,i);
+end
+
+

### 2017-12-18 articles-conferences/jcc2012/JCC2012/code-paper/RR.m

@@ -1,0 +1,12 @@
+
+function f = RR(X,Y,gamma,TD,N)
+
+f=zeros(N,1);
+b=zeros(TD,1);
+A = gamma*eye(TD);
+
+for i=1:N
+    f(i)=b'*(A\X(:,i));
+    A = A + X(:,i)*X(:,i)';
+    b = b + Y(i)*X(:,i);
+end

### 2017-12-18 articles-conferences/jcc2012/JCC2012/code-paper/SLAAR.m

@@ -1,0 +1,62 @@
+
+function f = SLAAR(X,Y,L,gamma,TD,N)
+
+f=zeros(N,1);
+b=zeros(TD,1);
+A = X(:,1:L)*X(:,1:L)'+gamma*eye(TD);
+b = X(:,1:L)*Y(1:L);
+invA = inv(A);
+
+for i=(L+1):N
+    f(i)=b'*inv(A)*X(:,i);
+    C = X(:,i)*X(:,i)'- X(:,i-L)*X(:,i-L)';
+    A = A + C;
+    invA = inv(A);
+    %Delta = C*invA;
+    %invA = invA*(eye(TD) - Delta + Delta^2);
+    %invA = 2*invA - invA*A*invA;
+    b = b + Y(i)*X(:,i) - Y(i-L)*X(:,i-L);
+end
+
+
+% function [f,f_r] = SLAAR(X,T,L,gamma,TD,N)
+% U=[];
+% d=[];
+% U_inv=zeros(TD,TD,N_tr);
+% c=0;
+% norma=[];
+% f=[];
+% f_r =[];
+% 
+% for i=1:N_tr
+%     if i <= L
+%         U = [X(:,i)'; U];
+%         d = [T(i);d];
+%         if i == L
+%             U_inv(:,:,i) = inv(U'*U + gamma*eye(TD));
+%             w = U_inv(:,:,i)*U'*d;
+%             w_r = U_inv(:,:,i)*U'*d;
+%             f(i)=w'*X(:,i);
+%             f_r(i)=w_r'*X(:,i);
+%         end
+%     else
+%         f(i)=w'*X(:,i);
+%         f_r(i)=w_r'*X(:,i);
+%         U=[X(:,i)';U(1:L-1,:)];                     %Updating U
+%         d=[T(i);d(1:L-1)];                          %Updating d
+%         
+%         A = X(:,i)*X(:,i)' - X(:,i-L)*X(:,i-L)';
+%         Delta = A*U_inv(:,:,i-1);
+%         Delt_inv = eye(TD)-Delta+Delta^2;             %Taylor expansion of (I+Delta)^{-1}
+%         B = U_inv(:,:,i-1)*Delt_inv;                  %Aproximation of inv(U(k+1)'*U(k+1))=B
+%         U_inv(:,:,i) = 2*B-B*(U'*U)*B;                %Improvement of the inverse calculation
+%         %U_inv(:,:,i)=inv(U'*U + gamma*eye(TD));
+%  
[…]

### 2017-12-18 articles-conferences/jcc2012/JCC2012/code-paper/importfile.m

@@ -1,0 +1,16 @@
+function importfile(fileToRead1)
+%IMPORTFILE(FILETOREAD1)
+%  Imports data from the specified file
+%  FILETOREAD1:  file to read
+
+%  Auto-generated by MATLAB on 05-Aug-2012 18:09:40
+
+% Import the file
+newData1 = importdata(fileToRead1);
+
+% Create new variables in the base workspace from those fields.
+vars = fieldnames(newData1);
+for i = 1:length(vars)
+    assignin('base', vars{i}, newData1.(vars{i}));
+end
+

### 2017-12-18 articles-conferences/jcc2012/JCC2012/code-paper/main.m

@@ -1,0 +1,109 @@
+
+clear all,
+close all
+clc
+
+
+%=========Parameters ==============
+gamma=0.1;
+x = 10:10:1000;                  %Length sliding windows
+error_f = zeros(length(x),1);
+i=1;
+nt=100;
+
+%======Format data================
+
+importfile('data_2000.csv');
+index = 46;          %indice del activo a predecir
+%data = data(1:200,1:10);
+[N TD]=size(data);
+data_n = zeros(N-1,TD);
+
+%Convirtiendo a retornos
+%dailyRV=100*diff(log(dailyPrices));
+
+for i=1:TD
+    data_n(:,i) = tick2ret(data(:,i));
+end
+data = data_n;
+N=N-1;
+TD=TD-1;
+
+fprintf('Stock,AAR error, RR error, SLAAR error, Best L\n');
+%for index=1:TD
+    X = data';
+    T=X(index,:)';       %Un precio de un activo se usará como target
+    X(index,:)=[];      %Se elimina la información del activo
+    
+    aar = AAR(X,T,gamma,TD,N);
+    rr = RR(X,T,gamma,TD,N);
+
+    e_aar = norm(aar(end-nt:end)-T(end-nt:end));
+    e_rr = norm(rr(end-nt:end)-T(end-nt:end));
+
+    best_error = 100;
+    best_L = 0;
+    best_f = zeros(N-1,1);
+    for L = x
+        f = SLAAR(X,T,L,gamma,TD,N);
+        error_f=norm(f(end-nt:end)-T(end-nt:end));
+        if error_f < best_error
+            best_error = error_f;
+            best_L=L;
+            best_f=f;
+        end
+    end
+    
+    fprintf('%s,%f,%f,%f,%d\n',char(colheaders(index)),e_aar,e_rr,best_error,best_L)
+%end
+
+results=[aar(end-nt:end) rr(end-nt:end) best_f(end-nt:end) T(end-nt:end)];
+nt=70;
+plot(1:nt+1,aar(end-nt:end),'s');
+hold on;
+plot(1:nt+1,rr(end-nt:end),'o');
+hold on;
+plot(1:nt+1,T(end-nt:end),'-');
+hold on;
+plot(best_f(end-nt:end),'*');
+legend('AAR','RR','Target','Our method');
[…]

### 2017-12-18 articles-conferences/jcc2012/JCC2012/code/AAR.m

@@ -1,0 +1,21 @@
+
+function f = AAR(X,T,L,gamma,TD,N_tr)
+U=[];
+d=[];
+U_inv=zeros(TD,TD,N_tr);
+c=0;
+norma=[];
+f=[];
+
+A = gamma*eye(TD);
+for i=1:N_tr
+    U = [X(:,i)'; U];
+    A = A + X(:,i)*X(:,i)';
+    d = [T(i);d];
+    w = A\U'*d;
+    if i >= L
+        f(i)=w'*X(:,i);
+    end
+end
+
+

### 2017-12-18 articles-conferences/jcc2012/JCC2012/code/RLMS.m

@@ -1,0 +1,42 @@
+
+function [f,f_r] = RLMS(X,T,L,gamma,TD,N_tr)
+U=[];
+d=[];
+U_inv=zeros(TD,TD,N_tr);
+c=0;
+norma=[];
+f=[];
+f_r =[];
+
+for i=1:N_tr
+    if i <= L
+        U = [X(:,i)'; U];
+        d = [T(i);d];
+        if i == L
+            U_inv(:,:,i) = inv(U'*U + gamma*eye(TD));
+            w = U_inv(:,:,i)*U'*d;
+            w_r = U_inv(:,:,i)*U'*d;
+            f(i)=w'*X(:,i);
+            f_r(i)=w_r'*X(:,i);
+        end
+    else
+        f(i)=w'*X(:,i);
+        f_r(i)=w_r'*X(:,i);
+        U=[X(:,i)';U(1:L-1,:)];                     %Updating U
+        d=[T(i);d(1:L-1)];                          %Updating d
+        
+        A = X(:,i)*X(:,i)' - X(:,i-L)*X(:,i-L)';
+        Delta = A*U_inv(:,:,i-1);
+        Delt_inv = eye(TD)-Delta+Delta^2;             %Taylor expansion of (I+Delta)^{-1}
+        B = U_inv(:,:,i-1)*Delt_inv;                  %Aproximation of inv(U(k+1)'*U(k+1))=B
+        U_inv(:,:,i) = 2*B-B*(U'*U)*B;                %Improvement of the inverse calculation
+        %U_inv(:,:,i)=inv(U'*U + gamma*eye(TD));
+ 
+        norma(i-L) = norm(U_inv(:,:,i) - inv(U'*U + gamma*eye(TD))); %error of the approximation
+        w = U_inv(:,:,i)*U'*d;                        %Calculo de w
+        w_r = (U'*U + gamma*eye(TD))\U'*d;
+
+    end
+end
+
+

### 2017-12-18 articles-conferences/jcc2012/JCC2012/code/RRR.m

@@ -1,0 +1,22 @@
+
+function f = RRM(X,T,L,gamma,TD,N_tr)
+U=[];
+d=[];
+U_inv=zeros(TD,TD,N_tr);
+c=0;
+norma=[];
+f=[];
+
+A = gamma*eye(TD);
+for i=1:N_tr
+    U = [X(:,i)'; U];
+    d = [T(i);d];
+    w = (U'*U+A)\U'*d;
+    if i >= L
+        f(i)=w'*X(:,i);
+        A = A + X(:,i)*X(:,i)'- X(:,i-L+1)*X(:,i-L+1)';
+        w = A\U'*d;
+    end
+end
+
+

### 2017-12-18 articles-conferences/jcc2012/JCC2012/code/format_data.m

@@ -1,0 +1,22 @@
+
+function [X,T,X_te,T_te] = format_data(TD,np,N_tr,N_te)
+load MK30                               %Load Mackey-Glass chaotic time series 5000 point
+MK30 = MK30+np*randn(size(MK30));       %Adding white noise mean 0 and std=np
+MK30 = MK30 - mean(MK30);               %Substracting
+MK30 = MK30(1500:end);                  %Eliminando primeros 1500 datos
+
+train_set = MK30(1:N_tr+TD);            
+test_set = MK30(N_tr+1:N_tr+N_te+TD);
+
+X = zeros(TD,N_tr);                     %Train Input vectors
+T = train_set(TD+1:TD+N_tr);            %Train Targets
+
+X_te = zeros(TD,N_te);                  %Test Input vectors
+T_te = test_set(TD+1:TD+N_te);          %Test Targets
+
+for k=1:N_tr
+    X(:,k) = train_set(k:k+TD-1)';
+end
+for k=1:N_te
+    X_te(:,k) = test_set(k:k+TD-1)';
+end

### 2017-12-18 articles-conferences/jcc2012/JCC2012/code/main.m

@@ -1,0 +1,76 @@
+
+clear all,
+close all
+clc
+
+%======Format data=======
+
+TD = 10;            %time delay (embedding) length
+np =.04;            %noise std
+N_tr = 1500;        %Training size
+N_te = 100;         %Test size
+
+[X,T,X_te,T_te] = format_data(TD,np,N_tr,N_te);
+
+%=========Regularized LMS===================
+
+%L=150;                                   %windows size
+%gamma=0.01;                            %regularized parameter
+%[f,f_r] = RLMS(X,T,L,gamma,TD,N_tr);
+%aar = AAR(X,T,L,gamma,TD,N_tr);
+
+T=T';
+gamma=0.01;
+x = 5:10:200;
+i=1;
+nt=80;
+% for L = x
+%     [f,f_r] = RLMS(X,T,L,gamma,TD,N_tr);
+%     aar = AAR(X,T,L,gamma,TD,N_tr);
+%     error_f(i)=norm(f(end-nt:end)-T(end-nt:end));
+%     error_aar(i)=norm(aar(end-nt:end)-T(end-nt:end));
+%     i=i+1;
+% end
+
+
+for L = x
+    f = RRR(X,T,L,gamma,TD,N_tr);
+    aar = AAR(X,T,L,gamma,TD,N_tr);
+    error_f(i)=norm(f(end-nt:end)-T(end-nt:end));
+    error_aar(i)=norm(aar(end-nt:end)-T(end-nt:end));
+    i=i+1;
+end
+
+
+
+
+plot(x,error_f,'-rs');hold on;plot(x,error_aar,'-bs');
+title('MSE');
+legend('RRG','AAR');
+xlabel('Windows size (L)');
+ylabel('MSE');
+
+
+% inicio=L;
+% fin=L+20;
+% 
+% plot(f(inicio:fin),'-rs');hold on;
+% %plot(f_r(inicio:fin),'-bs');hold on;
+% plot(aar(inicio:fin),'-ms');hold on;
+% plot(T(inicio:fin),'-gs');
+% legend('RRG','AAR','Target')
+
+
+% subplot(3,1,1);
+% plot(f(inicio:fin),'-rs');hold on;
+% plot(f_r(inicio:fin),'-bs');
+% title('F App Inverse v/s F Real Inverse');
+% 
+% subplot(3,1,2);
+% plot(f(inicio:fin),'-rs');hold on;
+% plot(T(inicio:fin),'-bs');
[…]

### 2018-10-22 src/1_Introduction.tex
correccion despues del examen

@@
 
-"An individual economic variable, viewed as a time series, can wander extensively
+``An individual economic variable, viewed as a time series, can wander extensively
 and yet some pairs of series may be expected to move so that they do not drift
-far apart."-Robert F. Engle and Clive W.J. Granger~\cite{engle1987}.\\
+far apart.''-Robert F. Engle and Clive W.J. Granger~\cite{engle1987}.\\
 In this introduction we present the scope, objectives, hypothesis and
@@
 The specific objectives of this research are,
-\begin{itemize}
+\begin{itemize}[noitemsep,topsep=0pt]
 \item $\Large \mathcal{O}_1$: \emph{A review of the literature on time series
@@
 \section{Research Hypothesis}
-
 Cointegration concept was introduced by Engle and Granger in 1987
@@
 non-stationary variables are stationary even though individually they are not.
-Moreover Stock and Watson in 1988 \cite{stock+watson1988} observed that
+Moreover, Stock and Watson in 1988 \cite{stock+watson1988} observed that
 cointegration reflects the common stochastic trends providing a useful way to

### 2018-10-22 src/3_TimeSeries.tex
Adding all before the exam
correccion despues del examen

@@
 
-Figure \ref{fig:returnacf} shows the ACF of the returns of SPY stock where the
+Figure~\ref{fig:returnacf} shows the ACF of the returns of SPY stock where the
 correlations are significant even for very long lags, this implies a long-memory
@@
 frequencies looks more like a normal distribution.
-In the figure \ref{fig:returndist} is shown the SPY returns distribution based
+In figure~\ref{fig:returndist} is shown the SPY returns distribution based
 on daily dates from the period 1st July 1998 to 4th April 2013. The distribution
@@
 processes have the following form:
-
+%
 \[
@@
 \]
-
+%
 \noindent where $t$ is the time trend and $\epsilon_{t}$ represents a stationary
@@
 \item If $f(t) = \alpha + \beta t$ we have a linear trend model which is widely used. 
-\item If $f(t)= \alpha \exp^rt$ we have an exponential growth curve.
+\item If $f(t)= \alpha \exp^{rt}$ we have an exponential growth curve.
 \item If $f(t) = c_1 + c_2t + c_3 t^2$ we have a quadratic trend model
-\item If $f(t) = \frac{1}{k+\alpha \beta ^t }$ we have a logistic curve
+\item If $f(t) = \displaystyle\frac{1}{k+\alpha \beta ^t }$ we have a logistic curve
 \end{itemize}
@@
 root or difference stationarity processes and have the following form:
-
 \[
@@
 A random walk with or without drift can be transformed to a stationary process
-by differencing the time serie once. The disadvantage of differencing is that
+by differencing the time series once. The disadvantage of differencing is that
 the process loses one observation each time the time series is differentiated.
@@
 In presence of spurious relationships the denominator of equation
-\ref{eq:rsquared} becomes very large because a large weight is placed on
+\eqref{eq:rsquared} becomes very large because a large weight is placed on
 extreme observations on either side of the mean $\bar{y}$.
@@
 
-Figure \ref{fig:spurious1} shows the time series $x_{1,t}$ and $x_{2,t}$ and
-figure \ref{fig:spurious2} shows the regression between them and how they seem
+Figure~\ref{fig:spurious1} shows the time series $x_{1,t}$ and $x_{2,t}$ and
+figure~\ref{fig:spurious2} shows the regression between them and how they seem
 to be related.
@@
 both series may drift but their residuals will not. A more formal definition of
-cointegration is given in the section \ref{sec:cointegration}.
+cointegration is given in the section~\ref{sec:cointegration}.
 
@@
 There are two statistics for cointegration under Johansen approach: the trace
-(shown in equation \ref{eq:johansentrace}) and the maximum-eigenvalue statistic
-(shown in equation \ref{eq:johansenmax}):
+(shown in equation~\eqref{eq:johansentrace}) and the maximum-eigenvalue statistic
+(shown in equation~\eqref{eq:johansenmax}):
 
@@
 
-\noindent where $r$ is the number of cointegration vectors and
-$\hat]{\lambda}_i$ is the estimated value for the ith ordered eigenvalue.
+\noindent where $r$ is the number of cointegration vectors, $T$ is the number of observations and
+$\hat{\lambda}_i$ is the estimated value for the ith ordered eigenvalue.
 $\lambda_{\text{trace}}$ null hypothesis is that the number of cointegration
@@
 
-If we have two-dimensional process $\mathbf{y}_t$, $t=1,\dots,T$ by:
+If we have a two-dimensional process $\mathbf{y}_t$, $t=1,\dots,T$ by:
 
@@
 
-Figure \ref{fig:cointex} shows the three time series examples.
+Figure~\ref{fig:cointex} shows the three time series examples.
 
@@
 method which gives with a certain probability the number of significant vectors.
-Figures \ref{fig:cointvectors3}, \ref{fig:cointvectors4}  and
-\ref{fig:cointvectors5}  shows the linear combinations obtained using
+Figures~\ref{fig:cointvectors3},~\ref{fig:cointvectors4}  and
+\eqref{fig:cointvectors5}  shows the linear combinations obtained using
 cointegration vectors given by Johansen method. The method confirm that only two
@@
 
-The $\text{ADF}_t$ statistic is based on the least squares estimates of equation \ref{eq:adftest}.
+The $\text{ADF}_t$ statistic is based on the least squares estimates of equation~\eqref{eq:adftest}.
 
@@
 \begin{eqnarray}
-r_{t}&=&100(\ln(p_{t,N}) - \ln(p_{t-1,N})) \label{eq: ret 1} \\
-\bar{\rho}(t) &=& \sum_{t=1}^T r_{t}^2  \label{eq: ret 2} \, .
+r_{t}&=&100(\ln(p_{t,N}) - \ln(p_{t-1,N})) \label{eq:ret1} \\
+\bar{\rho}(t) &=& \sum_{t=1}^T r_{t}^2  \label{eq:ret2} \, .
 \end{eqnarray}
-\noindent where equation~(\ref{eq: ret 1}) represents overnight 
+\noindent where equation~\eqref{eq:ret1} represents overnight 
 returns and the volatility as 
-equation~(\ref{eq: ret 2}), where $p_{t,N}$ is the last intraday 
+equation~\eqref{eq:ret2}, where $p_{t,N}$ is the last intraday 
 sample at day $t$. The scaled realised volatility $\rho(t)$ is 
@@
 majority of research carried out in the literature obtain the daily volatility
-as the daily squared returns as is shown in equation~(\ref{eq: ret 2}).
+as the daily squared returns as is shown in equation~\eqref{eq:ret2}.
 However, it has been proven that this measurement noise is too high for
@@
 volatility, as calculated by the cumulative sum of squared intraday returns and
-shown in equation (\ref{eq:rv}), is less noisy and doesn't lead to choosing an
+shown in equation~\eqref{eq:rv}, is less noisy and doesn't lead to choosing an
 inferior model.   
@@
 \noindent Therefore, the integral form of the stock price model shown in
-equation (\ref{eq:stockprice}) is:
+equation~\eqref{eq:stockprice} is:
 
@@
 is defined as:
-
+%
 \begin{equation}
@@
 
-The vector autoregressive VAR($p$) model \cite{sims1980} is one of the most easy
-to use, successful and flexible models for the analysis of multivariate time
-series. It is a natural extension of the univariate autoregressive (AR) model.
+There are several multivariate models widely used such as the multivariate ARCH model \cite{bollerslev1990modelling} and variations like the DCC-GARCH \cite{engle2002dynamic}, the BEKK model \cite{10.2307/3532933} and the VAR model  \cite{sims1980}.
+The vector autoregressive VAR($p$) model derives the vector correction model VECM, which studies the cointegration between time series and also is one of the most easy
+to use and flexible models for the analysis of multivariate time
+series.
+
+The VAR model is a natural extension of the univariate autoregressive (AR) model.
 The VAR model has proven to be useful for describing the dynamic behaviour of
@@
 Notice that, regarding our notation of the cointegration section
-(\ref{sec:cointegration}), we have here 
+\eqref{sec:cointegration}, we have here 
 $\mathbf{y}_t^0 = \mathbf{y}_t$,
@@
 $i=1,\dots,l$.
-Transposing each equation of the system (\ref{eq:var}) we can write
+Transposing each equation of the system~\eqref{eq:var} we can write
 the VAR($p$) model in block-matrix form as:
@@
 \end{alignat*}
-Taking into account the error term $\mathbf{E}$, equation \ref{eq:simplevareq}
+Taking into account the error term $\mathbf{E}$, equation~\eqref{eq:simplevareq}
 can be solved with respect to $\mathbf{X}$ using the ordinary least
-squares estimation (see section \ref{sec:OLS}).
+squares estimation (see section~\ref{sec:OLS}).
 
@@
 \cite{granger1974} when they introduced the concept of spurious regressions (see
-section \ref{sec:spurious}). In 1982, Nelson and Plosser \cite{nelson1982}
+section~\ref{sec:spurious}). In 1982, Nelson and Plosser \cite{nelson1982}
 showed that unit roots might be present in a wide variety of macroeconomic
@@
 
-VECM  is obtained re-writing equation \ref{eq:var} in terms of the new
+VECM  is obtained re-writing equation~\eqref{eq:var} in terms of the new
 variable $\Delta\mathbf{y}_t=\mathbf{y}_t-\mathbf{y}_{t-1}$.
@@
 $\boldsymbol{\Omega}$, expressed in terms of the matrices
-$\boldsymbol{\Phi}_i$ of the model VAR shown in equation \ref{eq:var}, are:
+$\boldsymbol{\Phi}_i$ of the model VAR shown in equation~\eqref{eq:var}, are:
 \begin{align*}
@@
 &:= -\left( \mathbb{I} - \boldsymbol{\Phi}_1 - \dots 
-    - \boldsymbol{\phi}_p \right)\,. 
+    - \boldsymbol{\Phi}_p \right)\,. 
 \end{align*}
@@
 $\boldsymbol{\alpha\beta}^\top=\boldsymbol{\alpha}^*(\boldsymbol{\beta}^*)^\top$.
-If cointegration exists, then equation (\ref{eq:vec}) can be written
+If cointegration exists, then equation~\eqref{eq:vec} can be written
 as follows:
@@
 
-Transposing each equation of the system (\ref{eq:vecfull}) we can write
+Transposing each equation of the system (\eqref{eq:vecfull}) we can write
 the VECM($p$) model in block-matrix form as:
@@
 
-\noindent where $\mathbf{Y}$ dimension is $((N-p)\times l)$, $\mathbf{A}$
+\noindent where $\mathbf{B}$ dimension is $((N-p)\times l)$, $\mathbf{A}$
 dimension is $((N-p)\times(r+(p-1)l +1))$, $\mathbf{X}$ dimension is $((r+(p-1)l
@@
 
-Taking into account the error term $\mathbf{E}$, equation~\ref{eq:simpleveceq}
+Taking into account the error term $\mathbf{E}$, equation~\eqref{eq:simpleveceq}
 can be solved with respect to $\mathbf{X}$ using the ordinary least
-squares estimation (see section \ref{sec:OLS}).
+squares estimation (see section~\ref{sec:OLS}).
 \section{Cointegration Tests}
@@
 method provides two tests based on an eigenvector decomposition of the matrix
-$\Omega = \alpha \beta^\top$ on equation~\ref{eq:vec} called trace and maximal
+$\boldsymbol{\Omega}=\boldsymbol{\alpha\beta}^\top$ on equation~\eqref{eq:vec} called trace and maximal
 eigenvalue statistic.  Once the number of cointegration number is determined,
 $r$ more significative eigenvectors provided by Johansen procedure are used to
-get the cointegration vector matrix $\beta^\top$.  Critical values for both
+get the cointegration matrix $\boldsymbol{\beta}^\top$.  Critical values for both
 tests are limited to 12 and 5 variables in the trace and maximal eigenvalue
@@
 
-When $\mathbf{A}$ is singular, solution to equation~(\ref{eq:simpleveceq}) is given
+When $\mathbf{A}$ is singular, solution to equation~\eqref{eq:simpleveceq} is given
 by the ordinary least squares (OLS) method. OLS consists of minimizing the sum
@@
 $rank(\mathbf{A})=k <  n \leq m$, $\mathbf{A}^\top \mathbf{A}$ is
-always singular and equation~(\ref{eq:pseudoinv}) cannot be used.
+always singular and equation~(\eqref{eq:pseudoinv}) cannot be used.
 More generally, the pseudo-inverse is best computed using the compact
@@
 
-More details can be found in section \ref{app:pscsvd}.
+More details can be found in section \eqref{app:pscsvd}.

### 2018-10-22 src/4_OnlineLearning.tex
correccion despues del examen

@@
 \section{Introduction}
-Machine Learning (ML) studies computer algorithms for learning something such as
-to complete a task, to make accurate predictions or to behave intelligently. The
+Machine Learning (ML) studies computer algorithms for learning patterns to complete a task, to make accurate predictions or to behave intelligently. The
 learning is always based on samples and the objective is about to do better in
@@
 supervised learning can be categorised into offline or batch learning and online
-learning (see section \ref{sec:onoffline}). 
+learning (see section~\ref{sec:onoffline}). 
 \item[Unsupervised learning] also known as clustering \cite{ben2005}. In this
@@
 
-%Supervised learning consists in finding a learning function $f: X \rightarrow Y$
-%which models a set of examples, called training set $S$, which have been drawn
-%randomly and independently according to a unknown joint distribution function
-%$p(x,y)$ on $X \times Y$:
-%
-%
-%$$
-%S = \{ (x_k,y_k)\in X\times Y : k=1,...,n \}
-%$$
-%
-%\noindent where $X \subseteq \mathbb{R}^m$ and $Y = \mathbb{R}$ in case of
-%regression problems and $Y = \{1,-1\}$  if it is a classification problem. 
-%
-%
-%A learning algorithm $\mathcal{A}$ takes as input a data set $S \in
-%\mathscr{T}=\mathscr{P}(X\times Y)$ and selects from $\mathscr{H}$ a function
-%$f$ such that $f(\mathbf{x})\approx y $ in a predictive way:
-%
-%\begin{eqnarray*}
-%\mathcal{A}: \mathscr{T} &\rightarrow & \mathscr{H} \\
-%S &\rightarrow &\mathcal{A}(S) = f
-%\end{eqnarray*}
-%
-%\noindent where $\mathscr{H}$ is called the {\em hypothesis space}, i.e. the space
-%of functions that the algorithm is allowed to search.
-
-%------------------------DRAFT PROFE -------------------------------
 The basic setting for supervised learning is a {\em data set\/} $X$, 
@@
 
-%------------------------FIN DRAFT PROFE -------------------------------
 {\em Supervised learning\/} consists then in finding a {\em learning function\/}
@@
 
-%\begin{eqnarray*}
-%\label{eq:lossfunction}
-% V & \rightarrow & \mathbb{R}^+ \\
-%  (f(\mathbf{x}), y) &\rightarrow & V(f(\mathbf{x}),y)
-%\end{eqnarray*}
-
 \begin{equation}\label{eq:lossfunction}
@@
 \label{eq:expetedrisk}
-%E[V(f(\mathbf{x}),y)] = \int_{X\times Y} V(f(\mathbf{x}),y)\;dp(x,y)\,,
 E\big[ V( f(\cdot),\cdot) \big]
-= \int_{X\times Y} V\big( f(\mathbf{x},y \big)\,dp(\mathbf{x},y)\,.
-\end{equation}
-
-For regression, the most common loss function is square loss or L2 loss function:
+= \int_{X\times Y} V\big( f(\mathbf{x},y) \big)\,dp(\mathbf{x},y)\,.
+\end{equation}
+
+For regression, the most common loss function is square loss or $L_{2}$ loss function:
 \begin{equation}
@@
 The ERM is a Riemann sum approximation of the expected error of $f$ shown in
-equation \ref{eq:expetedrisk}. By the law of large numbers it is known that:
-
-\begin{equation}
-%\frac{1}{n} \sum_{i=1}^n V(f(x_i),y_i) \xrightarrow{n \rightarrow \infty}  E[V(f(\mathbf{x}),y)]
-\frac1n\,\sum_{i=1}^n V\big( f(\bfx_i,y_i \big)
+equation~\eqref{eq:expetedrisk}. By the law of large numbers it is known that:
+
+\begin{equation}
+\frac1n\,\sum_{i=1}^n V\big( f(\bfx_i,y_i) \big)
 \overset{n\to\infty}{\longrightarrow} E\big[ V(f(\cdot),\cdot)\big]\,.
@@
 algorithm to generate the right output for an input instance outside the
-training set. It is measured by the expected risk (equation \ref{eq:expetedrisk}).
+training set. It is measured by the expected risk (equation~\eqref{eq:expetedrisk}).
 \item[The amount of training data] in most cases, generalisation accuracy
@@
 and error due to variance. Bias measures the prediction error of the model
-$\hat{f}$ from the real value, see equation \ref{eq:biastrade}. Variance is
+$\hat{f}$ from the real value, see equation~\eqref{eq:biastrade}. Variance is
 taken as the variability of a model prediction for a given data point or its
@@
 If we have a learning model $\hat{f}$, its expected generalisation error on a
-testing sample $x$ can be decomposed as shown in equation \ref{eq:bvtrade} (proof
+testing sample $x$ can be decomposed as shown in equation~\eqref{eq:bvtrade} (proof
 in \cite{geman1992neural}):
@@
 testing error can be decomposed into bias and variance components. 
-The bias shown in equation \ref{eq:biastrade} can be obtain through resampling
+The bias shown in equation~\eqref{eq:biastrade} can be obtain through resampling
 of the input data $x$, therefore an average of the predicted values can be
@@
 \noindent where $X \subseteq \mathbb{R}^m$ and $Y = \mathbb{R}$ , $V$ is the
-loss function defined in the equation \ref{eq:lossfunction} and $\mathcal{R}(f)$
+loss function defined in the equation~\eqref{eq:lossfunction} and $\mathcal{R}(f)$
 is the regulariser, a penalisation on $f$. One example of regularisation in
@@
 
-To solve equation \ref{eq:lsm} is equivalent to find a solution of:
+To solve equation~\eqref{eq:lsm} is equivalent to find a solution of:
 \begin{equation}
@@
 \underset{m \times n}{\mathbf{A}} \enskip \underset{n \times
-l}{\mathbf{\mathbf{X}}} = \underset{m \times l}{\mathbf{Y}}
+l}{\mathbf{\mathbf{X}}} = \underset{m \times l}{\mathbf{Y}} \;.
 \end{equation}
@@
 \label{eq:MP}
-\mathbf{\mathbf{X}}= \mathbf{A}^+ \mathbf{Y}
+\mathbf{\mathbf{X}}= \mathbf{A}^+ \mathbf{Y} \;,
 \end{equation}
@@
 \end{equation}
-
-%\textbf{Demo}\quad
-%
-%To solve equation~(\ref{eq:pseudoinverse}) is equivalent to solve the following
-%optimisation problem:
-%
-%
-%\begin{equation}
-%\label{eq:RRproblem2}
-%\underset{\mathbf{X}(\lambda)}{\text{min}} \quad \|
-%\mathbf{C}\mathbf{\mathbf{X}} - \mathbf{F} \|_2^2
-%\end{equation}
-%
-%\noindent where
-%
-%\begin{equation*}
-%	\mathbf{C}=
-%	\begin{bmatrix} \quad &\mathbf{A}& \quad \\ \hdashline \quad
-%& \sqrt\lambda\mathbb{I}&\quad  
-%\end{bmatrix} 
-%\quad \text{and} \quad
-%	\mathbf{F}=\begin{bmatrix} \quad &\mathbf{Y}& \quad \\ \hdashline \quad
-%& \mathbf{0}&\quad  \end{bmatrix} \, .
-%\end{equation*}
-%
-%Applying equation~(\ref{eq:MP}) and considering that $\mathbf{C}^\top
-%\mathbf{C} = \mathbf{A}^\top \mathbf{A} + \lambda \mathbb{I}$ and 
-%$\mathbf{C}^\top \mathbf{F}=\mathbf{A}^\top \mathbf{Y} $ we have:
-%
-%\begin{eqnarray*}
-%\mathbf{X}(\lambda)&=&(\mathbf{C}^\top
-%\mathbf{C})^{-1}\mathbf{C}^\top \mathbf{F} \\
-%&=& (\mathbf{A}^\top \mathbf{A} + \lambda \mathbb{I})^{-1} \mathbf{A}^\top \mathbf{Y}
-%\end{eqnarray*}
-%
-%$\blacksquare$
 
@@
     \underset{k \times k}{\mathbf{\Sigma_1}} \enskip
-    \underset{k \times n}{\mathbf{V_1}^\top}
+    \underset{k \times n}{\mathbf{V_1}^\top} \;,
 \end{equation}
@@
 \label{eq:pseudoinversesvd}
-\mathbf{A}^+ = \mathbf{V_1\Sigma_1^{-1}U_1^\top}
+\mathbf{A}^+ = \mathbf{V_1\Sigma_1^{-1}U_1^\top} \;.
 \end{equation}
@@
 J(\mathbf{\mathbf{X}}) =  \| \mathbf{A}\mathbf{\mathbf{X}} - \mathbf{Y} \|_2^2  + \lambda
- \| \mathbf{\mathbf{X}}\| ^2
+ \| \mathbf{\mathbf{X}}\| ^2 \;,
 \end{equation}
@@
 \label{eq:optsolRR}
-\mathbf{\mathbf{X}}_*=(\mathbf{A}^\top \mathbf{A}+\lambda \mathbb{I})^{-1}\mathbf{A}^\top y \, ,
+\mathbf{\mathbf{X}}_*=(\mathbf{A}^\top \mathbf{A}+\lambda \mathbb{I})^{-1}\mathbf{A}^\top y \;.
 \end{equation}
@@
 
-where $\sigma_1 \geq \sigma_2 \geq \dots \geq \sigma_n$.
+\noindent where $\sigma_1 \geq \sigma_2 \geq \dots \geq \sigma_n$.
 
@@
 regularization term affects the condition number as follows:
-
 \begin{eqnarray*}
@@
 \kappa_{ridge} &=& \|\mathbf{A}^\intercal \mathbf{A} + \lambda \mathbb{I}\| 
-\|(\mathbf{A}^\intercal \mathbf{A} + \lambda \mathbb{I})^{-1}\|=\frac{\sigma_1+\lambda}{\sigma_n + \lambda} \,
+\|(\mathbf{A}^\intercal \mathbf{A} + \lambda \mathbb{I})^{-1}\|=\frac{\sigma_1+\lambda}{\sigma_n + \lambda} \;.
 \end{eqnarray*}
@@
 
-One of the ways to determine parameter $\lambda$ is using the bias-variance tradeoff (see section \ref{sec:biasvar}). This parameter is crucial for ridge regression since it could
+One of the ways to determine parameter $\lambda$ is using the bias-variance tradeoff (see section~\ref{sec:biasvar}). This parameter is crucial for ridge regression since it could
 reduce the expected prediction error by reducing variance, considering a biased
@@
 would for these same observations.
-
+\newpage
 Algorithm~\ref{alg:onlinealg} shows the online learning algorithm structure:
-
 \begin{algorithm}[ht]
@@
 \end{algorithm}
-
 \noindent where $l$ is some loss function. Performance is later measured after
 $T$ trials as:
-
 \begin{equation*}
@@
 \end{equation*}
-
 The objective is to minimise this loss function for all instances.
@@
 
-The AAR, proposed by~\cite{vovk2001} (also known as Vovk-Azoury-Warmuth predictor ~\cite{azoury2001}), is an application of the aggregating
+The AAR, proposed by Vovk~\cite{vovk2001} (also known as Vovk-Azoury-Warmuth predictor ~\cite{azoury2001}), is an application of the aggregating
 algorithm to the problem of regression. The idea is introduce the new input
@@
 \label{eq:AARexpand}
-\mathbf{X}_{aar} = \displaystyle \big (\sum_{t=1}^{m+1}
-\mathbf{a}_t \mathbf{a}_t  ^\intercal + \gamma \mathbb{I}\big )^{-1}
+\mathbf{X}_{aar} = \displaystyle \left(\sum_{t=1}^{m+1}
+\mathbf{a}_t \mathbf{a}_t  ^\intercal + \gamma \mathbb{I} \right) ^{-1}
 \sum_{t=1}^m \mathbf{a}_t \mathbf{y}_t \, .

### 2018-10-22 src/5_1_Proposal.tex
correccion despues del examen

@@
 
-The AVECM algorithm is presented in section \ref{sec:51methodology}.
-In section \ref{sec:51results} we describe the tests carried on to assess the
+The AVECM algorithm is presented in section~\ref{sec:51methodology}.
+In section~\ref{sec:51results} we describe the tests carried on to assess the
 accuracy and the execution time of AVECM.  This section also includes a
-description of the test data.  Section \ref{sec:51conclusions} contains the
+description of the test data.  Section~\ref{sec:51conclusions} contains the
 conclusions and a discussion of future research.
@@
 
-Figure \ref{fig:hists} shows the distribution of the number of cointegration
+Figure~\ref{fig:hists} shows the distribution of the number of cointegration
 vectors given by the Johansen method for different values of $L=[2,6,10,14]$
@@
 
-From section \ref{sec:vecm} we know that $r=0$ means no cointegration and
+From section~\ref{sec:vecm} we know that $r=0$ means no cointegration and
 $r=l$ (we are using four rates, so $l=4$) reveals that no process is I(1) but
@@
 The goal of our next experiment was to find a relationship between the ratio $PC$
-and the performance measure MSE (see equation \ref{eq:MSE}). $L$ was defined
+and the performance measure MSE (see equation~\ref{eq:MSE}). $L$ was defined
 between $[2,14]$ hours, that corresponded to $[720, 5040]$ data points, and $p$
@@
 Execution time depends directly on $L$, $p$ and $nparams$ used, since they
-determine the size of matrix $\mathbf{A}$ (see equation \ref{eq:vecA}) and
+determine the size of matrix $\mathbf{A}$ (see equation~\ref{eq:vecA}) and
 therefore affect the OLS function execution time.  Therefore, if we try more
@@
 
-The superior figure in \ref{fig:extimes} shows that best performance accuracy
+The superior figure in~\ref{fig:extimes} shows that best performance accuracy
 measurements are achieved in times near or below 10 seconds in the parallel

### 2018-10-22 src/5_2_Proposal.tex
first draft ready to print
correccion despues del examen

@@
 significance level.
-\item The function \texttt{vecMatrix} returns the matrices~(\ref{eq:vecA})
-and~(\ref{eq:vecY}) that allows VECM to be solved.
+\item The function \texttt{vecMatrix} returns the matrices~\eqref{eq:vecA}
+and~\eqref{eq:vecY} that allows VECM to be solved.
 \item The function \texttt{vecMatrixOnline} returns the
-matrices~(\ref{eq:vecA}) and~(\ref{eq:vecY}) aggregating new data and removing
+matrices~\eqref{eq:vecA} and~\eqref{eq:vecY} aggregating new data and removing
 the old one, avoiding calculation of the matrix $\mathbf{A}$ from scratch.
@@
 \texttt{vecMatrixUpdate} only updates the corresponding columns of matrix
-$\mathbf{A}$ affected by those vectors (see equation~\ref{eq:vecA}).
+$\mathbf{A}$ affected by those vectors (see equation~\eqref{eq:vecA}).
 \end{itemize}
-\newpage
-
-A pseudocode of the algorithm \ref{alg:52proposal} is detailed in the figure\ref{fig:pseudocodigo}.
-\newline
 
@@
 }
+A pseudocode of the algorithm \ref{alg:52proposal} is detailed in the figure\ref{fig:pseudocodigo}.
 

### 2018-10-22 src/5_3_Proposal.tex
inicio revision capitulo 3x
correccion despues del examen

@@
 is detected by tracking the Mean Absolute Percentage Error (MAPE) defined in the
-equation \ref{eq:MAPE}, of the last $n$ in-sample forecasts.
+equation~\eqref{eq:MAPE}, of the last $n$ in-sample forecasts.
 
@@
 \noindent where matrix $\mathbf{B},\mathbf{X},\mathbf{E}$ and $\mathbf{A}$ were
-previously defined in equations \ref{eq:vecY} , \ref{eq:vecX} , \ref{eq:vecE} and
-\ref{eq:vecA} respectively.
+previously defined in equations~\eqref{eq:vecY} ,~\eqref{eq:vecX} ,~\eqref{eq:vecE} and
+\eqref{eq:vecA} respectively.
 
-In the traditional VECM the number of rows of matrix $\mathbf{A}$ increases with the number of
-observations $\mathbf{y}_t$. This proposal considers only the most recent $L$
-rows of matrices $\mathbf{A}$ and $\mathbf{B}$ defined as  $\mathbf{A}(t)$ and
-$\mathbf{B}(t)$:
+In the traditional VECM the number of rows of matrix $\mathbf{A}$ increases with
+the number of observations $\mathbf{y}_t$. This proposal considers only the most
+recent $L$ rows of matrices $\mathbf{A}$ and $\mathbf{B}$ defined as
+$\mathbf{A}(t)$ and $\mathbf{B}(t)$:
 
@@
 
-The RR solution $\mathbf{\hat{X}}(t)$ of equation \ref{eq:vareqt1} using the
+The RR solution $\mathbf{\hat{X}}(t)$ of equation~\eqref{eq:vareqt1} using the
 sliding window matrices defined above is:
@@
 \item The function \texttt{vecMatrix} returns VECM
-matrices $\mathbf{A}(t),\mathbf{B}(t)$ shown in equation~(\ref{eq:notation}). This
+matrices $\mathbf{A}(t),\mathbf{B}(t)$ shown in equation~\eqref{eq:notation}. This
 method is only required at the first step.
@@
 The proposal was compared against an optimal online algorithm which is a time varying
-VECM (OVECM-OPT). Algorithm ~\ref{alg:OOVECM} shows this modification:
+VECM (OVECM-OPT). Algorithm~\ref{alg:OOVECM} shows this modification:
 

### 2018-10-22 src/6_Conclusion.tex
correccion despues del examen

@@
 
+Financial time series are specially known for presenting high heteroscedasticity, common shocks in volatility could be wrongly interpreted as cointegration.
+To tackle this problem, a second step of the adaptive VECM algorithm could be implemented to the residuals.
+A good option could be a GARCH model, for example.
+
 Finally, it will be also interesting to improve the out-of-sample forecast
